---
title: "FADE — Directional Selection Analysis"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
params:
  json_dir:       "."
  direction:      "top"
  traitname:      "unknown_trait"
  bf_threshold:   100
  min_genes_hmap: 3
  output_dir:     "."
  fg_list_file:   NULL
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

<!-- 6.FADE_report.Rmd — Gene-level and site-level summary of the FADE results of one direction. -->
<!-- PhyloPhere | subworkflows/FADE/local/ -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  out.width = "100%"
)

suppressPackageStartupMessages({
  library(jsonlite)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggrepel)
  library(paletteer)
  library(DT)
  library(parallel)
})

# Tableau_10 colors, shared with the other reports
.tableau10 <- as.character(paletteer_d("ggthemes::Tableau_10"))

theme_set(
  theme_bw(base_size = 13) +
  theme(
    axis.title.y  = element_text(margin = ggplot2::margin(l = 20, r = 20)),
    axis.title.x  = element_text(margin = ggplot2::margin(t = 12)),
    axis.text.x   = element_text(margin = ggplot2::margin(t = 5)),
    plot.title    = element_text(face = "bold", hjust = 0, margin = ggplot2::margin(b = 6)),
    plot.subtitle = element_text(colour = "grey40", hjust = 0, margin = ggplot2::margin(b = 10))
  )
)
```

---

## Overview

This report summarises the results of **FADE** (FUBAR Approach to Directional
Evolution; Kosakovsky Pond _et al._ 2021) applied to **`r params$direction`**
foreground branches associated with the phenotype **`r params$traitname`**.

FADE quantifies, for each site in each protein alignment, whether the
**foreground branches** show a statistically elevated substitution bias towards
a particular amino acid, relative to the background rate modelled on the full
tree.  Evidence is measured by the **Bayes Factor (BF)** for each site × target
amino-acid combination.  A BF ≥ `r params$bf_threshold` is considered strong
evidence of directional selection at that site.

---

## Analysis Setup

```{r analysis-setup-fade, echo=FALSE}
has_fg <- !is.null(params$fg_list_file) &&
          nchar(params$fg_list_file %||% "") > 0 &&
          file.exists(params$fg_list_file)

if (has_fg) {
  fg_sp <- readLines(params$fg_list_file)
  fg_sp <- fg_sp[nzchar(trimws(fg_sp)) & !grepl("^#", fg_sp)]
  cat(sprintf("**Foreground species (%s direction): %d species**\n\n",
              params$direction, length(fg_sp)))
  knitr::kable(
    data.frame(Foreground_species = fg_sp),
    caption = "Species labeled as foreground branches in FADE"
  )
} else {
  cat("> Foreground species list not provided to this report.\n\n")
  cat("> Species are specified via `--branches` flags in the HyPhy command.\n")
}
```

---

```{r load-jsons}
# ── FADE JSON files ──────────────────────────────────────────────────────────
# The files are in json_dir or, when present, in its json/ subdirectory.
json_search_dir <- params$json_dir
if (dir.exists(file.path(params$json_dir, "json"))) {
  json_search_dir <- file.path(params$json_dir, "json")
}

json_files <- list.files(
  path    = json_search_dir,
  pattern = "\\.FADE\\.json$",
  full.names = TRUE
)

if (length(json_files) == 0) {
  message("No *.FADE.json files found in '", json_search_dir,
          "'. No FADE results for the ", params$direction, " direction ",
          "(genes may all have failed or no genes were significant).")
  knitr::knit_exit()
}

cat(sprintf("Found %d FADE result files.\n", length(json_files)))
```

```{r fade-json-branch-detect, echo=FALSE, results='asis'}
# Without fg_list_file, the foreground species are read from the first JSON file.
if (!has_fg && length(json_files) > 0) {
  js_sample <- tryCatch(
    fromJSON(json_files[1], simplifyVector = FALSE),
    error = function(e) NULL
  )
  if (!is.null(js_sample)) {
    message("FADE JSON top-level keys: ", paste(names(js_sample), collapse = ", "))
    fg_from_json <- tryCatch({
      branches <- js_sample[["tested"]] %||%
                  js_sample[["branch annotations"]] %||%
                  js_sample[["Branch Attributes"]]
      if (!is.null(branches) && is.list(branches)) {
        names(branches)[sapply(branches, function(b) {
          identical(b, "test") || identical(b, 1) || identical(b, "Foreground")
        })]
      } else character(0)
    }, error = function(e) character(0))
    if (length(fg_from_json) > 0) {
      cat(sprintf("\n**Foreground species detected from JSON (%s direction, first gene): %d species**\n\n",
                  params$direction, length(fg_from_json)))
      print(knitr::kable(
        data.frame(Species_from_JSON = fg_from_json),
        caption = "Foreground species detected in FADE JSON (first gene)"
      ))
    }
  }
}
```

```{r parse-fade, results='asis', message=TRUE}
# ── Parse each JSON into a per-gene summary row ──────────────────────────────
#
# Layout of the FADE JSON as read here:
#   $MLE$headers  : list of [col_name, description] pairs (same for all sites)
#   $MLE$content  : named list by amino acid (20 letters); each element is keyed by
#                   partition and holds one numeric vector per site, matching headers
#
# Each worker reduces its file to one gene-level row (gene, sites tested, number of
# significant site x AA pairs, maximum BF with its site and amino acid, per-AA counts)
# and keeps the significant sites as an attribute, so the (gene, site, AA) table is
# never materialized for all genes.

AA_LETTERS <- strsplit("ACDEFGHIKLMNPQRSTVWY", "")[[1]]
bf_thr <- params$bf_threshold

parse_fade_json <- function(path, bf_thr) {
  gene_id <- sub("\\.(top|bottom)\\.FADE\\.json$", "", basename(path))

  tryCatch({
    js <- fromJSON(path, simplifyVector = FALSE)

    if (is.null(js[["MLE"]]) || is.null(js[["MLE"]][["content"]])) {
      warning("Skipping ", gene_id, ": no MLE content in JSON")
      return(NULL)
    }

    # Column of the Bayes factor in the headers
    hdrs <- js[["MLE"]][["headers"]]
    bf_idx <- which(sapply(hdrs, function(h) {
      grepl("BayesFactor", h[[1]], ignore.case = TRUE)
    }))
    if (length(bf_idx) == 0) bf_idx <- 4L  # fallback: 4th column

    content <- js[["MLE"]][["content"]]

    # content[[aa]] is keyed by partition (a single key in these runs); the site index
    # is the position within that partition's list, not the key name.
    best_bf   <- -Inf
    best_site <- NA_integer_
    best_aa   <- NA_character_
    n_sig     <- 0L
    n_sites_tested <- 0L
    sig_aa_counts <- setNames(integer(length(AA_LETTERS)), AA_LETTERS)
    # Running maximum BF over the 20 target AAs for each site that reaches bf_thr
    # (one entry per significant site, not per site x AA).
    sig_site_bf <- numeric(0)   # names = 1-based site index (character)
    sig_site_aa <- character(0)

    for (aa in AA_LETTERS) {
      aa_data <- content[[aa]]
      if (is.null(aa_data) || length(aa_data) == 0) next
      for (site_key in names(aa_data)) {
        all_site_rows <- aa_data[[site_key]]
        if (is.null(all_site_rows) || length(all_site_rows) == 0) next
        n_sites_tested <- max(n_sites_tested, length(all_site_rows))

        bf_vals <- vapply(all_site_rows, function(rep_data) {
          rep_num <- as.numeric(unlist(rep_data))
          if (length(rep_num) >= bf_idx) rep_num[[bf_idx]] else NA_real_
        }, numeric(1))

        sig_idx <- which(!is.na(bf_vals) & bf_vals >= bf_thr)
        if (length(sig_idx) > 0) {
          n_sig <- n_sig + length(sig_idx)
          sig_aa_counts[aa] <- sig_aa_counts[aa] + length(sig_idx)
          for (si in sig_idx) {
            k <- as.character(si)
            if (!(k %in% names(sig_site_bf)) || bf_vals[si] > sig_site_bf[[k]]) {
              sig_site_bf[k] <- bf_vals[si]
              sig_site_aa[k] <- aa
            }
          }
        }

        finite_idx <- which(is.finite(bf_vals))
        if (length(finite_idx) > 0) {
          local_best <- finite_idx[which.max(bf_vals[finite_idx])]
          if (bf_vals[local_best] > best_bf) {
            best_bf   <- bf_vals[local_best]
            best_site <- local_best
            best_aa   <- aa
          }
        }
      }
    }

    if (!is.finite(best_bf)) return(NULL)

    out <- data.frame(
      gene           = gene_id,
      n_sites_tested = n_sites_tested,
      n_sig_site_aa  = n_sig,
      max_bf         = best_bf,
      top_target_aa  = if (best_bf >= bf_thr) best_aa   else NA_character_,
      top_site       = if (best_bf >= bf_thr) best_site else NA_integer_,
      stringsAsFactors = FALSE
    )
    for (aa in AA_LETTERS) out[[paste0("sigcount_", aa)]] <- sig_aa_counts[[aa]]

    # Significant sites of this gene (one row per site), attached to the gene-level row
    # as the "sig_sites" attribute.
    if (length(sig_site_bf) > 0) {
      si_int <- as.integer(names(sig_site_bf))
      attr(out, "sig_sites") <- data.frame(
        gene      = gene_id,
        site      = si_int,            # 1-based FADE site index
        position  = si_int - 1L,       # 0-based, as in the CAAS Position column
        max_bf    = as.numeric(sig_site_bf),
        target_aa = unname(sig_site_aa[names(sig_site_bf)]),
        stringsAsFactors = FALSE
      )
    }
    out

  }, error = function(e) {
    warning("Failed to parse ", gene_id, ": ", conditionMessage(e))
    NULL
  })
}

# Files are parsed in parallel across genes; FADE_REPORT_CORES (set by the Nextflow
# process) gives the number of workers.
n_report_cores <- max(1L, as.integer(Sys.getenv("FADE_REPORT_CORES", unset = detectCores())))
message(sprintf("Parsing %d FADE JSON files using %d core(s)...", length(json_files), n_report_cores))

all_results <- if (.Platform$OS.type == "unix" && n_report_cores > 1L) {
  mclapply(json_files, parse_fade_json, bf_thr = bf_thr, mc.cores = n_report_cores)
} else {
  lapply(json_files, parse_fade_json, bf_thr = bf_thr)
}
all_results <- Filter(Negate(is.null), all_results)

if (length(all_results) == 0) {
  message("✗ All FADE JSON files failed to parse (empty or corrupt). ",
          "No valid results for the ", params$direction, " direction.")
  knitr::knit_exit()
}

gene_summary <- bind_rows(all_results) %>%
  arrange(desc(n_sig_site_aa), desc(max_bf))

# Significant sites of all genes (one row per (gene, site)), from the "sig_sites"
# attribute of each gene-level row.
sig_sites <- bind_rows(lapply(all_results, function(x) attr(x, "sig_sites"))) %>%
  { if (nrow(.) == 0) . else arrange(., desc(max_bf)) }

cat(sprintf(
  "✓ Parsed %d genes | %d significant (gene, site) rows recovered.\n",
  nrow(gene_summary), nrow(sig_sites)
))
```

---

## Gene-level Summary

```{r gene-summary}
# Genes with at least one site x AA pair at or above bf_thr (gene_summary has one row per gene).
sig_genes <- gene_summary %>% filter(n_sig_site_aa > 0)

cat(sprintf(
  "✓ Gene-level summary created: %d genes tested, %d with ≥1 site (BF ≥ %g)\n",
  nrow(gene_summary), nrow(sig_genes), bf_thr
))
```

```{r gene-table}
DT::datatable(
  gene_summary %>%
    select(gene, n_sites_tested, n_sig_site_aa, max_bf, top_target_aa, top_site) %>%
    mutate(max_bf = round(max_bf, 2)) %>%
    rename(
      Gene              = gene,
      `Sites tested`    = n_sites_tested,
      `Sig. site×AA`   = n_sig_site_aa,
      `Max BF`          = max_bf,
      `Top target AA`   = top_target_aa,
      `Top site pos.`   = top_site
    ),
  filter    = "top",
  rownames  = FALSE,
  options   = list(pageLength = 25, scrollX = TRUE),
  caption   = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Gene-level FADE summary | BF threshold = %g | Direction: %s",
            bf_thr, toupper(params$direction))
  )
)
```

```{r write-summary-tsv}
# The table is written to the working directory, where Nextflow publishDir collects it;
# params$output_dir is not written to.

if (!exists("gene_summary")) {
  stop("ERROR: gene_summary variable does not exist!")
}

gene_summary_core <- gene_summary %>%
  select(gene, n_sites_tested, n_sig_site_aa, max_bf, top_target_aa, top_site)

if (nrow(gene_summary_core) == 0) {
  message("WARNING: gene_summary is empty, no TSV will be written")
} else {
  out_tsv <- paste0("fade_summary_", params$direction, ".tsv")
  write.table(gene_summary_core, out_tsv, sep = "\t", row.names = FALSE, quote = FALSE)

  # The file must exist after the write
  if (!file.exists(out_tsv)) {
    stop(sprintf("ERROR: Failed to create TSV file '%s'", out_tsv))
  }
}

invisible(NULL)
```

---

## Significant Sites (all)

```{r sig-sites-table}
if (nrow(sig_sites) == 0) {
  cat(sprintf("No sites cleared BF ≥ %g in the %s direction.\n",
              bf_thr, toupper(params$direction)))
} else {
  DT::datatable(
    sig_sites %>%
      transmute(
        Gene            = gene,
        `Site (1-based)` = site,
        `Position (0-based)` = position,
        `Max BF`        = round(max_bf, 2),
        `Target AA`     = target_aa
      ),
    filter   = "top",
    rownames = FALSE,
    options  = list(pageLength = 25, scrollX = TRUE),
    caption  = htmltools::tags$caption(
      style = "caption-side: top; text-align: left; font-weight: bold;",
      sprintf("All significant FADE sites | BF ≥ %g | Direction: %s | %d sites across %d genes",
              bf_thr, toupper(params$direction),
              nrow(sig_sites), length(unique(sig_sites$gene)))
    )
  )
}
```

```{r write-sites-tsv}
# Significant-site table, tab-delimited, with the columns of FADE_JSON_TO_CSV's
# fade_sites_<direction>.csv. The header is written even with no significant site.
out_sites_tsv <- paste0("fade_site_bf_", params$direction, ".tsv")
sites_core <- if (nrow(sig_sites) == 0) {
  data.frame(gene = character(), position = integer(),
             max_bf = numeric(), target_aa = character())
} else {
  sig_sites %>% select(gene, position, max_bf, target_aa)
}
write.table(sites_core, out_sites_tsv, sep = "\t", row.names = FALSE, quote = FALSE)
invisible(NULL)
```

---

## Visualisations

### Distribution of maximum Bayes Factors across genes

```{r bf-distribution, fig.height=5}
ggplot(gene_summary, aes(x = max_bf)) +
  geom_histogram(bins = 50, fill = .tableau10[1], colour = "white", alpha = 0.85) +
  geom_vline(xintercept = bf_thr, colour = "#d7191c", linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = bf_thr * 1.05, y = Inf, vjust = 1.5,
           label = paste("BF =", bf_thr), colour = "#d7191c", size = 3.5) +
  scale_x_log10(labels = scales::comma) +
  labs(
    title    = paste("Max Bayes Factor per gene —", toupper(params$direction), "direction"),
    subtitle = params$traitname,
    x        = "Max Bayes Factor (log10 scale)",
    y        = "Number of genes"
  ) +
  theme_bw(base_size = 13)
```

### Target amino-acid frequency among significant site × AA hits

```{r aa-frequency, fig.height=5}
if (nrow(sig_genes) == 0) {
  cat("No significant genes to plot.\n")
} else {
  # Per-AA counts of significant site x AA pairs come from the sigcount_<AA> columns,
  # summed over genes.
  aa_freq <- gene_summary %>%
    select(starts_with("sigcount_")) %>%
    summarise(across(everything(), sum)) %>%
    tidyr::pivot_longer(everything(), names_to = "target_aa", names_prefix = "sigcount_", values_to = "n") %>%
    filter(n > 0) %>%
    arrange(desc(n))

  ggplot(aa_freq, aes(x = reorder(target_aa, n), y = n)) +
    geom_col(fill = .tableau10[5], alpha = 0.85) +
    coord_flip() +
    labs(
      title = paste("Target amino-acid frequency in significant hits —",
                    toupper(params$direction)),
      x = "Target amino acid",
      y = "Number of significant site × AA combinations"
    ) +
    theme_bw(base_size = 16)
}
```


---

## Methods

**Tool:** FADE (FUBAR Approach to Directional Evolution) via HyPhy ≥ 2.5.

**Branch labelling:** Foreground branches correspond to species in the
**`r params$direction`** contrast group of the caastools traitfile for trait
`r params$traitname`. Background branches are the remaining species.
For a `r params$direction` analysis, contrast_group
`r ifelse(params$direction == "top", "= 1 (high-trait extreme)", "= 0 (low-trait extreme after FG/BG flip)")` species were labelled as foreground.

**Statistical cutoff:** Bayes Factor ≥ `r params$bf_threshold` was used as the
evidence threshold for directional selection at a site, following the convention
of Kosakovsky Pond _et al._ 2021 (_HyPhy 2.5_, MBE).

**Substitution model used for background rate:** `r params$traitname` (see
pipeline parameters for `fade_model`).

---

## Summary

This report summarizes the results of the **FADE** analysis.

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — FADE Module*
*Date of generation: `r Sys.Date()`*
