---
title: "FADE — Directional Selection Analysis"
subtitle: "`r paste0(params$traitname, ' | ', toupper(params$direction), ' direction')`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
params:
  json_dir:       "."
  direction:      "top"
  traitname:      "unknown_trait"
  bf_threshold:   100
  min_genes_hmap: 3
  output_dir:     "."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  out.width = "100%"
)

suppressPackageStartupMessages({
  library(jsonlite)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggrepel)
  library(DT)
})
```

---

## Overview

This report summarises the results of **FADE** (FUBAR Approach to Directional
Evolution; Kosakovsky Pond _et al._ 2021) applied to **`r params$direction`**
foreground branches associated with the phenotype **`r params$traitname`**.

FADE quantifies, for each site in each protein alignment, whether the
**foreground branches** show a statistically elevated substitution bias towards
a particular amino acid, relative to the background rate modelled on the full
tree.  Evidence is measured by the **Bayes Factor (BF)** for each site × target
amino-acid combination.  A BF ≥ `r params$bf_threshold` is considered strong
evidence of directional selection at that site.

```{r load-jsons}
# ── Locate all FADE JSON files in the working directory ──────────────────────
json_files <- list.files(
  path    = params$json_dir,
  pattern = "\\.FADE\\.json$",
  full.names = TRUE
)

if (length(json_files) == 0) {
  message("No *.FADE.json files found in '", params$json_dir,
          "'. No FADE results for the ", params$direction, " direction "
          "(genes may all have failed or no genes were significant).")
  knitr::knit_exit()
}

cat(sprintf("Found %d FADE result files.\n", length(json_files)))
```

```{r parse-fade, results='hide'}
# ── Parse each JSON: extract per-site × amino-acid Bayes Factors ─────────────
#
# Expected FADE JSON structure:
#   $MLE$headers  : list of [col_name, description] pairs (same for all sites)
#   $MLE$content  : named list, each element = one site (0-indexed)
#                   each site element = named list by amino acid (20 letters)
#                   each AA element   = numeric vector matching headers

AA_LETTERS <- strsplit("ACDEFGHIKLMNPQRSTVWY", "")[[1]]

parse_fade_json <- function(path) {
  gene_id <- sub("\\.FADE\\.json$", "", basename(path))

  tryCatch({
    js <- fromJSON(path, simplifyVector = FALSE)

    if (is.null(js[["MLE"]]) || is.null(js[["MLE"]][["content"]])) {
      warning("Skipping ", gene_id, ": no MLE content in JSON")
      return(NULL)
    }

    # Identify the column index of Bayes Factor
    hdrs <- js[["MLE"]][["headers"]]
    bf_idx <- which(sapply(hdrs, function(h) {
      grepl("Bayes Factor", h[[1]], ignore.case = TRUE)
    }))
    if (length(bf_idx) == 0) bf_idx <- 3L  # fallback: 3rd column

    sites <- js[["MLE"]][["content"]]

    rows <- lapply(seq_along(sites), function(i) {
      site_data <- sites[[i]]
      site_num  <- as.integer(names(sites)[i]) + 1L  # 1-indexed

      aa_rows <- lapply(AA_LETTERS, function(aa) {
        vals <- site_data[[aa]]
        if (is.null(vals)) return(NULL)
        bf_val <- tryCatch(as.numeric(vals[[bf_idx]]), error = function(e) NA_real_)
        data.frame(
          gene      = gene_id,
          site      = site_num,
          target_aa = aa,
          bayes_factor = bf_val,
          stringsAsFactors = FALSE
        )
      })
      do.call(rbind, Filter(Negate(is.null), aa_rows))
    })

    do.call(rbind, Filter(Negate(is.null), rows))

  }, error = function(e) {
    warning("Failed to parse ", gene_id, ": ", conditionMessage(e))
    NULL
  })
}

# Parse all files (may take a moment for large gene sets)
all_results <- lapply(json_files, parse_fade_json)
all_results <- Filter(Negate(is.null), all_results)

if (length(all_results) == 0) {
  message("All FADE JSON files failed to parse (empty or corrupt). "
          "No valid results for the ", params$direction, " direction.")
  knitr::knit_exit()
}

fade_df <- do.call(rbind, all_results)
cat(sprintf(
  "Parsed %d genes × %d site×AA combinations.\n",
  n_distinct(fade_df$gene),
  nrow(fade_df)
))
```

---

## Gene-level Summary

```{r gene-summary}
# ── For each gene: how many sites exceed the BF threshold? ───────────────────
bf_thr <- params$bf_threshold

gene_summary <- fade_df %>%
  filter(!is.na(bayes_factor)) %>%
  group_by(gene) %>%
  summarise(
    n_sites_tested   = n_distinct(site),
    n_sig_site_aa    = sum(bayes_factor >= bf_thr, na.rm = TRUE),
    max_bf           = max(bayes_factor, na.rm = TRUE),
    top_target_aa    = {
      sig <- filter(cur_data(), bayes_factor >= bf_thr)
      if (nrow(sig) == 0) NA_character_
      else sig$target_aa[which.max(sig$bayes_factor)]
    },
    top_site         = {
      sig <- filter(cur_data(), bayes_factor >= bf_thr)
      if (nrow(sig) == 0) NA_integer_
      else sig$site[which.max(sig$bayes_factor)]
    },
    .groups = "drop"
  ) %>%
  arrange(desc(n_sig_site_aa), desc(max_bf))

sig_genes <- gene_summary %>% filter(n_sig_site_aa > 0)

cat(sprintf(
  "%d / %d genes have ≥1 site with BF ≥ %g.\n",
  nrow(sig_genes), nrow(gene_summary), bf_thr
))
```

```{r gene-table}
DT::datatable(
  gene_summary %>%
    mutate(max_bf = round(max_bf, 2)) %>%
    rename(
      Gene              = gene,
      `Sites tested`    = n_sites_tested,
      `Sig. site×AA`   = n_sig_site_aa,
      `Max BF`          = max_bf,
      `Top target AA`   = top_target_aa,
      `Top site pos.`   = top_site
    ),
  filter    = "top",
  rownames  = FALSE,
  options   = list(pageLength = 25, scrollX = TRUE),
  caption   = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Gene-level FADE summary | BF threshold = %g | Direction: %s",
            bf_thr, toupper(params$direction))
  )
)
```

```{r write-summary-tsv}
out_tsv <- file.path(params$output_dir,
                     paste0("fade_summary_", params$direction, ".tsv"))
write.table(gene_summary, out_tsv, sep = "\t", row.names = FALSE, quote = FALSE)
invisible(NULL)
```

---

## Site × Amino-acid Detail

### Top significant site × AA combinations

```{r top-sites-table}
top_sites <- fade_df %>%
  filter(bayes_factor >= bf_thr) %>%
  arrange(desc(bayes_factor)) %>%
  mutate(bayes_factor = round(bayes_factor, 2)) %>%
  rename(
    Gene      = gene,
    Site      = site,
    `Target AA` = target_aa,
    `Bayes Factor` = bayes_factor
  )

if (nrow(top_sites) == 0) {
  cat("No site × AA combinations exceed BF ≥", bf_thr, "\n")
} else {
  DT::datatable(
    top_sites,
    filter   = "top",
    rownames = FALSE,
    options  = list(pageLength = 30, scrollX = TRUE)
  )
}
```

---

## Visualisations

### Distribution of maximum Bayes Factors across genes

```{r bf-distribution, fig.height=5}
ggplot(gene_summary, aes(x = max_bf)) +
  geom_histogram(bins = 50, fill = "#2c7bb6", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = bf_thr, colour = "#d7191c", linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = bf_thr * 1.05, y = Inf, vjust = 1.5,
           label = paste("BF =", bf_thr), colour = "#d7191c", size = 3.5) +
  scale_x_log10(labels = scales::comma) +
  labs(
    title    = paste("Max Bayes Factor per gene —", toupper(params$direction), "direction"),
    subtitle = params$traitname,
    x        = "Max Bayes Factor (log10 scale)",
    y        = "Number of genes"
  ) +
  theme_bw(base_size = 13)
```

### Target amino-acid frequency among significant site × AA hits

```{r aa-frequency, fig.height=5}
if (nrow(sig_genes) == 0) {
  cat("No significant genes to plot.\n")
} else {
  aa_freq <- fade_df %>%
    filter(bayes_factor >= bf_thr) %>%
    count(target_aa) %>%
    arrange(desc(n))

  ggplot(aa_freq, aes(x = reorder(target_aa, n), y = n)) +
    geom_col(fill = "#1a9641", alpha = 0.85) +
    coord_flip() +
    labs(
      title = paste("Target amino-acid frequency in significant hits —",
                    toupper(params$direction)),
      x = "Target amino acid",
      y = "Number of significant site × AA combinations"
    ) +
    theme_bw(base_size = 13)
}
```

### Heatmap: significant sites per gene × target amino acid

```{r heatmap, fig.height=8, fig.width=11}
n_sig <- nrow(sig_genes)

if (n_sig < params$min_genes_hmap) {
  cat(sprintf(
    "Only %d significant gene(s) found (threshold for heatmap = %d). Skipping heatmap.\n",
    n_sig, params$min_genes_hmap
  ))
} else {

  # Count how many sites each gene × AA pair has above threshold
  hm_data <- fade_df %>%
    filter(gene %in% sig_genes$gene, bayes_factor >= bf_thr) %>%
    group_by(gene, target_aa) %>%
    summarise(n_sig_sites = n(), max_bf = max(bayes_factor), .groups = "drop")

  # Pivot to wide and back to long for complete grid
  hm_wide <- hm_data %>%
    select(gene, target_aa, n_sig_sites) %>%
    pivot_wider(names_from = target_aa, values_from = n_sig_sites, values_fill = 0) %>%
    pivot_longer(-gene, names_to = "target_aa", values_to = "n_sig_sites")

  # Order genes by total significant sites
  gene_order <- hm_data %>%
    group_by(gene) %>% summarise(tot = sum(n_sig_sites)) %>%
    arrange(desc(tot)) %>% pull(gene)

  hm_wide <- hm_wide %>%
    mutate(
      gene      = factor(gene,      levels = rev(gene_order)),
      target_aa = factor(target_aa, levels = AA_LETTERS)
    )

  p <- ggplot(hm_wide, aes(x = target_aa, y = gene, fill = n_sig_sites)) +
    geom_tile(colour = "white", linewidth = 0.3) +
    scale_fill_gradient(low = "white", high = "#d7191c",
                        name = "# significant\nsites") +
    labs(
      title    = paste("Significant FADE sites per gene × target AA —",
                       toupper(params$direction)),
      subtitle = paste0("BF ≥ ", bf_thr, " | Trait: ", params$traitname),
      x = "Target amino acid",
      y = NULL
    ) +
    theme_bw(base_size = 12) +
    theme(
      axis.text.y = element_text(size = 8),
      legend.position = "right"
    )

  # Cap height at 80 genes to keep the plot readable
  if (n_sig > 80) {
    keep_genes <- gene_order[seq_len(80)]
    p <- p %+% filter(hm_wide, gene %in% keep_genes) +
      labs(caption = paste("Showing top 80 of", n_sig, "significant genes"))
  }

  print(p)
}
```

---

## Methods

**Tool:** FADE (FUBAR Approach to Directional Evolution) via HyPhy ≥ 2.5.

**Branch labelling:** Foreground branches correspond to species in the
**`r params$direction`** contrast group of the caastools traitfile for trait
`r params$traitname`. Background branches are the remaining species.
For a `r params$direction` analysis, contrast_group
`r ifelse(params$direction == "top", "= 1 (high-trait extreme)", "= 0 (low-trait extreme after FG/BG flip)")` species were labelled as foreground.

**Statistical cutoff:** Bayes Factor ≥ `r params$bf_threshold` was used as the
evidence threshold for directional selection at a site, following the convention
of Kosakovsky Pond _et al._ 2021 (_HyPhy 2.5_, MBE).

**Substitution model used for background rate:** `r params$traitname` (see
pipeline parameters for `fade_model`).

---

## Session Information

```{r session-info}
sessionInfo()
```
