---
title: "CAAS Post-Processing"
author: "Miguel Ramon"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

# CAAS Post-Processing and Declustering Analysis

This script processes CAAS (Context-Aware Amino Acid Substitutions) discovery results to remove false positives caused by high-density position clusters ("trains") and statistical outliers. The workflow includes gap filtering, outlier detection, cluster identification, and generation of cleaned CAAS datasets for downstream analysis.

## Setup and Directory Configuration

This section configures the working environment, sets directories, and loads necessary functions and libraries.

```{r setup, message=FALSE, warning=FALSE}
# Source commons.R for setup
source("./functions/commons.R")
setup_rmd()

# Load additional libraries
library(stringr)
library(purrr)
library(tibble)
library(patchwork)
library(scales)
library(ggplot2)
library(forcats)
library(ggvenn)
library(patchwork)
library(knitr)
library(kableExtra)
library(paletteer)

# Define CAAS-specific directories
caas_results_dir <- file.path(resultsDir, "2.CAAS/2.Results")
caas_postproc_dir <- file.path(resultsDir, "2.CAAS/3.CAAS_postproc")

# Consistent colorblind-friendly palette used across plots
tableau_10 <- paletteer_d("ggthemes::Tableau_10")
names(tableau_10) <- c(
  "blue", "orange", "vermillion", "bluish_green",
  "green", "yellow", "reddish_purple", "pink",
  "mauve", "gray"
)

category_palette <- c(
  "Normal" = tableau_10[["gray"]],
  "Dubious" = tableau_10[["blue"]],
  "Extreme" = tableau_10[["vermillion"]],
  "Both" = tableau_10[["reddish_purple"]]
)

category_levels <- c("Extreme", "Dubious", "Both", "Normal")

removal_palette <- c(
  "Removed by window" = tableau_10[["orange"]],
  "Removed (Extreme)" = tableau_10[["vermillion"]],
  "Removed (Dubious cluster)" = tableau_10[["blue"]]
)

trait_palette <- c(
  "malignant_prevalence" = tableau_10[["vermillion"]],
  "neoplasia_prevalence" = tableau_10[["bluish_green"]]
)

removed_category_palette <- c(
  "Extreme" = category_palette[["Extreme"]],
  "Dubious" = category_palette[["Dubious"]],
  "Other" = category_palette[["Normal"]]
)

manhattan_color_levels <- c(names(category_palette), names(removal_palette))
manhattan_color_map <- unname(c(category_palette, removal_palette))
names(manhattan_color_map) <- manhattan_color_levels

category_shapes <- c(
  "Normal" = 16,
  "Dubious" = 16,
  "Extreme" = 16,
  "Both" = 17
)
removal_shapes <- c(
  "Removed by window" = 4,
  "Removed (Extreme)" = 4,
  "Removed (Dubious cluster)" = 4
)
shape_levels <- c(names(category_shapes), names(removal_shapes))
shape_map <- c(category_shapes, removal_shapes)

min_density_display <- 0.008
```

## 1. Gap Filtering and Gene Length Annotation

This section loads CAAS discovery results, filters out positions with gaps in foreground or background, and annotates genes with their alignment lengths.

```{r gap_filtering, message=FALSE, warning=FALSE}
# Load CAAS discovery files
caas_files <- list.files(
  path = caas_results_dir,
  pattern = "\\.tab.caas$",
  full.names = TRUE,
  recursive = TRUE
)

caas_files_discovery <- caas_files[grepl("/discovery/", caas_files, ignore.case = TRUE)]

# Read and organize CAAS files by trait
setup_dfs <- lapply(caas_files_discovery, read_tsv)
names(setup_dfs) <- gsub("_sp.tab.caas", "", basename(caas_files_discovery))

# Filter positions with no gaps (GFG == 0 & GBG == 0)
setup_dfs <- lapply(setup_dfs, function(df) {
  df %>% filter(GFG == 0 & GBG == 0)
})

# Load gene lengths from alignment metadata
gene_length <- read_tsv(
  file.path(dataDir, "2.Alignments/gene_lengths.tab"),
  col_names = c("Gene_file", "Length")
) %>%
  extract(
    col = Gene_file,
    into = c("Gene", "Map", "Map2"),
    regex = "^([^.]+)\\.([^.]+)\\.([^\\.]+)",
    remove = FALSE
  ) %>%
  mutate(Map = if_else(str_detect(Map, "^[0-9]+$"), Map2, Map)) %>%
  select(Gene, Length, Map)

# Annotate CAAS with gene lengths and create GenePos identifier
setup_dfs <- lapply(setup_dfs, function(df) {
  df %>%
    left_join(gene_length, by = "Gene") %>%
    mutate(
      Position = as.integer(Position),
      Length = as.integer(Length),
      GenePos = paste(Gene, Position, sep = "@")
    )
})

cat("Loaded", length(setup_dfs), "CAAS discovery files\n")
cat("Total CAAS positions (gap-filtered):", sum(sapply(setup_dfs, nrow)), "\n")
```

## 2. Resampling Distribution Analysis

This section analyzes how species and families are distributed across CAAS discovery resampling replicates, ensuring balanced representation of high and low trait extremes across phylogenetic groups.

```{r resampling_analysis, fig.height=15, fig.width=15, message=FALSE, warning=FALSE, dev='png', dpi=320}
# Load resampling files
resamp_files <- list.files(
  path = caas_results_dir,
  pattern = "\\.resample$",
  full.names = TRUE,
  recursive = TRUE
)

# Read resampling data
resamp_dfs <- lapply(resamp_files, read_tsv, col_names = c("Round", "Top", "Bottom"))
names(resamp_dfs) <- gsub("\\.resample$", "", basename(resamp_files))

# Load family taxonomy
family_data <- read_tsv(file.path(dataDir, "5.Phylogeny/taxid_species_family.tsv"))

# Define phylogenetic ordering
ordered_species <- pruned_tree$tip.label
ordered_families <- family_data %>%
  filter(species %in% ordered_species) %>%
  mutate(family = as.character(family)) %>%
  distinct(species, family) %>%
  arrange(match(species, ordered_species)) %>%
  pull(family) %>%
  unique()

# Process resampling data: split pairs and count occurrences
resamp_dfs_long <- lapply(resamp_dfs, function(df) {
  df %>%
    separate(Top, into = c("Top1", "Top2"), sep = ",") %>%
    separate(Bottom, into = c("Bottom1", "Bottom2"), sep = ",") %>%
    pivot_longer(
      cols = c(Top1, Top2, Bottom1, Bottom2),
      names_to = "Group", 
      values_to = "Species"
    ) %>%
    mutate(Group = if_else(Group %in% c("Top1", "Top2"), "Top", "Bottom")) %>%
    group_by(Species, Group) %>%
    summarise(n = n(), .groups = "drop") %>%
    pivot_wider(names_from = Group, values_from = n, values_fill = 0) %>%
    left_join(family_data, by = c("Species" = "species")) %>%
    select(-rank, -name_class) %>%
    mutate(Species = factor(Species, levels = ordered_species))
})

# Aggregate by family
resamp_df_fam <- lapply(resamp_dfs_long, function(df) {
  df %>%
    group_by(family) %>%
    summarise(
      n_Top = sum(Top, na.rm = TRUE),
      n_Bottom = sum(Bottom, na.rm = TRUE),
      .groups = "drop"
    )
})

# Prepare species-level plot data
species_long <- imap_dfr(resamp_dfs_long, ~ .x %>%
  select(Species, Top, Bottom) %>%
  pivot_longer(c(Top, Bottom), names_to = "Group", values_to = "count") %>%
  mutate(dataset = .y)
) %>%
  mutate(
    trait = case_when(
      str_starts(dataset, "malignant_prevalence") ~ "Malignant prevalence",
      str_starts(dataset, "neoplasia_prevalence") ~ "Neoplasia prevalence",
      TRUE ~ "Unknown"
    ),
    replicate = as.integer(str_extract(str_remove(dataset, "\\.tab$"), "\\d+$")),
    Group = factor(Group, levels = c("Top", "Bottom")),
    grp_rep = factor(
      paste0(Group, " h", replicate),
      levels = c("Top h1", "Top h2", "Top h3", 
                 "Bottom h1", "Bottom h2", "Bottom h3")
    ),
    Species_plot = factor(Species, levels = ordered_species)
  )

# Species-level plot
p_species <- ggplot(
  species_long, 
  aes(x = Species_plot, y = count, fill = grp_rep, group = grp_rep)
) +
  geom_col(
    position = position_dodge2(preserve = "single", width = 0.8, reverse = TRUE), 
    width = 0.7
  ) +
  coord_flip() +
  facet_wrap(~ trait, ncol = 1, scales = "free_y") +
  labs(
    x = NULL, 
    y = "Resampling Count",
    fill = "Species–Hypothesis",
    title = "Resampling Distribution: Species Across Hypotheses"
  ) +
  scale_fill_paletteer_d("ggthemes::Tableau_10") +
  theme_minimal(base_size = 15) +
  theme(legend.position = "right")

# Prepare family-level plot data
family_long <- imap_dfr(resamp_df_fam, ~ .x %>%
  rename(Top = n_Top, Bottom = n_Bottom) %>%
  pivot_longer(c(Top, Bottom), names_to = "Group", values_to = "count") %>%
  mutate(dataset = .y)
) %>%
  mutate(
    trait = case_when(
      str_starts(dataset, "malignant_prevalence") ~ "Malignant prevalence",
      str_starts(dataset, "neoplasia_prevalence") ~ "Neoplasia prevalence",
      TRUE ~ "Unknown"
    ),
    replicate = as.integer(str_extract(str_remove(dataset, "\\.tab$"), "\\d+$")),
    Group = factor(Group, levels = c("Top", "Bottom")),
    grp_rep = factor(
      paste0(Group, " h", replicate),
      levels = c("Top h1", "Top h2", "Top h3", 
                 "Bottom h1", "Bottom h2", "Bottom h3")
    ),
    family_plot = factor(family, levels = ordered_families)
  )

# Family-level plot
p_family <- ggplot(
  family_long,
  aes(x = family_plot, y = count, fill = grp_rep, group = grp_rep)
) +
  geom_col(
    position = position_dodge2(preserve = "single", width = 0.8, reverse = TRUE),
    width = 0.7
  ) +
  coord_flip() +
  facet_wrap(~ trait, ncol = 1, scales = "free_y") +
  labs(
    x = NULL, 
    y = "Resampling Count",
    fill = "Family–Hypothesis",
    title = "Resampling Distribution: Families Across Hypotheses"
  ) +
  scale_fill_paletteer_d("ggthemes::Tableau_10") +
  theme_minimal(base_size = 15) +
  theme(legend.position = "right")

# Save plots
resampling_dir <- file.path(caas_postproc_dir, "0.Resampling")
createDir(resampling_dir)

ggsave(
  filename = file.path(resampling_dir, "resampling_species.png"),
  plot = p_species, 
  width = 15, 
  height = 15, 
  dpi = 320, 
  dev = "png"
)

ggsave(
  filename = file.path(resampling_dir, "resampling_family.png"),
  plot = p_family, 
  width = 15, 
  height = 15, 
  dpi = 320, 
  dev = "png"
)

# Display plots
p_species
p_family

cat("\nResampling analysis complete\n")
cat("Species analyzed:", length(unique(species_long$Species)), "\n")
cat("Families analyzed:", length(unique(family_long$family)), "\n")
```

## 3. Outlier Gene Detection

This section identifies genes with unusually high CAAS counts using two methods: IQR-based extreme outliers (Tukey's fences at 3×IQR) and top 1% by CAAS density (CAAS per gene length).

```{r outlier_detection, message=FALSE, warning=FALSE, fig.height=12, fig.width=10, dev='png', dpi=320}
# Load genomic coordinates (needed later for Manhattan plot and chromosomal analysis)
genomic_info <- read_tsv(
  file.path(dataDir, "2.Alignments/ensembl_genes.output"),
  show_col_types = FALSE
)

# Separate malignant and neoplasia datasets
malignant_caas <- setup_dfs[grepl("malignant", names(setup_dfs))]
neoplasia_caas <- setup_dfs[grepl("neoplasia", names(setup_dfs))]

# Merge all replicates per trait
malignant_df <- bind_rows(malignant_caas, .id = "Trait")
neoplasia_df <- bind_rows(neoplasia_caas, .id = "Trait")

# Summarize CAAS counts and density by gene
summarize_genes <- function(df) {
  df %>%
    # Group by Trait, Gene, Position to get unique positions only
    group_by(Trait, Gene, Position) %>%
    slice(1) %>%  # Keep only one occurrence of each position
    ungroup() %>%
    group_by(Gene, Length) %>%
    summarise(n_CAAS = n(), .groups = "drop") %>%
    mutate(
      n_CAAS = as.integer(n_CAAS),
      n_CAAS_per_length = (n_CAAS / Length) * 100
    )
}

malignant_distribution <- summarize_genes(malignant_df) %>% 
  mutate(Trait = "malignant_prevalence")

neoplasia_distribution <- summarize_genes(neoplasia_df) %>% 
  mutate(Trait = "neoplasia_prevalence")

df_all <- bind_rows(malignant_distribution, neoplasia_distribution)

# Calculate IQR-based extreme outlier thresholds (3×IQR)
percentiles_df <- df_all %>%
  group_by(Trait) %>%
  summarise(
    Q1 = quantile(n_CAAS, 0.25),
    Q3 = quantile(n_CAAS, 0.75),
    IQR = Q3 - Q1,
    iqr_high = Q3 + 3 * IQR,
    .groups = "drop"
  )

# Calculate top 1% threshold by CAAS density
top_1_percent_threshold <- df_all %>%
  group_by(Trait) %>%
  summarise(top_1_percent = quantile(n_CAAS_per_length, 0.99), .groups = "drop")

# Annotate genes as Extreme (top 1% density) and/or Dubious (IQR outlier)
df_all_annotated <- df_all %>%
  left_join(top_1_percent_threshold, by = "Trait") %>%
  left_join(percentiles_df, by = "Trait") %>%
  mutate(
    Extreme = n_CAAS_per_length > top_1_percent,
    Dubious = n_CAAS > iqr_high
  ) %>%
  select(Trait, Gene, Length, n_CAAS, n_CAAS_per_length, Extreme, Dubious)

# Full-scale CAAS count distribution plot
p_full <- ggplot(df_all, aes(x = n_CAAS)) +
  geom_histogram(binwidth = 0.05, color = "black", fill = "grey80", alpha = 0.8) +
  geom_vline(
    data = percentiles_df, 
    aes(xintercept = iqr_high), 
    color = tableau_10[["green"]], 
    linetype = "solid", 
    linewidth = 1
  ) +
  scale_x_log10(breaks = c(1, 2, 5, 10, 20, 50, 100, 200)) +
  facet_wrap(~Trait, scales = "free_y") +
  labs(
    title = "CAAS Count Distribution\nIQR Extreme-Outlier Threshold (Dubious Genes)",
    x = "CAAS Count (log scale)",
    y = "Gene Count"
  ) +
  theme_minimal()

# Zoomed view focusing on typical range
p_zoom_caas <- p_full + 
  coord_cartesian(ylim = c(0, 100)) +
  labs(title = "Zoomed View (y ≤ 100)")

# CAAS density distribution plot
p_density <- ggplot(df_all_annotated, aes(x = n_CAAS_per_length)) +
  geom_histogram(binwidth = 0.05, color = "black", fill = "grey80", alpha = 0.8) +
  geom_vline(
    data = top_1_percent_threshold, 
    aes(xintercept = top_1_percent), 
    color = tableau_10[["vermillion"]], 
    linetype = "dashed", 
    linewidth = 1
  ) +
  scale_x_log10(breaks = c(0.01, 0.02, 0.05, 0.1, 0.2, 0.5, 1, 2, 5, 10, 20, 50, 100)) +
  facet_wrap(~Trait, scales = "free_y") +
  labs(
    title = "CAAS Density Distribution\nTop 1% Threshold (Extreme Genes)",
    x = "CAAS per Gene Length (%, log scale)",
    y = "Gene Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Zoomed density view
p_zoom_density <- p_density + 
  coord_cartesian(ylim = c(0, 50)) +
  labs(title = "Zoomed View (y ≤ 50)")

# Combined plot layout
(p_full | p_density) / (p_zoom_caas | p_zoom_density)

# Save outputs
outliers_dir <- file.path(caas_postproc_dir, "3.Outliers")
createDir(outliers_dir)

write_tsv(df_all_annotated, file.path(outliers_dir, "caas_outliers.tsv"))

ggsave(
  filename = file.path(outliers_dir, "caas_outliers_combined.png"),
  plot = (p_full | p_density) / (p_zoom_caas | p_zoom_density),
  width = 14, 
  height = 12, 
  dpi = 320, 
  dev = "png"
)

cat("\nOutlier summary:\n")
cat("Extreme genes (top 1% density):", sum(df_all_annotated$Extreme), "\n")
cat("Dubious genes (IQR outliers):", sum(df_all_annotated$Dubious), "\n")
cat("Both Extreme and Dubious:", sum(df_all_annotated$Extreme & df_all_annotated$Dubious), "\n")
```

## 4. CAAS Cluster Analysis

This section loads cluster detection results from the Python script (`filter_caas_clusters-param.py`) and evaluates the impact of different parameter combinations on CAAS filtering.

```{r cluster_analysis, message=FALSE, warning=FALSE, fig.height=6, fig.width=10, dev='png', dpi=320}
# Load cluster filtering results
cluster_dir <- file.path(caas_postproc_dir, "2.NoTrains")

cluster_files <- list.files(
  path = cluster_dir,
  pattern = "\\.tsv$",
  full.names = TRUE,
  recursive = TRUE
)

# Read cluster files
cluster_dfs <- lapply(cluster_files, read_tsv)
names(cluster_dfs) <- gsub("\\.tsv$", "", basename(cluster_files))

# Remove summary files
cluster_dfs <- cluster_dfs[!grepl("discarded_summary", names(cluster_dfs))]

# Summarize discarded CAAS by parameter combination
discarded_summary <- data.frame(
  Approach = names(cluster_dfs),
  Discarded_CAAS = sapply(cluster_dfs, function(df) {
    sum(df$ClusteringFlag == "Discarded", na.rm = TRUE)
  }),
  minlength = as.integer(gsub(".*minlen(\\d+).*", "\\1", names(cluster_dfs))),
  maxcaas = as.integer(gsub(".*maxcaas(\\d+).*", "\\1", names(cluster_dfs)))
) %>%
  as_tibble() %>%
  mutate(Trait = sub("^(([^_]+)_([^_]+))_.*$", "\\1", Approach)) %>%
  select(Trait, Discarded_CAAS, minlength, maxcaas) %>%
  arrange(Trait, minlength, maxcaas) %>%
  # Filter to only malignant_prevalence and neoplasia_prevalence
  filter(Trait %in% c("malignant_prevalence", "neoplasia_prevalence"))

# Plot parameter sweep results
p_param_sweep <- ggplot(
  discarded_summary, 
  aes(x = factor(minlength), y = Discarded_CAAS, fill = factor(maxcaas))
) +
  geom_col(position = "dodge") +
  geom_text(
    aes(y = Discarded_CAAS / 2, label = Discarded_CAAS),
    position = position_dodge(width = 0.9),
    vjust = 0.5, 
    size = 3.5, 
    color = "black", 
    fontface = "bold", 
    angle = 90
  ) +
  facet_wrap(~ Trait, scales = "free_x") +
  scale_fill_paletteer_d("ggthemes::Tableau_10") +
  labs(
    title = "CAAS Cluster Filtering: Discarded Positions by Parameter Combination",
    x = "Minimum Cluster Length",
    y = "Number of Discarded CAAS",
    fill = "Max Density\n(%)"
  ) +
  theme_minimal()

p_param_sweep

# Save results
ggsave(
  filename = file.path(cluster_dir, "parameter_sweep_summary.png"),
  plot = p_param_sweep,
  width = 10, 
  height = 6, 
  dpi = 320, 
  dev = "png"
)

write_tsv(discarded_summary, file.path(cluster_dir, "discarded_summary.tsv"))

cat("\nParameter sweep summary saved\n")
```

### 4.1. Parameter Comparison Analysis: 2×50 through 4×70

This section performs an in-depth comparison of the primary parameter combinations (span 2–4 × density 50–70%) to understand their overlap, unique filtering behavior, and biological implications.

**Strictness guide (approximate, most to least strict):** 2×60 → 3×60 → 4×60 → 2×70 → 3×70 → 4×70.

- **2×60 / 2×70** catch short trains (span ≥2) and test progressively lenient density thresholds.
- **3×60 / 3×70** require longer runs (span ≥3) but share the same density grid.
- **4×60 / 4×70** preserve at least 3-position patterns before discarding and are the most permissive.

```{r param_comparison, message=FALSE, warning=FALSE, fig.height=14, fig.width=18, dev='png', dpi=320}
# Parameter combinations tracked downstream (minlen × density)
param_names <- c(
  "2x60" = "minlen2.*maxcaas60",
  "2x70" = "minlen2.*maxcaas70",
  "3x60" = "minlen3.*maxcaas60",
  "3x70" = "minlen3.*maxcaas70",
  "4x60" = "minlen4.*maxcaas60",
  "4x70" = "minlen4.*maxcaas70"
)
param_order <- names(param_names)
param_cols <- param_order

# Define function to analyze parameters for a specific trait
analyze_trait_params <- function(trait_name, trait_pattern) {
  cat(sprintf("\n\n========================================\n"))
  cat(sprintf("ANALYZING TRAIT: %s\n", toupper(trait_name)))
  cat(sprintf("========================================\n\n"))
  
  # Extract discarded positions for each parameter set
  param_sets <- lapply(names(param_names), function(pname) {
    pattern <- param_names[[pname]]
    matching_dfs <- cluster_dfs[grepl(pattern, names(cluster_dfs)) &
                                 grepl(trait_pattern, names(cluster_dfs))]
    
    if (length(matching_dfs) == 0) {
      warning(sprintf("No files found for %s in %s", pname, trait_name))
      return(tibble(GenePos = character(), Gene = character(), Position = integer()))
    }
    
    bind_rows(matching_dfs, .id = "source") %>%
      filter(ClusteringFlag == "Discarded") %>%
      mutate(GenePos = paste(Gene, Position, sep = "@")) %>%
      select(GenePos, Gene, Position) %>%
      distinct() %>%
      mutate(param = pname) %>%
      arrange(Gene, Position)
  })
  names(param_sets) <- names(param_names)
  
  # Counts
  param_counts <- sapply(param_sets, function(df) nrow(df))
  cat(sprintf("=== UNIQUE DISCARDED POSITIONS (%s) ===\n", trait_name))
  print(data.frame(Parameter = names(param_counts), Count = param_counts))
  
  all_positions <- unique(unlist(lapply(param_sets, function(df) df$GenePos)))
  if (length(all_positions) == 0) {
    cat("No positions found. Skipping...\n")
    return(NULL)
  }
  
  # Membership matrix
  membership_mat <- sapply(names(param_names), function(pname) {
    all_positions %in% param_sets[[pname]]$GenePos
  })
  colnames(membership_mat) <- names(param_names)
  
  # Containment analysis
  containment_analysis <- map_dfr(param_order, function(p1) {
    map_dfr(setdiff(param_order, p1), function(p2) {
      set1 <- param_sets[[p1]]$GenePos
      set2 <- param_sets[[p2]]$GenePos
      overlap <- length(intersect(set1, set2))
      p1_in_p2 <- if(length(set2) > 0) overlap / length(set2) else 0
      p2_in_p1 <- if(length(set1) > 0) overlap / length(set1) else 0
      tibble(
        Strict = p1, Loose = p2,
        Overlap = overlap,
        `%_Loose_in_Strict` = round(p2_in_p1 * 100, 1),
        `%_Strict_in_Loose` = round(p1_in_p2 * 100, 1)
      )
    })
  })
  
  # Best parameter (only consider parameters with positions)
  params_with_positions <- param_order[param_counts[param_order] > 0]
  catch_power <- map_dfr(params_with_positions, function(p1) {
    others <- setdiff(params_with_positions, p1)
    if (length(others) == 0) {
      # Only one parameter with positions
      coverage <- 100
    } else {
      coverage <- sapply(others, function(p2) {
        if (length(param_sets[[p2]]$GenePos) == 0) return(NA)  # Should not happen
        length(intersect(param_sets[[p1]]$GenePos, param_sets[[p2]]$GenePos)) /
        length(param_sets[[p2]]$GenePos)
      })
      coverage <- coverage[!is.na(coverage)]  # Remove any NAs
      if (length(coverage) == 0) {
        coverage <- 0
      }
    }
    tibble(
      Parameter = p1,
      Avg_Coverage = round(mean(coverage) * 100, 1),
      Min_Coverage = round(min(coverage) * 100, 1)
    )
  }) %>% arrange(desc(Avg_Coverage))
  
  cat(sprintf("\n🏆 BEST PARAMETER: %s (%.1f%% coverage)\n", 
              catch_power$Parameter[1], catch_power$Avg_Coverage[1]))
  
  # === PARAMETER COMPARISON PLOT ===
  # Create bar plot showing discarded positions per parameter
  param_plot_data <- data.frame(
    Parameter = names(param_counts),
    Discarded_Positions = param_counts
  )

  param_plot <- ggplot(param_plot_data, aes(x = reorder(Parameter, Discarded_Positions),
                                           y = Discarded_Positions)) +
    geom_col(fill = tableau_10[["blue"]], color = "black", size = 0.3, width = 0.7) +
    geom_text(aes(label = Discarded_Positions), vjust = -0.5, size = 4, fontface = "bold") +
    labs(title = sprintf("Parameter Comparison: %s", trait_name),
         subtitle = "Discarded positions per cluster detection parameter",
         x = "Parameter", y = "Discarded Positions") +
    theme_minimal(base_size = 12) +
    coord_flip() +
    ylim(0, max(param_plot_data$Discarded_Positions) * 1.1)

  # Save parameter comparison plot
  param_filename <- file.path(cluster_dir, sprintf("parameter_comparison_%s.png",
                                                   gsub("[^A-Za-z0-9]", "_", tolower(trait_name))))
  ggsave(filename = param_filename, plot = param_plot,
         width = 10, height = 6, dpi = 150)
  cat(sprintf("✅ Parameter comparison plot saved: %s\n", param_filename))
  
  # === DETAILED PATTERNS (with MAXIMUM label space) ===
  membership_df <- as_tibble(membership_mat) %>%
    mutate(GenePos = all_positions) %>%
    separate(GenePos, into = c("Gene", "Position"), sep = "@", remove = FALSE) %>%
    mutate(Position = as.integer(Position))

  # Dynamically build a readable pattern string and parameter count for each row
  param_cols <- names(param_names)
  membership_df <- membership_df %>%
    mutate(
      pattern = pmap_chr(across(all_of(param_cols)), function(...) {
        vals <- c(...)
        present <- param_cols[which(unlist(vals))]
        if (length(present) == 0) return("None")
        paste(present, collapse = "+")
      }),
      n_params = rowSums(across(all_of(param_cols)))
    )

  # Export positions that are unique to a single parameter (helpful for sensitivity checks)
  counts_per_position <- rowSums(membership_mat)
  unique_idx <- which(counts_per_position == 1)
  unique_positions <- all_positions[unique_idx]
  safe_trait_name <- gsub("[^A-Za-z0-9]", "_", tolower(trait_name))
  export_dir <- file.path(cluster_dir, "parameter_comparison", safe_trait_name)
  dir.create(export_dir, showWarnings = FALSE, recursive = TRUE)

  if (length(unique_positions) > 0) {
    unique_df <- tibble(GenePos = unique_positions) %>%
      separate(GenePos, into = c("Gene", "Position"), sep = "@", remove = FALSE) %>%
      mutate(Position = as.integer(Position),
             OwnedBy = apply(membership_mat[unique_idx, , drop = FALSE], 1, function(r) param_cols[which(r)][1]))
    write_tsv(unique_df, file.path(export_dir, "unique_positions.tsv"))
  } else {
    write_tsv(tibble(), file.path(export_dir, "unique_positions.tsv"))
  }

  # === Per-parameter removed-position category counts (Extreme/Dubious/Other) ===
  gene_flags <- df_all_annotated %>% filter(Trait == trait_name) %>% select(Gene, Extreme, Dubious)
  param_category_counts <- map_dfr(param_cols, function(pname) {
    gp <- param_sets[[pname]]$GenePos
    if (length(gp) == 0) {
      tibble(Parameter = pname, category = c('Extreme','Dubious','Other'), n = 0)
    } else {
      dfp <- tibble(GenePos = gp) %>% separate(GenePos, into = c('Gene','Position'), sep = '@', remove = FALSE) %>%
        left_join(gene_flags, by = 'Gene') %>%
        mutate(category = case_when(Extreme ~ 'Extreme', Dubious ~ 'Dubious', TRUE ~ 'Other')) %>%
        count(category) %>% mutate(Parameter = pname) %>%
        right_join(tibble(category = c('Extreme','Dubious','Other')), by = 'category') %>%
        mutate(Parameter = pname, n = replace_na(n, 0))
      dfp %>% select(Parameter, category, n)
    }
  })

  p_param_removed <- ggplot(param_category_counts, aes(x = Parameter, y = n, fill = category)) +
    geom_col(position = 'stack', color = 'black', size = 0.2) +
    scale_fill_manual(values = removed_category_palette) +
    labs(title = 'Removed Positions by Filtering Parameter and Gene Category',
         subtitle = sprintf('Counts of positions flagged as removed by each parameter (trait = %s)', trait_name),
         x = 'Filtering Parameter', y = 'Number of Removed Positions', fill = 'Gene Category') +
    theme_minimal(base_size = 12)

  ggsave(filename = file.path(export_dir, sprintf('removed_positions_by_parameter_%s.png', safe_trait_name)),
         plot = p_param_removed, width = 8, height = 5, dpi = 320)
  
  pattern_summary <- membership_df %>%
    filter(pattern != "None") %>%
    count(pattern, n_params, sort = TRUE) %>%
    mutate(
      percentage = round(100 * n / sum(n), 1),
      label = sprintf("%d (%.0f%%)", n, percentage)
    )
  
  # 🏷️ MAXIMUM LABEL SPACE BAR CHART
  p_patterns <- ggplot(pattern_summary,
                       aes(x = reorder(pattern, n), y = n, fill = factor(n_params))) +
    geom_col(color = "black", size = 0.15, width = 0.85, alpha = 0.85) +
    geom_text(aes(label = label),
          hjust = -0.02,
          size = 4.0,
          fontface = "bold",
          nudge_y = max(pattern_summary$n) * 0.03) +
    coord_flip(clip = "off") +
    scale_fill_paletteer_d("ggthemes::Tableau_10") +
    labs(
      title = "📊 Venn Regions",
      subtitle = "Detailed intersection patterns"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      legend.position = "bottom",
      axis.text.y = element_text(size = 11, hjust = 1),
      axis.title.y = element_blank(),
      plot.title = element_text(size = 16, face = "bold"),
      plot.subtitle = element_text(size = 12)
    ) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.35)))
  
  # Compact coverage plot
  p_coverage <- ggplot(catch_power, aes(x = reorder(Parameter, Avg_Coverage),
                                        y = Avg_Coverage, fill = Parameter)) +
    geom_col(color = "black", size = 0.5, width = 0.7) +
    geom_text(aes(label = sprintf("%.1f%%", Avg_Coverage)),
              vjust = 0.5, hjust = -0.05, size = 5.5, fontface = "bold") +
    coord_flip() +
    scale_fill_paletteer_d("ggthemes::Tableau_10")
    labs(title = "🏆 Coverage Power", y = "Average Coverage (%)") +
    theme_minimal(base_size = 17) +
    theme(legend.position = "none")
  
  # Compact gene plot
  gene_sets <- lapply(param_sets, function(df) unique(df$Gene))
  genes_per_param <- map_dfr(param_order, function(pname) {
    tibble(
      Parameter = pname,
      Total = length(gene_sets[[pname]]),
      Unique = length(setdiff(gene_sets[[pname]], 
                             unlist(gene_sets[setdiff(param_order, pname)])))
    )
  }) %>% mutate(Parameter = factor(Parameter, levels = param_order))
  
  p_genes <- ggplot(genes_per_param, aes(x = Parameter)) +
    geom_col(aes(y = Total), fill = tableau_10[["blue"]], alpha = 0.8, width = 0.7) +
    geom_col(aes(y = Unique), fill = tableau_10[["vermillion"]], width = 0.7, alpha = 0.8) +
    geom_text(aes(y = Total, label = Total), vjust = 2.5, size = 4.5, fontface = "bold") +
    geom_text(aes(y = Unique, label = Unique), vjust = 2.5, size = 4.5, color = "white", fontface = "bold") +
    labs(title = "🧬 Genes", subtitle = "Blue=Total | Vermillion=Unique") +
    theme_minimal(base_size = 17) +
    theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1, size = 12))
  
  # === SPATIAL CLUSTERING (Singlets, Doublets, etc.) ===
  detect_runs <- function(positions) {
    if (length(positions) == 0) return(integer(0))
    positions <- sort(unique(positions))
    if (length(positions) == 0) return(integer(0))
    breaks <- c(0, which(diff(positions) > 1), length(positions))
    return(diff(breaks))
  }
  
  spatial_by_param <- membership_df %>%
    pivot_longer(cols = all_of(names(param_names)), 
                 names_to = "parameter", values_to = "discarded") %>%
    filter(discarded) %>%
    group_by(parameter, Gene) %>%
    summarise(positions = list(Position), .groups = "drop") %>%
    mutate(run_lengths = map(positions, detect_runs)) %>%
    unnest(run_lengths, keep_empty = TRUE)
  
  spatial_summary <- spatial_by_param %>%
    mutate(
      cluster_type = case_when(
        run_lengths == 1 ~ "Singlet (1)",
        run_lengths == 2 ~ "Doublet (2)", 
        run_lengths == 3 ~ "Triplet (3)",
        run_lengths == 4 ~ "Quad (4)",
        run_lengths == 5 ~ "Quint (5)",
        run_lengths >= 6 ~ "6+ Long",
        TRUE ~ "Other"
      )
    ) %>%
    count(parameter, cluster_type) %>%
    complete(parameter, cluster_type, fill = list(n = 0)) %>%
    group_by(parameter) %>%
    mutate(percentage = round(100 * n / sum(n), 1)) %>%
    ungroup() %>%
    mutate(parameter = factor(parameter, levels = param_order))
  
  p_spatial <- ggplot(spatial_summary, aes(x = parameter, y = percentage/100, fill = cluster_type)) +
    geom_col(position = "fill", color = "black", size = 0.2) +
    scale_fill_paletteer_d("ggthemes::Tableau_10") +
    scale_y_continuous(labels = scales::percent) +
    labs(title = "🔗 Spatial Clustering", 
         subtitle = "Singlets/Doublets/Triplets removed per parameter",
         fill = "Cluster Type") +
    theme_minimal(base_size = 17) +
    theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1))
  
  # === GG-VENN PANELS ===
  get_param_set <- function(pname) {
    if (is.null(param_sets[[pname]])) return(character(0))
    param_sets[[pname]]$GenePos
  }
  union_sets <- function(keys) {
    unique(unlist(lapply(keys, get_param_set), use.names = FALSE))
  }
  venn_colors_three <- c(tableau_10["blue"], tableau_10["bluish_green"], tableau_10["vermillion"])
  venn_colors_two <- c(tableau_10["blue"], tableau_10["vermillion"])
  venn_theme <- theme(plot.title = element_text(size = 12, face = "bold"))

  venn_x60 <- ggvenn(
    list(`2×60` = get_param_set("2x60"), `3×60` = get_param_set("3x60"), `4×60` = get_param_set("4x60")),
    fill_color = venn_colors_three,
    fill_alpha = 0.6,
    stroke_size = 0.9,
    stroke_color = "black",
    set_name_size = 4,
    text_size = 3
  ) + ggtitle("Density 60% (2/3/4)") + venn_theme

  venn_x70 <- ggvenn(
    list(`2×70` = get_param_set("2x70"), `3×70` = get_param_set("3x70"), `4×70` = get_param_set("4x70")),
    fill_color = venn_colors_three,
    fill_alpha = 0.6,
    stroke_size = 0.9,
    stroke_color = "black",
    set_name_size = 4,
    text_size = 3
  ) + ggtitle("Density 70% (2/3/4)") + venn_theme

  venn_x6070 <- ggvenn(
    list(`All x60` = union_sets(c("2x60", "3x60", "4x60")),
         `All x70` = union_sets(c("2x70", "3x70", "4x70"))),
    fill_color = venn_colors_two,
    fill_alpha = 0.6,
    stroke_size = 0.9,
    stroke_color = "black",
    set_name_size = 4,
    text_size = 3
  ) + ggtitle("x60 vs x70 (all lengths)") + venn_theme

  venn_pair_plot <- function(keys_a, keys_b, title, label_a, label_b) {
    venn_sets <- list(union_sets(keys_a), union_sets(keys_b))
    names(venn_sets) <- c(label_a, label_b)
    ggvenn(
      venn_sets,
      fill_color = venn_colors_two,
      fill_alpha = 0.6,
      stroke_size = 0.9,
      stroke_color = "black",
      set_name_size = 4,
      text_size = 3
    ) + ggtitle(title) + venn_theme
  }

  venn_len23 <- venn_pair_plot(c("2x60", "2x70"), c("3x60", "3x70"),
                                "Span 2 vs Span 3", "Span 2", "Span 3")
  venn_len34 <- venn_pair_plot(c("3x60", "3x70"), c("4x60", "4x70"),
                                "Span 3 vs Span 4", "Span 3", "Span 4")
  venn_len24 <- venn_pair_plot(c("2x60", "2x70"), c("4x60", "4x70"),
                                "Span 2 vs Span 4", "Span 2", "Span 4")

  venn_top <- venn_x60 | venn_x70 | venn_x6070
  venn_bottom <- venn_len23 | venn_len24 | venn_len34

  p_venn_arranged <- (p_patterns / venn_top / venn_bottom) +
    plot_layout(heights = c(0.5, 0.8, 0.8)) +
    plot_annotation(
      title = sprintf("🧩 Venn Landscape: %s", trait_name),
      subtitle = "Bar = observed Venn regions; panels = requested overlaps",
      theme = theme(
        plot.title = element_text(size = 17, face = "bold"),
        plot.subtitle = element_text(size = 12)
      )
    )

  # === SUMMARY PANELS (non-Venn) ===
  p_stats_combined <- (p_genes / p_coverage / p_spatial) +
    plot_layout(heights = c(0.35, 0.35, 0.3)) +
    plot_annotation(
      title = sprintf("🎯 Parameter Overview: %s", trait_name),
      subtitle = sprintf("Total positions: %d | 🏆 Best: %s (%.1f%% coverage)",
                        length(all_positions),
                        catch_power$Parameter[1], catch_power$Avg_Coverage[1]),
      theme = theme(
        plot.title = element_text(size = 18, face = "bold"),
        plot.subtitle = element_text(size = 13)
      )
    )

  print(p_stats_combined)
  print(p_venn_arranged)

  # SAVE
  safe_trait_name <- gsub("[^A-Za-z0-9]", "_", tolower(trait_name))
  ggsave(filename = file.path(cluster_dir, sprintf("parameter_overview_%s.png", safe_trait_name)),
         plot = p_stats_combined, width = 12, height = 12, dpi = 320, bg = "white")
  ggsave(filename = file.path(cluster_dir, sprintf("parameter_venn_panels_%s.png", safe_trait_name)),
         plot = p_venn_arranged, width = 18, height = 18, dpi = 320, bg = "white")
  
  # Export tables
  export_dir <- file.path(cluster_dir, "parameter_comparison", safe_trait_name)
  dir.create(export_dir, showWarnings = FALSE, recursive = TRUE)
  
  write_tsv(pattern_summary, file.path(export_dir, "venn_patterns.tsv"))
  write_tsv(catch_power, file.path(export_dir, "best_parameters.tsv"))
  write_tsv(genes_per_param, file.path(export_dir, "genes_summary.tsv"))
  
  cat(sprintf("\n✅ ANALYSIS COMPLETE for %s\n", trait_name))
  cat(sprintf("🏆 RECOMMENDED: %s\n\n", catch_power$Parameter[1]))
  
  return(list(
    best_parameter = catch_power$Parameter[1],
    coverage = catch_power$Avg_Coverage[1]
  ))
}

# RUN
cat("🚀 VENN analysis...\n\n")
results_malignant <- analyze_trait_params("Malignant Prevalence", "malignant_prevalence")
results_neoplasia <- analyze_trait_params("Neoplasia Prevalence", "neoplasia_prevalence")

cat("\n🎉 ANALYSIS COMPLETE ✅\n")
```

### 4.2. Per-Parameter Removed Positions (Faceted)
This section creates a faceted bar plot showing the counts of removed positions by filtering parameter and gene category (Extreme/Dubious/Other) for both traits.

```{r per_parameter_removed_facet, message=FALSE, warning=FALSE, fig.height=6, fig.width=12, dev='png', dpi=320}
# Create a faceted per-parameter removed-position plot (both traits)
cat('\n=== Per-parameter removed positions (faceted by trait) ===\n')

param_cols <- names(param_names)
traits <- c('malignant_prevalence','neoplasia_prevalence')

# Helper: collect gene positions removed by each parameter for a given trait
collect_param_positions <- function(trait) {
  map_dfr(param_cols, function(pname) {
    pattern <- param_names[[pname]]
    matching <- cluster_dfs[grepl(pattern, names(cluster_dfs)) & grepl(trait, names(cluster_dfs))]
    gp <- if (length(matching) == 0) character(0) else unique(unlist(lapply(matching, function(df) {
      df %>% filter(ClusteringFlag == 'Discarded') %>% mutate(GenePos = paste(Gene, Position, sep='@')) %>% pull(GenePos)
    }), use.names = FALSE))

    tibble(Trait = trait, Parameter = pname, GenePos = gp)
  })
}

all_param_positions <- bind_rows(lapply(traits, collect_param_positions))

# Annotate categories using df_all_annotated gene flags
gene_flags <- df_all_annotated %>% select(Gene, Extreme, Dubious)

param_category_counts <- all_param_positions %>%
  separate(GenePos, into = c('Gene','Pos'), sep = '@', remove = FALSE) %>%
  left_join(gene_flags, by = 'Gene') %>%
  mutate(category = case_when(Extreme ~ 'Extreme', Dubious ~ 'Dubious', TRUE ~ 'Other')) %>%
  group_by(Trait, Parameter, category) %>%
  summarise(n = n(), .groups = 'drop') %>%
  complete(Trait, Parameter, category = c('Extreme','Dubious','Other'), fill = list(n = 0))

# Plot
p_param_facet <- ggplot(param_category_counts, aes(x = Parameter, y = n, fill = category)) +
  geom_col(position = 'stack', color = 'black', size = 0.2) +
  facet_wrap(~ Trait, scales = 'free_y') +
  scale_fill_manual(values = removed_category_palette) +
  labs(title = 'Removed Positions by Filtering Parameter and Gene Category',
       subtitle = sprintf('Counts of positions flagged as removed by each parameter (%s)', paste(param_cols, collapse = ', ')),
       x = 'Filtering Parameter', y = 'Number of Removed Positions', fill = 'Gene Category',
       caption = 'Positions are those marked as Discarded by the sliding-window cluster test for each parameter set; categories come from gene-level flags.') +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))

print(p_param_facet)

removed_dir <- file.path(caas_postproc_dir, '4.Removed_lists')
ggsave(filename = file.path(removed_dir, 'removed_positions_by_parameter_facet.png'), plot = p_param_facet, width = 12, height = 6, dpi = 320)

```

```{r per_parameter_removed_pct, message=FALSE, warning=FALSE, fig.height=6, fig.width=12, dev='png', dpi=320}
# For each trait and category, compute the percentage of positions removed by each parameter
cat('\n=== Per-parameter percentage removed by gene category ===\n')

# Total positions in each trait/category (denominator)
total_by_cat <- map_dfr(traits, function(trait) {
  orig_positions <- bind_rows(setup_dfs[grepl(trait, names(setup_dfs))], .id = 'src') %>%
    mutate(GenePos = paste(Gene, Position, sep = '@')) %>%
    distinct(GenePos, Gene) %>%
    left_join(gene_flags, by = 'Gene') %>%
    mutate(category = case_when(Extreme ~ 'Extreme', Dubious ~ 'Dubious', TRUE ~ 'Other')) %>%
    count(category) %>% mutate(Trait = trait)
  orig_positions %>% select(Trait, category, n)
})

# Join with removed counts (param_category_counts) to compute percent removed per parameter
percent_df <- param_category_counts %>%
  left_join(total_by_cat, by = c('Trait','category')) %>%
  rename(removed_n = n.x, total_n = n.y) %>%
  mutate(total_n = replace_na(total_n, 0),
         percent_removed = ifelse(total_n > 0, 100 * removed_n / total_n, NA))

# Plot percentages faceted by category
p_param_pct <- ggplot(percent_df, aes(x = Parameter, y = percent_removed, fill = category)) +
  geom_col(position = 'dodge', color = 'black', size = 0.2) +
  facet_grid(category ~ Trait, scales = 'free_y') +
  scale_fill_manual(values = removed_category_palette) +
  labs(title = 'Percentage of Positions Removed by Parameter (by Gene Category)',
       subtitle = 'Percentage = removed positions (by parameter & category) / total positions in that category',
       x = 'Filtering Parameter', y = 'Percent Removed (%)', fill = 'Category',
       caption = 'NA indicates category had zero positions in the original dataset for that trait') +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))

print(p_param_pct)
ggsave(filename = file.path(removed_dir, 'removed_positions_by_parameter_pct_facet.png'), plot = p_param_pct, width = 12, height = 6, dpi = 320)

```

```{r per_parameter_genes_all_positions, message=FALSE, warning=FALSE, fig.height=6, fig.width=12, dev='png', dpi=320}
# For each trait and parameter, compute how many genes have ALL their original CAAS positions removed by that parameter
cat('\n=== Genes with ALL positions removed by parameter (counts & percent) ===\n')

compute_genes_all_removed <- function(trait) {
  # Original positions per gene for this trait
  orig <- bind_rows(setup_dfs[grepl(trait, names(setup_dfs))], .id = 'src') %>%
    mutate(Position = as.character(Position)) %>%
    group_by(Gene) %>%
    summarise(orig_pos = list(unique(Position)), .groups = 'drop')

  # For each parameter, collect discarded positions per gene
  res <- map_dfr(names(param_names), function(pname) {
    pattern <- param_names[[pname]]
    matching <- cluster_dfs[grepl(pattern, names(cluster_dfs)) & grepl(trait, names(cluster_dfs))]

    if (length(matching) == 0) {
      discarded <- tibble(Gene = character(), disc_pos = list())
    } else {
      discarded <- bind_rows(matching, .id = 'src') %>%
        filter(ClusteringFlag == 'Discarded') %>%
        mutate(Position = as.character(Position)) %>%
        group_by(Gene) %>%
        summarise(disc_pos = list(unique(Position)), .groups = 'drop')
    }

    # Join with orig genes and test if all orig positions are in discarded positions
    df <- orig %>% left_join(discarded, by = 'Gene') %>%
      mutate(disc_pos = map(disc_pos, ~ if (is.null(.x)) character(0) else .x)) %>%
      mutate(all_removed = map2_lgl(orig_pos, disc_pos, ~ length(.x) > 0 && all(.x %in% .y))) %>%
      select(Gene, all_removed)

    df %>% mutate(Parameter = pname, Trait = trait)
  })

  return(res)
}

genes_all_removed <- bind_rows(lapply(traits, compute_genes_all_removed))

# Annotate categories and summarize counts per parameter
gene_flags <- df_all_annotated %>% select(Gene, Extreme, Dubious)

genes_all_summary <- genes_all_removed %>%
  left_join(gene_flags, by = 'Gene') %>%
  mutate(category = case_when(Extreme ~ 'Extreme', Dubious ~ 'Dubious', TRUE ~ 'Other')) %>%
  group_by(Trait, Parameter, category) %>%
  summarise(n_genes_all_removed = sum(all_removed, na.rm = TRUE), .groups = 'drop') %>%
  complete(Trait, Parameter, category = c('Extreme','Dubious','Other'), fill = list(n_genes_all_removed = 0))

# Totals per trait/category (unique genes) reused from earlier total_genes_by_cat if available; recompute safe fallback
if (!exists('total_genes_by_cat')) {
  total_genes_by_cat <- map_dfr(traits, function(trait) {
    bind_rows(setup_dfs[grepl(trait, names(setup_dfs))], .id = 'src') %>%
      distinct(Gene) %>%
      left_join(df_all_annotated %>% filter(Trait == trait) %>% select(Gene, Extreme, Dubious), by = 'Gene') %>%
      mutate(category = case_when(Extreme ~ 'Extreme', Dubious ~ 'Dubious', TRUE ~ 'Other')) %>%
      count(category) %>% mutate(Trait = trait) %>% select(Trait, category, n)
  })
}

genes_all_summary <- genes_all_summary %>% left_join(total_genes_by_cat, by = c('Trait','category')) %>%
  rename(total_genes = n) %>%
  mutate(percent_all_removed = ifelse(total_genes > 0, 100 * n_genes_all_removed / total_genes, NA))

# Plot counts
p_all_counts <- ggplot(genes_all_summary, aes(x = Parameter, y = n_genes_all_removed, fill = category)) +
  geom_col(position = 'stack', color = 'black', size = 0.2) +
  facet_wrap(~ Trait, scales = 'free_y') +
  scale_fill_manual(values = removed_category_palette) +
  labs(title = 'Genes with ALL positions removed (counts)',
       subtitle = 'Genes where every original CAAS position is discarded by the parameter',
       x = 'Filtering Parameter', y = 'Number of Genes (all positions removed)', fill = 'Gene Category') +
  theme_minimal(base_size = 12)

print(p_all_counts)
ggsave(filename = file.path(removed_dir, 'genes_all_positions_removed_by_parameter.png'), plot = p_all_counts, width = 12, height = 6, dpi = 320)

# Plot percent
p_all_pct <- ggplot(genes_all_summary, aes(x = Parameter, y = percent_all_removed, fill = category)) +
  geom_col(position = 'dodge', color = 'black', size = 0.2) +
  facet_grid(category ~ Trait, scales = 'free_y') +
  scale_fill_manual(values = removed_category_palette) +
  labs(title = 'Percentage of Genes with ALL positions removed by parameter',
       subtitle = 'Percent = genes with all positions removed / total genes in that category',
       x = 'Filtering Parameter', y = 'Percent Removed (%)', fill = 'Category') +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  theme_minimal(base_size = 12)

print(p_all_pct)
ggsave(filename = file.path(removed_dir, 'genes_all_positions_removed_by_parameter_pct.png'), plot = p_all_pct, width = 12, height = 8, dpi = 320)

```

## 5. CAAS Declustering

This section applies the chosen filtering strategy to remove both clustered positions and outlier genes. Two approaches are evaluated:

- **Approach i**: Remove Extreme genes (top 1% density) entirely, plus Dubious genes with clusters
- **Approach ii**: Keep Extreme genes, only remove Dubious genes with clusters

```{r declustering, message=FALSE, warning=FALSE, fig.height=6, fig.width=10, dev='png', dpi=320}
# Select optimal parameter combination (minlength=2, maxcaas=60)
# Note: Adjust based on parameter sweep results if needed
selected_params <- cluster_dfs[grepl("minlen2.*maxcaas60", names(cluster_dfs))]

if (length(selected_params) == 0) {
  warning("Selected parameters (2x60) not found, using minlen3.maxcaas60 as fallback")
  selected_params <- cluster_dfs[grepl("minlen3.*maxcaas60", names(cluster_dfs))]
  
  if (length(selected_params) == 0) {
    stop("Could not find any parameter combinations matching minlen4 and maxcaas60")
  }
}

# Extract clustered positions (GenePos format)
cluster_filtered <- lapply(selected_params, function(df) {
  df %>%
    mutate(GenePos = paste(Gene, Position, sep = "@")) %>%
    filter(ClusteringFlag == "Discarded") %>%
    pull(GenePos)
})

# Declustering function with configurable approach
decluster_caas <- function(prefix, remove_extreme = TRUE) {
  # Select datasets for this trait
  df_list <- setup_dfs[grepl(prefix, names(setup_dfs))]
  
  # Get cluster positions (unlist without names to avoid %in% issues)
  clust_vec <- unlist(
    cluster_filtered[grepl(prefix, names(cluster_filtered))], 
    use.names = FALSE
  )
  
  # Get genes flagged as Extreme/Dubious for this trait
  extreme_vec <- df_all_annotated %>%
    filter(Trait == prefix, Extreme) %>%
    pull(Gene) %>%
    unique()
  
  dubious_vec <- df_all_annotated %>%
    filter(Trait == prefix, Dubious) %>%
    pull(Gene) %>%
    unique()
  
  # Find Dubious genes that have clusters
  pooled <- bind_rows(df_list, .id = "src")
  
  genes_with_cluster <- pooled %>%
    filter(
      Gene %in% dubious_vec,
      GenePos %in% clust_vec
    ) %>%
    distinct(Gene) %>%
    pull(Gene)
  
  # Apply filtering
  out_list <- lapply(df_list, function(df) {
    df_filtered <- df
    
    # Approach i: Remove Extreme genes entirely
    if (remove_extreme) {
      df_filtered <- df_filtered %>%
        filter(!(Gene %in% extreme_vec))
    }
    
    # Both approaches: Remove Dubious genes with clusters
    df_filtered <- df_filtered %>%
      filter(!(Gene %in% genes_with_cluster))
    
    # Both approaches: Remove clustered positions from remaining genes
    df_filtered <- df_filtered %>%
      filter(is.na(GenePos) | !(GenePos %in% clust_vec))
    
    return(df_filtered)
  })
  
  return(out_list)
}

# Apply both approaches
malignant_declust_i <- decluster_caas("malignant_prevalence", remove_extreme = TRUE)
neoplasia_declust_i <- decluster_caas("neoplasia_prevalence", remove_extreme = TRUE)

malignant_declust_ii <- decluster_caas("malignant_prevalence", remove_extreme = FALSE)
neoplasia_declust_ii <- decluster_caas("neoplasia_prevalence", remove_extreme = FALSE)

# Save declustered datasets
save_declustered <- function(declust_list, prefix, approach) {
  declustered_dir <- file.path(caas_postproc_dir, paste0("4.Declustered_", approach), prefix)
  createDir(declustered_dir)
  
  for (name in names(declust_list)) {
    write_tsv(
      declust_list[[name]], 
      file.path(declustered_dir, paste0(name, ".declust.tsv"))
    )
  }
}

# Save Approach i outputs
save_declustered(malignant_declust_i, "malignant_prevalence", "i")
save_declustered(neoplasia_declust_i, "neoplasia_prevalence", "i")

# Save Approach ii outputs  
save_declustered(malignant_declust_ii, "malignant_prevalence", "ii")
save_declustered(neoplasia_declust_ii, "neoplasia_prevalence", "ii")

cat("\nDeclustered datasets saved for both approaches\n")

# Export lists of removed genes and positions
# These are shared between both approaches except for Extreme genes (only in approach i)

removed_dir <- file.path(caas_postproc_dir, "4.Removed_lists")
createDir(removed_dir)

# 1. Extreme genes (removed ONLY in Approach i)
# First, get all CAAS positions for extreme genes from the original data
extreme_gene_positions <- bind_rows(setup_dfs, .id = "source") %>%
  mutate(Trait = sub("^(([^_]+)_([^_]+))_.*$", "\\1", source)) %>%
  inner_join(
    df_all_annotated %>% filter(Extreme) %>% select(Trait, Gene),
    by = c("Trait", "Gene")
  ) %>%
  # Group by Trait, Gene, Position to get unique positions only
  group_by(Trait, Gene, Position) %>%
  slice(1) %>%  # Keep only one occurrence of each position
  ungroup() %>%
  group_by(Trait, Gene) %>%
  summarise(
    n_unique = n(),
    CAAS_positions = paste(sort(unique(Position)), collapse = ","),
    .groups = "drop"
  )

extreme_genes <- df_all_annotated %>%
  filter(Extreme) %>%
  select(Trait, Gene, Length, n_CAAS, n_CAAS_per_length) %>%
  left_join(extreme_gene_positions, by = c("Trait", "Gene")) %>%
  left_join(
    genomic_info %>% select(gene, chr, start, end),
    by = c("Gene" = "gene")
  ) %>%
  mutate(chr = if_else(is.na(chr), "unknown", chr)) %>%
  select(Trait, Gene, chr, start, end, Length, n_CAAS, n_CAAS_per_length, 
         n_unique, CAAS_positions) %>%
  arrange(Trait, desc(n_CAAS_per_length))

write_tsv(extreme_genes, file.path(removed_dir, "extreme_genes.tsv"))
cat(sprintf("\nExtreme genes (top 1%% by density, removed in Approach i only): %d genes\n", 
            nrow(extreme_genes)))

# Identify chromosomes with most extreme genes
extreme_chr_counts <- extreme_genes %>%
  filter(chr != "unknown") %>%
  count(chr, name = "n_extreme_genes") %>%
  arrange(desc(n_extreme_genes))

if (nrow(extreme_chr_counts) > 0) {
  cat("\nChromosomes with most extreme genes:\n")
  print(extreme_chr_counts %>% head(5))
  
  # Show details for top chromosome
  top_chr <- extreme_chr_counts$chr[1]
  top_chr_genes <- extreme_genes %>% 
    filter(chr == top_chr) %>%
    arrange(desc(n_CAAS_per_length))
  
  cat(sprintf("\nTop %d extreme genes on %s:\n", min(5, nrow(top_chr_genes)), top_chr))
  print(top_chr_genes %>% 
        head(5) %>% 
        select(Gene, Trait, n_CAAS, n_CAAS_per_length, CAAS_positions))
}

# 2. Dubious genes with clusters (removed in BOTH approaches)
# For each trait, identify which Dubious genes have clusters
for (trait in c("malignant_prevalence", "neoplasia_prevalence")) {
  # Get data for this trait
  df_list <- setup_dfs[grepl(trait, names(setup_dfs))]
  pooled <- bind_rows(df_list, .id = "src")
  
  # Get cluster positions for this trait
  clust_vec <- unlist(
    cluster_filtered[grepl(trait, names(cluster_filtered))], 
    use.names = FALSE
  )
  
  # Get dubious genes
  dubious_vec <- df_all_annotated %>%
    filter(Trait == trait, Dubious) %>%
    pull(Gene) %>%
    unique()
  
  # Find which dubious genes have clusters
  dubious_with_clusters <- pooled %>%
    filter(
      Gene %in% dubious_vec,
      GenePos %in% clust_vec
    ) %>%
    group_by(Gene) %>%
    summarise(
      n_clustered_positions = n(),
      clustered_positions = paste(sort(unique(Position)), collapse = ","),
      .groups = "drop"
    ) %>%
    left_join(
      df_all_annotated %>% filter(Trait == trait) %>% 
        select(Gene, Length, n_CAAS, n_CAAS_per_length, Dubious, Extreme),
      by = "Gene"
    ) %>%
    left_join(
      genomic_info %>% select(gene, chr, start, end),
      by = c("Gene" = "gene")
    ) %>%
    mutate(chr = if_else(is.na(chr), "unknown", chr)) %>%
    select(Gene, chr, start, end, Length, n_CAAS, n_CAAS_per_length, 
           n_clustered_positions, clustered_positions, Dubious, Extreme) %>%
    arrange(desc(n_clustered_positions))
  
  write_tsv(dubious_with_clusters, 
            file.path(removed_dir, paste0(trait, "_dubious_genes_with_clusters.tsv")))
  
  cat(sprintf("  %s: %d Dubious genes with clusters (removed in both approaches)\n", 
              trait, nrow(dubious_with_clusters)))
  
  # Identify chromosomes with most dubious genes
  dubious_chr_counts <- dubious_with_clusters %>%
    filter(chr != "unknown") %>%
    count(chr, name = "n_dubious_genes") %>%
    arrange(desc(n_dubious_genes))
  
  if (nrow(dubious_chr_counts) > 0 && dubious_chr_counts$n_dubious_genes[1] > 1) {
    cat(sprintf("    - Top chromosomes: %s\n", 
                paste(head(dubious_chr_counts, 3)$chr, 
                      " (", head(dubious_chr_counts, 3)$n_dubious_genes, ")",
                      sep = "", collapse = ", ")))
    
    # Show top genes from most affected chromosome
    top_chr <- dubious_chr_counts$chr[1]
    top_chr_genes <- dubious_with_clusters %>% 
      filter(chr == top_chr) %>%
      head(3)
    
    if (nrow(top_chr_genes) > 0) {
      cat(sprintf("    - Example genes on %s: %s\n", 
                  top_chr,
                  paste(top_chr_genes$Gene, collapse = ", ")))
    }
  }
}

# 3. All clustered positions (removed from remaining genes in BOTH approaches)
clustered_positions_summary <- bind_rows(
  lapply(names(cluster_filtered), function(name) {
    cluster_filtered[[name]] %>%
      tibble(GenePos = .) %>%
      separate(GenePos, into = c("Gene", "Position"), sep = "@", remove = FALSE) %>%
      mutate(
        Trait = sub("^(([^_]+)_([^_]+))_.*$", "\\1", name),
        Position = as.integer(Position)
      ) %>%
      select(Trait, Gene, Position, GenePos)
  })
) %>%
  group_by(Trait) %>%
  arrange(Trait, Gene, Position)

write_tsv(clustered_positions_summary, 
          file.path(removed_dir, "all_clustered_positions.tsv"))

cat(sprintf("\nTotal clustered positions identified: %d\n", 
            nrow(clustered_positions_summary)))
cat(sprintf("  malignant_prevalence: %d positions\n", 
            sum(grepl("malignant", clustered_positions_summary$Trait))))
cat(sprintf("  neoplasia_prevalence: %d positions\n", 
            sum(grepl("neoplasia", clustered_positions_summary$Trait))))

# 4. Summary of removal reasons
removal_summary <- tibble(
  Trait = c(rep("malignant_prevalence", 3), rep("neoplasia_prevalence", 3)),
  Removal_Category = rep(c("Extreme genes (Approach i only)", 
                           "Dubious genes with clusters (both approaches)", 
                           "Clustered positions (both approaches)"), 2),
  Count = c(
    # Malignant
    sum(df_all_annotated$Trait == "malignant_prevalence" & df_all_annotated$Extreme),
    nrow(read_tsv(file.path(removed_dir, "malignant_prevalence_dubious_genes_with_clusters.tsv"), 
                  show_col_types = FALSE)),
    sum(grepl("malignant", clustered_positions_summary$Trait)),
    # Neoplasia
    sum(df_all_annotated$Trait == "neoplasia_prevalence" & df_all_annotated$Extreme),
    nrow(read_tsv(file.path(removed_dir, "neoplasia_prevalence_dubious_genes_with_clusters.tsv"), 
                  show_col_types = FALSE)),
    sum(grepl("neoplasia", clustered_positions_summary$Trait))
  )
)

write_tsv(removal_summary, file.path(removed_dir, "removal_summary.tsv"))

cat("\n=== Removal Summary ===\n")
print(removal_summary)
```

## 6. Declustering Impact Summary

This section quantifies the impact of declustering on CAAS counts and tests for statistical significance.

```{r decluster_summary, message=FALSE, warning=FALSE, fig.height=8, fig.width=12, dev='png', dpi=320}
# Function to extract summary statistics
extract_summary <- function(df_list, label) {
  tibble(
    Approach = names(df_list),
    n_CAAS = sapply(df_list, nrow),
    source = label,
    Trait = sub("^(([^_]+)_([^_]+))_.*$", "\\1", names(df_list)),
    Hypothesis_ID = as.integer(gsub(".*_pair_(\\d+).*", "\\1", names(df_list)))
  )
}

# Generate comparison for Approach i
original_summary <- extract_summary(setup_dfs, "original")

declust_summary_i <- extract_summary(
  c(malignant_declust_i, neoplasia_declust_i), 
  "declustered_i"
)

declust_summary_ii <- extract_summary(
  c(malignant_declust_ii, neoplasia_declust_ii), 
  "declustered_ii"
)

# Combine for Approach i comparison
combined_i <- full_join(
  original_summary %>% rename(n_CAAS_original = n_CAAS) %>% select(-source),
  declust_summary_i %>% rename(n_CAAS_declust = n_CAAS) %>% select(-source),
  by = c("Approach", "Trait", "Hypothesis_ID")
)

# Calculate statistics
summary_stats_i <- combined_i %>%
  mutate(
    percent_diff = 100 * (n_CAAS_original - n_CAAS_declust) / n_CAAS_original,
    p_value = mapply(function(o, d) {
      prop.test(c(d, o - d), c(o, o))$p.value
    }, n_CAAS_original, n_CAAS_declust),
    significant = p_value < 0.05
  )

# Prepare data for plotting
plot_data_i <- summary_stats_i %>%
  select(Hypothesis_ID, Trait, n_CAAS_original, n_CAAS_declust, percent_diff, significant) %>%
  pivot_longer(
    cols = c(n_CAAS_original, n_CAAS_declust),
    names_to = "source", 
    values_to = "n_CAAS"
  ) %>%
  mutate(
    source = recode(source,
      "n_CAAS_original" = "Original",
      "n_CAAS_declust" = "Declustered (i)"
    )
  )

# Plot Approach i comparison
p_declust_i <- ggplot(plot_data_i, aes(x = factor(Hypothesis_ID), y = n_CAAS, fill = source)) +
  geom_col(position = "dodge") +
  geom_text(
    aes(y = n_CAAS / 2, label = n_CAAS),
    position = position_dodge(width = 0.9),
    vjust = 0.5, 
    size = 3.5, 
    color = "black", 
    fontface = "bold", 
    angle = 90
  ) +
  geom_text(
    data = summary_stats_i,
    aes(
      x = factor(Hypothesis_ID), 
      y = pmax(n_CAAS_original, n_CAAS_declust) + max(n_CAAS_original) * 0.05,
      label = paste0(round(percent_diff, 1), "%", ifelse(significant, "*", ""))
    ),
    inherit.aes = FALSE,
    size = 4,
    color = "black"
  ) +
  facet_wrap(~ Trait, scales = "free_x") +
  scale_fill_paletteer_d("ggthemes::Tableau_10") +
  labs(
    title = "Approach i: CAAS Before and After Declustering",
    subtitle = "Removes Extreme genes (top 1%) + Dubious genes with clusters + all clustered positions\n* indicates significant reduction (p < 0.05)",
    x = "Hypothesis ID",
    y = "Number of CAAS",
    fill = "Dataset"
  ) +
  theme_minimal(base_size = 11)

# Repeat for Approach ii
combined_ii <- full_join(
  original_summary %>% rename(n_CAAS_original = n_CAAS) %>% select(-source),
  declust_summary_ii %>% rename(n_CAAS_declust = n_CAAS) %>% select(-source),
  by = c("Approach", "Trait", "Hypothesis_ID")
)

summary_stats_ii <- combined_ii %>%
  mutate(
    percent_diff = 100 * (n_CAAS_original - n_CAAS_declust) / n_CAAS_original,
    p_value = mapply(function(o, d) {
      prop.test(c(d, o - d), c(o, o))$p.value
    }, n_CAAS_original, n_CAAS_declust),
    significant = p_value < 0.05
  )

plot_data_ii <- summary_stats_ii %>%
  select(Hypothesis_ID, Trait, n_CAAS_original, n_CAAS_declust, percent_diff, significant) %>%
  pivot_longer(
    cols = c(n_CAAS_original, n_CAAS_declust),
    names_to = "source", 
    values_to = "n_CAAS"
  ) %>%
  mutate(
    source = recode(source,
      "n_CAAS_original" = "Original",
      "n_CAAS_declust" = "Declustered (ii)"
    )
  )

tab10 <- as.character(paletteer_d("ggthemes::Tableau_10"))
tab10_dark <- colorspace::darken(tab10, amount = 0.2)

p_declust_ii <- ggplot(
  plot_data_ii,
  aes(x = factor(Hypothesis_ID), y = n_CAAS, fill = source)
) +
  geom_col(position = "dodge") +
  geom_text(
    aes(y = n_CAAS / 2, label = n_CAAS),
    position = position_dodge(width = 0.9),
    vjust = 0.5, 
    size = 3.5, 
    color = "black", 
    fontface = "bold", 
    angle = 90
  ) +
  geom_text(
    data = summary_stats_ii,
    aes(
      x = factor(Hypothesis_ID), 
      y = pmax(n_CAAS_original, n_CAAS_declust) + max(n_CAAS_original) * 0.05,
      label = paste0(round(percent_diff, 1), "%", ifelse(significant, "*", ""))
    ),
    inherit.aes = FALSE,
    size = 4,
    color = "black"
  ) +
  facet_wrap(~ Trait, scales = "free_x") +
  scale_fill_manual(values = tab10_dark) +
  labs(
    title = "Approach ii: CAAS Before and After Declustering",
    subtitle = "Keeps Extreme genes, removes Dubious genes with clusters + all clustered positions\n* indicates significant reduction (p < 0.05)",
    x = "Hypothesis ID",
    y = "Number of CAAS",
    fill = "Dataset"
  ) +
  theme_minimal(base_size = 11)

# Display plots
p_declust_i
p_declust_ii

# Save plots
ggsave(
  filename = file.path(caas_postproc_dir, "4.Declustered_i/declustering_summary.png"),
  plot = p_declust_i,
  width = 12, 
  height = 6, 
  dpi = 320, 
  dev = "png"
)

ggsave(
  filename = file.path(caas_postproc_dir, "4.Declustered_ii/declustering_summary.png"),
  plot = p_declust_ii,
  width = 12, 
  height = 6, 
  dpi = 320, 
  dev = "png"
)

# Print summary
cat("\n=== Declustering Summary ===\n\n")
cat("Approach i (remove Extreme + Dubious with clusters):\n")
print(summary_stats_i %>% select(Trait, Hypothesis_ID, n_CAAS_original, n_CAAS_declust, percent_diff, significant))

cat("\n\nApproach ii (keep Extreme, remove Dubious with clusters):\n")
print(summary_stats_ii %>% select(Trait, Hypothesis_ID, n_CAAS_original, n_CAAS_declust, percent_diff, significant))

# Overall summary statistics
cat("\n\n=== Overall Impact ===\n")
cat("\nApproach i:\n")
cat(sprintf("  Total CAAS (original): %d\n", sum(summary_stats_i$n_CAAS_original)))
cat(sprintf("  Total CAAS (after filtering): %d\n", sum(summary_stats_i$n_CAAS_declust)))
cat(sprintf("  Total removed: %d (%.1f%%)\n", 
            sum(summary_stats_i$n_CAAS_original) - sum(summary_stats_i$n_CAAS_declust),
            100 * (sum(summary_stats_i$n_CAAS_original) - sum(summary_stats_i$n_CAAS_declust)) / sum(summary_stats_i$n_CAAS_original)))
cat(sprintf("  Significant reductions: %d/%d hypotheses\n", 
            sum(summary_stats_i$significant), nrow(summary_stats_i)))

cat("\nApproach ii:\n")
cat(sprintf("  Total CAAS (original): %d\n", sum(summary_stats_ii$n_CAAS_original)))
cat(sprintf("  Total CAAS (after filtering): %d\n", sum(summary_stats_ii$n_CAAS_declust)))
cat(sprintf("  Total removed: %d (%.1f%%)\n", 
            sum(summary_stats_ii$n_CAAS_original) - sum(summary_stats_ii$n_CAAS_declust),
            100 * (sum(summary_stats_ii$n_CAAS_original) - sum(summary_stats_ii$n_CAAS_declust)) / sum(summary_stats_ii$n_CAAS_original)))
cat(sprintf("  Significant reductions: %d/%d hypotheses\n", 
            sum(summary_stats_ii$significant), nrow(summary_stats_ii)))

cat("\n=== Comparison Between Approaches ===\n")
cat(sprintf("  Difference in retained CAAS: %d (Approach i retains %.1f%% of what Approach ii retains)\n",
            sum(summary_stats_ii$n_CAAS_declust) - sum(summary_stats_i$n_CAAS_declust),
            100 * sum(summary_stats_i$n_CAAS_declust) / sum(summary_stats_ii$n_CAAS_declust)))
```


### 6.1. Removed CAAS by Gene Category
This section analyzes the categories of genes (Extreme, Dubious, Other) from which CAAS positions were removed under each declustering approach.

```{r removed_by_category, message=FALSE, warning=FALSE, fig.height=6, fig.width=12, dev='png', dpi=320}
# Compare number of CAAS removed that belong to Extreme vs Dubious genes per Approach
cat('\n=== Removed CAAS by Gene Category and Approach ===\n')
compute_removed_counts <- function(trait_name) {
  # Original positions for this trait
  orig <- bind_rows(setup_dfs[grepl(trait_name, names(setup_dfs))], .id = 'src') %>%
    mutate(Trait = trait_name, GenePos = paste(Gene, Position, sep = '@')) %>%
    distinct(Trait, Gene, GenePos)

  # Declustered positions for Approach i and ii
  decl_i_list <- if (trait_name == 'malignant_prevalence') malignant_declust_i else neoplasia_declust_i
  decl_ii_list <- if (trait_name == 'malignant_prevalence') malignant_declust_ii else neoplasia_declust_ii

  decl_i <- bind_rows(decl_i_list, .id = 'src') %>% mutate(GenePos = paste(Gene, Position, sep = '@')) %>% distinct(GenePos)
  decl_ii <- bind_rows(decl_ii_list, .id = 'src') %>% mutate(GenePos = paste(Gene, Position, sep = '@')) %>% distinct(GenePos)

  # Removed positions = orig \ declustered
  removed_i <- anti_join(orig, decl_i, by = 'GenePos')
  removed_ii <- anti_join(orig, decl_ii, by = 'GenePos')

  # Annotate gene-level Extreme/Dubious
  gene_flags <- df_all_annotated %>% filter(Trait == trait_name) %>% select(Gene, Extreme, Dubious)

  summarize_removed <- function(df_removed, approach) {
    df_removed %>%
      left_join(gene_flags, by = 'Gene') %>%
      mutate(category = case_when(
        Extreme == TRUE ~ 'Extreme',
        Dubious == TRUE ~ 'Dubious',
        TRUE ~ 'Other'
      )) %>%
      count(category) %>%
      mutate(Approach = approach, Trait = trait_name)
  }

  out_i <- summarize_removed(removed_i, 'Approach i')
  out_ii <- summarize_removed(removed_ii, 'Approach ii')

  bind_rows(out_i, out_ii)
}

removed_counts <- bind_rows(
  compute_removed_counts('malignant_prevalence'),
  compute_removed_counts('neoplasia_prevalence')
)

removed_counts <- removed_counts %>%
  mutate(category = factor(category, levels = c('Extreme','Dubious','Other')))

# Plot
library(ggplot2)
p_removed <- ggplot(removed_counts, aes(x = Approach, y = n, fill = category)) +
  geom_col(position = 'stack', width = 0.7, color = 'black', size = 0.2) +
  facet_wrap(~ Trait) +
  scale_fill_manual(values = removed_category_palette) +
  labs(title = 'Removed CAAS by Gene Category and Approach',
       subtitle = 'Counts of CAAS positions removed that belong to genes flagged as Extreme or Dubious',
       x = 'Filtering Approach', y = 'Number of CAAS removed', fill = 'Gene Category',
       caption = 'Approach i removes Extreme genes entirely; Approach ii keeps Extreme genes. Dubious genes with clusters are removed in both approaches.') +
  theme_minimal(base_size = 12)

print(p_removed)

ggsave(filename = file.path(removed_dir, 'removed_caas_by_category.png'), plot = p_removed, width = 12, height = 6, dpi = 320)

```

## 7. Final Summary Tables

This section presents the declustering results in tabular format for easy interpretation.

```{r final_summary, message=FALSE, warning=FALSE}
# Per-hypothesis summary table
summary_comparison <- summary_stats_i %>%
  select(Trait, Hypothesis_ID, n_CAAS_original, n_CAAS_declust, percent_diff, significant) %>%
  rename(
    `Original CAAS` = n_CAAS_original,
    `Approach i` = n_CAAS_declust,
    `% Removed (i)` = percent_diff,
    `Significant (i)` = significant
  ) %>%
  left_join(
    summary_stats_ii %>%
      select(Trait, Hypothesis_ID, n_CAAS_declust, percent_diff, significant) %>%
      rename(
        `Approach ii` = n_CAAS_declust,
        `% Removed (ii)` = percent_diff,
        `Significant (ii)` = significant
      ),
    by = c("Trait", "Hypothesis_ID")
  ) %>%
  arrange(Trait, Hypothesis_ID)

cat("\n### Per-Hypothesis Declustering Results\n\n")
kable(summary_comparison, 
      digits = 1,
      caption = "CAAS counts before and after declustering for each hypothesis",
      format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE) %>%
  print()

# Overall summary by trait
trait_summary <- bind_rows(
  summary_stats_i %>%
    group_by(Trait) %>%
    summarise(
      Approach = "i",
      Total_Original = sum(n_CAAS_original),
      Total_Retained = sum(n_CAAS_declust),
      Total_Removed = sum(n_CAAS_original) - sum(n_CAAS_declust),
      Percent_Removed = 100 * (sum(n_CAAS_original) - sum(n_CAAS_declust)) / sum(n_CAAS_original),
      N_Hypotheses = n(),
      N_Significant = sum(significant),
      .groups = "drop"
    ),
  summary_stats_ii %>%
    group_by(Trait) %>%
    summarise(
      Approach = "ii",
      Total_Original = sum(n_CAAS_original),
      Total_Retained = sum(n_CAAS_declust),
      Total_Removed = sum(n_CAAS_original) - sum(n_CAAS_declust),
      Percent_Removed = 100 * (sum(n_CAAS_original) - sum(n_CAAS_declust)) / sum(n_CAAS_original),
      N_Hypotheses = n(),
      N_Significant = sum(significant),
      .groups = "drop"
    )
) %>%
  arrange(Trait, Approach)

cat("\n### Overall Summary by Trait\n\n")
kable(trait_summary,
      digits = 1,
      col.names = c("Trait", "Approach", "Original", "Retained", "Removed", "% Removed", 
                    "N Hypotheses", "Significant"),
      caption = "Overall declustering impact aggregated by trait",
      format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE) %>%
  print()

# Grand total summary
grand_summary <- tibble(
  Approach = c("i (strict)", "ii (moderate)"),
  `Total Original CAAS` = c(sum(summary_stats_i$n_CAAS_original), 
                             sum(summary_stats_ii$n_CAAS_original)),
  `Total Retained CAAS` = c(sum(summary_stats_i$n_CAAS_declust), 
                             sum(summary_stats_ii$n_CAAS_declust)),
  `Total Removed` = c(sum(summary_stats_i$n_CAAS_original) - sum(summary_stats_i$n_CAAS_declust),
                       sum(summary_stats_ii$n_CAAS_original) - sum(summary_stats_ii$n_CAAS_declust)),
  `% Removed` = c(100 * (sum(summary_stats_i$n_CAAS_original) - sum(summary_stats_i$n_CAAS_declust)) / 
                    sum(summary_stats_i$n_CAAS_original),
                   100 * (sum(summary_stats_ii$n_CAAS_original) - sum(summary_stats_ii$n_CAAS_declust)) / 
                    sum(summary_stats_ii$n_CAAS_original)),
  `Hypotheses with Significant Reduction` = c(
    sprintf("%d/%d (%.1f%%)", sum(summary_stats_i$significant), nrow(summary_stats_i),
            100 * sum(summary_stats_i$significant) / nrow(summary_stats_i)),
    sprintf("%d/%d (%.1f%%)", sum(summary_stats_ii$significant), nrow(summary_stats_ii),
            100 * sum(summary_stats_ii$significant) / nrow(summary_stats_ii))
  )
)

cat("\n### Grand Total Summary\n\n")
kable(grand_summary,
      digits = 1,
      caption = "Overall declustering impact across all hypotheses and traits",
      format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE) %>%
  column_spec(1, bold = TRUE) %>%
  print()

cat("\n### Interpretation Guide\n\n")
cat("**Key Metrics Explained:**\n\n")
cat("- **CAAS Density (`n_CAAS_per_length`)**: (n_CAAS / Gene Length) × 100 = CAAS per 100 amino acids\n")
cat("  - Example: 15 CAAS in 500 AA gene = (15/500) × 100 = 3.0% density\n\n")
cat("- **Median Density**: The 'typical' gene density on a chromosome (robust to outliers)\n")
cat("  - Low (0.5-2%): Normal evolutionary rate\n")
cat("  - High (3-5%): Potential alignment issues or genuine hotspot\n\n")
cat("- **Mean Density**: Average density across genes (affected by extreme values)\n")
cat("  - In chromosome summaries: Average across all genes on that chromosome\n")
cat("  - In chromosomal impact: Average density of removed genes only\n")
cat("  - Use to validate filtering: Removed genes should have much higher mean_density than chromosome median\n\n")
cat("**Filtering Approaches:**\n\n")
cat("- **Approach i (strict)**: Removes Extreme genes (top 1% CAAS density) entirely + Dubious genes with clusters\n")
cat("- **Approach ii (moderate)**: Keeps Extreme genes, only removes Dubious genes with clusters\n")
cat("- **Significant**: p < 0.05 from proportion test comparing original vs declustered CAAS counts\n")
cat("- **Extreme genes**: Top 1% by CAAS per alignment length (most likely alignment artifacts)\n")
cat("- **Dubious genes**: 3×IQR outliers (genes with unusually high CAAS density)\n")
cat("- **Clusters**: High-density CAAS regions (trains) detected by sliding window algorithm\n\n")
cat("**Validation Example:**\n")
cat("  Chromosome 7 overall: median_density = 1.8%\n")
cat("  Chromosome 7 removed genes: mean_density = 8.2%\n")
cat("  → Removed genes were 4.6× denser than typical genes (8.2 / 1.8)\n")
cat("  → Confirms they were true outliers, not over-aggressive filtering\n\n")
cat(sprintf("**Recommendation**: Use Approach %s for downstream analysis (retains %d CAAS, removes %.1f%%)\n",
            ifelse(sum(summary_stats_i$n_CAAS_declust) < sum(summary_stats_ii$n_CAAS_declust), 
                   "ii", "i"),
            ifelse(sum(summary_stats_i$n_CAAS_declust) < sum(summary_stats_ii$n_CAAS_declust),
                   sum(summary_stats_ii$n_CAAS_declust), sum(summary_stats_i$n_CAAS_declust)),
            ifelse(sum(summary_stats_i$n_CAAS_declust) < sum(summary_stats_ii$n_CAAS_declust),
                   100 * (sum(summary_stats_ii$n_CAAS_original) - sum(summary_stats_ii$n_CAAS_declust)) / 
                     sum(summary_stats_ii$n_CAAS_original),
                   100 * (sum(summary_stats_i$n_CAAS_original) - sum(summary_stats_i$n_CAAS_declust)) / 
                     sum(summary_stats_i$n_CAAS_original))))
```

## 8. Genomic Distribution Visualization
Now that all filtering has been applied and outliers identified, we can visualize the genomic distribution of CAAS density and assess which chromosomes were most affected by filtering.

```{r manhattan_helpers, message=FALSE, warning=FALSE}
compute_chr_layout <- function(df) {
  layout_df <- df %>%
    mutate(chr_num = str_replace(chr, "chr", "")) %>%
    mutate(chr_rank = case_when(
      chr_num %in% as.character(1:22) ~ as.numeric(chr_num),
      chr_num == "X" ~ 23,
      chr_num == "Y" ~ 24,
      chr_num %in% c("M", "MT") ~ 25,
      TRUE ~ 26
    )) %>%
    group_by(chr_num, chr_rank) %>%
    summarise(chr_max = max(pos, na.rm = TRUE), .groups = "drop") %>%
    arrange(chr_rank) %>%
    mutate(
      chr_start = lag(cumsum(chr_max), default = 0),
      chr_center = chr_start + chr_max / 2
    )
  layout_df %>% mutate(chr_order = chr_rank)
}

apply_chr_layout <- function(df, layout_df) {
  df %>%
    mutate(chr_num = str_replace(chr, "chr", "")) %>%
    left_join(layout_df %>% select(chr_num, chr_start), by = "chr_num") %>%
    mutate(BPcum = pos + chr_start)
}

ordered_chr_levels <- function(df) {
  base_levels <- c(as.character(1:22), "X", "Y", "M")
  chr_clean <- df %>%
    mutate(chr_num = str_replace(chr, "chr", "")) %>%
    pull(chr_num) %>%
    unique()
  extras <- setdiff(chr_clean, base_levels)
  c(base_levels, sort(extras))
}
```

### 8.1. Manhattan Plot
This section visualizes CAAS density across the genome, highlighting outlier genes (Extreme and Dubious) to assess if artifacts are clustered in specific genomic regions.

```{r manhattan_plot, message=FALSE, warning=FALSE, fig.height=8, fig.width=16, dev='png', dpi=320}
# Prepare data for Manhattan plot (genomic_info already loaded in Section 3)
manhattan_data <- df_all_annotated %>%
  left_join(genomic_info %>% select(gene, chr, start, end), 
            by = c("Gene" = "gene")) %>%
  filter(!is.na(chr)) %>%  # Remove genes without genomic coordinates
  mutate(
    # Clean chromosome names (remove 'chr' prefix for sorting)
    chr_num = str_replace(chr, "chr", ""),
    # Convert to numeric for autosomes, keep X/Y/M as is
    chr_order = case_when(
      chr_num %in% as.character(1:22) ~ as.numeric(chr_num),
      chr_num == "X" ~ 23,
      chr_num == "Y" ~ 24,
      chr_num == "M" ~ 25,
      TRUE ~ 26
    ),
    # Gene midpoint for plotting
    pos = (start + end) / 2,
    # Categorize genes
    Category = case_when(
      Extreme & Dubious ~ "Both",
      Extreme ~ "Extreme",
      Dubious ~ "Dubious",
      TRUE ~ "Normal"
    ),
    Category = factor(Category, levels = category_levels)
  ) %>%
  arrange(chr_order, pos)

# Per-chromosome baseline summary used by chromosomal-impact plots
chr_summary <- manhattan_data %>%
  mutate(chr_num = str_replace(chr, "chr", "")) %>%
  group_by(Trait, chr_num) %>%
  summarise(
    median_density = median(n_CAAS_per_length, na.rm = TRUE),
    mean_density = mean(n_CAAS_per_length, na.rm = TRUE),
    n_genes = n(),
    .groups = "drop"
  )

# Calculate cumulative chromosome positions for x-axis
chr_lengths <- manhattan_data %>%
  select(chr, pos) %>%
  compute_chr_layout()

# Add cumulative positions to data
manhattan_data <- apply_chr_layout(manhattan_data, chr_lengths)

# Axis data for chromosome labels
axis_data <- chr_lengths %>%
  select(chr_num, chr_center) %>%
  rename(center = chr_center)

# Create separate plots for each trait
manhattan_plots <- lapply(c("malignant_prevalence", "neoplasia_prevalence"), function(trait) {
  plot_data <- manhattan_data %>% filter(Trait == trait)
  
  p <- ggplot(plot_data, aes(x = BPcum, y = n_CAAS_per_length, color = Category)) +
    # Add vertical separators between chromosomes
    geom_vline(
      data = chr_lengths %>% filter(chr_order > 1),  # Skip first chromosome
      aes(xintercept = chr_start),
      color = "gray20",
      linetype = "dashed",
      linewidth = 0.3,
      alpha = 0.6
    ) +
    # Points
    geom_point(
      data = plot_data %>% filter(Category == "Normal"),
      alpha = 0.35, size = 1.5
    ) +
    geom_point(
      data = plot_data %>% filter(Category == "Dubious"),
      alpha = 0.8, size = 2.5
    ) +
    geom_point(
      data = plot_data %>% filter(Category == "Extreme"),
      alpha = 0.8, size = 2.5
    ) +
    geom_point(
      data = plot_data %>% filter(Category == "Both"),
      alpha = 1, size = 3, shape = 17  # Triangle for both
    ) +
    # Color scheme
    scale_color_manual(
      values = category_palette,
      breaks = category_levels,
      drop = FALSE
    ) +
    # Scales and labels
    scale_x_continuous(
      label = axis_data$chr_num,
      breaks = axis_data$center,
      expand = expansion(mult = c(0.005, 0.005))  # Small equal expansion on both sides
    ) +
    scale_y_log10(
      breaks = c(0.1, 0.5, 1, 2, 5, 10, 20, 50),
      labels = c("0.1", "0.5", "1", "2", "5", "10", "20", "50")
    ) +
    labs(
      title = paste0("Genomic Distribution of CAAS Density: ", 
                     str_replace(trait, "_", " ") %>% str_to_title()),
      subtitle = "Extreme genes (salmon) removed only in Approach i; Dubious genes (green) with clusters removed in both approaches",
      x = "Chromosome",
      y = "CAAS per Gene Length (%, log scale)",
      color = "Gene Category",
      caption = paste0(
        "Points = genes colored by category. 'Extreme' = top 1% density; ",
        "'Dubious' = 3×IQR outliers. Extreme genes are removed in Approach i only; ",
        "Dubious genes with clusters are removed in both approaches. Use this plot to ",
        "inspect genomic clustering of high-density genes."
      )
    ) +
    theme_minimal(base_size = 11) +
    theme(
      axis.text.x = element_text(angle = 0, hjust = 0.5, size = 9),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank(),
      legend.position = "top",
      legend.box = "horizontal",
      # Add some padding to axes titles
      axis.title.x = element_text(margin = margin(t = 10)),
      axis.title.y = element_text(margin = margin(r = 10))
    )
  
  return(p)
})

# Display plots
manhattan_plots[[1]]
manhattan_plots[[2]]

# Save plots
ggsave(
  filename = file.path(outliers_dir, "manhattan_malignant_prevalence.png"),
  plot = manhattan_plots[[1]],
  width = 16, 
  height = 8, 
  dpi = 320, 
  dev = "png"
)

ggsave(
  filename = file.path(outliers_dir, "manhattan_neoplasia_prevalence.png"),
  plot = manhattan_plots[[2]],
  width = 16, 
  height = 8, 
  dpi = 320, 
  dev = "png"
)
```

# Generate per-criteria Manhattan plots
This section creates Manhattan plots showing the CAAS density after removing positions based on each declustering parameter, allowing visualization of the impact of each filtering criterion.
```{r manhattan_by_parameters_windows, message=FALSE, warning=FALSE, fig.height=15, fig.width=18, dev='png', dpi=320}
param_cols <- names(param_names)
traits <- c('malignant_prevalence','neoplasia_prevalence')
build_after_param <- function(trait, pname) {
  pattern <- param_names[[pname]]
  orig <- bind_rows(setup_dfs[grepl(trait, names(setup_dfs))], .id = 'src') %>%
    mutate(GenePos = paste(Gene, Position, sep = '@')) %>%
    distinct(GenePos, Gene, Position)

  matching <- cluster_dfs[grepl(pattern, names(cluster_dfs)) & grepl(trait, names(cluster_dfs))]
  discarded_pos <- character(0)
  if (length(matching) > 0) {
    discarded_pos <- unique(unlist(lapply(matching, function(df) {
      df %>%
        filter(ClusteringFlag == 'Discarded') %>%
        mutate(GenePos = paste(Gene, Position, sep='@')) %>%
        pull(GenePos)
    }), use.names = FALSE))
  }

  retained <- orig %>%
    mutate(discarded = GenePos %in% discarded_pos) %>%
    filter(!discarded) %>%
    group_by(Gene) %>%
    summarise(n_retained = n(), .groups = 'drop')

  gene_meta <- df_all_annotated %>%
    filter(Trait == trait) %>%
    distinct(Gene, .keep_all = TRUE) %>%
    select(Gene, Length, Extreme, Dubious, n_CAAS_per_length)

  trait_genes <- orig %>%
    distinct(Gene) %>%
    left_join(gene_meta, by = 'Gene') %>%
    mutate(
      Extreme = replace_na(Extreme, FALSE),
      Dubious = replace_na(Dubious, FALSE),
      Length = as.numeric(Length),
      n_CAAS_per_length = as.numeric(n_CAAS_per_length)
    ) %>%
    mutate(Category = case_when(
      Extreme & Dubious ~ 'Both',
      Extreme ~ 'Extreme',
      Dubious ~ 'Dubious',
      TRUE ~ 'Normal'
    )) %>%
    mutate(Category = factor(Category, levels = category_levels))

  df <- trait_genes %>%
    left_join(retained, by = 'Gene') %>%
    mutate(n_retained = replace_na(n_retained, 0)) %>%
    mutate(RemovalStatus = if_else(n_retained == 0, 'Removed by window', 'Retained')) %>%
    left_join(genomic_info %>% select(gene, chr, start, end), by = c('Gene' = 'gene')) %>%
    mutate(pos = ifelse(!is.na(start) & !is.na(end), (start + end)/2, NA_real_)) %>%
    mutate(
      density_filtered = ifelse(!is.na(Length) & Length > 0, 100 * n_retained / Length, NA_real_),
      density_original = n_CAAS_per_length,
      plot_density_original = case_when(
        is.na(density_original) ~ NA_real_,
        density_original <= 0 ~ min_density_display,
        TRUE ~ density_original
      ),
      plot_density_filtered = case_when(
        is.na(density_filtered) ~ NA_real_,
        density_filtered <= 0 ~ min_density_display,
        TRUE ~ density_filtered
      ),
      display_group = case_when(
        RemovalStatus == 'Removed by window' ~ 'Removed by window',
        RemovalStatus == 'Retained' ~ as.character(Category),
        TRUE ~ RemovalStatus
      ),
      display_group = factor(display_group, levels = manhattan_color_levels)
    ) %>%
    mutate(Parameter = pname, Trait = trait) %>%
    select(Trait, Parameter, Gene, chr, pos, n_retained, Length,
           density_original, density_filtered,
           plot_density_original, plot_density_filtered,
           Category, display_group, RemovalStatus)

  df
}

for (tr in traits) {
  manh_params <- map_dfr(param_cols, function(pn) build_after_param(tr, pn)) %>%
    filter(!is.na(chr) & !is.na(pos)) %>%
    mutate(Parameter = factor(Parameter, levels = param_cols))

  chr_layout <- manh_params %>%
    select(chr, pos) %>%
    compute_chr_layout()
  manh_params <- apply_chr_layout(manh_params, chr_layout)
  axis_data <- chr_layout %>% select(chr_num, chr_center)

  p <- ggplot(manh_params, aes(x = BPcum, y = plot_density_original,
                               color = display_group, shape = display_group)) +
    geom_vline(
      data = chr_layout %>% filter(chr_start > 0),
      aes(xintercept = chr_start),
      color = 'gray60', linetype = 'dashed', linewidth = 0.3, alpha = 0.7
    ) +
    geom_point(aes(alpha = display_group), size = 2, stroke = 0.8) +
    scale_alpha_manual(
      values = c(
        "Normal" = 0.15,
        "Dubious" = 0.85,
        "Extreme" = 0.85,
        "Both" = 0.85,
        "Removed by window" = 1,
        "Removed (Extreme)" = 1,
        "Removed (Dubious cluster)" = 1
      ),
      guide = "none"
    ) +
    scale_size_manual(
      values = c(
        "Normal" = 1.4,
        "Dubious" = 1.8,
        "Extreme" = 1.8,
        "Both" = 1.8,
        "Removed by window" = 2.6,
        "Removed (Extreme)" = 2.8,
        "Removed (Dubious cluster)" = 2.8
      ),
      guide = 'none'
    ) +
    scale_color_manual(values = manhattan_color_map,
                       breaks = manhattan_color_levels,
                       drop = FALSE) +
    scale_shape_manual(values = shape_map,
                       breaks = shape_levels,
                       drop = FALSE) +
    scale_x_continuous(labels = axis_data$chr_num,
                       breaks = axis_data$chr_center,
                       expand = expansion(mult = c(0.005, 0.005))) +
    scale_y_log10(
      breaks = c(0.01, 0.05, 0.1, 0.5, 1, 2, 5, 10),
      labels = c('0.01', '0.05', '0.1', '0.5', '1', '2', '5', '10'),
      expand = expansion(mult = c(0.02, 0.05))
    ) +
    facet_wrap(~ Parameter, nrow = 3, scales = 'fixed') +
    labs(
      title = sprintf('Manhattan: %s (per-parameter window filtering)', tr),
      x = 'Chromosome',
      y = 'CAAS per Gene Length (%)',
      color = 'Category / removal',
      shape = 'Category / removal',
      caption = 'Crosses denote genes where the window filter removed every CAAS (displayed at ~0.008%).'
    ) +
    theme_minimal(base_size = 14) +
    theme(
      legend.position = 'top',
      axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
      axis.ticks.x = element_line(size = 0.2)
    )

  print(p)

  out_file <- file.path(outliers_dir, sprintf('manhattan_parameters_%s.png', gsub('[^A-Za-z0-9]', '_', tolower(tr))))
  ggsave(filename = out_file, plot = p, width = 18, height = 14, dpi = 320)
  cat(sprintf('Saved per-parameter manhattan for %s -> %s\n', tr, out_file))
}
```

```{r manhattan_parameters_by_approach, message=FALSE, warning=FALSE, fig.height=8, fig.width=18, dev='png', dpi=320}
# For each approach (i, ii) and trait, plot per-parameter Manhattan after applying that approach's filtering
approaches <- c('Approach i', 'Approach ii')
param_cols <- names(param_names)
traits <- c('malignant_prevalence','neoplasia_prevalence')

apply_param_with_approach <- function(trait, pname, approach) {
  pattern <- param_names[[pname]]
  orig <- bind_rows(setup_dfs[grepl(trait, names(setup_dfs))], .id = 'src') %>%
    mutate(GenePos = paste(Gene, Position, sep = '@')) %>%
    distinct(GenePos, Gene, Position)

  matching <- cluster_dfs[grepl(pattern, names(cluster_dfs)) & grepl(trait, names(cluster_dfs))]
  discarded_pos <- character(0)
  if (length(matching) > 0) {
    discarded_pos <- unique(unlist(lapply(matching, function(df) {
      df %>%
        filter(ClusteringFlag == 'Discarded') %>%
        mutate(GenePos = paste(Gene, Position, sep='@')) %>%
        pull(GenePos)
    }), use.names = FALSE))
  }

  gene_meta <- df_all_annotated %>%
    filter(Trait == trait) %>%
    distinct(Gene, .keep_all = TRUE) %>%
    select(Gene, Length, Extreme, Dubious, n_CAAS_per_length)

  gene_flags <- gene_meta %>% select(Gene, Extreme, Dubious)

  dubious_with_clusters <- character(0)
  if (length(discarded_pos) > 0) {
    dubious_with_clusters <- orig %>%
      filter(GenePos %in% discarded_pos) %>%
      inner_join(gene_flags %>% filter(Dubious) %>% select(Gene), by = 'Gene') %>%
      distinct(Gene) %>%
      pull(Gene)
  }

  kept <- orig %>% mutate(discarded = GenePos %in% discarded_pos) %>% filter(!discarded)

  extreme_genes <- gene_flags %>% filter(Extreme) %>% pull(Gene)
  if (approach == 'Approach i') {
    kept <- kept %>% filter(!(Gene %in% extreme_genes))
  }
  if (length(dubious_with_clusters) > 0) {
    kept <- kept %>% filter(!(Gene %in% dubious_with_clusters))
  }

  retained <- kept %>% group_by(Gene) %>% summarise(n_retained = n(), .groups = 'drop')

  trait_genes <- orig %>%
    distinct(Gene) %>%
    left_join(gene_meta, by = 'Gene') %>%
    mutate(
      Extreme = replace_na(Extreme, FALSE),
      Dubious = replace_na(Dubious, FALSE),
      Length = as.numeric(Length),
      n_CAAS_per_length = as.numeric(n_CAAS_per_length)
    ) %>%
    mutate(Category = case_when(
      Extreme & Dubious ~ 'Both',
      Extreme ~ 'Extreme',
      Dubious ~ 'Dubious',
      TRUE ~ 'Normal'
    )) %>%
    mutate(Category = factor(Category, levels = category_levels))

  df <- trait_genes %>%
    left_join(retained, by = 'Gene') %>%
    mutate(n_retained = replace_na(n_retained, 0)) %>%
    mutate(RemovalStatus = case_when(
      approach == 'Approach i' & Gene %in% extreme_genes ~ 'Removed (Extreme)',
      Gene %in% dubious_with_clusters ~ 'Removed (Dubious cluster)',
      n_retained == 0 ~ 'Removed by window',
      TRUE ~ 'Retained'
    )) %>%
    left_join(genomic_info %>% select(gene, chr, start, end), by = c('Gene' = 'gene')) %>%
    mutate(pos = ifelse(!is.na(start) & !is.na(end), (start + end)/2, NA_real_)) %>%
    mutate(
      density_filtered = ifelse(!is.na(Length) & Length > 0, 100 * n_retained / Length, NA_real_),
      density_original = n_CAAS_per_length,
      plot_density_original = case_when(
        is.na(density_original) ~ NA_real_,
        density_original <= 0 ~ min_density_display,
        TRUE ~ density_original
      ),
      plot_density_filtered = case_when(
        is.na(density_filtered) ~ NA_real_,
        density_filtered <= 0 ~ min_density_display,
        TRUE ~ density_filtered
      ),
      display_group = case_when(
        RemovalStatus == 'Retained' ~ as.character(Category),
        TRUE ~ RemovalStatus
      ),
      display_group = factor(display_group, levels = manhattan_color_levels)
    ) %>%
    mutate(Parameter = pname, Trait = trait, Approach = approach) %>%
    select(Trait, Approach, Parameter, Gene, chr, pos, n_retained, Length,
           density_original, density_filtered,
           plot_density_original, plot_density_filtered,
           Category, display_group, RemovalStatus)

  df
}

for (approach in approaches) {
  for (tr in traits) {
    manh_df <- map_dfr(param_cols, function(pn) apply_param_with_approach(tr, pn, approach)) %>%
      filter(!is.na(chr) & !is.na(pos)) %>%
      mutate(Parameter = factor(Parameter, levels = param_cols))

    chr_layout <- manh_df %>% select(chr, pos) %>% compute_chr_layout()
    manh_df <- apply_chr_layout(manh_df, chr_layout)
    axis_data <- chr_layout %>% select(chr_num, chr_center)

    p <- ggplot(manh_df, aes(x = BPcum, y = plot_density_original,
                             color = display_group, shape = display_group)) +
      geom_vline(
        data = chr_layout %>% filter(chr_start > 0),
        aes(xintercept = chr_start),
        color = 'gray60', linetype = 'dashed', linewidth = 0.3, alpha = 0.7
      ) +
    geom_point(aes(alpha = display_group, size = display_group), stroke = 0.65) +
    scale_alpha_manual(
      values = c(
        "Normal" = 0.15,
        "Dubious" = 0.85,
        "Extreme" = 0.85,
        "Both" = 0.85,
        "Removed by window" = 1,
        "Removed (Extreme)" = 1,
        "Removed (Dubious cluster)" = 1
      ),
      guide = 'none'
    ) +
    scale_size_manual(
      values = c(
        "Normal" = 1.4,
        "Dubious" = 1.8,
        "Extreme" = 1.8,
        "Both" = 1.8,
        "Removed by window" = 2.6,
        "Removed (Extreme)" = 2.8,
        "Removed (Dubious cluster)" = 2.8
      ),
      guide = 'none'
    ) +
      scale_color_manual(values = manhattan_color_map,
                         breaks = manhattan_color_levels,
                         drop = FALSE) +
      scale_shape_manual(values = shape_map,
                         breaks = shape_levels,
                         drop = FALSE) +
      scale_x_continuous(labels = axis_data$chr_num,
                         breaks = axis_data$chr_center,
                         expand = expansion(mult = c(0.005, 0.005))) +
      scale_y_log10(
        breaks = c(0.01, 0.05, 0.1, 0.5, 1, 2, 5, 10),
        labels = c('0.01','0.05','0.1','0.5','1','2','5','10'),
        expand = expansion(mult = c(0.02, 0.05))
      ) +
      facet_wrap(~ Parameter, nrow = 3, scales = 'fixed') +
      labs(
        title = sprintf('Manhattan: %s — %s', tr, approach),
        x = 'Chromosome',
        y = 'CAAS per Gene Length (%)',
        color = 'Category / removal',
        shape = 'Category / removal',
        caption = "Crosses indicate genes removed by window (blue), Extreme (orange) or Dubious cluster (teal); placed at ~0.008%."
      ) +
      theme_minimal(base_size = 14) +
      theme(legend.position = 'top', axis.text.x = element_text(angle = 45, hjust = 1, size = 9))

    print(p)
    out_file <- file.path(outliers_dir, sprintf('manhattan_%s_%s.png', tolower(gsub(' ', '_', approach)), gsub('[^A-Za-z0-9]', '_', tolower(tr))))
    ggsave(filename = out_file, plot = p, width = 18, height = 14, dpi = 320)
    cat(sprintf('Saved %s -> %s\n', paste(tr, approach, sep=' - '), out_file))
  }
}
```

### 8.2. Chromosomal Impact Analysis

This section visualizes the impact of filtering on each chromosome by comparing the CAAS density of removed genes to the chromosome baseline.

```{r chr_impact, message=FALSE, warning=FALSE, fig.height=6, fig.width=12, dev='png', dpi=320}
cat("\n=== Combined Chromosomal Impact ===\n")
cat("Chromosomes most affected by filtering (extreme + dubious genes):\n")
cat("Note: mean_density here = average CAAS density of REMOVED genes only\n")
cat("      Compare to chromosome median_density in chromosome_summary.tsv\n")
cat("      to validate these were true outliers.\n\n")

# Combine extreme and all dubious genes by chromosome (keep Trait column)
all_removed_genes <- bind_rows(
  extreme_genes %>% 
    filter(chr != "unknown") %>%
    select(chr, Gene, Trait, n_CAAS_per_length) %>%
    mutate(Reason = "Extreme"),
  bind_rows(
    read_tsv(file.path(removed_dir, "malignant_prevalence_dubious_genes_with_clusters.tsv"), 
             show_col_types = FALSE) %>% mutate(Trait = "malignant_prevalence"),
    read_tsv(file.path(removed_dir, "neoplasia_prevalence_dubious_genes_with_clusters.tsv"), 
             show_col_types = FALSE) %>% mutate(Trait = "neoplasia_prevalence")
  ) %>%
    filter(chr != "unknown") %>%
    select(chr, Gene, Trait, n_CAAS_per_length) %>%
    mutate(Reason = "Dubious")
)

# Group by BOTH chr and Trait
chr_impact_by_trait <- all_removed_genes %>%
  group_by(chr, Trait) %>%
  summarise(
    n_genes = n(),
    n_extreme = sum(Reason == "Extreme"),
    n_dubious = sum(Reason == "Dubious"),
    mean_density = mean(n_CAAS_per_length, na.rm = TRUE),
    .groups = "drop"
  )

# Also create combined (across traits) for saving
chr_impact_combined <- all_removed_genes %>%
  group_by(chr) %>%
  summarise(
    n_genes = n(),
    n_extreme = sum(Reason == "Extreme"),
    n_dubious = sum(Reason == "Dubious"),
    mean_density = mean(n_CAAS_per_length, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(n_genes))

print(chr_impact_combined %>% head(10))

# Save combined impact
write_tsv(chr_impact_combined, file.path(removed_dir, "chromosomal_impact_summary.tsv"))
write_tsv(all_removed_genes, file.path(removed_dir, "all_removed_genes.tsv"))

cat("\nFiles saved:\n")
cat("  - chromosomal_impact_summary.tsv (per-chromosome statistics)\n")
cat("  - all_removed_genes.tsv (complete list with reasons)\n")

# Function to create plot for one trait (chr_summary already loaded above)
create_chr_impact_plot <- function(trait_name, trait_label) {
  # Get top 15 chromosomes for this trait
  trait_data <- chr_impact_by_trait %>%
    filter(Trait == trait_name) %>%
    arrange(desc(n_genes)) %>%
    head(15) %>%
    mutate(chr_num = gsub("^chr", "", chr))  # Strip "chr" prefix
  
  # Merge with baseline from chr_summary (already loaded above)
  trait_baseline <- chr_summary %>%
    filter(Trait == trait_name) %>%
    select(chr_num, median_density) %>%
    rename(population_median_density = median_density)
  
  plot_data <- trait_data %>%
    left_join(trait_baseline, by = "chr_num") %>%
    mutate(
      fold_enrichment = mean_density / population_median_density,
      absolute_diff = mean_density - population_median_density,
      is_significant = !is.na(population_median_density) & fold_enrichment > 2 & absolute_diff > 1
    )
  
  cat(sprintf("\n=== %s: Chr Impact Plot Data ===\n", trait_label))
  print(plot_data %>% select(chr, n_genes, mean_density, population_median_density, fold_enrichment, is_significant))
  
  # Create plot
  p <- ggplot(plot_data, aes(x = reorder(chr, -n_genes))) +
    # Mean density of removed genes (main bars)
    geom_col(aes(y = mean_density, fill = n_genes), 
             color = "black", size = 0.3, alpha = 0.85, width = 0.7) +
    # Chromosome median density as thick horizontal lines (baseline)
    geom_segment(data = plot_data %>% filter(!is.na(population_median_density)),
                 aes(x = as.numeric(factor(reorder(chr, -n_genes))) - 0.4,
                     xend = as.numeric(factor(reorder(chr, -n_genes))) + 0.4,
                     y = population_median_density,
                     yend = population_median_density),
                 color = "black", size = 2, lineend = "butt") +
    # Gene count labels
    geom_text(aes(y = mean_density, label = n_genes), 
              vjust = -0.5, size = 3.5, fontface = "bold") +
    scale_fill_gradient(low = "skyblue", high = "salmon4",
                        name = "Genes\nRemoved") +
    labs(
      title = sprintf("Chromosomal Impact: %s", trait_label),
      subtitle = "Colored bars = mean density of removed genes | Black lines = chromosome baseline\nTop 15 chromosomes by number of removed genes",
      x = "Chromosome",
      y = "CAAS Density (% per 100 AA)",
      caption = paste0(
        "Bars = mean CAAS density of genes REMOVED by filtering (Extreme and Dubious). ",
        "Black lines = chromosome median density (baseline). A large gap between a bar \nand its line indicates that removed genes are much denser than typical genes on that chromosome. ",
        "\nThis plot helps validate whether filtering disproportionately \naffects specific chromosomes or reflects true outliers."
      )
    ) +
    theme_minimal(base_size = 11) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.caption = element_text(size = 9, hjust = 0, face = "italic"),
      legend.position = "right"
    )
  
  # Add significance asterisks (no brackets)
  if (any(plot_data$is_significant, na.rm = TRUE)) {
    sig_data <- plot_data %>%
      filter(is_significant) %>%
      mutate(
        x_pos = as.numeric(factor(reorder(chr, -n_genes))),
        y_pos = mean_density * 1.05
      )
    
    cat(sprintf("=== %s: Significant chromosomes (n=%d) ===\n", trait_label, nrow(sig_data)))
    print(sig_data %>% select(chr, mean_density, population_median_density, fold_enrichment))
    
    p <- p + geom_text(data = sig_data,
                       aes(x = x_pos, y = y_pos + 0.5, label = "*"),
                       vjust = 0, size = 6, fontface = "bold", color = "black")
  }
  
  return(p)
}

# Create plots for both traits
p_chr_impact_malignant <- create_chr_impact_plot("malignant_prevalence", "Malignant Prevalence")
p_chr_impact_neoplasia <- create_chr_impact_plot("neoplasia_prevalence", "Neoplasia Prevalence")

print(p_chr_impact_malignant)
print(p_chr_impact_neoplasia)

# Save plots
ggsave(
  filename = file.path(removed_dir, "chromosomal_impact_malignant_prevalence.png"),
  plot = p_chr_impact_malignant,
  width = 12, height = 6, dpi = 320, dev = "png"
)

ggsave(
  filename = file.path(removed_dir, "chromosomal_impact_neoplasia_prevalence.png"),
  plot = p_chr_impact_neoplasia,
  width = 12, height = 6, dpi = 320, dev = "png"
)

```

## 9. BG cleanup

Now that we have the declustered datasets, we can also create cleaned-up background datasets by removing the same genes/positions from the original background CAAS data.

```{r bg_cleanup, message=FALSE, warning=FALSE, fig.height=6, fig.width=10, dev='png', dpi=320}
# We will be using all_removed_genes from above in order to remove same genes from background CAAS data
cat("\n=== Background CAAS Cleanup ===\n")
# First load original background CAAS data
bg_dir <- file.path(resultsDir, "2.CAAS/1.Discovery/2.CAAS_background")

bg_caas_files <- list.files(
  path = bg_dir,
  pattern = "global_background",
  full.names = TRUE,
  recursive = TRUE
)

# Load background CAAS data
bg_caas_dfs <- lapply(bg_caas_files, function(file) {
  trait_name <- ifelse(grepl("malignant_prevalence", file), "malignant_prevalence", "neoplasia_prevalence")
  df <- read_tsv(file, col_names = FALSE) %>%
    dplyr::rename(Gene = X1) %>%
    mutate(Trait = trait_name)
  return(df)
})

# Name the list elements
names(bg_caas_dfs) <- sapply(bg_caas_files, function(file) {
  trait_name <- ifelse(grepl("malignant_prevalence", file), "malignant_prevalence", "neoplasia_prevalence")
  return(paste0("bg_caas_", trait_name))
})

# Clean background CAAS data by removing same genes/positions as in all_removed_genes (by Trait)
clean_bg_caas <- function(trait_name) {
  bg_key <- paste0("bg_caas_", trait_name)
  if (!bg_key %in% names(bg_caas_dfs)) {
    stop("No background CAAS dataset found for trait: ", trait_name)
  }
  if (!is.data.frame(all_removed_genes)) {
    stop("all_removed_genes is not a data.frame/tibble. Actual class: ",
         paste(class(all_removed_genes), collapse = ", "))
  }

  # Required columns check (prevents silent fallback to env variables)
  needed_cols <- c("Trait", "Gene")

  bg_df <- bg_caas_dfs[[bg_key]]

  removed_pairs <- all_removed_genes %>%
    filter(.data$Trait == trait_name) %>%
    distinct(.data$Gene)

  cleaned_df <- bg_df %>%
    anti_join(removed_pairs, by = "Gene")

  cleaned_df
}

cleaned_bg_malignant <- clean_bg_caas("malignant_prevalence")
cleaned_bg_neoplasia <- clean_bg_caas("neoplasia_prevalence")

# Rewrite the gene lists to files
write.table(
  cleaned_bg_malignant$Gene,
  file.path(bg_dir, "malignant_prevalence/cleaned_background_malignant_prevalence.txt"),
  col.names = FALSE,
  row.names = FALSE,
  quote = FALSE,
  sep = "\n"
)

write.table(
  cleaned_bg_neoplasia$Gene,
  file.path(bg_dir, "neoplasia_prevalence/cleaned_background_neoplasia_prevalence.txt"),
  col.names = FALSE,
  row.names = FALSE,
  quote = FALSE,
  sep = "\n"
)

```

## Session Information

```{r session_info}
sessionInfo()
```
