---
title: "CAAS Signification and Organization for Downstream Analysis"
author: "Miguel Ramon"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

# CAAS Signification and Organization for Downstream Analysis

This script organizes CAAS (Context-Aware Amino Acid Substitutions) discovery results for downstream analyses (enrichment, meta-analysis, etc.). It applies **filtering criteria**—not statistical tests—to create organized subsets suitable for different analytical approaches.

**Purpose**: Prepare filtered CAAS datasets for downstream analysis

- **Hypergeometric p-values** (Hyp_05): Filter by biological variability/pattern strength
- **Permulation p-values** (Perm_05): Filter by strength of trait association
- **Combined filters** (Perm_Hyp_05): Strictest criteria for high-confidence CAAS

**Key Outputs:**

- Filtered CAAS tables organized by contrast, pattern, and filtering criteria
- Pattern-specific CAAS subsets (pattern1, pattern2, pattern3, and combinations)
- Gene lists for downstream enrichment analysis (e.g., GO, pathway)
- Summary statistics and visualizations
- Cross-contrast commonality tables for meta-analysis

**Note**: The actual downstream aggregation and statistical testing occurs in separate scripts. This script focuses on organizing and filtering discovery-stage CAAS calls.

## Setup and Directory Configuration

This section configures the working environment, sets directories, and loads necessary functions and libraries.

```{r setup, message=FALSE, warning=FALSE}
# Source commons.R for setup
source("./functions/commons.R")
setup_rmd()

# Load additional libraries specific to this analysis
library(reshape2)
library(tibble)
library(stringr)
library(ggplot2)
```

## 1. Data Import and Quality Control Filtering

This section loads CAAS discovery results from CAASTools and applies initial quality control filters.

**Quality Filters Applied:**
- **Gap filtering** (GFG=0, GBG=0): Safety check ensuring no-gap constraint from discovery phase
- **P-value filtering** (p≤0.05): Retains CAAS meeting nominal variability threshold

Note: These are filtering criteria for dataset organization, not statistical hypothesis tests.

```{r data_import, message=FALSE, warning=FALSE}
# Define input directory
caas_folder <- file.path(resultsDir, "2.CAAS/3.CAAS_postproc/4.Declustered_i/")

# Locate all CAAS files recursively
caas_files <- list.files(
  path = caas_folder, 
  pattern = "\\.tsv$", 
  full.names = TRUE, 
  recursive = TRUE
)

if (length(caas_files) == 0) {
  stop("No CAAS files found in ", caas_folder)
}

# Read all CAAS files into list
caas_raw <- lapply(caas_files, read_tsv, show_col_types = FALSE)
names(caas_raw) <- gsub("_sp.tab.caas_declustered.tsv", "", basename(caas_files))

# Summary of loaded data
cat("\n=== Data Import Summary ===\n")
cat(sprintf("Total CAAS files found: %d\n", length(caas_files)))
cat(sprintf("Total CAAS positions loaded: %s\n", format(sum(sapply(caas_raw, nrow)), big.mark = ",")))
cat("\nCAAS by contrast:\n")
print(data.frame(
  Contrast = names(caas_raw),
  Positions = sapply(caas_raw, nrow)
) %>% head(10))
if (length(caas_raw) > 10) {
  cat(sprintf("... and %d more contrasts\n", length(caas_raw) - 10))
}
```

```{r initial_filtering, message=FALSE, warning=FALSE}
# Apply gap filtering as quality control
# GFG (Gaps Foreground Group) and GBG (Gaps Background Group) from CAASTools
# This is a safety check: if no-gap constraint was enforced during discovery,
# this filter will be redundant (all values = 0)
filter_gaps <- function(df) {
  if (!all(c("GFG", "GBG") %in% colnames(df))) {
    warning("GFG/GBG columns not found in CAASTools output; skipping gap filter")
    return(df)
  }
  df %>% 
    filter(GFG == 0 & GBG == 0) %>%
    # Add GenePos identifier for merging later
    mutate(GenePos = paste0(Gene, "@", Position))
}

caas_gapless <- lapply(caas_raw, filter_gaps)
```

## 2. Permulation Test Integration

This section integrates permulation test results (bootstrap p-values) with CAAS calls. Permulation p-values assess **strength of association** between amino acid patterns and traits.

**Key Distinction:**
- **Hypergeometric p-values** (from CAASTools): Measure pattern variability/strength
- **Permulation p-values** (from RERConverge style "permulations"): Measure trait association strength

Both are used as **filtering criteria** (not statistical tests) to create different CAAS subsets for downstream analysis.

```{r bootstrap_import, message=FALSE, warning=FALSE}
boot_folder <- file.path(resultsDir, "2.CAAS/2.Results")

# Locate bootstrap CAAS files
boot_files <- list.files(
  path = boot_folder, 
  pattern = "\\.tab.caas$", 
  full.names = TRUE, 
  recursive = TRUE
)

boot_files <- boot_files[grepl("/bootstrap/", boot_files, ignore.case = TRUE)]

if (length(boot_files) == 0) {
  stop("No bootstrap files found in ", boot_folder)
}

# Read bootstrap data (no column names in original format)
boot_raw <- lapply(boot_files, read_tsv, col_names = FALSE, show_col_types = FALSE)
names(boot_raw) <- gsub("_sp.tab.caas", "", basename(boot_files))

cat("\n=== Permutation Test Data Import ===\n")
cat(sprintf("Bootstrap files loaded: %d\n", length(boot_files)))
cat(sprintf("Total positions tested: %s\n", format(sum(sapply(boot_raw, nrow)), big.mark = ",")))
```

```{r sig_processing, message=FALSE, warning=FALSE}
# Process permulation test p-values
process_bootstrap <- function(df) {
  df %>%
    as.data.frame() %>%
    # Correct zero p-values using pseudo-count: 1/(permulations+1)
    mutate(X4 = ifelse(X2 == 0, 1 / (X3 + 1), X4)) %>%
    rename(
      GenePos = X1,
      Counts = X2,
      Perms = X3,
      Pvalue.boot = X4
    ) %>%
    # Apply Benjamini-Hochberg correction
    mutate(Pvalue.boot.BH = p.adjust(Pvalue.boot, method = "BH"))
}

boot_processed <- lapply(boot_raw, process_bootstrap)
names(boot_processed) <- names(caas_gapless)

# Define p-value threshold
P_VALUE_THRESHOLD <- 0.05

# Merge permutation p-values with CAAS data early
# This ensures all downstream objects have Pvalue.boot available
caas_filtered <- lapply(names(caas_gapless), function(name) {
  caas_gapless[[name]] %>%
    left_join(
      boot_processed[[name]] %>% select(GenePos, Pvalue.boot),
      by = "GenePos"
    ) %>%
    # Add isSignificant flag: TRUE if BOTH p-values ≤ 0.05
    mutate(
      isSignificant = !is.na(Pvalue.boot) & 
                      as.numeric(Pvalue) <= P_VALUE_THRESHOLD & 
                      Pvalue.boot <= P_VALUE_THRESHOLD
    )
})
names(caas_filtered) <- names(boot_processed)

# Apply hypergeometric p-value filter
# This filters CAAS by pattern variability, NOT as a statistical test
# Used as criterion for downstream analysis inclusion
filter_pvalue <- function(df, threshold = P_VALUE_THRESHOLD) {
  df %>% filter(as.numeric(Pvalue) <= threshold)
}

caas_pval05 <- lapply(caas_filtered, filter_pvalue)

# Apply permutation p-value filter
boot_pval05 <- lapply(caas_filtered, function(df) {
  df %>% filter(!is.na(Pvalue.boot) & Pvalue.boot <= P_VALUE_THRESHOLD)
})

cat("\n=== Significance Processing Summary ===\n")
cat(sprintf("P-value threshold: %.2f\n\n", P_VALUE_THRESHOLD))
cat("Positions passing filtering criteria:\n")
cat(sprintf("  - Variability check (hypergeometric p ≤ %.2f): %s\n",
            P_VALUE_THRESHOLD,
            format(sum(sapply(caas_pval05, nrow)), big.mark = ",")))
cat(sprintf("  - Association check (permutation p ≤ %.2f): %s\n",
            P_VALUE_THRESHOLD,
            format(sum(sapply(boot_pval05, nrow)), big.mark = ",")))
cat(sprintf("  - Both criteria (isSignificant=TRUE): %s\n",
            format(sum(sapply(caas_filtered, function(df) sum(df$isSignificant, na.rm = TRUE))), big.mark = ",")))
cat(sprintf("\nPercentages relative to gap-filtered positions:\n"))
cat(sprintf("  - Variability: %.1f%%\n",
            100 * sum(sapply(caas_pval05, nrow)) / sum(sapply(caas_filtered, nrow))))
cat(sprintf("  - Association: %.1f%%\n",
            100 * sum(sapply(boot_pval05, nrow)) / sum(sapply(caas_filtered, nrow))))
cat(sprintf("  - Both: %.1f%%\n",
            100 * sum(sapply(caas_filtered, function(df) sum(df$isSignificant, na.rm = TRUE))) / sum(sapply(caas_filtered, nrow))))


# Apply hypergeometric p-value filter
# This filters CAAS by pattern variability, NOT as a statistical test
# Used as criterion for downstream analysis inclusion
P_VALUE_THRESHOLD <- 0.05

filter_pvalue <- function(df, threshold = P_VALUE_THRESHOLD) {
  df %>% filter(as.numeric(Pvalue) <= threshold)
}

caas_pval05 <- lapply(caas_filtered, filter_pvalue)


# Summary of filtering results
cat("\n=== Variability Control Filtering Summary ===\n")
cat(sprintf("P-value threshold: %.2f\n\n", P_VALUE_THRESHOLD))

filtering_summary <- data.frame(
  Stage = c("Raw CAAS", "After gap filter", "After hypergeometric filter", "After permutation filter", "Significant CAAS (both filters)"),
  Total_Positions = c(
    sum(sapply(caas_raw, nrow)),
    sum(sapply(caas_filtered, nrow)),
    sum(sapply(caas_pval05, nrow)),
    sum(sapply(boot_pval05, nrow)),
    sum(sapply(caas_filtered, function(df) sum(df$isSignificant, na.rm = TRUE)))
  ),
  Retained_Pct = c(
    100,
    round(100 * sum(sapply(caas_filtered, nrow)) / sum(sapply(caas_raw, nrow)), 1),
    round(100 * sum(sapply(caas_pval05, nrow)) / sum(sapply(caas_filtered, nrow)), 1),
    round(100 * sum(sapply(boot_pval05, nrow)) / sum(sapply(caas_filtered, nrow)), 1),
    round(100 * sum(sapply(caas_filtered, function(df) sum(df$isSignificant, na.rm = TRUE))) / sum(sapply(caas_filtered, nrow)), 1)
  )
)

print(filtering_summary)

cat(sprintf("\nSignificant positions (both filters): %s (%.1f%% of gap-filtered)\n",
            format(sum(sapply(caas_filtered, function(df) sum(df$isSignificant, na.rm = TRUE))), big.mark = ","),
            100 * sum(sapply(caas_filtered, function(df) sum(df$isSignificant, na.rm = TRUE))) / sum(sapply(caas_filtered, nrow))))
```

```{r merge_bootstrap, message=FALSE, warning=FALSE}
# Create four filtering strategy variants for downstream analysis:
# 1. caas_filtered: All gap-filtered CAAS (already has Pvalue.boot and isSignificant)
# 2. caas_pval05: Hypergeometric filter only
# 3. caas_boot (boot_pval05): Permutation filter only  
# 4. caas_pval05_boot: Combined filters (strictest - these have isSignificant=TRUE)

# caas_boot is just an alias for boot_pval05 (already created above)
caas_boot <- boot_pval05

# Combined filter: positions that pass BOTH filters (isSignificant=TRUE)
caas_pval05_boot <- lapply(caas_filtered, function(df) {
  df %>% filter(isSignificant == TRUE)
})

# Summary table of filtering strategies
cat("\n=== Filtering Strategy Comparison ===\n")
cat("Creating four CAAS subsets based on different filtering criteria:\n\n")

strategy_summary <- data.frame(
  Strategy = c("All (gap-filtered)", "Hypergeometric only", "Permutation only", "Combined (strictest)"),
  Description = c(
    "Quality control only",
    "Pattern variability filter",
    "Association strength filter",
    "Both filters applied"
  ),
  Positions = c(
    sum(sapply(caas_filtered, nrow)),
    sum(sapply(caas_pval05, nrow)),
    sum(sapply(caas_boot, nrow)),
    sum(sapply(caas_pval05_boot, nrow))
  )
)

print(strategy_summary)

cat(sprintf("\nStrictestfilter retains %.1f%% of hypergeometric-filtered positions\n",
            100 * sum(sapply(caas_pval05_boot, nrow)) / sum(sapply(caas_pval05, nrow))))
```

## 3. Pattern-Based Filtering

This section extracts CAAS by pattern type and creates combined pattern sets.
Each pattern represent (as defined by CAASTools):
- **pattern1**: Foreground and background groups have distinct amino acids, shared within groups
- **pattern2**: Foreground group shares the same amino acid; background group has distinct amino acids
- **pattern3**: Foreground group has distinct amino acids; background group shares the same amino acid
(Easter egg: pattern 4 is not considered as it represents no convergence, foreground and background share no amin
```{r pattern_filtering, message=FALSE, warning=FALSE}
# Unified pattern extraction function
extract_patterns <- function(df_list, patterns = c("pattern1", "pattern2", "pattern3")) {
  result <- list()
  
  # Extract individual patterns
  for (pattern in patterns) {
    result[[pattern]] <- lapply(names(df_list), function(name) {
      df_list[[name]] %>% filter(Pattern == pattern)
    })
    names(result[[pattern]]) <- names(df_list)
  }
  
  # Create combined patterns
  result[["pattern12"]] <- lapply(names(result[["pattern1"]]), function(name) {
    bind_rows(result[["pattern1"]][[name]], result[["pattern2"]][[name]])
  })
  names(result[["pattern12"]]) <- names(df_list)
  
  result[["pattern13"]] <- lapply(names(result[["pattern1"]]), function(name) {
    bind_rows(result[["pattern1"]][[name]], result[["pattern3"]][[name]])
  })
  names(result[["pattern13"]]) <- names(df_list)
  
  return(result)
}

# Extract patterns for all four CAAS variants
patterns_raw <- extract_patterns(caas_filtered)
patterns_pval05 <- extract_patterns(caas_pval05)
patterns_boot <- extract_patterns(caas_boot)
patterns_boot_pval05 <- extract_patterns(caas_pval05_boot)

# Pattern distribution summary
cat("\n=== Pattern Distribution Summary ===\n")
cat("CAAS patterns represent different types of amino acid convergence:\n")
cat("  - Pattern1: Distinct amino acids, shared within foreground/background\n")
cat("  - Pattern2: Foreground fixed, background variable\n")
cat("  - Pattern3: Foreground variable, background fixed\n")
cat("  - Pattern12/13: Combined pattern sets\n\n")

pattern_counts <- data.frame(
  Pattern = c("Pattern1", "Pattern2", "Pattern3", "Pattern12", "Pattern13"),
  All = c(
    sum(sapply(patterns_raw$pattern1, nrow)),
    sum(sapply(patterns_raw$pattern2, nrow)),
    sum(sapply(patterns_raw$pattern3, nrow)),
    sum(sapply(patterns_raw$pattern12, nrow)),
    sum(sapply(patterns_raw$pattern13, nrow))
  ),
  Hyp_05 = c(
    sum(sapply(patterns_pval05$pattern1, nrow)),
    sum(sapply(patterns_pval05$pattern2, nrow)),
    sum(sapply(patterns_pval05$pattern3, nrow)),
    sum(sapply(patterns_pval05$pattern12, nrow)),
    sum(sapply(patterns_pval05$pattern13, nrow))
  ),
  Combined = c(
    sum(sapply(patterns_boot_pval05$pattern1, nrow)),
    sum(sapply(patterns_boot_pval05$pattern2, nrow)),
    sum(sapply(patterns_boot_pval05$pattern3, nrow)),
    sum(sapply(patterns_boot_pval05$pattern12, nrow)),
    sum(sapply(patterns_boot_pval05$pattern13, nrow))
  )
)

print(pattern_counts)
```

## 4. Export Organized CAAS Tables

This section saves filtered CAAS tables in organized directory structures for downstream analyses (enrichment, meta-analysis, etc.). Each filtering strategy produces separate output files to support different analytical approaches.

```{r export_tables, message=FALSE, warning=FALSE}
# Set output directory
caas_list_dir <- file.path(resultsDir, "2.CAAS/3.5.CAAS_exploration")
createDir(caas_list_dir)

# Unified file export function
export_caas_tables <- function(df, name, filter_type, base_dir = caas_list_dir, subdir = "") {
  # Parse trait and scenario from name
  trait_name <- paste(strsplit(name, "_")[[1]][1:2], collapse = "_")
  file_name <- paste(strsplit(name, "_")[[1]][3:4], collapse = "_")
  
  # Create output path
  path <- file.path(base_dir, "1.CAAS_tables_filtered", trait_name, file_name, subdir)
  createDir(path)
  
  # Write file
  output_file <- file.path(path, paste0(filter_type, "_", file_name, ".output"))
  write.table(
    df, 
    output_file,
    quote = FALSE,
    row.names = FALSE,
    col.names = TRUE,
    sep = "\t"
  )
  
  return(df)
}

# Save all pattern variants
save_pattern_tables <- function(pattern_list, suffix, subdir = "") {
  results <- list()
  for (pattern_name in names(pattern_list)) {
    pattern_data <- pattern_list[[pattern_name]]
    file_prefix <- paste0(pattern_name, suffix)
    
    results[[pattern_name]] <- lapply(names(pattern_data), function(name) {
      export_caas_tables(pattern_data[[name]], name, file_prefix, subdir = subdir)
    })
    names(results[[pattern_name]]) <- names(pattern_data)
  }
  return(results)
}

# Save raw (unfiltered by p-value) patterns
saved_raw <- save_pattern_tables(patterns_raw, "")

# Save p-value filtered patterns
saved_pval05 <- save_pattern_tables(patterns_pval05, "_05")

# Save bootstrap filtered patterns
saved_boot <- save_pattern_tables(patterns_boot, "_boot", subdir = "boot")

# Save bootstrap + p-value filtered patterns
saved_boot_pval05 <- save_pattern_tables(patterns_boot_pval05, "_boot05", subdir = "boot")

# Export summary
cat("\n=== CAAS Table Export Summary ===\n")
cat(sprintf("Output directory: %s\n", caas_list_dir))
cat("\nFiles exported by filtering strategy:\n")

export_summary <- data.frame(
  Strategy = c("Raw patterns", "Hypergeometric filtered", "Permutation filtered", "Combined filtered"),
  Files_per_pattern = c(
    length(patterns_raw$pattern1),
    length(patterns_pval05$pattern1),
    length(patterns_boot$pattern1),
    length(patterns_boot_pval05$pattern1)
  ),
  Total_patterns = c(5, 5, 5, 5),
  Total_files = c(
    length(patterns_raw$pattern1) * 5,
    length(patterns_pval05$pattern1) * 5,
    length(patterns_boot$pattern1) * 5,
    length(patterns_boot_pval05$pattern1) * 5
  )
)

print(export_summary)
cat(sprintf("\nTotal files exported: %d\n", sum(export_summary$Total_files)))
```

## 5. Summary Statistics

This section computes summary counts and creates visualization-ready tables.

```{r summary_table, message=FALSE, warning=FALSE}
# Create comprehensive summary table
create_summary_table <- function(
  raw_list, pval05_list, boot_list, boot_pval05_list,
  pattern_raw, pattern_pval05, pattern_boot, pattern_boot_pval05
) {
  summary_df <- data.frame(
    File = names(raw_list),
    All = sapply(raw_list, nrow),
    Perm_05 = sapply(boot_list, nrow),
    Hyp_05 = sapply(pval05_list, nrow),
    Perm_Hyp_05 = sapply(boot_pval05_list, nrow)
  )
  
  # Add pattern counts
  for (pattern in c("pattern1", "pattern2", "pattern3", "pattern12", "pattern13")) {
    prefix <- gsub("pattern", "Pattern", pattern)
    summary_df[[paste0(prefix, "_All")]] <- sapply(pattern_raw[[pattern]], nrow)
    summary_df[[paste0(prefix, "_Perm_05")]] <- sapply(pattern_boot[[pattern]], nrow)
    summary_df[[paste0(prefix, "_Hyp_05")]] <- sapply(pattern_pval05[[pattern]], nrow)
    summary_df[[paste0(prefix, "_Perm_Hyp_05")]] <- sapply(pattern_boot_pval05[[pattern]], nrow)
  }
  
  return(summary_df)
}

summary_table <- create_summary_table(
  caas_filtered, caas_pval05, caas_boot, caas_pval05_boot,
  patterns_raw, patterns_pval05, patterns_boot, patterns_boot_pval05
)

# Add totals row
totals <- c("Total", colSums(summary_table[, -1]))
summary_table <- rbind(summary_table, totals)

# Add species pair information
malignant_pairs_file <- file.path(
  resultsDir, 
  "2.CAAS/1.Discovery/1.Traitfiles/malignant_prevalence/malignant_prevalence_paired_species.tab"
)
neoplasia_pairs_file <- file.path(
  resultsDir,
  "2.CAAS/1.Discovery/1.Traitfiles/neoplasia_prevalence/neoplasia_prevalence_paired_species.tab"
)

# Function to read and format species pairs
format_species_pairs <- function(file_path) {
  read_tsv(file_path, show_col_types = FALSE) %>%
    select(species, pair, contrast) %>%
    unique() %>%
    group_by(pair, contrast) %>%
    summarize(sp_pair = paste(species, collapse = ","), .groups = "drop") %>%
    group_by(contrast) %>%
    summarize(sp_pair_display = paste(sp_pair, collapse = "\n"), .groups = "drop")
}

if (file.exists(malignant_pairs_file) && file.exists(neoplasia_pairs_file)) {
  malignant_pairs <- format_species_pairs(malignant_pairs_file) %>%
    mutate(source = "malignant_pairs")
  
  neoplasia_pairs <- format_species_pairs(neoplasia_pairs_file) %>%
    mutate(source = "neoplasia_pairs")
  
  # Combine and merge with summary
  lookup_table <- bind_rows(malignant_pairs, neoplasia_pairs) %>%
    mutate(file_name = paste0(source, "_", contrast))
  
  summary_table_annotated <- summary_table %>%
    rownames_to_column(var = "RowNum") %>%
    mutate(
      contrast_num = str_extract(File, "\\d+"),
      source = case_when(
        str_detect(File, "^malignant_prevalence_pair_") ~ "malignant_pairs",
        str_detect(File, "^neoplasia_prevalence_pair_") ~ "neoplasia_pairs",
        TRUE ~ NA_character_
      ),
      file_name = paste0(source, "_", contrast_num)
    ) %>%
    left_join(
      select(lookup_table, file_name, sp_pair_display),
      by = "file_name"
    ) %>%
    mutate(
      File = case_when(
        !is.na(source) ~ paste0(source, "_", contrast_num),
        TRUE ~ File
      )
    ) %>%
    select(sp_pair_display, everything(), -RowNum, -contrast_num, -file_name)
} else {
  summary_table_annotated <- summary_table
}

# Display summary (show first 10 rows if large)
cat("\n=== CAAS Summary Statistics ===\n")
if (nrow(summary_table_annotated) > 11) {
  cat("Showing first 10 contrasts and totals:\n\n")
  print(rbind(
    head(summary_table_annotated, 10),
    tail(summary_table_annotated, 1)
  ))
  cat(sprintf("\n... %d contrasts omitted ...\n", nrow(summary_table_annotated) - 11))
} else {
  print(summary_table_annotated)
}

write.table(
  summary_table_annotated,
  file.path(caas_list_dir, "summary_table.txt"),
  row.names = FALSE,
  col.names = TRUE,
  quote = FALSE,
  sep = "\t"
)

cat(sprintf("\nFull summary table saved to: %s\n", 
            file.path(caas_list_dir, "summary_table.txt")))
```

## 6. Visualization

This section creates bar plots summarizing CAAS counts across contrasts and patterns.

```{r visualization, fig.width=16, fig.height=12, message=FALSE, warning=FALSE, dpi=320, dev="png"}
# Prepare plot data (exclude Total row)
plot_data <- summary_table_annotated %>%
  drop_na(sp_pair_display) %>%
  mutate(sp_pair_display = factor(sp_pair_display, levels = unique(sp_pair_display)))

# Define color palette
levels_palette <- c(
  "All" = "skyblue3", 
  "Hyp_05" = "salmon3", 
  "Perm_05" = "darkseagreen3", 
  "Perm_Hyp_05" = "tan1"
)

# Overall counts plot
overall_data <- plot_data %>%
  select(sp_pair_display, source, All, Hyp_05, Perm_05, Perm_Hyp_05) %>%
  pivot_longer(
    cols = c(All, Hyp_05, Perm_05, Perm_Hyp_05),
    names_to = "Category",
    values_to = "Count"
  ) %>%
  mutate(Count = as.numeric(Count))

p_overall <- ggplot(overall_data, aes(x = sp_pair_display, y = Count, fill = Category)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_grid(~ source, scales = "free_x") +
  theme_minimal() +
  labs(
    title = "CAAS Counts by Filtering Strategy",
    subtitle = "Hyp_05 = Hypergeometric (variability) | Perm_05 = Permulation (association) | Perm_Hyp_05 = Both",
    x = "Species Pair",
    y = "Count"
  ) +
  scale_fill_manual(values = levels_palette) +
  scale_y_continuous(breaks = seq(0, 11000, by = 500)) +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1, vjust = 1, size = 10),
    panel.grid.major = element_line(color = "darkgray", size = 0.5),
    panel.grid.minor = element_line(color = "lightgray", size = 0.5),
    axis.title.x = element_text(margin = margin(t = 20)),
    axis.title.y = element_text(margin = margin(r = 20))
  )

print(p_overall)

# Pattern-specific stacked plot
patterns <- c("1", "2", "3")
categories <- c("All", "Hyp_05", "Perm_05", "Perm_Hyp_05")

pattern_data_list <- list()

for (cat in categories) {
  suffix <- paste0("_", cat)
  pattern_cols <- paste0("Pattern", patterns, suffix)
  
  temp_data <- plot_data %>%
    select(sp_pair_display, source, all_of(pattern_cols))
  
  temp_long <- temp_data %>%
    pivot_longer(
      cols = all_of(pattern_cols),
      names_to = "Pattern",
      values_to = "Count"
    ) %>%
    mutate(
      Pattern = gsub("^Pattern", "", Pattern),
      Pattern = gsub(paste0(suffix, "$"), "", Pattern),
      Category = cat,
      Count = as.numeric(Count)
    )
  
  pattern_data_list[[cat]] <- temp_long
}

all_pattern_data <- bind_rows(pattern_data_list) %>%
  mutate(
    Category = factor(Category, levels = c("All", "Hyp_05", "Perm_05", "Perm_Hyp_05")),
    sp_pair_display = factor(sp_pair_display, levels = unique(plot_data$sp_pair_display))
  )

# Color palette for patterns
pattern_palette <- c(
  "1" = "skyblue3",
  "2" = "salmon3",
  "3" = "darkseagreen3"
)

p_stacked <- ggplot(all_pattern_data, aes(x = sp_pair_display, y = Count, fill = Pattern)) +
  geom_bar(stat = "identity") +
  facet_grid(Category ~ source, scales = "free") +
  theme_minimal() +
  labs(
    title = "CAAS Counts by Pattern Type and Filtering Strategy",
    x = "Hypothesis (Species Pair)",
    y = "Count",
    fill = "Pattern"
  ) +
  scale_fill_manual(values = pattern_palette) +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1, vjust = 1, size = 10),
    panel.grid.major = element_line(color = "darkgray", size = 0.5),
    panel.grid.minor = element_line(color = "lightgray", size = 0.5),
    axis.title.x = element_text(margin = margin(t = 20)),
    axis.title.y = element_text(margin = margin(r = 20))
  )

print(p_stacked)

# Save plots
ggsave(
  file.path(caas_list_dir, "overall_counts.png"),
  plot = p_overall,
  width = 15,
  height = 10,
  dpi = 320
)

ggsave(
  file.path(caas_list_dir, "stacked_counts.png"),
  plot = p_stacked,
  width = 10,
  height = 8,
  dpi = 320
)

cat("\n=== Visualization Export Summary ===\n")
cat(sprintf("Overall counts plot: %s\n", file.path(caas_list_dir, "overall_counts.png")))
cat(sprintf("Stacked patterns plot: %s\n", file.path(caas_list_dir, "stacked_counts.png")))
cat("\nPlots show CAAS distributions across:\n")
cat("  - Species pair contrasts\n")
cat("  - Filtering strategies (All, Hyp_05, Perm_05, Combined)\n")
cat("  - Pattern types (Pattern 1, 2, 3)\n")
```

## 7. Gene List Extraction for Enrichment Analysis

This section extracts unique gene lists from filtered CAAS datasets. These lists are formatted for downstream enrichment analysis tools (GO, KEGG, Reactome, etc.).

```{r gene_extraction, message=FALSE, warning=FALSE}
gene_path <- file.path(caas_list_dir, "3.Gene-tables")
createDir(gene_path)

# Unified gene extraction function
extract_gene_list <- function(df, name, criteria) {
  # Parse trait and scenario names
  trait_name <- paste(strsplit(name, "_")[[1]][1:2], collapse = "_")
  scenario_name <- gsub("\\.tab\\.caas", "", paste(strsplit(name, "_")[[1]][3:4], collapse = "_"))
  
  # Create output directory
  path <- file.path(gene_path, trait_name, scenario_name)
  createDir(path)
  
  # Extract unique genes
  genes <- unique(df$Gene)
  
  # Write to file
  output_file <- file.path(path, paste0(criteria, "_", name, ".genes"))
  write.table(
    genes,
    output_file,
    quote = FALSE,
    row.names = FALSE,
    col.names = FALSE,
    sep = "\n"
  )
  
  return(genes)
}

# Extract genes for all variants and patterns
extract_all_genes <- function(pattern_list, suffix) {
  results <- list()
  for (pattern_name in names(pattern_list)) {
    pattern_data <- pattern_list[[pattern_name]]
    criteria <- paste0(pattern_name, suffix)
    
    results[[pattern_name]] <- lapply(names(pattern_data), function(name) {
      extract_gene_list(pattern_data[[name]], name, criteria)
    })
    names(results[[pattern_name]]) <- names(pattern_data)
  }
  return(results)
}

# Extract for all pattern sets
genes_raw <- extract_all_genes(patterns_raw, "")
genes_pval05 <- extract_all_genes(patterns_pval05, "_05")
genes_boot <- extract_all_genes(patterns_boot, "_boot")
genes_boot_pval05 <- extract_all_genes(patterns_boot_pval05, "_boot05")

# Create aggregated gene lists per trait
aggregate_trait_genes <- function(gene_list, trait_name) {
  all_genes <- unique(unlist(gene_list))
  
  path <- file.path(gene_path, trait_name)
  createDir(path)
  
  output_file <- file.path(path, paste0("all_genes_", trait_name, ".txt"))
  write.table(
    sort(all_genes),
    output_file,
    quote = FALSE,
    row.names = FALSE,
    col.names = FALSE,
    sep = "\t"
  )
  
  return(all_genes)
}

# Aggregate by trait (malignant vs neoplasia)
for (trait in c("malignant_prevalence", "neoplasia_prevalence")) {
  trait_genes <- genes_raw[["pattern1"]][grepl(trait, names(genes_raw[["pattern1"]]))]
  aggregate_trait_genes(trait_genes, trait)
}

# Gene extraction summary
cat("\n=== Gene List Extraction Summary ===\n")
cat(sprintf("Output directory: %s\n", gene_path))
cat("\nUnique genes extracted by filtering strategy:\n")

gene_summary <- data.frame(
  Strategy = c("Raw", "Hypergeometric", "Permutation", "Combined"),
  Pattern1_genes = c(
    length(unique(unlist(lapply(genes_raw$pattern1, unique)))),
    length(unique(unlist(lapply(genes_pval05$pattern1, unique)))),
    length(unique(unlist(lapply(genes_boot$pattern1, unique)))),
    length(unique(unlist(lapply(genes_boot_pval05$pattern1, unique))))
  ),
  All_patterns_genes = c(
    length(unique(unlist(genes_raw))),
    length(unique(unlist(genes_pval05))),
    length(unique(unlist(genes_boot))),
    length(unique(unlist(genes_boot_pval05)))
  )
)

print(gene_summary)

cat("\nGene lists exported for:\n")
cat("  - Each contrast and pattern combination\n")
cat("  - Aggregated by trait (malignant_prevalence, neoplasia_prevalence)\n")
cat("  - Ready for enrichment analysis (GO, KEGG, Reactome, etc.)\n")
```

## 8. Cross-Contrast Commonality Analysis

This section identifies positions shared across multiple contrasts and creates aggregated datasets for meta-analysis.
"Significant" positions are determined based on the combined hypergeometric and permulation filtering.

```{r commonality_analysis, message=FALSE, warning=FALSE}
assign_caas_tags <- function(df) {
  set.seed(1998)
  
  n_rows <- nrow(df)
  tags <- sapply(1:n_rows, function(i) {
    paste0("CAAS_", 
           paste0(sample(LETTERS, 5, TRUE), collapse = ""),
           sprintf("%04d", sample(9999, 1)),
           sample(LETTERS, 1))
  })
  
  if ("Substitution" %in% colnames(df)) {
    df <- df %>% mutate(AminoConv = Substitution)
  }
  
  df %>%
    mutate(
      Trait = sub("^([^_]+_[^_]+)_.*", "\\1", source),
      Tag = tags
    ) %>%
    rename(Contrast = source) %>%
    # Extract only the final integer before the extension
    mutate(Contrast = sub(".*_(\\d+)\\..*$", "\\1", Contrast)) %>%
    select(-Trait)
}

# Aggregate by trait
aggregate_by_trait <- function(df_list) {
  malignant <- df_list[grepl("^malignant_", names(df_list))] %>%
    bind_rows(.id = "source") %>%
    assign_caas_tags()
  
  neoplasia <- df_list[grepl("^neoplasia_", names(df_list))] %>%
    bind_rows(.id = "source") %>%
    assign_caas_tags()
  
  return(list(
    malignant_prevalence = malignant,
    neoplasia_prevalence = neoplasia
  ))
}

# Create aggregated datasets
aggregated_raw <- aggregate_by_trait(caas_filtered)
aggregated_sig <- aggregate_by_trait(caas_pval05_boot)

# Extract position-level information
extract_position_info <- function(df) {
  df %>%
    mutate(GenePos = paste(Gene, Position, sep = "_")) %>%
    group_by(GenePos) %>%
    summarize(
      Contrast = paste(unique(Contrast), collapse = ","),
      Unique = !grepl(",", Contrast[1]),
      .groups = "drop"
    )
}

# Create unique/shared position tables
create_commonality_tables <- function(agg_df) {
  pos_info <- extract_position_info(agg_df)
  
  unique_positions <- pos_info %>%
    filter(Unique) %>%
    select(GenePos, Contrast)
  
  shared_positions <- pos_info %>%
    filter(!Unique) %>%
    select(GenePos, Contrast)
  
  merged_positions <- agg_df %>%
    mutate(GenePos = paste(Gene, Position, sep = "_")) %>%
    select(GenePos, Contrast, Tag)
  
  meta_caas <- agg_df %>%
    mutate(GenePos = paste(Gene, Position, sep = "_")) %>%
    select(GenePos, Tag, Contrast, Pattern, AminoConv, Pvalue, Pvalue.boot, isSignificant)
  
  return(list(
    unique_positions = unique_positions,
    shared_positions = shared_positions,
    merged_positions = merged_positions,
    meta_caas = meta_caas
  ))
}

# Process both raw and bootstrap aggregations
commonality_raw <- lapply(aggregated_raw, create_commonality_tables)
commonality_sig <- lapply(aggregated_sig, create_commonality_tables)

# Save commonality tables
save_commonality <- function(commonality_list, parent_name, subdir) {
  for (table_name in names(commonality_list)) {
    path <- file.path(caas_list_dir, "4.Commonalities", subdir, parent_name)
    createDir(path)
    
    output_file <- file.path(path, paste0(table_name, ".output"))
    write.table(
      commonality_list[[table_name]],
      output_file,
      quote = FALSE,
      row.names = FALSE,
      col.names = TRUE,
      sep = "\t"
    )
  }
}

# Save raw commonality tables
for (trait in names(commonality_raw)) {
  save_commonality(commonality_raw[[trait]], trait, "not-significant")
}

# Save significant commonality tables
for (trait in names(commonality_sig)) {
  save_commonality(commonality_sig[[trait]], trait, "significant")
}

# Commonality analysis summary
cat("\n=== Cross-Contrast Commonality Analysis Summary ===\n")
cat(sprintf("Output directory: %s\n", file.path(caas_list_dir, "4.Commonalities")))
cat("\nAnalysis identifies CAAS positions shared across multiple contrasts.\n")
cat("Useful for meta-analysis and identifying robust trait-associated positions.\n\n")

# Summary for each trait
for (trait in names(commonality_raw)) {
  cat(sprintf("Trait: %s\n", trait))
  cat(sprintf("  Total positions: %s\n", 
              format(nrow(commonality_raw[[trait]]$merged_positions), big.mark = ",")))
  cat(sprintf("  Unique to one contrast: %s\n", 
              format(nrow(commonality_raw[[trait]]$unique_positions), big.mark = ",")))
  cat(sprintf("  Shared across contrasts: %s\n", 
              format(nrow(commonality_raw[[trait]]$shared_positions), big.mark = ",")))
  
  if (nrow(commonality_raw[[trait]]$shared_positions) > 0) {
    cat(sprintf("  Sharing percentage: %.1f%%\n\n",
                100 * nrow(commonality_raw[[trait]]$shared_positions) / 
                  nrow(commonality_raw[[trait]]$merged_positions)))
  } else {
    cat("\n")
  }
}

cat("Tables exported:\n")
cat("  - unique_positions: Positions specific to single contrasts\n")
cat("  - shared_positions: Positions found in multiple contrasts\n")
cat("  - merged_positions: All positions with contrast assignments\n")
cat("  - meta_caas: Full CAAS metadata for meta-analysis\n")
```

## Session Information

```{r session_info}
sessionInfo()
```
