---
title: "Comparative Enrichment Across Modules — `r params$traitname`"
author: "PhyloPhere Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    code_folding: hide
    df_print: paged
    self_contained: true
    mathjax: null
    md_extensions: "-tex_math_dollars"   # gene-set names may contain '$' (V$HNF4_01)
params:
  fcs_dir:    "cmp_fcs"       # dir of per-module <module>_fcs_all.tsv
  string_dir: "cmp_string"    # dir of *_enrichment.tsv (optional)
  fdr_thr:    0.15
  pperm_thr:  0.025
  top_n:      20
  traitname:  "trait"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 7, dev = "png", dpi = 200)
suppressPackageStartupMessages({ library(tidyverse); library(knitr); library(kableExtra) })
filter <- dplyr::filter; select <- dplyr::select; rename <- dplyr::rename; mutate <- dplyr::mutate
has_pkg <- function(p) requireNamespace(p, quietly = TRUE)

QUAD_COLS <- c("Shared" = "#2166ac", "Top only" = "#d73027",
               "Bottom only" = "#4dac26", "Neither" = "gray80")
kbl_html <- function(df, height = NULL) {
  k <- kable(df, format = "html", escape = TRUE) %>%
    kable_styling(bootstrap_options = c("striped", "condensed", "hover"), full_width = FALSE)
  if (!is.null(height)) k <- scroll_box(k, height = height)
  k
}
```

```{r load, include=FALSE}
`%||%` <- function(a, b) if (!is.null(a)) a else b
REQ <- c("ranking", "database", "pathway", "stat", "p.adj")

# Per-module FCS results: cmp_fcs/<module>_fcs_all.tsv (caas/rer/fade/accum).
mod_files <- list.files(params$fcs_dir, pattern = "_fcs_all\\.tsv$", full.names = TRUE)
fcs_by_mod <- list()
for (f in mod_files) {
  mod <- sub("_fcs_all\\.tsv$", "", basename(f))
  d <- tryCatch(readr::read_tsv(f, show_col_types = FALSE), error = function(e) NULL)
  if (!is.null(d) && nrow(d) > 0 && all(REQ %in% names(d))) {
    fcs_by_mod[[mod]] <- dplyr::mutate(d, module = mod)
  }
}
fcs_all_mod <- if (length(fcs_by_mod)) dplyr::bind_rows(fcs_by_mod) else tibble()
modules_present <- names(fcs_by_mod)
has_multi <- length(fcs_by_mod) >= 1

# The CAAS module drives the legacy top-vs-bottom section (backward compatible).
fcs <- fcs_by_mod[["caas"]] %||% tibble()
has_fcs <- nrow(fcs) > 0 && all(REQ %in% names(fcs))
has_le <- FALSE; le <- tibble()   # leading edge not collected cross-module

string_files <- list.files(params$string_dir, pattern = "_enrichment\\.tsv$", full.names = TRUE)
string_all <- if (length(string_files) > 0)
  purrr::map_dfr(string_files, ~readr::read_tsv(.x, show_col_types = FALSE)) else tibble()
has_string <- nrow(string_all) > 0
```

# Overview

```{r overview, echo=FALSE}
mod_rows <- if (has_multi) sapply(fcs_by_mod, nrow) else integer(0)
tibble::tibble(
  Source = c(paste0("FCS module: ", names(mod_rows)), "STRING"),
  Rows   = c(unname(mod_rows), nrow(string_all))
) %>% kbl_html()
```

This report compares enrichment **across the evidence modules** (CAAS, RER, FADE,
accumulation) for `r params$traitname`. Because every module is scored by the same
FCS engine and dual-threshold significance, a gene set enriched in **multiple**
modules is convergent, independent evidence. A set is "significant" at BH < `r params$fdr_thr`
(and permulation p < threshold where available) with AUC > 0.5.

The report reads bottom-up: **(1)** what is significant *within each module on its
own*, then **(2)** which terms recur *across* modules (the convergence view), and
finally **(3)** the legacy CAAS top-vs-bottom and STRING directional contrasts.

# Per-module significance {.tabset .tabset-fade}

What does each module call significant on its own? This is the per-module baseline
that the cross-module convergence view (next section) builds on — a term recurring
across modules only means something once you know each module's own hit list.

`r if (!has_multi) "> **No module FCS results available — skipping.**"`

```{r permod_prep, eval=has_multi}
thr  <- params$fdr_thr
pthr <- params$pperm_thr %||% 1
# Per (module, database, pathway): best positively-enriched ranking (max AUC, min
# p.adj). Significance = BH < thr AND AUC > 0.5 AND (p.perm < pthr where present).
permod_best <- fcs_all_mod %>%
  filter(stat > 0) %>%
  mutate(sig = !is.na(p.adj) & p.adj < thr & stat > 0 &
               (!("p.perm" %in% names(.)) | is.na(p.perm) | p.perm < pthr)) %>%
  group_by(module, database, pathway) %>%
  summarise(AUC = max(stat, na.rm = TRUE) + 0.5,
            p.adj = min(p.adj, na.rm = TRUE),
            sig = any(sig, na.rm = TRUE),
            description = dplyr::first(dplyr::coalesce(description, pathway)),
            .groups = "drop")

permod_summary <- permod_best %>%
  group_by(module) %>%
  summarise(`Terms tested` = dplyr::n(),
            `Significant`  = sum(sig, na.rm = TRUE),
            `% significant` = round(100 * mean(sig, na.rm = TRUE), 1),
            .groups = "drop") %>%
  arrange(desc(Significant))

dir.create("compare_results", showWarnings = FALSE)
readr::write_tsv(permod_summary, "compare_results/per_module_significance_summary.tsv")
readr::write_tsv(permod_best,    "compare_results/per_module_best_per_term.tsv")
```

## Summary

```{r permod_summary_tab, eval=has_multi, results='asis'}
if (nrow(permod_summary) == 0) {
  cat("_No module produced enriched terms._\n")
} else {
  print(kbl_html(permod_summary))
}
```

```{r permod_tabs, eval=has_multi, results='asis'}
# One tab per module: its own top-N significant terms by AUC.
for (m in modules_present) {
  cat(sprintf("\n## %s\n\n", toupper(m)))
  d <- permod_best %>%
    filter(module == m, sig) %>%
    mutate(Description = stringr::str_trunc(description, 60),
           AUC = round(AUC, 3), p.adj = signif(p.adj, 3)) %>%
    arrange(desc(AUC)) %>%
    select(database, pathway, Description, AUC, p.adj) %>%
    head(params$top_n)
  if (nrow(d) == 0) {
    cat(sprintf("_No gene set is significant in %s at BH < %.2g._\n\n", toupper(m), thr))
  } else {
    print(kbl_html(d, height = if (nrow(d) > 15) "420px" else NULL))
  }
}
```

# Cross-module convergence {.tabset .tabset-fade}

`r if (!has_multi) "> **No module FCS results available — skipping.**"`

```{r xmod_prep, eval=has_multi}
thr <- params$fdr_thr
# Per (module, database, pathway): take the BEST positively-enriched ranking
# (smallest p.adj with AUC > 0.5). p.perm tightens significance when present.
best <- fcs_all_mod %>%
  mutate(sig = !is.na(p.adj) & p.adj < thr & stat > 0 &
               (!("p.perm" %in% names(.)) | is.na(p.perm) | p.perm < (params$pperm_thr %||% 1))) %>%
  filter(stat > 0) %>%
  group_by(module, database, pathway) %>%
  summarise(AUC = max(stat, na.rm = TRUE) + 0.5,
            p.adj = min(p.adj, na.rm = TRUE),
            sig = any(sig, na.rm = TRUE),
            description = dplyr::first(dplyr::coalesce(description, pathway)),
            .groups = "drop")

# Convergence matrix: which modules call each term significant.
conv <- best %>%
  filter(sig) %>%
  group_by(database, pathway) %>%
  summarise(n_modules = dplyr::n_distinct(module),
            modules = paste(sort(unique(module)), collapse = ", "),
            min_padj = min(p.adj, na.rm = TRUE),
            description = dplyr::first(description),
            .groups = "drop") %>%
  arrange(desc(n_modules), min_padj)

dir.create("compare_results", showWarnings = FALSE)
readr::write_tsv(conv, "compare_results/cross_module_convergence.tsv")
readr::write_tsv(best, "compare_results/cross_module_best_per_term.tsv")
```

## Convergent terms (≥2 modules)

```{r xmod_convergent, eval=has_multi, results='asis'}
cv <- conv %>% filter(n_modules >= 2)
if (nrow(cv) == 0) {
  cat(sprintf("_No gene set is significant in ≥2 modules at BH < %.2g._\n\n", thr))
} else {
  cat(sprintf("**%d** gene sets are significantly enriched in **≥2 modules** — the strongest, independent-evidence pathways.\n\n", nrow(cv)))
  disp <- cv %>%
    mutate(Description = stringr::str_trunc(description, 60), min_padj = signif(min_padj, 3)) %>%
    select(database, pathway, Description, n_modules, modules, min_padj) %>%
    head(params$top_n)
  print(kbl_html(disp, height = if (nrow(disp) > 15) "460px" else NULL))
}
```

## Term × module heatmap

```{r xmod_heatmap, eval=has_multi, fig.height=7}
top_terms <- conv %>% slice_head(n = 30) %>% select(database, pathway)
hm <- best %>%
  semi_join(top_terms, by = c("database", "pathway")) %>%
  mutate(label = stringr::str_trunc(description, 45),
         score = ifelse(sig, -log10(pmax(p.adj, 1e-300)), NA_real_))
if (nrow(hm) > 0) {
  p <- ggplot(hm, aes(x = module, y = forcats::fct_reorder(label, score, .fun = max, na.rm = TRUE), fill = score)) +
    geom_tile(colour = "grey90") +
    scale_fill_viridis_c(name = "-log10 BH", option = "C", na.value = "grey95") +
    labs(title = sprintf("Top convergent terms across modules — %s", params$traitname),
         x = NULL, y = NULL) +
    theme_minimal(base_size = 10)
  print(p)
} else {
  cat("_No significant terms to display._\n")
}
```

---

# FCS: Top vs Bottom (CAAS) {.tabset .tabset-fade}

The section below contrasts the **top** and **bottom** directional CAAS rankings:
which enriched gene sets are shared, and which are direction-specific.

`r if (!has_fcs) "> **No CAAS FCS results available — skipping.**"`

```{r fcs_prep, eval=has_fcs}
thr <- params$fdr_thr
desc_lookup <- fcs %>% distinct(database, pathway, description = dplyr::coalesce(description, pathway))
wide <- fcs %>%
  filter(ranking %in% c("top", "bottom")) %>%
  mutate(sig = !is.na(p.adj) & p.adj < thr & stat > 0) %>%
  select(database, pathway, ranking, AUC = stat, p.adj, sig) %>%
  mutate(AUC = AUC + 0.5) %>%
  pivot_wider(names_from = ranking, values_from = c(AUC, p.adj, sig),
              names_sep = "_") %>%
  left_join(desc_lookup, by = c("database", "pathway")) %>%
  mutate(
    sig_top    = dplyr::coalesce(sig_top, FALSE),
    sig_bottom = dplyr::coalesce(sig_bottom, FALSE),
    class = dplyr::case_when(
      sig_top & sig_bottom ~ "Shared",
      sig_top              ~ "Top only",
      sig_bottom           ~ "Bottom only",
      TRUE                 ~ "Neither"
    )
  )

# Calculate leading-edge percentages if available
wide_with_pcts <- wide
if (has_le) {
  flag_cols <- intersect(
    names(le),
    c("flag_gate_sig", "flag_gate_fdr", "pct_top10", "pct_top5", "pct_top1",
      "pct_bottom10", "pct_bottom5", "pct_bottom1", "flag_fade_top", "flag_fade_bottom",
      "flag_fade", "flag_rer_acc", "flag_rer_decc", "flag_accum")
  )
  
  le_pcts <- le %>%
    group_by(ranking, database, pathway) %>%
    summarise(
      across(all_of(flag_cols), ~ round(100 * mean(.x, na.rm = TRUE), 1)),
      .groups = "drop"
    )
    
  top_pcts <- le_pcts %>%
    filter(ranking == "top") %>%
    select(-ranking) %>%
    rename_with(~ paste0(.x, "_top"), -c(database, pathway))
    
  bot_pcts <- le_pcts %>%
    filter(ranking == "bottom") %>%
    select(-ranking) %>%
    rename_with(~ paste0(.x, "_bottom"), -c(database, pathway))
    
  wide_with_pcts <- wide %>%
    left_join(top_pcts, by = c("database", "pathway")) %>%
    left_join(bot_pcts, by = c("database", "pathway"))
}

dir.create("compare_results", showWarnings = FALSE)
readr::write_tsv(wide_with_pcts, "compare_results/fcs_top_vs_bottom.tsv")
```

## Summary

```{r fcs_summary, eval=has_fcs, results='asis'}
tab <- wide %>% filter(class != "Neither") %>% count(class, name = "n_sets")
if (nrow(tab) == 0) cat("_No sets reached BH significance in either direction._\n") else
  print(kbl_html(tab))
```

## Scatter (AUC top vs bottom)

```{r fcs_scatter, eval=has_fcs, fig.height=6}
pl <- wide %>% filter(!is.na(AUC_top) | !is.na(AUC_bottom)) %>%
  mutate(AUC_top = dplyr::coalesce(AUC_top, 0.5), AUC_bottom = dplyr::coalesce(AUC_bottom, 0.5))
p <- ggplot(pl, aes(AUC_top, AUC_bottom, colour = class)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(alpha = 0.7, size = 2) +
  scale_colour_manual(values = QUAD_COLS, name = NULL) +
  labs(title = sprintf("FCS top vs bottom AUC — %s", params$traitname),
       x = "AUC (top ranking)", y = "AUC (bottom ranking)") +
  theme_minimal(base_size = 12)
if (has_pkg("ggrepel")) {
  lab <- pl %>% filter(class %in% c("Top only", "Bottom only", "Shared")) %>%
    slice_max(abs(AUC_top - AUC_bottom), n = 12, with_ties = FALSE)
  if (nrow(lab) > 0) p <- p + ggrepel::geom_text_repel(
    data = lab, aes(label = stringr::str_trunc(description, 40)), size = 2.6, max.overlaps = 20, seed = 42)
}
print(p)
```

## Direction-specific sets

```{r fcs_tables, eval=has_fcs, results='asis'}
for (cl in c("Top only", "Bottom only", "Shared")) {
  if (has_le) {
    if (cl == "Top only") {
      d <- wide_with_pcts %>% filter(class == cl) %>%
        mutate(across(c(AUC_top, p.adj_top), ~round(.x, 3)),
               Description = stringr::str_trunc(description, 55)) %>%
        arrange(p.adj_top) %>%
        select(
          database, pathway, Description, AUC_top, p.adj_top,
          `gate_sig (%)` = flag_gate_sig_top,
          `gate_fdr (%)` = flag_gate_fdr_top,
          `top10% (%)` = pct_top10_top,
          `top5% (%)` = pct_top5_top,
          `top1% (%)` = pct_top1_top,
          `fade_top (%)` = flag_fade_top_top,
          `rer_acc (%)` = flag_rer_acc_top,
          `rer_decc (%)` = flag_rer_decc_top,
          `accum (%)` = flag_accum_top
        ) %>%
        head(params$top_n)
    } else if (cl == "Bottom only") {
      d <- wide_with_pcts %>% filter(class == cl) %>%
        mutate(across(c(AUC_bottom, p.adj_bottom), ~round(.x, 3)),
               Description = stringr::str_trunc(description, 55)) %>%
        arrange(p.adj_bottom) %>%
        select(
          database, pathway, Description, AUC_bottom, p.adj_bottom,
          `gate_sig (%)` = flag_gate_sig_bottom,
          `gate_fdr (%)` = flag_gate_fdr_bottom,
          `bot10% (%)` = pct_bottom10_bottom,
          `bot5% (%)` = pct_bottom5_bottom,
          `bot1% (%)` = pct_bottom1_bottom,
          `fade_bottom (%)` = flag_fade_bottom_bottom,
          `rer_acc (%)` = flag_rer_acc_bottom,
          `rer_decc (%)` = flag_rer_decc_bottom,
          `accum (%)` = flag_accum_bottom
        ) %>%
        head(params$top_n)
    } else {
      d <- wide_with_pcts %>% filter(class == cl) %>%
        mutate(across(c(AUC_top, AUC_bottom), ~round(.x, 3)),
               across(c(p.adj_top, p.adj_bottom), ~signif(.x, 3)),
               Description = stringr::str_trunc(description, 55)) %>%
        arrange(p.adj_top, p.adj_bottom) %>%
        select(
          database, pathway, Description,
          AUC_top, p.adj_top, AUC_bottom, p.adj_bottom,
          `gate_sig_top (%)` = flag_gate_sig_top,
          `fade_top (%)` = flag_fade_top_top,
          `rer_acc_top (%)` = flag_rer_acc_top,
          `accum_top (%)` = flag_accum_top,
          `gate_sig_bot (%)` = flag_gate_sig_bottom,
          `fade_bot (%)` = flag_fade_bottom_bottom,
          `rer_acc_bot (%)` = flag_rer_acc_bottom,
          `accum_bot (%)` = flag_accum_bottom
        ) %>%
        head(params$top_n)
    }
  } else {
    d <- wide %>% filter(class == cl) %>%
      mutate(across(c(AUC_top, AUC_bottom), ~round(.x, 3)),
             across(c(p.adj_top, p.adj_bottom), ~signif(.x, 3)),
             Description = stringr::str_trunc(description, 55)) %>%
      arrange(p.adj_top, p.adj_bottom) %>%
      select(database, pathway, Description, AUC_top, p.adj_top, AUC_bottom, p.adj_bottom) %>%
      head(params$top_n)
  }
  
  cat(sprintf("\n### %s (%d)\n\n", cl, nrow(d)))
  if (nrow(d) == 0) cat("_None._\n\n") else print(kbl_html(d, height = if (nrow(d) > 15) "420px" else NULL))
}
```

# STRING: Top vs Bottom {.tabset .tabset-fade}

`r if (!has_string) "> **No STRING results available — skipping.**"`

```{r string_compare, eval=has_string, results='asis'}
dir_col <- intersect(c("direction", "gene_list", "list_name"), names(string_all))[1]
if (is.na(dir_col)) {
  cat("_STRING results lack a direction column; showing combined table._\n\n")
  print(kbl_html(head(string_all, params$top_n), height = "420px"))
} else {
  sa <- string_all %>% mutate(direction = tolower(as.character(.data[[dir_col]])),
                              direction = dplyr::case_when(grepl("top", direction) ~ "top",
                                                           grepl("bottom|bot", direction) ~ "bottom",
                                                           TRUE ~ "other"))
  for (d in c("top", "bottom")) {
    sub <- sa %>% filter(direction == d) %>% head(params$top_n)
    cat(sprintf("\n## %s\n\n", d))
    if (nrow(sub) == 0) {
      cat("_None._\n\n")
      next
    }
    
    has_pct <- "pct_gate_sig" %in% names(sub)
    if (has_pct) {
      if (d == "top") {
        display_df <- sub %>%
          dplyr::select(
            Category = category,
            Term = description,
            FDR = fdr,
            `gate_sig (%)` = pct_gate_sig,
            `gate_fdr (%)` = pct_gate_fdr,
            `top10% (%)` = pct_top10,
            `top5% (%)` = pct_top5,
            `top1% (%)` = pct_top1,
            `fade_top (%)` = pct_fade_top,
            `rer_acc (%)` = pct_rer_acc,
            `rer_decc (%)` = pct_rer_decc,
            `accum (%)` = pct_accum
          )
      } else {
        display_df <- sub %>%
          dplyr::select(
            Category = category,
            Term = description,
            FDR = fdr,
            `gate_sig (%)` = pct_gate_sig,
            `gate_fdr (%)` = pct_gate_fdr,
            `bot10% (%)` = pct_bottom10,
            `bot5% (%)` = pct_bottom5,
            `bot1% (%)` = pct_bottom1,
            `fade_bottom (%)` = pct_fade_bottom,
            `rer_acc (%)` = pct_rer_acc,
            `rer_decc (%)` = pct_rer_decc,
            `accum (%)` = pct_accum
          )
      }
    } else {
      display_df <- sub
    }
    
    print(kbl_html(display_df, height = "420px"))
  }
}
```
