---
title: "Comparative Enrichment Across Modules — `r params$traitname`"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
    mathjax: null                       # no math in this report
    md_extensions: "-tex_math_dollars"  # gene-set names contain '$' (e.g. V$HNF4_01) — never parse as TeX math
params:
  fcs_dir:    "cmp_fcs"       # dir of per-module <module>_fcs_all.tsv
  fcs_le_dir: "cmp_fcs_le"    # dir of per-module <module>_leading_edge.tsv
  string_dir: "cmp_string"    # dir of *_enrichment.tsv (optional)
  fdr_thr:    0.15
  pperm_thr:  0.025
  top_n:      20
  traitname:  "trait"
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 7, dev = "png", dpi = 200)
suppressPackageStartupMessages({ library(tidyverse); library(knitr); library(kableExtra) })
filter <- dplyr::filter; select <- dplyr::select; rename <- dplyr::rename; mutate <- dplyr::mutate
has_pkg <- function(p) requireNamespace(p, quietly = TRUE)
source(file.path(dirname(normalizePath(knitr::current_input(), mustWork = FALSE)), "src", "orthogonal_score.R"))

QUAD_COLS <- c("Shared" = "#2166ac", "Top only" = "#d73027",
               "Bottom only" = "#4dac26", "Neither" = "gray80")
kbl_html <- function(df, height = NULL) {
  k <- kable(df, format = "html", escape = TRUE) %>%
    kable_styling(bootstrap_options = c("striped", "condensed", "hover"), full_width = FALSE)
  if (!is.null(height)) k <- scroll_box(k, height = height)
  k
}
```

```{r load, include=FALSE}
`%||%` <- function(a, b) if (!is.null(a)) a else b
REQ <- c("ranking", "database", "pathway", "stat", "p.adj")

# Per-module FCS results: cmp_fcs/<module>_fcs_all.tsv (caas/rer).
mod_files <- list.files(params$fcs_dir, pattern = "_fcs_all\\.tsv$", full.names = TRUE)
fcs_by_mod <- list()
for (f in mod_files) {
  mod <- sub("_fcs_all\\.tsv$", "", basename(f))
  d <- tryCatch(readr::read_tsv(f, show_col_types = FALSE), error = function(e) NULL)
  if (!is.null(d) && nrow(d) > 0 && all(REQ %in% names(d))) {
    fcs_by_mod[[mod]] <- dplyr::mutate(d, module = mod)
  }
}
fcs_all_mod <- if (length(fcs_by_mod)) dplyr::bind_rows(fcs_by_mod) else tibble()
modules_present <- names(fcs_by_mod)
has_multi <- length(fcs_by_mod) >= 1

# Direction facet (top/bottom), derived once here since every directional
# section below groups by it. Each module's `ranking` vocabulary differs:
# CAAS uses top/bottom directly; RER's accelerating/decelerating is the same
# two-way split framed around RER's own statistic. Modules with a real
# top/bottom ranking have their own 'global' ranking excluded from
# directional sections (it would double-count alongside their real
# top/bottom rows).
if (has_multi) {
  fcs_all_mod <- fcs_all_mod %>%
    mutate(.dir_raw = dplyr::case_when(
      ranking %in% c("top", "accelerating")    ~ "top",
      ranking %in% c("bottom", "decelerating") ~ "bottom",
      TRUE                                     ~ NA_character_
    ))
  fcs_dir_mod <- dplyr::bind_rows(
    fcs_all_mod %>% dplyr::filter(!is.na(.dir_raw)) %>% dplyr::rename(direction = .dir_raw),
    fcs_all_mod %>% dplyr::filter(is.na(.dir_raw), ranking == "global") %>%
      dplyr::select(-.dir_raw) %>% dplyr::mutate(direction = "top"),
    fcs_all_mod %>% dplyr::filter(is.na(.dir_raw), ranking == "global") %>%
      dplyr::select(-.dir_raw) %>% dplyr::mutate(direction = "bottom")
  )
  fcs_all_mod <- fcs_all_mod %>% dplyr::select(-.dir_raw)
} else {
  fcs_dir_mod <- fcs_all_mod
}

# Per-module leading-edge tables (cmp_fcs_le/<module>_leading_edge.tsv) feed the
# orthogonal composite score: each module's own directional percentile
# concentration + cross-module corroboration flags (FADE/RER/Accumulation,
# and PrimateAI-3D/COSMIC once wired in from posenrich). Percentile columns
# reflect CAAS's own ranking in every module's leading edge (per-module FCS
# reports are annotated with CAAS's fcs_stats.tsv as their overlay) — that's
# intentional, not a bug: it's the shared reference axis every module's
# leading edge is cross-checked against.
le_files <- list.files(params$fcs_le_dir, pattern = "_leading_edge\\.tsv$", full.names = TRUE)
le_by_mod <- list()
for (f in le_files) {
  mod <- sub("_leading_edge\\.tsv$", "", basename(f))
  d <- tryCatch(readr::read_tsv(f, show_col_types = FALSE), error = function(e) NULL)
  if (!is.null(d) && nrow(d) > 0) le_by_mod[[mod]] <- dplyr::mutate(d, module = mod)
}
le <- if (length(le_by_mod)) dplyr::bind_rows(le_by_mod) else tibble()
has_le <- nrow(le) > 0

string_files <- list.files(params$string_dir, pattern = "_enrichment\\.tsv$", full.names = TRUE)
string_all <- if (length(string_files) > 0)
  purrr::map_dfr(string_files, ~readr::read_tsv(.x, show_col_types = FALSE)) else tibble()
has_string <- nrow(string_all) > 0
```

# Overview

```{r overview, echo=FALSE}
mod_rows <- if (has_multi) sapply(fcs_by_mod, nrow) else integer(0)
tibble::tibble(
  Source = c(paste0("FCS module: ", names(mod_rows)), "STRING"),
  Rows   = c(unname(mod_rows), nrow(string_all))
) %>% kbl_html()
```

This report compares enrichment **across CAAS and RERconverge** (the two modules
with well-behaved FCS rankings and real permulation nulls) for `r params$traitname`.
FADE and Accumulation no longer run their own FCS ranking — both proved unreliable
for genome-wide gene-set enrichment (FADE's max-of-many-sites statistic and
Accumulation's missing permulation null each inflated false "significance") — but
both remain available as **cross-module corroboration**: a term's `signal` (the
orthogonal composite score) combines each module's own directional percentile
concentration with independent hypergeometric enrichment against FADE/RER/
Accumulation flags (and PrimateAI-3D/COSMIC once wired in). A set is "significant"
at BH < `r params$fdr_thr` (and permulation p < threshold where available) with
AUC > 0.5.

The report reads bottom-up: **(1)** what is significant *within each module on its
own*, then **(2)** which terms recur *across* modules (the convergence view, ordered
by `signal`), and finally **(3)** the STRING directional contrasts.

# Per-module significance {.tabset .tabset-fade}

What does each module call significant on its own? This is the per-module baseline
that the cross-module convergence view (next section) builds on — a term recurring
across modules only means something once you know each module's own hit list.

`r if (!has_multi) "> **No module FCS results available — skipping.**"`

```{r permod_prep, eval=has_multi}
thr  <- params$fdr_thr
pthr <- params$pperm_thr %||% 1
# Per (module, direction, database, pathway): best positively-enriched ranking
# within that direction (max AUC, min p.adj). Significance = BH < thr AND
# AUC > 0.5 AND (p.perm < pthr where present).
permod_best <- fcs_dir_mod %>%
  filter(stat > 0) %>%
  mutate(sig = !is.na(p.adj) & p.adj < thr & stat > 0 &
               (!("p.perm" %in% names(.)) | is.na(p.perm) | p.perm < pthr)) %>%
  group_by(module, direction, database, pathway) %>%
  summarise(AUC = max(stat, na.rm = TRUE) + 0.5,
            p.adj = min(p.adj, na.rm = TRUE),
            sig = any(sig, na.rm = TRUE),
            description = dplyr::first(dplyr::coalesce(description, pathway)),
            .groups = "drop")

permod_summary <- permod_best %>%
  group_by(direction, module) %>%
  summarise(`Terms tested` = dplyr::n(),
            `Significant`  = sum(sig, na.rm = TRUE),
            `% significant` = round(100 * mean(sig, na.rm = TRUE), 1),
            .groups = "drop") %>%
  arrange(direction, desc(Significant))

dir.create("compare_results", showWarnings = FALSE)
readr::write_tsv(permod_summary, "compare_results/per_module_significance_summary.tsv")
readr::write_tsv(permod_best,    "compare_results/per_module_best_per_term.tsv")
```

```{r permod_tabs, eval=has_multi, results='asis'}
# Direction is the primary split (per the report-wide per-direction design);
# module is a second-level tab within each direction.
for (dr in c("top", "bottom")) {
  cat(sprintf("\n## %s {.tabset}\n\n", tools::toTitleCase(dr)))

  dr_summary <- permod_summary %>% filter(direction == dr) %>% select(-direction)
  cat("### Summary\n\n")
  if (nrow(dr_summary) == 0) {
    cat("_No module produced enriched terms for this direction._\n\n")
  } else {
    print(kbl_html(dr_summary))
  }

  for (m in modules_present) {
    cat(sprintf("\n### %s\n\n", toupper(m)))
    d <- permod_best %>%
      filter(module == m, direction == dr, sig) %>%
      mutate(Description = stringr::str_trunc(description, 60),
             AUC = round(AUC, 3), p.adj = signif(p.adj, 3)) %>%
      arrange(desc(AUC)) %>%
      select(database, pathway, Description, AUC, p.adj) %>%
      head(params$top_n)
    if (nrow(d) == 0) {
      cat(sprintf("_No gene set is significant in %s (%s) at BH < %.2g._\n\n", toupper(m), dr, thr))
    } else {
      print(kbl_html(d, height = if (nrow(d) > 15) "420px" else NULL))
    }
  }
}
```

# Cross-module convergence {.tabset .tabset-fade}

`r if (!has_multi) "> **No module FCS results available — skipping.**"`

```{r xmod_prep, eval=has_multi}
thr <- params$fdr_thr

# Row identity for cross-module comparisons ("module_label"): CAAS keeps its
# own name. RER is kept as its two sub-rankings -- rer_accelerating and
# rer_decelerating -- as SEPARATE rows, both shown in BOTH direction tabs:
# RER's own accel/decel split is a different axis from which CAAS direction
# you happen to be looking at, so folding it into a single direction-dependent
# "RER" row would hide which of the two was actually driving any given cell.
# FADE/Accumulation no longer run their own FCS ranking (dropped module_label
# rows for them) — they still contribute as cross-module corroboration flags
# inside the orthogonal signal computed below.
xmod_labeled <- fcs_all_mod %>%
  mutate(module_label = dplyr::case_when(
    module == "rer" & ranking == "accelerating" ~ "rer_accelerating",
    module == "rer" & ranking == "decelerating" ~ "rer_decelerating",
    TRUE ~ module
  ))

xmod_dir <- dplyr::bind_rows(
  xmod_labeled %>% dplyr::filter(module == "caas", ranking == "top")    %>% dplyr::mutate(direction = "top"),
  xmod_labeled %>% dplyr::filter(module == "caas", ranking == "bottom") %>% dplyr::mutate(direction = "bottom"),
  xmod_labeled %>% dplyr::filter(module == "rer", ranking %in% c("accelerating", "decelerating")) %>% dplyr::mutate(direction = "top"),
  xmod_labeled %>% dplyr::filter(module == "rer", ranking %in% c("accelerating", "decelerating")) %>% dplyr::mutate(direction = "bottom")
)

# Per (module_label, direction, database, pathway): best positively-enriched
# ranking within that direction (smallest p.adj with AUC > 0.5). p.perm
# tightens significance when present.
best <- xmod_dir %>%
  mutate(sig = !is.na(p.adj) & p.adj < thr & stat > 0 &
               (!("p.perm" %in% names(.)) | is.na(p.perm) | p.perm < (params$pperm_thr %||% 1))) %>%
  filter(stat > 0) %>%
  group_by(module_label, direction, database, pathway) %>%
  summarise(AUC = max(stat, na.rm = TRUE) + 0.5,
            p.adj = min(p.adj, na.rm = TRUE),
            sig = any(sig, na.rm = TRUE),
            num.genes = suppressWarnings(max(num.genes, na.rm = TRUE)),
            description = dplyr::first(dplyr::coalesce(description, pathway)),
            .groups = "drop")

# ── Orthogonal composite signal ───────────────────────────────────────────
# Per (module, ranking): Lancaster-weighted composite of the module's own
# directional percentile concentration (from CAAS's percentile trio, the
# shared reference axis every per-module report is annotated against) plus
# independent cross-module flags -- excluding whichever flag would be
# self-referential for the CURRENT module (e.g. RER's own leading edge
# doesn't get to cite flag_rer_acc/flag_rer_decc as "corroboration").
# Background rates are estimated from each module's own observed
# leading-edge gene pool (this report doesn't carry the full universe file).
signal_by_mod <- tibble()
if (has_le) {
  self_ref <- list(caas = character(0), rer = c("flag_rer_acc", "flag_rer_decc"))
  signal_rows <- list()
  for (mod in unique(le$module)) {
    le_mod <- le %>% filter(module == mod)
    for (rk in unique(le_mod$ranking)) {
      le_rk <- le_mod %>% filter(ranking == rk)
      rk_dir <- dplyr::case_when(rk %in% c("top", "accelerating") ~ "top",
                                  rk %in% c("bottom", "decelerating") ~ "bottom",
                                  TRUE ~ "top")   # percentile trio falls back to top-style if undirected
      cols <- orth_cols_for_direction(rk_dir)
      cross_cols <- setdiff(cols$cross_cols, self_ref[[mod]] %||% character(0))
      uni_genes <- le_rk %>% distinct(gene, .keep_all = TRUE)
      n_uni <- nrow(uni_genes)
      if (n_uni < 10) next
      bg <- orth_background_rates(uni_genes, c(cols$percentile_cols, cross_cols))
      sc <- orthogonal_score(le_rk, bg, cols$percentile_cols, cross_cols, n_uni) %>%
        mutate(
          module_label = dplyr::case_when(
            mod == "rer" & rk == "accelerating" ~ "rer_accelerating",
            mod == "rer" & rk == "decelerating" ~ "rer_decelerating",
            TRUE ~ mod
          ),
          direction = rk_dir
        )
      signal_rows[[paste(mod, rk)]] <- sc
    }
  }
  signal_by_mod <- dplyr::bind_rows(signal_rows) %>%
    dplyr::select(module_label, direction, database, pathway, signal)
}
if (nrow(signal_by_mod) == 0) {
  signal_by_mod <- tibble(module_label = character(), direction = character(),
                           database = character(), pathway = character(), signal = double())
}
best <- best %>% dplyr::left_join(signal_by_mod, by = c("module_label", "direction", "database", "pathway"))

# Convergence: n_modules (out of up to 3 possible rows -- caas,
# rer_accelerating, rer_decelerating) and the strongest orthogonal signal
# among them. THIS pair is the ordering criterion for every table/plot in
# this section (n_modules first, signal as tiebreaker) -- not p.adj, which
# only says how confident any one module is, not how many independently
# agree, and not sum_AUC, which double-counts nested percentile evidence and
# says nothing about cross-module corroboration.
conv <- best %>%
  filter(sig) %>%
  group_by(direction, database, pathway) %>%
  summarise(n_modules = dplyr::n_distinct(module_label),
            modules = paste(sort(unique(module_label)), collapse = ", "),
            signal = max(signal, na.rm = TRUE),
            num.genes = suppressWarnings(max(num.genes, na.rm = TRUE)),
            description = dplyr::first(description),
            .groups = "drop") %>%
  mutate(signal = ifelse(is.finite(signal), signal, NA_real_)) %>%
  arrange(direction, desc(n_modules), desc(signal))

dir.create("compare_results", showWarnings = FALSE)
readr::write_tsv(conv, "compare_results/cross_module_convergence.tsv")
readr::write_tsv(best, "compare_results/cross_module_best_per_term.tsv")
```

## Convergent terms & dotplot {.tabset}

```{r xmod_convergent, eval=has_multi, results='asis'}
for (dr in c("top", "bottom")) {
  cat(sprintf("\n### %s\n\n", tools::toTitleCase(dr)))

  cv <- conv %>% filter(direction == dr, n_modules >= 2) %>% arrange(desc(n_modules), desc(signal)) %>% head(params$top_n)
  cat(sprintf("\n#### Convergent terms (≥2 of caas/rer_accelerating/rer_decelerating) — %s\n\n", tools::toTitleCase(dr)))
  cat(sprintf("_Ordered by **n_modules** (how many independently agree), then **signal** (orthogonal composite) as tiebreaker — not by p.adj. Top %d shown._\n\n", params$top_n))
  if (nrow(cv) == 0) {
    cat(sprintf("_No gene set is significant in ≥2 modules (%s) at BH < %.2g._\n\n", dr, thr))
  } else {
    disp <- cv %>%
      mutate(Description = stringr::str_trunc(description, 60), signal = round(signal, 2)) %>%
      select(database, pathway, Description, n_modules, modules, signal)
    print(kbl_html(disp, height = if (nrow(disp) > 15) "460px" else NULL))
  }

  # STRING-style bubble chart, replacing the old term x module heatmap:
  # x = orthogonal signal, y = term (ordered top-to-bottom by signal, matching
  # `cv`'s sort), size = term/set size (num.genes), colour = -log10(BH FDR).
  cat(sprintf("\n#### Convergent terms dotplot — %s\n\n", tools::toTitleCase(dr)))
  if (nrow(cv) > 0) {
    best_padj <- best %>%
      dplyr::filter(direction == dr) %>%
      dplyr::semi_join(cv %>% dplyr::select(database, pathway), by = c("database", "pathway")) %>%
      dplyr::group_by(database, pathway) %>%
      dplyr::summarise(p.adj = min(p.adj, na.rm = TRUE), .groups = "drop")
    dp <- cv %>%
      dplyr::left_join(best_padj, by = c("database", "pathway")) %>%
      mutate(label = stringr::str_trunc(description, 55),
             neglog10FDR = -log10(pmax(p.adj, 1e-300)))
    term_levels <- dp %>% arrange(signal) %>% pull(label) %>% unique()
    p <- ggplot(dp, aes(x = signal, y = factor(label, levels = term_levels),
                        size = num.genes, colour = neglog10FDR)) +
      geom_point(alpha = 0.85) +
      scale_colour_viridis_c(name = "-log10 BH", option = "C") +
      scale_size_continuous(name = "set genes") +
      labs(title = sprintf("Top convergent terms across modules — %s (%s)", params$traitname, dr),
           subtitle = "Rows ordered top-to-bottom by orthogonal signal",
           x = "signal (orthogonal composite)", y = NULL) +
      theme_minimal(base_size = 10) +
      theme(axis.text.y = element_text(colour = "black",
                                        size = pmax(8, pmin(11, 220 / length(term_levels)))))
    # Save to PNG + embed via an image tag rather than print()'ing the plot
    # object directly: inside a results='asis' loop, knitr flushes print()'d
    # plots through its plot hook AFTER the surrounding text stream, which
    # shifts them (and everything printed after, incl. the next tab's
    # heading) by one position -- exactly the "next tab renders underneath"
    # symptom. Same fix already used in 15.STRING_report.Rmd's network plots.
    .png_path <- file.path("compare_results", sprintf("xmod_dotplot_%s.png", dr))
    ggsave(.png_path, p, width = 9, height = max(4, length(term_levels) * 0.35 + 2), dpi = 150)
    cat(sprintf("![](%s)\n\n", .png_path))
  } else {
    cat("_No significant terms to display._\n\n")
  }
}
```

## Cross-module highlights: Top vs Bottom scatter

Each term's **best orthogonal signal** (percentile concentration + cross-module
corroboration, out of caas/rer_accelerating/rer_decelerating) plotted as
best signal in the **bottom** direction (y axis) against best signal in the
**top** direction (x axis). The colour class is driven by cross-module
convergence (≥2 rows significant in that direction) rather than by a single
module's own call. Only cross-module-significant terms are shown (Shared /
Top only / Bottom only), top `r params$top_n` of each class.

```{r xmod_scatter, eval=has_multi, fig.height=6}
xmod_wide <- best %>%
  group_by(direction, database, pathway) %>%
  summarise(signal = max(signal, na.rm = TRUE),
            n_modules_sig = dplyr::n_distinct(module_label[sig]),
            description = dplyr::first(description),
            .groups = "drop") %>%
  mutate(signal = ifelse(is.finite(signal), signal, NA_real_)) %>%
  tidyr::pivot_wider(id_cols = c(database, pathway, description),
                      names_from = direction, values_from = c(signal, n_modules_sig),
                      names_sep = "_")

req_cols <- c("signal_top", "signal_bottom", "n_modules_sig_top", "n_modules_sig_bottom")
for (rc in setdiff(req_cols, names(xmod_wide))) xmod_wide[[rc]] <- NA

xmod_wide <- xmod_wide %>%
  mutate(
    n_modules_sig_top    = dplyr::coalesce(n_modules_sig_top, 0L),
    n_modules_sig_bottom = dplyr::coalesce(n_modules_sig_bottom, 0L),
    class = dplyr::case_when(
      n_modules_sig_top >= 2 & n_modules_sig_bottom >= 2 ~ "Shared",
      n_modules_sig_top >= 2                              ~ "Top only",
      n_modules_sig_bottom >= 2                            ~ "Bottom only",
      TRUE                                                 ~ "Neither"
    )
  )

readr::write_tsv(xmod_wide, "compare_results/cross_module_top_vs_bottom.tsv")

# Only cross-module-significant terms, top_n per class (by n_modules_sig
# summed across directions, then combined signal as tiebreaker).
pl <- xmod_wide %>%
  filter(class != "Neither") %>%
  mutate(
    signal_top    = dplyr::coalesce(signal_top, 0),
    signal_bottom = dplyr::coalesce(signal_bottom, 0),
    rank_score = (n_modules_sig_top + n_modules_sig_bottom) + (signal_top + signal_bottom) / 1000
  ) %>%
  group_by(class) %>%
  slice_max(rank_score, n = params$top_n, with_ties = FALSE) %>%
  ungroup()

p <- ggplot(pl, aes(signal_top, signal_bottom, colour = class)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(alpha = 0.7, size = 2) +
  scale_colour_manual(values = QUAD_COLS, name = NULL) +
  labs(title = sprintf("Cross-module top vs bottom signal — %s", params$traitname),
       subtitle = sprintf("Cross-module-significant terms only, top %d per class", params$top_n),
       x = "Best signal (top)", y = "Best signal (bottom)") +
  theme_minimal(base_size = 12)
if (has_pkg("ggrepel") && nrow(pl) > 0) {
  p <- p + ggrepel::geom_text_repel(
    data = pl, aes(label = stringr::str_trunc(description, 40)), size = 2.6, max.overlaps = Inf, seed = 42)
}
print(p)
```

---

# STRING: Top vs Bottom {.tabset .tabset-fade}

`r if (!has_string) "> **No STRING results available — skipping.**"`

```{r string_compare, eval=has_string, results='asis'}
dir_col <- intersect(c("direction", "gene_list", "list_name"), names(string_all))[1]
if (is.na(dir_col)) {
  cat("_STRING results lack a direction column; showing combined table._\n\n")
  print(kbl_html(head(string_all, params$top_n), height = "420px"))
} else {
  sa <- string_all %>% mutate(direction = tolower(as.character(.data[[dir_col]])),
                              direction = dplyr::case_when(grepl("top", direction) ~ "top",
                                                           grepl("bottom|bot", direction) ~ "bottom",
                                                           TRUE ~ "other"))
  for (d in c("top", "bottom")) {
    sub <- sa %>% filter(direction == d) %>% head(params$top_n)
    cat(sprintf("\n## %s\n\n", d))
    if (nrow(sub) == 0) {
      cat("_None._\n\n")
      next
    }
    
    has_pct <- "pct_gate_sig" %in% names(sub)
    if (has_pct) {
      if (d == "top") {
        display_df <- sub %>%
          dplyr::select(
            Category = category,
            Term = description,
            FDR = fdr,
            `gate_sig (%)` = pct_gate_sig,
            `gate_fdr (%)` = pct_gate_fdr,
            `top10% (%)` = pct_top10,
            `top5% (%)` = pct_top5,
            `top1% (%)` = pct_top1,
            `fade_top (%)` = pct_fade_top,
            `rer_acc (%)` = pct_rer_acc,
            `rer_decc (%)` = pct_rer_decc,
            `accum (%)` = pct_accum
          )
      } else {
        display_df <- sub %>%
          dplyr::select(
            Category = category,
            Term = description,
            FDR = fdr,
            `gate_sig (%)` = pct_gate_sig,
            `gate_fdr (%)` = pct_gate_fdr,
            `bot10% (%)` = pct_bottom10,
            `bot5% (%)` = pct_bottom5,
            `bot1% (%)` = pct_bottom1,
            `fade_bottom (%)` = pct_fade_bottom,
            `rer_acc (%)` = pct_rer_acc,
            `rer_decc (%)` = pct_rer_decc,
            `accum (%)` = pct_accum
          )
      }
    } else {
      display_df <- sub
    }
    
    print(kbl_html(display_df, height = "420px"))
  }
}
```

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — Enrichment Module*
*Date of generation: `r Sys.Date()`*
