---
title: "FCS Enrichment (Multi-Method) - `r params$project_name`"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
    mathjax: null                       # no math in this report
    md_extensions: "-tex_math_dollars"  # gene-set names contain '$' (e.g. V$HNF4_01) - never parse as TeX math
params:
  stats_file:   NULL          # TSV: gene + one-or-more score_<ranking> cols + optional flag_<name> cols
  universe_file: NULL         # cleaned_background (one gene symbol per line)
  gmt_dir:      NULL          # directory of curated *.gmt files
  project_name: "FCS"
  num_g:        10            # min genes per set (fastwilcoxGMT num.g)
  max_g:        500           # max genes per set (0 / Inf = no limit)
  fdr_thr:      0.15          # BH (adjusted-p) cutoff - RERconverge categorical default (msae210)
  pperm_thr:    0.025         # permulation-p cutoff - applied in addition to fdr_thr when perms exist
  top_n:        20            # rows shown per database x ranking
  leading_edge_top_n: 100     # rows shown in the pooled (all-database) leading-edge
                              # composition table, deliberately more generous than
                              # top_n, which caps a single database's slice, not the
                              # whole significant set for a ranking. A term that IS
                              # significant but ranks outside this cap will silently
                              # disappear from the on-page table (though never from
                              # fcs_results/fcs_leading_edge.tsv, which is unbounded).
  traitname:    "trait"
  perms_file:   "NO_FILE"
  n_perms_sum:  10000          # permulation replicates for Path Sum Permulation (p-value floor = 1/(n+1))
  annot_file:   "NO_FILE"     # optional cross-module annotation overlay (SCORING fcs_stats.tsv):
                              # supplies flag_* + CAAS score_top/score_bottom for the leading edge.
                              # When NO_FILE, annotation falls back to the stats file itself.
  gene_lists_dir: NULL        # SCORING's published gene_lists/slice_<dir><pct>.tsv dir, when
                              # supplied, gene percentile membership is read from there instead of
                              # re-ranking score_top/score_bottom here. ONLY pass this for the CAAS
                              # report (SCORING_FCS_REPORT), never for RER's (RER_FCS_REPORT),
                              # whose score_top/score_bottom are RER's own ranking, not CAAS's.
  seed:         "1998"
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 7, dev = "png", dpi = 200)
suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
  library(kableExtra)
  library(patchwork)
})

theme_set(
  theme_minimal(base_size = 11) +
  theme(
    axis.title.y  = element_text(margin = ggplot2::margin(l = 20, r = 20)),
    axis.title.x  = element_text(margin = ggplot2::margin(t = 12)),
    axis.text.x   = element_text(margin = ggplot2::margin(t = 5)),
    plot.title    = element_text(face = "bold", hjust = 0, margin = ggplot2::margin(b = 6)),
    plot.subtitle = element_text(colour = "grey40", hjust = 0, margin = ggplot2::margin(b = 10)),
    legend.position = "right",
    legend.box.margin = ggplot2::margin(t = 5)
  )
)
filter <- dplyr::filter; select <- dplyr::select; rename <- dplyr::rename
mutate <- dplyr::mutate; arrange <- dplyr::arrange
`%||%` <- function(a, b) if (!is.null(a) && length(a) > 0 && nzchar(as.character(a)[1])) a else b

seed_val <- if (exists("params") && !is.null(params$seed) && nzchar(as.character(params$seed))) as.integer(params$seed) else 1998
set.seed(seed_val)

this_dir <- tryCatch(dirname(normalizePath(knitr::current_input(), mustWork = FALSE)),
                     error = function(e) ".")
src_candidates <- c(file.path(this_dir, "src", "fcs_enrich.R"), "src/fcs_enrich.R", file.path(this_dir, "fcs_enrich.R"), "fcs_enrich.R")
src <- src_candidates[file.exists(src_candidates)][1]
if (is.na(src)) stop("fcs_enrich.R not found under src/ or next to FCS_general.Rmd")
suppressMessages(invisible(capture.output(source(src))))

kbl_html <- function(df, height = NULL, escape = TRUE, caption = NULL, ...) {
  k <- kable(df, format = "html", escape = escape, caption = caption, ...) %>%
    kable_styling(bootstrap_options = c("striped", "condensed", "hover"), full_width = FALSE)
  if (!is.null(height)) k <- scroll_box(k, height = height)
  k
}
```

```{r load, include=FALSE}
stopifnot(!is.null(params$stats_file), file.exists(params$stats_file))
stopifnot(!is.null(params$universe_file), file.exists(params$universe_file))

stats <- readr::read_tsv(params$stats_file, show_col_types = FALSE)
if (!"gene" %in% names(stats)) {
  gcol <- intersect(c("Gene", "GENE"), names(stats))[1]
  if (!is.na(gcol)) stats <- dplyr::rename(stats, gene = !!gcol)
}
stopifnot("gene" %in% names(stats))

# Universe = cleaned_background (all tested genes). Fall back to the scored genes
# if no valid background was provided (e.g. a Nextflow sentinel placeholder).
universe <- character(0)
if (!is.null(params$universe_file) && file.exists(params$universe_file) &&
    !grepl("^NO_", basename(params$universe_file))) {
  universe <- unique(trimws(readLines(params$universe_file)))
  universe <- universe[nzchar(universe)]
}
if (length(universe) < 2) {
  message("No valid universe file - falling back to scored genes as the universe.")
  universe <- unique(stats$gene)
}

score_cols <- grep("^score_", names(stats), value = TRUE)
if (length(score_cols) == 0) stop("stats_file has no score_<ranking> columns")
ranking_names_raw <- sub("^score_", "", score_cols)
# Canonical tab order: Global > Top > Bottom > Accelerating > Decelerating, the
# same convention used throughout AMI/Comparison/Posenrich reports. Column order
# in stats_file is incidental and must not drive tab order. Any ranking name
# outside this known set is appended at the end, in its original order.
RANKING_ORDER <- c("global", "top", "bottom", "accelerating", "decelerating")
ranking_names <- c(intersect(RANKING_ORDER, ranking_names_raw),
                   setdiff(ranking_names_raw, RANKING_ORDER))

# Annotation overlay: cross-module leading-edge flags (+ CAAS directional scores for
# the top/bottom percentiles) come from a separate file when provided, e.g. SCORING's
# fcs_stats.tsv, so a per-module report (RER/FADE/accum) ranked by its OWN score can
# still be annotated with CAAS gates / FADE / accumulation membership. RANKINGS always
# come from `stats`; only annotation comes from `annot`. Falls back to `stats` itself
# (the CAAS report carries its own flags), so the un-annotated path is unchanged.
annot <- stats
if (!is.null(params$annot_file) && file.exists(params$annot_file) &&
    !grepl("^NO_", basename(params$annot_file))) {
  annot <- readr::read_tsv(params$annot_file, show_col_types = FALSE)
  if (!"gene" %in% names(annot)) {
    gcol <- intersect(c("Gene", "GENE"), names(annot))[1]
    if (!is.na(gcol)) annot <- dplyr::rename(annot, gene = !!gcol)
  }
}
flag_cols <- grep("^flag_", names(annot), value = TRUE)

gmt_dir <- params$gmt_dir %||% file.path(this_dir, "..", "dat")
invisible(capture.output(suppressMessages(gmts <- fcs_load_gmts(gmt_dir))))
desc_map <- suppressMessages(fcs_load_descriptions(gmt_dir))
```

# Overview

```{r overview, echo=FALSE}
tibble::tibble(
  Field = c("Project", "Trait", "Genes (stats)", "Universe genes", "Rankings",
            "Flags", "GMT databases", "num.g (min set)", "max.g (max set)", "BH threshold"),
  Value = c(params$project_name, params$traitname, nrow(stats), length(universe),
            paste(ranking_names, collapse = ", "),
            if (length(flag_cols)) paste(sub("^flag_", "", flag_cols), collapse = ", ") else "(none)",
            length(gmts), params$num_g,
            ifelse(is.null(params$max_g) || params$max_g <= 0 || !is.finite(params$max_g), "(none)", as.character(params$max_g)),
            params$fdr_thr)
) %>% kbl_html()
```

# Methods

This report tests every gene set in every GMT database with **three
complementary methods**, each sensitive to a different way a pathway can show
signal in the ranking:

* **Wilcoxon-AUC FCS** (`RERconverge::fastwilcoxGMTall`) - a rank-sum test:
  do this pathway's genes sit systematically higher in the ranking than the
  rest of the universe? Its statistic, **AUC** (`W.AUC`, `stat + 0.5`, range
  0–1), is the probability that a random pathway gene outranks a random
  background gene; 0.5 = no shift. Significance is BH-adjusted p (`W.adj`)
  below the threshold, and - when permulations are available - permulation p
  (`W.perm`) below `r params$pperm_thr` as well.
* **Lachenbruch Two-Part** - built for the zero-inflated shape of these
  rankings, where most genes carry no signal at all (score = 0). It splits
  the question in two: a Fisher exact test on *prevalence* (are pathway genes
  over-represented among the genes that scored above zero at all?) and a
  Wilcoxon test on *magnitude* (among only the genes that scored above zero,
  do pathway genes score higher?), combined into one χ²(df=2) p-value
  (`L.adj`). Its statistic, **`L.frac_mag`**, is the fraction of that
  combined signal coming from the magnitude half rather than the prevalence
  half - a set near 1 is driven by a few strongly-scoring genes, a set near 0
  by many weakly-scoring ones.
* **Path Sum Permulation** - a label-permutation test on the raw summed
  pathway score: the genes' score labels are shuffled `r params$n_perms_sum %||% 10000`
  times to build a null distribution for the pathway's total score, and the
  observed sum is compared against it. Its statistic, **NES**, is a z-score
  (`(observed − null mean) / null sd`); significance is BH-adjusted p
  (`P.adj`) below the threshold with `NES > 0`.

Each method casts one vote (its own FDR gate, plus direction where
applicable); a set passing **3/3** is labelled **Hard evidence**, **2/3**
**Supported**, **1/3** **Exploratory** (`evidence_count` / `evidence_label`).
Lachenbruch and Permulation only run for non-negative (zero-floored)
rankings - a signed ranking (e.g. RER's two-sided global statistic) is
Wilcoxon-only, since "score > 0" carries no "has signal" meaning there.

Running all three matters because each fails differently: Wilcoxon alone can
be diluted by the many zero-floored background genes in a large universe;
Lachenbruch's split exposes *why* a set is significant (a handful of extreme
genes vs. broad participation) but is exact/analytic rather than
permutation-based; Permulation is a direct simulation check that doesn't rely
on Wilcoxon's asymptotic approximation. A term that all three agree on is
much harder to explain away than one resting on a single test.

**Ranking convention.** Wherever this report ranks or orders significant
terms - the Multi-Method Evidence table, the per-database Enriched gene set
tables, the Overall dotplot, and the Leading-edge composition table - it uses
the same three-key convention: **evidence tier first** (Hard > Supported >
Exploratory), **then `evidence_score`**, **then alphabetical** by description
as a final tie-break.

`evidence_score` exists because evidence tier alone leaves every term in the
same tier tied - e.g. every "Supported" (2/3) term looks identical unless
something breaks the tie. Rather than average the three methods' raw
statistics (AUC and `L.frac_mag` are bounded to [0,1], but NES is an
unbounded z-score that can run into double digits, so it would silently
dominate any such average), each **passing** method's statistic is first
converted to its own percentile rank (0–1) among every term tested in that
ranking - putting all three on the same unitless scale - and a method that
didn't pass its own gate contributes 0. Summing the three gives a 0–3 score:
a continuous version of `evidence_count` that still rewards passing more
tests, but among terms tied on tier also rewards being relatively stronger
on the tests that did pass. It is a ranking heuristic, not a statistic in
its own right - the evidence tier, not `evidence_score`, is what carries the
actual significance claim.

```{r run, include=FALSE}
# One zero-floored ranking vector per score_<ranking> column.
rankings <- list()
for (sc in score_cols) {
  rk <- sub("^score_", "", sc)
  rankings[[rk]] <- fcs_build_vals(setNames(stats[[sc]], stats$gene), universe)
}
enrich <- suppressMessages(fcs_run_all(rankings, gmts, num_g = params$num_g,
                                       max_g        = params$max_g,
                                       perms_file   = params$perms_file,
                                       fdr_thr      = params$fdr_thr,
                                       p_perm_thr   = params$pperm_thr,
                                       n_perms_sum  = params$n_perms_sum %||% 10000)) %>%
  dplyr::left_join(desc_map, by = c("database", "pathway")) %>%
  dplyr::mutate(description = dplyr::coalesce(description, pathway))

# Per-gene attribute table: annotation flags + derived directional percentile flags.
# Sourced from `annot` (the overlay) so per-module reports inherit CAAS/FADE/accum
# annotation; `annot` defaults to `stats` when no overlay is supplied.
pct_flags <- fcs_percentile_flags(annot, gene_lists_dir = params$gene_lists_dir)
attr_df <- pct_flags
if (length(flag_cols)) attr_df <- dplyr::left_join(pct_flags, annot[, c("gene", flag_cols)], by = "gene")
attr_cols_all <- setdiff(names(attr_df), "gene")

# Restrict the leading edge to the genes that actually DRIVE each enrichment,
# i.e. pathway members with signal (score > 0) in that ranking. fastwilcoxGMT's
# gene.vals lists every member in the universe (incl. zero-floored background
# genes); summarising over those would dilute the composition with non-drivers
# and leave their flags undefined. score_long lets us keep only the drivers.
score_long <- stats %>%
  dplyr::select(gene, dplyr::all_of(score_cols)) %>%
  tidyr::pivot_longer(-gene, names_to = "ranking", values_to = "drv_score") %>%
  dplyr::mutate(ranking = sub("^score_", "", ranking))
leading <- fcs_annotate_leading_edge(enrich, attr_df) %>%
  dplyr::left_join(score_long, by = c("gene", "ranking")) %>%
  dplyr::filter(!is.na(drv_score) & drv_score > 0) %>%
  dplyr::select(-drv_score)

# Dual-threshold significance (RERconverge categorical convention, msae210):
#   adjusted p < fdr_thr AND - when permulations are available - permulation p <
#   pperm_thr. The permulation p corrects for non-independence in gene ranks; a low
#   value indicates the BH significance is driven by the phenotype, not other
#   correlation in the data. Where no permulations exist (perms_file = NO_FILE), the
#   permulation clause is vacuous and significance falls back to the BH cutoff alone.
#
#   CAVEAT - read p.perm as "beyond a NAIVE random relabeling". The CAAS
#   permulation null permutes the phenotype without preserving the observed
#   matched-sister-pair design: the permuted fg/bg species are paired at RANDOM
#   across the tree, so pair MRCAs are deep and the ASR path-score's
#   isolation/independence geometry differs from the observed (which uses shallow
#   matched pairs). We cannot preserve matched pairs under permutation - a random
#   labeling almost never yields N independent matched contrasts - so this is by
#   design; the up-side is a homogeneous, well-defined null. But it means p.perm
#   tests "phenotype + matched-pair design" jointly (mildly anti-conservative on
#   the MRCA axes), not the phenotype in isolation. A within-pair fg/bg swap would
#   isolate the phenotype but only spans 2^n labelings (too coarse unless n is
#   large). See docs / scoring_caas_perms.R.
# evidence_count / evidence_label / sig_* already computed by fcs_run_all().
# sig = evidence_count >= 1 (any method passed), used to gate dotplots + leading edge.
# For the Wilcoxon-only gate we still need p.perm-awareness in the tables below.
#
# evidence_rank / evidence_score: the shared ranking convention explained in
# "Methods" above (evidence tier, then evidence_score, then alphabetical
# description), computed once here so every table/plot below orders
# identically instead of drifting.
#
# evidence_score sums each PASSING method's own statistic converted to a
# percentile rank (0-1) among all terms tested in the same ranking, so the
# three methods contribute on a common, unitless scale instead of raw AUC
# (0-1) / L.frac_mag (0-1) / NES (unbounded, easily >10) being mixed
# directly, a term with only a huge NES could otherwise outrank a 3/3
# term whose AUC and L.frac_mag are both near their own maximum. A method
# that didn't pass its own significance gate for this term contributes 0,
# so evidence_score (range 0-3) is a continuous refinement of
# evidence_count (0-3, integer), same idea, just without the ties a plain
# pass-count leaves between every term that clears the same number of gates.
enrich <- enrich %>%
  dplyr::group_by(ranking) %>%
  dplyr::mutate(
    .auc_pct = dplyr::percent_rank(stat + 0.5),
    .lfm_pct = dplyr::percent_rank(lach_frac_magnitude),
    .nes_pct = dplyr::percent_rank(perm_nes)
  ) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    sig = evidence_count >= 1L,
    evidence_rank = dplyr::case_when(
      evidence_label == "Hard evidence" ~ 1L,
      evidence_label == "Supported"     ~ 2L,
      evidence_label == "Exploratory"   ~ 3L,
      TRUE                              ~ 4L
    ),
    evidence_score = dplyr::if_else(sig_wilcoxon %in% TRUE,    dplyr::coalesce(.auc_pct, 0), 0) +
                     dplyr::if_else(sig_lachenbruch %in% TRUE, dplyr::coalesce(.lfm_pct, 0), 0) +
                     dplyr::if_else(sig_permulation %in% TRUE, dplyr::coalesce(.nes_pct, 0), 0)
  ) %>%
  dplyr::select(-.auc_pct, -.lfm_pct, -.nes_pct)
has_perm_any <- any(!is.na(enrich$p.perm))
n_sig <- sum(enrich$sig, na.rm = TRUE)

dir.create("fcs_results", showWarnings = FALSE)
readr::write_tsv(enrich %>% select(-any_of("gene.vals")), "fcs_results/fcs_all_results.tsv")
readr::write_tsv(leading, "fcs_results/fcs_leading_edge.tsv")
```

```{r banner, echo=FALSE, results='asis'}
n_hard  <- sum(enrich$evidence_count == 3L, na.rm = TRUE)
n_supp  <- sum(enrich$evidence_count == 2L, na.rm = TRUE)
n_expl  <- sum(enrich$evidence_count == 1L, na.rm = TRUE)
n_dbs   <- dplyr::n_distinct(enrich$database)
cat(sprintf(
  "**%d hard-evidence** (3/3) | **%d supported** (2/3) | **%d exploratory** (1/3) gene sets across %d databases.\n\n",
  n_hard, n_supp, n_expl, n_dbs))
cat("_Methods: Wilcoxon-AUC FCS &nbsp;|&nbsp; Lachenbruch Two-Part &nbsp;|&nbsp; Path Sum Permulation. FDR threshold: ",
    params$fdr_thr, "_\n\n")
```

# Multi-Method Evidence {.tabset .tabset-fade}

All terms where **at least one test** passed its significance gate, grouped by
direction. Columns: `W.AUC` (Wilcoxon effect size), `W.adj` (Wilcoxon BH p), `W.perm` (Wilcoxon permulation p),
`L.adj` (Lachenbruch BH p), `L.frac_mag` (fraction of Lachenbruch signal from
magnitude vs prevalence), `P.adj` (Path Sum Permulation BH p), `NES` (Permulation
normalised enrichment score). NA = test not applicable (signed ranking) or
not run. Green = Hard evidence (3/3), orange = Supported (2/3), grey = Exploratory (1/3).
Rows follow the ranking convention from Methods above: evidence tier, then
mean(`W.AUC`, `L.frac_mag`, `NES`), then alphabetical.

```{r evidence_table, results='asis'}
evidence_colours <- c(
  "Hard evidence" = "#27ae60",
  "Supported"     = "#e67e22",
  "Exploratory"   = "#7f8c8d"
)
for (rk in ranking_names) {
  cat(sprintf("\n## %s\n\n", rk))
  has_w_perm <- any(!is.na(enrich$p.perm))
  ev <- enrich %>%
    dplyr::filter(ranking == rk, evidence_count >= 1L) %>%
    dplyr::arrange(evidence_rank, dplyr::desc(evidence_score), description) %>%
    dplyr::mutate(
      Description  = stringr::str_trunc(description, 60),
      W.AUC        = round(stat + 0.5, 3),
      W.adj        = signif(p.adj, 3),
      W.perm       = if (has_w_perm) signif(p.perm, 3) else NA_real_,
      L.adj        = signif(lach_p.adj, 3),
      L.frac_mag   = round(lach_frac_magnitude, 2),
      P.adj        = signif(perm_p.adj, 3),
      NES          = round(perm_nes, 2),
      Wilcoxon     = ifelse(dplyr::coalesce(sig_wilcoxon, FALSE), "X", ""),
      Lachenbruch  = ifelse(dplyr::coalesce(sig_lachenbruch, FALSE), "X", ""),
      Permulation  = ifelse(dplyr::coalesce(sig_permulation, FALSE), "X", "")
    ) %>%
    dplyr::select(evidence_label, pathway, Description, num.genes,
                  Wilcoxon, Lachenbruch, Permulation,
                  W.AUC, W.adj, any_of("W.perm"), L.adj, L.frac_mag, P.adj, NES)
  if (nrow(ev) == 0) { cat("_No terms passed any significance gate._\n"); next }
  # Colour the evidence_label column via kableExtra cell_spec
  ev$evidence_label <- kableExtra::cell_spec(
    ev$evidence_label, "html",
    color = "white",
    background = evidence_colours[ev$evidence_label],
    bold = TRUE, extra_css = "padding: 2px 6px; border-radius: 3px;"
  )
  print(kbl_html(ev, escape = FALSE, height = if (nrow(ev) > 15) "460px" else NULL))
}
```

# Enriched gene sets {.tabset .tabset-fade}

Displays the top `r params$top_n` positively-enriched sets per database (`stat` > 0, i.e. Wilcoxon AUC > 0.5), ordered by the ranking convention from Methods above (evidence tier, then mean of the methods' own statistics, then alphabetical). Significance is evaluated independently by three complementary tests: **Wilcoxon** (rank-sum AUC / BH p), **Lachenbruch** (two-part test), and **Permulation** (path-sum null model), each against its own threshold. The **Wilcoxon**, **Lachenbruch**, and **Permulation** columns indicate which tests passed (`X`).

```{r tables, results='asis'}
# `top_sel[[rk]]` captures the exact rows (full precision, table order) shown per
# database below, so the Overall dotplot section can pool from precisely the same
# selection instead of re-deriving its own.
top_sel <- list()
for (rk in ranking_names) {
  cat(sprintf("\n## %s {.tabset}\n", rk))
  sub <- enrich %>% filter(ranking == rk, stat > 0)
  if (nrow(sub) == 0) { cat("\n_No positively-enriched sets._\n"); top_sel[[rk]] <- sub; next }

  # Ranking convention from Methods above: evidence tier, then evidence_score desc,
  # then alphabetical description.
  sub <- sub %>% dplyr::arrange(evidence_rank, dplyr::desc(evidence_score), description)

  db_sel <- sub %>% dplyr::group_by(database) %>% dplyr::slice_head(n = params$top_n) %>% dplyr::ungroup()
  top_sel[[rk]] <- db_sel

  for (db in unique(sub$database)) {
    dbres <- db_sel %>% filter(database == db) %>%
      mutate(AUC = round(stat + 0.5, 3),
             Description = stringr::str_trunc(description, 70),
             pval = signif(pval, 3),
             p.adj = signif(p.adj, 3))
    n_hard_db  <- sum(dbres$evidence_count == 3L, na.rm = TRUE)
    n_supp_db  <- sum(dbres$evidence_count == 2L, na.rm = TRUE)
    n_expl_db  <- sum(dbres$evidence_count == 1L, na.rm = TRUE)

    has_lach <- any(!is.na(dbres$lach_p.adj))
    has_perm_rk <- any(!is.na(dbres$perm_p.adj))
    has_w_perm_db <- any(!is.na(dbres$p.perm))

    disp <- dbres %>%
      mutate(evidence = evidence_label,
             W.AUC = round(stat + 0.5, 3),
             W.adj = signif(p.adj, 3),
             W.perm = if (has_w_perm_db) signif(p.perm, 3) else NA_real_,
             L.adj = if (has_lach) signif(lach_p.adj, 3) else NA_real_,
             L.frac_mag = if (has_lach) round(lach_frac_magnitude, 2) else NA_real_,
             P.adj = if (has_perm_rk) signif(perm_p.adj, 3) else NA_real_,
             NES   = if (has_perm_rk) round(perm_nes, 2) else NA_real_,
             Wilcoxon    = ifelse(dplyr::coalesce(sig_wilcoxon, FALSE), "X", ""),
             Lachenbruch = ifelse(dplyr::coalesce(sig_lachenbruch, FALSE), "X", ""),
             Permulation = ifelse(dplyr::coalesce(sig_permulation, FALSE), "X", "")) %>%
      select(evidence, pathway, Description, num.genes, Wilcoxon, Lachenbruch, Permulation,
             W.AUC, W.adj, any_of(c("W.perm", "L.adj", "L.frac_mag", "P.adj", "NES")))

    cat(sprintf("\n### %s\n\n", db))
    cat(sprintf("_%d shown - hard: %d | supported: %d | exploratory: %d._\n\n",
                nrow(disp), n_hard_db, n_supp_db, n_expl_db))
    print(kbl_html(disp, height = if (nrow(disp) > 15) "420px" else NULL))
  }
}
```

# Overall dotplot {.tabset .tabset-fade}

One dotplot per direction, pooling across all databases and restricted to the
top `r params$top_n` significant terms, ordered top-to-bottom by the same
ranking convention as the tables above (evidence tier, then evidence_score,
then alphabetical). Each term shows up to three horizontal arms - one per
method that ran - laid out on a **shared x-axis of `-log10(BH-adjusted p)`**
(so all three methods' significance is directly comparable on one scale: the
BH threshold is the same dashed line for all of them). An arm's **colour**
encodes that method's own effect-size statistic (AUC for Wilcoxon, `L.frac_mag`
for Lachenbruch, NES for Permulation) - light-to-dark within that method's own
colour family, darker meaning a larger statistic among the terms shown. Faint
(low-alpha) arms mark a method that ran but did not pass its own significance
gate for that term. The **tip marker** sits at the term's single
most-significant arm, drawn in a dark, saturated shade of that arm's colour
family (bigger than the arm dots) so the strongest single piece of evidence
per term is immediately visible; its shape encodes the evidence tier
(circle = Hard, triangle = Supported, square = Exploratory) and its size the
gene-set size.

```{r overall_dotplot_prep, include=FALSE}
DIR_MAP <- c(top = "top", accelerating = "top", bottom = "bottom", decelerating = "bottom")
evidence_shapes <- c("Hard evidence" = 16, "Supported" = 17, "Exploratory" = 15)

# Darkens a hex colour toward black by `amount` (0-1), used so the tip
# marker reads as "this arm's own colour, just deeper" rather than jumping to
# an unrelated fixed anchor shade.
.darken_hex <- function(hex, amount = 0.3) {
  rgb <- grDevices::col2rgb(hex) / 255
  rgb <- rgb * (1 - amount)
  grDevices::rgb(rgb[1, ], rgb[2, ], rgb[3, ])
}

# A standalone light->dark gradient bar for one method, used as that method's
# legend key, scale_colour_identity() can only draw flat swatches, so the
# arm-colour ramp itself is otherwise invisible in the legend.
.method_gradient_legend <- function(title, light, dark) {
  df <- data.frame(y = 1, v = seq(0, 1, length.out = 100))
  ggplot(df, aes(x = v, y = y, fill = v)) +
    geom_tile() +
    scale_fill_gradientn(colours = c(light, dark), guide = "none") +
    scale_x_continuous(breaks = c(0, 1), labels = c("low", "high")) +
    scale_y_continuous(breaks = NULL) +
    labs(title = title, x = NULL, y = NULL) +
    theme_minimal(base_size = 8) +
    theme(plot.title = element_text(size = 8, hjust = 0.5),
          panel.grid = element_blank(),
          plot.margin = margin(2, 2, 2, 2))
}

# Dark/light colour anchors per method, the arm colour interpolates between
# them by that method's own statistic (normalised within the terms actually
# shown); the tip marker always uses the dark anchor of its winning method, so
# "tip colour" reads as "which method, at maximum confidence" regardless of
# that method's own statistic value.
.test_hue_dark  <- c(Wilcoxon = "#08306b", Lachenbruch = "#8c2d04", Permulation = "#4a148c")
.test_hue_light <- c(Wilcoxon = "#d1e5f0", Lachenbruch = "#fee0b6", Permulation = "#e7d4e8")

# One row per (term, method that ran): x = -log10(method's own BH p) --
# a shared axis across methods, since every method produces its own adjusted
# p. colour = that method's own effect-size statistic (AUC / L.frac_mag / NES),
# mapped light-to-dark within the method's own colour family and normalised
# within the rows passed in (so colour is relative to what's on THIS plot).
.fcs_test_dotplot_data <- function(df) {
  test_specs <- tibble::tribble(
    ~test,          ~p_col,        ~sig_col,          ~raw_col,
    "Wilcoxon",     "p.adj",       "sig_wilcoxon",     "stat",
    "Lachenbruch",  "lach_p.adj",  "sig_lachenbruch",  "lach_frac_magnitude",
    "Permulation",  "perm_p.adj",  "sig_permulation",  "perm_nes"
  )
  out <- purrr::map_dfr(seq_len(nrow(test_specs)), function(i) {
    sp <- test_specs[i, ]
    if (!sp$p_col %in% names(df)) return(tibble::tibble())
    p_val   <- df[[sp$p_col]]
    sig_val <- if (sp$sig_col %in% names(df)) dplyr::coalesce(df[[sp$sig_col]], FALSE) else FALSE
    raw     <- if (sp$raw_col %in% names(df)) df[[sp$raw_col]] else NA_real_
    stat_value <- if (sp$test == "Wilcoxon") raw + 0.5 else raw
    df %>%
      dplyr::mutate(test = sp$test, test_p_adj = p_val, test_sig = sig_val,
                    test_signal = -log10(pmax(test_p_adj, 1e-300)),
                    stat_value = stat_value) %>%
      dplyr::filter(!is.na(test_p_adj), is.finite(test_signal))
  })
  if (nrow(out) == 0) return(out)
  out %>%
    dplyr::group_by(test) %>%
    dplyr::mutate(
      # .min/.rng are per-group constants, but inside a grouped mutate they are
      # stored as columns recycled to the group's length. Any test applied to them
      # must therefore be VECTORISED: a scalar `if (...) ... else ...` receives a
      # length-n logical here and, from R 4.3 on, errors outright
      # ("'length = n' in coercion to 'logical(1)'") for every group holding more
      # than one term.
      .min  = suppressWarnings(min(stat_value, na.rm = TRUE)),
      .rng  = suppressWarnings(diff(range(stat_value, na.rm = TRUE))),
      # Two groups cannot be normalised and take the mid-ramp colour instead:
      #   • no finite statistic at all - range(na.rm = TRUE) over an all-NA vector
      #     falls back to (Inf, -Inf), so .rng is -Inf rather than NA. Lachenbruch
      #     reaches this whenever every term has chi_total == 0, since
      #     fcs_enrich.R then writes lach_frac_magnitude = NA.
      #   • a single distinct value (.rng == 0), where the ramp position is
      #     arbitrary.
      .frac = dplyr::if_else(is.finite(.rng) & .rng > 0,
                             (stat_value - .min) / .rng,
                             0.7),
      # A lone NA statistic inside an otherwise-normalisable group would otherwise
      # index the colour ramp with NA below.
      .frac = dplyr::if_else(is.finite(.frac), .frac, 0.7)
    ) %>%
    dplyr::ungroup() %>%
    dplyr::rowwise() %>%
    dplyr::mutate(arm_colour = grDevices::colorRampPalette(
      c(.test_hue_light[[test]], .test_hue_dark[[test]]))(101)[max(1, min(101, round(.frac * 100) + 1))]) %>%
    dplyr::ungroup() %>%
    dplyr::select(-.rng, -.frac, -.min)
}
```

```{r overall_dotplot, results='asis', fig.height=6}
# knitr opens the figure device for a chunk ONCE, using the fig.height in
# effect at chunk start - opts_current$set(fig.height = ...) called mid-chunk
# (e.g. inside this per-direction loop) does NOT change the size of the image
# actually written out, verified empirically, it's a known knitr limitation,
# not a working per-plot override. Render+embed each plot manually instead,
# which is the only reliable way to vary height per plot in one asis chunk.
.emit_sized_plot <- function(p, height_in, width_in = knitr::opts_chunk$get("fig.width"),
                              dpi = knitr::opts_chunk$get("dpi")) {
  f <- tempfile(fileext = ".png")
  on.exit(unlink(f), add = TRUE)
  ggplot2::ggsave(f, plot = p, width = width_in, height = height_in, dpi = dpi, units = "in")
  cat(sprintf('<img src="data:image/png;base64,%s" style="max-width:100%%;" />\n\n',
              base64enc::base64encode(f)))
}

for (dr in c("global", "top", "bottom")) {
  cat(sprintf("\n## %s\n\n", tools::toTitleCase(dr)))
  rks <- if (dr == "global") {
    intersect("global", ranking_names)
  } else {
    intersect(names(DIR_MAP)[DIR_MAP == dr], ranking_names)
  }
  if (length(rks) == 0) { cat("_No ranking for this direction in this report._\n\n"); next }

  sig_terms <- enrich %>% filter(ranking %in% rks, sig)
  if (nrow(sig_terms) == 0) { cat(sprintf("_No significant terms (%s)._\n\n", dr)); next }

  # One representative row per (database, pathway): its single best-evidence
  # ranking, by the same convention as everywhere else in this report, when
  # a direction pools two rankings (e.g. top + accelerating), a term can only
  # appear once here.
  term_best <- sig_terms %>%
    dplyr::group_by(database, pathway) %>%
    dplyr::slice_min(evidence_rank, n = 1, with_ties = TRUE) %>%
    dplyr::slice_max(evidence_score, n = 1, with_ties = FALSE) %>%
    dplyr::ungroup()

  dp <- term_best %>%
    dplyr::arrange(evidence_rank, dplyr::desc(evidence_score), description) %>%
    head(params$top_n) %>%
    dplyr::mutate(label = stringr::str_trunc(dplyr::coalesce(description, pathway), 55))

  if (nrow(dp) == 0) { cat("_No terms to display._\n\n"); next }

  # y-axis index 1 = bottom of plot, reverse the ranking order so the
  # best-ranked term (tier 1, highest evidence_score) ends up at the TOP.
  term_levels <- rev(unique(dp$label))
  dp <- dp %>% dplyr::mutate(y_base = as.numeric(factor(label, levels = term_levels)))

  line_df <- .fcs_test_dotplot_data(dp) %>%
    dplyr::mutate(y = y_base + dplyr::case_when(test == "Wilcoxon" ~ 0.22,
                                                test == "Lachenbruch" ~ 0,
                                                TRUE ~ -0.22))
  if (nrow(line_df) == 0) { cat("_No test results to plot for this direction._\n\n"); next }

  # Tip marker: this term's single most-significant arm (smallest BH p across
  # the three methods), centred on the term's row and drawn in a darkened
  # shade of that same arm's own colour (not a fixed per-method anchor), so
  # the tip reads as "this arm, just deeper" rather than an unrelated shade.
  cap_df <- line_df %>%
    dplyr::group_by(pathway) %>%
    dplyr::slice_min(test_p_adj, n = 1, with_ties = FALSE) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(tip_colour = .darken_hex(arm_colour),
                  ev_shape = factor(evidence_label, levels = c("Hard evidence", "Supported", "Exploratory")))

  p <- ggplot(line_df, aes(y = y)) +
    geom_segment(aes(x = 0, xend = test_signal, yend = y, colour = arm_colour, alpha = test_sig),
                 linewidth = 1.1, lineend = "round") +
    geom_point(data = cap_df,
               aes(x = test_signal, y = y, size = num.genes, shape = ev_shape, colour = tip_colour),
               alpha = 1) +
    scale_y_continuous(breaks = seq_along(term_levels), labels = term_levels) +
    scale_alpha_manual(name = "test passed", values = c(`TRUE` = 1, `FALSE` = 0.25), drop = FALSE) +
    scale_colour_identity() +
    scale_size_continuous(name = "set genes") +
    scale_shape_manual(name = "evidence", values = evidence_shapes, drop = FALSE) +
    # Bottom legend strip (evidence / test passed / set genes, one row each).
    # A patchwork guide_area() + plot_layout(guides = "collect") placement was
    # tried first, to put the method gradient bars in a single right-hand
    # column with these guides - the collected guides render as their own
    # block beside the gradient bars instead of stacked above them, so the
    # two never actually shared one column. Bottom placement sidesteps that:
    # these guides render as part of p itself (native ggplot legend.position),
    # gradient bars are appended as a further row below the whole plot.
    guides(shape = guide_legend(order = 1, nrow = 1), alpha = guide_legend(order = 2, nrow = 1),
           size = guide_legend(order = 3, nrow = 1)) +
    geom_vline(xintercept = -log10(params$fdr_thr), linetype = "dashed", colour = "grey50") +
    labs(title = sprintf("%s - %s", params$project_name, dr),
         subtitle = "Wilcoxon, Lachenbruch, Permulation",
         x = expression(-log[10]~"BH-adjusted p"), y = NULL) +
    theme_minimal(base_size = 10) +
    theme(axis.text.y = element_text(colour = "black",
                                      size = pmax(8, pmin(11, 220 / length(term_levels)))),
          plot.subtitle = element_text(margin = margin(b = 12)),
          legend.position = "bottom",
          legend.box = "horizontal")

  # Method legend: a gradient bar per method (light = weak statistic, dark =
  # strong), scale_colour_identity() above can only draw flat swatches, so
  # the ramp itself needs its own mini legend. One row, placed directly under
  # p (which already carries its own evidence/test passed/set genes legend at
  # its bottom via legend.position = "bottom" above), so the whole legend
  # reads top-to-bottom as one column: plot legend, then gradient bars.
  method_stat_labels <- c(Wilcoxon = "AUC", Lachenbruch = "L.frac_mag", Permulation = "NES")
  gradient_row <- purrr::reduce(names(.test_hue_dark), function(acc, m) {
    g <- .method_gradient_legend(sprintf("%s (%s)", m, method_stat_labels[[m]]),
                                  .test_hue_light[[m]], .test_hue_dark[[m]])
    if (is.null(acc)) g else acc | g
  }, .init = NULL)

  # Fixed absolute height for the gradient row (not a proportion of the whole
  # figure) - a relative split (e.g. heights = c(6, 1)) makes the bar grow
  # right along with the main plot on term-heavy reports, ending up huge;
  # a thin colour ramp needs the same ~0.3in regardless of plot height.
  combo <- p / gradient_row + patchwork::plot_layout(heights = grid::unit(c(1, 0.18), c("null", "in")))
  .emit_sized_plot(combo, height_in = pmax(5, pmin(22, 2.6 + 0.4 * length(term_levels))))
}
```

# Leading-edge composition {.tabset .tabset-fade}

For each significantly-enriched set (BH < `r params$fdr_thr``r if (has_perm_any) sprintf(" ∩ permulation p < %.3g", params$pperm_thr) else ""`), the
**percentage of its leading-edge genes** carrying each attribute - significance
gates, directional score percentiles, and cross-module support. Pooled across
databases per direction, **ordered by the same convention as every other table
in this report** (evidence tier, then evidence_score, then alphabetical - see
Methods above, and the Overall dotplot, whose term order matches this table's
row order) and capped at `r params$leading_edge_top_n %||% 100` rows; if a
direction has more significant sets than that, the table says so explicitly
rather than silently dropping the low-ranked ones.

```{r leadingedge, results='asis'}
# Friendly display names per attribute column.
disp_names <- c(
  pct_top25 = "top25%", pct_top10 = "top10%", pct_top5 = "top5%", pct_top1 = "top1%",
  pct_bottom25 = "bot25%", pct_bottom10 = "bot10%", pct_bottom5 = "bot5%", pct_bottom1 = "bot1%",
  flag_fade = "fade", flag_fade_top = "fade_top", flag_fade_bottom = "fade_bottom",
  flag_rer_acc = "rer_acc", flag_rer_decc = "rer_decc",
  flag_accum = "accum", flag_accum_top = "accum_top", flag_accum_bottom = "accum_bottom"
)
# Attribute columns shown per ranking (direction-aware). flag_accum (global) is
# the non-directional accumulation_all_* test; flag_accum_top/flag_accum_bottom
# are the direction-restricted accumulation_top_*/accumulation_bottom_* tests,
# same convention as flag_fade/flag_fade_top/flag_fade_bottom. Percentile
# quartet (25/10/5/1%), not a trio: 25% is a real, tested tier elsewhere
# (posenrich's own Fisher test, AMI's DOMINO search), so it belongs here too.
cols_by_rk <- list(
  global = c("pct_top25","pct_top10","pct_top5","pct_top1",
             "pct_bottom25","pct_bottom10","pct_bottom5","pct_bottom1",
             "flag_fade","flag_rer_acc","flag_rer_decc","flag_accum"),
  top    = c("pct_top25","pct_top10","pct_top5","pct_top1",
             "flag_fade_top","flag_rer_acc","flag_rer_decc","flag_accum_top"),
  bottom = c("pct_bottom25","pct_bottom10","pct_bottom5","pct_bottom1",
             "flag_fade_bottom","flag_rer_acc","flag_rer_decc","flag_accum_bottom")
)

le_comp_all <- list()  # collect leading-edge composition tables for export

for (rk in ranking_names) {
  cat(sprintf("\n## %s\n\n", rk))
  want <- cols_by_rk[[rk]] %||% attr_cols_all

  sig_terms_rk <- enrich %>% filter(ranking == rk, sig)
  if (nrow(sig_terms_rk) == 0) { cat("_No sets passed the significance thresholds for this direction._\n"); next }

  le_rk_all <- leading %>% filter(ranking == rk) %>%
    semi_join(sig_terms_rk, by = c("ranking", "database", "pathway"))
  if (nrow(le_rk_all) == 0) { cat("_No sets passed the significance thresholds for this direction._\n"); next }

  # Order by the shared convention (evidence tier, then evidence_score desc, then
  # alphabetical), computed over EVERY significant term first (not just a
  # truncated subset), so a term's rank is fixed before the table is cut.
  # Same convention as the Overall dotplot above, so a term's dotplot
  # position and its row here agree.
  ltn <- params$leading_edge_top_n %||% 100
  n_sig_rk <- nrow(sig_terms_rk)
  term_order <- sig_terms_rk %>%
    dplyr::arrange(evidence_rank, dplyr::desc(evidence_score), description) %>%
    head(ltn) %>%
    dplyr::mutate(.rank = dplyr::row_number()) %>%
    dplyr::select(database, pathway, .rank)

  le_rk <- le_rk_all %>% semi_join(term_order, by = c("database", "pathway"))

  if (n_sig_rk > ltn) {
    cat(sprintf("_Showing the top %d of %d significant sets for this direction (ranking convention above); the rest are truncated here but remain in `fcs_results/fcs_leading_edge.tsv` (unbounded) and, if within a database's own top %d, in the Enriched gene sets tables above._\n\n",
                ltn, n_sig_rk, params$top_n))
  }

  summ <- fcs_leading_edge_summary(le_rk, want) %>%
    dplyr::left_join(term_order, by = c("database", "pathway")) %>%
    dplyr::arrange(.rank) %>%
    dplyr::select(-.rank) %>%
    dplyr::left_join(sig_terms_rk %>% dplyr::select(database, pathway, evidence_label, sig_wilcoxon, sig_lachenbruch, sig_permulation), by = c("database", "pathway")) %>%
    left_join(desc_map, by = c("database", "pathway")) %>%
    mutate(Description = stringr::str_trunc(coalesce(description, pathway), 50),
           Wilcoxon    = ifelse(dplyr::coalesce(sig_wilcoxon, FALSE), "X", ""),
           Lachenbruch = ifelse(dplyr::coalesce(sig_lachenbruch, FALSE), "X", ""),
           Permulation = ifelse(dplyr::coalesce(sig_permulation, FALSE), "X", "")) %>%
    select(database, pathway, Description, n_le, any_of("evidence_label"),
           Wilcoxon, Lachenbruch, Permulation,
           any_of(intersect(want, names(.))))
  # rename attribute columns to friendly labels
  present_attr <- intersect(want, names(summ))
  names(summ)[match(present_attr, names(summ))] <- disp_names[present_attr]
  print(kbl_html(summ, height = if (nrow(summ) > 15) "480px" else NULL))

  # Collect for export (add ranking column for identification)
  le_comp_all[[rk]] <- summ %>% dplyr::mutate(ranking = rk, .before = 1)
}

# Export the full leading-edge composition table (all rankings) for
# downstream consumption by 15.Comparison_report.Rmd.
le_comp_export <- dplyr::bind_rows(le_comp_all)
if (nrow(le_comp_export) > 0) {
  readr::write_tsv(le_comp_export, "fcs_results/fcs_leading_edge_composition.tsv")
}
```

# Method

```{r method, echo=FALSE}
cat("Engines    : (1) Wilcoxon-AUC FCS | (2) Lachenbruch Two-Part | (3) Path Sum Permulation\n")
cat("Evidence   : 3/3 Hard | 2/3 Supported | 1/3 Exploratory - FDR <", params$fdr_thr, "per method\n")
cat("Wilcoxon   : RERconverge::fastwilcoxGMTall; 'greater' for zero-floored rankings, 'two.sided' for signed rankings; BH per GMT\n")
cat("Lachenbruch: Fisher exact (prevalence) + Wilcoxon (magnitude), chi2(df=2) combined; zero-floored rankings only\n")
cat("Permulation: label-permuted score-sum; NES = (obs-mu)/sd; p-floor = 1/(n_perms+1); zero-floored rankings only\n")
cat("Universe   : cleaned_background; no-signal genes floored to 0 (zero-floor)\n")
cat("MTC        : Benjamini-Hochberg, per GMT database, per method\n")
cat("Ranking    : evidence tier, then evidence_score (per-method percentile rank, zeroed when not significant, summed), then alphabetical (see Methods above)\n")
cat("n_perms_sum:", params$n_perms_sum %||% 10000, "(Path Sum Permulation replicates)\n")
```

---

## Summary

This report presents multi-method Functional Class Scoring (FCS) enrichment
results for **`r params$traitname`** across `r length(gmts)` GMT database(s)
and `r length(ranking_names)` ranking(s) (`r paste(ranking_names, collapse = ", ")`),
combining Wilcoxon-AUC, Lachenbruch Two-Part, and Path Sum Permulation tests
as described in Methods above. `r n_sig` term(s) passed at least one test's
significance gate.

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere - Enrichment Module*
*Date of generation: `r Sys.Date()`*
