---
title: "FCS Enrichment (Wilcoxon-AUC) — `r params$project_name`"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
    mathjax: null                       # no math in this report
    md_extensions: "-tex_math_dollars"  # gene-set names contain '$' (e.g. V$HNF4_01) — never parse as TeX math
params:
  stats_file:   NULL          # TSV: gene + one-or-more score_<ranking> cols + optional flag_<name> cols
  universe_file: NULL         # cleaned_background (one gene symbol per line)
  gmt_dir:      NULL          # directory of curated *.gmt files
  project_name: "FCS"
  num_g:        10            # min genes per set (fastwilcoxGMT num.g)
  fdr_thr:      0.15          # BH (adjusted-p) cutoff — RERconverge categorical default (msae210)
  pperm_thr:    0.025         # permulation-p cutoff — applied in addition to fdr_thr when perms exist
  top_n:        20            # rows shown per database x ranking
  leading_edge_top_n: 100     # rows shown in the pooled (all-database) leading-edge
                              # composition table -- deliberately more generous than
                              # top_n, which caps a single database's slice, not the
                              # whole significant set for a ranking. A term that IS
                              # significant but ranks outside this cap will silently
                              # disappear from the on-page table (though never from
                              # fcs_results/fcs_leading_edge.tsv, which is unbounded).
  traitname:    "trait"
  enrich:       FALSE         # RER-only permutation enrichment toggle
  perms_file:   "NO_FILE"
  annot_file:   "NO_FILE"     # optional cross-module annotation overlay (SCORING fcs_stats.tsv):
                              # supplies flag_* + CAAS score_top/score_bottom for the leading edge.
                              # When NO_FILE, annotation falls back to the stats file itself.
  gene_lists_dir: NULL        # SCORING's published gene_lists/slice_<dir><pct>.tsv dir -- when
                              # supplied, gene percentile membership is read from there instead of
                              # re-ranking score_top/score_bottom here. ONLY pass this for the CAAS
                              # report (SCORING_FCS_REPORT) -- never for RER's (RER_FCS_REPORT),
                              # whose score_top/score_bottom are RER's own ranking, not CAAS's.
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 7, dev = "png", dpi = 200)
suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
  library(kableExtra)
})
filter <- dplyr::filter; select <- dplyr::select; rename <- dplyr::rename
mutate <- dplyr::mutate; arrange <- dplyr::arrange
`%||%` <- function(a, b) if (!is.null(a) && length(a) > 0 && nzchar(as.character(a)[1])) a else b

this_dir <- tryCatch(dirname(normalizePath(knitr::current_input(), mustWork = FALSE)),
                     error = function(e) ".")
src_candidates <- c(file.path(this_dir, "src", "fcs_enrich.R"), "src/fcs_enrich.R", file.path(this_dir, "fcs_enrich.R"), "fcs_enrich.R")
src <- src_candidates[file.exists(src_candidates)][1]
if (is.na(src)) stop("fcs_enrich.R not found under src/ or next to FCS_general.Rmd")
suppressMessages(invisible(capture.output(source(src))))

orth_candidates <- c(file.path(this_dir, "src", "orthogonal_score.R"), "src/orthogonal_score.R",
                     file.path(this_dir, "orthogonal_score.R"), "orthogonal_score.R")
orth_src <- orth_candidates[file.exists(orth_candidates)][1]
if (is.na(orth_src)) stop("orthogonal_score.R not found under src/ or next to FCS_general.Rmd")
source(orth_src)

kbl_html <- function(df, height = NULL) {
  k <- kable(df, format = "html", escape = TRUE) %>%
    kable_styling(bootstrap_options = c("striped", "condensed", "hover"), full_width = FALSE)
  if (!is.null(height)) k <- scroll_box(k, height = height)
  k
}
```

```{r load, include=FALSE}
stopifnot(!is.null(params$stats_file), file.exists(params$stats_file))
stopifnot(!is.null(params$universe_file), file.exists(params$universe_file))

stats <- readr::read_tsv(params$stats_file, show_col_types = FALSE)
if (!"gene" %in% names(stats)) {
  gcol <- intersect(c("Gene", "GENE"), names(stats))[1]
  if (!is.na(gcol)) stats <- dplyr::rename(stats, gene = !!gcol)
}
stopifnot("gene" %in% names(stats))

# Universe = cleaned_background (all tested genes). Fall back to the scored genes
# if no valid background was provided (e.g. a Nextflow sentinel placeholder).
universe <- character(0)
if (!is.null(params$universe_file) && file.exists(params$universe_file) &&
    !grepl("^NO_", basename(params$universe_file))) {
  universe <- unique(trimws(readLines(params$universe_file)))
  universe <- universe[nzchar(universe)]
}
if (length(universe) < 2) {
  message("No valid universe file — falling back to scored genes as the universe.")
  universe <- unique(stats$gene)
}

score_cols <- grep("^score_", names(stats), value = TRUE)
if (length(score_cols) == 0) stop("stats_file has no score_<ranking> columns")
ranking_names_raw <- sub("^score_", "", score_cols)
# Canonical tab order: Global > Top > Bottom > Accelerating > Decelerating (the
# convention used throughout AMI/Comparison/Posenrich reports) -- previously this
# vector was left in whatever order score_<ranking> columns happened to appear in
# stats_file (an incidental upstream detail: column order in a TSV, not a
# deliberate choice), so e.g. a stats_file with score_bottom before score_top
# rendered tabs as Bottom/Top instead of Top/Bottom. Any ranking name outside this
# known set (shouldn't happen, but don't silently drop it) is appended at the end
# in its original order.
RANKING_ORDER <- c("global", "top", "bottom", "accelerating", "decelerating")
ranking_names <- c(intersect(RANKING_ORDER, ranking_names_raw),
                   setdiff(ranking_names_raw, RANKING_ORDER))

# Annotation overlay: cross-module leading-edge flags (+ CAAS directional scores for
# the top/bottom percentiles) come from a separate file when provided — e.g. SCORING's
# fcs_stats.tsv — so a per-module report (RER/FADE/accum) ranked by its OWN score can
# still be annotated with CAAS gates / FADE / accumulation membership. RANKINGS always
# come from `stats`; only annotation comes from `annot`. Falls back to `stats` itself
# (the CAAS report carries its own flags), so the un-annotated path is unchanged.
annot <- stats
if (!is.null(params$annot_file) && file.exists(params$annot_file) &&
    !grepl("^NO_", basename(params$annot_file))) {
  annot <- readr::read_tsv(params$annot_file, show_col_types = FALSE)
  if (!"gene" %in% names(annot)) {
    gcol <- intersect(c("Gene", "GENE"), names(annot))[1]
    if (!is.na(gcol)) annot <- dplyr::rename(annot, gene = !!gcol)
  }
}
flag_cols <- grep("^flag_", names(annot), value = TRUE)

gmt_dir <- params$gmt_dir %||% file.path(this_dir, "..", "dat")
invisible(capture.output(suppressMessages(gmts <- fcs_load_gmts(gmt_dir))))
desc_map <- suppressMessages(fcs_load_descriptions(gmt_dir))
```

# Overview

```{r overview, echo=FALSE}
tibble::tibble(
  Field = c("Project", "Trait", "Genes (stats)", "Universe genes", "Rankings",
            "Flags", "GMT databases", "num.g (min set)", "BH threshold"),
  Value = c(params$project_name, params$traitname, nrow(stats), length(universe),
            paste(ranking_names, collapse = ", "),
            if (length(flag_cols)) paste(sub("^flag_", "", flag_cols), collapse = ", ") else "(none)",
            length(gmts), params$num_g, params$fdr_thr)
) %>% kbl_html()
```

```{r run, include=FALSE}
# One zero-floored ranking vector per score_<ranking> column.
rankings <- list()
for (sc in score_cols) {
  rk <- sub("^score_", "", sc)
  rankings[[rk]] <- fcs_build_vals(setNames(stats[[sc]], stats$gene), universe)
}
enrich <- suppressMessages(fcs_run_all(rankings, gmts, num_g = params$num_g, perms_file = params$perms_file)) %>%
  dplyr::left_join(desc_map, by = c("database", "pathway")) %>%
  dplyr::mutate(description = dplyr::coalesce(description, pathway))

# Per-gene attribute table: annotation flags + derived directional percentile flags.
# Sourced from `annot` (the overlay) so per-module reports inherit CAAS/FADE/accum
# annotation; `annot` defaults to `stats` when no overlay is supplied.
pct_flags <- fcs_percentile_flags(annot, gene_lists_dir = params$gene_lists_dir)
attr_df <- pct_flags
if (length(flag_cols)) attr_df <- dplyr::left_join(pct_flags, annot[, c("gene", flag_cols)], by = "gene")
attr_cols_all <- setdiff(names(attr_df), "gene")

# Restrict the leading edge to the genes that actually DRIVE each enrichment —
# i.e. pathway members with signal (score > 0) in that ranking. fastwilcoxGMT's
# gene.vals lists every member in the universe (incl. zero-floored background
# genes); summarising over those would dilute the composition with non-drivers
# and leave their flags undefined. score_long lets us keep only the drivers.
score_long <- stats %>%
  dplyr::select(gene, dplyr::all_of(score_cols)) %>%
  tidyr::pivot_longer(-gene, names_to = "ranking", values_to = "drv_score") %>%
  dplyr::mutate(ranking = sub("^score_", "", ranking))
leading <- fcs_annotate_leading_edge(enrich, attr_df) %>%
  dplyr::left_join(score_long, by = c("gene", "ranking")) %>%
  dplyr::filter(!is.na(drv_score) & drv_score > 0) %>%
  dplyr::select(-drv_score)

dir.create("fcs_results", showWarnings = FALSE)
readr::write_tsv(enrich %>% select(-any_of("gene.vals")), "fcs_results/fcs_all_results.tsv")
readr::write_tsv(leading, "fcs_results/fcs_leading_edge.tsv")

# Dual-threshold significance (RERconverge categorical convention, msae210):
#   adjusted p < fdr_thr AND — when permulations are available — permulation p <
#   pperm_thr. The permulation p corrects for non-independence in gene ranks; a low
#   value indicates the BH significance is driven by the phenotype, not other
#   correlation in the data. Where no permulations exist (perms_file = NO_FILE), the
#   permulation clause is vacuous and significance falls back to the BH cutoff alone.
#
#   CAVEAT — read p.perm as "beyond a NAIVE random relabeling". The CAAS
#   permulation null permutes the phenotype without preserving the observed
#   matched-sister-pair design: the permuted fg/bg species are paired at RANDOM
#   across the tree, so pair MRCAs are deep and the ASR path-score's
#   isolation/independence geometry differs from the observed (which uses shallow
#   matched pairs). We cannot preserve matched pairs under permutation — a random
#   labeling almost never yields N independent matched contrasts — so this is by
#   design; the up-side is a homogeneous, well-defined null. But it means p.perm
#   tests "phenotype + matched-pair design" jointly (mildly anti-conservative on
#   the MRCA axes), not the phenotype in isolation. A within-pair fg/bg swap would
#   isolate the phenotype but only spans 2^n labelings (too coarse unless n is
#   large). See docs / scoring_caas_perms.R.
enrich <- enrich %>% dplyr::mutate(
  sig = stat > 0 & !is.na(p.adj) & p.adj < params$fdr_thr &
        (is.na(p.perm) | p.perm < params$pperm_thr)
)
has_perm_any <- any(!is.na(enrich$p.perm))
n_sig <- sum(enrich$sig, na.rm = TRUE)
```

`r if (has_perm_any) sprintf("**%d** gene sets are significantly enriched (BH < %.2g **and** permulation p < %.3g) across %d databases.", n_sig, params$fdr_thr, params$pperm_thr, dplyr::n_distinct(enrich$database)) else sprintf("**%d** gene sets are positively enriched at BH < %.2g across %d databases. _(No permulations: permulation-p threshold not applied.)_", n_sig, params$fdr_thr, dplyr::n_distinct(enrich$database))`

# Enriched gene sets {.tabset .tabset-fade}

Positive `stat` (AUC − 0.5 > 0) means the ranking's genes sit systematically
**high** for that set. Top `r params$top_n` positively-enriched sets per database,
ordered by permulation p (when available) then BH p — the same selection and order
used by the dotplots below, so every point there has a matching row here (rows
shown are not all significant; see the per-tab pass count and the dashed
significance line on the dotplots).

```{r tables, results='asis'}
# `top_sel[[rk]]` captures the exact rows (full precision, table order) shown per
# database below, so the "Enrichment dotplots" section can plot precisely the same
# selection instead of re-deriving its own (previously it re-sorted by p.adj alone
# and ignored params$top_n, which is why points could appear there with no matching
# table row, or vice versa).
top_sel <- list()
for (rk in ranking_names) {
  cat(sprintf("\n## %s {.tabset}\n", rk))
  sub <- enrich %>% filter(ranking == rk, stat > 0)
  if (nrow(sub) == 0) { cat("\n_No positively-enriched sets._\n"); top_sel[[rk]] <- sub; next }

  # Check if permutations exist for this ranking
  has_perm <- "p.perm" %in% names(sub) && any(!is.na(sub$p.perm))

  # Arrange by p.perm if available, otherwise by p.adj
  if (has_perm) {
    sub <- sub %>% arrange(p.perm, p.adj, desc(stat))
  } else {
    sub <- sub %>% arrange(p.adj, desc(stat))
  }

  db_sel <- sub %>% dplyr::group_by(database) %>% dplyr::slice_head(n = params$top_n) %>% dplyr::ungroup()
  top_sel[[rk]] <- db_sel

  for (db in unique(sub$database)) {
    dbres <- db_sel %>% filter(database == db) %>%
      mutate(AUC = round(stat + 0.5, 3),
             Description = stringr::str_trunc(description, 70),
             pval = signif(pval, 3),
             p.adj = signif(p.adj, 3))
    n_pass <- sum(dbres$sig, na.rm = TRUE)   # dual-threshold significance among shown sets

    if (has_perm) {
      disp <- dbres %>% mutate(p.perm = signif(p.perm, 3)) %>%
        select(pathway, Description, AUC, num.genes, pval, p.adj, p.perm)
    } else {
      disp <- dbres %>% select(pathway, Description, AUC, num.genes, pval, p.adj)
    }

    cat(sprintf("\n### %s\n\n", db))
    if (has_perm) {
      cat(sprintf("_%d shown, %d significant (BH < %.2g ∩ permulation p < %.3g)._\n\n",
                  nrow(disp), n_pass, params$fdr_thr, params$pperm_thr))
    } else {
      cat(sprintf("_%d shown, %d significant (BH < %.2g)._\n\n",
                  nrow(disp), n_pass, params$fdr_thr))
    }
    print(kbl_html(disp, height = if (nrow(disp) > 15) "420px" else NULL))
  }
}
```

# Enrichment dotplots {.tabset .tabset-fade}

Same top `r params$top_n` sets per database as the table above (gene-set
**descriptions** on the y-axis, point colour = AUC, size = set genes, dashed line
= BH threshold). Points left of the dashed line, or otherwise missing a checkmark
in the pass count above, are not significant — the dotplot deliberately still
shows them so near-misses are visible relative to the threshold. Ranking →
database tabs.

```{r dotplots, results='asis', fig.height=5}
for (rk in ranking_names) {
  cat(sprintf("\n## %s {.tabset}\n", rk))
  subrk <- top_sel[[rk]]
  if (is.null(subrk) || nrow(subrk) == 0) { cat("\n_No positively-enriched sets._\n"); next }
  subrk <- subrk %>% filter(!is.na(p.adj))
  if (nrow(subrk) == 0) { cat("\n_No positively-enriched sets._\n"); next }
  for (db in unique(subrk$database)) {
    sub <- subrk %>% filter(database == db) %>%
      mutate(label = stringr::str_trunc(description, 55),
             label = forcats::fct_reorder(label, -log10(p.adj)))
    cat(sprintf("\n### %s\n\n", db))
    p <- ggplot(sub, aes(x = -log10(p.adj), y = label, size = num.genes, colour = stat + 0.5)) +
      geom_point() +
      scale_colour_viridis_c(name = "AUC", option = "C", limits = c(0.5, 1)) +
      scale_size_continuous(name = "set genes") +
      geom_vline(xintercept = -log10(params$fdr_thr), linetype = "dashed", colour = "grey50") +
      labs(title = sprintf("%s — %s", db, rk), x = expression(-log[10]~"BH p"), y = NULL) +
      theme_minimal(base_size = 11)
    print(p); cat("\n")
  }
}
```

# Overall dotplot {.tabset .tabset-fade}

Pooled across all databases, ranked by the **orthogonal composite signal**
(this ranking's own directional percentile concentration + independent
cross-module corroboration — FADE, RER, Accumulation, and PrimateAI-3D/COSMIC
where available) rather than raw BH p. One dotplot per direction, restricted
to the top `r params$top_n` significant terms.

```{r overall_dotplot, results='asis', fig.height=6}
DIR_MAP <- c(top = "top", accelerating = "top", bottom = "bottom", decelerating = "bottom")
n_universe <- length(universe)
bg_all <- orth_background_rates(attr_df, c("pct_top25", "pct_top10", "pct_top5", "pct_top1",
                                            "pct_bottom25", "pct_bottom10", "pct_bottom5", "pct_bottom1",
                                            ORTH_CROSS_COLS))

for (dr in c("global", "top", "bottom")) {
  cat(sprintf("\n## %s\n\n", tools::toTitleCase(dr)))
  # "global" rankings (e.g. CAAS's undirected score) aren't in DIR_MAP -- they're
  # matched by ranking name directly, same convention as cols_by_rk$global in the
  # Leading-edge section below. orth_cols_for_direction() already has a
  # non-top/bottom fallback (cross-source flags only, no percentile trio) built
  # for exactly this case.
  rks <- if (dr == "global") {
    intersect("global", ranking_names)
  } else {
    intersect(names(DIR_MAP)[DIR_MAP == dr], ranking_names)
  }
  if (length(rks) == 0) { cat("_No ranking for this direction in this report._\n\n"); next }

  sig_terms <- enrich %>% filter(ranking %in% rks, sig)
  if (nrow(sig_terms) == 0) { cat(sprintf("_No significant terms (%s)._\n\n", dr)); next }

  le_sub <- leading %>% filter(ranking %in% rks) %>%
    semi_join(sig_terms, by = c("ranking", "database", "pathway"))
  if (nrow(le_sub) == 0) { cat("_No leading-edge data for significant terms._\n\n"); next }

  cols <- orth_cols_for_direction(dr)
  sc <- orthogonal_score(le_sub, bg_all, cols$percentile_cols, cols$cross_cols, n_universe)

  term_meta <- sig_terms %>%
    dplyr::group_by(database, pathway) %>%
    dplyr::summarise(p.adj = min(p.adj, na.rm = TRUE),
                      num.genes = suppressWarnings(max(num.genes, na.rm = TRUE)),
                      description = dplyr::first(description),
                      .groups = "drop")

  dp <- sc %>%
    dplyr::left_join(term_meta, by = c("database", "pathway")) %>%
    arrange(desc(signal)) %>%
    head(params$top_n) %>%
    mutate(label = stringr::str_trunc(dplyr::coalesce(description, pathway), 55),
           neglog10FDR = -log10(pmax(p.adj, 1e-300)))

  if (nrow(dp) == 0) { cat("_No terms to display._\n\n"); next }

  term_levels <- dp %>% arrange(signal) %>% pull(label) %>% unique()
  p <- ggplot(dp, aes(x = signal, y = factor(label, levels = term_levels),
                      size = num.genes, colour = neglog10FDR)) +
    geom_point(alpha = 0.85) +
    scale_colour_viridis_c(name = "-log10 BH", option = "C") +
    scale_size_continuous(name = "set genes") +
    labs(title = sprintf("%s — %s (pooled across databases)", params$project_name, dr),
         subtitle = "Ordered top-to-bottom by orthogonal signal",
         x = "signal (orthogonal composite)", y = NULL) +
    theme_minimal(base_size = 11) +
    theme(axis.text.y = element_text(colour = "black",
                                      size = pmax(8, pmin(11, 220 / length(term_levels)))))
  print(p); cat("\n")
}
```

# Leading-edge composition {.tabset .tabset-fade}

For each significantly-enriched set (BH < `r params$fdr_thr``r if (has_perm_any) sprintf(" ∩ permulation p < %.3g", params$pperm_thr) else ""`), the
**percentage of its leading-edge genes** carrying each attribute — significance gates,
directional score percentiles, and cross-module support — plus `Orthog. signal`,
the same orthogonal-support composite plotted in the Overall dotplot above.
Pooled across databases per direction, **ordered by `Orthog. signal`** (highest
corroboration first — same ordering as the Overall dotplot, so a term's position
there and its row here agree) and capped at `r params$leading_edge_top_n %||% 100`
rows; if a direction has more significant sets than that, the table says so
explicitly rather than silently dropping the low-ranked ones.

`Orthog. signal` is `-log10(p)` of a Lancaster-weighted combination of one-sided
hypergeometric enrichment tests (see `orthogonal_score()` in `orthogonal_score.R`):
for each attribute, does this term's leading edge carry *more* of that flag than
the flag's own background rate predicts, given the leading edge's actual size
(`n_le`). Because it is a hypergeometric test rather than a raw percentage, it is
**already normalised for term size** — the same 50% flag rate is far more surprising,
and yields a higher signal, in a 150-gene leading edge than in a 20-gene one, since
a larger draw gives the test more power to distinguish real enrichment from noise.
Reading the raw `%` columns alongside `n_le` without this in mind can be misleading
for exactly that reason — always weigh a percentage against its `n_le`.

```{r leadingedge, results='asis'}
# Friendly display names per attribute column.
disp_names <- c(
  flag_gate_sig = "gate_sig",
  pct_top25 = "top25%", pct_top10 = "top10%", pct_top5 = "top5%", pct_top1 = "top1%",
  pct_bottom25 = "bot25%", pct_bottom10 = "bot10%", pct_bottom5 = "bot5%", pct_bottom1 = "bot1%",
  flag_fade = "fade", flag_fade_top = "fade_top", flag_fade_bottom = "fade_bottom",
  flag_rer_acc = "rer_acc", flag_rer_decc = "rer_decc",
  flag_accum = "accum", flag_accum_top = "accum_top", flag_accum_bottom = "accum_bottom"
)
# Attribute columns shown per ranking (direction-aware). flag_accum (global) is
# the non-directional accumulation_all_* test; flag_accum_top/flag_accum_bottom
# are the direction-restricted accumulation_top_*/accumulation_bottom_* tests —
# same convention as flag_fade/flag_fade_top/flag_fade_bottom. Percentile
# quartet (25/10/5/1%), not a trio -- 25% is a real, tested tier elsewhere
# (posenrich's own Fisher test, AMI's DOMINO search), so it belongs here too.
cols_by_rk <- list(
  global = c("flag_gate_sig","pct_top25","pct_top10","pct_top5","pct_top1",
             "pct_bottom25","pct_bottom10","pct_bottom5","pct_bottom1",
             "flag_fade","flag_rer_acc","flag_rer_decc","flag_accum"),
  top    = c("flag_gate_sig","pct_top25","pct_top10","pct_top5","pct_top1",
             "flag_fade_top","flag_rer_acc","flag_rer_decc","flag_accum_top"),
  bottom = c("flag_gate_sig","pct_bottom25","pct_bottom10","pct_bottom5","pct_bottom1",
             "flag_fade_bottom","flag_rer_acc","flag_rer_decc","flag_accum_bottom")
)

le_comp_all <- list()  # collect leading-edge composition tables for export

for (rk in ranking_names) {
  cat(sprintf("\n## %s\n\n", rk))
  want <- cols_by_rk[[rk]] %||% attr_cols_all

  sig_terms_rk <- enrich %>% filter(ranking == rk, sig)
  if (nrow(sig_terms_rk) == 0) { cat("_No sets passed the significance thresholds for this direction._\n"); next }

  le_rk_all <- leading %>% filter(ranking == rk) %>%
    semi_join(sig_terms_rk, by = c("ranking", "database", "pathway"))
  if (nrow(le_rk_all) == 0) { cat("_No sets passed the significance thresholds for this direction._\n"); next }

  # Orthogonal-support composite for EVERY significant term first (not just a
  # p.adj/p.perm-truncated subset) -- ordering/truncation below is BY signal,
  # so a term's rank here has to be computed before it can be cut. Same
  # bg_all/n_universe/DIR_MAP as the Overall dotplot above, so a term's
  # dotplot position and its row here carry the same number.
  dr_rk <- unname(DIR_MAP[rk]); if (is.na(dr_rk)) dr_rk <- "global"
  cols_rk <- orth_cols_for_direction(dr_rk)
  sc_rk_all <- orthogonal_score(le_rk_all, bg_all, cols_rk$percentile_cols, cols_rk$cross_cols, n_universe)

  # Order by `Orthog. signal` (highest corroboration first) -- the same
  # ordering as the Overall dotplot, not by this term's own p.adj/p.perm --
  # so if the table has to truncate, it keeps the most cross-corroborated
  # terms, not just the ones most significant on their own axis.
  ltn <- params$leading_edge_top_n %||% 100
  n_sig_rk <- nrow(sc_rk_all)
  term_order <- sc_rk_all %>% dplyr::arrange(dplyr::desc(signal)) %>% head(ltn) %>%
    dplyr::mutate(.rank = dplyr::row_number()) %>%
    dplyr::select(database, pathway, .rank)

  le_rk <- le_rk_all %>% semi_join(term_order, by = c("database", "pathway"))

  if (n_sig_rk > ltn) {
    cat(sprintf("_Showing the top %d of %d significant sets for this direction (ordered by Orthog. signal); the rest are truncated here but remain in `fcs_results/fcs_leading_edge.tsv` (unbounded) and, if within a database's own top %d, in the tables/dotplots above._\n\n",
                ltn, n_sig_rk, params$top_n))
  }

  summ <- fcs_leading_edge_summary(le_rk, want) %>%
    dplyr::left_join(term_order, by = c("database", "pathway")) %>%
    dplyr::arrange(.rank) %>%
    dplyr::select(-.rank) %>%
    dplyr::left_join(sc_rk_all %>% dplyr::select(database, pathway, signal), by = c("database", "pathway")) %>%
    left_join(desc_map, by = c("database", "pathway")) %>%
    mutate(Description = stringr::str_trunc(coalesce(description, pathway), 50),
           p.adj = signif(p.adj, 3), AUC = round(stat + 0.5, 3),
           `Orthog. signal` = round(signal, 2)) %>%
    select(database, pathway, Description, n_le, AUC, p.adj, `Orthog. signal`,
           any_of(intersect(want, names(.))))
  # rename attribute columns to friendly labels
  present_attr <- intersect(want, names(summ))
  names(summ)[match(present_attr, names(summ))] <- disp_names[present_attr]
  print(kbl_html(summ, height = if (nrow(summ) > 15) "480px" else NULL))

  # Collect for export (add ranking column for identification)
  le_comp_all[[rk]] <- summ %>% dplyr::mutate(ranking = rk, .before = 1)
}

# Export the full leading-edge composition table (all rankings) for
# downstream consumption by 15.Comparison_report.Rmd.
le_comp_export <- dplyr::bind_rows(le_comp_all)
if (nrow(le_comp_export) > 0) {
  readr::write_tsv(le_comp_export, "fcs_results/fcs_leading_edge_composition.tsv")
}
```

# Method

```{r method, echo=FALSE}
cat("Engine    : RERconverge::fastwilcoxGMTall (Wilcoxon rank-sum / Mann-Whitney AUC)\n")
cat("Universe  : cleaned_background; no-signal genes floored to 0 (zero-floor)\n")
cat("Direction : membership rankings (top = {top,both}, bottom = {bottom,both})\n")
cat("MTC       : Benjamini-Hochberg, per GMT database\n")
cat("Highlights: significance gates + directional percentile + cross-module flags\n")
```

---

## Summary

This report summarizes the results of the **Enrichment** analysis.

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — Enrichment Module*
*Date of generation: `r Sys.Date()`*
