---
title: "AMI Analysis - DOMINO Active Modules"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
    mathjax: null                       # no math in this report
    md_extensions: "-tex_math_dollars"  # gene-set names contain '$' (e.g. V$HNF4_01) - never parse as TeX math
params:
  background_file: NULL
  background_basename: NULL
  gene_list_files: NULL
  project_name: "AMI_Analysis"
  species: 9606
  gene_scores_file: NULL
  string_db_dir: NULL
  string_cache_dir: NULL
  min_cluster_genes: 3
  domino_network_sif: NULL
  domino_modules_dir: NULL
  domino_network_score_thr: 700
  domino_edge_scores_file: NULL  # gene1/gene2/combined_score TSV alongside domino_network_sif
                                  # (STRING's own combined_score, discarded from network.sif
                                  # itself since DOMINO's tools expect a plain 3-col SIF) --
                                  # optional: NULL/absent degrades to fixed-width edges.
  # FADE and RER each get their own DOMINO network + gene lists, built against
  # their own gene universe (verified to differ meaningfully from CAAS's
  # cleaned_background and from each other on real data, see
  # ENRICHMENT workflow / bugfix_fcs_universe_wrong_background.md). NULL when
  # that tool didn't run this invocation; that tool's section is then skipped.
  fade_gene_lists_dir: NULL          # dir containing fade_{top,bottom,global}_significant.txt
  fade_background_file: NULL        # union of both directions' own background.txt
  fade_domino_network_sif: NULL
  fade_domino_modules_dir: NULL
  fade_domino_edge_scores_file: NULL
  rer_gene_lists_dir: NULL          # dir containing rer_{significant,accelerating,decelerating}.txt
  rer_background_file: NULL        # RER's own background.txt
  rer_domino_network_sif: NULL
  rer_domino_modules_dir: NULL
  rer_domino_edge_scores_file: NULL
  seed: "1998"
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

<!-- 13.AMI_analysis.Rmd — DOMINO active-module report with STRING functional labels, one network per gene list and tool. -->
<!-- PhyloPhere | subworkflows/ENRICHMENT/local/ -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7,
  dev = 'png',
  dpi = 320
)

# Connection timeouts for RCurl and httr, so that a stalled STRING request cannot hang the render.
options(timeout = 30)
options(RCurlOptions = list(connecttimeout = 10, timeout = 30))
if (requireNamespace("httr", quietly = TRUE)) {
  httr::set_config(httr::config(connecttimeout = 10, timeout = 30))
}

# Minimum module size (genes) kept in the module description table and the network
# plot; smaller modules are labeled "Too small" and dropped from both.
.min_cluster_genes <- if (!is.null(params$min_cluster_genes)) as.integer(params$min_cluster_genes) else 3L

suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
  library(kableExtra)
  library(STRINGdb)
  library(paletteer)
})

# Seed of the layout and of any random jitter.
seed_val <- if (exists("params") && !is.null(params$seed) && nzchar(as.character(params$seed))) as.integer(params$seed) else 1998
set.seed(seed_val)

theme_set(
  theme_minimal(base_size = 13) +
  theme(
    axis.title.y  = element_text(margin = ggplot2::margin(l = 20, r = 20)),
    axis.title.x  = element_text(margin = ggplot2::margin(t = 12)),
    axis.text.x   = element_text(margin = ggplot2::margin(t = 5)),
    plot.title    = element_text(face = "bold", hjust = 0, margin = ggplot2::margin(b = 6)),
    plot.subtitle = element_text(colour = "grey40", hjust = 0, margin = ggplot2::margin(b = 10)),
    legend.position = "right",
    legend.box.margin = ggplot2::margin(t = 5)
  )
)

# ── STRING protein id to symbol cache ─────────────────────────────────────────
# Filled by safe_string_map_ids(); used to translate protein ids back to symbols.
global_id2gene <- new.env(parent = emptyenv())

# ── Local functional enrichment ───────────────────────────────────────────────
# Offline STRING term enrichment (hypergeometric test per category). enrich_terms_db is
# loaded below; when it stays NULL, describe_clusters() queries the STRING API.
enrich_terms_db <- NULL

find_enrichment_terms_file <- function(species, cache_dir) {
  fname <- paste0(species, ".protein.enrichment.terms.v12.0.txt.gz")
  
  # Search order: string_db_dir (parameter), then the cache directory.
  paths <- c(
    if (!is.null(params$string_db_dir) && nzchar(params$string_db_dir)) file.path(params$string_db_dir, fname) else NULL,
    file.path(cache_dir, fname)
  )
  
  for (p in paths) {
    if (!is.null(p) && nzchar(p) && file.exists(p)) {
      return(normalizePath(p))
    }
  }
  NULL
}

load_local_enrichment_db <- function(species, cache_dir) {
  terms_file <- find_enrichment_terms_file(species, cache_dir)
  
  if (is.null(terms_file)) {
    # Not found locally: try to download it.
    message(sprintf("[local_enrichment] Local file not found. Attempting to download for species %s...", species))
    dest_file <- file.path(cache_dir, paste0(species, ".protein.enrichment.terms.v12.0.txt.gz"))
    url <- sprintf("https://stringdb-downloads.org/download/protein.enrichment.terms.v12.0/%s.protein.enrichment.terms.v12.0.txt.gz", species)
    
    download_ok <- tryCatch({
      download.file(url, dest_file, quiet = TRUE)
      TRUE
    }, error = function(e) {
      message(sprintf("[local_enrichment] Download failed: %s", e$message))
      FALSE
    })
    
    if (download_ok && file.exists(dest_file)) {
      terms_file <- dest_file
    } else {
      stop("Local enrichment terms file not found and download failed.")
    }
  }
  
  message(sprintf("[local_enrichment] Loading enrichment database from: %s", terms_file))
  db <- readr::read_tsv(
    terms_file,
    col_types = "cccc",
    comment = "#",
    col_names = c("string_protein_id", "category", "term", "description")
  )
  db
}

get_enrichment_local <- function(fg_ids, bg_ids = NULL, fdr_threshold = 0.1) {
  if (is.null(enrich_terms_db)) return(NULL)
  if (length(fg_ids) == 0) return(tibble::tibble())
  
  # With a missing or tiny background, use all proteins of the terms database.
  if (is.null(bg_ids) || length(bg_ids) < 5) {
    bg_ids <- unique(enrich_terms_db$string_protein_id)
  }
  
  # The foreground must lie inside the background.
  fg_ids <- intersect(fg_ids, bg_ids)
  if (length(fg_ids) == 0) return(tibble::tibble())
  
  # Term sizes in the background (the test requires at least 2 background genes per term).
  bg_data <- enrich_terms_db %>% dplyr::filter(string_protein_id %in% bg_ids)
  
  term_bg_counts <- bg_data %>%
    dplyr::group_by(category, term, description) %>%
    dplyr::summarise(M = dplyr::n(), .groups = "drop") %>%
    dplyr::filter(M >= 2)
    
  # Term counts in the foreground.
  fg_data <- enrich_terms_db %>% dplyr::filter(string_protein_id %in% fg_ids)
  fg_counts <- fg_data %>%
    dplyr::group_by(category, term, description) %>%
    dplyr::summarise(k = dplyr::n(), .groups = "drop")
    
  N <- length(bg_ids)
  n <- length(fg_ids)
  
  # Hypergeometric test, P(X >= k) = phyper(k - 1, M, N - M, n, lower.tail = FALSE);
  # BH correction within each category.
  res <- term_bg_counts %>%
    dplyr::inner_join(fg_counts, by = c("category", "term", "description")) %>%
    dplyr::mutate(
      pvalue = phyper(k - 1, M, N - M, n, lower.tail = FALSE)
    ) %>%
    dplyr::group_by(category) %>%
    dplyr::mutate(fdr = p.adjust(pvalue, method = "fdr")) %>%
    dplyr::ungroup() %>%
    dplyr::filter(fdr <= fdr_threshold) %>%
    dplyr::arrange(category, fdr)
    
  if (nrow(res) == 0) return(tibble::tibble())
  
  # Foreground protein ids to symbols through the global cache.
  fg_mapped <- fg_data %>%
    dplyr::mutate(gene = sapply(string_protein_id, function(id) {
      sym <- global_id2gene[[id]]
      if (is.null(sym)) id else sym
    }))
    
  # preferredNames: comma-separated symbols of the foreground genes of each term.
  term_genes <- fg_mapped %>%
    dplyr::group_by(category, term) %>%
    dplyr::summarise(
      preferredNames = paste(sort(unique(gene)), collapse = ","),
      .groups = "drop"
    )
    
  res <- res %>%
    dplyr::left_join(term_genes, by = c("category", "term")) %>%
    dplyr::rename(
      number_of_genes = k,
      number_of_genes_in_background = M,
      p_value = pvalue
    )
    
  res
}

# ── Palette and file helpers ──────────────────────────────────────────────────
# Tableau 10 palette, interpolated when more than 10 colors are needed.
.pal10_base <- as.character(paletteer_d("ggthemes::Tableau_10"))
.safe_pal   <- function(n) if (n <= 10L) .pal10_base[seq_len(n)] else colorRampPalette(.pal10_base)(n)

# Fallback operator: b when a is NULL, empty or all NA.
`%||%` <- function(a, b) {
  if (!is.null(a) && length(a) > 0 && !all(is.na(a))) a else b
}

createDir <- function(directory) {
  if (!dir.exists(directory)) dir.create(directory, recursive = TRUE, showWarnings = FALSE)
}

# Download with retries and a long timeout (the STRING files are large).
download_with_retry <- function(url, destfile, tries = 4L, timeout_sec = 1800L) {
  old_timeout <- getOption("timeout")
  on.exit(options(timeout = old_timeout), add = TRUE)
  options(timeout = max(as.integer(old_timeout), as.integer(timeout_sec), 600L))

  for (i in seq_len(tries)) {
    ok <- tryCatch({
      utils::download.file(url, destfile = destfile, mode = "wb", quiet = FALSE, method = "libcurl")
      TRUE
    }, error = function(e) {
      message(sprintf("Download failed (%d/%d): %s", i, tries, e$message))
      FALSE
    })

    if (isTRUE(ok) && file.exists(destfile) && file.size(destfile) > 0) return(invisible(TRUE))
    if (i < tries) Sys.sleep(min(30, i * 5))
  }

  stop(sprintf("Failed to download after %d attempts: %s", tries, url))
}

# Make the four STRING files of the species (aliases, info, links, physical links)
# available in input_directory: copied from string_db_dir when present, else downloaded.
ensure_string_cache <- function(species, version = "12.0", input_directory = "string_cache") {
  createDir(input_directory)

  species <- as.integer(species)
  base <- "https://stringdb-downloads.org/download"
  files <- c(
    sprintf("protein.aliases.v%s/%d.protein.aliases.v%s.txt.gz", version, species, version),
    sprintf("protein.info.v%s/%d.protein.info.v%s.txt.gz", version, species, version),
    sprintf("protein.links.v%s/%d.protein.links.v%s.txt.gz", version, species, version),
    sprintf("protein.physical.links.v%s/%d.protein.physical.links.v%s.txt.gz", version, species, version)
  )

  for (rel in files) {
    url <- sprintf("%s/%s", base, rel)
    fname <- basename(rel)
    dest <- file.path(input_directory, fname)
    if (!file.exists(dest) || file.size(dest) == 0) {
      # string_db_dir first, the download as the fallback.
      local_src <- if (!is.null(params$string_db_dir) && nzchar(params$string_db_dir)) file.path(params$string_db_dir, fname) else NULL
      if (!is.null(local_src) && file.exists(local_src) && file.size(local_src) > 0) {
        message(sprintf("[ensure_string_cache] Using local pre-cached file: %s", local_src))
        file.copy(local_src, dest, overwrite = TRUE)
      } else {
        message(sprintf("Caching STRING file from online: %s", fname))
        tryCatch({
          download_with_retry(url, destfile = dest)
        }, error = function(e) {
          stop(sprintf("Failed to retrieve required STRING cache file %s: %s", fname, e$message))
        })
      }
    }
  }

  invisible(normalizePath(input_directory, winslash = "/", mustWork = FALSE))
}

# One gene symbol per line; blank lines dropped, duplicates removed.
read_gene_vector <- function(path) {
  x <- readr::read_lines(path)
  x <- trimws(x)
  x <- x[x != ""]
  unique(x)
}

# ── Output directories and STRING ─────────────────────────────────────────────
ami_root <- "."
ami_results_dir <- "ami_results"
ami_summary_dir <- "ami_summary"
ami_plots_dir   <- "ami_plots"
string_cache_dir   <- if (!is.null(params$string_cache_dir) && nzchar(params$string_cache_dir)) params$string_cache_dir else "string_cache"

createDir(ami_results_dir)
createDir(ami_summary_dir)
createDir(ami_plots_dir)
createDir(string_cache_dir)

# Local enrichment database; on failure the online queries take over.
enrich_terms_db <- tryCatch({
  load_local_enrichment_db(as.integer(params$species), string_cache_dir)
}, error = function(e) {
  warning(sprintf("Could not load local enrichment database: %s. Will fallback to online queries.", e$message))
  NULL
})

# Cache the STRING files up front, with retries, so that a transient network failure
# cannot interrupt the mapping or enrichment calls.
ensure_string_cache(
  species = as.integer(params$species),
  version = "12.0",
  input_directory = string_cache_dir
)

# ── STRINGdb instance ─────────────────────────────────────────────────────────
# STRING serves only the id mapping ($map()), the background and the functional term
# enrichment (get_enrichment_local() or get_enrichment()). Module finding and the
# per-module significance come from DOMINO, whose test accepts a custom background;
# STRING's get_ppi_enrichment() does not (foreground plus background are capped at
# 2000 proteins, while the cleaned background has about 10,000 genes).
string_db <- STRINGdb$new(
  version         = "12.0",
  species         = as.integer(params$species),
  input_directory = string_cache_dir
)

ami_networks_dir <- "ami_networks"
createDir(ami_networks_dir)

# ── DOMINO module loading ─────────────────────────────────────────────────────
# DOMINO modules and their Bonferroni-corrected hypergeometric p-values are computed
# upstream by DOMINO_RUN_MODULES (subworkflows/ENRICHMENT/domino.nf), which calls
# DOMINO's Python API through src/run_domino_modules.py so that the p-value is kept
# (the `domino` CLI discards it). This function reads the two TSVs of a gene list:
# <list>_domino_modules.tsv (node, cluster) and <list>_domino_module_stats.tsv.
# It returns NULL elements when either file is missing.
load_domino_modules <- function(list_name, modules_dir) {
  mod_file   <- file.path(modules_dir, sprintf("%s_domino_modules.tsv", list_name))
  stats_file <- file.path(modules_dir, sprintf("%s_domino_module_stats.tsv", list_name))
  if (!file.exists(mod_file) || !file.exists(stats_file)) {
    return(list(clusters_df = NULL, module_stats = NULL))
  }
  list(
    clusters_df  = readr::read_tsv(mod_file, col_types = "ci"),
    module_stats = readr::read_tsv(stats_file, col_types = "iicdd")
  )
}

# ── Module-aware layout ───────────────────────────────────────────────────────
# Fruchterman-Reingold layout seeded with the nodes of each module placed around one
# centroid per module (on a circle), so that members of a module start close together
# and modules appear as visual blobs.
cluster_layout <- function(net, clusters_df, seed = seed_val) {
  set.seed(seed)
  n <- igraph::vcount(net)
  if (is.null(clusters_df) || n < 2) return(igraph::layout_with_fr(net))

  vnames  <- igraph::V(net)$name
  cl_map  <- setNames(clusters_df$cluster, clusters_df$node)
  clusters <- cl_map[vnames]
  clusters[is.na(clusters)] <- max(clusters_df$cluster, na.rm = TRUE) + 1L

  unique_cl <- sort(unique(clusters))
  n_cl <- length(unique_cl)
  if (n_cl <= 1) return(igraph::layout_with_fr(net))

  theta  <- seq(0, 2 * pi * (1 - 1 / n_cl), length.out = n_cl)
  spread <- max(5, sqrt(n) * 0.9)
  cx     <- spread * cos(theta)
  cy     <- spread * sin(theta)

  init <- matrix(0, nrow = n, ncol = 2)
  for (i in seq_along(unique_cl)) {
    mask  <- clusters == unique_cl[i]
    n_sub <- sum(mask)
    sd_jit <- max(0.5, sqrt(n_sub) * 0.35)
    init[mask, 1] <- cx[i] + rnorm(n_sub, 0, sd_jit)
    init[mask, 2] <- cy[i] + rnorm(n_sub, 0, sd_jit)
  }
  igraph::layout_with_fr(net, coords = init, niter = 300)
}

# ── Module functional descriptions ────────────────────────────────────────────
# Runs the STRING term enrichment on the genes of each DOMINO module and labels the
# module with its top terms. The module significance comes from module_stats (DOMINO's
# p_value and p_adj, joined at the end), not from this function. The columns are
# cluster, n_genes, genes (at most 8, for display), all_genes (every member) and
# description ("Too small" and "No enrichment" are placeholders).
describe_clusters <- function(sdb, clusters_df, id2gene, module_stats = NULL, top_terms = 3,
                               min_cluster_genes = 3, bg_string_ids = NULL, active_genes = NULL) {
  # With no modules the result keeps the column schema of the populated path: callers
  # reference n_genes before any nrow() > 0 guard, and arrange() cannot resolve a
  # missing column.
  if (is.null(clusters_df) || nrow(clusters_df) == 0) {
    return(tibble(cluster = integer(), n_genes = integer(), genes = character(),
                  all_genes = character(), description = character()))
  }
  # bg_string_ids defaults to the CAAS background mapped in the parse_background chunk.
  # compute_gene_list_result() passes the mapped background of each tool (CAAS, FADE
  # and RER enrich against their own universe, as their DOMINO networks do).
  if (is.null(bg_string_ids)) bg_string_ids <- get("bg_string_ids", envir = .GlobalEnv)
  desc <- purrr::map_dfr(sort(unique(clusters_df$cluster)), function(cl_id) {
    cl_nodes <- clusters_df$node[clusters_df$cluster == cl_id]
    genes    <- id2gene[cl_nodes]
    genes_sorted <- sort(genes[!is.na(genes)])
    # The listed genes are restricted to the active (scored) input genes. DOMINO's
    # connector genes belong to the module (they count for the min_cluster_genes gate and
    # the STRING enrichment below) but they are not genes that the list hit, so the
    # table does not report them. The network plot still draws them (see .n_connector).
    display_genes <- if (!is.null(active_genes)) intersect(genes_sorted, active_genes) else genes_sorted
    # `genes` is capped at 8 to keep the display table readable; `all_genes` keeps every
    # member, so the written *_domino_module_descriptions.tsv files and the roll-ups keep
    # the full membership. 15.Comparison_report.Rmd reads all_genes: the capped list
    # would undercount the support of a module larger than 8 genes.
    gene_str     <- paste(head(display_genes, 8), collapse = ", ")
    all_gene_str <- paste(display_genes, collapse = ", ")

    if (length(cl_nodes) < min_cluster_genes) {
      return(tibble(cluster = cl_id, n_genes = length(display_genes),
                    genes = gene_str, all_genes = all_gene_str, description = "Too small"))
    }
    message(sprintf("[describe_clusters] Querying STRING enrichment for module %d (%d genes)...", cl_id, length(cl_nodes)))
    # cl_nodes and genes are gene symbols (the DOMINO node ids), while
    # get_enrichment_local() and $get_enrichment() of STRINGdb work with STRING protein
    # ids, as does the background (bg_string_ids). The symbols are therefore mapped
    # first: otherwise the foreground and background never intersect and every module
    # would get "No enrichment".
    cl_string_ids <- safe_string_map_ids(unique(stats::na.omit(genes)))
    enrich <- if (length(cl_string_ids) == 0) {
      NULL
    } else if (!is.null(enrich_terms_db)) {
      get_enrichment_local(cl_string_ids, bg_string_ids)
    } else {
      # No local database: query the STRING API.
      message(sprintf("[describe_clusters] Querying STRING online enrichment for module %d (%d genes)...", cl_id, length(cl_nodes)))
      tryCatch(
        sdb$get_enrichment(cl_string_ids),
        error = function(e) {
          message(sprintf("[describe_clusters] Warning: STRING enrichment failed for module %d: %s", cl_id, e$message))
          NULL
        }
      )
    }
    if (is.null(enrich) || nrow(enrich) == 0 || "error" %in% names(enrich)) {
      return(tibble(cluster = cl_id, n_genes = length(display_genes),
                    genes = gene_str, all_genes = all_gene_str, description = "No enrichment"))
    }
    pvcol   <- intersect(c("fdr", "p_value", "pvalue", "p.adjust"), names(enrich))[1]
    if (!is.na(pvcol)) enrich <- enrich[order(enrich[[pvcol]]), ]
    dcol    <- intersect(c("description", "term", "name"), names(enrich))[1]
    top     <- head(as.character(enrich[[dcol]]), top_terms)
    tibble(cluster = cl_id, n_genes = length(display_genes),
           genes = gene_str, all_genes = all_gene_str, description = paste(top, collapse = " / "))
  })

  if (!is.null(module_stats) && nrow(module_stats) > 0) {
    desc <- desc %>%
      dplyr::left_join(
        module_stats %>% dplyr::select(cluster, domino_p_value = p_value, domino_p_adj = p_adj),
        by = "cluster"
      )
  }
  desc
}

# ── Custom igraph vertex shapes ───────────────────────────────────────────────
# igraph provides circle, square and a few others (igraph::shapes()), which covers the
# FADE overlay; diamond, hexagon and star are registered here. Each is drawn with the
# symbols(stars = ) glyph of base graphics, a polygon with one radius per point: equal
# radii at 4 or 6 points give a diamond or a hexagon, alternating long and short radii
# give a star. The clip function of the circle is reused for the edges, which only need
# an approximate bounding shape.
.make_polygon_shape_fn <- function(radii) {
  force(radii)
  function(coords, v = NULL, params) {
    vertex.color <- params("vertex", "color")
    if (length(vertex.color) != 1 && !is.null(v)) vertex.color <- vertex.color[v]
    vertex.size <- (1 / 150) * params("vertex", "size")
    if (length(vertex.size) != 1 && !is.null(v)) vertex.size <- vertex.size[v]
    graphics::symbols(
      x = coords[, 1], y = coords[, 2], bg = vertex.color,
      stars = outer(vertex.size, radii),
      add = TRUE, inches = FALSE
    )
  }
}
if (!"diamond" %in% names(igraph::shapes())) {
  igraph::add_shape("diamond", clip = igraph::shapes("circle")$clip, plot = .make_polygon_shape_fn(rep(1, 4)))
}
if (!"hexagon" %in% names(igraph::shapes())) {
  igraph::add_shape("hexagon", clip = igraph::shapes("circle")$clip, plot = .make_polygon_shape_fn(rep(1, 6)))
}
if (!"star" %in% names(igraph::shapes())) {
  igraph::add_shape("star", clip = igraph::shapes("circle")$clip, plot = .make_polygon_shape_fn(rep(c(1, 0.4), 5)))
}

# ── Network plot ──────────────────────────────────────────────────────────────
# Cross-tool corroboration shape, used in the CAAS network plots only (the FADE and RER
# sections use circles). FADE significance matches the direction of the CAAS list on
# screen (top: flag_fade_top, bottom: flag_fade_bottom, otherwise flag_fade); RER
# acceleration and deceleration are shown whatever the direction of the list. They are
# mutually exclusive, so the only overlap is FADE with either RER flag, drawn as one
# combined shape (star).
compute_gene_shapes <- function(dir_label, attr_df) {
  if (is.null(attr_df)) return(NULL)
  fade_flag <- if (dir_label == "top") attr_df$flag_fade_top
               else if (dir_label == "bottom") attr_df$flag_fade_bottom
               else attr_df$flag_fade
  rer_flag <- attr_df$flag_rer_acc | attr_df$flag_rer_decc
  shape <- dplyr::case_when(
    fade_flag & rer_flag ~ "star",
    fade_flag            ~ "square",
    attr_df$flag_rer_acc  ~ "diamond",
    attr_df$flag_rer_decc ~ "hexagon",
    TRUE                  ~ "circle"
  )
  setNames(shape, attr_df$gene)
}

plot_cluster_network <- function(net, clusters_df, title, id_to_gene = NULL,
                                  cluster_labels = NULL, node_signal = NULL,
                                  active_genes = NULL, vertex_shape = NULL) {
  n_clust <- if (!is.null(clusters_df)) length(unique(clusters_df$cluster)) else 1L
  pal     <- .safe_pal(max(n_clust, 2L))

  # Edge width and opacity encode the STRING combined_score, an edge attribute attached
  # by build_tool_network() from the edge-scores file (network.sif itself has no score).
  # Without a varying score (a single edge, or no attribute) a fixed width and color
  # are used.
  .escore <- igraph::E(net)$combined_score
  if (!is.null(.escore) && length(unique(stats::na.omit(.escore))) > 1) {
    edge_widths <- scales::rescale(.escore, to = c(0.4, 3))
    edge_alphas <- scales::rescale(.escore, to = c(0.25, 0.95))
    edge_colors <- scales::alpha("#888888", edge_alphas)
  } else {
    edge_widths <- 0.7
    edge_colors <- "#aaaaaa"
  }

  if (!is.null(clusters_df) && nrow(clusters_df) > 0) {
    cl_map  <- setNames(clusters_df$cluster, clusters_df$node)
    v_cl    <- cl_map[igraph::V(net)$name]
    v_cl[is.na(v_cl)] <- n_clust + 1L
    cl_ids  <- sort(unique(clusters_df$cluster))
    vertex_colors <- pal[match(v_cl, c(cl_ids, n_clust + 1L))]
  } else {
    cl_ids        <- integer(0)
    vertex_colors <- rep(pal[1], igraph::vcount(net))
  }

  labels <- if (!is.null(id_to_gene)) id_to_gene[igraph::V(net)$name] else igraph::V(net)$name
  labels[is.na(labels)] <- igraph::V(net)$name[is.na(labels)]

  # Node size encodes the primary statistic of the tool that owns the section (module
  # membership owns the color): the CAAS gene score, the RER significance
  # (-log10 p) or the FADE maximum Bayes factor (log10); see caas_node_score(),
  # rer_node_score() and fade_node_score(). It is not a cross-tool composite: cross-tool
  # corroboration is shown by the node shape (compute_gene_shapes()). Without a
  # signal, or when it does not vary, the size is flat.
  .base_vsize <- pmin(18, pmax(4, 55 / sqrt(igraph::vcount(net))))
  vertex_sizes <- if (!is.null(node_signal)) {
    sig <- node_signal[igraph::V(net)$name]
    sig[is.na(sig)] <- 0
    if (length(unique(sig)) > 1) .base_vsize * scales::rescale(sig, to = c(0.6, 1.7)) else .base_vsize
  } else {
    .base_vsize
  }

  # Border and fill alpha separate the active input genes from the connector genes:
  # module members that the prize-collecting Steiner tree step of DOMINO added only to
  # link active genes, not because they were active themselves. Without the distinction
  # a well-connected module can look like scattered nodes, since the edges that justify
  # it run through the connectors. Connectors are dimmed, not only outlined, so that
  # they read as secondary at a glance. Without active_genes every node counts as active.
  is_active <- if (!is.null(active_genes)) igraph::V(net)$name %in% active_genes else rep(TRUE, igraph::vcount(net))
  vertex_colors        <- ifelse(is_active, vertex_colors, scales::alpha(vertex_colors, 0.35))
  vertex_frame_colors  <- ifelse(is_active, "black", "grey55")

  vertex_frame_widths  <- ifelse(is_active, 1.4, 0.6)

  # Node shape: cross-tool corroboration (CAAS plots only, see compute_gene_shapes());
  # circle otherwise.
  vertex_shapes <- if (!is.null(vertex_shape)) {
    vs <- vertex_shape[igraph::V(net)$name]
    vs[is.na(vs)] <- "circle"
    vs
  } else {
    "circle"
  }

  lo <- cluster_layout(net, clusters_df)

  old_par <- par(mar = c(0, 0, 2, 5))
  on.exit(par(old_par), add = TRUE)
  igraph::plot.igraph(
    net,
    main               = title,
    layout             = lo,
    vertex.color       = vertex_colors,
    vertex.size        = vertex_sizes,
    vertex.shape       = vertex_shapes,
    vertex.frame.color = vertex_frame_colors,
    vertex.frame.width = vertex_frame_widths,
    vertex.label       = labels,
    vertex.label.cex   = max(0.55, min(1.1, 10 / sqrt(igraph::vcount(net)))),
    vertex.label.color = "black",
    vertex.label.font  = 2,
    edge.arrow.size    = 0,
    edge.width         = edge_widths,
    edge.color         = edge_colors
  )
  if (length(cl_ids) > 1) {
    leg_labels <- if (!is.null(cluster_labels)) {
      vapply(as.character(cl_ids), function(k) {
        lbl <- cluster_labels[k]
        if (is.na(lbl)) paste0("Cluster ", k) else paste0("C", k, ": ", stringr::str_trunc(lbl, 35))
      }, character(1))
    } else {
      paste0("Cluster ", cl_ids)
    }
    legend("topright", legend = leg_labels,
           fill = pal[seq_along(cl_ids)], cex = 0.55, bty = "n", title = "Cluster")
  }

  # Text legend for the encodings without a discrete swatch (size, edge width and
  # opacity, border, shape), always in the same place.
  encoding_lines <- c("Node size = this tool's own gene score (CAAS score / RER -log10(p) / FADE log10(max BF))",
                      "Solid + black border = active input gene",
                      "Faded + grey border = DOMINO connector gene (not in input list)")
  if (!is.null(igraph::E(net)$combined_score)) {
    encoding_lines <- c(encoding_lines, "Edge width/opacity = STRING interaction confidence")
  }
  if (!is.null(vertex_shape)) {
    encoding_lines <- c(encoding_lines,
      "Shape: circle=none, square=FADE-sig, diamond=RER-accelerating,",
      "  hexagon=RER-decelerating, star=FADE+RER overlap")
  }
  legend("bottomleft", legend = encoding_lines, bty = "n", cex = 0.5, text.col = "grey30")
}

# Map gene symbols to STRING ids. Returns the unique STRING_id values and fills the
# global protein id to symbol cache.
safe_string_map_ids <- function(genes) {
  if (length(genes) == 0) return(character(0))
  df_in <- data.frame(gene = genes, stringsAsFactors = FALSE)
  mapped <- tryCatch(
    string_db$map(df_in, "gene", removeUnmappedRows = TRUE, quiet = TRUE),
    error = function(e) {
      message(sprintf("STRING ID mapping failed: %s", e$message))
      NULL
    }
  )
  if (is.null(mapped) || nrow(mapped) == 0) return(character(0))
  
  # Fill the global protein id to symbol cache.
  for (i in 1:nrow(mapped)) {
    global_id2gene[[mapped$STRING_id[i]]] <- mapped$gene[i]
  }
  
  unique(as.character(mapped$STRING_id))
}

# ── Inputs ────────────────────────────────────────────────────────────────────
# Background file: the parameter, or its basename in the work directory; absolute path.
bg_file <- params$background_file
if (!file.exists(bg_file) && !is.null(params$background_basename)) {
  if (file.exists(params$background_basename)) bg_file <- params$background_basename
}

if (is.null(bg_file) || !file.exists(bg_file)) {
  stop("Error: background_file not found")
}
bg_file <- normalizePath(bg_file, mustWork = TRUE)
bg_basename <- basename(bg_file)

# Gene list files: the parameter, or every *.txt of the working directory except the
# background file. Paths are made absolute so that later reads do not depend on the
# working directory of the render.
gene_list_files <- params$gene_list_files
if (is.null(gene_list_files) || length(gene_list_files) == 0) {
  gene_list_files <- list.files(pattern = "\\.txt$", full.names = TRUE)
  gene_list_files <- gene_list_files[basename(gene_list_files) != bg_basename]
}
gene_list_files <- gene_list_files[file.exists(gene_list_files)]
gene_list_files <- normalizePath(gene_list_files, mustWork = FALSE)
gene_list_files <- gene_list_files[file.exists(gene_list_files)]
if (length(gene_list_files) == 0) stop("Error: no valid gene list files found")

# Tab order: direction (global, top, bottom), then descending percentile width (25, 10,
# 5, 1: the broadest slice first). The alphabetical order of list.files() would
# interleave directions and put "10" before "5". Names that do not match
# <global|top|bottom><pct> come last, in alphabetical order.
.ami_list_order <- function(paths) {
  base <- tools::file_path_sans_ext(basename(paths))
  dir_rank <- dplyr::case_when(
    grepl("^global", base, ignore.case = TRUE) ~ 0L,
    grepl("^top", base, ignore.case = TRUE)    ~ 1L,
    grepl("^bottom", base, ignore.case = TRUE) ~ 2L,
    TRUE                                       ~ 3L
  )
  pct <- suppressWarnings(as.numeric(sub("^(global|top|bottom)", "", base, ignore.case = TRUE)))
  pct[is.na(pct)] <- -Inf  # unrecognized names sort last within their direction
  paths[order(dir_rank, -pct, base)]
}
gene_list_files <- .ami_list_order(gene_list_files)

```

# AMI analysis configuration

```{r config_overview}
cat("=== AMI (Active Module Identification) configuration ===\n")
cat("Background file (cleaned_background):", bg_file, "\n")
cat("Gene list files:", length(gene_list_files), "\n")
cat("Species (NCBI taxon ID):", params$species, "\n")
cat("DOMINO network score threshold:", params$domino_network_score_thr, "\n")
cat("Module finding + significance: DOMINO (Shamir-Lab)\n")
cat("Module functional labels: STRING term enrichment (untouched functional-enrichment path)\n")
```

# Map background to STRING IDs

> **Note:** A custom background (`cleaned_background`) constrains enrichment to the experimental proteome (all tested genes).
> When background genes are provided, they are mapped to STRING identifiers and set
> on the `STRINGdb` object via `set_background()`, which applies to all subsequent
> `get_enrichment()` calls.

```{r parse_background}
bg_symbols  <- read_gene_vector(bg_file)
bg_string_ids <- safe_string_map_ids(bg_symbols)

cat("Background symbols supplied:", length(bg_symbols), "\n")
cat("Background STRING IDs successfully mapped:", length(bg_string_ids), "\n")

if (length(bg_string_ids) < 5) {
  warning(
    sprintf(
      "Only %d background genes mapped to STRING IDs. ",
      length(bg_string_ids)
    ),
    "Proceeding without a custom background (STRING will use the full proteome)."
  )
  use_custom_background <- FALSE
} else {
  # Set the background on the STRINGdb object so all subsequent enrichment
  # calls are constrained to the experimental proteome.
  use_custom_background <- tryCatch({
    string_db$set_background(bg_string_ids)
    TRUE
  }, error = function(e) {
    warning(
      sprintf("Could not set custom STRING background: %s", e$message),
      " Proceeding with STRING default background."
    )
    FALSE
  })

  if (isTRUE(use_custom_background)) {
    cat("Custom background set on STRINGdb object (", length(bg_string_ids), " proteins).\n")
  }
}
```

# Load gene scores for overlap checking

```{r load_gene_scores}
# Minimum number of STRING-mapped genes for a gene list to get a network.
MIN_MAPPED <- 5L

gene_scores_df <- NULL
if (!is.null(params$gene_scores_file) && file.exists(params$gene_scores_file)) {
  gene_scores_df <- readr::read_tsv(params$gene_scores_file, show_col_types = FALSE)
  cat("Loaded gene_scores_file:", params$gene_scores_file, "with", nrow(gene_scores_df), "genes.\n")
} else {
  cat("No gene_scores_file provided or file not found. Overlap checks will be skipped.\n")
}

prepare_gene_attributes <- function(df) {
  if (is.null(df) || nrow(df) == 0) return(NULL)
  
  # Percentile rank among the genes with a score above 0: about 0 for the best gene and
  # 1 for the worst. Two consequences for anything downstream:
  #   - the value depends on how many genes score above 0 in this run, so percentiles
  #     are not comparable across runs, traits or module subsets;
  #   - zeros are excluded and not ranked last, so a gene without signal gets NA, not
  #     1, and filtering on `pct < x` drops it.
  # Ranks are used because the resampling statistics are floor-limited at 1/N, so raw
  # values pile up at the floor and ranks recover resolution.
  pct_of <- function(v) {
    ok <- !is.na(v) & v > 0
    p <- rep(NA_real_, length(v))
    if (any(ok)) p[ok] <- 1 - (rank(v[ok], ties.method = "max") - 1) / sum(ok)
    p
  }
  
  .col <- function(c, default = NA) if (c %in% names(df)) df[[c]] else rep(default, nrow(df))
  .istrue <- function(x) !is.na(x) & x %in% c(TRUE, "TRUE", "True", "true", 1, "1")
  
  out <- tibble::tibble(gene = df$Gene)
  
  # CAAS directional scores behind the percentiles.
  score_top <- suppressWarnings(as.numeric(.col("gene_caas_score_top")))
  
  score_bot <- suppressWarnings(as.numeric(.col("gene_caas_score_bottom")))
  
  # Cumulative top 10, 5 and 1% flags per direction.
  breaks <- c(0.10, 0.05, 0.01)
  if (!all(is.na(score_top))) {
    pt <- pct_of(score_top)
    out$pct_top10 <- !is.na(pt) & pt <= breaks[1]
    out$pct_top5  <- !is.na(pt) & pt <= breaks[2]
    out$pct_top1  <- !is.na(pt) & pt <= breaks[3]
  } else {
    out$pct_top10 <- FALSE; out$pct_top5 <- FALSE; out$pct_top1 <- FALSE
  }
  
  if (!all(is.na(score_bot))) {
    pb <- pct_of(score_bot)
    out$pct_bottom10 <- !is.na(pb) & pb <= breaks[1]
    out$pct_bottom5  <- !is.na(pb) & pb <= breaks[2]
    out$pct_bottom1  <- !is.na(pb) & pb <= breaks[3]
  } else {
    out$pct_bottom10 <- FALSE; out$pct_bottom5 <- FALSE; out$pct_bottom1 <- FALSE
  }

  # FADE
  out$flag_fade_top    <- .istrue(.col("fade_significant_top"))
  out$flag_fade_bottom <- .istrue(.col("fade_significant_bottom"))
  out$flag_fade        <- out$flag_fade_top | out$flag_fade_bottom
  
  # RER
  .rer_dir <- tolower(as.character(.col("rer_acceleration")))
  out$flag_rer_acc     <- .istrue(.col("rer_significant")) & grepl("acc", .rer_dir)
  out$flag_rer_decc    <- .istrue(.col("rer_significant")) & grepl("dec", .rer_dir)
  
  # Accumulation: flag_accum is the non-directional pooled test; flag_accum_top and
  # flag_accum_bottom are the direction-restricted tests, as for FADE above.
  out$flag_accum        <- .istrue(.col("accum_significant"))
  out$flag_accum_top    <- .istrue(.col("accum_significant_top"))
  out$flag_accum_bottom <- .istrue(.col("accum_significant_bottom"))

  out
}

compute_term_overlap_percentages <- function(preferred_names, attr_df, dir_label) {
  if (is.na(preferred_names) || preferred_names == "" || is.null(attr_df)) {
    return(tibble::tibble(
      top10 = NA_real_, top5 = NA_real_, top1 = NA_real_,
      bot10 = NA_real_, bot5 = NA_real_, bot1 = NA_real_,
      fade = NA_real_, fade_top = NA_real_, fade_bottom = NA_real_,
      rer_acc = NA_real_, rer_decc = NA_real_,
      accum = NA_real_, accum_top = NA_real_, accum_bottom = NA_real_
    ))
  }
  genes <- strsplit(preferred_names, "[,;]")[[1]]
  sub_attr <- attr_df %>% dplyr::filter(gene %in% genes)
  if (nrow(sub_attr) == 0) {
    return(tibble::tibble(
      top10 = 0, top5 = 0, top1 = 0,
      bot10 = 0, bot5 = 0, bot1 = 0,
      fade = 0, fade_top = 0, fade_bottom = 0,
      rer_acc = 0, rer_decc = 0,
      accum = 0, accum_top = 0, accum_bottom = 0
    ))
  }

  # Percentage of the genes carrying each flag.
  round_pct <- function(x) round(100 * mean(x, na.rm = TRUE), 1)

  tibble::tibble(
    top10    = round_pct(sub_attr$pct_top10),
    top5     = round_pct(sub_attr$pct_top5),
    top1     = round_pct(sub_attr$pct_top1),
    bot10    = round_pct(sub_attr$pct_bottom10),
    bot5     = round_pct(sub_attr$pct_bottom5),
    bot1     = round_pct(sub_attr$pct_bottom1),
    fade     = round_pct(sub_attr$flag_fade),
    fade_top = round_pct(sub_attr$flag_fade_top),
    fade_bottom = round_pct(sub_attr$flag_fade_bottom),
    rer_acc  = round_pct(sub_attr$flag_rer_acc),
    rer_decc = round_pct(sub_attr$flag_rer_decc),
    accum_top    = round_pct(sub_attr$flag_accum_top),
    accum_bottom = round_pct(sub_attr$flag_accum_bottom),
    accum    = round_pct(sub_attr$flag_accum)
  )
}
```

# Shared DOMINO helpers (CAAS/FADE/RER) {-}

```{r domino_shared_helpers}
suppressPackageStartupMessages(library(igraph))

# Build the DOMINO network of one tool from its network.sif and an optional edge-scores
# file (gene1, gene2, combined_score). Each tool (CAAS, FADE, RER) has its own network,
# built against its own gene universe (see domino.nf).
build_tool_network <- function(sif_path, edge_scores_path = NULL) {
  .sif <- readr::read_tsv(sif_path, col_names = c("from", "interaction", "to"), col_types = "ccc")
  .edge_sif <- .sif[, c("from", "to")]
  if (!is.null(edge_scores_path) && !is.na(edge_scores_path) && file.exists(edge_scores_path)) {
    .edge_scores <- readr::read_tsv(edge_scores_path, show_col_types = FALSE)
    .scores_fwd <- .edge_scores %>% dplyr::rename(from = gene1, to = gene2)
    .scores_rev <- .edge_scores %>% dplyr::rename(from = gene2, to = gene1) %>%
      dplyr::select(from, to, combined_score)
    .score_lookup <- dplyr::bind_rows(.scores_fwd, .scores_rev) %>%
      dplyr::distinct(from, to, .keep_all = TRUE)
    .edge_sif <- .edge_sif %>% dplyr::left_join(.score_lookup, by = c("from", "to"))
  }
  igraph::graph_from_data_frame(.edge_sif, directed = FALSE)
}

# Per-gene flag table used by the module description tables (top10%, fade_top%, ...)
# and by the shape overlay (compute_gene_shapes()). It holds membership flags only;
# the node size comes from the builders below.
.attr_df_global <- if (!is.null(gene_scores_df)) prepare_gene_attributes(gene_scores_df) else NULL

# Node size builders. Each returns a named numeric vector (gene to size value) with the
# primary statistic of one tool: the CAAS gene score (gene_caas_score_top or
# _bottom, by direction), the RER significance (rer_min_pval, -log10 so that a
# smaller p gives a bigger node) or the FADE maximum Bayes factor (fade_max_bf_top or
# _bottom, log10 because Bayes factors span orders of magnitude). The columns come from
# gene_scores.tsv (gene_scores_df). NULL or NA where a column is missing:
# plot_cluster_network() then uses a flat size.
.col_num <- function(df, col) if (col %in% names(df)) suppressWarnings(as.numeric(df[[col]])) else rep(NA_real_, nrow(df))

caas_node_score <- function(dir_label) {
  if (is.null(gene_scores_df)) return(NULL)
  col <- if (dir_label == "top") "gene_caas_score_top" else if (dir_label == "bottom") "gene_caas_score_bottom" else NA
  if (is.na(col)) return(NULL)
  setNames(.col_num(gene_scores_df, col), gene_scores_df$Gene)
}

rer_node_score <- function(dir_label) {
  if (is.null(gene_scores_df) || !"rer_min_pval" %in% names(gene_scores_df)) return(NULL)
  p <- .col_num(gene_scores_df, "rer_min_pval")
  setNames(-log10(pmax(p, 1e-300)), gene_scores_df$Gene)
}

fade_node_score <- function(dir_label) {
  if (is.null(gene_scores_df)) return(NULL)
  top <- .col_num(gene_scores_df, "fade_max_bf_top")
  bot <- .col_num(gene_scores_df, "fade_max_bf_bottom")
  v <- if (dir_label == "top") top
       else if (dir_label == "bottom") bot
       else pmax(top, bot, na.rm = TRUE)  # Global: whichever direction has a BF for this gene
  setNames(log10(pmax(v, 1)), gene_scores_df$Gene)
}

# Direction classifiers. A gene list name decides (a) the direction bucket under which
# its modules are rolled up in the tabs, and (b) which CAAS percentile and flag columns
# annotate its modules and which direction the node score builders read. CAAS and FADE
# share the top and bottom directions. RER has accelerating and decelerating, which do
# not map onto the CAAS top and bottom percentiles, so RER always uses the "global"
# annotation (both columns) and its own Acc, Decc and Global roll-up. The match is a
# substring, so "fade_top_significant" matches "top".
top_bottom_pct_classifier <- function(nm) {
  if (grepl("top", nm, ignore.case = TRUE)) "top"
  else if (grepl("bottom", nm, ignore.case = TRUE)) "bottom"
  else "global"
}
top_bottom_roll_classifier <- function(other_label = "Other") {
  force(other_label)
  function(nm) {
    if (grepl("^global", nm, ignore.case = TRUE)) "Global"
    else if (grepl("top", nm, ignore.case = TRUE)) "Top"
    else if (grepl("bottom", nm, ignore.case = TRUE)) "Bottom"
    else other_label
  }
}
always_global_pct_classifier <- function(nm) "global"
rer_roll_classifier <- function(nm) {
  if (grepl("accelerat", nm, ignore.case = TRUE)) "Acc"
  else if (grepl("decelerat", nm, ignore.case = TRUE)) "Decc"
  else "Global"
}

# ── One gene list ─────────────────────────────────────────────────────────────
# Computes everything for one gene list (modules, STRING enrichment, network PNG) and
# returns print-ready pieces (strings, data frames, file paths). Printing is separate
# (print_gene_list_result()) so that the expensive work runs once per list although
# the nested tabs show each list under its own direction and under "Global".
# use_shape_overlay is TRUE only for CAAS (cross-tool FADE and RER shapes, see
# compute_gene_shapes()); the FADE and RER sections use circles.
compute_gene_list_result <- function(.gf, net_full, modules_dir, prefix,
                                      roll_dir_classifier, pct_dir_classifier,
                                      use_shape_overlay = FALSE, bg_string_ids = NULL,
                                      node_score_fn = NULL) {
  .list_name <- tools::file_path_sans_ext(basename(.gf))
  .genes_sym <- read_gene_vector(.gf)
  .mapped    <- intersect(.genes_sym, igraph::V(net_full)$name)
  .direction <- roll_dir_classifier(.list_name)

  res <- list(list_name = .list_name, direction = .direction, skip_message = NULL,
              connectivity_line = NULL, modules_line = NULL, display_df = NULL,
              png_path = NULL, clust_desc = NULL, network_summary_row = NULL)

  if (length(.mapped) < MIN_MAPPED) {
    res$skip_message <- sprintf("> Too few mapped genes for %s (need >= %d).", .list_name, MIN_MAPPED)
    return(res)
  }

  # The modules are loaded before the plotted network is built and are not restricted to
  # the active input genes: DOMINO adds connector genes (not active themselves, but
  # between active genes in the network), which is what separates network-based module
  # discovery from plain set overlap.
  .domino <- load_domino_modules(.list_name, modules_dir)
  .clusters_df <- if (!is.null(.domino$clusters_df)) {
    .domino$clusters_df %>% dplyr::filter(node %in% igraph::V(net_full)$name)
  } else NULL
  .module_stats <- .domino$module_stats

  # Modules are renumbered by size (genes, descending) and capped to the 20 largest, which
  # keeps the legend and description table readable. module_stats is renumbered the same
  # way so that the p_value and p_adj join stays aligned.
  .n_mod_total <- if (!is.null(.clusters_df)) length(unique(.clusters_df$cluster)) else NA_integer_
  if (!is.null(.clusters_df) && nrow(.clusters_df) > 0) {
    .cl_counts <- .clusters_df %>%
      dplyr::group_by(cluster) %>%
      dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
      dplyr::arrange(desc(n), cluster)
    .cl_map <- setNames(seq_len(nrow(.cl_counts)), .cl_counts$cluster)
    .clusters_df$cluster <- .cl_map[as.character(.clusters_df$cluster)]
    if (!is.null(.module_stats)) {
      .module_stats <- .module_stats %>%
        dplyr::filter(cluster %in% as.integer(names(.cl_map))) %>%
        dplyr::mutate(cluster = .cl_map[as.character(cluster)])
    }
    .clusters_df <- .clusters_df %>% dplyr::filter(cluster <= 20)
  }
  .n_mod <- if (!is.null(.clusters_df)) length(unique(.clusters_df$cluster)) else NA_integer_

  # Plotted network: the active input genes plus the member genes of every retained
  # module, so that the connector genes are visible.
  .plot_gene_pool <- union(.genes_sym, if (!is.null(.clusters_df)) .clusters_df$node else character(0))
  .net <- igraph::induced_subgraph(net_full, igraph::V(net_full)[igraph::V(net_full)$name %in% .plot_gene_pool])
  if (igraph::vcount(.net) == 0) {
    res$skip_message <- sprintf("> No network retrieved for %s.", .list_name)
    return(res)
  }

  # Keep the connected nodes only (at least one edge at the score threshold).
  .connected <- igraph::V(.net)[igraph::degree(.net) > 0]
  .net_conn  <- igraph::induced_subgraph(.net, .connected)
  .n_nodes   <- igraph::vcount(.net_conn)
  .n_edges   <- igraph::ecount(.net_conn)
  .n_connector <- sum(!igraph::V(.net_conn)$name %in% .genes_sym)

  res$connectivity_line <- sprintf(
    "Connected nodes: **%d** (%d input genes + %d DOMINO connector genes) | Edges: **%d** | Input genes in network: **%d**",
    .n_nodes, .n_nodes - .n_connector, .n_connector, .n_edges, length(.mapped))

  if (.n_nodes < 3) {
    res$skip_message <- "> Too few connected nodes - network plot skipped."
    res$network_summary_row <- tibble(
      gene_list = .list_name, n_input = length(.mapped), n_connected = .n_nodes,
      n_edges = .n_edges, n_modules = .n_mod_total)
    return(res)
  }

  .pct_dir_label <- pct_dir_classifier(.list_name)

  if (!is.null(.clusters_df)) {
    # The DOMINO nodes are gene symbols already, so node and gene are equal.
    .clusters_df$gene <- .clusters_df$node

    .clusters_out <- .clusters_df
    attr_df <- .attr_df_global  # one table for every list and tool
    if (!is.null(attr_df)) {
      .clusters_out <- .clusters_out %>%
        dplyr::left_join(attr_df, by = c("gene" = "gene"))
    }

    .clust_file <- file.path(ami_networks_dir, sprintf("%s_%s_domino_clusters.tsv", prefix, .list_name))
    readr::write_tsv(.clusters_out %>% dplyr::arrange(cluster, gene), .clust_file)
    res$modules_line <- sprintf("Modules found: **%d** of **%d** total (showing the 20 largest; written to `%s`)",
                                 .n_mod, .n_mod_total, basename(.clust_file))

    .clust_desc <- describe_clusters(
      string_db, .clusters_df, setNames(.clusters_df$node, .clusters_df$node),
      module_stats = .module_stats, min_cluster_genes = .min_cluster_genes,
      bg_string_ids = bg_string_ids, active_genes = .genes_sym
    )

    if (!is.null(gene_scores_df) && !is.null(attr_df) && nrow(.clust_desc) > 0) {
      .cl_pcts <- purrr::map_dfr(sort(unique(.clusters_df$cluster)), function(cl_id) {
        cl_genes <- .clusters_df$gene[.clusters_df$cluster == cl_id]
        compute_term_overlap_percentages(paste(cl_genes, collapse = ","), attr_df, .pct_dir_label) %>%
          mutate(cluster = cl_id)
      })
      .clust_desc <- .clust_desc %>% dplyr::left_join(.cl_pcts, by = "cluster")
    }

    # Module descriptions, with the DOMINO p-values, written for downstream readers.
    .clust_desc_file <- file.path(ami_networks_dir, sprintf("%s_%s_domino_module_descriptions.tsv", prefix, .list_name))
    readr::write_tsv(.clust_desc %>% dplyr::arrange(desc(n_genes)), .clust_desc_file)

    if (nrow(.clust_desc) > 0) {
      res$clust_desc <- .clust_desc %>%
        dplyr::mutate(gene_list = .list_name, direction = .direction, .before = 1)
    }

    .clust_desc_display <- .clust_desc %>%
      dplyr::filter(!description %in% c("Too small", "No enrichment")) %>%
      dplyr::arrange(desc(n_genes))

    if (nrow(.clust_desc_display) > 0) {
      if (!is.null(gene_scores_df)) {
        if (.pct_dir_label == "top") {
          display_df <- .clust_desc_display %>%
            dplyr::select(
              Module = cluster, Genes = n_genes,
              `Representative genes` = genes, `Top enriched terms` = description,
              `top10% (%)` = top10, `top5% (%)` = top5, `top1% (%)` = top1,
              `fade_top (%)` = fade_top, `rer_acc (%)` = rer_acc, `rer_decc (%)` = rer_decc,
              `accum_top (%)` = accum_top
            )
        } else if (.pct_dir_label == "bottom") {
          display_df <- .clust_desc_display %>%
            dplyr::select(
              Module = cluster, Genes = n_genes,
              `Representative genes` = genes, `Top enriched terms` = description,
              `bot10% (%)` = bot10, `bot5% (%)` = bot5, `bot1% (%)` = bot1,
              `fade_bottom (%)` = fade_bottom, `rer_acc (%)` = rer_acc, `rer_decc (%)` = rer_decc,
              `accum_bottom (%)` = accum_bottom
            )
        } else {
          display_df <- .clust_desc_display %>%
            dplyr::select(
              Module = cluster, Genes = n_genes,
              `Representative genes` = genes, `Top enriched terms` = description,
              `top10% (%)` = top10, `top5% (%)` = top5, `top1% (%)` = top1,
              `bot10% (%)` = bot10, `bot5% (%)` = bot5, `bot1% (%)` = bot1,
              `fade (%)` = fade, `rer_acc (%)` = rer_acc, `rer_decc (%)` = rer_decc,
              `accum (%)` = accum
            )
        }
      } else {
        display_df <- .clust_desc_display %>%
          dplyr::select(Module = cluster, Genes = n_genes,
                         `Representative genes` = genes, `Top enriched terms` = description)
      }

      # DOMINO per-module significance, appended by position (display_df keeps the row
      # order of .clust_desc_display).
      if (all(c("domino_p_value", "domino_p_adj") %in% names(.clust_desc_display))) {
        display_df <- dplyr::bind_cols(
          display_df,
          .clust_desc_display %>% dplyr::transmute(
            `DOMINO p`     = signif(domino_p_value, 3),
            `DOMINO p.adj` = signif(domino_p_adj, 3)
          )
        )
      }

      res$display_df <- collapsible_table(display_df)
    }

    .cl_labels <- setNames(.clust_desc$description, as.character(.clust_desc$cluster))
    .cl_labels <- .cl_labels[!.cl_labels %in% c("Too small", "No enrichment")]

    .clusters_df_viz <- .clusters_df %>%
      dplyr::filter(cluster %in% .clust_desc$cluster[.clust_desc$n_genes >= .min_cluster_genes])
  } else {
    .cl_labels <- NULL
    .clusters_df_viz <- .clusters_df
  }

  # The plotted graph keeps only the nodes of the shown modules (the 20 largest that
  # pass min_cluster_genes).
  .net_plot <- if (!is.null(.clusters_df_viz) && nrow(.clusters_df_viz) > 0) {
    igraph::induced_subgraph(
      .net_conn,
      igraph::V(.net_conn)[igraph::V(.net_conn)$name %in% .clusters_df_viz$node]
    )
  } else {
    .net_conn
  }

  # Node signal: node_score_fn (caas_node_score, rer_node_score or fade_node_score)
  # evaluated for the direction of this gene list.
  .node_signal <- if (!is.null(node_score_fn)) node_score_fn(.pct_dir_label) else NULL

  .node_shape <- if (use_shape_overlay) compute_gene_shapes(.pct_dir_label, .attr_df_global) else NULL

  .png_path <- file.path(ami_networks_dir, sprintf("%s_%s_domino_network.png", prefix, .list_name))
  png(.png_path, width = 1800, height = 1500, res = 160)
  plot_cluster_network(
    .net_plot, .clusters_df_viz,
    title          = sprintf("%s - DOMINO modules (score >= %s, top 20 modules)", .list_name, params$domino_network_score_thr),
    cluster_labels = .cl_labels,
    node_signal    = .node_signal,
    active_genes   = .genes_sym,
    vertex_shape   = .node_shape
  )
  dev.off()
  res$png_path <- .png_path

  res$network_summary_row <- tibble(
    gene_list = .list_name, n_input = length(.mapped), n_connected = .n_nodes,
    n_edges = .n_edges, n_modules = .n_mod
  )
  res
}

# Print a compute_gene_list_result() without recomputing it: cat() and kable() on the
# cached strings, data frames and file paths. Called once per direction tab that shows
# the list (its own direction and "Global").
print_gene_list_result <- function(res) {
  cat(sprintf("\n\n##### %s\n\n", res$list_name))
  if (!is.null(res$skip_message)) { cat(res$skip_message, "\n\n"); return(invisible(NULL)) }
  if (!is.null(res$connectivity_line)) cat(res$connectivity_line, "\n\n")
  if (!is.null(res$modules_line)) cat(res$modules_line, "\n\n")
  if (!is.null(res$display_df)) {
    print(knitr::kable(
      res$display_df,
      caption = sprintf("Module functional descriptions - %s", res$list_name),
      escape = FALSE
    ) %>% kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"), full_width = TRUE))
  }
  if (!is.null(res$png_path)) cat(sprintf("![](%s)\n\n", res$png_path))
  invisible(NULL)
}

# ── Roll-up across gene lists ─────────────────────────────────────────────────
# Term x gene-list heatmap and tables; generic over the module descriptions of any tool.
.normalize_term <- function(x) {
  x <- tolower(x)
  x <- gsub("[^a-z0-9 ]", " ", x)
  gsub("\\s+", " ", trimws(x))
}
# The description of a module recurs, worded slightly differently, across nested gene
# lists that share most of their genes. Two kinds of duplicate are merged into one row:
# near-duplicate strings (case, punctuation, minor rewording), found with a normalized
# edit distance (utils::adist()), and nested terms, whose full word set is contained in
# that of another term ("NADPH oxidase complex" inside "phagocyte NADPH oxidase
# complex"). The edit distance alone rejects the nested pair because the extra words
# dominate the ratio, so containment is tested explicitly and forces the merge.
# Different terms (no containment, low similarity) stay separate.
.term_contains <- function(a, b) {
  wa <- strsplit(a, " ")[[1]]; wb <- strsplit(b, " ")[[1]]
  if (length(wa) == 0 || length(wb) == 0) return(FALSE)
  short <- if (length(wa) <= length(wb)) wa else wb
  long  <- if (length(wa) <= length(wb)) wb else wa
  all(short %in% long)
}
.cluster_terms <- function(terms, sim_thr = 0.85) {
  norm <- vapply(terms, .normalize_term, character(1))
  reps_norm <- character(0); reps_orig <- character(0)
  assign <- integer(length(terms))
  for (i in seq_along(terms)) {
    if (length(reps_norm) == 0) { reps_norm <- norm[i]; reps_orig <- terms[i]; assign[i] <- 1L; next }
    d <- utils::adist(norm[i], reps_norm)[1, ]
    sim <- 1 - d / pmax(nchar(norm[i]), nchar(reps_norm), 1)
    contained <- vapply(reps_norm, function(r) .term_contains(norm[i], r), logical(1))
    sim[contained] <- 1
    best <- which.max(sim)
    if (sim[best] >= sim_thr) {
      assign[i] <- best
      # When a compound description folds into a shorter one (containment), the shorter
      # text becomes the representative, being the more canonical wording. Plain
      # near-duplicate merges keep the term seen first.
      if (contained[best] && nchar(terms[i]) < nchar(reps_orig[best])) {
        reps_orig[best] <- terms[i]
      }
    } else {
      reps_norm <- c(reps_norm, norm[i]); reps_orig <- c(reps_orig, terms[i])
      assign[i] <- length(reps_norm)
    }
  }
  reps_orig[assign]
}

# ── Collapsible long cells ────────────────────────────────────────────────────
# Character values longer than char_thr are wrapped in a native <details>/<summary>
# element (no JavaScript), collapsed to a preview, so that long gene or term lists do not
# stretch a table row. The input is always HTML-escaped first (short values come back
# escaped and unwrapped), so a table can be passed to kable(escape = FALSE) without HTML
# injection from gene symbols or term text.
.html_escape <- function(x) {
  x <- gsub("&", "&amp;", x, fixed = TRUE)
  x <- gsub("<", "&lt;",  x, fixed = TRUE)
  x <- gsub(">", "&gt;",  x, fixed = TRUE)
  x <- gsub('"', "&quot;", x, fixed = TRUE)
  x
}
collapsible_cell <- function(x, char_thr = 180, preview_chars = 100) {
  x <- as.character(x)
  out <- rep(NA_character_, length(x))
  ok <- !is.na(x)
  long <- ok & nchar(x) > char_thr
  if (any(long)) {
    out[long] <- sprintf(
      '<details><summary>%s&hellip; <span style="color:#888;font-size:0.85em;">(click to expand)</span></summary>%s</details>',
      .html_escape(substr(x[long], 1, preview_chars)), .html_escape(x[long])
    )
  }
  short <- ok & !long
  out[short] <- .html_escape(x[short])
  out
}
# Apply collapsible_cell() to every character column of a data frame (other columns
# cannot carry HTML), so that a whole table can use escape = FALSE safely.
collapsible_table <- function(df) {
  df %>% dplyr::mutate(dplyr::across(dplyr::where(is.character), collapsible_cell))
}

.render_rollup <- function(df, caption) {
  if (is.null(df) || nrow(df) == 0) { cat(sprintf("_No modules to show (%s)._\n\n", caption)); return(invisible(NULL)) }
  df <- df %>% dplyr::filter(!description %in% c("Too small", "No enrichment"))
  if (nrow(df) == 0) { cat(sprintf("_No described modules to show (%s)._\n\n", caption)); return(invisible(NULL)) }

  if ("domino_p_adj" %in% names(df)) {
    df <- df %>% dplyr::arrange(domino_p_adj, dplyr::desc(n_genes))
  } else {
    df <- df %>% dplyr::arrange(dplyr::desc(n_genes))
  }

  disp <- df %>% dplyr::transmute(
    `Gene list` = gene_list, Module = cluster, Genes = n_genes,
    `Representative genes` = genes,
    `Top enriched terms` = description,
    `DOMINO p` = if ("domino_p_value" %in% names(df)) signif(domino_p_value, 3) else NA,
    `DOMINO p.adj` = if ("domino_p_adj" %in% names(df)) signif(domino_p_adj, 3) else NA
  ) %>% dplyr::select(dplyr::where(~ !all(is.na(.))))

  print(knitr::kable(collapsible_table(disp), caption = caption, escape = FALSE) %>%
          kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE))
}

.render_term_heatmap <- function(term_rows, gene_list_order, title) {
  if (is.null(term_rows) || nrow(term_rows) == 0) { cat(sprintf("_No described terms to show (%s)._\n\n", title)); return(invisible(NULL)) }
  term_rows <- term_rows %>% dplyr::filter(gene_list %in% gene_list_order)
  if (nrow(term_rows) == 0) { cat(sprintf("_No described terms for these gene lists (%s)._\n\n", title)); return(invisible(NULL)) }

  pres <- term_rows %>% dplyr::distinct(term, gene_list)
  # Terms by the number of lists in which they appear (most frequent first), then
  # alphabetically.
  term_counts <- pres %>% dplyr::count(term, name = "n_thresholds") %>%
    dplyr::arrange(dplyr::desc(n_thresholds), term)
  pres <- pres %>%
    dplyr::mutate(gene_list = factor(gene_list, levels = gene_list_order),
                  term = factor(term, levels = rev(term_counts$term)))

  n_terms <- nrow(term_counts)
  p <- ggplot(pres, aes(x = gene_list, y = term)) +
    geom_tile(fill = "#2c7fb8", colour = "white", linewidth = 0.6) +
    scale_x_discrete(drop = FALSE) +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          axis.text.y = element_text(size = pmax(6, pmin(10, 260 / max(n_terms, 1))))) +
    labs(title = sprintf("Described terms by gene list - %s", title),
         x = "gene list", y = NULL)
  print(p)
  cat("\n\n")
}

.render_term_recurrence_table <- function(term_rows, gene_list_order, title) {
  if (is.null(term_rows) || nrow(term_rows) == 0) { cat(sprintf("_No described terms to show (%s)._\n\n", title)); return(invisible(NULL)) }
  term_rows <- term_rows %>% dplyr::filter(gene_list %in% gene_list_order)
  if (nrow(term_rows) == 0) { cat(sprintf("_No described terms for these gene lists (%s)._\n\n", title)); return(invisible(NULL)) }

  pres <- term_rows %>% dplyr::distinct(term, gene_list)
  term_counts <- pres %>% dplyr::count(term, name = "n_thresholds") %>%
    dplyr::arrange(dplyr::desc(n_thresholds), term)
  pres <- pres %>%
    dplyr::mutate(gene_list = factor(gene_list, levels = gene_list_order))

  disp <- pres %>%
    dplyr::mutate(gene_list = as.character(gene_list)) %>%
    dplyr::arrange(term, gene_list) %>%
    dplyr::group_by(term) %>%
    dplyr::summarise(`N lists` = dplyr::n(),
                      `Gene lists` = paste(gene_list, collapse = ", "),
                      .groups = "drop") %>%
    dplyr::arrange(dplyr::desc(`N lists`), term) %>%
    dplyr::rename(Term = term)
  print(knitr::kable(collapsible_table(disp), caption = sprintf("Term -> gene-list membership - %s", title),
                      escape = FALSE) %>%
          kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE))
  cat("\n\n")
}

# ── Nested tabs of one tool ───────────────────────────────────────────────────
# The outer tab is the direction (Global, Top, Bottom, or Global, Acc, Decc for RER),
# with four sub-tabs: Modules (the networks of the gene lists of that direction),
# Recurrence Heatmap (term x gene list), Recurrence Table (term to gene-list membership)
# and Gene lists (the module description roll-up).
#
# direction_tabs: named list, tab label to the `direction` value that filters the lists
# (NULL = no filter, as in the Global tab); the direction of a list comes from
# roll_dir_classifier(). gene_list_orders: named list with the same names, the ordered
# gene-list names of the heatmap of each tab. write_all_lists_file = TRUE (CAAS) writes
# module_descriptions_all_lists.tsv and term_threshold_membership.tsv without a tool
# prefix; the other tools write prefixed names.
#
# compute_gene_list_result() runs once per list; print_gene_list_result() re-emits the
# cached result under each tab.
render_tool_section_nested <- function(gene_list_files, net_full, modules_dir, prefix,
                                        roll_dir_classifier, pct_dir_classifier,
                                        direction_tabs, gene_list_orders,
                                        use_shape_overlay = FALSE, bg_string_ids = NULL,
                                        node_score_fn = NULL, write_all_lists_file = FALSE) {
  results <- lapply(gene_list_files, compute_gene_list_result,
                     net_full = net_full, modules_dir = modules_dir, prefix = prefix,
                     roll_dir_classifier = roll_dir_classifier, pct_dir_classifier = pct_dir_classifier,
                     use_shape_overlay = use_shape_overlay, bg_string_ids = bg_string_ids,
                     node_score_fn = node_score_fn)

  clust_desc_all <- dplyr::bind_rows(lapply(results, `[[`, "clust_desc"))
  if (nrow(clust_desc_all) == 0) clust_desc_all <- NULL
  network_summary_all <- dplyr::bind_rows(lapply(results, `[[`, "network_summary_row"))

  term_rows_all <- if (!is.null(clust_desc_all)) {
    clust_desc_all %>%
      dplyr::filter(!description %in% c("Too small", "No enrichment")) %>%
      dplyr::mutate(term_raw = strsplit(description, "\\s*/\\s*")) %>%
      tidyr::unnest(term_raw) %>%
      dplyr::mutate(term_raw = trimws(term_raw)) %>%
      dplyr::filter(nzchar(term_raw))
  } else NULL
  if (!is.null(term_rows_all) && nrow(term_rows_all) > 0) {
    term_rows_all$term <- .cluster_terms(term_rows_all$term_raw)
  }
  assign(sprintf(".ami_term_rows_%s", prefix), term_rows_all, envir = .GlobalEnv)

  for (dir_label in names(direction_tabs)) {
    filter_val <- direction_tabs[[dir_label]]
    cat(sprintf("\n\n### %s {.tabset}\n\n", dir_label))

    sel <- if (is.null(filter_val)) results else Filter(function(r) r$direction == filter_val, results)

    cat("\n\n#### Modules {.tabset}\n\n")
    if (length(sel) == 0) {
      cat("_No gene lists for this direction._\n\n")
    } else {
      for (r in sel) print_gene_list_result(r)
    }

    cat("\n\n#### Recurrence Heatmap\n\n")
    .render_term_heatmap(term_rows_all, gene_list_orders[[dir_label]], dir_label)

    cat("\n\n#### Recurrence Table\n\n")
    .render_term_recurrence_table(term_rows_all, gene_list_orders[[dir_label]], dir_label)

    cat("\n\n#### Gene lists\n\n")
    df <- if (is.null(filter_val)) clust_desc_all
          else if (!is.null(clust_desc_all)) dplyr::filter(clust_desc_all, direction == filter_val)
          else NULL
    .render_rollup(df, sprintf("%s gene lists", dir_label))
  }

  if (!is.null(network_summary_all) && nrow(network_summary_all) > 0) {
    readr::write_tsv(network_summary_all, file.path(ami_networks_dir, sprintf("%s_network_summary.tsv", prefix)))
    cat("\n\n### Network summary\n\n")
    print(knitr::kable(network_summary_all, caption = "DOMINO network summary") %>%
            kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")))
  }

  desc_file <- if (write_all_lists_file) "module_descriptions_all_lists.tsv" else sprintf("%s_module_descriptions_all_lists.tsv", prefix)
  term_file <- if (write_all_lists_file) "term_threshold_membership.tsv" else sprintf("%s_term_threshold_membership.tsv", prefix)
  if (!is.null(clust_desc_all)) {
    readr::write_tsv(clust_desc_all, file.path(ami_networks_dir, desc_file))
  }
  if (!is.null(term_rows_all) && nrow(term_rows_all) > 0) {
    readr::write_tsv(term_rows_all, file.path(ami_networks_dir, term_file))
  }
  invisible(clust_desc_all)
}

.GLOBAL_THRESHOLDS <- c("global25", "global10", "global5", "global1")
.TOP_THRESHOLDS    <- c("top25", "top10", "top5", "top1")
.BOTTOM_THRESHOLDS <- c("bottom25", "bottom10", "bottom5", "bottom1")
.ALL_THRESHOLDS    <- c(.GLOBAL_THRESHOLDS, .TOP_THRESHOLDS, .BOTTOM_THRESHOLDS)
```

# Active Network Modules (AMI) {.tabset .tabset-fade}

## CAAS {.tabset}

Network score threshold: **`r params$domino_network_score_thr`** (= `r params$domino_network_score_thr / 1000` on the 0–1 STRING scale).
Only nodes with at least one edge at this confidence level (both endpoints in `cleaned_background`) are in DOMINO's input network.
Module finding + significance: **DOMINO** (Shamir-Lab) - Bonferroni-corrected hypergeometric p-value per module, computed by DOMINO itself.

Network plots additionally mark cross-tool corroboration by node **shape**: circle = none, square = FADE-significant, diamond = RER-accelerating, hexagon = RER-decelerating, star = FADE-significant overlapping either RER direction.

Each direction below (**Global**/**Top**/**Bottom**) has four sub-tabs: **Modules**, **Recurrence Heatmap**, **Recurrence Table**, and **Gene lists**.

```{r caas_domino_networks, fig.width=10, fig.height=8, results='asis'}
.caas_net_full <- build_tool_network(params$domino_network_sif, params$domino_edge_scores_file)
CAAS_clust_desc_all <- render_tool_section_nested(
  gene_list_files, .caas_net_full, params$domino_modules_dir, "caas",
  roll_dir_classifier = top_bottom_roll_classifier("Other"),
  pct_dir_classifier  = top_bottom_pct_classifier,
  direction_tabs      = list(Global = NULL, Top = "Top", Bottom = "Bottom"),
  gene_list_orders    = list(Global = .ALL_THRESHOLDS, Top = .TOP_THRESHOLDS, Bottom = .BOTTOM_THRESHOLDS),
  use_shape_overlay   = TRUE,
  bg_string_ids       = bg_string_ids,  # CAAS background, mapped in parse_background
  node_score_fn       = caas_node_score,
  write_all_lists_file = TRUE
)
```

## RER {.tabset}

RER's own DOMINO network, built against RER's own gene universe. Skipped entirely when RER didn't run this invocation. Same nested layout as CAAS above: each direction tab (**Global**/**Top**/**Bottom**) holds **Modules** / **Recurrence Heatmap** / **Recurrence Table** / **Gene lists** sub-tabs.

```{r rer_domino_networks, fig.width=10, fig.height=8, results='asis'}
RER_clust_desc_all <- NULL
if (!is.null(params$rer_domino_network_sif) && !is.null(params$rer_gene_lists_dir)) {
  .rer_net_full <- build_tool_network(params$rer_domino_network_sif, params$rer_domino_edge_scores_file)
  .rer_gene_list_files <- list.files(params$rer_gene_lists_dir, pattern = "\\.txt$", full.names = TRUE)
  # Order of the lists in the Modules sub-tab: global (rer_significant), then accelerating, then decelerating.
  .rer_rank <- dplyr::case_when(
    grepl("accelerat", basename(.rer_gene_list_files), ignore.case = TRUE) ~ 1L,
    grepl("decelerat", basename(.rer_gene_list_files), ignore.case = TRUE) ~ 2L,
    TRUE ~ 0L
  )
  .rer_gene_list_files <- .rer_gene_list_files[order(.rer_rank)]

  if (length(.rer_gene_list_files) > 0) {
    .rer_bg_string_ids <- if (!is.null(params$rer_background_file) && file.exists(params$rer_background_file)) {
      safe_string_map_ids(read_gene_vector(params$rer_background_file))
    } else bg_string_ids
    RER_clust_desc_all <- render_tool_section_nested(
      .rer_gene_list_files, .rer_net_full, params$rer_domino_modules_dir, "rer",
      roll_dir_classifier = rer_roll_classifier,
      pct_dir_classifier  = always_global_pct_classifier,
      direction_tabs      = list(Global = NULL, Top = "Acc", Bottom = "Decc"),
      gene_list_orders    = list(
        Global = c("rer_significant", "rer_accelerating", "rer_decelerating"),
        Top    = "rer_accelerating",
        Bottom = "rer_decelerating"
      ),
      use_shape_overlay   = FALSE,
      bg_string_ids       = .rer_bg_string_ids,
      node_score_fn       = rer_node_score
    )
  } else {
    cat("_No RER gene list files found._\n\n")
  }
} else {
  cat("_RER did not run this invocation - section skipped._\n\n")
}
```

## FADE {.tabset}

FADE's own DOMINO network, built against FADE's own gene universe. Skipped entirely when FADE didn't run this invocation. Same nested layout as CAAS above: each direction tab (**Global**/**Top**/**Bottom**) holds **Modules** / **Recurrence Heatmap** / **Recurrence Table** / **Gene lists** sub-tabs.

```{r fade_domino_networks, fig.width=10, fig.height=8, results='asis'}
FADE_clust_desc_all <- NULL
if (!is.null(params$fade_domino_network_sif) && !is.null(params$fade_gene_lists_dir)) {
  .fade_net_full <- build_tool_network(params$fade_domino_network_sif, params$fade_domino_edge_scores_file)
  .fade_gene_list_files <- list.files(params$fade_gene_lists_dir, pattern = "\\.txt$", full.names = TRUE)
  .fade_rank <- dplyr::case_when(
    grepl("global", basename(.fade_gene_list_files), ignore.case = TRUE) ~ 0L,
    grepl("top",    basename(.fade_gene_list_files), ignore.case = TRUE) ~ 1L,
    grepl("bottom", basename(.fade_gene_list_files), ignore.case = TRUE) ~ 2L,
    TRUE ~ 3L
  )
  .fade_gene_list_files <- .fade_gene_list_files[order(.fade_rank)]

  if (length(.fade_gene_list_files) > 0) {
    .fade_bg_string_ids <- if (!is.null(params$fade_background_file) && file.exists(params$fade_background_file)) {
      safe_string_map_ids(read_gene_vector(params$fade_background_file))
    } else bg_string_ids
    FADE_clust_desc_all <- render_tool_section_nested(
      .fade_gene_list_files, .fade_net_full, params$fade_domino_modules_dir, "fade",
      roll_dir_classifier = top_bottom_roll_classifier("Global"),
      pct_dir_classifier  = top_bottom_pct_classifier,
      direction_tabs      = list(Global = NULL, Top = "Top", Bottom = "Bottom"),
      gene_list_orders    = list(
        Global = c("fade_global_significant", "fade_top_significant", "fade_bottom_significant"),
        Top    = "fade_top_significant",
        Bottom = "fade_bottom_significant"
      ),
      use_shape_overlay   = FALSE,
      bg_string_ids       = .fade_bg_string_ids,
      node_score_fn       = fade_node_score
    )
  } else {
    cat("_No FADE gene list files found._\n\n")
  }
} else {
  cat("_FADE did not run this invocation - section skipped._\n\n")
}
```

```{r export_all_tool_tables, include=FALSE}
.ami_all_desc <- dplyr::bind_rows(
  if (!is.null(CAAS_clust_desc_all)) dplyr::mutate(CAAS_clust_desc_all, tool = "CAAS", .before = 1),
  if (!is.null(RER_clust_desc_all))  dplyr::mutate(RER_clust_desc_all,  tool = "RER",  .before = 1),
  if (!is.null(FADE_clust_desc_all)) dplyr::mutate(FADE_clust_desc_all, tool = "FADE", .before = 1)
)
.ami_all_terms <- dplyr::bind_rows(
  if (exists(".ami_term_rows_caas", envir = .GlobalEnv) && !is.null(.ami_term_rows_caas)) {
    dplyr::mutate(.ami_term_rows_caas, tool = "CAAS", .before = 1)
  },
  if (exists(".ami_term_rows_rer", envir = .GlobalEnv) && !is.null(.ami_term_rows_rer)) {
    dplyr::mutate(.ami_term_rows_rer, tool = "RER", .before = 1)
  },
  if (exists(".ami_term_rows_fade", envir = .GlobalEnv) && !is.null(.ami_term_rows_fade)) {
    dplyr::mutate(.ami_term_rows_fade, tool = "FADE", .before = 1)
  }
)
if (nrow(.ami_all_desc) > 0) {
  readr::write_tsv(.ami_all_desc, file.path(ami_networks_dir, "ami_module_descriptions_all_tools.tsv"))
}
if (nrow(.ami_all_terms) > 0) {
  readr::write_tsv(.ami_all_terms, file.path(ami_networks_dir, "ami_term_threshold_membership_all_tools.tsv"))
}
```

---

## Summary

This report summarizes the results of the **AMI (Active Module Identification)**
analysis: DOMINO-detected network modules per gene list, with STRING-derived
functional labels and DOMINO's own per-module significance.

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere - AMI Analysis Module*
*Date of generation: `r Sys.Date()`*
