---
title: "ORA Analysis (clusterProfiler)"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
    mathjax: null                       # no math in this report
    md_extensions: "-tex_math_dollars"  # gene-set names contain '$' (e.g. V$HNF4_01) — never parse as TeX math
params:
  background_file: NULL
  background_basename: NULL
  gene_list_files: NULL
  output_dir: "ora"
  project_name: "ORA_clusterProfiler"
  organism: "hsapiens"
  ora_databases: "geneontology_Biological_Process_noRedundant,geneontology_Cellular_Component_noRedundant,geneontology_Molecular_Function_noRedundant,pathway_KEGG,pathway_Reactome,pathway_Wikipathway"
  gmt_dir: NULL
  min_num: 5
  max_num: 500
  fdr_thr: 0.1
  report_num: 20
  top_thr: 15
  n_threads: 8
  enable_overlap_heatmap: true
  enable_enrichment_map: false
  overlap_top_terms: 50
  compare_metric: "overlap"
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7,
  dev = 'png',
  dpi = 320
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
  library(kableExtra)
  library(clusterProfiler)
  library(enrichplot)
  library(org.Hs.eg.db)
  library(paletteer)
})

# ── Explicit namespace guards ─────────────────────────────────────────────────
# BioConductor packages (clusterProfiler, org.Hs.eg.db, S4Vectors, AnnotationDbi)
# mask several tidyverse functions.  Re-pin the ones we use to their dplyr/tidyr
# originals so pipe chains behave as expected regardless of load order.
filter   <- dplyr::filter
select   <- dplyr::select
rename   <- dplyr::rename
mutate   <- dplyr::mutate
arrange  <- dplyr::arrange
group_by <- dplyr::group_by
count    <- dplyr::count

# ── Shared colour palette (mirrors 9.Characterization_report.Rmd for consistency) ──────────
tableau_10 <- paletteer_d("ggthemes::Tableau_10")
names(tableau_10) <- c(
  "blue", "orange", "vermillion", "bluish_green",
  "green", "yellow", "reddish_purple", "pink",
  "brown", "gray"
)

has_pkg <- function(pkg) requireNamespace(pkg, quietly = TRUE)

`%||%` <- function(a, b) {
  if (!is.null(a) && length(a) > 0 && !all(is.na(a))) a else b
}

createDir <- function(directory) {
  if (!dir.exists(directory)) dir.create(directory, recursive = TRUE, showWarnings = FALSE)
}

read_gene_vector <- function(path) {
  x <- readr::read_lines(path)
  x <- trimws(x)
  x <- x[x != ""]
  unique(x)
}

safe_symbol_to_entrez <- function(genes) {
  if (length(genes) == 0) return(character(0))
  mapped <- tryCatch(
    suppressWarnings(
      bitr(
        genes,
        fromType = "SYMBOL",
        toType = "ENTREZID",
        OrgDb = org.Hs.eg.db,
        drop = TRUE
      )
    ),
    error = function(e) NULL
  )
  if (is.null(mapped) || nrow(mapped) == 0) return(character(0))
  unique(as.character(mapped$ENTREZID))
}

bg_file <- params$background_file
if (!file.exists(bg_file) && !is.null(params$background_basename)) {
  if (file.exists(params$background_basename)) bg_file <- params$background_basename
}

if (is.null(bg_file) || !file.exists(bg_file)) {
  stop("Error: background_file not found")
}
# Resolve background to absolute path so it survives any working-directory changes
bg_file <- normalizePath(bg_file, mustWork = TRUE)
bg_basename <- basename(bg_file)

# Auto-detect gene list files from the current working directory, excluding the
# background file.  Using normalizePath() ensures that subsequent reads succeed
# even if the working directory changes during report rendering.
gene_list_files <- params$gene_list_files
if (is.null(gene_list_files) || length(gene_list_files) == 0) {
  gene_list_files <- list.files(pattern = "\\.txt$", full.names = TRUE)
  gene_list_files <- gene_list_files[basename(gene_list_files) != bg_basename]
}
gene_list_files <- gene_list_files[file.exists(gene_list_files)]
# Always normalise to absolute paths
gene_list_files <- normalizePath(gene_list_files, mustWork = FALSE)
gene_list_files <- gene_list_files[file.exists(gene_list_files)]
no_gene_lists <- length(gene_list_files) == 0

dbs <- strsplit(as.character(params$ora_databases), ",")[[1]]
dbs <- trimws(dbs)
dbs <- dbs[dbs != ""]
if (length(dbs) == 0) stop("Error: ora_databases is empty")

ora_root <- "."
ora_results_dir <- "ora_results"
ora_summary_dir <- "ora_summary"
ora_plots_dir <- "ora_plots"
ora_files_dir <- "ORA_files"
cache_dir <- file.path(ora_files_dir, ".cache")

createDir(ora_root)
createDir(ora_results_dir)
createDir(ora_summary_dir)
createDir(ora_plots_dir)
createDir(ora_files_dir)
createDir(cache_dir)

# Normalise all directory paths to ABSOLUTE paths immediately after creation.
# rmarkdown/knitr may change the working directory during HTML-report generation,
# so relative paths can resolve to wrong locations once the working directory
# drifts.  Capture the absolute working dir first, then build absolute paths so
# every subsequent file.path() call is wd-independent.
ora_work_dir    <- normalizePath(".", mustWork = TRUE)
ora_results_dir <- file.path(ora_work_dir, ora_results_dir)
ora_summary_dir <- file.path(ora_work_dir, ora_summary_dir)
ora_plots_dir   <- file.path(ora_work_dir, ora_plots_dir)
ora_files_dir   <- file.path(ora_work_dir, ora_files_dir)
cache_dir       <- file.path(ora_work_dir, "ORA_files", ".cache")
```

# ORA configuration

```{r config_overview}
cat("=== ORA configuration ===\n")
cat("Background file:", bg_file, "\n")
cat("Gene list files:", length(gene_list_files), "\n")
cat("Organism:", params$organism, "\n")
cat("Databases:", paste(dbs, collapse = ", "), "\n")
cat("Thresholds: min_num=", params$min_num, ", max_num=", params$max_num, ", fdr_thr=", params$fdr_thr, "\n", sep = "")
```

`r if (no_gene_lists) "> **Note:** No gene list files were passed to this ORA run (gene lists may be empty due to insufficient data). Skipping enrichment analysis."`

# Parse background and convert to ENTREZ

```{r parse_background, eval=!no_gene_lists}
bg_symbols <- read_gene_vector(bg_file)
cat("Background symbols:", length(bg_symbols), "\n")

if (length(bg_symbols) < params$min_num) {
  stop("Error: background symbols set is too small for ORA")
}

# ── Load GMT files dynamically ──────────────────────────────────────────────
gmt_dir <- if (!is.null(params$gmt_dir) && nzchar(params$gmt_dir)) {
  params$gmt_dir
} else {
  file.path(dirname(normalizePath(knitr::current_input(), mustWork = FALSE)), "..", "dat")
}
gmt_files <- list.files(gmt_dir, pattern = "\\.gmt$", full.names = TRUE)

term2gene_list <- list()
for (f in gmt_files) {
  gmt_data <- tryCatch(clusterProfiler::read.gmt(f), error = function(e) NULL)
  if (!is.null(gmt_data) && nrow(gmt_data) > 0) {
    if ("term" %in% colnames(gmt_data)) {
      colnames(gmt_data)[colnames(gmt_data) == "term"] <- "ont"
    }
    # Derive a clean database label from the file name, version-agnostically:
    #   MSigDB   c2.all.v2026.1.Hs.symbols.gmt -> c2
    #   WebGestalt pathway_KEGG.symbols.gmt    -> pathway_KEGG
    db_name <- basename(f)
    db_name <- sub("\\.gmt$", "", db_name)
    db_name <- sub("\\.v[0-9.]+\\.Hs\\.symbols$", "", db_name)
    db_name <- sub("\\.symbols$", "", db_name)
    db_name <- sub("\\.all$", "", db_name)
    gmt_data$database <- db_name
    term2gene_list[[f]] <- gmt_data
  }
}
term2gene_all <- dplyr::bind_rows(term2gene_list)
term_db_map <- unique(term2gene_all[, c("ont", "database")])
```

# Run ORA per gene list

```{r run_ora, eval=!no_gene_lists, include=FALSE}
run_single_ora <- function(gene_file, bg_symbols, term2gene_all, term_db_map) {
  genes_sym <- read_gene_vector(gene_file)
  list_name <- tools::file_path_sans_ext(basename(gene_file))
  out_dir <- file.path(ora_results_dir, list_name)
  createDir(out_dir)

  if (length(genes_sym) < params$min_num) {
    return(list(
      list_name = list_name,
      input_n = length(genes_sym),
      mapped_n = length(genes_sym),
      status = "skipped_too_few_genes",
      n_rows = 0,
      results = tibble()
    ))
  }

  res <- tryCatch({
    clusterProfiler::enricher(
      gene = genes_sym,
      universe = bg_symbols,
      pvalueCutoff = params$fdr_thr,
      pAdjustMethod = "BH",
      minGSSize = params$min_num,
      maxGSSize = params$max_num,
      TERM2GENE = term2gene_all[, c("ont", "gene")]
    )
  }, error = function(e) {
    message(sprintf("enricher failed for %s: %s", list_name, e$message))
    NULL
  })

  if (is.null(res)) {
    return(list(
      list_name = list_name,
      input_n = length(genes_sym),
      mapped_n = length(genes_sym),
      status = "failed",
      n_rows = 0,
      results = tibble()
    ))
  }

  res_df <- as.data.frame(res)

  if (nrow(res_df) == 0) {
    return(list(
      list_name = list_name,
      input_n = length(genes_sym),
      mapped_n = length(genes_sym),
      status = "no_results",
      n_rows = 0,
      results = tibble()
    ))
  }

  # Translate clusterProfiler columns to WebGestalt format for downstream code compatibility
  res2 <- res_df %>%
    left_join(term_db_map, by = c("ID" = "ont")) %>%
    mutate(
      geneSet = ID,
      description = Description,
      link = paste0("https://www.gsea-msigdb.org/gsea/msigdb/human/geneset/", ID),
      size = as.integer(gsub("/.*", "", BgRatio)),
      overlap = Count,
      gene_ratio_val = sapply(GeneRatio, function(x) {
        nums <- as.numeric(strsplit(x, "/")[[1]])
        nums[1] / nums[2]
      }),
      bg_ratio_val = sapply(BgRatio, function(x) {
        nums <- as.numeric(strsplit(x, "/")[[1]])
        nums[1] / nums[2]
      }),
      enrichmentRatio = gene_ratio_val / bg_ratio_val,
      pValue = pvalue,
      FDR = p.adjust,
      userId = geneID,
      gene_list = list_name
    ) %>%
    select(geneSet, description, link, size, overlap, enrichmentRatio, pValue, FDR, userId, database, gene_list)

  readr::write_tsv(res2, file.path(ora_summary_dir, paste0(list_name, "_results.tsv")))

  # Save clusterProfiler enrichResult object
  saveRDS(res, file.path(out_dir, paste0(list_name, "_enrichResult.rds")))

  list(
    list_name = list_name,
    input_n = length(genes_sym),
    mapped_n = length(genes_sym),
    status = "ok",
    n_rows = nrow(res2),
    results = res2
  )
}

ora_runs <- lapply(gene_list_files, run_single_ora, bg_symbols = bg_symbols, term2gene_all = term2gene_all, term_db_map = term_db_map)
```

# Summary and plots {.tabset .tabset-fade .tabset-pills}

```{r summary_and_plots, results='asis', eval=!no_gene_lists}
# Helper: build a fill/color scale that expands Tableau_10 when n > 10
.pal10_base <- as.character(paletteer::paletteer_d("ggthemes::Tableau_10"))
.safe_pal   <- function(n) if (n <= 10L) .pal10_base[seq_len(n)] else colorRampPalette(.pal10_base)(n)
safe_fill   <- function(n) scale_fill_manual(values  = .safe_pal(n))
safe_color  <- function(n) scale_color_manual(values = .safe_pal(n))

run_summary <- purrr::map_dfr(ora_runs, ~tibble(
  gene_list = .x$list_name,
  input_genes = .x$input_n,
  mapped_entrez = .x$mapped_n,
  status = .x$status,
  n_results = .x$n_rows
))

all_results <- purrr::map_dfr(ora_runs, "results")

readr::write_tsv(run_summary, file.path(ora_summary_dir, "ora_run_summary.tsv"))
readr::write_tsv(all_results, file.path(ora_summary_dir, "ora_all_results.tsv"))

kable(run_summary, caption = "ORA run summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

if (nrow(all_results) > 0) {
  fdr_col <- c("FDR", "fdr", "adjP", "fdrBH", "fdr_bh")
  fdr_col <- fdr_col[fdr_col %in% names(all_results)][1] %||% "FDR"
  db_col <- c("database", "enrichmentCategory", "category")
  db_col <- db_col[db_col %in% names(all_results)][1] %||% "database"
  desc_col <- c("description", "geneSet", "term", "name")
  desc_col <- desc_col[desc_col %in% names(all_results)][1] %||% "description"

  std_results <- all_results %>%
    mutate(
      FDR_std = suppressWarnings(as.numeric(.data[[fdr_col]])),
      FDR_std = pmax(FDR_std, .Machine$double.xmin, na.rm = TRUE),
      database_std = as.character(.data[[db_col]]),
      term_std = as.character(.data[[desc_col]]),
      neglog10FDR = -log10(FDR_std)
    ) %>%
    filter(!is.na(FDR_std), !is.na(term_std), term_std != "")

  # ── Parse scenario (global / top / bottom) and gate (sig / fdr) ─────────────
  std_results <- std_results %>%
    mutate(
      scenario = coalesce(
        stringr::str_match(gene_list, "_(top|bottom)(?:_|$)")[, 2],
        "global"
      ),
      gate = coalesce(
        stringr::str_match(gene_list, "_(sig|fdr)_?")[, 2],
        "unknown"
      )
    )
  std_results_all   <- std_results
  present_scenarios <- intersect(c("global", "top", "bottom"), unique(std_results$scenario))
  cat(sprintf("Scenarios detected: %s\n\n", paste(present_scenarios, collapse = ", ")))

  # Nested tabs: direction (##) → gate (###)
  for (.scen in present_scenarios) {
    cat(sprintf("\n## %s {.tabset .tabset-fade}\n\n", tools::toTitleCase(.scen)))

    gates_in_scen <- intersect(c("sig", "fdr", "unknown"),
                                unique(std_results_all$gate[std_results_all$scenario == .scen]))

    for (.gate in gates_in_scen) {
      std_results <- std_results_all %>% filter(scenario == .scen, gate == .gate)
      if (nrow(std_results) == 0) next

      gate_label <- if (.gate == "unknown") "All" else toupper(.gate)
      cat(sprintf("\n### %s\n\n", gate_label))

      # ── Top hits bar chart ─────────────────────────────────────────────────
      top_hits <- std_results %>%
        arrange(FDR_std) %>%
        slice_head(n = params$top_thr %||% 15L)

      p_hits <- ggplot(top_hits,
                       aes(x = reorder(term_std, neglog10FDR), y = neglog10FDR,
                           fill = database_std)) +
        geom_col() +
        coord_flip() +
        safe_fill(n_distinct(top_hits$database_std)) +
        theme_minimal() +
        labs(
          title = sprintf("Top ORA hits — %s / %s gate",
                          tools::toTitleCase(.scen), gate_label),
          x = NULL, y = "-log10(FDR)", fill = "Database"
        )

      print(p_hits)
      ggsave(file.path(ora_plots_dir,
               paste0("top_ora_hits_", .scen, "_", .gate, ".png")),
             p_hits, width = 14, height = 9, dpi = 320)

      # ── Enriched terms by database ─────────────────────────────────────────
      db_counts <- std_results %>% count(database_std, sort = TRUE)

      p_db <- ggplot(db_counts,
                     aes(x = reorder(database_std, n), y = n, fill = database_std)) +
        geom_col() +
        coord_flip() +
        safe_fill(n_distinct(db_counts$database_std)) +
        guides(fill = "none") +
        theme_minimal() +
        labs(title = "Enriched terms by database", x = NULL, y = "N terms")

      print(p_db)
      ggsave(file.path(ora_plots_dir,
               paste0("enriched_terms_by_database_", .scen, "_", .gate, ".png")),
             p_db, width = 12, height = 7, dpi = 320)

      # ── Top enriched terms table ───────────────────────────────────────────
      top_terms_df <- std_results %>%
        arrange(FDR_std) %>%
        slice_head(n = params$report_num %||% 20L) %>%
        mutate(term_short = stringr::str_trunc(term_std, 65)) %>%
        select(Database = database_std, Term = term_short,
               `-log10(FDR)` = neglog10FDR, FDR = FDR_std)

      readr::write_tsv(
        std_results %>% arrange(FDR_std),
        file.path(ora_summary_dir,
                  paste0("ora_top_terms_", .scen, "_", .gate, ".tsv"))
      )

      print(
        kable(top_terms_df, digits = 4,
              caption = sprintf("Top enriched terms — %s / %s gate",
                                tools::toTitleCase(.scen), gate_label)) %>%
          kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                        full_width = FALSE)
      )
      cat("\n")
    } # end gate loop
  } # end scenario loop
  std_results <- std_results_all   # restore for enrichment_heatmap / upset_plot chunks
}
```

# Enrichment signal heatmap {.tabset .tabset-fade .tabset-pills}

Rows are the union of the top-N enriched terms per gene list; columns are the
gene list scenarios. Cell colour encodes **-log10(FDR)**; zero (white) means the
term was not enriched in that scenario. Hierarchical clustering (when ≥ 3
scenarios) reveals groups of scenarios with similar functional landscapes.
A separate per-database heatmap is produced for each functional category so
terms from different ontologies are compared on the same scale.

```{r enrichment_heatmap, results='asis', eval=!no_gene_lists}
for (.scen in (if (exists("present_scenarios")) present_scenarios else "global")) {
  if (exists("std_results_all")) std_results <- std_results_all %>% dplyr::filter(scenario == .scen)
  cat(sprintf("\n## %s\n\n", tools::toTitleCase(.scen)))

if (exists("std_results") && nrow(std_results) > 0 && isTRUE(params$enable_overlap_heatmap)) {

  hm_top_n <- as.integer(params$overlap_top_terms %||% 30)
  hm_top_n <- if (is.na(hm_top_n) || hm_top_n <= 0) 30L else min(hm_top_n, 50L)

  # Union of top enriched terms across all lists
  hm_terms <- std_results %>%
    arrange(FDR_std) %>%
    group_by(gene_list) %>%
    slice_head(n = hm_top_n) %>%
    ungroup() %>%
    distinct(term_std) %>%
    pull(term_std)

  if (length(hm_terms) > 0) {

    # Build term × list signal matrix (-log10 FDR; 0 = not enriched)
    hm_wide <- std_results %>%
      filter(term_std %in% hm_terms) %>%
      group_by(gene_list, term_std) %>%
      summarise(signal = max(neglog10FDR, na.rm = TRUE), .groups = "drop") %>%
      tidyr::pivot_wider(names_from = gene_list, values_from = signal, values_fill = 0) %>%
      tibble::column_to_rownames("term_std")

    rownames(hm_wide) <- stringr::str_trunc(rownames(hm_wide), 65)
    hm_mat <- as.matrix(hm_wide)
    hm_mat <- hm_mat[order(rowSums(hm_mat), decreasing = TRUE), , drop = FALSE]

    cat(sprintf("Enrichment heatmap: %d terms × %d gene lists\n", nrow(hm_mat), ncol(hm_mat)))

    # ── render function (pheatmap or ggplot fallback) ──────────────────────
    draw_signal_heatmap <- function(mat, title, palette_high, out_png) {
      if (has_pkg("pheatmap")) {
        library(pheatmap)
        p_out <- pheatmap(
          mat,
          color        = colorRampPalette(c("white", palette_high))(100),
          cluster_rows = nrow(mat) > 2,
          cluster_cols = ncol(mat) > 2,
          fontsize_row = max(8, min(12, 200 / nrow(mat))),
          fontsize_col = 9,
          main         = title,
          angle_col    = 45,
          border_color = NA,
          silent       = TRUE
        )
        png(out_png,
            width  = max(1200, ncol(mat) * 130 + 800),
            height = max(900,  nrow(mat) * 30  + 250), res = 160)
        grid::grid.newpage()
        grid::grid.draw(p_out$gtable)
        dev.off()
        grid::grid.newpage()
        grid::grid.draw(p_out$gtable)
      } else {
        long_df <- as.data.frame(mat) %>%
          tibble::rownames_to_column("term") %>%
          tidyr::pivot_longer(-term, names_to = "gene_list", values_to = "signal") %>%
          mutate(term = factor(term, levels = rev(rownames(mat))))
        p_gg <- ggplot(long_df, aes(x = gene_list, y = term, fill = signal)) +
          geom_tile(color = "grey90", linewidth = 0.15) +
          scale_fill_gradient(low = "white", high = palette_high, name = "-log10(FDR)") +
          theme_minimal(base_size = 16) +
          labs(title = title,
               subtitle = "-log10(FDR) | rows = enriched terms, cols = scenarios",
               x = NULL, y = NULL) +
          theme(
            axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
            axis.text.y = element_text(colour = "black",
                                        size = max(8, min(12, 180 / nrow(mat))))
          )
        print(p_gg)
        ggsave(out_png, p_gg,
               width  = max(8, ncol(mat) * 1.5 + 3),
               height = max(6, nrow(mat) * 0.32 + 2), dpi = 300)
      }
    }

    # Global heatmap (all databases pooled)
    draw_signal_heatmap(
      hm_mat,
      title       = "Enrichment signal: -log10(FDR) per term × scenario (all databases)",
      palette_high = as.character(tableau_10[["blue"]]),
      out_png     = file.path(ora_plots_dir, paste0("ora_enrichment_heatmap_", .scen, ".png"))
    )

    # Per-database heatmaps
    for (db_i in unique(std_results$database_std)) {
      db_res <- std_results %>% filter(database_std == db_i)
      db_terms <- db_res %>%
        arrange(FDR_std) %>%
        group_by(gene_list) %>%
        slice_head(n = 20) %>%
        ungroup() %>%
        distinct(term_std) %>%
        pull(term_std)
      if (length(db_terms) < 2) next

      db_wide <- db_res %>%
        filter(term_std %in% db_terms) %>%
        group_by(gene_list, term_std) %>%
        summarise(signal = max(neglog10FDR, na.rm = TRUE), .groups = "drop") %>%
        tidyr::pivot_wider(names_from = gene_list, values_from = signal, values_fill = 0) %>%
        tibble::column_to_rownames("term_std")

      rownames(db_wide) <- stringr::str_trunc(rownames(db_wide), 60)
      db_mat <- as.matrix(db_wide)
      db_mat <- db_mat[order(rowSums(db_mat), decreasing = TRUE), , drop = FALSE]
      db_slug <- gsub("[^A-Za-z0-9_]", "_", db_i)

      draw_signal_heatmap(
        db_mat,
        title        = paste0("-log10(FDR) | ", db_i),
        palette_high  = as.character(tableau_10[["vermillion"]]),
        out_png      = file.path(ora_plots_dir, paste0("ora_heatmap_", db_slug, "_", .scen, ".png"))
      )
    }

  } else {
    cat("No enriched terms found — enrichment heatmap skipped.\n")
  }
} else {
  cat("Enrichment heatmap disabled or no results available.\n")
}
} # end for (.scen)
if (exists("std_results_all")) std_results <- std_results_all
```

# UpSet plot: term intersections across scenarios {.tabset .tabset-fade .tabset-pills}

Each gene list / scenario is treated as a **set of enriched terms**. The UpSet
plot decomposes those sets into their exact intersections — revealing terms
significantly enriched in every scenario, terms unique to a single scenario, and
all gradations in between. This is more informative than a pairwise Jaccard score
because it shows the precise multi-way structure of the overlap.

Requires the **UpSetR** (preferred) or **ComplexUpset** package.
Install with `install.packages("UpSetR")`.

```{r upset_plot, fig.width=14, fig.height=8, results='asis', eval=!no_gene_lists}
if (exists("std_results") && nrow(std_results) > 0 && isTRUE(params$enable_enrichment_map)) {

for (.scen in present_scenarios) {
  std_results <- std_results_all %>% filter(scenario == .scen)
  cat(sprintf("\n\n## %s\n\n", .scen))

  upset_top_n <- as.integer(params$overlap_top_terms %||% 50)
  upset_top_n <- if (is.na(upset_top_n) || upset_top_n <= 0) 50L else min(upset_top_n, 75L)

  upset_terms <- std_results %>%
    arrange(FDR_std) %>%
    group_by(gene_list) %>%
    slice_head(n = upset_top_n) %>%
    ungroup() %>%
    distinct(gene_list, term_std)

  all_terms_u <- unique(upset_terms$term_std)
  glists_u    <- unique(upset_terms$gene_list)

  if (length(glists_u) >= 2 && length(all_terms_u) > 0) {

    # Binary membership matrix: rows = terms, cols = gene lists
    upset_mat <- as.data.frame(
      setNames(
        lapply(glists_u, function(gl) {
          as.integer(all_terms_u %in% upset_terms$term_std[upset_terms$gene_list == gl])
        }),
        glists_u
      )
    )
    rownames(upset_mat) <- all_terms_u
    upset_mat <- upset_mat[rowSums(upset_mat) >= 1, , drop = FALSE]

    cat(sprintf("UpSet input: %d terms × %d gene lists\n", nrow(upset_mat), ncol(upset_mat)))

    if (has_pkg("UpSetR")) {
      library(UpSetR)
      upset_args <- list(
        data            = upset_mat,
        sets            = rev(glists_u),
        order.by        = "freq",
        decreasing      = TRUE,
        mb.ratio        = c(0.60, 0.40),
        text.scale      = c(1.5, 1.2, 1.1, 1.0, 1.2, 1.0),
        mainbar.y.label = "# shared enriched terms",
        sets.x.label    = "Total enriched terms per scenario",
        main.bar.color  = "#4393c3",
        sets.bar.color  = "#2166ac",
        matrix.color    = "#08306b",
        point.size      = 3.0,
        line.size       = 0.8,
        nintersects     = min(40L, 2L ^ length(glists_u))
      )
      # Save to file
      png(
        file.path(ora_plots_dir, paste0("ora_upset_terms_", .scen, ".png")),
        width  = max(1600, length(glists_u) * 200 + 600),
        height = max(900,  length(glists_u) * 80  + 400),
        res    = 150
      )
      do.call(upset, upset_args)
      dev.off()
      # Render inline
      do.call(upset, upset_args)

    } else if (has_pkg("ComplexUpset")) {
      library(ComplexUpset)
      upset_tbl <- upset_mat %>%
        tibble::rownames_to_column("term") %>%
        dplyr::mutate(dplyr::across(all_of(glists_u), as.logical))
      p_upset_cx <- upset(
        upset_tbl, glists_u,
        name        = "Gene list",
        width_ratio = 0.25,
        min_size    = 1,
        sort_sets   = FALSE,
        base_annotations = list(
          "Shared terms" = intersection_size(
            counts = TRUE,
            mapping = aes(fill = "bar")
          ) + scale_fill_manual(values = c("bar" = "#4393c3"), guide = "none")
        )
      ) +
        labs(
          title    = "UpSet: intersection of enriched terms across gene lists",
          subtitle = sprintf("Top %d enriched terms per scenario", upset_top_n)
        )
      print(p_upset_cx)
      ggsave(
        file.path(ora_plots_dir, paste0("ora_upset_terms_", .scen, ".png")), p_upset_cx,
        width  = max(12, length(glists_u) * 2 + 6),
        height = max(8,  length(glists_u) * 1.2 + 4), dpi = 200
      )

    } else {
      # Fallback: manual intersection decomposition rendered as a bar chart
      message("UpSetR not found — falling back to intersection bar chart. Install with: install.packages('UpSetR')")
      n_lists <- length(glists_u)
      if (n_lists <= 6) {
        all_combos <- do.call(c, lapply(seq_len(n_lists), function(k) combn(glists_u, k, simplify = FALSE)))
        combo_df <- purrr::map_dfr(all_combos, function(combo) {
          in_all   <- rowSums(upset_mat[, combo, drop = FALSE]) == length(combo)
          out_rest <- if (length(combo) < n_lists) {
            rowSums(upset_mat[, setdiff(glists_u, combo), drop = FALSE]) == 0
          } else {
            rep(TRUE, nrow(upset_mat))
          }
          tibble::tibble(
            intersection = paste(sort(combo), collapse = " ∩ "),
            n_terms      = sum(in_all & out_rest),
            n_sets       = length(combo)
          )
        }) %>%
          filter(n_terms > 0) %>%
          arrange(desc(n_terms))

        p_fallback <- ggplot(
          combo_df %>% slice_head(n = 30),
          aes(x = reorder(intersection, n_terms), y = n_terms, fill = factor(n_sets))
        ) +
          geom_col() +
          coord_flip() +
          scale_fill_paletteer_d("ggthemes::Tableau_10", name = "# lists") +
          theme_minimal() +
          labs(
            title    = "Enriched term intersections across gene lists",
            subtitle = "(UpSetR not installed — simplified view)",
            x        = NULL,
            y        = "# shared terms"
          )
        print(p_fallback)
        ggsave(
          file.path(ora_plots_dir, paste0("ora_upset_fallback_", .scen, ".png")), p_fallback,
          width = 12, height = 8, dpi = 300
        )
      } else {
        cat("Too many gene lists for the fallback intersection chart (> 6). Please install UpSetR.\n")
      }
    }

  } else {
    cat("UpSet plot requires ≥ 2 gene lists with enrichment results — skipping.\n")
  }
} # end for (.scen in present_scenarios)
if (exists("std_results_all")) std_results <- std_results_all
} else {
  cat("UpSet plot disabled or no results available.\n")
}
```

---

## Summary

This report summarizes the results of the **Enrichment** analysis.

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — Enrichment Module*
*Date of generation: `r Sys.Date()`*
