---
title: "Position-level Enrichment (Fisher, fixed cutoffs) — `r params$traitname`"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
    mathjax: null                       # no math in this report
    md_extensions: "-tex_math_dollars"  # pathway names contain '$' (e.g. V$HNF4_01) — never parse as TeX math
params:
  results_file: NULL
  leading_edge_file: NULL
  traitname: "trait"
  padj_thr: 0.15
  fold_thr: 1.5
  table_top_n: 200
  # Position Characterisation (moved here from 11.Scoring_report.Rmd): PrimateAI-3D
  # + COSMIC validation and regional accumulation, all restricted to the US scheme.
  # NULL/absent inputs degrade gracefully (has_pos_char is FALSE, section skipped).
  position_scores_file: NULL   # position_scores.tsv (per-position CAAS scores)
  # SCORING's published position_lists/slice_<dir><pct>.tsv dir (scoring_compute.R)
  # -- when supplied, the 10/5/1% membership used throughout Position
  # Characterisation is read from there instead of an independent quantile()
  # cut computed here (see the has_pos_char setup chunk's test_glob_*/test_top_*/
  # test_bot_* closures). Falls back to quantile() per direction when NULL or a
  # slice file is missing.
  position_lists_dir: NULL
  gene_scores_file:     NULL   # gene_scores.tsv (background gene fallback)
  vep_primateai_file:   NULL   # PrimateAI-3D score TSV
  vep_cosmic_file:      NULL   # COSMIC Mutant Census score TSV
  genomic_info_file:    NULL   # gene, chr, start, end, length (no header assumed)
  # FADE_top_sig/FADE_bottom_sig position group (gene,position,max_bf,target_aa;
  # BF >= fade_bf_threshold only). Direction-specific, unlike PAI3D/COSMIC.
  fade_sites_top_file:    NULL
  fade_sites_bottom_file: NULL
  # SCORING's fcs_stats.tsv (gene + flag_* columns) -- same file already
  # consumed as --annot-file by posenrich_enrich.py, reused here for the
  # Overall dotplot's orthogonal-support composite background rates.
  fcs_stats_file: NULL
  # cleaned_background_main.txt -- the SAME universe_file 12.FCS_general_report.Rmd
  # uses. Pins the Overall dotplot's gene-level composite background to the
  # identical universe everywhere it's computed, rather than inferring a gene
  # count from nrow(fcs_stats.tsv) (a different file that usually, but isn't
  # guaranteed to, agree in size -- see fcs_stats.tsv's own full_join-vs-CAAS
  # history). NULL falls back to nrow(fcs_stats_df) as before.
  universe_file: NULL
  top_pct:      0.10
  window_size_bp: 1000000
  output_dir:   "."            # regional_accumulation.tsv export target
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6, dev = "png", dpi = 150)
suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
  library(kableExtra)
  library(DT)
  library(scales)
})

PADJ <- params$padj_thr
FOLD <- params$fold_thr
TOP_N <- params$table_top_n

this_dir <- tryCatch(dirname(normalizePath(knitr::current_input(), mustWork = FALSE)),
                     error = function(e) ".")
orth_candidates <- c(file.path(this_dir, "src", "orthogonal_score.R"), "src/orthogonal_score.R",
                     file.path(this_dir, "orthogonal_score.R"), "orthogonal_score.R")
orth_src <- orth_candidates[file.exists(orth_candidates)][1]
if (is.na(orth_src)) stop("orthogonal_score.R not found under src/ or next to 14.Position_enrichment_report.Rmd")
source(orth_src)

kbl_html <- function(df, caption = NULL) {
  kable(df, format = "html", escape = TRUE, caption = caption) %>%
    kable_styling(bootstrap_options = c("striped", "condensed", "hover"),
                  full_width = FALSE)
}

# Safe correlation helper: returns NA (not an error) when inputs have fewer
# than 3 complete pairs -- avoids crashes in sparse optional-module sections.
safe_cor <- function(x, y, method = "pearson") {
  ok <- complete.cases(x, y)
  if (sum(ok) < 3) return(NA_real_)
  cor(x[ok], y[ok], method = method)
}
```

```{r load, include=FALSE}
stopifnot(!is.null(params$results_file), file.exists(params$results_file))
res <- readr::read_tsv(params$results_file, show_col_types = FALSE)
# Drop 100%/50% cutoffs unconditionally -- retired from the analysis (100% is
# a no-op foreground = full background, 50% is too coarse to be informative)
# but older results_file/leading_edge_file runs may still carry them.
if (!is.null(res) && "top_frac" %in% names(res)) {
  res <- res %>% dplyr::filter(!top_frac %in% c(1.0, 0.5))
}
has_res <- !is.null(res) && nrow(res) > 0
# Canonical tab order everywhere `res$ranking`/`sig_res$ranking` drives a tab
# loop in this report -- Global > Top > Bottom, never left to depend on
# incidental row/column order in the upstream results file.
RANKING_ORDER <- c("global", "top", "bottom")
if (has_res && !"sig" %in% names(res)) {
  # No fold-enrichment gate by design — p.adj alone decides significance;
  # fold_enrichment stays available as a displayed column to filter on
  # downstream rather than being baked into what counts as "significant".
  res$sig <- res$p_adj < PADJ
}

# Truncates a "gene1, gene2, ..." string to the first n_show entries, with
# the remainder tucked behind an inline <details>/<summary> toggle -- a row
# can otherwise carry hundreds of comma-separated gene:positions.
truncate_gene_list <- function(x, n_show = 5) {
  vapply(x, function(s) {
    if (is.na(s) || !nzchar(s)) return("")
    items <- strsplit(s, ", ", fixed = TRUE)[[1]]
    if (length(items) <= n_show) return(s)
    shown <- paste(items[seq_len(n_show)], collapse = ", ")
    rest  <- paste(items[(n_show + 1):length(items)], collapse = ", ")
    sprintf(
      '%s <details style="display:inline"><summary style="display:inline; cursor:pointer; color:#2166ac;">&hellip; (%d more)</summary> %s</details>',
      shown, length(items) - n_show, rest
    )
  }, character(1), USE.NAMES = FALSE)
}

# Observed gene:position members of the overlap, significant terms only
# (posenrich_leading_edge.tsv — long format, one row per member). Optional.
le_raw <- NULL
if (!is.null(params$leading_edge_file) && file.exists(params$leading_edge_file)) {
  le_raw <- tryCatch(readr::read_tsv(params$leading_edge_file, show_col_types = FALSE),
                     error = function(e) NULL)
  if (!is.null(le_raw) && "top_frac" %in% names(le_raw)) {
    le_raw <- le_raw %>% dplyr::filter(!top_frac %in% c(1.0, 0.5))
  }
  if (!is.null(le_raw) && nrow(le_raw) == 0) le_raw <- NULL
}
le_summary <- if (!is.null(le_raw)) {
  le_raw %>%
    dplyr::group_by(ranking, database, pathway, top_frac) %>%
    dplyr::summarise(`Genes/positions` = paste(sort(unique(gene_position)), collapse = ", "),
                     `N genes` = dplyr::n_distinct(gene),
                     .groups = "drop")
} else NULL

# Direction-aware corroboration columns (mirrors 12.FCS_general_report.Rmd's
# cols_by_rk): fade_bottom only under `bottom`, fade_top only under `top` --
# no combined "fade" column under `global` (redundant with _top/_bottom, which
# already cover both directions). pct_accum_top/pct_accum_bottom are the
# direction-restricted accumulation tests (same convention as fade); pct_accum
# (non-directional, accumulation_all_* pooled) is reserved for `global`. Only
# columns actually present (i.e. an --annot-file was supplied to
# posenrich_enrich.py) are ever shown.
flag_cols_by_rk <- list(
  global = c("pct_gate_sig", "pct_gate_fdr", "pct_rer_acc", "pct_rer_decc", "pct_accum"),
  top    = c("pct_gate_sig", "pct_gate_fdr", "pct_fade_top", "pct_rer_acc", "pct_rer_decc", "pct_accum_top"),
  bottom = c("pct_gate_sig", "pct_gate_fdr", "pct_fade_bottom", "pct_rer_acc", "pct_rer_decc", "pct_accum_bottom")
)

# SCORING's fcs_stats.tsv -- universe-level flag_* rates for the Overall
# dotplot's orthogonal-support composite (lancaster_combine(), sourced
# above). Optional: the composite section degrades to raw -log10(p_adj)
# ordering when absent (see overall-dotplot chunk below). Defined here rather
# than reusing the later .file_ok() (Position Characterisation, further down
# in this file) since this chunk runs first.
.fcs_stats_file_ok <- function(p) !is.null(p) && !is.na(p) && nchar(p) > 0 && file.exists(p)
has_fcs_stats <- .fcs_stats_file_ok(params$fcs_stats_file)
fcs_stats_df <- if (has_fcs_stats) readr::read_tsv(params$fcs_stats_file, show_col_types = FALSE) else NULL

# Gene universe for the Overall dotplot's gene-level composite -- same
# cleaned_background universe_file 12.FCS_general_report.Rmd loads, so
# n_universe is identical everywhere the gene-level composite is computed
# instead of incidentally inferred from nrow(fcs_stats_df). Falls back to
# nrow(fcs_stats_df) when universe_file wasn't wired in (older runs).
n_universe_genes <- NA_integer_
if (.fcs_stats_file_ok(params$universe_file) && !grepl("^NO_", basename(params$universe_file))) {
  .universe_genes <- unique(trimws(readLines(params$universe_file)))
  .universe_genes <- .universe_genes[nzchar(.universe_genes)]
  if (length(.universe_genes) >= 2) n_universe_genes <- length(.universe_genes)
}
if (is.na(n_universe_genes) && has_fcs_stats) n_universe_genes <- nrow(fcs_stats_df)

# FADE_top_sig/FADE_bottom_sig: gene,position,max_bf,target_aa, direction-specific
# (never cross-checked -- a TOP-direction position is only ever validated against
# fade_sites_top, never fade_sites_bottom, and vice versa). Position is already in
# CAAS's own coordinate space (see build_position_gmt.py), so no map_cache lookup.
# Hoisted here (was previously loaded only inside the Position Characterisation
# section, further down) so the Overall dotplot's position-native FADE-overlap
# component can use it too -- position-granular, not the gene-level flag_fade_*.
has_fade_sites_top    <- .fcs_stats_file_ok(params$fade_sites_top_file)
has_fade_sites_bottom <- .fcs_stats_file_ok(params$fade_sites_bottom_file)
fade_sites_top_df <- if (has_fade_sites_top) {
  readr::read_csv(params$fade_sites_top_file, show_col_types = FALSE) %>%
    dplyr::rename(Gene = gene, Position = position) %>%
    dplyr::mutate(Position = suppressWarnings(as.integer(Position)))
} else NULL
fade_sites_bottom_df <- if (has_fade_sites_bottom) {
  readr::read_csv(params$fade_sites_bottom_file, show_col_types = FALSE) %>%
    dplyr::rename(Gene = gene, Position = position) %>%
    dplyr::mutate(Position = suppressWarnings(as.integer(Position)))
} else NULL
```

# Overview {.tabset .tabset-fade}

This report summarises **position-level enrichment** for `r params$traitname`. Unlike
the gene-level FCS (a Wilcoxon-AUC rank-shift test), the position background is a
sea of invariant columns that can never carry a CAAS change — a rank-shift test
degenerates there. Instead we use **Fisher's exact test** at four fixed
foreground cutoffs (top 25 / 10 / 5 / 1% of scored positions, per direction), run
against every pathway across 9 GMT-style databases plus the broad functional
`characterization` layers (UCR core/flank, FUBAR positive/purifying) — all on
the same mechanism.

A set is flagged **significant** when `p.adj < `r PADJ`` (BH-adjusted per
ranking × database). No fold-enrichment magnitude gate is applied here —
`fold_enrichment` is still reported alongside every result so a magnitude
cutoff can be applied downstream, but it does not affect which sets are
marked significant in this report. Fisher/hypergeometric tests gain enormous
power at this background size (~1.47M positions), so a barely-there deviation
from the null ratio can still reach a "significant" p-value on its own — keep
this in mind when filtering (see Methods).

```{r overview, echo=FALSE, results='asis'}
if (!has_res) {
  cat("No enrichment results were produced.\n")
} else {
  for (rk in intersect(RANKING_ORDER, unique(res$ranking))) {
    cat(sprintf("\n## %s\n\n", toupper(rk)))
    ov <- res %>% filter(ranking == rk) %>%
      group_by(database) %>%
      summarise(`Sets tested` = n_distinct(pathway),
                `Sig (p.adj)` = n_distinct(pathway[sig]),
                .groups = "drop") %>%
      arrange(database)
    print(kbl_html(ov, caption = sprintf(
      "%s — sets tested and significant at p.adj < %.3g.",
      toupper(rk), PADJ)))
    cat("\n\n")
  }
}
```

# Enriched position sets {.tabset .tabset-fade}

Foreground = the top 25 / 10 / 5 / 1% of scored positions by `CAAS_score` in
that direction (`top`/`bottom` restrict to `change_side` first, matching the
SCORING module's own percentile convention; no `global` ranking — dropped in
favour of top/bottom only, consistent with every other report in this
pipeline). Background = the full honest tested
position pool for every test — never restricted to the scored subset, so
tested-and-negative positions are never silently dropped from the null.
`fold > 1` = enriched, `< 1` = depleted. UCR core and UCR flank (and their
per-orthogroup counterparts) are tested as independent, potentially-overlapping
layers, not a forced partition — nothing here requires them to be disjoint.
When available, `%gate_sig`/`%gate_fdr`/`%fade_*`/`%rer_*`/`%accum` columns
report the percentage of *distinct genes* contributing to the overlap (never
raw positions — one gene must not dominate) that are independently flagged by
SCORING's own significance/FADE/RER/accumulation criteria — a cross-module
corroboration check for whether a hit is broadly supported or isolated.

```{r enrichment-tables, echo=FALSE, results='asis'}
if (!has_res) {
  cat("_No results._\n\n")
} else {
  for (rk in intersect(RANKING_ORDER, unique(res$ranking))) {
    cat(sprintf("\n## %s {.tabset}\n\n", toupper(rk)))
    rk_df <- res %>% filter(ranking == rk)
    for (db in unique(rk_df$database)) {
      cat(sprintf("\n### %s {.tabset}\n\n", db))
      db_df <- rk_df %>% filter(database == db)

      for (tf in sort(unique(db_df$top_frac), decreasing = TRUE)) {
        cat(sprintf("\n#### Top %s%%\n\n", format(100 * tf, trim = TRUE)))
        tf_all <- db_df %>% filter(top_frac == tf) %>% arrange(desc(sig), p_adj, p_value)
        n_total_tf <- nrow(tf_all)
        tf_src <- tf_all %>% head(TOP_N)
        if (n_total_tf > TOP_N) {
          cat(sprintf("_Showing top %d of %d tested pathways (sorted by significance)._\n\n", TOP_N, n_total_tf))
        }
        tf_df <- tf_src %>%
          transmute(Pathway = pathway, Description = description,
                    `Layer size` = layer_size, `Foreground` = foreground_size,
                    Observed = observed, Expected = round(expected, 1),
                    Fold = round(fold_enrichment, 3), Direction = direction,
                    `p (Fisher)` = signif(p_value, 3),
                    `p.adj` = signif(p_adj, 3), Sig = ifelse(sig, "*", ""))
        # Cross-module corroboration: % of the DISTINCT genes in the overlap
        # (never % of raw positions — one gene must not dominate) carrying
        # each SCORING flag (gate_sig/gate_fdr/fade/rer/accum). Only present
        # when POSENRICH was run with --annot-file. Direction-restricted via
        # flag_cols_by_rk (same convention as the Leading edge section below) --
        # posenrich_enrich.py computes pct_<flag> for every flag regardless of
        # direction, so without this restriction a `bottom` table would show
        # `%accum_top`/`%fade_top` alongside its own `%accum_bottom`/`%fade_bottom`.
        want_rk <- flag_cols_by_rk[[rk]]
        if (is.null(want_rk)) want_rk <- character(0)
        flag_cols <- intersect(want_rk, names(tf_src))
        if (length(flag_cols) > 0) {
          flag_df <- tf_src %>% transmute(across(all_of(flag_cols), ~round(.x, 1)))
          names(flag_df) <- paste0("%", sub("^pct_", "", flag_cols))
          tf_df <- dplyr::bind_cols(tf_df, flag_df)
        }
        id <- paste("dt", rk, db, gsub("\\.", "_", as.character(tf)), sep = "_")
        cat(knitr::knit_child(text = c(
          paste0("```{r ", id, ", echo=FALSE}"),
          "DT::datatable(tf_df, options = list(pageLength = 10, scrollX = TRUE), rownames = FALSE)",
          "```"), envir = environment(), quiet = TRUE))
        cat("\n\n")
      }
    }
  }
}
```

# Overall dotplot {.tabset .tabset-fade}

Pooled across all databases and threshold cutoffs, ranked by an **orthogonal
composite signal** computed at **position granularity** — this report's own
native unit, not genes (see Methods). Each term's evidence is tested at
whatever grain it's actually native to and Lancaster-combined:

- **Top-fraction recurrence** (the percentile-trio analogue): of this term's
  pooled position members, how many recur in the tighter 10%/5%/1% cutoffs —
  nested exactly like the gene-level percentile trio, same fractional-weight
  discount so the family contributes one test's worth, not three.
- **FADE-site overlap**: tested *position-by-position* against the honest
  position universe (is this exact position a FADE-significant site, not just
  "is the gene FADE-significant somewhere").
- **RER acceleration/deceleration and Accumulation significance**: gene-level
  calls by nature, so tested at **distinct-gene** resolution against the gene
  universe — testing them position-by-position would overstate the evidence,
  since every position in the same gene shares that gene's flag value and
  isn't an independent draw with respect to it.

`gate_sig`/`gate_fdr` are excluded for the same reason as elsewhere
(near-ceiling among already-significant terms, no discriminating power). One
dotplot per direction, restricted to the top `r params$table_top_n` terms.
Falls back to plain `-log10(p.adj)` ordering when `fcs_stats_file` wasn't
supplied.

```{r overall-dotplot, echo=FALSE, results='asis', fig.height=6}
.overall_dotplot_rows <- list()  # accumulated across rankings, written to disk after the loop
                                  # -- the Comparison report builds its merged/
                                  # fuzzy-matched summary table from this file.
if (!has_res || !any(res$sig, na.rm = TRUE)) {
  cat("_No significant sets._\n\n")
} else if (is.null(le_raw)) {
  cat("_No leading-edge data (posenrich_leading_edge.tsv missing) -- composite signal unavailable, falling back to raw significance elsewhere in this report._\n\n")
} else {
  # background_n: the exact N posenrich_enrich.py used for THIS row's own
  # Fisher test (respects per-database coverage restriction, e.g. COSMIC/
  # PAI3D) -- the position universe for the FADE-overlap component (FADE sites
  # aren't restricted to CAAS-scored positions, so they're tested against the
  # full honest background). n_scored is the SEPARATE, much smaller, SCORED-
  # position population top_frac cutoffs are actually defined over (see
  # posenrich_enrich.py: foreground_by_frac's n = round(frac * len(scored_sorted)))
  # -- using background_n there instead would test the wrong null (a position's
  # true chance of landing in the top 10% is ~10% of n_scored, not ~10% of the
  # much larger honest background).
  .has_bg_n <- "background_n" %in% names(res)
  .bg_n_fallback <- if (.has_bg_n) suppressWarnings(max(res$background_n, na.rm = TRUE)) else NA_integer_
  .has_n_scored <- "n_scored" %in% names(res)

  for (rk in intersect(RANKING_ORDER, unique(res$ranking[res$sig]))) {
    cat(sprintf("\n## %s\n\n", toupper(rk)))
    rk_sig <- res %>% filter(ranking == rk, sig)
    if (nrow(rk_sig) == 0) { cat("_No significant sets for this direction._\n\n"); next }

    le_rk <- le_raw %>% filter(ranking == rk)
    if (nrow(le_rk) == 0) { cat("_No leading-edge data for this direction._\n\n"); next }

    # Direction-matched gene-inherited flags + FADE site table -- same
    # convention as flag_cols_by_rk above (global gets the pooled/non-
    # directional accum flag; top/bottom get their own direction's).
    gene_flag_cols <- switch(rk,
      top    = c(rer_acc = "flag_rer_acc", rer_decc = "flag_rer_decc", accum = "flag_accum_top"),
      bottom = c(rer_acc = "flag_rer_acc", rer_decc = "flag_rer_decc", accum = "flag_accum_bottom"),
      c(rer_acc = "flag_rer_acc", rer_decc = "flag_rer_decc", accum = "flag_accum")
    )
    fade_df_rk   <- switch(rk, top = fade_sites_top_df, bottom = fade_sites_bottom_df, NULL)
    n_fade_sites <- if (!is.null(fade_df_rk)) nrow(dplyr::distinct(fade_df_rk, Gene, Position)) else 0L
    # Constant within this ranking (posenrich_enrich.py computes it once per
    # direction, before the database loop) -- take the first non-NA value
    # rather than re-deriving it per term.
    n_scored_rk <- if (.has_n_scored) suppressWarnings(dplyr::first(stats::na.omit(rk_sig$n_scored))) else NA_integer_
    if (length(n_scored_rk) == 0) n_scored_rk <- NA_integer_

    # list_threshold/n_occurrences: which top_frac cutoffs this term reached
    # significance at, and how many -- already folded into `signal` above (the
    # nested top-fraction-recurrence component), kept here as a separate
    # human-readable/auditable column rather than making the reader reverse-
    # engineer it from the composite number. The Comparison report displays
    # both this and the merged signal side by side.
    term_meta <- rk_sig %>%
      dplyr::group_by(database, pathway) %>%
      dplyr::summarise(p_adj = min(p_adj, na.rm = TRUE),
                        background_n = if (.has_bg_n) dplyr::first(background_n) else NA_integer_,
                        description = dplyr::first(dplyr::coalesce(description, pathway)),
                        list_threshold = paste(sort(100 * unique(top_frac), decreasing = TRUE), "%", sep = "", collapse = ", "),
                        n_occurrences = dplyr::n_distinct(top_frac),
                        .groups = "drop")

    sc_rows <- vector("list", nrow(term_meta))
    for (i in seq_len(nrow(term_meta))) {
      db <- term_meta$database[i]; pw <- term_meta$pathway[i]
      le_dp <- le_rk %>% dplyr::filter(database == db, pathway == pw)
      n_universe_positions <- if (!is.na(term_meta$background_n[i])) term_meta$background_n[i] else .bg_n_fallback
      fade_bg_rate <- if (n_fade_sites > 0 && !is.na(n_universe_positions) && n_universe_positions > 0) {
        n_fade_sites / n_universe_positions
      } else NA_real_
      sc_rows[[i]] <- posenrich_position_signal(
        le_dp, top_fracs = c(0.10, 0.05, 0.01), n_universe_scored = n_scored_rk,
        gene_flag_cols = gene_flag_cols,
        fcs_stats_df = fcs_stats_df, n_universe_genes = n_universe_genes,
        fade_df = fade_df_rk, n_universe_positions = n_universe_positions,
        fade_bg_rate = fade_bg_rate
      ) %>% dplyr::mutate(database = db, pathway = pw)
    }
    sc <- dplyr::bind_rows(sc_rows) %>%
      dplyr::left_join(term_meta, by = c("database", "pathway"))

    dp <- sc %>%
      dplyr::filter(!is.na(signal)) %>%
      arrange(desc(signal)) %>%
      head(TOP_N) %>%
      mutate(
        label       = stringr::str_trunc(sprintf("%s (%s)", description, database), 60),
        neglog10FDR = -log10(pmax(p_adj, 1e-300))
      )

    if (nrow(dp) == 0) { cat("_No terms to display._\n\n"); next }

    .overall_dotplot_rows[[rk]] <- dp %>% dplyr::mutate(ranking = rk, .before = 1)

    term_levels <- dp %>% arrange(signal) %>% pull(label) %>% unique()
    p <- ggplot(dp, aes(x = signal, y = factor(label, levels = term_levels),
                        size = n_le, colour = neglog10FDR)) +
      geom_point(alpha = 0.85) +
      scale_colour_viridis_c(name = "-log10 p.adj", option = "C") +
      scale_size_continuous(name = "N positions") +
      labs(title = sprintf("%s — %s (pooled across databases)", params$traitname, toupper(rk)),
           subtitle = "Ordered by position-granularity orthogonal composite signal",
           x = "signal (orthogonal composite)", y = NULL) +
      theme_minimal(base_size = 11) +
      theme(axis.text.y = element_text(colour = "black",
                                        size = pmax(8, pmin(11, 220 / length(term_levels)))))
    print(p); cat("\n")
  }
}
if (length(.overall_dotplot_rows) > 0) {
  readr::write_tsv(dplyr::bind_rows(.overall_dotplot_rows), "posenrich_overall_dotplot.tsv")
}
```

# Top vs Bottom (orthogonal signal)

One point per (database, pathway): x = **top**-direction orthogonal composite
signal, y = **bottom**-direction signal (each pooled across cutoffs and
computed at position granularity — the same `signal` as the Overall dotplot
above, not a separate statistic and not log2 fold at fixed thresholds/colours
like this plot used to be). Points off the diagonal favour one direction over
the other. Terms present in **both** directions' dotplot data (i.e.
independently significant in top and in bottom) are highlighted and labelled
— the point of this plot is spotting that recurrence, not re-deriving
significance from scratch.

```{r signal_top_vs_bottom, echo=FALSE, fig.height=7}
sc_all <- if (length(.overall_dotplot_rows) > 0) dplyr::bind_rows(.overall_dotplot_rows) else tibble::tibble()
# OR, not AND: a direction with zero significant/leading-edge-qualifying terms
# must not blank the whole plot when the other direction has plenty of data --
# missing side is coalesced to 0 below (signal_top/signal_bottom), so one-sided
# data still plots meaningfully (all points pinned to that one axis).
if (nrow(sc_all) == 0 || !any(c("top", "bottom") %in% sc_all$ranking)) {
  cat("_No top/bottom signal data to compare (need at least one direction's Overall dotplot data)._\n\n")
} else {
  sc <- sc_all %>%
    dplyr::filter(ranking %in% c("top", "bottom")) %>%
    tidyr::pivot_wider(
      id_cols = c(database, pathway, description),
      names_from = ranking, values_from = signal, names_prefix = "signal_"
    )

  req_cols <- c("signal_top", "signal_bottom")
  for (rc in setdiff(req_cols, names(sc))) sc[[rc]] <- NA_real_

  # Both signal_top and signal_bottom are only non-NA when that (database,
  # pathway) was significant AND scored in that direction's own Overall
  # dotplot -- so "present in both" is exactly "recurs significant in both
  # directions", the same recurrence concept the old fixed-cutoff version
  # tracked, just derived from the pooled signal instead of raw presence.
  sc <- sc %>%
    dplyr::filter(!is.na(signal_top) | !is.na(signal_bottom)) %>%
    dplyr::mutate(
      signal_top    = dplyr::coalesce(signal_top, 0),
      signal_bottom = dplyr::coalesce(signal_bottom, 0),
      recurs_both   = signal_top > 0 & signal_bottom > 0
    )

  if (nrow(sc) == 0) {
    cat("_No plottable points._\n\n")
  } else {
    p <- ggplot(sc, aes(signal_top, signal_bottom, colour = recurs_both)) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
      geom_point(alpha = 0.7, size = 2) +
      scale_colour_manual(values = c(`TRUE` = "#c0392b", `FALSE` = "grey55"),
                           name = "Recurs in both", labels = c(`TRUE` = "yes", `FALSE` = "no")) +
      labs(title = sprintf("Position enrichment — top vs bottom orthogonal signal — %s", params$traitname),
           x = "signal (orthogonal composite, top)",
           y = "signal (orthogonal composite, bottom)") +
      theme_minimal(base_size = 12)

    lab <- sc %>% dplyr::filter(recurs_both)
    if (nrow(lab) > 0) {
      p <- p + ggrepel::geom_text_repel(
        data = lab, aes(label = stringr::str_trunc(description, 45)), colour = "black",
        size = 3.6, fontface = "bold", max.overlaps = Inf, seed = 42,
        box.padding = 0.6, point.padding = 0.4,
        force = 4, force_pull = 0.5,
        max.time = 3, max.iter = 100000,
        min.segment.length = 0, segment.size = 0.3, segment.alpha = 0.6,
        bg.color = "white", bg.r = 0.12, show.legend = FALSE)
    }
    print(p)
  }
}
```

# Leading edge {.tabset .tabset-fade}

For every **significant** (`p.adj < `r PADJ``) pathway, the
observed `gene:position` members of the overlap, plus the same cross-module
corroboration columns as above (`%gate_sig`/`%gate_fdr`/`%fade_*`/`%rer_*`/
`%accum`, direction-aware — `%fade_bottom` only appears under `BOTTOM`,
`%fade_top` only under `TOP`, plain `%fade` under `GLOBAL`). The full member
list is also written to `posenrich_leading_edge.tsv` alongside this report.

```{r leading-edge, echo=FALSE, results='asis'}
if (!has_res || !any(res$sig, na.rm = TRUE)) {
  cat("_No significant sets._\n\n")
} else {
  sig_res <- res %>% filter(sig)
  for (rk in intersect(RANKING_ORDER, unique(sig_res$ranking))) {
    cat(sprintf("\n## %s\n\n", toupper(rk)))
    rk_sig <- sig_res %>% filter(ranking == rk) %>% arrange(database, p_adj)
    if (nrow(rk_sig) == 0) { cat("_No significant sets for this direction._\n\n"); next }

    want <- flag_cols_by_rk[[rk]]
    if (is.null(want)) want <- character(0)
    present_flags <- intersect(want, names(rk_sig))

    if (!is.null(le_summary)) {
      rk_sig <- rk_sig %>%
        dplyr::left_join(le_summary, by = c("ranking", "database", "pathway", "top_frac"))
    }

    le_df <- rk_sig %>%
      transmute(Database = database, Pathway = pathway, Description = description,
                `Top frac` = sprintf("%s%%", format(100 * top_frac, trim = TRUE)),
                Fold = round(fold_enrichment, 3), Direction = direction,
                `p.adj` = signif(p_adj, 3))
    if (length(present_flags) > 0) {
      flag_df <- rk_sig %>% transmute(across(all_of(present_flags), ~round(.x, 1)))
      names(flag_df) <- paste0("%", sub("^pct_", "", present_flags))
      le_df <- dplyr::bind_cols(le_df, flag_df)
    }
    esc_cols <- TRUE
    if ("Genes/positions" %in% names(rk_sig)) {
      le_df <- dplyr::bind_cols(le_df, rk_sig %>% dplyr::select(`N genes`, `Genes/positions`))
      # Full member list is still in posenrich_leading_edge.tsv -- here, show
      # only the first few inline with a <details> toggle for the rest, since
      # a row can otherwise carry hundreds of comma-separated gene:positions.
      le_df[["Genes/positions"]] <- truncate_gene_list(le_df[["Genes/positions"]])
      # DT column indices for `escape` are 1-based over the data columns when
      # rownames = FALSE (no offset needed) -- verified empirically, since
      # DT's own docs are ambiguous on this point and getting it wrong throws
      # "attempt to select less than one element in integerOneIndex" deep
      # inside DT:::escapeData(), aborting the whole render.
      esc_cols <- which(names(le_df) != "Genes/positions")
    }

    id <- paste("dt_le", rk, sep = "_")
    assign(paste0(".esc_cols_", rk), esc_cols, envir = environment())
    cat(knitr::knit_child(text = c(
      paste0("```{r ", id, ", echo=FALSE}"),
      sprintf("DT::datatable(le_df, options = list(pageLength = 10, scrollX = TRUE), rownames = FALSE, escape = .esc_cols_%s)", rk),
      "```"), envir = environment(), quiet = TRUE))
    cat("\n\n")
  }
}
```

# Module summary across gene lists {.tabset .tabset-fade}

Every significant term/pathway, pooled across this direction's four percentile
cutoffs (`top25`/`top10`/`top5`/`top1` under **Top**, the `bottom*` equivalents
under **Bottom**, both sets under **Global**) so a term can be compared without
switching cutoff tabs above — mirrors the AMI report's "Module summary across
gene lists" rollup, with DOMINO modules replaced by posenrich's own enriched
terms (posenrich has no module-finding step of its own). Sorted by best
(smallest) `p.adj`, then by how many cutoffs recovered it. Corroboration
columns (`%gate_sig`/`%gate_fdr`/`%fade_*`/`%rer_*`/`%accum*`) are
direction-matched via the same `flag_cols_by_rk` used in the Leading edge
section above — a `bottom`-direction row never carries `%accum_top`/`%fade_top`.

Underneath each table, a **term x percentile-cutoff** heatmap (ordered
25%->10%->5%->1%, matching this report's own cutoff convention): a term
recurring across several cutoffs as the foreground tightens is a stronger
candidate than one appearing at a single cutoff.

```{r module-rollup-setup, include=FALSE}
# Term-similarity clustering — ported verbatim from 13.AMI_analysis.Rmd's
# Module summary rollup. Different GMT databases can describe the same
# biological concept with slightly different wording (case, punctuation,
# minor rewording -- collapsed via normalized edit-distance, utils::adist()),
# or one description can be a compound/nested restatement of another (e.g.
# "Primary ciliary dyskinesia" inside "Axonemal dynein complex and Primary
# ciliary dyskinesia") -- character-level edit distance alone rejects these
# (the extra words dominate the ratio even though one term is wholly
# contained in the other), so containment is checked explicitly and forces a
# merge regardless of the edit-distance threshold. As in AMI, this only
# collapses the HEATMAP's term axis below -- the recurrence table stays one
# row per (database, pathway), same as AMI's per-module rollup table stays
# one row per DOMINO module.
.normalize_term <- function(x) {
  x <- tolower(x)
  x <- gsub("[^a-z0-9 ]", " ", x)
  gsub("\\s+", " ", trimws(x))
}
.term_contains <- function(a, b) {
  wa <- strsplit(a, " ")[[1]]; wb <- strsplit(b, " ")[[1]]
  if (length(wa) == 0 || length(wb) == 0) return(FALSE)
  short <- if (length(wa) <= length(wb)) wa else wb
  long  <- if (length(wa) <= length(wb)) wb else wa
  all(short %in% long)
}
.cluster_terms <- function(terms, sim_thr = 0.85) {
  norm <- vapply(terms, .normalize_term, character(1))
  reps_norm <- character(0); reps_orig <- character(0)
  assign <- integer(length(terms))
  for (i in seq_along(terms)) {
    if (length(reps_norm) == 0) { reps_norm <- norm[i]; reps_orig <- terms[i]; assign[i] <- 1L; next }
    d <- utils::adist(norm[i], reps_norm)[1, ]
    sim <- 1 - d / pmax(nchar(norm[i]), nchar(reps_norm), 1)
    contained <- vapply(reps_norm, function(r) .term_contains(norm[i], r), logical(1))
    sim[contained] <- 1
    best <- which.max(sim)
    if (sim[best] >= sim_thr) {
      assign[i] <- best
      if (contained[best] && nchar(terms[i]) < nchar(reps_orig[best])) {
        reps_orig[best] <- terms[i]
      }
    } else {
      reps_norm <- c(reps_norm, norm[i]); reps_orig <- c(reps_orig, terms[i])
      assign[i] <- length(reps_norm)
    }
  }
  reps_orig[assign]
}

.render_posenrich_heatmap <- function(rk_sig, cluster_map, title) {
  if (nrow(rk_sig) == 0) { cat(sprintf("_No terms to show (%s)._\n\n", title)); return(invisible(NULL)) }
  pres <- rk_sig %>%
    dplyr::mutate(description = dplyr::coalesce(description, pathway),
                  description = unname(cluster_map[description])) %>%
    dplyr::distinct(description, top_frac) %>%
    dplyr::mutate(cutoff = sprintf("%s%%", format(100 * top_frac, trim = TRUE)))
  cutoff_levels <- sprintf("%s%%", format(sort(unique(100 * rk_sig$top_frac), decreasing = TRUE), trim = TRUE))

  # Term order recomputed post-clustering (most cutoffs recovered first) --
  # a term merged from several pathways can now recur at more cutoffs than
  # any single contributing pathway did on its own.
  term_counts <- pres %>% dplyr::count(description, name = "n_cutoffs") %>%
    dplyr::arrange(dplyr::desc(n_cutoffs), description)
  term_order <- term_counts$description

  pres <- pres %>%
    dplyr::mutate(cutoff = factor(cutoff, levels = cutoff_levels),
                  description = factor(description, levels = rev(term_order)))
  n_terms <- length(term_order)
  p <- ggplot(pres, aes(x = cutoff, y = description)) +
    geom_tile(fill = "#2c7fb8", colour = "white", linewidth = 0.6) +
    scale_x_discrete(drop = FALSE) +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 0),
          axis.text.y = element_text(size = pmax(6, pmin(10, 260 / max(n_terms, 1))))) +
    labs(title = sprintf("Recovered terms by percentile cutoff — %s", title),
         x = "cutoff (% of scored positions)", y = NULL)
  print(p)
  cat("\n\n")

  disp <- pres %>%
    dplyr::mutate(cutoff = as.character(cutoff)) %>%
    dplyr::arrange(description, cutoff) %>%
    dplyr::group_by(description) %>%
    dplyr::summarise(`N cutoffs` = dplyr::n(),
                      `Cutoffs` = paste(cutoff, collapse = ", "),
                      .groups = "drop") %>%
    dplyr::arrange(dplyr::desc(`N cutoffs`), description) %>%
    dplyr::rename(Term = description)
  print(kbl_html(disp, caption = sprintf("Term -> cutoff membership — %s", title)))
  cat("\n\n")
}
```

```{r module-rollup, echo=FALSE, results='asis'}
if (!has_res || !any(res$sig, na.rm = TRUE)) {
  cat("_No significant sets._\n\n")
} else {
  sig_res <- res %>% filter(sig)
  # Direction is the OUTER tab, each with two H3 sub-tabs -- Recurrence Table /
  # Recurrence Heatmap -- mirroring 13.AMI_analysis.Rmd's nested-tab layout
  # (## direction {.tabset} -> ### Modules/Recurrence Table/Recurrence Heatmap).
  # Posenrich has no DOMINO module-finding step of its own, so no "Modules"
  # sub-tab here; the other two are otherwise identical in structure.
  for (rk in intersect(RANKING_ORDER, unique(sig_res$ranking))) {
    cat(sprintf("\n## %s {.tabset}\n\n", toupper(rk)))
    rk_sig <- sig_res %>% filter(ranking == rk)
    if (nrow(rk_sig) == 0) { cat("_No significant sets for this direction._\n\n"); next }

    want <- flag_cols_by_rk[[rk]]
    if (is.null(want)) want <- character(0)
    present_flags <- intersect(want, names(rk_sig))

    n_genes_tbl <- if (!is.null(le_raw)) {
      le_raw %>% dplyr::filter(ranking == rk) %>%
        dplyr::group_by(database, pathway) %>%
        dplyr::summarise(`N genes` = dplyr::n_distinct(gene), .groups = "drop")
    } else NULL

    term_meta <- rk_sig %>%
      dplyr::group_by(database, pathway) %>%
      dplyr::summarise(
        description = dplyr::first(dplyr::coalesce(description, pathway)),
        `p.adj (min)` = signif(min(p_adj, na.rm = TRUE), 3),
        `N cutoffs` = dplyr::n_distinct(top_frac),
        dplyr::across(dplyr::all_of(present_flags), ~round(mean(.x, na.rm = TRUE), 1)),
        .groups = "drop"
      ) %>%
      { if (!is.null(n_genes_tbl)) dplyr::left_join(., n_genes_tbl, by = c("database", "pathway")) else . } %>%
      dplyr::arrange(`p.adj (min)`, dplyr::desc(`N cutoffs`))

    names(term_meta)[names(term_meta) %in% present_flags] <-
      paste0("%", sub("^pct_", "", names(term_meta)[names(term_meta) %in% present_flags]))

    # Cluster map (near-duplicate/nested description text -> canonical label),
    # built once per direction and reused by both the table's display label and
    # the heatmap below, so a pathway shows the same collapsed term text in
    # both places. Only the LABEL is collapsed here -- the table stays one row
    # per (database, pathway); the heatmap (which reads rk_sig directly, not
    # this table) is where cutoff-presence actually gets merged across
    # differently-worded pathways that map to the same cluster.
    desc_levels <- unique(term_meta$description)
    cluster_map <- setNames(.cluster_terms(desc_levels), desc_levels)

    disp <- term_meta %>%
      dplyr::mutate(description = unname(cluster_map[description])) %>%
      dplyr::rename(Database = database, Term = description) %>%
      dplyr::select(-pathway)

    cat("\n\n### Recurrence Table\n\n")
    id <- paste("dt_rollup", rk, sep = "_")
    cat(knitr::knit_child(text = c(
      paste0("```{r ", id, ", echo=FALSE}"),
      "DT::datatable(disp, options = list(pageLength = 10, scrollX = TRUE), rownames = FALSE)",
      "```"), envir = environment(), quiet = TRUE))
    cat("\n\n")

    cat("\n\n### Recurrence Heatmap\n\n")
    .render_posenrich_heatmap(rk_sig, cluster_map, toupper(rk))
  }
}
```

---

```{r poschar-setup, include=FALSE}
.file_ok <- function(p) !is.null(p) && !is.na(p) && nchar(p) > 0 && file.exists(p)

# Tableau_10 colour swatch for pipeline-wide consistency (same palette used
# throughout the Scoring/Characterization reports).
if (requireNamespace("paletteer", quietly = TRUE)) {
  .tableau10 <- as.character(paletteer::paletteer_d("ggthemes::Tableau_10"))
} else {
  .tableau10 <- c(
    "#4E79A7", "#F28E2B", "#E15759", "#76B7B2", "#59A14F",
    "#EDC948", "#B07AA1", "#FF9DA7", "#9C755F", "#BAB0AC"
  )
}

has_position_scores <- .file_ok(params$position_scores_file)
has_gene_scores      <- .file_ok(params$gene_scores_file)
pos_df  <- if (has_position_scores) readr::read_tsv(params$position_scores_file, show_col_types = FALSE) else NULL
gene_df <- if (has_gene_scores)     readr::read_tsv(params$gene_scores_file,     show_col_types = FALSE) else NULL

has_vep_primateai <- .file_ok(params$vep_primateai_file)
has_vep_cosmic    <- .file_ok(params$vep_cosmic_file)
has_genomic       <- .file_ok(params$genomic_info_file)
# has_fade_sites_top/bottom + fade_sites_top_df/bottom_df already loaded in the
# `load` chunk near the top of this file (hoisted so the Overall dotplot's
# position-native FADE-overlap component can use them too).
# Position scores are required for every sub-tab below (PAI3D/COSMIC/FADE joins,
# regional accumulation) -- gate the whole section on them, not just on which
# external database happens to be supplied.
has_pos_char <- has_position_scores &&
  (has_vep_primateai || has_vep_cosmic || has_genomic || has_fade_sites_top || has_fade_sites_bottom)

if (has_vep_primateai) {
  vep_pai_df <- readr::read_tsv(params$vep_primateai_file, show_col_types = FALSE)
  if ("Position" %in% names(vep_pai_df)) vep_pai_df$Position <- suppressWarnings(as.integer(vep_pai_df$Position))
} else {
  vep_pai_df <- NULL
}

if (has_vep_cosmic) {
  vep_cosmic_df <- readr::read_tsv(params$vep_cosmic_file, show_col_types = FALSE)
  if ("Position" %in% names(vep_cosmic_df)) vep_cosmic_df$Position <- suppressWarnings(as.integer(vep_cosmic_df$Position))
} else {
  vep_cosmic_df <- NULL
}

if (has_genomic) {
  # Format: gene, chr, start, end, length (no header assumed; detect if present)
  genomic_raw <- readr::read_tsv(params$genomic_info_file,
                          col_names = c("Gene", "chr", "gene_start", "gene_end", "gene_length"),
                          show_col_types = FALSE)
  if (tolower(genomic_raw$Gene[1]) == "gene") genomic_raw <- genomic_raw[-1, ]
  genomic_raw <- genomic_raw %>%
    dplyr::mutate(
      gene_start  = as.integer(gene_start),
      gene_end    = as.integer(gene_end),
      gene_length = as.integer(gene_length),
      gene_mid    = as.integer((gene_start + gene_end) / 2L)
    )
} else {
  genomic_raw <- NULL
}

# Join scored positions with genomic coordinates via their gene, and lay
# out chromosome-cumulative offsets for the regional-accumulation section
# below -- each position inherits its gene's genomic midpoint as x-coordinate.
if (has_genomic && has_position_scores) {
  pos_genomic <- pos_df %>%
    dplyr::inner_join(genomic_raw, by = "Gene") %>%
    dplyr::mutate(
      chr_label = sub("^chr", "", chr),
      chr_num   = suppressWarnings(as.integer(chr_label)),
      chr_order = dplyr::case_when(
        !is.na(chr_num)              ~ chr_num,
        chr_label == "X"             ~ 23L,
        chr_label == "Y"             ~ 24L,
        chr_label %in% c("MT", "M")  ~ 25L,
        TRUE                         ~ 26L
      )
    ) %>%
    dplyr::arrange(chr_order, gene_mid)

  pos_chr_offsets <- pos_genomic %>%
    dplyr::group_by(chr_order, chr_label) %>%
    dplyr::summarise(chr_max = max(gene_end, na.rm = TRUE), .groups = "drop") %>%
    dplyr::arrange(chr_order) %>%
    dplyr::mutate(offset = cumsum(dplyr::lag(chr_max, default = 0L)) + (chr_order - 1L) * 5e6L)

  pos_genomic <- pos_genomic %>%
    dplyr::left_join(pos_chr_offsets %>% dplyr::select(chr_order, offset), by = "chr_order") %>%
    dplyr::mutate(
      manhattan_x = gene_mid + offset,
      chr_even    = chr_order %% 2 == 0
    )
} else {
  pos_genomic <- NULL
  pos_chr_offsets <- NULL
}

if (has_pos_char) {
  top_pct  <- params$top_pct %||% 0.10
  top5_pct <- 0.05
  top1_pct <- 0.01

  pos_top    <- pos_df %>% dplyr::filter(change_side %in% c("top",    "both"))
  pos_bottom <- pos_df %>% dplyr::filter(change_side %in% c("bottom", "both"))

  # Percentile membership: prefer SCORING's published position_lists/
  # slice_<dir><pct>.tsv (scoring_compute.R) -- the same canonical top/bottom/
  # global x 10/5/1% cutoff 15.Comparison_report.Rmd and posenrich both read
  # -- over an independently re-derived quantile() cut. Each .pos_pct_test()
  # call returns a CLOSURE (not a bare vector) so it can be applied to
  # whichever population later needs it (pos_score_meta, pos_genomic_top/bot
  # are different row subsets of pos_df, so the SAME membership set/threshold
  # must be evaluated against each one individually, not computed once and
  # reused positionally). Falls back to quantile() on `pop` (the direction's
  # own population) when position_lists_dir is NULL or a slice file is
  # missing -- all 3 files for a direction, never a mix.
  .pos_slice_ids <- function(dir_key, pct) {
    if (is.null(params$position_lists_dir)) return(NULL)
    f <- file.path(params$position_lists_dir, sprintf("slice_%s%d.tsv", dir_key, pct))
    if (!file.exists(f)) return(NULL)
    d <- tryCatch(readr::read_tsv(f, show_col_types = FALSE), error = function(e) NULL)
    if (is.null(d) || !all(c("Gene", "Position") %in% names(d))) return(NULL)
    unique(paste(d$Gene, d$Position))
  }
  .pos_pct_test <- function(dir_key, pct_frac, pop) {
    ids_set <- .pos_slice_ids(dir_key, round(pct_frac * 100))
    if (!is.null(ids_set)) {
      function(df) paste(df$Gene, df$Position) %in% ids_set
    } else {
      thr <- quantile(pop$CAAS_score, 1 - pct_frac, na.rm = TRUE)
      function(df) df$CAAS_score >= thr
    }
  }

  # top25_pct: only used by the PAI3D/COSMIC/FADE "characterization by CAAS
  # threshold" Fisher-OR sweep tables below, replacing their old ad hoc 50%
  # row (quantile(0.50), no SCORING equivalent) with a 25% row that DOES have
  # one (position_lists/slice_<dir>25.tsv) -- so every percentage row in
  # those tables except the trivial 100% baseline now comes from the same
  # published source. Not used for in_top_Xpct (only ever a 10/5/1% trio
  # elsewhere in this report).
  top25_pct <- 0.25

  test_glob_25 <- .pos_pct_test("global", top25_pct, pos_df)
  test_glob_10 <- .pos_pct_test("global", top_pct,  pos_df)
  test_glob_5  <- .pos_pct_test("global", top5_pct, pos_df)
  test_glob_1  <- .pos_pct_test("global", top1_pct, pos_df)
  test_top_25  <- .pos_pct_test("top",    top25_pct, pos_top)
  test_top_10  <- .pos_pct_test("top",    top_pct,  pos_top)
  test_top_5   <- .pos_pct_test("top",    top5_pct, pos_top)
  test_top_1   <- .pos_pct_test("top",    top1_pct, pos_top)
  test_bot_25  <- .pos_pct_test("bottom", top25_pct, pos_bottom)
  test_bot_10  <- .pos_pct_test("bottom", top_pct,  pos_bottom)
  test_bot_5   <- .pos_pct_test("bottom", top5_pct, pos_bottom)
  test_bot_1   <- .pos_pct_test("bottom", top1_pct, pos_bottom)

  # thr_10: cosmetic-only scalar for the geom_vline reference line further
  # below (a plot annotation has no equivalent to a discrete membership set) --
  # always a plain quantile, independent of whether the tests above used the
  # published path.
  thr_10 <- quantile(pos_df$CAAS_score, 1 - top_pct, na.rm = TRUE)
}
```
---

```{r pos-char-header, results='asis'}
if (has_pos_char) cat(r"(# Position Characterisation {.tabset #position-characterisation}

This section examines how ranked positions behave in relation to external
signals that do **not** contribute to the `CAAS_score`. PrimateAI-3D, COSMIC
Mutant Census, and FADE branch-site selection evidence are treated as
independent characterisation layers.

PrimateAI-3D and COSMIC both describe specific amino acid transitions, so the
biochemical Grantham Score (GS) groupings are out of scope for those two (as
they group transitions by biochemical radicalness rather than modeling the
specific substitution directly). We restrict those two layers strictly to the
**Unweighted Scheme (US)**, with all scores and mutation overlaps derived
solely from US transitions. FADE evidence is not transition-specific (it is a
per-site test of dN/dS shift along phenotype-associated branches) and so is
not restricted to the US scheme.

- **`pai_char_score`**: The pathogenicity score of mutating the ancestral human reference amino acid to the derived US alternative amino acid(s) in PrimateAI-3D.
- **COSMIC Overlap**: The presence and recurrence of somatic mutations mapped to the specific codon under the US transition.
- **FADE Overlap**: Whether the position falls within the direction-specific FADE_top_sig/FADE_bottom_sig significant site set (BF ≥ `fade_bf_threshold`), and its maximum Bayes Factor.

)")
```

```{r vep-setup-common}
# Computed against pos_df directly (test_glob_*() need Gene/Position/CAAS_score,
# all present there) -- select() below only drops columns, never reorders/
# filters rows, so assigning these vectors positionally into pos_score_meta
# stays row-aligned with pos_df.
.glob10 <- test_glob_10(pos_df); .glob5 <- test_glob_5(pos_df); .glob1 <- test_glob_1(pos_df)

pos_score_meta <- pos_df %>%
  select(Gene, Position, CAAS_score, biochem_weight_sum, change_side, gate_sig, gate_fdr) %>%
  mutate(
    caas_decile  = dplyr::ntile(CAAS_score, 10),
    in_top_10pct = .glob10,
    in_top_5pct  = .glob5,
    in_top_1pct  = .glob1
  )

has_background <- .file_ok(params$background_file)
if (has_background) {
  bg_lines <- readLines(params$background_file)
  bg_genes <- bg_lines[nzchar(trimws(bg_lines)) & !grepl("^#", bg_lines)]
} else {
  bg_genes <- unique(gene_df$Gene)
}
```

```{r vep-pai-subsec, results='asis'}
if (has_vep_primateai) cat("## VEP: PrimateAI-3D\n\n")
```

```{r vep-pai-setup, eval=has_vep_primateai}
# Detect PrimateAI score column (varies by database version)
.pai_score_col <- intersect(
  c("score_PAI3D", "score", "primateai_score", "pathogenicity_score", "PrimateAI_score"),
  names(vep_pai_df)
)[1]
if (is.na(.pai_score_col)) {
  warning("PrimateAI: no recognised score column found — skipping.")
  has_vep_primateai <- FALSE
}
```

```{r vep-pai-join, eval=has_vep_primateai}
# Restrict to shared genes present in both our results (background clean) and database
shared_genes_pai <- intersect(bg_genes, unique(vep_pai_df$Gene))
pos_pai_shared <- pos_score_meta %>% filter(Gene %in% shared_genes_pai)

# Under US-only design, pai_char_score is simply the pathogenicity score of the US change.
# If multiple alternative AAs map, we keep the maximum score.
# Pathogenic status is defined exclusively by the prediction column (containing
# "pathogenic", case-insensitive) — never by a raw score cutoff.
pai_by_prot <- vep_pai_df %>%
  filter(caap_group == "US", Gene %in% shared_genes_pai) %>%
  mutate(
    pai_score = as.numeric(.data[[.pai_score_col]]),
    is_pathogenic = grepl("pathogenic", prediction, ignore.case = TRUE)
  ) %>%
  filter(is.finite(pai_score)) %>%
  group_by(Gene, Position) %>%
  summarise(
    pai_char_score = max(pai_score, na.rm = TRUE),
    pai_pathogenic = any(is_pathogenic, na.rm = TRUE),
    n_pai_schemes  = 1L,
    .groups = "drop"
  )

pos_vep_pai <- pai_by_prot %>%
  inner_join(pos_pai_shared, by = c("Gene", "Position")) %>%
  distinct(Gene, Position, .keep_all = TRUE)

# Join back caas_change and other details for display table
pos_vep_pai_table_source <- vep_pai_df %>%
  filter(caap_group == "US", Gene %in% shared_genes_pai) %>%
  inner_join(pos_pai_shared, by = c("Gene", "Position")) %>%
  distinct(Gene, Position, .keep_all = TRUE)

has_vep_primateai <- nrow(pos_vep_pai) > 0

# Directional PAI subsets: top+both and bottom+both
pos_vep_pai_top    <- pos_vep_pai %>% filter(change_side %in% c("top",    "both"))
pos_vep_pai_bottom <- pos_vep_pai %>% filter(change_side %in% c("bottom", "both"))
```

`r if (exists("pos_vep_pai") && nrow(pos_vep_pai) == 0) "> **Note:** PrimateAI-3D inputs were provided, but no scored positions could be matched to the VEP/PAI3D annotations for this direction."`

```{r vep-pai-violin, eval=has_vep_primateai, fig.height=7}
# pai_char_score violin per CAAS decile, split by direction (TOP and BOTTOM separately).
pai_dir_df <- bind_rows(
  pos_vep_pai_top    %>% mutate(pai_direction = "TOP (top + both)"),
  pos_vep_pai_bottom %>% mutate(pai_direction = "BOTTOM (bottom + both)")
) %>%
  filter(!is.na(caas_decile), !is.na(pai_char_score))

# Per-direction median for horizontal reference lines
pai_medians <- pai_dir_df %>%
  group_by(pai_direction) %>%
  summarise(median_pai = median(pai_char_score, na.rm = TRUE), .groups = "drop")

ggplot(pai_dir_df, aes(x = factor(caas_decile, levels = 1:10), y = pai_char_score)) +
  geom_violin(aes(fill = pai_direction), alpha = 0.55, trim = TRUE, scale = "width",
              show.legend = FALSE) +
  geom_boxplot(width = 0.14, outlier.size = 0.4, alpha = 0.9,
               fill = "white", colour = "grey30") +
  geom_hline(data = pai_medians,
             aes(yintercept = median_pai),
             linetype = "dashed", colour = "grey20", linewidth = 0.7) +
  geom_text(data = pai_medians,
            aes(x = 10.6, y = median_pai,
                label = sprintf("median = %.3f", median_pai)),
            hjust = 0, vjust = -0.4, size = 3, colour = "grey20",
            inherit.aes = FALSE) +
  scale_fill_manual(values = c("TOP (top + both)" = .tableau10[1],
                               "BOTTOM (bottom + both)" = .tableau10[3])) +
  scale_x_discrete(
    drop   = FALSE,
    labels = setNames(c("1\n(low)", as.character(2:9), "10\n(high)"), as.character(1:10))
  ) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.08))) +
  facet_wrap(~pai_direction, ncol = 1) +
  labs(
    title    = "PAI_TOP and PAI_BOTTOM: PrimateAI-3D Score by CAAS Decile",
    subtitle = sprintf(
      "PAI_TOP: %d positions | PAI_BOTTOM: %d positions | dashed = per-direction median",
      nrow(pos_vep_pai_top), nrow(pos_vep_pai_bottom)
    ),
    x = "CAAS Score Decile", y = "PrimateAI-3D Score"
  ) +
  theme(plot.title = element_text(face = "bold"),
        strip.text = element_text(face = "bold", size = 11))
```

```{r vep-pai-scatter, eval=has_vep_primateai, fig.height=5}
# CAAS_score vs pai_char_score, coloured by direction.
# Normalise both axes for Pearson r (rank / n).
pos_vep_pai_dir <- bind_rows(
  pos_vep_pai_top    %>% mutate(pai_direction = "TOP"),
  pos_vep_pai_bottom %>% mutate(pai_direction = "BOTTOM")
) %>% distinct(Gene, Position, .keep_all = TRUE)

.pai_caas_r_top  <- safe_cor(
  rank(pos_vep_pai_top$CAAS_score,    na.last = "keep") / sum(!is.na(pos_vep_pai_top$CAAS_score)),
  rank(pos_vep_pai_top$pai_char_score, na.last = "keep") / sum(!is.na(pos_vep_pai_top$pai_char_score))
)
.pai_caas_r_bot  <- safe_cor(
  rank(pos_vep_pai_bottom$CAAS_score,    na.last = "keep") / sum(!is.na(pos_vep_pai_bottom$CAAS_score)),
  rank(pos_vep_pai_bottom$pai_char_score, na.last = "keep") / sum(!is.na(pos_vep_pai_bottom$pai_char_score))
)

ggplot(pos_vep_pai_dir, aes(x = CAAS_score, y = pai_char_score,
                             colour = pai_direction)) +
  geom_point(alpha = 0.35, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8, alpha = 0.15) +
  geom_vline(xintercept = thr_10, linetype = "dashed",
             colour = "grey40", linewidth = 0.5) +
  scale_colour_manual(values = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[3]),
                      name = "Direction") +
  labs(
    title    = "CAAS Score vs PrimateAI-3D Score by Direction",
    subtitle = sprintf("Pearson r (rank-norm): TOP = %.3f | BOTTOM = %.3f | dashed = top-10%% CAAS",
                       .pai_caas_r_top, .pai_caas_r_bot),
    x = "CAAS Score", y = "PrimateAI-3D Score"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r vep-pai-pathogenic-bar, eval=has_vep_primateai, fig.height=5}
# Count of pathogenic vs non-pathogenic positions, TOP and BOTTOM.
pai_pathogenic_counts <- bind_rows(
  pos_vep_pai_top    %>% mutate(pai_direction = "TOP"),
  pos_vep_pai_bottom %>% mutate(pai_direction = "BOTTOM")
) %>%
  mutate(pathogenic = ifelse(pai_pathogenic, "pathogenic", "non-pathogenic")) %>%
  count(pai_direction, pathogenic)

ggplot(pai_pathogenic_counts, aes(x = pathogenic, y = n, fill = pai_direction)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = n), position = position_dodge(width = 0.8), vjust = -0.3, size = 3.5) +
  scale_fill_manual(values = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[3])) +
  labs(
    title    = "Pathogenic vs Non-Pathogenic Positions (PrimateAI-3D)",
    subtitle = "Pathogenic defined by prediction column",
    x = NULL, y = "N positions", fill = "Direction"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r vep-pai-phenaxis-scatter, eval=has_vep_primateai && "phen_score" %in% names(pos_vep_pai_top), fig.height=5}
.pai_phen_r_top <- safe_cor(
  rank(pos_vep_pai_top$phen_score,      na.last = "keep") / sum(!is.na(pos_vep_pai_top$phen_score)),
  rank(pos_vep_pai_top$pai_char_score,  na.last = "keep") / sum(!is.na(pos_vep_pai_top$pai_char_score))
)
.pai_phen_r_bot <- safe_cor(
  rank(pos_vep_pai_bottom$phen_score,      na.last = "keep") / sum(!is.na(pos_vep_pai_bottom$phen_score)),
  rank(pos_vep_pai_bottom$pai_char_score,  na.last = "keep") / sum(!is.na(pos_vep_pai_bottom$pai_char_score))
)

ggplot(pos_vep_pai_dir, aes(x = phen_score, y = pai_char_score,
                             colour = pai_direction)) +
  geom_point(alpha = 0.35, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8, alpha = 0.15) +
  scale_colour_manual(values = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[3]),
                      name = "Direction") +
  labs(
    title    = "Phenotype Evolution Score vs PrimateAI-3D Score by Direction",
    subtitle = sprintf("Pearson r (rank-norm): TOP = %.3f | BOTTOM = %.3f",
                       .pai_phen_r_top, .pai_phen_r_bot),
    x = "phen_score (1 - pvalue_boot)", y = "PrimateAI-3D Score"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r vep-pai-table, eval=has_vep_primateai}
pai_table <- pos_vep_pai_table_source %>%
  arrange(desc(CAAS_score)) %>%
  select(Gene, Position, hg38_ref_aa, caas_alt_aas, caas_change, CAAS_score, change_side, score_PAI3D, percentile_PAI3D, prediction) %>%
  rename(
    `Ref AA` = hg38_ref_aa,
    `Alt AA(s)` = caas_alt_aas,
    `CAAS Change` = caas_change,
    `CAAS Score` = CAAS_score,
    `Change Side` = change_side,
    `PAI3D Score` = score_PAI3D,
    `PAI3D Percentile` = percentile_PAI3D,
    `Prediction` = prediction
  )

DT::datatable(
  pai_table,
  caption = "CAAS positions overlapping with PrimateAI-3D pathogenicity predictions, ordered by CAAS score",
  options = list(pageLength = 10, scrollX = TRUE),
  rownames = FALSE
)
```

```{r vep-pai-fisher, eval=has_vep_primateai}
# Odds ratio & P-value interpretation:
# - Odds Ratio: Measures the enrichment of pathogenic residues (by the prediction column) within the specified CAAS threshold compared to the rest of the positions. An odds ratio > 1 indicates enrichment.
# - P-value: Reaches significance (via Fisher's exact test) when the proportion of pathogenic mutations is significantly higher inside the CAAS threshold.
# `N positions` IS the Fisher test's actual population (PAI-scored, shared-gene
# positions) — mirrors run_cosmic_fisher_dir below. Previously this column showed
# a broader shared-gene pool count (via a separate coverage_df) that did not match
# what the test was actually computed over; fixed to report the true population.
run_pai_fisher_dir <- function(df, label) {
  df   <- df %>% mutate(high_pai = !is.na(pai_pathogenic) & pai_pathogenic)
  n_all <- nrow(df)

  # Build named threshold list: CAAS-flag-based thresholds + percentage-based
  # rows. 100% is trivial (the full population, no computation needed); every
  # other row -- 25/10/5/1% -- comes from SCORING's published position_lists
  # slices (test_glob_25/in_top_Xpct), not an independently re-derived
  # quantile() cut. 25% replaces the old ad hoc 50% row (quantile(0.50) had no
  # published equivalent; SCORING only ever publishes 25/10/5/1%).
  pct_rows <- list(
    list(col = NULL, lbl = "100% (all positions)",    flag = rep(TRUE, n_all)),
    list(col = NULL, lbl = "25% (top 25% CAAS)",      flag = test_glob_25(df)),
    list(col = "in_top_10pct", lbl = sprintf("Top %g%%", top_pct  * 100), flag = NULL),
    list(col = "in_top_5pct",  lbl = sprintf("Top %g%%", top5_pct * 100), flag = NULL),
    list(col = "in_top_1pct",  lbl = sprintf("Top %g%%", top1_pct * 100), flag = NULL)
  )

  purrr::map_dfr(pct_rows, function(thr) {
    in_top   <- if (!is.null(thr$flag)) thr$flag else df[[thr$col]]
    in_top   <- replace_na(in_top, FALSE)

    n_top    <- sum(in_top)
    n_top_h  <- sum(in_top & df$high_pai, na.rm = TRUE)
    n_rest   <- sum(!in_top)
    n_rest_h <- sum(!in_top & df$high_pai, na.rm = TRUE)
    # Column order is [flag, not-flag] (not [not-flag, flag]) so fisher.test's
    # odds ratio reads as "odds of pathogenic in-threshold vs out-of-threshold"
    # the right way round -- verified against an unambiguous 90%-vs-10%
    # synthetic case (OR=77.3, not the inverted 0.013 the other order gives).
    mat <- matrix(c(n_top_h, n_rest_h, n_top - n_top_h, n_rest - n_rest_h), nrow = 2)
    ft  <- tryCatch(fisher.test(mat),
                    error = function(e) list(estimate = NA, conf.int = c(NA, NA), p.value = NA))

    pct_sig <- if ("gate_sig" %in% names(df)) sprintf("%.1f%%", 100 * mean(df$gate_sig[in_top], na.rm = TRUE)) else "0.0%"
    pct_fdr <- if ("gate_fdr" %in% names(df)) sprintf("%.1f%%", 100 * mean(df$gate_fdr[in_top], na.rm = TRUE)) else "0.0%"

    tibble(
      Direction            = label,
      Threshold            = thr$lbl,
      `N positions`        = n_top,
      `N pathogenic`       = n_top_h,
      `% pathogenic`       = sprintf("%.1f%%", 100 * n_top_h / max(n_top, 1)),
      `Odds ratio`         = sprintf("%.2f", ft$estimate),
      `P value`            = formatC(ft$p.value, format = "e", digits = 2)
    )
  })
}

pai_fisher <- bind_rows(
  run_pai_fisher_dir(pos_vep_pai_top,    "TOP"),
  run_pai_fisher_dir(pos_vep_pai_bottom, "BOTTOM")
)

knitr::kable(
  pai_fisher,
  caption = "PAI3D enrichment of pathogenic positions (defined by prediction column) by CAAS threshold and direction. N positions = the actual test population (shared-gene, PAI3D-scored positions) at that threshold. 100% row = all directional positions (baseline); 25/10/5/1% rows use SCORING's published position percentile slices; Fisher OR and P value compare top-subset vs remainder, entirely within PAI-covered positions.",
  align = c("l", "l", rep("r", ncol(pai_fisher) - 2))
)
```


```{r vep-cosmic-subsec, results='asis'}
if (has_vep_cosmic) cat("## VEP: COSMIC Mutant Census\n\n")
```

```{r vep-cosmic-setup, eval=has_vep_cosmic}
# Restrict to shared genes present in both our results (background clean) and database
shared_genes_cosmic <- intersect(bg_genes, unique(vep_cosmic_df$Gene))
pos_cosmic_shared <- pos_score_meta %>% filter(Gene %in% shared_genes_cosmic)

# Filter cosmic annotations for US only if caap_group is present
if ("caap_group" %in% names(vep_cosmic_df)) {
  vep_cosmic_df <- vep_cosmic_df %>% filter(caap_group == "US", Gene %in% shared_genes_cosmic)
} else {
  vep_cosmic_df <- vep_cosmic_df %>% filter(Gene %in% shared_genes_cosmic)
}

# Join cosmic annotations with pos_cosmic_shared (keeping all rows to count recurrence)
cosmic_joined <- vep_cosmic_df %>%
  inner_join(pos_cosmic_shared, by = c("Gene", "Position"))

has_vep_cosmic <- nrow(cosmic_joined) > 0
```

`r if (exists("cosmic_joined") && nrow(cosmic_joined) == 0) "> **Note:** COSMIC Mutant Census mutations were mapped, but no scored positions could be matched to the COSMIC annotations for this direction."`

```{r vep-cosmic-stats, eval=has_vep_cosmic}
cosmic_table <- cosmic_joined %>%
  group_by(Gene, Position, hg38_ref_aa, caas_alt_aas, caas_change, CAAS_score, change_side) %>%
  summarise(
    Recurrence = n(),
    `Mutation AA(s)` = paste(unique(MUTATION_AA), collapse = "; "),
    `Description(s)` = paste(unique(MUTATION_DESCRIPTION), collapse = "; "),
    `Somatic Status(es)` = paste(unique(MUTATION_SOMATIC_STATUS), collapse = "; "),
    .groups = "drop"
  ) %>%
  arrange(desc(Recurrence)) %>%
  rename(
    `Ref AA` = hg38_ref_aa,
    `Alt AA(s)` = caas_alt_aas,
    `CAAS Change` = caas_change,
    `CAAS Score` = CAAS_score,
    `Change Side` = change_side,
    `Mutation AA(s)` = `Mutation AA(s)`,
    `Description` = `Description(s)`,
    `Somatic Status` = `Somatic Status(es)`
  )

DT::datatable(
  cosmic_table,
  caption = "CAAS positions overlapping with COSMIC cancer-related mutations, ordered by mutation recurrence",
  options = list(pageLength = 10, scrollX = TRUE),
  rownames = FALSE
)
```

```{r vep-cosmic-somatic-bar, eval=has_vep_cosmic, fig.height=5}
# Count of COSMIC mutations by Somatic Status and direction.
cosmic_counts <- bind_rows(
  cosmic_joined %>% filter(change_side %in% c("top",    "both")) %>% mutate(cosmic_direction = "TOP"),
  cosmic_joined %>% filter(change_side %in% c("bottom", "both")) %>% mutate(cosmic_direction = "BOTTOM")
) %>%
  filter(!is.na(MUTATION_SOMATIC_STATUS) & MUTATION_SOMATIC_STATUS != "") %>%
  count(cosmic_direction, MUTATION_SOMATIC_STATUS)

ggplot(cosmic_counts, aes(x = MUTATION_SOMATIC_STATUS, y = n, fill = cosmic_direction)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = n), position = position_dodge(width = 0.8), vjust = -0.3, size = 3.5) +
  scale_fill_manual(values = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[3])) +
  labs(
    title    = "COSMIC Mutation Somatic Status Distribution",
    x = NULL, y = "N mutations", fill = "Direction"
  ) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r vep-cosmic-fisher, eval=has_vep_cosmic}
# Odds ratio & P-value interpretation:
# - Odds Ratio: Measures the enrichment of positions with somatic mutations reported in the COSMIC database within the specified CAAS threshold compared to the rest of the positions. An odds ratio > 1 indicates enrichment.
# - P-value: Reaches significance (via Fisher's exact test) when the proportion of cancer somatic mutations is significantly higher inside the CAAS threshold.
pos_cosmic <- pos_cosmic_shared %>%
  mutate(has_cosmic = paste(Gene, Position, sep = "_") %in% paste(vep_cosmic_df$Gene, vep_cosmic_df$Position, sep = "_"))

pos_cosmic_top    <- pos_cosmic %>% filter(change_side %in% c("top",    "both"))
pos_cosmic_bottom <- pos_cosmic %>% filter(change_side %in% c("bottom", "both"))

run_cosmic_fisher_dir <- function(df, label) {
  n_all <- nrow(df)
  pct_rows <- list(
    list(col = NULL, lbl = "100% (all positions)",    flag = rep(TRUE, n_all)),
    list(col = NULL, lbl = "25% (top 25% CAAS)",      flag = test_glob_25(df)),
    list(col = "in_top_10pct", lbl = sprintf("Top %g%%", top_pct  * 100), flag = NULL),
    list(col = "in_top_5pct",  lbl = sprintf("Top %g%%", top5_pct * 100), flag = NULL),
    list(col = "in_top_1pct",  lbl = sprintf("Top %g%%", top1_pct * 100), flag = NULL)
  )

  purrr::map_dfr(pct_rows, function(thr) {
    in_top   <- if (!is.null(thr$flag)) thr$flag else df[[thr$col]]
    in_top   <- replace_na(in_top, FALSE)
    n_top    <- sum(in_top)
    n_top_c  <- sum(in_top & df$has_cosmic, na.rm = TRUE)
    n_rest   <- sum(!in_top)
    n_rest_c <- sum(!in_top & df$has_cosmic, na.rm = TRUE)
    # Column order is [flag, not-flag] (not [not-flag, flag]) so fisher.test's
    # odds ratio reads as "odds of COSMIC overlap in-threshold vs out-of-
    # threshold" the right way round -- verified against an unambiguous
    # 90%-vs-10% synthetic case (OR=77.3, not the inverted 0.013 the other
    # order gives).
    mat <- matrix(c(n_top_c, n_rest_c, n_top - n_top_c, n_rest - n_rest_c), nrow = 2)
    ft  <- tryCatch(fisher.test(mat),
                    error = function(e) list(estimate = NA, conf.int = c(NA, NA), p.value = NA))
    
    pct_sig <- if ("gate_sig" %in% names(df)) sprintf("%.1f%%", 100 * mean(df$gate_sig[in_top], na.rm = TRUE)) else "0.0%"
    pct_fdr <- if ("gate_fdr" %in% names(df)) sprintf("%.1f%%", 100 * mean(df$gate_fdr[in_top], na.rm = TRUE)) else "0.0%"

    tibble(
      Direction            = label,
      Threshold            = thr$lbl,
      `N positions`        = n_top,
      `N COSMIC`           = n_top_c,
      `% COSMIC`           = sprintf("%.1f%%", 100 * n_top_c / max(n_top, 1)),
      `% gate_sig`         = pct_sig,
      `% gate_fdr`         = pct_fdr,
      `Odds ratio`         = sprintf("%.2f", ft$estimate),
      `P value`            = formatC(ft$p.value, format = "e", digits = 2)
    )
  })
}

cosmic_fisher <- bind_rows(
  run_cosmic_fisher_dir(pos_cosmic_top,    "TOP"),
  run_cosmic_fisher_dir(pos_cosmic_bottom, "BOTTOM")
)

knitr::kable(
  cosmic_fisher,
  caption = "Fisher exact test for COSMIC somatic mutations overlap enrichment within CAAS score thresholds",
  align = c("l", "l", rep("r", ncol(cosmic_fisher) - 2))
)
```


```{r fade-subsec, results='asis'}
if (has_fade_sites_top || has_fade_sites_bottom) cat("## FADE Site Overlap\n\n")
```

```{r fade-setup, eval=has_fade_sites_top || has_fade_sites_bottom}
# Direction-specific by construction: TOP positions are only ever checked
# against fade_sites_top, BOTTOM positions only against fade_sites_bottom --
# never cross-checked (mirrors the FADE_top_sig/FADE_bottom_sig GMT layers in
# build_position_gmt.py). Unlike COSMIC/PAI3D, FADE ran on the full analysed
# gene set (not a partial external database), so no shared-gene / coverage
# restriction is needed here -- the universe is simply the direction-specific
# CAAS position set itself (pos_score_meta, already carries in_top_Xpct flags).
pos_fade_top <- if (has_fade_sites_top) {
  pos_score_meta %>%
    dplyr::filter(change_side %in% c("top", "both")) %>%
    dplyr::left_join(
      fade_sites_top_df %>% dplyr::select(Gene, Position, max_bf, target_aa),
      by = c("Gene", "Position")
    ) %>%
    dplyr::mutate(has_fade = !is.na(max_bf))
} else NULL

pos_fade_bottom <- if (has_fade_sites_bottom) {
  pos_score_meta %>%
    dplyr::filter(change_side %in% c("bottom", "both")) %>%
    dplyr::left_join(
      fade_sites_bottom_df %>% dplyr::select(Gene, Position, max_bf, target_aa),
      by = c("Gene", "Position")
    ) %>%
    dplyr::mutate(has_fade = !is.na(max_bf))
} else NULL
```

`r if ((has_fade_sites_top && sum(pos_fade_top$has_fade) == 0) || (has_fade_sites_bottom && sum(pos_fade_bottom$has_fade) == 0)) "> **Note:** FADE site evidence was provided, but no scored positions in at least one direction overlapped a FADE-significant site."`

```{r fade-table, eval=has_fade_sites_top || has_fade_sites_bottom}
fade_table <- dplyr::bind_rows(
  if (!is.null(pos_fade_top))    pos_fade_top    %>% dplyr::filter(has_fade) %>% dplyr::mutate(Direction = "TOP"),
  if (!is.null(pos_fade_bottom)) pos_fade_bottom %>% dplyr::filter(has_fade) %>% dplyr::mutate(Direction = "BOTTOM")
) %>%
  dplyr::arrange(dplyr::desc(max_bf)) %>%
  dplyr::select(Gene, Position, Direction, CAAS_score, change_side, target_aa, max_bf) %>%
  dplyr::rename(
    `CAAS Score`      = CAAS_score,
    `Change Side`     = change_side,
    `FADE Target AA`  = target_aa,
    `Max Bayes Factor` = max_bf
  )

DT::datatable(
  fade_table,
  caption = "CAAS positions overlapping with FADE-significant sites (BF >= fade_bf_threshold), ordered by Bayes Factor",
  options = list(pageLength = 10, scrollX = TRUE),
  rownames = FALSE
)
```

```{r fade-bf-violin, eval=has_fade_sites_top || has_fade_sites_bottom, fig.height=5}
fade_bf_df <- dplyr::bind_rows(
  if (!is.null(pos_fade_top))    pos_fade_top    %>% dplyr::filter(has_fade) %>% dplyr::mutate(fade_direction = "TOP"),
  if (!is.null(pos_fade_bottom)) pos_fade_bottom %>% dplyr::filter(has_fade) %>% dplyr::mutate(fade_direction = "BOTTOM")
)

ggplot(fade_bf_df, aes(x = fade_direction, y = max_bf, fill = fade_direction)) +
  geom_violin(alpha = 0.55, trim = TRUE, scale = "width", show.legend = FALSE) +
  geom_boxplot(width = 0.14, outlier.size = 0.4, alpha = 0.9, fill = "white", colour = "grey30") +
  scale_fill_manual(values = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[3])) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.08))) +
  labs(
    title    = "Bayes Factor of FADE-significant Positions by Direction",
    subtitle = sprintf("TOP: %d sites | BOTTOM: %d sites",
                       sum(fade_bf_df$fade_direction == "TOP"), sum(fade_bf_df$fade_direction == "BOTTOM")),
    x = NULL, y = "Max Bayes Factor"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r fade-fisher, eval=has_fade_sites_top || has_fade_sites_bottom}
# Odds ratio & P-value interpretation:
# - Odds Ratio: Measures the enrichment of FADE-significant sites (BF >= fade_bf_threshold)
#   within the specified CAAS threshold compared to the rest of the direction's positions.
#   An odds ratio > 1 indicates enrichment.
# - P-value: Reaches significance (via Fisher's exact test) when the proportion of
#   FADE-significant sites is significantly higher inside the CAAS threshold.
run_fade_fisher_dir <- function(df, label) {
  n_all <- nrow(df)
  pct_rows <- list(
    list(col = NULL, lbl = "100% (all positions)",    flag = rep(TRUE, n_all)),
    list(col = NULL, lbl = "25% (top 25% CAAS)",      flag = test_glob_25(df)),
    list(col = "in_top_10pct", lbl = sprintf("Top %g%%", top_pct  * 100), flag = NULL),
    list(col = "in_top_5pct",  lbl = sprintf("Top %g%%", top5_pct * 100), flag = NULL),
    list(col = "in_top_1pct",  lbl = sprintf("Top %g%%", top1_pct * 100), flag = NULL)
  )

  purrr::map_dfr(pct_rows, function(thr) {
    in_top   <- if (!is.null(thr$flag)) thr$flag else df[[thr$col]]
    in_top   <- replace_na(in_top, FALSE)
    n_top    <- sum(in_top)
    n_top_f  <- sum(in_top & df$has_fade, na.rm = TRUE)
    n_rest   <- sum(!in_top)
    n_rest_f <- sum(!in_top & df$has_fade, na.rm = TRUE)
    # Column order is [flag, not-flag] (not [not-flag, flag] as in the sibling
    # PAI3D/COSMIC Fisher helpers above) so fisher.test's odds ratio reads as
    # "odds of FADE-sig in-threshold vs out-of-threshold" the right way round --
    # verified against an unambiguous 90%-vs-10% synthetic case (OR=77.3, not 0.013).
    mat <- matrix(c(n_top_f, n_rest_f, n_top - n_top_f, n_rest - n_rest_f), nrow = 2)
    ft  <- tryCatch(fisher.test(mat),
                    error = function(e) list(estimate = NA, conf.int = c(NA, NA), p.value = NA))

    pct_sig <- if ("gate_sig" %in% names(df)) sprintf("%.1f%%", 100 * mean(df$gate_sig[in_top], na.rm = TRUE)) else "0.0%"
    pct_fdr <- if ("gate_fdr" %in% names(df)) sprintf("%.1f%%", 100 * mean(df$gate_fdr[in_top], na.rm = TRUE)) else "0.0%"

    tibble(
      Direction     = label,
      Threshold     = thr$lbl,
      `N positions` = n_top,
      `N FADE-sig`  = n_top_f,
      `% FADE-sig`  = sprintf("%.1f%%", 100 * n_top_f / max(n_top, 1)),
      `% gate_sig`  = pct_sig,
      `% gate_fdr`  = pct_fdr,
      `Odds ratio`  = sprintf("%.2f", ft$estimate),
      `P value`     = formatC(ft$p.value, format = "e", digits = 2)
    )
  })
}

fade_fisher <- dplyr::bind_rows(
  if (!is.null(pos_fade_top))    run_fade_fisher_dir(pos_fade_top,    "TOP"),
  if (!is.null(pos_fade_bottom)) run_fade_fisher_dir(pos_fade_bottom, "BOTTOM")
)

knitr::kable(
  fade_fisher,
  caption = "Fisher exact test for FADE-significant site overlap enrichment within CAAS score thresholds. Direction-specific: TOP positions tested only against fade_sites_top, BOTTOM only against fade_sites_bottom.",
  align = c("l", "l", rep("r", ncol(fade_fisher) - 2))
)
```

```{r reg-accum-subsec, results='asis'}
if (has_genomic) cat("## Regional Accumulation\n\n")
```

```{r position-region-accum, eval=has_genomic, fig.height=10, fig.width=14}
# Build directional window summaries — TOP and BOTTOM use direction-specific
# membership tests (test10/test5/test1: closures from test_top_*/test_bot_*
# above, published-slice-preferring with a quantile() fallback -- pos_sub here
# is pos_genomic_top/pos_genomic_bot, a DIFFERENT row subset than pos_top/
# pos_bottom the closures were built against, so they must be re-evaluated
# against pos_sub's own rows, not reused positionally).
.build_region_wins <- function(pos_sub, test10, test5, test1, chr_offsets_df, dir_label) {
  pos_sub %>%
    mutate(
      win_id    = floor(gene_mid / params$window_size_bp),
      win_start = win_id * params$window_size_bp,
      win_end   = win_start + params$window_size_bp - 1L,
      win_mid   = win_start + params$window_size_bp / 2L,
      in_top10  = test10(pos_sub),
      in_top5   = test5(pos_sub),
      in_top1   = test1(pos_sub)
    ) %>%
    group_by(chr_order, chr_label, win_id, win_start, win_end, win_mid) %>%
    summarise(
      n_positions        = n(),
      n_top10_positions  = sum(in_top10, na.rm = TRUE),
      n_top5_positions   = sum(in_top5,  na.rm = TRUE),
      n_top1_positions   = sum(in_top1,  na.rm = TRUE),
      frac_top_positions = n_top10_positions / n_positions,
      mean_caas          = mean(CAAS_score, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    left_join(chr_offsets_df %>% select(chr_order, offset), by = "chr_order") %>%
    mutate(
      direction    = dir_label,
      win_x        = win_mid + offset,
      region_label = sprintf("chr%s:%0.1f-%0.1f Mb",
                             chr_label, win_start / 1e6, win_end / 1e6)
    )
}

# pos_genomic already has chr_order and gene_mid; split by direction
pos_genomic_top <- pos_genomic %>% filter(change_side %in% c("top",    "both"))
pos_genomic_bot <- pos_genomic %>% filter(change_side %in% c("bottom", "both"))

pos_region_top <- .build_region_wins(pos_genomic_top, test_top_10, test_top_5, test_top_1,
                                      pos_chr_offsets, "TOP")
pos_region_bot <- .build_region_wins(pos_genomic_bot, test_bot_10, test_bot_5, test_bot_1,
                                      pos_chr_offsets, "BOTTOM")

# Combined (global thresholds) for overview panel
pos_region_wins <- pos_genomic %>%
  mutate(
    win_id    = floor(gene_mid / params$window_size_bp),
    win_start = win_id * params$window_size_bp,
    win_end   = win_start + params$window_size_bp - 1L,
    win_mid   = win_start + params$window_size_bp / 2L,
    in_top10  = test_glob_10(pos_genomic),
    in_top5   = test_glob_5(pos_genomic),
    in_top1   = test_glob_1(pos_genomic)
  ) %>%
  group_by(chr_order, chr_label, win_id, win_start, win_end, win_mid) %>%
  summarise(
    n_positions        = n(),
    n_top10_positions  = sum(in_top10, na.rm = TRUE),
    n_top5_positions   = sum(in_top5,  na.rm = TRUE),
    n_top1_positions   = sum(in_top1,  na.rm = TRUE),
    frac_top_positions = n_top10_positions / n_positions,
    mean_caas          = mean(CAAS_score, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(pos_chr_offsets %>% select(chr_order, offset), by = "chr_order") %>%
  mutate(
    win_x        = win_mid + offset,
    region_label = sprintf("chr%s:%0.1f-%0.1f Mb",
                           chr_label, win_start / 1e6, win_end / 1e6)
  )

# Dotplots removed — tables below provide the same information with GenePos detail.
```

```{r position-region-accum-table, eval=has_genomic}
.make_region_tbl <- function(wins_df) {
  wins_df %>%
    filter(n_top10_positions > 0) %>%
    arrange(desc(n_positions), desc(frac_top_positions)) %>%
    slice_head(n = 30) %>%
    mutate(
      frac_top10 = sprintf("%.1f%%", 100 * n_top10_positions / n_positions),
      frac_top5  = sprintf("%.1f%%", 100 * n_top5_positions  / n_positions),
      frac_top1  = sprintf("%.1f%%", 100 * n_top1_positions  / n_positions)
    ) %>%
    select(
      Region              = region_label,
      `N positions`       = n_positions,
      `N top-10%`         = n_top10_positions,
      `% top-10%`         = frac_top10,
      `N top-5%`          = n_top5_positions,
      `% top-5%`          = frac_top5,
      `N top-1%`          = n_top1_positions,
      `% top-1%`          = frac_top1,
      `Mean CAAS`         = mean_caas
    ) %>%
    mutate(`Mean CAAS` = round(`Mean CAAS`, 4))
}

knitr::kable(.make_region_tbl(pos_region_top),
  caption = "TOP-direction: genomic windows with high-CAAS positions, ordered by N positions (top 30; direction-specific thresholds)")
knitr::kable(.make_region_tbl(pos_region_bot),
  caption = "BOTTOM-direction: genomic windows with high-CAAS positions, ordered by N positions (top 30; direction-specific thresholds)")

# Export directional regional accumulation tables as TSV
region_export <- bind_rows(
  .make_region_tbl(pos_region_top) %>% mutate(Direction = "TOP"),
  .make_region_tbl(pos_region_bot) %>% mutate(Direction = "BOTTOM")
) %>% select(Direction, everything())
readr::write_tsv(region_export,
  file.path(params$output_dir, "regional_accumulation.tsv"))
```

---

# Methods

- **Test**: Fisher's exact test, one 2x2 table per (ranking, database, pathway,
  top_frac): foreground = the top 25/10/5/1% of scored positions (by `CAAS_score`,
  `scoring` module `position_scores.tsv`) in that direction; background = the
  full honest tested position universe (all non-gapped "selected" alignment
  columns of the cleaned-background genes) — never restricted to the scored
  subset. Run separately for the **global / top / bottom** directions
  (`change_side`); `top`/`bottom` filter to `change_side` first (matching the
  `scoring` module's own percentile convention — see `11.Scoring_report.Rmd`),
  `global` does not.
- **Two background conventions.** Every database above tests against the
  **full honest background** *except* `cosmic_orthogroups` and
  `pai3d_orthogroups`, which test against a **coverage-restricted background**
  (foreground *and* background both intersected with the set of genes the
  external database itself could ever annotate — see
  `cosmic_coverage_genes.txt` / `pai3d_coverage_genes.txt`, emitted by
  `build_position_gmt.py`). The distinction is *why* a position is absent from
  a term: `orthogroups`/`pfam_domains`/`selection_*`/`ucr_*` are internal
  deterministic computations with complete coverage by construction, so an
  absent position is genuine negative evidence. COSMIC and PAI3D are external
  observational/curated databases with incomplete real-world coverage — a gene
  COSMIC never sequenced, or PAI3D never scored, is absent from the term for a
  reason that has nothing to do with the trait being tested. Testing those two
  GMTs against the full ~1.47M-position background would dilute the signal
  with genes/positions the database could never have annotated regardless of
  outcome; restricting both foreground and background to the coverage subset
  keeps the 2x2 table's implicit universe honest to what was actually tested.
- **Databases**: 10 GMT-style position-set databases (Pfam domains/clans,
  COSMIC and PAI3D and general orthogroups, genomic locations,
  positive/purifying-selection orthogroups, UCR core/flank orthogroups) plus
  the broad `characterization` layers (UCR core/flank, FUBAR
  positive/purifying) — unified onto the same Fisher mechanism. `min`/`max`
  set-size filtering applies only to the GMT-style databases; the broad layers
  were never size-filtered. PAI3D membership is restricted to variants the
  `prediction` column calls pathogenic (substring match, no score-cutoff
  fallback — the same convention as `11.Scoring_report.Rmd`'s
  `is_pathogenic`); COSMIC has no such filter since every somatic mutation
  counts as evidence.
- **`p.adj`**: BH-adjusted per (ranking, database) — a handful of huge broad
  layers and up to thousands of small GMT terms carry very different
  multiple-testing burdens, so they are not pooled into one BH family.
- **Cross-module corroboration (`%gate_sig`, `%gate_fdr`, `%fade_*`, `%rer_*`,
  `%accum`)**: optional, present when POSENRICH is run with an `--annot-file`
  (SCORING's `fcs_stats.tsv`). For each overlap, the distinct genes
  contributing to it (never raw positions — a single gene contributing many
  overlapping positions must not dominate the percentage) are looked up
  against SCORING's own `flag_*` columns, and the percentage carrying each
  flag is reported alongside the Fisher result — the same idea as the
  gene-level FCS report's leading-edge composition, applied to positions.
- **Significance is `p.adj < `r PADJ`` alone** — no fold-enrichment magnitude
  gate is applied. `fold_enrichment` is still computed and reported for every
  result so a minimum-effect-size cutoff can be applied downstream; be aware
  that at this background size (~1.47M positions) even a biologically
  negligible deviation from the expected ratio (e.g. fold ≈ 0.95) can reach a
  "significant" p-value on p-value alone.
- **No permulation (`p.perm`) at position level.** The CAAS permulation null is a
  *naive random relabeling*: when the phenotype is permuted, the fg/bg species are
  paired at **random across the tree**, not as the observed **matched sister
  pairs** — so pair MRCAs are deep and the ASR path-score's isolation/independence
  geometry differs from the observed. That null therefore tests "phenotype + a
  broken matched-pair design" jointly rather than isolating the phenotype, and it
  is anti-conservative on the MRCA axes. The **gene-level FCS** still uses
  `p.perm`; see its report for the same pairing caveat.

*Report generated `r Sys.time()`.*

---

## Summary

This report summarizes the results of the **Enrichment** analysis.

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — Enrichment Module*
*Date of generation: `r Sys.Date()`*
