---
title: "Position-level Enrichment (Path Sum Permulation) - `r params$traitname`"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
    mathjax: null                       # no math in this report
    md_extensions: "-tex_math_dollars"  # pathway names contain '$' (e.g. V$HNF4_01) - never parse as TeX math
params:
  results_file: NULL
  leading_edge_file: NULL
  traitname: "trait"
  padj_thr: 0.15
  table_top_n: 200
  # Position Characterisation: PrimateAI-3D
  # + COSMIC validation and regional accumulation, all restricted to the US scheme.
  # NULL/absent inputs degrade gracefully (has_pos_char is FALSE, section skipped).
  position_scores_file: NULL   # position_scores.tsv (per-position CAAS scores)
  # SCORING's published position_lists/slice_<dir><pct>.tsv dir (scoring_compute.R),
  # when supplied, the 10/5/1% membership used throughout Position
  # Characterisation is read from there instead of an independent quantile()
  # cut computed here (see the has_pos_char setup chunk's test_glob_*/test_top_*/
  # test_bot_* closures). Falls back to quantile() per direction when NULL or a
  # slice file is missing.
  position_lists_dir: NULL
  gene_scores_file:     NULL   # gene_scores.tsv (background gene fallback)
  vep_primateai_file:   NULL   # PrimateAI-3D score TSV
  vep_cosmic_file:      NULL   # COSMIC Mutant Census score TSV
  genomic_info_file:    NULL   # gene, chr, start, end, length (no header assumed)
  # FADE_top_sig/FADE_bottom_sig position group (gene,position,max_bf,target_aa;
  # BF >= fade_bf_threshold only). Direction-specific, unlike PAI3D/COSMIC.
  fade_sites_top_file:    NULL
  fade_sites_bottom_file: NULL
  # SCORING's fcs_stats.tsv (gene + flag_* columns), same file already
  # consumed as --annot-file by posenrich_enrich.py. Unused by this report
  # directly (kept declared for pipeline-wiring compatibility; the Overall
  # dotplot no longer computes a cross-module composite that needed it).
  fcs_stats_file: NULL
  # cleaned_background_main.txt: the honest tested-gene background used to
  # restrict the PrimateAI-3D/COSMIC join sections to genes actually in scope
  # (shared_genes_pai/shared_genes_cosmic below), rather than falling back to
  # whichever genes happen to have a gene_scores.tsv row.
  universe_file: NULL
  top_pct:      0.10
  window_size_bp: 1000000
  output_dir:   "."            # regional_accumulation.tsv export target
  seed:         "1998"
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

<!-- 14.Position_enrichment_report.Rmd — Position-level path-sum permulation enrichment and characterization of the CAAS positions. -->
<!-- PhyloPhere | subworkflows/ENRICHMENT/local/ -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6, dev = "png", dpi = 150)

# ── Dependencies and theme ────────────────────────────────────────────────────
suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
  library(kableExtra)
  library(DT)
  theme_set(
    theme_minimal(base_size = 13) +
    theme(
      axis.title.y  = element_text(margin = ggplot2::margin(l = 20, r = 20)),
      axis.title.x  = element_text(margin = ggplot2::margin(t = 12)),
      axis.text.x   = element_text(margin = ggplot2::margin(t = 5)),
      plot.title    = element_text(face = "bold", hjust = 0, margin = ggplot2::margin(b = 6)),
      plot.subtitle = element_text(colour = "grey40", hjust = 0, margin = ggplot2::margin(b = 10)),
      legend.position = "right",
      legend.box.margin = ggplot2::margin(t = 5)
    )
  )

  seed_val <- if (exists("params") && !is.null(params$seed) && nzchar(as.character(params$seed))) as.integer(params$seed) else 1998
  set.seed(seed_val)
  library(scales)
  library(patchwork)
})

# ── Parameters and helpers ────────────────────────────────────────────────────
# BH threshold of significance and rows shown per table.
PADJ <- params$padj_thr
TOP_N <- params$table_top_n

# Directory of this Rmd.
this_dir <- tryCatch(dirname(normalizePath(knitr::current_input(), mustWork = FALSE)),
                     error = function(e) ".")

# HTML table with the report style; optional scroll box of fixed height.
kbl_html <- function(df, caption = NULL, height = NULL, escape = TRUE, ...) {
  k <- kable(df, format = "html", escape = escape, caption = caption, ...) %>%
    kable_styling(bootstrap_options = c("striped", "condensed", "hover"), full_width = FALSE)
  if (!is.null(height)) k <- scroll_box(k, height = height)
  k
}

# Correlation that returns NA, not an error, with fewer than 3 complete pairs, so that
# sparse optional sections do not abort the render.
safe_cor <- function(x, y, method = "pearson") {
  ok <- complete.cases(x, y)
  if (sum(ok) < 3) return(NA_real_)
  cor(x[ok], y[ok], method = method)
}
```

```{r load, include=FALSE}
# ── Results table ─────────────────────────────────────────────────────────────
stopifnot(!is.null(params$results_file), file.exists(params$results_file))
res <- readr::read_tsv(params$results_file, show_col_types = FALSE)
if (!is.null(res) && "top_frac" %in% names(res)) {
  res <- res %>% dplyr::filter(!top_frac %in% c(1.0, 0.5))
} else if (!is.null(res)) {
  res$top_frac <- NA_real_
}
if (!is.null(res)) {
  if (!"foreground_size" %in% names(res)) res$foreground_size <- if ("n_pos_with_score" %in% names(res)) res$n_pos_with_score else NA_real_
  if (!"observed" %in% names(res)) res$observed <- if ("obs_sum" %in% names(res)) res$obs_sum else NA_real_
  if (!"expected" %in% names(res)) res$expected <- if ("null_mean" %in% names(res)) res$null_mean else NA_real_
  if (!"fold_enrichment" %in% names(res)) res$fold_enrichment <- if ("perm_nes" %in% names(res)) res$perm_nes else NA_real_
}
has_res <- !is.null(res) && nrow(res) > 0
RANKING_ORDER <- c("global", "top", "bottom")
if (has_res && !"sig" %in% names(res)) {
  # Significant: p_adj < padj_thr and NES > 0.
  res$sig <- res$p_adj < PADJ & (!"perm_nes" %in% names(res) | res$perm_nes > 0)
}

# Truncate a "gene1, gene2, ..." string to its first n_show entries; the remainder sits
# behind an inline <details>/<summary> toggle.
truncate_gene_list <- function(x, n_show = 5) {
  vapply(x, function(s) {
    if (is.na(s) || !nzchar(s)) return("")
    items <- strsplit(s, ", ", fixed = TRUE)[[1]]
    if (length(items) <= n_show) return(s)
    shown <- paste(items[seq_len(n_show)], collapse = ", ")
    rest  <- paste(items[(n_show + 1):length(items)], collapse = ", ")
    sprintf(
      '%s <details style="display:inline"><summary style="display:inline; cursor:pointer; color:#2166ac;">&hellip; (%d more)</summary> %s</details>',
      shown, length(items) - n_show, rest
    )
  }, character(1), USE.NAMES = FALSE)
}

# ── Leading edge and flags ────────────────────────────────────────────────────
# Observed gene:position members of the overlap, significant terms only.
le_raw <- NULL
if (!is.null(params$leading_edge_file) && file.exists(params$leading_edge_file)) {
  le_raw <- tryCatch(readr::read_tsv(params$leading_edge_file, show_col_types = FALSE),
                     error = function(e) NULL)
  if (!is.null(le_raw) && "top_frac" %in% names(le_raw)) {
    le_raw <- le_raw %>% dplyr::filter(!top_frac %in% c(1.0, 0.5))
  } else if (!is.null(le_raw)) {
    le_raw$top_frac <- NA_real_
  }
  if (!is.null(le_raw) && nrow(le_raw) == 0) le_raw <- NULL
}
# Pooled across the top_frac cutoffs: one row per ranking x database x pathway.
le_summary <- if (!is.null(le_raw)) {
  le_raw %>%
    dplyr::group_by(ranking, database, pathway) %>%
    dplyr::summarise(`Genes/positions` = paste(sort(unique(gene_position)), collapse = ", "),
                     `N genes` = dplyr::n_distinct(gene),
                     .groups = "drop")
} else NULL

# Direction-aware corroboration columns (as cols_by_rk of 12.FCS_general_report.Rmd):
# pct_fade_bottom only under `bottom`, pct_fade_top only under `top`, and no combined
# fade column under `global`. pct_accum_top and pct_accum_bottom are the
# direction-restricted accumulation tests, as for FADE; the non-directional pct_accum is
# reserved for `global`. Only the columns present in the results are shown.
flag_cols_by_rk <- list(
  global = c("pct_rer_acc", "pct_rer_decc", "pct_accum"),
  top    = c("pct_fade_top", "pct_rer_acc", "pct_rer_decc", "pct_accum_top"),
  bottom = c("pct_fade_bottom", "pct_rer_acc", "pct_rer_decc", "pct_accum_bottom")
)

.fcs_stats_file_ok <- function(p) !is.null(p) && !is.na(p) && nchar(p) > 0 && file.exists(p)

# FADE_top_sig and FADE_bottom_sig sites (gene, position, max_bf, target_aa). They are
# direction-specific: a top position is only checked against fade_sites_top and a
# bottom position against fade_sites_bottom. Position is already in the coordinate space
# of the CAAS positions, so no coordinate mapping is needed.
has_fade_sites_top    <- .fcs_stats_file_ok(params$fade_sites_top_file)
has_fade_sites_bottom <- .fcs_stats_file_ok(params$fade_sites_bottom_file)
fade_sites_top_df <- if (has_fade_sites_top) {
  readr::read_csv(params$fade_sites_top_file, show_col_types = FALSE) %>%
    dplyr::rename(Gene = gene, Position = position) %>%
    dplyr::mutate(Position = suppressWarnings(as.integer(Position)))
} else NULL
fade_sites_bottom_df <- if (has_fade_sites_bottom) {
  readr::read_csv(params$fade_sites_bottom_file, show_col_types = FALSE) %>%
    dplyr::rename(Gene = gene, Position = position) %>%
    dplyr::mutate(Position = suppressWarnings(as.integer(Position)))
} else NULL
```

# Overview {.tabset .tabset-fade}

> [!WARNING]
> **Proof of concept - not validated for standalone significance claims.** This
> module has not had a real end-to-end production run, and its GMT construction,
> its handling of the very large (~1.47M-position) background, and its
> significance gating (`p_adj < padj_thr` with `NES > 0`) have not been
> checked against any known-truth signal. Treat everything in this report as
> descriptive characterization only (e.g. "does this term show up near
> CAAS-high positions") - do not cite it as evidence for a biological claim on
> its own, and do not build downstream decisions on top of its "significant"
> term counts without independent corroboration.

This report summarises **position-level enrichment** for `r params$traitname`. Unlike
the gene-level FCS (a Wilcoxon-AUC rank-shift test over genes), the position background is a
sea of invariant columns where <0.2% of positions carry a non-zero CAAS score. A rank-shift test
degenerates there, and fixed-cutoff Fisher tests suffer from extreme sample-size sensitivity at $N \approx 1.47\text{M}$.
Instead, we use **Position-Level Path Sum Permulation**: accumulating continuous CAAS score magnitudes per term ($T_\text{obs} = \sum_{p \in \text{term}} s_p$)
and comparing against a label-permuted null distribution across the full ~1.47M position background ($N_\text{perms} = 2,000$).

A set is flagged **significant** when `p.adj < `r PADJ`` (BH-adjusted per
ranking × database) and `NES > 0`.

```{r overview, echo=FALSE, results='asis'}
if (!has_res) {
  cat("No enrichment results were produced.\n")
} else {
  for (rk in intersect(RANKING_ORDER, unique(res$ranking))) {
    cat(sprintf("\n## %s\n\n", toupper(rk)))
    ov <- res %>% filter(ranking == rk) %>%
      group_by(database) %>%
      summarise(`Sets tested` = n_distinct(pathway),
                `Sig (p.adj)` = n_distinct(pathway[sig]),
                .groups = "drop") %>%
      arrange(database)
    print(kbl_html(ov, caption = sprintf(
      "%s - sets tested and significant at p.adj < %.3g.",
      toupper(rk), PADJ)))
    cat("\n\n")
  }
}
```

# Enriched position sets {.tabset .tabset-fade}

Foreground = the top 25 / 10 / 5 / 1% of scored positions by `CAAS_score` in
that direction (`top`/`bottom` restrict to `side` first, matching the
SCORING module's own percentile convention; no `global` ranking - dropped in
favour of top/bottom only, consistent with every other report in this
pipeline). Background = the full honest tested
position pool for every test - never restricted to the scored subset, so
tested-and-negative positions are never silently dropped from the null.
`fold > 1` = enriched, `< 1` = depleted. UCR core and UCR flank (and their
per-orthogroup counterparts) are tested as independent, potentially-overlapping
layers, not a forced partition - nothing here requires them to be disjoint.
When available, `%gate_sig`/`%fade_*`/`%rer_*`/`%accum` columns
report the percentage of *distinct genes* contributing to the overlap (never
raw positions - one gene must not dominate) that are independently flagged by
SCORING's own significance/FADE/RER/accumulation criteria - a cross-module
corroboration check for whether a hit is broadly supported or isolated.

```{r enrichment-tables, echo=FALSE, results='asis'}
if (!has_res) {
  cat("_No results._\n\n")
} else {
  for (rk in intersect(RANKING_ORDER, unique(res$ranking))) {
    cat(sprintf("\n## %s {.tabset}\n\n", toupper(rk)))
    rk_df <- res %>% filter(ranking == rk)
    for (db in unique(rk_df$database)) {
      cat(sprintf("\n### %s {.tabset}\n\n", db))
      db_df <- rk_df %>% filter(database == db)

      has_tf_vals <- "top_frac" %in% names(db_df) && !all(is.na(db_df$top_frac))
      tf_vals <- if (has_tf_vals) sort(unique(db_df$top_frac), decreasing = TRUE) else NA_real_
      for (tf in tf_vals) {
        if (!is.na(tf)) cat(sprintf("\n#### Top %s%%\n\n", format(100 * tf, trim = TRUE)))
        tf_all <- if (!is.na(tf)) db_df %>% filter(top_frac == tf) else db_df
        tf_all <- tf_all %>% arrange(desc(sig), p_adj, p_value)
        n_total_tf <- nrow(tf_all)
        tf_src <- tf_all %>% head(TOP_N)
        if (n_total_tf > TOP_N) {
          cat(sprintf("_Showing top %d of %d tested pathways (sorted by significance)._\n\n", TOP_N, n_total_tf))
        }
        if ("perm_nes" %in% names(tf_src)) {
          tf_df <- tf_src %>%
            transmute(Pathway = pathway, Description = description,
                      `Layer size` = layer_size, `N scored` = n_pos_with_score,
                      `Observed sum` = round(obs_sum, 2), `Null mean` = round(null_mean, 2),
                      NES = round(perm_nes, 3), Direction = direction,
                      `p (Perm)` = signif(p_value, 3),
                      `p.adj` = signif(p_adj, 3), Sig = ifelse(sig, "*", ""))
        } else {
          tf_df <- tf_src %>%
            transmute(Pathway = pathway, Description = description,
                      `Layer size` = layer_size, `Foreground` = foreground_size,
                      Observed = observed, Expected = round(expected, 1),
                      Fold = round(fold_enrichment, 3), Direction = direction,
                      `p (Fisher)` = signif(p_value, 3),
                      `p.adj` = signif(p_adj, 3), Sig = ifelse(sig, "*", ""))
        }
        # Cross-module corroboration: % of the DISTINCT genes in the overlap
        # (never % of raw positions - one gene must not dominate) carrying
        # each SCORING flag (gate_sig/fade/rer/accum). Only present
        # when POSENRICH was run with --annot-file. Direction-restricted via
        # flag_cols_by_rk (same convention as the Leading edge section below) --
        # posenrich_enrich.py computes pct_<flag> for every flag regardless of
        # direction, so without this restriction a `bottom` table would show
        # `%accum_top`/`%fade_top` alongside its own `%accum_bottom`/`%fade_bottom`.
        want_rk <- flag_cols_by_rk[[rk]]
        if (is.null(want_rk)) want_rk <- character(0)
        flag_cols <- intersect(want_rk, names(tf_src))
        if (length(flag_cols) > 0) {
          flag_df <- tf_src %>% transmute(across(all_of(flag_cols), ~round(.x, 1)))
          names(flag_df) <- paste0("%", sub("^pct_", "", flag_cols))
          tf_df <- dplyr::bind_cols(tf_df, flag_df)
        }
        id <- paste("dt", rk, db, gsub("\\.", "_", as.character(tf)), sep = "_")
        cat(knitr::knit_child(text = c(
          paste0("```{r ", id, ", echo=FALSE}"),
          "DT::datatable(tf_df, options = list(pageLength = 10, scrollX = TRUE), rownames = FALSE)",
          "```"), envir = environment(), quiet = TRUE))
        cat("\n\n")
      }
    }
  }
}
```

```{r term-clustering-setup, include=FALSE}
# Term-similarity clustering, the same as in 13.AMI_analysis.Rmd. Different GMT databases
# describe the same concept with slightly different wording (case, punctuation, minor
# rewording), merged through a normalized edit distance (utils::adist()), or one
# description nests in another ("Primary ciliary dyskinesia" inside "Axonemal dynein
# complex and Primary ciliary dyskinesia"). The edit distance alone rejects the nested
# pair because the extra words dominate the ratio, so containment is tested explicitly
# and forces the merge. The functions are defined before the Overall dotplot, the
# Direction triangle and the Term recurrence sections so that all three use the same
# canonical Term per (database, pathway).
.normalize_term <- function(x) {
  x <- tolower(x)
  x <- gsub("[^a-z0-9 ]", " ", x)
  gsub("\\s+", " ", trimws(x))
}
.term_contains <- function(a, b) {
  wa <- strsplit(a, " ")[[1]]; wb <- strsplit(b, " ")[[1]]
  if (length(wa) == 0 || length(wb) == 0) return(FALSE)
  short <- if (length(wa) <= length(wb)) wa else wb
  long  <- if (length(wa) <= length(wb)) wb else wa
  all(short %in% long)
}
.cluster_terms <- function(terms, sim_thr = 0.85) {
  norm <- vapply(terms, .normalize_term, character(1))
  reps_norm <- character(0); reps_orig <- character(0)
  assign <- integer(length(terms))
  for (i in seq_along(terms)) {
    if (length(reps_norm) == 0) { reps_norm <- norm[i]; reps_orig <- terms[i]; assign[i] <- 1L; next }
    d <- utils::adist(norm[i], reps_norm)[1, ]
    sim <- 1 - d / pmax(nchar(norm[i]), nchar(reps_norm), 1)
    contained <- vapply(reps_norm, function(r) .term_contains(norm[i], r), logical(1))
    sim[contained] <- 1
    best <- which.max(sim)
    if (sim[best] >= sim_thr) {
      assign[i] <- best
      if (contained[best] && nchar(terms[i]) < nchar(reps_orig[best])) {
        reps_orig[best] <- terms[i]
      }
    } else {
      reps_norm <- c(reps_norm, norm[i]); reps_orig <- c(reps_orig, terms[i])
      assign[i] <- length(reps_norm)
    }
  }
  reps_orig[assign]
}
```

# Overall dotplot {.tabset .tabset-fade}

Pooled across all databases and threshold cutoffs, ranked by BH-adjusted p
(the one test this report runs - Position-Level Path Sum Permulation - so
there is no multi-method vote to combine, unlike the gene-level FCS report).
One dotplot per direction, restricted to the top `r params$table_top_n`
terms, on a **shared x-axis of `-log10(BH-adjusted p)`**. Each term's segment
colour encodes its own **NES** (light-to-dark, darker = larger NES among the
terms shown); the **tip marker** is drawn in a fixed dark shade - always
darker than any point on the segment's own NES gradient - so the
significance endpoint stands out regardless of that term's NES. `list
_threshold`/`n_occurrences` (also in the table export) record which
10%/5%/1% cutoffs recovered the term and how many, as a simple, auditable
recurrence count - not a weighted score.

```{r overall-dotplot, echo=FALSE, results='asis', fig.height=6}
.NES_HUE_LIGHT <- "#e7d4e8"  # color family of the Permulation method in 12.FCS_general_report.Rmd,
.NES_HUE_DARK  <- "#4a148c"  # the one test of this report.

# Darken a hex color toward black by `amount` (0-1); the tip marker takes a deeper shade
# of the color of its own segment.
.darken_hex <- function(hex, amount = 0.3) {
  rgb <- grDevices::col2rgb(hex) / 255
  rgb <- rgb * (1 - amount)
  grDevices::rgb(rgb[1, ], rgb[2, ], rgb[3, ])
}

# Light-to-dark gradient bar used as the legend key: scale_colour_identity() draws only
# flat swatches, so the color ramp would otherwise be absent from the legend.
.method_gradient_legend <- function(title, light, dark) {
  df <- data.frame(y = 1, v = seq(0, 1, length.out = 100))
  ggplot(df, aes(x = v, y = y, fill = v)) +
    geom_tile() +
    scale_fill_gradientn(colours = c(light, dark), guide = "none") +
    scale_x_continuous(breaks = c(0, 1), labels = c("low", "high")) +
    scale_y_continuous(breaks = NULL) +
    labs(title = title, x = NULL, y = NULL) +
    theme_minimal(base_size = 8) +
    theme(plot.title = element_text(size = 8, hjust = 0.5),
          panel.grid = element_blank(),
          plot.margin = margin(2, 2, 2, 2))
}

# knitr opens the figure device once per chunk with the fig.height in effect at the
# start, so setting it inside the per-direction loop does not resize the image written
# out. Each plot is rendered and embedded as a base64 PNG instead, which allows a
# different height per plot in one asis chunk.
.emit_sized_plot <- function(p, height_in, width_in = knitr::opts_chunk$get("fig.width"),
                              dpi = knitr::opts_chunk$get("dpi")) {
  f <- tempfile(fileext = ".png")
  on.exit(unlink(f), add = TRUE)
  ggplot2::ggsave(f, plot = p, width = width_in, height = height_in, dpi = dpi, units = "in")
  cat(sprintf('<img src="data:image/png;base64,%s" style="max-width:100%%;" />\n\n',
              base64enc::base64encode(f)))
}

.overall_dotplot_rows <- list()  # one table per ranking, written to posenrich_overall_dotplot.tsv
                                  # after the loop (read by the Comparison report).
if (!has_res || !any(res$sig, na.rm = TRUE)) {
  cat("_No significant sets._\n\n")
} else {
  for (rk in intersect(RANKING_ORDER, unique(res$ranking[res$sig]))) {
    cat(sprintf("\n## %s\n\n", toupper(rk)))
    rk_sig <- res %>% filter(ranking == rk, sig)
    if (nrow(rk_sig) == 0) { cat("_No significant sets for this direction._\n\n"); next }

    # list_threshold and n_occurrences: the top_frac cutoffs at which the term is
    # significant and their number, a plain recurrence count.
    .has_tf_col <- "top_frac" %in% names(rk_sig) && !all(is.na(rk_sig$top_frac))
    term_meta <- rk_sig %>%
      dplyr::group_by(database, pathway) %>%
      dplyr::summarise(p_adj = min(p_adj, na.rm = TRUE),
                        perm_nes = { v <- stats::na.omit(perm_nes); if (length(v)) max(v) else NA_real_ },
                        description = dplyr::first(dplyr::coalesce(description, pathway)),
                        list_threshold = if (.has_tf_col) paste(sort(100 * unique(na.omit(top_frac)), decreasing = TRUE), "%", sep = "", collapse = ", ") else "N/A",
                        n_occurrences = if (.has_tf_col) dplyr::n_distinct(top_frac, na.rm = TRUE) else 1L,
                        .groups = "drop")

    # Number of positions in the leading edge (dot size); descriptive, it does not enter
    # the ranking.
    if (!is.null(le_raw)) {
      n_le_tbl <- le_raw %>% dplyr::filter(ranking == rk) %>%
        dplyr::group_by(database, pathway) %>%
        dplyr::summarise(n_le = dplyr::n_distinct(gene_position), .groups = "drop")
      term_meta <- term_meta %>% dplyr::left_join(n_le_tbl, by = c("database", "pathway"))
    } else {
      term_meta$n_le <- NA_integer_
    }

    dp_all <- term_meta %>%
      dplyr::filter(!is.na(p_adj)) %>%
      dplyr::arrange(p_adj) %>%
      dplyr::mutate(label = stringr::str_trunc(sprintf("%s (%s)", description, database), 60))

    # Every significant term of this direction is kept for the triangle plot and the
    # exported TSV; only the dotplot is limited to TOP_N.
    .overall_dotplot_rows[[rk]] <- dp_all %>% dplyr::mutate(ranking = rk, .before = 1)

    dp <- dp_all %>% head(TOP_N)
    if (nrow(dp) == 0) { cat("_No terms to display._\n\n"); next }

    # y index 1 is the bottom of the plot: reversing puts the term with the smallest p_adj at the top.
    term_levels <- rev(unique(dp$label))
    dp <- dp %>%
      dplyr::mutate(
        y_base = as.numeric(factor(label, levels = term_levels)),
        test_signal = -log10(pmax(p_adj, 1e-300)),
        .nes_pct = dplyr::percent_rank(perm_nes),
        arm_colour = grDevices::colorRampPalette(c(.NES_HUE_LIGHT, .NES_HUE_DARK))(101)[
          pmax(1, pmin(101, round(dplyr::coalesce(.nes_pct, 0.7) * 100) + 1))],
        # Tip marker: darkened shade of the color of the row, tying the dot to its bar.
        tip_colour = .darken_hex(arm_colour)
      )

    p <- ggplot(dp, aes(y = y_base)) +
      geom_segment(aes(x = 0, xend = test_signal, yend = y_base, colour = arm_colour),
                   linewidth = 1.1, lineend = "round", alpha = 0.85) +
      geom_point(aes(x = test_signal, size = n_le, colour = tip_colour), alpha = 1) +
      scale_y_continuous(breaks = seq_along(term_levels), labels = term_levels) +
      scale_colour_identity() +
      scale_size_continuous(name = "N positions") +
      geom_vline(xintercept = -log10(PADJ), linetype = "dashed", colour = "grey50") +
      labs(title = sprintf("%s - %s (pooled across databases)", params$traitname, toupper(rk)),
           subtitle = "Position-Level Path Sum Permulation - tip at the BH-adjusted p",
           x = expression(-log[10]~"BH-adjusted p"), y = NULL) +
      theme_minimal(base_size = 11) +
      theme(axis.text.y = element_text(colour = "black",
                                        size = pmax(8, pmin(11, 220 / length(term_levels)))),
            plot.subtitle = element_text(margin = margin(b = 12)),
            legend.position = "bottom",
            legend.box = "horizontal")

    # Method legend: the gradient bar of the one test of this report (Permulation NES),
    # directly under p, which carries its own "N positions" size legend at the bottom
    # (legend.position = "bottom"), so the whole legend reads as one column.
    gradient_bar <- .method_gradient_legend("Permulation (NES)", .NES_HUE_LIGHT, .NES_HUE_DARK)
    # The gradient bar has a fixed absolute height (0.18 in): a relative split would make
    # it grow with the main plot on reports with many terms.
    combo <- p / gradient_bar + patchwork::plot_layout(heights = grid::unit(c(1, 0.18), c("null", "in")))
    .emit_sized_plot(combo, height_in = pmax(5, pmin(22, 2.6 + 0.4 * length(term_levels))))
  }
}
if (length(.overall_dotplot_rows) > 0) {
  readr::write_tsv(dplyr::bind_rows(.overall_dotplot_rows), "posenrich_overall_dotplot.tsv")
}
```

# Direction triangle (NES)

One point per **term** - merged across near-duplicate descriptions the same
way as the Term recurrence table below (so a term split across two GMT
databases, or two slightly-differently-worded entries, is one point here
too, not two incomplete ones) - positioned by **which direction(s)** it is
significant in and **how strongly** (NES, clipped to ≥ 0) in each: **Global**
at the top vertex, **Bottom** at the left vertex, **Top** at the right
vertex. A term significant in only one direction sits exactly at that
vertex (there is no second direction to weigh it against, so its exact NES
value cannot move it off the vertex); a term significant in two sits
somewhere along the edge between them, positioned by the relative size of
the two NES values; a term significant in **all three** sits inside the
triangle. Colour marks which of the seven possible combinations a term
falls into (three single-direction, three two-direction edges, one
three-direction interior).

Terms sharing an exact vertex/edge position (most commonly every
single-direction term, which all collapse to the same vertex) are fanned out
around it rather than randomly jittered, **ordered by total raw NES**
(stronger terms sit closer to the vertex, weaker ones further along the
fan) - the fan position is not itself a magnitude axis, but its *order* is,
so two overlapping dots are never placed arbitrarily relative to each other.
Only terms significant in **two or three** directions are labelled (edge and
interior points) - single-direction terms sit exactly on a vertex, where
they're already identified by that vertex, so labelling them too would just
clutter the densest part of the plot without adding information.

```{r direction_triangle, echo=FALSE, results='asis'}
sc_all <- if (length(.overall_dotplot_rows) > 0) dplyr::bind_rows(.overall_dotplot_rows) else tibble::tibble()
if (nrow(sc_all) == 0) {
  cat("_No significant terms to compare across directions._\n\n")
} else {
  # Near-duplicate descriptions are merged before pivoting to directions, so that a term
  # split across two databases or two wordings is one point, as in the Recurrence Table
  # and Heatmap below. Rows that merge into the same Term and ranking (two databases
  # both significant under "top") keep the higher NES, the "best across merged aliases"
  # convention of 15.Comparison_report.Rmd.
  desc_levels <- unique(dplyr::coalesce(sc_all$description, sc_all$pathway))
  cluster_map <- setNames(.cluster_terms(desc_levels), desc_levels)
  sc_all <- sc_all %>%
    dplyr::mutate(Term = unname(cluster_map[dplyr::coalesce(description, pathway)])) %>%
    dplyr::group_by(Term, ranking) %>%
    # max(na.rm = TRUE) of an all-NA group returns -Inf with a warning. The guard keeps NA
    # for a pathway with p_adj but no recorded NES.
    dplyr::summarise(perm_nes = if (all(is.na(perm_nes))) NA_real_ else max(perm_nes, na.rm = TRUE),
                      .groups = "drop")

  sc <- sc_all %>%
    tidyr::pivot_wider(
      id_cols = Term,
      names_from = ranking, values_from = perm_nes, names_prefix = "nes_"
    )
  for (rc in setdiff(c("nes_global", "nes_top", "nes_bottom"), names(sc))) sc[[rc]] <- NA_real_

  sc <- sc %>%
    dplyr::mutate(
      has_g = !is.na(nes_global), has_t = !is.na(nes_top), has_b = !is.na(nes_bottom),
      w_g = pmax(dplyr::coalesce(nes_global, 0), 0),
      w_t = pmax(dplyr::coalesce(nes_top, 0), 0),
      w_b = pmax(dplyr::coalesce(nes_bottom, 0), 0)
    ) %>%
    dplyr::filter(has_g | has_t | has_b) %>%
    dplyr::mutate(
      # A direction with NES <= 0 still counts for the category (it passed its own gate)
      # but has zero barycentric weight. When all weights are 0 the weight is split
      # equally among the significant directions, so that the point does not collapse
      # to the origin.
      w_sum = w_g + w_t + w_b,
      w_g = dplyr::if_else(w_sum > 0, w_g / w_sum, as.numeric(has_g) / pmax(has_g + has_t + has_b, 1)),
      w_t = dplyr::if_else(w_sum > 0, w_t / w_sum, as.numeric(has_t) / pmax(has_g + has_t + has_b, 1)),
      w_b = dplyr::if_else(w_sum > 0, w_b / w_sum, as.numeric(has_b) / pmax(has_g + has_t + has_b, 1)),
      x = w_t * 1 + w_g * 0.5,
      y = w_g * (sqrt(3) / 2),
      Category = dplyr::case_when(
        has_g & has_t & has_b ~ "all",
        has_g & has_t         ~ "global_top",
        has_g & has_b         ~ "global_bottom",
        has_t & has_b         ~ "top_bottom",
        has_g                 ~ "global",
        has_t                 ~ "top",
        has_b                 ~ "bottom"
      ),
      Category = factor(Category, levels = c("global", "top", "bottom", "top_bottom",
                                              "global_bottom", "global_top", "all")),
      n_dirs = has_g + has_t + has_b,
      # Total NES over the significant directions. It only orders the points that land on
      # the same vertex or edge position; it is not a coordinate.
      strength = pmax(dplyr::coalesce(nes_global, 0), 0) +
                 pmax(dplyr::coalesce(nes_top, 0), 0) +
                 pmax(dplyr::coalesce(nes_bottom, 0), 0)
    )

  if (nrow(sc) == 0) {
    cat("_No plottable points._\n\n")
  } else {
    tri <- tibble::tibble(x = c(0.5, 0, 1, 0.5), y = c(sqrt(3) / 2, 0, 0, sqrt(3) / 2))
    tri_cx <- 0.5; tri_cy <- sqrt(3) / 6  # triangle centroid
    cat_colours <- c(global = "#f1c40f", top = "#2980b9", bottom = "#27ae60",
                     top_bottom = "#8e44ad", global_bottom = "#16a085",
                     global_top = "#d35400", all = "#e74c3c")

    # Deterministic spread of coincident points. The barycentric coordinates put every
    # single-direction term on the same vertex (and terms can coincide on an edge), with
    # no magnitude axis to separate them. Random jitter would suggest a meaning that the
    # offset does not have, so colliding points are fanned out along the tangent to their
    # shared position (perpendicular to the ray from the centroid of the triangle) in the
    # order of `strength`: stronger terms stay closer to the true position. The spacing
    # shrinks as a cluster grows, which keeps the fan near its vertex.
    sc <- sc %>%
      dplyr::group_by(x, y) %>%
      dplyr::mutate(
        .n_here = dplyr::n(),
        .rank_here = rank(-strength, ties.method = "first"),
        .radial_x = x - tri_cx, .radial_y = y - tri_cy,
        .rlen = pmax(sqrt(.radial_x^2 + .radial_y^2), 1e-6),
        .tan_x = -.radial_y / .rlen, .tan_y = .radial_x / .rlen,
        .spacing = pmin(0.035, 0.5 / pmax(.n_here - 1, 1)),
        .offset = dplyr::if_else(.n_here > 1, (.rank_here - (.n_here + 1) / 2) * .spacing, 0),
        x = x + .tan_x * .offset,
        y = y + .tan_y * .offset
      ) %>%
      dplyr::ungroup() %>%
      dplyr::select(-.n_here, -.rank_here, -.radial_x, -.radial_y, -.rlen, -.tan_x, -.tan_y, -.spacing, -.offset)

    p <- ggplot(sc, aes(x, y)) +
      geom_path(data = tri, colour = "grey50", linetype = "dashed") +
      geom_point(aes(colour = Category), alpha = 0.8, size = 2.2) +
      annotate("text", x = 0.5, y = sqrt(3) / 2 + 0.04, label = "Global", fontface = "bold") +
      annotate("text", x = -0.05, y = -0.04, label = "Bottom", fontface = "bold", hjust = 1) +
      annotate("text", x = 1.05, y = -0.04, label = "Top", fontface = "bold", hjust = 0) +
      scale_colour_manual(values = cat_colours, name = "Significant in", drop = FALSE) +
      coord_equal(clip = "off") +
      labs(title = sprintf("Position enrichment - direction triangle (NES) - %s", params$traitname),
           subtitle = "Vertex = single direction, edge = two directions, interior = all three, overlapping terms fanned out by total NES, not jittered") +
      theme_void(base_size = 12) +
      theme(plot.subtitle = element_text(margin = margin(b = 12), colour = "grey40"),
            legend.position = "right",
            plot.margin = margin(20, 40, 10, 40))

    # Only terms shared by two or three directions (edge and interior points) are
    # labeled; single-direction terms are identified by their vertex.
    shared_sc <- dplyr::filter(sc, n_dirs >= 2)
    if (nrow(shared_sc) > 0) {
      p <- p + ggrepel::geom_text_repel(
        data = shared_sc, aes(label = stringr::str_trunc(Term, 45)), colour = "black",
        size = 2.6, fontface = "bold", max.overlaps = Inf, seed = seed_val,
        box.padding = 0.3, point.padding = 0.2,
        force = 3, force_pull = 0.5,
        max.time = 5, max.iter = 200000,
        min.segment.length = 0, segment.size = 0.3, segment.alpha = 0.6,
        bg.color = "white", bg.r = 0.1, show.legend = FALSE)
    }
    # The height grows with the number of points; the plot is embedded with
    # .emit_sized_plot() as in the Overall dotplot.
    .emit_sized_plot(p, height_in = pmax(6, pmin(20, 4 + 0.03 * nrow(sc))))
  }
}
```

# Leading edge {.tabset .tabset-fade}

One row per **significant** (`p.adj < `r PADJ``) pathway - pooled across the
10%/5%/1% cutoffs it was tested at, not one row per cutoff - with the union
of observed `gene:position` members across those cutoffs, plus the same
cross-module corroboration columns as above (`%gate_sig`/`%fade_*`/`%rer_*`/
`%accum`, direction-aware - `%fade_bottom` only appears under `BOTTOM`,
`%fade_top` only under `TOP`, plain `%fade` under `GLOBAL`). `N cutoffs`/
`Cutoffs` record which of the tighter cutoffs recovered the term as a plain
audit trail. The full per-cutoff member list is also written to
`posenrich_leading_edge.tsv`, and this pooled table to
`posenrich_leading_edge_summary.tsv`, both alongside this report - the
Comparison report imports the latter directly.

```{r leading-edge, echo=FALSE, results='asis'}
.leading_edge_summary_rows <- list()
if (!has_res || !any(res$sig, na.rm = TRUE)) {
  cat("_No significant sets._\n\n")
} else {
  sig_res <- res %>% filter(sig)
  for (rk in intersect(RANKING_ORDER, unique(sig_res$ranking))) {
    cat(sprintf("\n## %s\n\n", toupper(rk)))
    rk_sig_raw <- sig_res %>% filter(ranking == rk)
    if (nrow(rk_sig_raw) == 0) { cat("_No significant sets for this direction._\n\n"); next }

    want <- flag_cols_by_rk[[rk]]
    if (is.null(want)) want <- character(0)
    present_flags <- intersect(want, names(rk_sig_raw))
    has_nes <- "perm_nes" %in% names(rk_sig_raw)
    .has_tf <- "top_frac" %in% names(rk_sig_raw) && !all(is.na(rk_sig_raw$top_frac))

    # One row per pathway: the smallest p.adj, the largest NES or fold over the cutoffs at
    # which it recurs, and the mean of the flag percentages over those cutoffs.
    rk_sig <- rk_sig_raw %>%
      dplyr::group_by(database, pathway) %>%
      dplyr::summarise(
        description = dplyr::first(dplyr::coalesce(description, pathway)),
        direction = dplyr::first(direction),
        p_adj = min(p_adj, na.rm = TRUE),
        perm_nes = if (has_nes) { v <- stats::na.omit(perm_nes); if (length(v)) max(v) else NA_real_ } else NA_real_,
        fold_enrichment = if (!has_nes) { v <- stats::na.omit(fold_enrichment); if (length(v)) max(v) else NA_real_ } else NA_real_,
        n_occurrences = if (.has_tf) dplyr::n_distinct(top_frac, na.rm = TRUE) else 1L,
        `Cutoffs` = if (.has_tf) paste(sort(100 * unique(na.omit(top_frac)), decreasing = TRUE), "%", sep = "", collapse = ", ") else "N/A",
        dplyr::across(dplyr::all_of(present_flags), ~round(mean(.x, na.rm = TRUE), 1)),
        .groups = "drop"
      ) %>%
      dplyr::arrange(p_adj)

    if (!is.null(le_summary)) {
      le_rk_summary <- le_summary %>% dplyr::filter(ranking == rk) %>% dplyr::select(-ranking)
      rk_sig <- rk_sig %>% dplyr::left_join(le_rk_summary, by = c("database", "pathway"))
    }

    if (has_nes) {
      le_df <- rk_sig %>%
        transmute(Database = database, Pathway = pathway, Description = description,
                  NES = round(perm_nes, 3), Direction = direction,
                  `p.adj` = signif(p_adj, 3), `N cutoffs` = n_occurrences, Cutoffs)
    } else {
      le_df <- rk_sig %>%
        transmute(Database = database, Pathway = pathway, Description = description,
                  Fold = round(fold_enrichment, 3), Direction = direction,
                  `p.adj` = signif(p_adj, 3), `N cutoffs` = n_occurrences, Cutoffs)
    }
    if (length(present_flags) > 0) {
      flag_df <- rk_sig %>% dplyr::select(dplyr::all_of(present_flags))
      names(flag_df) <- paste0("%", sub("^pct_", "", present_flags))
      le_df <- dplyr::bind_cols(le_df, flag_df)
    }
    esc_cols <- TRUE
    if ("Genes/positions" %in% names(rk_sig)) {
      le_df <- dplyr::bind_cols(le_df, rk_sig %>% dplyr::select(`N genes`, `Genes/positions`))
      .leading_edge_summary_rows[[rk]] <- le_df %>% dplyr::mutate(Ranking = rk, .before = 1)
      # posenrich_leading_edge_summary.tsv keeps the full member list; the table shows
      # the first few entries and a <details> toggle for the rest, since a row can carry
      # hundreds of gene:positions.
      le_df[["Genes/positions"]] <- truncate_gene_list(le_df[["Genes/positions"]])
      # Column indices of DT's `escape` are 1-based over the data columns when
      # rownames = FALSE (no offset). A wrong index makes DT:::escapeData() fail with
      # "attempt to select less than one element in integerOneIndex" and aborts the render.
      esc_cols <- which(names(le_df) != "Genes/positions")
    } else {
      .leading_edge_summary_rows[[rk]] <- le_df %>% dplyr::mutate(Ranking = rk, .before = 1)
    }

    id <- paste("dt_le", rk, sep = "_")
    assign(paste0(".esc_cols_", rk), esc_cols, envir = environment())
    cat(knitr::knit_child(text = c(
      paste0("```{r ", id, ", echo=FALSE}"),
      sprintf("DT::datatable(le_df, options = list(pageLength = 10, scrollX = TRUE), rownames = FALSE, escape = .esc_cols_%s)", rk),
      "```"), envir = environment(), quiet = TRUE))
    cat("\n\n")
  }
}
if (length(.leading_edge_summary_rows) > 0) {
  readr::write_tsv(dplyr::bind_rows(.leading_edge_summary_rows), "posenrich_leading_edge_summary.tsv")
}
```

# Term recurrence across direction lists

Every significant term/pathway, pooled across the three direction lists
(**GLOBAL**/**TOP**/**BOTTOM**) into one table - not one table per direction.
Recurrence *within* a single direction's own nested percentile cutoffs
(`top25` ⊃ `top10` ⊃ `top5` ⊃ `top1`) isn't an independent signal: a term is
mechanically less likely to survive the tighter, smaller foreground, so
"recovered at fewer cutoffs" mostly just restates the nesting rather than
telling you anything new. Recurrence *across* the three direction lists is
the meaningful comparison - a term significant under both **Top** and
**Bottom** (or all three) is corroborated by genuinely independent
foreground constructions, not by a subset relationship.

`N_lists`/`Lists` say how many of, and which of, GLOBAL/TOP/BOTTOM recovered
the term (at any of its own percentile cutoffs) - no other column, since
nothing else is comparable across three direction lists built from different
score signs. Underneath, a **term x direction-list** heatmap makes the same
recurrence pattern visually scannable.

```{r module-rollup-setup, include=FALSE}
# .normalize_term, .term_contains and .cluster_terms are defined in the
# term-clustering-setup chunk (above the Overall dotplot), so that the Overall dotplot,
# the Direction triangle and the Recurrence Table and Heatmap merge near-duplicate
# descriptions identically.

# Term x direction-list heatmap: one tile per (canonical Term, GLOBAL/TOP/BOTTOM) under
# which the term is recovered. Recurrence across the cutoffs of one direction is not an
# independent signal (top1 is a subset of top5, of top10 and of top25; see the section
# introduction). It takes `rec_pairs` after clustering (canonical `Term`, one row per
# database, pathway and list), so its merge is the one of the Recurrence Table.
.render_posenrich_list_heatmap <- function(rec_pairs) {
  if (nrow(rec_pairs) == 0) { cat("_No terms to show._\n\n"); return(invisible(NULL)) }
  pres <- rec_pairs %>% dplyr::distinct(Term, list_name)

  list_levels <- intersect(c("GLOBAL", "TOP", "BOTTOM"), unique(pres$list_name))

  # Term order: most lists first. A term merged from several pathways can recur across
  # more lists than any one of its pathways.
  term_counts <- pres %>% dplyr::count(Term, name = "n_lists") %>%
    dplyr::arrange(dplyr::desc(n_lists), Term)
  term_order <- term_counts$Term

  pres <- pres %>%
    dplyr::mutate(list_name = factor(list_name, levels = list_levels),
                  Term = factor(Term, levels = rev(term_order)))
  n_terms <- length(term_order)
  p <- ggplot(pres, aes(x = list_name, y = Term)) +
    geom_tile(fill = "#2c7fb8", colour = "white", linewidth = 0.6) +
    scale_x_discrete(drop = FALSE) +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 0),
          axis.text.y = element_text(size = pmax(6, pmin(10, 260 / max(n_terms, 1))))) +
    labs(title = "Recovered terms by direction list", x = "direction list", y = NULL)
  print(p)
  cat("\n\n")
}
```

```{r module-rollup, echo=FALSE, results='asis'}
if (!has_res || !any(res$sig, na.rm = TRUE)) {
  cat("_No significant sets._\n\n")
} else {
  sig_res <- res %>% filter(sig)

  # One row per (database, pathway, direction list that recovers it), the unit before the
  # similarity clustering, pooled over the percentile cutoffs of the list. The clustering
  # below unions a term split across databases or near-duplicate GMT entries into one row.
  rec_pairs <- sig_res %>%
    dplyr::mutate(description = dplyr::coalesce(description, pathway),
                  list_name = toupper(ranking)) %>%
    dplyr::distinct(database, pathway, description, list_name)

  if (nrow(rec_pairs) == 0) {
    cat("_No significant sets._\n\n")
  } else {
    # The cluster map (near-duplicate or nested description to canonical label) drives
    # the merge: (database, pathway) pairs that map to one Term become a single row whose
    # Database, Lists and N_lists are pooled over the merge.
    desc_levels <- unique(rec_pairs$description)
    cluster_map <- setNames(.cluster_terms(desc_levels), desc_levels)
    rec_pairs <- rec_pairs %>% dplyr::mutate(Term = unname(cluster_map[description]))

    disp <- rec_pairs %>%
      dplyr::group_by(Term) %>%
      dplyr::summarise(
        Database = paste(sort(unique(database)), collapse = ", "),
        Lists = paste(sort(unique(list_name)), collapse = ", "),
        N_lists = dplyr::n_distinct(list_name),
        .groups = "drop"
      ) %>%
      dplyr::arrange(dplyr::desc(N_lists), Term) %>%
      dplyr::select(Database, Term, N_lists, Lists)

    cat("\n\n## Recurrence Table\n\n")
    cat(knitr::knit_child(text = c(
      "```{r dt_rollup_all, echo=FALSE}",
      "DT::datatable(disp, options = list(pageLength = 10, scrollX = TRUE), rownames = FALSE)",
      "```"), envir = environment(), quiet = TRUE))
    cat("\n\n")

    cat("\n\n## Recurrence Heatmap\n\n")
    .render_posenrich_list_heatmap(rec_pairs)
  }
}
```

---

```{r poschar-setup, include=FALSE}
# ── Position characterization inputs ──────────────────────────────────────────
# TRUE when the path is a non-empty string of an existing file.
.file_ok <- function(p) !is.null(p) && !is.na(p) && nchar(p) > 0 && file.exists(p)

# Tableau 10 palette, shared with the other reports.
if (requireNamespace("paletteer", quietly = TRUE)) {
  .tableau10 <- as.character(paletteer::paletteer_d("ggthemes::Tableau_10"))
} else {
  .tableau10 <- c(
    "#4E79A7", "#F28E2B", "#E15759", "#76B7B2", "#59A14F",
    "#EDC948", "#B07AA1", "#FF9DA7", "#9C755F", "#BAB0AC"
  )
}

has_position_scores <- .file_ok(params$position_scores_file)
has_gene_scores      <- .file_ok(params$gene_scores_file)
pos_df  <- if (has_position_scores) readr::read_tsv(params$position_scores_file, show_col_types = FALSE) else NULL
gene_df <- if (has_gene_scores)     readr::read_tsv(params$gene_scores_file,     show_col_types = FALSE) else NULL

has_vep_primateai <- .file_ok(params$vep_primateai_file)
has_vep_cosmic    <- .file_ok(params$vep_cosmic_file)
has_genomic       <- .file_ok(params$genomic_info_file)
# has_fade_sites_top/bottom and fade_sites_top_df/bottom_df are loaded in the `load` chunk.
# Position scores are required by every sub-tab below (PAI3D, COSMIC and FADE joins,
# regional accumulation), so they gate the whole section together with at least one
# external layer.
has_pos_char <- has_position_scores &&
  (has_vep_primateai || has_vep_cosmic || has_genomic || has_fade_sites_top || has_fade_sites_bottom)

if (has_vep_primateai) {
  vep_pai_df <- readr::read_tsv(params$vep_primateai_file, show_col_types = FALSE)
  if ("Position" %in% names(vep_pai_df)) vep_pai_df$Position <- suppressWarnings(as.integer(vep_pai_df$Position))
} else {
  vep_pai_df <- NULL
}

if (has_vep_cosmic) {
  vep_cosmic_df <- readr::read_tsv(params$vep_cosmic_file, show_col_types = FALSE)
  if ("Position" %in% names(vep_cosmic_df)) vep_cosmic_df$Position <- suppressWarnings(as.integer(vep_cosmic_df$Position))
} else {
  vep_cosmic_df <- NULL
}

if (has_genomic) {
  # Columns: gene, chr, start, end, length; a header line is dropped when present.
  genomic_raw <- readr::read_tsv(params$genomic_info_file,
                          col_names = c("Gene", "chr", "gene_start", "gene_end", "gene_length"),
                          show_col_types = FALSE)
  if (tolower(genomic_raw$Gene[1]) == "gene") genomic_raw <- genomic_raw[-1, ]
  genomic_raw <- genomic_raw %>%
    dplyr::mutate(
      gene_start  = as.integer(gene_start),
      gene_end    = as.integer(gene_end),
      gene_length = as.integer(gene_length),
      gene_mid    = as.integer((gene_start + gene_end) / 2L)
    )
} else {
  genomic_raw <- NULL
}

# Scored positions are joined to genomic coordinates through their gene: each position
# inherits the genomic midpoint of its gene as x coordinate, shifted by chromosome-
# cumulative offsets (plus a 5 Mb gap) for the regional accumulation plot.
if (has_genomic && has_position_scores) {
  pos_genomic <- pos_df %>%
    dplyr::inner_join(genomic_raw, by = "Gene") %>%
    dplyr::mutate(
      chr_label = sub("^chr", "", chr),
      chr_num   = suppressWarnings(as.integer(chr_label)),
      chr_order = dplyr::case_when(
        !is.na(chr_num)              ~ chr_num,
        chr_label == "X"             ~ 23L,
        chr_label == "Y"             ~ 24L,
        chr_label %in% c("MT", "M")  ~ 25L,
        TRUE                         ~ 26L
      )
    ) %>%
    dplyr::arrange(chr_order, gene_mid)

  pos_chr_offsets <- pos_genomic %>%
    dplyr::group_by(chr_order, chr_label) %>%
    dplyr::summarise(chr_max = max(gene_end, na.rm = TRUE), .groups = "drop") %>%
    dplyr::arrange(chr_order) %>%
    dplyr::mutate(offset = cumsum(dplyr::lag(chr_max, default = 0L)) + (chr_order - 1L) * 5e6L)

  pos_genomic <- pos_genomic %>%
    dplyr::left_join(pos_chr_offsets %>% dplyr::select(chr_order, offset), by = "chr_order") %>%
    dplyr::mutate(
      manhattan_x = gene_mid + offset,
      chr_even    = chr_order %% 2 == 0
    )
} else {
  pos_genomic <- NULL
  pos_chr_offsets <- NULL
}

if (has_pos_char) {
  top_pct  <- params$top_pct %||% 0.10
  top5_pct <- 0.05
  top1_pct <- 0.01

  pos_top    <- pos_df %>% dplyr::filter(side == "top")
  pos_bottom <- pos_df %>% dplyr::filter(side == "bottom")

  # Percentile membership comes from the position_lists/slice_<dir><pct>.tsv files
  # published by scoring_compute.R when position_lists_dir is given, so that the
  # reports agree on the top percentiles by construction. Otherwise it is a quantile()
  # cut on the population `pop` of the direction. .pos_pct_test() returns a closure
  # because the same membership set or threshold must be applied to different row
  # subsets of pos_df (pos_score_meta, pos_genomic_top/bot), not computed once and
  # reused by position.
  .pos_slice_ids <- function(dir_key, pct) {
    if (is.null(params$position_lists_dir)) return(NULL)
    f <- file.path(params$position_lists_dir, sprintf("slice_%s%d.tsv", dir_key, pct))
    if (!file.exists(f)) return(NULL)
    d <- tryCatch(readr::read_tsv(f, show_col_types = FALSE), error = function(e) NULL)
    if (is.null(d) || !all(c("Gene", "Position") %in% names(d))) return(NULL)
    unique(paste(d$Gene, d$Position))
  }
  .pos_pct_test <- function(dir_key, pct_frac, pop) {
    ids_set <- .pos_slice_ids(dir_key, round(pct_frac * 100))
    if (!is.null(ids_set)) {
      function(df) paste(df$Gene, df$Position) %in% ids_set
    } else {
      thr <- quantile(pop$CAAS_score, 1 - pct_frac, na.rm = TRUE)
      function(df) df$CAAS_score >= thr
    }
  }

  # top25_pct is used only by the Fisher odds-ratio tables of PAI3D, COSMIC and FADE
  # (rows 100, 25, 10, 5 and 1%); every row except the 100% baseline comes from the
  # published slices. The in_top_Xpct flags cover 10, 5 and 1% only.
  top25_pct <- 0.25

  test_glob_25 <- .pos_pct_test("global", top25_pct, pos_df)
  test_glob_10 <- .pos_pct_test("global", top_pct,  pos_df)
  test_glob_5  <- .pos_pct_test("global", top5_pct, pos_df)
  test_glob_1  <- .pos_pct_test("global", top1_pct, pos_df)
  test_top_25  <- .pos_pct_test("top",    top25_pct, pos_top)
  test_top_10  <- .pos_pct_test("top",    top_pct,  pos_top)
  test_top_5   <- .pos_pct_test("top",    top5_pct, pos_top)
  test_top_1   <- .pos_pct_test("top",    top1_pct, pos_top)
  test_bot_25  <- .pos_pct_test("bottom", top25_pct, pos_bottom)
  test_bot_10  <- .pos_pct_test("bottom", top_pct,  pos_bottom)
  test_bot_5   <- .pos_pct_test("bottom", top5_pct, pos_bottom)
  test_bot_1   <- .pos_pct_test("bottom", top1_pct, pos_bottom)

  # thr_10: scalar for the reference line of the scatter plot, always a plain quantile
  # (a line has no equivalent of a membership set), whether or not the tests above use
  # the published slices.
  thr_10 <- quantile(pos_df$CAAS_score, 1 - top_pct, na.rm = TRUE)
}
```
---

```{r pos-char-header, results='asis'}
if (has_pos_char) cat(r"(# Position Characterisation {.tabset #position-characterisation}

This section examines how ranked positions behave in relation to external
signals that do **not** contribute to the `CAAS_score`. PrimateAI-3D, COSMIC
Mutant Census, and FADE branch-site selection evidence are treated as
independent characterisation layers.

PrimateAI-3D and COSMIC both describe specific amino acid transitions, so the
biochemical Grantham Score (GS) groupings are out of scope for those two (as
they group transitions by biochemical radicalness rather than modeling the
specific substitution directly). We restrict those two layers strictly to the
**Unweighted Scheme (US)**, with all scores and mutation overlaps derived
solely from US transitions. FADE evidence is not transition-specific (it is a
per-site test of dN/dS shift along phenotype-associated branches) and so is
not restricted to the US scheme.

- **`pai_char_score`**: The pathogenicity score of mutating the ancestral human reference amino acid to the derived US alternative amino acid(s) in PrimateAI-3D.
- **COSMIC Overlap**: The presence and recurrence of somatic mutations mapped to the specific codon under the US transition.
- **FADE Overlap**: Whether the position falls within the direction-specific FADE_top_sig/FADE_bottom_sig significant site set (BF ≥ `fade_bf_threshold`), and its maximum Bayes Factor.

)")
```

```{r vep-setup-common}
# The tests are evaluated on pos_df (they need Gene, Position and CAAS_score). select()
# below only drops columns and never reorders or filters rows, so the vectors stay
# row-aligned with pos_df when assigned to pos_score_meta.
.glob10 <- test_glob_10(pos_df); .glob5 <- test_glob_5(pos_df); .glob1 <- test_glob_1(pos_df)

pos_score_meta <- pos_df %>%
  select(Gene, Position, CAAS_score, n_schemes, side,
         any_of(c("n_hypotheses", "participating_hypotheses", "supporting_hypotheses"))) %>%
  mutate(
    caas_decile  = dplyr::ntile(CAAS_score, 10),
    in_top_10pct = .glob10,
    in_top_5pct  = .glob5,
    in_top_1pct  = .glob1
  )
if (!"supporting_hypotheses" %in% names(pos_score_meta) && "participating_hypotheses" %in% names(pos_score_meta)) {
  pos_score_meta$supporting_hypotheses <- pos_score_meta$participating_hypotheses
}

has_background <- .file_ok(params$universe_file)
if (has_background) {
  bg_lines <- readLines(params$universe_file)
  bg_genes <- bg_lines[nzchar(trimws(bg_lines)) & !grepl("^#", bg_lines)]
} else {
  bg_genes <- unique(gene_df$Gene)
}
```

```{r vep-pai-subsec, results='asis'}
if (has_vep_primateai) cat("## VEP: PrimateAI-3D\n\n")
```

```{r vep-pai-setup, eval=has_vep_primateai}
# The PrimateAI score column name varies with the database version.
.pai_score_col <- intersect(
  c("score_PAI3D", "score", "primateai_score", "pathogenicity_score", "PrimateAI_score"),
  names(vep_pai_df)
)[1]
if (is.na(.pai_score_col)) {
  warning("PrimateAI: no recognised score column found - skipping.")
  has_vep_primateai <- FALSE
}
```

```{r vep-pai-join, eval=has_vep_primateai}
# Restrict to the genes present in both the background and the database.
shared_genes_pai <- intersect(bg_genes, unique(vep_pai_df$Gene))
pos_pai_shared <- pos_score_meta %>% filter(Gene %in% shared_genes_pai)

# The layer is restricted to the US scheme, so pai_char_score is the pathogenicity score
# of the US change; with several alternative amino acids the maximum is kept. Pathogenic
# status comes only from the prediction column (containing "pathogenic", case-insensitive),
# never from a score cutoff.
pai_by_prot <- vep_pai_df %>%
  filter(caap_group == "US", Gene %in% shared_genes_pai) %>%
  mutate(
    pai_score = as.numeric(.data[[.pai_score_col]]),
    is_pathogenic = grepl("pathogenic", prediction, ignore.case = TRUE)
  ) %>%
  filter(is.finite(pai_score)) %>%
  group_by(Gene, Position) %>%
  summarise(
    pai_char_score = max(pai_score, na.rm = TRUE),
    pai_pathogenic = any(is_pathogenic, na.rm = TRUE),
    n_pai_schemes  = 1L,
    .groups = "drop"
  )

pos_vep_pai <- pai_by_prot %>%
  inner_join(pos_pai_shared, by = c("Gene", "Position")) %>%
  distinct(Gene, Position, .keep_all = TRUE)

# Keep caas_change and the other details for the display table.
pos_vep_pai_table_source <- vep_pai_df %>%
  filter(caap_group == "US", Gene %in% shared_genes_pai) %>%
  inner_join(pos_pai_shared, by = c("Gene", "Position")) %>%
  distinct(Gene, Position, .keep_all = TRUE)

has_vep_primateai <- nrow(pos_vep_pai) > 0

# Directional PAI subsets (side top or bottom).
pos_vep_pai_top    <- pos_vep_pai %>% filter(side == "top")
pos_vep_pai_bottom <- pos_vep_pai %>% filter(side == "bottom")
```

`r if (exists("pos_vep_pai") && nrow(pos_vep_pai) == 0) "> **Note:** PrimateAI-3D inputs were provided, but no scored positions could be matched to the VEP/PAI3D annotations for this direction."`

```{r vep-pai-violin, eval=has_vep_primateai, fig.height=7}
# pai_char_score violin per CAAS decile, split by direction (TOP and BOTTOM separately).
pai_dir_df <- bind_rows(
  pos_vep_pai_top    %>% mutate(pai_direction = "TOP (top + both)"),
  pos_vep_pai_bottom %>% mutate(pai_direction = "BOTTOM (bottom + both)")
) %>%
  filter(!is.na(caas_decile), !is.na(pai_char_score))

# Median per direction, for the reference lines.
pai_medians <- pai_dir_df %>%
  group_by(pai_direction) %>%
  summarise(median_pai = median(pai_char_score, na.rm = TRUE), .groups = "drop")

ggplot(pai_dir_df, aes(x = factor(caas_decile, levels = 1:10), y = pai_char_score)) +
  geom_violin(aes(fill = pai_direction), alpha = 0.55, trim = TRUE, scale = "width",
              show.legend = FALSE) +
  geom_boxplot(width = 0.14, outlier.size = 0.4, alpha = 0.9,
               fill = "white", colour = "grey30") +
  geom_hline(data = pai_medians,
             aes(yintercept = median_pai),
             linetype = "dashed", colour = "grey20", linewidth = 0.7) +
  geom_text(data = pai_medians,
            aes(x = 10.6, y = median_pai,
                label = sprintf("median = %.3f", median_pai)),
            hjust = 0, vjust = -0.4, size = 3, colour = "grey20",
            inherit.aes = FALSE) +
  scale_fill_manual(values = c("TOP (top + both)" = .tableau10[1],
                               "BOTTOM (bottom + both)" = .tableau10[3])) +
  scale_x_discrete(
    drop   = FALSE,
    labels = setNames(c("1\n(low)", as.character(2:9), "10\n(high)"), as.character(1:10))
  ) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.08))) +
  facet_wrap(~pai_direction, ncol = 1) +
  labs(
    title    = "PAI_TOP and PAI_BOTTOM: PrimateAI-3D Score by CAAS Decile",
    subtitle = sprintf(
      "PAI_TOP: %d positions | PAI_BOTTOM: %d positions | dashed = per-direction median",
      nrow(pos_vep_pai_top), nrow(pos_vep_pai_bottom)
    ),
    x = "CAAS Score Decile", y = "PrimateAI-3D Score"
  ) +
  theme(plot.title = element_text(face = "bold"),
        strip.text = element_text(face = "bold", size = 11))
```

```{r vep-pai-scatter, eval=has_vep_primateai, fig.height=5}
# CAAS_score against pai_char_score, colored by direction.
#
# The reported r is a Pearson correlation of the within-run percentile ranks (rank / n)
# of both axes, i.e. a Spearman-type rank correlation, and its magnitude is not
# comparable to an r on raw scores. Ranks are used because the resampling statistics
# are floor-limited at 1/N, so raw values tie at the floor. The same caveat applies to
# every "(rank-norm)" subtitle: values are interpretable only relative to the other
# positions of this run, not across runs or traits.
pos_vep_pai_dir <- bind_rows(
  pos_vep_pai_top    %>% mutate(pai_direction = "TOP"),
  pos_vep_pai_bottom %>% mutate(pai_direction = "BOTTOM")
) %>% distinct(Gene, Position, .keep_all = TRUE)

.pai_caas_r_top  <- safe_cor(
  rank(pos_vep_pai_top$CAAS_score,    na.last = "keep") / sum(!is.na(pos_vep_pai_top$CAAS_score)),
  rank(pos_vep_pai_top$pai_char_score, na.last = "keep") / sum(!is.na(pos_vep_pai_top$pai_char_score))
)
.pai_caas_r_bot  <- safe_cor(
  rank(pos_vep_pai_bottom$CAAS_score,    na.last = "keep") / sum(!is.na(pos_vep_pai_bottom$CAAS_score)),
  rank(pos_vep_pai_bottom$pai_char_score, na.last = "keep") / sum(!is.na(pos_vep_pai_bottom$pai_char_score))
)

ggplot(pos_vep_pai_dir, aes(x = CAAS_score, y = pai_char_score,
                             colour = pai_direction)) +
  geom_point(alpha = 0.35, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8, alpha = 0.15) +
  geom_vline(xintercept = thr_10, linetype = "dashed",
             colour = "grey40", linewidth = 0.5) +
  scale_colour_manual(values = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[3]),
                      name = "Direction") +
  labs(
    title    = "CAAS Score vs PrimateAI-3D Score by Direction",
    subtitle = sprintf("Pearson r (rank-norm): TOP = %.3f | BOTTOM = %.3f | dashed = top-10%% CAAS",
                       .pai_caas_r_top, .pai_caas_r_bot),
    x = "CAAS Score", y = "PrimateAI-3D Score"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r vep-pai-pathogenic-bar, eval=has_vep_primateai, fig.height=5}
# Pathogenic and non-pathogenic positions per direction.
pai_pathogenic_counts <- bind_rows(
  pos_vep_pai_top    %>% mutate(pai_direction = "TOP"),
  pos_vep_pai_bottom %>% mutate(pai_direction = "BOTTOM")
) %>%
  mutate(pathogenic = ifelse(pai_pathogenic, "pathogenic", "non-pathogenic")) %>%
  count(pai_direction, pathogenic)

ggplot(pai_pathogenic_counts, aes(x = pathogenic, y = n, fill = pai_direction)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = n), position = position_dodge(width = 0.8), vjust = -0.3, size = 3.5) +
  scale_fill_manual(values = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[3])) +
  labs(
    title    = "Pathogenic vs Non-Pathogenic Positions (PrimateAI-3D)",
    subtitle = "Pathogenic defined by prediction column",
    x = NULL, y = "N positions", fill = "Direction"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r vep-pai-table, eval=has_vep_primateai}
pai_table <- pos_vep_pai_table_source %>%
  arrange(desc(CAAS_score)) %>%
  select(Gene, Position, hg38_ref_aa, caas_alt_aas, caas_change, CAAS_score, side, score_PAI3D, percentile_PAI3D, prediction) %>%
  rename(
    `Ref AA` = hg38_ref_aa,
    `Alt AA(s)` = caas_alt_aas,
    `CAAS Change` = caas_change,
    `CAAS Score` = CAAS_score,
    `Change Side` = side,
    `PAI3D Score` = score_PAI3D,
    `PAI3D Percentile` = percentile_PAI3D,
    `Prediction` = prediction
  )

DT::datatable(
  pai_table,
  caption = "CAAS positions overlapping with PrimateAI-3D pathogenicity predictions, ordered by CAAS score",
  options = list(pageLength = 10, scrollX = TRUE),
  rownames = FALSE
)
```

```{r vep-pai-fisher, eval=has_vep_primateai}
# Odds ratio: enrichment of pathogenic residues (by the prediction column) inside the CAAS
# threshold relative to the other positions; a value above 1 is enrichment. P value:
# Fisher's exact test of a higher proportion of pathogenic residues inside the threshold.
# `N positions` is the population of the test (PAI-scored positions of shared genes), as
# in run_cosmic_fisher_dir below.
run_pai_fisher_dir <- function(df, label) {
  df   <- df %>% mutate(high_pai = !is.na(pai_pathogenic) & pai_pathogenic)
  n_all <- nrow(df)

  # Threshold rows: 100% is the whole population; 25, 10, 5 and 1% come from the
  # published position_lists slices (test_glob_25, in_top_Xpct).
  pct_rows <- list(
    list(col = NULL, lbl = "100% (all positions)",    flag = rep(TRUE, n_all)),
    list(col = NULL, lbl = "25% (top 25% CAAS)",      flag = test_glob_25(df)),
    list(col = "in_top_10pct", lbl = sprintf("Top %g%%", top_pct  * 100), flag = NULL),
    list(col = "in_top_5pct",  lbl = sprintf("Top %g%%", top5_pct * 100), flag = NULL),
    list(col = "in_top_1pct",  lbl = sprintf("Top %g%%", top1_pct * 100), flag = NULL)
  )

  purrr::map_dfr(pct_rows, function(thr) {
    in_top   <- if (!is.null(thr$flag)) thr$flag else df[[thr$col]]
    in_top   <- replace_na(in_top, FALSE)

    n_top    <- sum(in_top)
    n_top_h  <- sum(in_top & df$high_pai, na.rm = TRUE)
    n_rest   <- sum(!in_top)
    n_rest_h <- sum(!in_top & df$high_pai, na.rm = TRUE)
    # The columns are ordered [flag, not flag] so that the odds ratio of fisher.test()
    # is the odds of pathogenic inside the threshold over the odds outside it (the
    # reverse order would give its inverse).
    mat <- matrix(c(n_top_h, n_rest_h, n_top - n_top_h, n_rest - n_rest_h), nrow = 2)
    ft  <- tryCatch(fisher.test(mat),
                    error = function(e) list(estimate = NA, conf.int = c(NA, NA), p.value = NA))

    tibble(
      Direction            = label,
      Threshold            = thr$lbl,
      `N positions`        = n_top,
      `N pathogenic`       = n_top_h,
      `% pathogenic`       = sprintf("%.1f%%", 100 * n_top_h / max(n_top, 1)),
      `Odds ratio`         = sprintf("%.2f", ft$estimate),
      `P value`            = formatC(ft$p.value, format = "e", digits = 2)
    )
  })
}

pai_fisher <- bind_rows(
  run_pai_fisher_dir(pos_vep_pai_top,    "TOP"),
  run_pai_fisher_dir(pos_vep_pai_bottom, "BOTTOM")
)

knitr::kable(
  pai_fisher,
  caption = "PAI3D enrichment of pathogenic positions (defined by prediction column) by CAAS threshold and direction. N positions = the actual test population (shared-gene, PAI3D-scored positions) at that threshold. 100% row = all directional positions (baseline); 25/10/5/1% rows use SCORING's published position percentile slices; Fisher OR and P value compare top-subset vs remainder, entirely within PAI-covered positions.",
  align = c("l", "l", rep("r", ncol(pai_fisher) - 2))
)
```


```{r vep-cosmic-subsec, results='asis'}
if (has_vep_cosmic) cat("## VEP: COSMIC Mutant Census\n\n")
```

```{r vep-cosmic-setup, eval=has_vep_cosmic}
# Restrict to the genes present in both the background and the database.
shared_genes_cosmic <- intersect(bg_genes, unique(vep_cosmic_df$Gene))
pos_cosmic_shared <- pos_score_meta %>% filter(Gene %in% shared_genes_cosmic)

# US scheme only, when the table has caap_group.
if ("caap_group" %in% names(vep_cosmic_df)) {
  vep_cosmic_df <- vep_cosmic_df %>% filter(caap_group == "US", Gene %in% shared_genes_cosmic)
} else {
  vep_cosmic_df <- vep_cosmic_df %>% filter(Gene %in% shared_genes_cosmic)
}

# All rows are kept so that the recurrence of each position can be counted.
cosmic_joined <- vep_cosmic_df %>%
  inner_join(pos_cosmic_shared, by = c("Gene", "Position"))

has_vep_cosmic <- nrow(cosmic_joined) > 0
```

`r if (exists("cosmic_joined") && nrow(cosmic_joined) == 0) "> **Note:** COSMIC Mutant Census mutations were mapped, but no scored positions could be matched to the COSMIC annotations for this direction."`

```{r vep-cosmic-stats, eval=has_vep_cosmic}
cosmic_table <- cosmic_joined %>%
  group_by(Gene, Position, hg38_ref_aa, caas_alt_aas, caas_change, CAAS_score, side) %>%
  summarise(
    Recurrence = n(),
    `Mutation AA(s)` = paste(unique(MUTATION_AA), collapse = "; "),
    `Description(s)` = paste(unique(MUTATION_DESCRIPTION), collapse = "; "),
    `Somatic Status(es)` = paste(unique(MUTATION_SOMATIC_STATUS), collapse = "; "),
    .groups = "drop"
  ) %>%
  arrange(desc(Recurrence)) %>%
  rename(
    `Ref AA` = hg38_ref_aa,
    `Alt AA(s)` = caas_alt_aas,
    `CAAS Change` = caas_change,
    `CAAS Score` = CAAS_score,
    `Change Side` = side,
    `Mutation AA(s)` = `Mutation AA(s)`,
    `Description` = `Description(s)`,
    `Somatic Status` = `Somatic Status(es)`
  )

DT::datatable(
  cosmic_table,
  caption = "CAAS positions overlapping with COSMIC cancer-related mutations, ordered by mutation recurrence",
  options = list(pageLength = 10, scrollX = TRUE),
  rownames = FALSE
)
```

```{r vep-cosmic-somatic-bar, eval=has_vep_cosmic, fig.height=5}
# COSMIC mutations by somatic status and direction.
cosmic_counts <- bind_rows(
  cosmic_joined %>% filter(side == "top") %>% mutate(cosmic_direction = "TOP"),
  cosmic_joined %>% filter(side == "bottom") %>% mutate(cosmic_direction = "BOTTOM")
) %>%
  filter(!is.na(MUTATION_SOMATIC_STATUS) & MUTATION_SOMATIC_STATUS != "") %>%
  count(cosmic_direction, MUTATION_SOMATIC_STATUS)

ggplot(cosmic_counts, aes(x = MUTATION_SOMATIC_STATUS, y = n, fill = cosmic_direction)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = n), position = position_dodge(width = 0.8), vjust = -0.3, size = 3.5) +
  scale_fill_manual(values = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[3])) +
  labs(
    title    = "COSMIC Mutation Somatic Status Distribution",
    x = NULL, y = "N mutations", fill = "Direction"
  ) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r vep-cosmic-fisher, eval=has_vep_cosmic}
# Odds ratio: enrichment of positions with COSMIC somatic mutations inside the CAAS
# threshold relative to the other positions; a value above 1 is enrichment. P value:
# Fisher's exact test of a higher proportion of such positions inside the threshold.
pos_cosmic <- pos_cosmic_shared %>%
  mutate(has_cosmic = paste(Gene, Position, sep = "_") %in% paste(vep_cosmic_df$Gene, vep_cosmic_df$Position, sep = "_"))

pos_cosmic_top    <- pos_cosmic %>% filter(side == "top")
pos_cosmic_bottom <- pos_cosmic %>% filter(side == "bottom")

run_cosmic_fisher_dir <- function(df, label) {
  n_all <- nrow(df)
  pct_rows <- list(
    list(col = NULL, lbl = "100% (all positions)",    flag = rep(TRUE, n_all)),
    list(col = NULL, lbl = "25% (top 25% CAAS)",      flag = test_glob_25(df)),
    list(col = "in_top_10pct", lbl = sprintf("Top %g%%", top_pct  * 100), flag = NULL),
    list(col = "in_top_5pct",  lbl = sprintf("Top %g%%", top5_pct * 100), flag = NULL),
    list(col = "in_top_1pct",  lbl = sprintf("Top %g%%", top1_pct * 100), flag = NULL)
  )

  purrr::map_dfr(pct_rows, function(thr) {
    in_top   <- if (!is.null(thr$flag)) thr$flag else df[[thr$col]]
    in_top   <- replace_na(in_top, FALSE)
    n_top    <- sum(in_top)
    n_top_c  <- sum(in_top & df$has_cosmic, na.rm = TRUE)
    n_rest   <- sum(!in_top)
    n_rest_c <- sum(!in_top & df$has_cosmic, na.rm = TRUE)
    # The columns are ordered [flag, not flag] so that the odds ratio of fisher.test()
    # is the odds of COSMIC overlap inside the threshold over the odds outside it.
    mat <- matrix(c(n_top_c, n_rest_c, n_top - n_top_c, n_rest - n_rest_c), nrow = 2)
    ft  <- tryCatch(fisher.test(mat),
                    error = function(e) list(estimate = NA, conf.int = c(NA, NA), p.value = NA))
    
    tibble(
      Direction            = label,
      Threshold            = thr$lbl,
      `N positions`        = n_top,
      `N COSMIC`           = n_top_c,
      `% COSMIC`           = sprintf("%.1f%%", 100 * n_top_c / max(n_top, 1)),
      `Odds ratio`         = sprintf("%.2f", ft$estimate),
      `P value`            = formatC(ft$p.value, format = "e", digits = 2)
    )
  })
}

cosmic_fisher <- bind_rows(
  run_cosmic_fisher_dir(pos_cosmic_top,    "TOP"),
  run_cosmic_fisher_dir(pos_cosmic_bottom, "BOTTOM")
)

knitr::kable(
  cosmic_fisher,
  caption = "Fisher exact test for COSMIC somatic mutations overlap enrichment within CAAS score thresholds",
  align = c("l", "l", rep("r", ncol(cosmic_fisher) - 2))
)
```


```{r fade-subsec, results='asis'}
if (has_fade_sites_top || has_fade_sites_bottom) cat("## FADE Site Overlap\n\n")
```

```{r fade-setup, eval=has_fade_sites_top || has_fade_sites_bottom}
# Direction-specific: top positions are checked only against fade_sites_top and bottom
# positions only against fade_sites_bottom (as the FADE_top_sig and FADE_bottom_sig
# layers of build_position_gmt.py). FADE covers the full analyzed gene set, unlike the
# partial external databases COSMIC and PAI3D, so no shared-gene restriction applies: the
# universe is the CAAS position set of the direction (pos_score_meta, with the
# in_top_Xpct flags).
pos_fade_top <- if (has_fade_sites_top) {
  pos_score_meta %>%
    dplyr::filter(side == "top") %>%
    dplyr::left_join(
      fade_sites_top_df %>% dplyr::select(Gene, Position, max_bf, target_aa),
      by = c("Gene", "Position")
    ) %>%
    dplyr::mutate(has_fade = !is.na(max_bf))
} else NULL

pos_fade_bottom <- if (has_fade_sites_bottom) {
  pos_score_meta %>%
    dplyr::filter(side == "bottom") %>%
    dplyr::left_join(
      fade_sites_bottom_df %>% dplyr::select(Gene, Position, max_bf, target_aa),
      by = c("Gene", "Position")
    ) %>%
    dplyr::mutate(has_fade = !is.na(max_bf))
} else NULL
```

`r if ((has_fade_sites_top && sum(pos_fade_top$has_fade) == 0) || (has_fade_sites_bottom && sum(pos_fade_bottom$has_fade) == 0)) "> **Note:** FADE site evidence was provided, but no scored positions in at least one direction overlapped a FADE-significant site."`

```{r fade-table, eval=has_fade_sites_top || has_fade_sites_bottom}
fade_table <- dplyr::bind_rows(
  if (!is.null(pos_fade_top))    pos_fade_top    %>% dplyr::filter(has_fade) %>% dplyr::mutate(Direction = "TOP"),
  if (!is.null(pos_fade_bottom)) pos_fade_bottom %>% dplyr::filter(has_fade) %>% dplyr::mutate(Direction = "BOTTOM")
) %>%
  dplyr::arrange(dplyr::desc(max_bf)) %>%
  dplyr::select(Gene, Position, Direction, CAAS_score, dplyr::any_of(c("n_hypotheses", "supporting_hypotheses")), side, target_aa, max_bf) %>%
  dplyr::rename(
    `CAAS Score`        = CAAS_score,
    `N Hypotheses`      = dplyr::any_of("n_hypotheses"),
    `Hypotheses`        = dplyr::any_of("supporting_hypotheses"),
    `Change Side`       = side,
    `FADE Target AA`    = target_aa,
    `Max Bayes Factor`   = max_bf
  )

DT::datatable(
  fade_table,
  caption = "CAAS positions overlapping with FADE-significant sites (BF >= fade_bf_threshold), ordered by Bayes Factor",
  options = list(pageLength = 10, scrollX = TRUE),
  rownames = FALSE
)
```

```{r fade-bf-violin, eval=has_fade_sites_top || has_fade_sites_bottom, fig.height=5}
fade_bf_df <- dplyr::bind_rows(
  if (!is.null(pos_fade_top))    pos_fade_top    %>% dplyr::filter(has_fade) %>% dplyr::mutate(fade_direction = "TOP"),
  if (!is.null(pos_fade_bottom)) pos_fade_bottom %>% dplyr::filter(has_fade) %>% dplyr::mutate(fade_direction = "BOTTOM")
)

ggplot(fade_bf_df, aes(x = fade_direction, y = max_bf, fill = fade_direction)) +
  geom_violin(alpha = 0.55, trim = TRUE, scale = "width", show.legend = FALSE) +
  geom_boxplot(width = 0.14, outlier.size = 0.4, alpha = 0.9, fill = "white", colour = "grey30") +
  scale_fill_manual(values = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[3])) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.08))) +
  labs(
    title    = "Bayes Factor of FADE-significant Positions by Direction",
    subtitle = sprintf("TOP: %d sites | BOTTOM: %d sites",
                       sum(fade_bf_df$fade_direction == "TOP"), sum(fade_bf_df$fade_direction == "BOTTOM")),
    x = NULL, y = "Max Bayes Factor"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r fade-fisher, eval=has_fade_sites_top || has_fade_sites_bottom}
# Odds ratio: enrichment of FADE-significant sites (BF >= fade_bf_threshold) inside the
# CAAS threshold relative to the other positions of the direction; a value above 1 is
# enrichment. P value: Fisher's exact test of a higher proportion of such sites inside
# the threshold.
run_fade_fisher_dir <- function(df, label) {
  n_all <- nrow(df)
  pct_rows <- list(
    list(col = NULL, lbl = "100% (all positions)",    flag = rep(TRUE, n_all)),
    list(col = NULL, lbl = "25% (top 25% CAAS)",      flag = test_glob_25(df)),
    list(col = "in_top_10pct", lbl = sprintf("Top %g%%", top_pct  * 100), flag = NULL),
    list(col = "in_top_5pct",  lbl = sprintf("Top %g%%", top5_pct * 100), flag = NULL),
    list(col = "in_top_1pct",  lbl = sprintf("Top %g%%", top1_pct * 100), flag = NULL)
  )

  purrr::map_dfr(pct_rows, function(thr) {
    in_top   <- if (!is.null(thr$flag)) thr$flag else df[[thr$col]]
    in_top   <- replace_na(in_top, FALSE)
    n_top    <- sum(in_top)
    n_top_f  <- sum(in_top & df$has_fade, na.rm = TRUE)
    n_rest   <- sum(!in_top)
    n_rest_f <- sum(!in_top & df$has_fade, na.rm = TRUE)
    # The columns are ordered [flag, not flag] so that the odds ratio of fisher.test()
    # is the odds of a FADE-significant site inside the threshold over the odds outside it.
    mat <- matrix(c(n_top_f, n_rest_f, n_top - n_top_f, n_rest - n_rest_f), nrow = 2)
    ft  <- tryCatch(fisher.test(mat),
                    error = function(e) list(estimate = NA, conf.int = c(NA, NA), p.value = NA))

    tibble(
      Direction     = label,
      Threshold     = thr$lbl,
      `N positions` = n_top,
      `N FADE-sig`  = n_top_f,
      `% FADE-sig`  = sprintf("%.1f%%", 100 * n_top_f / max(n_top, 1)),
      `Odds ratio`  = sprintf("%.2f", ft$estimate),
      `P value`     = formatC(ft$p.value, format = "e", digits = 2)
    )
  })
}

fade_fisher <- dplyr::bind_rows(
  if (!is.null(pos_fade_top))    run_fade_fisher_dir(pos_fade_top,    "TOP"),
  if (!is.null(pos_fade_bottom)) run_fade_fisher_dir(pos_fade_bottom, "BOTTOM")
)

knitr::kable(
  fade_fisher,
  caption = "Fisher exact test for FADE-significant site overlap enrichment within CAAS score thresholds. Direction-specific: TOP positions tested only against fade_sites_top, BOTTOM only against fade_sites_bottom.",
  align = c("l", "l", rep("r", ncol(fade_fisher) - 2))
)
```

```{r reg-accum-subsec, results='asis'}
if (has_genomic) cat("## Regional Accumulation\n\n")
```

```{r position-region-accum, eval=has_genomic, fig.height=10, fig.width=14}
# Window summaries per direction. test10, test5 and test1 are the closures test_top_* or
# test_bot_* (published slices, quantile() fallback). pos_sub is pos_genomic_top or
# pos_genomic_bot, a different row subset than pos_top and pos_bottom (which the closures
# were built on), so the closures are evaluated on its own rows.
.build_region_wins <- function(pos_sub, test10, test5, test1, chr_offsets_df, dir_label) {
  pos_sub %>%
    mutate(
      win_id    = floor(gene_mid / params$window_size_bp),
      win_start = win_id * params$window_size_bp,
      win_end   = win_start + params$window_size_bp - 1L,
      win_mid   = win_start + params$window_size_bp / 2L,
      in_top10  = test10(pos_sub),
      in_top5   = test5(pos_sub),
      in_top1   = test1(pos_sub)
    ) %>%
    group_by(chr_order, chr_label, win_id, win_start, win_end, win_mid) %>%
    summarise(
      n_positions        = n(),
      n_top10_positions  = sum(in_top10, na.rm = TRUE),
      n_top5_positions   = sum(in_top5,  na.rm = TRUE),
      n_top1_positions   = sum(in_top1,  na.rm = TRUE),
      frac_top_positions = n_top10_positions / n_positions,
      mean_caas          = mean(CAAS_score, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    left_join(chr_offsets_df %>% select(chr_order, offset), by = "chr_order") %>%
    mutate(
      direction    = dir_label,
      win_x        = win_mid + offset,
      region_label = sprintf("chr%s:%0.1f-%0.1f Mb",
                             chr_label, win_start / 1e6, win_end / 1e6)
    )
}

# pos_genomic carries chr_order and gene_mid; split by direction.
pos_genomic_top <- pos_genomic %>% filter(side == "top")
pos_genomic_bot <- pos_genomic %>% filter(side == "bottom")

pos_region_top <- .build_region_wins(pos_genomic_top, test_top_10, test_top_5, test_top_1,
                                      pos_chr_offsets, "TOP")
pos_region_bot <- .build_region_wins(pos_genomic_bot, test_bot_10, test_bot_5, test_bot_1,
                                      pos_chr_offsets, "BOTTOM")
```

```{r position-region-accum-table, eval=has_genomic}
.make_region_tbl <- function(wins_df) {
  wins_df %>%
    filter(n_top10_positions > 0) %>%
    arrange(desc(n_positions), desc(frac_top_positions)) %>%
    slice_head(n = 30) %>%
    mutate(
      frac_top10 = sprintf("%.1f%%", 100 * n_top10_positions / n_positions),
      frac_top5  = sprintf("%.1f%%", 100 * n_top5_positions  / n_positions),
      frac_top1  = sprintf("%.1f%%", 100 * n_top1_positions  / n_positions)
    ) %>%
    select(
      Region              = region_label,
      `N positions`       = n_positions,
      `N top-10%`         = n_top10_positions,
      `% top-10%`         = frac_top10,
      `N top-5%`          = n_top5_positions,
      `% top-5%`          = frac_top5,
      `N top-1%`          = n_top1_positions,
      `% top-1%`          = frac_top1,
      `Mean CAAS`         = mean_caas
    ) %>%
    mutate(`Mean CAAS` = round(`Mean CAAS`, 4))
}

knitr::kable(.make_region_tbl(pos_region_top),
  caption = "TOP-direction: genomic windows with high-CAAS positions, ordered by N positions (top 30; direction-specific thresholds)")
knitr::kable(.make_region_tbl(pos_region_bot),
  caption = "BOTTOM-direction: genomic windows with high-CAAS positions, ordered by N positions (top 30; direction-specific thresholds)")

# Export of the regional accumulation tables (regional_accumulation.tsv).
region_export <- bind_rows(
  .make_region_tbl(pos_region_top) %>% mutate(Direction = "TOP"),
  .make_region_tbl(pos_region_bot) %>% mutate(Direction = "BOTTOM")
) %>% select(Direction, everything())
readr::write_tsv(region_export,
  file.path(params$output_dir, "regional_accumulation.tsv"))
```

---

# Methods

- **Test**: Position-Level Path Sum Permulation test. Raw CAAS score magnitude
  is summed for every position in a set ($T_\text{obs} = \sum_{p \in \text{term}} s_p$),
  and compared against a null built from the real cycles of the CAAS
  permulation null (`perm_pos_cycle_caas.tsv.gz`) when supplied — same
  preference gene-level FCS's Path Sum Permulation test gives that null
  (`12.FCS_general_report.Rmd`). Without a CAAS null the null-based values
  (NES, p-values) are `NA` and no set is called significant; a label shuffle of
  the non-zero position scores runs only when requested (`--allow-label-shuffle`).
  Normalized Enrichment Score
  ($\text{NES} = (T_\text{obs} - \mu_\text{null}) / \sigma_\text{null}$) and
  empirical p-values ($p = \frac{\#\{T_\text{null} \ge T_\text{obs}\} + 1}{N_\text{draws} + 1}$)
  are computed per term against whichever null was used. Run separately for
  the **global / top / bottom** directions (`side`).
- **Two background conventions.** Every database above tests against the
  **full honest background** *except* `cosmic_orthogroups` and
  `pai3d_orthogroups`, which test against a **coverage-restricted background**
  (background intersected with the set of genes the external database itself could ever annotate - see
  `cosmic_coverage_genes.txt` / `pai3d_coverage_genes.txt`, emitted by
  `build_position_gmt.py`).
- **Databases**: 10 GMT-style position-set databases (Pfam domains/clans,
  COSMIC and PAI3D and general orthogroups, genomic locations,
  positive/purifying-selection orthogroups, UCR core/flank orthogroups) plus
  the broad `characterization` layers (UCR core/flank, FUBAR
  positive/purifying). `min`/`max` set-size filtering applies only to the GMT-style databases.
- **`p.adj`**: BH-adjusted per (ranking, database).
- **Significance**: `p.adj < `r PADJ`` and `NES > 0`.

*Report generated `r Sys.time()`.*

---

## Summary

This report presents position-level enrichment results for
**`r params$traitname`** via Position-Level Path Sum Permulation, testing
`r if (has_res) dplyr::n_distinct(res$database) else 0` position-set
database(s) across the global/top/bottom directions.
`r if (has_res) sum(res$sig, na.rm = TRUE) else 0` (ranking, database,
pathway, cutoff) combinations passed at `p.adj < ``r PADJ`` with NES > 0.

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere - Enrichment Module*
*Date of generation: `r Sys.Date()`*
