---
title: "Position-level Enrichment (Path Sum Permulation) - `r params$traitname`"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
    mathjax: null                       # no math in this report
    md_extensions: "-tex_math_dollars"  # pathway names contain '$' (e.g. V$HNF4_01) - never parse as TeX math
params:
  results_file: NULL
  leading_edge_file: NULL
  traitname: "trait"
  padj_thr: 0.15
  table_top_n: 200
  # Position Characterisation: PrimateAI-3D
  # + COSMIC validation and regional accumulation, all restricted to the US scheme.
  # NULL/absent inputs degrade gracefully (has_pos_char is FALSE, section skipped).
  position_scores_file: NULL   # position_scores.tsv (per-position CAAS scores)
  # SCORING's published position_lists/slice_<dir><pct>.tsv dir (scoring_compute.R),
  # when supplied, the 10/5/1% membership used throughout Position
  # Characterisation is read from there instead of an independent quantile()
  # cut computed here (see the has_pos_char setup chunk's test_glob_*/test_top_*/
  # test_bot_* closures). Falls back to quantile() per direction when NULL or a
  # slice file is missing.
  position_lists_dir: NULL
  gene_scores_file:     NULL   # gene_scores.tsv (background gene fallback)
  vep_primateai_file:   NULL   # PrimateAI-3D score TSV
  vep_cosmic_file:      NULL   # COSMIC Mutant Census score TSV
  genomic_info_file:    NULL   # gene, chr, start, end, length (no header assumed)
  # FADE_top_sig/FADE_bottom_sig position group (gene,position,max_bf,target_aa;
  # BF >= fade_bf_threshold only). Direction-specific, unlike PAI3D/COSMIC.
  fade_sites_top_file:    NULL
  fade_sites_bottom_file: NULL
  # SCORING's fcs_stats.tsv (gene + flag_* columns), same file already
  # consumed as --annot-file by posenrich_enrich.py. Unused by this report
  # directly (kept declared for pipeline-wiring compatibility; the Overall
  # dotplot no longer computes a cross-module composite that needed it).
  fcs_stats_file: NULL
  # cleaned_background_main.txt: the honest tested-gene background used to
  # restrict the PrimateAI-3D/COSMIC join sections to genes actually in scope
  # (shared_genes_pai/shared_genes_cosmic below), rather than falling back to
  # whichever genes happen to have a gene_scores.tsv row.
  universe_file: NULL
  top_pct:      0.10
  window_size_bp: 1000000
  output_dir:   "."            # regional_accumulation.tsv export target
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6, dev = "png", dpi = 150)
suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
  library(kableExtra)
  library(DT)
  theme_set(
    theme_minimal(base_size = 13) +
    theme(
      axis.title.y  = element_text(margin = ggplot2::margin(l = 20, r = 20)),
      axis.title.x  = element_text(margin = ggplot2::margin(t = 12)),
      axis.text.x   = element_text(margin = ggplot2::margin(t = 5)),
      plot.title    = element_text(face = "bold", hjust = 0, margin = ggplot2::margin(b = 6)),
      plot.subtitle = element_text(colour = "grey40", hjust = 0, margin = ggplot2::margin(b = 10)),
      legend.position = "right",
      legend.box.margin = ggplot2::margin(t = 5)
    )
  )
  library(scales)
  library(patchwork)
})

PADJ <- params$padj_thr
TOP_N <- params$table_top_n

this_dir <- tryCatch(dirname(normalizePath(knitr::current_input(), mustWork = FALSE)),
                     error = function(e) ".")

kbl_html <- function(df, caption = NULL, height = NULL, escape = TRUE, ...) {
  k <- kable(df, format = "html", escape = escape, caption = caption, ...) %>%
    kable_styling(bootstrap_options = c("striped", "condensed", "hover"), full_width = FALSE)
  if (!is.null(height)) k <- scroll_box(k, height = height)
  k
}

# Safe correlation helper: returns NA (not an error) when inputs have fewer
# than 3 complete pairs, avoids crashes in sparse optional-module sections.
safe_cor <- function(x, y, method = "pearson") {
  ok <- complete.cases(x, y)
  if (sum(ok) < 3) return(NA_real_)
  cor(x[ok], y[ok], method = method)
}
```

```{r load, include=FALSE}
stopifnot(!is.null(params$results_file), file.exists(params$results_file))
res <- readr::read_tsv(params$results_file, show_col_types = FALSE)
if (!is.null(res) && "top_frac" %in% names(res)) {
  res <- res %>% dplyr::filter(!top_frac %in% c(1.0, 0.5))
} else if (!is.null(res)) {
  res$top_frac <- NA_real_
}
if (!is.null(res)) {
  if (!"foreground_size" %in% names(res)) res$foreground_size <- if ("n_pos_with_score" %in% names(res)) res$n_pos_with_score else NA_real_
  if (!"observed" %in% names(res)) res$observed <- if ("obs_sum" %in% names(res)) res$obs_sum else NA_real_
  if (!"expected" %in% names(res)) res$expected <- if ("null_mean" %in% names(res)) res$null_mean else NA_real_
  if (!"fold_enrichment" %in% names(res)) res$fold_enrichment <- if ("perm_nes" %in% names(res)) res$perm_nes else NA_real_
}
has_res <- !is.null(res) && nrow(res) > 0
RANKING_ORDER <- c("global", "top", "bottom")
if (has_res && !"sig" %in% names(res)) {
  # p.adj < padj_thr AND NES > 0
  res$sig <- res$p_adj < PADJ & (!"perm_nes" %in% names(res) | res$perm_nes > 0)
}

# Truncates a "gene1, gene2, ..." string to the first n_show entries, with
# the remainder tucked behind an inline <details>/<summary> toggle
truncate_gene_list <- function(x, n_show = 5) {
  vapply(x, function(s) {
    if (is.na(s) || !nzchar(s)) return("")
    items <- strsplit(s, ", ", fixed = TRUE)[[1]]
    if (length(items) <= n_show) return(s)
    shown <- paste(items[seq_len(n_show)], collapse = ", ")
    rest  <- paste(items[(n_show + 1):length(items)], collapse = ", ")
    sprintf(
      '%s <details style="display:inline"><summary style="display:inline; cursor:pointer; color:#2166ac;">&hellip; (%d more)</summary> %s</details>',
      shown, length(items) - n_show, rest
    )
  }, character(1), USE.NAMES = FALSE)
}

# Observed gene:position members of the overlap, significant terms only
le_raw <- NULL
if (!is.null(params$leading_edge_file) && file.exists(params$leading_edge_file)) {
  le_raw <- tryCatch(readr::read_tsv(params$leading_edge_file, show_col_types = FALSE),
                     error = function(e) NULL)
  if (!is.null(le_raw) && "top_frac" %in% names(le_raw)) {
    le_raw <- le_raw %>% dplyr::filter(!top_frac %in% c(1.0, 0.5))
  } else if (!is.null(le_raw)) {
    le_raw$top_frac <- NA_real_
  }
  if (!is.null(le_raw) && nrow(le_raw) == 0) le_raw <- NULL
}
# Pooled across top_frac cutoffs (one row per ranking x database x pathway,
# not per cutoff)
le_summary <- if (!is.null(le_raw)) {
  le_raw %>%
    dplyr::group_by(ranking, database, pathway) %>%
    dplyr::summarise(`Genes/positions` = paste(sort(unique(gene_position)), collapse = ", "),
                     `N genes` = dplyr::n_distinct(gene),
                     .groups = "drop")
} else NULL

# Direction-aware corroboration columns (mirrors 12.FCS_general_report.Rmd's
# cols_by_rk): fade_bottom only under `bottom`, fade_top only under `top` --
# no combined "fade" column under `global` (redundant with _top/_bottom, which
# already cover both directions). pct_accum_top/pct_accum_bottom are the
# direction-restricted accumulation tests (same convention as fade); pct_accum
# (non-directional, accumulation_all_* pooled) is reserved for `global`. Only
# columns actually present are ever shown.
flag_cols_by_rk <- list(
  global = c("pct_gate_sig", "pct_rer_acc", "pct_rer_decc", "pct_accum"),
  top    = c("pct_gate_sig", "pct_fade_top", "pct_rer_acc", "pct_rer_decc", "pct_accum_top"),
  bottom = c("pct_gate_sig", "pct_fade_bottom", "pct_rer_acc", "pct_rer_decc", "pct_accum_bottom")
)

.fcs_stats_file_ok <- function(p) !is.null(p) && !is.na(p) && nchar(p) > 0 && file.exists(p)

# FADE_top_sig/FADE_bottom_sig: gene,position,max_bf,target_aa, direction-specific
# (never cross-checked, a TOP-direction position is only ever validated against
# fade_sites_top, never fade_sites_bottom, and vice versa). Position is already in
# CAAS's own coordinate space (see build_position_gmt.py), so no map_cache lookup.
has_fade_sites_top    <- .fcs_stats_file_ok(params$fade_sites_top_file)
has_fade_sites_bottom <- .fcs_stats_file_ok(params$fade_sites_bottom_file)
fade_sites_top_df <- if (has_fade_sites_top) {
  readr::read_csv(params$fade_sites_top_file, show_col_types = FALSE) %>%
    dplyr::rename(Gene = gene, Position = position) %>%
    dplyr::mutate(Position = suppressWarnings(as.integer(Position)))
} else NULL
fade_sites_bottom_df <- if (has_fade_sites_bottom) {
  readr::read_csv(params$fade_sites_bottom_file, show_col_types = FALSE) %>%
    dplyr::rename(Gene = gene, Position = position) %>%
    dplyr::mutate(Position = suppressWarnings(as.integer(Position)))
} else NULL
```

# Overview {.tabset .tabset-fade}

> [!WARNING]
> **Proof of concept - not validated for standalone significance claims.** This
> module has not had a real end-to-end production run, and its GMT construction,
> its handling of the very large (~1.47M-position) background, and its
> significance gating (`p_adj < padj_thr` with `NES > 0`) have not been
> checked against any known-truth signal. Treat everything in this report as
> descriptive characterization only (e.g. "does this term show up near
> CAAS-high positions") - do not cite it as evidence for a biological claim on
> its own, and do not build downstream decisions on top of its "significant"
> term counts without independent corroboration.

This report summarises **position-level enrichment** for `r params$traitname`. Unlike
the gene-level FCS (a Wilcoxon-AUC rank-shift test over genes), the position background is a
sea of invariant columns where <0.2% of positions carry a non-zero CAAS score. A rank-shift test
degenerates there, and fixed-cutoff Fisher tests suffer from extreme sample-size sensitivity at $N \approx 1.47\text{M}$.
Instead, we use **Position-Level Path Sum Permulation**: accumulating continuous CAAS score magnitudes per term ($T_\text{obs} = \sum_{p \in \text{term}} s_p$)
and comparing against a label-permuted null distribution across the full ~1.47M position background ($N_\text{perms} = 2,000$).

A set is flagged **significant** when `p.adj < `r PADJ`` (BH-adjusted per
ranking × database) and `NES > 0`.

```{r overview, echo=FALSE, results='asis'}
if (!has_res) {
  cat("No enrichment results were produced.\n")
} else {
  for (rk in intersect(RANKING_ORDER, unique(res$ranking))) {
    cat(sprintf("\n## %s\n\n", toupper(rk)))
    ov <- res %>% filter(ranking == rk) %>%
      group_by(database) %>%
      summarise(`Sets tested` = n_distinct(pathway),
                `Sig (p.adj)` = n_distinct(pathway[sig]),
                .groups = "drop") %>%
      arrange(database)
    print(kbl_html(ov, caption = sprintf(
      "%s - sets tested and significant at p.adj < %.3g.",
      toupper(rk), PADJ)))
    cat("\n\n")
  }
}
```

# Enriched position sets {.tabset .tabset-fade}

Foreground = the top 25 / 10 / 5 / 1% of scored positions by `CAAS_score` in
that direction (`top`/`bottom` restrict to `change_side` first, matching the
SCORING module's own percentile convention; no `global` ranking - dropped in
favour of top/bottom only, consistent with every other report in this
pipeline). Background = the full honest tested
position pool for every test - never restricted to the scored subset, so
tested-and-negative positions are never silently dropped from the null.
`fold > 1` = enriched, `< 1` = depleted. UCR core and UCR flank (and their
per-orthogroup counterparts) are tested as independent, potentially-overlapping
layers, not a forced partition - nothing here requires them to be disjoint.
When available, `%gate_sig`/`%fade_*`/`%rer_*`/`%accum` columns
report the percentage of *distinct genes* contributing to the overlap (never
raw positions - one gene must not dominate) that are independently flagged by
SCORING's own significance/FADE/RER/accumulation criteria - a cross-module
corroboration check for whether a hit is broadly supported or isolated.

```{r enrichment-tables, echo=FALSE, results='asis'}
if (!has_res) {
  cat("_No results._\n\n")
} else {
  for (rk in intersect(RANKING_ORDER, unique(res$ranking))) {
    cat(sprintf("\n## %s {.tabset}\n\n", toupper(rk)))
    rk_df <- res %>% filter(ranking == rk)
    for (db in unique(rk_df$database)) {
      cat(sprintf("\n### %s {.tabset}\n\n", db))
      db_df <- rk_df %>% filter(database == db)

      has_tf_vals <- "top_frac" %in% names(db_df) && !all(is.na(db_df$top_frac))
      tf_vals <- if (has_tf_vals) sort(unique(db_df$top_frac), decreasing = TRUE) else NA_real_
      for (tf in tf_vals) {
        if (!is.na(tf)) cat(sprintf("\n#### Top %s%%\n\n", format(100 * tf, trim = TRUE)))
        tf_all <- if (!is.na(tf)) db_df %>% filter(top_frac == tf) else db_df
        tf_all <- tf_all %>% arrange(desc(sig), p_adj, p_value)
        n_total_tf <- nrow(tf_all)
        tf_src <- tf_all %>% head(TOP_N)
        if (n_total_tf > TOP_N) {
          cat(sprintf("_Showing top %d of %d tested pathways (sorted by significance)._\n\n", TOP_N, n_total_tf))
        }
        if ("perm_nes" %in% names(tf_src)) {
          tf_df <- tf_src %>%
            transmute(Pathway = pathway, Description = description,
                      `Layer size` = layer_size, `N scored` = n_pos_with_score,
                      `Observed sum` = round(obs_sum, 2), `Null mean` = round(null_mean, 2),
                      NES = round(perm_nes, 3), Direction = direction,
                      `p (Perm)` = signif(p_value, 3),
                      `p.adj` = signif(p_adj, 3), Sig = ifelse(sig, "*", ""))
        } else {
          tf_df <- tf_src %>%
            transmute(Pathway = pathway, Description = description,
                      `Layer size` = layer_size, `Foreground` = foreground_size,
                      Observed = observed, Expected = round(expected, 1),
                      Fold = round(fold_enrichment, 3), Direction = direction,
                      `p (Fisher)` = signif(p_value, 3),
                      `p.adj` = signif(p_adj, 3), Sig = ifelse(sig, "*", ""))
        }
        # Cross-module corroboration: % of the DISTINCT genes in the overlap
        # (never % of raw positions - one gene must not dominate) carrying
        # each SCORING flag (gate_sig/fade/rer/accum). Only present
        # when POSENRICH was run with --annot-file. Direction-restricted via
        # flag_cols_by_rk (same convention as the Leading edge section below) --
        # posenrich_enrich.py computes pct_<flag> for every flag regardless of
        # direction, so without this restriction a `bottom` table would show
        # `%accum_top`/`%fade_top` alongside its own `%accum_bottom`/`%fade_bottom`.
        want_rk <- flag_cols_by_rk[[rk]]
        if (is.null(want_rk)) want_rk <- character(0)
        flag_cols <- intersect(want_rk, names(tf_src))
        if (length(flag_cols) > 0) {
          flag_df <- tf_src %>% transmute(across(all_of(flag_cols), ~round(.x, 1)))
          names(flag_df) <- paste0("%", sub("^pct_", "", flag_cols))
          tf_df <- dplyr::bind_cols(tf_df, flag_df)
        }
        id <- paste("dt", rk, db, gsub("\\.", "_", as.character(tf)), sep = "_")
        cat(knitr::knit_child(text = c(
          paste0("```{r ", id, ", echo=FALSE}"),
          "DT::datatable(tf_df, options = list(pageLength = 10, scrollX = TRUE), rownames = FALSE)",
          "```"), envir = environment(), quiet = TRUE))
        cat("\n\n")
      }
    }
  }
}
```

```{r term-clustering-setup, include=FALSE}
# Term-similarity clustering - ported verbatim from 13.AMI_analysis.Rmd's
# Module summary rollup. Different GMT databases can describe the same
# biological concept with slightly different wording (case, punctuation,
# minor rewording, collapsed via normalized edit-distance, utils::adist()),
# or one description can be a compound/nested restatement of another (e.g.
# "Primary ciliary dyskinesia" inside "Axonemal dynein complex and Primary
# ciliary dyskinesia"), character-level edit distance alone rejects these
# (the extra words dominate the ratio even though one term is wholly
# contained in the other), so containment is checked explicitly and forces a
# merge regardless of the edit-distance threshold. Defined here (ahead of the
# Overall dotplot / Direction triangle / Term recurrence sections below) so
# all three consume the SAME canonical Term per (database, pathway) --
# otherwise a term split across two near-duplicate GMT entries would merge
# in one section but not another.
.normalize_term <- function(x) {
  x <- tolower(x)
  x <- gsub("[^a-z0-9 ]", " ", x)
  gsub("\\s+", " ", trimws(x))
}
.term_contains <- function(a, b) {
  wa <- strsplit(a, " ")[[1]]; wb <- strsplit(b, " ")[[1]]
  if (length(wa) == 0 || length(wb) == 0) return(FALSE)
  short <- if (length(wa) <= length(wb)) wa else wb
  long  <- if (length(wa) <= length(wb)) wb else wa
  all(short %in% long)
}
.cluster_terms <- function(terms, sim_thr = 0.85) {
  norm <- vapply(terms, .normalize_term, character(1))
  reps_norm <- character(0); reps_orig <- character(0)
  assign <- integer(length(terms))
  for (i in seq_along(terms)) {
    if (length(reps_norm) == 0) { reps_norm <- norm[i]; reps_orig <- terms[i]; assign[i] <- 1L; next }
    d <- utils::adist(norm[i], reps_norm)[1, ]
    sim <- 1 - d / pmax(nchar(norm[i]), nchar(reps_norm), 1)
    contained <- vapply(reps_norm, function(r) .term_contains(norm[i], r), logical(1))
    sim[contained] <- 1
    best <- which.max(sim)
    if (sim[best] >= sim_thr) {
      assign[i] <- best
      if (contained[best] && nchar(terms[i]) < nchar(reps_orig[best])) {
        reps_orig[best] <- terms[i]
      }
    } else {
      reps_norm <- c(reps_norm, norm[i]); reps_orig <- c(reps_orig, terms[i])
      assign[i] <- length(reps_norm)
    }
  }
  reps_orig[assign]
}
```

# Overall dotplot {.tabset .tabset-fade}

Pooled across all databases and threshold cutoffs, ranked by BH-adjusted p
(the one test this report runs - Position-Level Path Sum Permulation - so
there is no multi-method vote to combine, unlike the gene-level FCS report).
One dotplot per direction, restricted to the top `r params$table_top_n`
terms, on a **shared x-axis of `-log10(BH-adjusted p)`**. Each term's segment
colour encodes its own **NES** (light-to-dark, darker = larger NES among the
terms shown); the **tip marker** is drawn in a fixed dark shade - always
darker than any point on the segment's own NES gradient - so the
significance endpoint stands out regardless of that term's NES. `list
_threshold`/`n_occurrences` (also in the table export) record which
10%/5%/1% cutoffs recovered the term and how many, as a simple, auditable
recurrence count - not a weighted score.

```{r overall-dotplot, echo=FALSE, results='asis', fig.height=6}
.NES_HUE_LIGHT <- "#e7d4e8"  # same Permulation colour family as 12.FCS_general_report.Rmd,
.NES_HUE_DARK  <- "#4a148c"  # since this report's one test IS Path Sum Permulation.

# Darkens a hex colour toward black by `amount` (0-1), used so the tip
# marker reads as "this arm's own colour, just deeper" rather than a fixed
# anchor shade unrelated to the term's own NES.
.darken_hex <- function(hex, amount = 0.3) {
  rgb <- grDevices::col2rgb(hex) / 255
  rgb <- rgb * (1 - amount)
  grDevices::rgb(rgb[1, ], rgb[2, ], rgb[3, ])
}

# A standalone light->dark gradient bar used as the legend key, scale_
# colour_identity() can only draw a flat swatch, so the ramp itself is
# otherwise invisible in the legend.
.method_gradient_legend <- function(title, light, dark) {
  df <- data.frame(y = 1, v = seq(0, 1, length.out = 100))
  ggplot(df, aes(x = v, y = y, fill = v)) +
    geom_tile() +
    scale_fill_gradientn(colours = c(light, dark), guide = "none") +
    scale_x_continuous(breaks = c(0, 1), labels = c("low", "high")) +
    scale_y_continuous(breaks = NULL) +
    labs(title = title, x = NULL, y = NULL) +
    theme_minimal(base_size = 8) +
    theme(plot.title = element_text(size = 8, hjust = 0.5),
          panel.grid = element_blank(),
          plot.margin = margin(2, 2, 2, 2))
}

# knitr opens the figure device for a chunk ONCE, using the fig.height in
# effect at chunk start - opts_current$set(fig.height = ...) called mid-chunk
# (e.g. inside this per-direction loop) does NOT change the size of the image
# actually written out, verified empirically, it's a known knitr limitation,
# not a working per-plot override. Render+embed each plot manually instead,
# which is the only reliable way to vary height per plot in one asis chunk.
.emit_sized_plot <- function(p, height_in, width_in = knitr::opts_chunk$get("fig.width"),
                              dpi = knitr::opts_chunk$get("dpi")) {
  f <- tempfile(fileext = ".png")
  on.exit(unlink(f), add = TRUE)
  ggplot2::ggsave(f, plot = p, width = width_in, height = height_in, dpi = dpi, units = "in")
  cat(sprintf('<img src="data:image/png;base64,%s" style="max-width:100%%;" />\n\n',
              base64enc::base64encode(f)))
}

.overall_dotplot_rows <- list()  # accumulated across rankings, written to disk after
                                  # the loop; the Comparison report imports this file directly.
if (!has_res || !any(res$sig, na.rm = TRUE)) {
  cat("_No significant sets._\n\n")
} else {
  for (rk in intersect(RANKING_ORDER, unique(res$ranking[res$sig]))) {
    cat(sprintf("\n## %s\n\n", toupper(rk)))
    rk_sig <- res %>% filter(ranking == rk, sig)
    if (nrow(rk_sig) == 0) { cat("_No significant sets for this direction._\n\n"); next }

    # list_threshold/n_occurrences: which top_frac cutoffs this term reached
    # significance at, and how many, a plain recurrence count, independent
    # of any composite score.
    .has_tf_col <- "top_frac" %in% names(rk_sig) && !all(is.na(rk_sig$top_frac))
    term_meta <- rk_sig %>%
      dplyr::group_by(database, pathway) %>%
      dplyr::summarise(p_adj = min(p_adj, na.rm = TRUE),
                        perm_nes = { v <- stats::na.omit(perm_nes); if (length(v)) max(v) else NA_real_ },
                        description = dplyr::first(dplyr::coalesce(description, pathway)),
                        list_threshold = if (.has_tf_col) paste(sort(100 * unique(na.omit(top_frac)), decreasing = TRUE), "%", sep = "", collapse = ", ") else "N/A",
                        n_occurrences = if (.has_tf_col) dplyr::n_distinct(top_frac, na.rm = TRUE) else 1L,
                        .groups = "drop")

    # N positions in the leading edge (dot size), purely descriptive, not
    # part of the ranking.
    if (!is.null(le_raw)) {
      n_le_tbl <- le_raw %>% dplyr::filter(ranking == rk) %>%
        dplyr::group_by(database, pathway) %>%
        dplyr::summarise(n_le = dplyr::n_distinct(gene_position), .groups = "drop")
      term_meta <- term_meta %>% dplyr::left_join(n_le_tbl, by = c("database", "pathway"))
    } else {
      term_meta$n_le <- NA_integer_
    }

    dp_all <- term_meta %>%
      dplyr::filter(!is.na(p_adj)) %>%
      dplyr::arrange(p_adj) %>%
      dplyr::mutate(label = stringr::str_trunc(sprintf("%s (%s)", description, database), 60))

    # Every significant/scored term for this direction, kept for the Triangle
    # plot below and the exported TSV; the visual dotplot is TOP_N-limited only.
    .overall_dotplot_rows[[rk]] <- dp_all %>% dplyr::mutate(ranking = rk, .before = 1)

    dp <- dp_all %>% head(TOP_N)
    if (nrow(dp) == 0) { cat("_No terms to display._\n\n"); next }

    # y-axis index 1 = bottom of plot, reverse so the best (smallest p_adj) term is at the top.
    term_levels <- rev(unique(dp$label))
    dp <- dp %>%
      dplyr::mutate(
        y_base = as.numeric(factor(label, levels = term_levels)),
        test_signal = -log10(pmax(p_adj, 1e-300)),
        .nes_pct = dplyr::percent_rank(perm_nes),
        arm_colour = grDevices::colorRampPalette(c(.NES_HUE_LIGHT, .NES_HUE_DARK))(101)[
          pmax(1, pmin(101, round(dplyr::coalesce(.nes_pct, 0.7) * 100) + 1))],
        # Tip marker: darkened shade of this row's own arm colour, not the
        # fixed anchor, keeps the dot visually tied to its own bar.
        tip_colour = .darken_hex(arm_colour)
      )

    p <- ggplot(dp, aes(y = y_base)) +
      geom_segment(aes(x = 0, xend = test_signal, yend = y_base, colour = arm_colour),
                   linewidth = 1.1, lineend = "round", alpha = 0.85) +
      geom_point(aes(x = test_signal, size = n_le, colour = tip_colour), alpha = 1) +
      scale_y_continuous(breaks = seq_along(term_levels), labels = term_levels) +
      scale_colour_identity() +
      scale_size_continuous(name = "N positions") +
      geom_vline(xintercept = -log10(PADJ), linetype = "dashed", colour = "grey50") +
      labs(title = sprintf("%s - %s (pooled across databases)", params$traitname, toupper(rk)),
           subtitle = "Position-Level Path Sum Permulation - tip at the BH-adjusted p",
           x = expression(-log[10]~"BH-adjusted p"), y = NULL) +
      theme_minimal(base_size = 11) +
      theme(axis.text.y = element_text(colour = "black",
                                        size = pmax(8, pmin(11, 220 / length(term_levels)))),
            plot.subtitle = element_text(margin = margin(b = 12)),
            legend.position = "bottom",
            legend.box = "horizontal")

    # Method legend: a gradient bar for the one test this report runs
    # (Permulation NES), placed directly under p (which already carries its
    # own "N positions" size legend at its bottom via legend.position =
    # "bottom" above) - a patchwork guide_area() + guides = "collect"
    # placement was tried first, to share one right-hand column with that
    # size legend, but the collected guide rendered as its own block beside
    # the gradient bar instead of stacked above it. Bottom placement avoids
    # that entirely: the whole legend reads top-to-bottom as one column.
    gradient_bar <- .method_gradient_legend("Permulation (NES)", .NES_HUE_LIGHT, .NES_HUE_DARK)
    # Fixed absolute height for the gradient bar (not a proportion of the
    # whole figure) - a relative split makes it grow right along with the
    # main plot on term-heavy reports, ending up huge; a thin colour ramp
    # needs the same ~0.3in regardless of plot height.
    combo <- p / gradient_bar + patchwork::plot_layout(heights = grid::unit(c(1, 0.18), c("null", "in")))
    .emit_sized_plot(combo, height_in = pmax(5, pmin(22, 2.6 + 0.4 * length(term_levels))))
  }
}
if (length(.overall_dotplot_rows) > 0) {
  readr::write_tsv(dplyr::bind_rows(.overall_dotplot_rows), "posenrich_overall_dotplot.tsv")
}
```

# Direction triangle (NES)

One point per **term** - merged across near-duplicate descriptions the same
way as the Term recurrence table below (so a term split across two GMT
databases, or two slightly-differently-worded entries, is one point here
too, not two incomplete ones) - positioned by **which direction(s)** it is
significant in and **how strongly** (NES, clipped to ≥ 0) in each: **Global**
at the top vertex, **Bottom** at the left vertex, **Top** at the right
vertex. A term significant in only one direction sits exactly at that
vertex (there is no second direction to weigh it against, so its exact NES
value cannot move it off the vertex); a term significant in two sits
somewhere along the edge between them, positioned by the relative size of
the two NES values; a term significant in **all three** sits inside the
triangle. Colour marks which of the seven possible combinations a term
falls into (three single-direction, three two-direction edges, one
three-direction interior).

Terms sharing an exact vertex/edge position (most commonly every
single-direction term, which all collapse to the same vertex) are fanned out
around it rather than randomly jittered, **ordered by total raw NES**
(stronger terms sit closer to the vertex, weaker ones further along the
fan) - the fan position is not itself a magnitude axis, but its *order* is,
so two overlapping dots are never placed arbitrarily relative to each other.
Only terms significant in **two or three** directions are labelled (edge and
interior points) - single-direction terms sit exactly on a vertex, where
they're already identified by that vertex, so labelling them too would just
clutter the densest part of the plot without adding information.

```{r direction_triangle, echo=FALSE, results='asis'}
sc_all <- if (length(.overall_dotplot_rows) > 0) dplyr::bind_rows(.overall_dotplot_rows) else tibble::tibble()
if (nrow(sc_all) == 0) {
  cat("_No significant terms to compare across directions._\n\n")
} else {
  # Merge near-duplicate term descriptions BEFORE pivoting to directions --
  # otherwise a term split across two databases (or two differently-worded
  # GMT entries) would survive as two separate, incomplete points instead of
  # the one term the Recurrence Table/Heatmap below reports. Two rows that
  # merge into the same Term *and* the same ranking (e.g. two databases both
  # significant under "top") keep the stronger (higher) NES, mirroring
  # 15.Comparison_report.Rmd's "best across merged aliases" convention.
  desc_levels <- unique(dplyr::coalesce(sc_all$description, sc_all$pathway))
  cluster_map <- setNames(.cluster_terms(desc_levels), desc_levels)
  sc_all <- sc_all %>%
    dplyr::mutate(Term = unname(cluster_map[dplyr::coalesce(description, pathway)])) %>%
    dplyr::group_by(Term, ranking) %>%
    # max(..., na.rm=TRUE) over an all-NA group returns -Inf with a warning,
    # not NA, guard explicitly so an all-NA perm_nes group (a pathway with
    # p_adj but no recorded NES) stays NA rather than becoming a bogus -Inf.
    dplyr::summarise(perm_nes = if (all(is.na(perm_nes))) NA_real_ else max(perm_nes, na.rm = TRUE),
                      .groups = "drop")

  sc <- sc_all %>%
    tidyr::pivot_wider(
      id_cols = Term,
      names_from = ranking, values_from = perm_nes, names_prefix = "nes_"
    )
  for (rc in setdiff(c("nes_global", "nes_top", "nes_bottom"), names(sc))) sc[[rc]] <- NA_real_

  sc <- sc %>%
    dplyr::mutate(
      has_g = !is.na(nes_global), has_t = !is.na(nes_top), has_b = !is.na(nes_bottom),
      w_g = pmax(dplyr::coalesce(nes_global, 0), 0),
      w_t = pmax(dplyr::coalesce(nes_top, 0), 0),
      w_b = pmax(dplyr::coalesce(nes_bottom, 0), 0)
    ) %>%
    dplyr::filter(has_g | has_t | has_b) %>%
    dplyr::mutate(
      # A direction with NES <= 0 still "counts" for the category (it cleared
      # its own significance gate) even though it contributes zero barycentric
      # weight, fall back to an equal split among the significant
      # direction(s) so the point doesn't collapse to the origin.
      w_sum = w_g + w_t + w_b,
      w_g = dplyr::if_else(w_sum > 0, w_g / w_sum, as.numeric(has_g) / pmax(has_g + has_t + has_b, 1)),
      w_t = dplyr::if_else(w_sum > 0, w_t / w_sum, as.numeric(has_t) / pmax(has_g + has_t + has_b, 1)),
      w_b = dplyr::if_else(w_sum > 0, w_b / w_sum, as.numeric(has_b) / pmax(has_g + has_t + has_b, 1)),
      x = w_t * 1 + w_g * 0.5,
      y = w_g * (sqrt(3) / 2),
      Category = dplyr::case_when(
        has_g & has_t & has_b ~ "all",
        has_g & has_t         ~ "global_top",
        has_g & has_b         ~ "global_bottom",
        has_t & has_b         ~ "top_bottom",
        has_g                 ~ "global",
        has_t                 ~ "top",
        has_b                 ~ "bottom"
      ),
      Category = factor(Category, levels = c("global", "top", "bottom", "top_bottom",
                                              "global_bottom", "global_top", "all")),
      n_dirs = has_g + has_t + has_b,
      # Total raw NES across whichever direction(s) cleared significance --
      # used only to ORDER points that land on the exact same vertex/edge
      # position, not as a literal coordinate (there is no magnitude axis).
      strength = pmax(dplyr::coalesce(nes_global, 0), 0) +
                 pmax(dplyr::coalesce(nes_top, 0), 0) +
                 pmax(dplyr::coalesce(nes_bottom, 0), 0)
    )

  if (nrow(sc) == 0) {
    cat("_No plottable points._\n\n")
  } else {
    tri <- tibble::tibble(x = c(0.5, 0, 1, 0.5), y = c(sqrt(3) / 2, 0, 0, sqrt(3) / 2))
    tri_cx <- 0.5; tri_cy <- sqrt(3) / 6  # triangle centroid
    cat_colours <- c(global = "#f1c40f", top = "#2980b9", bottom = "#27ae60",
                     top_bottom = "#8e44ad", global_bottom = "#16a085",
                     global_top = "#d35400", all = "#e74c3c")

    # Deterministic, order-preserving spread: exact barycentric coordinates
    # collapse every single-direction term onto the same vertex (and can
    # coincide on an edge too), with no magnitude axis to separate them.
    # Random jitter would imply a meaning the offset doesn't have, so instead
    # fan colliding points out along the TANGENT to their shared position
    # (perpendicular to the ray from the triangle centroid through that
    # point), ranked by `strength`, stronger terms end up closer to the
    # true vertex/edge position, weaker ones further out along the fan.
    # Spacing shrinks as a cluster grows so the fan stays bounded near its
    # vertex rather than spilling arbitrarily far across the plot.
    sc <- sc %>%
      dplyr::group_by(x, y) %>%
      dplyr::mutate(
        .n_here = dplyr::n(),
        .rank_here = rank(-strength, ties.method = "first"),
        .radial_x = x - tri_cx, .radial_y = y - tri_cy,
        .rlen = pmax(sqrt(.radial_x^2 + .radial_y^2), 1e-6),
        .tan_x = -.radial_y / .rlen, .tan_y = .radial_x / .rlen,
        .spacing = pmin(0.035, 0.5 / pmax(.n_here - 1, 1)),
        .offset = dplyr::if_else(.n_here > 1, (.rank_here - (.n_here + 1) / 2) * .spacing, 0),
        x = x + .tan_x * .offset,
        y = y + .tan_y * .offset
      ) %>%
      dplyr::ungroup() %>%
      dplyr::select(-.n_here, -.rank_here, -.radial_x, -.radial_y, -.rlen, -.tan_x, -.tan_y, -.spacing, -.offset)

    p <- ggplot(sc, aes(x, y)) +
      geom_path(data = tri, colour = "grey50", linetype = "dashed") +
      geom_point(aes(colour = Category), alpha = 0.8, size = 2.2) +
      annotate("text", x = 0.5, y = sqrt(3) / 2 + 0.04, label = "Global", fontface = "bold") +
      annotate("text", x = -0.05, y = -0.04, label = "Bottom", fontface = "bold", hjust = 1) +
      annotate("text", x = 1.05, y = -0.04, label = "Top", fontface = "bold", hjust = 0) +
      scale_colour_manual(values = cat_colours, name = "Significant in", drop = FALSE) +
      coord_equal(clip = "off") +
      labs(title = sprintf("Position enrichment - direction triangle (NES) - %s", params$traitname),
           subtitle = "Vertex = single direction, edge = two directions, interior = all three, overlapping terms fanned out by total NES, not jittered") +
      theme_void(base_size = 12) +
      theme(plot.subtitle = element_text(margin = margin(b = 12), colour = "grey40"),
            legend.position = "right",
            plot.margin = margin(20, 40, 10, 40))

    # Only terms shared between two or three directions are labelled (edge
    # and interior points) - single-direction terms already read as "which
    # vertex" without a label, so ggrepel spends its budget on the points
    # that actually need disambiguating.
    shared_sc <- dplyr::filter(sc, n_dirs >= 2)
    if (nrow(shared_sc) > 0) {
      p <- p + ggrepel::geom_text_repel(
        data = shared_sc, aes(label = stringr::str_trunc(Term, 45)), colour = "black",
        size = 2.6, fontface = "bold", max.overlaps = Inf, seed = 42,
        box.padding = 0.3, point.padding = 0.2,
        force = 3, force_pull = 0.5,
        max.time = 5, max.iter = 200000,
        min.segment.length = 0, segment.size = 0.3, segment.alpha = 0.6,
        bg.color = "white", bg.r = 0.1, show.legend = FALSE)
    }
    # Height scales with how many points are on the plot, same convention as
    # the Overall dotplot above (manual render+embed - opts_current$set()
    # mid-chunk does not actually resize the figure device, see .emit_sized_plot).
    .emit_sized_plot(p, height_in = pmax(6, pmin(20, 4 + 0.03 * nrow(sc))))
  }
}
```

# Leading edge {.tabset .tabset-fade}

One row per **significant** (`p.adj < `r PADJ``) pathway - pooled across the
10%/5%/1% cutoffs it was tested at, not one row per cutoff - with the union
of observed `gene:position` members across those cutoffs, plus the same
cross-module corroboration columns as above (`%gate_sig`/`%fade_*`/`%rer_*`/
`%accum`, direction-aware - `%fade_bottom` only appears under `BOTTOM`,
`%fade_top` only under `TOP`, plain `%fade` under `GLOBAL`). `N cutoffs`/
`Cutoffs` record which of the tighter cutoffs recovered the term as a plain
audit trail. The full per-cutoff member list is also written to
`posenrich_leading_edge.tsv`, and this pooled table to
`posenrich_leading_edge_summary.tsv`, both alongside this report - the
Comparison report imports the latter directly.

```{r leading-edge, echo=FALSE, results='asis'}
.leading_edge_summary_rows <- list()
if (!has_res || !any(res$sig, na.rm = TRUE)) {
  cat("_No significant sets._\n\n")
} else {
  sig_res <- res %>% filter(sig)
  for (rk in intersect(RANKING_ORDER, unique(sig_res$ranking))) {
    cat(sprintf("\n## %s\n\n", toupper(rk)))
    rk_sig_raw <- sig_res %>% filter(ranking == rk)
    if (nrow(rk_sig_raw) == 0) { cat("_No significant sets for this direction._\n\n"); next }

    want <- flag_cols_by_rk[[rk]]
    if (is.null(want)) want <- character(0)
    present_flags <- intersect(want, names(rk_sig_raw))
    has_nes <- "perm_nes" %in% names(rk_sig_raw)
    .has_tf <- "top_frac" %in% names(rk_sig_raw) && !all(is.na(rk_sig_raw$top_frac))

    # Collapse to one row per pathway: best (min) p.adj, strongest (max)
    # NES/fold across the cutoffs it recurred at, cutoff flags averaged
    # (already a %, so the mean of several cutoffs' % is itself meaningful).
    rk_sig <- rk_sig_raw %>%
      dplyr::group_by(database, pathway) %>%
      dplyr::summarise(
        description = dplyr::first(dplyr::coalesce(description, pathway)),
        direction = dplyr::first(direction),
        p_adj = min(p_adj, na.rm = TRUE),
        perm_nes = if (has_nes) { v <- stats::na.omit(perm_nes); if (length(v)) max(v) else NA_real_ } else NA_real_,
        fold_enrichment = if (!has_nes) { v <- stats::na.omit(fold_enrichment); if (length(v)) max(v) else NA_real_ } else NA_real_,
        n_occurrences = if (.has_tf) dplyr::n_distinct(top_frac, na.rm = TRUE) else 1L,
        `Cutoffs` = if (.has_tf) paste(sort(100 * unique(na.omit(top_frac)), decreasing = TRUE), "%", sep = "", collapse = ", ") else "N/A",
        dplyr::across(dplyr::all_of(present_flags), ~round(mean(.x, na.rm = TRUE), 1)),
        .groups = "drop"
      ) %>%
      dplyr::arrange(p_adj)

    if (!is.null(le_summary)) {
      le_rk_summary <- le_summary %>% dplyr::filter(ranking == rk) %>% dplyr::select(-ranking)
      rk_sig <- rk_sig %>% dplyr::left_join(le_rk_summary, by = c("database", "pathway"))
    }

    if (has_nes) {
      le_df <- rk_sig %>%
        transmute(Database = database, Pathway = pathway, Description = description,
                  NES = round(perm_nes, 3), Direction = direction,
                  `p.adj` = signif(p_adj, 3), `N cutoffs` = n_occurrences, Cutoffs)
    } else {
      le_df <- rk_sig %>%
        transmute(Database = database, Pathway = pathway, Description = description,
                  Fold = round(fold_enrichment, 3), Direction = direction,
                  `p.adj` = signif(p_adj, 3), `N cutoffs` = n_occurrences, Cutoffs)
    }
    if (length(present_flags) > 0) {
      flag_df <- rk_sig %>% dplyr::select(dplyr::all_of(present_flags))
      names(flag_df) <- paste0("%", sub("^pct_", "", present_flags))
      le_df <- dplyr::bind_cols(le_df, flag_df)
    }
    esc_cols <- TRUE
    if ("Genes/positions" %in% names(rk_sig)) {
      le_df <- dplyr::bind_cols(le_df, rk_sig %>% dplyr::select(`N genes`, `Genes/positions`))
      .leading_edge_summary_rows[[rk]] <- le_df %>% dplyr::mutate(Ranking = rk, .before = 1)
      # Full member list is unabridged in posenrich_leading_edge_summary.tsv --
      # here, show only the first few inline with a <details> toggle for the
      # rest, since a row can otherwise carry hundreds of comma-separated
      # gene:positions.
      le_df[["Genes/positions"]] <- truncate_gene_list(le_df[["Genes/positions"]])
      # DT column indices for `escape` are 1-based over the data columns when
      # rownames = FALSE (no offset needed), verified empirically, since
      # DT's own docs are ambiguous on this point and getting it wrong throws
      # "attempt to select less than one element in integerOneIndex" deep
      # inside DT:::escapeData(), aborting the whole render.
      esc_cols <- which(names(le_df) != "Genes/positions")
    } else {
      .leading_edge_summary_rows[[rk]] <- le_df %>% dplyr::mutate(Ranking = rk, .before = 1)
    }

    id <- paste("dt_le", rk, sep = "_")
    assign(paste0(".esc_cols_", rk), esc_cols, envir = environment())
    cat(knitr::knit_child(text = c(
      paste0("```{r ", id, ", echo=FALSE}"),
      sprintf("DT::datatable(le_df, options = list(pageLength = 10, scrollX = TRUE), rownames = FALSE, escape = .esc_cols_%s)", rk),
      "```"), envir = environment(), quiet = TRUE))
    cat("\n\n")
  }
}
if (length(.leading_edge_summary_rows) > 0) {
  readr::write_tsv(dplyr::bind_rows(.leading_edge_summary_rows), "posenrich_leading_edge_summary.tsv")
}
```

# Term recurrence across direction lists

Every significant term/pathway, pooled across the three direction lists
(**GLOBAL**/**TOP**/**BOTTOM**) into one table - not one table per direction.
Recurrence *within* a single direction's own nested percentile cutoffs
(`top25` ⊃ `top10` ⊃ `top5` ⊃ `top1`) isn't an independent signal: a term is
mechanically less likely to survive the tighter, smaller foreground, so
"recovered at fewer cutoffs" mostly just restates the nesting rather than
telling you anything new. Recurrence *across* the three direction lists is
the meaningful comparison - a term significant under both **Top** and
**Bottom** (or all three) is corroborated by genuinely independent
foreground constructions, not by a subset relationship.

`N_lists`/`Lists` say how many of, and which of, GLOBAL/TOP/BOTTOM recovered
the term (at any of its own percentile cutoffs) - no other column, since
nothing else is comparable across three direction lists built from different
score signs. Underneath, a **term x direction-list** heatmap makes the same
recurrence pattern visually scannable.

```{r module-rollup-setup, include=FALSE}
# .normalize_term/.term_contains/.cluster_terms are defined earlier
# (term-clustering-setup, above the Overall dotplot section) so the Overall
# dotplot, Direction triangle, and this section's Recurrence Table/Heatmap
# all merge near-duplicate term descriptions the same way.

# Term x direction-list heatmap, one tile per (canonical Term, GLOBAL/TOP/
# BOTTOM) it was recovered under, replacing the old term x percentile-cutoff
# heatmap: cutoff-level recurrence within one direction is a tautology of
# top1 ⊂ top5 ⊂ top10 ⊂ top25, not an independent signal (see section intro
# above). Takes `rec_pairs` post-clustering (already carries the canonical
# `Term` column, one row per database/pathway/list) so the heatmap's merge
# is identical to the Recurrence Table's, not a second independent one.
.render_posenrich_list_heatmap <- function(rec_pairs) {
  if (nrow(rec_pairs) == 0) { cat("_No terms to show._\n\n"); return(invisible(NULL)) }
  pres <- rec_pairs %>% dplyr::distinct(Term, list_name)

  list_levels <- intersect(c("GLOBAL", "TOP", "BOTTOM"), unique(pres$list_name))

  # Term order (most lists recovered first), a term merged from several
  # pathways can now recur across more lists than any single contributing
  # pathway did on its own.
  term_counts <- pres %>% dplyr::count(Term, name = "n_lists") %>%
    dplyr::arrange(dplyr::desc(n_lists), Term)
  term_order <- term_counts$Term

  pres <- pres %>%
    dplyr::mutate(list_name = factor(list_name, levels = list_levels),
                  Term = factor(Term, levels = rev(term_order)))
  n_terms <- length(term_order)
  p <- ggplot(pres, aes(x = list_name, y = Term)) +
    geom_tile(fill = "#2c7fb8", colour = "white", linewidth = 0.6) +
    scale_x_discrete(drop = FALSE) +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 0),
          axis.text.y = element_text(size = pmax(6, pmin(10, 260 / max(n_terms, 1))))) +
    labs(title = "Recovered terms by direction list", x = "direction list", y = NULL)
  print(p)
  cat("\n\n")
}
```

```{r module-rollup, echo=FALSE, results='asis'}
if (!has_res || !any(res$sig, na.rm = TRUE)) {
  cat("_No significant sets._\n\n")
} else {
  sig_res <- res %>% filter(sig)

  # One row per (database, pathway, direction list actually recovering it) --
  # the base unit BEFORE similarity clustering, so a term split across two
  # databases (or two near-duplicate GMT entries) is unioned into ONE row
  # below rather than surviving as several rows that merely display the same
  # text. Pooled across that list's own percentile cutoffs.
  rec_pairs <- sig_res %>%
    dplyr::mutate(description = dplyr::coalesce(description, pathway),
                  list_name = toupper(ranking)) %>%
    dplyr::distinct(database, pathway, description, list_name)

  if (nrow(rec_pairs) == 0) {
    cat("_No significant sets._\n\n")
  } else {
    # Cluster map (near-duplicate/nested description text -> canonical
    # label) drives the actual merge below, two (database, pathway) pairs
    # mapping to the same canonical Term become ONE row, with their
    # Database/Lists/N_lists all pooled across the merge, not just relabelled.
    desc_levels <- unique(rec_pairs$description)
    cluster_map <- setNames(.cluster_terms(desc_levels), desc_levels)
    rec_pairs <- rec_pairs %>% dplyr::mutate(Term = unname(cluster_map[description]))

    disp <- rec_pairs %>%
      dplyr::group_by(Term) %>%
      dplyr::summarise(
        Database = paste(sort(unique(database)), collapse = ", "),
        Lists = paste(sort(unique(list_name)), collapse = ", "),
        N_lists = dplyr::n_distinct(list_name),
        .groups = "drop"
      ) %>%
      dplyr::arrange(dplyr::desc(N_lists), Term) %>%
      dplyr::select(Database, Term, N_lists, Lists)

    cat("\n\n## Recurrence Table\n\n")
    cat(knitr::knit_child(text = c(
      "```{r dt_rollup_all, echo=FALSE}",
      "DT::datatable(disp, options = list(pageLength = 10, scrollX = TRUE), rownames = FALSE)",
      "```"), envir = environment(), quiet = TRUE))
    cat("\n\n")

    cat("\n\n## Recurrence Heatmap\n\n")
    .render_posenrich_list_heatmap(rec_pairs)
  }
}
```

---

```{r poschar-setup, include=FALSE}
.file_ok <- function(p) !is.null(p) && !is.na(p) && nchar(p) > 0 && file.exists(p)

# Tableau_10 colour swatch for pipeline-wide consistency (same palette used
# throughout the Scoring/Characterization reports).
if (requireNamespace("paletteer", quietly = TRUE)) {
  .tableau10 <- as.character(paletteer::paletteer_d("ggthemes::Tableau_10"))
} else {
  .tableau10 <- c(
    "#4E79A7", "#F28E2B", "#E15759", "#76B7B2", "#59A14F",
    "#EDC948", "#B07AA1", "#FF9DA7", "#9C755F", "#BAB0AC"
  )
}

has_position_scores <- .file_ok(params$position_scores_file)
has_gene_scores      <- .file_ok(params$gene_scores_file)
pos_df  <- if (has_position_scores) readr::read_tsv(params$position_scores_file, show_col_types = FALSE) else NULL
gene_df <- if (has_gene_scores)     readr::read_tsv(params$gene_scores_file,     show_col_types = FALSE) else NULL

has_vep_primateai <- .file_ok(params$vep_primateai_file)
has_vep_cosmic    <- .file_ok(params$vep_cosmic_file)
has_genomic       <- .file_ok(params$genomic_info_file)
# has_fade_sites_top/bottom + fade_sites_top_df/bottom_df already loaded in the
# `load` chunk near the top of this file (hoisted so the Overall dotplot's
# position-native FADE-overlap component can use them too).
# Position scores are required for every sub-tab below (PAI3D/COSMIC/FADE joins,
# regional accumulation), gate the whole section on them, not just on which
# external database happens to be supplied.
has_pos_char <- has_position_scores &&
  (has_vep_primateai || has_vep_cosmic || has_genomic || has_fade_sites_top || has_fade_sites_bottom)

if (has_vep_primateai) {
  vep_pai_df <- readr::read_tsv(params$vep_primateai_file, show_col_types = FALSE)
  if ("Position" %in% names(vep_pai_df)) vep_pai_df$Position <- suppressWarnings(as.integer(vep_pai_df$Position))
} else {
  vep_pai_df <- NULL
}

if (has_vep_cosmic) {
  vep_cosmic_df <- readr::read_tsv(params$vep_cosmic_file, show_col_types = FALSE)
  if ("Position" %in% names(vep_cosmic_df)) vep_cosmic_df$Position <- suppressWarnings(as.integer(vep_cosmic_df$Position))
} else {
  vep_cosmic_df <- NULL
}

if (has_genomic) {
  # Format: gene, chr, start, end, length (no header assumed; detect if present)
  genomic_raw <- readr::read_tsv(params$genomic_info_file,
                          col_names = c("Gene", "chr", "gene_start", "gene_end", "gene_length"),
                          show_col_types = FALSE)
  if (tolower(genomic_raw$Gene[1]) == "gene") genomic_raw <- genomic_raw[-1, ]
  genomic_raw <- genomic_raw %>%
    dplyr::mutate(
      gene_start  = as.integer(gene_start),
      gene_end    = as.integer(gene_end),
      gene_length = as.integer(gene_length),
      gene_mid    = as.integer((gene_start + gene_end) / 2L)
    )
} else {
  genomic_raw <- NULL
}

# Join scored positions with genomic coordinates via their gene, and lay
# out chromosome-cumulative offsets for the regional-accumulation section
# below, each position inherits its gene's genomic midpoint as x-coordinate.
if (has_genomic && has_position_scores) {
  pos_genomic <- pos_df %>%
    dplyr::inner_join(genomic_raw, by = "Gene") %>%
    dplyr::mutate(
      chr_label = sub("^chr", "", chr),
      chr_num   = suppressWarnings(as.integer(chr_label)),
      chr_order = dplyr::case_when(
        !is.na(chr_num)              ~ chr_num,
        chr_label == "X"             ~ 23L,
        chr_label == "Y"             ~ 24L,
        chr_label %in% c("MT", "M")  ~ 25L,
        TRUE                         ~ 26L
      )
    ) %>%
    dplyr::arrange(chr_order, gene_mid)

  pos_chr_offsets <- pos_genomic %>%
    dplyr::group_by(chr_order, chr_label) %>%
    dplyr::summarise(chr_max = max(gene_end, na.rm = TRUE), .groups = "drop") %>%
    dplyr::arrange(chr_order) %>%
    dplyr::mutate(offset = cumsum(dplyr::lag(chr_max, default = 0L)) + (chr_order - 1L) * 5e6L)

  pos_genomic <- pos_genomic %>%
    dplyr::left_join(pos_chr_offsets %>% dplyr::select(chr_order, offset), by = "chr_order") %>%
    dplyr::mutate(
      manhattan_x = gene_mid + offset,
      chr_even    = chr_order %% 2 == 0
    )
} else {
  pos_genomic <- NULL
  pos_chr_offsets <- NULL
}

if (has_pos_char) {
  top_pct  <- params$top_pct %||% 0.10
  top5_pct <- 0.05
  top1_pct <- 0.01

  pos_top    <- pos_df %>% dplyr::filter(change_side %in% c("top",    "both"))
  pos_bottom <- pos_df %>% dplyr::filter(change_side %in% c("bottom", "both"))

  # Percentile membership: prefer SCORING's published position_lists/
  # slice_<dir><pct>.tsv (scoring_compute.R), the same canonical top/bottom/
  # global x 10/5/1% cutoff 15.Comparison_report.Rmd and posenrich both read,
  # over an independently re-derived quantile() cut. Each .pos_pct_test()
  # call returns a CLOSURE (not a bare vector) so it can be applied to
  # whichever population later needs it (pos_score_meta, pos_genomic_top/bot
  # are different row subsets of pos_df, so the SAME membership set/threshold
  # must be evaluated against each one individually, not computed once and
  # reused positionally). Falls back to quantile() on `pop` (the direction's
  # own population) when position_lists_dir is NULL or a slice file is
  # missing, all 3 files for a direction, never a mix.
  .pos_slice_ids <- function(dir_key, pct) {
    if (is.null(params$position_lists_dir)) return(NULL)
    f <- file.path(params$position_lists_dir, sprintf("slice_%s%d.tsv", dir_key, pct))
    if (!file.exists(f)) return(NULL)
    d <- tryCatch(readr::read_tsv(f, show_col_types = FALSE), error = function(e) NULL)
    if (is.null(d) || !all(c("Gene", "Position") %in% names(d))) return(NULL)
    unique(paste(d$Gene, d$Position))
  }
  .pos_pct_test <- function(dir_key, pct_frac, pop) {
    ids_set <- .pos_slice_ids(dir_key, round(pct_frac * 100))
    if (!is.null(ids_set)) {
      function(df) paste(df$Gene, df$Position) %in% ids_set
    } else {
      thr <- quantile(pop$CAAS_score, 1 - pct_frac, na.rm = TRUE)
      function(df) df$CAAS_score >= thr
    }
  }

  # top25_pct: only used by the PAI3D/COSMIC/FADE "characterization by CAAS
  # threshold" Fisher-OR sweep tables below, replacing their old ad hoc 50%
  # row (quantile(0.50), no SCORING equivalent) with a 25% row that DOES have
  # one (position_lists/slice_<dir>25.tsv), so every percentage row in
  # those tables except the trivial 100% baseline now comes from the same
  # published source. Not used for in_top_Xpct (only ever a 10/5/1% trio
  # elsewhere in this report).
  top25_pct <- 0.25

  test_glob_25 <- .pos_pct_test("global", top25_pct, pos_df)
  test_glob_10 <- .pos_pct_test("global", top_pct,  pos_df)
  test_glob_5  <- .pos_pct_test("global", top5_pct, pos_df)
  test_glob_1  <- .pos_pct_test("global", top1_pct, pos_df)
  test_top_25  <- .pos_pct_test("top",    top25_pct, pos_top)
  test_top_10  <- .pos_pct_test("top",    top_pct,  pos_top)
  test_top_5   <- .pos_pct_test("top",    top5_pct, pos_top)
  test_top_1   <- .pos_pct_test("top",    top1_pct, pos_top)
  test_bot_25  <- .pos_pct_test("bottom", top25_pct, pos_bottom)
  test_bot_10  <- .pos_pct_test("bottom", top_pct,  pos_bottom)
  test_bot_5   <- .pos_pct_test("bottom", top5_pct, pos_bottom)
  test_bot_1   <- .pos_pct_test("bottom", top1_pct, pos_bottom)

  # thr_10: cosmetic-only scalar for the geom_vline reference line further
  # below (a plot annotation has no equivalent to a discrete membership set) --
  # always a plain quantile, independent of whether the tests above used the
  # published path.
  thr_10 <- quantile(pos_df$CAAS_score, 1 - top_pct, na.rm = TRUE)
}
```
---

```{r pos-char-header, results='asis'}
if (has_pos_char) cat(r"(# Position Characterisation {.tabset #position-characterisation}

This section examines how ranked positions behave in relation to external
signals that do **not** contribute to the `CAAS_score`. PrimateAI-3D, COSMIC
Mutant Census, and FADE branch-site selection evidence are treated as
independent characterisation layers.

PrimateAI-3D and COSMIC both describe specific amino acid transitions, so the
biochemical Grantham Score (GS) groupings are out of scope for those two (as
they group transitions by biochemical radicalness rather than modeling the
specific substitution directly). We restrict those two layers strictly to the
**Unweighted Scheme (US)**, with all scores and mutation overlaps derived
solely from US transitions. FADE evidence is not transition-specific (it is a
per-site test of dN/dS shift along phenotype-associated branches) and so is
not restricted to the US scheme.

- **`pai_char_score`**: The pathogenicity score of mutating the ancestral human reference amino acid to the derived US alternative amino acid(s) in PrimateAI-3D.
- **COSMIC Overlap**: The presence and recurrence of somatic mutations mapped to the specific codon under the US transition.
- **FADE Overlap**: Whether the position falls within the direction-specific FADE_top_sig/FADE_bottom_sig significant site set (BF ≥ `fade_bf_threshold`), and its maximum Bayes Factor.

)")
```

```{r vep-setup-common}
# Computed against pos_df directly (test_glob_*() need Gene/Position/CAAS_score,
# all present there), select() below only drops columns, never reorders/
# filters rows, so assigning these vectors positionally into pos_score_meta
# stays row-aligned with pos_df.
.glob10 <- test_glob_10(pos_df); .glob5 <- test_glob_5(pos_df); .glob1 <- test_glob_1(pos_df)

pos_score_meta <- pos_df %>%
  select(Gene, Position, CAAS_score, n_schemes, change_side, gate_sig) %>%
  mutate(
    caas_decile  = dplyr::ntile(CAAS_score, 10),
    in_top_10pct = .glob10,
    in_top_5pct  = .glob5,
    in_top_1pct  = .glob1
  )

has_background <- .file_ok(params$universe_file)
if (has_background) {
  bg_lines <- readLines(params$universe_file)
  bg_genes <- bg_lines[nzchar(trimws(bg_lines)) & !grepl("^#", bg_lines)]
} else {
  bg_genes <- unique(gene_df$Gene)
}
```

```{r vep-pai-subsec, results='asis'}
if (has_vep_primateai) cat("## VEP: PrimateAI-3D\n\n")
```

```{r vep-pai-setup, eval=has_vep_primateai}
# Detect PrimateAI score column (varies by database version)
.pai_score_col <- intersect(
  c("score_PAI3D", "score", "primateai_score", "pathogenicity_score", "PrimateAI_score"),
  names(vep_pai_df)
)[1]
if (is.na(.pai_score_col)) {
  warning("PrimateAI: no recognised score column found - skipping.")
  has_vep_primateai <- FALSE
}
```

```{r vep-pai-join, eval=has_vep_primateai}
# Restrict to shared genes present in both our results (background clean) and database
shared_genes_pai <- intersect(bg_genes, unique(vep_pai_df$Gene))
pos_pai_shared <- pos_score_meta %>% filter(Gene %in% shared_genes_pai)

# Under US-only design, pai_char_score is simply the pathogenicity score of the US change.
# If multiple alternative AAs map, we keep the maximum score.
# Pathogenic status is defined exclusively by the prediction column (containing
# "pathogenic", case-insensitive) - never by a raw score cutoff.
pai_by_prot <- vep_pai_df %>%
  filter(caap_group == "US", Gene %in% shared_genes_pai) %>%
  mutate(
    pai_score = as.numeric(.data[[.pai_score_col]]),
    is_pathogenic = grepl("pathogenic", prediction, ignore.case = TRUE)
  ) %>%
  filter(is.finite(pai_score)) %>%
  group_by(Gene, Position) %>%
  summarise(
    pai_char_score = max(pai_score, na.rm = TRUE),
    pai_pathogenic = any(is_pathogenic, na.rm = TRUE),
    n_pai_schemes  = 1L,
    .groups = "drop"
  )

pos_vep_pai <- pai_by_prot %>%
  inner_join(pos_pai_shared, by = c("Gene", "Position")) %>%
  distinct(Gene, Position, .keep_all = TRUE)

# Join back caas_change and other details for display table
pos_vep_pai_table_source <- vep_pai_df %>%
  filter(caap_group == "US", Gene %in% shared_genes_pai) %>%
  inner_join(pos_pai_shared, by = c("Gene", "Position")) %>%
  distinct(Gene, Position, .keep_all = TRUE)

has_vep_primateai <- nrow(pos_vep_pai) > 0

# Directional PAI subsets: top+both and bottom+both
pos_vep_pai_top    <- pos_vep_pai %>% filter(change_side %in% c("top",    "both"))
pos_vep_pai_bottom <- pos_vep_pai %>% filter(change_side %in% c("bottom", "both"))
```

`r if (exists("pos_vep_pai") && nrow(pos_vep_pai) == 0) "> **Note:** PrimateAI-3D inputs were provided, but no scored positions could be matched to the VEP/PAI3D annotations for this direction."`

```{r vep-pai-violin, eval=has_vep_primateai, fig.height=7}
# pai_char_score violin per CAAS decile, split by direction (TOP and BOTTOM separately).
pai_dir_df <- bind_rows(
  pos_vep_pai_top    %>% mutate(pai_direction = "TOP (top + both)"),
  pos_vep_pai_bottom %>% mutate(pai_direction = "BOTTOM (bottom + both)")
) %>%
  filter(!is.na(caas_decile), !is.na(pai_char_score))

# Per-direction median for horizontal reference lines
pai_medians <- pai_dir_df %>%
  group_by(pai_direction) %>%
  summarise(median_pai = median(pai_char_score, na.rm = TRUE), .groups = "drop")

ggplot(pai_dir_df, aes(x = factor(caas_decile, levels = 1:10), y = pai_char_score)) +
  geom_violin(aes(fill = pai_direction), alpha = 0.55, trim = TRUE, scale = "width",
              show.legend = FALSE) +
  geom_boxplot(width = 0.14, outlier.size = 0.4, alpha = 0.9,
               fill = "white", colour = "grey30") +
  geom_hline(data = pai_medians,
             aes(yintercept = median_pai),
             linetype = "dashed", colour = "grey20", linewidth = 0.7) +
  geom_text(data = pai_medians,
            aes(x = 10.6, y = median_pai,
                label = sprintf("median = %.3f", median_pai)),
            hjust = 0, vjust = -0.4, size = 3, colour = "grey20",
            inherit.aes = FALSE) +
  scale_fill_manual(values = c("TOP (top + both)" = .tableau10[1],
                               "BOTTOM (bottom + both)" = .tableau10[3])) +
  scale_x_discrete(
    drop   = FALSE,
    labels = setNames(c("1\n(low)", as.character(2:9), "10\n(high)"), as.character(1:10))
  ) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.08))) +
  facet_wrap(~pai_direction, ncol = 1) +
  labs(
    title    = "PAI_TOP and PAI_BOTTOM: PrimateAI-3D Score by CAAS Decile",
    subtitle = sprintf(
      "PAI_TOP: %d positions | PAI_BOTTOM: %d positions | dashed = per-direction median",
      nrow(pos_vep_pai_top), nrow(pos_vep_pai_bottom)
    ),
    x = "CAAS Score Decile", y = "PrimateAI-3D Score"
  ) +
  theme(plot.title = element_text(face = "bold"),
        strip.text = element_text(face = "bold", size = 11))
```

```{r vep-pai-scatter, eval=has_vep_primateai, fig.height=5}
# CAAS_score vs pai_char_score, coloured by direction.
# Normalise both axes for Pearson r (rank / n).
#
# NOTE the reported r is computed on WITHIN-RUN PERCENTILE RANKS of both axes, not
# on their raw values, so it is a Spearman-style rank correlation reported under a
# Pearson call, and its magnitude is not comparable to an r computed on raw scores.
# Ranking is deliberate: CAAS_score's own phenotype component is already a
# percent_rank (see scoring_compute.R sec. 2f / 11.Scoring_report.Rmd), and the raw
# statistics underneath are floor-limited by the resampling resolution (1/N), so raw
# correlations would be dominated by ties at the floor. The same caveat applies to
# every "(rank-norm)" subtitle in this report: values are only interpretable relative
# to other positions in THIS run, never across runs or traits.
pos_vep_pai_dir <- bind_rows(
  pos_vep_pai_top    %>% mutate(pai_direction = "TOP"),
  pos_vep_pai_bottom %>% mutate(pai_direction = "BOTTOM")
) %>% distinct(Gene, Position, .keep_all = TRUE)

.pai_caas_r_top  <- safe_cor(
  rank(pos_vep_pai_top$CAAS_score,    na.last = "keep") / sum(!is.na(pos_vep_pai_top$CAAS_score)),
  rank(pos_vep_pai_top$pai_char_score, na.last = "keep") / sum(!is.na(pos_vep_pai_top$pai_char_score))
)
.pai_caas_r_bot  <- safe_cor(
  rank(pos_vep_pai_bottom$CAAS_score,    na.last = "keep") / sum(!is.na(pos_vep_pai_bottom$CAAS_score)),
  rank(pos_vep_pai_bottom$pai_char_score, na.last = "keep") / sum(!is.na(pos_vep_pai_bottom$pai_char_score))
)

ggplot(pos_vep_pai_dir, aes(x = CAAS_score, y = pai_char_score,
                             colour = pai_direction)) +
  geom_point(alpha = 0.35, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8, alpha = 0.15) +
  geom_vline(xintercept = thr_10, linetype = "dashed",
             colour = "grey40", linewidth = 0.5) +
  scale_colour_manual(values = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[3]),
                      name = "Direction") +
  labs(
    title    = "CAAS Score vs PrimateAI-3D Score by Direction",
    subtitle = sprintf("Pearson r (rank-norm): TOP = %.3f | BOTTOM = %.3f | dashed = top-10%% CAAS",
                       .pai_caas_r_top, .pai_caas_r_bot),
    x = "CAAS Score", y = "PrimateAI-3D Score"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r vep-pai-pathogenic-bar, eval=has_vep_primateai, fig.height=5}
# Count of pathogenic vs non-pathogenic positions, TOP and BOTTOM.
pai_pathogenic_counts <- bind_rows(
  pos_vep_pai_top    %>% mutate(pai_direction = "TOP"),
  pos_vep_pai_bottom %>% mutate(pai_direction = "BOTTOM")
) %>%
  mutate(pathogenic = ifelse(pai_pathogenic, "pathogenic", "non-pathogenic")) %>%
  count(pai_direction, pathogenic)

ggplot(pai_pathogenic_counts, aes(x = pathogenic, y = n, fill = pai_direction)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = n), position = position_dodge(width = 0.8), vjust = -0.3, size = 3.5) +
  scale_fill_manual(values = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[3])) +
  labs(
    title    = "Pathogenic vs Non-Pathogenic Positions (PrimateAI-3D)",
    subtitle = "Pathogenic defined by prediction column",
    x = NULL, y = "N positions", fill = "Direction"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r vep-pai-phenaxis-scatter, eval=has_vep_primateai && "phen_score" %in% names(pos_vep_pai_top), fig.height=5}
.pai_phen_r_top <- safe_cor(
  rank(pos_vep_pai_top$phen_score,      na.last = "keep") / sum(!is.na(pos_vep_pai_top$phen_score)),
  rank(pos_vep_pai_top$pai_char_score,  na.last = "keep") / sum(!is.na(pos_vep_pai_top$pai_char_score))
)
.pai_phen_r_bot <- safe_cor(
  rank(pos_vep_pai_bottom$phen_score,      na.last = "keep") / sum(!is.na(pos_vep_pai_bottom$phen_score)),
  rank(pos_vep_pai_bottom$pai_char_score,  na.last = "keep") / sum(!is.na(pos_vep_pai_bottom$pai_char_score))
)

ggplot(pos_vep_pai_dir, aes(x = phen_score, y = pai_char_score,
                             colour = pai_direction)) +
  geom_point(alpha = 0.35, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8, alpha = 0.15) +
  scale_colour_manual(values = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[3]),
                      name = "Direction") +
  labs(
    title    = "Phenotype Evolution Score vs PrimateAI-3D Score by Direction",
    subtitle = sprintf("Pearson r (rank-norm): TOP = %.3f | BOTTOM = %.3f",
                       .pai_phen_r_top, .pai_phen_r_bot),
    x = "phen_score (1 - pvalue_boot)", y = "PrimateAI-3D Score"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r vep-pai-table, eval=has_vep_primateai}
pai_table <- pos_vep_pai_table_source %>%
  arrange(desc(CAAS_score)) %>%
  select(Gene, Position, hg38_ref_aa, caas_alt_aas, caas_change, CAAS_score, change_side, score_PAI3D, percentile_PAI3D, prediction) %>%
  rename(
    `Ref AA` = hg38_ref_aa,
    `Alt AA(s)` = caas_alt_aas,
    `CAAS Change` = caas_change,
    `CAAS Score` = CAAS_score,
    `Change Side` = change_side,
    `PAI3D Score` = score_PAI3D,
    `PAI3D Percentile` = percentile_PAI3D,
    `Prediction` = prediction
  )

DT::datatable(
  pai_table,
  caption = "CAAS positions overlapping with PrimateAI-3D pathogenicity predictions, ordered by CAAS score",
  options = list(pageLength = 10, scrollX = TRUE),
  rownames = FALSE
)
```

```{r vep-pai-fisher, eval=has_vep_primateai}
# Odds ratio & P-value interpretation:
# - Odds Ratio: Measures the enrichment of pathogenic residues (by the prediction column) within the specified CAAS threshold compared to the rest of the positions. An odds ratio > 1 indicates enrichment.
# - P-value: Reaches significance (via Fisher's exact test) when the proportion of pathogenic mutations is significantly higher inside the CAAS threshold.
# `N positions` IS the Fisher test's actual population (PAI-scored, shared-gene
# positions), mirrors run_cosmic_fisher_dir below.
run_pai_fisher_dir <- function(df, label) {
  df   <- df %>% mutate(high_pai = !is.na(pai_pathogenic) & pai_pathogenic)
  n_all <- nrow(df)

  # Build named threshold list: CAAS-flag-based thresholds + percentage-based
  # rows. 100% is trivial (the full population, no computation needed); every
  # other row, 25/10/5/1%, comes from SCORING's published position_lists
  # slices (test_glob_25/in_top_Xpct), not an independently re-derived
  # quantile() cut. 25% replaces the old ad hoc 50% row (quantile(0.50) had no
  # published equivalent; SCORING only ever publishes 25/10/5/1%).
  pct_rows <- list(
    list(col = NULL, lbl = "100% (all positions)",    flag = rep(TRUE, n_all)),
    list(col = NULL, lbl = "25% (top 25% CAAS)",      flag = test_glob_25(df)),
    list(col = "in_top_10pct", lbl = sprintf("Top %g%%", top_pct  * 100), flag = NULL),
    list(col = "in_top_5pct",  lbl = sprintf("Top %g%%", top5_pct * 100), flag = NULL),
    list(col = "in_top_1pct",  lbl = sprintf("Top %g%%", top1_pct * 100), flag = NULL)
  )

  purrr::map_dfr(pct_rows, function(thr) {
    in_top   <- if (!is.null(thr$flag)) thr$flag else df[[thr$col]]
    in_top   <- replace_na(in_top, FALSE)

    n_top    <- sum(in_top)
    n_top_h  <- sum(in_top & df$high_pai, na.rm = TRUE)
    n_rest   <- sum(!in_top)
    n_rest_h <- sum(!in_top & df$high_pai, na.rm = TRUE)
    # Column order is [flag, not-flag] (not [not-flag, flag]) so fisher.test's
    # odds ratio reads as "odds of pathogenic in-threshold vs out-of-threshold"
    # the right way round, verified against an unambiguous 90%-vs-10%
    # synthetic case (OR=77.3, not the inverted 0.013 the other order gives).
    mat <- matrix(c(n_top_h, n_rest_h, n_top - n_top_h, n_rest - n_rest_h), nrow = 2)
    ft  <- tryCatch(fisher.test(mat),
                    error = function(e) list(estimate = NA, conf.int = c(NA, NA), p.value = NA))

    pct_sig <- if ("gate_sig" %in% names(df)) sprintf("%.1f%%", 100 * mean(df$gate_sig[in_top], na.rm = TRUE)) else "0.0%"

    tibble(
      Direction            = label,
      Threshold            = thr$lbl,
      `N positions`        = n_top,
      `N pathogenic`       = n_top_h,
      `% pathogenic`       = sprintf("%.1f%%", 100 * n_top_h / max(n_top, 1)),
      `% gate_sig`         = pct_sig,
      `Odds ratio`         = sprintf("%.2f", ft$estimate),
      `P value`            = formatC(ft$p.value, format = "e", digits = 2)
    )
  })
}

pai_fisher <- bind_rows(
  run_pai_fisher_dir(pos_vep_pai_top,    "TOP"),
  run_pai_fisher_dir(pos_vep_pai_bottom, "BOTTOM")
)

knitr::kable(
  pai_fisher,
  caption = "PAI3D enrichment of pathogenic positions (defined by prediction column) by CAAS threshold and direction. N positions = the actual test population (shared-gene, PAI3D-scored positions) at that threshold. 100% row = all directional positions (baseline); 25/10/5/1% rows use SCORING's published position percentile slices; Fisher OR and P value compare top-subset vs remainder, entirely within PAI-covered positions.",
  align = c("l", "l", rep("r", ncol(pai_fisher) - 2))
)
```


```{r vep-cosmic-subsec, results='asis'}
if (has_vep_cosmic) cat("## VEP: COSMIC Mutant Census\n\n")
```

```{r vep-cosmic-setup, eval=has_vep_cosmic}
# Restrict to shared genes present in both our results (background clean) and database
shared_genes_cosmic <- intersect(bg_genes, unique(vep_cosmic_df$Gene))
pos_cosmic_shared <- pos_score_meta %>% filter(Gene %in% shared_genes_cosmic)

# Filter cosmic annotations for US only if caap_group is present
if ("caap_group" %in% names(vep_cosmic_df)) {
  vep_cosmic_df <- vep_cosmic_df %>% filter(caap_group == "US", Gene %in% shared_genes_cosmic)
} else {
  vep_cosmic_df <- vep_cosmic_df %>% filter(Gene %in% shared_genes_cosmic)
}

# Join cosmic annotations with pos_cosmic_shared (keeping all rows to count recurrence)
cosmic_joined <- vep_cosmic_df %>%
  inner_join(pos_cosmic_shared, by = c("Gene", "Position"))

has_vep_cosmic <- nrow(cosmic_joined) > 0
```

`r if (exists("cosmic_joined") && nrow(cosmic_joined) == 0) "> **Note:** COSMIC Mutant Census mutations were mapped, but no scored positions could be matched to the COSMIC annotations for this direction."`

```{r vep-cosmic-stats, eval=has_vep_cosmic}
cosmic_table <- cosmic_joined %>%
  group_by(Gene, Position, hg38_ref_aa, caas_alt_aas, caas_change, CAAS_score, change_side) %>%
  summarise(
    Recurrence = n(),
    `Mutation AA(s)` = paste(unique(MUTATION_AA), collapse = "; "),
    `Description(s)` = paste(unique(MUTATION_DESCRIPTION), collapse = "; "),
    `Somatic Status(es)` = paste(unique(MUTATION_SOMATIC_STATUS), collapse = "; "),
    .groups = "drop"
  ) %>%
  arrange(desc(Recurrence)) %>%
  rename(
    `Ref AA` = hg38_ref_aa,
    `Alt AA(s)` = caas_alt_aas,
    `CAAS Change` = caas_change,
    `CAAS Score` = CAAS_score,
    `Change Side` = change_side,
    `Mutation AA(s)` = `Mutation AA(s)`,
    `Description` = `Description(s)`,
    `Somatic Status` = `Somatic Status(es)`
  )

DT::datatable(
  cosmic_table,
  caption = "CAAS positions overlapping with COSMIC cancer-related mutations, ordered by mutation recurrence",
  options = list(pageLength = 10, scrollX = TRUE),
  rownames = FALSE
)
```

```{r vep-cosmic-somatic-bar, eval=has_vep_cosmic, fig.height=5}
# Count of COSMIC mutations by Somatic Status and direction.
cosmic_counts <- bind_rows(
  cosmic_joined %>% filter(change_side %in% c("top",    "both")) %>% mutate(cosmic_direction = "TOP"),
  cosmic_joined %>% filter(change_side %in% c("bottom", "both")) %>% mutate(cosmic_direction = "BOTTOM")
) %>%
  filter(!is.na(MUTATION_SOMATIC_STATUS) & MUTATION_SOMATIC_STATUS != "") %>%
  count(cosmic_direction, MUTATION_SOMATIC_STATUS)

ggplot(cosmic_counts, aes(x = MUTATION_SOMATIC_STATUS, y = n, fill = cosmic_direction)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = n), position = position_dodge(width = 0.8), vjust = -0.3, size = 3.5) +
  scale_fill_manual(values = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[3])) +
  labs(
    title    = "COSMIC Mutation Somatic Status Distribution",
    x = NULL, y = "N mutations", fill = "Direction"
  ) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r vep-cosmic-fisher, eval=has_vep_cosmic}
# Odds ratio & P-value interpretation:
# - Odds Ratio: Measures the enrichment of positions with somatic mutations reported in the COSMIC database within the specified CAAS threshold compared to the rest of the positions. An odds ratio > 1 indicates enrichment.
# - P-value: Reaches significance (via Fisher's exact test) when the proportion of cancer somatic mutations is significantly higher inside the CAAS threshold.
pos_cosmic <- pos_cosmic_shared %>%
  mutate(has_cosmic = paste(Gene, Position, sep = "_") %in% paste(vep_cosmic_df$Gene, vep_cosmic_df$Position, sep = "_"))

pos_cosmic_top    <- pos_cosmic %>% filter(change_side %in% c("top",    "both"))
pos_cosmic_bottom <- pos_cosmic %>% filter(change_side %in% c("bottom", "both"))

run_cosmic_fisher_dir <- function(df, label) {
  n_all <- nrow(df)
  pct_rows <- list(
    list(col = NULL, lbl = "100% (all positions)",    flag = rep(TRUE, n_all)),
    list(col = NULL, lbl = "25% (top 25% CAAS)",      flag = test_glob_25(df)),
    list(col = "in_top_10pct", lbl = sprintf("Top %g%%", top_pct  * 100), flag = NULL),
    list(col = "in_top_5pct",  lbl = sprintf("Top %g%%", top5_pct * 100), flag = NULL),
    list(col = "in_top_1pct",  lbl = sprintf("Top %g%%", top1_pct * 100), flag = NULL)
  )

  purrr::map_dfr(pct_rows, function(thr) {
    in_top   <- if (!is.null(thr$flag)) thr$flag else df[[thr$col]]
    in_top   <- replace_na(in_top, FALSE)
    n_top    <- sum(in_top)
    n_top_c  <- sum(in_top & df$has_cosmic, na.rm = TRUE)
    n_rest   <- sum(!in_top)
    n_rest_c <- sum(!in_top & df$has_cosmic, na.rm = TRUE)
    # Column order is [flag, not-flag] (not [not-flag, flag]) so fisher.test's
    # odds ratio reads as "odds of COSMIC overlap in-threshold vs out-of-
    # threshold" the right way round, verified against an unambiguous
    # 90%-vs-10% synthetic case (OR=77.3, not the inverted 0.013 the other
    # order gives).
    mat <- matrix(c(n_top_c, n_rest_c, n_top - n_top_c, n_rest - n_rest_c), nrow = 2)
    ft  <- tryCatch(fisher.test(mat),
                    error = function(e) list(estimate = NA, conf.int = c(NA, NA), p.value = NA))
    
    pct_sig <- if ("gate_sig" %in% names(df)) sprintf("%.1f%%", 100 * mean(df$gate_sig[in_top], na.rm = TRUE)) else "0.0%"

    tibble(
      Direction            = label,
      Threshold            = thr$lbl,
      `N positions`        = n_top,
      `N COSMIC`           = n_top_c,
      `% COSMIC`           = sprintf("%.1f%%", 100 * n_top_c / max(n_top, 1)),
      `% gate_sig`         = pct_sig,
      `Odds ratio`         = sprintf("%.2f", ft$estimate),
      `P value`            = formatC(ft$p.value, format = "e", digits = 2)
    )
  })
}

cosmic_fisher <- bind_rows(
  run_cosmic_fisher_dir(pos_cosmic_top,    "TOP"),
  run_cosmic_fisher_dir(pos_cosmic_bottom, "BOTTOM")
)

knitr::kable(
  cosmic_fisher,
  caption = "Fisher exact test for COSMIC somatic mutations overlap enrichment within CAAS score thresholds",
  align = c("l", "l", rep("r", ncol(cosmic_fisher) - 2))
)
```


```{r fade-subsec, results='asis'}
if (has_fade_sites_top || has_fade_sites_bottom) cat("## FADE Site Overlap\n\n")
```

```{r fade-setup, eval=has_fade_sites_top || has_fade_sites_bottom}
# Direction-specific by construction: TOP positions are only ever checked
# against fade_sites_top, BOTTOM positions only against fade_sites_bottom --
# never cross-checked (mirrors the FADE_top_sig/FADE_bottom_sig GMT layers in
# build_position_gmt.py). Unlike COSMIC/PAI3D, FADE ran on the full analysed
# gene set (not a partial external database), so no shared-gene / coverage
# restriction is needed here, the universe is simply the direction-specific
# CAAS position set itself (pos_score_meta, already carries in_top_Xpct flags).
pos_fade_top <- if (has_fade_sites_top) {
  pos_score_meta %>%
    dplyr::filter(change_side %in% c("top", "both")) %>%
    dplyr::left_join(
      fade_sites_top_df %>% dplyr::select(Gene, Position, max_bf, target_aa),
      by = c("Gene", "Position")
    ) %>%
    dplyr::mutate(has_fade = !is.na(max_bf))
} else NULL

pos_fade_bottom <- if (has_fade_sites_bottom) {
  pos_score_meta %>%
    dplyr::filter(change_side %in% c("bottom", "both")) %>%
    dplyr::left_join(
      fade_sites_bottom_df %>% dplyr::select(Gene, Position, max_bf, target_aa),
      by = c("Gene", "Position")
    ) %>%
    dplyr::mutate(has_fade = !is.na(max_bf))
} else NULL
```

`r if ((has_fade_sites_top && sum(pos_fade_top$has_fade) == 0) || (has_fade_sites_bottom && sum(pos_fade_bottom$has_fade) == 0)) "> **Note:** FADE site evidence was provided, but no scored positions in at least one direction overlapped a FADE-significant site."`

```{r fade-table, eval=has_fade_sites_top || has_fade_sites_bottom}
fade_table <- dplyr::bind_rows(
  if (!is.null(pos_fade_top))    pos_fade_top    %>% dplyr::filter(has_fade) %>% dplyr::mutate(Direction = "TOP"),
  if (!is.null(pos_fade_bottom)) pos_fade_bottom %>% dplyr::filter(has_fade) %>% dplyr::mutate(Direction = "BOTTOM")
) %>%
  dplyr::arrange(dplyr::desc(max_bf)) %>%
  dplyr::select(Gene, Position, Direction, CAAS_score, change_side, target_aa, max_bf) %>%
  dplyr::rename(
    `CAAS Score`      = CAAS_score,
    `Change Side`     = change_side,
    `FADE Target AA`  = target_aa,
    `Max Bayes Factor` = max_bf
  )

DT::datatable(
  fade_table,
  caption = "CAAS positions overlapping with FADE-significant sites (BF >= fade_bf_threshold), ordered by Bayes Factor",
  options = list(pageLength = 10, scrollX = TRUE),
  rownames = FALSE
)
```

```{r fade-bf-violin, eval=has_fade_sites_top || has_fade_sites_bottom, fig.height=5}
fade_bf_df <- dplyr::bind_rows(
  if (!is.null(pos_fade_top))    pos_fade_top    %>% dplyr::filter(has_fade) %>% dplyr::mutate(fade_direction = "TOP"),
  if (!is.null(pos_fade_bottom)) pos_fade_bottom %>% dplyr::filter(has_fade) %>% dplyr::mutate(fade_direction = "BOTTOM")
)

ggplot(fade_bf_df, aes(x = fade_direction, y = max_bf, fill = fade_direction)) +
  geom_violin(alpha = 0.55, trim = TRUE, scale = "width", show.legend = FALSE) +
  geom_boxplot(width = 0.14, outlier.size = 0.4, alpha = 0.9, fill = "white", colour = "grey30") +
  scale_fill_manual(values = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[3])) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.08))) +
  labs(
    title    = "Bayes Factor of FADE-significant Positions by Direction",
    subtitle = sprintf("TOP: %d sites | BOTTOM: %d sites",
                       sum(fade_bf_df$fade_direction == "TOP"), sum(fade_bf_df$fade_direction == "BOTTOM")),
    x = NULL, y = "Max Bayes Factor"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r fade-fisher, eval=has_fade_sites_top || has_fade_sites_bottom}
# Odds ratio & P-value interpretation:
# - Odds Ratio: Measures the enrichment of FADE-significant sites (BF >= fade_bf_threshold)
#   within the specified CAAS threshold compared to the rest of the direction's positions.
#   An odds ratio > 1 indicates enrichment.
# - P-value: Reaches significance (via Fisher's exact test) when the proportion of
#   FADE-significant sites is significantly higher inside the CAAS threshold.
run_fade_fisher_dir <- function(df, label) {
  n_all <- nrow(df)
  pct_rows <- list(
    list(col = NULL, lbl = "100% (all positions)",    flag = rep(TRUE, n_all)),
    list(col = NULL, lbl = "25% (top 25% CAAS)",      flag = test_glob_25(df)),
    list(col = "in_top_10pct", lbl = sprintf("Top %g%%", top_pct  * 100), flag = NULL),
    list(col = "in_top_5pct",  lbl = sprintf("Top %g%%", top5_pct * 100), flag = NULL),
    list(col = "in_top_1pct",  lbl = sprintf("Top %g%%", top1_pct * 100), flag = NULL)
  )

  purrr::map_dfr(pct_rows, function(thr) {
    in_top   <- if (!is.null(thr$flag)) thr$flag else df[[thr$col]]
    in_top   <- replace_na(in_top, FALSE)
    n_top    <- sum(in_top)
    n_top_f  <- sum(in_top & df$has_fade, na.rm = TRUE)
    n_rest   <- sum(!in_top)
    n_rest_f <- sum(!in_top & df$has_fade, na.rm = TRUE)
    # Column order is [flag, not-flag] (not [not-flag, flag] as in the sibling
    # PAI3D/COSMIC Fisher helpers above) so fisher.test's odds ratio reads as
    # "odds of FADE-sig in-threshold vs out-of-threshold" the right way round --
    # verified against an unambiguous 90%-vs-10% synthetic case (OR=77.3, not 0.013).
    mat <- matrix(c(n_top_f, n_rest_f, n_top - n_top_f, n_rest - n_rest_f), nrow = 2)
    ft  <- tryCatch(fisher.test(mat),
                    error = function(e) list(estimate = NA, conf.int = c(NA, NA), p.value = NA))

    pct_sig <- if ("gate_sig" %in% names(df)) sprintf("%.1f%%", 100 * mean(df$gate_sig[in_top], na.rm = TRUE)) else "0.0%"

    tibble(
      Direction     = label,
      Threshold     = thr$lbl,
      `N positions` = n_top,
      `N FADE-sig`  = n_top_f,
      `% FADE-sig`  = sprintf("%.1f%%", 100 * n_top_f / max(n_top, 1)),
      `% gate_sig`  = pct_sig,
      `Odds ratio`  = sprintf("%.2f", ft$estimate),
      `P value`     = formatC(ft$p.value, format = "e", digits = 2)
    )
  })
}

fade_fisher <- dplyr::bind_rows(
  if (!is.null(pos_fade_top))    run_fade_fisher_dir(pos_fade_top,    "TOP"),
  if (!is.null(pos_fade_bottom)) run_fade_fisher_dir(pos_fade_bottom, "BOTTOM")
)

knitr::kable(
  fade_fisher,
  caption = "Fisher exact test for FADE-significant site overlap enrichment within CAAS score thresholds. Direction-specific: TOP positions tested only against fade_sites_top, BOTTOM only against fade_sites_bottom.",
  align = c("l", "l", rep("r", ncol(fade_fisher) - 2))
)
```

```{r reg-accum-subsec, results='asis'}
if (has_genomic) cat("## Regional Accumulation\n\n")
```

```{r position-region-accum, eval=has_genomic, fig.height=10, fig.width=14}
# Build directional window summaries - TOP and BOTTOM use direction-specific
# membership tests (test10/test5/test1: closures from test_top_*/test_bot_*
# above, published-slice-preferring with a quantile() fallback, pos_sub here
# is pos_genomic_top/pos_genomic_bot, a DIFFERENT row subset than pos_top/
# pos_bottom the closures were built against, so they must be re-evaluated
# against pos_sub's own rows, not reused positionally).
.build_region_wins <- function(pos_sub, test10, test5, test1, chr_offsets_df, dir_label) {
  pos_sub %>%
    mutate(
      win_id    = floor(gene_mid / params$window_size_bp),
      win_start = win_id * params$window_size_bp,
      win_end   = win_start + params$window_size_bp - 1L,
      win_mid   = win_start + params$window_size_bp / 2L,
      in_top10  = test10(pos_sub),
      in_top5   = test5(pos_sub),
      in_top1   = test1(pos_sub)
    ) %>%
    group_by(chr_order, chr_label, win_id, win_start, win_end, win_mid) %>%
    summarise(
      n_positions        = n(),
      n_top10_positions  = sum(in_top10, na.rm = TRUE),
      n_top5_positions   = sum(in_top5,  na.rm = TRUE),
      n_top1_positions   = sum(in_top1,  na.rm = TRUE),
      frac_top_positions = n_top10_positions / n_positions,
      mean_caas          = mean(CAAS_score, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    left_join(chr_offsets_df %>% select(chr_order, offset), by = "chr_order") %>%
    mutate(
      direction    = dir_label,
      win_x        = win_mid + offset,
      region_label = sprintf("chr%s:%0.1f-%0.1f Mb",
                             chr_label, win_start / 1e6, win_end / 1e6)
    )
}

# pos_genomic already has chr_order and gene_mid; split by direction
pos_genomic_top <- pos_genomic %>% filter(change_side %in% c("top",    "both"))
pos_genomic_bot <- pos_genomic %>% filter(change_side %in% c("bottom", "both"))

pos_region_top <- .build_region_wins(pos_genomic_top, test_top_10, test_top_5, test_top_1,
                                      pos_chr_offsets, "TOP")
pos_region_bot <- .build_region_wins(pos_genomic_bot, test_bot_10, test_bot_5, test_bot_1,
                                      pos_chr_offsets, "BOTTOM")
```

```{r position-region-accum-table, eval=has_genomic}
.make_region_tbl <- function(wins_df) {
  wins_df %>%
    filter(n_top10_positions > 0) %>%
    arrange(desc(n_positions), desc(frac_top_positions)) %>%
    slice_head(n = 30) %>%
    mutate(
      frac_top10 = sprintf("%.1f%%", 100 * n_top10_positions / n_positions),
      frac_top5  = sprintf("%.1f%%", 100 * n_top5_positions  / n_positions),
      frac_top1  = sprintf("%.1f%%", 100 * n_top1_positions  / n_positions)
    ) %>%
    select(
      Region              = region_label,
      `N positions`       = n_positions,
      `N top-10%`         = n_top10_positions,
      `% top-10%`         = frac_top10,
      `N top-5%`          = n_top5_positions,
      `% top-5%`          = frac_top5,
      `N top-1%`          = n_top1_positions,
      `% top-1%`          = frac_top1,
      `Mean CAAS`         = mean_caas
    ) %>%
    mutate(`Mean CAAS` = round(`Mean CAAS`, 4))
}

knitr::kable(.make_region_tbl(pos_region_top),
  caption = "TOP-direction: genomic windows with high-CAAS positions, ordered by N positions (top 30; direction-specific thresholds)")
knitr::kable(.make_region_tbl(pos_region_bot),
  caption = "BOTTOM-direction: genomic windows with high-CAAS positions, ordered by N positions (top 30; direction-specific thresholds)")

# Export directional regional accumulation tables as TSV
region_export <- bind_rows(
  .make_region_tbl(pos_region_top) %>% mutate(Direction = "TOP"),
  .make_region_tbl(pos_region_bot) %>% mutate(Direction = "BOTTOM")
) %>% select(Direction, everything())
readr::write_tsv(region_export,
  file.path(params$output_dir, "regional_accumulation.tsv"))
```

---

# Methods

- **Test**: Position-Level Path Sum Permulation test. Raw CAAS score magnitude
  is summed for every position in a set ($T_\text{obs} = \sum_{p \in \text{term}} s_p$).
  The non-zero position scores ($K \approx 3{,}000$) are randomly permuted $N_\text{perms} = 2,000$ times
  across the full ~1.47M position background to build an exact empirical null distribution.
  Normalized Enrichment Score ($\text{NES} = (T_\text{obs} - \mu_\text{null}) / \sigma_\text{null}$)
  and empirical p-values ($p_\text{perm} = \frac{\#\{T_\text{null} \ge T_\text{obs}\} + 1}{N_\text{perms} + 1}$)
  are computed per term. Run separately for the **global / top / bottom** directions (`change_side`).
- **Two background conventions.** Every database above tests against the
  **full honest background** *except* `cosmic_orthogroups` and
  `pai3d_orthogroups`, which test against a **coverage-restricted background**
  (background intersected with the set of genes the external database itself could ever annotate - see
  `cosmic_coverage_genes.txt` / `pai3d_coverage_genes.txt`, emitted by
  `build_position_gmt.py`).
- **Databases**: 10 GMT-style position-set databases (Pfam domains/clans,
  COSMIC and PAI3D and general orthogroups, genomic locations,
  positive/purifying-selection orthogroups, UCR core/flank orthogroups) plus
  the broad `characterization` layers (UCR core/flank, FUBAR
  positive/purifying). `min`/`max` set-size filtering applies only to the GMT-style databases.
- **`p.adj`**: BH-adjusted per (ranking, database).
- **Significance**: `p.adj < `r PADJ`` and `NES > 0`.
- **No permulation (`p.perm`) at position level.** The CAAS permulation null is a
  *naive random relabeling*: when the phenotype is permuted, the fg/bg species are
  paired at **random across the tree**, not as the observed **matched sister
  pairs** - so pair MRCAs are deep and the ASR path-score's isolation/independence
  geometry differs from the observed. That null therefore tests "phenotype + a
  broken matched-pair design" jointly rather than isolating the phenotype, and it
  is anti-conservative on the MRCA axes. The **gene-level FCS** still uses
  `p.perm`; see its report for the same pairing caveat.

*Report generated `r Sys.time()`.*

---

## Summary

This report presents position-level enrichment results for
**`r params$traitname`** via Position-Level Path Sum Permulation, testing
`r if (has_res) dplyr::n_distinct(res$database) else 0` position-set
database(s) across the global/top/bottom directions.
`r if (has_res) sum(res$sig, na.rm = TRUE) else 0` (ranking, database,
pathway, cutoff) combinations passed at `p.adj < ``r PADJ`` with NES > 0.

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere - Enrichment Module*
*Date of generation: `r Sys.Date()`*
