---
title: "AMI Analysis — DOMINO Active Modules"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
    mathjax: null                       # no math in this report
    md_extensions: "-tex_math_dollars"  # gene-set names contain '$' (e.g. V$HNF4_01) — never parse as TeX math
params:
  background_file: NULL
  background_basename: NULL
  gene_list_files: NULL
  output_dir: "string"
  project_name: "AMI_Analysis"
  species: 9606
  required_score: 400
  gene_scores_file: NULL
  string_db_dir: NULL
  min_cluster_genes: 3
  domino_network_sif: NULL
  domino_modules_dir: NULL
  domino_network_score_thr: 700
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7,
  dev = 'png',
  dpi = 320
)

# Prevent hangs by setting connection timeouts for RCurl and httr
options(timeout = 30)
options(RCurlOptions = list(connecttimeout = 10, timeout = 30))
if (requireNamespace("httr", quietly = TRUE)) {
  httr::set_config(httr::config(connecttimeout = 10, timeout = 30))
}

# Minimum cluster size (genes) to keep in the cluster description table and
# the network visualisation; smaller clusters are labelled "Too small" and
# dropped from both.
.min_cluster_genes <- if (!is.null(params$min_cluster_genes)) as.integer(params$min_cluster_genes) else 3L

suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
  library(kableExtra)
  library(STRINGdb)
  library(paletteer)
})

# ── Global ENSP -> Symbol Mapping Cache ──────────────────────────────────────
global_id2gene <- new.env(parent = emptyenv())

# ── Local offline enrichment implementation ───────────────────────────────────
enrich_terms_db <- NULL

find_enrichment_terms_file <- function(species, cache_dir) {
  fname <- paste0(species, ".protein.enrichment.terms.v12.0.txt.gz")
  
  # Search paths in order of preference, starting with parameterized string_db_dir
  paths <- c(
    if (!is.null(params$string_db_dir) && nzchar(params$string_db_dir)) file.path(params$string_db_dir, fname) else NULL,
    file.path(cache_dir, fname)
  )
  
  for (p in paths) {
    if (!is.null(p) && nzchar(p) && file.exists(p)) {
      return(normalizePath(p))
    }
  }
  NULL
}

load_local_enrichment_db <- function(species, cache_dir) {
  terms_file <- find_enrichment_terms_file(species, cache_dir)
  
  if (is.null(terms_file)) {
    # If not found locally, try downloading it (fallback if internet is available)
    message(sprintf("[local_enrichment] Local file not found. Attempting to download for species %s...", species))
    dest_file <- file.path(cache_dir, paste0(species, ".protein.enrichment.terms.v12.0.txt.gz"))
    url <- sprintf("https://stringdb-downloads.org/download/protein.enrichment.terms.v12.0/%s.protein.enrichment.terms.v12.0.txt.gz", species)
    
    download_ok <- tryCatch({
      download.file(url, dest_file, quiet = TRUE)
      TRUE
    }, error = function(e) {
      message(sprintf("[local_enrichment] Download failed: %s", e$message))
      FALSE
    })
    
    if (download_ok && file.exists(dest_file)) {
      terms_file <- dest_file
    } else {
      stop("Local enrichment terms file not found and download failed.")
    }
  }
  
  message(sprintf("[local_enrichment] Loading enrichment database from: %s", terms_file))
  db <- readr::read_tsv(
    terms_file,
    col_types = "cccc",
    comment = "#",
    col_names = c("string_protein_id", "category", "term", "description")
  )
  db
}

get_enrichment_local <- function(fg_ids, bg_ids = NULL, fdr_threshold = 0.1) {
  if (is.null(enrich_terms_db)) return(NULL)
  if (length(fg_ids) == 0) return(tibble::tibble())
  
  # If bg_ids is too small/empty, default to all proteins present in the terms db
  if (is.null(bg_ids) || length(bg_ids) < 5) {
    bg_ids <- unique(enrich_terms_db$string_protein_id)
  }
  
  # Ensure fg_ids is constrained by background
  fg_ids <- intersect(fg_ids, bg_ids)
  if (length(fg_ids) == 0) return(tibble::tibble())
  
  # Filter database to background proteins
  bg_data <- enrich_terms_db %>% dplyr::filter(string_protein_id %in% bg_ids)
  
  # Count term sizes in background (STRING requires >= 2 genes in background for test)
  term_bg_counts <- bg_data %>%
    dplyr::group_by(category, term, description) %>%
    dplyr::summarise(M = dplyr::n(), .groups = "drop") %>%
    dplyr::filter(M >= 2)
    
  # Filter database to foreground and count
  fg_data <- enrich_terms_db %>% dplyr::filter(string_protein_id %in% fg_ids)
  fg_counts <- fg_data %>%
    dplyr::group_by(category, term, description) %>%
    dplyr::summarise(k = dplyr::n(), .groups = "drop")
    
  N <- length(bg_ids)
  n <- length(fg_ids)
  
  # Run hypergeometric test: phyper(k-1, M, N-M, n, lower.tail = FALSE) -> P(X >= k)
  res <- term_bg_counts %>%
    dplyr::inner_join(fg_counts, by = c("category", "term", "description")) %>%
    dplyr::mutate(
      pvalue = phyper(k - 1, M, N - M, n, lower.tail = FALSE)
    ) %>%
    dplyr::group_by(category) %>%
    dplyr::mutate(fdr = p.adjust(pvalue, method = "fdr")) %>%
    dplyr::ungroup() %>%
    dplyr::filter(fdr <= fdr_threshold) %>%
    dplyr::arrange(category, fdr)
    
  if (nrow(res) == 0) return(tibble::tibble())
  
  # Map protein IDs in foreground to symbols using our global cache
  fg_mapped <- fg_data %>%
    dplyr::mutate(gene = sapply(string_protein_id, function(id) {
      sym <- global_id2gene[[id]]
      if (is.null(sym)) id else sym
    }))
    
  # Append preferredNames (comma-separated list of symbols)
  term_genes <- fg_mapped %>%
    dplyr::group_by(category, term) %>%
    dplyr::summarise(
      preferredNames = paste(sort(unique(gene)), collapse = ","),
      .groups = "drop"
    )
    
  res <- res %>%
    dplyr::left_join(term_genes, by = c("category", "term")) %>%
    dplyr::rename(
      number_of_genes = k,
      number_of_genes_in_background = M,
      p_value = pvalue
    )
    
  res
}

# ── Palette helpers (overflow-safe Tableau_10) ────────────────────────────────
.pal10_base <- as.character(paletteer_d("ggthemes::Tableau_10"))
.safe_pal   <- function(n) if (n <= 10L) .pal10_base[seq_len(n)] else colorRampPalette(.pal10_base)(n)
safe_fill   <- function(n) scale_fill_manual(values  = .safe_pal(n))
safe_color  <- function(n) scale_color_manual(values = .safe_pal(n))

has_pkg <- function(pkg) requireNamespace(pkg, quietly = TRUE)

`%||%` <- function(a, b) {
  if (!is.null(a) && length(a) > 0 && !all(is.na(a))) a else b
}

createDir <- function(directory) {
  if (!dir.exists(directory)) dir.create(directory, recursive = TRUE, showWarnings = FALSE)
}

download_with_retry <- function(url, destfile, tries = 4L, timeout_sec = 1800L) {
  old_timeout <- getOption("timeout")
  on.exit(options(timeout = old_timeout), add = TRUE)
  options(timeout = max(as.integer(old_timeout), as.integer(timeout_sec), 600L))

  for (i in seq_len(tries)) {
    ok <- tryCatch({
      utils::download.file(url, destfile = destfile, mode = "wb", quiet = FALSE, method = "libcurl")
      TRUE
    }, error = function(e) {
      message(sprintf("Download failed (%d/%d): %s", i, tries, e$message))
      FALSE
    })

    if (isTRUE(ok) && file.exists(destfile) && file.size(destfile) > 0) return(invisible(TRUE))
    if (i < tries) Sys.sleep(min(30, i * 5))
  }

  stop(sprintf("Failed to download after %d attempts: %s", tries, url))
}

ensure_string_cache <- function(species, version = "12.0", input_directory = "string_cache") {
  createDir(input_directory)

  species <- as.integer(species)
  base <- "https://stringdb-downloads.org/download"
  files <- c(
    sprintf("protein.aliases.v%s/%d.protein.aliases.v%s.txt.gz", version, species, version),
    sprintf("protein.info.v%s/%d.protein.info.v%s.txt.gz", version, species, version),
    sprintf("protein.links.v%s/%d.protein.links.v%s.txt.gz", version, species, version),
    sprintf("protein.physical.links.v%s/%d.protein.physical.links.v%s.txt.gz", version, species, version)
  )

  for (rel in files) {
    url <- sprintf("%s/%s", base, rel)
    fname <- basename(rel)
    dest <- file.path(input_directory, fname)
    if (!file.exists(dest) || file.size(dest) == 0) {
      # Try finding in params$string_db_dir first
      local_src <- if (!is.null(params$string_db_dir) && nzchar(params$string_db_dir)) file.path(params$string_db_dir, fname) else NULL
      if (!is.null(local_src) && file.exists(local_src) && file.size(local_src) > 0) {
        message(sprintf("[ensure_string_cache] Using local pre-cached file: %s", local_src))
        file.copy(local_src, dest, overwrite = TRUE)
      } else {
        message(sprintf("Caching STRING file from online: %s", fname))
        tryCatch({
          download_with_retry(url, destfile = dest)
        }, error = function(e) {
          stop(sprintf("Failed to retrieve required STRING cache file %s: %s", fname, e$message))
        })
      }
    }
  }

  invisible(normalizePath(input_directory, winslash = "/", mustWork = FALSE))
}

read_gene_vector <- function(path) {
  x <- readr::read_lines(path)
  x <- trimws(x)
  x <- x[x != ""]
  unique(x)
}

string_root <- "."
string_results_dir <- "string_results"
string_summary_dir <- "string_summary"
string_plots_dir   <- "string_plots"
string_cache_dir   <- "string_cache"

createDir(string_results_dir)
createDir(string_summary_dir)
createDir(string_plots_dir)
createDir(string_cache_dir)

# Load local enrichment database
enrich_terms_db <- tryCatch({
  load_local_enrichment_db(as.integer(params$species), string_cache_dir)
}, error = function(e) {
  warning(sprintf("Could not load local enrichment database: %s. Will fallback to online queries.", e$message))
  NULL
})

# Pre-cache STRING DB files with retry/timeout guards to avoid transient
# network failures during set_background()/enrichment calls.
ensure_string_cache(
  species = as.integer(params$species),
  version = "12.0",
  input_directory = string_cache_dir
)

# ── STRINGdb instance ─────────────────────────────────────────────────────────
# STRING is used ONLY for ID mapping ($map()), background caching, and the
# functional/term enrichment path (get_enrichment_local()/get_enrichment())
# below. Module finding + per-module significance is DOMINO's job (see the
# "DOMINO Active Modules" section) — STRINGdb's walktrap clustering and
# get_ppi_enrichment() (both per-cluster and whole-list) have been removed
# entirely: get_ppi_enrichment() cannot honour a custom background above 2000
# proteins combined (confirmed empirically against the live API), while our
# cleaned_background is ~10k genes.
string_db <- STRINGdb$new(
  version         = "12.0",
  species         = as.integer(params$species),
  score_threshold = as.integer(params$required_score),
  input_directory = string_cache_dir
)

string_networks_dir <- "string_networks"
createDir(string_networks_dir)
# ── DOMINO module loading ─────────────────────────────────────────────────────
# DOMINO (Shamir-Lab) replaces STRING's walktrap clustering here: it recovers
# known complexes more fully than walktrap at our typical list sizes (10-40
# genes), validated in a prior investigation session against AMINE/SIMBA on
# real bottom-direction CAAS gene lists. Modules + their DOMINO-native
# Bonferroni-corrected hypergeometric p-value were precomputed upstream
# (subworkflows/ENRICHMENT/domino.nf's DOMINO_RUN_MODULES, via
# src/run_domino_modules.py calling DOMINO's Python API directly so the
# p-value survives — the `domino` CLI itself discards it). This just reads
# those two TSVs back in; column names (node/cluster) are kept identical to
# the old walktrap clusters_df shape so cluster_layout()/plot_cluster_network()
# below are unchanged.
load_domino_modules <- function(list_name, modules_dir) {
  mod_file   <- file.path(modules_dir, sprintf("%s_domino_modules.tsv", list_name))
  stats_file <- file.path(modules_dir, sprintf("%s_domino_module_stats.tsv", list_name))
  if (!file.exists(mod_file) || !file.exists(stats_file)) {
    return(list(clusters_df = NULL, module_stats = NULL))
  }
  list(
    clusters_df  = readr::read_tsv(mod_file, col_types = "ci"),
    module_stats = readr::read_tsv(stats_file, col_types = "iicdd")
  )
}

# ── Cluster-aware FR layout ───────────────────────────────────────────────────
# Seeds node positions by cluster centroid before Fruchterman-Reingold so that
# same-cluster nodes start near each other and the final layout mirrors the
# STRING web interface (clusters appear as visual blobs, not scattered).
cluster_layout <- function(net, clusters_df, seed = 42) {
  set.seed(seed)
  n <- igraph::vcount(net)
  if (is.null(clusters_df) || n < 2) return(igraph::layout_with_fr(net))

  vnames  <- igraph::V(net)$name
  cl_map  <- setNames(clusters_df$cluster, clusters_df$node)
  clusters <- cl_map[vnames]
  clusters[is.na(clusters)] <- max(clusters_df$cluster, na.rm = TRUE) + 1L

  unique_cl <- sort(unique(clusters))
  n_cl <- length(unique_cl)
  if (n_cl <= 1) return(igraph::layout_with_fr(net))

  theta  <- seq(0, 2 * pi * (1 - 1 / n_cl), length.out = n_cl)
  spread <- max(5, sqrt(n) * 0.9)
  cx     <- spread * cos(theta)
  cy     <- spread * sin(theta)

  init <- matrix(0, nrow = n, ncol = 2)
  for (i in seq_along(unique_cl)) {
    mask  <- clusters == unique_cl[i]
    n_sub <- sum(mask)
    sd_jit <- max(0.5, sqrt(n_sub) * 0.35)
    init[mask, 1] <- cx[i] + rnorm(n_sub, 0, sd_jit)
    init[mask, 2] <- cy[i] + rnorm(n_sub, 0, sd_jit)
  }
  igraph::layout_with_fr(net, coords = init, niter = 300)
}

# ── Cluster functional descriptions ──────────────────────────────────────────
# Runs STRING enrichment (untouched functional-enrichment path) on each
# DOMINO module's genes to derive a descriptive label from the top significant
# GO/pathway term. Significance itself is no longer computed here — it comes
# straight from DOMINO's own module_stats (p_value/p_adj, joined in below),
# replacing the old per-cluster get_ppi_enrichment() density test.
describe_clusters <- function(sdb, clusters_df, id2gene, module_stats = NULL, top_terms = 3,
                               min_cluster_genes = 3) {
  if (is.null(clusters_df) || nrow(clusters_df) == 0) return(tibble())
  desc <- purrr::map_dfr(sort(unique(clusters_df$cluster)), function(cl_id) {
    cl_nodes <- clusters_df$node[clusters_df$cluster == cl_id]
    genes    <- id2gene[cl_nodes]
    gene_str <- paste(head(sort(genes[!is.na(genes)]), 8), collapse = ", ")

    if (length(cl_nodes) < min_cluster_genes) {
      return(tibble(cluster = cl_id, n_genes = length(cl_nodes),
                    genes = gene_str, description = "Too small"))
    }
    message(sprintf("[describe_clusters] Querying STRING enrichment for module %d (%d genes)...", cl_id, length(cl_nodes)))
    enrich <- if (!is.null(enrich_terms_db)) {
      get_enrichment_local(cl_nodes, bg_string_ids)
    } else {
      # Fallback to online API
      message(sprintf("[describe_clusters] Querying STRING online enrichment for module %d (%d genes)...", cl_id, length(cl_nodes)))
      tryCatch(
        sdb$get_enrichment(cl_nodes),
        error = function(e) {
          message(sprintf("[describe_clusters] Warning: STRING enrichment failed for module %d: %s", cl_id, e$message))
          NULL
        }
      )
    }
    if (is.null(enrich) || nrow(enrich) == 0 || "error" %in% names(enrich)) {
      return(tibble(cluster = cl_id, n_genes = length(cl_nodes),
                    genes = gene_str, description = "No enrichment"))
    }
    pvcol   <- intersect(c("fdr", "p_value", "pvalue", "p.adjust"), names(enrich))[1]
    if (!is.na(pvcol)) enrich <- enrich[order(enrich[[pvcol]]), ]
    dcol    <- intersect(c("description", "term", "name"), names(enrich))[1]
    top     <- head(as.character(enrich[[dcol]]), top_terms)
    tibble(cluster = cl_id, n_genes = length(cl_nodes),
           genes = gene_str, description = paste(top, collapse = " / "))
  })

  if (!is.null(module_stats) && nrow(module_stats) > 0) {
    desc <- desc %>%
      dplyr::left_join(
        module_stats %>% dplyr::select(cluster, domino_p_value = p_value, domino_p_adj = p_adj),
        by = "cluster"
      )
  }
  desc
}

# ── Network plot (cluster-aware layout, no isolated nodes) ───────────────────
# Edge width/opacity encode STRING's combined_score (already an edge attribute
# on every graph loaded via STRINGdb$load() -- graph.data.frame() preserves it
# automatically, it was just never read here), mirroring STRING's own network
# images where thicker/more solid lines mean higher-confidence interactions.
plot_cluster_network <- function(net, clusters_df, title, id_to_gene = NULL,
                                  cluster_labels = NULL) {
  n_clust <- if (!is.null(clusters_df)) length(unique(clusters_df$cluster)) else 1L
  pal     <- .safe_pal(max(n_clust, 2L))

  # Scale edge width/opacity by combined_score when present and varying;
  # fall back to the previous fixed width/color otherwise (e.g. a subgraph
  # with only one edge, or one built without that attribute).
  .escore <- igraph::E(net)$combined_score
  if (!is.null(.escore) && length(unique(stats::na.omit(.escore))) > 1) {
    edge_widths <- scales::rescale(.escore, to = c(0.4, 3))
    edge_alphas <- scales::rescale(.escore, to = c(0.25, 0.95))
    edge_colors <- scales::alpha("#888888", edge_alphas)
  } else {
    edge_widths <- 0.7
    edge_colors <- "#aaaaaa"
  }

  if (!is.null(clusters_df) && nrow(clusters_df) > 0) {
    cl_map  <- setNames(clusters_df$cluster, clusters_df$node)
    v_cl    <- cl_map[igraph::V(net)$name]
    v_cl[is.na(v_cl)] <- n_clust + 1L
    cl_ids  <- sort(unique(clusters_df$cluster))
    vertex_colors <- pal[match(v_cl, c(cl_ids, n_clust + 1L))]
  } else {
    cl_ids        <- integer(0)
    vertex_colors <- rep(pal[1], igraph::vcount(net))
  }

  labels <- if (!is.null(id_to_gene)) id_to_gene[igraph::V(net)$name] else igraph::V(net)$name
  labels[is.na(labels)] <- igraph::V(net)$name[is.na(labels)]

  lo <- cluster_layout(net, clusters_df)

  old_par <- par(mar = c(0, 0, 2, 5))
  on.exit(par(old_par), add = TRUE)
  igraph::plot.igraph(
    net,
    main               = title,
    layout             = lo,
    vertex.color       = vertex_colors,
    vertex.size        = pmin(18, pmax(4, 55 / sqrt(igraph::vcount(net)))),
    vertex.label       = labels,
    vertex.label.cex   = max(0.55, min(1.1, 10 / sqrt(igraph::vcount(net)))),
    vertex.label.color = "black",
    vertex.label.font  = 2,
    edge.arrow.size    = 0,
    edge.width         = edge_widths,
    edge.color         = edge_colors
  )
  if (length(cl_ids) > 1) {
    leg_labels <- if (!is.null(cluster_labels)) {
      vapply(as.character(cl_ids), function(k) {
        lbl <- cluster_labels[k]
        if (is.na(lbl)) paste0("Cluster ", k) else paste0("C", k, ": ", stringr::str_trunc(lbl, 35))
      }, character(1))
    } else {
      paste0("Cluster ", cl_ids)
    }
    legend("topright", legend = leg_labels,
           fill = pal[seq_along(cl_ids)], cex = 0.55, bty = "n", title = "Cluster")
  }
}

# Map a character vector of gene SYMBOLs to STRING identifiers.
# Returns a character vector of STRING_id values.
safe_string_map_ids <- function(genes) {
  if (length(genes) == 0) return(character(0))
  df_in <- data.frame(gene = genes, stringsAsFactors = FALSE)
  mapped <- tryCatch(
    string_db$map(df_in, "gene", removeUnmappedRows = TRUE, quiet = TRUE),
    error = function(e) {
      message(sprintf("STRING ID mapping failed: %s", e$message))
      NULL
    }
  )
  if (is.null(mapped) || nrow(mapped) == 0) return(character(0))
  
  # Populate the global ENSP -> symbol cache
  for (i in 1:nrow(mapped)) {
    global_id2gene[[mapped$STRING_id[i]]] <- mapped$gene[i]
  }
  
  unique(as.character(mapped$STRING_id))
}

# Resolve background file path and normalize to absolute path
bg_file <- params$background_file
if (!file.exists(bg_file) && !is.null(params$background_basename)) {
  if (file.exists(params$background_basename)) bg_file <- params$background_basename
}

if (is.null(bg_file) || !file.exists(bg_file)) {
  stop("Error: background_file not found")
}
bg_file <- normalizePath(bg_file, mustWork = TRUE)
bg_basename <- basename(bg_file)

# Auto-detect gene list files from the current working directory, excluding the
# background file.  Using normalizePath() ensures that subsequent reads succeed
# even if the working directory changes during report rendering.
gene_list_files <- params$gene_list_files
if (is.null(gene_list_files) || length(gene_list_files) == 0) {
  gene_list_files <- list.files(pattern = "\\.txt$", full.names = TRUE)
  gene_list_files <- gene_list_files[basename(gene_list_files) != bg_basename]
}
gene_list_files <- gene_list_files[file.exists(gene_list_files)]
# Always normalize to absolute paths
gene_list_files <- normalizePath(gene_list_files, mustWork = FALSE)
gene_list_files <- gene_list_files[file.exists(gene_list_files)]
if (length(gene_list_files) == 0) stop("Error: no valid gene list files found")


```

# AMI analysis configuration

```{r config_overview}
cat("=== AMI (Active Module Identification) configuration ===\n")
cat("Background file (cleaned_background):", bg_file, "\n")
cat("Gene list files:", length(gene_list_files), "\n")
cat("Species (NCBI taxon ID):", params$species, "\n")
cat("STRING interaction score (network edges):", params$required_score, "\n")
cat("DOMINO network score threshold:", params$domino_network_score_thr, "\n")
cat("Module finding + significance: DOMINO (Shamir-Lab)\n")
cat("Module functional labels: STRING term enrichment (untouched functional-enrichment path)\n")
```

# Map background to STRING IDs

> **Note:** A custom background (`cleaned_background`) constrains enrichment to the experimental proteome (all tested genes).
> When background genes are provided, they are mapped to STRING identifiers and set
> on the `STRINGdb` object via `set_background()`, which applies to all subsequent
> `get_enrichment()` calls.

```{r parse_background}
bg_symbols  <- read_gene_vector(bg_file)
bg_string_ids <- safe_string_map_ids(bg_symbols)

cat("Background symbols supplied:", length(bg_symbols), "\n")
cat("Background STRING IDs successfully mapped:", length(bg_string_ids), "\n")

if (length(bg_string_ids) < 5) {
  warning(
    sprintf(
      "Only %d background genes mapped to STRING IDs. ",
      length(bg_string_ids)
    ),
    "Proceeding without a custom background (STRING will use the full proteome)."
  )
  use_custom_background <- FALSE
} else {
  # Set the background on the STRINGdb object so all subsequent enrichment
  # calls are constrained to the experimental proteome.
  use_custom_background <- tryCatch({
    string_db$set_background(bg_string_ids)
    TRUE
  }, error = function(e) {
    warning(
      sprintf("Could not set custom STRING background: %s", e$message),
      " Proceeding with STRING default background."
    )
    FALSE
  })

  if (isTRUE(use_custom_background)) {
    cat("Custom background set on STRINGdb object (", length(bg_string_ids), " proteins).\n")
  }
}
```

# Load gene scores for overlap checking

```{r load_gene_scores}
# Minimum STRING-mapped genes for a gene list to get a network at all (used by
# the DOMINO Active Modules section below).
MIN_MAPPED <- 5L

gene_scores_df <- NULL
if (!is.null(params$gene_scores_file) && file.exists(params$gene_scores_file)) {
  gene_scores_df <- readr::read_tsv(params$gene_scores_file, show_col_types = FALSE)
  cat("Loaded gene_scores_file:", params$gene_scores_file, "with", nrow(gene_scores_df), "genes.\n")
} else {
  cat("No gene_scores_file provided or file not found. Overlap checks will be skipped.\n")
}

prepare_gene_attributes <- function(df) {
  if (is.null(df) || nrow(df) == 0) return(NULL)
  
  # Helper to compute percentile flags
  pct_of <- function(v) {
    ok <- !is.na(v) & v > 0
    p <- rep(NA_real_, length(v))
    if (any(ok)) p[ok] <- 1 - (rank(v[ok], ties.method = "max") - 1) / sum(ok)
    p
  }
  
  .col <- function(c, default = NA) if (c %in% names(df)) df[[c]] else rep(default, nrow(df))
  .istrue <- function(x) !is.na(x) & x %in% c(TRUE, "TRUE", "True", "true", 1, "1")
  
  out <- tibble::tibble(gene = df$Gene)
  
  # CAAS score columns for percentiles
  score_top <- suppressWarnings(as.numeric(.col("gene_caas_score_top_all")))
  if (all(is.na(score_top))) score_top <- suppressWarnings(as.numeric(.col("gene_caas_score_top")))
  
  score_bot <- suppressWarnings(as.numeric(.col("gene_caas_score_bottom_all")))
  if (all(is.na(score_bot))) score_bot <- suppressWarnings(as.numeric(.col("gene_caas_score_bottom")))
  
  # Percentiles
  breaks <- c(0.10, 0.05, 0.01)
  if (!all(is.na(score_top))) {
    pt <- pct_of(score_top)
    out$pct_top10 <- !is.na(pt) & pt <= breaks[1]
    out$pct_top5  <- !is.na(pt) & pt <= breaks[2]
    out$pct_top1  <- !is.na(pt) & pt <= breaks[3]
  } else {
    out$pct_top10 <- FALSE; out$pct_top5 <- FALSE; out$pct_top1 <- FALSE
  }
  
  if (!all(is.na(score_bot))) {
    pb <- pct_of(score_bot)
    out$pct_bottom10 <- !is.na(pb) & pb <= breaks[1]
    out$pct_bottom5  <- !is.na(pb) & pb <= breaks[2]
    out$pct_bottom1  <- !is.na(pb) & pb <= breaks[3]
  } else {
    out$pct_bottom10 <- FALSE; out$pct_bottom5 <- FALSE; out$pct_bottom1 <- FALSE
  }
  
  # Significance gates (gene has >=1 position passing position-level gate_sig/gate_fdr)
  out$flag_gate_sig <- .istrue(.col("gene_caas_gate_sig"))
  out$flag_gate_fdr <- .istrue(.col("gene_caas_gate_fdr"))
  
  # FADE
  out$flag_fade_top    <- .istrue(.col("fade_significant_top"))
  out$flag_fade_bottom <- .istrue(.col("fade_significant_bottom"))
  out$flag_fade        <- out$flag_fade_top | out$flag_fade_bottom
  
  # RER
  .rer_dir <- tolower(as.character(.col("rer_acceleration")))
  out$flag_rer_acc     <- .istrue(.col("rer_significant")) & grepl("acc", .rer_dir)
  out$flag_rer_decc    <- .istrue(.col("rer_significant")) & grepl("dec", .rer_dir)
  
  # Accumulation
  out$flag_accum       <- .istrue(.col("accum_significant"))
  
  out
}

compute_term_overlap_percentages <- function(preferred_names, attr_df, dir_label) {
  if (is.na(preferred_names) || preferred_names == "" || is.null(attr_df)) {
    return(tibble::tibble(
      gate_sig = NA_real_, gate_fdr = NA_real_,
      top10 = NA_real_, top5 = NA_real_, top1 = NA_real_,
      bot10 = NA_real_, bot5 = NA_real_, bot1 = NA_real_,
      fade = NA_real_, fade_top = NA_real_, fade_bottom = NA_real_,
      rer_acc = NA_real_, rer_decc = NA_real_, accum = NA_real_
    ))
  }
  genes <- strsplit(preferred_names, "[,;]")[[1]]
  sub_attr <- attr_df %>% dplyr::filter(gene %in% genes)
  if (nrow(sub_attr) == 0) {
    return(tibble::tibble(
      gate_sig = 0, gate_fdr = 0,
      top10 = 0, top5 = 0, top1 = 0,
      bot10 = 0, bot5 = 0, bot1 = 0,
      fade = 0, fade_top = 0, fade_bottom = 0,
      rer_acc = 0, rer_decc = 0, accum = 0
    ))
  }
  
  # Calculate percentages (mean * 100)
  round_pct <- function(x) round(100 * mean(x, na.rm = TRUE), 1)
  
  tibble::tibble(
    gate_sig = round_pct(sub_attr$flag_gate_sig),
    gate_fdr = round_pct(sub_attr$flag_gate_fdr),
    top10    = round_pct(sub_attr$pct_top10),
    top5     = round_pct(sub_attr$pct_top5),
    top1     = round_pct(sub_attr$pct_top1),
    bot10    = round_pct(sub_attr$pct_bottom10),
    bot5     = round_pct(sub_attr$pct_bottom5),
    bot1     = round_pct(sub_attr$pct_bottom1),
    fade     = round_pct(sub_attr$flag_fade),
    fade_top = round_pct(sub_attr$flag_fade_top),
    fade_bottom = round_pct(sub_attr$flag_fade_bottom),
    rer_acc  = round_pct(sub_attr$flag_rer_acc),
    rer_decc = round_pct(sub_attr$flag_rer_decc),
    accum    = round_pct(sub_attr$flag_accum)
  )
}
```

# DOMINO Active Modules {.tabset}

Network score threshold: **`r params$domino_network_score_thr`** (= `r params$domino_network_score_thr / 1000` on the 0–1 STRING scale).
Only nodes with at least one edge at this confidence level (both endpoints in `cleaned_background`) are in DOMINO's input network.
Module finding + significance: **DOMINO** (Shamir-Lab) — Bonferroni-corrected hypergeometric p-value per module, computed by DOMINO itself
(not a STRING PPI-density test, which cannot honour a custom background above 2000 proteins).

```{r domino_networks, fig.width=10, fig.height=8, results='asis'}
suppressPackageStartupMessages(library(igraph))

# Whole network is built once (nodes are already gene symbols -- no STRING ID
# round-trip needed here, unlike the functional-enrichment path above).
.domino_sif <- readr::read_tsv(
  params$domino_network_sif,
  col_names = c("from", "interaction", "to"),
  col_types = "ccc"
)
.net_full <- igraph::graph_from_data_frame(.domino_sif[, c("from", "to")], directed = FALSE)

network_summary_rows <- list()

for (.gf in gene_list_files) {
  .list_name <- tools::file_path_sans_ext(basename(.gf))
  .genes_sym <- read_gene_vector(.gf)
  .mapped    <- intersect(.genes_sym, igraph::V(.net_full)$name)

  if (length(.mapped) < MIN_MAPPED) next

  cat(sprintf("\n\n## %s\n\n", .list_name))

  .net <- igraph::induced_subgraph(.net_full, igraph::V(.net_full)[igraph::V(.net_full)$name %in% .genes_sym])
  if (igraph::vcount(.net) == 0) {
    cat(sprintf("> No network retrieved for %s.\n\n", .list_name))
    next
  }

  # Keep only connected nodes (nodes with >=1 edge at the score threshold)
  .connected <- igraph::V(.net)[igraph::degree(.net) > 0]
  .net_conn  <- igraph::induced_subgraph(.net, .connected)
  .n_nodes   <- igraph::vcount(.net_conn)
  .n_edges   <- igraph::ecount(.net_conn)

  cat(sprintf("Connected nodes: **%d** | Edges: **%d** | Input genes in network: **%d**\n\n",
              .n_nodes, .n_edges, length(.mapped)))

  if (.n_nodes < 3) {
    cat("> Too few connected nodes — network plot skipped.\n\n")
    network_summary_rows[[length(network_summary_rows) + 1]] <- tibble(
      gene_list = .list_name, n_input = length(.mapped), n_connected = .n_nodes,
      n_edges = .n_edges, n_modules = NA_integer_
    )
    next
  }

  .domino <- load_domino_modules(.list_name, params$domino_modules_dir)
  .net_vnames  <- igraph::V(.net_conn)$name
  .clusters_df <- if (!is.null(.domino$clusters_df)) {
    .domino$clusters_df %>% dplyr::filter(node %in% .net_vnames)
  } else NULL
  .module_stats <- .domino$module_stats

  # Reorder modules by size (number of genes) descending, capped to the 20
  # largest -- keeps the network plot legend/description table readable (same
  # rationale as the old walktrap cap: a 500-5000+ gene list can otherwise
  # produce many modules that were never going to be individually
  # interpretable). Renumbering is applied identically to module_stats so the
  # p_value/p_adj join below stays aligned.
  .n_mod_total <- if (!is.null(.clusters_df)) length(unique(.clusters_df$cluster)) else NA_integer_
  if (!is.null(.clusters_df) && nrow(.clusters_df) > 0) {
    .cl_counts <- .clusters_df %>%
      dplyr::group_by(cluster) %>%
      dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
      dplyr::arrange(desc(n), cluster)
    .cl_map <- setNames(seq_len(nrow(.cl_counts)), .cl_counts$cluster)
    .clusters_df$cluster <- .cl_map[as.character(.clusters_df$cluster)]
    if (!is.null(.module_stats)) {
      .module_stats <- .module_stats %>%
        dplyr::filter(cluster %in% as.integer(names(.cl_map))) %>%
        dplyr::mutate(cluster = .cl_map[as.character(cluster)])
    }
    .clusters_df <- .clusters_df %>% dplyr::filter(cluster <= 20)
  }

  .n_mod <- if (!is.null(.clusters_df)) length(unique(.clusters_df$cluster)) else NA_integer_

  if (!is.null(.clusters_df)) {
    # DOMINO's nodes ARE gene symbols already, so node == gene here.
    .clusters_df$gene <- .clusters_df$node

    .clusters_out <- .clusters_df
    attr_df <- NULL
    if (!is.null(gene_scores_df)) {
      attr_df <- prepare_gene_attributes(gene_scores_df)
      if (!is.null(attr_df)) {
        .clusters_out <- .clusters_out %>%
          dplyr::left_join(attr_df, by = c("gene" = "gene"))
      }
    }

    .clust_file <- file.path(string_networks_dir, sprintf("%s_domino_clusters.tsv", .list_name))
    readr::write_tsv(.clusters_out %>% dplyr::arrange(cluster, gene), .clust_file)
    cat(sprintf("Modules found: **%d** of **%d** total (showing the 20 largest; written to `%s`)\n\n",
                .n_mod, .n_mod_total, basename(.clust_file)))

    .clust_desc <- describe_clusters(
      string_db, .clusters_df, setNames(.clusters_df$node, .clusters_df$node),
      module_stats = .module_stats, min_cluster_genes = .min_cluster_genes
    )

    if (!is.null(gene_scores_df) && !is.null(attr_df) && nrow(.clust_desc) > 0) {
      .dir_label <- if (grepl("top", .list_name)) "top" else if (grepl("bottom", .list_name)) "bottom" else "global"

      .cl_pcts <- purrr::map_dfr(sort(unique(.clusters_df$cluster)), function(cl_id) {
        cl_genes <- .clusters_df$gene[.clusters_df$cluster == cl_id]
        compute_term_overlap_percentages(paste(cl_genes, collapse = ","), attr_df, .dir_label) %>%
          mutate(cluster = cl_id)
      })

      .clust_desc <- .clust_desc %>% dplyr::left_join(.cl_pcts, by = "cluster")
    }

    # Persist module-level descriptions (incl. DOMINO's own p-value) as a
    # durable artifact downstream consumers can read.
    .clust_desc_file <- file.path(string_networks_dir, sprintf("%s_domino_module_descriptions.tsv", .list_name))
    readr::write_tsv(.clust_desc %>% dplyr::arrange(desc(n_genes)), .clust_desc_file)

    .clust_desc_display <- .clust_desc %>%
      dplyr::filter(!description %in% c("Too small", "No enrichment")) %>%
      dplyr::arrange(desc(n_genes))

    if (nrow(.clust_desc_display) > 0) {
      if (!is.null(gene_scores_df)) {
        .dir_label <- if (grepl("top", .list_name)) "top" else if (grepl("bottom", .list_name)) "bottom" else "global"

        if (.dir_label == "top") {
          display_df <- .clust_desc_display %>%
            dplyr::select(
              Module = cluster,
              Genes = n_genes,
              `Representative genes` = genes,
              `Top enriched terms` = description,
              `gate_sig (%)` = gate_sig,
              `gate_fdr (%)` = gate_fdr,
              `top10% (%)` = top10,
              `top5% (%)` = top5,
              `top1% (%)` = top1,
              `fade_top (%)` = fade_top,
              `rer_acc (%)` = rer_acc,
              `rer_decc (%)` = rer_decc,
              `accum (%)` = accum
            )
        } else if (.dir_label == "bottom") {
          display_df <- .clust_desc_display %>%
            dplyr::select(
              Module = cluster,
              Genes = n_genes,
              `Representative genes` = genes,
              `Top enriched terms` = description,
              `gate_sig (%)` = gate_sig,
              `gate_fdr (%)` = gate_fdr,
              `bot10% (%)` = bot10,
              `bot5% (%)` = bot5,
              `bot1% (%)` = bot1,
              `fade_bottom (%)` = fade_bottom,
              `rer_acc (%)` = rer_acc,
              `rer_decc (%)` = rer_decc,
              `accum (%)` = accum
            )
        } else {
          display_df <- .clust_desc_display %>%
            dplyr::select(
              Module = cluster,
              Genes = n_genes,
              `Representative genes` = genes,
              `Top enriched terms` = description,
              `gate_sig (%)` = gate_sig,
              `gate_fdr (%)` = gate_fdr,
              `top10% (%)` = top10,
              `top5% (%)` = top5,
              `top1% (%)` = top1,
              `bot10% (%)` = bot10,
              `bot5% (%)` = bot5,
              `bot1% (%)` = bot1,
              `fade (%)` = fade,
              `rer_acc (%)` = rer_acc,
              `rer_decc (%)` = rer_decc,
              `accum (%)` = accum
            )
        }
      } else {
        display_df <- .clust_desc_display %>%
          dplyr::select(
            Module = cluster,
            Genes = n_genes,
            `Representative genes` = genes,
            `Top enriched terms` = description
          )
      }

      # Append DOMINO's own per-module significance (row-aligned: display_df
      # derives from .clust_desc_display via select(), preserving row order).
      if (all(c("domino_p_value", "domino_p_adj") %in% names(.clust_desc_display))) {
        display_df <- dplyr::bind_cols(
          display_df,
          .clust_desc_display %>% dplyr::transmute(
            `DOMINO p`     = signif(domino_p_value, 3),
            `DOMINO p.adj` = signif(domino_p_adj, 3)
          )
        )
      }

      print(knitr::kable(
        display_df,
        caption = sprintf("Module functional descriptions — %s", .list_name)
      ) %>% kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"), full_width = TRUE))
    }

    .cl_labels <- setNames(.clust_desc$description, as.character(.clust_desc$cluster))
    .cl_labels <- .cl_labels[!.cl_labels %in% c("Too small", "No enrichment")]

    .clusters_df_viz <- .clusters_df %>%
      dplyr::filter(cluster %in% .clust_desc$cluster[.clust_desc$n_genes >= .min_cluster_genes])
  } else {
    .cl_labels <- NULL
    .clusters_df_viz <- .clusters_df
  }

  # Restrict the PLOTTED graph to nodes in the shown (top-20,
  # min_cluster_genes-passing) modules -- not just their colouring.
  .net_plot <- if (!is.null(.clusters_df_viz) && nrow(.clusters_df_viz) > 0) {
    igraph::induced_subgraph(
      .net_conn,
      igraph::V(.net_conn)[igraph::V(.net_conn)$name %in% .clusters_df_viz$node]
    )
  } else {
    .net_conn
  }

  .png_path <- file.path(string_networks_dir, sprintf("%s_domino_network.png", .list_name))
  png(.png_path, width = 1800, height = 1500, res = 160)
  plot_cluster_network(
    .net_plot, .clusters_df_viz,
    title          = sprintf("%s — DOMINO modules (score >= %s, top 20)", .list_name, params$domino_network_score_thr),
    cluster_labels = .cl_labels
  )
  dev.off()
  # Emit image via cat() so it lands in the text stream at the correct tab position.
  cat(sprintf("![](%s)\n\n", .png_path))

  network_summary_rows[[length(network_summary_rows) + 1]] <- tibble(
    gene_list = .list_name, n_input = length(.mapped), n_connected = .n_nodes,
    n_edges = .n_edges, n_modules = .n_mod
  )
}

if (length(network_summary_rows) > 0) {
  .net_summary <- dplyr::bind_rows(network_summary_rows)
  readr::write_tsv(.net_summary, file.path(string_networks_dir, "network_summary.tsv"))
  cat("\n\n## Network summary\n\n")
  print(knitr::kable(.net_summary, caption = "DOMINO network summary") %>%
          kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")))
}
```

---

## Summary

This report summarizes the results of the **AMI (Active Module Identification)**
analysis: DOMINO-detected network modules per gene list, with STRING-derived
functional labels and DOMINO's own per-module significance.

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — AMI Analysis Module*
*Date of generation: `r Sys.Date()`*
