---
title: "Comparative Enrichment Across Modules - `r params$traitname`"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
    mathjax: null                       # no math in this report
    md_extensions: "-tex_math_dollars"  # gene-set names contain '$' (e.g. V$HNF4_01) - never parse as TeX math
params:
  fcs_dir:    "cmp_fcs"       # dir of per-module <module>_fcs_all.tsv
  fcs_le_dir: "cmp_fcs_le"    # dir of per-module <module>_leading_edge.tsv
  fdr_thr:    0.15
  pperm_thr:  0.025
  domino_module_thr: 0.05  # DOMINO's own module-significance cutoff (domino.nf's domino_module_thr)
  top_n:      250
  traitname:  "trait"
  ami_module_desc_file:       NULL  # AMI's ami_networks/module_descriptions_all_lists.tsv (optional)
  ami_term_membership_file:   NULL  # AMI's ami_networks/term_threshold_membership.tsv (optional, recurrence table)
  posenrich_dotplot_file:     NULL  # posenrich's posenrich_overall_dotplot.tsv (optional)
  # posenrich_leading_edge.tsv (gene:position members of significant posenrich
  # terms, already position-granular), drives the Interesting Genes/Positions tables' posenrich angle
  # AND the Interesting Genes/Interesting Positions tables below.
  posenrich_leading_edge_file: NULL
  # posenrich's pooled-per-term leading-edge display table (one row per
  # significant term, not per cutoff), 14.Position_enrichment_report.Rmd's
  # own Leading edge section export, posenrich_leading_edge_summary.tsv.
  posenrich_leading_edge_summary_file: NULL
  # SCORING's own gene-level table (percentile flags + FADE/RER/Accumulation
  # significance), same file 12.FCS_general_report.Rmd/14.Position_enrichment_report.Rmd
  # consume, reused here for the Interesting Genes/Positions tables.
  fcs_stats_file: NULL
  # Position-level CAAS scores + FADE-site/VEP annotations, same optional
  # inputs 14.Position_enrichment_report.Rmd's Position Characterisation
  # section consumes, reused here for the Interesting Positions table.
  position_scores_file:  NULL
  vep_primateai_file:    NULL
  vep_cosmic_file:       NULL
  fade_sites_top_file:    NULL
  fade_sites_bottom_file: NULL
  # posenrich's own per-position characterization layers (PFAM domain/clan,
  # UCR core/flank region + per-position variability, FUBAR selection call),
  # build_position_gmt.py's position_characterization.tsv - the same GMT-source
  # layers posenrich tests, flattened for a direct Gene/Position join here.
  position_characterization_file: NULL
  # SCORING's published percentile slices (scoring_compute.R), when
  # supplied, gene/position 25/10/5/1% membership is read from these
  # directories instead of being re-derived independently in this report (see
  # fcs_percentile_flags()'s gene_lists_dir param / .position_pct_flags()).
  gene_lists_dir:     NULL
  position_lists_dir: NULL
  seed:               "1998"
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 7, dev = "png", dpi = 200)
suppressPackageStartupMessages({ library(igraph); library(tidyverse); library(knitr); library(kableExtra) })

theme_set(
  theme_minimal(base_size = 13) +
  theme(
    axis.title.y  = element_text(margin = ggplot2::margin(l = 20, r = 20)),
    axis.title.x  = element_text(margin = ggplot2::margin(t = 12)),
    axis.text.x   = element_text(margin = ggplot2::margin(t = 5)),
    plot.title    = element_text(face = "bold", hjust = 0, margin = ggplot2::margin(b = 6)),
    plot.subtitle = element_text(colour = "grey40", hjust = 0, margin = ggplot2::margin(b = 10)),
    legend.position = "right",
    legend.box.margin = ggplot2::margin(t = 5)
  )
)
filter <- dplyr::filter; select <- dplyr::select; rename <- dplyr::rename; mutate <- dplyr::mutate

seed_val <- if (exists("params") && !is.null(params$seed) && nzchar(as.character(params$seed))) as.integer(params$seed) else 1998
set.seed(seed_val)

.src_dir <- file.path(dirname(normalizePath(knitr::current_input(), mustWork = FALSE)), "src")
# fcs_percentile_flags() (used by the Interesting Genes/Positions tables) lives
# in percentile_flags.R, sourced directly rather than via fcs_enrich.R, which
# would drag in a hard RERconverge dependency this report never otherwise
# needs (fcs_enrich.R's fastwilcoxGMT/fcs_run_all machinery is
# report 12/RER-report-specific).
source(file.path(.src_dir, "percentile_flags.R"))

kbl_html <- function(df, height = NULL, escape = TRUE, caption = NULL, ...) {
  k <- kable(df, format = "html", escape = escape, caption = caption, ...) %>%
    kable_styling(bootstrap_options = c("striped", "condensed", "hover"), full_width = FALSE)
  # Wide tables (long term names/descriptions, many columns) always get a
  # horizontal scrollbar instead of overflowing the page; the vertical
  # scroll_box (height) only kicks in on top of that for tall tables.
  k <- scroll_box(k, width = "100%", height = height)
  k
}

# Evidence-tier weight/colour, same convention as 12.FCS_general_report.Rmd's
# Multi-Method Evidence table (green/orange/grey badges), reused everywhere
# an evidence_label appears in this report (Per-module significance tables,
# evidence_label_caas/evidence_label_rer in Interesting Genes/Positions), so
# the same tier always reads the same colour everywhere. Any table using
# .evidence_badge() must be rendered via kbl_html(..., escape = FALSE).
.exploratory_labels <- c("Exploratory")
.evidence_weight_of <- function(lbl) dplyr::case_when(
  lbl == "Supported" ~ 2L, lbl %in% .exploratory_labels ~ 1L, TRUE ~ 0L)
.evidence_colours <- c("Supported" = "#e67e22", "Exploratory" = "#7f8c8d", "None" = "#dcdcdc")
.evidence_badge <- function(x) {
  x <- dplyr::coalesce(x, "None")
  kableExtra::cell_spec(x, "html",
    color = ifelse(x == "None", "#888888", "white"),
    background = .evidence_colours[x], bold = x != "None",
    extra_css = "padding: 2px 6px; border-radius: 3px;")
}

# Fallback tool classifier from a gene-list filename, used when an AMI export
# predates the explicit `tool` column. Defined here (not next to its main use
# in ami_term_membership_tabs below) because cross_angle_gene_hits, an earlier
# and unconditionally-evaluated chunk, also needs it as a fallback.
.ami_tool_classifier <- function(gl) {
  dplyr::case_when(
    grepl("^rer_",  gl, ignore.case = TRUE) ~ "RER",
    grepl("^fade_", gl, ignore.case = TRUE) ~ "FADE",
    TRUE                                    ~ "CAAS"
  )
}
```

```{r load, include=FALSE}
`%||%` <- function(a, b) if (!is.null(a)) a else b
REQ <- c("ranking", "database", "pathway", "stat", "p.adj")

# Per-module FCS results: cmp_fcs/<module>_fcs_all.tsv (caas/rer).
mod_files <- list.files(params$fcs_dir, pattern = "_fcs_all\\.tsv$", full.names = TRUE)
fcs_by_mod <- list()
for (f in mod_files) {
  mod <- sub("_fcs_all\\.tsv$", "", basename(f))
  d <- tryCatch(readr::read_tsv(f, show_col_types = FALSE), error = function(e) NULL)
  if (!is.null(d) && nrow(d) > 0 && all(REQ %in% names(d))) {
    fcs_by_mod[[mod]] <- dplyr::mutate(d, module = mod)
  }
}
fcs_all_mod <- if (length(fcs_by_mod)) dplyr::bind_rows(fcs_by_mod) else tibble()
modules_present <- names(fcs_by_mod)
has_multi <- length(fcs_by_mod) >= 1

# Direction facet (Global/top/bottom), derived once here since every
# directional section below groups by it. Each module's `ranking` vocabulary
# differs: CAAS uses top/bottom/global directly; RER's accelerating/
# decelerating is the same two-way split framed around RER's own statistic
# (RER carries no separate undirected ranking, so it contributes no "global"
# rows). CAAS's "global" ranking is its OWN direction here, kept fully
# separate from top/bottom: a "global" row can never share a (module,
# direction, database, pathway) group key with a "top" or "bottom" row, so
# permod_prep's grouping below never blends the real directional stat and
# the undirected global stat for the same pathway into one summarised row.
if (has_multi) {
  fcs_dir_mod <- fcs_all_mod %>%
    mutate(direction = dplyr::case_when(
      tolower(ranking) %in% c("top", "accelerating")    ~ "top",
      tolower(ranking) %in% c("bottom", "decelerating") ~ "bottom",
      tolower(ranking) == "global"                      ~ "global",
      TRUE                                              ~ NA_character_
    )) %>%
    dplyr::filter(!is.na(direction))
} else {
  fcs_dir_mod <- fcs_all_mod
}

# Per-module leading-edge tables (cmp_fcs_le/<module>_leading_edge.tsv) feed
# the Interesting Genes/Positions criteria below: each module's own
# directional percentile concentration + cross-module corroboration flags
# (FADE/RER/Accumulation, and PrimateAI-3D/COSMIC once wired in from
# posenrich). Percentile columns reflect CAAS's own ranking in every module's
# leading edge (per-module FCS
# reports are annotated with CAAS's fcs_stats.tsv as their overlay) - that's
# intentional, not a bug: it's the shared reference axis every module's
# leading edge is cross-checked against.
le_files <- list.files(params$fcs_le_dir, pattern = "_leading_edge\\.tsv$", full.names = TRUE)
le_by_mod <- list()
for (f in le_files) {
  mod <- sub("_leading_edge\\.tsv$", "", basename(f))
  d <- tryCatch(readr::read_tsv(f, show_col_types = FALSE), error = function(e) NULL)
  if (!is.null(d) && nrow(d) > 0) le_by_mod[[mod]] <- dplyr::mutate(d, module = mod)
}
le <- if (length(le_by_mod)) dplyr::bind_rows(le_by_mod) else tibble()
has_le <- nrow(le) > 0

# ── Cross-angle inputs: AMI (network modules) + posenrich (position-level) ──
# Neither is gene-set-ranked the way CAAS/RER FCS is, so they don't fold into
# fcs_all_mod/le above, each gets its own section below, and all three feed
# the Interesting Genes/Interesting Positions tables via the genes they involve.
.file_ok <- function(p) !is.null(p) && !is.na(p) && nchar(p) > 0 && file.exists(p)

has_ami <- .file_ok(params$ami_module_desc_file)
ami_df  <- if (has_ami) readr::read_tsv(params$ami_module_desc_file, show_col_types = FALSE) else NULL
ami_disp <- NULL

# AMI term membership recurrence table (term_threshold_membership.tsv from
# 13.AMI_analysis.Rmd's render_tool_section_nested()) - imported directly
# rather than re-derived, so the table shown here matches the AMI report exactly.
has_ami_term_membership <- .file_ok(params$ami_term_membership_file)
ami_term_membership_df  <- if (has_ami_term_membership) readr::read_tsv(params$ami_term_membership_file, show_col_types = FALSE) else NULL

# Per-module leading-edge composition tables (fcs_leading_edge_composition.tsv
# from 12.FCS_general_report.Rmd) - imported directly so the table shown here
# matches the FCS reports exactly.
le_comp_files <- list.files(params$fcs_le_dir, pattern = "_leading_edge_composition\\.tsv$", full.names = TRUE)
le_comp_by_mod <- list()
for (f in le_comp_files) {
  mod <- sub("_leading_edge_composition\\.tsv$", "", basename(f))
  d <- tryCatch(readr::read_tsv(f, show_col_types = FALSE), error = function(e) NULL)
  if (!is.null(d) && nrow(d) > 0) le_comp_by_mod[[mod]] <- dplyr::mutate(d, module = mod)
}
le_comp_all_mod <- if (length(le_comp_by_mod)) dplyr::bind_rows(le_comp_by_mod) else tibble()
has_le_comp <- nrow(le_comp_all_mod) > 0

has_posenrich_dotplot <- .file_ok(params$posenrich_dotplot_file)
posenrich_dotplot_df  <- if (has_posenrich_dotplot) readr::read_tsv(params$posenrich_dotplot_file, show_col_types = FALSE) else NULL

# posenrich_leading_edge.tsv: gene:position members of significant posenrich
# terms (already position-granular, ranking, database, pathway, top_frac,
# gene, gene_position). Drives the Interesting Genes/Interesting Positions tables below.
has_posenrich_le <- .file_ok(params$posenrich_leading_edge_file)
posenrich_le_df  <- if (has_posenrich_le) readr::read_tsv(params$posenrich_leading_edge_file, show_col_types = FALSE) else NULL

has_posenrich_le_summary <- .file_ok(params$posenrich_leading_edge_summary_file)
posenrich_le_summary_df  <- if (has_posenrich_le_summary) readr::read_tsv(params$posenrich_leading_edge_summary_file, show_col_types = FALSE) else NULL

# ── Gene/position-level inputs for the Interesting Genes/Positions tables ──
# Same optional files 12.FCS_general_report.Rmd (fcs_stats) and
# 14.Position_enrichment_report.Rmd (position_scores, VEP, FADE sites) already
# consume, reused here rather than re-derived.
has_fcs_stats_cmp <- .file_ok(params$fcs_stats_file)
fcs_stats_cmp_df  <- if (has_fcs_stats_cmp) readr::read_tsv(params$fcs_stats_file, show_col_types = FALSE) else NULL

has_position_scores_cmp <- .file_ok(params$position_scores_file)
position_scores_cmp_df  <- if (has_position_scores_cmp) readr::read_tsv(params$position_scores_file, show_col_types = FALSE) else NULL
if (!is.null(position_scores_cmp_df) && "Position" %in% names(position_scores_cmp_df)) {
  position_scores_cmp_df$Position <- suppressWarnings(as.integer(position_scores_cmp_df$Position))
}
if (!is.null(position_scores_cmp_df) && "Gene" %in% names(position_scores_cmp_df)) {
  position_scores_cmp_df$Gene <- as.character(position_scores_cmp_df$Gene)
}
if (!is.null(position_scores_cmp_df) && "gene" %in% names(position_scores_cmp_df)) {
  position_scores_cmp_df$gene <- as.character(position_scores_cmp_df$gene)
}
if (!is.null(position_scores_cmp_df) && "side" %in% names(position_scores_cmp_df)) {
  position_scores_cmp_df$side <- as.character(position_scores_cmp_df$side)
}

has_vep_primateai_cmp <- .file_ok(params$vep_primateai_file)
vep_primateai_cmp_df  <- if (has_vep_primateai_cmp) readr::read_tsv(params$vep_primateai_file, show_col_types = FALSE) else NULL

has_vep_cosmic_cmp <- .file_ok(params$vep_cosmic_file)
vep_cosmic_cmp_df  <- if (has_vep_cosmic_cmp) readr::read_tsv(params$vep_cosmic_file, show_col_types = FALSE) else NULL

has_fade_sites_top_cmp    <- .file_ok(params$fade_sites_top_file)
has_fade_sites_bottom_cmp <- .file_ok(params$fade_sites_bottom_file)
fade_sites_top_cmp_df <- if (has_fade_sites_top_cmp) {
  readr::read_csv(params$fade_sites_top_file, show_col_types = FALSE) %>%
    dplyr::rename(gene = gene, Position = position) %>%
    dplyr::mutate(Position = suppressWarnings(as.integer(Position)))
} else NULL
fade_sites_bottom_cmp_df <- if (has_fade_sites_bottom_cmp) {
  readr::read_csv(params$fade_sites_bottom_file, show_col_types = FALSE) %>%
    dplyr::rename(gene = gene, Position = position) %>%
    dplyr::mutate(Position = suppressWarnings(as.integer(Position)))
} else NULL

# posenrich's per-position PFAM domain/clan, UCR core/flank region + per-position
# variability, and FUBAR selection call (build_position_gmt.py's
# position_characterization.tsv) - characterization-only, like COSMIC/PAI3D
# below, so joined straight in without any percentile/significance gating.
has_position_char_cmp <- .file_ok(params$position_characterization_file)
position_char_cmp_df  <- if (has_position_char_cmp) {
  # A header-only file (no gene tested any position, plausible on a small
  # universe) gives readr nothing to infer types from, so every column -
  # Position included - comes back character; force it back to integer so
  # the later left_join by c("gene", "Position") against position_scores_cmp_df's
  # numeric Position doesn't fail on incompatible types (same fix already
  # applied to fade_sites_top/bottom_cmp_df above for the same reason).
  readr::read_tsv(params$position_characterization_file, show_col_types = FALSE) %>%
    dplyr::mutate(Position = suppressWarnings(as.integer(Position)))
} else NULL

# Per-gene attribute table: percentile flags (from fcs_stats' own score_top/
# score_bottom, same derivation as 12.FCS_general_report.Rmd) + the gene-level
# flag_* columns fcs_stats.tsv already carries (gate/FADE/RER/Accumulation,
# direction-aware: flag_accum_top/flag_accum_bottom are the direction-restricted
# tests, not the non-directional flag_accum).
gene_attr_df <- if (has_fcs_stats_cmp) {
  pf <- fcs_percentile_flags(fcs_stats_cmp_df, gene_lists_dir = params$gene_lists_dir)
  flag_cols_cmp <- grep("^flag_", names(fcs_stats_cmp_df), value = TRUE)
  dplyr::left_join(pf, fcs_stats_cmp_df[, c("gene", flag_cols_cmp)], by = "gene")
} else NULL
```

```{r cross_angle_gene_hits, include=FALSE}
# ── Cross-angle gene-level hit sets ───────────────────────────────────────
# Computed once here so the per-module significance section, the AMI module
# description section, and the Interesting Genes/Positions tables below all
# read the same per-gene FCS/posenrich corroboration, one shared definition,
# not three drifting copies.
#
# Each angle uses ITS OWN module's significance definition (not "any
# leading-edge/described membership"), FCS/RER's dual p.adj+p.perm gate,
# posenrich's own p_adj gate, DOMINO's own module-significance cutoff.
# FCS and posenrich both carry a genuine "global" ranking (CAAS's undirected
# composite; posenrich's own directions = ["global","top","bottom"]), so
# fcs_genes/posenrich_genes have real direction=="Global" rows. AMI has no
# equivalent, DOMINO only ever runs on Top/Bottom percentile gene lists
# (.TOP_THRESHOLDS/.BOTTOM_THRESHOLDS in 13.AMI_analysis.Rmd, no undirected
# list), so ami_genes' direction column is only ever "Top"/"Bottom"; a
# "Global" row for AMI never exists here. This means the Interesting Genes/Positions tables' Global
# tab can never carry an AMI contribution, unlike .ami_desc_for(dir_filter = NULL), used by the
# Interesting Genes/Positions AMI cells, which pools Top+Bottom explicitly
# for its own "Global" view instead of relying on a literal direction match.

# FCS/RER own significance: BH < fdr_thr AND (p.perm < pperm_thr when
# present), same formula permod_prep below applies, computed once here off
# fcs_dir_mod so this chunk and permod_prep agree on what "significant" means.
# fcs_leading_edge.tsv (`le`) is NOT itself restricted to significant terms
# (12.FCS_general_report.Rmd's leading-edge export only requires drv_score >
# 0, i.e. "this gene drives some signal in this term", not "this term
# cleared BH"), inner_join against fcs_sig_terms is what actually restricts
# fcs_genes to genes belonging to a term each module itself calls significant.
fcs_sig_terms <- if (has_multi) {
  fcs_dir_mod %>%
    dplyr::filter(stat > 0, !is.na(p.adj), p.adj < params$fdr_thr,
                  !("p.perm" %in% names(.)) | is.na(p.perm) | p.perm < (params$pperm_thr %||% 1)) %>%
    dplyr::mutate(description = dplyr::coalesce(description, pathway)) %>%
    dplyr::distinct(module, direction, database, pathway, description)
} else {
  tibble::tibble(module = character(), direction = character(), database = character(),
                 pathway = character(), description = character())
}

fcs_genes <- if (has_le && has_multi) {
  le %>%
    dplyr::mutate(direction = dplyr::case_when(
      ranking %in% c("top", "accelerating")    ~ "top",
      ranking %in% c("bottom", "decelerating") ~ "bottom",
      ranking == "global"                      ~ "global",
      TRUE                                      ~ NA_character_
    )) %>%
    dplyr::filter(!is.na(direction)) %>%
    dplyr::inner_join(fcs_sig_terms, by = c("module", "direction", "database", "pathway")) %>%
    dplyr::mutate(direction = tools::toTitleCase(direction)) %>%
    dplyr::distinct(gene, direction) %>%
    dplyr::mutate(fcs_hit = TRUE)
} else {
  tibble::tibble(gene = character(), direction = character(), fcs_hit = logical())
}

# AMI angle: DOMINO's own module-significance cutoff (domino_module_thr,
# domino.nf's own default 0.05) gates which described modules count --
# "described" alone (not Too small/No enrichment) was never a significance
# claim, just "STRING could summarise it". Expands each significant module's
# FULL membership into one row per gene: `all_genes` (every member gene,
# uncapped, describe_clusters() in 13.AMI_analysis.Rmd) when present, else
# the older capped `genes` column (top-8 representatives) for
# module_descriptions_all_lists.tsv files written before all_genes existed.
# direction is only ever "Top"/"Bottom" here (DOMINO never runs on an
# undirected gene list, so there's no "Global" value to filter for, see
# this chunk's opening comment).
.normalize_term <- function(x) {
  x <- tolower(x)
  x <- gsub("[^a-z0-9 ]", " ", x)
  gsub("\\s+", " ", trimws(x))
}
.cluster_terms <- function(terms, sim_thr = 0.85) {
  if (length(terms) == 0) return(character(0))
  norm <- vapply(terms, .normalize_term, character(1))
  reps_norm <- character(0); reps_orig <- character(0)
  assign <- integer(length(terms))
  for (i in seq_along(terms)) {
    if (length(reps_norm) == 0) { reps_norm <- norm[i]; reps_orig <- terms[i]; assign[i] <- 1L; next }
    d <- utils::adist(norm[i], reps_norm)[1, ]
    sim <- 1 - d / pmax(nchar(norm[i]), nchar(reps_norm), 1)
    best <- which.max(sim)
    if (sim[best] >= sim_thr) {
      assign[i] <- best
    } else {
      reps_norm <- c(reps_norm, norm[i]); reps_orig <- c(reps_orig, terms[i])
      assign[i] <- length(reps_norm)
    }
  }
  reps_orig[assign]
}

ami_disp <- if (has_ami && !is.null(ami_df) && nrow(ami_df) > 0) {
  d <- ami_df %>%
    dplyr::mutate(
      .tool = if ("tool" %in% names(.)) as.character(tool) else .ami_tool_classifier(gene_list),
      direction = if ("direction" %in% names(.)) dplyr::case_when(
        direction %in% c("Acc", "accelerating", "Accelerating") ~ "Top",
        direction %in% c("Decc", "decelerating", "Decelerating") ~ "Bottom",
        TRUE ~ as.character(direction)
      ) else NA_character_
    )
  if ("description" %in% names(d)) {
    d <- d %>% dplyr::filter(!description %in% c("Too small", "No enrichment"))
  }
  if ("domino_p_adj" %in% names(d)) {
    d <- d %>% dplyr::filter(!is.na(domino_p_adj), domino_p_adj < (params$domino_module_thr %||% 0.05))
  }
  if (nrow(d) > 0 && "description" %in% names(d)) {
    d$term <- .cluster_terms(d$description)
  } else if (nrow(d) > 0) {
    d$term <- character(nrow(d))
  } else {
    d$term <- character(0)
  }
  d
} else NULL

ami_genes <- if (has_ami && !is.null(ami_df) && nrow(ami_df) > 0) {
  .ami_gene_col <- if ("all_genes" %in% names(ami_df)) "all_genes" else if ("genes" %in% names(ami_df)) "genes" else NULL
  if (is.null(.ami_gene_col)) {
    tibble::tibble(gene = character(), direction = character(), ami_hit = logical())
  } else {
    d <- ami_df %>%
      dplyr::mutate(
        .tool = if ("tool" %in% names(.)) as.character(tool) else .ami_tool_classifier(gene_list),
        direction = if ("direction" %in% names(.)) dplyr::case_when(
          direction %in% c("Acc", "accelerating", "Accelerating") ~ "Top",
          direction %in% c("Decc", "decelerating", "Decelerating") ~ "Bottom",
          TRUE ~ as.character(direction)
        ) else NA_character_
      )
    if ("description" %in% names(d)) {
      d <- d %>% dplyr::filter(!description %in% c("Too small", "No enrichment"))
    }
    if ("direction" %in% names(d)) {
      d <- d %>% dplyr::filter(direction %in% c("Top", "Bottom"))
    }
    if ("domino_p_adj" %in% names(d)) {
      d <- d %>% dplyr::filter(!is.na(domino_p_adj), domino_p_adj < (params$domino_module_thr %||% 0.05))
    }
    if (nrow(d) > 0) {
      d %>%
        dplyr::mutate(gene = strsplit(.data[[.ami_gene_col]], ",\\s*")) %>%
        tidyr::unnest(gene) %>%
        dplyr::mutate(gene = trimws(gene)) %>%
        dplyr::distinct(gene, direction) %>%
        dplyr::mutate(ami_hit = TRUE)
    } else {
      tibble::tibble(gene = character(), direction = character(), ami_hit = logical())
    }
  }
} else {
  tibble::tibble(gene = character(), direction = character(), ami_hit = logical())
}

# Posenrich angle: posenrich_leading_edge.tsv is ALREADY restricted to
# significant terms (Lachenbruch gates of posenrich_enrich.py)
# itself (only rows with r["sig"] get written), no additional gate needed
# here, just direction mapping (now including "global"). Also carries the
# actual Position(s) behind each hit (kept for the Interesting Genes/Positions tables' Position
# column below, lets a reader see whether the specific position posenrich
# flagged for a gene is the same one driving that gene's FCS/AMI signal, not
# just that the gene appears in both).
posenrich_genes <- if (has_posenrich_le) {
  posenrich_le_df %>%
    dplyr::mutate(direction = dplyr::case_when(
      ranking %in% c("top", "accelerating")    ~ "top",
      ranking %in% c("bottom", "decelerating") ~ "bottom",
      ranking == "global"                      ~ "global",
      TRUE                                      ~ NA_character_
    ),
    Position = sub(".*:", "", gene_position)) %>%
    dplyr::filter(!is.na(direction)) %>%
    dplyr::mutate(direction = tools::toTitleCase(direction)) %>%
    dplyr::group_by(gene, direction) %>%
    dplyr::summarise(Position = paste(sort(unique(Position)), collapse = ", "), .groups = "drop") %>%
    dplyr::mutate(posenrich_hit = TRUE)
} else {
  tibble::tibble(gene = character(), direction = character(), Position = character(), posenrich_hit = logical())
}

```

# Overview

```{r overview, echo=FALSE}
mod_rows <- if (has_multi) sapply(fcs_by_mod, nrow) else integer(0)
tibble::tibble(
  Source = paste0("FCS module: ", names(mod_rows)),
  Rows   = unname(mod_rows)
) %>% kbl_html()
```

This report compares enrichment **across CAAS and RERconverge** (the two modules
with well-behaved FCS rankings and real permulation nulls) for `r params$traitname`.
FADE and Accumulation do not run their own FCS ranking: FADE's statistic is a
max over many sites and Accumulation has no permulation null, so neither
supports a reliable genome-wide gene-set enrichment test on its own. Both
remain available as **cross-module corroboration**, surfaced as flag columns
(`%fade_*`/`%rer_*`/`%accum*`) in the imported leading-edge composition tables
below and as weighted criteria in the Interesting Genes/Positions tables further
down. A gene set is "significant" at BH < `r params$fdr_thr` (and permulation p
< threshold where available) with AUC > 0.5 - see
[12.FCS_general_report.Rmd](12.FCS_general_report.Rmd)'s Methods section for the
two-method definition (Wilcoxon-AUC / Lachenbruch) and the
`evidence_label`/`evidence_score` ranking convention this report's own tables follow.

DOMINO active-module (AMI) results and posenrich's position-level enrichment are
neither gene-set-ranked the same way as FCS, so they get their own sections below.

# Per-module significance {.tabset .tabset-fade}

Leading-edge composition from each module's own FCS report
([12.FCS_general_report.Rmd](12.FCS_general_report.Rmd)), imported directly so the
tables shown here match the per-module reports exactly. For each significantly-
enriched set, the **percentage of its leading-edge genes** carrying each attribute:
significance gates, directional score percentiles, and cross-module support.
Rows follow the same ranking convention as the source FCS report: **evidence
tier first** (Supported > Exploratory evidence, from the Wilcoxon-AUC /
Lachenbruch vote), **then `evidence_score`** - each *passing*
method's own statistic (AUC, `L.frac_mag`) converted to a percentile
rank so the two are comparable on one scale, zeroed for a method that
didn't pass, then summed - **then alphabetical** as a tie-break, already
baked into the row order of the imported table, not recomputed here. Faceted
Global/Top/Bottom, each module in its own sub-tab, capped at `r params$top_n`
rows.

`r if (!has_le_comp && !has_multi) "> **No module FCS results available - skipping.**"`

```{r permod_prep, eval=has_multi}
dir.create("compare_results", showWarnings = FALSE)
```

```{r permod_le_comp_tabs, eval=has_le_comp, results='asis'}
# Direction mapping for the ranking column in the imported tables.
.le_comp_dir_map <- c(top = "Top", accelerating = "Top", bottom = "Bottom",
                       decelerating = "Bottom", global = "Global")

le_comp_all_mod <- le_comp_all_mod %>%
  dplyr::mutate(direction = dplyr::coalesce(.le_comp_dir_map[tolower(ranking)], "Global"))

for (dr in c("Global", "Top", "Bottom")) {
  cat(sprintf("\n## %s {.tabset}\n\n", dr))

  mods_to_show <- union(modules_present, unique(le_comp_all_mod$module))
  for (m in mods_to_show) {
    cat(sprintf("\n### %s\n\n", toupper(m)))
    d <- le_comp_all_mod %>%
      dplyr::filter(module == m, direction == dr) %>%
      dplyr::select(-module, -direction) %>%
      head(params$top_n)
    if (nrow(d) == 0) {
      cat(sprintf("_No leading-edge composition for %s (%s)._\n\n", toupper(m), dr))
    } else {
      if ("evidence_label" %in% names(d)) d$evidence_label <- .evidence_badge(d$evidence_label)
      print(kbl_html(d, height = if (nrow(d) > 15) "480px" else NULL, escape = FALSE))
    }
  }
}
```

```{r permod_tabs_fallback, eval=has_multi && !has_le_comp, results='asis'}
# Fallback: recompute per-module significance when imported leading-edge
# composition tables are not available (e.g. older pipeline runs). Same
# ranking convention as 12.FCS_general_report.Rmd: evidence tier, then
# evidence_score, then alphabetical, recomputed here from the raw
# per-module fcs_all.tsv columns (percentile-ranked within each module x
# direction tab, the population these tables actually compare within)
# rather than reusing the source report's own evidence_rank/evidence_score
# directly, since older exports may predate them.
thr <- params$fdr_thr
.evidence_rank_of <- function(lbl) dplyr::case_when(
  lbl == "Supported" ~ 1L, lbl %in% .exploratory_labels ~ 2L, TRUE ~ 3L)

if (!"evidence_label" %in% names(fcs_dir_mod)) fcs_dir_mod$evidence_label <- NA_character_
for (cl in c("lach_frac_magnitude")) {
  if (!cl %in% names(fcs_dir_mod)) fcs_dir_mod[[cl]] <- NA_real_
}
for (cl in c("sig_wilcoxon", "sig_lachenbruch")) {
  if (!cl %in% names(fcs_dir_mod)) fcs_dir_mod[[cl]] <- FALSE
}

permod_best <- fcs_dir_mod %>%
  filter(stat > 0) %>%
  mutate(evidence_label = dplyr::coalesce(evidence_label, "Exploratory")) %>%
  group_by(module, direction, database, pathway) %>%
  summarise(AUC = max(stat, na.rm = TRUE) + 0.5,
            p.adj = min(p.adj, na.rm = TRUE),
            lach_frac_magnitude = suppressWarnings({ v <- na.omit(lach_frac_magnitude); if (length(v)) v[1] else NA_real_ }),
            sig_wilcoxon    = any(sig_wilcoxon %in% TRUE),
            sig_lachenbruch = any(sig_lachenbruch %in% TRUE),
            evidence_label = dplyr::first(evidence_label),
            description = dplyr::first(dplyr::coalesce(description, pathway)),
            .groups = "drop") %>%
  # evidence_score: each PASSING method's statistic converted to its own
  # percentile rank within this (module, direction) tab, zeroed when that
  # method didn't pass, see 12.FCS_general_report.Rmd's Methods section
  # for why (raw AUC and L.frac_mag are on incompatible scales).
  dplyr::group_by(module, direction) %>%
  dplyr::mutate(.auc_pct = dplyr::percent_rank(AUC),
                .lfm_pct = dplyr::percent_rank(lach_frac_magnitude)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    evidence_rank = .evidence_rank_of(evidence_label),
    evidence_score = dplyr::if_else(sig_wilcoxon,    dplyr::coalesce(.auc_pct, 0), 0) +
                     dplyr::if_else(sig_lachenbruch, dplyr::coalesce(.lfm_pct, 0), 0)
  ) %>%
  dplyr::select(-.auc_pct, -.lfm_pct) %>%
  dplyr::left_join(fcs_sig_terms %>% dplyr::select(-description) %>% dplyr::mutate(sig = TRUE),
                    by = c("module", "direction", "database", "pathway")) %>%
  dplyr::mutate(sig = dplyr::coalesce(sig, FALSE))

for (dr in c("global", "top", "bottom")) {
  cat(sprintf("\n## %s {.tabset}\n\n", tools::toTitleCase(dr)))
  for (m in modules_present) {
    cat(sprintf("\n### %s\n\n", toupper(m)))
    d <- permod_best %>%
      filter(module == m, direction == dr, sig) %>%
      arrange(evidence_rank, dplyr::desc(evidence_score), description) %>%
      mutate(Description = stringr::str_trunc(description, 60),
             AUC = round(AUC, 3), p.adj = signif(p.adj, 3),
             L.frac_mag = round(lach_frac_magnitude, 2)) %>%
      select(database, pathway, Description, evidence_label, AUC, p.adj, any_of("L.frac_mag")) %>%
      head(params$top_n)
    if (nrow(d) == 0) {
      cat(sprintf("_No gene set is significant in %s (%s) at BH < %.2g._\n\n", toupper(m), dr, thr))
    } else {
      d$evidence_label <- .evidence_badge(d$evidence_label)
      print(kbl_html(d, height = if (nrow(d) > 15) "420px" else NULL, escape = FALSE))
    }
  }
}
```

---

# AMI module descriptions {.tabset .tabset-fade}

`r if (!has_ami && !has_ami_term_membership) "> **No AMI module descriptions available (AMI/DOMINO didn't run, or the file wasn't wired in) - skipping.**"`

DOMINO active modules ([13.AMI_analysis.Rmd](13.AMI_analysis.Rmd)) aren't gene-set-ranked
like CAAS/RER FCS, so they don't fold into the FCS tables above. The recurrence table
below shows each described term and which gene lists it was recovered under - imported
directly from the AMI report so the table shown here matches that report exactly.

```{r ami_term_membership_tabs, eval=has_ami_term_membership, results='asis'}
# Imported term membership recurrence table from AMI report.
# .ami_tool_classifier is defined in the setup chunk above (needed earlier by
# cross_angle_gene_hits too).
.ami_tm_direction <- function(gl) {
  dplyr::case_when(
    grepl("global|significant$", gl, ignore.case = TRUE) & !grepl("top|bottom|accelerating|decelerating", gl, ignore.case = TRUE) ~ "Global",
    grepl("top|accelerating", gl, ignore.case = TRUE) ~ "Top",
    grepl("bottom|decelerating", gl, ignore.case = TRUE) ~ "Bottom",
    TRUE ~ "Global"
  )
}

ami_tm <- ami_term_membership_df %>%
  dplyr::mutate(
    .tool      = if ("tool" %in% names(.)) as.character(tool) else .ami_tool_classifier(gene_list),
    .direction = dplyr::case_when(
      "direction" %in% names(.) & direction %in% c("Acc", "accelerating", "Accelerating") ~ "Top",
      "direction" %in% names(.) & direction %in% c("Decc", "decelerating", "Decelerating") ~ "Bottom",
      "direction" %in% names(.) & direction %in% c("Top", "Bottom", "Global") ~ as.character(direction),
      TRUE ~ .ami_tm_direction(gene_list)
    )
  )

for (tl in c("CAAS", "RER", "FADE")) {
  cat(sprintf("\n## %s {.tabset}\n\n", tl))
  tl_df <- ami_tm %>% dplyr::filter(.tool == tl)
  if (nrow(tl_df) == 0) {
    cat(sprintf("_No AMI active modules for %s._\n\n", tl))
    next
  }

  for (dr in c("Global", "Top", "Bottom")) {
    cat(sprintf("\n### %s\n\n", dr))
    sub <- if (dr == "Global") tl_df else tl_df %>% dplyr::filter(.direction == dr)
    if (nrow(sub) == 0) {
      cat(sprintf("_No described terms for %s (%s)._\n\n", tl, dr))
      next
    }

    disp <- sub %>%
      dplyr::distinct(term, gene_list) %>%
      dplyr::group_by(term) %>%
      dplyr::summarise(`N lists` = dplyr::n(),
                        `Gene lists` = paste(sort(unique(gene_list)), collapse = ", "),
                        .groups = "drop") %>%
      dplyr::arrange(dplyr::desc(`N lists`), term) %>%
      dplyr::rename(Term = term) %>%
      head(params$top_n)
    print(kbl_html(disp, height = if (nrow(disp) > 15) "420px" else NULL))
  }
}
```

```{r ami_desc_fallback_prep, eval=has_ami && !has_ami_term_membership}
# Fallback: re-derive when the imported term membership file isn't available.
.build_ami_rollup <- function(df) {
  df %>%
    dplyr::group_by(term) %>%
    dplyr::summarise(
      `N modules`    = dplyr::n(),
      `Gene lists`   = paste(sort(unique(gene_list)), collapse = ", "),
      `DOMINO p.adj` = { v <- if ("domino_p_adj" %in% names(.data)) domino_p_adj[!is.na(domino_p_adj)] else numeric(0); if (length(v)) signif(min(v), 3) else NA_real_ },
      Genes          = suppressWarnings(max(n_genes, na.rm = TRUE)),
      .groups = "drop"
    ) %>%
    dplyr::rename(Term = term) %>%
    dplyr::arrange(`DOMINO p.adj`)
}
```

```{r ami_desc_fallback_tabs, eval=has_ami && !has_ami_term_membership, results='asis'}
for (tl in c("CAAS", "RER", "FADE")) {
  cat(sprintf("\n## %s {.tabset}\n\n", tl))
  if (!exists("ami_disp") || is.null(ami_disp) || nrow(ami_disp) == 0) {
    cat(sprintf("_No described modules for %s._\n\n", tl))
    next
  }
  tl_disp <- if (".tool" %in% names(ami_disp)) ami_disp %>% dplyr::filter(.tool == tl) else ami_disp
  if (nrow(tl_disp) == 0) {
    cat(sprintf("_No described modules for %s._\n\n", tl))
    next
  }
  for (dr in c("Global", "Top", "Bottom")) {
    cat(sprintf("\n### %s\n\n", dr))
    sub <- if (dr == "Global") tl_disp else if ("direction" %in% names(tl_disp)) tl_disp %>% dplyr::filter(direction == dr) else tl_disp
    if (nrow(sub) == 0) { cat(sprintf("_No described modules for %s (%s)._\n\n", tl, dr)); next }
    disp <- .build_ami_rollup(sub) %>% head(params$top_n)
    print(kbl_html(disp, height = if (nrow(disp) > 15) "420px" else NULL))
  }
}
```

---

# Posenrich position-level enrichment {.tabset .tabset-fade}

<details>
<summary>Click to show/hide</summary>

`r if (!has_posenrich_le_summary && !has_posenrich_le) "> **No posenrich outputs available (--posenrich was off, or the files weren't wired in) - skipping.**"`

Position-level enrichment ([14.Position_enrichment_report.Rmd](14.Position_enrichment_report.Rmd))
operates on individual CAAS positions rather than genes, so it gets its own section
here rather than folding into the gene-set FCS tables above. Its significant terms
feed the Interesting Genes/Interesting Positions tables further down (individual
gene- and position-level corroboration, in full detail). Faceted by direction
below (`r params$top_n` rows each, not one shared budget split across both) so a
direction with fewer but stronger hits isn't crowded out of a single pooled
table by the other.

## Overall dotplot terms - merged summary {.tabset}

One row per significant term, **merged across name variants** that are
themselves `/`-joined synonyms in their own description (e.g. a term
described as `"V$HNF4_01 / HNF4A"` is treated as the same underlying term as
one independently named `"HNF4A"` elsewhere) - connected via shared segments,
not just identical full strings, so synonym spelling drift across databases
doesn't fragment one biological term into several rows. `Merged from` lists
every original (database, pathway) pair folded into the row, so the merge is
auditable rather than a silent black box.

`Cutoffs`/`N cutoffs` show which top-fraction cutoffs (10/5/1%) the term
recurred significant at *before* merging - a plain recurrence count, unioned
across merged aliases. `χ²`/`p.adj` (best across merged aliases: highest
χ², lowest p.adj) come straight from
[14.Position_enrichment_report.Rmd](14.Position_enrichment_report.Rmd)'s own
Leading edge table (`posenrich_leading_edge_summary.tsv`, imported below);
this report runs one test (Lachenbruch two-part test on the position scores), so there's
no multi-method vote to combine here, unlike the gene-level FCS tables above.

```{r posenrich_dotplot_prep, eval=has_posenrich_le_summary}
# Sourced from the SAME imported table as the Leading edge sub-tab below
# (posenrich_leading_edge_summary.tsv), this section is a synonym-merged
# ROLLUP of that table, not an independent derivation, so the two never drift.
posenrich_dotplot_disp <- posenrich_le_summary_df %>%
  dplyr::rename(database = Database, pathway = Pathway, p_adj = `p.adj`) %>%
  dplyr::mutate(.direction = tools::toTitleCase(Ranking))

# Fuzzy '/'-synonym merge: two rows merge iff they share at least one trimmed
# '/'-segment of their own Description, within the same direction (never
# across direction, a top-direction term and a bottom-direction term merging
# would conflate two different biological claims). Connected-components over
# a segment-sharing graph, not exact-string grouping, so "HNF4A" and
# "V$HNF4_01 / HNF4A" merge even though neither string contains the other.
fuzzy_merge_terms <- function(df) {
  if (nrow(df) == 0) return(df %>% dplyr::mutate(.merge_id = integer(0)))
  df <- df %>% dplyr::mutate(.row_id = dplyr::row_number())
  segs <- df %>%
    dplyr::mutate(seg = strsplit(Description, "\\s*/\\s*")) %>%
    tidyr::unnest(seg) %>%
    dplyr::mutate(seg = trimws(tolower(seg))) %>%
    dplyr::filter(nzchar(seg))
  if (nrow(segs) == 0 || dplyr::n_distinct(segs$seg) == nrow(segs)) {
    # Nothing shares a segment with anything else, every row is its own component.
    return(df %>% dplyr::mutate(.merge_id = .row_id))
  }
  edges <- segs %>%
    dplyr::inner_join(segs, by = "seg", suffix = c("_a", "_b"), relationship = "many-to-many") %>%
    dplyr::filter(.row_id_a < .row_id_b) %>%
    dplyr::distinct(.row_id_a, .row_id_b)
  g <- igraph::graph_from_data_frame(edges, directed = FALSE, vertices = df$.row_id)
  comp <- igraph::components(g)
  df %>% dplyr::mutate(.merge_id = comp$membership[as.character(.row_id)])
}
```

```{r posenrich_dotplot_tabs, eval=has_posenrich_le_summary, results='asis'}
has_chi_col <- "χ²" %in% names(posenrich_dotplot_disp)
for (dr in c("Global", "Top", "Bottom")) {
  cat(sprintf("\n### %s\n\n", dr))
  dr_df <- posenrich_dotplot_disp %>% dplyr::filter(.direction == dr)
  if (nrow(dr_df) == 0) { cat("_No terms for this direction._\n\n"); next }

  merged <- fuzzy_merge_terms(dr_df) %>%
    dplyr::group_by(.merge_id) %>%
    dplyr::summarise(
      Description  = dplyr::first(Description[order(p_adj)]),
      `Merged from` = paste(unique(sprintf("%s: %s", database, pathway)), collapse = "; "),
      Cutoffs       = paste(unique(unlist(strsplit(Cutoffs, ",\\s*"))), collapse = ", "),
      `N cutoffs`   = suppressWarnings(max(`N cutoffs`, na.rm = TRUE)),
      `N merged`    = dplyr::n(),
      `χ²`          = if (has_chi_col) round(suppressWarnings(max(`χ²`, na.rm = TRUE)), 3) else NA_real_,
      `p.adj`       = signif(suppressWarnings(min(p_adj, na.rm = TRUE)), 3),
      .groups = "drop"
    ) %>%
    # Same convention as 14.Position_enrichment_report.Rmd's "Term -> cutoff
    # membership" table: recurrence (N cutoffs) first, alphabetical tie-break.
    dplyr::arrange(dplyr::desc(`N cutoffs`), Description) %>%
    dplyr::select(-.merge_id) %>%
    { if (!has_chi_col) dplyr::select(., -`χ²`) else . } %>%
    head(params$top_n)

  print(kbl_html(merged, height = if (nrow(merged) > 15) "460px" else NULL))
}
```

## Leading edge {.tabset .tabset-fade}

Imported directly from [14.Position_enrichment_report.Rmd](14.Position_enrichment_report.Rmd)'s
own Leading edge section (`posenrich_leading_edge_summary.tsv`) so the table
shown here matches that report exactly - one row per significant term
(pooled across the 10/5/1% cutoffs it was tested at), with the union of
observed `gene:position` members, cross-module corroboration columns, and
`N cutoffs`/`Cutoffs` as a plain recurrence audit trail. Faceted by
direction, capped at `r params$top_n` rows.

```{r posenrich_le_summary_tabs, eval=has_posenrich_le_summary, results='asis'}
for (dr in c("GLOBAL", "TOP", "BOTTOM")) {
  cat(sprintf("\n### %s\n\n", tools::toTitleCase(dr)))
  d <- posenrich_le_summary_df %>%
    dplyr::filter(toupper(Ranking) == dr) %>%
    dplyr::select(-Ranking) %>%
    head(params$top_n)
  if (nrow(d) == 0) {
    cat(sprintf("_No significant terms for %s._\n\n", tools::toTitleCase(dr)))
  } else {
    print(kbl_html(d, height = if (nrow(d) > 15) "480px" else NULL))
  }
}
```

`r if (!has_posenrich_le_summary) "> **posenrich_leading_edge_summary_file not supplied - Leading edge table skipped.**"`

</details>

---

# Interesting Genes {.tabset .tabset-fade}

Prioritized gene candidates faceted by direction (**Global**, **Top**, **Bottom**). Columns are organized into four functional bands:

- **Metadatos** (grey): Gene identifier (`Gene`).
- **Pondera** (blue): Criteria directly contributing to `Composite score`:
  - *CAAS percentiles* (`25%`, `10%`, `5%`, `1%`): collapsed to deepest nested tier (0–4).
  - *CAAS FCS evidence* (`CAAS evidence`): strongest FCS evidence tier reached (Supported = 2, Exploratory = 1, None = 0).
  - *Accumulation* (`Accum_dir`): directional US accumulation significance (0–1).
  - *RER direction* (`RER_dir`): direction-matching acceleration/deceleration (0–1).
  - *RER FCS evidence* (`RER evidence`): strongest RER evidence tier reached (Supported = 2, Exploratory = 1, None = 0).
  - *FADE direction* (`FADE_dir`): directional excess divergence significance (0–1).
- **Descriptivas** (amber): Characterization-only annotations that do not contribute to weighting (FCS significant term descriptions, AMI/DOMINO module descriptions, and position-enrichment terms). Neither `posenrich` nor `AMI` contributes to scoring.
- **Resultados** (green): Family subtotal scores (`caas_family`, `rer_family`, `fade_family`), `N criteria` (count of weighting criteria satisfied), and `Composite score` (total sum).

Only columns corresponding to analyses that were executed are displayed.

```{r interesting_shared_helpers}
# Shared by both Interesting Genes AND Interesting Positions below, defined
# unconditionally (not gated on has_fcs_stats_cmp/has_position_scores_cmp)
# because either table's own gate can be TRUE independently of the other's,
# and build_interesting_positions() below calls these same helpers. Gating
# them behind only one of the two flags would leave them undefined whenever
# that flag was FALSE but the other (needing them) was TRUE.

# Dependency-free HTML escaping (avoids requiring htmltools be attached) and
# a collapsible-<details> renderer for semicolon-joined multi-term cells --
# any kable() call including a column built by .collapsible_list() MUST be
# passed escape = FALSE, or the markup renders as literal text.
.html_escape <- function(x) {
  x <- gsub("&", "&amp;", x, fixed = TRUE)
  x <- gsub("<", "&lt;",  x, fixed = TRUE)
  gsub(">", "&gt;", x, fixed = TRUE)
}
.collapsible_list <- function(joined, sep = "; ") {
  vapply(joined, function(x) {
    if (is.na(x) || !nzchar(x)) return("")
    items <- strsplit(x, sep, fixed = TRUE)[[1]]
    summary_txt <- if (length(items) == 1) .html_escape(stringr::str_trunc(items[1], 40)) else sprintf("%d terms", length(items))
    sprintf("<details><summary>%s</summary>%s</details>",
            summary_txt,
            paste(sprintf("<div>%s</div>", .html_escape(items)), collapse = ""))
  }, character(1), USE.NAMES = FALSE)
}

.collapsible_hypotheses <- function(joined) {
  vapply(joined, function(x) {
    if (is.na(x) || !nzchar(x)) return("")
    items <- unlist(strsplit(as.character(x), "[,;]\\s*"))
    items <- items[nzchar(items)]
    if (length(items) == 0) return("")
    summary_txt <- if (length(items) == 1) .html_escape(items[1]) else sprintf("%d hypotheses", length(items))
    sprintf("<details><summary>%s</summary>%s</details>",
            summary_txt,
            paste(sprintf("<div>%s</div>", .html_escape(items)), collapse = ""))
  }, character(1), USE.NAMES = FALSE)
}

# Per-gene best evidence tier per module (Supported=2/Exploratory=1/
# none=0), alongside the existing term-description text, evidence_label_caas/
# evidence_label_rer surface in Interesting Genes/Positions below as their own
# weighted criterion, distinct from the plain CAAS_FCS_sig/RER_FCS_sig boolean
# (which only says "belongs to >=1 significant term", not how strong the best
# one is).
.fcs_terms_for <- function(rk_set) {
  empty <- tibble::tibble(gene = character(), caas_fcs_terms = character(), rer_fcs_terms = character(),
                          evidence_label_caas = character(), evidence_label_rer = character())
  if (!has_le || !has_multi) return(empty)
  res <- le %>%
    dplyr::filter(is.null(rk_set) | ranking %in% rk_set) %>%
    dplyr::mutate(direction = dplyr::case_when(
      tolower(ranking) %in% c("top", "accelerating")    ~ "top",
      tolower(ranking) %in% c("bottom", "decelerating") ~ "bottom",
      tolower(ranking) == "global"                      ~ "global",
      TRUE                                              ~ NA_character_
    )) %>%
    dplyr::filter(!is.na(direction)) %>%
    dplyr::inner_join(fcs_sig_terms, by = c("module", "direction", "database", "pathway")) %>%
    dplyr::left_join(fcs_dir_mod %>% dplyr::distinct(module, direction, database, pathway, evidence_label),
                      by = c("module", "direction", "database", "pathway"))

  .best_evidence <- function(lbl) {
    lbl <- lbl[!is.na(lbl)]
    if (length(lbl) == 0) return(NA_character_)
    lbl[which.max(.evidence_weight_of(lbl))]
  }

  caas <- res %>% dplyr::filter(module == "caas") %>%
    dplyr::group_by(gene) %>%
    dplyr::summarise(caas_fcs_terms = paste(sort(unique(stringr::str_trunc(description, 50))), collapse = "; "),
                      evidence_label_caas = .best_evidence(evidence_label), .groups = "drop")

  rer <- res %>% dplyr::filter(module == "rer") %>%
    dplyr::group_by(gene) %>%
    dplyr::summarise(rer_fcs_terms = paste(sort(unique(stringr::str_trunc(description, 50))), collapse = "; "),
                      evidence_label_rer = .best_evidence(evidence_label), .groups = "drop")

  dplyr::full_join(caas, rer, by = "gene")
}

# Nested percentile quartet collapsed to its single deepest tier (1% subset
# of 5% subset of 10% subset of 25%), counts as ONE 0-4 score, not four
# separate booleans, so the same underlying signal can't be counted four
# times over in the grouped composite below.
.pct_tier_score <- function(p25, p10, p5, p1) dplyr::case_when(
  dplyr::coalesce(p1, FALSE)  ~ 4L, dplyr::coalesce(p5, FALSE)  ~ 3L,
  dplyr::coalesce(p10, FALSE) ~ 2L, dplyr::coalesce(p25, FALSE) ~ 1L, TRUE ~ 0L)

# Gene composite score: CAAS percentiles + directional Accumulation + CAAS FCS evidence tier +
# direction-matching RER + RER FCS evidence tier + directional FADE.
# Neither posenrich nor AMI contributes to the score.
.gene_composite <- function(df, has_accum, has_rer, has_fade, has_fcs_caas, has_fcs_rer) {
  df %>%
    dplyr::mutate(
      caas_family = .pct_tier_score(`25%`, `10%`, `5%`, `1%`) +
        (if (has_accum) as.integer(Accum_dir) else 0L) +
        (if (has_fcs_caas) weight_caas else 0L),
      rer_family  = if (has_rer) as.integer(RER_dir) + (if (has_fcs_rer) weight_rer else 0L) else 0L,
      fade_family = if (has_fade) as.integer(FADE_dir) else 0L,
      `Composite score` = caas_family + rer_family + fade_family,
      `N criteria` = as.integer(`25%`) + as.integer(`10%`) + as.integer(`5%`) + as.integer(`1%`) +
        (if (has_accum) as.integer(Accum_dir) else 0L) +
        (if (has_fcs_caas) as.integer(CAAS_FCS_sig) else 0L) +
        (if (has_rer) as.integer(RER_dir) + (if (has_fcs_rer) as.integer(RER_FCS_sig) else 0L) else 0L) +
        (if (has_fade) as.integer(FADE_dir) else 0L)
    )
}

# Position composite score: strictly position-native (CAAS percentiles + FADE site overlap).
# Zero gene-level weights (RER, Accumulation, FCS, AMI). Posenrich is purely descriptive.
.position_composite <- function(df, has_fade) {
  df %>%
    dplyr::mutate(
      caas_family = .pct_tier_score(`25%`, `10%`, `5%`, `1%`),
      fade_family = if (has_fade) as.integer(FADE_overlap) else 0L,
      `Composite score` = caas_family + fade_family,
      `N criteria` = as.integer(`25%`) + as.integer(`10%`) + as.integer(`5%`) + as.integer(`1%`) +
        (if (has_fade) as.integer(FADE_overlap) else 0L)
    )
}

# Banded table renderer: renders interactive tables styled into four functional color bands:
# Metadatos (Charcoal #4A5568), Pondera (Blue #2B6CB0), Descriptivas (Amber #C05621), Resultados (Green #2F855A).
render_banded_table <- function(res_obj, height = NULL) {
  if (is.null(res_obj) || !is.list(res_obj) || is.null(res_obj$data) || nrow(res_obj$data) == 0) return(NULL)
  df <- res_obj$data
  meta_cols <- intersect(res_obj$meta_cols, names(df))
  pond_cols <- intersect(res_obj$pond_cols, names(df))
  desc_cols <- intersect(res_obj$desc_cols, names(df))
  res_cols  <- intersect(res_obj$res_cols,  names(df))

  ordered_cols <- c(meta_cols, pond_cols, desc_cols, res_cols)
  df <- df[, ordered_cols, drop = FALSE]

  header_spec <- c()
  header_colors <- c()
  col_tints <- list()
  borders <- c()
  curr_idx <- 0

  if (length(meta_cols) > 0) {
    header_spec <- c(header_spec, "Metadatos" = length(meta_cols))
    header_colors <- c(header_colors, "#4A5568")
    col_tints[[length(col_tints) + 1]] <- list(idx = (curr_idx + 1):(curr_idx + length(meta_cols)), bg = "#F8F9FA")
    curr_idx <- curr_idx + length(meta_cols)
    borders <- c(borders, curr_idx)
  }
  if (length(pond_cols) > 0) {
    header_spec <- c(header_spec, "Pondera" = length(pond_cols))
    header_colors <- c(header_colors, "#2B6CB0")
    col_tints[[length(col_tints) + 1]] <- list(idx = (curr_idx + 1):(curr_idx + length(pond_cols)), bg = "#EFF6FF")
    curr_idx <- curr_idx + length(pond_cols)
    borders <- c(borders, curr_idx)
  }
  if (length(desc_cols) > 0) {
    header_spec <- c(header_spec, "Descriptivas" = length(desc_cols))
    header_colors <- c(header_colors, "#C05621")
    col_tints[[length(col_tints) + 1]] <- list(idx = (curr_idx + 1):(curr_idx + length(desc_cols)), bg = "#FFFBEB")
    curr_idx <- curr_idx + length(desc_cols)
    borders <- c(borders, curr_idx)
  }
  if (length(res_cols) > 0) {
    header_spec <- c(header_spec, "Resultados" = length(res_cols))
    header_colors <- c(header_colors, "#2F855A")
    col_tints[[length(col_tints) + 1]] <- list(idx = (curr_idx + 1):(curr_idx + length(res_cols)), bg = "#F0FDF4")
    curr_idx <- curr_idx + length(res_cols)
  }
  borders <- borders[borders < ncol(df)]

  k <- kable(df, format = "html", escape = FALSE) %>%
    kable_styling(bootstrap_options = c("condensed", "hover"), full_width = FALSE)

  if (length(header_spec) > 0) {
    k <- k %>% add_header_above(header_spec, color = "white", background = header_colors, bold = TRUE)
  }
  for (t in col_tints) {
    k <- k %>% column_spec(t$idx, background = t$bg)
  }
  if (length(borders) > 0) {
    k <- k %>% column_spec(borders, border_right = "2px solid #CBD5E0")
  }
  if (!is.null(height)) {
    k <- k %>% scroll_box(width = "100%", height = height)
  } else {
    k <- k %>% scroll_box(width = "100%")
  }
  k
}

.ami_desc_for <- function(dir_filter) {
  if (!has_ami || !exists("ami_disp") || is.null(ami_disp) || nrow(ami_disp) == 0) {
    return(tibble::tibble(gene = character(), caas_ami_desc = character(), rer_ami_desc = character(), fade_ami_desc = character()))
  }
  .ami_gene_col <- if ("all_genes" %in% names(ami_disp)) "all_genes" else if ("genes" %in% names(ami_disp)) "genes" else NULL
  if (is.null(.ami_gene_col) || !"term" %in% names(ami_disp)) {
    return(tibble::tibble(gene = character(), caas_ami_desc = character(), rer_ami_desc = character(), fade_ami_desc = character()))
  }
  
  d <- ami_disp
  if ("direction" %in% names(d) && !is.null(dir_filter)) {
    d <- d %>% dplyr::filter(direction %in% dir_filter)
  }
  if ("domino_p_adj" %in% names(d)) {
    d <- d %>% dplyr::filter(!is.na(domino_p_adj), domino_p_adj < (params$domino_module_thr %||% 0.05))
  }
  if (nrow(d) == 0) {
    return(tibble::tibble(gene = character(), caas_ami_desc = character(), rer_ami_desc = character(), fade_ami_desc = character()))
  }
  
  d_unnest <- d %>%
    dplyr::mutate(
      tool = if (".tool" %in% names(.)) .tool else .ami_tool_classifier(gene_list),
      gene = strsplit(.data[[.ami_gene_col]], ",\\s*"),
      hit_label = sprintf("%s (%s: %s)", term, direction, gene_list)
    ) %>%
    tidyr::unnest(gene) %>%
    dplyr::mutate(gene = trimws(gene))
  
  caas <- d_unnest %>% dplyr::filter(tool == "CAAS") %>%
    dplyr::group_by(gene) %>%
    dplyr::summarise(caas_ami_desc = paste(unique(hit_label), collapse = "; "), .groups = "drop")
  
  rer <- d_unnest %>% dplyr::filter(tool == "RER") %>%
    dplyr::group_by(gene) %>%
    dplyr::summarise(rer_ami_desc = paste(unique(hit_label), collapse = "; "), .groups = "drop")
  
  fade <- d_unnest %>% dplyr::filter(tool == "FADE") %>%
    dplyr::group_by(gene) %>%
    dplyr::summarise(fade_ami_desc = paste(unique(hit_label), collapse = "; "), .groups = "drop")
  
  caas %>%
    dplyr::full_join(rer, by = "gene") %>%
    dplyr::full_join(fade, by = "gene")
}

.posenrich_for <- function(rk_set) {
  if (!has_posenrich_le) return(tibble::tibble(gene = character(), pe_position = character(), pe_terms = character()))
  pe <- posenrich_le_df %>%
    dplyr::filter(is.null(rk_set) | ranking %in% rk_set) %>%
    dplyr::mutate(Position = sub(".*:", "", gene_position))
  if (has_posenrich_dotplot) {
    pe <- pe %>%
      dplyr::left_join(
        posenrich_dotplot_df %>% dplyr::mutate(description = dplyr::coalesce(description, pathway)) %>%
          dplyr::distinct(ranking, database, pathway, description),
        by = c("ranking", "database", "pathway")
      ) %>%
      dplyr::mutate(description = dplyr::coalesce(description, pathway))
  } else {
    pe$description <- pe$pathway
  }
  pe %>%
    dplyr::group_by(gene) %>%
    dplyr::summarise(pe_position = paste(sort(unique(Position)), collapse = ", "),
                      pe_terms    = paste(sort(unique(stringr::str_trunc(description, 50))), collapse = "; "),
                      .groups = "drop")
}

.pos_slice_positions <- function(dir_key, pct) {
  if (is.null(params$position_lists_dir)) return(NULL)
  f <- file.path(params$position_lists_dir, sprintf("slice_%s%d.tsv", dir_key, pct))
  if (!file.exists(f)) return(NULL)
  d <- tryCatch(readr::read_tsv(f, show_col_types = FALSE), error = function(e) NULL)
  if (is.null(d) || !all(c("Gene", "Position") %in% names(d))) return(NULL)
  unique(paste(d$Gene, d$Position))
}
.position_pct_flags <- function(base, dir_key) {
  pub25 <- .pos_slice_positions(dir_key, 25)
  pub10 <- .pos_slice_positions(dir_key, 10)
  pub5  <- .pos_slice_positions(dir_key, 5)
  pub1  <- .pos_slice_positions(dir_key, 1)
  if (!is.null(pub25) && !is.null(pub10) && !is.null(pub5) && !is.null(pub1)) {
    ids <- paste(base$Gene, base$Position)
    base %>%
      dplyr::mutate(gene = Gene,
                    `25%` = ids %in% pub25, `10%` = ids %in% pub10,
                    `5%` = ids %in% pub5, `1%` = ids %in% pub1)
  } else {
    thr <- quantile(base$CAAS_score, c(0.75, 0.90, 0.95, 0.99), na.rm = TRUE)
    base %>%
      dplyr::mutate(gene = Gene,
                    `25%` = CAAS_score >= thr[1], `10%` = CAAS_score >= thr[2],
                    `5%` = CAAS_score >= thr[3], `1%` = CAAS_score >= thr[4])
  }
}

.gcol <- function(df, col) if (col %in% names(df)) df[[col]] else rep(FALSE, nrow(df))
```

```{r interesting_genes_prep, eval=has_fcs_stats_cmp}
build_interesting_genes <- function(direction) {
  rk_set   <- switch(direction, Top = c("top", "accelerating"), Bottom = c("bottom", "decelerating"), NULL)
  ami_dirf <- switch(direction, Top = "Top", Bottom = "Bottom", NULL)

  base <- gene_attr_df
  if (direction == "Top") {
    base <- base %>% dplyr::transmute(gene, `25%` = pct_top25, `10%` = pct_top10, `5%` = pct_top5, `1%` = pct_top1,
                                       FADE_dir = .gcol(gene_attr_df, "flag_fade_top"),
                                       Accum_dir = .gcol(gene_attr_df, "flag_accum_top"))
  } else if (direction == "Bottom") {
    base <- base %>% dplyr::transmute(gene, `25%` = pct_bottom25, `10%` = pct_bottom10, `5%` = pct_bottom5, `1%` = pct_bottom1,
                                       FADE_dir = .gcol(gene_attr_df, "flag_fade_bottom"),
                                       Accum_dir = .gcol(gene_attr_df, "flag_accum_bottom"))
  } else {
    base <- base %>% dplyr::transmute(
      gene,
      `25%` = dplyr::coalesce(pct_top25, FALSE) | dplyr::coalesce(pct_bottom25, FALSE),
      `10%` = dplyr::coalesce(pct_top10, FALSE) | dplyr::coalesce(pct_bottom10, FALSE),
      `5%`  = dplyr::coalesce(pct_top5,  FALSE) | dplyr::coalesce(pct_bottom5,  FALSE),
      `1%`  = dplyr::coalesce(pct_top1,  FALSE) | dplyr::coalesce(pct_bottom1,  FALSE),
      FADE_dir  = .gcol(gene_attr_df, "flag_fade"),
      Accum_dir = .gcol(gene_attr_df, "flag_accum"))
  }
  base$RER_acc  <- .gcol(gene_attr_df, "flag_rer_acc")
  base$RER_decc <- .gcol(gene_attr_df, "flag_rer_decc")
  base$RER_dir <- switch(direction, Top = base$RER_acc, Bottom = base$RER_decc, base$RER_acc | base$RER_decc)

  # Check active modules for genes
  has_accum_gene <- any(grepl("^flag_accum", names(gene_attr_df))) &&
                    any(.gcol(gene_attr_df, "flag_accum") | .gcol(gene_attr_df, "flag_accum_top") | .gcol(gene_attr_df, "flag_accum_bottom"), na.rm = TRUE)
  has_rer_gene   <- any(grepl("^flag_rer", names(gene_attr_df))) ||
                    (has_multi && any(fcs_dir_mod$module == "rer"))
  has_fade_gene  <- any(grepl("^flag_fade", names(gene_attr_df))) &&
                    any(.gcol(gene_attr_df, "flag_fade") | .gcol(gene_attr_df, "flag_fade_top") | .gcol(gene_attr_df, "flag_fade_bottom"), na.rm = TRUE)
  has_fcs_caas   <- has_le && has_multi && any(fcs_dir_mod$module == "caas")
  has_fcs_rer    <- has_le && has_multi && any(fcs_dir_mod$module == "rer")
  has_ami_gene   <- has_ami && !is.null(ami_disp) && nrow(ami_disp) > 0
  has_pe_gene    <- has_posenrich_le && !is.null(posenrich_le_df) && nrow(posenrich_le_df) > 0

  all_genes <- base
  if (has_fcs_caas || has_fcs_rer) {
    all_genes <- all_genes %>% dplyr::left_join(.fcs_terms_for(rk_set), by = "gene")
  } else {
    all_genes$caas_fcs_terms <- NA_character_
    all_genes$rer_fcs_terms  <- NA_character_
    all_genes$evidence_label_caas <- "None"
    all_genes$evidence_label_rer  <- "None"
  }
  if (has_ami_gene) {
    all_genes <- all_genes %>% dplyr::left_join(.ami_desc_for(ami_dirf), by = "gene")
  } else {
    all_genes$caas_ami_desc <- NA_character_
    all_genes$rer_ami_desc  <- NA_character_
    all_genes$fade_ami_desc <- NA_character_
  }
  if (has_pe_gene) {
    all_genes <- all_genes %>% dplyr::left_join(.posenrich_for(rk_set), by = "gene")
  } else {
    all_genes$pe_position <- NA_character_
    all_genes$pe_terms    <- NA_character_
  }

  all_genes <- all_genes %>%
    dplyr::mutate(
      dplyr::across(c(`25%`, `10%`, `5%`, `1%`, FADE_dir, Accum_dir, RER_acc, RER_decc, RER_dir), ~ dplyr::coalesce(.x, FALSE)),
      evidence_label_caas = dplyr::coalesce(evidence_label_caas, "None"),
      evidence_label_rer  = dplyr::coalesce(evidence_label_rer, "None"),
      weight_caas  = .evidence_weight_of(evidence_label_caas),
      weight_rer   = .evidence_weight_of(evidence_label_rer),
      CAAS_FCS_sig = !is.na(caas_fcs_terms) & nzchar(caas_fcs_terms),
      RER_FCS_sig  = !is.na(rer_fcs_terms)  & nzchar(rer_fcs_terms),
      CAAS_AMI_net = !is.na(caas_ami_desc)  & nzchar(caas_ami_desc),
      RER_AMI_net  = !is.na(rer_ami_desc)   & nzchar(rer_ami_desc),
      FADE_AMI_net = !is.na(fade_ami_desc)  & nzchar(fade_ami_desc),
      PE_sig       = !is.na(pe_terms)       & nzchar(pe_terms)
    )

  # Grouped composite score: strictly CAAS + RER + FADE. Neither posenrich nor AMI weights!
  out <- all_genes %>%
    .gene_composite(has_accum = has_accum_gene, has_rer = has_rer_gene, has_fade = has_fade_gene,
                    has_fcs_caas = has_fcs_caas, has_fcs_rer = has_fcs_rer)

  # Filter to genes with any active signal
  out <- out %>%
    dplyr::filter(
      `25%` | `10%` | `5%` | `1%` |
      (if (has_fade_gene) FADE_dir else FALSE) |
      (if (has_accum_gene) Accum_dir else FALSE) |
      (if (has_rer_gene) RER_dir else FALSE) |
      (if (has_fcs_caas) CAAS_FCS_sig else FALSE) |
      (if (has_fcs_rer) RER_FCS_sig else FALSE) |
      (if (has_ami_gene) (CAAS_AMI_net | RER_AMI_net | FADE_AMI_net) else FALSE) |
      (if (has_pe_gene) PE_sig else FALSE)
    )
  if (nrow(out) == 0) return(NULL)

  out <- out %>%
    dplyr::mutate(
      `CAAS FCS term(s)`    = .collapsible_list(caas_fcs_terms),
      `RER FCS term(s)`     = .collapsible_list(rer_fcs_terms),
      `CAAS AMI desc(s)`    = .collapsible_list(caas_ami_desc),
      `RER AMI desc(s)`     = .collapsible_list(rer_ami_desc),
      `FADE AMI desc(s)`    = .collapsible_list(fade_ami_desc),
      `PE term(s)`          = .collapsible_list(pe_terms),
      `CAAS evidence`       = .evidence_badge(evidence_label_caas),
      `RER evidence`        = .evidence_badge(evidence_label_rer)
    )

  # Define the 4 tramos dynamically based on what actually ran
  meta_cols <- c("Gene")

  pond_cols <- c("25%", "10%", "5%", "1%")
  if (has_fcs_caas)   pond_cols <- c(pond_cols, "CAAS evidence")
  if (has_accum_gene) pond_cols <- c(pond_cols, "Accum_dir")
  if (has_rer_gene)   pond_cols <- c(pond_cols, "RER_dir")
  if (has_fcs_rer)    pond_cols <- c(pond_cols, "RER evidence")
  if (has_fade_gene)  pond_cols <- c(pond_cols, "FADE_dir")

  desc_cols <- c()
  if (has_fcs_caas) desc_cols <- c(desc_cols, "CAAS FCS term(s)")
  if (has_fcs_rer)  desc_cols <- c(desc_cols, "RER FCS term(s)")
  if (has_ami_gene) {
    desc_cols <- c(desc_cols, "CAAS AMI desc(s)")
    if (has_rer_gene)  desc_cols <- c(desc_cols, "RER AMI desc(s)")
    if (has_fade_gene) desc_cols <- c(desc_cols, "FADE AMI desc(s)")
  }
  if (has_pe_gene) {
    desc_cols <- c(desc_cols, "PE position", "PE term(s)")
  }

  res_cols <- c("caas_family")
  if (has_rer_gene)  res_cols <- c(res_cols, "rer_family")
  if (has_fade_gene) res_cols <- c(res_cols, "fade_family")
  res_cols <- c(res_cols, "N criteria", "Composite score")

  out <- out %>%
    dplyr::rename(Gene = gene, `PE position` = pe_position) %>%
    dplyr::arrange(dplyr::desc(`Composite score`)) %>%
    dplyr::select(dplyr::all_of(c(meta_cols, pond_cols, desc_cols, res_cols))) %>%
    head(params$top_n * 3)

  list(
    data      = out,
    meta_cols = meta_cols,
    pond_cols = pond_cols,
    desc_cols = desc_cols,
    res_cols  = res_cols
  )
}
```

```{r interesting_genes_tabs, eval=has_fcs_stats_cmp, results='asis'}
for (dr in c("Global", "Top", "Bottom")) {
  cat(sprintf("\n## %s\n\n", dr))
  res <- build_interesting_genes(dr)
  if (is.null(res) || is.null(res$data) || nrow(res$data) == 0) {
    cat("_No genes meet any criterion for this direction._\n\n")
    next
  }
  print(render_banded_table(res, height = if (nrow(res$data) > 15) "480px" else NULL))
}
```

`r if (!has_fcs_stats_cmp) "> **fcs_stats_file not supplied - Interesting Genes table skipped.**"`

---

# Interesting Positions {.tabset .tabset-fade}

Prioritized amino-acid position candidates evaluated strictly at **position-native resolution**, faceted by direction (**Global**, **Top**, **Bottom**). Gene-level criteria (RER, Accumulation, FCS, AMI) are omitted to prevent gene-wide signals from confounding position ranking. Columns are organized into four functional bands:

- **Metadatos** (grey): Position coordinates (`Gene`, `Position`), ancestral and derived states (`CAAS`, lineage residues, conserved pairs), supporting hypotheses, and continuous `CAAS Score`.
- **Pondera** (blue): Position-native criteria contributing to `Composite score`:
  - *CAAS percentiles* (`25%`, `10%`, `5%`, `1%`): collapsed to deepest nested tier (0–4).
  - *FADE site overlap* (`FADE_overlap`): exact codon site identified by FADE (0–1).
- **Descriptivas** (amber): Characterization-only annotations excluded from scoring:
  - *Biochemical profile*: `GS_schemes` (detection across Grantham Score groups GS1–GS4) and `is_biochem` (hit in ≥1 GS scheme).
  - *PosEnrich*: position-level functional enrichment terms (`PE_sig`, `PE term(s)`). Does not weight.
  - *Empirical statistics*: permutation p-values (`p.emp`, `p.adj_bh`).
  - *External annotations*: `COSMIC` somatic hits, `PAI3D_pathogenic` PrimateAI-3D pathogenicity calls.
  - *Domain & selection*: `Pfam domain`/`Pfam clan`, `UCR region`, `Position variability`, and `FUBAR selection`.
- **Resultados** (green): Family subtotal scores (`caas_family`, `fade_family`), `N criteria`, and `Composite score` (sum: 0–5).

Only columns corresponding to analyses that were executed are displayed.

```{r interesting_positions_prep, eval=has_position_scores_cmp}
build_interesting_positions <- function(direction) {
  rk_set    <- switch(direction, Top = c("top", "accelerating"), Bottom = c("bottom", "decelerating"), NULL)
  side_vals <- switch(direction, Top = "top", Bottom = "bottom", NULL)

  base <- position_scores_cmp_df
  if (direction == "Top") {
    base <- base %>% dplyr::filter(side == "top")
  } else if (direction == "Bottom") {
    base <- base %>% dplyr::filter(side == "bottom")
  } else {
    base <- base %>%
      dplyr::arrange(
        dplyr::desc(CAAS_score),
        dplyr::desc(if ("n_hypotheses" %in% names(.)) n_hypotheses else 0)
      ) %>%
      dplyr::distinct(Gene, Position, .keep_all = TRUE)
  }
  if (nrow(base) == 0) return(NULL)

  # Harmonize hypothesis column names from position_scores.tsv / upstream discovery
  if (!"participating_hypotheses" %in% names(base) && "supporting_hypotheses" %in% names(base)) {
    base$participating_hypotheses <- base$supporting_hypotheses
  }
  if (!"supporting_hypotheses" %in% names(base) && "participating_hypotheses" %in% names(base)) {
    base$supporting_hypotheses <- base$participating_hypotheses
  }
  if (!"n_hypotheses" %in% names(base) && "supporting_hypotheses" %in% names(base)) {
    base$n_hypotheses <- vapply(strsplit(as.character(base$supporting_hypotheses), "[,;]\\s*"), function(x) length(x[nzchar(x)]), integer(1))
  }

  base <- .position_pct_flags(base, tolower(direction))
  base <- base %>% dplyr::distinct(gene, Position, .keep_all = TRUE)

  # Position-native FADE-site overlap: check if FADE site analysis was run
  has_fade_pos <- (has_fade_sites_top_cmp && !is.null(fade_sites_top_cmp_df) && nrow(fade_sites_top_cmp_df) > 0) ||
                  (has_fade_sites_bottom_cmp && !is.null(fade_sites_bottom_cmp_df) && nrow(fade_sites_bottom_cmp_df) > 0)
  fade_df <- switch(direction,
    Top = fade_sites_top_cmp_df, Bottom = fade_sites_bottom_cmp_df,
    dplyr::bind_rows(fade_sites_top_cmp_df, fade_sites_bottom_cmp_df))
  base$FADE_overlap <- if (has_fade_pos && !is.null(fade_df) && nrow(fade_df) > 0) {
    paste(base$gene, base$Position) %in% paste(fade_df$gene, fade_df$Position)
  } else FALSE

  # Position-native posenrich: check if posenrich was run
  has_pe_pos <- FALSE
  if (has_posenrich_le && !is.null(posenrich_le_df) && nrow(posenrich_le_df) > 0) {
    pe <- posenrich_le_df %>%
      dplyr::filter(is.null(rk_set) | ranking %in% rk_set) %>%
      dplyr::mutate(gene = sub(":.*", "", gene_position), Position = suppressWarnings(as.integer(sub(".*:", "", gene_position)))) %>%
      dplyr::group_by(gene, Position) %>%
      dplyr::summarise(pe_terms = paste(sort(unique(pathway)), collapse = "; "), .groups = "drop")
    base <- base %>% dplyr::left_join(pe, by = c("gene", "Position"))
    base$PE_sig <- !is.na(base$pe_terms) & nzchar(base$pe_terms)
    has_pe_pos <- any(base$PE_sig, na.rm = TRUE)
  } else {
    base$pe_terms <- NA_character_
    base$PE_sig   <- FALSE
  }

  # Biochemical profile flags (GS_schemes and is_biochem)
  if (!"GS_schemes" %in% names(base)) {
    if ("scheme_set" %in% names(base)) {
      base$GS_schemes <- gsub("\\+?US\\+?", "", base$scheme_set)
      base$GS_schemes <- gsub("^\\+|\\+$", "", base$GS_schemes)
      base$is_biochem <- grepl("GS", base$GS_schemes)
    } else {
      base$GS_schemes <- NA_character_
      base$is_biochem <- FALSE
    }
  }
  base$is_biochem <- dplyr::coalesce(as.logical(base$is_biochem), FALSE)
  base$GS_schemes <- dplyr::coalesce(as.character(base$GS_schemes), "")

  # External characterization: PrimateAI-3D
  has_pai3d_pos <- FALSE
  if (has_vep_primateai_cmp && !is.null(vep_primateai_cmp_df) && nrow(vep_primateai_cmp_df) > 0) {
    .pai <- vep_primateai_cmp_df
    if ("caap_group" %in% names(.pai)) .pai <- dplyr::filter(.pai, caap_group == "US")
    .pred_col <- intersect(c("prediction"), names(.pai))[1]
    if (!is.na(.pred_col)) {
      .pai_pos <- .pai %>%
        dplyr::mutate(.patho = grepl("pathogenic", .data[[.pred_col]], ignore.case = TRUE)) %>%
        dplyr::group_by(Gene, Position) %>% dplyr::summarise(.patho = any(.patho), .groups = "drop")
      .m <- .pai_pos[match(paste(base$gene, base$Position), paste(.pai_pos$Gene, .pai_pos$Position)), ]
      base$PAI3D_pathogenic <- dplyr::case_when(is.na(.m$.patho) ~ "missing", .m$.patho ~ "pathogenic", TRUE ~ "benign")
      has_pai3d_pos <- any(base$PAI3D_pathogenic != "missing")
    }
  }
  if (!has_pai3d_pos) base$PAI3D_pathogenic <- rep("missing", nrow(base))

  # External characterization: COSMIC
  has_cosmic_pos <- FALSE
  if (has_vep_cosmic_cmp && !is.null(vep_cosmic_cmp_df) && nrow(vep_cosmic_cmp_df) > 0) {
    .cosmic <- if ("caap_group" %in% names(vep_cosmic_cmp_df)) dplyr::filter(vep_cosmic_cmp_df, caap_group == "US") else vep_cosmic_cmp_df
    .cosmic_hit <- paste(base$gene, base$Position) %in% paste(.cosmic$Gene, .cosmic$Position)
    .checkable <- if (has_pai3d_pos) base$PAI3D_pathogenic != "missing" else rep(TRUE, nrow(base))
    base$COSMIC <- dplyr::case_when(.cosmic_hit ~ "hit", .checkable ~ "no_hit", TRUE ~ "missing")
    has_cosmic_pos <- any(base$COSMIC != "missing")
  }
  if (!has_cosmic_pos) base$COSMIC <- rep("missing", nrow(base))

  # Position characterization GMTs: Pfam, UCR, variability, FUBAR
  has_pfam_pos  <- FALSE
  has_ucr_pos   <- FALSE
  has_var_pos   <- FALSE
  has_fubar_pos <- FALSE
  if (has_position_char_cmp && !is.null(position_char_cmp_df) && nrow(position_char_cmp_df) > 0) {
    base <- base %>%
      dplyr::left_join(
        position_char_cmp_df %>%
          dplyr::distinct(Gene, Position, .keep_all = TRUE) %>%
          dplyr::select(gene = Gene, Position, dplyr::any_of(c(
            "pfam_domain", "pfam_clan", "ucr_region", "position_variability", "fubar_selection"))),
        by = c("gene", "Position"))
  }
  for (.col in c("pfam_domain", "pfam_clan", "ucr_region", "position_variability", "fubar_selection")) {
    if (!.col %in% names(base)) base[[.col]] <- NA
  }
  has_pfam_pos  <- any(!is.na(base$pfam_domain) & nzchar(as.character(base$pfam_domain))) ||
                   any(!is.na(base$pfam_clan)   & nzchar(as.character(base$pfam_clan)))
  has_ucr_pos   <- any(!is.na(base$ucr_region)  & nzchar(as.character(base$ucr_region)))
  has_var_pos   <- any(!is.na(base$position_variability) & nzchar(as.character(base$position_variability)))
  has_fubar_pos <- any(!is.na(base$fubar_selection) & nzchar(as.character(base$fubar_selection)))

  # Empirical p-values
  has_pemp <- "p.emp" %in% names(base) && any(!is.na(base$p.emp))
  has_padj <- "p.adj_bh" %in% names(base) && any(!is.na(base$p.adj_bh))

  # Hypotheses
  has_hypotheses <- ("supporting_hypotheses" %in% names(base)) && any(!is.na(base$supporting_hypotheses) & nzchar(as.character(base$supporting_hypotheses)))

  # Format residues
  has_top_res <- "top_species_residues" %in% names(base) && any(!is.na(base$top_species_residues) & nzchar(as.character(base$top_species_residues)))
  has_bot_res <- "bottom_species_residues" %in% names(base) && any(!is.na(base$bottom_species_residues) & nzchar(as.character(base$bottom_species_residues)))
  has_pairs   <- "n_conserved_pairs" %in% names(base) && any(!is.na(base$n_conserved_pairs) & nzchar(as.character(base$n_conserved_pairs)))

  all_positions <- base %>%
    dplyr::mutate(
      dplyr::across(c(`25%`, `10%`, `5%`, `1%`, FADE_overlap, PE_sig), ~ dplyr::coalesce(.x, FALSE))
    )

  # Position composite: strictly CAAS + FADE site overlap. Zero gene-level weights!
  out <- all_positions %>%
    .position_composite(has_fade = has_fade_pos)

  # Filter to positions meeting at least one active criterion
  out <- out %>%
    dplyr::filter(
      `Composite score` > 0 | `25%` | is_biochem | (if (has_pe_pos) PE_sig else FALSE) |
      (if (has_cosmic_pos) COSMIC == "hit" else FALSE) |
      (if (has_pai3d_pos) PAI3D_pathogenic == "pathogenic" else FALSE)
    )
  if (nrow(out) == 0) return(NULL)

  out <- out %>%
    dplyr::mutate(
      `N Hypotheses`        = if ("n_hypotheses" %in% names(.)) as.integer(n_hypotheses) else NA_integer_,
      `Hypotheses`          = if ("supporting_hypotheses" %in% names(.)) .collapsible_hypotheses(supporting_hypotheses) else "",
      `PE term(s)`          = .collapsible_list(pe_terms),
      `CAAS`                = if ("caas" %in% names(.)) caas else NA_character_,
      `Top species residues`    = {
        .ts <- if ("top_species_residues" %in% names(.)) top_species_residues else NA_character_
        .cs <- if ("caas" %in% names(.)) caas else NA_character_
        dplyr::if_else(is.na(.ts) | !nzchar(.ts) | .ts == "NA",
                       dplyr::if_else(grepl("/", .cs), sub("/.*", "", .cs), .cs),
                       .ts)
      },
      `Conserved pairs` = if ("n_conserved_pairs" %in% names(.)) dplyr::na_if(as.character(n_conserved_pairs), "") else NA_character_,
      `Bottom species residues` = {
        .bs <- if ("bottom_species_residues" %in% names(.)) bottom_species_residues else NA_character_
        .cs <- if ("caas" %in% names(.)) caas else NA_character_
        dplyr::if_else(is.na(.bs) | !nzchar(.bs) | .bs == "NA",
                       dplyr::if_else(grepl("/", .cs), sub(".*/", "", .cs), NA_character_),
                       .bs)
      }
    )

  # Define the 4 tramos dynamically
  meta_cols <- c("Gene", "Position", "CAAS")
  if (has_top_res) meta_cols <- c(meta_cols, "Top species residues")
  if (has_bot_res) meta_cols <- c(meta_cols, "Bottom species residues")
  if (has_pairs)   meta_cols <- c(meta_cols, "Conserved pairs")
  if (has_hypotheses) meta_cols <- c(meta_cols, "N Hypotheses", "Hypotheses")
  meta_cols <- c(meta_cols, "CAAS Score")

  pond_cols <- c("25%", "10%", "5%", "1%")
  if (has_fade_pos) pond_cols <- c(pond_cols, "FADE_overlap")

  desc_cols <- c("GS_schemes", "is_biochem")
  if (has_pe_pos)    desc_cols <- c(desc_cols, "PE_sig", "PE term(s)")
  if (has_pemp)      desc_cols <- c(desc_cols, "p.emp")
  if (has_padj)      desc_cols <- c(desc_cols, "p.adj_bh")
  if (has_cosmic_pos) desc_cols <- c(desc_cols, "COSMIC")
  if (has_pai3d_pos)  desc_cols <- c(desc_cols, "PAI3D_pathogenic")
  if (has_pfam_pos) {
    if (any(!is.na(base$pfam_domain) & nzchar(as.character(base$pfam_domain)))) desc_cols <- c(desc_cols, "Pfam domain")
    if (any(!is.na(base$pfam_clan) & nzchar(as.character(base$pfam_clan)))) desc_cols <- c(desc_cols, "Pfam clan")
  }
  if (has_ucr_pos)   desc_cols <- c(desc_cols, "UCR region")
  if (has_var_pos)   desc_cols <- c(desc_cols, "Position variability")
  if (has_fubar_pos) desc_cols <- c(desc_cols, "FUBAR selection")

  res_cols <- c("caas_family")
  if (has_fade_pos) res_cols <- c(res_cols, "fade_family")
  res_cols <- c(res_cols, "N criteria", "Composite score")

  out$gene <- NULL
  if ("CAAS_score" %in% names(out)) out$`CAAS Score` <- out$CAAS_score
  if ("pfam_domain" %in% names(out)) out$`Pfam domain` <- out$pfam_domain
  if ("pfam_clan" %in% names(out)) out$`Pfam clan` <- out$pfam_clan
  if ("ucr_region" %in% names(out)) out$`UCR region` <- out$ucr_region
  if ("position_variability" %in% names(out)) out$`Position variability` <- out$position_variability
  if ("fubar_selection" %in% names(out)) out$`FUBAR selection` <- out$fubar_selection

  out <- out %>%
    dplyr::arrange(dplyr::desc(`Composite score`), dplyr::desc(`CAAS Score`)) %>%
    dplyr::select(dplyr::all_of(c(meta_cols, pond_cols, desc_cols, res_cols))) %>%
    head(params$top_n * 3)

  list(
    data      = out,
    meta_cols = meta_cols,
    pond_cols = pond_cols,
    desc_cols = desc_cols,
    res_cols  = res_cols
  )
}
```

```{r interesting_positions_tabs, eval=has_position_scores_cmp, results='asis'}
for (dr in c("Global", "Top", "Bottom")) {
  cat(sprintf("\n## %s\n\n", dr))
  res <- build_interesting_positions(dr)
  if (is.null(res) || is.null(res$data) || nrow(res$data) == 0) {
    cat("_No positions meet any criterion for this direction._\n\n")
    next
  }
  print(render_banded_table(res, height = if (nrow(res$data) > 15) "480px" else NULL))
}
```

`r if (!has_position_scores_cmp) "> **position_scores_file not supplied - Interesting Positions table skipped.**"`

---

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere - Enrichment Module*
*Date of generation: `r Sys.Date()`*
