---
title: "Comparative Enrichment Across Modules - `r params$traitname`"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
    mathjax: null                       # no math in this report
    md_extensions: "-tex_math_dollars"  # gene-set names contain '$' (e.g. V$HNF4_01) - never parse as TeX math
params:
  fcs_dir:    "cmp_fcs"       # dir of per-module <module>_fcs_all.tsv
  fcs_le_dir: "cmp_fcs_le"    # dir of per-module <module>_leading_edge.tsv
  fdr_thr:    0.15
  pperm_thr:  0.025
  domino_module_thr: 0.05  # DOMINO's own module-significance cutoff (domino.nf's domino_module_thr)
  top_n:      250
  traitname:  "trait"
  ami_module_desc_file:       NULL  # AMI's ami_networks/module_descriptions_all_lists.tsv (optional)
  ami_term_membership_file:   NULL  # AMI's ami_networks/term_threshold_membership.tsv (optional, recurrence table)
  posenrich_dotplot_file:     NULL  # posenrich's posenrich_overall_dotplot.tsv (optional)
  # posenrich_leading_edge.tsv (gene:position members of significant posenrich
  # terms, already position-granular), drives the Interesting Genes/Positions tables' posenrich angle
  # AND the Interesting Genes/Interesting Positions tables below.
  posenrich_leading_edge_file: NULL
  # posenrich's pooled-per-term leading-edge display table (one row per
  # significant term, not per cutoff), 14.Position_enrichment_report.Rmd's
  # own Leading edge section export, posenrich_leading_edge_summary.tsv.
  posenrich_leading_edge_summary_file: NULL
  # SCORING's own gene-level table (percentile flags + FADE/RER/Accumulation
  # significance), same file 12.FCS_general_report.Rmd/14.Position_enrichment_report.Rmd
  # consume, reused here for the Interesting Genes/Positions tables.
  fcs_stats_file: NULL
  # Position-level CAAS scores + FADE-site/VEP annotations, same optional
  # inputs 14.Position_enrichment_report.Rmd's Position Characterisation
  # section consumes, reused here for the Interesting Positions table.
  position_scores_file:  NULL
  vep_primateai_file:    NULL
  vep_cosmic_file:       NULL
  fade_sites_top_file:    NULL
  fade_sites_bottom_file: NULL
  # SCORING's published percentile slices (scoring_compute.R), when
  # supplied, gene/position 25/10/5/1% membership is read from these
  # directories instead of being re-derived independently in this report (see
  # fcs_percentile_flags()'s gene_lists_dir param / .position_pct_flags()).
  gene_lists_dir:     NULL
  position_lists_dir: NULL
  seed:               "1998"
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 7, dev = "png", dpi = 200)
suppressPackageStartupMessages({ library(igraph); library(tidyverse); library(knitr); library(kableExtra) })

theme_set(
  theme_minimal(base_size = 13) +
  theme(
    axis.title.y  = element_text(margin = ggplot2::margin(l = 20, r = 20)),
    axis.title.x  = element_text(margin = ggplot2::margin(t = 12)),
    axis.text.x   = element_text(margin = ggplot2::margin(t = 5)),
    plot.title    = element_text(face = "bold", hjust = 0, margin = ggplot2::margin(b = 6)),
    plot.subtitle = element_text(colour = "grey40", hjust = 0, margin = ggplot2::margin(b = 10)),
    legend.position = "right",
    legend.box.margin = ggplot2::margin(t = 5)
  )
)
filter <- dplyr::filter; select <- dplyr::select; rename <- dplyr::rename; mutate <- dplyr::mutate

seed_val <- if (exists("params") && !is.null(params$seed) && nzchar(as.character(params$seed))) as.integer(params$seed) else 1998
set.seed(seed_val)

.src_dir <- file.path(dirname(normalizePath(knitr::current_input(), mustWork = FALSE)), "src")
# fcs_percentile_flags() (used by the Interesting Genes/Positions tables) lives
# in percentile_flags.R, sourced directly rather than via fcs_enrich.R, which
# would drag in a hard RERconverge dependency this report never otherwise
# needs (fcs_enrich.R's fastwilcoxGMT/fcs_run_all machinery is
# report 12/RER-report-specific).
source(file.path(.src_dir, "percentile_flags.R"))

kbl_html <- function(df, height = NULL, escape = TRUE, caption = NULL, ...) {
  k <- kable(df, format = "html", escape = escape, caption = caption, ...) %>%
    kable_styling(bootstrap_options = c("striped", "condensed", "hover"), full_width = FALSE)
  # Wide tables (long term names/descriptions, many columns) always get a
  # horizontal scrollbar instead of overflowing the page; the vertical
  # scroll_box (height) only kicks in on top of that for tall tables.
  k <- scroll_box(k, width = "100%", height = height)
  k
}

# Evidence-tier weight/colour, same convention as 12.FCS_general_report.Rmd's
# Multi-Method Evidence table (green/orange/grey badges), reused everywhere
# an evidence_label appears in this report (Per-module significance tables,
# evidence_label_caas/evidence_label_rer in Interesting Genes/Positions), so
# the same tier always reads the same colour everywhere. Any table using
# .evidence_badge() must be rendered via kbl_html(..., escape = FALSE).
.evidence_weight_of <- function(lbl) dplyr::case_when(
  lbl == "Hard evidence" ~ 3L, lbl == "Supported" ~ 2L, lbl == "Exploratory" ~ 1L, TRUE ~ 0L)
.evidence_colours <- c("Hard evidence" = "#27ae60", "Supported" = "#e67e22",
                       "Exploratory" = "#7f8c8d", "None" = "#dcdcdc")
.evidence_badge <- function(x) {
  x <- dplyr::coalesce(x, "None")
  kableExtra::cell_spec(x, "html",
    color = ifelse(x == "None", "#888888", "white"),
    background = .evidence_colours[x], bold = x != "None",
    extra_css = "padding: 2px 6px; border-radius: 3px;")
}

# Fallback tool classifier from a gene-list filename, used when an AMI export
# predates the explicit `tool` column. Defined here (not next to its main use
# in ami_term_membership_tabs below) because cross_angle_gene_hits, an earlier
# and unconditionally-evaluated chunk, also needs it as a fallback.
.ami_tool_classifier <- function(gl) {
  dplyr::case_when(
    grepl("^rer_",  gl, ignore.case = TRUE) ~ "RER",
    grepl("^fade_", gl, ignore.case = TRUE) ~ "FADE",
    TRUE                                    ~ "CAAS"
  )
}
```

```{r load, include=FALSE}
`%||%` <- function(a, b) if (!is.null(a)) a else b
REQ <- c("ranking", "database", "pathway", "stat", "p.adj")

# Per-module FCS results: cmp_fcs/<module>_fcs_all.tsv (caas/rer).
mod_files <- list.files(params$fcs_dir, pattern = "_fcs_all\\.tsv$", full.names = TRUE)
fcs_by_mod <- list()
for (f in mod_files) {
  mod <- sub("_fcs_all\\.tsv$", "", basename(f))
  d <- tryCatch(readr::read_tsv(f, show_col_types = FALSE), error = function(e) NULL)
  if (!is.null(d) && nrow(d) > 0 && all(REQ %in% names(d))) {
    fcs_by_mod[[mod]] <- dplyr::mutate(d, module = mod)
  }
}
fcs_all_mod <- if (length(fcs_by_mod)) dplyr::bind_rows(fcs_by_mod) else tibble()
modules_present <- names(fcs_by_mod)
has_multi <- length(fcs_by_mod) >= 1

# Direction facet (Global/top/bottom), derived once here since every
# directional section below groups by it. Each module's `ranking` vocabulary
# differs: CAAS uses top/bottom/global directly; RER's accelerating/
# decelerating is the same two-way split framed around RER's own statistic
# (RER carries no separate undirected ranking, so it contributes no "global"
# rows). CAAS's "global" ranking is its OWN direction here, kept fully
# separate from top/bottom: a "global" row can never share a (module,
# direction, database, pathway) group key with a "top" or "bottom" row, so
# permod_prep's grouping below never blends the real directional stat and
# the undirected global stat for the same pathway into one summarised row.
if (has_multi) {
  fcs_dir_mod <- fcs_all_mod %>%
    mutate(direction = dplyr::case_when(
      tolower(ranking) %in% c("top", "accelerating")    ~ "top",
      tolower(ranking) %in% c("bottom", "decelerating") ~ "bottom",
      tolower(ranking) == "global"                      ~ "global",
      TRUE                                              ~ NA_character_
    )) %>%
    dplyr::filter(!is.na(direction))
} else {
  fcs_dir_mod <- fcs_all_mod
}

# Per-module leading-edge tables (cmp_fcs_le/<module>_leading_edge.tsv) feed
# the Interesting Genes/Positions criteria below: each module's own
# directional percentile concentration + cross-module corroboration flags
# (FADE/RER/Accumulation, and PrimateAI-3D/COSMIC once wired in from
# posenrich). Percentile columns reflect CAAS's own ranking in every module's
# leading edge (per-module FCS
# reports are annotated with CAAS's fcs_stats.tsv as their overlay) - that's
# intentional, not a bug: it's the shared reference axis every module's
# leading edge is cross-checked against.
le_files <- list.files(params$fcs_le_dir, pattern = "_leading_edge\\.tsv$", full.names = TRUE)
le_by_mod <- list()
for (f in le_files) {
  mod <- sub("_leading_edge\\.tsv$", "", basename(f))
  d <- tryCatch(readr::read_tsv(f, show_col_types = FALSE), error = function(e) NULL)
  if (!is.null(d) && nrow(d) > 0) le_by_mod[[mod]] <- dplyr::mutate(d, module = mod)
}
le <- if (length(le_by_mod)) dplyr::bind_rows(le_by_mod) else tibble()
has_le <- nrow(le) > 0

# ── Cross-angle inputs: AMI (network modules) + posenrich (position-level) ──
# Neither is gene-set-ranked the way CAAS/RER FCS is, so they don't fold into
# fcs_all_mod/le above, each gets its own section below, and all three feed
# the Interesting Genes/Interesting Positions tables via the genes they involve.
.file_ok <- function(p) !is.null(p) && !is.na(p) && nchar(p) > 0 && file.exists(p)

has_ami <- .file_ok(params$ami_module_desc_file)
ami_df  <- if (has_ami) readr::read_tsv(params$ami_module_desc_file, show_col_types = FALSE) else NULL
ami_disp <- NULL

# AMI term membership recurrence table (term_threshold_membership.tsv from
# 13.AMI_analysis.Rmd's render_tool_section_nested()) - imported directly
# rather than re-derived, so the table shown here matches the AMI report exactly.
has_ami_term_membership <- .file_ok(params$ami_term_membership_file)
ami_term_membership_df  <- if (has_ami_term_membership) readr::read_tsv(params$ami_term_membership_file, show_col_types = FALSE) else NULL

# Per-module leading-edge composition tables (fcs_leading_edge_composition.tsv
# from 12.FCS_general_report.Rmd) - imported directly so the table shown here
# matches the FCS reports exactly.
le_comp_files <- list.files(params$fcs_le_dir, pattern = "_leading_edge_composition\\.tsv$", full.names = TRUE)
le_comp_by_mod <- list()
for (f in le_comp_files) {
  mod <- sub("_leading_edge_composition\\.tsv$", "", basename(f))
  d <- tryCatch(readr::read_tsv(f, show_col_types = FALSE), error = function(e) NULL)
  if (!is.null(d) && nrow(d) > 0) le_comp_by_mod[[mod]] <- dplyr::mutate(d, module = mod)
}
le_comp_all_mod <- if (length(le_comp_by_mod)) dplyr::bind_rows(le_comp_by_mod) else tibble()
has_le_comp <- nrow(le_comp_all_mod) > 0

has_posenrich_dotplot <- .file_ok(params$posenrich_dotplot_file)
posenrich_dotplot_df  <- if (has_posenrich_dotplot) readr::read_tsv(params$posenrich_dotplot_file, show_col_types = FALSE) else NULL

# posenrich_leading_edge.tsv: gene:position members of significant posenrich
# terms (already position-granular, ranking, database, pathway, top_frac,
# gene, gene_position). Drives the Interesting Genes/Interesting Positions tables below.
has_posenrich_le <- .file_ok(params$posenrich_leading_edge_file)
posenrich_le_df  <- if (has_posenrich_le) readr::read_tsv(params$posenrich_leading_edge_file, show_col_types = FALSE) else NULL

has_posenrich_le_summary <- .file_ok(params$posenrich_leading_edge_summary_file)
posenrich_le_summary_df  <- if (has_posenrich_le_summary) readr::read_tsv(params$posenrich_leading_edge_summary_file, show_col_types = FALSE) else NULL

# ── Gene/position-level inputs for the Interesting Genes/Positions tables ──
# Same optional files 12.FCS_general_report.Rmd (fcs_stats) and
# 14.Position_enrichment_report.Rmd (position_scores, VEP, FADE sites) already
# consume, reused here rather than re-derived.
has_fcs_stats_cmp <- .file_ok(params$fcs_stats_file)
fcs_stats_cmp_df  <- if (has_fcs_stats_cmp) readr::read_tsv(params$fcs_stats_file, show_col_types = FALSE) else NULL

has_position_scores_cmp <- .file_ok(params$position_scores_file)
position_scores_cmp_df  <- if (has_position_scores_cmp) readr::read_tsv(params$position_scores_file, show_col_types = FALSE) else NULL

has_vep_primateai_cmp <- .file_ok(params$vep_primateai_file)
vep_primateai_cmp_df  <- if (has_vep_primateai_cmp) readr::read_tsv(params$vep_primateai_file, show_col_types = FALSE) else NULL

has_vep_cosmic_cmp <- .file_ok(params$vep_cosmic_file)
vep_cosmic_cmp_df  <- if (has_vep_cosmic_cmp) readr::read_tsv(params$vep_cosmic_file, show_col_types = FALSE) else NULL

has_fade_sites_top_cmp    <- .file_ok(params$fade_sites_top_file)
has_fade_sites_bottom_cmp <- .file_ok(params$fade_sites_bottom_file)
fade_sites_top_cmp_df <- if (has_fade_sites_top_cmp) {
  readr::read_csv(params$fade_sites_top_file, show_col_types = FALSE) %>%
    dplyr::rename(gene = gene, Position = position) %>%
    dplyr::mutate(Position = suppressWarnings(as.integer(Position)))
} else NULL
fade_sites_bottom_cmp_df <- if (has_fade_sites_bottom_cmp) {
  readr::read_csv(params$fade_sites_bottom_file, show_col_types = FALSE) %>%
    dplyr::rename(gene = gene, Position = position) %>%
    dplyr::mutate(Position = suppressWarnings(as.integer(Position)))
} else NULL

# Per-gene attribute table: percentile flags (from fcs_stats' own score_top/
# score_bottom, same derivation as 12.FCS_general_report.Rmd) + the gene-level
# flag_* columns fcs_stats.tsv already carries (gate/FADE/RER/Accumulation,
# direction-aware: flag_accum_top/flag_accum_bottom are the direction-restricted
# tests, not the non-directional flag_accum).
gene_attr_df <- if (has_fcs_stats_cmp) {
  pf <- fcs_percentile_flags(fcs_stats_cmp_df, gene_lists_dir = params$gene_lists_dir)
  flag_cols_cmp <- grep("^flag_", names(fcs_stats_cmp_df), value = TRUE)
  dplyr::left_join(pf, fcs_stats_cmp_df[, c("gene", flag_cols_cmp)], by = "gene")
} else NULL
```

```{r cross_angle_gene_hits, include=FALSE}
# ── Cross-angle gene-level hit sets ───────────────────────────────────────
# Computed once here so the per-module significance section, the AMI module
# description section, and the Interesting Genes/Positions tables below all
# read the same per-gene FCS/posenrich corroboration, one shared definition,
# not three drifting copies.
#
# Each angle uses ITS OWN module's significance definition (not "any
# leading-edge/described membership"), FCS/RER's dual p.adj+p.perm gate,
# posenrich's own p_adj gate, DOMINO's own module-significance cutoff.
# FCS and posenrich both carry a genuine "global" ranking (CAAS's undirected
# composite; posenrich's own directions = ["global","top","bottom"]), so
# fcs_genes/posenrich_genes have real direction=="Global" rows. AMI has no
# equivalent, DOMINO only ever runs on Top/Bottom percentile gene lists
# (.TOP_THRESHOLDS/.BOTTOM_THRESHOLDS in 13.AMI_analysis.Rmd, no undirected
# list), so ami_genes' direction column is only ever "Top"/"Bottom"; a
# "Global" row for AMI never exists here. This means the Interesting Genes/Positions tables' Global
# tab can never carry an AMI contribution, unlike .ami_desc_for(dir_filter = NULL), used by the
# Interesting Genes/Positions AMI cells, which pools Top+Bottom explicitly
# for its own "Global" view instead of relying on a literal direction match.

# FCS/RER own significance: BH < fdr_thr AND (p.perm < pperm_thr when
# present), same formula permod_prep below applies, computed once here off
# fcs_dir_mod so this chunk and permod_prep agree on what "significant" means.
# fcs_leading_edge.tsv (`le`) is NOT itself restricted to significant terms
# (12.FCS_general_report.Rmd's leading-edge export only requires drv_score >
# 0, i.e. "this gene drives some signal in this term", not "this term
# cleared BH"), inner_join against fcs_sig_terms is what actually restricts
# fcs_genes to genes belonging to a term each module itself calls significant.
fcs_sig_terms <- if (has_multi) {
  fcs_dir_mod %>%
    dplyr::filter(stat > 0, !is.na(p.adj), p.adj < params$fdr_thr,
                  !("p.perm" %in% names(.)) | is.na(p.perm) | p.perm < (params$pperm_thr %||% 1)) %>%
    dplyr::mutate(description = dplyr::coalesce(description, pathway)) %>%
    dplyr::distinct(module, direction, database, pathway, description)
} else {
  tibble::tibble(module = character(), direction = character(), database = character(),
                 pathway = character(), description = character())
}

fcs_genes <- if (has_le && has_multi) {
  le %>%
    dplyr::mutate(direction = dplyr::case_when(
      ranking %in% c("top", "accelerating")    ~ "top",
      ranking %in% c("bottom", "decelerating") ~ "bottom",
      ranking == "global"                      ~ "global",
      TRUE                                      ~ NA_character_
    )) %>%
    dplyr::filter(!is.na(direction)) %>%
    dplyr::inner_join(fcs_sig_terms, by = c("module", "direction", "database", "pathway")) %>%
    dplyr::mutate(direction = tools::toTitleCase(direction)) %>%
    dplyr::distinct(gene, direction) %>%
    dplyr::mutate(fcs_hit = TRUE)
} else {
  tibble::tibble(gene = character(), direction = character(), fcs_hit = logical())
}

# AMI angle: DOMINO's own module-significance cutoff (domino_module_thr,
# domino.nf's own default 0.05) gates which described modules count --
# "described" alone (not Too small/No enrichment) was never a significance
# claim, just "STRING could summarise it". Expands each significant module's
# FULL membership into one row per gene: `all_genes` (every member gene,
# uncapped, describe_clusters() in 13.AMI_analysis.Rmd) when present, else
# the older capped `genes` column (top-8 representatives) for
# module_descriptions_all_lists.tsv files written before all_genes existed.
# direction is only ever "Top"/"Bottom" here (DOMINO never runs on an
# undirected gene list, so there's no "Global" value to filter for, see
# this chunk's opening comment).
.normalize_term <- function(x) {
  x <- tolower(x)
  x <- gsub("[^a-z0-9 ]", " ", x)
  gsub("\\s+", " ", trimws(x))
}
.cluster_terms <- function(terms, sim_thr = 0.85) {
  if (length(terms) == 0) return(character(0))
  norm <- vapply(terms, .normalize_term, character(1))
  reps_norm <- character(0); reps_orig <- character(0)
  assign <- integer(length(terms))
  for (i in seq_along(terms)) {
    if (length(reps_norm) == 0) { reps_norm <- norm[i]; reps_orig <- terms[i]; assign[i] <- 1L; next }
    d <- utils::adist(norm[i], reps_norm)[1, ]
    sim <- 1 - d / pmax(nchar(norm[i]), nchar(reps_norm), 1)
    best <- which.max(sim)
    if (sim[best] >= sim_thr) {
      assign[i] <- best
    } else {
      reps_norm <- c(reps_norm, norm[i]); reps_orig <- c(reps_orig, terms[i])
      assign[i] <- length(reps_norm)
    }
  }
  reps_orig[assign]
}

ami_disp <- if (has_ami && !is.null(ami_df) && nrow(ami_df) > 0) {
  d <- ami_df %>%
    dplyr::mutate(
      .tool = if ("tool" %in% names(.)) as.character(tool) else .ami_tool_classifier(gene_list),
      direction = if ("direction" %in% names(.)) dplyr::case_when(
        direction %in% c("Acc", "accelerating", "Accelerating") ~ "Top",
        direction %in% c("Decc", "decelerating", "Decelerating") ~ "Bottom",
        TRUE ~ as.character(direction)
      ) else NA_character_
    )
  if ("description" %in% names(d)) {
    d <- d %>% dplyr::filter(!description %in% c("Too small", "No enrichment"))
  }
  if ("domino_p_adj" %in% names(d)) {
    d <- d %>% dplyr::filter(!is.na(domino_p_adj), domino_p_adj < (params$domino_module_thr %||% 0.05))
  }
  if (nrow(d) > 0 && "description" %in% names(d)) {
    d$term <- .cluster_terms(d$description)
  } else if (nrow(d) > 0) {
    d$term <- character(nrow(d))
  } else {
    d$term <- character(0)
  }
  d
} else NULL

ami_genes <- if (has_ami && !is.null(ami_df) && nrow(ami_df) > 0) {
  .ami_gene_col <- if ("all_genes" %in% names(ami_df)) "all_genes" else if ("genes" %in% names(ami_df)) "genes" else NULL
  if (is.null(.ami_gene_col)) {
    tibble::tibble(gene = character(), direction = character(), ami_hit = logical())
  } else {
    d <- ami_df %>%
      dplyr::mutate(
        .tool = if ("tool" %in% names(.)) as.character(tool) else .ami_tool_classifier(gene_list),
        direction = if ("direction" %in% names(.)) dplyr::case_when(
          direction %in% c("Acc", "accelerating", "Accelerating") ~ "Top",
          direction %in% c("Decc", "decelerating", "Decelerating") ~ "Bottom",
          TRUE ~ as.character(direction)
        ) else NA_character_
      )
    if ("description" %in% names(d)) {
      d <- d %>% dplyr::filter(!description %in% c("Too small", "No enrichment"))
    }
    if ("direction" %in% names(d)) {
      d <- d %>% dplyr::filter(direction %in% c("Top", "Bottom"))
    }
    if ("domino_p_adj" %in% names(d)) {
      d <- d %>% dplyr::filter(!is.na(domino_p_adj), domino_p_adj < (params$domino_module_thr %||% 0.05))
    }
    if (nrow(d) > 0) {
      d %>%
        dplyr::mutate(gene = strsplit(.data[[.ami_gene_col]], ",\\s*")) %>%
        tidyr::unnest(gene) %>%
        dplyr::mutate(gene = trimws(gene)) %>%
        dplyr::distinct(gene, direction) %>%
        dplyr::mutate(ami_hit = TRUE)
    } else {
      tibble::tibble(gene = character(), direction = character(), ami_hit = logical())
    }
  }
} else {
  tibble::tibble(gene = character(), direction = character(), ami_hit = logical())
}

# Posenrich angle: posenrich_leading_edge.tsv is ALREADY restricted to
# significant (p_adj < posenrich's own padj_thr) terms by posenrich_enrich.py
# itself (only rows with r["sig"] get written), no additional gate needed
# here, just direction mapping (now including "global"). Also carries the
# actual Position(s) behind each hit (kept for the Interesting Genes/Positions tables' Position
# column below, lets a reader see whether the specific position posenrich
# flagged for a gene is the same one driving that gene's FCS/AMI signal, not
# just that the gene appears in both).
posenrich_genes <- if (has_posenrich_le) {
  posenrich_le_df %>%
    dplyr::mutate(direction = dplyr::case_when(
      ranking %in% c("top", "accelerating")    ~ "top",
      ranking %in% c("bottom", "decelerating") ~ "bottom",
      ranking == "global"                      ~ "global",
      TRUE                                      ~ NA_character_
    ),
    Position = sub(".*:", "", gene_position)) %>%
    dplyr::filter(!is.na(direction)) %>%
    dplyr::mutate(direction = tools::toTitleCase(direction)) %>%
    dplyr::group_by(gene, direction) %>%
    dplyr::summarise(Position = paste(sort(unique(Position)), collapse = ", "), .groups = "drop") %>%
    dplyr::mutate(posenrich_hit = TRUE)
} else {
  tibble::tibble(gene = character(), direction = character(), Position = character(), posenrich_hit = logical())
}

```

# Overview

```{r overview, echo=FALSE}
mod_rows <- if (has_multi) sapply(fcs_by_mod, nrow) else integer(0)
tibble::tibble(
  Source = paste0("FCS module: ", names(mod_rows)),
  Rows   = unname(mod_rows)
) %>% kbl_html()
```

This report compares enrichment **across CAAS and RERconverge** (the two modules
with well-behaved FCS rankings and real permulation nulls) for `r params$traitname`.
FADE and Accumulation do not run their own FCS ranking: FADE's statistic is a
max over many sites and Accumulation has no permulation null, so neither
supports a reliable genome-wide gene-set enrichment test on its own. Both
remain available as **cross-module corroboration**, surfaced as flag columns
(`%fade_*`/`%rer_*`/`%accum*`) in the imported leading-edge composition tables
below and as weighted criteria in the Interesting Genes/Positions tables further
down. A gene set is "significant" at BH < `r params$fdr_thr` (and permulation p
< threshold where available) with AUC > 0.5 - see
[12.FCS_general_report.Rmd](12.FCS_general_report.Rmd)'s Methods section for the
full three-method definition (Wilcoxon-AUC / Lachenbruch / Permulation) and the
`evidence_label`/`evidence_score` ranking convention this report's own tables follow.

DOMINO active-module (AMI) results and posenrich's position-level enrichment are
neither gene-set-ranked the same way as FCS, so they get their own sections below.

# Per-module significance {.tabset .tabset-fade}

Leading-edge composition from each module's own FCS report
([12.FCS_general_report.Rmd](12.FCS_general_report.Rmd)), imported directly so the
tables shown here match the per-module reports exactly. For each significantly-
enriched set, the **percentage of its leading-edge genes** carrying each attribute:
significance gates, directional score percentiles, and cross-module support.
Rows follow the same ranking convention as the source FCS report: **evidence
tier first** (Hard > Supported > Exploratory evidence, from the Wilcoxon-AUC /
Lachenbruch / Permulation vote), **then `evidence_score`** - each *passing*
method's own statistic (AUC, `L.frac_mag`, NES) converted to a percentile
rank so the three are comparable on one scale, zeroed for a method that
didn't pass, then summed - **then alphabetical** as a tie-break, already
baked into the row order of the imported table, not recomputed here. Faceted
Global/Top/Bottom, each module in its own sub-tab, capped at `r params$top_n`
rows.

`r if (!has_le_comp && !has_multi) "> **No module FCS results available - skipping.**"`

```{r permod_prep, eval=has_multi}
dir.create("compare_results", showWarnings = FALSE)
```

```{r permod_le_comp_tabs, eval=has_le_comp, results='asis'}
# Direction mapping for the ranking column in the imported tables.
.le_comp_dir_map <- c(top = "Top", accelerating = "Top", bottom = "Bottom",
                       decelerating = "Bottom", global = "Global")

le_comp_all_mod <- le_comp_all_mod %>%
  dplyr::mutate(direction = dplyr::coalesce(.le_comp_dir_map[tolower(ranking)], "Global"))

for (dr in c("Global", "Top", "Bottom")) {
  cat(sprintf("\n## %s {.tabset}\n\n", dr))

  mods_to_show <- union(modules_present, unique(le_comp_all_mod$module))
  for (m in mods_to_show) {
    cat(sprintf("\n### %s\n\n", toupper(m)))
    d <- le_comp_all_mod %>%
      dplyr::filter(module == m, direction == dr) %>%
      dplyr::select(-module, -direction) %>%
      head(params$top_n)
    if (nrow(d) == 0) {
      cat(sprintf("_No leading-edge composition for %s (%s)._\n\n", toupper(m), dr))
    } else {
      if ("evidence_label" %in% names(d)) d$evidence_label <- .evidence_badge(d$evidence_label)
      print(kbl_html(d, height = if (nrow(d) > 15) "480px" else NULL, escape = FALSE))
    }
  }
}
```

```{r permod_tabs_fallback, eval=has_multi && !has_le_comp, results='asis'}
# Fallback: recompute per-module significance when imported leading-edge
# composition tables are not available (e.g. older pipeline runs). Same
# ranking convention as 12.FCS_general_report.Rmd: evidence tier, then
# evidence_score, then alphabetical, recomputed here from the raw
# per-module fcs_all.tsv columns (percentile-ranked within each module x
# direction tab, the population these tables actually compare within)
# rather than reusing the source report's own evidence_rank/evidence_score
# directly, since older exports may predate them.
thr <- params$fdr_thr
.evidence_rank_of <- function(lbl) dplyr::case_when(
  lbl == "Hard evidence" ~ 1L, lbl == "Supported" ~ 2L, lbl == "Exploratory" ~ 3L, TRUE ~ 4L)

if (!"evidence_label" %in% names(fcs_dir_mod)) fcs_dir_mod$evidence_label <- NA_character_
for (cl in c("lach_frac_magnitude", "perm_nes")) {
  if (!cl %in% names(fcs_dir_mod)) fcs_dir_mod[[cl]] <- NA_real_
}
for (cl in c("sig_wilcoxon", "sig_lachenbruch", "sig_permulation")) {
  if (!cl %in% names(fcs_dir_mod)) fcs_dir_mod[[cl]] <- FALSE
}

permod_best <- fcs_dir_mod %>%
  filter(stat > 0) %>%
  mutate(evidence_label = dplyr::coalesce(evidence_label, "Exploratory")) %>%
  group_by(module, direction, database, pathway) %>%
  summarise(AUC = max(stat, na.rm = TRUE) + 0.5,
            p.adj = min(p.adj, na.rm = TRUE),
            lach_frac_magnitude = suppressWarnings({ v <- na.omit(lach_frac_magnitude); if (length(v)) v[1] else NA_real_ }),
            perm_nes = suppressWarnings({ v <- na.omit(perm_nes); if (length(v)) v[1] else NA_real_ }),
            sig_wilcoxon    = any(sig_wilcoxon %in% TRUE),
            sig_lachenbruch = any(sig_lachenbruch %in% TRUE),
            sig_permulation = any(sig_permulation %in% TRUE),
            evidence_label = dplyr::first(evidence_label),
            description = dplyr::first(dplyr::coalesce(description, pathway)),
            .groups = "drop") %>%
  # evidence_score: each PASSING method's statistic converted to its own
  # percentile rank within this (module, direction) tab, zeroed when that
  # method didn't pass, see 12.FCS_general_report.Rmd's Methods section
  # for why (raw AUC/L.frac_mag/NES are on incompatible scales).
  dplyr::group_by(module, direction) %>%
  dplyr::mutate(.auc_pct = dplyr::percent_rank(AUC),
                .lfm_pct = dplyr::percent_rank(lach_frac_magnitude),
                .nes_pct = dplyr::percent_rank(perm_nes)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    evidence_rank = .evidence_rank_of(evidence_label),
    evidence_score = dplyr::if_else(sig_wilcoxon,    dplyr::coalesce(.auc_pct, 0), 0) +
                     dplyr::if_else(sig_lachenbruch, dplyr::coalesce(.lfm_pct, 0), 0) +
                     dplyr::if_else(sig_permulation, dplyr::coalesce(.nes_pct, 0), 0)
  ) %>%
  dplyr::select(-.auc_pct, -.lfm_pct, -.nes_pct) %>%
  dplyr::left_join(fcs_sig_terms %>% dplyr::select(-description) %>% dplyr::mutate(sig = TRUE),
                    by = c("module", "direction", "database", "pathway")) %>%
  dplyr::mutate(sig = dplyr::coalesce(sig, FALSE))

for (dr in c("global", "top", "bottom")) {
  cat(sprintf("\n## %s {.tabset}\n\n", tools::toTitleCase(dr)))
  for (m in modules_present) {
    cat(sprintf("\n### %s\n\n", toupper(m)))
    d <- permod_best %>%
      filter(module == m, direction == dr, sig) %>%
      arrange(evidence_rank, dplyr::desc(evidence_score), description) %>%
      mutate(Description = stringr::str_trunc(description, 60),
             AUC = round(AUC, 3), p.adj = signif(p.adj, 3),
             L.frac_mag = round(lach_frac_magnitude, 2), NES = round(perm_nes, 2)) %>%
      select(database, pathway, Description, evidence_label, AUC, p.adj, any_of(c("L.frac_mag", "NES"))) %>%
      head(params$top_n)
    if (nrow(d) == 0) {
      cat(sprintf("_No gene set is significant in %s (%s) at BH < %.2g._\n\n", toupper(m), dr, thr))
    } else {
      d$evidence_label <- .evidence_badge(d$evidence_label)
      print(kbl_html(d, height = if (nrow(d) > 15) "420px" else NULL, escape = FALSE))
    }
  }
}
```

---

# AMI module descriptions {.tabset .tabset-fade}

`r if (!has_ami && !has_ami_term_membership) "> **No AMI module descriptions available (AMI/DOMINO didn't run, or the file wasn't wired in) - skipping.**"`

DOMINO active modules ([13.AMI_analysis.Rmd](13.AMI_analysis.Rmd)) aren't gene-set-ranked
like CAAS/RER FCS, so they don't fold into the FCS tables above. The recurrence table
below shows each described term and which gene lists it was recovered under - imported
directly from the AMI report so the table shown here matches that report exactly.

```{r ami_term_membership_tabs, eval=has_ami_term_membership, results='asis'}
# Imported term membership recurrence table from AMI report.
# .ami_tool_classifier is defined in the setup chunk above (needed earlier by
# cross_angle_gene_hits too).
.ami_tm_direction <- function(gl) {
  dplyr::case_when(
    grepl("global|significant$", gl, ignore.case = TRUE) & !grepl("top|bottom|accelerating|decelerating", gl, ignore.case = TRUE) ~ "Global",
    grepl("top|accelerating", gl, ignore.case = TRUE) ~ "Top",
    grepl("bottom|decelerating", gl, ignore.case = TRUE) ~ "Bottom",
    TRUE ~ "Global"
  )
}

ami_tm <- ami_term_membership_df %>%
  dplyr::mutate(
    .tool      = if ("tool" %in% names(.)) as.character(tool) else .ami_tool_classifier(gene_list),
    .direction = dplyr::case_when(
      "direction" %in% names(.) & direction %in% c("Acc", "accelerating", "Accelerating") ~ "Top",
      "direction" %in% names(.) & direction %in% c("Decc", "decelerating", "Decelerating") ~ "Bottom",
      "direction" %in% names(.) & direction %in% c("Top", "Bottom", "Global") ~ as.character(direction),
      TRUE ~ .ami_tm_direction(gene_list)
    )
  )

for (tl in c("CAAS", "RER", "FADE")) {
  cat(sprintf("\n## %s {.tabset}\n\n", tl))
  tl_df <- ami_tm %>% dplyr::filter(.tool == tl)
  if (nrow(tl_df) == 0) {
    cat(sprintf("_No AMI active modules for %s._\n\n", tl))
    next
  }

  for (dr in c("Global", "Top", "Bottom")) {
    cat(sprintf("\n### %s\n\n", dr))
    sub <- if (dr == "Global") tl_df else tl_df %>% dplyr::filter(.direction == dr)
    if (nrow(sub) == 0) {
      cat(sprintf("_No described terms for %s (%s)._\n\n", tl, dr))
      next
    }

    disp <- sub %>%
      dplyr::distinct(term, gene_list) %>%
      dplyr::group_by(term) %>%
      dplyr::summarise(`N lists` = dplyr::n(),
                        `Gene lists` = paste(sort(unique(gene_list)), collapse = ", "),
                        .groups = "drop") %>%
      dplyr::arrange(dplyr::desc(`N lists`), term) %>%
      dplyr::rename(Term = term) %>%
      head(params$top_n)
    print(kbl_html(disp, height = if (nrow(disp) > 15) "420px" else NULL))
  }
}
```

```{r ami_desc_fallback_prep, eval=has_ami && !has_ami_term_membership}
# Fallback: re-derive when the imported term membership file isn't available.
.build_ami_rollup <- function(df) {
  df %>%
    dplyr::group_by(term) %>%
    dplyr::summarise(
      `N modules`    = dplyr::n(),
      `Gene lists`   = paste(sort(unique(gene_list)), collapse = ", "),
      `DOMINO p.adj` = { v <- if ("domino_p_adj" %in% names(.data)) domino_p_adj[!is.na(domino_p_adj)] else numeric(0); if (length(v)) signif(min(v), 3) else NA_real_ },
      Genes          = suppressWarnings(max(n_genes, na.rm = TRUE)),
      .groups = "drop"
    ) %>%
    dplyr::rename(Term = term) %>%
    dplyr::arrange(`DOMINO p.adj`)
}
```

```{r ami_desc_fallback_tabs, eval=has_ami && !has_ami_term_membership, results='asis'}
for (tl in c("CAAS", "RER", "FADE")) {
  cat(sprintf("\n## %s {.tabset}\n\n", tl))
  if (!exists("ami_disp") || is.null(ami_disp) || nrow(ami_disp) == 0) {
    cat(sprintf("_No described modules for %s._\n\n", tl))
    next
  }
  tl_disp <- if (".tool" %in% names(ami_disp)) ami_disp %>% dplyr::filter(.tool == tl) else ami_disp
  if (nrow(tl_disp) == 0) {
    cat(sprintf("_No described modules for %s._\n\n", tl))
    next
  }
  for (dr in c("Global", "Top", "Bottom")) {
    cat(sprintf("\n### %s\n\n", dr))
    sub <- if (dr == "Global") tl_disp else if ("direction" %in% names(tl_disp)) tl_disp %>% dplyr::filter(direction == dr) else tl_disp
    if (nrow(sub) == 0) { cat(sprintf("_No described modules for %s (%s)._\n\n", tl, dr)); next }
    disp <- .build_ami_rollup(sub) %>% head(params$top_n)
    print(kbl_html(disp, height = if (nrow(disp) > 15) "420px" else NULL))
  }
}
```

---

# Posenrich position-level enrichment {.tabset .tabset-fade}

<details>
<summary>Click to show/hide</summary>

`r if (!has_posenrich_le_summary && !has_posenrich_le) "> **No posenrich outputs available (--posenrich was off, or the files weren't wired in) - skipping.**"`

Position-level enrichment ([14.Position_enrichment_report.Rmd](14.Position_enrichment_report.Rmd))
operates on individual CAAS positions rather than genes, so it gets its own section
here rather than folding into the gene-set FCS tables above. Its significant terms
feed the Interesting Genes/Interesting Positions tables further down (individual
gene- and position-level corroboration, in full detail). Faceted by direction
below (`r params$top_n` rows each, not one shared budget split across both) so a
direction with fewer but stronger hits isn't crowded out of a single pooled
table by the other.

## Overall dotplot terms - merged summary {.tabset}

One row per significant term, **merged across name variants** that are
themselves `/`-joined synonyms in their own description (e.g. a term
described as `"V$HNF4_01 / HNF4A"` is treated as the same underlying term as
one independently named `"HNF4A"` elsewhere) - connected via shared segments,
not just identical full strings, so synonym spelling drift across databases
doesn't fragment one biological term into several rows. `Merged from` lists
every original (database, pathway) pair folded into the row, so the merge is
auditable rather than a silent black box.

`Cutoffs`/`N cutoffs` show which top-fraction cutoffs (10/5/1%) the term
recurred significant at *before* merging - a plain recurrence count, unioned
across merged aliases. `NES`/`p.adj` (best across merged aliases: highest
NES, lowest p.adj) come straight from
[14.Position_enrichment_report.Rmd](14.Position_enrichment_report.Rmd)'s own
Leading edge table (`posenrich_leading_edge_summary.tsv`, imported below);
this report runs one test (Position-Level Path Sum Permulation), so there's
no multi-method vote to combine here, unlike the gene-level FCS tables above.

```{r posenrich_dotplot_prep, eval=has_posenrich_le_summary}
# Sourced from the SAME imported table as the Leading edge sub-tab below
# (posenrich_leading_edge_summary.tsv), this section is a synonym-merged
# ROLLUP of that table, not an independent derivation, so the two never drift.
posenrich_dotplot_disp <- posenrich_le_summary_df %>%
  dplyr::rename(database = Database, pathway = Pathway, p_adj = `p.adj`) %>%
  dplyr::mutate(.direction = tools::toTitleCase(Ranking))

# Fuzzy '/'-synonym merge: two rows merge iff they share at least one trimmed
# '/'-segment of their own Description, within the same direction (never
# across direction, a top-direction term and a bottom-direction term merging
# would conflate two different biological claims). Connected-components over
# a segment-sharing graph, not exact-string grouping, so "HNF4A" and
# "V$HNF4_01 / HNF4A" merge even though neither string contains the other.
fuzzy_merge_terms <- function(df) {
  if (nrow(df) == 0) return(df %>% dplyr::mutate(.merge_id = integer(0)))
  df <- df %>% dplyr::mutate(.row_id = dplyr::row_number())
  segs <- df %>%
    dplyr::mutate(seg = strsplit(Description, "\\s*/\\s*")) %>%
    tidyr::unnest(seg) %>%
    dplyr::mutate(seg = trimws(tolower(seg))) %>%
    dplyr::filter(nzchar(seg))
  if (nrow(segs) == 0 || dplyr::n_distinct(segs$seg) == nrow(segs)) {
    # Nothing shares a segment with anything else, every row is its own component.
    return(df %>% dplyr::mutate(.merge_id = .row_id))
  }
  edges <- segs %>%
    dplyr::inner_join(segs, by = "seg", suffix = c("_a", "_b"), relationship = "many-to-many") %>%
    dplyr::filter(.row_id_a < .row_id_b) %>%
    dplyr::distinct(.row_id_a, .row_id_b)
  g <- igraph::graph_from_data_frame(edges, directed = FALSE, vertices = df$.row_id)
  comp <- igraph::components(g)
  df %>% dplyr::mutate(.merge_id = comp$membership[as.character(.row_id)])
}
```

```{r posenrich_dotplot_tabs, eval=has_posenrich_le_summary, results='asis'}
has_nes_col <- "NES" %in% names(posenrich_dotplot_disp)
for (dr in c("Global", "Top", "Bottom")) {
  cat(sprintf("\n### %s\n\n", dr))
  dr_df <- posenrich_dotplot_disp %>% dplyr::filter(.direction == dr)
  if (nrow(dr_df) == 0) { cat("_No terms for this direction._\n\n"); next }

  merged <- fuzzy_merge_terms(dr_df) %>%
    dplyr::group_by(.merge_id) %>%
    dplyr::summarise(
      Description  = dplyr::first(Description[order(p_adj)]),
      `Merged from` = paste(unique(sprintf("%s: %s", database, pathway)), collapse = "; "),
      Cutoffs       = paste(unique(unlist(strsplit(Cutoffs, ",\\s*"))), collapse = ", "),
      `N cutoffs`   = suppressWarnings(max(`N cutoffs`, na.rm = TRUE)),
      `N merged`    = dplyr::n(),
      NES           = if (has_nes_col) round(suppressWarnings(max(NES, na.rm = TRUE)), 3) else NA_real_,
      `p.adj`       = signif(suppressWarnings(min(p_adj, na.rm = TRUE)), 3),
      .groups = "drop"
    ) %>%
    # Same convention as 14.Position_enrichment_report.Rmd's "Term -> cutoff
    # membership" table: recurrence (N cutoffs) first, alphabetical tie-break.
    dplyr::arrange(dplyr::desc(`N cutoffs`), Description) %>%
    dplyr::select(-.merge_id) %>%
    { if (!has_nes_col) dplyr::select(., -NES) else . } %>%
    head(params$top_n)

  print(kbl_html(merged, height = if (nrow(merged) > 15) "460px" else NULL))
}
```

## Leading edge {.tabset .tabset-fade}

Imported directly from [14.Position_enrichment_report.Rmd](14.Position_enrichment_report.Rmd)'s
own Leading edge section (`posenrich_leading_edge_summary.tsv`) so the table
shown here matches that report exactly - one row per significant term
(pooled across the 10/5/1% cutoffs it was tested at), with the union of
observed `gene:position` members, cross-module corroboration columns, and
`N cutoffs`/`Cutoffs` as a plain recurrence audit trail. Faceted by
direction, capped at `r params$top_n` rows.

```{r posenrich_le_summary_tabs, eval=has_posenrich_le_summary, results='asis'}
for (dr in c("GLOBAL", "TOP", "BOTTOM")) {
  cat(sprintf("\n### %s\n\n", tools::toTitleCase(dr)))
  d <- posenrich_le_summary_df %>%
    dplyr::filter(toupper(Ranking) == dr) %>%
    dplyr::select(-Ranking) %>%
    head(params$top_n)
  if (nrow(d) == 0) {
    cat(sprintf("_No significant terms for %s._\n\n", tools::toTitleCase(dr)))
  } else {
    print(kbl_html(d, height = if (nrow(d) > 15) "480px" else NULL))
  }
}
```

`r if (!has_posenrich_le_summary) "> **posenrich_leading_edge_summary_file not supplied - Leading edge table skipped.**"`

</details>

---

# Interesting Genes {.tabset .tabset-fade}

One row per gene carrying at least one of the criteria below, faceted
Global/Top/Bottom. Each criterion is a plain boolean (or bool + supporting
list, e.g. which enriched term(s) a gene belongs to) - deliberately not
collapsed into a single opaque score, so every claim in this table is
independently checkable. `FCS term(s)`/`AMI description(s)`/`PE term(s)` show
each term's **description** (not its bare database id), restricted to terms
that module itself calls significant (FCS/RER's own BH+p.perm gate, DOMINO's
own module-significance cutoff - posenrich's leading edge is already
significance-restricted at the source), and collapse behind a `<details>`
toggle when a gene carries more than one. `CAAS evidence`/`RER evidence`
(`evidence_label_caas`/`evidence_label_rer`) show the **strongest** FCS
evidence tier the gene's best CAAS/RER gene set reached - Hard (weight 3) >
Supported (2) > Exploratory (1) > None (0), same tiers and colours as
[12.FCS_general_report.Rmd](12.FCS_general_report.Rmd)'s Multi-Method
Evidence table - a strictly more informative sibling of the plain
`CAAS_FCS_sig`/`RER_FCS_sig` booleans (which only say "belongs to ≥1
significant term", not how strong the best one is).

`Composite score` is a **plain sum of four independent evidence families**:
no background-rate weighting, no hypergeometric test, just addition: **CAAS**
(the 25/10/5/1% percentile quartet collapsed to its single deepest tier, 0–4,
since 1% ⊂ 5% ⊂ 10% ⊂ 25% is one nested signal, not four - plus Accumulation
significance, plus the CAAS evidence-tier weight above), **RER**
(direction-matching acceleration/deceleration, plus the RER evidence-tier
weight), **AMI + FADE** (DOMINO network membership for CAAS/RER/FADE gene
lists, plus direct FADE-site significance - grouped together because AMI's
own FADE-seeded gene list and FADE's direct call are two views of the same
underlying evidence, not independent), and **posenrich** (position-level
corroboration). The four family subtotals (`caas_family`/`rer_family`/
`ami_fade_family`/`posenrich_family`) are shown alongside the total so the
composite is never a black box. Criteria are always restricted to the
table's own direction - e.g. under **Top**, only `flag_fade_top`/
`flag_accum_top`/RER *acceleration* count; a Bottom-direction signal never
leaks into a Top-direction gene's score (and vice versa; **Global** has no
direction of its own, so either side counts there).

```{r interesting_shared_helpers}
# Shared by both Interesting Genes AND Interesting Positions below, defined
# unconditionally (not gated on has_fcs_stats_cmp/has_position_scores_cmp)
# because either table's own gate can be TRUE independently of the other's,
# and build_interesting_positions() below calls these same helpers. Gating
# them behind only one of the two flags would leave them undefined whenever
# that flag was FALSE but the other (needing them) was TRUE.

# Dependency-free HTML escaping (avoids requiring htmltools be attached) and
# a collapsible-<details> renderer for semicolon-joined multi-term cells --
# any kable() call including a column built by .collapsible_list() MUST be
# passed escape = FALSE, or the markup renders as literal text.
.html_escape <- function(x) {
  x <- gsub("&", "&amp;", x, fixed = TRUE)
  x <- gsub("<", "&lt;",  x, fixed = TRUE)
  gsub(">", "&gt;", x, fixed = TRUE)
}
.collapsible_list <- function(joined, sep = "; ") {
  vapply(joined, function(x) {
    if (is.na(x) || !nzchar(x)) return("")
    items <- strsplit(x, sep, fixed = TRUE)[[1]]
    summary_txt <- if (length(items) == 1) .html_escape(stringr::str_trunc(items[1], 40)) else sprintf("%d terms", length(items))
    sprintf("<details><summary>%s</summary>%s</details>",
            summary_txt,
            paste(sprintf("<div>%s</div>", .html_escape(items)), collapse = ""))
  }, character(1), USE.NAMES = FALSE)
}

# Per-gene best evidence tier per module (Hard=3/Supported=2/Exploratory=1/
# none=0), alongside the existing term-description text, evidence_label_caas/
# evidence_label_rer surface in Interesting Genes/Positions below as their own
# weighted criterion, distinct from the plain CAAS_FCS_sig/RER_FCS_sig boolean
# (which only says "belongs to >=1 significant term", not how strong the best
# one is).
.fcs_terms_for <- function(rk_set) {
  empty <- tibble::tibble(gene = character(), caas_fcs_terms = character(), rer_fcs_terms = character(),
                          evidence_label_caas = character(), evidence_label_rer = character())
  if (!has_le || !has_multi) return(empty)
  res <- le %>%
    dplyr::filter(is.null(rk_set) | ranking %in% rk_set) %>%
    dplyr::mutate(direction = dplyr::case_when(
      tolower(ranking) %in% c("top", "accelerating")    ~ "top",
      tolower(ranking) %in% c("bottom", "decelerating") ~ "bottom",
      tolower(ranking) == "global"                      ~ "global",
      TRUE                                              ~ NA_character_
    )) %>%
    dplyr::filter(!is.na(direction)) %>%
    dplyr::inner_join(fcs_sig_terms, by = c("module", "direction", "database", "pathway")) %>%
    dplyr::left_join(fcs_dir_mod %>% dplyr::distinct(module, direction, database, pathway, evidence_label),
                      by = c("module", "direction", "database", "pathway"))

  .best_evidence <- function(lbl) {
    lbl <- lbl[!is.na(lbl)]
    if (length(lbl) == 0) return(NA_character_)
    lbl[which.max(.evidence_weight_of(lbl))]
  }

  caas <- res %>% dplyr::filter(module == "caas") %>%
    dplyr::group_by(gene) %>%
    dplyr::summarise(caas_fcs_terms = paste(sort(unique(stringr::str_trunc(description, 50))), collapse = "; "),
                      evidence_label_caas = .best_evidence(evidence_label), .groups = "drop")

  rer <- res %>% dplyr::filter(module == "rer") %>%
    dplyr::group_by(gene) %>%
    dplyr::summarise(rer_fcs_terms = paste(sort(unique(stringr::str_trunc(description, 50))), collapse = "; "),
                      evidence_label_rer = .best_evidence(evidence_label), .groups = "drop")

  dplyr::full_join(caas, rer, by = "gene")
}

# Nested percentile quartet collapsed to its single deepest tier (1% subset
# of 5% subset of 10% subset of 25%), counts as ONE 0-4 score, not four
# separate booleans, so the same underlying signal can't be counted four
# times over in the grouped composite below.
.pct_tier_score <- function(p25, p10, p5, p1) dplyr::case_when(
  dplyr::coalesce(p1, FALSE)  ~ 4L, dplyr::coalesce(p5, FALSE)  ~ 3L,
  dplyr::coalesce(p10, FALSE) ~ 2L, dplyr::coalesce(p25, FALSE) ~ 1L, TRUE ~ 0L)

# Simple grouped-sum composite score: four INDEPENDENT evidence families,
# CAAS, RER, AMI+FADE (network/module + direct FADE-site evidence travel
# together, since AMI's own gene lists are themselves seeded from FADE
# significance), and posenrich, each contributing the plain count of
# criteria it satisfies (weighted 0-3 by evidence tier where a tier exists),
# summed across families with NO further weighting. `fade_col` is "FADE_dir"
# (genes) or "FADE_overlap" (positions), same shape, different column name
# at the two granularities.
.grouped_composite <- function(df, fade_col) {
  df %>%
    dplyr::mutate(
      caas_family      = .pct_tier_score(`25%`, `10%`, `5%`, `1%`) + as.integer(Accum_dir) + weight_caas,
      rer_family        = as.integer(RER_dir) + weight_rer,
      ami_fade_family    = as.integer(.data[[fade_col]]) + as.integer(CAAS_AMI_net) +
                            as.integer(RER_AMI_net) + as.integer(FADE_AMI_net),
      posenrich_family  = as.integer(PE_sig),
      `Composite score` = caas_family + rer_family + ami_fade_family + posenrich_family,
      `N criteria` = as.integer(`25%`) + as.integer(`10%`) + as.integer(`5%`) + as.integer(`1%`) +
        as.integer(Accum_dir) + as.integer(.data[[fade_col]]) + as.integer(RER_dir) +
        as.integer(CAAS_FCS_sig) + as.integer(RER_FCS_sig) +
        as.integer(CAAS_AMI_net) + as.integer(RER_AMI_net) + as.integer(FADE_AMI_net) + as.integer(PE_sig)
    )
}

.ami_desc_for <- function(dir_filter) {
  if (!has_ami || !exists("ami_disp") || is.null(ami_disp) || nrow(ami_disp) == 0) {
    return(tibble::tibble(gene = character(), caas_ami_desc = character(), rer_ami_desc = character(), fade_ami_desc = character()))
  }
  .ami_gene_col <- if ("all_genes" %in% names(ami_disp)) "all_genes" else if ("genes" %in% names(ami_disp)) "genes" else NULL
  if (is.null(.ami_gene_col) || !"term" %in% names(ami_disp)) {
    return(tibble::tibble(gene = character(), caas_ami_desc = character(), rer_ami_desc = character(), fade_ami_desc = character()))
  }
  
  d <- ami_disp
  if ("direction" %in% names(d) && !is.null(dir_filter)) {
    d <- d %>% dplyr::filter(direction %in% dir_filter)
  }
  if ("domino_p_adj" %in% names(d)) {
    d <- d %>% dplyr::filter(!is.na(domino_p_adj), domino_p_adj < (params$domino_module_thr %||% 0.05))
  }
  if (nrow(d) == 0) {
    return(tibble::tibble(gene = character(), caas_ami_desc = character(), rer_ami_desc = character(), fade_ami_desc = character()))
  }
  
  d_unnest <- d %>%
    dplyr::mutate(
      tool = if (".tool" %in% names(.)) .tool else .ami_tool_classifier(gene_list),
      gene = strsplit(.data[[.ami_gene_col]], ",\\s*"),
      hit_label = sprintf("%s (%s: %s)", term, direction, gene_list)
    ) %>%
    tidyr::unnest(gene) %>%
    dplyr::mutate(gene = trimws(gene))
  
  caas <- d_unnest %>% dplyr::filter(tool == "CAAS") %>%
    dplyr::group_by(gene) %>%
    dplyr::summarise(caas_ami_desc = paste(unique(hit_label), collapse = "; "), .groups = "drop")
  
  rer <- d_unnest %>% dplyr::filter(tool == "RER") %>%
    dplyr::group_by(gene) %>%
    dplyr::summarise(rer_ami_desc = paste(unique(hit_label), collapse = "; "), .groups = "drop")
  
  fade <- d_unnest %>% dplyr::filter(tool == "FADE") %>%
    dplyr::group_by(gene) %>%
    dplyr::summarise(fade_ami_desc = paste(unique(hit_label), collapse = "; "), .groups = "drop")
  
  caas %>%
    dplyr::full_join(rer, by = "gene") %>%
    dplyr::full_join(fade, by = "gene")
}

.posenrich_for <- function(rk_set) {
  if (!has_posenrich_le) return(tibble::tibble(gene = character(), pe_position = character(), pe_terms = character()))
  pe <- posenrich_le_df %>%
    dplyr::filter(is.null(rk_set) | ranking %in% rk_set) %>%
    dplyr::mutate(Position = sub(".*:", "", gene_position))
  if (has_posenrich_dotplot) {
    pe <- pe %>%
      dplyr::left_join(
        posenrich_dotplot_df %>% dplyr::mutate(description = dplyr::coalesce(description, pathway)) %>%
          dplyr::distinct(ranking, database, pathway, description),
        by = c("ranking", "database", "pathway")
      ) %>%
      dplyr::mutate(description = dplyr::coalesce(description, pathway))
  } else {
    pe$description <- pe$pathway
  }
  pe %>%
    dplyr::group_by(gene) %>%
    dplyr::summarise(pe_position = paste(sort(unique(Position)), collapse = ", "),
                      pe_terms    = paste(sort(unique(stringr::str_trunc(description, 50))), collapse = "; "),
                      .groups = "drop")
}

.pos_slice_positions <- function(dir_key, pct) {
  if (is.null(params$position_lists_dir)) return(NULL)
  f <- file.path(params$position_lists_dir, sprintf("slice_%s%d.tsv", dir_key, pct))
  if (!file.exists(f)) return(NULL)
  d <- tryCatch(readr::read_tsv(f, show_col_types = FALSE), error = function(e) NULL)
  if (is.null(d) || !all(c("Gene", "Position") %in% names(d))) return(NULL)
  unique(paste(d$Gene, d$Position))
}
.position_pct_flags <- function(base, dir_key) {
  pub25 <- .pos_slice_positions(dir_key, 25)
  pub10 <- .pos_slice_positions(dir_key, 10)
  pub5  <- .pos_slice_positions(dir_key, 5)
  pub1  <- .pos_slice_positions(dir_key, 1)
  if (!is.null(pub25) && !is.null(pub10) && !is.null(pub5) && !is.null(pub1)) {
    ids <- paste(base$Gene, base$Position)
    base %>%
      dplyr::transmute(gene = Gene, Position, CAAS_score,
                        `25%` = ids %in% pub25, `10%` = ids %in% pub10,
                        `5%` = ids %in% pub5, `1%` = ids %in% pub1)
  } else {
    thr <- quantile(base$CAAS_score, c(0.75, 0.90, 0.95, 0.99), na.rm = TRUE)
    base %>%
      dplyr::transmute(gene = Gene, Position, CAAS_score,
                        `25%` = CAAS_score >= thr[1], `10%` = CAAS_score >= thr[2],
                        `5%` = CAAS_score >= thr[3], `1%` = CAAS_score >= thr[4])
  }
}

.gcol <- function(df, col) if (col %in% names(df)) df[[col]] else rep(FALSE, nrow(df))
```

```{r interesting_genes_prep, eval=has_fcs_stats_cmp}
build_interesting_genes <- function(direction) {
  rk_set   <- switch(direction, Top = c("top", "accelerating"), Bottom = c("bottom", "decelerating"), NULL)
  ami_dirf <- switch(direction, Top = "Top", Bottom = "Bottom", NULL)

  base <- gene_attr_df
  if (direction == "Top") {
    base <- base %>% dplyr::transmute(gene, `25%` = pct_top25, `10%` = pct_top10, `5%` = pct_top5, `1%` = pct_top1,
                                       FADE_dir = .gcol(gene_attr_df, "flag_fade_top"),
                                       Accum_dir = .gcol(gene_attr_df, "flag_accum_top"))
  } else if (direction == "Bottom") {
    base <- base %>% dplyr::transmute(gene, `25%` = pct_bottom25, `10%` = pct_bottom10, `5%` = pct_bottom5, `1%` = pct_bottom1,
                                       FADE_dir = .gcol(gene_attr_df, "flag_fade_bottom"),
                                       Accum_dir = .gcol(gene_attr_df, "flag_accum_bottom"))
  } else {
    base <- base %>% dplyr::transmute(
      gene,
      `25%` = dplyr::coalesce(pct_top25, FALSE) | dplyr::coalesce(pct_bottom25, FALSE),
      `10%` = dplyr::coalesce(pct_top10, FALSE) | dplyr::coalesce(pct_bottom10, FALSE),
      `5%`  = dplyr::coalesce(pct_top5,  FALSE) | dplyr::coalesce(pct_bottom5,  FALSE),
      `1%`  = dplyr::coalesce(pct_top1,  FALSE) | dplyr::coalesce(pct_bottom1,  FALSE),
      FADE_dir  = .gcol(gene_attr_df, "flag_fade"),
      Accum_dir = .gcol(gene_attr_df, "flag_accum"))
  }
  base$RER_acc  <- .gcol(gene_attr_df, "flag_rer_acc")
  base$RER_decc <- .gcol(gene_attr_df, "flag_rer_decc")
  # RER_dir: the ONE direction-matching RER flag, Top only counts
  # acceleration, Bottom only deceleration, so a Bottom-direction gene never
  # gets credit for an unrelated acceleration call (and vice versa). Global
  # has no direction of its own, so either counts. RER_acc/RER_decc themselves
  # stay in the display below (both are still shown, just not both scored).
  base$RER_dir <- switch(direction, Top = base$RER_acc, Bottom = base$RER_decc, base$RER_acc | base$RER_decc)

  all_genes <- base %>%
    dplyr::left_join(.fcs_terms_for(rk_set), by = "gene") %>%
    dplyr::left_join(.ami_desc_for(ami_dirf), by = "gene") %>%
    dplyr::left_join(.posenrich_for(rk_set), by = "gene") %>%
    dplyr::mutate(
      dplyr::across(c(`25%`, `10%`, `5%`, `1%`, FADE_dir, Accum_dir, RER_acc, RER_decc, RER_dir), ~ dplyr::coalesce(.x, FALSE)),
      evidence_label_caas = dplyr::coalesce(evidence_label_caas, "None"),
      evidence_label_rer  = dplyr::coalesce(evidence_label_rer, "None"),
      weight_caas  = .evidence_weight_of(evidence_label_caas),
      weight_rer   = .evidence_weight_of(evidence_label_rer),
      CAAS_FCS_sig = !is.na(caas_fcs_terms) & nzchar(caas_fcs_terms),
      RER_FCS_sig  = !is.na(rer_fcs_terms)  & nzchar(rer_fcs_terms),
      CAAS_AMI_net = !is.na(caas_ami_desc)  & nzchar(caas_ami_desc),
      RER_AMI_net  = !is.na(rer_ami_desc)   & nzchar(rer_ami_desc),
      FADE_AMI_net = !is.na(fade_ami_desc)  & nzchar(fade_ami_desc),
      PE_sig       = !is.na(pe_terms)       & nzchar(pe_terms)
    )

  out <- all_genes %>%
    dplyr::filter(`25%` | `10%` | `5%` | `1%` | FADE_dir | Accum_dir | RER_dir |
                  CAAS_FCS_sig | RER_FCS_sig | CAAS_AMI_net | RER_AMI_net | FADE_AMI_net | PE_sig)
  if (nrow(out) == 0) return(out)

  out <- out %>%
    .grouped_composite(fade_col = "FADE_dir") %>%
    dplyr::mutate(
      `CAAS FCS term(s)`    = .collapsible_list(caas_fcs_terms),
      `RER FCS term(s)`     = .collapsible_list(rer_fcs_terms),
      `CAAS AMI desc(s)`    = .collapsible_list(caas_ami_desc),
      `RER AMI desc(s)`     = .collapsible_list(rer_ami_desc),
      `FADE AMI desc(s)`    = .collapsible_list(fade_ami_desc),
      `PE term(s)`          = .collapsible_list(pe_terms),
      `CAAS evidence`       = .evidence_badge(evidence_label_caas),
      `RER evidence`        = .evidence_badge(evidence_label_rer)
    )

  out %>%
    dplyr::arrange(dplyr::desc(`Composite score`)) %>%
    dplyr::select(Gene = gene, `25%`, `10%`, `5%`, `1%`, FADE_dir, RER_acc, RER_decc, Accum_dir,
                  CAAS_FCS_sig, `CAAS FCS term(s)`, `CAAS evidence`,
                  RER_FCS_sig, `RER FCS term(s)`, `RER evidence`,
                  CAAS_AMI_net, `CAAS AMI desc(s)`, RER_AMI_net, `RER AMI desc(s)`, FADE_AMI_net, `FADE AMI desc(s)`,
                  PE_sig, `PE position` = pe_position, `PE term(s)`,
                  caas_family, rer_family, ami_fade_family, posenrich_family,
                  `N criteria`, `Composite score`) %>%
    head(params$top_n * 3)
}
```

```{r interesting_genes_tabs, eval=has_fcs_stats_cmp, results='asis'}
for (dr in c("Global", "Top", "Bottom")) {
  cat(sprintf("\n## %s\n\n", dr))
  tbl <- build_interesting_genes(dr)
  if (nrow(tbl) == 0) { cat("_No genes meet any criterion for this direction._\n\n"); next }
  print(kbl_html(tbl, height = if (nrow(tbl) > 15) "480px" else NULL, escape = FALSE))
}
```

`r if (!has_fcs_stats_cmp) "> **fcs_stats_file not supplied - Interesting Genes table skipped.**"`

---

# Interesting Positions {.tabset .tabset-fade}

One row per position carrying at least one of the criteria below, faceted
Global/Top/Bottom. Same philosophy as Interesting Genes, but at the position's
own native resolution - each criterion is either **position-native** (this
exact position: CAAS percentile, FADE-site overlap, posenrich significant-term
membership) or **gene-inherited** (every position in a flagged gene carries
the same value: RER acceleration/deceleration, Accumulation significance, FCS
gene-set membership, AMI module membership). That distinction matters - a
gene can be FADE-significant overall while this specific position isn't the
site driving it, so gene-level and position-level FADE overlap are genuinely
different claims, not duplicated information. `Composite score` is the same
plain four-family sum (CAAS / RER / AMI+FADE / posenrich, family subtotals
shown alongside the total) as Interesting Genes above - see that section for
the full breakdown; `CAAS evidence`/`RER evidence` are the same gene-inherited
evidence-tier columns too. `COSMIC`/`PAI3D_pathogenic` are
**EXTRAS**: shown for characterisation only, excluded from `Composite score`
(they describe a different kind of evidence - external database overlap
rather than corroboration from this pipeline's own modules - mixing them into
the same sum would conflate the two). Both are reported as 3-state
tags rather than TRUE/FALSE, so a real negative is never conflated with "never
checked": `PAI3D_pathogenic` is `pathogenic` / `benign` / `missing`; `COSMIC`
is `hit` / `no_hit` / `missing` (COSMIC's own file only ever lists positive
hits, so "checked, no record" is inferred from PrimateAI-3D's exhaustive
per-position coverage as the "was this position checkable at all" reference).

```{r interesting_positions_prep, eval=has_position_scores_cmp}
build_interesting_positions <- function(direction) {
  rk_set    <- switch(direction, Top = c("top", "accelerating"), Bottom = c("bottom", "decelerating"), NULL)
  side_vals <- switch(direction, Top = c("top", "both"), Bottom = c("bottom", "both"), NULL)

  base <- position_scores_cmp_df
  if (!is.null(side_vals)) base <- base %>% dplyr::filter(change_side %in% side_vals)
  if (nrow(base) == 0) return(base)

  base <- .position_pct_flags(base, tolower(direction))

  # .position_pct_flags transmutes base down to gene/Position/CAAS_score/pcts, so
  # re-attach the raw-AA descriptors (position_scores.tsv carries them per position).
  .desc_cols <- intersect(
    c("derived_residues", "top_residue_support", "bottom_residue_support",
      "n_conserved_pairs", "convergence_schemes"),
    names(position_scores_cmp_df))
  if (length(.desc_cols)) {
    base <- base %>% dplyr::left_join(
      position_scores_cmp_df %>%
        dplyr::transmute(gene = Gene, Position, dplyr::across(dplyr::all_of(.desc_cols))),
      by = c("gene", "Position"))
  }

  # Position-native FADE-site overlap.
  fade_df <- switch(direction,
    Top = fade_sites_top_cmp_df, Bottom = fade_sites_bottom_cmp_df,
    dplyr::bind_rows(fade_sites_top_cmp_df, fade_sites_bottom_cmp_df))
  base$FADE_overlap <- if (!is.null(fade_df)) {
    paste(base$gene, base$Position) %in% paste(fade_df$gene, fade_df$Position)
  } else FALSE

  # Gene-inherited flags (every position in a flagged gene shares its value).
  # gene_attr_df can be NULL here even when has_position_scores_cmp is TRUE --
  # the two gates (fcs_stats_file vs position_scores_file) are independent, so
  # this table must degrade gracefully (RER/Accum flags just come back FALSE
  # via .gcol) rather than assume fcs_stats_file was also supplied.
  if (!is.null(gene_attr_df)) {
    base <- base %>%
      dplyr::left_join(gene_attr_df %>% dplyr::select(gene, dplyr::any_of(c(
        "flag_rer_acc", "flag_rer_decc", "flag_accum_top", "flag_accum_bottom", "flag_accum"))),
        by = "gene")
  }
  base$RER_acc  <- .gcol(base, "flag_rer_acc")
  base$RER_decc <- .gcol(base, "flag_rer_decc")
  # RER_dir: direction-matching flag only, same reasoning as
  # build_interesting_genes() above (a Bottom position never gets credit for
  # an unrelated acceleration call, and vice versa).
  base$RER_dir <- switch(direction, Top = base$RER_acc, Bottom = base$RER_decc, base$RER_acc | base$RER_decc)
  base$Accum_dir <- switch(direction,
    Top = .gcol(base, "flag_accum_top"), Bottom = .gcol(base, "flag_accum_bottom"), .gcol(base, "flag_accum"))

  ami_dirf <- switch(direction, Top = "Top", Bottom = "Bottom", NULL)
  base <- base %>%
    dplyr::left_join(.fcs_terms_for(rk_set), by = "gene") %>%
    dplyr::left_join(.ami_desc_for(ami_dirf), by = "gene") %>%
    dplyr::mutate(
      evidence_label_caas = dplyr::coalesce(evidence_label_caas, "None"),
      evidence_label_rer  = dplyr::coalesce(evidence_label_rer, "None"),
      weight_caas  = .evidence_weight_of(evidence_label_caas),
      weight_rer   = .evidence_weight_of(evidence_label_rer),
      CAAS_FCS_sig = !is.na(caas_fcs_terms) & nzchar(caas_fcs_terms),
      RER_FCS_sig  = !is.na(rer_fcs_terms)  & nzchar(rer_fcs_terms),
      CAAS_AMI_net = !is.na(caas_ami_desc)  & nzchar(caas_ami_desc),
      RER_AMI_net  = !is.na(rer_ami_desc)   & nzchar(rer_ami_desc),
      FADE_AMI_net = !is.na(fade_ami_desc)  & nzchar(fade_ami_desc)
    )

  # Position-native posenrich significance: exact Gene:Position match, not
  # just gene, a gene can have OTHER significant positions without this one
  # being among them.
  if (has_posenrich_le) {
    pe <- posenrich_le_df %>%
      dplyr::filter(is.null(rk_set) | ranking %in% rk_set) %>%
      dplyr::mutate(gene = sub(":.*", "", gene_position), Position = suppressWarnings(as.integer(sub(".*:", "", gene_position)))) %>%
      dplyr::group_by(gene, Position) %>%
      dplyr::summarise(pe_terms = paste(sort(unique(pathway)), collapse = "; "), .groups = "drop")
    base <- base %>% dplyr::left_join(pe, by = c("gene", "Position"))
  } else {
    base$pe_terms <- NA_character_
  }
  base$PE_sig <- !is.na(base$pe_terms) & nzchar(base$pe_terms)

  # EXTRAS, characterisation only, never in the score (see section text).
  # Both reported as 3-state tags (not TRUE/FALSE) so "checked, negative" is
  # never conflated with "never checked":
  #   PAI3D_pathogenic: "pathogenic" / "benign" / "missing"
  #   COSMIC:           "hit" / "no_hit" / "missing"

  # PrimateAI-3D's own table is exhaustive (one row per position it scored,
  # `prediction` always benign/pathogenic), a position simply absent from it
  # was never scored by PrimateAI at all, a genuine "missing", not a benign
  # call. Also doubles as the "was this position even checkable" reference for
  # COSMIC below (COSMIC's own file is hits-only, see there).
  .pai_pos <- NULL
  if (has_vep_primateai_cmp) {
    .pai <- vep_primateai_cmp_df
    if ("caap_group" %in% names(.pai)) .pai <- dplyr::filter(.pai, caap_group == "US")
    .pred_col <- intersect(c("prediction"), names(.pai))[1]
    if (!is.na(.pred_col)) {
      .pai_pos <- .pai %>%
        dplyr::mutate(.patho = grepl("pathogenic", .data[[.pred_col]], ignore.case = TRUE)) %>%
        dplyr::group_by(Gene, Position) %>% dplyr::summarise(.patho = any(.patho), .groups = "drop")
    }
  }
  base$PAI3D_pathogenic <- if (!is.null(.pai_pos)) {
    .m <- .pai_pos[match(paste(base$gene, base$Position), paste(.pai_pos$Gene, .pai_pos$Position)), ]
    dplyr::case_when(is.na(.m$.patho) ~ "missing", .m$.patho ~ "pathogenic", TRUE ~ "benign")
  } else rep("missing", nrow(base))

  # COSMIC's own file (cosmic_scores.tsv) is HITS-ONLY, every row IS a
  # documented COSMIC mutation, unlike PrimateAI's exhaustive per-position
  # table. A position absent from it could mean either "checked, no COSMIC
  # record" (a real negative) or "never genomically checkable at all" (e.g.
  # liftover/coordinate mapping failed), the file alone can't tell the two
  # apart. PrimateAI-3D's own position set is used as the "was this position
  # checkable" reference instead (verified against real data: 116/118 COSMIC
  # hits also appear in PrimateAI's set, i.e. the two databases draw from
  # essentially the same genomically-mapped position universe).
  base$COSMIC <- if (has_vep_cosmic_cmp) {
    .cosmic <- if ("caap_group" %in% names(vep_cosmic_cmp_df)) dplyr::filter(vep_cosmic_cmp_df, caap_group == "US") else vep_cosmic_cmp_df
    .cosmic_hit <- paste(base$gene, base$Position) %in% paste(.cosmic$Gene, .cosmic$Position)
    .checkable <- if (!is.null(.pai_pos)) {
      paste(base$gene, base$Position) %in% paste(.pai_pos$Gene, .pai_pos$Position)
    } else rep(TRUE, nrow(base))  # no PrimateAI reference available, can't distinguish no_hit from missing
    dplyr::case_when(.cosmic_hit ~ "hit", .checkable ~ "no_hit", TRUE ~ "missing")
  } else rep("missing", nrow(base))

  all_positions <- base %>%
    dplyr::mutate(dplyr::across(c(`25%`, `10%`, `5%`, `1%`, FADE_overlap, RER_acc, RER_decc, RER_dir, Accum_dir,
                                 CAAS_FCS_sig, RER_FCS_sig, CAAS_AMI_net, RER_AMI_net, FADE_AMI_net, PE_sig),
                                 ~ dplyr::coalesce(.x, FALSE)))

  out <- all_positions %>%
    dplyr::filter(`25%` | `10%` | `5%` | `1%` | FADE_overlap | Accum_dir | RER_dir |
                  CAAS_FCS_sig | RER_FCS_sig | CAAS_AMI_net | RER_AMI_net | FADE_AMI_net | PE_sig)
  if (nrow(out) == 0) return(out)

  # Collapsible <details> rendering + grouped composite score, see
  # build_interesting_genes() above (same four-family sum, "FADE_overlap"
  # here vs "FADE_dir" there is the only difference). interesting_positions_tabs
  # below must render this table with escape = FALSE.
  out <- out %>%
    .grouped_composite(fade_col = "FADE_overlap") %>%
    dplyr::mutate(
      `Hypotheses`          = if ("supporting_hypotheses" %in% names(.)) .collapsible_list(supporting_hypotheses) else NA_character_,
      `CAAS FCS term(s)`    = .collapsible_list(caas_fcs_terms),
      `RER FCS term(s)`     = .collapsible_list(rer_fcs_terms),
      `CAAS AMI desc(s)`    = .collapsible_list(caas_ami_desc),
      `RER AMI desc(s)`     = .collapsible_list(rer_ami_desc),
      `FADE AMI desc(s)`    = .collapsible_list(fade_ami_desc),
      `PE term(s)`          = .collapsible_list(pe_terms),
      `CAAS evidence`       = .evidence_badge(evidence_label_caas),
      `RER evidence`        = .evidence_badge(evidence_label_rer),
      `Top support`         = if ("top_residue_support" %in% names(.)) top_residue_support else NA_character_,
      `Bottom support`      = if ("bottom_residue_support" %in% names(.)) bottom_residue_support else NA_character_
    )

  out %>%
    dplyr::arrange(dplyr::desc(`Composite score`)) %>%
    dplyr::select(Gene = gene, Position, `CAAS Score` = CAAS_score,
                  `Top/Bottom residues` = dplyr::any_of("derived_residues"),
                  `Top support`, `Bottom support`,
                  `N conserved pairs` = dplyr::any_of("n_conserved_pairs"),
                  `Conv. schemes` = dplyr::any_of("convergence_schemes"),
                  `N Hypotheses` = dplyr::any_of("n_hypotheses"),
                  `Hypotheses`,
                  `25%`, `10%`, `5%`, `1%`,
                  FADE_overlap, RER_acc, RER_decc, Accum_dir,
                  CAAS_FCS_sig, `CAAS FCS term(s)`, `CAAS evidence`,
                  RER_FCS_sig, `RER FCS term(s)`, `RER evidence`,
                  CAAS_AMI_net, `CAAS AMI desc(s)`, RER_AMI_net, `RER AMI desc(s)`, FADE_AMI_net, `FADE AMI desc(s)`,
                  PE_sig, `PE term(s)`,
                  COSMIC, PAI3D_pathogenic,
                  caas_family, rer_family, ami_fade_family, posenrich_family,
                  `N criteria`, `Composite score`) %>%
    head(params$top_n * 3)
}
```

```{r interesting_positions_tabs, eval=has_position_scores_cmp, results='asis'}
for (dr in c("Global", "Top", "Bottom")) {
  cat(sprintf("\n## %s\n\n", dr))
  tbl <- build_interesting_positions(dr)
  if (nrow(tbl) == 0) { cat("_No positions meet any criterion for this direction._\n\n"); next }
  print(kbl_html(tbl, height = if (nrow(tbl) > 15) "480px" else NULL, escape = FALSE))
}
```

`r if (!has_position_scores_cmp) "> **position_scores_file not supplied - Interesting Positions table skipped.**"`

---

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere - Enrichment Module*
*Date of generation: `r Sys.Date()`*
