---
title: "Comparative Enrichment Across Modules — `r params$traitname`"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
    mathjax: null                       # no math in this report
    md_extensions: "-tex_math_dollars"  # gene-set names contain '$' (e.g. V$HNF4_01) — never parse as TeX math
params:
  fcs_dir:    "cmp_fcs"       # dir of per-module <module>_fcs_all.tsv
  fcs_le_dir: "cmp_fcs_le"    # dir of per-module <module>_leading_edge.tsv
  fdr_thr:    0.15
  pperm_thr:  0.025
  domino_module_thr: 0.05  # DOMINO's own module-significance cutoff (domino.nf's domino_module_thr)
  top_n:      20
  traitname:  "trait"
  ami_module_desc_file:       NULL  # AMI's ami_networks/module_descriptions_all_lists.tsv (optional)
  posenrich_dotplot_file:     NULL  # posenrich's posenrich_overall_dotplot.tsv (optional)
  # posenrich_leading_edge.tsv (gene:position members of significant posenrich
  # terms, already position-granular) -- drives the scorecard's posenrich angle
  # AND the Interesting Genes/Interesting Positions tables below.
  posenrich_leading_edge_file: NULL
  # SCORING's own gene-level table (percentile flags + FADE/RER/Accumulation
  # significance) -- same file 12.FCS_general_report.Rmd/14.Position_enrichment_report.Rmd
  # consume, reused here for the Interesting Genes/Positions tables.
  fcs_stats_file: NULL
  # Position-level CAAS scores + FADE-site/VEP annotations -- same optional
  # inputs 14.Position_enrichment_report.Rmd's Position Characterisation
  # section consumes, reused here for the Interesting Positions table.
  position_scores_file:  NULL
  vep_primateai_file:    NULL
  vep_cosmic_file:       NULL
  fade_sites_top_file:    NULL
  fade_sites_bottom_file: NULL
  # SCORING's published percentile slices (scoring_compute.R) -- when
  # supplied, gene/position 25/10/5/1% membership is read from these
  # directories instead of being re-derived independently in this report (see
  # fcs_percentile_flags()'s gene_lists_dir param / .position_pct_flags()).
  gene_lists_dir:     NULL
  position_lists_dir: NULL
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 7, dev = "png", dpi = 200)
suppressPackageStartupMessages({ library(igraph); library(tidyverse); library(knitr); library(kableExtra) })
filter <- dplyr::filter; select <- dplyr::select; rename <- dplyr::rename; mutate <- dplyr::mutate
has_pkg <- function(p) requireNamespace(p, quietly = TRUE)
.src_dir <- file.path(dirname(normalizePath(knitr::current_input(), mustWork = FALSE)), "src")
# fcs_percentile_flags() (used by the Interesting Genes/Positions tables) now
# lives in orthogonal_score.R itself -- deliberately NOT sourcing fcs_enrich.R
# here, which would drag in a hard RERconverge dependency this report never
# otherwise needs (fcs_enrich.R's fastwilcoxGMT/fcs_run_all machinery is
# report 12/RER-report-specific).
source(file.path(.src_dir, "orthogonal_score.R"))

kbl_html <- function(df, height = NULL, escape = TRUE) {
  k <- kable(df, format = "html", escape = escape) %>%
    kable_styling(bootstrap_options = c("striped", "condensed", "hover"), full_width = FALSE)
  if (!is.null(height)) k <- scroll_box(k, height = height)
  k
}
```

```{r load, include=FALSE}
`%||%` <- function(a, b) if (!is.null(a)) a else b
REQ <- c("ranking", "database", "pathway", "stat", "p.adj")

# Per-module FCS results: cmp_fcs/<module>_fcs_all.tsv (caas/rer).
mod_files <- list.files(params$fcs_dir, pattern = "_fcs_all\\.tsv$", full.names = TRUE)
fcs_by_mod <- list()
for (f in mod_files) {
  mod <- sub("_fcs_all\\.tsv$", "", basename(f))
  d <- tryCatch(readr::read_tsv(f, show_col_types = FALSE), error = function(e) NULL)
  if (!is.null(d) && nrow(d) > 0 && all(REQ %in% names(d))) {
    fcs_by_mod[[mod]] <- dplyr::mutate(d, module = mod)
  }
}
fcs_all_mod <- if (length(fcs_by_mod)) dplyr::bind_rows(fcs_by_mod) else tibble()
modules_present <- names(fcs_by_mod)
has_multi <- length(fcs_by_mod) >= 1

# Direction facet (Global/top/bottom), derived once here since every
# directional section below groups by it. Each module's `ranking` vocabulary
# differs: CAAS uses top/bottom/global directly; RER's accelerating/
# decelerating is the same two-way split framed around RER's own statistic
# (RER carries no separate undirected ranking, so it contributes no "global"
# rows). CAAS's "global" ranking is its OWN direction here, kept fully
# separate from top/bottom -- it used to be duplicated into BOTH the top and
# bottom buckets under the same (module, direction, database, pathway) group
# key permod_prep groups by below, which silently blended two different
# statistical tests (the real directional stat and the undirected global stat
# for the same pathway) into one summarised row. That's now impossible: a
# "global" row can never share a group key with a "top" or "bottom" row.
if (has_multi) {
  fcs_dir_mod <- fcs_all_mod %>%
    mutate(direction = dplyr::case_when(
      ranking %in% c("top", "accelerating")    ~ "top",
      ranking %in% c("bottom", "decelerating") ~ "bottom",
      ranking == "global"                      ~ "global",
      TRUE                                      ~ NA_character_
    )) %>%
    dplyr::filter(!is.na(direction))
} else {
  fcs_dir_mod <- fcs_all_mod
}

# Per-module leading-edge tables (cmp_fcs_le/<module>_leading_edge.tsv) feed the
# orthogonal composite score: each module's own directional percentile
# concentration + cross-module corroboration flags (FADE/RER/Accumulation,
# and PrimateAI-3D/COSMIC once wired in from posenrich). Percentile columns
# reflect CAAS's own ranking in every module's leading edge (per-module FCS
# reports are annotated with CAAS's fcs_stats.tsv as their overlay) — that's
# intentional, not a bug: it's the shared reference axis every module's
# leading edge is cross-checked against.
le_files <- list.files(params$fcs_le_dir, pattern = "_leading_edge\\.tsv$", full.names = TRUE)
le_by_mod <- list()
for (f in le_files) {
  mod <- sub("_leading_edge\\.tsv$", "", basename(f))
  d <- tryCatch(readr::read_tsv(f, show_col_types = FALSE), error = function(e) NULL)
  if (!is.null(d) && nrow(d) > 0) le_by_mod[[mod]] <- dplyr::mutate(d, module = mod)
}
le <- if (length(le_by_mod)) dplyr::bind_rows(le_by_mod) else tibble()
has_le <- nrow(le) > 0

# ── Cross-angle inputs: AMI (network modules) + posenrich (position-level) ──
# Neither is gene-set-ranked the way CAAS/RER FCS is, so they don't fold into
# fcs_all_mod/le above -- each gets its own section below, and all three feed
# the Integrated gene scorecard via the genes they involve.
.file_ok <- function(p) !is.null(p) && !is.na(p) && nchar(p) > 0 && file.exists(p)

has_ami <- .file_ok(params$ami_module_desc_file)
ami_df  <- if (has_ami) readr::read_tsv(params$ami_module_desc_file, show_col_types = FALSE) else NULL

has_posenrich_dotplot <- .file_ok(params$posenrich_dotplot_file)
posenrich_dotplot_df  <- if (has_posenrich_dotplot) readr::read_tsv(params$posenrich_dotplot_file, show_col_types = FALSE) else NULL

# posenrich_leading_edge.tsv: gene:position members of significant posenrich
# terms (already position-granular -- ranking, database, pathway, top_frac,
# gene, gene_position). Drives the scorecard's posenrich angle AND the
# Interesting Genes/Interesting Positions tables below.
has_posenrich_le <- .file_ok(params$posenrich_leading_edge_file)
posenrich_le_df  <- if (has_posenrich_le) readr::read_tsv(params$posenrich_leading_edge_file, show_col_types = FALSE) else NULL

# ── Gene/position-level inputs for the Interesting Genes/Positions tables ──
# Same optional files 12.FCS_general_report.Rmd (fcs_stats) and
# 14.Position_enrichment_report.Rmd (position_scores, VEP, FADE sites) already
# consume -- reused here rather than re-derived.
has_fcs_stats_cmp <- .file_ok(params$fcs_stats_file)
fcs_stats_cmp_df  <- if (has_fcs_stats_cmp) readr::read_tsv(params$fcs_stats_file, show_col_types = FALSE) else NULL

has_position_scores_cmp <- .file_ok(params$position_scores_file)
position_scores_cmp_df  <- if (has_position_scores_cmp) readr::read_tsv(params$position_scores_file, show_col_types = FALSE) else NULL

has_vep_primateai_cmp <- .file_ok(params$vep_primateai_file)
vep_primateai_cmp_df  <- if (has_vep_primateai_cmp) readr::read_tsv(params$vep_primateai_file, show_col_types = FALSE) else NULL

has_vep_cosmic_cmp <- .file_ok(params$vep_cosmic_file)
vep_cosmic_cmp_df  <- if (has_vep_cosmic_cmp) readr::read_tsv(params$vep_cosmic_file, show_col_types = FALSE) else NULL

has_fade_sites_top_cmp    <- .file_ok(params$fade_sites_top_file)
has_fade_sites_bottom_cmp <- .file_ok(params$fade_sites_bottom_file)
fade_sites_top_cmp_df <- if (has_fade_sites_top_cmp) {
  readr::read_csv(params$fade_sites_top_file, show_col_types = FALSE) %>%
    dplyr::rename(gene = gene, Position = position) %>%
    dplyr::mutate(Position = suppressWarnings(as.integer(Position)))
} else NULL
fade_sites_bottom_cmp_df <- if (has_fade_sites_bottom_cmp) {
  readr::read_csv(params$fade_sites_bottom_file, show_col_types = FALSE) %>%
    dplyr::rename(gene = gene, Position = position) %>%
    dplyr::mutate(Position = suppressWarnings(as.integer(Position)))
} else NULL

# Per-gene attribute table: percentile flags (from fcs_stats' own score_top/
# score_bottom, same derivation as 12.FCS_general_report.Rmd) + the gene-level
# flag_* columns fcs_stats.tsv already carries (gate/FADE/RER/Accumulation,
# direction-aware -- see orthogonal_score.R's ORTH_CROSS_COLS for the
# flag_accum_top/flag_accum_bottom convention).
gene_attr_df <- if (has_fcs_stats_cmp) {
  pf <- fcs_percentile_flags(fcs_stats_cmp_df, gene_lists_dir = params$gene_lists_dir)
  flag_cols_cmp <- grep("^flag_", names(fcs_stats_cmp_df), value = TRUE)
  dplyr::left_join(pf, fcs_stats_cmp_df[, c("gene", flag_cols_cmp)], by = "gene")
} else NULL
```

```{r cross_angle_gene_hits, include=FALSE}
# ── Cross-angle gene-level hit sets ───────────────────────────────────────
# Hoisted here (was previously computed only inside the Integrated gene
# scorecard, at the bottom of the report) because the per-module significance
# and AMI module description sections above the scorecard now also need
# per-gene FCS/posenrich corroboration -- one shared definition, reused by
# three sections, rather than three drifting copies.
#
# Each angle uses ITS OWN module's significance definition (not "any
# leading-edge/described membership") -- FCS/RER's dual p.adj+p.perm gate,
# posenrich's own p_adj gate, DOMINO's own module-significance cutoff.
# FCS and posenrich both carry a genuine "global" ranking (CAAS's undirected
# composite; posenrich's own directions = ["global","top","bottom"]), so
# fcs_genes/posenrich_genes have real direction=="Global" rows. AMI has no
# equivalent -- DOMINO only ever runs on Top/Bottom percentile gene lists
# (.TOP_THRESHOLDS/.BOTTOM_THRESHOLDS in 13.AMI_analysis.Rmd, no undirected
# list), so ami_genes' direction column is only ever "Top"/"Bottom"; a
# "Global" row for AMI never exists here. This means the scorecard's Global
# tab (a plain full_join keyed on gene+direction) can never carry an AMI
# contribution -- unlike .ami_desc_for(dir_filter = NULL), used by the
# Interesting Genes/Positions AMI cells, which pools Top+Bottom explicitly
# for its own "Global" view instead of relying on a literal direction match.

# FCS/RER own significance: BH < fdr_thr AND (p.perm < pperm_thr when
# present) -- same formula permod_prep below applies, computed once here off
# fcs_dir_mod so this chunk and permod_prep agree on what "significant" means.
# fcs_leading_edge.tsv (`le`) is NOT itself restricted to significant terms
# (12.FCS_general_report.Rmd's leading-edge export only requires drv_score >
# 0, i.e. "this gene drives some signal in this term", not "this term
# cleared BH") -- inner_join against fcs_sig_terms is what actually restricts
# fcs_genes to genes belonging to a term each module itself calls significant.
fcs_sig_terms <- if (has_multi) {
  fcs_dir_mod %>%
    dplyr::filter(stat > 0, !is.na(p.adj), p.adj < params$fdr_thr,
                  !("p.perm" %in% names(.)) | is.na(p.perm) | p.perm < (params$pperm_thr %||% 1)) %>%
    dplyr::mutate(description = dplyr::coalesce(description, pathway)) %>%
    dplyr::distinct(module, direction, database, pathway, description)
} else {
  tibble::tibble(module = character(), direction = character(), database = character(),
                 pathway = character(), description = character())
}

fcs_genes <- if (has_le && has_multi) {
  le %>%
    dplyr::mutate(direction = dplyr::case_when(
      ranking %in% c("top", "accelerating")    ~ "top",
      ranking %in% c("bottom", "decelerating") ~ "bottom",
      ranking == "global"                      ~ "global",
      TRUE                                      ~ NA_character_
    )) %>%
    dplyr::filter(!is.na(direction)) %>%
    dplyr::inner_join(fcs_sig_terms, by = c("module", "direction", "database", "pathway")) %>%
    dplyr::mutate(direction = tools::toTitleCase(direction)) %>%
    dplyr::distinct(gene, direction) %>%
    dplyr::mutate(fcs_hit = TRUE)
} else {
  tibble::tibble(gene = character(), direction = character(), fcs_hit = logical())
}

# AMI angle: DOMINO's own module-significance cutoff (domino_module_thr,
# domino.nf's own default 0.05) gates which described modules count --
# "described" alone (not Too small/No enrichment) was never a significance
# claim, just "STRING could summarise it". Expands each significant module's
# FULL membership into one row per gene: `all_genes` (every member gene,
# uncapped -- describe_clusters() in 13.AMI_analysis.Rmd) when present, else
# the older capped `genes` column (top-8 representatives) for
# module_descriptions_all_lists.tsv files written before all_genes existed.
# direction is only ever "Top"/"Bottom" here (DOMINO never runs on an
# undirected gene list, so there's no "Global" value to filter for -- see
# this chunk's opening comment).
ami_genes <- if (has_ami) {
  .ami_gene_col <- if ("all_genes" %in% names(ami_df)) "all_genes" else "genes"
  ami_df %>%
    dplyr::filter(!description %in% c("Too small", "No enrichment"),
                  direction %in% c("Top", "Bottom"),
                  !is.na(domino_p_adj), domino_p_adj < params$domino_module_thr) %>%
    dplyr::mutate(gene = strsplit(.data[[.ami_gene_col]], ",\\s*")) %>%
    tidyr::unnest(gene) %>%
    dplyr::mutate(gene = trimws(gene)) %>%
    dplyr::distinct(gene, direction) %>%
    dplyr::mutate(ami_hit = TRUE)
} else {
  tibble::tibble(gene = character(), direction = character(), ami_hit = logical())
}

# Posenrich angle: posenrich_leading_edge.tsv is ALREADY restricted to
# significant (p_adj < posenrich's own padj_thr) terms by posenrich_enrich.py
# itself (only rows with r["sig"] get written) -- no additional gate needed
# here, just direction mapping (now including "global"). Also carries the
# actual Position(s) behind each hit (kept for the scorecard's Position
# column below -- lets a reader see whether the specific position posenrich
# flagged for a gene is the same one driving that gene's FCS/AMI signal, not
# just that the gene appears in both).
posenrich_genes <- if (has_posenrich_le) {
  posenrich_le_df %>%
    dplyr::mutate(direction = dplyr::case_when(
      ranking %in% c("top", "accelerating")    ~ "top",
      ranking %in% c("bottom", "decelerating") ~ "bottom",
      ranking == "global"                      ~ "global",
      TRUE                                      ~ NA_character_
    ),
    Position = sub(".*:", "", gene_position)) %>%
    dplyr::filter(!is.na(direction)) %>%
    dplyr::mutate(direction = tools::toTitleCase(direction)) %>%
    dplyr::group_by(gene, direction) %>%
    dplyr::summarise(Position = paste(sort(unique(Position)), collapse = ", "), .groups = "drop") %>%
    dplyr::mutate(posenrich_hit = TRUE)
} else {
  tibble::tibble(gene = character(), direction = character(), Position = character(), posenrich_hit = logical())
}

# ── Orthogonal composite signal, per (module, direction, database, pathway) ──
# Lancaster-weighted composite of each module's own directional percentile
# concentration (from CAAS's percentile quartet, 25/10/5/1%, the shared
# reference axis every per-module report is annotated against) plus
# independent cross-module flags
# -- excluding whichever flag would be self-referential for the CURRENT module
# (e.g. RER's own leading edge doesn't get to cite flag_rer_acc/flag_rer_decc
# as "corroboration"). Background rates are estimated from each module's own
# observed leading-edge gene pool (this report doesn't carry the full universe
# file). Computed once here, ahead of every section that needs it (Per-module
# significance's `Orthog. signal` column).
signal_by_mod <- tibble()
if (has_le) {
  self_ref <- list(caas = character(0), rer = c("flag_rer_acc", "flag_rer_decc"))
  signal_rows <- list()
  for (mod in unique(le$module)) {
    le_mod <- le %>% filter(module == mod)
    for (rk in unique(le_mod$ranking)) {
      le_rk <- le_mod %>% filter(ranking == rk)
      # "global" (CAAS's undirected ranking) is kept as its own direction, NOT
      # coerced to "top" -- orth_cols_for_direction() already has a correct
      # non-directional branch (no percentile quartet, both directions' cross
      # flags) for any direction string other than "top"/"bottom"; passing
      # "global" straight through reaches it instead of silently scoring the
      # undirected ranking as if it were the top direction.
      rk_dir <- dplyr::case_when(rk %in% c("top", "accelerating") ~ "top",
                                  rk %in% c("bottom", "decelerating") ~ "bottom",
                                  TRUE ~ "global")
      cols <- orth_cols_for_direction(rk_dir)
      cross_cols <- setdiff(cols$cross_cols, self_ref[[mod]] %||% character(0))
      uni_genes <- le_rk %>% distinct(gene, .keep_all = TRUE)
      n_uni <- nrow(uni_genes)
      if (n_uni < 10) next
      bg <- orth_background_rates(uni_genes, c(cols$percentile_cols, cross_cols))
      sc <- orthogonal_score(le_rk, bg, cols$percentile_cols, cross_cols, n_uni) %>%
        mutate(
          module_label = dplyr::case_when(
            mod == "rer" & rk == "accelerating" ~ "rer_accelerating",
            mod == "rer" & rk == "decelerating" ~ "rer_decelerating",
            TRUE ~ mod
          ),
          direction = rk_dir
        )
      signal_rows[[paste(mod, rk)]] <- sc
    }
  }
  signal_by_mod <- dplyr::bind_rows(signal_rows) %>%
    dplyr::select(module_label, direction, database, pathway, signal)
}
if (nrow(signal_by_mod) == 0) {
  signal_by_mod <- tibble(module_label = character(), direction = character(),
                           database = character(), pathway = character(), signal = double())
}

# module + direction -> module_label, for joining signal_by_mod (keyed by
# module_label, which splits RER into rer_accelerating/rer_decelerating) onto
# tables keyed by the plain (module, direction) pair -- e.g. permod_best below.
mod_dir_to_label <- tibble::tribble(
  ~module, ~direction, ~module_label,
  "caas",  "top",      "caas",
  "caas",  "bottom",   "caas",
  "caas",  "global",   "caas",
  "rer",   "top",      "rer_accelerating",
  "rer",   "bottom",   "rer_decelerating"
  # RER has no undirected ranking, so no "rer"/"global" row.
)
```

# Overview

```{r overview, echo=FALSE}
mod_rows <- if (has_multi) sapply(fcs_by_mod, nrow) else integer(0)
tibble::tibble(
  Source = paste0("FCS module: ", names(mod_rows)),
  Rows   = unname(mod_rows)
) %>% kbl_html()
```

This report compares enrichment **across CAAS and RERconverge** (the two modules
with well-behaved FCS rankings and real permulation nulls) for `r params$traitname`.
FADE and Accumulation no longer run their own FCS ranking — both proved unreliable
for genome-wide gene-set enrichment (FADE's max-of-many-sites statistic and
Accumulation's missing permulation null each inflated false "significance") — but
both remain available as **cross-module corroboration**: a term's `signal` (the
orthogonal composite score) combines each module's own directional percentile
concentration with independent hypergeometric enrichment against FADE/RER/
Accumulation flags (and PrimateAI-3D/COSMIC once wired in). A set is "significant"
at BH < `r params$fdr_thr` (and permulation p < threshold where available) with
AUC > 0.5.

DOMINO active-module (AMI) results and posenrich's position-level enrichment are
neither gene-set-ranked the same way as FCS, so they get their own sections below
— but all three angles (FCS, AMI, posenrich) are cross-referenced by gene in the
Integrated gene scorecard at the end.

# Per-module significance {.tabset .tabset-fade}

What does each module call significant on its own, faceted Global/Top/Bottom?
Tables are sorted by this module's own AUC (its own evidence), but each row also
carries `Orthog. signal` — a Lancaster-weighted, hypergeometric orthogonal-support
composite (percentile concentration + independent cross-module corroboration
against FADE/RER/Accumulation flags), so a term that looks marginal on its own AUC
but is already well-corroborated by other modules stands out here too.

`r if (!has_multi) "> **No module FCS results available — skipping.**"`

```{r permod_prep, eval=has_multi}
thr <- params$fdr_thr
# Per (module, direction, database, pathway): best positively-enriched ranking
# within that direction (max AUC, min p.adj). Significance reuses
# fcs_sig_terms (cross_angle_gene_hits chunk above) -- the SAME BH+p.perm gate
# fcs_genes filters its leading-edge membership by, so this table and the
# scorecard's FCS angle never disagree on what "significant" means.
permod_best <- fcs_dir_mod %>%
  filter(stat > 0) %>%
  group_by(module, direction, database, pathway) %>%
  summarise(AUC = max(stat, na.rm = TRUE) + 0.5,
            p.adj = min(p.adj, na.rm = TRUE),
            description = dplyr::first(dplyr::coalesce(description, pathway)),
            .groups = "drop") %>%
  dplyr::left_join(fcs_sig_terms %>% dplyr::select(-description) %>% dplyr::mutate(sig = TRUE),
                    by = c("module", "direction", "database", "pathway")) %>%
  dplyr::mutate(sig = dplyr::coalesce(sig, FALSE)) %>%
  # signal_by_mod is keyed by module_label (splits RER into
  # rer_accelerating/rer_decelerating) -- mod_dir_to_label resolves the plain
  # (module, direction) pair here to the right module_label before the join.
  dplyr::left_join(mod_dir_to_label, by = c("module", "direction")) %>%
  dplyr::left_join(signal_by_mod, by = c("module_label", "direction", "database", "pathway")) %>%
  dplyr::select(-module_label)

permod_summary <- permod_best %>%
  group_by(direction, module) %>%
  summarise(`Terms tested` = dplyr::n(),
            `Significant`  = sum(sig, na.rm = TRUE),
            `% significant` = round(100 * mean(sig, na.rm = TRUE), 1),
            .groups = "drop") %>%
  arrange(direction, desc(Significant))

dir.create("compare_results", showWarnings = FALSE)
readr::write_tsv(permod_summary, "compare_results/per_module_significance_summary.tsv")
readr::write_tsv(permod_best,    "compare_results/per_module_best_per_term.tsv")
```

```{r permod_tabs, eval=has_multi, results='asis'}
# Direction is the primary split (per the report-wide per-direction design,
# Global/Top/Bottom order matching every other tabset in this report); module
# is a second-level tab within each direction. "global" is CAAS's own
# undirected ranking (RER never produces one, so its Global tab is always
# empty for RER) -- a genuine third direction now, not folded into top/bottom.
for (dr in c("global", "top", "bottom")) {
  cat(sprintf("\n## %s {.tabset}\n\n", tools::toTitleCase(dr)))

  dr_summary <- permod_summary %>% filter(direction == dr) %>% select(-direction)
  cat("### Summary\n\n")
  if (nrow(dr_summary) == 0) {
    cat("_No module produced enriched terms for this direction._\n\n")
  } else {
    print(kbl_html(dr_summary))
  }

  for (m in modules_present) {
    cat(sprintf("\n### %s\n\n", toupper(m)))
    d <- permod_best %>%
      filter(module == m, direction == dr, sig) %>%
      mutate(Description = stringr::str_trunc(description, 60),
             AUC = round(AUC, 3), p.adj = signif(p.adj, 3),
             `Orthog. signal` = round(signal, 2)) %>%
      arrange(desc(AUC)) %>%
      select(database, pathway, Description, AUC, p.adj, `Orthog. signal`) %>%
      head(params$top_n)
    if (nrow(d) == 0) {
      cat(sprintf("_No gene set is significant in %s (%s) at BH < %.2g._\n\n", toupper(m), dr, thr))
    } else {
      print(kbl_html(d, height = if (nrow(d) > 15) "420px" else NULL))
    }
  }
}
```

---

# AMI module descriptions {.tabset .tabset-fade}

`r if (!has_ami) "> **No AMI module descriptions available (AMI/DOMINO didn't run, or the file wasn't wired in) — skipping.**"`

DOMINO active modules ([13.AMI_analysis.Rmd](13.AMI_analysis.Rmd)) aren't gene-set-ranked
like CAAS/RER FCS, so they don't fold into the FCS tables above — but a described
module's best term and DOMINO significance are useful context here, and genes inside
a significant module (DOMINO p.adj < `r params$domino_module_thr`) feed the
Integrated gene scorecard below. Rows here are **collapsed across gene lists**, the
same near-duplicate-term clustering [13.AMI_analysis.Rmd](13.AMI_analysis.Rmd)'s own
"Module summary across gene lists" section already applies (normalized edit-distance
over each module's description, e.g. `"V$HNF4_01 / HNF4A"` and `"HNF4A"` merge into
one term) — so a term recurring, worded slightly differently, across several nested
gene lists shows up as ONE row rather than fragmenting into near-duplicates.

```{r ami_desc_prep, eval=has_ami}
# Restricted to DOMINO-significant modules (domino_p_adj < domino_module_thr)
# -- the SAME gate ami_genes/.ami_desc_for() apply (cross_angle_gene_hits /
# interesting_shared_helpers chunks), so the visible table here, the
# scorecard's AMI angle, and the Interesting Genes/Positions AMI cells never
# disagree on which modules count as significant. Mirrors Per-module
# significance above, which likewise only displays `sig` rows, not every
# tested term.
ami_disp <- ami_df %>%
  dplyr::filter(!description %in% c("Too small", "No enrichment"),
                !is.na(domino_p_adj), domino_p_adj < params$domino_module_thr)

# Verbatim copy of 13.AMI_analysis.Rmd's .normalize_term()/.cluster_terms() --
# kept in sync with that copy if this clustering logic ever changes. Collapses
# near-duplicate description strings (case, punctuation, minor rewording) to
# one row via normalized edit-distance (base R's own utils::adist()), not
# just exact string equality.
.normalize_term <- function(x) {
  x <- tolower(x)
  x <- gsub("[^a-z0-9 ]", " ", x)
  gsub("\\s+", " ", trimws(x))
}
.cluster_terms <- function(terms, sim_thr = 0.85) {
  norm <- vapply(terms, .normalize_term, character(1))
  reps_norm <- character(0); reps_orig <- character(0)
  assign <- integer(length(terms))
  for (i in seq_along(terms)) {
    if (length(reps_norm) == 0) { reps_norm <- norm[i]; reps_orig <- terms[i]; assign[i] <- 1L; next }
    d <- utils::adist(norm[i], reps_norm)[1, ]
    sim <- 1 - d / pmax(nchar(norm[i]), nchar(reps_norm), 1)
    best <- which.max(sim)
    if (sim[best] >= sim_thr) {
      assign[i] <- best
    } else {
      reps_norm <- c(reps_norm, norm[i]); reps_orig <- c(reps_orig, terms[i])
      assign[i] <- length(reps_norm)
    }
  }
  reps_orig[assign]
}

# Clustered ONCE over the full pool (not re-clustered per Global/Top/Bottom
# tab below) so a term's cluster assignment never depends on which tab is
# looking at it -- mirrors 13.AMI_analysis.Rmd's render_module_rollup_tabset(),
# which clusters term_rows_all once and only filters per tab afterward.
ami_disp$term <- .cluster_terms(ami_disp$description)

.build_ami_rollup <- function(df) {
  df %>%
    dplyr::group_by(term) %>%
    dplyr::summarise(
      `N modules`    = dplyr::n(),
      `Gene lists`   = paste(sort(unique(gene_list)), collapse = ", "),
      `DOMINO p.adj` = { v <- domino_p_adj[!is.na(domino_p_adj)]; if (length(v)) signif(min(v), 3) else NA_real_ },
      Genes          = suppressWarnings(max(n_genes, na.rm = TRUE)),
      .groups = "drop"
    ) %>%
    dplyr::rename(Term = term) %>%
    dplyr::arrange(`DOMINO p.adj`)
}
```

```{r ami_desc_tabs, eval=has_ami, results='asis'}
for (dr in c("Global", "Top", "Bottom")) {
  cat(sprintf("\n## %s\n\n", dr))
  sub <- if (dr == "Global") ami_disp else ami_disp %>% dplyr::filter(direction == dr)
  if (nrow(sub) == 0) { cat("_No described modules for this direction._\n\n"); next }
  disp <- .build_ami_rollup(sub) %>% head(params$top_n)
  print(kbl_html(disp, height = if (nrow(disp) > 15) "420px" else NULL))
}
```

---

# Posenrich position-level enrichment {.tabset .tabset-fade}

`r if (!has_posenrich_dotplot && !has_posenrich_le) "> **No posenrich outputs available (--posenrich was off, or the files weren't wired in) — skipping.**"`

Position-level enrichment ([14.Position_enrichment_report.Rmd](14.Position_enrichment_report.Rmd))
operates on individual CAAS positions rather than genes, so it gets its own section
here rather than folding into the gene-set FCS tables above. Its most-corroborated terms
(Overall dotplot) feed the Integrated gene scorecard below via the genes they
involve; the Interesting Genes/Interesting Positions tables further down cover
individual gene- and position-level corroboration in full detail. Faceted by
direction below (`r params$top_n` rows each, not one shared budget split
across both) so a direction with fewer but stronger hits isn't crowded out of
a single pooled table by the other.

## Overall dotplot terms — merged summary {.tabset}

One row per significant term, **merged across name variants** that are
themselves `/`-joined synonyms in their own description (e.g. a term
described as `"V$HNF4_01 / HNF4A"` is treated as the same underlying term as
one independently named `"HNF4A"` elsewhere) — connected via shared segments,
not just identical full strings, so synonym spelling drift across databases
doesn't fragment one biological term into several rows. `Merged from` lists
every original (database, pathway) pair folded into the row, so the merge is
auditable rather than a silent black box.

`list_threshold` and `n occurrences` show which top-fraction cutoffs (25/10/5/1%)
the term recurred significant at *before* merging — this recurrence already
feeds into `Orthog. signal` itself (position-granularity, computed in
[14.Position_enrichment_report.Rmd](14.Position_enrichment_report.Rmd): nested
top-fraction recurrence + gene-inherited RER/Accumulation flags at
distinct-gene resolution + position-native FADE-site overlap, Lancaster-
combined — same weighted-by-term-size logic as the gene-level FCS reports, see
that report's Overall dotplot section for the full explanation), so a term
recurring across many cutoffs already scores higher; these two columns are
kept as a human-readable audit trail alongside the number, not a second
scoring mechanism. When a term is merged, `Orthog. signal` takes the
**maximum** across its aliases (the strongest evidence among the synonym
variants), and `list_threshold`/`n occurrences` are unioned across them.

```{r posenrich_dotplot_prep, eval=has_posenrich_dotplot}
posenrich_dotplot_disp <- posenrich_dotplot_df %>%
  dplyr::mutate(Description = dplyr::coalesce(description, pathway),
                signal = round(signal, 2),
                .direction = dplyr::case_when(
                  ranking %in% c("top", "accelerating")    ~ "Top",
                  ranking %in% c("bottom", "decelerating") ~ "Bottom",
                  TRUE                                      ~ tools::toTitleCase(ranking)
                ))

# Fuzzy '/'-synonym merge: two rows merge iff they share at least one trimmed
# '/'-segment of their own Description, within the same direction (never
# across direction -- a top-direction term and a bottom-direction term merging
# would conflate two different biological claims). Connected-components over
# a segment-sharing graph, not exact-string grouping, so "HNF4A" and
# "V$HNF4_01 / HNF4A" merge even though neither string contains the other.
fuzzy_merge_terms <- function(df) {
  if (nrow(df) == 0) return(df %>% dplyr::mutate(.merge_id = integer(0)))
  df <- df %>% dplyr::mutate(.row_id = dplyr::row_number())
  segs <- df %>%
    dplyr::mutate(seg = strsplit(Description, "\\s*/\\s*")) %>%
    tidyr::unnest(seg) %>%
    dplyr::mutate(seg = trimws(tolower(seg))) %>%
    dplyr::filter(nzchar(seg))
  if (nrow(segs) == 0 || dplyr::n_distinct(segs$seg) == nrow(segs)) {
    # Nothing shares a segment with anything else -- every row is its own component.
    return(df %>% dplyr::mutate(.merge_id = .row_id))
  }
  edges <- segs %>%
    dplyr::inner_join(segs, by = "seg", suffix = c("_a", "_b"), relationship = "many-to-many") %>%
    dplyr::filter(.row_id_a < .row_id_b) %>%
    dplyr::distinct(.row_id_a, .row_id_b)
  g <- igraph::graph_from_data_frame(edges, directed = FALSE, vertices = df$.row_id)
  comp <- igraph::components(g)
  df %>% dplyr::mutate(.merge_id = comp$membership[as.character(.row_id)])
}
```

```{r posenrich_dotplot_tabs, eval=has_posenrich_dotplot, results='asis'}
for (dr in c("Global", "Top", "Bottom")) {
  cat(sprintf("\n### %s\n\n", dr))
  dr_df <- posenrich_dotplot_disp %>% dplyr::filter(.direction == dr)
  if (nrow(dr_df) == 0) { cat("_No terms for this direction._\n\n"); next }

  merged <- fuzzy_merge_terms(dr_df) %>%
    dplyr::group_by(.merge_id) %>%
    dplyr::summarise(
      Description   = dplyr::first(Description[order(-signal)]),
      `Merged from`  = paste(unique(sprintf("%s: %s", database, pathway)), collapse = "; "),
      list_threshold = paste(unique(unlist(strsplit(list_threshold, ",\\s*"))), collapse = ", "),
      `n occurrences` = suppressWarnings(max(n_occurrences, na.rm = TRUE)),
      `N merged`      = dplyr::n(),
      `Orthog. signal` = max(signal, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::arrange(dplyr::desc(`Orthog. signal`)) %>%
    dplyr::select(-.merge_id) %>%
    head(params$top_n)

  print(kbl_html(merged, height = if (nrow(merged) > 15) "460px" else NULL))
}
```

---

# Interesting Genes {.tabset .tabset-fade}

One row per gene carrying at least one of the criteria below, faceted
Global/Top/Bottom. Each criterion is a plain boolean (or bool + supporting
list, e.g. which enriched term(s) a gene belongs to) — deliberately not
collapsed into a single opaque score, so every claim in this table is
independently checkable. `FCS term(s)`/`AMI description(s)`/`PE term(s)` show
each term's **description** (not its bare database id), restricted to terms
that module itself calls significant (FCS/RER's own BH+p.perm gate, DOMINO's
own module-significance cutoff — posenrich's leading edge is already
significance-restricted at the source), and collapse behind a `<details>`
toggle when a gene carries more than one. `Composite score` still ranks rows (surprisal-
weighted: rare criteria count for more than common ones, `-log2` of each
criterion's own background rate across **all** genes in this direction, not
just the ones shown here — a criterion's rarity has to be measured against
the true population, not the already-enriched subset that cleared the filter
below) — the 25/10/5/1% percentile quartet is nested (1% ⊂ 5% ⊂ 10% ⊂ 25%), so
each of the four contributes at **1/4 weight** rather than full weight, the
same one-family-shouldn't-quadruple-count idea as the Lancaster discount used
elsewhere in this pipeline, just expressed as a plain fraction here to match
the simpler surprisal-sum mechanism (not a hypergeometric test). 25% is
included alongside 10/5/1% because it's a real, tested tier elsewhere in this
pipeline (posenrich's own Fisher test, AMI's DOMINO module search both run at
25%), not an arbitrary extra cutoff.

```{r interesting_shared_helpers}
# Shared by both Interesting Genes AND Interesting Positions below -- defined
# unconditionally (not gated on has_fcs_stats_cmp/has_position_scores_cmp)
# because either table's own gate can be TRUE independently of the other's,
# and build_interesting_positions() below calls these same helpers. Gating
# them behind only one of the two flags would leave them undefined whenever
# that flag was FALSE but the other (needing them) was TRUE.

# Dependency-free HTML escaping (avoids requiring htmltools be attached) and
# a collapsible-<details> renderer for semicolon-joined multi-term cells --
# any kable() call including a column built by .collapsible_list() MUST be
# passed escape = FALSE, or the markup renders as literal text.
.html_escape <- function(x) {
  x <- gsub("&", "&amp;", x, fixed = TRUE)
  x <- gsub("<", "&lt;",  x, fixed = TRUE)
  gsub(">", "&gt;", x, fixed = TRUE)
}
.collapsible_list <- function(joined, sep = "; ") {
  vapply(joined, function(x) {
    if (is.na(x) || !nzchar(x)) return("")
    items <- strsplit(x, sep, fixed = TRUE)[[1]]
    if (length(items) <= 1) return(.html_escape(x))
    sprintf("<details><summary>%d terms</summary>%s</details>",
            length(items),
            paste(sprintf("<div>%s</div>", .html_escape(items)), collapse = ""))
  }, character(1), USE.NAMES = FALSE)
}

# FCS term membership per gene, direction-matched ranking set -- restricted to
# terms FCS/RER itself calls significant (fcs_sig_terms, cross_angle_gene_hits
# chunk above: BH < fdr_thr AND p.perm < pperm_thr when present), using each
# term's description (not the bare pathway id). `le` alone is NOT already
# significance-restricted (12.FCS_general_report.Rmd's leading-edge export
# only requires drv_score > 0) -- the inner_join against fcs_sig_terms is
# what actually restricts this to genuinely enriched terms.
.fcs_terms_for <- function(rk_set) {
  if (!has_le || !has_multi) return(tibble::tibble(gene = character(), fcs_terms = character()))
  le %>%
    dplyr::filter(is.null(rk_set) | ranking %in% rk_set) %>%
    dplyr::mutate(direction = dplyr::case_when(
      ranking %in% c("top", "accelerating")    ~ "top",
      ranking %in% c("bottom", "decelerating") ~ "bottom",
      ranking == "global"                      ~ "global",
      TRUE                                      ~ NA_character_
    )) %>%
    dplyr::filter(!is.na(direction)) %>%
    dplyr::inner_join(fcs_sig_terms, by = c("module", "direction", "database", "pathway")) %>%
    dplyr::group_by(gene) %>%
    dplyr::summarise(fcs_terms = paste(sort(unique(stringr::str_trunc(description, 50))), collapse = "; "),
                      .groups = "drop")
}
# AMI module descriptions per gene, direction-matched -- restricted to
# DOMINO-significant modules (domino_p_adj < domino_module_thr, same gate
# ami_genes above applies), using the COLLAPSED/clustered description
# (ami_disp$term, task: AMI collapsed rollup above) rather than each module's
# raw near-duplicate-worded description, and tagged with which direction and
# gene-list threshold (e.g. top10, bottom5) it was found at.
.ami_desc_for <- function(dir_filter) {
  if (!has_ami) return(tibble::tibble(gene = character(), ami_desc = character()))
  ami_disp %>%
    dplyr::filter(is.null(dir_filter) | direction %in% dir_filter,
                  !is.na(domino_p_adj), domino_p_adj < params$domino_module_thr) %>%
    dplyr::mutate(gene = strsplit(.data[[.ami_gene_col]], ",\\s*"),
                  hit_label = sprintf("%s (%s: %s)", term, direction, gene_list)) %>%
    tidyr::unnest(gene) %>%
    dplyr::mutate(gene = trimws(gene)) %>%
    dplyr::group_by(gene) %>%
    dplyr::summarise(ami_desc = paste(unique(hit_label), collapse = "; "), .groups = "drop")
}
# Posenrich significant-term positions per gene, direction-matched ranking
# set. posenrich_leading_edge.tsv is ALREADY significance-restricted by
# posenrich_enrich.py itself (only sig rows get written), so no additional
# filter is needed here -- just resolving each term's description (not the
# bare pathway id) via posenrich's own dotplot table when available.
.posenrich_for <- function(rk_set) {
  if (!has_posenrich_le) return(tibble::tibble(gene = character(), pe_position = character(), pe_terms = character()))
  pe <- posenrich_le_df %>%
    dplyr::filter(is.null(rk_set) | ranking %in% rk_set) %>%
    dplyr::mutate(Position = sub(".*:", "", gene_position))
  if (has_posenrich_dotplot) {
    pe <- pe %>%
      dplyr::left_join(
        posenrich_dotplot_df %>% dplyr::mutate(description = dplyr::coalesce(description, pathway)) %>%
          dplyr::distinct(ranking, database, pathway, description),
        by = c("ranking", "database", "pathway")
      ) %>%
      dplyr::mutate(description = dplyr::coalesce(description, pathway))
  } else {
    pe$description <- pe$pathway
  }
  pe %>%
    dplyr::group_by(gene) %>%
    dplyr::summarise(pe_position = paste(sort(unique(Position)), collapse = ", "),
                      pe_terms    = paste(sort(unique(stringr::str_trunc(description, 50))), collapse = "; "),
                      .groups = "drop")
}

# Position-level percentile membership, from SCORING's published
# position_lists/slice_<dir><pct>.tsv (scoring_compute.R) when
# params$position_lists_dir is supplied and all 4 files for a direction
# exist -- so this table's 25/10/5/1% agrees, by construction, with whatever
# else in the pipeline reads the same published cutoff, rather than
# re-deriving its own quantile() cut over base$CAAS_score at render time
# (the previous behaviour, kept as the fallback below). dir_key is the
# lowercase direction ("global"/"top"/"bottom"), matching the published
# filenames exactly. Quartet, not a trio -- 25% is a real, tested tier
# elsewhere (posenrich's own Fisher test, AMI's DOMINO search).
.pos_slice_positions <- function(dir_key, pct) {
  if (is.null(params$position_lists_dir)) return(NULL)
  f <- file.path(params$position_lists_dir, sprintf("slice_%s%d.tsv", dir_key, pct))
  if (!file.exists(f)) return(NULL)
  d <- tryCatch(readr::read_tsv(f, show_col_types = FALSE), error = function(e) NULL)
  if (is.null(d) || !all(c("Gene", "Position") %in% names(d))) return(NULL)
  unique(paste(d$Gene, d$Position))
}
.position_pct_flags <- function(base, dir_key) {
  pub25 <- .pos_slice_positions(dir_key, 25)
  pub10 <- .pos_slice_positions(dir_key, 10)
  pub5  <- .pos_slice_positions(dir_key, 5)
  pub1  <- .pos_slice_positions(dir_key, 1)
  if (!is.null(pub25) && !is.null(pub10) && !is.null(pub5) && !is.null(pub1)) {
    ids <- paste(base$Gene, base$Position)
    base %>%
      dplyr::transmute(gene = Gene, Position, CAAS_score,
                        `25%` = ids %in% pub25, `10%` = ids %in% pub10,
                        `5%` = ids %in% pub5, `1%` = ids %in% pub1)
  } else {
    thr <- quantile(base$CAAS_score, c(0.75, 0.90, 0.95, 0.99), na.rm = TRUE)
    base %>%
      dplyr::transmute(gene = Gene, Position, CAAS_score,
                        `25%` = CAAS_score >= thr[1], `10%` = CAAS_score >= thr[2],
                        `5%` = CAAS_score >= thr[3], `1%` = CAAS_score >= thr[4])
  }
}

# Safe column getter: FALSE for every row when the column is absent (older
# fcs_stats.tsv without flag_accum_top/bottom, etc.) rather than erroring --
# `.data[[col]]` throws on a genuinely missing column, it does not return
# NULL for `%||%` to catch.
.gcol <- function(df, col) if (col %in% names(df)) df[[col]] else rep(FALSE, nrow(df))
```

```{r interesting_genes_prep, eval=has_fcs_stats_cmp}
build_interesting_genes <- function(direction) {
  rk_set   <- switch(direction, Top = c("top", "accelerating"), Bottom = c("bottom", "decelerating"), NULL)
  ami_dirf <- switch(direction, Top = "Top", Bottom = "Bottom", NULL)

  base <- gene_attr_df
  if (direction == "Top") {
    base <- base %>% dplyr::transmute(gene, `25%` = pct_top25, `10%` = pct_top10, `5%` = pct_top5, `1%` = pct_top1,
                                       FADE_dir = .gcol(gene_attr_df, "flag_fade_top"),
                                       Accum_dir = .gcol(gene_attr_df, "flag_accum_top"))
  } else if (direction == "Bottom") {
    base <- base %>% dplyr::transmute(gene, `25%` = pct_bottom25, `10%` = pct_bottom10, `5%` = pct_bottom5, `1%` = pct_bottom1,
                                       FADE_dir = .gcol(gene_attr_df, "flag_fade_bottom"),
                                       Accum_dir = .gcol(gene_attr_df, "flag_accum_bottom"))
  } else {
    base <- base %>% dplyr::transmute(
      gene,
      `25%` = dplyr::coalesce(pct_top25, FALSE) | dplyr::coalesce(pct_bottom25, FALSE),
      `10%` = dplyr::coalesce(pct_top10, FALSE) | dplyr::coalesce(pct_bottom10, FALSE),
      `5%`  = dplyr::coalesce(pct_top5,  FALSE) | dplyr::coalesce(pct_bottom5,  FALSE),
      `1%`  = dplyr::coalesce(pct_top1,  FALSE) | dplyr::coalesce(pct_bottom1,  FALSE),
      FADE_dir  = .gcol(gene_attr_df, "flag_fade"),
      Accum_dir = .gcol(gene_attr_df, "flag_accum"))
  }
  base$RER_acc  <- .gcol(gene_attr_df, "flag_rer_acc")
  base$RER_decc <- .gcol(gene_attr_df, "flag_rer_decc")

  all_genes <- base %>%
    dplyr::left_join(.fcs_terms_for(rk_set), by = "gene") %>%
    dplyr::left_join(.ami_desc_for(ami_dirf), by = "gene") %>%
    dplyr::left_join(.posenrich_for(rk_set), by = "gene") %>%
    dplyr::mutate(
      dplyr::across(c(`25%`, `10%`, `5%`, `1%`, FADE_dir, Accum_dir, RER_acc, RER_decc), ~ dplyr::coalesce(.x, FALSE)),
      FCS_sig  = !is.na(fcs_terms) & nzchar(fcs_terms),
      AMI_net  = !is.na(ami_desc)  & nzchar(ami_desc),
      PE_sig   = !is.na(pe_terms)  & nzchar(pe_terms)
    )

  # Surprisal-weighted composite: percentile quartet at 1/4 weight each
  # (nested, 1% subset of 5% subset of 10% subset of 25% -- see the section
  # text above), every other criterion at full weight. 25% is included
  # alongside 10/5/1% because it's a real, tested tier elsewhere in this
  # pipeline (posenrich's own Fisher test, AMI's DOMINO search both run at
  # 25%; SCORING now publishes a 25% gene list too), not an arbitrary extra
  # cutoff. Background rates computed from the FULL, UNFILTERED gene pool
  # (all_genes, before restricting to "meets >=1 criterion") -- using the
  # filtered/displayed subset here would inflate every criterion's apparent
  # background rate (since it's already an enriched subpopulation) and
  # understate how rare/surprising each one really is, defeating the whole
  # point of the surprisal weighting.
  crit_full  <- c("FADE_dir", "Accum_dir", "RER_acc", "RER_decc", "FCS_sig", "AMI_net", "PE_sig")
  crit_quart <- c("25%", "10%", "5%", "1%")
  bg_full  <- vapply(crit_full,  function(cl) mean(all_genes[[cl]], na.rm = TRUE), numeric(1))
  bg_quart <- vapply(crit_quart, function(cl) mean(all_genes[[cl]], na.rm = TRUE), numeric(1))

  out <- all_genes %>%
    dplyr::filter(`25%` | `10%` | `5%` | `1%` | FADE_dir | Accum_dir | RER_acc | RER_decc | FCS_sig | AMI_net | PE_sig)
  if (nrow(out) == 0) return(out)

  w_full  <- -log2(pmax(bg_full, 1e-6))
  w_quart <- -log2(pmax(bg_quart, 1e-6)) / 4
  out$`Composite score` <- round(as.numeric(
    as.matrix(out[, crit_full])  %*% w_full +
    as.matrix(out[, crit_quart]) %*% w_quart), 2)
  out$`N criteria` <- rowSums(out[, c(crit_full, crit_quart)])

  # Collapsible <details> rendering for the multi-term text columns -- applied
  # AFTER FCS_sig/AMI_net/PE_sig and the composite score are already computed
  # from the raw semicolon-joined text (the HTML wrapping below must never
  # feed back into those). interesting_genes_tabs below must render this
  # table with escape = FALSE for the markup to actually collapse.
  out <- out %>%
    dplyr::mutate(
      `FCS term(s)`         = .collapsible_list(fcs_terms),
      `AMI description(s)`  = .collapsible_list(ami_desc),
      `PE term(s)`          = .collapsible_list(pe_terms)
    )

  out %>%
    dplyr::arrange(dplyr::desc(`Composite score`)) %>%
    dplyr::select(Gene = gene, `25%`, `10%`, `5%`, `1%`, FADE_dir, RER_acc, RER_decc, Accum_dir,
                  FCS_sig, `FCS term(s)`, AMI_net, `AMI description(s)`,
                  PE_sig, `PE position` = pe_position, `PE term(s)`,
                  `N criteria`, `Composite score`) %>%
    head(params$top_n * 3)
}
```

```{r interesting_genes_tabs, eval=has_fcs_stats_cmp, results='asis'}
for (dr in c("Global", "Top", "Bottom")) {
  cat(sprintf("\n## %s\n\n", dr))
  tbl <- build_interesting_genes(dr)
  if (nrow(tbl) == 0) { cat("_No genes meet any criterion for this direction._\n\n"); next }
  print(kbl_html(tbl, height = if (nrow(tbl) > 15) "480px" else NULL, escape = FALSE))
}
```

`r if (!has_fcs_stats_cmp) "> **fcs_stats_file not supplied — Interesting Genes table skipped.**"`

---

# Interesting Positions {.tabset .tabset-fade}

One row per position carrying at least one of the criteria below, faceted
Global/Top/Bottom. Same philosophy as Interesting Genes, but at the position's
own native resolution — each criterion is either **position-native** (this
exact position: CAAS percentile, FADE-site overlap, posenrich significant-term
membership) or **gene-inherited** (every position in a flagged gene carries
the same value: RER acceleration/deceleration, Accumulation significance, FCS
gene-set membership, AMI module membership). That distinction matters — a
gene can be FADE-significant overall while this specific position isn't the
site driving it, so gene-level and position-level FADE overlap are genuinely
different claims, not duplicated information. `COSMIC`/`PAI3D_pathogenic` are
**EXTRAS**: shown for characterisation only, excluded from `Composite score`
(they describe a different kind of evidence — external database overlap
rather than corroboration from this pipeline's own modules — mixing them into
the same weighted sum would conflate the two). Both are reported as 3-state
tags rather than TRUE/FALSE, so a real negative is never conflated with "never
checked": `PAI3D_pathogenic` is `pathogenic` / `benign` / `missing`; `COSMIC`
is `hit` / `no_hit` / `missing` (COSMIC's own file only ever lists positive
hits, so "checked, no record" is inferred from PrimateAI-3D's exhaustive
per-position coverage as the "was this position checkable at all" reference).

```{r interesting_positions_prep, eval=has_position_scores_cmp}
build_interesting_positions <- function(direction) {
  rk_set    <- switch(direction, Top = c("top", "accelerating"), Bottom = c("bottom", "decelerating"), NULL)
  side_vals <- switch(direction, Top = c("top", "both"), Bottom = c("bottom", "both"), NULL)

  base <- position_scores_cmp_df
  if (!is.null(side_vals)) base <- base %>% dplyr::filter(change_side %in% side_vals)
  if (nrow(base) == 0) return(base)

  base <- .position_pct_flags(base, tolower(direction))

  # Position-native FADE-site overlap.
  fade_df <- switch(direction,
    Top = fade_sites_top_cmp_df, Bottom = fade_sites_bottom_cmp_df,
    dplyr::bind_rows(fade_sites_top_cmp_df, fade_sites_bottom_cmp_df))
  base$FADE_overlap <- if (!is.null(fade_df)) {
    paste(base$gene, base$Position) %in% paste(fade_df$gene, fade_df$Position)
  } else FALSE

  # Gene-inherited flags (every position in a flagged gene shares its value).
  # gene_attr_df can be NULL here even when has_position_scores_cmp is TRUE --
  # the two gates (fcs_stats_file vs position_scores_file) are independent, so
  # this table must degrade gracefully (RER/Accum flags just come back FALSE
  # via .gcol) rather than assume fcs_stats_file was also supplied.
  if (!is.null(gene_attr_df)) {
    base <- base %>%
      dplyr::left_join(gene_attr_df %>% dplyr::select(gene, dplyr::any_of(c(
        "flag_rer_acc", "flag_rer_decc", "flag_accum_top", "flag_accum_bottom", "flag_accum"))),
        by = "gene")
  }
  base$RER_acc  <- .gcol(base, "flag_rer_acc")
  base$RER_decc <- .gcol(base, "flag_rer_decc")
  base$Accum_dir <- switch(direction,
    Top = .gcol(base, "flag_accum_top"), Bottom = .gcol(base, "flag_accum_bottom"), .gcol(base, "flag_accum"))

  ami_dirf <- switch(direction, Top = "Top", Bottom = "Bottom", NULL)
  base <- base %>%
    dplyr::left_join(.fcs_terms_for(rk_set), by = "gene") %>%
    dplyr::left_join(.ami_desc_for(ami_dirf), by = "gene") %>%
    dplyr::mutate(FCS_sig = !is.na(fcs_terms) & nzchar(fcs_terms),
                  AMI_net = !is.na(ami_desc)  & nzchar(ami_desc))

  # Position-native posenrich significance: exact Gene:Position match, not
  # just gene -- a gene can have OTHER significant positions without this one
  # being among them.
  if (has_posenrich_le) {
    pe <- posenrich_le_df %>%
      dplyr::filter(is.null(rk_set) | ranking %in% rk_set) %>%
      dplyr::mutate(gene = sub(":.*", "", gene_position), Position = suppressWarnings(as.integer(sub(".*:", "", gene_position)))) %>%
      dplyr::group_by(gene, Position) %>%
      dplyr::summarise(pe_terms = paste(sort(unique(pathway)), collapse = "; "), .groups = "drop")
    base <- base %>% dplyr::left_join(pe, by = c("gene", "Position"))
  } else {
    base$pe_terms <- NA_character_
  }
  base$PE_sig <- !is.na(base$pe_terms) & nzchar(base$pe_terms)

  # EXTRAS -- characterisation only, never in the score (see section text).
  # Both reported as 3-state tags (not TRUE/FALSE) so "checked, negative" is
  # never conflated with "never checked":
  #   PAI3D_pathogenic: "pathogenic" / "benign" / "missing"
  #   COSMIC:           "hit" / "no_hit" / "missing"

  # PrimateAI-3D's own table is exhaustive (one row per position it scored,
  # `prediction` always benign/pathogenic) -- a position simply absent from it
  # was never scored by PrimateAI at all, a genuine "missing", not a benign
  # call. Also doubles as the "was this position even checkable" reference for
  # COSMIC below (COSMIC's own file is hits-only -- see there).
  .pai_pos <- NULL
  if (has_vep_primateai_cmp) {
    .pai <- vep_primateai_cmp_df
    if ("caap_group" %in% names(.pai)) .pai <- dplyr::filter(.pai, caap_group == "US")
    .pred_col <- intersect(c("prediction"), names(.pai))[1]
    if (!is.na(.pred_col)) {
      .pai_pos <- .pai %>%
        dplyr::mutate(.patho = grepl("pathogenic", .data[[.pred_col]], ignore.case = TRUE)) %>%
        dplyr::group_by(Gene, Position) %>% dplyr::summarise(.patho = any(.patho), .groups = "drop")
    }
  }
  base$PAI3D_pathogenic <- if (!is.null(.pai_pos)) {
    .m <- .pai_pos[match(paste(base$gene, base$Position), paste(.pai_pos$Gene, .pai_pos$Position)), ]
    dplyr::case_when(is.na(.m$.patho) ~ "missing", .m$.patho ~ "pathogenic", TRUE ~ "benign")
  } else rep("missing", nrow(base))

  # COSMIC's own file (cosmic_scores.tsv) is HITS-ONLY -- every row IS a
  # documented COSMIC mutation, unlike PrimateAI's exhaustive per-position
  # table. A position absent from it could mean either "checked, no COSMIC
  # record" (a real negative) or "never genomically checkable at all" (e.g.
  # liftover/coordinate mapping failed) -- the file alone can't tell the two
  # apart. PrimateAI-3D's own position set is used as the "was this position
  # checkable" reference instead (verified against real data: 116/118 COSMIC
  # hits also appear in PrimateAI's set, i.e. the two databases draw from
  # essentially the same genomically-mapped position universe).
  base$COSMIC <- if (has_vep_cosmic_cmp) {
    .cosmic <- if ("caap_group" %in% names(vep_cosmic_cmp_df)) dplyr::filter(vep_cosmic_cmp_df, caap_group == "US") else vep_cosmic_cmp_df
    .cosmic_hit <- paste(base$gene, base$Position) %in% paste(.cosmic$Gene, .cosmic$Position)
    .checkable <- if (!is.null(.pai_pos)) {
      paste(base$gene, base$Position) %in% paste(.pai_pos$Gene, .pai_pos$Position)
    } else rep(TRUE, nrow(base))  # no PrimateAI reference available -- can't distinguish no_hit from missing
    dplyr::case_when(.cosmic_hit ~ "hit", .checkable ~ "no_hit", TRUE ~ "missing")
  } else rep("missing", nrow(base))

  all_positions <- base %>%
    dplyr::mutate(dplyr::across(c(`25%`, `10%`, `5%`, `1%`, FADE_overlap, RER_acc, RER_decc, Accum_dir, FCS_sig, AMI_net, PE_sig),
                                 ~ dplyr::coalesce(.x, FALSE)))

  # Background rates from the FULL, UNFILTERED position pool for this
  # direction (all_positions) -- see build_interesting_genes() above for why
  # computing them from the filtered/displayed subset would be wrong. Quartet
  # (25/10/5/1%), not a trio -- same reasoning as build_interesting_genes().
  crit_full  <- c("FADE_overlap", "RER_acc", "RER_decc", "Accum_dir", "FCS_sig", "AMI_net", "PE_sig")
  crit_quart <- c("25%", "10%", "5%", "1%")
  bg_full  <- vapply(crit_full,  function(cl) mean(all_positions[[cl]], na.rm = TRUE), numeric(1))
  bg_quart <- vapply(crit_quart, function(cl) mean(all_positions[[cl]], na.rm = TRUE), numeric(1))

  out <- all_positions %>%
    dplyr::filter(`25%` | `10%` | `5%` | `1%` | FADE_overlap | RER_acc | RER_decc | Accum_dir | FCS_sig | AMI_net | PE_sig)
  if (nrow(out) == 0) return(out)

  w_full  <- -log2(pmax(bg_full, 1e-6))
  w_quart <- -log2(pmax(bg_quart, 1e-6)) / 4
  out$`Composite score` <- round(as.numeric(
    as.matrix(out[, crit_full]) %*% w_full + as.matrix(out[, crit_quart]) %*% w_quart), 2)
  out$`N criteria` <- rowSums(out[, c(crit_full, crit_quart)])

  # Collapsible <details> rendering, applied AFTER FCS_sig/AMI_net/PE_sig and
  # the composite score are already computed from the raw text -- see
  # build_interesting_genes() above. interesting_positions_tabs below must
  # render this table with escape = FALSE.
  out <- out %>%
    dplyr::mutate(
      `FCS term(s)`         = .collapsible_list(fcs_terms),
      `AMI description(s)`  = .collapsible_list(ami_desc),
      `PE term(s)`          = .collapsible_list(pe_terms)
    )

  out %>%
    dplyr::arrange(dplyr::desc(`Composite score`)) %>%
    dplyr::select(Gene = gene, Position, `CAAS Score` = CAAS_score, `25%`, `10%`, `5%`, `1%`,
                  FADE_overlap, RER_acc, RER_decc, Accum_dir,
                  FCS_sig, `FCS term(s)`, AMI_net, `AMI description(s)`,
                  PE_sig, `PE term(s)`,
                  COSMIC, PAI3D_pathogenic,
                  `N criteria`, `Composite score`) %>%
    head(params$top_n * 3)
}
```

```{r interesting_positions_tabs, eval=has_position_scores_cmp, results='asis'}
for (dr in c("Global", "Top", "Bottom")) {
  cat(sprintf("\n## %s\n\n", dr))
  tbl <- build_interesting_positions(dr)
  if (nrow(tbl) == 0) { cat("_No positions meet any criterion for this direction._\n\n"); next }
  print(kbl_html(tbl, height = if (nrow(tbl) > 15) "480px" else NULL, escape = FALSE))
}
```

`r if (!has_position_scores_cmp) "> **position_scores_file not supplied — Interesting Positions table skipped.**"`

---

# Integrated gene scorecard {.tabset .tabset-fade}

One row per gene that shows up in at least one of the three independent angles
analysed across this pipeline: **FCS** (leading-edge member of a significant
CAAS/RER gene set, above), **AMI** (member of a described DOMINO module,
[13.AMI_analysis.Rmd](13.AMI_analysis.Rmd)), and **posenrich** (source of an
"interesting" CAAS position, [14.Position_enrichment_report.Rmd](14.Position_enrichment_report.Rmd)).
Ranked by a surprisal-weighted composite of which angles corroborate it — rare
*combinations* outweigh common ones, same philosophy as the per-position/per-gene
composite scores elsewhere in this pipeline (`surprisal_weighted_scores()` in
`orthogonal_score.R`) — not a p-value, a ranking heuristic. `N angles` (the raw
count, out of 3) is kept alongside for a human-readable summary. Each TRUE flag
is paired with a collapsible description cell (`FCS term(s)`/`AMI description(s)`/
`PE term(s)`) — the same significance-restricted, description-based lookups the
Interesting Genes/Positions tables use above, so a reader never has to take a
flag on faith. `Position` shows the actual posenrich-flagged position(s) for
that gene (when it has a posenrich hit) — so a reader can see whether the
specific position driving the posenrich signal is one worth cross-checking
directly in [14.Position_enrichment_report.Rmd](14.Position_enrichment_report.Rmd),
rather than just knowing the gene overlaps in both angles.

```{r scorecard_prep}
# fcs_genes / ami_genes / posenrich_genes are computed once, earlier in the
# document (cross_angle_gene_hits chunk, right after `load`) -- reused as-is
# here (and by the Per-module significance and AMI sections above).
scorecard <- dplyr::full_join(fcs_genes, ami_genes, by = c("gene", "direction")) %>%
  dplyr::full_join(posenrich_genes, by = c("gene", "direction")) %>%
  dplyr::mutate(dplyr::across(c(fcs_hit, ami_hit, posenrich_hit), ~ dplyr::coalesce(.x, FALSE)),
                Position = dplyr::coalesce(Position, "")) %>%
  dplyr::filter(fcs_hit | ami_hit | posenrich_hit)

has_scorecard <- nrow(scorecard) > 0
if (has_scorecard) {
  .sc_bg <- c(fcs_hit = mean(scorecard$fcs_hit), ami_hit = mean(scorecard$ami_hit),
              posenrich_hit = mean(scorecard$posenrich_hit))
  scorecard$`N angles` <- rowSums(scorecard[, c("fcs_hit", "ami_hit", "posenrich_hit")])
  scorecard$`Composite score` <- round(
    surprisal_weighted_scores(scorecard, c("fcs_hit", "ami_hit", "posenrich_hit"), .sc_bg), 2
  )
  scorecard <- scorecard %>%
    dplyr::arrange(direction, dplyr::desc(`Composite score`), dplyr::desc(`N angles`))
}
```

```{r scorecard_tabs, results='asis'}
for (dr in c("Global", "Top", "Bottom")) {
  cat(sprintf("\n## %s\n\n", dr))
  if (!has_scorecard) { cat("_No cross-angle data available (need at least one of FCS leading-edge, AMI module descriptions, or posenrich interesting-positions)._\n\n"); next }
  sub <- scorecard %>% dplyr::filter(direction == dr) %>% head(params$top_n * 3)
  if (nrow(sub) == 0) { cat("_No genes for this direction._\n\n"); next }

  # Same direction-scoped, significance-restricted, description-based helpers
  # as the Interesting Genes/Positions tables above -- so a TRUE flag here is
  # always backed by an actual description, not just a bare boolean.
  rk_set   <- switch(dr, Top = c("top", "accelerating"), Bottom = c("bottom", "decelerating"), NULL)
  ami_dirf <- switch(dr, Top = "Top", Bottom = "Bottom", NULL)
  disp <- sub %>%
    dplyr::left_join(.fcs_terms_for(rk_set), by = "gene") %>%
    dplyr::left_join(.ami_desc_for(ami_dirf), by = "gene") %>%
    dplyr::left_join(.posenrich_for(rk_set) %>% dplyr::select(gene, pe_terms), by = "gene") %>%
    dplyr::mutate(
      `FCS term(s)`        = .collapsible_list(fcs_terms),
      `AMI description(s)` = .collapsible_list(ami_desc),
      `PE term(s)`         = .collapsible_list(pe_terms)
    ) %>%
    dplyr::transmute(Gene = gene, FCS = fcs_hit, `FCS term(s)`,
                      AMI = ami_hit, `AMI description(s)`,
                      Posenrich = posenrich_hit, `PE term(s)`,
                      Position, `N angles`, `Composite score`)
  print(kbl_html(disp, height = if (nrow(disp) > 15) "480px" else NULL, escape = FALSE))
}
```

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — Enrichment Module*
*Date of generation: `r Sys.Date()`*
