---
title: "Position-level Enrichment (Fisher, fixed cutoffs) — `r params$traitname`"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
    mathjax: null                       # no math in this report
    md_extensions: "-tex_math_dollars"  # pathway names contain '$' (e.g. V$HNF4_01) — never parse as TeX math
params:
  results_file: NULL
  leading_edge_file: NULL
  traitname: "trait"
  padj_thr: 0.15
  fold_thr: 1.5
  table_top_n: 200
  # Position Characterisation (moved here from 11.Scoring_report.Rmd): PrimateAI-3D
  # + COSMIC validation and regional accumulation, all restricted to the US scheme.
  # NULL/absent inputs degrade gracefully (has_pos_char is FALSE, section skipped).
  position_scores_file: NULL   # position_scores.tsv (per-position CAAS scores)
  gene_scores_file:     NULL   # gene_scores.tsv (background gene fallback)
  vep_primateai_file:   NULL   # PrimateAI-3D score TSV
  vep_cosmic_file:      NULL   # COSMIC Mutant Census score TSV
  genomic_info_file:    NULL   # gene, chr, start, end, length (no header assumed)
  top_pct:      0.10
  window_size_bp: 1000000
  output_dir:   "."            # regional_accumulation.tsv export target
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6, dev = "png", dpi = 150)
suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
  library(kableExtra)
  library(DT)
  library(scales)
})

PADJ <- params$padj_thr
FOLD <- params$fold_thr
TOP_N <- params$table_top_n

kbl_html <- function(df, caption = NULL) {
  kable(df, format = "html", escape = TRUE, caption = caption) %>%
    kable_styling(bootstrap_options = c("striped", "condensed", "hover"),
                  full_width = FALSE)
}
```

```{r load, include=FALSE}
stopifnot(!is.null(params$results_file), file.exists(params$results_file))
res <- readr::read_tsv(params$results_file, show_col_types = FALSE)
# Drop 100%/50% cutoffs unconditionally -- retired from the analysis (100% is
# a no-op foreground = full background, 50% is too coarse to be informative)
# but older results_file/leading_edge_file runs may still carry them.
if (!is.null(res) && "top_frac" %in% names(res)) {
  res <- res %>% dplyr::filter(!top_frac %in% c(1.0, 0.5))
}
has_res <- !is.null(res) && nrow(res) > 0
if (has_res && !"sig" %in% names(res)) {
  # No fold-enrichment gate by design — p.adj alone decides significance;
  # fold_enrichment stays available as a displayed column to filter on
  # downstream rather than being baked into what counts as "significant".
  res$sig <- res$p_adj < PADJ
}

# Truncates a "gene1, gene2, ..." string to the first n_show entries, with
# the remainder tucked behind an inline <details>/<summary> toggle -- a row
# can otherwise carry hundreds of comma-separated gene:positions.
truncate_gene_list <- function(x, n_show = 5) {
  vapply(x, function(s) {
    if (is.na(s) || !nzchar(s)) return("")
    items <- strsplit(s, ", ", fixed = TRUE)[[1]]
    if (length(items) <= n_show) return(s)
    shown <- paste(items[seq_len(n_show)], collapse = ", ")
    rest  <- paste(items[(n_show + 1):length(items)], collapse = ", ")
    sprintf(
      '%s <details style="display:inline"><summary style="display:inline; cursor:pointer; color:#2166ac;">&hellip; (%d more)</summary> %s</details>',
      shown, length(items) - n_show, rest
    )
  }, character(1), USE.NAMES = FALSE)
}

# Observed gene:position members of the overlap, significant terms only
# (posenrich_leading_edge.tsv — long format, one row per member). Optional.
le_raw <- NULL
if (!is.null(params$leading_edge_file) && file.exists(params$leading_edge_file)) {
  le_raw <- tryCatch(readr::read_tsv(params$leading_edge_file, show_col_types = FALSE),
                     error = function(e) NULL)
  if (!is.null(le_raw) && "top_frac" %in% names(le_raw)) {
    le_raw <- le_raw %>% dplyr::filter(!top_frac %in% c(1.0, 0.5))
  }
  if (!is.null(le_raw) && nrow(le_raw) == 0) le_raw <- NULL
}
le_summary <- if (!is.null(le_raw)) {
  le_raw %>%
    dplyr::group_by(ranking, database, pathway, top_frac) %>%
    dplyr::summarise(`Genes/positions` = paste(sort(unique(gene_position)), collapse = ", "),
                     `N genes` = dplyr::n_distinct(gene),
                     .groups = "drop")
} else NULL

# Direction-aware corroboration columns (mirrors 13.FCS_general_report.Rmd's
# cols_by_rk): fade_bottom only under `bottom`, fade_top only under `top` --
# no combined "fade" column under `global` (redundant with _top/_bottom, which
# already cover both directions). Only columns actually present (i.e. an
# --annot-file was supplied to posenrich_enrich.py) are ever shown.
flag_cols_by_rk <- list(
  global = c("pct_gate_sig", "pct_gate_fdr", "pct_rer_acc", "pct_rer_decc", "pct_accum"),
  top    = c("pct_gate_sig", "pct_gate_fdr", "pct_fade_top", "pct_rer_acc", "pct_rer_decc", "pct_accum"),
  bottom = c("pct_gate_sig", "pct_gate_fdr", "pct_fade_bottom", "pct_rer_acc", "pct_rer_decc", "pct_accum")
)
```

# Overview {.tabset .tabset-fade}

This report summarises **position-level enrichment** for `r params$traitname`. Unlike
the gene-level FCS (a Wilcoxon-AUC rank-shift test), the position background is a
sea of invariant columns that can never carry a CAAS change — a rank-shift test
degenerates there. Instead we use **Fisher's exact test** at four fixed
foreground cutoffs (top 25 / 10 / 5 / 1% of scored positions, per direction), run
against every pathway across 9 GMT-style databases plus the broad functional
`characterization` layers (UCR core/flank, FUBAR positive/purifying) — all on
the same mechanism.

A set is flagged **significant** when `p.adj < `r PADJ`` (BH-adjusted per
ranking × database). No fold-enrichment magnitude gate is applied here —
`fold_enrichment` is still reported alongside every result so a magnitude
cutoff can be applied downstream, but it does not affect which sets are
marked significant in this report. Fisher/hypergeometric tests gain enormous
power at this background size (~1.47M positions), so a barely-there deviation
from the null ratio can still reach a "significant" p-value on its own — keep
this in mind when filtering (see Methods).

```{r overview, echo=FALSE, results='asis'}
if (!has_res) {
  cat("No enrichment results were produced.\n")
} else {
  for (rk in unique(res$ranking)) {
    cat(sprintf("\n## %s\n\n", toupper(rk)))
    ov <- res %>% filter(ranking == rk) %>%
      group_by(database) %>%
      summarise(`Sets tested` = n_distinct(pathway),
                `Sig (p.adj)` = n_distinct(pathway[sig]),
                .groups = "drop") %>%
      arrange(database)
    print(kbl_html(ov, caption = sprintf(
      "%s — sets tested and significant at p.adj < %.3g.",
      toupper(rk), PADJ)))
    cat("\n\n")
  }
}
```

# Enriched position sets {.tabset .tabset-fade}

Foreground = the top 25 / 10 / 5 / 1% of scored positions by `CAAS_score` in
that direction (`top`/`bottom` restrict to `change_side` first, matching the
SCORING module's own percentile convention; no `global` ranking — dropped in
favour of top/bottom only, consistent with every other report in this
pipeline). Background = the full honest tested
position pool for every test — never restricted to the scored subset, so
tested-and-negative positions are never silently dropped from the null.
`fold > 1` = enriched, `< 1` = depleted. UCR core and UCR flank (and their
per-orthogroup counterparts) are tested as independent, potentially-overlapping
layers, not a forced partition — nothing here requires them to be disjoint.
When available, `%gate_sig`/`%gate_fdr`/`%fade_*`/`%rer_*`/`%accum` columns
report the percentage of *distinct genes* contributing to the overlap (never
raw positions — one gene must not dominate) that are independently flagged by
SCORING's own significance/FADE/RER/accumulation criteria — a cross-module
corroboration check for whether a hit is broadly supported or isolated.

```{r enrichment-tables, echo=FALSE, results='asis'}
if (!has_res) {
  cat("_No results._\n\n")
} else {
  for (rk in unique(res$ranking)) {
    cat(sprintf("\n## %s {.tabset}\n\n", toupper(rk)))
    rk_df <- res %>% filter(ranking == rk)
    for (db in unique(rk_df$database)) {
      cat(sprintf("\n### %s {.tabset}\n\n", db))
      db_df <- rk_df %>% filter(database == db)

      for (tf in sort(unique(db_df$top_frac), decreasing = TRUE)) {
        cat(sprintf("\n#### Top %s%%\n\n", format(100 * tf, trim = TRUE)))
        tf_all <- db_df %>% filter(top_frac == tf) %>% arrange(desc(sig), p_adj, p_value)
        n_total_tf <- nrow(tf_all)
        tf_src <- tf_all %>% head(TOP_N)
        if (n_total_tf > TOP_N) {
          cat(sprintf("_Showing top %d of %d tested pathways (sorted by significance)._\n\n", TOP_N, n_total_tf))
        }
        tf_df <- tf_src %>%
          transmute(Pathway = pathway, Description = description,
                    `Layer size` = layer_size, `Foreground` = foreground_size,
                    Observed = observed, Expected = round(expected, 1),
                    Fold = round(fold_enrichment, 3), Direction = direction,
                    `p (Fisher)` = signif(p_value, 3),
                    `p.adj` = signif(p_adj, 3), Sig = ifelse(sig, "*", ""))
        # Cross-module corroboration: % of the DISTINCT genes in the overlap
        # (never % of raw positions — one gene must not dominate) carrying
        # each SCORING flag (gate_sig/gate_fdr/fade/rer/accum). Only present
        # when POSENRICH was run with --annot-file.
        flag_cols <- grep("^pct_", names(tf_src), value = TRUE)
        if (length(flag_cols) > 0) {
          flag_df <- tf_src %>% transmute(across(all_of(flag_cols), ~round(.x, 1)))
          names(flag_df) <- paste0("%", sub("^pct_", "", flag_cols))
          tf_df <- dplyr::bind_cols(tf_df, flag_df)
        }
        id <- paste("dt", rk, db, gsub("\\.", "_", as.character(tf)), sep = "_")
        cat(knitr::knit_child(text = c(
          paste0("```{r ", id, ", echo=FALSE}"),
          "DT::datatable(tf_df, options = list(pageLength = 10, scrollX = TRUE), rownames = FALSE)",
          "```"), envir = environment(), quiet = TRUE))
        cat("\n\n")
      }
    }
  }
}
```

# Enrichment dotplots {.tabset .tabset-fade}

Each point is a tested pathway: x = log2 fold enrichment, y = −log10 `p.adj`,
faceted by top-fraction cutoff within each direction × database. Red points
pass the `p.adj` significance threshold (dashed line); the dotted fold lines
are shown for reference only and do not gate significance here.

```{r enrichment-dotplots, echo=FALSE, results='asis', fig.height=5}
if (!has_res) {
  cat("_No results._\n\n")
} else {
  for (rk in unique(res$ranking)) {
    cat(sprintf("\n## %s {.tabset}\n\n", toupper(rk)))
    rk_df <- res %>% filter(ranking == rk)
    for (db in unique(rk_df$database)) {
      cat(sprintf("\n### %s\n\n", db))
      db_df <- rk_df %>% filter(database == db)
      d <- db_df %>% filter(!is.na(fold_enrichment), fold_enrichment > 0) %>%
        mutate(l2f = log2(fold_enrichment), neglogp = -log10(pmax(p_adj, 1e-300)),
               frac_lab = sprintf("Top %s%%", format(100 * top_frac, trim = TRUE)))
      if (nrow(d) == 0) { cat("_No plottable points._\n\n"); next }
      lab <- d %>% filter(sig) %>% arrange(p_adj) %>% group_by(top_frac) %>% slice_head(n = 8) %>% ungroup()
      p <- ggplot(d, aes(l2f, neglogp)) +
        geom_hline(yintercept = -log10(PADJ), linetype = "dashed", colour = "grey50") +
        geom_vline(xintercept = c(-log2(FOLD), log2(FOLD)), linewidth = 0.3, colour = "grey70", linetype = "dotted") +
        geom_point(aes(colour = sig), alpha = 0.75) +
        { if (nrow(lab) > 0) ggrepel::geom_text_repel(data = lab, aes(label = pathway), size = 3, max.overlaps = 12, show.legend = FALSE) } +
        facet_wrap(~frac_lab) +
        scale_colour_manual(values = c(`FALSE` = "grey55", `TRUE` = "#c0392b"),
                            name = sprintf("p.adj<%.2g", PADJ)) +
        labs(x = expression(log[2]~fold), y = expression(-log[10]~p.adj),
             title = sprintf("%s — %s", toupper(rk), db)) +
        theme_minimal(base_size = 12)
      print(p)
      cat("\n\n")
    }
  }
}
```

# Top vs Bottom (log2 fold enrichment)

Mirrors the Comparison report's top-vs-bottom scatter: x = log2 fold enrichment
under the **top** ranking, y = log2 fold enrichment under the **bottom**
ranking, one point per (database, pathway, top-fraction cutoff) — restricted
to the three tightest cutoffs (10/5/1%; 25% dropped here for readability),
coloured by cutoff rather than by direction (direction is already the two
axes). Points off the diagonal favour one direction over the other; points
significant (`p.adj < 0.05`) in either direction are labelled
`description_percentage`, as are terms that recur significantly at **more
than one** of the three cutoffs (every term is tested at all three by
construction, so raw presence isn't informative — recurring *significance*
is; when a term qualifies, all of its cutoff points are labelled together so
the recurring pattern reads as a group, not just its individually-significant
points).

```{r log2fold_top_vs_bottom, echo=FALSE, fig.height=7}
if (!has_res) {
  cat("_No results._\n\n")
} else {
  sc <- res %>%
    filter(ranking %in% c("top", "bottom"), !is.na(fold_enrichment), fold_enrichment > 0,
           top_frac %in% c(0.10, 0.05, 0.01)) %>%
    mutate(l2f = log2(fold_enrichment),
           frac_lab = sprintf("%s%%", format(100 * top_frac, trim = TRUE))) %>%
    tidyr::pivot_wider(
      id_cols = c(database, pathway, description, top_frac, frac_lab),
      names_from = ranking, values_from = c(l2f, p_adj),
      names_sep = "_"
    )

  req_cols <- c("l2f_top", "l2f_bottom", "p_adj_top", "p_adj_bottom")
  for (rc in setdiff(req_cols, names(sc))) sc[[rc]] <- NA

  sc <- sc %>%
    mutate(
      l2f_top    = dplyr::coalesce(l2f_top, 0),
      l2f_bottom = dplyr::coalesce(l2f_bottom, 0),
      is_sig     = dplyr::coalesce(p_adj_top < 0.05, FALSE) | dplyr::coalesce(p_adj_bottom < 0.05, FALSE),
      highlight_tag = sprintf("%s_%s", description, frac_lab)
    ) %>%
    # Every term is tested at all three cutoffs by construction, so raw
    # presence carries no information -- recurring *significance* does.
    # A term crossing p.adj < 0.05 at >=2 of the three cutoffs gets ALL of
    # its points labelled (not just the individually-significant ones), so
    # the recurring pattern is visible as a connected group in the plot.
    group_by(database, pathway) %>%
    mutate(n_sig_fracs = sum(is_sig, na.rm = TRUE)) %>%
    ungroup() %>%
    mutate(highlight = is_sig | n_sig_fracs > 1)

  if (nrow(sc) == 0) {
    cat("_No plottable points._\n\n")
  } else {
    p <- ggplot(sc, aes(l2f_top, l2f_bottom, colour = frac_lab)) +
      geom_hline(yintercept = 0, linetype = "dashed", colour = "grey70") +
      geom_vline(xintercept = 0, linetype = "dashed", colour = "grey70") +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
      geom_point(alpha = 0.6, size = 1.8) +
      scale_colour_manual(name = "Top fraction",
                           # viridis' 3-stop scale with its yellow endpoint (1%)
                           # swapped for orange -- yellow is near-invisible
                           # against the white panel background here.
                           values = c(`10%` = "#440154", `5%` = "#21908C", `1%` = "#E67E22"),
                           limits = c("10%", "5%", "1%")) +
      labs(title = sprintf("Position enrichment — top vs bottom, 10/5/1%% cutoffs — %s", params$traitname),
           x = expression(log[2]~fold~enrichment~(top)),
           y = expression(log[2]~fold~enrichment~(bottom))) +
      theme_minimal(base_size = 12)

    lab <- sc %>% filter(highlight)
    if (nrow(lab) > 0) {
      # Larger labels need more clearance to avoid (a) overlapping each other
      # and (b) sitting on top of dense point clusters -- higher force/padding
      # and more iterations give the algorithm room to push labels clear of
      # both, not just of the point they're anchored to.
      p <- p + ggrepel::geom_text_repel(
        data = lab, aes(label = stringr::str_trunc(highlight_tag, 45)),
        size = 3.6, fontface = "bold", max.overlaps = Inf, seed = 42,
        box.padding = 0.6, point.padding = 0.4,
        force = 4, force_pull = 0.5,
        max.time = 3, max.iter = 100000,
        min.segment.length = 0, segment.size = 0.3, segment.alpha = 0.6,
        bg.color = "white", bg.r = 0.12, show.legend = FALSE)
    }
    print(p)
  }
}
```

# Leading edge {.tabset .tabset-fade}

For every **significant** (`p.adj < `r PADJ``) pathway, the
observed `gene:position` members of the overlap, plus the same cross-module
corroboration columns as above (`%gate_sig`/`%gate_fdr`/`%fade_*`/`%rer_*`/
`%accum`, direction-aware — `%fade_bottom` only appears under `BOTTOM`,
`%fade_top` only under `TOP`, plain `%fade` under `GLOBAL`). The full member
list is also written to `posenrich_leading_edge.tsv` alongside this report.

```{r leading-edge, echo=FALSE, results='asis'}
if (!has_res || !any(res$sig, na.rm = TRUE)) {
  cat("_No significant sets._\n\n")
} else {
  sig_res <- res %>% filter(sig)
  for (rk in unique(sig_res$ranking)) {
    cat(sprintf("\n## %s\n\n", toupper(rk)))
    rk_sig <- sig_res %>% filter(ranking == rk) %>% arrange(database, p_adj)
    if (nrow(rk_sig) == 0) { cat("_No significant sets for this direction._\n\n"); next }

    want <- flag_cols_by_rk[[rk]]
    if (is.null(want)) want <- character(0)
    present_flags <- intersect(want, names(rk_sig))

    if (!is.null(le_summary)) {
      rk_sig <- rk_sig %>%
        dplyr::left_join(le_summary, by = c("ranking", "database", "pathway", "top_frac"))
    }

    le_df <- rk_sig %>%
      transmute(Database = database, Pathway = pathway, Description = description,
                `Top frac` = sprintf("%s%%", format(100 * top_frac, trim = TRUE)),
                Fold = round(fold_enrichment, 3), Direction = direction,
                `p.adj` = signif(p_adj, 3))
    if (length(present_flags) > 0) {
      flag_df <- rk_sig %>% transmute(across(all_of(present_flags), ~round(.x, 1)))
      names(flag_df) <- paste0("%", sub("^pct_", "", present_flags))
      le_df <- dplyr::bind_cols(le_df, flag_df)
    }
    esc_cols <- TRUE
    if ("Genes/positions" %in% names(rk_sig)) {
      le_df <- dplyr::bind_cols(le_df, rk_sig %>% dplyr::select(`N genes`, `Genes/positions`))
      # Full member list is still in posenrich_leading_edge.tsv -- here, show
      # only the first few inline with a <details> toggle for the rest, since
      # a row can otherwise carry hundreds of comma-separated gene:positions.
      le_df[["Genes/positions"]] <- truncate_gene_list(le_df[["Genes/positions"]])
      # DT column indices for `escape` are 1-based over the data columns when
      # rownames = FALSE (no offset needed) -- verified empirically, since
      # DT's own docs are ambiguous on this point and getting it wrong throws
      # "attempt to select less than one element in integerOneIndex" deep
      # inside DT:::escapeData(), aborting the whole render.
      esc_cols <- which(names(le_df) != "Genes/positions")
    }

    id <- paste("dt_le", rk, sep = "_")
    assign(paste0(".esc_cols_", rk), esc_cols, envir = environment())
    cat(knitr::knit_child(text = c(
      paste0("```{r ", id, ", echo=FALSE}"),
      sprintf("DT::datatable(le_df, options = list(pageLength = 10, scrollX = TRUE), rownames = FALSE, escape = .esc_cols_%s)", rk),
      "```"), envir = environment(), quiet = TRUE))
    cat("\n\n")
  }
}
```

---

# Position Characterisation

Moved here from the Scoring report (was a companion-report cross-reference;
now lives directly alongside the position-level enrichment analysis it
complements). Logic is unchanged from the original — same US-scheme
restriction, same CAAS-score-threshold framing.

```{r poschar-setup, include=FALSE}
.file_ok <- function(p) !is.null(p) && !is.na(p) && nchar(p) > 0 && file.exists(p)

# Tableau_10 colour swatch for pipeline-wide consistency (same palette used
# throughout the Scoring/Characterization reports).
if (requireNamespace("paletteer", quietly = TRUE)) {
  .tableau10 <- as.character(paletteer::paletteer_d("ggthemes::Tableau_10"))
} else {
  .tableau10 <- c(
    "#4E79A7", "#F28E2B", "#E15759", "#76B7B2", "#59A14F",
    "#EDC948", "#B07AA1", "#FF9DA7", "#9C755F", "#BAB0AC"
  )
}

has_position_scores <- .file_ok(params$position_scores_file)
has_gene_scores      <- .file_ok(params$gene_scores_file)
pos_df  <- if (has_position_scores) readr::read_tsv(params$position_scores_file, show_col_types = FALSE) else NULL
gene_df <- if (has_gene_scores)     readr::read_tsv(params$gene_scores_file,     show_col_types = FALSE) else NULL

has_vep_primateai <- .file_ok(params$vep_primateai_file)
has_vep_cosmic    <- .file_ok(params$vep_cosmic_file)
has_genomic       <- .file_ok(params$genomic_info_file)
# Position scores are required for every sub-tab below (PAI3D/COSMIC joins,
# regional accumulation) -- gate the whole section on them, not just on which
# external database happens to be supplied.
has_pos_char <- has_position_scores && (has_vep_primateai || has_vep_cosmic || has_genomic)

if (has_vep_primateai) {
  vep_pai_df <- readr::read_tsv(params$vep_primateai_file, show_col_types = FALSE)
  if ("Position" %in% names(vep_pai_df)) vep_pai_df$Position <- suppressWarnings(as.integer(vep_pai_df$Position))
} else {
  vep_pai_df <- NULL
}

if (has_vep_cosmic) {
  vep_cosmic_df <- readr::read_tsv(params$vep_cosmic_file, show_col_types = FALSE)
  if ("Position" %in% names(vep_cosmic_df)) vep_cosmic_df$Position <- suppressWarnings(as.integer(vep_cosmic_df$Position))
} else {
  vep_cosmic_df <- NULL
}

if (has_genomic) {
  # Format: gene, chr, start, end, length (no header assumed; detect if present)
  genomic_raw <- readr::read_tsv(params$genomic_info_file,
                          col_names = c("Gene", "chr", "gene_start", "gene_end", "gene_length"),
                          show_col_types = FALSE)
  if (tolower(genomic_raw$Gene[1]) == "gene") genomic_raw <- genomic_raw[-1, ]
  genomic_raw <- genomic_raw %>%
    dplyr::mutate(
      gene_start  = as.integer(gene_start),
      gene_end    = as.integer(gene_end),
      gene_length = as.integer(gene_length),
      gene_mid    = as.integer((gene_start + gene_end) / 2L)
    )
} else {
  genomic_raw <- NULL
}

if (has_pos_char) {
  top_pct  <- params$top_pct %||% 0.10
  top5_pct <- 0.05
  top1_pct <- 0.01

  pos_top    <- pos_df %>% dplyr::filter(change_side %in% c("top",    "both"))
  pos_bottom <- pos_df %>% dplyr::filter(change_side %in% c("bottom", "both"))

  # Global thresholds (used throughout this section for CAAS filters)
  thr_10 <- quantile(pos_df$CAAS_score, 1 - top_pct,  na.rm = TRUE)
  thr_5  <- quantile(pos_df$CAAS_score, 1 - top5_pct, na.rm = TRUE)
  thr_1  <- quantile(pos_df$CAAS_score, 1 - top1_pct, na.rm = TRUE)

  # Direction-specific thresholds (used for is_Xpct flags)
  .thr_top_10 <- quantile(pos_top$CAAS_score, 1 - top_pct,  na.rm = TRUE)
  .thr_top_5  <- quantile(pos_top$CAAS_score, 1 - top5_pct, na.rm = TRUE)
  .thr_top_1  <- quantile(pos_top$CAAS_score, 1 - top1_pct, na.rm = TRUE)
  .thr_bot_10 <- quantile(pos_bottom$CAAS_score, 1 - top_pct,  na.rm = TRUE)
  .thr_bot_5  <- quantile(pos_bottom$CAAS_score, 1 - top5_pct, na.rm = TRUE)
  .thr_bot_1  <- quantile(pos_bottom$CAAS_score, 1 - top1_pct, na.rm = TRUE)
}
```
---

```{r pos-char-header, results='asis'}
if (has_pos_char) cat(r"(## Position Characterisation {.tabset #position-characterisation}

This section examines how ranked positions behave in relation to external
signals that do **not** contribute to the `CAAS_score`. Both PrimateAI-3D and
COSMIC Mutant Census are treated as independent characterisation layers.

Since both databases describe specific amino acid transitions, the biochemical
Grantham Score (GS) groupings are out of scope (as they group transitions by
biochemical radicalness rather than modeling the specific substitution directly).
Therefore, we restrict the characterisation strictly to the **Unweighted Scheme (US)**,
with all scores and mutation overlaps derived solely from US transitions.

- **`pai_char_score`**: The pathogenicity score of mutating the ancestral human reference amino acid to the derived US alternative amino acid(s) in PrimateAI-3D.
- **COSMIC Overlap**: The presence and recurrence of somatic mutations mapped to the specific codon under the US transition.

)")
```

```{r vep-setup-common}
pos_score_meta <- pos_df %>%
  select(Gene, Position, CAAS_score, biochem_weight_sum, change_side, gate_sig, gate_fdr) %>%
  mutate(
    caas_decile  = dplyr::ntile(CAAS_score, 10),
    in_top_10pct = CAAS_score >= thr_10,
    in_top_5pct  = CAAS_score >= thr_5,
    in_top_1pct  = CAAS_score >= thr_1
  )

has_background <- .file_ok(params$background_file)
if (has_background) {
  bg_lines <- readLines(params$background_file)
  bg_genes <- bg_lines[nzchar(trimws(bg_lines)) & !grepl("^#", bg_lines)]
} else {
  bg_genes <- unique(gene_df$Gene)
}
```

```{r vep-pai-subsec, results='asis'}
if (has_vep_primateai) cat("### VEP: PrimateAI-3D\n\n")
```

```{r vep-pai-setup, eval=has_vep_primateai}
# Detect PrimateAI score column (varies by database version)
.pai_score_col <- intersect(
  c("score_PAI3D", "score", "primateai_score", "pathogenicity_score", "PrimateAI_score"),
  names(vep_pai_df)
)[1]
if (is.na(.pai_score_col)) {
  warning("PrimateAI: no recognised score column found — skipping.")
  has_vep_primateai <- FALSE
}
```

```{r vep-pai-join, eval=has_vep_primateai}
# Restrict to shared genes present in both our results (background clean) and database
shared_genes_pai <- intersect(bg_genes, unique(vep_pai_df$Gene))
pos_pai_shared <- pos_score_meta %>% filter(Gene %in% shared_genes_pai)

# Under US-only design, pai_char_score is simply the pathogenicity score of the US change.
# If multiple alternative AAs map, we keep the maximum score.
# Pathogenic status is defined exclusively by the prediction column (containing
# "pathogenic", case-insensitive) — never by a raw score cutoff.
pai_by_prot <- vep_pai_df %>%
  filter(caap_group == "US", Gene %in% shared_genes_pai) %>%
  mutate(
    pai_score = as.numeric(.data[[.pai_score_col]]),
    is_pathogenic = grepl("pathogenic", prediction, ignore.case = TRUE)
  ) %>%
  filter(is.finite(pai_score)) %>%
  group_by(Gene, Position) %>%
  summarise(
    pai_char_score = max(pai_score, na.rm = TRUE),
    pai_pathogenic = any(is_pathogenic, na.rm = TRUE),
    n_pai_schemes  = 1L,
    .groups = "drop"
  )

pos_vep_pai <- pai_by_prot %>%
  inner_join(pos_pai_shared, by = c("Gene", "Position")) %>%
  distinct(Gene, Position, .keep_all = TRUE)

# Join back caas_change and other details for display table
pos_vep_pai_table_source <- vep_pai_df %>%
  filter(caap_group == "US", Gene %in% shared_genes_pai) %>%
  inner_join(pos_pai_shared, by = c("Gene", "Position")) %>%
  distinct(Gene, Position, .keep_all = TRUE)

has_vep_primateai <- nrow(pos_vep_pai) > 0

# Directional PAI subsets: top+both and bottom+both
pos_vep_pai_top    <- pos_vep_pai %>% filter(change_side %in% c("top",    "both"))
pos_vep_pai_bottom <- pos_vep_pai %>% filter(change_side %in% c("bottom", "both"))
```

`r if (exists("pos_vep_pai") && nrow(pos_vep_pai) == 0) "> **Note:** PrimateAI-3D inputs were provided, but no scored positions could be matched to the VEP/PAI3D annotations for this direction."`

```{r vep-pai-violin, eval=has_vep_primateai, fig.height=7}
# pai_char_score violin per CAAS decile, split by direction (TOP and BOTTOM separately).
pai_dir_df <- bind_rows(
  pos_vep_pai_top    %>% mutate(pai_direction = "TOP (top + both)"),
  pos_vep_pai_bottom %>% mutate(pai_direction = "BOTTOM (bottom + both)")
) %>%
  filter(!is.na(caas_decile), !is.na(pai_char_score))

# Per-direction median for horizontal reference lines
pai_medians <- pai_dir_df %>%
  group_by(pai_direction) %>%
  summarise(median_pai = median(pai_char_score, na.rm = TRUE), .groups = "drop")

ggplot(pai_dir_df, aes(x = factor(caas_decile, levels = 1:10), y = pai_char_score)) +
  geom_violin(aes(fill = pai_direction), alpha = 0.55, trim = TRUE, scale = "width",
              show.legend = FALSE) +
  geom_boxplot(width = 0.14, outlier.size = 0.4, alpha = 0.9,
               fill = "white", colour = "grey30") +
  geom_hline(data = pai_medians,
             aes(yintercept = median_pai),
             linetype = "dashed", colour = "grey20", linewidth = 0.7) +
  geom_text(data = pai_medians,
            aes(x = 10.6, y = median_pai,
                label = sprintf("median = %.3f", median_pai)),
            hjust = 0, vjust = -0.4, size = 3, colour = "grey20",
            inherit.aes = FALSE) +
  scale_fill_manual(values = c("TOP (top + both)" = .tableau10[1],
                               "BOTTOM (bottom + both)" = .tableau10[3])) +
  scale_x_discrete(
    drop   = FALSE,
    labels = setNames(c("1\n(low)", as.character(2:9), "10\n(high)"), as.character(1:10))
  ) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.08))) +
  facet_wrap(~pai_direction, ncol = 1) +
  labs(
    title    = "PAI_TOP and PAI_BOTTOM: PrimateAI-3D Score by CAAS Decile",
    subtitle = sprintf(
      "PAI_TOP: %d positions | PAI_BOTTOM: %d positions | dashed = per-direction median",
      nrow(pos_vep_pai_top), nrow(pos_vep_pai_bottom)
    ),
    x = "CAAS Score Decile", y = "PrimateAI-3D Score"
  ) +
  theme(plot.title = element_text(face = "bold"),
        strip.text = element_text(face = "bold", size = 11))
```

```{r vep-pai-scatter, eval=has_vep_primateai, fig.height=5}
# CAAS_score vs pai_char_score, coloured by direction.
# Normalise both axes for Pearson r (rank / n).
pos_vep_pai_dir <- bind_rows(
  pos_vep_pai_top    %>% mutate(pai_direction = "TOP"),
  pos_vep_pai_bottom %>% mutate(pai_direction = "BOTTOM")
) %>% distinct(Gene, Position, .keep_all = TRUE)

.pai_caas_r_top  <- safe_cor(
  rank(pos_vep_pai_top$CAAS_score,    na.last = "keep") / sum(!is.na(pos_vep_pai_top$CAAS_score)),
  rank(pos_vep_pai_top$pai_char_score, na.last = "keep") / sum(!is.na(pos_vep_pai_top$pai_char_score))
)
.pai_caas_r_bot  <- safe_cor(
  rank(pos_vep_pai_bottom$CAAS_score,    na.last = "keep") / sum(!is.na(pos_vep_pai_bottom$CAAS_score)),
  rank(pos_vep_pai_bottom$pai_char_score, na.last = "keep") / sum(!is.na(pos_vep_pai_bottom$pai_char_score))
)

ggplot(pos_vep_pai_dir, aes(x = CAAS_score, y = pai_char_score,
                             colour = pai_direction)) +
  geom_point(alpha = 0.35, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8, alpha = 0.15) +
  geom_vline(xintercept = thr_10, linetype = "dashed",
             colour = "grey40", linewidth = 0.5) +
  scale_colour_manual(values = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[3]),
                      name = "Direction") +
  labs(
    title    = "CAAS Score vs PrimateAI-3D Score by Direction",
    subtitle = sprintf("Pearson r (rank-norm): TOP = %.3f | BOTTOM = %.3f | dashed = top-10%% CAAS",
                       .pai_caas_r_top, .pai_caas_r_bot),
    x = "CAAS Score", y = "PrimateAI-3D Score"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r vep-pai-pathogenic-bar, eval=has_vep_primateai, fig.height=5}
# Count of pathogenic vs non-pathogenic positions, TOP and BOTTOM.
pai_pathogenic_counts <- bind_rows(
  pos_vep_pai_top    %>% mutate(pai_direction = "TOP"),
  pos_vep_pai_bottom %>% mutate(pai_direction = "BOTTOM")
) %>%
  mutate(pathogenic = ifelse(pai_pathogenic, "pathogenic", "non-pathogenic")) %>%
  count(pai_direction, pathogenic)

ggplot(pai_pathogenic_counts, aes(x = pathogenic, y = n, fill = pai_direction)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = n), position = position_dodge(width = 0.8), vjust = -0.3, size = 3.5) +
  scale_fill_manual(values = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[3])) +
  labs(
    title    = "Pathogenic vs Non-Pathogenic Positions (PrimateAI-3D)",
    subtitle = "Pathogenic defined by prediction column",
    x = NULL, y = "N positions", fill = "Direction"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r vep-pai-phenaxis-scatter, eval=has_vep_primateai && "phen_score" %in% names(pos_vep_pai_top), fig.height=5}
.pai_phen_r_top <- safe_cor(
  rank(pos_vep_pai_top$phen_score,      na.last = "keep") / sum(!is.na(pos_vep_pai_top$phen_score)),
  rank(pos_vep_pai_top$pai_char_score,  na.last = "keep") / sum(!is.na(pos_vep_pai_top$pai_char_score))
)
.pai_phen_r_bot <- safe_cor(
  rank(pos_vep_pai_bottom$phen_score,      na.last = "keep") / sum(!is.na(pos_vep_pai_bottom$phen_score)),
  rank(pos_vep_pai_bottom$pai_char_score,  na.last = "keep") / sum(!is.na(pos_vep_pai_bottom$pai_char_score))
)

ggplot(pos_vep_pai_dir, aes(x = phen_score, y = pai_char_score,
                             colour = pai_direction)) +
  geom_point(alpha = 0.35, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8, alpha = 0.15) +
  scale_colour_manual(values = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[3]),
                      name = "Direction") +
  labs(
    title    = "Phenotype Evolution Score vs PrimateAI-3D Score by Direction",
    subtitle = sprintf("Pearson r (rank-norm): TOP = %.3f | BOTTOM = %.3f",
                       .pai_phen_r_top, .pai_phen_r_bot),
    x = "phen_score (1 - pvalue_boot)", y = "PrimateAI-3D Score"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r vep-pai-table, eval=has_vep_primateai}
pai_table <- pos_vep_pai_table_source %>%
  arrange(desc(CAAS_score)) %>%
  select(Gene, Position, hg38_ref_aa, caas_alt_aas, caas_change, CAAS_score, change_side, score_PAI3D, percentile_PAI3D, prediction) %>%
  rename(
    `Ref AA` = hg38_ref_aa,
    `Alt AA(s)` = caas_alt_aas,
    `CAAS Change` = caas_change,
    `CAAS Score` = CAAS_score,
    `Change Side` = change_side,
    `PAI3D Score` = score_PAI3D,
    `PAI3D Percentile` = percentile_PAI3D,
    `Prediction` = prediction
  )

DT::datatable(
  pai_table,
  caption = "CAAS positions overlapping with PrimateAI-3D pathogenicity predictions, ordered by CAAS score",
  options = list(pageLength = 10, scrollX = TRUE),
  rownames = FALSE
)
```

```{r vep-pai-fisher, eval=has_vep_primateai}
# Odds ratio & P-value interpretation:
# - Odds Ratio: Measures the enrichment of pathogenic residues (by the prediction column) within the specified CAAS threshold compared to the rest of the positions. An odds ratio > 1 indicates enrichment.
# - P-value: Reaches significance (via Fisher's exact test) when the proportion of pathogenic mutations is significantly higher inside the CAAS threshold.
# `N positions` IS the Fisher test's actual population (PAI-scored, shared-gene
# positions) — mirrors run_cosmic_fisher_dir below. Previously this column showed
# a broader shared-gene pool count (via a separate coverage_df) that did not match
# what the test was actually computed over; fixed to report the true population.
run_pai_fisher_dir <- function(df, label) {
  df   <- df %>% mutate(high_pai = !is.na(pai_pathogenic) & pai_pathogenic)
  n_all <- nrow(df)

  # Build named threshold list: CAAS-flag-based thresholds + percentage-based rows
  # Percentage rows use in_top_Xpct flags when available; for 100% and 50% we
  # derive a logical vector from the CAAS score quantile.
  thr_50_val <- quantile(df$CAAS_score, 0.50, na.rm = TRUE)
  pct_rows <- list(
    list(col = NULL, lbl = "100% (all positions)",    flag = rep(TRUE, n_all)),
    list(col = NULL, lbl = "50% (top 50% CAAS)",      flag = df$CAAS_score >= thr_50_val),
    list(col = "in_top_10pct", lbl = sprintf("Top %g%%", top_pct  * 100), flag = NULL),
    list(col = "in_top_5pct",  lbl = sprintf("Top %g%%", top5_pct * 100), flag = NULL),
    list(col = "in_top_1pct",  lbl = sprintf("Top %g%%", top1_pct * 100), flag = NULL)
  )

  purrr::map_dfr(pct_rows, function(thr) {
    in_top   <- if (!is.null(thr$flag)) thr$flag else df[[thr$col]]
    in_top   <- replace_na(in_top, FALSE)

    n_top    <- sum(in_top)
    n_top_h  <- sum(in_top & df$high_pai, na.rm = TRUE)
    n_rest   <- sum(!in_top)
    n_rest_h <- sum(!in_top & df$high_pai, na.rm = TRUE)
    mat <- matrix(c(n_top - n_top_h, n_rest - n_rest_h, n_top_h, n_rest_h), nrow = 2)
    ft  <- tryCatch(fisher.test(mat),
                    error = function(e) list(estimate = NA, conf.int = c(NA, NA), p.value = NA))

    pct_sig <- if ("gate_sig" %in% names(df)) sprintf("%.1f%%", 100 * mean(df$gate_sig[in_top], na.rm = TRUE)) else "0.0%"
    pct_fdr <- if ("gate_fdr" %in% names(df)) sprintf("%.1f%%", 100 * mean(df$gate_fdr[in_top], na.rm = TRUE)) else "0.0%"

    tibble(
      Direction            = label,
      Threshold            = thr$lbl,
      `N positions`        = n_top,
      `N pathogenic`       = n_top_h,
      `% pathogenic`       = sprintf("%.1f%%", 100 * n_top_h / max(n_top, 1)),
      `Odds ratio`         = sprintf("%.2f", ft$estimate),
      `P value`            = formatC(ft$p.value, format = "e", digits = 2)
    )
  })
}

pai_fisher <- bind_rows(
  run_pai_fisher_dir(pos_vep_pai_top,    "TOP"),
  run_pai_fisher_dir(pos_vep_pai_bottom, "BOTTOM")
)

knitr::kable(
  pai_fisher,
  caption = "PAI3D enrichment of pathogenic positions (defined by prediction column) by CAAS threshold and direction. N positions = the actual test population (shared-gene, PAI3D-scored positions) at that threshold. 100% row = all directional positions (baseline); 50% row = top half by CAAS score; Fisher OR and P value compare top-subset vs remainder, entirely within PAI-covered positions.",
  align = c("l", "l", rep("r", ncol(pai_fisher) - 2))
)
```


```{r vep-cosmic-subsec, results='asis'}
if (has_vep_cosmic) cat("### VEP: COSMIC Mutant Census\n\n")
```

```{r vep-cosmic-setup, eval=has_vep_cosmic}
# Restrict to shared genes present in both our results (background clean) and database
shared_genes_cosmic <- intersect(bg_genes, unique(vep_cosmic_df$Gene))
pos_cosmic_shared <- pos_score_meta %>% filter(Gene %in% shared_genes_cosmic)

# Filter cosmic annotations for US only if caap_group is present
if ("caap_group" %in% names(vep_cosmic_df)) {
  vep_cosmic_df <- vep_cosmic_df %>% filter(caap_group == "US", Gene %in% shared_genes_cosmic)
} else {
  vep_cosmic_df <- vep_cosmic_df %>% filter(Gene %in% shared_genes_cosmic)
}

# Join cosmic annotations with pos_cosmic_shared (keeping all rows to count recurrence)
cosmic_joined <- vep_cosmic_df %>%
  inner_join(pos_cosmic_shared, by = c("Gene", "Position"))

has_vep_cosmic <- nrow(cosmic_joined) > 0
```

`r if (exists("cosmic_joined") && nrow(cosmic_joined) == 0) "> **Note:** COSMIC Mutant Census mutations were mapped, but no scored positions could be matched to the COSMIC annotations for this direction."`

```{r vep-cosmic-stats, eval=has_vep_cosmic}
cosmic_table <- cosmic_joined %>%
  group_by(Gene, Position, hg38_ref_aa, caas_alt_aas, caas_change, CAAS_score, change_side) %>%
  summarise(
    Recurrence = n(),
    `Mutation AA(s)` = paste(unique(MUTATION_AA), collapse = "; "),
    `Description(s)` = paste(unique(MUTATION_DESCRIPTION), collapse = "; "),
    `Somatic Status(es)` = paste(unique(MUTATION_SOMATIC_STATUS), collapse = "; "),
    .groups = "drop"
  ) %>%
  arrange(desc(Recurrence)) %>%
  rename(
    `Ref AA` = hg38_ref_aa,
    `Alt AA(s)` = caas_alt_aas,
    `CAAS Change` = caas_change,
    `CAAS Score` = CAAS_score,
    `Change Side` = change_side,
    `Mutation AA(s)` = `Mutation AA(s)`,
    `Description` = `Description(s)`,
    `Somatic Status` = `Somatic Status(es)`
  )

DT::datatable(
  cosmic_table,
  caption = "CAAS positions overlapping with COSMIC cancer-related mutations, ordered by mutation recurrence",
  options = list(pageLength = 10, scrollX = TRUE),
  rownames = FALSE
)
```

```{r vep-cosmic-somatic-bar, eval=has_vep_cosmic, fig.height=5}
# Count of COSMIC mutations by Somatic Status and direction.
cosmic_counts <- bind_rows(
  cosmic_joined %>% filter(change_side %in% c("top",    "both")) %>% mutate(cosmic_direction = "TOP"),
  cosmic_joined %>% filter(change_side %in% c("bottom", "both")) %>% mutate(cosmic_direction = "BOTTOM")
) %>%
  filter(!is.na(MUTATION_SOMATIC_STATUS) & MUTATION_SOMATIC_STATUS != "") %>%
  count(cosmic_direction, MUTATION_SOMATIC_STATUS)

ggplot(cosmic_counts, aes(x = MUTATION_SOMATIC_STATUS, y = n, fill = cosmic_direction)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = n), position = position_dodge(width = 0.8), vjust = -0.3, size = 3.5) +
  scale_fill_manual(values = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[3])) +
  labs(
    title    = "COSMIC Mutation Somatic Status Distribution",
    x = NULL, y = "N mutations", fill = "Direction"
  ) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r vep-cosmic-fisher, eval=has_vep_cosmic}
# Odds ratio & P-value interpretation:
# - Odds Ratio: Measures the enrichment of positions with somatic mutations reported in the COSMIC database within the specified CAAS threshold compared to the rest of the positions. An odds ratio > 1 indicates enrichment.
# - P-value: Reaches significance (via Fisher's exact test) when the proportion of cancer somatic mutations is significantly higher inside the CAAS threshold.
pos_cosmic <- pos_cosmic_shared %>%
  mutate(has_cosmic = paste(Gene, Position, sep = "_") %in% paste(vep_cosmic_df$Gene, vep_cosmic_df$Position, sep = "_"))

pos_cosmic_top    <- pos_cosmic %>% filter(change_side %in% c("top",    "both"))
pos_cosmic_bottom <- pos_cosmic %>% filter(change_side %in% c("bottom", "both"))

run_cosmic_fisher_dir <- function(df, label) {
  n_all <- nrow(df)
  thr_50_val <- quantile(df$CAAS_score, 0.50, na.rm = TRUE)
  pct_rows <- list(
    list(col = NULL, lbl = "100% (all positions)",    flag = rep(TRUE, n_all)),
    list(col = NULL, lbl = "50% (top 50% CAAS)",      flag = df$CAAS_score >= thr_50_val),
    list(col = "in_top_10pct", lbl = sprintf("Top %g%%", top_pct  * 100), flag = NULL),
    list(col = "in_top_5pct",  lbl = sprintf("Top %g%%", top5_pct * 100), flag = NULL),
    list(col = "in_top_1pct",  lbl = sprintf("Top %g%%", top1_pct * 100), flag = NULL)
  )

  purrr::map_dfr(pct_rows, function(thr) {
    in_top   <- if (!is.null(thr$flag)) thr$flag else df[[thr$col]]
    in_top   <- replace_na(in_top, FALSE)
    n_top    <- sum(in_top)
    n_top_c  <- sum(in_top & df$has_cosmic, na.rm = TRUE)
    n_rest   <- sum(!in_top)
    n_rest_c <- sum(!in_top & df$has_cosmic, na.rm = TRUE)
    mat <- matrix(c(n_top - n_top_c, n_rest - n_rest_c, n_top_c, n_rest_c), nrow = 2)
    ft  <- tryCatch(fisher.test(mat),
                    error = function(e) list(estimate = NA, conf.int = c(NA, NA), p.value = NA))
    
    pct_sig <- if ("gate_sig" %in% names(df)) sprintf("%.1f%%", 100 * mean(df$gate_sig[in_top], na.rm = TRUE)) else "0.0%"
    pct_fdr <- if ("gate_fdr" %in% names(df)) sprintf("%.1f%%", 100 * mean(df$gate_fdr[in_top], na.rm = TRUE)) else "0.0%"

    tibble(
      Direction            = label,
      Threshold            = thr$lbl,
      `N positions`        = n_top,
      `N COSMIC`           = n_top_c,
      `% COSMIC`           = sprintf("%.1f%%", 100 * n_top_c / max(n_top, 1)),
      `% gate_sig`         = pct_sig,
      `% gate_fdr`         = pct_fdr,
      `Odds ratio`         = sprintf("%.2f", ft$estimate),
      `P value`            = formatC(ft$p.value, format = "e", digits = 2)
    )
  })
}

cosmic_fisher <- bind_rows(
  run_cosmic_fisher_dir(pos_cosmic_top,    "TOP"),
  run_cosmic_fisher_dir(pos_cosmic_bottom, "BOTTOM")
)

knitr::kable(
  cosmic_fisher,
  caption = "Fisher exact test for COSMIC somatic mutations overlap enrichment within CAAS score thresholds",
  align = c("l", "l", rep("r", ncol(cosmic_fisher) - 2))
)
```

```{r reg-accum-subsec, results='asis'}
if (has_genomic) cat("### Regional Accumulation\n\n")
```

```{r position-region-accum, eval=has_genomic, fig.height=10, fig.width=14}
# Build directional window summaries — TOP and BOTTOM use direction-specific thresholds
.build_region_wins <- function(pos_sub, thr10, thr5, thr1, chr_offsets_df, dir_label) {
  pos_sub %>%
    mutate(
      win_id    = floor(gene_mid / params$window_size_bp),
      win_start = win_id * params$window_size_bp,
      win_end   = win_start + params$window_size_bp - 1L,
      win_mid   = win_start + params$window_size_bp / 2L,
      in_top10  = CAAS_score >= thr10,
      in_top5   = CAAS_score >= thr5,
      in_top1   = CAAS_score >= thr1
    ) %>%
    group_by(chr_order, chr_label, win_id, win_start, win_end, win_mid) %>%
    summarise(
      n_positions        = n(),
      n_top10_positions  = sum(in_top10, na.rm = TRUE),
      n_top5_positions   = sum(in_top5,  na.rm = TRUE),
      n_top1_positions   = sum(in_top1,  na.rm = TRUE),
      frac_top_positions = n_top10_positions / n_positions,
      mean_caas          = mean(CAAS_score, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    left_join(chr_offsets_df %>% select(chr_order, offset), by = "chr_order") %>%
    mutate(
      direction    = dir_label,
      win_x        = win_mid + offset,
      region_label = sprintf("chr%s:%0.1f-%0.1f Mb",
                             chr_label, win_start / 1e6, win_end / 1e6)
    )
}

# pos_genomic already has chr_order and gene_mid; split by direction
pos_genomic_top <- pos_genomic %>% filter(change_side %in% c("top",    "both"))
pos_genomic_bot <- pos_genomic %>% filter(change_side %in% c("bottom", "both"))

pos_region_top <- .build_region_wins(pos_genomic_top, .thr_top_10, .thr_top_5, .thr_top_1,
                                      pos_chr_offsets, "TOP")
pos_region_bot <- .build_region_wins(pos_genomic_bot, .thr_bot_10, .thr_bot_5, .thr_bot_1,
                                      pos_chr_offsets, "BOTTOM")

# Combined (global thresholds) for overview panel
pos_region_wins <- pos_genomic %>%
  mutate(
    win_id    = floor(gene_mid / params$window_size_bp),
    win_start = win_id * params$window_size_bp,
    win_end   = win_start + params$window_size_bp - 1L,
    win_mid   = win_start + params$window_size_bp / 2L,
    in_top10  = CAAS_score >= thr_10,
    in_top5   = CAAS_score >= thr_5,
    in_top1   = CAAS_score >= thr_1
  ) %>%
  group_by(chr_order, chr_label, win_id, win_start, win_end, win_mid) %>%
  summarise(
    n_positions        = n(),
    n_top10_positions  = sum(in_top10, na.rm = TRUE),
    n_top5_positions   = sum(in_top5,  na.rm = TRUE),
    n_top1_positions   = sum(in_top1,  na.rm = TRUE),
    frac_top_positions = n_top10_positions / n_positions,
    mean_caas          = mean(CAAS_score, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(pos_chr_offsets %>% select(chr_order, offset), by = "chr_order") %>%
  mutate(
    win_x        = win_mid + offset,
    region_label = sprintf("chr%s:%0.1f-%0.1f Mb",
                           chr_label, win_start / 1e6, win_end / 1e6)
  )

# Dotplots removed — tables below provide the same information with GenePos detail.
```

```{r position-region-accum-table, eval=has_genomic}
.make_region_tbl <- function(wins_df) {
  wins_df %>%
    filter(n_top10_positions > 0) %>%
    arrange(desc(n_positions), desc(frac_top_positions)) %>%
    slice_head(n = 30) %>%
    mutate(
      frac_top10 = sprintf("%.1f%%", 100 * n_top10_positions / n_positions),
      frac_top5  = sprintf("%.1f%%", 100 * n_top5_positions  / n_positions),
      frac_top1  = sprintf("%.1f%%", 100 * n_top1_positions  / n_positions)
    ) %>%
    select(
      Region              = region_label,
      `N positions`       = n_positions,
      `N top-10%`         = n_top10_positions,
      `% top-10%`         = frac_top10,
      `N top-5%`          = n_top5_positions,
      `% top-5%`          = frac_top5,
      `N top-1%`          = n_top1_positions,
      `% top-1%`          = frac_top1,
      `Mean CAAS`         = mean_caas
    ) %>%
    mutate(`Mean CAAS` = round(`Mean CAAS`, 4))
}

knitr::kable(.make_region_tbl(pos_region_top),
  caption = "TOP-direction: genomic windows with high-CAAS positions, ordered by N positions (top 30; direction-specific thresholds)")
knitr::kable(.make_region_tbl(pos_region_bot),
  caption = "BOTTOM-direction: genomic windows with high-CAAS positions, ordered by N positions (top 30; direction-specific thresholds)")

# Export directional regional accumulation tables as TSV
region_export <- bind_rows(
  .make_region_tbl(pos_region_top) %>% mutate(Direction = "TOP"),
  .make_region_tbl(pos_region_bot) %>% mutate(Direction = "BOTTOM")
) %>% select(Direction, everything())
readr::write_tsv(region_export,
  file.path(params$output_dir, "regional_accumulation.tsv"))
```


# Methods

- **Test**: Fisher's exact test, one 2x2 table per (ranking, database, pathway,
  top_frac): foreground = the top 25/10/5/1% of scored positions (by `CAAS_score`,
  `scoring` module `position_scores.tsv`) in that direction; background = the
  full honest tested position universe (all non-gapped "selected" alignment
  columns of the cleaned-background genes) — never restricted to the scored
  subset. Run separately for the **global / top / bottom** directions
  (`change_side`); `top`/`bottom` filter to `change_side` first (matching the
  `scoring` module's own percentile convention — see `11.Scoring_report.Rmd`),
  `global` does not.
- **Two background conventions.** Every database above tests against the
  **full honest background** *except* `cosmic_orthogroups` and
  `pai3d_orthogroups`, which test against a **coverage-restricted background**
  (foreground *and* background both intersected with the set of genes the
  external database itself could ever annotate — see
  `cosmic_coverage_genes.txt` / `pai3d_coverage_genes.txt`, emitted by
  `build_position_gmt.py`). The distinction is *why* a position is absent from
  a term: `orthogroups`/`pfam_domains`/`selection_*`/`ucr_*` are internal
  deterministic computations with complete coverage by construction, so an
  absent position is genuine negative evidence. COSMIC and PAI3D are external
  observational/curated databases with incomplete real-world coverage — a gene
  COSMIC never sequenced, or PAI3D never scored, is absent from the term for a
  reason that has nothing to do with the trait being tested. Testing those two
  GMTs against the full ~1.47M-position background would dilute the signal
  with genes/positions the database could never have annotated regardless of
  outcome; restricting both foreground and background to the coverage subset
  keeps the 2x2 table's implicit universe honest to what was actually tested.
- **Databases**: 10 GMT-style position-set databases (Pfam domains/clans,
  COSMIC and PAI3D and general orthogroups, genomic locations,
  positive/purifying-selection orthogroups, UCR core/flank orthogroups) plus
  the broad `characterization` layers (UCR core/flank, FUBAR
  positive/purifying) — unified onto the same Fisher mechanism. `min`/`max`
  set-size filtering applies only to the GMT-style databases; the broad layers
  were never size-filtered. PAI3D membership is restricted to variants the
  `prediction` column calls pathogenic (substring match, no score-cutoff
  fallback — the same convention as `11.Scoring_report.Rmd`'s
  `is_pathogenic`); COSMIC has no such filter since every somatic mutation
  counts as evidence.
- **`p.adj`**: BH-adjusted per (ranking, database) — a handful of huge broad
  layers and up to thousands of small GMT terms carry very different
  multiple-testing burdens, so they are not pooled into one BH family.
- **Cross-module corroboration (`%gate_sig`, `%gate_fdr`, `%fade_*`, `%rer_*`,
  `%accum`)**: optional, present when POSENRICH is run with an `--annot-file`
  (SCORING's `fcs_stats.tsv`). For each overlap, the distinct genes
  contributing to it (never raw positions — a single gene contributing many
  overlapping positions must not dominate the percentage) are looked up
  against SCORING's own `flag_*` columns, and the percentage carrying each
  flag is reported alongside the Fisher result — the same idea as the
  gene-level FCS report's leading-edge composition, applied to positions.
- **Significance is `p.adj < `r PADJ`` alone** — no fold-enrichment magnitude
  gate is applied. `fold_enrichment` is still computed and reported for every
  result so a minimum-effect-size cutoff can be applied downstream; be aware
  that at this background size (~1.47M positions) even a biologically
  negligible deviation from the expected ratio (e.g. fold ≈ 0.95) can reach a
  "significant" p-value on p-value alone.
- **No permulation (`p.perm`) at position level.** The CAAS permulation null is a
  *naive random relabeling*: when the phenotype is permuted, the fg/bg species are
  paired at **random across the tree**, not as the observed **matched sister
  pairs** — so pair MRCAs are deep and the ASR path-score's isolation/independence
  geometry differs from the observed. That null therefore tests "phenotype + a
  broken matched-pair design" jointly rather than isolating the phenotype, and it
  is anti-conservative on the MRCA axes. The **gene-level FCS** still uses
  `p.perm`; see its report for the same pairing caveat.

*Report generated `r Sys.time()`.*

---

## Summary

This report summarizes the results of the **Enrichment** analysis.

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — Enrichment Module*
*Date of generation: `r Sys.Date()`*
