---
title: "Position-level Enrichment (Fisher, fixed cutoffs) — `r params$traitname`"
author: "PhyloPhere Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    code_folding: hide
    df_print: paged
    self_contained: true
params:
  results_file: NULL
  traitname: "trait"
  padj_thr: 0.15
  fold_thr: 1.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6, dev = "png", dpi = 150)
suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
  library(kableExtra)
  library(DT)
  library(scales)
})

PADJ <- params$padj_thr
FOLD <- params$fold_thr

kbl_html <- function(df, caption = NULL) {
  kable(df, format = "html", escape = TRUE, caption = caption) %>%
    kable_styling(bootstrap_options = c("striped", "condensed", "hover"),
                  full_width = FALSE)
}
```

```{r load, include=FALSE}
stopifnot(!is.null(params$results_file), file.exists(params$results_file))
res <- readr::read_tsv(params$results_file, show_col_types = FALSE)
has_res <- !is.null(res) && nrow(res) > 0
if (has_res && !"sig" %in% names(res)) {
  res$sig <- res$p_adj < PADJ & !is.na(res$fold_enrichment) &
    (res$fold_enrichment >= FOLD | res$fold_enrichment <= 1 / FOLD)
}
```

# Overview

This report summarises **position-level enrichment** for `r params$traitname`. Unlike
the gene-level FCS (a Wilcoxon-AUC rank-shift test), the position background is a
sea of invariant columns that can never carry a CAAS change — a rank-shift test
degenerates there. Instead we use **Fisher's exact test** at three fixed
foreground cutoffs (top 10 / 5 / 1% of scored positions, per direction), run
against every pathway across 9 GMT-style databases plus the broad functional
`characterization` layers (UCR core/flank, FUBAR positive/purifying) — all on
the same mechanism.

A set is flagged **significant** only when it clears **both** `p.adj < `r PADJ``
(BH-adjusted per ranking × database) **and** a minimum fold-enrichment
magnitude (`fold >= `r FOLD`` for enrichment, or `<= `r round(1/FOLD, 3)`` for
depletion). The fold gate matters because Fisher/hypergeometric tests gain
enormous power at this background size (~1.47M positions) — a barely-there
deviation from the null ratio can otherwise reach a "significant" p-value on
its own (see Methods).

```{r overview, echo=FALSE}
if (has_res) {
  ov <- res %>%
    group_by(ranking, database) %>%
    summarise(`Sets tested` = n_distinct(pathway),
              `Sig (p.adj & fold)` = n_distinct(pathway[sig]),
              .groups = "drop") %>%
    arrange(ranking, database)
  kbl_html(ov, caption = sprintf(
    "Sets tested and significant at p.adj < %.3g AND fold outside [%.3g, %.3g], by direction and database.",
    PADJ, 1 / FOLD, FOLD))
} else {
  cat("No enrichment results were produced.\n")
}
```

# Enrichment by direction and database {.tabset .tabset-fade}

Foreground = the top 10 / 5 / 1% of scored positions by `CAAS_score` in that
direction (`top`/`bottom` restrict to `change_side` first, matching the
SCORING module's own percentile convention; `global` uses every scored
position, unrestricted by `change_side`). Background = the full honest tested
position pool for every test — never restricted to the scored subset, so
tested-and-negative positions are never silently dropped from the null.
`fold > 1` = enriched, `< 1` = depleted. UCR core and UCR flank (and their
per-orthogroup counterparts) are tested as independent, potentially-overlapping
layers, not a forced partition — nothing here requires them to be disjoint.
When available, `%gate_sig`/`%gate_fdr`/`%fade_*`/`%rer_*`/`%accum` columns
report the percentage of *distinct genes* contributing to the overlap (never
raw positions — one gene must not dominate) that are independently flagged by
SCORING's own significance/FADE/RER/accumulation criteria — a cross-module
corroboration check for whether a hit is broadly supported or isolated.

```{r enrichment, echo=FALSE, results='asis'}
if (!has_res) {
  cat("_No results._\n\n")
} else {
  for (rk in unique(res$ranking)) {
    cat(sprintf("\n## %s {.tabset}\n\n", toupper(rk)))
    rk_df <- res %>% filter(ranking == rk)
    for (db in unique(rk_df$database)) {
      cat(sprintf("\n### %s {.tabset}\n\n", db))
      db_df <- rk_df %>% filter(database == db)

      for (tf in sort(unique(db_df$top_frac), decreasing = TRUE)) {
        cat(sprintf("\n#### Top %s%%\n\n", format(100 * tf)))
        tf_src <- db_df %>% filter(top_frac == tf) %>% arrange(desc(sig), p_adj, p_value)
        tf_df <- tf_src %>%
          transmute(Pathway = pathway, Description = description,
                    `Layer size` = layer_size, `Foreground` = foreground_size,
                    Observed = observed, Expected = round(expected, 1),
                    Fold = round(fold_enrichment, 3), Direction = direction,
                    `p (Fisher)` = signif(p_value, 3),
                    `p.adj` = signif(p_adj, 3), Sig = ifelse(sig, "*", ""))
        # Cross-module corroboration: % of the DISTINCT genes in the overlap
        # (never % of raw positions — one gene must not dominate) carrying
        # each SCORING flag (gate_sig/gate_fdr/fade/rer/accum). Only present
        # when POSENRICH was run with --annot-file.
        flag_cols <- grep("^pct_", names(tf_src), value = TRUE)
        if (length(flag_cols) > 0) {
          flag_df <- tf_src %>% transmute(across(all_of(flag_cols), ~round(.x, 1)))
          names(flag_df) <- paste0("%", sub("^pct_", "", flag_cols))
          tf_df <- dplyr::bind_cols(tf_df, flag_df)
        }
        id <- paste("dt", rk, db, gsub("\\.", "_", as.character(tf)), sep = "_")
        cat(knitr::knit_child(text = c(
          paste0("```{r ", id, ", echo=FALSE}"),
          "DT::datatable(tf_df, options = list(pageLength = 10, scrollX = TRUE), rownames = FALSE)",
          "```"), envir = environment(), quiet = TRUE))
        cat("\n\n")
      }

      # One dotplot per (ranking, database), faceted by top_frac.
      d <- db_df %>% filter(!is.na(fold_enrichment), fold_enrichment > 0) %>%
        mutate(l2f = log2(fold_enrichment), neglogp = -log10(pmax(p_adj, 1e-300)),
               frac_lab = sprintf("Top %s%%", format(100 * top_frac)))
      if (nrow(d) > 0) {
        lab <- d %>% filter(sig) %>% arrange(p_adj) %>% group_by(top_frac) %>% slice_head(n = 8) %>% ungroup()
        p <- ggplot(d, aes(l2f, neglogp)) +
          geom_hline(yintercept = -log10(PADJ), linetype = "dashed", colour = "grey50") +
          geom_vline(xintercept = c(-log2(FOLD), log2(FOLD)), linewidth = 0.3, colour = "grey70", linetype = "dotted") +
          geom_point(aes(colour = sig), alpha = 0.75) +
          { if (nrow(lab) > 0) ggrepel::geom_text_repel(data = lab, aes(label = pathway), size = 3, max.overlaps = 12) } +
          facet_wrap(~frac_lab) +
          scale_colour_manual(values = c(`FALSE` = "grey55", `TRUE` = "#c0392b"),
                              name = sprintf("p.adj<%.2g & fold-gated", PADJ)) +
          labs(x = expression(log[2]~fold), y = expression(-log[10]~p.adj),
               title = sprintf("%s — %s", toupper(rk), db)) +
          theme_minimal(base_size = 12)
        print(p)
        cat("\n\n")
      }
    }
  }
}
```

# Methods

- **Test**: Fisher's exact test, one 2x2 table per (ranking, database, pathway,
  top_frac): foreground = the top 10/5/1% of scored positions (by `CAAS_score`,
  `scoring` module `position_scores.tsv`) in that direction; background = the
  full honest tested position universe (all non-gapped "selected" alignment
  columns of the cleaned-background genes) — never restricted to the scored
  subset. Run separately for the **global / top / bottom** directions
  (`change_side`); `top`/`bottom` filter to `change_side` first (matching the
  `scoring` module's own percentile convention — see `11.Scoring_report.Rmd`),
  `global` does not.
- **Databases**: 9 GMT-style position-set databases (Pfam domains/clans, COSMIC
  and general orthogroups, genomic locations, positive/purifying-selection
  orthogroups, UCR core/flank orthogroups) plus the broad `characterization`
  layers (UCR core/flank, FUBAR positive/purifying) — unified onto the same
  Fisher mechanism. `min`/`max` set-size filtering applies only to the 9
  GMT-style databases; the broad layers were never size-filtered.
- **`p.adj`**: BH-adjusted per (ranking, database) — a handful of huge broad
  layers and up to thousands of small GMT terms carry very different
  multiple-testing burdens, so they are not pooled into one BH family.
- **Cross-module corroboration (`%gate_sig`, `%gate_fdr`, `%fade_*`, `%rer_*`,
  `%accum`)**: optional, present when POSENRICH is run with an `--annot-file`
  (SCORING's `fcs_stats.tsv`). For each overlap, the distinct genes
  contributing to it (never raw positions — a single gene contributing many
  overlapping positions must not dominate the percentage) are looked up
  against SCORING's own `flag_*` columns, and the percentage carrying each
  flag is reported alongside the Fisher result — the same idea as the
  gene-level FCS report's leading-edge composition, applied to positions.
- **Significance is dual-gated**: `p.adj < `r PADJ`` **and** fold-enrichment
  clearing `r FOLD`x (or `r round(1/FOLD, 3)`x for depletion). The fold gate
  exists because, at this background size (~1.47M positions), even a
  biologically negligible deviation from the expected ratio (e.g. fold ≈ 0.95)
  can reach a "significant" p-value on p-value alone — a minimum-effect-size
  floor is required alongside significance, mirroring the dual-threshold
  design the gene-level FCS report already uses (`p.adj` AND `p.perm`).
- **No permulation (`p.perm`) at position level.** The CAAS permulation null is a
  *naive random relabeling*: when the phenotype is permuted, the fg/bg species are
  paired at **random across the tree**, not as the observed **matched sister
  pairs** — so pair MRCAs are deep and the ASR path-score's isolation/independence
  geometry differs from the observed. That null therefore tests "phenotype + a
  broken matched-pair design" jointly rather than isolating the phenotype, and it
  is anti-conservative on the MRCA axes. The **gene-level FCS** still uses
  `p.perm`; see its report for the same pairing caveat.

*Report generated `r Sys.time()`.*
