---
title: "FCS Enrichment (Wilcoxon-AUC) — `r params$project_name`"
author: "PhyloPhere Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    code_folding: hide
    df_print: paged
    self_contained: true
    mathjax: null                       # no math in this report
    md_extensions: "-tex_math_dollars"  # gene-set names contain '$' (e.g. V$HNF4_01) — never parse as TeX math
params:
  stats_file:   NULL          # TSV: gene + one-or-more score_<ranking> cols + optional flag_<name> cols
  universe_file: NULL         # cleaned_background (one gene symbol per line)
  gmt_dir:      NULL          # directory of curated *.gmt files
  project_name: "FCS"
  num_g:        10            # min genes per set (fastwilcoxGMT num.g)
  fdr_thr:      0.15          # BH (adjusted-p) cutoff — RERconverge categorical default (msae210)
  pperm_thr:    0.025         # permulation-p cutoff — applied in addition to fdr_thr when perms exist
  top_n:        20            # rows shown per database x ranking
  traitname:    "trait"
  enrich:       FALSE         # RER-only permutation enrichment toggle
  perms_file:   "NO_FILE"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 7, dev = "png", dpi = 200)
suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
  library(kableExtra)
})
filter <- dplyr::filter; select <- dplyr::select; rename <- dplyr::rename
mutate <- dplyr::mutate; arrange <- dplyr::arrange
`%||%` <- function(a, b) if (!is.null(a) && length(a) > 0 && nzchar(as.character(a)[1])) a else b

this_dir <- tryCatch(dirname(normalizePath(knitr::current_input(), mustWork = FALSE)),
                     error = function(e) ".")
src_candidates <- c(file.path(this_dir, "fcs_enrich.R"), "fcs_enrich.R")
src <- src_candidates[file.exists(src_candidates)][1]
if (is.na(src)) stop("fcs_enrich.R not found next to FCS_general.Rmd")
suppressMessages(invisible(capture.output(source(src))))

kbl_html <- function(df, height = NULL) {
  k <- kable(df, format = "html", escape = TRUE) %>%
    kable_styling(bootstrap_options = c("striped", "condensed", "hover"), full_width = FALSE)
  if (!is.null(height)) k <- scroll_box(k, height = height)
  k
}
```

```{r load, include=FALSE}
stopifnot(!is.null(params$stats_file), file.exists(params$stats_file))
stopifnot(!is.null(params$universe_file), file.exists(params$universe_file))

stats <- readr::read_tsv(params$stats_file, show_col_types = FALSE)
if (!"gene" %in% names(stats)) {
  gcol <- intersect(c("Gene", "GENE"), names(stats))[1]
  if (!is.na(gcol)) stats <- dplyr::rename(stats, gene = !!gcol)
}
stopifnot("gene" %in% names(stats))

# Universe = cleaned_background (all tested genes). Fall back to the scored genes
# if no valid background was provided (e.g. a Nextflow sentinel placeholder).
universe <- character(0)
if (!is.null(params$universe_file) && file.exists(params$universe_file) &&
    !grepl("^NO_", basename(params$universe_file))) {
  universe <- unique(trimws(readLines(params$universe_file)))
  universe <- universe[nzchar(universe)]
}
if (length(universe) < 2) {
  message("No valid universe file — falling back to scored genes as the universe.")
  universe <- unique(stats$gene)
}

score_cols <- grep("^score_", names(stats), value = TRUE)
flag_cols  <- grep("^flag_",  names(stats), value = TRUE)
if (length(score_cols) == 0) stop("stats_file has no score_<ranking> columns")
ranking_names <- sub("^score_", "", score_cols)

gmt_dir <- params$gmt_dir %||% file.path(this_dir, "..", "dat")
invisible(capture.output(suppressMessages(gmts <- fcs_load_gmts(gmt_dir))))
desc_map <- suppressMessages(fcs_load_descriptions(gmt_dir))
```

# Overview

```{r overview, echo=FALSE}
tibble::tibble(
  Field = c("Project", "Trait", "Genes (stats)", "Universe genes", "Rankings",
            "Flags", "GMT databases", "num.g (min set)", "BH threshold"),
  Value = c(params$project_name, params$traitname, nrow(stats), length(universe),
            paste(ranking_names, collapse = ", "),
            if (length(flag_cols)) paste(sub("^flag_", "", flag_cols), collapse = ", ") else "(none)",
            length(gmts), params$num_g, params$fdr_thr)
) %>% kbl_html()
```

```{r run, include=FALSE}
# One zero-floored ranking vector per score_<ranking> column.
rankings <- list()
for (sc in score_cols) {
  rk <- sub("^score_", "", sc)
  rankings[[rk]] <- fcs_build_vals(setNames(stats[[sc]], stats$gene), universe)
}
enrich <- suppressMessages(fcs_run_all(rankings, gmts, num_g = params$num_g, perms_file = params$perms_file)) %>%
  dplyr::left_join(desc_map, by = c("database", "pathway")) %>%
  dplyr::mutate(description = dplyr::coalesce(description, pathway))

# Per-gene attribute table: stats flags + derived directional percentile flags.
pct_flags <- fcs_percentile_flags(stats)
attr_df <- pct_flags
if (length(flag_cols)) attr_df <- dplyr::left_join(pct_flags, stats[, c("gene", flag_cols)], by = "gene")
attr_cols_all <- setdiff(names(attr_df), "gene")

# Restrict the leading edge to the genes that actually DRIVE each enrichment —
# i.e. pathway members with signal (score > 0) in that ranking. fastwilcoxGMT's
# gene.vals lists every member in the universe (incl. zero-floored background
# genes); summarising over those would dilute the composition with non-drivers
# and leave their flags undefined. score_long lets us keep only the drivers.
score_long <- stats %>%
  dplyr::select(gene, dplyr::all_of(score_cols)) %>%
  tidyr::pivot_longer(-gene, names_to = "ranking", values_to = "drv_score") %>%
  dplyr::mutate(ranking = sub("^score_", "", ranking))
leading <- fcs_annotate_leading_edge(enrich, attr_df) %>%
  dplyr::left_join(score_long, by = c("gene", "ranking")) %>%
  dplyr::filter(!is.na(drv_score) & drv_score > 0) %>%
  dplyr::select(-drv_score)

dir.create("fcs_results", showWarnings = FALSE)
readr::write_tsv(enrich %>% select(-any_of("gene.vals")), "fcs_results/fcs_all_results.tsv")
readr::write_tsv(leading, "fcs_results/fcs_leading_edge.tsv")

# Dual-threshold significance (RERconverge categorical convention, msae210):
#   adjusted p < fdr_thr AND — when permulations are available — permulation p <
#   pperm_thr. The permulation p corrects for non-independence in gene ranks; a low
#   value indicates the BH significance is driven by the phenotype, not other
#   correlation in the data. Where no permulations exist (perms_file = NO_FILE), the
#   permulation clause is vacuous and significance falls back to the BH cutoff alone.
enrich <- enrich %>% dplyr::mutate(
  sig = stat > 0 & !is.na(p.adj) & p.adj < params$fdr_thr &
        (is.na(p.perm) | p.perm < params$pperm_thr)
)
has_perm_any <- any(!is.na(enrich$p.perm))
n_sig <- sum(enrich$sig, na.rm = TRUE)
```

`r if (has_perm_any) sprintf("**%d** gene sets are significantly enriched (BH < %.2g **and** permulation p < %.3g) across %d databases.", n_sig, params$fdr_thr, params$pperm_thr, dplyr::n_distinct(enrich$database)) else sprintf("**%d** gene sets are positively enriched at BH < %.2g across %d databases. _(No permulations: permulation-p threshold not applied.)_", n_sig, params$fdr_thr, dplyr::n_distinct(enrich$database))`

# Enriched gene sets {.tabset .tabset-fade}

Positive `stat` (AUC − 0.5 > 0) means the ranking's genes sit systematically
**high** for that set. Top `r params$top_n` positively-enriched sets per database,
ordered by BH p. Each ranking is a tab; databases are nested tabs.

```{r tables, results='asis'}
for (rk in ranking_names) {
  cat(sprintf("\n## %s {.tabset}\n", rk))
  sub <- enrich %>% filter(ranking == rk, stat > 0)
  if (nrow(sub) == 0) { cat("\n_No positively-enriched sets._\n"); next }
  
  # Check if permutations exist for this ranking
  has_perm <- "p.perm" %in% names(sub) && any(!is.na(sub$p.perm))
  
  # Arrange by p.perm if available, otherwise by p.adj
  if (has_perm) {
    sub <- sub %>% arrange(p.perm, p.adj, desc(stat))
  } else {
    sub <- sub %>% arrange(p.adj, desc(stat))
  }
  
  for (db in unique(sub$database)) {
    dbres <- sub %>% filter(database == db) %>% head(params$top_n) %>%
      mutate(AUC = round(stat + 0.5, 3),
             Description = stringr::str_trunc(description, 70),
             pval = signif(pval, 3),
             p.adj = signif(p.adj, 3))
    n_pass <- sum(dbres$sig, na.rm = TRUE)   # dual-threshold significance among shown sets

    if (has_perm) {
      disp <- dbres %>% mutate(p.perm = signif(p.perm, 3)) %>%
        select(pathway, Description, AUC, num.genes, pval, p.adj, p.perm)
    } else {
      disp <- dbres %>% select(pathway, Description, AUC, num.genes, pval, p.adj)
    }

    cat(sprintf("\n### %s\n\n", db))
    if (has_perm) {
      cat(sprintf("_%d shown, %d significant (BH < %.2g ∩ permulation p < %.3g)._\n\n",
                  nrow(disp), n_pass, params$fdr_thr, params$pperm_thr))
    } else {
      cat(sprintf("_%d shown, %d significant (BH < %.2g)._\n\n",
                  nrow(disp), n_pass, params$fdr_thr))
    }
    print(kbl_html(disp, height = if (nrow(disp) > 15) "420px" else NULL))
  }
}
```

# Enrichment dotplots {.tabset .tabset-fade}

Top enriched sets per database (gene-set **descriptions** on the y-axis, point
colour = AUC, size = set genes, dashed line = BH threshold). Ranking → database tabs.

```{r dotplots, results='asis', fig.height=5}
for (rk in ranking_names) {
  cat(sprintf("\n## %s {.tabset}\n", rk))
  subrk <- enrich %>% filter(ranking == rk, stat > 0, !is.na(p.adj))
  if (nrow(subrk) == 0) { cat("\n_No positively-enriched sets._\n"); next }
  for (db in unique(subrk$database)) {
    sub <- subrk %>% filter(database == db) %>% arrange(p.adj) %>% head(20) %>%
      mutate(label = stringr::str_trunc(description, 55),
             label = forcats::fct_reorder(label, -log10(p.adj)))
    cat(sprintf("\n### %s\n\n", db))
    p <- ggplot(sub, aes(x = -log10(p.adj), y = label, size = num.genes, colour = stat + 0.5)) +
      geom_point() +
      scale_colour_viridis_c(name = "AUC", option = "C", limits = c(0.5, 1)) +
      scale_size_continuous(name = "set genes") +
      geom_vline(xintercept = -log10(params$fdr_thr), linetype = "dashed", colour = "grey50") +
      labs(title = sprintf("%s — %s", db, rk), x = expression(-log[10]~"BH p"), y = NULL) +
      theme_minimal(base_size = 11)
    print(p); cat("\n")
  }
}
```

# Leading-edge composition {.tabset .tabset-fade}

For each significantly-enriched set (BH < `r params$fdr_thr``r if (has_perm_any) sprintf(" ∩ permulation p < %.3g", params$pperm_thr) else ""`), the
**percentage of its leading-edge genes** carrying each attribute — significance gates,
directional score percentiles, and cross-module support. One table per direction.

```{r leadingedge, results='asis'}
# Friendly display names per attribute column.
disp_names <- c(
  flag_gate_sig = "gate_sig", flag_gate_fdr = "gate_fdr",
  pct_top10 = "top10%", pct_top5 = "top5%", pct_top1 = "top1%",
  pct_bottom10 = "bot10%", pct_bottom5 = "bot5%", pct_bottom1 = "bot1%",
  flag_fade = "fade", flag_fade_top = "fade_top", flag_fade_bottom = "fade_bottom",
  flag_rer_acc = "rer_acc", flag_rer_decc = "rer_decc", flag_accum = "accum"
)
# Attribute columns shown per ranking (direction-aware).
cols_by_rk <- list(
  global = c("flag_gate_sig","flag_gate_fdr","pct_top10","pct_top5","pct_top1",
             "pct_bottom10","pct_bottom5","pct_bottom1",
             "flag_fade","flag_rer_acc","flag_rer_decc","flag_accum"),
  top    = c("flag_gate_sig","flag_gate_fdr","pct_top10","pct_top5","pct_top1",
             "flag_fade_top","flag_rer_acc","flag_rer_decc","flag_accum"),
  bottom = c("flag_gate_sig","flag_gate_fdr","pct_bottom10","pct_bottom5","pct_bottom1",
             "flag_fade_bottom","flag_rer_acc","flag_rer_decc","flag_accum")
)

for (rk in ranking_names) {
  cat(sprintf("\n## %s\n\n", rk))
  want <- cols_by_rk[[rk]] %||% attr_cols_all
  le_rk <- leading %>% filter(ranking == rk) %>%
    semi_join(enrich %>% filter(ranking == rk, sig),
              by = c("ranking", "database", "pathway"))
  if (nrow(le_rk) == 0) { cat("_No sets passed the significance thresholds for this direction._\n"); next }
  summ <- fcs_leading_edge_summary(le_rk, want) %>%
    left_join(desc_map, by = c("database", "pathway")) %>%
    mutate(Description = stringr::str_trunc(coalesce(description, pathway), 50),
           p.adj = signif(p.adj, 3), AUC = round(stat + 0.5, 3)) %>%
    select(database, pathway, Description, n_le, AUC, p.adj,
           any_of(intersect(want, names(.)))) %>%
    head(60)
  # rename attribute columns to friendly labels
  present_attr <- intersect(want, names(summ))
  names(summ)[match(present_attr, names(summ))] <- disp_names[present_attr]
  print(kbl_html(summ, height = if (nrow(summ) > 15) "480px" else NULL))
}
```

# Method

```{r method, echo=FALSE}
cat("Engine    : RERconverge::fastwilcoxGMTall (Wilcoxon rank-sum / Mann-Whitney AUC)\n")
cat("Universe  : cleaned_background; no-signal genes floored to 0 (zero-floor)\n")
cat("Direction : membership rankings (top = {top,both}, bottom = {bottom,both})\n")
cat("MTC       : Benjamini-Hochberg, per GMT database\n")
cat("Highlights: significance gates + directional percentile + cross-module flags\n")
```
