---
title: "Position-level Enrichment (XL-mHG) — `r params$traitname`"
author: "PhyloPhere Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    code_folding: hide
    df_print: paged
    self_contained: true
params:
  results_file: NULL
  characterization_file: NULL
  traitname: "trait"
  padj_thr: 0.05
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6, dev = "png", dpi = 150)
suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
  library(kableExtra)
  library(DT)
  library(scales)
})

PADJ  <- params$padj_thr

kbl_html <- function(df, caption = NULL) {
  kable(df, format = "html", escape = TRUE, caption = caption) %>%
    kable_styling(bootstrap_options = c("striped", "condensed", "hover"),
                  full_width = FALSE)
}

sig_stars <- function(p) {
  ifelse(is.na(p), "",
  ifelse(p < 0.001, "***",
  ifelse(p < 0.01,  "**",
  ifelse(p < 0.05,  "*", ""))))
}
```

```{r load, include=FALSE}
stopifnot(!is.null(params$results_file), file.exists(params$results_file))
res <- readr::read_tsv(params$results_file, show_col_types = FALSE)

# characterization is optional
char <- NULL
if (!is.null(params$characterization_file) &&
    file.exists(params$characterization_file)) {
  char <- tryCatch(readr::read_tsv(params$characterization_file, show_col_types = FALSE),
                   error = function(e) NULL)
  if (!is.null(char) && nrow(char) == 0) char <- NULL
}

has_res <- !is.null(res) && nrow(res) > 0
```

# Overview

This report summarises **position-level enrichment** for `r params$traitname`. Unlike
the gene-level FCS (a Wilcoxon-AUC rank-shift test), the position background is a
sea of invariant columns that can never carry a CAAS change — a rank-shift test
degenerates there. Instead we use the **XL-mHG (minimum-hypergeometric) test**: for
each position set it finds the `CAAS_score` cutoff at which the set is most
over-represented at the *top* of the ranking, so it is robust to the tied
zero-mass and remains powerful. `p.adj` is the BH-adjusted **exact XL-mHG p-value**
(per direction × database) — the exact XL-mHG p already corrects the
cutoff-optimisation, so this is the only significance used here (no permulation;
see Methods).

```{r overview, echo=FALSE}
if (has_res) {
  ov <- res %>%
    group_by(ranking, database) %>%
    summarise(`Sets tested` = n(),
              `Sig (p.adj)` = sum(p_adj < PADJ, na.rm = TRUE),
              .groups = "drop") %>%
    arrange(ranking, database)
  kbl_html(ov, caption = sprintf(
    "Sets tested and significant at p.adj < %.3g, by direction and database.", PADJ))
} else {
  cat("No enrichment results were produced.\n")
}
```

# Global Characterization {.tabset .tabset-fade}

Large functional layers (UCR cores/flanks, FUBAR positive/purifying sites) are too
big to rank as ordinary sets, so they are characterised with a **hypergeometric /
Fisher overlap**: are the top-scoring CAAS positions enriched (or depleted) inside
each layer, relative to the whole position universe? Foreground = the top 10 / 5 /
1 % of scored positions by global `CAAS_score`. `fold > 1` = enriched, `< 1` =
depleted; stars mark Fisher significance (`* p<0.05, ** p<0.01, *** p<0.001`).

```{r characterization, echo=FALSE, results='asis', fig.height=4}
if (is.null(char)) {
  cat("_No characterization layers were available._\n\n")
} else {
  char <- char %>%
    mutate(group = ifelse(grepl("^UCR", layer), "UCR",
                   ifelse(grepl("^FUBAR", layer), "Selection", "Other")),
           log2fold = log2(pmax(fold_enrichment, 1e-6)),
           star = sig_stars(p_adj))
  for (grp in c("UCR", "Selection")) {
    g_df <- char %>% filter(group == grp)
    if (nrow(g_df) == 0) next
    cat(sprintf("\n## %s {.tabset}\n\n", grp))
    for (tf in sort(unique(g_df$top_frac))) {
      cat(sprintf("\n### Top %s%% {.tabset}\n\n", format(100 * tf)))
      tf_df <- g_df %>% filter(top_frac == tf) %>%
        arrange(p_adj)
      tab <- tf_df %>%
        transmute(Layer = layer, Description = description,
                  `Layer size` = layer_size, `Foreground` = foreground_size,
                  Observed = observed, Expected = round(expected, 1),
                  `Fold` = round(fold_enrichment, 3),
                  Direction = direction,
                  `p (Fisher)` = signif(p_value, 3),
                  `p.adj` = signif(p_adj, 3), Sig = star)
      print(kbl_html(tab))
      cat("\n\n")
      p <- ggplot(tf_df, aes(x = reorder(layer, log2fold), y = log2fold,
                             fill = direction)) +
        geom_col() +
        geom_text(aes(label = star), vjust = ifelse(tf_df$log2fold >= 0, -0.2, 1.1)) +
        geom_hline(yintercept = 0, linewidth = 0.3) +
        scale_fill_manual(values = c(enriched = "#c0392b", depleted = "#2c7fb8")) +
        coord_flip() +
        labs(x = NULL, y = expression(log[2]~fold~enrichment),
             title = sprintf("%s — top %s%% CAAS foreground", grp, format(100 * tf))) +
        theme_minimal(base_size = 12)
      print(p)
      cat("\n\n")
    }
  }
}
```

# Enriched position sets {.tabset .tabset-fade}

XL-mHG enrichment per direction and database. `fold` and `leading edge` are taken
at the mHG optimal cutoff; `n scored` is how many of the set's positions carry any
CAAS signal, and `top 10/5/1%` how many sit in the top tiers of the score
distribution.

```{r enriched, echo=FALSE, results='asis'}
if (!has_res) {
  cat("_No results._\n\n")
} else {
  for (rk in unique(res$ranking)) {
    cat(sprintf("\n## %s {.tabset}\n\n", toupper(rk)))
    rk_df <- res %>% filter(ranking == rk)
    for (db in unique(rk_df$database)) {
      cat(sprintf("\n### %s\n\n", db))
      db_df <- rk_df %>% filter(database == db) %>%
        arrange(p_adj, pval) %>%
        transmute(Pathway = pathway, Description = description,
                  Fold = round(fold, 3), Cutoff = cutoff,
                  `p (mHG)` = signif(pval, 3), `p.adj` = signif(p_adj, 3),
                  N = set_size,
                  `n scored` = n_scored_members, `LE` = leading_edge_size,
                  `LE genes` = n_le_genes,
                  `top10` = n_top10, `top5` = n_top5, `top1` = n_top1)
      id <- paste("dt", rk, db, sep = "_")
      cat(knitr::knit_child(text = c(
        paste0("```{r ", id, ", echo=FALSE}"),
        "DT::datatable(db_df, options = list(pageLength = 10, scrollX = TRUE), rownames = FALSE)",
        "```"), envir = environment(), quiet = TRUE))
      cat("\n\n")
    }
  }
}
```

# Enrichment dotplots {.tabset .tabset-fade}

Each point is a position set: x = log2 fold at the mHG peak, y = −log10 `p.adj`,
size = leading-edge size. Red points are significant at `p.adj < `r PADJ``
(the dashed line).

```{r dotplots, echo=FALSE, results='asis', fig.height=5}
if (!has_res) {
  cat("_No results._\n\n")
} else {
  for (rk in unique(res$ranking)) {
    cat(sprintf("\n## %s {.tabset}\n\n", toupper(rk)))
    rk_df <- res %>% filter(ranking == rk)
    for (db in unique(rk_df$database)) {
      cat(sprintf("\n### %s\n\n", db))
      d <- rk_df %>% filter(database == db, !is.na(fold), fold > 0) %>%
        mutate(l2f = log2(fold), neglogp = -log10(pmax(p_adj, 1e-300)))
      if (nrow(d) == 0) { cat("_No plottable sets._\n\n"); next }
      d$sig <- d$p_adj < PADJ
      col_name <- sprintf("p.adj < %.3g", PADJ)
      lab <- d %>% arrange(p_adj) %>% head(8)
      p <- ggplot(d, aes(l2f, neglogp)) +
        geom_hline(yintercept = -log10(PADJ), linetype = "dashed", colour = "grey50") +
        geom_vline(xintercept = 0, linewidth = 0.3, colour = "grey70") +
        geom_point(aes(size = leading_edge_size, colour = sig), alpha = 0.75) +
        ggrepel::geom_text_repel(data = lab, aes(label = pathway), size = 3, max.overlaps = 12) +
        scale_colour_manual(values = c(`FALSE` = "grey55", `TRUE` = "#c0392b"),
                            name = col_name) +
        scale_size_continuous(name = "leading edge") +
        labs(x = expression(log[2]~fold), y = expression(-log[10]~p.adj),
             title = sprintf("%s — %s", toupper(rk), db)) +
        theme_minimal(base_size = 12)
      print(p)
      cat("\n\n")
    }
  }
}
```

# Leading-edge composition {.tabset .tabset-fade}

For the significant sets (`p.adj < `r PADJ``), what fraction of their positions sit
in the top **10 / 5 / 1 %** of the CAAS-score distribution (the scoring module's
percentile convention). High top-1% fractions flag sets driven by a few very
strong residues. **`LE genes`** is the number of *distinct genes* in the leading
edge — a low value (1–2) means the enrichment is carried by within-gene clumping
rather than a broad signal (the non-independence a permutation null would guard
against; here it is surfaced directly instead — see Methods).

```{r leadingedge, echo=FALSE, results='asis'}
if (!has_res) {
  cat("_No results._\n\n")
} else {
  for (rk in unique(res$ranking)) {
    le <- res %>% filter(ranking == rk, p_adj < PADJ, set_size > 0)
    cat(sprintf("\n## %s {.tabset}\n\n", toupper(rk)))
    if (nrow(le) == 0) { cat("_No sets pass p.adj._\n\n"); next }
    for (db in unique(le$database)) {
      cat(sprintf("\n### %s\n\n", db))
      db_le <- le %>% filter(database == db) %>%
        arrange(p_adj) %>%
        transmute(Pathway = pathway, N = set_size,
                  `n scored` = n_scored_members,
                  `LE` = leading_edge_size, `LE genes` = n_le_genes,
                  `% top10` = round(100 * n_top10 / set_size, 1),
                  `% top5`  = round(100 * n_top5  / set_size, 1),
                  `% top1`  = round(100 * n_top1  / set_size, 1),
                  `p.adj` = signif(p_adj, 3))
      print(kbl_html(db_le))
      cat("\n\n")
    }
  }
}
```

# Methods

- **Test**: XL-mHG (`xlmhglite`), positions ranked by `CAAS_score` (scoring
  module `position_scores.tsv`), separately for the **global / top / bottom**
  directions (`change_side`). `X = ` the `--posenrich_min_overlap` parameter
  (minimum set members above the cutoff; ≥2 neutralises one-hit-wonders); `L =`
  the scored region.
- **Background**: the honest testable position universe — all non-gapped
  ("selected") alignment columns of the cleaned-background genes.
- **`p.adj`**: BH-adjusted exact XL-mHG p-value, per direction × database. This is
  the **only** significance used here — the exact XL-mHG p already corrects the
  cutoff-optimisation, so no permulation is applied.
- **No permulation (`p.perm`) at position level.** The CAAS permulation null is a
  *naive random relabeling*: when the phenotype is permuted, the fg/bg species are
  paired at **random across the tree**, not as the observed **matched sister
  pairs** — so pair MRCAs are deep and the ASR path-score's isolation/independence
  geometry differs from the observed. That null therefore tests "phenotype + a
  broken matched-pair design" jointly rather than isolating the phenotype, and it
  is anti-conservative on the MRCA axes. (For a rank-based enrichment the offset
  is roughly constant across positions and largely cancels — but rather than lean
  on that, POSENRICH simply uses the exact analytic `p.adj`.) `LE genes` flags the
  within-gene clumping a null would otherwise guard against. The **gene-level FCS**
  still uses `p.perm`; see its report for the same pairing caveat.
- **Characterization**: Fisher's exact overlap of the top 10/5/1% scored
  positions against each global layer (UCR core/flank, FUBAR positive/purifying).

*Report generated `r Sys.time()`.*
```
