---
title: "STRING Enrichment Analysis"
author: "PhyloPhere Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    code_folding: hide
    df_print: paged
    self_contained: true
params:
  background_file: NULL
  background_basename: NULL
  gene_list_files: NULL
  output_dir: "string"
  project_name: "STRING_Analysis"
  species: 9606
  required_score: 400
  network_score_thr: 400
  mcl_inflation: 3
  fdr_thr: 0.1
  top_thr: 15
  report_num: 20
  enable_ppi_enrichment: false
  gene_scores_file: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7,
  dev = 'png',
  dpi = 320
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
  library(kableExtra)
  library(STRINGdb)
  library(paletteer)
})

# ── Palette helpers (overflow-safe Tableau_10) ────────────────────────────────
.pal10_base <- as.character(paletteer_d("ggthemes::Tableau_10"))
.safe_pal   <- function(n) if (n <= 10L) .pal10_base[seq_len(n)] else colorRampPalette(.pal10_base)(n)
safe_fill   <- function(n) scale_fill_manual(values  = .safe_pal(n))
safe_color  <- function(n) scale_color_manual(values = .safe_pal(n))

has_pkg <- function(pkg) requireNamespace(pkg, quietly = TRUE)

`%||%` <- function(a, b) {
  if (!is.null(a) && length(a) > 0 && !all(is.na(a))) a else b
}

createDir <- function(directory) {
  if (!dir.exists(directory)) dir.create(directory, recursive = TRUE, showWarnings = FALSE)
}

download_with_retry <- function(url, destfile, tries = 4L, timeout_sec = 1800L) {
  old_timeout <- getOption("timeout")
  on.exit(options(timeout = old_timeout), add = TRUE)
  options(timeout = max(as.integer(old_timeout), as.integer(timeout_sec), 600L))

  for (i in seq_len(tries)) {
    ok <- tryCatch({
      utils::download.file(url, destfile = destfile, mode = "wb", quiet = FALSE, method = "libcurl")
      TRUE
    }, error = function(e) {
      message(sprintf("Download failed (%d/%d): %s", i, tries, e$message))
      FALSE
    })

    if (isTRUE(ok) && file.exists(destfile) && file.size(destfile) > 0) return(invisible(TRUE))
    if (i < tries) Sys.sleep(min(30, i * 5))
  }

  stop(sprintf("Failed to download after %d attempts: %s", tries, url))
}

ensure_string_cache <- function(species, version = "12.0", input_directory = "string_cache") {
  createDir(input_directory)

  species <- as.integer(species)
  base <- "https://stringdb-downloads.org/download"
  files <- c(
    sprintf("protein.aliases.v%s/%d.protein.aliases.v%s.txt.gz", version, species, version),
    sprintf("protein.info.v%s/%d.protein.info.v%s.txt.gz", version, species, version),
    sprintf("protein.links.v%s/%d.protein.links.v%s.txt.gz", version, species, version),
    sprintf("protein.physical.links.v%s/%d.protein.physical.links.v%s.txt.gz", version, species, version)
  )

  for (rel in files) {
    url <- sprintf("%s/%s", base, rel)
    dest <- file.path(input_directory, basename(rel))
    if (!file.exists(dest) || file.size(dest) == 0) {
      message(sprintf("Caching STRING file: %s", basename(rel)))
      download_with_retry(url, destfile = dest)
    }
  }

  invisible(normalizePath(input_directory, winslash = "/", mustWork = FALSE))
}

read_gene_vector <- function(path) {
  x <- readr::read_lines(path)
  x <- trimws(x)
  x <- x[x != ""]
  unique(x)
}

string_root <- "."
string_results_dir <- "string_results"
string_summary_dir <- "string_summary"
string_plots_dir   <- "string_plots"
string_cache_dir   <- "string_cache"

createDir(string_results_dir)
createDir(string_summary_dir)
createDir(string_plots_dir)
createDir(string_cache_dir)

# Pre-cache STRING DB files with retry/timeout guards to avoid transient
# network failures during set_background()/enrichment calls.
ensure_string_cache(
  species = as.integer(params$species),
  version = "12.0",
  input_directory = string_cache_dir
)

# ── STRINGdb instances ────────────────────────────────────────────────────────
# Enrichment instance: uses required_score (default 400); background/enrichment calls.
string_db <- STRINGdb$new(
  version         = "12.0",
  species         = as.integer(params$species),
  score_threshold = as.integer(params$required_score),
  input_directory = string_cache_dir
)

# Network instance: uses network_score_thr (default 700 = 0.7) for PPI graphs.
# Higher threshold → only high-confidence interactions → sparser, more meaningful networks.
string_db_network <- STRINGdb$new(
  version         = "12.0",
  species         = as.integer(params$species),
  score_threshold = as.integer(params$network_score_thr),
  input_directory = string_cache_dir
)

# Physical subnetwork instance: uses only experimentally-determined physical interactions.
string_db_physical <- tryCatch(
  STRINGdb$new(
    version         = "12.0",
    species         = as.integer(params$species),
    score_threshold = as.integer(params$network_score_thr),
    network_type    = "physical",
    input_directory = string_cache_dir
  ),
  error = function(e) {
    message(sprintf("Physical STRINGdb instance failed: %s — physical networks skipped.", e$message))
    NULL
  }
)

string_networks_dir <- "string_networks"
createDir(string_networks_dir)

# ── STRING clustering (walktrap algorithm) ────────────────────────────────────
# STRINGdb$get_clusters() wraps igraph community detection; available algorithms
# are fastgreedy, walktrap, spinglass, and edge.betweenness.
# Walktrap is closest to STRING's own MCL: both simulate random diffusion on the
# weighted network — walks stay within dense communities, matching MCL's
# expansion/inflation dynamic.
run_string_clustering <- function(sdb, string_ids, algorithm = "walktrap") {
  if (length(string_ids) < 3) return(NULL)
  tryCatch({
    cl_list <- sdb$get_clusters(string_ids, algorithm = algorithm)
    if (length(cl_list) == 0) return(NULL)
    purrr::imap_dfr(cl_list, ~data.frame(node = .x, cluster = .y, stringsAsFactors = FALSE))
  }, error = function(e) {
    message(sprintf("STRING get_clusters (%s): %s — no clustering available", algorithm, e$message))
    NULL
  })
}

# ── Cluster-aware FR layout ───────────────────────────────────────────────────
# Seeds node positions by cluster centroid before Fruchterman-Reingold so that
# same-cluster nodes start near each other and the final layout mirrors the
# STRING web interface (clusters appear as visual blobs, not scattered).
cluster_layout <- function(net, clusters_df, seed = 42) {
  set.seed(seed)
  n <- igraph::vcount(net)
  if (is.null(clusters_df) || n < 2) return(igraph::layout_with_fr(net))

  vnames  <- igraph::V(net)$name
  cl_map  <- setNames(clusters_df$cluster, clusters_df$node)
  clusters <- cl_map[vnames]
  clusters[is.na(clusters)] <- max(clusters_df$cluster, na.rm = TRUE) + 1L

  unique_cl <- sort(unique(clusters))
  n_cl <- length(unique_cl)
  if (n_cl <= 1) return(igraph::layout_with_fr(net))

  theta  <- seq(0, 2 * pi * (1 - 1 / n_cl), length.out = n_cl)
  spread <- max(5, sqrt(n) * 0.9)
  cx     <- spread * cos(theta)
  cy     <- spread * sin(theta)

  init <- matrix(0, nrow = n, ncol = 2)
  for (i in seq_along(unique_cl)) {
    mask  <- clusters == unique_cl[i]
    n_sub <- sum(mask)
    sd_jit <- max(0.5, sqrt(n_sub) * 0.35)
    init[mask, 1] <- cx[i] + rnorm(n_sub, 0, sd_jit)
    init[mask, 2] <- cy[i] + rnorm(n_sub, 0, sd_jit)
  }
  igraph::layout_with_fr(net, coords = init, niter = 300)
}

# ── Cluster functional descriptions ──────────────────────────────────────────
# Runs STRING enrichment on each cluster's nodes to derive a functional label
# from the top significant GO / pathway term.
describe_clusters <- function(sdb, clusters_df, id2gene, top_terms = 3) {
  if (is.null(clusters_df) || nrow(clusters_df) == 0) return(tibble())
  purrr::map_dfr(sort(unique(clusters_df$cluster)), function(cl_id) {
    cl_nodes <- clusters_df$node[clusters_df$cluster == cl_id]
    genes    <- id2gene[cl_nodes]
    gene_str <- paste(head(sort(genes[!is.na(genes)]), 8), collapse = ", ")
    if (length(cl_nodes) < 3) {
      return(tibble(cluster = cl_id, n_genes = length(cl_nodes),
                    genes = gene_str, description = "Too small"))
    }
    enrich <- tryCatch(sdb$get_enrichment(cl_nodes), error = function(e) NULL)
    if (is.null(enrich) || nrow(enrich) == 0 || "error" %in% names(enrich)) {
      return(tibble(cluster = cl_id, n_genes = length(cl_nodes),
                    genes = gene_str, description = "No enrichment"))
    }
    pvcol   <- intersect(c("fdr", "p_value", "pvalue", "p.adjust"), names(enrich))[1]
    if (!is.na(pvcol)) enrich <- enrich[order(enrich[[pvcol]]), ]
    dcol    <- intersect(c("description", "term", "name"), names(enrich))[1]
    top     <- head(as.character(enrich[[dcol]]), top_terms)
    tibble(cluster = cl_id, n_genes = length(cl_nodes),
           genes = gene_str, description = paste(top, collapse = " / "))
  })
}

# ── Network plot (cluster-aware layout, no isolated nodes) ───────────────────
plot_cluster_network <- function(net, clusters_df, title, id_to_gene = NULL,
                                  cluster_labels = NULL) {
  n_clust <- if (!is.null(clusters_df)) length(unique(clusters_df$cluster)) else 1L
  pal     <- .safe_pal(max(n_clust, 2L))

  if (!is.null(clusters_df) && nrow(clusters_df) > 0) {
    cl_map  <- setNames(clusters_df$cluster, clusters_df$node)
    v_cl    <- cl_map[igraph::V(net)$name]
    v_cl[is.na(v_cl)] <- n_clust + 1L
    cl_ids  <- sort(unique(clusters_df$cluster))
    vertex_colors <- pal[match(v_cl, c(cl_ids, n_clust + 1L))]
  } else {
    cl_ids        <- integer(0)
    vertex_colors <- rep(pal[1], igraph::vcount(net))
  }

  labels <- if (!is.null(id_to_gene)) id_to_gene[igraph::V(net)$name] else igraph::V(net)$name
  labels[is.na(labels)] <- igraph::V(net)$name[is.na(labels)]

  lo <- cluster_layout(net, clusters_df)

  old_par <- par(mar = c(0, 0, 2, 5))
  on.exit(par(old_par), add = TRUE)
  igraph::plot.igraph(
    net,
    main               = title,
    layout             = lo,
    vertex.color       = vertex_colors,
    vertex.size        = pmin(18, pmax(4, 55 / sqrt(igraph::vcount(net)))),
    vertex.label       = labels,
    vertex.label.cex   = max(0.55, min(1.1, 10 / sqrt(igraph::vcount(net)))),
    vertex.label.color = "black",
    vertex.label.font  = 2,
    edge.arrow.size    = 0,
    edge.width         = 0.7,
    edge.color         = "#aaaaaa"
  )
  if (length(cl_ids) > 1) {
    leg_labels <- if (!is.null(cluster_labels)) {
      vapply(as.character(cl_ids), function(k) {
        lbl <- cluster_labels[k]
        if (is.na(lbl)) paste0("Cluster ", k) else paste0("C", k, ": ", stringr::str_trunc(lbl, 35))
      }, character(1))
    } else {
      paste0("Cluster ", cl_ids)
    }
    legend("topright", legend = leg_labels,
           fill = pal[seq_along(cl_ids)], cex = 0.55, bty = "n", title = "Cluster")
  }
}

# Map a character vector of gene SYMBOLs to STRING identifiers.
# Returns a character vector of STRING_id values.
safe_string_map_ids <- function(genes) {
  if (length(genes) == 0) return(character(0))
  df_in <- data.frame(gene = genes, stringsAsFactors = FALSE)
  mapped <- tryCatch(
    string_db$map(df_in, "gene", removeUnmappedRows = TRUE, quiet = TRUE),
    error = function(e) {
      message(sprintf("STRING ID mapping failed: %s", e$message))
      NULL
    }
  )
  if (is.null(mapped) || nrow(mapped) == 0) return(character(0))
  unique(as.character(mapped$STRING_id))
}

# Resolve background file path and normalize to absolute path
bg_file <- params$background_file
if (!file.exists(bg_file) && !is.null(params$background_basename)) {
  if (file.exists(params$background_basename)) bg_file <- params$background_basename
}

if (is.null(bg_file) || !file.exists(bg_file)) {
  stop("Error: background_file not found")
}
bg_file <- normalizePath(bg_file, mustWork = TRUE)
bg_basename <- basename(bg_file)

# Auto-detect gene list files from the current working directory, excluding the
# background file.  Using normalizePath() ensures that subsequent reads succeed
# even if the working directory changes during report rendering.
gene_list_files <- params$gene_list_files
if (is.null(gene_list_files) || length(gene_list_files) == 0) {
  gene_list_files <- list.files(pattern = "\\.txt$", full.names = TRUE)
  gene_list_files <- gene_list_files[basename(gene_list_files) != bg_basename]
}
gene_list_files <- gene_list_files[file.exists(gene_list_files)]
# Always normalize to absolute paths
gene_list_files <- normalizePath(gene_list_files, mustWork = FALSE)
gene_list_files <- gene_list_files[file.exists(gene_list_files)]
if (length(gene_list_files) == 0) stop("Error: no valid gene list files found")


```

# STRING analysis configuration

```{r config_overview}
cat("=== STRING configuration ===\n")
cat("Background file (cleaned_background):", bg_file, "\n")
cat("Gene list files:", length(gene_list_files), "\n")
cat("Species (NCBI taxon ID):", params$species, "\n")
cat("Required interaction score:", params$required_score, "\n")
cat("FDR threshold:", params$fdr_thr, "\n")
cat("Top terms per list:", params$top_thr, "\n")
cat("PPI enrichment test:", params$enable_ppi_enrichment, "\n")
cat("Network score threshold:", params$network_score_thr, "\n")
cat("Clustering algorithm: walktrap\n")
```

# Map background to STRING IDs

> **Note:** A custom background (`cleaned_background`) constrains enrichment to the experimental proteome (all tested genes).
> When background genes are provided, they are mapped to STRING identifiers and set
> on the `STRINGdb` object via `set_background()`, which applies to all subsequent
> `get_enrichment()` calls.

```{r parse_background}
bg_symbols  <- read_gene_vector(bg_file)
bg_string_ids <- safe_string_map_ids(bg_symbols)

cat("Background symbols supplied:", length(bg_symbols), "\n")
cat("Background STRING IDs successfully mapped:", length(bg_string_ids), "\n")

if (length(bg_string_ids) < 5) {
  warning(
    sprintf(
      "Only %d background genes mapped to STRING IDs. ",
      length(bg_string_ids)
    ),
    "Proceeding without a custom background (STRING will use the full proteome)."
  )
  use_custom_background <- FALSE
} else {
  # Set the background on the STRINGdb object so all subsequent enrichment
  # calls are constrained to the experimental proteome.
  use_custom_background <- tryCatch({
    string_db$set_background(bg_string_ids)
    TRUE
  }, error = function(e) {
    warning(
      sprintf("Could not set custom STRING background: %s", e$message),
      " Proceeding with STRING default background."
    )
    FALSE
  })

  if (isTRUE(use_custom_background)) {
    cat("Custom background set on STRINGdb object (", length(bg_string_ids), " proteins).\n")
  }
}
```

# Load gene scores for overlap checking

```{r load_gene_scores}
gene_scores_df <- NULL
if (!is.null(params$gene_scores_file) && file.exists(params$gene_scores_file)) {
  gene_scores_df <- readr::read_tsv(params$gene_scores_file, show_col_types = FALSE)
  cat("Loaded gene_scores_file:", params$gene_scores_file, "with", nrow(gene_scores_df), "genes.\n")
} else {
  cat("No gene_scores_file provided or file not found. Overlap checks will be skipped.\n")
}
```

# Run STRING enrichment per gene list

```{r run_string_enrichment}
MIN_MAPPED <- 5L

run_single_string <- function(gene_file, use_custom_background) {
  genes_sym   <- read_gene_vector(gene_file)
  list_name   <- tools::file_path_sans_ext(basename(gene_file))

  # Map gene list symbols to STRING IDs
  gene_string_ids <- safe_string_map_ids(genes_sym)

  if (length(gene_string_ids) < MIN_MAPPED) {
    return(list(
      list_name      = list_name,
      input_n        = length(genes_sym),
      mapped_n       = length(gene_string_ids),
      status         = "skipped_too_few_mapped_genes",
      n_rows         = 0L,
      results        = tibble(),
      ppi_result     = NULL
    ))
  }

  # ── Functional enrichment ──────────────────────────────────────────────────
  # get_enrichment() returns a flat data.frame with a 'category' column.
  # The background is already set globally on string_db via set_background().
  enrich_df <- tryCatch(
    {
      res <- string_db$get_enrichment(gene_string_ids)
      if (!is.null(res) && nrow(res) > 0) {
        # STRINGdb may return an error frame with columns 'error'/'message'
        # (e.g. background_error when mapped genes fall outside the custom
        # background set) instead of real enrichment rows.  Detect and skip.
        if ("error" %in% names(res)) {
          message(sprintf(
            "STRING enrichment API error for %s: %s",
            list_name,
            paste(as.character(res[["message"]]), collapse = "; ")
          ))
          tibble()
        } else {
          res %>% mutate(gene_list = list_name)
        }
      } else {
        tibble()
      }
    },
    error = function(e) {
      message(sprintf("STRING enrichment failed for %s: %s", list_name, e$message))
      tibble()
    }
  )

  if (nrow(enrich_df) == 0) {
    return(list(
      list_name  = list_name,
      input_n    = length(genes_sym),
      mapped_n   = length(gene_string_ids),
      status     = "no_results",
      n_rows     = 0L,
      results    = tibble(),
      ppi_result = NULL
    ))
  }

  # Standardise FDR / p-value column (STRINGdb uses 'fdr' and 'p_value')
  pval_col <- c("fdr", "p_value", "pvalue", "p.adjust", "FDR")
  pval_col <- pval_col[pval_col %in% names(enrich_df)][1] %||% NA_character_

  if (!is.na(pval_col)) {
    enrich_df <- enrich_df %>%
      filter(suppressWarnings(as.numeric(.data[[pval_col]])) <= params$fdr_thr)
    # Rename to canonical 'fdr' so all downstream consumers (all_results,
    # plots chunk, accumulation) see a consistent column name regardless of
    # the STRINGdb version that produced it.
    if (pval_col != "fdr") {
      enrich_df <- enrich_df %>% dplyr::rename(fdr = dplyr::all_of(pval_col))
    }
  }

  # ── Compute Top Overlap Metrics for STRING enriched terms ──
  enrich_df$n_top_10 <- NA_integer_
  enrich_df$n_top_5  <- NA_integer_
  enrich_df$n_top_1  <- NA_integer_
  enrich_df$n_fade   <- NA_integer_
  enrich_df$n_rer    <- NA_integer_
  enrich_df$n_accum  <- NA_integer_

  if (!is.null(gene_scores_df) && nrow(enrich_df) > 0) {
    dir_label <- if (grepl("top", list_name)) "top" else if (grepl("bottom", list_name)) "bottom" else "global"
    score_col <- if (dir_label == "top") "gene_caas_score_top" else if (dir_label == "bottom") "gene_caas_score_bottom" else "gene_caas_score"

    if (score_col %in% names(gene_scores_df)) {
      ranked_df <- gene_scores_df %>%
        filter(!is.na(.data[[score_col]])) %>%
        select(Gene, score = all_of(score_col),
               any_of(c("fade_significant_top", "fade_significant_bottom",
                        "rer_significant", "accum_significant")))

      ranked_df <- ranked_df[order(ranked_df$score, decreasing = TRUE), ]

      n_total <- nrow(ranked_df)
      if (n_total > 0) {
        top_10_genes <- head(ranked_df$Gene, max(1, round(n_total * 0.10)))
        top_5_genes  <- head(ranked_df$Gene, max(1, round(n_total * 0.05)))
        top_1_genes  <- head(ranked_df$Gene, max(1, round(n_total * 0.01)))

        fade_sig_col <- if (dir_label == "bottom") "fade_significant_bottom" else "fade_significant_top"
        fade_pos  <- ranked_df$Gene[ranked_df[[fade_sig_col]] %in% TRUE]
        rer_pos   <- ranked_df$Gene[ranked_df$rer_significant %in% TRUE]
        accum_pos <- ranked_df$Gene[ranked_df$accum_significant %in% TRUE]

        for (i in 1:nrow(enrich_df)) {
          term_pref <- as.character(enrich_df$preferredNames[i])
          if (!is.na(term_pref) && term_pref != "") {
            term_genes <- strsplit(term_pref, ",")[[1]]
            enrich_df$n_top_10[i] <- length(intersect(term_genes, top_10_genes))
            enrich_df$n_top_5[i]  <- length(intersect(term_genes, top_5_genes))
            enrich_df$n_top_1[i]  <- length(intersect(term_genes, top_1_genes))
            enrich_df$n_fade[i]   <- length(intersect(term_genes, fade_pos))
            enrich_df$n_rer[i]    <- length(intersect(term_genes, rer_pos))
            enrich_df$n_accum[i]  <- length(intersect(term_genes, accum_pos))
          }
        }
      }
    }
  }

  readr::write_tsv(
    enrich_df,
    file.path(string_summary_dir, paste0(list_name, "_results.tsv"))
  )

  # ── PPI density enrichment ─────────────────────────────────────────────────
  ppi_res <- NULL
  if (isTRUE(params$enable_ppi_enrichment)) {
    if (length(gene_string_ids) > 500L) {
      message(sprintf(
        "PPI enrichment skipped for %s: gene set too large (%d > 500 STRING IDs).",
        list_name, length(gene_string_ids)
      ))
    } else {
      ppi_res <- tryCatch(
        string_db$get_ppi_enrichment(gene_string_ids),
        error = function(e) {
          message(sprintf("PPI enrichment failed for %s: %s", list_name, e$message))
          NULL
        }
      )
    }

    if (!is.null(ppi_res)) {
      # Replace any NULL fields with NA before coercing to tibble; STRINGdb
      # returns NULL for 'enrichment' and 'lambda' when no interactions are
      # found, which makes as_tibble() abort with "column is NULL".
      ppi_lst <- lapply(as.list(ppi_res), function(x) if (is.null(x)) NA else x)
      ppi_df  <- as_tibble(ppi_lst) %>% mutate(gene_list = list_name)
      readr::write_tsv(
        ppi_df,
        file.path(string_summary_dir, paste0(list_name, "_ppi.tsv"))
      )
    }
  }

  list(
    list_name  = list_name,
    input_n    = length(genes_sym),
    mapped_n   = length(gene_string_ids),
    status     = "ok",
    n_rows     = nrow(enrich_df),
    results    = enrich_df,
    ppi_result = ppi_res
  )
}

string_runs <- lapply(
  gene_list_files,
  run_single_string,
  use_custom_background = use_custom_background
)
```

# Summary tables

```{r summary_tables}
run_summary <- purrr::map_dfr(string_runs, ~tibble(
  gene_list     = .x$list_name,
  input_genes   = .x$input_n,
  mapped_string = .x$mapped_n,
  status        = .x$status,
  n_results     = .x$n_rows
))

all_results <- purrr::map_dfr(string_runs, "results")

readr::write_tsv(run_summary, file.path(string_summary_dir, "string_run_summary.tsv"))
readr::write_tsv(all_results, file.path(string_summary_dir, "string_all_results.tsv"))

kable(run_summary, caption = "STRING enrichment run summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Render top enriched terms with overlaps
if (nrow(all_results) > 0 && !is.null(gene_scores_df)) {
  for (.gl in unique(all_results$gene_list)) {
    .gl_dat <- all_results %>%
      dplyr::filter(gene_list == .gl) %>%
      dplyr::arrange(fdr) %>%
      dplyr::slice_head(n = 10)

    if (nrow(.gl_dat) > 0) {
      display_df <- .gl_dat %>%
        dplyr::select(
          Category = category,
          Term = description,
          FDR = fdr,
          `Top 10% (N)` = n_top_10,
          `Top 5% (N)`  = n_top_5,
          `Top 1% (N)`  = n_top_1,
          `FADE (N)`    = n_fade,
          `RER (N)`     = n_rer,
          `Accum. (N)`  = n_accum
        )

      cat(sprintf("\n\n#### Top Enriched STRING Terms for %s\n\n", .gl))
      print(
        knitr::kable(display_df, digits = 4, caption = sprintf("Top Enriched STRING Terms and Gene Overlaps for %s", .gl)) %>%
          kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
      )
      cat("\n\n")
    }
  }
}
```

```{r ppi_summary, eval=isTRUE(params$enable_ppi_enrichment)}
ppi_rows <- purrr::keep(string_runs, ~!is.null(.x$ppi_result))
if (length(ppi_rows) > 0) {
  ppi_summary <- purrr::map_dfr(ppi_rows, function(x) {
    pr <- as.list(x$ppi_result)
    tibble(
      gene_list                   = x$list_name,
      number_of_nodes             = pr$number_of_nodes %||% NA_integer_,
      number_of_edges             = pr$number_of_edges %||% NA_integer_,
      average_node_degree         = pr$average_node_degree %||% NA_real_,
      local_clustering_coefficient = pr$local_clustering_coefficient %||% NA_real_,
      expected_number_of_edges    = pr$expected_number_of_edges %||% NA_integer_,
      p_value                     = pr$p_value %||% NA_real_
    )
  })

  readr::write_tsv(ppi_summary, file.path(string_summary_dir, "string_ppi_summary.tsv"))

  kable(ppi_summary, caption = "STRING PPI enrichment (density test) summary") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
}
```

# STRING Protein Networks {.tabset}

Network score threshold: **`r params$network_score_thr`** (= `r params$network_score_thr / 1000` on the 0–1 STRING scale).
Only nodes with at least one edge at this confidence level are shown.
Clustering algorithm: **walktrap** (random-walk community detection; closest available method to STRING's own MCL).

```{r string_networks, fig.width=10, fig.height=8, results='asis'}
suppressPackageStartupMessages(library(igraph))

# Map all gene symbols to STRING IDs up-front so we can reverse-annotate network nodes.
build_id2gene <- function(genes_sym) {
  if (length(genes_sym) == 0) return(character(0))
  df_in  <- data.frame(gene = genes_sym, stringsAsFactors = FALSE)
  mapped <- tryCatch(
    string_db$map(df_in, "gene", removeUnmappedRows = TRUE, quiet = TRUE),
    error = function(e) NULL
  )
  if (is.null(mapped) || nrow(mapped) == 0) return(character(0))
  setNames(mapped$gene, mapped$STRING_id)
}

network_summary_rows <- list()

for (.gf in gene_list_files) {
  .list_name   <- tools::file_path_sans_ext(basename(.gf))
  .genes_sym   <- read_gene_vector(.gf)
  .string_ids  <- safe_string_map_ids(.genes_sym)

  if (length(.string_ids) < MIN_MAPPED) next

  .id2gene <- build_id2gene(.genes_sym)
  cat(sprintf("\n\n## %s {.tabset}\n\n", .list_name))

  for (.net_type in c("full", "physical")) {
    .sdb <- if (.net_type == "full") string_db_network else string_db_physical
    if (is.null(.sdb)) {
      cat(sprintf("> Physical STRINGdb unavailable — skipping physical network for %s.\n\n", .list_name))
      next
    }

    .net <- tryCatch(
      .sdb$get_subnetwork(.string_ids),
      error = function(e) {
        message(sprintf("[%s][%s] get_subnetwork failed: %s", .list_name, .net_type, e$message))
        NULL
      }
    )
    if (is.null(.net) || igraph::vcount(.net) == 0) {
      cat(sprintf("> No %s network retrieved for %s.\n\n", .net_type, .list_name))
      next
    }

    # Keep only connected nodes (nodes with ≥1 edge at the score threshold)
    .connected <- igraph::V(.net)[igraph::degree(.net) > 0]
    .net_conn  <- igraph::induced_subgraph(.net, .connected)
    .n_nodes   <- igraph::vcount(.net_conn)
    .n_edges   <- igraph::ecount(.net_conn)

    cat(sprintf("### %s\n\n", tools::toTitleCase(.net_type)))
    cat(sprintf("Connected nodes: **%d** | Edges: **%d** | Input genes mapped: **%d**\n\n",
                .n_nodes, .n_edges, length(.string_ids)))

    if (.n_nodes < 3) {
      cat("> Too few connected nodes — network plot skipped.\n\n")
      network_summary_rows[[length(network_summary_rows) + 1]] <- tibble(
        gene_list = .list_name, net_type = .net_type,
        n_input = length(.string_ids), n_connected = .n_nodes,
        n_edges = .n_edges, n_clusters = NA_integer_
      )
      next
    }

    # Walktrap clustering via STRINGdb (closest to STRING's own MCL)
    # Run on all mapped IDs so the algorithm sees the full graph topology,
    # then restrict results to nodes that survive the connectivity filter.
    .clusters_raw <- run_string_clustering(.sdb, .string_ids)
    .net_vnames   <- igraph::V(.net_conn)$name
    .clusters_df  <- if (!is.null(.clusters_raw)) {
      .clusters_raw %>% dplyr::filter(node %in% .net_vnames)
    } else NULL
    .n_clust <- if (!is.null(.clusters_df)) length(unique(.clusters_df$cluster)) else NA_integer_

    # Annotate clusters with gene symbols and write TSV
    if (!is.null(.clusters_df)) {
      .clusters_df$gene <- .id2gene[.clusters_df$node]
      .clusters_df$gene[is.na(.clusters_df$gene)] <- .clusters_df$node[is.na(.clusters_df$gene)]
      .clust_file <- file.path(
        string_networks_dir,
        sprintf("%s_%s_clusters.tsv", .list_name, .net_type)
      )
      readr::write_tsv(.clusters_df %>% dplyr::arrange(cluster, gene), .clust_file)
      cat(sprintf("Clusters found: **%d** (written to `%s`)\n\n", .n_clust, basename(.clust_file)))

      # Cluster descriptions from top enriched term per cluster
      .clust_desc <- describe_clusters(string_db, .clusters_df, .id2gene)
      .clust_desc_display <- .clust_desc %>%
        dplyr::filter(!description %in% c("Too small", "No enrichment"))
      if (nrow(.clust_desc_display) > 0) {
        print(knitr::kable(
          .clust_desc_display,
          caption = sprintf("Cluster functional descriptions — %s %s", .list_name, .net_type),
          col.names = c("Cluster", "Genes", "Representative genes", "Top enriched terms")
        ) %>% kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"), full_width = TRUE))
      }

      # Named vector for legend labels: cluster id → short description
      # Exclude uninformative labels so they don't clutter the legend
      .cl_labels <- setNames(.clust_desc$description, as.character(.clust_desc$cluster))
      .cl_labels <- .cl_labels[!.cl_labels %in% c("Too small", "No enrichment")]
    } else {
      .cl_labels <- NULL
    }

    # Network plot saved to PNG and rendered inline
    .png_path <- file.path(
      string_networks_dir,
      sprintf("%s_%s_network.png", .list_name, .net_type)
    )
    png(.png_path, width = 1800, height = 1500, res = 160)
    plot_cluster_network(
      .net_conn, .clusters_df,
      title          = sprintf("%s — %s network (score ≥ %s)", .list_name, .net_type, params$network_score_thr),
      id_to_gene     = .id2gene,
      cluster_labels = .cl_labels
    )
    dev.off()
    # Emit image via cat() so it lands in the text stream at the correct tab position.
    # Calling plot_cluster_network() inline here causes knitr to flush the plot
    # through the plot hook *after* the current text buffer, shifting it one tab forward.
    cat(sprintf("![](%s)\n\n", .png_path))

    network_summary_rows[[length(network_summary_rows) + 1]] <- tibble(
      gene_list  = .list_name, net_type = .net_type,
      n_input    = length(.string_ids), n_connected = .n_nodes,
      n_edges    = .n_edges, n_clusters = .n_clust
    )
  }
}

if (length(network_summary_rows) > 0) {
  .net_summary <- dplyr::bind_rows(network_summary_rows)
  readr::write_tsv(.net_summary, file.path(string_networks_dir, "network_summary.tsv"))
  cat("\n\n## Network summary\n\n")
  print(knitr::kable(.net_summary, caption = "STRING network summary") %>%
          kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")))
}
```

# Visualisations {.tabset}

```{r plots, fig.width=12, fig.height=8, results='asis'}
if (nrow(all_results) > 0) {

  # Detect key columns (STRING may return 'description' or 'term').
  # 'p.adjust' is included for STRINGdb versions that use that name; the
  # run_string_enrichment chunk normalises it to 'fdr' at source, but we
  # keep the full list here as a belt-and-suspenders safeguard.
  pval_col <- c("fdr", "p_value", "pvalue", "p.adjust", "FDR")
  pval_col <- pval_col[pval_col %in% names(all_results)][1]
  if (is.na(pval_col) || length(pval_col) == 0) {
    warning(sprintf(
      "No recognised FDR/p-value column found in all_results. Available columns: %s. Plots may be empty.",
      paste(names(all_results), collapse = ", ")
    ))
    pval_col <- names(all_results)[1]  # last-resort: use first column
  }
  desc_col <- c("description", "term", "name")
  desc_col <- desc_col[desc_col %in% names(all_results)][1]
  if (is.na(desc_col) || length(desc_col) == 0) desc_col <- "description"

  std_results <- all_results %>%
    mutate(
      FDR_std       = suppressWarnings(as.numeric(.data[[pval_col]])),
      FDR_std       = pmax(FDR_std, .Machine$double.xmin, na.rm = TRUE),
      term_std      = as.character(.data[[desc_col]]),
      neglog10FDR   = -log10(FDR_std)
    ) %>%
    filter(!is.na(FDR_std), !is.na(term_std), term_std != "")

  std_results <- std_results %>%
    mutate(scenario = {
      # Match _top_ or _bottom_ as a delimited token anywhere in the filename stem.
      # Files without either token (i.e. global lists) fall back to "global".
      s <- stringr::str_match(gene_list, "_(top|bottom)(?:_|$)")[, 2]
      dplyr::if_else(is.na(s), "global", s)
    })
  std_results_all    <- std_results
  present_scenarios  <- intersect(c("global", "top", "bottom"), unique(std_results$scenario))

  for (.scen in present_scenarios) {
    std_results <- std_results_all %>% dplyr::filter(scenario == .scen)
    if (length(present_scenarios) > 1) {
      cat(sprintf("\n\n## %s {.tabset}\n\n", tools::toTitleCase(.scen)))
      .gl_hdr <- "###"
    } else {
      .gl_hdr <- "##"
    }

  # ── Per-list enrichment barplots (typical STRINGdb style) ──────────────
  # One plot per gene list; terms coloured by STRING functional category.
  # Mirrors the canonical barplot shown in the STRINGdb Bioconductor vignette.
  for (.gl in unique(std_results$gene_list)) {
    .gl_dat <- std_results %>%
      dplyr::filter(gene_list == .gl) %>%
      dplyr::arrange(FDR_std) %>%
      dplyr::slice_head(n = as.integer(params$top_thr)) %>%
      dplyr::mutate(term_short = stringr::str_trunc(term_std, 65))

    if (nrow(.gl_dat) == 0) next

    cat(sprintf("\n\n%s %s\n\n", .gl_hdr, .gl))

    .n_cats <- dplyr::n_distinct(.gl_dat$category)
    .p_bar <- ggplot(
      .gl_dat,
      aes(x = reorder(term_short, neglog10FDR), y = neglog10FDR, fill = category)
    ) +
      geom_col() +
      coord_flip() +
      safe_fill(.n_cats) +
      theme_minimal(base_size = 11) +
      labs(
        title    = .gl,
        subtitle = sprintf("Top %d enriched STRING terms (FDR ≤ %s)", nrow(.gl_dat), params$fdr_thr),
        x        = NULL,
        y        = expression(-log[10](FDR)),
        fill     = "Category"
      ) +
      theme(plot.title = element_text(face = "bold"), axis.text.y = element_text(size = 9))
    print(.p_bar)
    ggsave(
      file.path(string_plots_dir, sprintf("string_enrichment_%s_%s.png", .gl, .scen)),
      .p_bar,
      width  = 13,
      height = max(5, nrow(.gl_dat) * 0.35 + 2),
      dpi    = 320
    )
  } # end for (.gl in unique(std_results$gene_list))

  } # end for (.scen in present_scenarios)
  if (exists("std_results_all")) std_results <- std_results_all
}
```

## Session Info

```{r session_info}
sessionInfo()
```
