---
title: "WebGestalt ORA Analysis"
author: "PhyloPhere Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    code_folding: hide
    df_print: paged
params:
  background_file: NULL
  background_basename: NULL
  gene_list_files: NULL
  output_dir: "ora"
  project_name: "ORA_WebGestalt"
  organism: "hsapiens"
  ora_databases: "geneontology_Biological_Process_noRedundant,geneontology_Cellular_Component_noRedundant,geneontology_Molecular_Function_noRedundant,pathway_KEGG,pathway_Reactome,pathway_WikiPathways"
  min_num: 5
  max_num: 500
  fdr_thr: 0.1
  report_num: 20
  top_thr: 15
  n_threads: 8
  enable_spider_plot: false
  enable_overlap_heatmap: true
  enable_enrichment_map: true
  overlap_top_terms: 50
  compare_metric: "jaccard"
  enable_enrichment_heatmap: true
  enable_upset_plot: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7,
  dev = 'png',
  dpi = 320
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
  library(kableExtra)
  library(clusterProfiler)
  library(org.Hs.eg.db)
  library(WebGestaltR)
  library(paletteer)
})

# ── Explicit namespace guards ─────────────────────────────────────────────────
# BioConductor packages (clusterProfiler, org.Hs.eg.db, S4Vectors, AnnotationDbi)
# mask several tidyverse functions.  Re-pin the ones we use to their dplyr/tidyr
# originals so pipe chains behave as expected regardless of load order.
filter   <- dplyr::filter
select   <- dplyr::select
rename   <- dplyr::rename
mutate   <- dplyr::mutate
arrange  <- dplyr::arrange
group_by <- dplyr::group_by
count    <- dplyr::count

# ── Shared colour palette (mirrors CT_postproc.Rmd for consistency) ──────────
tableau_10 <- paletteer_d("ggthemes::Tableau_10")
names(tableau_10) <- c(
  "blue", "orange", "vermillion", "bluish_green",
  "green", "yellow", "reddish_purple", "pink",
  "brown", "gray"
)

has_pkg <- function(pkg) requireNamespace(pkg, quietly = TRUE)

`%||%` <- function(a, b) {
  if (!is.null(a) && length(a) > 0 && !all(is.na(a))) a else b
}

createDir <- function(directory) {
  if (!dir.exists(directory)) dir.create(directory, recursive = TRUE, showWarnings = FALSE)
}

read_gene_vector <- function(path) {
  x <- readr::read_lines(path)
  x <- trimws(x)
  x <- x[x != ""]
  unique(x)
}

safe_symbol_to_entrez <- function(genes) {
  if (length(genes) == 0) return(character(0))
  mapped <- suppressWarnings(
    bitr(
      genes,
      fromType = "SYMBOL",
      toType = "ENTREZID",
      OrgDb = org.Hs.eg.db,
      drop = TRUE
    )
  )
  if (is.null(mapped) || nrow(mapped) == 0) return(character(0))
  unique(as.character(mapped$ENTREZID))
}

bg_file <- params$background_file
if (!file.exists(bg_file) && !is.null(params$background_basename)) {
  if (file.exists(params$background_basename)) bg_file <- params$background_basename
}

if (is.null(bg_file) || !file.exists(bg_file)) {
  stop("Error: background_file not found")
}
# Resolve background to absolute path so it survives any working-directory changes
bg_file <- normalizePath(bg_file, mustWork = TRUE)
bg_basename <- basename(bg_file)

# Auto-detect gene list files from the current working directory, excluding the
# background file.  Using normalizePath() ensures that subsequent reads succeed
# even when WebGestaltR changes the working directory during ORA execution.
gene_list_files <- params$gene_list_files
if (is.null(gene_list_files) || length(gene_list_files) == 0) {
  gene_list_files <- list.files(pattern = "\\.txt$", full.names = TRUE)
  gene_list_files <- gene_list_files[basename(gene_list_files) != bg_basename]
}
gene_list_files <- gene_list_files[file.exists(gene_list_files)]
# Always normalise to absolute paths
gene_list_files <- normalizePath(gene_list_files, mustWork = FALSE)
gene_list_files <- gene_list_files[file.exists(gene_list_files)]
if (length(gene_list_files) == 0) stop("Error: no valid gene list files found")

dbs <- strsplit(as.character(params$ora_databases), ",")[[1]]
dbs <- trimws(dbs)
dbs <- dbs[dbs != ""]
if (length(dbs) == 0) stop("Error: ora_databases is empty")

ora_root <- "."
ora_results_dir <- "ora_results"
ora_summary_dir <- "ora_summary"
ora_plots_dir <- "ora_plots"
ora_files_dir <- "ORA_files"
cache_dir <- file.path(ora_files_dir, ".cache")

createDir(ora_root)
createDir(ora_results_dir)
createDir(ora_summary_dir)
createDir(ora_plots_dir)
createDir(ora_files_dir)
createDir(cache_dir)

# Normalise all directory paths to ABSOLUTE paths immediately after creation.
# WebGestaltR calls setwd() internally during HTML-report generation even with
# isParallel=FALSE, so relative paths can resolve to wrong locations once the
# working directory drifts.  Capture the absolute working dir first, then build
# absolute paths so every subsequent file.path() call is wd-independent.
ora_work_dir    <- normalizePath(".", mustWork = TRUE)
ora_results_dir <- file.path(ora_work_dir, ora_results_dir)
ora_summary_dir <- file.path(ora_work_dir, ora_summary_dir)
ora_plots_dir   <- file.path(ora_work_dir, ora_plots_dir)
ora_files_dir   <- file.path(ora_work_dir, ora_files_dir)
cache_dir       <- file.path(ora_work_dir, "ORA_files", ".cache")
```

# ORA configuration

```{r config_overview}
cat("=== ORA configuration ===\n")
cat("Background file:", bg_file, "\n")
cat("Gene list files:", length(gene_list_files), "\n")
cat("Organism:", params$organism, "\n")
cat("Databases:", paste(dbs, collapse = ", "), "\n")
cat("Thresholds: min_num=", params$min_num, ", max_num=", params$max_num, ", fdr_thr=", params$fdr_thr, "\n", sep = "")
cat("Spider plot:", params$enable_spider_plot, "\n")
cat("Overlap heatmap:", params$enable_overlap_heatmap, "\n")
cat("Enrichment map:", params$enable_enrichment_map, "\n")
cat("Overlap top terms:", params$overlap_top_terms, "\n")
cat("Overlap metric:", params$compare_metric, "\n")
```

# Parse background and convert to ENTREZ

```{r parse_background}
bg_symbols <- read_gene_vector(bg_file)
bg_entrez <- safe_symbol_to_entrez(bg_symbols)

cat("Background symbols:", length(bg_symbols), "\n")
cat("Background ENTREZ mapped:", length(bg_entrez), "\n")

if (length(bg_entrez) < params$min_num) {
  stop("Error: background ENTREZ set is too small for ORA")
}
```

# Run ORA per gene list

```{r run_ora}
run_single_ora <- function(gene_file, bg_entrez, dbs) {
  genes_sym <- read_gene_vector(gene_file)
  genes_entrez <- safe_symbol_to_entrez(genes_sym)

  list_name <- tools::file_path_sans_ext(basename(gene_file))
  out_dir <- file.path(ora_results_dir, list_name)
  createDir(out_dir)

  if (length(genes_entrez) < params$min_num) {
    return(list(
      list_name = list_name,
      input_n = length(genes_sym),
      mapped_n = length(genes_entrez),
      status = "skipped_too_few_mapped_genes",
      n_rows = 0,
      results = tibble()
    ))
  }

  res <- tryCatch({
    WebGestaltR(
      enrichMethod = "ORA",
      organism = params$organism,
      enrichDatabase = dbs,
      interestGene = genes_entrez,
      interestGeneType = "entrezgene",
      referenceGene = bg_entrez,
      referenceGeneType = "entrezgene",
      minNum = params$min_num,
      maxNum = params$max_num,
      sigMethod = "fdr",
      fdrMethod = "BH",
      fdrThr = params$fdr_thr,
      topThr = params$top_thr,
      reportNum = params$report_num,
      outputDirectory = out_dir,
      projectName = list_name,
      dagColor = "continuous",
      setCoverNum = 10,
      # isParallel = TRUE is disabled: WebGestaltR worker threads each call
      # setwd() internally, which is a global R operation.  With parallel workers
      # the threads race on the working directory, causing relative-path resolution
      # to break and results to be silently lost.  Sequential execution is safe.
      nThreads = 1L,
      cache = cache_dir,
      isParallel = FALSE
    )
  }, error = function(e) {
    message(sprintf("ORA failed for %s: %s", list_name, e$message))
    NULL
  })

  # WebGestaltR with multiple databases may return a *list* of data frames
  # (one per database) rather than a single data frame.  Coerce to one frame.
  if (!is.null(res) && is.list(res) && !is.data.frame(res)) {
    res <- tryCatch(dplyr::bind_rows(res), error = function(e) NULL)
  }

  # WebGestaltR writes enrichment_results_*.txt to disk *before* returning.
  # If the R call errored out (e.g. during HTML-report generation) or returned
  # an empty frame, recover results by reading the file that was already written.
  if (is.null(res) || (is.data.frame(res) && nrow(res) == 0)) {
    enr_file <- file.path(out_dir,
                          paste0("Project_", list_name),
                          paste0("enrichment_results_", list_name, ".txt"))
    if (file.exists(enr_file)) {
      res_disk <- tryCatch(
        readr::read_tsv(enr_file,
                        col_types = readr::cols(.default = "c"),
                        show_col_types = FALSE),
        error = function(e) NULL
      )
      if (!is.null(res_disk) && nrow(res_disk) > 0) {
        message(sprintf("[ORA] Recovered %d result(s) from disk for %s",
                        nrow(res_disk), list_name))
        res <- res_disk %>%
          dplyr::mutate(
            pValue          = suppressWarnings(as.numeric(pValue)),
            FDR             = suppressWarnings(as.numeric(FDR)),
            enrichmentRatio = suppressWarnings(as.numeric(enrichmentRatio)),
            size            = suppressWarnings(as.integer(size)),
            overlap         = suppressWarnings(as.integer(overlap))
          )
      }
    }
  }

  if (is.null(res) || nrow(res) == 0) {
    return(list(
      list_name = list_name,
      input_n = length(genes_sym),
      mapped_n = length(genes_entrez),
      status = "no_results",
      n_rows = 0,
      results = tibble()
    ))
  }

  res2 <- res %>% mutate(gene_list = list_name)
  readr::write_tsv(res2, file.path(ora_summary_dir, paste0(list_name, "_results.tsv")))

  list(
    list_name = list_name,
    input_n = length(genes_sym),
    mapped_n = length(genes_entrez),
    status = "ok",
    n_rows = nrow(res2),
    results = res2
  )
}

ora_runs <- lapply(gene_list_files, run_single_ora, bg_entrez = bg_entrez, dbs = dbs)
```

# Summary and plots

```{r summary_and_plots}
# Helper: build a fill/color scale that expands Tableau_10 when n > 10
.pal10_base <- as.character(paletteer::paletteer_d("ggthemes::Tableau_10"))
.safe_pal   <- function(n) if (n <= 10L) .pal10_base[seq_len(n)] else colorRampPalette(.pal10_base)(n)
safe_fill   <- function(n) scale_fill_manual(values  = .safe_pal(n))
safe_color  <- function(n) scale_color_manual(values = .safe_pal(n))

run_summary <- purrr::map_dfr(ora_runs, ~tibble(
  gene_list = .x$list_name,
  input_genes = .x$input_n,
  mapped_entrez = .x$mapped_n,
  status = .x$status,
  n_results = .x$n_rows
))

all_results <- purrr::map_dfr(ora_runs, "results")

readr::write_tsv(run_summary, file.path(ora_summary_dir, "ora_run_summary.tsv"))
readr::write_tsv(all_results, file.path(ora_summary_dir, "ora_all_results.tsv"))

kable(run_summary, caption = "ORA run summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

if (nrow(all_results) > 0) {
  fdr_col <- c("FDR", "fdr", "adjP", "fdrBH", "fdr_bh")
  fdr_col <- fdr_col[fdr_col %in% names(all_results)][1] %||% "FDR"
  db_col <- c("database", "enrichmentCategory", "category")
  db_col <- db_col[db_col %in% names(all_results)][1] %||% "database"
  desc_col <- c("description", "geneSet", "term", "name")
  desc_col <- desc_col[desc_col %in% names(all_results)][1] %||% "description"

  std_results <- all_results %>%
    mutate(
      FDR_std = suppressWarnings(as.numeric(.data[[fdr_col]])),
      FDR_std = pmax(FDR_std, .Machine$double.xmin, na.rm = TRUE),
      database_std = as.character(.data[[db_col]]),
      term_std = as.character(.data[[desc_col]]),
      neglog10FDR = -log10(FDR_std)
    ) %>%
    filter(!is.na(FDR_std), !is.na(term_std), term_std != "")

  top_hits <- std_results %>%
    arrange(FDR_std) %>%
    group_by(gene_list) %>%
    slice_head(n = 10) %>%
    ungroup()

  p_hits <- ggplot(top_hits, aes(x = reorder(term_std, neglog10FDR), y = neglog10FDR, fill = gene_list)) +
    geom_col(position = "dodge") +
    coord_flip() +
    safe_fill(n_distinct(top_hits$gene_list)) +
    theme_minimal() +
    labs(title = "Top ORA hits per gene list", x = "Gene set", y = "-log10(FDR)")

  print(p_hits)
  ggsave(file.path(ora_plots_dir, "top_ora_hits_per_list.png"), p_hits, width = 14, height = 9, dpi = 320)

  db_counts <- std_results %>%
    count(gene_list, database_std, sort = TRUE)

  p_db <- ggplot(db_counts, aes(x = gene_list, y = n, fill = database_std)) +
    geom_col(position = "stack") +
    safe_fill(n_distinct(db_counts$database_std)) +
    theme_minimal() +
    labs(title = "Number of enriched terms by database", x = "Gene list", y = "N terms") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  print(p_db)
  ggsave(file.path(ora_plots_dir, "enriched_terms_by_database.png"), p_db, width = 12, height = 7, dpi = 320)

  # Overall characterization: recurrence of enriched terms across lists
  term_recurrence <- std_results %>%
    distinct(gene_list, term_std, database_std, FDR_std) %>%
    group_by(term_std, database_std) %>%
    summarise(
      n_lists = n_distinct(gene_list),
      best_fdr = min(FDR_std, na.rm = TRUE),
      mean_fdr = mean(FDR_std, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(desc(n_lists), best_fdr)

  readr::write_tsv(term_recurrence, file.path(ora_summary_dir, "ora_term_recurrence.tsv"))

  top_recurrent <- term_recurrence %>%
    # slice_head() already returns all available rows when n > nrow(data)
    # (using n() here errors in recent dplyr because n() is only valid
    # inside data-masking verbs such as mutate/summarise/filter).
    slice_head(n = 30L) %>%
    mutate(term_short = stringr::str_trunc(term_std, 70))

  if (nrow(top_recurrent) > 0) {
    p_recurrent <- ggplot(top_recurrent, aes(x = reorder(term_short, n_lists), y = n_lists, fill = database_std)) +
      geom_col() +
      coord_flip() +
      safe_fill(n_distinct(top_recurrent$database_std)) +
      theme_minimal() +
      labs(
        title = "Most recurrent enriched functions across gene lists",
        x = "Term",
        y = "# gene lists where term is enriched"
      )

    print(p_recurrent)
    ggsave(file.path(ora_plots_dir, "ora_recurrent_terms.png"), p_recurrent, width = 13, height = 10, dpi = 320)
  }

  # Pairwise overlap characterization (Jaccard / overlap coefficient)
  if (isTRUE(params$enable_overlap_heatmap)) {
    top_n <- as.integer(params$overlap_top_terms)
    top_n <- ifelse(is.na(top_n) || top_n <= 0, 50, top_n)

    term_sets <- std_results %>%
      arrange(FDR_std) %>%
      group_by(gene_list) %>%
      slice_head(n = top_n) %>%
      summarise(term_set = list(unique(term_std)), .groups = "drop")

    glists <- term_sets$gene_list
    overlap_mat <- matrix(0, nrow = length(glists), ncol = length(glists), dimnames = list(glists, glists))

    for (i in seq_along(glists)) {
      for (j in seq_along(glists)) {
        a <- term_sets$term_set[[i]]
        b <- term_sets$term_set[[j]]
        inter <- length(intersect(a, b))
        union_n <- length(unique(c(a, b)))
        min_n <- min(length(a), length(b))

        metric <- tolower(as.character(params$compare_metric %||% "jaccard"))
        overlap_val <- if (metric == "overlap") {
          ifelse(min_n > 0, inter / min_n, 0)
        } else {
          ifelse(union_n > 0, inter / union_n, 0)
        }

        overlap_mat[i, j] <- overlap_val
      }
    }

    overlap_df <- as.data.frame(as.table(overlap_mat)) %>%
      rename(gene_list_x = Var1, gene_list_y = Var2, overlap = Freq)

    readr::write_tsv(overlap_df, file.path(ora_summary_dir, "ora_overlap_matrix.tsv"))

    p_overlap <- ggplot(overlap_df, aes(x = gene_list_x, y = gene_list_y, fill = overlap)) +
      geom_tile(color = "white") +
      geom_text(aes(label = sprintf("%.2f", overlap)), size = 3) +
      scale_fill_viridis_c(option = "C") +
      theme_minimal() +
      labs(
        title = sprintf("Pairwise enrichment overlap (%s)", params$compare_metric),
        subtitle = sprintf("Computed from top %d enriched terms per list", top_n),
        x = NULL,
        y = NULL
      ) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))

    print(p_overlap)
    ggsave(file.path(ora_plots_dir, "ora_overlap_heatmap.png"), p_overlap, width = 10, height = 8, dpi = 320)
  }

  # clusterProfiler-style pooled comparison (dotplot-like)
  pooled_dot <- std_results %>%
    arrange(FDR_std) %>%
    group_by(gene_list) %>%
    slice_head(n = 15) %>%
    ungroup() %>%
    mutate(term_short = stringr::str_trunc(term_std, 60))

  if (nrow(pooled_dot) > 0) {
    p_dot <- ggplot(pooled_dot, aes(x = gene_list, y = term_short, size = neglog10FDR, color = database_std)) +
      geom_point(alpha = 0.85) +
      safe_color(n_distinct(pooled_dot$database_std)) +
      theme_minimal() +
      labs(
        title = "Pooled ORA characterization across lists",
        subtitle = "Top enriched terms per list (dot size = -log10(FDR))",
        x = "Gene list",
        y = "Enriched term"
      ) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))

    print(p_dot)
    ggsave(file.path(ora_plots_dir, "ora_pooled_dotplot.png"), p_dot, width = 14, height = 11, dpi = 320)
  }

  # Optional enrichment map/network visualization if dependencies are available
  if (isTRUE(params$enable_enrichment_map) && has_pkg("igraph")) {
    term_links <- std_results %>%
      distinct(gene_list, term_std) %>%
      group_by(term_std) %>%
      summarise(gene_lists = list(sort(unique(gene_list))), n_lists = n(), .groups = "drop") %>%
      filter(n_lists > 1)

    if (nrow(term_links) > 0) {
      edges <- term_links %>%
        mutate(pairs = purrr::map(gene_lists, ~{
          x <- .x
          if (length(x) < 2) return(tibble(from = character(0), to = character(0)))
          as_tibble(t(combn(x, 2)), .name_repair = "minimal") %>%
            setNames(c("from", "to"))
        })) %>%
        select(term_std, pairs) %>%
        tidyr::unnest(pairs) %>%
        count(from, to, name = "shared_terms")

      if (nrow(edges) > 0) {
        p_network <- ggplot(edges, aes(x = from, y = to, fill = shared_terms)) +
          geom_tile(color = "white") +
          scale_fill_viridis_c(option = "B") +
          theme_minimal() +
          labs(
            title = "Shared enriched functions between gene lists",
            subtitle = "Cell value = number of overlapping enriched terms",
            x = NULL,
            y = NULL
          ) +
          theme(axis.text.x = element_text(angle = 45, hjust = 1))

        print(p_network)
        ggsave(file.path(ora_plots_dir, "ora_shared_term_network_matrix.png"), p_network, width = 10, height = 8, dpi = 320)
      }
    }
  }
}
```

# Enrichment signal heatmap

Rows are the union of the top-N enriched terms per gene list; columns are the
gene list scenarios. Cell colour encodes **-log10(FDR)**; zero (white) means the
term was not enriched in that scenario. Hierarchical clustering (when ≥ 3
scenarios) reveals groups of scenarios with similar functional landscapes.
A separate per-database heatmap is produced for each functional category so
terms from different ontologies are compared on the same scale.

```{r enrichment_heatmap}
if (exists("std_results") && nrow(std_results) > 0 && isTRUE(params$enable_enrichment_heatmap)) {

  hm_top_n <- as.integer(params$overlap_top_terms %||% 30)
  hm_top_n <- if (is.na(hm_top_n) || hm_top_n <= 0) 30L else min(hm_top_n, 50L)

  # Union of top enriched terms across all lists
  hm_terms <- std_results %>%
    arrange(FDR_std) %>%
    group_by(gene_list) %>%
    slice_head(n = hm_top_n) %>%
    ungroup() %>%
    distinct(term_std) %>%
    pull(term_std)

  if (length(hm_terms) > 0) {

    # Build term × list signal matrix (-log10 FDR; 0 = not enriched)
    hm_wide <- std_results %>%
      filter(term_std %in% hm_terms) %>%
      group_by(gene_list, term_std) %>%
      summarise(signal = max(neglog10FDR, na.rm = TRUE), .groups = "drop") %>%
      tidyr::pivot_wider(names_from = gene_list, values_from = signal, values_fill = 0) %>%
      tibble::column_to_rownames("term_std")

    rownames(hm_wide) <- stringr::str_trunc(rownames(hm_wide), 65)
    hm_mat <- as.matrix(hm_wide)
    hm_mat <- hm_mat[order(rowSums(hm_mat), decreasing = TRUE), , drop = FALSE]

    cat(sprintf("Enrichment heatmap: %d terms × %d gene lists\n", nrow(hm_mat), ncol(hm_mat)))

    # ── render function (pheatmap or ggplot fallback) ──────────────────────
    draw_signal_heatmap <- function(mat, title, palette_high, out_png) {
      if (has_pkg("pheatmap")) {
        library(pheatmap)
        p_out <- pheatmap(
          mat,
          color        = colorRampPalette(c("white", palette_high))(100),
          cluster_rows = nrow(mat) > 2,
          cluster_cols = ncol(mat) > 2,
          fontsize_row = max(5, min(10, 200 / nrow(mat))),
          fontsize_col = 9,
          main         = title,
          angle_col    = 45,
          border_color = NA,
          silent       = TRUE
        )
        png(out_png,
            width  = max(1200, ncol(mat) * 130 + 800),
            height = max(800,  nrow(mat) * 24  + 250), res = 160)
        grid::grid.newpage()
        grid::grid.draw(p_out$gtable)
        dev.off()
        grid::grid.newpage()
        grid::grid.draw(p_out$gtable)
      } else {
        long_df <- as.data.frame(mat) %>%
          tibble::rownames_to_column("term") %>%
          tidyr::pivot_longer(-term, names_to = "gene_list", values_to = "signal") %>%
          mutate(term = factor(term, levels = rev(rownames(mat))))
        p_gg <- ggplot(long_df, aes(x = gene_list, y = term, fill = signal)) +
          geom_tile(color = "grey90", linewidth = 0.15) +
          scale_fill_gradient(low = "white", high = palette_high, name = "-log10(FDR)") +
          theme_minimal(base_size = 16) +
          labs(title = title,
               subtitle = "-log10(FDR) | rows = enriched terms, cols = scenarios",
               x = NULL, y = NULL) +
          theme(
            axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
            axis.text.y = element_text(size = max(5, min(8, 180 / nrow(mat))))
          )
        print(p_gg)
        ggsave(out_png, p_gg,
               width  = max(8, ncol(mat) * 1.5 + 3),
               height = max(6, nrow(mat) * 0.28 + 2), dpi = 300)
      }
    }

    # Global heatmap (all databases pooled)
    draw_signal_heatmap(
      hm_mat,
      title       = "Enrichment signal: -log10(FDR) per term × scenario (all databases)",
      palette_high = as.character(tableau_10[["blue"]]),
      out_png     = file.path(ora_plots_dir, "ora_enrichment_heatmap.png")
    )

    # Per-database heatmaps
    for (db_i in unique(std_results$database_std)) {
      db_res <- std_results %>% filter(database_std == db_i)
      db_terms <- db_res %>%
        arrange(FDR_std) %>%
        group_by(gene_list) %>%
        slice_head(n = 20) %>%
        ungroup() %>%
        distinct(term_std) %>%
        pull(term_std)
      if (length(db_terms) < 2) next

      db_wide <- db_res %>%
        filter(term_std %in% db_terms) %>%
        group_by(gene_list, term_std) %>%
        summarise(signal = max(neglog10FDR, na.rm = TRUE), .groups = "drop") %>%
        tidyr::pivot_wider(names_from = gene_list, values_from = signal, values_fill = 0) %>%
        tibble::column_to_rownames("term_std")

      rownames(db_wide) <- stringr::str_trunc(rownames(db_wide), 60)
      db_mat <- as.matrix(db_wide)
      db_mat <- db_mat[order(rowSums(db_mat), decreasing = TRUE), , drop = FALSE]
      db_slug <- gsub("[^A-Za-z0-9_]", "_", db_i)

      draw_signal_heatmap(
        db_mat,
        title        = paste0("-log10(FDR) | ", db_i),
        palette_high  = as.character(tableau_10[["vermillion"]]),
        out_png      = file.path(ora_plots_dir, paste0("ora_heatmap_", db_slug, ".png"))
      )
    }

  } else {
    cat("No enriched terms found — enrichment heatmap skipped.\n")
  }
} else {
  cat("Enrichment heatmap disabled or no results available.\n")
}
```

# UpSet plot: term intersections across scenarios

Each gene list / scenario is treated as a **set of enriched terms**. The UpSet
plot decomposes those sets into their exact intersections — revealing terms
significantly enriched in every scenario, terms unique to a single scenario, and
all gradations in between. This is more informative than a pairwise Jaccard score
because it shows the precise multi-way structure of the overlap.

Requires the **UpSetR** (preferred) or **ComplexUpset** package.
Install with `install.packages("UpSetR")`.

```{r upset_plot, fig.width=14, fig.height=8}
if (exists("std_results") && nrow(std_results) > 0 && isTRUE(params$enable_upset_plot)) {

  upset_top_n <- as.integer(params$overlap_top_terms %||% 50)
  upset_top_n <- if (is.na(upset_top_n) || upset_top_n <= 0) 50L else min(upset_top_n, 75L)

  upset_terms <- std_results %>%
    arrange(FDR_std) %>%
    group_by(gene_list) %>%
    slice_head(n = upset_top_n) %>%
    ungroup() %>%
    distinct(gene_list, term_std)

  all_terms_u <- unique(upset_terms$term_std)
  glists_u    <- unique(upset_terms$gene_list)

  if (length(glists_u) >= 2 && length(all_terms_u) > 0) {

    # Binary membership matrix: rows = terms, cols = gene lists
    upset_mat <- as.data.frame(
      setNames(
        lapply(glists_u, function(gl) {
          as.integer(all_terms_u %in% upset_terms$term_std[upset_terms$gene_list == gl])
        }),
        glists_u
      )
    )
    rownames(upset_mat) <- all_terms_u
    upset_mat <- upset_mat[rowSums(upset_mat) >= 1, , drop = FALSE]

    cat(sprintf("UpSet input: %d terms × %d gene lists\n", nrow(upset_mat), ncol(upset_mat)))

    if (has_pkg("UpSetR")) {
      library(UpSetR)
      upset_args <- list(
        data            = upset_mat,
        sets            = rev(glists_u),
        order.by        = "freq",
        decreasing      = TRUE,
        mb.ratio        = c(0.60, 0.40),
        text.scale      = c(1.5, 1.2, 1.1, 1.0, 1.2, 1.0),
        mainbar.y.label = "# shared enriched terms",
        sets.x.label    = "Total enriched terms per scenario",
        main.bar.color  = "#4393c3",
        sets.bar.color  = "#2166ac",
        matrix.color    = "#08306b",
        point.size      = 3.0,
        line.size       = 0.8,
        nintersects     = min(40L, 2L ^ length(glists_u))
      )
      # Save to file
      png(
        file.path(ora_plots_dir, "ora_upset_terms.png"),
        width  = max(1600, length(glists_u) * 200 + 600),
        height = max(900,  length(glists_u) * 80  + 400),
        res    = 150
      )
      do.call(upset, upset_args)
      dev.off()
      # Render inline
      do.call(upset, upset_args)

    } else if (has_pkg("ComplexUpset")) {
      library(ComplexUpset)
      upset_tbl <- upset_mat %>%
        tibble::rownames_to_column("term") %>%
        dplyr::mutate(dplyr::across(all_of(glists_u), as.logical))
      p_upset_cx <- upset(
        upset_tbl, glists_u,
        name        = "Gene list",
        width_ratio = 0.25,
        min_size    = 1,
        sort_sets   = FALSE,
        base_annotations = list(
          "Shared terms" = intersection_size(
            counts = TRUE,
            mapping = aes(fill = "bar")
          ) + scale_fill_manual(values = c("bar" = "#4393c3"), guide = "none")
        )
      ) +
        labs(
          title    = "UpSet: intersection of enriched terms across gene lists",
          subtitle = sprintf("Top %d enriched terms per scenario", upset_top_n)
        )
      print(p_upset_cx)
      ggsave(
        file.path(ora_plots_dir, "ora_upset_terms.png"), p_upset_cx,
        width  = max(12, length(glists_u) * 2 + 6),
        height = max(8,  length(glists_u) * 1.2 + 4), dpi = 200
      )

    } else {
      # Fallback: manual intersection decomposition rendered as a bar chart
      message("UpSetR not found — falling back to intersection bar chart. Install with: install.packages('UpSetR')")
      n_lists <- length(glists_u)
      if (n_lists <= 6) {
        all_combos <- do.call(c, lapply(seq_len(n_lists), function(k) combn(glists_u, k, simplify = FALSE)))
        combo_df <- purrr::map_dfr(all_combos, function(combo) {
          in_all   <- rowSums(upset_mat[, combo, drop = FALSE]) == length(combo)
          out_rest <- if (length(combo) < n_lists) {
            rowSums(upset_mat[, setdiff(glists_u, combo), drop = FALSE]) == 0
          } else {
            rep(TRUE, nrow(upset_mat))
          }
          tibble::tibble(
            intersection = paste(sort(combo), collapse = " ∩ "),
            n_terms      = sum(in_all & out_rest),
            n_sets       = length(combo)
          )
        }) %>%
          filter(n_terms > 0) %>%
          arrange(desc(n_terms))

        p_fallback <- ggplot(
          combo_df %>% slice_head(n = 30),
          aes(x = reorder(intersection, n_terms), y = n_terms, fill = factor(n_sets))
        ) +
          geom_col() +
          coord_flip() +
          scale_fill_paletteer_d("ggthemes::Tableau_10", name = "# lists") +
          theme_minimal() +
          labs(
            title    = "Enriched term intersections across gene lists",
            subtitle = "(UpSetR not installed — simplified view)",
            x        = NULL,
            y        = "# shared terms"
          )
        print(p_fallback)
        ggsave(
          file.path(ora_plots_dir, "ora_upset_fallback.png"), p_fallback,
          width = 12, height = 8, dpi = 300
        )
      } else {
        cat("Too many gene lists for the fallback intersection chart (> 6). Please install UpSetR.\n")
      }
    }

  } else {
    cat("UpSet plot requires ≥ 2 gene lists with enrichment results — skipping.\n")
  }
} else {
  cat("UpSet plot disabled or no results available.\n")
}
```

## Session Info

```{r session_info}
sessionInfo()
```
