---
title: "STRING Enrichment Analysis"
author: "PhyloPhere Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    code_folding: hide
    df_print: paged
params:
  background_file: NULL
  background_basename: NULL
  gene_list_files: NULL
  output_dir: "string"
  project_name: "STRING_Analysis"
  species: 9606
  required_score: 400
  fdr_thr: 0.1
  top_thr: 15
  report_num: 20
  enable_spider_plot: true
  enable_overlap_heatmap: true
  enable_enrichment_map: true
  enable_ppi_enrichment: true
  overlap_top_terms: 50
  compare_metric: "jaccard"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7,
  dev = 'png',
  dpi = 320
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
  library(kableExtra)
  library(STRINGdb)
})

has_pkg <- function(pkg) requireNamespace(pkg, quietly = TRUE)

`%||%` <- function(a, b) {
  if (!is.null(a) && length(a) > 0 && !all(is.na(a))) a else b
}

createDir <- function(directory) {
  if (!dir.exists(directory)) dir.create(directory, recursive = TRUE, showWarnings = FALSE)
}

download_with_retry <- function(url, destfile, tries = 4L, timeout_sec = 1800L) {
  old_timeout <- getOption("timeout")
  on.exit(options(timeout = old_timeout), add = TRUE)
  options(timeout = max(as.integer(old_timeout), as.integer(timeout_sec), 600L))

  for (i in seq_len(tries)) {
    ok <- tryCatch({
      utils::download.file(url, destfile = destfile, mode = "wb", quiet = FALSE, method = "libcurl")
      TRUE
    }, error = function(e) {
      message(sprintf("Download failed (%d/%d): %s", i, tries, e$message))
      FALSE
    })

    if (isTRUE(ok) && file.exists(destfile) && file.size(destfile) > 0) return(invisible(TRUE))
    if (i < tries) Sys.sleep(min(30, i * 5))
  }

  stop(sprintf("Failed to download after %d attempts: %s", tries, url))
}

ensure_string_cache <- function(species, version = "12.0", input_directory = "string_cache") {
  createDir(input_directory)

  species <- as.integer(species)
  base <- "https://stringdb-downloads.org/download"
  files <- c(
    sprintf("protein.aliases.v%s/%d.protein.aliases.v%s.txt.gz", version, species, version),
    sprintf("protein.info.v%s/%d.protein.info.v%s.txt.gz", version, species, version),
    sprintf("protein.links.v%s/%d.protein.links.v%s.txt.gz", version, species, version)
  )

  for (rel in files) {
    url <- sprintf("%s/%s", base, rel)
    dest <- file.path(input_directory, basename(rel))
    if (!file.exists(dest) || file.size(dest) == 0) {
      message(sprintf("Caching STRING file: %s", basename(rel)))
      download_with_retry(url, destfile = dest)
    }
  }

  invisible(normalizePath(input_directory, winslash = "/", mustWork = FALSE))
}

read_gene_vector <- function(path) {
  x <- readr::read_lines(path)
  x <- trimws(x)
  x <- x[x != ""]
  unique(x)
}

string_root <- "."
string_results_dir <- "string_results"
string_summary_dir <- "string_summary"
string_plots_dir   <- "string_plots"
string_cache_dir   <- "string_cache"

createDir(string_results_dir)
createDir(string_summary_dir)
createDir(string_plots_dir)
createDir(string_cache_dir)

# Pre-cache STRING DB files with retry/timeout guards to avoid transient
# network failures during set_background()/enrichment calls.
ensure_string_cache(
  species = as.integer(params$species),
  version = "12.0",
  input_directory = string_cache_dir
)

# Initialise a STRINGdb instance (R5 reference class).
# score_threshold is set low here; filtering by required_score happens in
# get_interactions() / plot_network() calls, not in enrichment.
string_db <- STRINGdb$new(
  version          = "12.0",
  species          = as.integer(params$species),
  score_threshold  = as.integer(params$required_score),
  input_directory  = string_cache_dir
)

# Map a character vector of gene SYMBOLs to STRING identifiers.
# Returns a character vector of STRING_id values.
safe_string_map_ids <- function(genes) {
  if (length(genes) == 0) return(character(0))
  df_in <- data.frame(gene = genes, stringsAsFactors = FALSE)
  mapped <- tryCatch(
    string_db$map(df_in, "gene", removeUnmappedRows = TRUE, quiet = TRUE),
    error = function(e) {
      message(sprintf("STRING ID mapping failed: %s", e$message))
      NULL
    }
  )
  if (is.null(mapped) || nrow(mapped) == 0) return(character(0))
  unique(as.character(mapped$STRING_id))
}

# Resolve background file path
bg_file <- params$background_file
if (!file.exists(bg_file) && !is.null(params$background_basename)) {
  if (file.exists(params$background_basename)) bg_file <- params$background_basename
}

if (is.null(bg_file) || !file.exists(bg_file)) {
  stop("Error: background_file not found")
}

gene_list_files <- params$gene_list_files
if (is.null(gene_list_files) || length(gene_list_files) == 0) {
  gene_list_files <- list.files(pattern = "\\.txt$", full.names = TRUE)
}
gene_list_files <- gene_list_files[file.exists(gene_list_files)]
if (length(gene_list_files) == 0) stop("Error: no valid gene list files found")


```

# STRING analysis configuration

```{r config_overview}
cat("=== STRING configuration ===\n")
cat("Background file:", bg_file, "\n")
cat("Gene list files:", length(gene_list_files), "\n")
cat("Species (NCBI taxon ID):", params$species, "\n")
cat("Required interaction score:", params$required_score, "\n")
cat("FDR threshold:", params$fdr_thr, "\n")
cat("Top terms per list:", params$top_thr, "\n")
cat("PPI enrichment test:", params$enable_ppi_enrichment, "\n")
cat("Spider plot:", params$enable_spider_plot, "\n")
cat("Overlap heatmap:", params$enable_overlap_heatmap, "\n")
cat("Enrichment map:", params$enable_enrichment_map, "\n")
cat("Overlap top terms:", params$overlap_top_terms, "\n")
cat("Overlap metric:", params$compare_metric, "\n")
```

# Map background to STRING IDs

> **Note:** A custom background constrains enrichment to the experimental proteome.
> When background genes are provided, they are mapped to STRING identifiers and set
> on the `STRINGdb` object via `set_background()`, which applies to all subsequent
> `get_enrichment()` calls.

```{r parse_background}
bg_symbols  <- read_gene_vector(bg_file)
bg_string_ids <- safe_string_map_ids(bg_symbols)

cat("Background symbols supplied:", length(bg_symbols), "\n")
cat("Background STRING IDs successfully mapped:", length(bg_string_ids), "\n")

if (length(bg_string_ids) < 5) {
  warning(
    sprintf(
      "Only %d background genes mapped to STRING IDs. ",
      length(bg_string_ids)
    ),
    "Proceeding without a custom background (STRING will use the full proteome)."
  )
  use_custom_background <- FALSE
} else {
  # Set the background on the STRINGdb object so all subsequent enrichment
  # calls are constrained to the experimental proteome.
  use_custom_background <- tryCatch({
    string_db$set_background(bg_string_ids)
    TRUE
  }, error = function(e) {
    warning(
      sprintf("Could not set custom STRING background: %s", e$message),
      " Proceeding with STRING default background."
    )
    FALSE
  })

  if (isTRUE(use_custom_background)) {
    cat("Custom background set on STRINGdb object (", length(bg_string_ids), " proteins).\n")
  }
}
```

# Run STRING enrichment per gene list

```{r run_string_enrichment}
MIN_MAPPED <- 5L

run_single_string <- function(gene_file, use_custom_background) {
  genes_sym   <- read_gene_vector(gene_file)
  list_name   <- tools::file_path_sans_ext(basename(gene_file))

  # Map gene list symbols to STRING IDs
  gene_string_ids <- safe_string_map_ids(genes_sym)

  if (length(gene_string_ids) < MIN_MAPPED) {
    return(list(
      list_name      = list_name,
      input_n        = length(genes_sym),
      mapped_n       = length(gene_string_ids),
      status         = "skipped_too_few_mapped_genes",
      n_rows         = 0L,
      results        = tibble(),
      ppi_result     = NULL
    ))
  }

  # в”Ђв”Ђ Functional enrichment в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
  # get_enrichment() returns a flat data.frame with a 'category' column.
  # The background is already set globally on string_db via set_background().
  enrich_df <- tryCatch(
    {
      res <- string_db$get_enrichment(gene_string_ids)
      if (!is.null(res) && nrow(res) > 0) {
        # STRINGdb may return an error frame with columns 'error'/'message'
        # (e.g. background_error when mapped genes fall outside the custom
        # background set) instead of real enrichment rows.  Detect and skip.
        if ("error" %in% names(res)) {
          message(sprintf(
            "STRING enrichment API error for %s: %s",
            list_name,
            paste(as.character(res[["message"]]), collapse = "; ")
          ))
          tibble()
        } else {
          res %>% mutate(gene_list = list_name)
        }
      } else {
        tibble()
      }
    },
    error = function(e) {
      message(sprintf("STRING enrichment failed for %s: %s", list_name, e$message))
      tibble()
    }
  )

  if (nrow(enrich_df) == 0) {
    return(list(
      list_name  = list_name,
      input_n    = length(genes_sym),
      mapped_n   = length(gene_string_ids),
      status     = "no_results",
      n_rows     = 0L,
      results    = tibble(),
      ppi_result = NULL
    ))
  }

  # Standardise FDR / p-value column (STRINGdb uses 'fdr' and 'p_value')
  pval_col <- c("fdr", "p_value", "pvalue", "p.adjust", "FDR")
  pval_col <- pval_col[pval_col %in% names(enrich_df)][1] %||% NA_character_

  if (!is.na(pval_col)) {
    enrich_df <- enrich_df %>%
      filter(suppressWarnings(as.numeric(.data[[pval_col]])) <= params$fdr_thr)
    # Rename to canonical 'fdr' so all downstream consumers (all_results,
    # plots chunk, accumulation) see a consistent column name regardless of
    # the STRINGdb version that produced it.
    if (pval_col != "fdr") {
      enrich_df <- enrich_df %>% dplyr::rename(fdr = dplyr::all_of(pval_col))
    }
  }

  readr::write_tsv(
    enrich_df,
    file.path(string_summary_dir, paste0(list_name, "_results.tsv"))
  )

  # в”Ђв”Ђ PPI density enrichment в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
  ppi_res <- NULL
  if (isTRUE(params$enable_ppi_enrichment)) {
    ppi_res <- tryCatch(
      string_db$get_ppi_enrichment(gene_string_ids),
      error = function(e) {
        message(sprintf("PPI enrichment failed for %s: %s", list_name, e$message))
        NULL
      }
    )

    if (!is.null(ppi_res)) {
      # Replace any NULL fields with NA before coercing to tibble; STRINGdb
      # returns NULL for 'enrichment' and 'lambda' when no interactions are
      # found, which makes as_tibble() abort with "column is NULL".
      ppi_lst <- lapply(as.list(ppi_res), function(x) if (is.null(x)) NA else x)
      ppi_df  <- as_tibble(ppi_lst) %>% mutate(gene_list = list_name)
      readr::write_tsv(
        ppi_df,
        file.path(string_summary_dir, paste0(list_name, "_ppi.tsv"))
      )
    }
  }

  list(
    list_name  = list_name,
    input_n    = length(genes_sym),
    mapped_n   = length(gene_string_ids),
    status     = "ok",
    n_rows     = nrow(enrich_df),
    results    = enrich_df,
    ppi_result = ppi_res
  )
}

string_runs <- lapply(
  gene_list_files,
  run_single_string,
  use_custom_background = use_custom_background
)
```

# Summary tables

```{r summary_tables}
run_summary <- purrr::map_dfr(string_runs, ~tibble(
  gene_list     = .x$list_name,
  input_genes   = .x$input_n,
  mapped_string = .x$mapped_n,
  status        = .x$status,
  n_results     = .x$n_rows
))

all_results <- purrr::map_dfr(string_runs, "results")

readr::write_tsv(run_summary, file.path(string_summary_dir, "string_run_summary.tsv"))
readr::write_tsv(all_results, file.path(string_summary_dir, "string_all_results.tsv"))

kable(run_summary, caption = "STRING enrichment run summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

```{r ppi_summary, eval=isTRUE(params$enable_ppi_enrichment)}
ppi_rows <- purrr::keep(string_runs, ~!is.null(.x$ppi_result))
if (length(ppi_rows) > 0) {
  ppi_summary <- purrr::map_dfr(ppi_rows, function(x) {
    pr <- as.list(x$ppi_result)
    tibble(
      gene_list                   = x$list_name,
      number_of_nodes             = pr$number_of_nodes %||% NA_integer_,
      number_of_edges             = pr$number_of_edges %||% NA_integer_,
      average_node_degree         = pr$average_node_degree %||% NA_real_,
      local_clustering_coefficient = pr$local_clustering_coefficient %||% NA_real_,
      expected_number_of_edges    = pr$expected_number_of_edges %||% NA_integer_,
      p_value                     = pr$p_value %||% NA_real_
    )
  })

  readr::write_tsv(ppi_summary, file.path(string_summary_dir, "string_ppi_summary.tsv"))

  kable(ppi_summary, caption = "STRING PPI enrichment (density test) summary") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
}
```

# Visualisations

```{r plots, fig.width=12, fig.height=8}
if (nrow(all_results) > 0) {

  # Detect key columns (STRING may return 'description' or 'term').
  # 'p.adjust' is included for STRINGdb versions that use that name; the
  # run_string_enrichment chunk normalises it to 'fdr' at source, but we
  # keep the full list here as a belt-and-suspenders safeguard.
  pval_col <- c("fdr", "p_value", "pvalue", "p.adjust", "FDR")
  pval_col <- pval_col[pval_col %in% names(all_results)][1]
  if (is.na(pval_col) || length(pval_col) == 0) {
    warning(sprintf(
      "No recognised FDR/p-value column found in all_results. Available columns: %s. Plots may be empty.",
      paste(names(all_results), collapse = ", ")
    ))
    pval_col <- names(all_results)[1]  # last-resort: use first column
  }
  desc_col <- c("description", "term", "name")
  desc_col <- desc_col[desc_col %in% names(all_results)][1]
  if (is.na(desc_col) || length(desc_col) == 0) desc_col <- "description"

  std_results <- all_results %>%
    mutate(
      FDR_std       = suppressWarnings(as.numeric(.data[[pval_col]])),
      FDR_std       = pmax(FDR_std, .Machine$double.xmin, na.rm = TRUE),
      term_std      = as.character(.data[[desc_col]]),
      neglog10FDR   = -log10(FDR_std)
    ) %>%
    filter(!is.na(FDR_std), !is.na(term_std), term_std != "")

  # в”Ђв”Ђ Top hits per list в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
  top_hits <- std_results %>%
    arrange(FDR_std) %>%
    group_by(gene_list) %>%
    slice_head(n = as.integer(params$top_thr)) %>%
    ungroup()

  p_hits <- ggplot(
    top_hits,
    aes(x = reorder(stringr::str_trunc(term_std, 60), neglog10FDR),
        y = neglog10FDR,
        fill = gene_list)
  ) +
    geom_col(position = "dodge") +
    coord_flip() +
    theme_minimal() +
    labs(
      title   = sprintf("Top STRING enrichment hits per gene list (FDR \u2264 %s)", params$fdr_thr),
      x       = "Term",
      y       = "-log10(FDR)"
    )
  print(p_hits)
  ggsave(file.path(string_plots_dir, "top_string_hits_per_list.png"),
         p_hits, width = 14, height = 9, dpi = 320)

  # в”Ђв”Ђ Enriched terms by STRING category в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
  cat_counts <- std_results %>%
    count(gene_list, category, sort = TRUE)

  p_cat <- ggplot(cat_counts, aes(x = gene_list, y = n, fill = category)) +
    geom_col(position = "stack") +
    theme_minimal() +
    labs(
      title = "Number of enriched terms by STRING category",
      x     = "Gene list",
      y     = "N terms"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  print(p_cat)
  ggsave(file.path(string_plots_dir, "string_enriched_terms_by_category.png"),
         p_cat, width = 12, height = 7, dpi = 320)

  # в”Ђв”Ђ Spider / radar-like profile в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
  if (isTRUE(params$enable_spider_plot)) {
    db_profile <- std_results %>%
      group_by(gene_list, category) %>%
      summarise(
        min_fdr    = min(FDR_std, na.rm = TRUE),
        n_terms    = n_distinct(term_std),
        signal     = -log10(pmax(min_fdr, .Machine$double.xmin)),
        .groups    = "drop"
      ) %>%
      group_by(category) %>%
      mutate(
        signal_scaled = ifelse(max(signal, na.rm = TRUE) > 0,
                                signal / max(signal, na.rm = TRUE), 0)
      ) %>%
      ungroup()

    # geom_line + coord_polar crashes when any group has only 1 point
    # (dist_polar returns a 0-row data frame that cannot receive a 1-element
    # assignment).  Keep only gene_lists that span >= 2 categories.
    db_profile_spider <- db_profile %>%
      group_by(gene_list) %>%
      filter(n() >= 2) %>%
      ungroup()

    if (nrow(db_profile_spider) > 0) {
      p_spider <- ggplot(
        db_profile_spider,
        aes(x = category, y = signal_scaled, group = gene_list, color = gene_list)
      ) +
        geom_polygon(aes(fill = gene_list), alpha = 0.08, linewidth = 0.4) +
        geom_line(linewidth = 0.8) +
        geom_point(size = 2) +
        coord_polar() +
        theme_minimal() +
        labs(
          title    = "STRING functional profile (radar-like) by gene list",
          subtitle = "Scaled -log10(min FDR) across STRING categories",
          x        = NULL,
          y        = "Scaled signal"
        ) +
        theme(axis.text.x = element_text(size = 9))
      print(p_spider)
      ggsave(file.path(string_plots_dir, "string_spider_profile.png"),
             p_spider, width = 10, height = 9, dpi = 320)
    }
  }

  # в”Ђв”Ђ Term recurrence across lists в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
  term_recurrence <- std_results %>%
    distinct(gene_list, term_std, category, FDR_std) %>%
    group_by(term_std, category) %>%
    summarise(
      n_lists  = n_distinct(gene_list),
      best_fdr = min(FDR_std, na.rm = TRUE),
      mean_fdr = mean(FDR_std, na.rm = TRUE),
      .groups  = "drop"
    ) %>%
    arrange(desc(n_lists), best_fdr)

  readr::write_tsv(term_recurrence,
                   file.path(string_summary_dir, "string_term_recurrence.tsv"))

  top_recurrent <- term_recurrence %>%
    # slice_head() already returns all available rows when n > nrow(data)
    # (using n() here errors in recent dplyr because n() is only valid
    # inside data-masking verbs such as mutate/summarise/filter).
    slice_head(n = 30L) %>%
    mutate(term_short = stringr::str_trunc(term_std, 70))

  if (nrow(top_recurrent) > 0) {
    p_recurrent <- ggplot(
      top_recurrent,
      aes(x = reorder(term_short, n_lists), y = n_lists, fill = category)
    ) +
      geom_col() +
      coord_flip() +
      theme_minimal() +
      labs(
        title = "Most recurrent STRING terms across gene lists",
        x     = "Term",
        y     = "# gene lists where term is enriched"
      )
    print(p_recurrent)
    ggsave(file.path(string_plots_dir, "string_recurrent_terms.png"),
           p_recurrent, width = 13, height = 10, dpi = 320)
  }

  # в”Ђв”Ђ Pairwise overlap heatmap в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
  if (isTRUE(params$enable_overlap_heatmap)) {
    top_n     <- as.integer(params$overlap_top_terms)
    top_n     <- ifelse(is.na(top_n) || top_n <= 0L, 50L, top_n)
    metric    <- tolower(as.character(params$compare_metric %||% "jaccard"))

    term_sets <- std_results %>%
      arrange(FDR_std) %>%
      group_by(gene_list) %>%
      slice_head(n = top_n) %>%
      summarise(term_set = list(unique(term_std)), .groups = "drop")

    glists <- term_sets$gene_list
    overlap_mat <- matrix(
      0,
      nrow = length(glists), ncol = length(glists),
      dimnames = list(glists, glists)
    )

    for (i in seq_along(glists)) {
      for (j in seq_along(glists)) {
        a <- term_sets$term_set[[i]]
        b <- term_sets$term_set[[j]]
        inter  <- length(intersect(a, b))
        union_n <- length(unique(c(a, b)))
        min_n  <- min(length(a), length(b))
        overlap_mat[i, j] <- if (metric == "overlap") {
          ifelse(min_n > 0, inter / min_n, 0)
        } else {
          ifelse(union_n > 0, inter / union_n, 0)
        }
      }
    }

    overlap_df <- as.data.frame(as.table(overlap_mat)) %>%
      rename(gene_list_x = Var1, gene_list_y = Var2, overlap = Freq)
    readr::write_tsv(overlap_df,
                     file.path(string_summary_dir, "string_overlap_matrix.tsv"))

    p_overlap <- ggplot(overlap_df,
                        aes(x = gene_list_x, y = gene_list_y, fill = overlap)) +
      geom_tile(color = "white") +
      geom_text(aes(label = sprintf("%.2f", overlap)), size = 3) +
      scale_fill_viridis_c(option = "C") +
      theme_minimal() +
      labs(
        title    = sprintf("Pairwise STRING enrichment overlap (%s)", params$compare_metric),
        subtitle = sprintf("Computed from top %d enriched terms per list", top_n),
        x = NULL, y = NULL
      ) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
    print(p_overlap)
    ggsave(file.path(string_plots_dir, "string_overlap_heatmap.png"),
           p_overlap, width = 10, height = 8, dpi = 320)
  }

  # в”Ђв”Ђ Pooled dotplot в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
  pooled_dot <- std_results %>%
    arrange(FDR_std) %>%
    group_by(gene_list) %>%
    slice_head(n = as.integer(params$top_thr)) %>%
    ungroup() %>%
    mutate(term_short = stringr::str_trunc(term_std, 60))

  if (nrow(pooled_dot) > 0) {
    p_dot <- ggplot(
      pooled_dot,
      aes(x = gene_list, y = term_short, size = neglog10FDR, color = category)
    ) +
      geom_point(alpha = 0.85) +
      theme_minimal() +
      labs(
        title    = "Pooled STRING characterisation across gene lists",
        subtitle = "Top enriched terms per list (dot size = -log10(FDR))",
        x = "Gene list", y = "Enriched term"
      ) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
    print(p_dot)
    ggsave(file.path(string_plots_dir, "string_pooled_dotplot.png"),
           p_dot, width = 14, height = 11, dpi = 320)
  }

  # в”Ђв”Ђ Shared-term network matrix в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
  if (isTRUE(params$enable_enrichment_map) && has_pkg("igraph")) {
    term_links <- std_results %>%
      distinct(gene_list, term_std) %>%
      group_by(term_std) %>%
      summarise(
        gene_lists = list(sort(unique(gene_list))),
        n_lists    = n(),
        .groups    = "drop"
      ) %>%
      filter(n_lists > 1)

    if (nrow(term_links) > 0) {
      edges <- term_links %>%
        mutate(pairs = purrr::map(gene_lists, ~{
          x <- .x
          if (length(x) < 2) return(tibble(from = character(0), to = character(0)))
          as_tibble(t(combn(x, 2)), .name_repair = "minimal") %>%
            setNames(c("from", "to"))
        })) %>%
        select(term_std, pairs) %>%
        tidyr::unnest(pairs) %>%
        count(from, to, name = "shared_terms")

      if (nrow(edges) > 0) {
        p_network <- ggplot(edges, aes(x = from, y = to, fill = shared_terms)) +
          geom_tile(color = "white") +
          scale_fill_viridis_c(option = "B") +
          theme_minimal() +
          labs(
            title    = "Shared STRING enriched functions between gene lists",
            subtitle = "Cell value = number of overlapping enriched terms",
            x = NULL, y = NULL
          ) +
          theme(axis.text.x = element_text(angle = 45, hjust = 1))
        print(p_network)
        ggsave(file.path(string_plots_dir, "string_shared_term_network_matrix.png"),
               p_network, width = 10, height = 8, dpi = 320)
      }
    }
  }
}
```

## Session Info

```{r session_info}
sessionInfo()
```
