---
title: "CAAS Significance Report"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
params:
  global_meta_input: NULL
  position_scores_input: NULL
  gene_scores_input: NULL
  output_dir: "."
  p_emp_thr: 0.05
  seed: "1998"
---

<!--
This report is a DISTINCT, LATER pipeline stage than 7.CAAS_pattern_annotation.Rmd.
CT_META_CAAS (which renders 7.CAAS_pattern_annotation.Rmd) runs upstream of
SCORING in the live DAG -- it cannot join position_scores.tsv/gene_scores.tsv
because those files don't exist yet when it runs. This report re-reads
CT_META_CAAS's already-published meta_caas output (one row per
Gene, Position, pattern, caap_group) together with SCORING's published
position_scores.tsv/gene_scores.tsv to annotate that pattern breakdown with
the permulation-null significance columns (p.emp/p.adj_bh/p.adj_sam,
gene_caas_pperm/gene_caas_pperm_adj) that only exist once SCORING has run.
-->

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
  library(kableExtra)
  library(scales)
  library(data.table)
  library(paletteer)
})

createDir <- function(path) {
  if (!dir.exists(path)) dir.create(path, recursive = TRUE, showWarnings = FALSE)
}

# Same house palette as 7.CAAS_pattern_annotation.Rmd, for visual continuity.
tableau_10 <- paletteer_d("ggthemes::Tableau_10")
names(tableau_10) <- c(
  "blue", "orange", "vermillion", "bluish_green",
  "green", "yellow", "reddish_purple", "pink",
  "brown", "gray"
)

sig_palette <- c(
  "Significant"     = tableau_10[["bluish_green"]],
  "Not significant" = tableau_10[["gray"]]
)

seed_val <- if (!is.null(params$seed) && nzchar(as.character(params$seed))) as.integer(params$seed) else 1998
set.seed(seed_val)

p_emp_thr <- suppressWarnings(as.numeric(params$p_emp_thr))
if (is.na(p_emp_thr)) p_emp_thr <- 0.05
cat(sprintf("Using p.adj_bh / p.adj_sam / gene_caas_pperm_adj significance threshold: %s\n", p_emp_thr))
```

## 1. Data Import

Load CT_META_CAAS's published meta_caas table and SCORING's published
position/gene score tables, all from the same pipeline run.

```{r data_import}
if (is.null(params$global_meta_input) || !file.exists(params$global_meta_input)) {
  stop("global_meta_input (meta_caas / global_meta_caas.tsv) not found: ", params$global_meta_input)
}
if (is.null(params$position_scores_input) || !file.exists(params$position_scores_input)) {
  stop("position_scores_input (position_scores.tsv) not found: ", params$position_scores_input)
}
if (is.null(params$gene_scores_input) || !file.exists(params$gene_scores_input)) {
  stop("gene_scores_input (gene_scores.tsv) not found: ", params$gene_scores_input)
}

meta_caas <- read_tsv(params$global_meta_input, show_col_types = FALSE)
position_scores <- read_tsv(params$position_scores_input, show_col_types = FALSE)
gene_scores <- read_tsv(params$gene_scores_input, show_col_types = FALSE)

cat(sprintf("Loaded meta_caas: %d rows from %s\n", nrow(meta_caas), basename(params$global_meta_input)))
cat(sprintf("Loaded position_scores: %d rows from %s\n", nrow(position_scores), basename(params$position_scores_input)))
cat(sprintf("Loaded gene_scores: %d rows from %s\n", nrow(gene_scores), basename(params$gene_scores_input)))

required_meta_cols <- c("Gene", "Position")
missing_meta_cols <- setdiff(required_meta_cols, names(meta_caas))
if (length(missing_meta_cols) > 0) {
  stop("Missing required columns in meta_caas: ", paste(missing_meta_cols, collapse = ", "))
}
has_caap_group <- "caap_group" %in% names(meta_caas)
type_col <- if ("convergence_type" %in% names(meta_caas)) {
  "convergence_type"
} else if ("pattern" %in% names(meta_caas)) {
  "pattern"
} else {
  NULL
}
type_label <- if (identical(type_col, "convergence_type")) "Convergence Type" else if (identical(type_col, "pattern")) "Pattern" else "Classification"
```

## 2. Join SCORING's null-based significance onto the meta_caas rows

### 2a. Position-level: p.emp / p.adj_bh / p.adj_sam

`position_scores.tsv` carries one row per (Gene, Position, side); `p.emp` and
its adjustments `p.adj_bh` (BH over the permutation family) and `p.adj_sam`
(permutation FDR) are already pooled across sides (max-over-sides "all" axis --
docs/scoring_v2_p_emp.md §1 / scoring_compute.R §2f-ter), so
both side-rows for a given (Gene, Position) carry the identical p.emp value.
We deduplicate to one row per (Gene, Position) BEFORE joining -- joining on
the undeduped table (or on `side`, which meta_caas doesn't have) would
silently duplicate every meta_caas row once per side.

```{r position_join}
has_p_emp_cols <- all(c("p.emp", "p.adj_bh", "p.adj_sam") %in% names(position_scores))
if (!has_p_emp_cols) {
  warning("position_scores.tsv has no p.emp/p.adj_bh/p.adj_sam columns -- was it produced by ",
          "a stale scoring_compute.R build? Significance columns will be all-NA.")
  pos_dedup <- tibble(Gene = character(), Position = integer(),
                       p.emp = numeric(), p.adj_bh = numeric(), p.adj_sam = numeric())
} else {
  pos_dedup <- position_scores %>%
    mutate(Position = as.integer(Position)) %>%
    distinct(Gene, Position, p.emp, p.adj_bh, p.adj_sam)
  # Guard against a corrupt/undeduped upstream file: if the same (Gene,Position)
  # carries >1 distinct p.emp value, the pooling assumption above is violated --
  # surface it loudly rather than silently duplicating rows downstream.
  n_dupe_keys <- pos_dedup %>% count(Gene, Position) %>% filter(n > 1) %>% nrow()
  if (n_dupe_keys > 0) {
    warning(sprintf(paste0("%d (Gene,Position) keys have >1 distinct p.emp value in ",
                            "position_scores.tsv -- the max-over-sides pooling assumption ",
                            "this report relies on does not hold for this run. Keeping the ",
                            "first row per key; p.emp/p.adj_bh/p.adj_sam may be unreliable."), n_dupe_keys))
    pos_dedup <- pos_dedup %>% distinct(Gene, Position, .keep_all = TRUE)
  }
}

meta_caas <- meta_caas %>%
  mutate(Position = as.integer(Position)) %>%
  left_join(pos_dedup, by = c("Gene", "Position")) %>%
  mutate(
    position_significant_raw = !is.na(p.emp) & p.emp <= p_emp_thr,
    position_significant_bh  = !is.na(p.adj_bh) & p.adj_bh <= p_emp_thr,
    position_significant_sam = !is.na(p.adj_sam) & p.adj_sam <= p_emp_thr,
    position_significant     = position_significant_bh
  )

.n_meta_rows <- nrow(meta_caas)
.n_pos_matched <- meta_caas %>% filter(!is.na(p.emp)) %>% nrow()
.n_unique_pos <- n_distinct(meta_caas$Position)
.n_unique_pos_matched <- meta_caas %>% filter(!is.na(p.emp)) %>% distinct(Position) %>% nrow()
.pos_rate <- if (.n_meta_rows > 0) .n_pos_matched / .n_meta_rows else 0
cat(sprintf("  p.emp position join: matched %d/%d rows (%d/%d unique positions, %.1f%%)\n",
            .n_pos_matched, .n_meta_rows, .n_unique_pos_matched, .n_unique_pos, 100 * .pos_rate))
if (.pos_rate < 0.5 && !("convergence_type" %in% names(meta_caas))) {
  cat(sprintf(paste0("  WARNING: p.emp join rate %.1f%% < 50%% -- the meta_caas positions ",
                     "and position_scores.tsv positions are likely on different coordinate ",
                     "systems or from different pipeline runs. Treat position_significant as ",
                     "unreliable.\n"), 100 * .pos_rate), file = stderr())
}
```

### 2b. Gene-level: gene_caas_pperm / gene_caas_pperm_adj

`gene_scores.tsv` already has one row per Gene (no dedup needed). Only the
global/pooled `gene_caas_pperm`/`gene_caas_pperm_adj` columns are used --
meta_caas rows carry no side/direction column, so joining the `_top`/
`_bottom` variants would be meaningless at this join's grain (which
top/bottom pair would a given meta_caas row match against?).

```{r gene_join}
has_gene_pperm_cols <- all(c("gene_caas_pperm", "gene_caas_pperm_adj") %in% names(gene_scores))
if (!has_gene_pperm_cols) {
  warning("gene_scores.tsv has no gene_caas_pperm/gene_caas_pperm_adj columns -- ",
          "significance columns will be all-NA.")
  gene_dedup <- tibble(Gene = character(), gene_caas_pperm = numeric(), gene_caas_pperm_adj = numeric())
} else {
  gene_dedup <- gene_scores %>%
    distinct(Gene, gene_caas_pperm, gene_caas_pperm_adj)
}

meta_caas <- meta_caas %>%
  left_join(gene_dedup, by = "Gene") %>%
  mutate(
    gene_significant_raw = !is.na(gene_caas_pperm) & gene_caas_pperm <= p_emp_thr,
    gene_significant_adj = !is.na(gene_caas_pperm_adj) & gene_caas_pperm_adj <= p_emp_thr,
    gene_significant     = gene_significant_adj
  )

.n_genes_meta <- n_distinct(meta_caas$Gene)
.n_genes_matched <- meta_caas %>% filter(!is.na(gene_caas_pperm)) %>% distinct(Gene) %>% nrow()
.gene_rate <- if (.n_genes_meta > 0) .n_genes_matched / .n_genes_meta else 0
cat(sprintf("  gene_caas_pperm gene join: matched %d/%d genes (%.1f%%)\n",
            .n_genes_matched, .n_genes_meta, 100 * .gene_rate))
if (.gene_rate < 0.5) {
  cat(sprintf(paste0("  WARNING: gene_caas_pperm join rate %.1f%% < 50%% -- the meta_caas ",
                     "genes and gene_scores.tsv genes are likely from different pipeline runs. ",
                     "Treat gene_significant as unreliable.\n"), 100 * .gene_rate), file = stderr())
}
```

## 3. Prioritization summary by `r type_label`

**This is a prioritization summary, not a hypothesis test.**
We report, per `r tolower(type_label)` (and per caap_group when present), simple descriptive
counts driven by both uncorrected (`p.emp <= `r p_emp_thr``, `gene_caas_pperm <= `r p_emp_thr``)
and FDR-adjusted thresholds: BH (`p.adj_bh <= `r p_emp_thr``, `gene_caas_pperm_adj <= `r p_emp_thr``)
and, for positions, permutation FDR (`p.adj_sam <= `r p_emp_thr``), along with the single most
extreme drivers (`min_p_emp`, `min_p_adj_bh`, `min_p_adj_sam`) -- never a combined test statistic.
`position_significant` follows BH.

```{r pattern_summary_table, results='asis'}
group_vars <- c()
if (has_caap_group) group_vars <- c(group_vars, "caap_group")
if (!is.null(type_col)) group_vars <- c(group_vars, type_col)
if (length(group_vars) == 0) group_vars <- "Gene"

pattern_summary <- meta_caas %>%
  group_by(across(all_of(group_vars))) %>%
  summarize(
    n_positions = n(),
    n_positions_significant_raw = sum(position_significant_raw, na.rm = TRUE),
    pct_positions_significant_raw = 100 * n_positions_significant_raw / n_positions,
    n_positions_significant_bh = sum(position_significant_bh, na.rm = TRUE),
    pct_positions_significant_bh = 100 * n_positions_significant_bh / n_positions,
    n_positions_significant_sam = sum(position_significant_sam, na.rm = TRUE),
    pct_positions_significant_sam = 100 * n_positions_significant_sam / n_positions,
    min_p_emp = suppressWarnings(min(p.emp, na.rm = TRUE)),
    min_p_adj_bh = suppressWarnings(min(p.adj_bh, na.rm = TRUE)),
    min_p_adj_sam = suppressWarnings(min(p.adj_sam, na.rm = TRUE)),
    n_genes = n_distinct(Gene),
    n_genes_significant_raw = n_distinct(Gene[gene_significant_raw %in% TRUE]),
    pct_genes_significant_raw = 100 * n_genes_significant_raw / n_genes,
    n_genes_significant_adj = n_distinct(Gene[gene_significant_adj %in% TRUE]),
    pct_genes_significant_adj = 100 * n_genes_significant_adj / n_genes,
    .groups = "drop"
  ) %>%
  mutate(
    min_p_emp = ifelse(is.infinite(min_p_emp), NA_real_, min_p_emp),
    min_p_adj_bh = ifelse(is.infinite(min_p_adj_bh), NA_real_, min_p_adj_bh),
    min_p_adj_sam = ifelse(is.infinite(min_p_adj_sam), NA_real_, min_p_adj_sam)
  ) %>%
  arrange(across(all_of(group_vars)))

kable(pattern_summary, digits = 4, format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE)
```

```{r pattern_summary_plot}
x_col <- if (!is.null(type_col)) type_col else "Gene"

plot_df_stacked <- pattern_summary %>%
  mutate(
    n_nonsig_raw = n_positions - n_positions_significant_raw,
    n_nonsig_bh = n_positions - n_positions_significant_bh,
    n_nonsig_sam = n_positions - n_positions_significant_sam
  ) %>%
  select(
    all_of(group_vars),
    `Uncorrected (p.emp): Significant` = n_positions_significant_raw,
    `Uncorrected (p.emp): Non-significant` = n_nonsig_raw,
    `BH (p.adj_bh): Significant` = n_positions_significant_bh,
    `BH (p.adj_bh): Non-significant` = n_nonsig_bh,
    `Permutation FDR (p.adj_sam): Significant` = n_positions_significant_sam,
    `Permutation FDR (p.adj_sam): Non-significant` = n_nonsig_sam
  ) %>%
  pivot_longer(
    cols = -all_of(group_vars),
    names_to = c("correction", "status"),
    names_sep = ": ",
    values_to = "n_positions_count"
  ) %>%
  mutate(
    status = factor(status, levels = c("Non-significant", "Significant")),
    correction = factor(correction, levels = c("Uncorrected (p.emp)", "BH (p.adj_bh)",
                                               "Permutation FDR (p.adj_sam)"))
  )

p_counts_sig <- ggplot(plot_df_stacked, aes(x = factor(.data[[x_col]]), y = n_positions_count, fill = status)) +
  geom_bar(stat = "identity", position = "stack", width = 0.7) +
  facet_grid(if (has_caap_group) caap_group ~ correction else ~ correction, scales = "free_y") +
  scale_fill_manual(values = c("Significant" = "#59A14F", "Non-significant" = "#BAB0AC")) +
  theme_minimal() +
  labs(
    title = sprintf("Position Significance Counts by %s", type_label),
    subtitle = sprintf("Uncorrected (p.emp <= %s), BH-adjusted (p.adj_bh <= %s) and permutation-FDR (p.adj_sam <= %s) significant and non-significant counts", p_emp_thr, p_emp_thr, p_emp_thr),
    x = type_label,
    y = "Number of positions",
    fill = "Status"
  ) +
  theme(
    axis.title.y = element_text(margin = ggplot2::margin(l = 20, r = 20)),
    axis.title.x = element_text(margin = ggplot2::margin(t = 12)),
    plot.title = element_text(face = "bold", hjust = 0, margin = ggplot2::margin(b = 6)),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    panel.grid.major = element_line(color = "darkgray", linewidth = 0.3),
    legend.position = "bottom"
  )

print(p_counts_sig)
```

## 4. Row-level significance table

```{r rowlevel_table}
key_cols <- intersect(
  c("Gene", "Position", "caap_group", "convergence_type", "pattern", "side",
    "p.emp", "p.adj_bh", "p.adj_sam", "position_significant", "gene_caas_pperm", "gene_caas_pperm_adj"),
  names(meta_caas)
)
other_cols <- setdiff(names(meta_caas), key_cols)
meta_caas_display <- meta_caas %>%
  select(all_of(c(key_cols, other_cols))) %>%
  mutate(across(where(is.numeric), ~ round(., 4)))

if (requireNamespace("DT", quietly = TRUE)) {
  DT::datatable(
    meta_caas_display,
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 15, scrollX = TRUE, autoWidth = TRUE),
    caption = htmltools::tags$caption(
      style = "caption-side: top; text-align: left; font-weight: bold;",
      sprintf("Row-level CAAS significance annotations (%d total positions; full table exported to significance/meta_caas_significance.tsv)", nrow(meta_caas))
    )
  )
} else {
  meta_caas_display
}
```

## 5. Export

```{r export}
sig_dir <- "significance"
createDir(sig_dir)

write_tsv(meta_caas, file.path(sig_dir, "meta_caas_significance.tsv"))
write_tsv(pattern_summary, file.path(sig_dir, "pattern_significance_summary.tsv"))

cat(sprintf("Wrote %s (%d rows)\n", file.path(sig_dir, "meta_caas_significance.tsv"), nrow(meta_caas)))
cat(sprintf("Wrote %s (%d rows)\n", file.path(sig_dir, "pattern_significance_summary.tsv"), nrow(pattern_summary)))
```

---

## Summary

This report annotates the CAAS positions evaluated by SCORING with
permulation-null significance columns, reporting both uncorrected (`p.emp`,
`gene_caas_pperm`), BH-adjusted (`p.adj_bh`, `gene_caas_pperm_adj`) and, for positions,
permutation-FDR (`p.adj_sam`) values thresholded at `p_emp_thr = `r p_emp_thr``. Group-level numbers are a prioritization
summary (counts and rates), not a combined hypothesis test.

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — CT Meta-CAAS Module (post-SCORING significance annotation)*
*Date of generation: `r Sys.Date()`*
