---
title: "CAAS Position p-value Report"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
params:
  global_meta_input: NULL
  position_scores_input: NULL
  output_dir: "."
  seed: "1998"
  max_table_rows: 2000
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

<!-- 16.CAAS_significance_report.Rmd — Annotation of the CAAS pattern table with the empirical position p-values of SCORING. -->
<!-- PhyloPhere | subworkflows/CT_META_CAAS/local/ -->

<!--
Rendered by CAAS_SIGNIFICANCE_REPORT (ctpp_meta_caas.nf), after SCORING. The per-position CAAS table
(global_meta_input: global_meta_caas.tsv, or the post-processed discovery table SCORING evaluated; one row per
Gene, Position and, when present, pattern and caap_group) is joined with position_scores.tsv (p.emp, p.adj_bh, p.emp_fact, p.adj_bh_fact)
of the same run. It cannot be merged into 7.CAAS_pattern_annotation.Rmd, which runs before SCORING, when that
table does not exist yet.
Outputs: significance/meta_caas_significance.tsv and significance/pattern_significance_summary.tsv.
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
  library(kableExtra)
  library(scales)
  library(data.table)
  library(paletteer)
})

createDir <- function(path) {
  if (!dir.exists(path)) dir.create(path, recursive = TRUE, showWarnings = FALSE)
}

# Same palette as 7.CAAS_pattern_annotation.Rmd.
tableau_10 <- paletteer_d("ggthemes::Tableau_10")
names(tableau_10) <- c(
  "blue", "orange", "vermillion", "bluish_green",
  "green", "yellow", "reddish_purple", "pink",
  "brown", "gray"
)

seed_val <- if (!is.null(params$seed) && nzchar(as.character(params$seed))) as.integer(params$seed) else 1998
set.seed(seed_val)
```

## 1. Data Import

Load CT_META_CAAS's published meta_caas table and SCORING's published
position score table, both from the same pipeline run.

```{r data_import}
if (is.null(params$global_meta_input) || !file.exists(params$global_meta_input)) {
  stop("global_meta_input (meta_caas / global_meta_caas.tsv) not found: ", params$global_meta_input)
}
if (is.null(params$position_scores_input) || !file.exists(params$position_scores_input)) {
  stop("position_scores_input (position_scores.tsv) not found: ", params$position_scores_input)
}

meta_caas <- read_tsv(params$global_meta_input, show_col_types = FALSE)
position_scores <- read_tsv(params$position_scores_input, show_col_types = FALSE)

cat(sprintf("Loaded meta_caas: %d rows from %s\n", nrow(meta_caas), basename(params$global_meta_input)))
cat(sprintf("Loaded position_scores: %d rows from %s\n", nrow(position_scores), basename(params$position_scores_input)))

required_meta_cols <- c("Gene", "Position")
missing_meta_cols <- setdiff(required_meta_cols, names(meta_caas))
if (length(missing_meta_cols) > 0) {
  stop("Missing required columns in meta_caas: ", paste(missing_meta_cols, collapse = ", "))
}
has_caap_group <- "caap_group" %in% names(meta_caas)
type_col <- if ("convergence_type" %in% names(meta_caas)) {
  "convergence_type"
} else if ("pattern" %in% names(meta_caas)) {
  "pattern"
} else {
  NULL
}
type_label <- if (identical(type_col, "convergence_type")) "Convergence Type" else if (identical(type_col, "pattern")) "Pattern" else "Classification"
```

## 2. Join SCORING's position p-values onto the meta_caas rows

The p-values are informative: no threshold or significance call is applied to them.

- `p.emp` is the empirical permulation p of the position: the proportion of the N permuted-labeling cycles that
  detect the position (on either phenotype side) and give it a CAAS score at least as high as the observed one,
  counting the observed run as one more cycle, (k + 1)/(N + 1). Its minimum is 1/(N + 1), shown as `< 1/N`.
  A position with an observed score of 0 has `p.emp` = 1. `p.adj_bh` is its Benjamini-Hochberg adjustment.
- `p.emp_fact` is the same quantity factorized: the proportion of cycles that detect the position times the
  proportion of the detections of positions of a similar propensity to be detected that reach its score. It is not
  limited by the 1/(N + 1) floor. `p.adj_bh_fact` is its Benjamini-Hochberg adjustment.

### Position p-values

`position_scores.tsv` carries one row per (Gene, Position, side); `p.emp` and
its adjustment `p.adj_bh` (BH over the permutation family) are already pooled across sides (max-over-sides "all" axis --
docs/scoring_v2_p_emp.md §1 / scoring_compute.R §2f-ter), so
both side-rows for a given (Gene, Position) carry the identical p.emp value.
We deduplicate to one row per (Gene, Position) BEFORE joining -- joining on
the undeduped table (or on `side`, which meta_caas doesn't have) would
silently duplicate every meta_caas row once per side.

```{r position_join}
has_p_emp_cols <- all(c("p.emp", "p.adj_bh") %in% names(position_scores))
if (has_p_emp_cols) {
  for (.c in c("p.emp_fact", "p.adj_bh_fact")) if (!.c %in% names(position_scores)) position_scores[[.c]] <- NA_real_
}
# p.emp is add-one smoothed: its minimum is 1/(N + 1). A value at that floor is displayed as "< 1/N" (Saputra et al.
# 2021: with 1000 permulations the lowest reportable p is 0.001 and a p of 0 is written "<0.001"); the TSV keeps the number.
.p_floor <- if (has_p_emp_cols) suppressWarnings(min(position_scores$p.emp, na.rm = TRUE)) else NA_real_
.n_perm  <- if (is.finite(.p_floor) && .p_floor > 0) round(1 / .p_floor) - 1 else NA_real_
.fmt_p <- function(p) vapply(p, function(x) {
  if (is.na(x)) return(NA_character_)
  if (is.finite(.n_perm) && x <= .p_floor * (1 + 1e-9)) return(sprintf("<%s", format(1 / .n_perm, digits = 3)))
  format(signif(x, 3), scientific = FALSE, trim = TRUE)
}, character(1))
if (!has_p_emp_cols) {
  warning("position_scores.tsv has no p.emp/p.adj_bh columns -- was it produced by ",
          "a stale scoring_compute.R build? The p-value columns will be all-NA.")
  pos_dedup <- tibble(Gene = character(), Position = integer(),
                       p.emp = numeric(), p.adj_bh = numeric(), p.emp_fact = numeric(), p.adj_bh_fact = numeric())
} else {
  pos_dedup <- position_scores %>%
    mutate(Position = as.integer(Position)) %>%
    distinct(Gene, Position, p.emp, p.adj_bh, p.emp_fact, p.adj_bh_fact)
  # Check the pooling assumption of the text above: a (Gene, Position) with more than one
  # distinct p.emp value would duplicate meta_caas rows in the join, so warn and keep one.
  n_dupe_keys <- pos_dedup %>% count(Gene, Position) %>% filter(n > 1) %>% nrow()
  if (n_dupe_keys > 0) {
    warning(sprintf(paste0("%d (Gene,Position) keys have >1 distinct p.emp value in ",
                            "position_scores.tsv -- the max-over-sides pooling assumption ",
                            "this report relies on does not hold for this run. Keeping the ",
                            "first row per key; p.emp/p.adj_bh may be unreliable."), n_dupe_keys))
    pos_dedup <- pos_dedup %>% distinct(Gene, Position, .keep_all = TRUE)
  }
}

meta_caas <- meta_caas %>%
  mutate(Position = as.integer(Position)) %>%
  left_join(pos_dedup, by = c("Gene", "Position"))

.n_meta_rows <- nrow(meta_caas)
.n_pos_matched <- meta_caas %>% filter(!is.na(p.emp)) %>% nrow()
.n_unique_pos <- n_distinct(meta_caas$Position)
.n_unique_pos_matched <- meta_caas %>% filter(!is.na(p.emp)) %>% distinct(Position) %>% nrow()
.pos_rate <- if (.n_meta_rows > 0) .n_pos_matched / .n_meta_rows else 0
cat(sprintf("  p.emp position join: matched %d/%d rows (%d/%d unique positions, %.1f%%)\n",
            .n_pos_matched, .n_meta_rows, .n_unique_pos_matched, .n_unique_pos, 100 * .pos_rate))
if (.pos_rate < 0.5 && !("convergence_type" %in% names(meta_caas))) {
  cat(sprintf(paste0("  WARNING: p.emp join rate %.1f%% < 50%% -- the meta_caas positions ",
                     "and position_scores.tsv positions are likely on different coordinate ",
                     "systems or from different pipeline runs. Treat the p-values as ",
                     "unreliable.\n"), 100 * .pos_rate), file = stderr())
}
```

## 3. Position p-values by `r type_label`

Descriptive summary per `r tolower(type_label)` (and per caap_group when present): number of positions, how many
have a p-value, how many sit at the `p.emp` floor, the smallest p-values and the median of `p.emp_fact`. These are
not tests and no group-level statistic is combined from them.

```{r pattern_summary_table, results='asis'}
group_vars <- c()
if (has_caap_group) group_vars <- c(group_vars, "caap_group")
if (!is.null(type_col)) group_vars <- c(group_vars, type_col)
if (length(group_vars) == 0) group_vars <- "Gene"

.min_na <- function(x) { m <- suppressWarnings(min(x, na.rm = TRUE)); if (is.infinite(m)) NA_real_ else m }
.med_na <- function(x) if (all(is.na(x))) NA_real_ else median(x, na.rm = TRUE)

pattern_summary <- meta_caas %>%
  group_by(across(all_of(group_vars))) %>%
  summarize(
    n_positions = n(),
    n_positions_with_p = sum(!is.na(p.emp)),
    n_at_p_emp_floor = sum(!is.na(p.emp) & p.emp <= .p_floor * (1 + 1e-9)),
    min_p_emp = .min_na(p.emp),               median_p_emp = .med_na(p.emp),
    min_p_adj_bh = .min_na(p.adj_bh),         median_p_adj_bh = .med_na(p.adj_bh),
    min_p_emp_fact = .min_na(p.emp_fact),     median_p_emp_fact = .med_na(p.emp_fact),
    min_p_adj_bh_fact = .min_na(p.adj_bh_fact), median_p_adj_bh_fact = .med_na(p.adj_bh_fact),
    n_genes = n_distinct(Gene),
    .groups = "drop"
  ) %>%
  arrange(across(all_of(group_vars)))

kable(pattern_summary %>% mutate(min_p_emp = .fmt_p(min_p_emp), median_p_emp = .fmt_p(median_p_emp)),
      digits = 4, format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE)
```

```{r pattern_summary_plot, fig.height=8}
x_col <- if (!is.null(type_col)) type_col else "Gene"

plot_df <- meta_caas %>%
  select(all_of(c(x_col, if (has_caap_group) "caap_group")), p.emp, p.adj_bh, p.emp_fact, p.adj_bh_fact) %>%
  pivot_longer(c(p.emp, p.adj_bh, p.emp_fact, p.adj_bh_fact), names_to = "p_value", values_to = "p") %>%
  filter(!is.na(p)) %>%
  mutate(
    neg_log10_p = -log10(pmax(p, 1e-300)),
    p_value = factor(p_value, levels = c("p.emp", "p.adj_bh", "p.emp_fact", "p.adj_bh_fact"))
  )

if (nrow(plot_df) > 0) {
  p_box <- ggplot(plot_df, aes(x = factor(.data[[x_col]]), y = neg_log10_p)) +
    geom_boxplot(fill = "#BAB0AC", outlier.size = 0.8) +
    theme_minimal() +
    labs(
      title = sprintf("Position p-values by %s", type_label),
      subtitle = "Empirical p (p.emp), factorized p (p.emp_fact) and their Benjamini-Hochberg adjustments (p.adj_bh, p.adj_bh_fact)",
      x = type_label,
      y = "-log10(p)"
    ) +
    theme(
      axis.title.y = element_text(margin = ggplot2::margin(l = 20, r = 20)),
      axis.title.x = element_text(margin = ggplot2::margin(t = 12)),
      plot.title = element_text(face = "bold", hjust = 0, margin = ggplot2::margin(b = 6)),
      axis.text.x = element_text(angle = 0, hjust = 0.5),
      panel.grid.major = element_line(color = "darkgray", linewidth = 0.3)
    )
  p_box <- if (has_caap_group) p_box + facet_grid(caap_group ~ p_value, scales = "free_y") else p_box + facet_wrap(~ p_value, scales = "free_y")
  print(p_box)
} else {
  cat("No p-values to plot.\n")
}
```

## 4. Row-level p-value table

```{r rowlevel_table}
key_cols <- intersect(
  c("Gene", "Position", "caap_group", "convergence_type", "pattern", "side",
    "p.emp", "p.adj_bh", "p.emp_fact", "p.adj_bh_fact"),
  names(meta_caas)
)
other_cols <- setdiff(names(meta_caas), key_cols)
# The widget embeds every row as JSON in the HTML, so only the most significant
# rows are shown; the full table is exported below.
sort_col <- intersect(c("p.emp", "p.adj_bh"), names(meta_caas))[1]
n_show <- min(nrow(meta_caas), as.integer(params$max_table_rows))
meta_caas_display <- meta_caas %>%
  { if (is.na(sort_col)) . else arrange(., .data[[sort_col]]) } %>%
  slice_head(n = n_show) %>%
  select(all_of(c(key_cols, other_cols))) %>%
  mutate(across(where(is.numeric), ~ round(., 4)), across(any_of("p.emp"), .fmt_p))

if (requireNamespace("DT", quietly = TRUE)) {
  DT::datatable(
    meta_caas_display,
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 15, scrollX = TRUE, autoWidth = TRUE),
    caption = htmltools::tags$caption(
      style = "caption-side: top; text-align: left; font-weight: bold;",
      sprintf("Row-level CAAS position p-values (top %d of %d positions by %s; full table exported to significance/meta_caas_significance.tsv)", n_show, nrow(meta_caas), if (is.na(sort_col)) "file order" else sort_col)
    )
  )
} else {
  meta_caas_display
}
```

## 5. Export

```{r export}
sig_dir <- "significance"
createDir(sig_dir)

write_tsv(meta_caas, file.path(sig_dir, "meta_caas_significance.tsv"))
write_tsv(pattern_summary, file.path(sig_dir, "pattern_significance_summary.tsv"))

cat(sprintf("Wrote %s (%d rows)\n", file.path(sig_dir, "meta_caas_significance.tsv"), nrow(meta_caas)))
cat(sprintf("Wrote %s (%d rows)\n", file.path(sig_dir, "pattern_significance_summary.tsv"), nrow(pattern_summary)))
```

---

## Summary

This report annotates the CAAS positions evaluated by SCORING with their empirical position p-values (`p.emp`,
`p.emp_fact`) and the Benjamini-Hochberg adjustments of both (`p.adj_bh`, `p.adj_bh_fact`). The p-values are
informative and no threshold is applied. Group-level numbers are descriptive.

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — CT Meta-CAAS Module (post-SCORING position p-value annotation)*
*Date of generation: `r Sys.Date()`*
