---
title: "CAAS Pattern Annotation"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
params:
  discovery_input: NULL
  background_input: NULL
  caap_mode: FALSE
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

<!-- 7.CAAS_pattern_annotation.Rmd — Distribution of the CAAS positions over contrast patterns and caap_group. -->
<!-- PhyloPhere | subworkflows/CT_META_CAAS/local/ -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7
)

# ── Dependencies ──────────────────────────────────────────────────────────────
suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
  library(scales)
  library(data.table)
  library(paletteer)
})


# ── Helpers ───────────────────────────────────────────────────────────────────

# Creates `path`, and its parents, when it does not exist.
createDir <- function(path) {
  if (!dir.exists(path)) dir.create(path, recursive = TRUE, showWarnings = FALSE)
}


# ── Palettes ──────────────────────────────────────────────────────────────────

# Tableau 10 under colourblind-safe names; the category palette below is built from it.
tableau_10 <- paletteer_d("ggthemes::Tableau_10")
names(tableau_10) <- c(
  "blue", "orange", "vermillion", "bluish_green",
  "green", "yellow", "reddish_purple", "pink",
  "brown", "gray"
)

category_palette <- c(
  "Normal" = tableau_10[["gray"]],
  "Dubious" = tableau_10[["blue"]],
  "Extreme" = tableau_10[["vermillion"]],
  "Both" = tableau_10[["reddish_purple"]]
)

removal_palette <- c(
  Discarded = tableau_10[["vermillion"]],
  Good = tableau_10[["green"]]
)
```

## 1. Data Import and Validation

This section loads the filtered CAAS data and validates input files.

```{r data_import, message=FALSE}
# ── Filtered discovery ────────────────────────────────────────────────────────

filtered_discovery_path <- params$discovery_input
if (is.null(filtered_discovery_path)) {
  stop("Filtered discovery file parameter is NULL")
}

cat(sprintf("Attempting to load: %s\n", filtered_discovery_path))
cat(sprintf("  File exists: %s\n", file.exists(filtered_discovery_path)))
cat(sprintf("  Working directory: %s\n", getwd()))

if (!file.exists(filtered_discovery_path)) {
  # Show the working directory to help locate a misplaced input.
  cat("Files in current directory:\n")
  print(list.files(".", full.names = TRUE))
  stop("Filtered discovery file not found: ", filtered_discovery_path)
}

discovery_df <- read_tsv(filtered_discovery_path, show_col_types = FALSE)

# The columns of discovery.tab are lowercase (gene, mode, trait, position, pattern, ffgn, ...).
# Only the structural keys are renamed to the CamelCase names this report joins on (Gene,
# Position; GenePos = paste(Gene, Position)). The other columns, including the `pattern`
# contrast index, keep their lowercase names.
.disc_keymap <- c(Gene = "gene", Position = "position")
for (.new in names(.disc_keymap)) {
  .old <- .disc_keymap[[.new]]
  if (.old %in% names(discovery_df) && !(.new %in% names(discovery_df))) {
    names(discovery_df)[names(discovery_df) == .old] <- .new
  }
}

# A table without caap_group is treated as the US scheme.
if (!"caap_group" %in% names(discovery_df)) {
  discovery_df <- discovery_df %>% mutate(caap_group = "US")
}
cat(sprintf("Loaded discovery file: %s\n", basename(filtered_discovery_path)))
cat(sprintf("  Total CAAS positions: %s\n", format(nrow(discovery_df), big.mark = ",")))

# TRUE after the step above; the groups present are listed below.
has_caap_group <- 'caap_group' %in% names(discovery_df)
cat(sprintf("  CAAP mode detected: %s\n", has_caap_group))

if (has_caap_group) {
  groups <- unique(discovery_df$caap_group)
  cat(sprintf("  Groups present: %s\n", paste(groups, collapse = ", ")))
}


# ── Filtered background ───────────────────────────────────────────────────────

filtered_background_path <- params$background_input
if (is.null(filtered_background_path) || !file.exists(filtered_background_path)) {
  stop("Filtered background file not found: ", filtered_background_path)
}

background_df <- read_tsv(filtered_background_path, show_col_types = FALSE)
cat(sprintf("Loaded background file: %s\n", basename(filtered_background_path)))
cat(sprintf("Total genes in background: %s\n", format(nrow(background_df), big.mark = ",")))


# ── Required columns ──────────────────────────────────────────────────────────

required_cols <- c("Gene", "Position", "pattern")
missing_cols <- setdiff(required_cols, names(discovery_df))
if (length(missing_cols) > 0) {
  stop("Missing required columns in discovery: ", paste(missing_cols, collapse = ", "))
}

```

## 2. Pattern-Based Analysis

Analyze CAAS patterns across discovered positions.

```{r pattern_analysis}
cat("\n=== Pattern-Based Analysis ===\n")

# Columns the meta_caas export reads; filled with NA when the table does not carry them.
if (!"is_conserved_meta" %in% names(discovery_df)) {
  discovery_df <- discovery_df %>% mutate(is_conserved_meta = NA)
}
if (!"conserved_pair" %in% names(discovery_df)) {
  discovery_df <- discovery_df %>% mutate(conserved_pair = NA_character_)
}

caas_for_export <- discovery_df

# ── Pattern distribution ──────────────────────────────────────────────────────

# Contrast patterns present in the table.
patterns_present <- unique(caas_for_export$pattern)
cat(sprintf("Patterns detected: %s\n", paste(patterns_present, collapse = ", ")))

# One panel per caap_group when the table has the column, one plot otherwise.
if (has_caap_group) {
  pattern_summary <- caas_for_export %>%
    group_by(caap_group, pattern) %>%
    summarize(
      Total = n(),
      .groups = "drop"
    ) %>%
    arrange(caap_group, pattern)
  
  cat("\nPattern Distribution by Group:\n")
  print(kable(pattern_summary, format.args = list(big.mark = ",")))
  
  p_pattern <- ggplot(pattern_summary, aes(x = factor(pattern), y = Total, fill = factor(pattern))) +
    geom_bar(stat = "identity") +
    facet_wrap(~ caap_group, scales = "free_y") +
    theme_minimal() +
    labs(
      title = "CAAS Pattern Distribution by Group",
      x = "Pattern",
      y = "Candidate Count",
      fill = "Pattern"
    ) +
    theme(
      axis.title.y = element_text(margin = ggplot2::margin(l = 20, r = 20)),
      axis.title.x = element_text(margin = ggplot2::margin(t = 12)),
      plot.title = element_text(face = "bold", hjust = 0, margin = ggplot2::margin(b = 6)),
      axis.text.x = element_text(angle = 0, hjust = 0.5),
      panel.grid.major = element_line(color = "darkgray", size = 0.5),
      legend.position = "none"
    )
  
  print(p_pattern)
} else {
  pattern_summary <- caas_for_export %>%
    group_by(pattern) %>%
    summarize(
      Total = n(),
      .groups = "drop"
    )
  
  cat("\nPattern Distribution:\n")
  print(kable(pattern_summary, format.args = list(big.mark = ",")))
  
  p_pattern <- ggplot(pattern_summary, aes(x = factor(pattern), y = Total, fill = factor(pattern))) +
    geom_bar(stat = "identity") +
    theme_minimal() +
    labs(
      title = "CAAS Pattern Distribution",
      x = "Pattern",
      y = "Candidate Count"
    ) +
    theme(
      axis.title.y = element_text(margin = ggplot2::margin(l = 20, r = 20)),
      axis.title.x = element_text(margin = ggplot2::margin(t = 12)),
      plot.title = element_text(face = "bold", hjust = 0, margin = ggplot2::margin(b = 6)),
      axis.text.x = element_text(angle = 0, hjust = 0.5),
      panel.grid.major = element_line(color = "darkgray", size = 0.5),
      legend.position = "none"
    )
  
  print(p_pattern)
}
```

---

## Summary

This report summarizes the results of the **CAAS Pattern Annotation** step.

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — CAAS Pattern Annotation (CT_META_CAAS module)*
*Date of generation: `r Sys.Date()`*
