---
title: "CT Signification Analysis"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    code_folding: hide
    df_print: paged
params:
  filtered_discovery: NULL
  filtered_background: NULL
  bootstrap_file: NULL
  output_dir: "."
  caap_mode: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7
)

# Load required libraries
suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
  library(scales)
  library(data.table)
  library(paletteer)
})

# Directory management
createDir <- function(path) {
  if (!dir.exists(path)) dir.create(path, recursive = TRUE, showWarnings = FALSE)
}

# Define color palettes, including a colorblind-friendly palette for categories
tableau_10 <- paletteer_d("ggthemes::Tableau_10")
names(tableau_10) <- c(
  "blue", "orange", "vermillion", "bluish_green",
  "green", "yellow", "reddish_purple", "pink",
  "brown", "gray"
)

# Consistent colorblind-friendly palette
category_palette <- c(
  "Normal" = tableau_10[["gray"]],
  "Dubious" = tableau_10[["blue"]],
  "Extreme" = tableau_10[["vermillion"]],
  "Both" = tableau_10[["reddish_purple"]]
)

removal_palette <- c(
  Discarded = tableau_10[["vermillion"]],
  Good = tableau_10[["green"]]
)
```

## 1. Data Import and Validation

This section loads the filtered CAAS data and validates input files.

```{r data_import, message=FALSE}
cat("=== Loading Filtered CAAS Data ===\n")

# Load filtered discovery file
filtered_discovery_path <- params$filtered_discovery
if (is.null(filtered_discovery_path)) {
  stop("Filtered discovery file parameter is NULL")
}

cat(sprintf("Attempting to load: %s\n", filtered_discovery_path))
cat(sprintf("  File exists: %s\n", file.exists(filtered_discovery_path)))
cat(sprintf("  Working directory: %s\n", getwd()))

if (!file.exists(filtered_discovery_path)) {
  # List files in current directory for debugging
  cat("Files in current directory:\n")
  print(list.files(".", full.names = TRUE))
  stop("Filtered discovery file not found: ", filtered_discovery_path)
}

discovery_df <- read_tsv(filtered_discovery_path, show_col_types = FALSE)

# Normalize discovery schema (legacy CAAS files may not include CAAP_Group)
if (!"CAAP_Group" %in% names(discovery_df)) {
  discovery_df <- discovery_df %>% mutate(CAAP_Group = "US")
}
cat(sprintf("Loaded discovery file: %s\n", basename(filtered_discovery_path)))
cat(sprintf("  Total CAAS positions: %s\n", format(nrow(discovery_df), big.mark = ",")))

# Detect CAAP mode (after normalization)
has_caap_group <- 'CAAP_Group' %in% names(discovery_df)
cat(sprintf("  CAAP mode detected: %s\n", has_caap_group))

if (has_caap_group) {
  groups <- unique(discovery_df$CAAP_Group)
  cat(sprintf("  Groups present: %s\n", paste(groups, collapse = ", ")))
}

# Load filtered background
filtered_background_path <- params$filtered_background
if (is.null(filtered_background_path) || !file.exists(filtered_background_path)) {
  stop("Filtered background file not found: ", filtered_background_path)
}

background_df <- read_tsv(filtered_background_path, show_col_types = FALSE)
cat(sprintf("\nLoaded background file: %s\n", basename(filtered_background_path)))
cat(sprintf("  Total genes in background: %s\n", format(nrow(background_df), big.mark = ",")))

# Validate required columns
required_cols <- c("Gene", "Position", "Pattern", "Pvalue")
missing_cols <- setdiff(required_cols, names(discovery_df))
if (length(missing_cols) > 0) {
  stop("Missing required columns in discovery: ", paste(missing_cols, collapse = ", "))
}

cat("\n=== Data Validation Complete ===\n")
```

## 2. Bootstrap Permutation Integration

Load bootstrap permutation file to calculate empirical p-values.

```{r bootstrap_integration, message=FALSE}
cat("\n=== Loading Bootstrap File ===\n")

# Load bootstrap file
if (is.null(params$bootstrap_file)) {
  stop("Bootstrap file parameter is NULL")
}

bootstrap_file_path <- params$bootstrap_file
if (!file.exists(bootstrap_file_path)) {
  stop("Bootstrap file not found: ", bootstrap_file_path)
}

cat(sprintf("Loading bootstrap file: %s\n", basename(bootstrap_file_path)))

# Load bootstrap data (headerless expected, but tolerate legacy variants)
# Supported formats:
#   5 cols: GenePos CAAP_Group Occurrences Total Pvalue.boot
#   4 cols: GenePos Occurrences Total Pvalue.boot   (legacy CAAS)
bootstrap_raw <- tryCatch({
  read_tsv(
    bootstrap_file_path,
    col_names = FALSE,
    show_col_types = FALSE
  )
}, error = function(e) {
  stop("Failed to load bootstrap file: ", e$message)
})

if (ncol(bootstrap_raw) < 4) {
  stop(sprintf("Bootstrap file has %d columns; expected at least 4", ncol(bootstrap_raw)))
}

if (ncol(bootstrap_raw) >= 5) {
  names(bootstrap_raw)[1:5] <- c("GenePos", "CAAP_Group", "Occurrences", "Total", "Pvalue.boot")
} else {
  names(bootstrap_raw)[1:4] <- c("GenePos", "Occurrences", "Total", "Pvalue.boot")
  bootstrap_raw <- bootstrap_raw %>% mutate(CAAP_Group = "US")
}

# Drop accidental textual header rows if present
bootstrap_raw <- bootstrap_raw %>%
  filter(!is.na(GenePos), GenePos != "Gene@Position")

# Normalize types
bootstrap_raw <- bootstrap_raw %>%
  mutate(
    CAAP_Group = if_else(is.na(CAAP_Group) | CAAP_Group == "", "US", CAAP_Group),
    Occurrences = suppressWarnings(as.integer(Occurrences)),
    Total = suppressWarnings(as.integer(Total)),
    Pvalue.boot = suppressWarnings(as.numeric(Pvalue.boot))
  )

# Split GenePos into Gene and Position
bootstrap_df <- bootstrap_raw %>%
  separate(GenePos, into = c("Gene", "Position"), sep = "@", remove = FALSE) %>%
  mutate(Position = as.integer(Position))

# Final safeguard for expected columns
if (!"Pvalue.boot" %in% names(bootstrap_df)) {
  stop("Bootstrap integration requires Pvalue.boot column after normalization")
}

cat(sprintf("Loaded bootstrap data: %s permutations\n", 
            format(nrow(bootstrap_df), big.mark = ",")))

# Check if bootstrap has CAAP groups
bootstrap_has_groups <- 'CAAP_Group' %in% names(bootstrap_df) && 
                        any(!is.na(bootstrap_df$CAAP_Group))

if (has_caap_group && !bootstrap_has_groups) {
  warning("Bootstrap file missing CAAP_Group data, but discovery has CAAP mode enabled")
  warning("Bootstrap results may not be group-aware")
}

if (bootstrap_has_groups) {
  boot_groups <- unique(bootstrap_df$CAAP_Group)
  cat(sprintf("  Bootstrap groups: %s\n", paste(boot_groups, collapse = ", ")))
}

# Calculate empirical p-values
cat("\nIntegrating bootstrap p-values...\n")

# Bootstrap file already contains Pvalue.boot, just need to join
if (has_caap_group && bootstrap_has_groups) {
  # Group-aware permutation p-values
  discovery_with_perm <- discovery_df %>%
    left_join(
      bootstrap_df %>%
        select(Gene, Position, CAAP_Group, Pvalue.boot, Occurrences, Total),
      by = c("Gene", "Position", "CAAP_Group")
    )
} else if (has_caap_group && !bootstrap_has_groups) {
  # Discovery has groups but bootstrap doesn't - join without CAAP_Group
  warning("Joining bootstrap data without CAAP_Group matching")
  discovery_with_perm <- discovery_df %>%
    left_join(
      bootstrap_df %>%
        select(Gene, Position, Pvalue.boot, Occurrences, Total) %>%
        distinct(Gene, Position, .keep_all = TRUE),
      by = c("Gene", "Position")
    )
} else {
  # Standard permutation p-values (no CAAP groups)
  discovery_with_perm <- discovery_df %>%
    left_join(
      bootstrap_df %>%
        select(Gene, Position, Pvalue.boot, Occurrences, Total),
      by = c("Gene", "Position")
    )
}

# Handle missing permutation p-values
discovery_with_perm <- discovery_with_perm %>%
  mutate(
    Pvalue.boot = if_else(is.na(Pvalue.boot), Pvalue, Pvalue.boot),
    Occurrences = if_else(is.na(Occurrences), 0L, as.integer(Occurrences)),
    Total = if_else(is.na(Total), 0L, as.integer(Total))
  )

# Normalize optional conserved columns for downstream meta export
if (!"ConservedPair" %in% names(discovery_with_perm)) {
  discovery_with_perm <- discovery_with_perm %>% mutate(ConservedPair = NA_character_)
}
if (!"ConservedPairs" %in% names(discovery_with_perm)) {
  discovery_with_perm <- discovery_with_perm %>% mutate(ConservedPairs = NA_character_)
}

cat(sprintf("Integrated bootstrap p-values for %d positions\n", 
            sum(!is.na(discovery_with_perm$Pvalue.boot))))

cat("\n=== Bootstrap Integration Complete ===\n")
```

## 3. Significance Filtering

Apply hypergeometric and permutation significance thresholds.

```{r significance_filtering}
cat("\n=== Applying Significance Thresholds ===\n")

alpha <- 0.05

# Apply filters
caas_filtered <- discovery_with_perm %>%
  mutate(
    isSignificant = Pvalue <= alpha | Pvalue.boot <= alpha,
    sig_hyp = Pvalue <= alpha,
    sig_perm = Pvalue.boot <= alpha,
    sig_both = sig_hyp & sig_perm
  )

# Summary statistics
if (has_caap_group) {
  # Group-aware summary
  summary_stats <- caas_filtered %>%
    group_by(CAAP_Group) %>%
    summarize(
      Total = n(),
      Hyp_05 = sum(sig_hyp, na.rm = TRUE),
      Perm_05 = sum(sig_perm, na.rm = TRUE),
      Both_05 = sum(sig_both, na.rm = TRUE),
      .groups = "drop"
    )
  
  cat("\nSignificance Summary by Group:\n")
  print(kable(summary_stats, format.args = list(big.mark = ",")))

  caas_for_export <- caas_filtered

} else {
  # Standard summary
  summary_stats <- caas_filtered %>%
    summarize(
      Total = n(),
      Hyp_05 = sum(sig_hyp, na.rm = TRUE),
      Perm_05 = sum(sig_perm, na.rm = TRUE),
      Both_05 = sum(sig_both, na.rm = TRUE)
    )
  
  cat("\nSignificance Summary:\n")
  print(kable(summary_stats, format.args = list(big.mark = ",")))
  
  caas_for_export <- caas_filtered
}

cat("\n=== Significance Filtering Complete ===\n")
```

## 4. Pattern-Based Analysis

Analyze CAAS patterns and their significance profiles.

```{r pattern_analysis}
cat("\n=== Pattern-Based Analysis ===\n")

# Detect available patterns
patterns_present <- unique(caas_filtered$Pattern)
cat(sprintf("Patterns detected: %s\n", paste(patterns_present, collapse = ", ")))

# Pattern distribution
if (has_caap_group) {
  pattern_summary <- caas_filtered %>%
    group_by(CAAP_Group, Pattern) %>%
    summarize(
      Total = n(),
      Hyp_05 = sum(sig_hyp, na.rm = TRUE),
      Perm_05 = sum(sig_perm, na.rm = TRUE),
      Both_05 = sum(sig_both, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(CAAP_Group, Pattern)
  
  cat("\nPattern Distribution by Group:\n")
  print(kable(pattern_summary, format.args = list(big.mark = ",")))
  
  # Visualization: faceted by group
  plot_data <- pattern_summary %>%
    pivot_longer(
      cols = c(Total, Hyp_05, Perm_05, Both_05),
      names_to = "Filter",
      values_to = "Count"
    ) %>%
    mutate(
      Filter = factor(Filter, levels = c("Total", "Hyp_05", "Perm_05", "Both_05")),
      Pattern = factor(Pattern)
    )
  
  p_pattern <- ggplot(plot_data, aes(x = Pattern, y = Count, fill = Filter)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~ CAAP_Group, scales = "free_y") +
    theme_minimal() +
    labs(
      title = "CAAS Pattern Distribution by Significance Filter",
      subtitle = "Hyp_05 = Hypergeometric | Perm_05 = Permutation | Both_05 = Combined",
      x = "Pattern",
      y = "Count"
    ) +
    scale_fill_manual(values = c(
      "Total" = tableau_10[["gray"]],
      "Hyp_05" = tableau_10[["blue"]],
      "Perm_05" = tableau_10[["green"]],
      "Both_05" = tableau_10[["vermillion"]]
    )) +
    theme(
      axis.text.x = element_text(angle = 0, hjust = 0.5),
      panel.grid.major = element_line(color = "darkgray", size = 0.5),
      legend.position = "bottom"
    )
  
  print(p_pattern)
} else {
  pattern_summary <- caas_filtered %>%
    group_by(Pattern) %>%
    summarize(
      Total = n(),
      Hyp_05 = sum(sig_hyp, na.rm = TRUE),
      Perm_05 = sum(sig_perm, na.rm = TRUE),
      Both_05 = sum(sig_both, na.rm = TRUE),
      .groups = "drop"
    )
  
  cat("\nPattern Distribution:\n")
  print(kable(pattern_summary, format.args = list(big.mark = ",")))
  
  # Visualization
  plot_data <- pattern_summary %>%
    pivot_longer(
      cols = c(Total, Hyp_05, Perm_05, Both_05),
      names_to = "Filter",
      values_to = "Count"
    ) %>%
    mutate(
      Filter = factor(Filter, levels = c("Total", "Hyp_05", "Perm_05", "Both_05")),
      Pattern = factor(Pattern)
    )
  
  p_pattern <- ggplot(plot_data, aes(x = Pattern, y = Count, fill = Filter)) +
    geom_bar(stat = "identity", position = "dodge") +
    theme_minimal() +
    labs(
      title = "CAAS Pattern Distribution by Significance Filter",
      subtitle = "Hyp_05 = Hypergeometric | Perm_05 = Permutation | Both_05 = Combined",
      x = "Pattern",
      y = "Count"
    ) +
    scale_fill_manual(values = c(
      "Total" = tableau_10[["gray"]],
      "Hyp_05" = tableau_10[["blue"]],
      "Perm_05" = tableau_10[["green"]],
      "Both_05" = tableau_10[["vermillion"]]
    )) +
    theme(
      axis.text.x = element_text(angle = 0, hjust = 0.5),
      panel.grid.major = element_line(color = "darkgray", size = 0.5),
      legend.position = "bottom"
    )
  
  print(p_pattern)
}

cat("\n=== Pattern Analysis Complete ===\n")
```

## 5. Gene List Export for Enrichment

Export gene lists for downstream enrichment analysis (GO, KEGG, Reactome).

```{r gene_export}
cat("\n=== Exporting Gene Lists ===\n")

# Use local directory for Nextflow to capture
gene_list_dir <- "gene_lists"
createDir(gene_list_dir)

# Export function
export_gene_list <- function(df, name, dir) {
  genes <- unique(df$Gene)
  output_file <- file.path(dir, paste0(name, ".genes"))
  write.table(
    genes,
    output_file,
    quote = FALSE,
    row.names = FALSE,
    col.names = FALSE
  )
  return(length(genes))
}

# Export by significance criteria
if (has_caap_group) {
  # Per-group exports (excluding gs0)
  for (grp in unique(caas_for_export$CAAP_Group)) {
    grp_dir <- file.path(gene_list_dir, grp)
    createDir(grp_dir)
    
    grp_data <- caas_for_export %>% filter(CAAP_Group == grp)
    
    # All genes
    n_all <- export_gene_list(grp_data, "all", grp_dir)
    
    # Hypergeometric significant
    n_hyp <- export_gene_list(
      grp_data %>% filter(sig_hyp),
      "hypergeometric_p05",
      grp_dir
    )
    
    # Permutation significant
    n_perm <- export_gene_list(
      grp_data %>% filter(sig_perm),
      "permutation_p05",
      grp_dir
    )
    
    # Both significant
    n_both <- export_gene_list(
      grp_data %>% filter(sig_both),
      "combined_p05",
      grp_dir
    )
    
    cat(sprintf("  %s: All=%d | Hyp=%d | Perm=%d | Both=%d\n",
                grp, n_all, n_hyp, n_perm, n_both))
  }
  
  # Global exports (all groups except gs0)
  global_dir <- file.path(gene_list_dir, "global")
  createDir(global_dir)
  
  n_all <- export_gene_list(caas_for_export, "all", global_dir)
  n_hyp <- export_gene_list(
    caas_for_export %>% filter(sig_hyp),
    "hypergeometric_p05",
    global_dir
  )
  n_perm <- export_gene_list(
    caas_for_export %>% filter(sig_perm),
    "permutation_p05",
    global_dir
  )
  n_both <- export_gene_list(
    caas_for_export %>% filter(sig_both),
    "combined_p05",
    global_dir
  )
  
  cat(sprintf("  Global (excluding gs0): All=%d | Hyp=%d | Perm=%d | Both=%d\n",
              n_all, n_hyp, n_perm, n_both))
} else {
  # Standard exports
  n_all <- export_gene_list(caas_for_export, "all", gene_list_dir)
  n_hyp <- export_gene_list(
    caas_for_export %>% filter(sig_hyp),
    "hypergeometric_p05",
    gene_list_dir
  )
  n_perm <- export_gene_list(
    caas_for_export %>% filter(sig_perm),
    "permutation_p05",
    gene_list_dir
  )
  n_both <- export_gene_list(
    caas_for_export %>% filter(sig_both),
    "combined_p05",
    gene_list_dir
  )
  
  cat(sprintf("  All=%d | Hyp=%d | Perm=%d | Both=%d\n",
              n_all, n_hyp, n_perm, n_both))
}

cat(sprintf("\nGene lists exported to: %s\n", gene_list_dir))
cat("\n=== Gene List Export Complete ===\n")
```

## 6. Meta-CAAS Extraction

Extract comprehensive CAAS metadata for meta-analysis and cross-study comparison.

```{r meta_caas_export}
cat("\n=== Exporting Meta-CAAS Data ===\n")

# Function to assign random CAAS tags
assign_caas_tags <- function(n_rows) {
  set.seed(1234)  # For reproducibility
  
  tags <- sapply(1:n_rows, function(i) {
    paste0("CAAS_", 
           paste0(sample(LETTERS, 5, TRUE), collapse = ""),
           sprintf("%04d", sample(9999, 1)),
           sample(LETTERS, 1))
  })
  
  return(tags)
}

# Use local directory for Nextflow to capture
meta_dir <- "meta_caas"
createDir(meta_dir)

if (has_caap_group) {
  # Per-group meta-CAAS
  for (grp in unique(caas_for_export$CAAP_Group)) {
    grp_data <- caas_for_export %>% filter(CAAP_Group == grp)
    
    grp_meta <- grp_data %>%
      mutate(
        GenePos = paste(Gene, Position, sep = "_"),
        SignificanceLevel = case_when(
          sig_both ~ "Both",
          sig_hyp ~ "Hypergeometric",
          sig_perm ~ "Permutation",
          TRUE ~ "Not Significant"
        ),
        Tag = assign_caas_tags(n()),
        IsConserved = if_else(!is.na(ConservedPair) & ConservedPair != "", TRUE, FALSE),
        ConservedPair = if_else(
          !is.na(ConservedPairs) & ConservedPairs != "",
          sub("^\\d+:", "", ConservedPairs),
          NA_character_
        ),
        AminoConv = if ("Substitution" %in% names(.)) Substitution else NA_character_,
        AminoEncoded = if ("Encoded" %in% names(.)) Encoded else NA_character_
      ) %>%
      select(
        Tag, GenePos, Gene, Position, Pattern, CAAP_Group,
        Pvalue, Pvalue.boot, Occurrences, Total,
        sig_hyp, sig_perm, sig_both, SignificanceLevel,
        IsConserved, ConservedPair, AminoConv, AminoEncoded
      )
    
    output_file <- file.path(meta_dir, paste0(grp, "_meta_caas.tsv"))
    write_tsv(grp_meta, output_file)
    
    cat(sprintf("  %s: %d positions\n", grp, nrow(grp_meta)))
  }
  
  # Global meta-CAAS
  global_meta <- caas_for_export %>%
    mutate(
      GenePos = paste(Gene, Position, sep = "_"),
      SignificanceLevel = case_when(
        sig_both ~ "Both",
        sig_hyp ~ "Hypergeometric",
        sig_perm ~ "Permutation",
        TRUE ~ "Not Significant"
      ),
      Tag = assign_caas_tags(n()),
      IsConserved = if_else(!is.na(ConservedPair) & ConservedPair != "", TRUE, FALSE),
      ConservedPair = if_else(
        !is.na(ConservedPairs) & ConservedPairs != "",
        sub("^\\d+:", "", ConservedPairs),
        NA_character_
      ),
      AminoConv = if ("Substitution" %in% names(.)) Substitution else NA_character_,
      AminoEncoded = if ("Encoded" %in% names(.)) Encoded else NA_character_
    ) %>%
    select(
      Tag, GenePos, Gene, Position, Pattern, CAAP_Group,
      Pvalue, Pvalue.boot, Occurrences, Total,
      sig_hyp, sig_perm, sig_both, SignificanceLevel,
      IsConserved, ConservedPair, AminoConv, AminoEncoded
    )
  
  output_file <- file.path(meta_dir, "global_meta_caas.tsv")
  write_tsv(global_meta, output_file)
  
  cat(sprintf("  Global: %d positions\n", nrow(global_meta)))
} else {
  # Standard meta-CAAS
  meta_caas <- caas_for_export %>%
    mutate(
      GenePos = paste(Gene, Position, sep = "_"),
      SignificanceLevel = case_when(
        sig_both ~ "Both",
        sig_hyp ~ "Hypergeometric",
        sig_perm ~ "Permutation",
        TRUE ~ "Not Significant"
      ),
      Tag = assign_caas_tags(n()),
      IsConserved = if_else(!is.na(ConservedPair) & ConservedPair != "", TRUE, FALSE),
      ConservedPair = if_else(
        !is.na(ConservedPairs) & ConservedPairs != "",
        sub("^\\d+:", "", ConservedPairs),
        NA_character_
      ),
      AminoConv = if ("Substitution" %in% names(.)) Substitution else NA_character_,
      AminoEncoded = if ("Encoded" %in% names(.)) Encoded else NA_character_
    ) %>%
    select(
      Tag, GenePos, Gene, Position, Pattern,
      Pvalue, Pvalue.boot, Occurrences, Total,
      sig_hyp, sig_perm, sig_both, SignificanceLevel,
      IsConserved, ConservedPair, AminoConv, AminoEncoded
    )
  
  output_file <- file.path(meta_dir, "meta_caas.tsv")
  write_tsv(meta_caas, output_file)
  
  cat(sprintf("  Total: %d positions\n", nrow(meta_caas)))
}

cat(sprintf("\nMeta-CAAS data exported to: %s\n", meta_dir))
cat("\n=== Meta-CAAS Export Complete ===\n")
```

## Session Information

```{r session_info}
sessionInfo()
```
