---
title: "CT Signification Analysis"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    code_folding: hide
    df_print: paged
params:
  discovery_input: NULL
  background_input: NULL
  bootstrap_file: NULL
  output_dir: "."
  caap_mode: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7
)

# Load required libraries
suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
  library(scales)
  library(data.table)
  library(paletteer)
})

# Directory management
createDir <- function(path) {
  if (!dir.exists(path)) dir.create(path, recursive = TRUE, showWarnings = FALSE)
}

# Define color palettes, including a colorblind-friendly palette for categories
tableau_10 <- paletteer_d("ggthemes::Tableau_10")
names(tableau_10) <- c(
  "blue", "orange", "vermillion", "bluish_green",
  "green", "yellow", "reddish_purple", "pink",
  "brown", "gray"
)

# Consistent colorblind-friendly palette
category_palette <- c(
  "Normal" = tableau_10[["gray"]],
  "Dubious" = tableau_10[["blue"]],
  "Extreme" = tableau_10[["vermillion"]],
  "Both" = tableau_10[["reddish_purple"]]
)

removal_palette <- c(
  Discarded = tableau_10[["vermillion"]],
  Good = tableau_10[["green"]]
)
```

## 1. Data Import and Validation

This section loads the filtered CAAS data and validates input files.

```{r data_import, message=FALSE}
cat("=== Loading Filtered CAAS Data ===\n")

# Load filtered discovery file
filtered_discovery_path <- params$discovery_input
if (is.null(filtered_discovery_path)) {
  stop("Filtered discovery file parameter is NULL")
}

cat(sprintf("Attempting to load: %s\n", filtered_discovery_path))
cat(sprintf("  File exists: %s\n", file.exists(filtered_discovery_path)))
cat(sprintf("  Working directory: %s\n", getwd()))

if (!file.exists(filtered_discovery_path)) {
  # List files in current directory for debugging
  cat("Files in current directory:\n")
  print(list.files(".", full.names = TRUE))
  stop("Filtered discovery file not found: ", filtered_discovery_path)
}

discovery_df <- read_tsv(filtered_discovery_path, show_col_types = FALSE)

# discovery.tab is now fully lowercase (gene/mode/trait/position/pattern/ffgn/…).
# Map ONLY the structural keys to the CamelCase names this report uses internally
# (joins on Gene/Position, GenePos = paste(Gene, Position)). Concept columns —
# including the cardinal `pattern` contrast index — stay lowercase (canonical), so
# they are no longer re-capitalised here.
.disc_keymap <- c(Gene = "gene", Position = "position")
for (.new in names(.disc_keymap)) {
  .old <- .disc_keymap[[.new]]
  if (.old %in% names(discovery_df) && !(.new %in% names(discovery_df))) {
    names(discovery_df)[names(discovery_df) == .old] <- .new
  }
}

# Normalize discovery schema (legacy CAAS files may not include caap_group)
if (!"caap_group" %in% names(discovery_df)) {
  discovery_df <- discovery_df %>% mutate(caap_group = "US")
}
cat(sprintf("Loaded discovery file: %s\n", basename(filtered_discovery_path)))
cat(sprintf("  Total CAAS positions: %s\n", format(nrow(discovery_df), big.mark = ",")))

# Detect CAAP mode (after normalization)
has_caap_group <- 'caap_group' %in% names(discovery_df)
cat(sprintf("  CAAP mode detected: %s\n", has_caap_group))

if (has_caap_group) {
  groups <- unique(discovery_df$caap_group)
  cat(sprintf("  Groups present: %s\n", paste(groups, collapse = ", ")))
}

# Load filtered background
filtered_background_path <- params$background_input
if (is.null(filtered_background_path) || !file.exists(filtered_background_path)) {
  stop("Filtered background file not found: ", filtered_background_path)
}

background_df <- read_tsv(filtered_background_path, show_col_types = FALSE)
cat(sprintf("\nLoaded background file: %s\n", basename(filtered_background_path)))
cat(sprintf("  Total genes in background: %s\n", format(nrow(background_df), big.mark = ",")))

# Validate required columns
required_cols <- c("Gene", "Position", "pattern", "pvalue")
missing_cols <- setdiff(required_cols, names(discovery_df))
if (length(missing_cols) > 0) {
  stop("Missing required columns in discovery: ", paste(missing_cols, collapse = ", "))
}

cat("\n=== Data Validation Complete ===\n")
```

## 2. Bootstrap Permutation Integration

Load bootstrap permutation file to calculate empirical p-values.

```{r bootstrap_integration, message=FALSE}
cat("\n=== Loading Bootstrap File ===\n")

# Load bootstrap file
if (is.null(params$bootstrap_file)) {
  stop("Bootstrap file parameter is NULL")
}

bootstrap_file_path <- params$bootstrap_file
if (!file.exists(bootstrap_file_path)) {
  stop("Bootstrap file not found: ", bootstrap_file_path)
}

cat(sprintf("Loading bootstrap file: %s\n", basename(bootstrap_file_path)))

# Load bootstrap data (headerless expected, but tolerate legacy variants)
# Supported formats:
#   5 cols: GenePos caap_group occurrences total pvalue_boot
#   4 cols: GenePos occurrences total pvalue_boot   (legacy CAAS)
bootstrap_raw <- tryCatch({
  read_tsv(
    bootstrap_file_path,
    col_names = FALSE,
    show_col_types = FALSE
  )
}, error = function(e) {
  stop("Failed to load bootstrap file: ", e$message)
})

if (ncol(bootstrap_raw) < 4) {
  stop(sprintf("Bootstrap file has %d columns; expected at least 4", ncol(bootstrap_raw)))
}

if (ncol(bootstrap_raw) >= 5) {
  names(bootstrap_raw)[1:5] <- c("GenePos", "caap_group", "occurrences", "total", "pvalue_boot")
} else {
  names(bootstrap_raw)[1:4] <- c("GenePos", "occurrences", "total", "pvalue_boot")
  bootstrap_raw <- bootstrap_raw %>% mutate(caap_group = "US")
}

# Drop accidental textual header rows if present
bootstrap_raw <- bootstrap_raw %>%
  filter(!is.na(GenePos), GenePos != "gene@position", GenePos != "Gene@Position")

# Normalize types
bootstrap_raw <- bootstrap_raw %>%
  mutate(
    caap_group = if_else(is.na(caap_group) | caap_group == "", "US", caap_group),
    occurrences = suppressWarnings(as.integer(occurrences)),
    total = suppressWarnings(as.integer(total)),
    pvalue_boot = suppressWarnings(as.numeric(pvalue_boot))
  )

# Split GenePos into Gene and Position
bootstrap_df <- bootstrap_raw %>%
  separate(GenePos, into = c("Gene", "Position"), sep = "@", remove = FALSE) %>%
  mutate(Position = as.integer(Position))

# Final safeguard for expected columns
if (!"pvalue_boot" %in% names(bootstrap_df)) {
  stop("Bootstrap integration requires pvalue_boot column after normalization")
}

cat(sprintf("Loaded bootstrap data: %s permutations\n", 
            format(nrow(bootstrap_df), big.mark = ",")))

# Check if bootstrap has CAAP groups
bootstrap_has_groups <- 'caap_group' %in% names(bootstrap_df) && 
                        any(!is.na(bootstrap_df$caap_group))

if (has_caap_group && !bootstrap_has_groups) {
  warning("Bootstrap file missing caap_group data, but discovery has CAAP mode enabled")
  warning("Bootstrap results may not be group-aware")
}

if (bootstrap_has_groups) {
  boot_groups <- unique(bootstrap_df$caap_group)
  cat(sprintf("  Bootstrap groups: %s\n", paste(boot_groups, collapse = ", ")))
}

# Calculate empirical p-values
cat("\nIntegrating bootstrap p-values...\n")

# Bootstrap file already contains pvalue_boot, just need to join
if (has_caap_group && bootstrap_has_groups) {
  # Group-aware permutation p-values
  discovery_with_perm <- discovery_df %>%
    left_join(
      bootstrap_df %>%
        select(Gene, Position, caap_group, pvalue_boot, occurrences, total),
      by = c("Gene", "Position", "caap_group")
    )
} else if (has_caap_group && !bootstrap_has_groups) {
  # Discovery has groups but bootstrap doesn't - join without caap_group
  warning("Joining bootstrap data without caap_group matching")
  discovery_with_perm <- discovery_df %>%
    left_join(
      bootstrap_df %>%
        select(Gene, Position, pvalue_boot, occurrences, total) %>%
        distinct(Gene, Position, .keep_all = TRUE),
      by = c("Gene", "Position")
    )
} else {
  # Standard permutation p-values (no CAAP groups)
  discovery_with_perm <- discovery_df %>%
    left_join(
      bootstrap_df %>%
        select(Gene, Position, pvalue_boot, occurrences, total),
      by = c("Gene", "Position")
    )
}

# Handle missing permutation p-values
discovery_with_perm <- discovery_with_perm %>%
  mutate(
    pvalue_boot = if_else(is.na(pvalue_boot), pvalue, pvalue_boot),
    occurrences = if_else(is.na(occurrences), 0L, as.integer(occurrences)),
    total = if_else(is.na(total), 0L, as.integer(total))
  )

# Normalize optional conserved columns for downstream meta export
if (!"is_conserved_meta" %in% names(discovery_with_perm)) {
  discovery_with_perm <- discovery_with_perm %>% mutate(is_conserved_meta = NA)
}
if (!"conserved_pair" %in% names(discovery_with_perm)) {
  discovery_with_perm <- discovery_with_perm %>% mutate(conserved_pair = NA_character_)
}

cat(sprintf("Integrated bootstrap p-values for %d positions\n", 
            sum(!is.na(discovery_with_perm$pvalue_boot))))

cat("\n=== Bootstrap Integration Complete ===\n")
```

## 3. Significance Filtering

Apply hypergeometric and permutation significance thresholds.

```{r significance_filtering}
cat("\n=== Applying Significance Thresholds ===\n")

alpha <- 0.05

# Apply filters.
# Raw thresholds (sig_hyp/sig_perm/sig_both) are kept as the primary calls.
# Benjamini-Hochberg FDR is added alongside as a multiplicity-controlled
# validation: q-values (pvalue_fdr / pvalue_boot_fdr) and BH-significant flags
# (sig_*_fdr). BH is applied *within each biochemical scheme* (caap_group) — each
# scheme is its own family of position-level tests — so positions are not
# multiply-counted across schemes. (The permutation "p-value" is not a calibrated
# p-value under its null; it is BH-adjusted here for a comparable figure, with
# that caveat.)
caas_filtered <- discovery_with_perm
if (has_caap_group) caas_filtered <- caas_filtered %>% group_by(caap_group)
caas_filtered <- caas_filtered %>%
  mutate(
    pvalue_fdr      = p.adjust(pvalue,      method = "BH"),
    pvalue_boot_fdr = p.adjust(pvalue_boot, method = "BH"),
    is_significant = pvalue <= alpha & pvalue_boot <= alpha,
    sig_hyp = pvalue <= alpha,
    sig_perm = pvalue_boot <= alpha,
    sig_hyp_fdr  = pvalue_fdr      <= alpha,
    sig_perm_fdr = pvalue_boot_fdr <= alpha,
    is_significant_fdr = sig_hyp_fdr & sig_perm_fdr
  ) %>%
  ungroup()

# Summary statistics
if (has_caap_group) {
  # Group-aware summary
  summary_stats <- caas_filtered %>%
    group_by(caap_group) %>%
    summarize(
      Total = n(),
      Hyp_05 = sum(sig_hyp, na.rm = TRUE),
      Perm_05 = sum(sig_perm, na.rm = TRUE),
      Both_05 = sum(is_significant, na.rm = TRUE),
      Hyp_FDR = sum(sig_hyp_fdr, na.rm = TRUE),
      Perm_FDR = sum(sig_perm_fdr, na.rm = TRUE),
      Both_FDR = sum(is_significant_fdr, na.rm = TRUE),
      .groups = "drop"
    )

  cat("\nSignificance Summary by Group (raw alpha=0.05 vs BH-FDR q<=0.05):\n")
  print(kable(summary_stats, format.args = list(big.mark = ",")))

  caas_for_export <- caas_filtered

} else {
  # Standard summary
  summary_stats <- caas_filtered %>%
    summarize(
      Total = n(),
      Hyp_05 = sum(sig_hyp, na.rm = TRUE),
      Perm_05 = sum(sig_perm, na.rm = TRUE),
      Both_05 = sum(is_significant, na.rm = TRUE),
      Hyp_FDR = sum(sig_hyp_fdr, na.rm = TRUE),
      Perm_FDR = sum(sig_perm_fdr, na.rm = TRUE),
      Both_FDR = sum(is_significant_fdr, na.rm = TRUE)
    )

  cat("\nSignificance Summary (raw alpha=0.05 vs BH-FDR q<=0.05):\n")
  print(kable(summary_stats, format.args = list(big.mark = ",")))
  
  caas_for_export <- caas_filtered
}

cat("\n=== Significance Filtering Complete ===\n")
```

## 4. Pattern-Based Analysis

Analyze CAAS patterns and their significance profiles.

```{r pattern_analysis}
cat("\n=== Pattern-Based Analysis ===\n")

# Detect available patterns
patterns_present <- unique(caas_filtered$pattern)
cat(sprintf("Patterns detected: %s\n", paste(patterns_present, collapse = ", ")))

# Pattern distribution
if (has_caap_group) {
  pattern_summary <- caas_filtered %>%
    group_by(caap_group, pattern) %>%
    summarize(
      Total = n(),
      Hyp_05 = sum(sig_hyp, na.rm = TRUE),
      Perm_05 = sum(sig_perm, na.rm = TRUE),
      Both_05 = sum(is_significant, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(caap_group, pattern)
  
  cat("\nPattern Distribution by Group:\n")
  print(kable(pattern_summary, format.args = list(big.mark = ",")))
  
  # Visualization: faceted by group
  plot_data <- pattern_summary %>%
    pivot_longer(
      cols = c(Total, Hyp_05, Perm_05, Both_05),
      names_to = "Filter",
      values_to = "Count"
    ) %>%
    mutate(
      Filter = factor(Filter, levels = c("Total", "Hyp_05", "Perm_05", "Both_05")),
      pattern = factor(pattern)
    )
  
  p_pattern <- ggplot(plot_data, aes(x = pattern, y = Count, fill = Filter)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~ caap_group, scales = "free_y") +
    theme_minimal() +
    labs(
      title = "CAAS Pattern Distribution by Significance Filter",
      subtitle = "Hyp_05 = Hypergeometric | Perm_05 = Permutation | Both_05 = Combined",
      x = "Pattern",
      y = "Count"
    ) +
    scale_fill_manual(values = c(
      "Total" = tableau_10[["gray"]],
      "Hyp_05" = tableau_10[["blue"]],
      "Perm_05" = tableau_10[["green"]],
      "Both_05" = tableau_10[["vermillion"]]
    )) +
    theme(
      axis.text.x = element_text(angle = 0, hjust = 0.5),
      panel.grid.major = element_line(color = "darkgray", size = 0.5),
      legend.position = "bottom"
    )
  
  print(p_pattern)
} else {
  pattern_summary <- caas_filtered %>%
    group_by(pattern) %>%
    summarize(
      Total = n(),
      Hyp_05 = sum(sig_hyp, na.rm = TRUE),
      Perm_05 = sum(sig_perm, na.rm = TRUE),
      Both_05 = sum(is_significant, na.rm = TRUE),
      .groups = "drop"
    )
  
  cat("\nPattern Distribution:\n")
  print(kable(pattern_summary, format.args = list(big.mark = ",")))
  
  # Visualization
  plot_data <- pattern_summary %>%
    pivot_longer(
      cols = c(Total, Hyp_05, Perm_05, Both_05),
      names_to = "Filter",
      values_to = "Count"
    ) %>%
    mutate(
      Filter = factor(Filter, levels = c("Total", "Hyp_05", "Perm_05", "Both_05")),
      pattern = factor(pattern)
    )
  
  p_pattern <- ggplot(plot_data, aes(x = pattern, y = Count, fill = Filter)) +
    geom_bar(stat = "identity", position = "dodge") +
    theme_minimal() +
    labs(
      title = "CAAS Pattern Distribution by Significance Filter",
      subtitle = "Hyp_05 = Hypergeometric | Perm_05 = Permutation | Both_05 = Combined",
      x = "Pattern",
      y = "Count"
    ) +
    scale_fill_manual(values = c(
      "Total" = tableau_10[["gray"]],
      "Hyp_05" = tableau_10[["blue"]],
      "Perm_05" = tableau_10[["green"]],
      "Both_05" = tableau_10[["vermillion"]]
    )) +
    theme(
      axis.text.x = element_text(angle = 0, hjust = 0.5),
      panel.grid.major = element_line(color = "darkgray", size = 0.5),
      legend.position = "bottom"
    )
  
  print(p_pattern)
}

cat("\n=== Pattern Analysis Complete ===\n")
```

## 5. Meta-CAAS Extraction

Extract comprehensive CAAS metadata for meta-analysis and cross-study comparison.

```{r meta_caas_export}
cat("\n=== Exporting Meta-CAAS Data ===\n")

# Function to assign random CAAS tags
assign_caas_tags <- function(n_rows) {
  set.seed(1234)  # For reproducibility
  
  tags <- sapply(1:n_rows, function(i) {
    paste0("CAAS_", 
           paste0(sample(LETTERS, 5, TRUE), collapse = ""),
           sprintf("%04d", sample(9999, 1)),
           sample(LETTERS, 1))
  })
  
  return(tags)
}

# Use local directory for Nextflow to capture
meta_dir <- "meta_caas"
createDir(meta_dir)

if (has_caap_group) {
  # Per-group meta-CAAS
  for (grp in unique(caas_for_export$caap_group)) {
    grp_data <- caas_for_export %>% filter(caap_group == grp)
    
    grp_meta <- grp_data %>%
      mutate(
        GenePos = paste(Gene, Position, sep = "_"),
        significance_level = case_when(
          is_significant ~ "Both",
          sig_hyp ~ "Hypergeometric",
          sig_perm ~ "Permutation",
          TRUE ~ "Not Significant"
        ),
        tag = assign_caas_tags(n()),
        is_conserved_meta = if_else(!is.na(is_conserved_meta) & toupper(trimws(as.character(is_conserved_meta))) == "TRUE", TRUE, FALSE),
        conserved_pair = if_else(
          !is.na(conserved_pair) & conserved_pair != "",
          sub("^\\d+:", "", conserved_pair),
          NA_character_
        ),
        caas = if ("caas" %in% names(.)) caas else NA_character_,
        amino_encoded = if ("amino_encoded" %in% names(.)) amino_encoded else NA_character_
      ) %>%
      select(
        tag, GenePos, Gene, Position, pattern, caap_group,
        pvalue, pvalue_boot, pvalue_fdr, pvalue_boot_fdr, occurrences, total,
        sig_hyp, sig_perm, is_significant,
        sig_hyp_fdr, sig_perm_fdr, is_significant_fdr, significance_level,
        is_conserved_meta, conserved_pair, caas, amino_encoded
      )

    output_file <- file.path(meta_dir, paste0(grp, "_meta_caas.tsv"))
    write_tsv(grp_meta, output_file)
    
    cat(sprintf("  %s: %d positions\n", grp, nrow(grp_meta)))
  }
  
  # Global meta-CAAS
  global_meta <- caas_for_export %>%
    mutate(
      GenePos = paste(Gene, Position, sep = "_"),
      significance_level = case_when(
        is_significant ~ "Both",
        sig_hyp ~ "Hypergeometric",
        sig_perm ~ "Permutation",
        TRUE ~ "Not Significant"
      ),
      tag = assign_caas_tags(n()),
      is_conserved_meta = if_else(!is.na(is_conserved_meta) & toupper(trimws(as.character(is_conserved_meta))) == "TRUE", TRUE, FALSE),
      conserved_pair = if_else(
        !is.na(conserved_pair) & conserved_pair != "",
        sub("^\\d+:", "", conserved_pair),
        NA_character_
      ),
      caas = if ("caas" %in% names(.)) caas else NA_character_,
      amino_encoded = if ("amino_encoded" %in% names(.)) amino_encoded else NA_character_
    ) %>%
    select(
      tag, GenePos, Gene, Position, pattern, caap_group,
      pvalue, pvalue_boot, pvalue_fdr, pvalue_boot_fdr, occurrences, total,
      sig_hyp, sig_perm, is_significant,
      sig_hyp_fdr, sig_perm_fdr, is_significant_fdr, significance_level,
      is_conserved_meta, conserved_pair, caas, amino_encoded
    )

  output_file <- file.path(meta_dir, "global_meta_caas.tsv")
  write_tsv(global_meta, output_file)
  
  cat(sprintf("  Global: %d positions\n", nrow(global_meta)))
} else {
  # Standard meta-CAAS
  meta_caas <- caas_for_export %>%
    mutate(
      GenePos = paste(Gene, Position, sep = "_"),
      significance_level = case_when(
        is_significant ~ "Both",
        sig_hyp ~ "Hypergeometric",
        sig_perm ~ "Permutation",
        TRUE ~ "Not Significant"
      ),
      tag = assign_caas_tags(n()),
      is_conserved_meta = if_else(!is.na(is_conserved_meta) & toupper(trimws(as.character(is_conserved_meta))) == "TRUE", TRUE, FALSE),
      conserved_pair = if_else(
        !is.na(conserved_pair) & conserved_pair != "",
        sub("^\\d+:", "", conserved_pair),
        NA_character_
      ),
      caas = if ("caas" %in% names(.)) caas else NA_character_,
      amino_encoded = if ("amino_encoded" %in% names(.)) amino_encoded else NA_character_
    ) %>%
    select(
      tag, GenePos, Gene, Position, pattern,
      pvalue, pvalue_boot, pvalue_fdr, pvalue_boot_fdr, occurrences, total,
      sig_hyp, sig_perm, is_significant,
      sig_hyp_fdr, sig_perm_fdr, is_significant_fdr, significance_level,
      is_conserved_meta, conserved_pair, caas, amino_encoded
    )

  output_file <- file.path(meta_dir, "meta_caas.tsv")
  write_tsv(meta_caas, output_file)
  
  cat(sprintf("  Total: %d positions\n", nrow(meta_caas)))
}

cat(sprintf("\nMeta-CAAS data exported to: %s\n", meta_dir))
cat("\n=== Meta-CAAS Export Complete ===\n")
```

## Session Information

```{r session_info}
sessionInfo()
```
