---
title: "FADE — Directional Positive Selection Report"
subtitle: "`r paste0(params$traitname, ' | ', toupper(params$direction), ' direction')`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
params:
  # ── required ────────────────────────────────────────────────────────────────
  traitname:      "unknown_trait"
  direction:      "top"          # top | bottom
  json_dir:       "."            # directory of per-gene FADE JSON outputs
  output_dir:     "."
  # ── optional ────────────────────────────────────────────────────────────────
  gene_list:      NULL           # path to gene list for overlap annotation
  bf_threshold:   10             # Bayes Factor threshold for site-level calls
---

<!-- FADE_report.Rmd — Summarises FADE site-level positive selection results per direction. -->
<!-- PhyloPhere | subworkflows/FADE/local/ -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo       = TRUE,
    message    = FALSE,
    warning    = FALSE,
    fig.align  = "center"
)

# ── Dependencies ───────────────────────────────────────────────────────────────
suppressPackageStartupMessages({
    library(dplyr)
    library(ggplot2)
    library(knitr)
    library(DT)
    library(jsonlite)
})

# ── Pipeline palette & theme (sourced from shared utility) ─────────────────────
# source(file.path(params$output_dir, "obj", "palettes.R"))

# Use inline fallback if obj/ is not available
PALETTE_DIRECTION <- c(top = "#E63946", bottom = "#457B9D")
```

## Overview {.tabset}

> **Trait:** `r params$traitname`  |  **Direction:** `r toupper(params$direction)`
>
> FADE (Fast Ancestral Sequence reconstruction with empirical bayes Dropout Elimination)
> identifies sites evolving under positive selection in the foreground branches.
> Sites with Bayes Factor ≥ `r params$bf_threshold` are considered significant.

### Site-level results

```{r load-fade}
# Load all per-gene FADE JSON files from json_dir
json_files <- list.files(params$json_dir, pattern = "\\.json$", full.names = TRUE)

if (length(json_files) == 0) {
    knitr::knit_exit("No FADE JSON files found in json_dir.")
}

fade_raw <- lapply(json_files, function(f) {
    gene <- tools::file_path_sans_ext(basename(f))
    tryCatch({
        dat <- jsonlite::fromJSON(f, flatten = TRUE)
        # Extract site-level BF from the standard FADE JSON schema
        sites <- as.data.frame(dat[["MLE"]][["content"]])
        sites$gene <- gene
        sites
    }, error = function(e) {
        message("Skipping ", gene, ": ", conditionMessage(e))
        NULL
    })
}) |> dplyr::bind_rows()
```

```{r site-table}
# Sites above the BF threshold
sig_sites <- fade_raw |>
    dplyr::filter(`Bayes Factor` >= params$bf_threshold) |>
    dplyr::arrange(dplyr::desc(`Bayes Factor`))

DT::datatable(sig_sites,
    caption = paste0("Sites with BF ≥ ", params$bf_threshold,
                     " — ", params$direction, " direction"),
    options = list(pageLength = 20, scrollX = TRUE)
)
```

### Gene-level summary

```{r gene-summary}
gene_summary <- fade_raw |>
    dplyr::group_by(gene) |>
    dplyr::summarise(
        n_sites_tested = dplyr::n(),
        n_sites_sig    = sum(`Bayes Factor` >= params$bf_threshold, na.rm = TRUE),
        max_bf         = max(`Bayes Factor`, na.rm = TRUE),
        .groups = "drop"
    ) |>
    dplyr::arrange(dplyr::desc(max_bf))

knitr::kable(gene_summary, digits = 2,
    caption = "Gene-level FADE summary (sorted by max Bayes Factor)")
```

## Session info {.tabset}

```{r session}
sessionInfo()
```
