---
title: "Dataset Exploration for Trait Analysis"
author: "PhyloPhere Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    code_folding: hide
    df_print: paged
params:
  trait_file: NULL
  tree_file: NULL
  output_dir: !r getwd()
  seed: ""
  clade_name: "clade"
  taxon_of_interest: "family"
  traitname: "trait"
  n_trait: ""
  c_trait: ""
  tax_id: ""
  secondary_trait: ""
  branch_trait: ""
  discrete_method: "decile"
  top_quantile: "0.90"
  bottom_quantile: "0.10"
---

```{r header}

#  ██████╗ ██╗  ██╗██╗   ██╗██╗      ██████╗ ██████╗ ██╗  ██╗███████╗██████╗ ███████╗
#  ██╔══██╗██║  ██║╚██╗ ██╔╝██║     ██╔═══██╗██╔══██╗██║  ██║██╔════╝██╔══██╗██╔════╝
#  ██████╔╝███████║ ╚████╔╝ ██║     ██║   ██║██████╔╝███████║█████╗  ██████╔╝█████╗
#  ██╔═══╝ ██╔══██║  ╚██╔╝  ██║     ██║   ██║██╔═══╝ ██╔══██║██╔══╝  ██╔══██╗██╔══╝
#  ██║     ██║  ██║   ██║   ███████╗╚██████╔╝██║     ██║  ██║███████╗██║  ██║███████╗
#  ╚═╝     ╚═╝  ╚═╝   ╚═╝   ╚══════╝ ╚═════╝ ╚═╝     ╚═╝  ╚═╝╚══════╝╚═╝  ╚═╝╚══════╝

## PHYLOPHERE: A Nextflow pipeline including a complete set
## of phylogenetic comparative tools and analyses for Phenome-Genome studies

### Github: https://github.com/nozerorma/caastools/nf-phylophere
### Author: Miguel Ramon (miguel.ramon@upf.edu)

#### File: dataset_exploration.Rmd

```

# Data Visualization and Processing

Script to visualize and process data. It generates family distribution plots, summary tables, and a processed dataset.

---

## Setup and Directory Configuration
This section configures the working environment, sets directories, and loads necessary functions and libraries.

```{r setup, message=FALSE, warning=FALSE}

# Make params available in environment before sourcing commons.R
if (!exists("params")) {
  stop("params object not found - ensure this Rmd is rendered with rmarkdown::render()")
}
assign("params", params, envir = .GlobalEnv)

# Call the setup function from commons.R
source("./obj/commons.R")
setup_rmd()

# Load necessary libraries
library(dplyr)
library(ggplot2)
library(readr)
library(reshape2)
library(paletteer)

# Objects
color_palette <- paste0(clade_name, "_palette") # Color palette for the clade
capitalized_taxon <- tools::toTitleCase(taxon_of_interest) # Capitalized taxon name
capitalized_trait <- tools::toTitleCase(gsub("_", " ", trait)) # Capitalized and deconvoluted trait name
capitalized_clade <- tools::toTitleCase(gsub("_", " ", clade_name)) # Capitalized and deconvoluted clade name

createDir(species_distribution_dir)

```

## 1. Dataset Visualization
This section visualizes the distribution of species across families to provide an overview of the dataset.

```{r dataset_visualization, fig.height=7, fig.width=7, dev='png', dpi=320}

# 1. DATA VISUALIZATION  ------------------------------------------------------
# Calculate species count per taxon_of_interest
# Note that for this section we use the original trait dataframe (trait_df_ori), and will generate two plots
# : one for the pruned tree species, and another for all species in the original dataframe.

taxa_distribution <- trait_df_ori %>%
  dplyr::rename(taxa = !!taxon_of_interest) %>%
  dplyr::group_by(taxa) %>%
  dplyr::summarise(count = n())

# Ensure species order matches phylogeny
ordered_species <-  tidytree::as_tibble(pruned_tree) %>%
  dplyr::arrange(desc(.data$node)) %>%
  # Remove internal nodes by filtering only rows where label is not NA
  dplyr::filter(!is.na(.data$label)) %>%
  dplyr::pull(.data$label)

print(ordered_species)

# Define ordered taxa for plotting based on phylogeny order
ordered_taxa <- ordered_species %>%
  sapply(function(sp) {
    idx <- which(trait_df_ori$species == sp)
    if (length(idx) == 0) return(NA_character_)
    trait_df_ori[[taxon_of_interest]][idx[1]]
  }) %>%
  unname() %>%
  na.omit() %>%
  unique()

print(ordered_taxa)

## PLOT FUNCTION --------------------------------------------------------------
taxa_plot.f <- function(df, ordered_taxa_levels, caption_text = NULL){
  max_count <- max(df$count, na.rm = TRUE)
  plot_profile <- species_plot_profile(
    n_species = sum(df$count, na.rm = TRUE),
    n_taxa = length(ordered_taxa_levels)
  )$contrast
  palette_values <- resolve_taxa_palette(ordered_taxa_levels, palette_name = color_palette)
  count_label_size <- clamp_value(5.0 - 0.06 * length(ordered_taxa_levels), 3.2, 5.0)
  right_margin <- clamp_value(45 + 3 * nchar(as.character(max_count)), 50, 90)
  count_limit <- max(1, max_count * 1.12 + 0.8)
  debug_log("taxa_plot.f rows = %d, mapped_taxa = %d", nrow(df), ifelse(is.null(palette_values), 0, length(palette_values)))
  fill_scale <- if (is.null(palette_values) || length(palette_values) == 0) {
    ggplot2::scale_fill_discrete()
  } else {
    ggplot2::scale_fill_manual(
      values = palette_values,
      breaks = rev(ordered_taxa_levels),
      limits = ordered_taxa_levels,
      drop = FALSE
    )
  }
<<<<<<< HEAD
  ggplot2::ggplot(df, 
                aes(x = taxa, y = count, fill = taxa)) +
  scale_x_discrete(limits = rev(ordered_taxa_levels)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = count), 
            vjust = 0.5, hjust = -1.5, 
            color = "black", fontface = "bold", 
            size = 6) +
  labs(x = paste(capitalized_taxon, "-", clade_name), 
       y = "Number of species", 
       title = paste(capitalized_taxon, "distribution in", clade_name)) +
  fill_scale + 
  scale_y_continuous(limits = c(0, max_count+2), expand = c(0, 0)) +
  theme_minimal(base_size = 16) +
  theme(
    axis.text.y = element_text(hjust = 1, size = 18, face = "bold", color = "black"),
    axis.text.x = element_blank(),
    plot.title = element_text(size = 20, hjust = 0.5, margin = margin(t = 20), face = "bold", color = "black"),
    axis.title.x = element_text(size = 18, hjust = 0.5, margin = margin(t = 20), face = "bold", color = "black"),
    axis.title.y = element_blank(),
    legend.position = "none"
=======
  plot <- ggplot2::ggplot(
    df,
    aes(x = taxa, y = count, fill = taxa)
>>>>>>> asr
  ) +
    scale_x_discrete(limits = rev(ordered_taxa_levels), drop = FALSE) +
    geom_col(width = 0.75) +
    geom_text(
      aes(label = count),
      hjust = -0.15,
      color = "black",
      fontface = "bold",
      size = count_label_size
    ) +
    labs(
      x = paste(capitalized_taxon, "-", clade_name),
      y = "Number of species",
      title = paste(capitalized_taxon, "distribution in", clade_name),
      caption = caption_text
    ) +
    fill_scale +
    scale_y_continuous(
      limits = c(0, count_limit),
      expand = expansion(mult = c(0, 0.02))
    ) +
    theme_minimal(base_size = 16) +
    theme(
      axis.text.y = element_text(
        hjust = 1,
        size = clamp_value(18 - 0.4 * length(ordered_taxa_levels), 10, 18),
        face = "bold",
        color = "black"
      ),
      axis.text.x = element_blank(),
      plot.title = element_text(size = plot_profile$title_size, hjust = 0.5, margin = margin(t = 20), face = "bold", color = "black"),
      axis.title.x = element_text(size = plot_profile$axis_title, hjust = 0.5, margin = margin(t = 20), face = "bold", color = "black"),
      axis.title.y = element_blank(),
      plot.caption = element_text(size = plot_profile$caption_size, hjust = 0, face = "italic", color = "grey30", margin = margin(t = 14)),
      plot.margin = margin(20, right_margin, 35, 20),
      legend.position = "none"
    ) +
    coord_flip(clip = "off")

  attr(plot, "plot_dims") <- list(
    width = clamp_value(plot_profile$width - 1.0, 8.0, 10.0),
    height = clamp_value(4.8 + 0.38 * length(ordered_taxa_levels), 6.0, 13.5)
  )
  plot
}


# Order taxa for the full dataset using original data order
ordered_taxa_all <- unique(taxa_distribution$taxa)

# Create bar plot for taxa distribution (all taxa)
taxa_distribution_all <- taxa_distribution %>%
  dplyr::mutate(taxa = factor(taxa, levels = ordered_taxa_all)) %>%
  dplyr::arrange(taxa, .by_group = TRUE)

p1 <- taxa_plot.f(
  taxa_distribution_all,
  ordered_taxa_all,
  caption_text = "Ordering note: this non-pruned plot is shown in alphabetical order\nbecause some taxa may be absent from the provided phylogeny."
)
plot_dims_all <- attr(p1, "plot_dims")
knitr::opts_current$set(fig.width = plot_dims_all$width, fig.height = plot_dims_all$height)

# Save the plot to the results directory
ggplot2::ggsave(file.path(species_distribution_dir, paste0(clade_name, "_", taxon_of_interest, "_distribution_original.png")), p1, 
                device = "png", width = plot_dims_all$width, height = plot_dims_all$height, dpi = "retina")

# Display the plot
p1

# Prune taxa distribution to only include species in the pruned tree
pruned_taxa_distribution <- taxa_distribution %>%
  dplyr::filter(taxa %in% (trait_df %>%
                             dplyr::pull(!!sym(taxon_of_interest)) %>%
                             unique()))

# Create bar plot for pruned taxa distribution
pruned_taxa_distribution <- pruned_taxa_distribution %>%
  dplyr::mutate(taxa = factor(taxa, levels = ordered_taxa)) %>%
  dplyr::arrange(taxa, .by_group = TRUE)

p2 <- taxa_plot.f(
  pruned_taxa_distribution,
  ordered_taxa,
  caption_text = "Ordering note: this pruned plot follows the phylogenetic order\nof taxa represented in the provided tree."
)
plot_dims_pruned <- attr(p2, "plot_dims")
knitr::opts_current$set(fig.width = plot_dims_pruned$width, fig.height = plot_dims_pruned$height)

# Save the plot to the results directory
ggplot2::ggsave(file.path(species_distribution_dir, paste0(clade_name, "_", taxon_of_interest, "_distribution_pruned.png")), p2, 
                device = "png", width = plot_dims_pruned$width, height = plot_dims_pruned$height, dpi = "retina")

# Display the plot
p2

```

## 2. Data Processing and statistical overview

```{r stats_overview}

# Apply stats.f to trait_df
# Here we directly use trait_df from the global environment, already loaded and pruned using the phylogeny
stats_df <- stats.f(trait_df)

# Save the melted dataframe
save_path <- file.path(species_distribution_dir, "trait_stats.csv")
if (!is.null(stats_df)) {
  write.csv(stats_df, save_path, row.names = FALSE)
}

# Print an excerpt
if (!is.null(stats_df)) {
  print(head(stats_df))
}

```

## Session Information

```{r debug_dump, results='asis'}

if (length(phylo_debug_log) > 0) {
  cat("### Debug log\n")
  cat(paste0("[DEBUG] ", phylo_debug_log, "\n"))
}

```

```{r session_info}

sessionInfo()

```
