---
title: "Dataset Exploration for Trait Analysis"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
params:
  trait_file: NULL
  tree_file: NULL
  output_dir: !r getwd()
  seed: ""
  clade_name: "clade"
  taxon_of_interest: "family"
  sp_colname: "species"
  traitname: "trait"
  n_trait: ""
  c_trait: ""
  secondary_trait: ""
  branch_trait: ""
  pss_top_pct: "0.05"
  perm_strategy: "best_model"
  trait_type: ""
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

# Data Visualization and Processing

Script to visualize and process data. It generates family distribution plots, summary tables, and a processed dataset.

---

<!-- 1.Dataset_exploration.Rmd — Species counts per taxon and the per-species trait statistics table. -->
<!-- PhyloPhere | subworkflows/TRAIT_ANALYSIS/local/ -->

## Setup and Directory Configuration
This section configures the working environment, sets directories, and loads necessary functions and libraries.

```{r setup, message=FALSE, warning=FALSE}

if (exists("params")) {
  tryCatch(assign("params", params, envir = .GlobalEnv), error = function(e) NULL)
}

# Call the setup function from commons.R
source("./src/commons.R")
setup_rmd()

# ── Dependencies ──────────────────────────────────────────────────────────────

library(dplyr)
library(ggplot2)
library(readr)
library(reshape2)
library(paletteer)


# ── Labels and output directory ───────────────────────────────────────────────

color_palette <- paste0(clade_name, "_palette") # name of the clade palette defined in palettes.R
capitalized_taxon <- tools::toTitleCase(taxon_of_interest)
capitalized_trait <- tools::toTitleCase(gsub("_", " ", trait)) # underscores shown as spaces
capitalized_clade <- tools::toTitleCase(gsub("_", " ", clade_name))

createDir(species_distribution_dir)

```

## 1. Dataset Visualization
This section visualizes the distribution of species across families to provide an overview of the dataset.

```{r dataset_visualization, fig.height=7, fig.width=7, dev='png', dpi=320}

# ── Species per taxon ─────────────────────────────────────────────────────────

# Two bar plots are drawn: all species of the original trait table (trait_df_ori, the table
# before pruning) and the species kept after pruning (trait_df).

taxa_distribution <- trait_df_ori %>%
  dplyr::rename(taxa = !!taxon_of_interest) %>%
  dplyr::group_by(taxa) %>%
  dplyr::summarise(count = n())

# Species in tree order, so the taxa follow the phylogeny.
ordered_species <-  tidytree::as_tibble(pruned_tree) %>%
  dplyr::arrange(desc(.data$node)) %>%
  # Tip nodes are numbered 1..Ntip (ape convention). Filtering on `!is.na(label)` alone
  # would also admit internal-node support labels, which repeat across nodes and give
  # factor(levels =) duplicates.
  dplyr::filter(.data$node <= length(pruned_tree$tip.label)) %>%
  dplyr::pull(.data$label)

# Taxa ordered by the first species of each taxon in tree order.
ordered_taxa <- ordered_species %>%
  sapply(function(sp) {
    idx <- which(trait_df_ori$species == sp)
    if (length(idx) == 0) return(NA_character_)
    trait_df_ori[[taxon_of_interest]][idx[1]]
  }) %>%
  unname() %>%
  na.omit() %>%
  unique()


# ── Bar plot ──────────────────────────────────────────────────────────────────

# Horizontal bars with the species count per taxon, in the order of `ordered_taxa_levels`.
# Label size, margins and canvas size follow the number of taxa and the largest count.
taxa_plot.f <- function(df, ordered_taxa_levels, caption_text = NULL){
  max_count <- max(df$count, na.rm = TRUE)
  plot_profile <- species_plot_profile(
    n_species = sum(df$count, na.rm = TRUE),
    n_taxa = length(ordered_taxa_levels)
  )$contrast
  palette_values <- resolve_taxa_palette(ordered_taxa_levels, palette_name = color_palette)
  count_label_size <- clamp_value(5.0 - 0.06 * length(ordered_taxa_levels), 3.2, 5.0)
  right_margin <- clamp_value(45 + 3 * nchar(as.character(max_count)), 50, 90)
  count_limit <- max(1, max_count * 1.12 + 0.8)
  debug_log("taxa_plot.f rows = %d, mapped_taxa = %d", nrow(df), ifelse(is.null(palette_values), 0, length(palette_values)))
  fill_scale <- if (is.null(palette_values) || length(palette_values) == 0) {
    ggplot2::scale_fill_discrete()
  } else {
    ggplot2::scale_fill_manual(
      values = palette_values,
      breaks = rev(ordered_taxa_levels),
      limits = ordered_taxa_levels,
      drop = FALSE
    )
  }
  plot <- ggplot2::ggplot(
    df,
    aes(x = taxa, y = count, fill = taxa)
  ) +
    scale_x_discrete(limits = rev(ordered_taxa_levels), drop = FALSE) +
    geom_col(width = 0.75) +
    geom_text(
      aes(label = count),
      hjust = -0.15,
      color = "black",
      fontface = "bold",
      size = count_label_size
    ) +
    labs(
      x = paste(capitalized_taxon, "-", clade_name),
      y = "Number of species",
      title = paste(capitalized_taxon, "distribution in", clade_name),
      caption = caption_text
    ) +
    fill_scale +
    scale_y_continuous(
      limits = c(0, count_limit),
      expand = expansion(mult = c(0, 0.02))
    ) +
    theme_minimal(base_size = 16) +
    theme(
      axis.text.y = element_text(
        hjust = 1,
        size = clamp_value(18 - 0.4 * length(ordered_taxa_levels), 10, 18),
        face = "bold",
        color = "black"
      ),
      axis.text.x = element_blank(),
      plot.title = element_text(size = plot_profile$title_size, hjust = 0, margin = ggplot2::margin(t = 20, b = 6), face = "bold", color = "black"),
      axis.title.x = element_text(size = plot_profile$axis_title, margin = ggplot2::margin(t = 12), face = "bold", color = "black"),
      axis.title.y = element_blank(),
      plot.caption = element_text(size = plot_profile$caption_size, hjust = 0, face = "italic", color = "grey30", margin = margin(t = 14)),
      plot.margin = margin(20, right_margin, 35, 20),
      legend.position = "none"
    ) +
    coord_flip(clip = "off")

  attr(plot, "plot_dims") <- list(
    width = clamp_value(plot_profile$width - 1.0, 8.0, 10.0),
    height = clamp_value(4.8 + 0.38 * length(ordered_taxa_levels), 6.0, 13.5)
  )
  plot
}


# All species: taxa in the order of the original table, since some may be absent from the tree.
ordered_taxa_all <- unique(taxa_distribution$taxa)

taxa_distribution_all <- taxa_distribution %>%
  dplyr::mutate(taxa = factor(taxa, levels = ordered_taxa_all)) %>%
  dplyr::arrange(taxa, .by_group = TRUE)

p1 <- taxa_plot.f(
  taxa_distribution_all,
  ordered_taxa_all,
  caption_text = "Ordering note: this non-pruned plot is shown in alphabetical\norder because some taxa may be absent from the provided\nphylogeny."
)
plot_dims_all <- attr(p1, "plot_dims")
knitr::opts_current$set(fig.width = plot_dims_all$width, fig.height = plot_dims_all$height)

# Saved to the results directory.
ggplot2::ggsave(file.path(species_distribution_dir, paste0(clade_name, "_", taxon_of_interest, "_distribution_original.png")), p1, 
                device = "png", width = plot_dims_all$width, height = plot_dims_all$height, dpi = "retina")

p1

# Pruned species: the counts of trait_df, in tree order of the taxa.
pruned_taxa_distribution <- trait_df %>%
  dplyr::rename(taxa = !!sym(taxon_of_interest)) %>%
  dplyr::group_by(taxa) %>%
  dplyr::summarise(count = n(), .groups = "drop")

pruned_taxa_distribution <- pruned_taxa_distribution %>%
  dplyr::mutate(taxa = factor(taxa, levels = ordered_taxa)) %>%
  dplyr::arrange(taxa, .by_group = TRUE)

p2 <- taxa_plot.f(
  pruned_taxa_distribution,
  ordered_taxa,
  caption_text = "Ordering note: this pruned plot follows the phylogenetic\norder of taxa represented in the provided tree."
)
plot_dims_pruned <- attr(p2, "plot_dims")
knitr::opts_current$set(fig.width = plot_dims_pruned$width, fig.height = plot_dims_pruned$height)

# Saved to the results directory.
ggplot2::ggsave(file.path(species_distribution_dir, paste0(clade_name, "_", taxon_of_interest, "_distribution_pruned.png")), p2, 
                device = "png", width = plot_dims_pruned$width, height = plot_dims_pruned$height, dpi = "retina")

p2

```

## 2. Data Processing and statistical overview
Stats excerpt (head only)

```{r stats_overview}

# Per-species statistics (stats.f in stats.R) of trait_df, the table already matched to the tree.
# trait_stats.csv is read by 2.Phenotype_exploration.Rmd.
stats_df <- stats.f(trait_df)

save_path <- file.path(species_distribution_dir, "trait_stats.csv")
if (!is.null(stats_df)) {
  write.csv(stats_df, save_path, row.names = FALSE)
}

if (!is.null(stats_df)) {
  print(head(stats_df))
}

```

---

## Summary

This report summarizes the results of the **Trait Analysis** analysis.

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — Trait Analysis Module*
*Date of generation: `r Sys.Date()`*
