---
title: "Phenotype Exploration for Trait Analysis"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
params:
  trait_file: NULL
  tree_file: NULL
  output_dir: !r getwd()
  seed: ""
  clade_name: "clade"
  taxon_of_interest: "family"
  sp_colname: "species"
  traitname: "trait"
  n_trait: ""
  c_trait: ""
  secondary_trait: ""
  branch_trait: ""
  pss_top_pct: "0.05"
  perm_strategy: "best_model"
  trait_type: ""
  max_contrasts: "0"
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

<!-- 2.Phenotype_exploration.Rmd — Contrast, violin, ancestral-state and fan-tree plots of the trait. -->
<!-- PhyloPhere | subworkflows/TRAIT_ANALYSIS/local/ -->

```{r header}

#  ██████╗ ██╗  ██╗██╗   ██╗██╗      ██████╗ ██████╗ ██╗  ██╗███████╗██████╗ ███████╗
#  ██╔══██╗██║  ██║╚██╗ ██╔╝██║     ██╔═══██╗██╔══██╗██║  ██║██╔════╝██╔══██╗██╔════╝
#  ██████╔╝███████║ ╚████╔╝ ██║     ██║   ██║██████╔╝███████║█████╗  ██████╔╝█████╗
#  ██╔═══╝ ██╔══██║  ╚██╔╝  ██║     ██║   ██║██╔═══╝ ██╔══██║██╔══╝  ██╔══██╗██╔══╝
#  ██║     ██║  ██║   ██║   ███████╗╚██████╔╝██║     ██║  ██║███████╗██║  ██║███████╗
#  ╚═╝     ╚═╝  ╚═╝   ╚═╝   ╚══════╝ ╚═════╝ ╚═╝     ╚═╝  ╚═╝╚══════╝╚═╝  ╚═╝╚══════╝

## PHYLOPHERE: A Nextflow pipeline including a complete set
## of phylogenetic comparative tools and analyses for Phenome-Genome studies

### Github: https://github.com/nozerorma/caastools/nf-phylophere
### Author: Miguel Ramon (miguel.ramon@upf.edu)

#### File: 2.Phenotype_exploration.Rmd

```

# Phenotype Exploration
This script allows for the visualization of trait distributions and phylogenetic patterns. The script includes contrast plots and phylogenetic trees to assess trait relationships.

## Setup and Directory Configuration
This section configures the working environment, sets directories, and loads necessary functions and libraries.

```{r setup, message=FALSE, warning=FALSE}

if (exists("params")) {
  tryCatch(assign("params", params, envir = .GlobalEnv), error = function(e) NULL)
}

# Call the setup function from commons.R
source("./src/commons.R")
setup_rmd()

# Fallback logger, used only when commons.R has not defined debug_log().
if (is.null(getOption("phylo_debug"))) {
  options(phylo_debug = TRUE)
}
if (!exists("phylo_debug_log", envir = .GlobalEnv)) {
  phylo_debug_log <- character()
}
if (!exists("debug_log", inherits = TRUE)) {
  debug_log <- function(...) {
    msg <- sprintf(...)
    phylo_debug_log <<- c(phylo_debug_log, msg)
    cat("[DEBUG] ", msg, "\n", sep = "")
  }
}

# ── Dependencies ──────────────────────────────────────────────────────────────

library(dplyr)
library(ggplot2)
library(readr)
library(reshape2)
library(ggrepel)
library(ggtree)
library(ggnewscale)
library(ggstar)
library(colorspace)
library(treeio)
library(tidytree)
library(phylolm)
library(ggtreeExtra)
library(geiger)
library(rphylopic)
library(paletteer)



# ── Labels, palettes and output directories ───────────────────────────────────

color_palette <- paste0(clade_name, "_palette") # name of the clade palette defined in palettes.R
capitalized_taxon <- tools::toTitleCase(taxon_of_interest)
capitalized_trait <- tools::toTitleCase(gsub("_", " ", trait)) # underscores shown as spaces
capitalized_clade <- tools::toTitleCase(gsub("_", " ", clade_name))

# Tableau 10 under colorblind-safe names.
tableau_10 <- paletteer_d("ggthemes::Tableau_10")
names(tableau_10) <- c(
  "blue", "orange", "vermillion", "bluish_green",
  "green", "yellow", "reddish_purple", "pink",
  "brown", "gray"
)

createDir(extreme_plots_dir)
createDir(asr_trees)
createDir(phylo_distribution_dir)

```

## 1. Data Import
This section imports trait summary data from the dataset exploration stage.

```{r trait_import, message=FALSE, warning=FALSE}

# Per-species statistics written by 1.Dataset_exploration.Rmd.
stats_path <- file.path(species_distribution_dir, "trait_stats.csv")
if (!file.exists(stats_path)) {
  stop("Missing trait stats file: ", stats_path)
}
stats_df <- read.csv(stats_path, stringsAsFactors = FALSE)

# When the table lacks the quantile and outlier columns the plots need, it is recomputed with stats.f().
.required_summary_cols <- c("g_q10", "g_q20", "g_q80", "g_q90", "outlier")
if (!is.data.frame(stats_df) || !all(.required_summary_cols %in% names(stats_df))) {
  message("Recomputing stats from trait_df (summary columns missing from loaded CSV).")
  stats_df <- stats.f(trait_df)
  if (is.null(stats_df)) stop("stats.f returned NULL; check trait_df validity.")
}

```

## 1. Load Melted Trait Data
This section loads pre-processed trait data in a long (melted) format and adds common names for species.

```{r, stat_visualization, message=FALSE, warning=FALSE}

# ── Trait statistics ──────────────────────────────────────────────────────────

head(stats_df)

stats_df <- stats_df %>%
  dplyr::mutate(
    value = .data[[trait]],
    taxa = .data[[taxon_of_interest]],
    common_name = gsub("_", " ", species)
  )

if(isTRUE(has.n)){
  stats_df <- stats_df %>%   
    mutate(N = .data[[n_trait]])
  debug_log("contrast_plot.f using n_trait = %s, missing N = %d", n_trait, sum(is.na(stats_df$N)))
}

# Species in tree order, so the plot rows follow the phylogeny. Tip nodes are numbered
# 1..Ntip (ape convention, kept by tidytree), so `node <= Ntip` selects the tips.
# Filtering on `!is.na(label)` alone would also admit internal-node support labels, which
# repeat across nodes and give factor(levels =) duplicates.
ordered_species <-  tidytree::as_tibble(pruned_tree) %>%
  dplyr::arrange(desc(.data$node)) %>%
  dplyr::filter(.data$node <= length(pruned_tree$tip.label)) %>%
  dplyr::pull(.data$label)

stats_df$species <- factor(stats_df$species, levels = ordered_species)
debug_log("contrast_plot.f ordered_species = %d", length(ordered_species))

# Taxa ordered by the first species of each taxon in tree order.
ordered_taxa <- ordered_species %>%
  sapply(function(sp) {
    idx <- which(stats_df$species == sp)
    if (length(idx) == 0) return(NA_character_)
    stats_df$taxa[idx[1]]
  }) %>%
  unname() %>%
  na.omit() %>%
  unique()

stats_df <- stats_df %>%
  dplyr::mutate(taxa = factor(taxa, levels = ordered_taxa)) %>%
  dplyr::arrange(taxa, .by_group = TRUE)

```

# 3. Contrast Plots
This section generates contrast plots to visualize the trait across taxa, highlighting extreme values and outliers.

```{r phenoplot, fig.height=10, fig.width=15, message=FALSE, warning=FALSE, dev='png', dpi=320}

# ── Contrast plot ─────────────────────────────────────────────────────────────

# Trait values by taxon: one filled circle per species, the global median as a dashed line
# and the per-taxon median as a bar. The foreground/background partition is not drawn
# here (4.Independent_contrasts.Rmd selects it). The sizes come from species_plot_profile().
contrast_plot.f <- function(df, palette_values = NULL) {
  stopifnot(all(c("species", taxon_of_interest, trait) %in% names(df)))
  plot_profile <- species_plot_profile(
    n_species = dplyr::n_distinct(df$species),
    n_taxa = dplyr::n_distinct(df$taxa)
  )
  contrast_dims <- plot_profile$contrast
  if (is.null(palette_values)) {
    palette_values <- get_palette_values()
  }
  debug_log("contrast_plot.f rows = %d, trait = %s, taxon = %s", nrow(df), trait, taxon_of_interest)
  debug_log("contrast_plot.f columns: %s", paste(names(df), collapse = ", "))
  debug_log("color_palette = %s", color_palette)
  debug_log("palette_values is NULL: %s", is.null(palette_values))
  debug_log("palette_values length: %d", ifelse(is.null(palette_values), 0, length(palette_values)))
  if (!is.null(palette_values)) {
    debug_log("palette_values: %s", paste(palette_values, collapse = ", "))
  }
  
  # Without a clade palette the plot falls back to the default discrete scales.
  if (is.null(palette_values) || length(palette_values) == 0) {
    debug_log("Using discrete color scales (no palette available)")
    palette_values <- NULL  # normalizes an empty palette to NULL
  }
  fill_scale <- if (is.null(palette_values)) {
    ggplot2::scale_fill_discrete()
  } else {
    ggplot2::scale_fill_manual(values = palette_values)
  }
  color_scale <- if (is.null(palette_values)) {
    ggplot2::scale_color_discrete()
  } else {
    ggplot2::scale_color_manual(values = palette_values)
  }

  global_median <- df$g_median[1]

  # One uniform point layer per species.
  p <- ggplot(df, aes(x = value, y = taxa)) +
    ggplot2::scale_y_discrete(limits = rev(ordered_taxa)) +
    ggplot2::geom_point(
      aes(color = taxa), shape = 16, size = contrast_dims$point_size
    )

  # breaks = 0 matches no level, which hides the legends and keeps the colors.
  if (!is.null(palette_values) && length(palette_values) > 0) {
    p <- p +
      ggplot2::scale_fill_manual(values = palette_values, breaks = 0) +
      ggplot2::scale_color_manual(values = palette_values, breaks = 0)
  }

  # Reference marks: the global median and the per-taxon medians.
  p <- p +
    ggplot2::geom_vline(xintercept = global_median, linetype = "longdash", linewidth = 1.2, color = tableau_10["vermillion"]) +
    ggplot2::stat_summary(fun = median, geom = "errorbar",
                          aes(xmax = after_stat(x), xmin = after_stat(x), y = taxa),
                          linewidth = 1.2, color = tableau_10["blue"], alpha = 0.8) +
    ggplot2::labs(x = capitalized_trait,
                  y = paste(capitalized_taxon, "-", capitalized_clade),
                  title = paste("Trait distribution for trait:", capitalized_trait, "in", clade_name),
                  caption = "Global median: longdash red line. Per-taxa median: blue error bars.") +
    ggplot2::theme_minimal() +
    ggplot2::theme(
      plot.title = ggplot2::element_text(size = contrast_dims$title_size, hjust = 0, margin = ggplot2::margin(t = 20, b = 6), face = "bold"),
      plot.subtitle = ggplot2::element_text(size = contrast_dims$subtitle_size, hjust = 0, color = "grey40", margin = ggplot2::margin(b = 10)),
      axis.text.y = ggplot2::element_text(size = contrast_dims$axis_text_y),
      axis.title.x = ggplot2::element_text(size = contrast_dims$axis_title, margin = ggplot2::margin(t = 12)),
      axis.title.y = ggplot2::element_text(size = contrast_dims$axis_title, angle = 90, margin = ggplot2::margin(l = 20, r = 20)),
      plot.caption = ggplot2::element_text(size = contrast_dims$caption_size, hjust = 0, margin = ggplot2::margin(t = 20)),
      legend.text = ggplot2::element_text(size = 15),
      legend.position = "bottom"
    )
  attr(p, "plot_dims") <- contrast_dims[c("width", "height")]
  p
}

plot_ready <- !is.null(stats_df) && nrow(stats_df) > 0
required_cols <- c("species", taxon_of_interest, trait)
plot_ready <- plot_ready && all(required_cols %in% names(stats_df))

if (!plot_ready) {
  message("Skipping contrast plot: missing or empty trait stats.")
} else {
  # get_palette_values() first, then a direct lookup of the clade palette as a fallback.
  palette_for_plot <- get_palette_values()
  if (is.null(palette_for_plot) && exists("color_palette") && nzchar(color_palette)) {
    if (exists(color_palette, envir = .GlobalEnv)) {
      palette_for_plot <- get(color_palette, envir = .GlobalEnv)
      debug_log("Retrieved palette '%s' directly from GlobalEnv: length = %d", color_palette, length(palette_for_plot))
    }
  }
  debug_log("Calling contrast_plot.f with palette length: %d", ifelse(is.null(palette_for_plot), 0, length(palette_for_plot)))
  contrast_graph <- contrast_plot.f(stats_df, palette_for_plot)
  contrast_dims <- attr(contrast_graph, "plot_dims")
  knitr::opts_current$set(fig.width = contrast_dims$width, fig.height = contrast_dims$height)
  ggplot2::ggsave(
    file.path(extreme_plots_dir, paste0(trait, "_contrast_plot.png")),
    contrast_graph, device = "png", width = contrast_dims$width, height = contrast_dims$height, dpi = "retina"
  )
  contrast_graph
}

```

# 4. Violin Plots
This section generates violin plots to show trait distributions across taxa.

```{r violin_plot, fig.height=10, fig.width=8, message=FALSE, warning=FALSE, dev='png', dpi=320}

# ── Violin plot ───────────────────────────────────────────────────────────────

# Trait distribution by taxon: a violin per taxon, species as jittered points (outliers
# filled) and the global median as a vertical line.
violin_extremes.f <- function(df, trait, taxon_of_interest) {
  stopifnot(all(c("species", taxon_of_interest, trait) %in% names(df)))
  plot_profile <- species_plot_profile(
    n_species = dplyr::n_distinct(df$species),
    n_taxa = dplyr::n_distinct(df$taxa)
  )
  violin_dims <- plot_profile$violin
  palette_values <- get_palette_values()
  debug_log("violin_extremes.f rows = %d, trait = %s, taxon = %s", nrow(df), trait, taxon_of_interest)
  fill_scale <- if (is.null(palette_values)) {
    ggplot2::scale_fill_discrete()
  } else {
    ggplot2::scale_fill_manual(values = palette_values)
  }
  dark_palette_values <- get_palette_values(paste0("dark_", color_palette))
  fill_scale_dark <- if (is.null(dark_palette_values)) {
    fill_scale
  } else {
    ggplot2::scale_fill_manual(values = dark_palette_values)
  }
  color_scale_dark <- if (is.null(dark_palette_values)) {
    ggplot2::scale_color_discrete()
  } else {
    ggplot2::scale_color_manual(values = dark_palette_values)
  }

  # Species in tree order.
  ordered_species <- pruned_tree$tip.label
  df$species <- factor(df$species, levels = ordered_species)
  debug_log("violin_extremes.f ordered_species = %d", length(ordered_species))
  
  # Taxa in tree order: the global ordered_taxa when it exists, else rebuilt from the tree.
  if (!exists("ordered_taxa", inherits = TRUE) || length(ordered_taxa) == 0) {
    ordered_taxa <- unique(pruned_tree$tip.label %>%
      sapply(function(sp) df$taxa[df$species == sp]) %>%
      na.omit())
    if (length(ordered_taxa) == 0) {
      ordered_taxa <- unique(df$taxa[!is.na(df$taxa)])
    }
  }
  
  median_value <- median(df$value, na.rm = TRUE)
  
  plot <- ggplot2::ggplot(data = df, aes(x = value, y = taxa)) +
    ggplot2::scale_y_discrete(limits = rev(ordered_taxa)) +
    ggplot2::geom_violin(
      aes(fill = taxa), color = "black", width = 0.5, trim = TRUE, scale = "width", alpha = 0.15
    ) +
    fill_scale_dark +
    ggplot2::geom_point(
      data = subset(df, outlier == "normal"),
      aes(color = taxa), shape = 16,
      size = violin_dims$point_size, position = ggplot2::position_jitter(height = violin_dims$jitter_height)
    ) +
    ggplot2::geom_point(
      data = subset(df, outlier != "normal"),
      aes(color = taxa, fill = taxa),
      shape = 21, size = violin_dims$point_size, stroke = violin_dims$stroke,
      position = ggplot2::position_jitter(height = violin_dims$jitter_height)
    ) +
    color_scale_dark +
    ggplot2::geom_vline(
      aes(xintercept = median_value), linewidth = 1, color = tableau_10["vermillion"], alpha = 0.5
    ) +
    ggplot2::stat_summary(
      fun = median, geom = "errorbar",
      aes(xmax = after_stat(x), xmin = after_stat(x)),
      linewidth = 1.5, color = "black", alpha = 0.6, width = 0.75
    ) +
    ggplot2::labs(x = capitalized_trait) +
    ggplot2::theme_minimal() +
    ggplot2::theme(
      axis.text.y = ggplot2::element_text(size = violin_dims$axis_text_y, hjust = 0.5, face = "bold"),
      axis.title.x = ggplot2::element_text(size = violin_dims$axis_title, face = "bold", margin = ggplot2::margin(t = 12)),
      axis.title.y = ggplot2::element_blank(),
      legend.text = ggplot2::element_text(size = 15),
      legend.position = "none"
    )

  attr(plot, "plot_dims") <- violin_dims[c("width", "height")]
  plot
}

if (!plot_ready) {
  message("Skipping violin plot: missing or empty trait stats.")
} else {
  violin_plot_path <- file.path(extreme_plots_dir, paste0(trait, "_violin_plot.png"))
  violin_graph <- violin_extremes.f(stats_df, trait, taxon_of_interest)
  violin_dims <- attr(violin_graph, "plot_dims")
  knitr::opts_current$set(fig.width = violin_dims$width, fig.height = violin_dims$height)
  ggplot2::ggsave(
    violin_plot_path,
    violin_graph, device = "png", width = violin_dims$width, height = violin_dims$height, dpi = "retina"
  )
  violin_graph
}
  
```

# 5. Ancestral State Reconstruction (ASR) Plots
This section generates ASR plots to visualize the evolutionary history of the trait across the phylogenetic
```{r asr_plot, fig.height=20, fig.width=17, message=FALSE, warning=FALSE, dev='png', dpi=320}

# ── Ancestral state reconstruction ────────────────────────────────────────────

# Continuous ASR of the trait on the species tree. BM and OU are fitted with geiger and the
# OU model is kept when it improves AICc by at least 2 and alpha is above 1e-4. Node
# estimates and their 95% CI (ape::ace) are then used to flag tips that lie outside the CI
# of their immediate parent ("derived") or of all their ancestors ("cumulatively derived").
# Returns the node and tip tables, the model and a ggtree plot (element `plot`).
asr_tree.f <- function(df, tree, trait, taxon_of_interest) {
  
  # Trait vector, named by species, without missing or non-finite values.
  trait_vec <- df[[trait]]
  names(trait_vec) <- df$species
  trait_vec <- trait_vec[!is.na(trait_vec) & is.finite(trait_vec)]
  if(length(trait_vec) < 3) stop("Need >=3 non-missing species.")
  debug_log("asr_tree.f input rows = %d, trait_vec = %d", nrow(df), length(trait_vec))
  

  tree <- ape::drop.tip(tree, setdiff(tree$tip.label, names(trait_vec)))
  # ape::drop.tip() can leave a $root.edge on the pruned tree even when the source Newick
  # had none (when the root keeps a single child, its branch length is folded in as a root
  # edge). The root edge is removed before the fit, rescaling and ASR steps.
  tree$root.edge <- NULL
  tree <- ape::reorder.phylo(tree, "cladewise")
  Ntip <- length(tree$tip.label)
  Nnode <- tree$Nnode
  asr_dims <- species_plot_profile(n_species = Ntip, n_taxa = dplyr::n_distinct(df[[taxon_of_interest]]))$asr
  debug_log("asr_tree.f tree tips after prune = %d, nodes = %d", Ntip, Nnode)
  node_indices <- (Ntip+1):(Ntip+Nnode)

  # Trait vector and data frame in the tip order of the pruned tree.
  trait_vec <- trait_vec[tree$tip.label]

  df <- df[df$species %in% tree$tip.label, ]
  df$species <- factor(df$species, levels = tree$tip.label)

  debug_log("asr_tree.f trait_vec after prune = %d", length(trait_vec))
  debug_log("asr_tree.f df after prune = %d", nrow(df))
  
  # Fit BM and OU; sigma2 is the rate and alpha the OU pull (0 under BM).
  fit_model <- function(tree, trait, model = c("BM", "OU")) {
    model <- match.arg(model)
    fit <- if (model == "OU") {
      # alpha upper bound scaled to tree height, as in pss_core.R::ou_alpha_bounds
      suppressWarnings(geiger::fitContinuous(
        tree, trait, model = "OU", ncores = 1,
        bounds = list(alpha = c(exp(-500), 100 / max(ape::node.depth.edgelength(tree))))
      ))
    } else {
      suppressWarnings(geiger::fitContinuous(tree, trait, model = "BM", ncores = 1))
    }
    aicc <- fit$opt$aicc
    list(
      lnL = fit$opt$lnL,
      sigma2 = fit$opt$sigsq,
      alpha = if (model == "OU") fit$opt$alpha else 0,
      aicc = aicc,
      raw = fit
    )
  }
  
  fit_bm <- fit_model(tree, trait_vec, "BM")
  fit_ou <- fit_model(tree, trait_vec, "OU")
  debug_log("asr_tree.f fit_bm aicc = %.3f, fit_ou aicc = %.3f", fit_bm$aicc, fit_ou$aicc)
  
  delta_aicc <- fit_bm$aicc - fit_ou$aicc
  chosen_model <- if (delta_aicc >= 2.0 && !is.null(fit_ou$alpha) && fit_ou$alpha > 1e-4) "OU" else "BM"
  debug_log("asr_tree.f chosen_model = %s", chosen_model)
  
  # Under OU the tree is rescaled with the fitted alpha before the ASR.
  if (chosen_model == "OU") {
    alpha_hat <- fit_ou$alpha
    debug_log("asr_tree.f alpha_hat = %.4f", alpha_hat)
    sigma2_used <- fit_ou$sigma2
    rescaled_tree <- phytools::rescale(tree, "OU", alpha = alpha_hat)
  } else {
    sigma2_used <- fit_bm$sigma2
    rescaled_tree <- tree
  }
  debug_log("asr_tree.f sigma2_used = %.6f", sigma2_used)
  
  # Node indices of the rescaled tree.
  Ntip_rescaled <- length(rescaled_tree$tip.label)
  Nnode_rescaled <- rescaled_tree$Nnode
  node_indices <- (Ntip_rescaled+1):(Ntip_rescaled+Nnode_rescaled)

  # ACE (REML, then ML) on the rescaled tree; when both fail, ACE on the original tree
  # (the model becomes BM), and finally fastAnc without CIs.
  ace_try <- function(tree_obj) {
    tryCatch(
      ape::ace(x = trait_vec, phy = tree_obj, type = "continuous", method = "REML"),
      error = function(e) tryCatch(
        ape::ace(x = trait_vec, phy = tree_obj, type = "continuous", method = "ML"),
        error = function(e2) NULL
      )
    )
  }
  ace_res <- ace_try(rescaled_tree)
  model_used <- chosen_model
  if (is.null(ace_res)) {
    message("ACE failed on rescaled tree; retrying on original tree.")
    ace_res <- ace_try(tree)
    if (is.null(ace_res)) {
      message("ACE failed on original tree; falling back to fastAnc without CIs.")
      anc_est <- tryCatch(
        phytools::fastAnc(rescaled_tree, trait_vec),
        error = function(e) NULL
      )
      if (is.null(anc_est)) {
        debug_log("asr_tree.f fastAnc failed")
      } else {
        debug_log("asr_tree.f fastAnc ok, n = %d", length(anc_est))
      }
      if (is.null(anc_est)) {
        stop("ACE and fastAnc failed.")
      }
      ace_res <- list(
        ace = anc_est,
        CI95 = cbind(anc_est, anc_est)
      )
    } else {
      rescaled_tree <- tree
      Ntip_rescaled <- length(rescaled_tree$tip.label)
      Nnode_rescaled <- rescaled_tree$Nnode
      node_indices <- (Ntip_rescaled+1):(Ntip_rescaled+Nnode_rescaled)
      model_used <- "BM"
    }
  }
  
  node_est <- ace_res$ace
  tip_est <- trait_vec[rescaled_tree$tip.label]
  debug_log("asr_tree.f node_est = %d, tip_est = %d", length(node_est), length(tip_est))
  
  # Node and tip tables. Node CIs are the analytical 95% CIs of ace().
  node_ci <- ace_res$CI95
  node_df <- data.frame(
    node = node_indices,
    estimate = as.numeric(node_est),
    ci_lower = as.numeric(node_ci[,1]),
    ci_upper = as.numeric(node_ci[,2]),
    stringsAsFactors = FALSE
  )
  # Tips carry no CI in the ace output: the observed value is used as both bounds.
  tip_df <- data.frame(
    tip = rescaled_tree$tip.label,
    estimate = as.numeric(tip_est[rescaled_tree$tip.label]),
    ci_lower = as.numeric(tip_est[rescaled_tree$tip.label]),
    ci_upper = as.numeric(tip_est[rescaled_tree$tip.label]),
    immediate_parent = NA,
    cumulative_derived = NA,
    direction = NA,
    parent_node_id = NA,
    parent_ci_lower = NA,
    parent_ci_upper = NA,
    stringsAsFactors = FALSE
  )
  
  # Derivedness tests: against the immediate parent and against all ancestors.
  # parent_map[i] is the parent node of node i (NA for the root).
  parent_map <- integer(Ntip_rescaled + Nnode_rescaled)
  parent_map[] <- NA_integer_
  for (j in seq_len(nrow(rescaled_tree$edge))) {
    parent_map[rescaled_tree$edge[j,2]] <- rescaled_tree$edge[j,1]
  }
  
  # A tip is derived when its value lies outside the CI of its parent node
  # (direction "up" or "down").
  for (i in seq_len(nrow(tip_df))) {
    tip_label <- tip_df$tip[i]
    tip_est_val <- tip_df$estimate[i]
    tip_idx <- which(rescaled_tree$tip.label == tip_label)
    parent_id <- parent_map[tip_idx]
    
    if (!is.na(parent_id)) {
      tip_df$parent_node_id[i] <- parent_id
      
      if (parent_id <= Ntip_rescaled) {
        parent_row <- tip_df[tip_df$tip == rescaled_tree$tip.label[parent_id], ]
      } else {
        parent_row <- node_df[node_df$node == parent_id, ]
      }
      parent_ci_low <- parent_row$ci_lower
      parent_ci_high <- parent_row$ci_upper
      
      tip_df$immediate_parent[i] <- (tip_est_val > parent_ci_high) || (tip_est_val < parent_ci_low)
      
      if (tip_est_val > parent_ci_high) {
        tip_df$direction[i] <- "up"
      } else if (tip_est_val < parent_ci_low) {
        tip_df$direction[i] <- "down"
      } else {
        tip_df$direction[i] <- "none"
      }
      
      if (tip_df$immediate_parent[i]) {
        tip_df$parent_ci_lower[i] <- parent_ci_low
        tip_df$parent_ci_upper[i] <- parent_ci_high
      }
    }
    
    # Cumulatively derived: outside the CI of every ancestor node.
    anc_list <- integer(0)
    cur <- tip_idx
    while (!is.na(parent_map[cur])) {
      cur <- parent_map[cur]
      anc_list <- c(anc_list, cur)
    }
    anc_nodes <- anc_list[anc_list > Ntip_rescaled]
    
    if (length(anc_nodes) == 0) {
      tip_df$cumulative_derived[i] <- FALSE
    } else {
      anc_rows <- node_df[node_df$node %in% anc_nodes, ]
      if (nrow(anc_rows) == 0) {
        tip_df$cumulative_derived[i] <- NA
      } else {
        tip_df$cumulative_derived[i] <- all((tip_est_val > anc_rows$ci_upper) | (tip_est_val < anc_rows$ci_lower))
      }
    }
  }
  
  # Plot (ggtree). Branches are colored by the node estimate; tips carry the species
  # name, with the estimate in brackets when the tip is derived from its parent. Text and
  # symbol sizes shrink slowly with the number of tips, while the canvas size comes from
  # species_plot_profile()$asr.
  all_vals <- rbind(
    data.frame(node = seq_len(Ntip_rescaled), estimate = as.numeric(tip_est[rescaled_tree$tip.label])),
    data.frame(node = node_df$node, estimate = node_df$estimate)
  )
  tip_meta <- tip_df
  tip_meta$node <- seq_len(Ntip_rescaled)
  tip_meta$is_up <- !is.na(tip_meta$direction) & tip_meta$direction == "up"
  tip_meta$is_down <- !is.na(tip_meta$direction) & tip_meta$direction == "down"
  tip_meta$species_annotated <- ifelse(
    tip_meta$immediate_parent %in% TRUE,
    sprintf("plain(%s)~bold(\"[%.2f]\")", vapply(tip_meta$tip, deparse, character(1)), tip_meta$estimate),
    sprintf("plain(%s)", vapply(tip_meta$tip, deparse, character(1)))
  )

  # Parent nodes of derived tips get a label with the estimate and its CI. The labels are
  # joined into plot_df because geom_nodelab() needs the tree-attached data (x, y and isTip
  # from the %<+% join); a standalone data frame without those columns fails in its
  # internal `!isTip` check.
  parent_nodes_with_derived_tips <- unique(tip_df$parent_node_id[tip_df$immediate_parent %in% TRUE & !is.na(tip_df$parent_node_id)])
  node_df$node_label <- ifelse(
    node_df$node %in% parent_nodes_with_derived_tips,
    sprintf("%.2f (%.2f-%.2f)", node_df$estimate, node_df$ci_lower, node_df$ci_upper),
    NA_character_
  )

  plot_df <- dplyr::left_join(
    all_vals,
    tip_meta[, c("node", "is_up", "is_down", "cumulative_derived", "species_annotated")],
    by = "node"
  )
  plot_df$is_up[is.na(plot_df$is_up)] <- FALSE
  plot_df$is_down[is.na(plot_df$is_down)] <- FALSE
  plot_df$cumulative_derived[is.na(plot_df$cumulative_derived)] <- FALSE
  plot_df <- dplyr::left_join(plot_df, node_df[, c("node", "node_label")], by = "node")

  root_est_str <- sprintf("Root: %.2f (%.2f-%.2f)", node_df$estimate[1], node_df$ci_lower[1], node_df$ci_upper[1])

  # Text and symbol sizes decrease linearly with the number of tips, down to a floor,
  # so the labels do not overlap vertically in large trees.
  tip_lab_size   <- max(2.2, 4.5 - 0.008 * Ntip_rescaled)
  symbol_size    <- max(2.5, 5.0 - 0.008 * Ntip_rescaled)
  node_lab_size  <- max(2.0, 4.0 - 0.008 * Ntip_rescaled)

  vermillion_col <- as.character(tableau_10["vermillion"])
  # "green" rather than "bluish_green": the teal of the latter is hard to see against the
  # white background in the down-triangle.
  green_col <- as.character(tableau_10["green"])
  blue_col <- as.character(tableau_10["blue"])

  gg_base <- ggtree::ggtree(rescaled_tree, layout = "rectangular", color = "black", linewidth = 1.2) %<+% plot_df
  tree_depth <- max(gg_base$data$x, na.rm = TRUE)
  # The tip rows of the joined data position the symbol layers at explicit (x, y), because
  # the tiplab offset and position_nudge only move along the branch axis.
  tip_data <- gg_base$data[gg_base$data$isTip, ]

  p <- gg_base +
    ggtree::geom_tree(ggplot2::aes(color = estimate), linewidth = 0.7) +
    ggplot2::scale_color_gradient(
      low = "white", high = vermillion_col,
      name = paste0(trait, " (", model_used, ")"),
      guide = ggplot2::guide_colorbar(
        barwidth = ggplot2::unit(20, "lines"),
        barheight = ggplot2::unit(1.2, "lines"),
        title.position = "top",
        title.hjust = 0.5
      )
    ) +
    ggtree::geom_tiplab(ggplot2::aes(label = species_annotated), parse = TRUE, size = tip_lab_size, offset = 0.08 * tree_depth, color = "black") +
    ggtree::theme_tree2() +
    # hexpand(direction = 1) pads the x-range on the label side only, which frees room on
    # the right for tip labels, values and symbols that would be clipped at the panel edge.
    ggtree::hexpand(0.28, direction = 1) +
    ggplot2::theme(
      legend.position = "none",  # the color bar is added below the caption with cowplot
      plot.margin = ggplot2::margin(10, 60, 10, 10),
      plot.subtitle = ggplot2::element_text(size = 16, color = "black", face = "bold"),
      plot.caption = ggplot2::element_text(size = 13, color = "black", hjust = 0.35),
      axis.text = ggplot2::element_text(size = 12, color = "black")
    ) +
    ggplot2::labs(
      subtitle = root_est_str,
      caption = paste0("Symbols: ▲ Tip Up / ▼ Tip Down (derived vs. immediate parent node), ",
                        "■ Cumulatively derived (derived vs. ALL ancestor nodes); ",
                        "parent node values shown for derived tips.")
    )

  # Derivation symbols sit on the tip row, to the right of the tip: the up/down triangle at
  # the first offset and the cumulative-derived square at the second offset when a triangle
  # is also present, at the first offset otherwise.
  sym_offset1 <- 0.025 * tree_depth
  sym_offset2 <- 0.052 * tree_depth
  has_tri <- tip_data$is_up | tip_data$is_down
  tip_data$sq_x <- ifelse(has_tri, tip_data$x + sym_offset2, tip_data$x + sym_offset1)

  if (any(tip_data$is_up)) {
    p <- p + ggplot2::geom_point(data = subset(tip_data, is_up),
                                  ggplot2::aes(x = x + sym_offset1, y = y),
                                  shape = 24, fill = vermillion_col, color = "black", size = symbol_size)
  }
  if (any(tip_data$is_down)) {
    p <- p + ggplot2::geom_point(data = subset(tip_data, is_down),
                                  ggplot2::aes(x = x + sym_offset1, y = y),
                                  shape = 25, fill = green_col, color = "black", size = symbol_size)
  }
  if (any(tip_data$cumulative_derived)) {
    p <- p + ggplot2::geom_point(data = subset(tip_data, cumulative_derived),
                                  ggplot2::aes(x = sq_x, y = y),
                                  shape = 22, fill = blue_col, color = "black", size = symbol_size)
  }
  if (any(!is.na(plot_df$node_label))) {
    p <- p + ggtree::geom_nodelab(ggplot2::aes(subset = !is.na(node_label), label = node_label),
                                   size = node_lab_size, fontface = "bold", hjust = -0.1, color = "black")
  }

  # The color bar is extracted from the plot and stacked below the whole panel, caption
  # included; ggplot2's own bottom legend would sit between the panel and the caption.
  legend_grob <- cowplot::get_legend(
    p + ggplot2::theme(
      legend.position = "bottom",
      legend.title = ggplot2::element_text(size = 14, color = "black", margin = ggplot2::margin(b = 6)),
      legend.text = ggplot2::element_text(size = 12, color = "black")
    )
  )
  p <- cowplot::plot_grid(p, legend_grob, ncol = 1, rel_heights = c(1, 0.06))

  list(
    chosen_model=chosen_model,
    node_table=node_df,
    tip_table=tip_df,
    sigma2_used=sigma2_used,
    ace=ace_res,
    rescaled_tree=rescaled_tree,
    plot_dims=asr_dims[c("width", "height")],
    plot=p
  )
}

if (!plot_ready) {
  plot.new()
  text(0.5, 0.5, "ASR plot skipped: missing or empty trait stats.", cex = 1.2)
} else {
  asr_plot_path <- file.path(asr_trees, paste0(trait, "_asr_tree.png"))
  asr_dims <- species_plot_profile(
    n_species = sum(!is.na(stats_df[[trait]]) & is.finite(stats_df[[trait]])),
    n_taxa = dplyr::n_distinct(stats_df[[taxon_of_interest]])
  )$asr
  knitr::opts_current$set(fig.width = asr_dims$width, fig.height = asr_dims$height)
  # One call renders and saves the plot, so the fit runs once. A failure is drawn as a
  # placeholder in the report, because message() is suppressed by message=FALSE and the
  # section would otherwise be empty with no visible reason.
  asr_result <- tryCatch(
    asr_tree.f(stats_df, pruned_tree, trait, taxon_of_interest),
    error = function(e) {
      plot.new()
      text(0.5, 0.5, paste("ASR plot failed:", conditionMessage(e)), cex = 1.1, col = "firebrick")
      NULL
    }
  )
  # asr_tree.f() returns a ggplot object, which is printed explicitly for the HTML output.
  # The PNG is written in both cases: the ggplot on success, a copy of the placeholder
  # device on failure, so the ASR_trees folder is never empty.
  if (!is.null(asr_result) && !is.null(asr_result$plot)) {
    print(asr_result$plot)
    tryCatch(
      ggplot2::ggsave(asr_plot_path, plot = asr_result$plot,
                       width = asr_dims$width, height = asr_dims$height,
                       dpi = 320, bg = "white", limitsize = FALSE),
      error = function(e) message("ASR plot file save failed: ", conditionMessage(e))
    )
  } else {
    tryCatch({
      grDevices::dev.copy(grDevices::png, filename = asr_plot_path,
                           width = asr_dims$width, height = asr_dims$height,
                           res = 320, bg = "white", units = "in")
      grDevices::dev.off()
    }, error = function(e) message("ASR plot file save failed: ", conditionMessage(e)))
  }
}

```

## 6. Fan Plot for Phylogenetic Distribution

This section creates a fan-shaped phylogenetic tree.

```{r visual_distribution, fig.height=15, fig.width=15, dev="png", dpi=320, message=FALSE, warning=FALSE}

# ── Fan tree and trait rings ──────────────────────────────────────────────────

palette_values <- get_palette_values()
# Fallback: the clade palette from the global environment.
if (is.null(palette_values) && exists("color_palette") && nzchar(color_palette)) {
  if (exists(color_palette, envir = .GlobalEnv)) {
    palette_values <- get(color_palette, envir = .GlobalEnv)
    debug_log("Retrieved palette '%s' directly from GlobalEnv for visual_distribution: length = %d", color_palette, length(palette_values))
  }
}
fill_scale <- if (is.null(palette_values)) {
  ggplot2::scale_fill_discrete()
} else {
  ggplot2::scale_fill_manual(values = palette_values)
}
color_scale <- if (is.null(palette_values)) {
  ggplot2::scale_color_discrete()
} else {
  ggplot2::scale_color_manual(values = palette_values)
}

# Fan layout, saved as a bare tree first.
tree_profile_base <- species_plot_profile(
  n_species = length(pruned_tree$tip.label),
  n_taxa = dplyr::n_distinct(stats_df[[taxon_of_interest]]),
  n_rings = 0
)$tree
knitr::opts_current$set(fig.width = tree_profile_base$diagnostic_size, fig.height = tree_profile_base$diagnostic_size)
base_fan_tree <- ggtree(pruned_tree,
                        layout="fan",
                        open.angle=15, 
                        size=tree_profile_base$tree_line_width)

ggsave(file.path(phylo_distribution_dir, "big_tree.png"), base_fan_tree, device = "png", width = tree_profile_base$diagnostic_size, height = tree_profile_base$diagnostic_size, dpi = "retina")

ordered_species <- rev(pruned_tree$tip.label)

# The fan plots keep the species of the pruned tree that have a finite primary trait value
# (when at least 3 remain).
tree_plot_trait_df <- stats_df %>%
  dplyr::filter(species %in% pruned_tree$tip.label)

valid_primary_species <- tree_plot_trait_df %>%
  dplyr::filter(!is.na(value) & is.finite(value)) %>%
  dplyr::pull(species) %>%
  unique()

if (length(valid_primary_species) >= 3) {
  circular_tree <- ape::drop.tip(pruned_tree, setdiff(pruned_tree$tip.label, valid_primary_species))
  tree_plot_trait_df <- tree_plot_trait_df %>%
    dplyr::filter(species %in% circular_tree$tip.label)
  debug_log("visual_distribution: circular_tree tips after trait pruning = %d", length(circular_tree$tip.label))
} else {
  circular_tree <- pruned_tree
  tree_plot_trait_df <- tree_plot_trait_df %>%
    dplyr::filter(species %in% circular_tree$tip.label)
  debug_log("visual_distribution: insufficient non-missing trait species (%d); using pruned_tree", length(valid_primary_species))
}

n_tips <- length(circular_tree$tip.label)
n_taxa_tree <- dplyr::n_distinct(tree_plot_trait_df$taxa)
n_fruit_rings <- if (isTRUE(has.secondary)) 2L else 1L
tree_profile <- species_plot_profile(n_species = n_tips, n_taxa = n_taxa_tree, n_rings = n_fruit_rings)$tree
trait_axis_nbreak <- as.integer(max(1, round(tree_profile$axis_nbreak)))
trait_axis_text_size <- as.numeric(tree_profile$axis_text_size)
secondary_axis_text_size <- as.numeric(tree_profile$axis_text_size * 0.8)
knitr::opts_current$set(fig.width = tree_profile$width, fig.height = tree_profile$height)

# One row per species and taxon with the summed trait (and N or the secondary trait when present).
tree_trait_data <- tree_plot_trait_df %>%
  group_by(species, taxa)

# Branch color trait: its ancestral states come from branch_trait_node_values()
# (plotting_fun.R), which tolerates duplicated species rows and partial coverage. When it
# returns NULL the branches are drawn plain.
if (isTRUE(has.branch)) {
  combined_br_data <- branch_trait_node_values(
    circular_tree,
    tree_trait_data$species,
    tree_trait_data[[branch_trait]]
  )
  if (is.null(combined_br_data)) {
    debug_log("visual_distribution: branch trait '%s' unusable for ASR; drawing plain branches", branch_trait)
    has.branch <- FALSE
  }
}
  
if (isTRUE(has.n) && isTRUE(has.secondary)) { # N and secondary trait
  tree_trait_data <- tree_trait_data %>%
    dplyr::rename(secondary_value = .data[[secondary_trait]]) %>%
    summarize(
      trait_sum = sum(value, na.rm = TRUE),
      n_sum = sum(N, na.rm = TRUE),
      secondary_sum = sum(secondary_value, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    ungroup()
} else if (isTRUE(has.n)) {
  tree_trait_data <- tree_trait_data %>%
    summarize(
      trait_sum = sum(value, na.rm = TRUE),
      n_sum = sum(N, na.rm = TRUE),
      .groups = "drop") %>%
    ungroup()
} else if (isTRUE(has.secondary)) { # secondary trait only
  tree_trait_data <- tree_trait_data %>%
    dplyr::rename(secondary_value = .data[[secondary_trait]]) %>%
    summarize(
      trait_sum = sum(value, na.rm = TRUE),
      secondary_sum = sum(secondary_value, na.rm = TRUE),
      .groups = "drop") %>%
    ungroup()
} else { # neither N nor secondary trait
  tree_trait_data <- tree_trait_data %>%
    summarize(
      trait_sum = sum(value, na.rm = TRUE),
      .groups = "drop") %>%
    ungroup()
}

primary_ring_limits <- compute_ring_axis_limits(tree_trait_data$trait_sum, min_limit = 1)
secondary_ring_limits <- if (isTRUE(has.secondary)) compute_ring_axis_limits(tree_trait_data$secondary_sum, min_limit = 1) else c(0, 1)

# Tree as a tibble joined with the trait data.
tree_tibble <- tidytree::as_tibble(circular_tree) %>%
  dplyr::rename(species = label)

tree_with_traits <- full_join(tree_tibble, tree_trait_data, by = 'species')

# Node values of the branch trait.
if (isTRUE(has.branch)) {
  tree_with_traits <- full_join(tree_with_traits, combined_br_data, by = 'node')
}

# Debug export of the joined table.
write.csv(tree_with_traits, file.path(phylo_distribution_dir, "tree_with_traits.csv"), row.names = FALSE)

# treedata object for ggtree.
tree_data_object <- tidytree::as.treedata(tree_with_traits)

# Diagnostic tree with node numbers, colored by taxon.
diagnostic_tree <- ggtree(tree_data_object,
                          layout = "fan",
                          open.angle = 15,
                          size = tree_profile$tree_line_width) +
  geom_text2(aes(label = node), hjust = -0.5, vjust = -0.5, size = tree_profile$node_text_size) +
  aes(color = taxa) +
  color_scale

ggsave(file.path(phylo_distribution_dir, "diagnostic_tree.png"), diagnostic_tree, device = "png", width = tree_profile$diagnostic_size, height = tree_profile$diagnostic_size, dpi = "retina")

# If branch color trait is present, add it to the tree plot

# Base tree, with the branch trait (column BR) as a color gradient when present.
if (isTRUE(has.branch)) {
  capitalized_branch_trait <- tools::toTitleCase(gsub("_", " ", branch_trait))
  tree_plot_step1 <- ggtree(tree_data_object, aes(color = BR),
                                  layout="fan",
                                  open.angle=15, 
                                  size=tree_profile$tree_line_width) +
    scale_color_gradient(
      name = capitalized_branch_trait,
      low = "#F4EBD0", high = "#2B2D42",
      # Branches that subtend too few covered species have no reconstruction and are grey.
      na.value = "grey70",
                        guide = guide_colorbar(order = 3)) +  # legend order: branch trait last
    new_scale_color()
} else {
  tree_plot_step1 <- ggtree(tree_data_object,
                                  layout="fan",
                                  open.angle=15, 
                                  size=tree_profile$tree_line_width)
}

tree_plot_step1 <- tree_plot_step1 +
  # Ring of trait bars
  geom_fruit(
    geom = geom_col,
    mapping = aes(y = species, x = trait_sum, fill = trait_sum, width = 0.5),
    alpha = 0.5,
    show.legend = TRUE,
    pwidth = tree_profile$fruit_pwidth,
    color = "black",
    size = 0.7,
    axis.params = list(
      axis = 'x',
      text.size = trait_axis_text_size,
      nbreak = trait_axis_nbreak,
      text.angle = 270,
      vjust = 0.5,
      hjust = 0,
      limits = primary_ring_limits
    ),
    grid.params = list()
  ) +
  scale_fill_gradient2(
    name = capitalized_trait, 
    low = "white", 
    high = as.character(tableau_10["blue"]),
    guide = guide_colorbar(order = 1),
    aesthetics = "fill"
  ) +
  new_scale_fill()

# Second ring for the secondary trait.
if (isTRUE(has.secondary)) {
  capitalized_secondary_trait <- tools::toTitleCase(gsub("_", " ", secondary_trait))

  tree_plot_step1 <- tree_plot_step1 +
    geom_fruit(
      geom = geom_col,
      offset = -tree_profile$fruit_pwidth,
      mapping = aes(y = species, x = secondary_sum, fill = secondary_sum, width = 0.5),
      alpha = 0.5,
      show.legend = TRUE,
      pwidth = tree_profile$fruit_pwidth,
      color = "black",
      size = 0.7,
      axis.params = list(
        axis = 'x',
        text.size = 0,
        nbreak = 0,
        text.angle = 270,
        vjust = 0.5,
        hjust = 0,
        limits = secondary_ring_limits
      ),
      grid.params = list()
    ) +
    scale_fill_gradient2(
      name = capitalized_secondary_trait, 
      low = "white", 
      high = as.character(tableau_10["vermillion"]),  # Different gradient color for distinction
      guide = guide_colorbar(order = 2),
      aesthetics = "fill"
    ) +
    new_scale_fill()
}

# geom_fruit sizes the trait rings to `fruit_pwidth` and scales the bars against
# `axis.params$limits`, but a bar can still extend past that band (an outlier value, the
# axis tick labels), depending on where the data sit relative to the limits. Instead of a
# fixed margin, the end of the ring stack is measured with fan_label_offsets() (as for the
# label ring), and only the part not covered by the nominal offset of the taxa ring is added.
trait_ring_measure <- fan_label_offsets(tree_plot_step1, labels = character(), ring_gap = 0, image_gap = 0)
trait_ring_span_fraction <- if (trait_ring_measure$tip_radius > 0) {
  (trait_ring_measure$outer_radius - trait_ring_measure$tip_radius) / trait_ring_measure$tip_radius
} else {
  tree_profile$fruit_pwidth
}
taxa_bar_offset_effective <- tree_profile$taxa_bar_offset +
  max(0, trait_ring_span_fraction - tree_profile$fruit_pwidth)

# Ring that marks the taxon of each species.
tree_plot_step2 <- tree_plot_step1 +
  geom_fruit(geom = geom_col,
             mapping = aes(y = species, x = 1, fill = taxa),
             pwidth = tree_profile$taxa_bar_pwidth, color = "black", linewidth = 0.5, offset = taxa_bar_offset_effective)

if (!is.null(palette_values) && length(palette_values) > 0) {
  tree_plot_step2 <- tree_plot_step2 + scale_fill_manual(values = palette_values, breaks = 0)
}

tree_plot_step2 <- tree_plot_step2 +
  new_scale_fill() +
  theme_tree() +
  theme(panel.background = element_rect(fill = "transparent", colour = NA),  
        plot.background = element_rect(fill = "transparent", colour = NA),
        legend.position = "none",
        plot.title = element_text(size = 17, margin = margin(t = 0, r = 0, b = -1.25, l = 0, unit = "cm")),
        plot.subtitle = element_text(size = 15, margin = margin(t = 0, r = 0, b = -1.5, l = 0, unit = "cm"))
  )

# Sample size (N) labels, when available.
if(isTRUE(has.n)){
  tree_plot_step3 <- tree_plot_step2 +
    geom_text(aes(label = n_sum), nudge_x = tree_profile$n_label_nudge, fontface = "bold", size = tree_profile$n_label_size, vjust = 0.5)
} else {
  tree_plot_step3 <- tree_plot_step2
}

dyn_height <- tree_profile$height
dyn_width <- tree_profile$width

```
```{r visual_distribution_final, fig.height=dyn_height, fig.width=dyn_width, dev="png", dpi=320, message=FALSE, warning=FALSE}

# ── Taxon labels and silhouettes ──────────────────────────────────────────────

# Most recent common ancestor node of each taxon.
tip_df <- find_taxon_mrca(circular_tree, tree_with_traits)

# One phylopic silhouette (rphylopic UUID) per taxon name.
tip_df <- tip_df %>%
  rowwise() %>%
  mutate(phylopic_UUID = rphylopic::get_uuid(name = taxa, n = 1))

# Debug export.
write.csv(tip_df, file.path(phylo_distribution_dir, "taxon_node_mapping.csv"), row.names = FALSE)

# The label rings are placed relative to where the fruit rings end (see plotting_fun.R).
label_offsets <- fan_label_offsets(
  tree_plot_step3,
  labels = tip_df$taxa,
  label_fontsize = tree_profile$branch_label_size,
  canvas_in = dyn_height,
  ring_gap = tree_profile$label_ring_gap,
  image_gap = tree_profile$label_image_gap,
  image_size = tree_profile$image_size
)
cladelab_offset_text <- label_offsets$text
cladelab_offset_phylopic <- label_offsets$phylopic

# Silhouettes at the MRCA of each taxon, then the taxon names.
final_tree_plot <- tree_plot_step3 +
  geom_cladelab(data = tip_df,
                mapping = aes(
                  node = mrca_node,
                  label = taxa,
                  image = phylopic_UUID,
                  color = taxa),
                geom="phylopic",
                barsize = NA,
                offset = cladelab_offset_phylopic,
                imagesize = tree_profile$image_size,
                alpha = 0.75)

if (!is.null(palette_values) && length(palette_values) > 0) {
  final_tree_plot <- final_tree_plot + scale_color_manual(values = palette_values, breaks = 0)
}

final_tree_plot <- final_tree_plot +
  geom_cladelab(data = tip_df,
                mapping = aes(
                  node = mrca_node,
                  label = taxa),
                show.legend = FALSE,
                          color = "black",
                angle = "auto",
                horizontal = TRUE,
                offset = cladelab_offset_text,
                barsize = NA,
                fontsize = tree_profile$branch_label_size,
                fontface = "bold") +
  theme_tree() +
  theme(panel.background = element_rect(fill = "transparent", colour = NA),  
        plot.background = element_rect(fill = "transparent", colour = NA),
        legend.position = "right",
        # Key size and spacing are absolute lengths, so they scale with the canvas
        # together with the legend text.
        legend.spacing.y = unit(tree_profile$legend_spacing_cm, 'cm'),
        legend.title = element_text(
          size = tree_profile$legend_title_size, face = "bold",
          margin = margin(b = 15)),
        legend.key.size = unit(tree_profile$legend_key_cm, 'cm'),
        legend.text = element_text(size = tree_profile$legend_text_size),
        plot.title = element_text(size = 17, margin = margin(t = 0, r = 0, b = -1.25, l = 0, unit = "cm")),
        plot.subtitle = element_text(size = 15, margin = margin(t = 0, r = 0, b = -1.5, l = 0, unit = "cm"))
  )

final_tree_plot

# One PNG per construction step of the tree.
ggsave(file.path(phylo_distribution_dir, "tree_with_trait_bars.png"), tree_plot_step1, device = "png", width = dyn_width, height = dyn_height, dpi = "retina")
ggsave(file.path(phylo_distribution_dir, "tree_with_taxa_bars.png"), tree_plot_step2, device = "png", width = dyn_width, height = dyn_height, dpi = "retina")
if(isTRUE(has.n)){
  ggsave(file.path(phylo_distribution_dir, "tree_with_n_labels.png"), tree_plot_step3, device = "png", width = dyn_width, height = dyn_height, dpi = "retina")
}
ggsave(file.path(phylo_distribution_dir, "final_tree_plot.png"), final_tree_plot, device = "png", width = dyn_width, height = dyn_height, dpi = "retina")

```

---

## Summary

This report summarizes the results of the **Trait Analysis** analysis.

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — Trait Analysis Module*
*Date of generation: `r Sys.Date()`*
