---
title: "Trait Composition Analysis"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
params:
  trait_file: NULL
  tree_file: NULL
  output_dir: !r getwd()
  seed: ""
  clade_name: "clade"
  taxon_of_interest: "family"
  sp_colname: "species"
  traitname: "trait"
  n_trait: ""
  c_trait: ""
  secondary_trait: ""
  branch_trait: ""
  pss_top_pct: "0.05"
  perm_strategy: "best_model"
  trait_type: ""
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

<!-- 3.CI-composition.Rmd — Jeffreys intervals, pair scores and the candidate foreground/background pool. -->
<!-- PhyloPhere | subworkflows/TRAIT_ANALYSIS/local/ -->

```{r header}

#  ██████╗ ██╗  ██╗██╗   ██╗██╗      ██████╗ ██████╗ ██╗  ██╗███████╗██████╗ ███████╗
#  ██╔══██╗██║  ██║╚██╗ ██╔╝██║     ██╔═══██╗██╔══██╗██║  ██║██╔════╝██╔══██╗██╔════╝
#  ██████╔╝███████║ ╚████╔╝ ██║     ██║   ██║██████╔╝███████║█████╗  ██████╔╝█████╗
#  ██╔═══╝ ██╔══██║  ╚██╔╝  ██║     ██║   ██║██╔═══╝ ██╔══██║██╔══╝  ██╔══██╗██╔══╝
#  ██║     ██║  ██║   ██║   ███████╗╚██████╔╝██║     ██║  ██║███████╗██║  ██║███████╗
#  ╚═╝     ╚═╝  ╚═╝   ╚═╝   ╚══════╝ ╚═════╝ ╚═╝     ╚═╝  ╚═╝╚══════╝╚═╝  ╚═╝╚══════╝

## PHYLOPHERE: A Nextflow pipeline including a complete set
## of phylogenetic comparative tools and analyses for Phenome-Genome studies

### Github: https://github.com/nozerorma/caastools/nf-phylophere
### Author: Miguel Ramon (miguel.ramon@upf.edu)

#### File: 3.CI-composition.Rmd

```

# Trait Composition Analysis
This script computes contrasting species pairs for the trait of interest. The rule that makes a pair eligible depends on the trait type:

- **Count trait** (`n_trait` and `c_trait` columns exist): computes 95% Jeffreys credible intervals with a Beta(0.5, 0.5) prior; a pair is eligible when its intervals do not overlap.
- **Ordinal trait** (coded levels): the highest level is *top*, the lowest is *bottom* and the levels in between are *intermediate*; a pair is eligible when one species is *top* and the other *bottom*, so intermediate species form no pair.
- **Continuous trait**: the median splits *top* from *bottom*; a pair is eligible when it falls in the top `pss_top_pct` of the pairs by Phylogenetic Shift Score (PSS).

---

## Setup and Directory Configuration

This section configures the working environment, sets directories, and loads necessary functions and libraries.

```{r setup, message=FALSE, warning=FALSE}
if (exists("params")) {
  tryCatch(assign("params", params, envir = .GlobalEnv), error = function(e) NULL)
}

# Call the setup function from commons.R
source("./src/commons.R")
setup_rmd()

# PSS engine (Fabio Barteri's phyloq, vendored); ct_ci.nf copies it into src/.
source("./src/pss_core.R")

# commons.R sets pss_top_pct and perm_strategy from params.
debug_log("pss_top_pct = %.4f, perm_strategy = %s", pss_top_pct, perm_strategy)

# Fallback logger, used only when commons.R has not defined debug_log().
if (is.null(getOption("phylo_debug"))) {
  options(phylo_debug = TRUE)
}
if (!exists("phylo_debug_log", envir = .GlobalEnv)) {
  phylo_debug_log <- character()
}
if (!exists("debug_log", inherits = TRUE)) {
  debug_log <- function(...) {
    msg <- sprintf(...)
    phylo_debug_log <<- c(phylo_debug_log, msg)
    cat("[DEBUG] ", msg, "\n", sep = "")
  }
}

# ── Dependencies ──────────────────────────────────────────────────────────────

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(patchwork)
library(paletteer)


# ── Labels, palettes and output directory ─────────────────────────────────────

color_palette <- paste0(clade_name, "_palette") # name of the clade palette defined in palettes.R
capitalized_taxon <- tools::toTitleCase(taxon_of_interest)
capitalized_trait <- tools::toTitleCase(gsub("_", " ", trait)) # underscores shown as spaces
capitalized_clade <- tools::toTitleCase(gsub("_", " ", clade_name))

# Tableau 10 under colorblind-safe names.
tableau_10 <- paletteer_d("ggthemes::Tableau_10")
names(tableau_10) <- c(
  "blue", "orange", "vermillion", "bluish_green",
  "green", "yellow", "reddish_purple", "pink",
  "mauve", "gray"
)

createDir(ci_dir)

```

## 1. Composition Analysis

When population count columns (`n_trait` / `c_trait`) are present, 95% Jeffreys credibility intervals are computed. For continuous traits, candidate divergence is evaluated analytically using the Phylogenetic Shift Score (PSS) under the best-fitting evolutionary model (OU / BM).

```{r compute_CI, message=FALSE, warning=FALSE}
# ── Trait categories and intervals ────────────────────────────────────────────

# Three paths: count data (Jeffreys intervals), ordinal coded levels, or a continuous trait.
# `category` is top / bottom / intermediate in all of them.
use_count_data <- has.n && has.c
# Ordinal trait (2 to 5 integer levels, or trait_type = ordinal): the extremes are the
# coded levels themselves.
is_ordinal <- !use_count_data &&
  exists("is_ordinal_trait", inherits = TRUE) &&
  is_ordinal_trait(trait_df[[trait]], if (exists("trait_type", inherits = TRUE)) trait_type else "auto")
debug_log("use_count_data = %s, is_ordinal = %s", use_count_data, is_ordinal)

if (use_count_data) {
  # Count path: 95% credible interval of the proportion n_cases / n_population (Jeffreys
  # prior, Beta(0.5, 0.5)); top and bottom are the species above and below the median.
  debug_log("compute_CI: Jeffreys CI method (n_trait = %s, c_trait = %s)", n_trait, c_trait)
  library(binom)

  # One bound of the interval (column 5 lower, 6 upper of binom.confint), NA for invalid counts.
  safe_binom_bound <- function(k, n, bound_col) {
    if (is.na(k) || is.na(n) || n <= 0 || k < 0 || k > n) {
      return(NA_real_)
    }
    res <- tryCatch(
      binom::binom.confint(k, n, method = "bayes", priors = c(0.5, 0.5), conf.level = 0.95),
      error = function(e) NULL
    )
    if (is.null(res)) return(NA_real_)
    return(res[[bound_col]])
  }

  g_med <- median(trait_df[[trait]], na.rm = TRUE)

  trait_ci <- trait_df %>%
    group_by(species) %>%
    dplyr::rename(n_population = .data[[n_trait]],
                  n_cases      = .data[[c_trait]],
                  trait        = .data[[trait]]) %>%
    dplyr::mutate(
      trait_ci_lb = safe_binom_bound(n_cases, n_population, 5),
      trait_ci_ub = safe_binom_bound(n_cases, n_population, 6),
      trait_ci = ifelse(is.na(trait_ci_lb) | is.na(trait_ci_ub),
                        NA_character_,
                        paste0("[", round(trait_ci_lb, 2), "-", round(trait_ci_ub, 2), "]")),
      category = dplyr::case_when(
        trait > g_med ~ "top",
        trait < g_med ~ "bottom",
        TRUE ~ "intermediate"
      )
    ) %>%
    ungroup()

  cat("Jeffreys Bayesian 95% Credible Intervals computed for count data.\n")

} else if (is_ordinal) {
  # Ordinal path: the highest coded level is top, the lowest is bottom, the rest intermediate.
  debug_log("compute_CI: Ordinal trait mode (coded levels)")
  .levels <- sort(unique(stats::na.omit(trait_df[[trait]])))
  .lo_lvl <- .levels[1]
  .hi_lvl <- .levels[length(.levels)]
  g_med <- median(trait_df[[trait]], na.rm = TRUE)

  trait_ci <- trait_df %>%
    dplyr::rename(trait = .data[[trait]]) %>%
    dplyr::mutate(
      category = dplyr::case_when(
        trait >= .hi_lvl ~ "top",
        trait <= .lo_lvl ~ "bottom",
        TRUE             ~ "intermediate"
      ),
      n_population = NA_real_,
      n_cases      = NA_real_,
      trait_ci_lb  = NA_real_,
      trait_ci_ub  = NA_real_,
      trait_ci     = NA_character_
    )

  cat("Ordinal trait mode: foreground = highest coded level, background = lowest; middle levels are intermediate (excluded from contrasts).\n")

} else {
  # Continuous path: the median splits top from bottom; the candidate pairs come from the
  # PSS gate below.
  debug_log("compute_CI: Continuous trait mode (PSS top pct = %.4f)", pss_top_pct)
  g_med <- median(trait_df[[trait]], na.rm = TRUE)

  trait_ci <- trait_df %>%
    dplyr::rename(trait = .data[[trait]]) %>%
    dplyr::mutate(
      category     = dplyr::case_when(
        trait > g_med ~ "top",
        trait < g_med ~ "bottom",
        TRUE          ~ "intermediate"
      ),
      n_population = NA_real_,
      n_cases      = NA_real_,
      trait_ci_lb  = NA_real_,
      trait_ci_ub  = NA_real_,
      trait_ci     = NA_character_
    )

  cat("Continuous trait mode: candidate pairs are selected via continuous Phylogenetic Shift Score (PSS) under the best-fitting evolutionary model.\n")
}

# Deduplicate by species so each species tree tip has exactly one row.
trait_ci <- trait_ci %>%
  dplyr::distinct(species, .keep_all = TRUE)

write.csv(trait_ci, file.path(ci_dir, "trait_CI.csv"),
          quote = FALSE, row.names = FALSE)

head(trait_ci)
```

## 3. Pairwise Comparison Preparation

This section prepares data for comparing trait values between all pairs of species.

A pair is an eligible contrast candidate when it passes the trait-type non-overlap test: **count** — Jeffreys credible intervals that do not overlap; **ordinal** — one species at the top coded level, the other at the bottom; **continuous** — the top `pss_top_pct` of hi>lo pairs by Phylogenetic Shift Score (PSS) under the AIC-selected OU/BM model (continuous has no other significance test). For every trait type the PSS is computed and used to **rank** the eligible candidates for the independent-contrast selection in the next report.

```{r prep_CI, paged.print=TRUE, message=FALSE, warning=FALSE}
# Table of all ordered species pairs (species1, species2) with the trait values, the
# categories, the intervals, the PSS score and `trait_overlap`. A pair with
# trait_overlap == FALSE is a candidate contrast: species1 has the higher trait value.
create_pairwise_data <- function(ci_data) {
  ci_subset <- ci_data %>%
    dplyr::distinct(species, .keep_all = TRUE) %>%
    dplyr::select(species, n_population, n_cases, trait,
                  trait_ci_lb, trait_ci_ub, category)

  pss_df <- data.frame(species1 = character(), species2 = character(),
                       pss_score = numeric(), S_score = numeric(), stringsAsFactors = FALSE)
  pss_top_pairs <- character(0)

  # Phylogenetic Shift Score (PSS), computed for every trait type with the engine of
  # src/pss_core.R (fit BM and OU, select by AIC: OU if AIC_OU + 2 < AIC_BM, else BM;
  # score every species pair as S * normDelta / normPatristic). pss_score is the pair
  # divergence signal that ranks the candidates; for a continuous trait the top
  # `pss_top_pct` of pairs is also a gate (below). When the fit fails, no score is
  # stored and the selection falls back to the type gate and patristic-distance ranking.
  .force <- if (exists("perm_strategy", inherits = TRUE) &&
                tolower(perm_strategy) %in% c("bm", "ou")) toupper(perm_strategy) else NULL
  .pct   <- if (exists("pss_top_pct", inherits = TRUE)) as.numeric(pss_top_pct) else 0.05
  if (exists("tree") && !is.null(tree) && length(tree$tip.label) >= 2) {
    tryCatch({
      .pss <- phylogenetic_shift_score(
        data = as.data.frame(ci_subset[, c("species", "trait")]),
        tree = tree, trait = "trait", species_col = "species",
        verbose = FALSE, force_model = .force
      )
      sc <- .pss$scores
      sc_S <- if (identical(.pss$selected_model, "BM")) sc$Sbm else sc$Sou
      pss_df <- rbind(
        data.frame(species1 = sc$Species1, species2 = sc$Species2,
                   pss_score = round(sc$FinalScore, 6), S_score = round(sc_S, 6),
                   stringsAsFactors = FALSE),
        data.frame(species1 = sc$Species2, species2 = sc$Species1,
                   pss_score = round(sc$FinalScore, 6), S_score = round(sc_S, 6),
                   stringsAsFactors = FALSE)
      )
      if (exists("ci_dir", inherits = TRUE) && dir.exists(ci_dir)) {
        write.table(.pss$model_fit, file.path(ci_dir, "evo_model.tsv"),
                    sep = "\t", row.names = FALSE, quote = FALSE)
      }
      debug_log("PSS: model=%s (AIC BM=%.2f OU=%.2f), %d scored pairs",
                .pss$selected_model, sc$AIC_BM[1], sc$AIC_OU[1], nrow(sc))
    }, error = function(e) {
      debug_log("PSS calculation failed: %s", conditionMessage(e))
    })
  }
  have_pss <- nrow(pss_df) > 0

  # All ordered species pairs, with the data of each member.
  pairwise <- expand.grid(species1 = ci_subset$species,
                          species2 = ci_subset$species,
                          stringsAsFactors = FALSE)

  pairwise <- pairwise %>%
    left_join(ci_subset %>% rename_with(~paste0(.x, "1"), .cols = -species),
              by = c("species1" = "species")) %>%
    left_join(ci_subset %>% rename_with(~paste0(.x, "2"), .cols = -species),
              by = c("species2" = "species"))

  if (nrow(pss_df) > 0) {
    pairwise <- pairwise %>%
      left_join(pss_df, by = c("species1", "species2"))
  } else {
    pairwise$pss_score <- NA_real_
    pairwise$S_score <- NA_real_
  }

  # Candidate eligibility (trait_overlap == FALSE):
  #   count      -> the Jeffreys intervals do not overlap. This is the gate; PSS only
  #                 ranks the survivors (4.Independent_contrasts.Rmd).
  #   ordinal    -> one species at the top coded level, the other at the bottom.
  #   continuous -> there is no separate significance test, so the top `pss_top_pct`
  #                 of the hi>lo pairs by PSS is the gate. A pss_top_pct outside (0, 1)
  #                 disables it: every hi>lo pair is kept and PSS only ranks.
  # When the PSS fit failed for a continuous trait, every hi>lo pair is kept and the
  # patristic-distance ranking takes over.
  pairwise <- pairwise %>%
    mutate(
      trait_diff = trait1 - trait2,
      .type_nonoverlap = if (use_count_data) {
        !ci_overlap(trait_ci_lb1, trait_ci_ub1, trait_ci_lb2, trait_ci_ub2)
      } else if (isTRUE(is_ordinal)) {
        (category1 %in% c("top", "bottom")) &
          (category2 %in% c("top", "bottom")) &
          (category1 != category2)
      } else {
        TRUE
      }
    )

  .stage1 <- which(pairwise$.type_nonoverlap & pairwise$trait_diff > 0 &
                   !is.na(pairwise$trait_diff))
  pairwise$trait_overlap <- TRUE
  if (length(.stage1) > 0) {
    .pct  <- if (exists("pss_top_pct", inherits = TRUE)) as.numeric(pss_top_pct) else 0.05
    .continuous_pss_gate <- !use_count_data && !isTRUE(is_ordinal) &&
      isTRUE(have_pss) && any(is.finite(pairwise$pss_score[.stage1])) &&
      is.finite(.pct) && .pct > 0 && .pct < 1
    if (.continuous_pss_gate) {
      .keep <- min(max(1L, ceiling(length(.stage1) * .pct)), length(.stage1))
      .ord  <- .stage1[order(pairwise$pss_score[.stage1], decreasing = TRUE)]
      pairwise$trait_overlap[.ord[seq_len(.keep)]] <- FALSE
      debug_log("Continuous PSS gate: %d pairs -> top %.1f%% -> %d candidates",
                length(.stage1), 100 * .pct, .keep)
    } else {
      pairwise$trait_overlap[.stage1] <- FALSE
      debug_log("Type gate: %d candidate pairs (PSS ranks, does not gate)", length(.stage1))
    }
  }
  pairwise$diff_value  <- pairwise$trait_diff
  pairwise$overlap_use <- pairwise$trait_overlap
  pairwise$.type_nonoverlap <- NULL

  return(pairwise)
}

pairwise_data <- create_pairwise_data(trait_ci)

# pairwise_data is written whole (pairwise_data.csv, read by 4.Independent_contrasts.Rmd);
# the plots use a copy without the rows that have NA trait values.
pairwise_data_viz <- pairwise_data %>%
  dplyr::filter(!is.na(trait1), !is.na(trait2), !is.na(diff_value), !is.na(overlap_use))

trait_ci_extreme <- trait_ci %>%
  dplyr::filter(category %in% c("top", "bottom"))

extreme_categories_present <- sort(unique(trait_ci_extreme$category))
has_heatmap_extremes <- nrow(trait_ci_extreme) >= 2 &&
  all(c("top", "bottom") %in% extreme_categories_present)

pairwise_data_viz_extreme <- pairwise_data_viz %>%
  dplyr::filter(
    category1 %in% c("top", "bottom"),
    category2 %in% c("top", "bottom")
  )

if (nrow(pairwise_data_viz) == 0) {
  warning("No valid non-NA rows available for pairwise visualization")
}

if (!has_heatmap_extremes) {
  if (use_count_data) {
    warning("Skipping CI heatmap visualization: fewer than two extreme species or only one extreme class remains.")
  } else {
    warning("Skipping phenotype heatmap visualization: fewer than two extreme species or only one extreme class remains.")
  }
}

write.csv(pairwise_data, file.path(ci_dir, "pairwise_data.csv"),
          quote = FALSE, row.names = FALSE)

# ── Candidate species for FADE ────────────────────────────────────────────────

# Union of the members of every candidate pair, taken before the independence selection
# (modified Dunn index) of 4.Independent_contrasts.Rmd. FADE tests selection on the
# foreground branches and does not need mutually independent pairs, so it uses this full
# pool. Written without header as species<TAB>contrast_group<TAB>pair (1 = top, 0 = bottom;
# `pair` is a running index), the format that EXTRACT_EXTREME_SPECIES in SELECTION_PREP reads.
.nonoverlap  <- pairwise_data %>%
  dplyr::filter(trait_overlap == FALSE, !is.na(species1), !is.na(species2))
.cand_species <- unique(c(.nonoverlap$species1, .nonoverlap$species2))
.cat_by_sp   <- setNames(as.character(trait_ci$category), trait_ci$species)
.fg_species  <- .cand_species[which(.cat_by_sp[.cand_species] == "top")]
.bg_species  <- .cand_species[which(.cat_by_sp[.cand_species] == "bottom")]
candidate_species <- rbind(
  if (length(.fg_species)) data.frame(species = .fg_species, contrast_group = 1L, stringsAsFactors = FALSE) else NULL,
  if (length(.bg_species)) data.frame(species = .bg_species, contrast_group = 0L, stringsAsFactors = FALSE) else NULL
)
if (!is.null(candidate_species) && nrow(candidate_species) > 0) {
  candidate_species$pair <- seq_len(nrow(candidate_species))
} else {
  candidate_species <- data.frame(species = character(), contrast_group = integer(), pair = integer())
}
write.table(candidate_species, file.path(ci_dir, "candidate_species.tab"),
            quote = FALSE, row.names = FALSE, col.names = FALSE, sep = "\t")
debug_log("candidate_species: %d foreground, %d background (pre-Dunn pool)",
          length(.fg_species), length(.bg_species))

head(pairwise_data)
```

## 4. Difference Matrix Creation

This section transforms pairwise data into matrix format for visualization.

```{r diff_matrix, paged.print=TRUE, message=FALSE, warning=FALSE}
# Wide matrix of the trait differences (rows species1, columns species2).
build_diff_matrix <- function(pairwise_data_input) {
  pairwise_data_input %>%
    dplyr::distinct(species1, species2, .keep_all = TRUE) %>%
    dplyr::select(species1, species2, diff_value) %>%
    tidyr::pivot_wider(names_from = species2, values_from = diff_value, values_fn = mean) %>%
    tibble::column_to_rownames("species1") %>%
    as.matrix()
}

# All species, and top/bottom species only when both classes are present.
matrix <- build_diff_matrix(pairwise_data_viz)
matrix_extreme <- if (has_heatmap_extremes) build_diff_matrix(pairwise_data_viz_extreme) else NULL

write.csv(matrix, file.path(ci_dir, "diff_matrix-CI.csv"), 
          quote = FALSE, row.names = TRUE)
if (!is.null(matrix_extreme)) {
  write.csv(matrix_extreme, file.path(ci_dir, "diff_matrix-CI-extreme_only.csv"),
            quote = FALSE, row.names = TRUE)
}

head(matrix)
```

## 5. Heatmap Visualization

This section creates heatmaps to visualize pairwise differences in prevalence. Species labels are colored based on prevalence quantiles.

```{r plot_CI, fig.height=12, fig.width=15, message=FALSE, warning=FALSE, dev="ragg_png", dpi=320}
# Plain-text species labels (underscores shown as spaces) with their trait and category.
create_species_labels <- function(trait_data) {
  trait_data %>%
    distinct(species, trait, category) %>%
    mutate(
      label_text = gsub("_", " ", species)
    ) %>%
    arrange(species)
}

# Canvas size, font sizes and margins of the heatmap by number of species; cell values
# are printed up to 28 species.
heatmap_layout_profile <- function(n_species) {
  if (n_species <= 10) {
    width <- clamp_value(9.5 + 0.55 * n_species, 12, 16)
    height <- clamp_value(9.0 + 0.60 * n_species, 12, 16)
    axis_label_size <- clamp_value(16.0 - 0.15 * n_species, 13.5, 16.0)
    cell_label_size <- clamp_value(6.8 - 0.08 * n_species, 5.4, 6.8)
    rotate_angle <- 35
    bottom_margin <- 90
    left_margin <- 85
    caption_size <- 16
  } else if (n_species <= 18) {
    width <- clamp_value(11.0 + 0.45 * n_species, 15, 22)
    height <- clamp_value(10.5 + 0.50 * n_species, 15, 22)
    axis_label_size <- clamp_value(14.0 - 0.12 * n_species, 10.5, 13.5)
    cell_label_size <- clamp_value(5.8 - 0.06 * n_species, 4.2, 5.8)
    rotate_angle <- 40
    bottom_margin <- 105
    left_margin <- 100
    caption_size <- 15
  } else {
    base_profile <- species_plot_profile(n_species = n_species, n_taxa = n_species)$contrast
    width  <- clamp_value(base_profile$width  + 0.26 * n_species, 18, 30)
    height <- clamp_value(base_profile$height + 0.24 * n_species, 18, 30)
    axis_label_size <- clamp_value(12.0 - 0.12 * n_species, 7.0, 10.5)
    cell_label_size <- clamp_value(5.2 - 0.04 * n_species, 3.2, 5.2)
    rotate_angle <- 45
    bottom_margin <- clamp_value(105 + 1.6 * n_species, 120, 165)
    left_margin <- clamp_value(95 + 1.3 * n_species, 115, 155)
    caption_size <- 14
  }

  list(
    width = width,
    height = height,
    axis_label_size = axis_label_size,
    cell_label_size = cell_label_size,
    tile_border = clamp_value(0.30 - 0.002 * n_species, 0.08, 0.30),
    highlight_border = clamp_value(1.5 - 0.010 * n_species, 0.45, 1.5),
    show_cell_labels = n_species <= 28,
    rotate_angle = rotate_angle,
    bottom_margin = bottom_margin,
    left_margin = left_margin,
    caption_margin = 22,
    caption_size = caption_size
  )
}

# Heatmap of the absolute pairwise differences; with `highlight_nonoverlap` the candidate
# pairs (overlap_use == FALSE) get a bold border.
create_heatmap <- function(pairwise_data, species_labels, 
                          highlight_nonoverlap = TRUE) {
  if (!all(c("species", "label_text") %in% names(species_labels))) {
    stop("species_labels must contain species and label_text columns.")
  }
  species_levels <- species_labels$species
  n_species <- length(species_levels)
  heatmap_dims <- heatmap_layout_profile(n_species)
  label_lookup <- setNames(as.character(species_labels$label_text), as.character(species_labels$species))

  pairwise_plot <- pairwise_data %>%
    mutate(
      species1 = factor(species1, levels = rev(species_levels)),
      species2 = factor(species2, levels = species_levels)
    )

  overlap_label <- if (use_count_data) "Non-overlapping CIs" else "Non-overlapping phenotypes"
  overlap_caption <- if (use_count_data) {
    "Bold border: non-overlapping CIs."
  } else {
    "Bold border: non-overlapping phenotypes."
  }

  p <- ggplot(pairwise_plot, aes(x = species2, y = species1, fill = abs(diff_value))) +
    geom_tile(color = "black", linewidth = heatmap_dims$tile_border) +
    scale_fill_gradient(low = "white", high = as.character(tableau_10["vermillion"]), name = "Difference") +
    scale_x_discrete(
      labels = function(x) unname(label_lookup[as.character(x)]),
      expand = c(0, 0)
    ) +
    scale_y_discrete(
      labels = function(y) unname(label_lookup[as.character(y)]),
      expand = c(0, 0)
    ) +
    theme_minimal(base_size = 16) +
    theme(
      axis.text.x = element_text(
        size = heatmap_dims$axis_label_size,
        angle = heatmap_dims$rotate_angle,
        hjust = 1,
        vjust = 1,
        colour = "#222222"
      ),
      axis.text.y = element_text(
        size = heatmap_dims$axis_label_size,
        hjust = 1,
        vjust = 0.5,
        colour = "#222222"
      ),
      axis.title = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank(),
      legend.position = "none",
      plot.margin = margin(30, 25, heatmap_dims$bottom_margin, heatmap_dims$left_margin),
      plot.caption = element_text(size = heatmap_dims$caption_size, hjust = 0, margin = margin(t = heatmap_dims$caption_margin))
    )

  if (isTRUE(heatmap_dims$show_cell_labels)) {
    p <- p + geom_text(aes(label = sprintf("%.2f", abs(diff_value))), size = heatmap_dims$cell_label_size)
  }
  
  if (highlight_nonoverlap) {
    p <- p + 
      geom_tile(data = filter(pairwise_plot, !overlap_use), 
                fill = NA, color = "black", linewidth = heatmap_dims$highlight_border) +
      labs(
        x = NULL, y = NULL,
        title = paste("Pairwise Differences in trait: ", capitalized_trait, " with ", overlap_label, sep = ""),
        caption = overlap_caption
      )
  } else {
    p <- p +
      labs(
        x = NULL, y = NULL,
        title = paste("Pairwise Differences in trait: ", capitalized_trait, sep = ""),
        caption = overlap_caption
      )
  }

  attr(p, "heatmap_dims") <- heatmap_dims[c("width", "height")]
  return(p)
}

if (has_heatmap_extremes) {
  species_labels <- create_species_labels(trait_ci_extreme)

  # Label table, kept for inspection.
  write.csv(species_labels, file.path(ci_dir, "my_labels_colored.csv"), 
            quote = FALSE, row.names = FALSE)

  overlap_heatmap <- create_heatmap(pairwise_data_viz_extreme, species_labels, highlight_nonoverlap = TRUE)
  heatmap_dims <- attr(overlap_heatmap, "heatmap_dims")

  p_overlap <- overlap_heatmap

  knitr::opts_current$set(fig.width = heatmap_dims$width, fig.height = heatmap_dims$height)

  ggsave(file.path(ci_dir, "visual_matrix_no_overlap.png"), p_overlap,
         width = heatmap_dims$width, height = heatmap_dims$height, dpi = "retina",
         device = ragg::agg_png)

  print(p_overlap)
} else {
  if (use_count_data) {
    cat("CI heatmap skipped: fewer than two extreme species remain, or only one extreme class is present.\n")
  } else {
    cat("Phenotype heatmap skipped: fewer than two extreme species remain, or only one extreme class is present.\n")
  }
}
```

## 6. Polished Tables Preparation

This section creates readable summary tables for the trait and pairwise comparisons.

```{r polished_tables, message=FALSE, warning=FALSE}
# The count columns exist only on the count path.
count_cols_select  <- if (use_count_data) c("n_population", "n_cases") else character(0)

# ── Tables ────────────────────────────────────────────────────────────────────

# Trait table with readable column names.
trait_ci_polished <- trait_ci %>%
  mutate(
    trait = round(trait, 3),
  ) %>%
  dplyr::select(species, all_of(count_cols_select), category,
                trait, trait_ci_lb, trait_ci_ub, trait_ci) %>%
  dplyr::rename(
    "Category"           = category,
    "Trait"              = trait,
    "Trait CI Lower Bound" = trait_ci_lb,
    "Trait CI Upper Bound" = trait_ci_ub,
    "Trait CI / Group"   = trait_ci
  )

write.csv(trait_ci_polished, 
          file.path(ci_dir, "trait_CI_polished.csv"), 
          quote = FALSE, row.names = FALSE)

# Pairwise table rebuilt from the trait table: absolute difference and the two intervals.
trait_diff <- trait_ci %>%
  dplyr::select(species, all_of(count_cols_select), category, trait,
                trait_ci_lb, trait_ci_ub, trait_ci)

pairwise_data_full <- expand.grid(species1 = trait_diff$species,
                                   species2 = trait_diff$species,
                                   stringsAsFactors = FALSE) %>%
  left_join(trait_diff %>% rename_with(~paste0(.x, "1"), .cols = -species),
            by = c("species1" = "species")) %>%
  left_join(trait_diff %>% rename_with(~paste0(.x, "2"), .cols = -species),
            by = c("species2" = "species"))

pairwise_data_end_polished <- pairwise_data_full %>%
  dplyr::mutate(
    trait_diff = abs(trait1 - trait2),
    trait_diff = round(trait_diff, 3),
    # Interval as text, "[lower-upper]".
    trait_CI1 = sprintf("[%.2f-%.2f]", 
                                 as.numeric(trait_ci_lb1), 
                                 as.numeric(trait_ci_ub1)),
    trait_CI2 = sprintf("[%.2f-%.2f]", 
                                 as.numeric(trait_ci_lb2), 
                                 as.numeric(trait_ci_ub2)),
    diff_value = trait_diff
  ) %>%
  dplyr::select(species1, species2, 
         trait_diff, trait_CI1, trait_CI2) %>%
  dplyr::rename(
    "Trait Difference" = trait_diff,
    "Trait CI1" = trait_CI1,        
    "Trait CI2" = trait_CI2
  )

write.csv(pairwise_data_end_polished,
          file.path(ci_dir, "pairwise_data_end_polished.csv"),
          quote = FALSE, row.names = FALSE)

# Candidate pairs only, each pair once (species1 < species2).
pairwise_nonoverlap <- pairwise_data_end_polished %>%
  dplyr::left_join(
    pairwise_data %>% dplyr::select(species1, species2, overlap_use),
    by = c("species1", "species2")
  ) %>%
  dplyr::filter(!overlap_use, species1 < species2) %>%
  dplyr::select(-overlap_use)

pairwise_nonoverlap

```

---

## Summary

This report summarizes the results of the **Trait Analysis** analysis.

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — Trait Analysis Module*
*Date of generation: `r Sys.Date()`*
