---
title: "Trait Composition Analysis"
author: "PhyloPhere Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    code_folding: hide
    df_print: paged
params:
  trait_file: NULL
  tree_file: NULL
  output_dir: !r getwd()
  seed: ""
  clade_name: "clade"
  taxon_of_interest: "family"
  traitname: "trait"
  n_trait: ""
  c_trait: ""
  tax_id: ""
  secondary_trait: ""
  branch_trait: ""
  discrete_method: "decile"
  top_quantile: "0.90"
  bottom_quantile: "0.10"
---

```{r header}

#  ██████╗ ██╗  ██╗██╗   ██╗██╗      ██████╗ ██████╗ ██╗  ██╗███████╗██████╗ ███████╗
#  ██╔══██╗██║  ██║╚██╗ ██╔╝██║     ██╔═══██╗██╔══██╗██║  ██║██╔════╝██╔══██╗██╔════╝
#  ██████╔╝███████║ ╚████╔╝ ██║     ██║   ██║██████╔╝███████║█████╗  ██████╔╝█████╗
#  ██╔═══╝ ██╔══██║  ╚██╔╝  ██║     ██║   ██║██╔═══╝ ██╔══██║██╔══╝  ██╔══██╗██╔══╝
#  ██║     ██║  ██║   ██║   ███████╗╚██████╔╝██║     ██║  ██║███████╗██║  ██║███████╗
#  ╚═╝     ╚═╝  ╚═╝   ╚═╝   ╚══════╝ ╚═════╝ ╚═╝     ╚═╝  ╚═╝╚══════╝╚═╝  ╚═╝╚══════╝

## PHYLOPHERE: A Nextflow pipeline including a complete set
## of phylogenetic comparative tools and analyses for Phenome-Genome studies

### Github: https://github.com/nozerorma/caastools/nf-phylophere
### Author: Miguel Ramon (miguel.ramon@upf.edu)

#### File: CI-composition.Rmd

```

# Trait Composition Analysis
This script computes contrasting species pairs for the trait of interest using one of two methods:

- **Jeffreys CI method** (when `n_trait` and `c_trait` count columns exist): computes 95% credibility intervals using a Beta(0.5,0.5) prior and identifies species pairs whose CIs do not overlap.
- **Discrete categorization** (fallback when count columns are absent): assigns each species to *top*, *bottom*, or *intermediate* groups using one of several threshold strategies (`quartile`, `quintile`, `decile`, `median_sd`, `parameterized`), and treats top–bottom pairs as non-overlapping.

---

## Setup and Directory Configuration

This section configures the working environment, sets directories, and loads necessary functions and libraries.

```{r setup, message=TRUE, warning=TRUE}
# Make params available in environment before sourcing commons.R
if (!exists("params")) {
  stop("params object not found - ensure this Rmd is rendered with rmarkdown::render()")
}
assign("params", params, envir = .GlobalEnv)

# Call the setup function from commons.R
source("./obj/commons.R")
setup_rmd()

# discrete_method, top_quantile, bottom_quantile are set by commons.R from params
debug_log("discrete_method = %s, top_quantile = %s, bottom_quantile = %s",
          discrete_method, top_quantile, bottom_quantile)

# Debug helper (prints into HTML output)
if (is.null(getOption("phylo_debug"))) {
  options(phylo_debug = TRUE)
}
if (!exists("phylo_debug_log", envir = .GlobalEnv)) {
  phylo_debug_log <- character()
}
if (!exists("debug_log", inherits = TRUE)) {
  debug_log <- function(...) {
    msg <- sprintf(...)
    phylo_debug_log <<- c(phylo_debug_log, msg)
    cat("[DEBUG] ", msg, "\n", sep = "")
  }
}

# Load additional libraries
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(patchwork)
library(ggtext)
library(paletteer)

color_palette <- paste0(clade_name, "_palette") # Color palette for the clade
capitalized_taxon <- tools::toTitleCase(taxon_of_interest) # Capitalized taxon name
capitalized_trait <- tools::toTitleCase(gsub("_", " ", trait)) # Capitalized and deconvoluted trait name
capitalized_clade <- tools::toTitleCase(gsub("_", " ", clade_name)) # Capitalized and deconvoluted clade name

# Define color palettes, including a colorblind-friendly palette for categories
tableau_10 <- paletteer_d("ggthemes::Tableau_10")
names(tableau_10) <- c(
  "blue", "orange", "vermillion", "bluish_green",
  "green", "yellow", "reddish_purple", "pink",
  "mauve", "gray"
)

createDir(ci_dir)

```

## 1. Composition Analysis

When population count columns (`n_trait` / `c_trait`) are present, 95% Jeffreys credibility intervals are computed. Otherwise, species are categorised into *top*, *bottom*, and *intermediate* groups using the selected `discrete_method`.

```{r compute_CI, message=TRUE, warning=TRUE}
# Determine which analysis path to follow
use_count_data <- has.n && has.c
debug_log("use_count_data = %s", use_count_data)

if (use_count_data) {
  # ---- Jeffreys CI path ----
  debug_log("compute_CI: Jeffreys CI method (n_trait = %s, c_trait = %s)", n_trait, c_trait)
  library(binom)

  trait_ci <- trait_df %>%
    group_by(species) %>%
    dplyr::rename(n_population = .data[[n_trait]],
                  n_cases      = .data[[c_trait]],
                  trait        = .data[[trait]]) %>%
    dplyr::mutate(
      # Jeffreys beta method CIs
      trait_ci_lb = binom::binom.confint(n_cases, n_population,
                                         method = "bayes", priors = c(0.5, 0.5),
                                         conf.level = 0.95)[, 5],
      trait_ci_ub = binom::binom.confint(n_cases, n_population,
                                         method = "bayes", priors = c(0.5, 0.5),
                                         conf.level = 0.95)[, 6],
      trait_ci = paste0("[", round(trait_ci_lb, 2), "-", round(trait_ci_ub, 2), "]"),
      # Category assigned from quartiles of the trait for colouring
      category = dplyr::case_when(
        trait >= quantile(trait, 0.75, na.rm = TRUE) ~ "top",
        trait <= quantile(trait, 0.25, na.rm = TRUE) ~ "bottom",
        TRUE ~ "intermediate"
      )
    ) %>%
    ungroup()

} else {
  # ---- Discrete categorization path ----
  debug_log("compute_CI: discrete method = %s", discrete_method)

  trait_vals <- trait_df[[trait]]

  thresholds <- switch(
    discrete_method,
    "quartile" = list(
      lower = quantile(trait_vals, 0.25, na.rm = TRUE),
      upper = quantile(trait_vals, 0.75, na.rm = TRUE)
    ),
    "quintile" = list(
      lower = quantile(trait_vals, 0.20, na.rm = TRUE),
      upper = quantile(trait_vals, 0.80, na.rm = TRUE)
    ),
    "median_sd" = list(
      lower = median(trait_vals, na.rm = TRUE) - sd(trait_vals, na.rm = TRUE),
      upper = median(trait_vals, na.rm = TRUE) + sd(trait_vals, na.rm = TRUE)
    ),
    "decile" = list(
      lower = quantile(trait_vals, 0.10, na.rm = TRUE),
      upper = quantile(trait_vals, 0.90, na.rm = TRUE)
    ),
    "parameterized" = list(
      lower = quantile(trait_vals, as.numeric(bottom_quantile), na.rm = TRUE),
      upper = quantile(trait_vals, as.numeric(top_quantile),    na.rm = TRUE)
    ),
    stop(sprintf(
      "Unknown discrete_method: '%s'. Choose one of: quartile, quintile, decile, median_sd, parameterized",
      discrete_method
    ))
  )

  lower_thresh <- as.numeric(thresholds$lower)
  upper_thresh <- as.numeric(thresholds$upper)
  debug_log("compute_CI: lower_thresh = %.4f, upper_thresh = %.4f", lower_thresh, upper_thresh)

  trait_ci <- trait_df %>%
    dplyr::rename(trait = .data[[trait]]) %>%
    dplyr::mutate(
      category     = dplyr::case_when(
        trait >= upper_thresh ~ "top",
        trait <= lower_thresh ~ "bottom",
        TRUE                  ~ "intermediate"
      ),
      n_population = NA_real_,
      n_cases      = NA_real_,
      # Pseudo-bounds: the thresholds themselves, so downstream columns are consistent
      trait_ci_lb  = lower_thresh,
      trait_ci_ub  = upper_thresh,
      trait_ci     = category
    )

  cat("Discrete categorization summary:\n")
  print(table(trait_ci$category))
  cat(sprintf("Method: %s | lower threshold = %.4f | upper threshold = %.4f\n",
              discrete_method, lower_thresh, upper_thresh))
}

# Save results
write.csv(trait_ci, file.path(ci_dir, "trait_CI.csv"),
          quote = FALSE, row.names = FALSE)

# Display preview
head(trait_ci)
```

## 3. Pairwise Comparison Preparation

This section prepares data for comparing trait values between all pairs of species.
For the **Jeffreys CI** path, two species are considered non-overlapping when their credibility intervals do not overlap.
For the **discrete** path, a pair is considered non-overlapping only when one species is in the *top* group and the other in the *bottom* group.

```{r prep_CI, paged.print=TRUE, message=TRUE, warning=TRUE}
# Helper function to create pairwise comparisons
create_pairwise_data <- function(ci_data) {
  ci_subset <- ci_data %>%
    dplyr::select(species, n_population, n_cases, trait,
                  trait_ci_lb, trait_ci_ub, category)

  # Generate all pairwise species combinations
  pairwise <- expand.grid(species1 = ci_subset$species,
                          species2 = ci_subset$species,
                          stringsAsFactors = FALSE)

  # Merge species data for each pair
  pairwise <- pairwise %>%
    left_join(ci_subset %>% rename_with(~paste0(.x, "1"), .cols = -species),
              by = c("species1" = "species")) %>%
    left_join(ci_subset %>% rename_with(~paste0(.x, "2"), .cols = -species),
              by = c("species2" = "species"))

  pairwise <- pairwise %>%
    mutate(
      trait_diff = trait1 - trait2,
      trait_overlap = if (use_count_data) {
        # CI-based: overlapping when intervals share any range
        ci_overlap(trait_ci_lb1, trait_ci_ub1, trait_ci_lb2, trait_ci_ub2)
      } else {
        # Discrete: non-overlapping only for top vs bottom contrasts
        !(category1 %in% c("top", "bottom") &
          category2 %in% c("top", "bottom") &
          category1 != category2)
      },
      diff_value  = trait_diff,
      overlap_use = trait_overlap
    )

  return(pairwise)
}

# Create pairwise data
pairwise_data <- create_pairwise_data(trait_ci)

# Keep full pairwise_data for downstream contrast selection files,
# but generate a visualization-only copy that removes rows involving NA trait values.
pairwise_data_viz <- pairwise_data %>%
  dplyr::filter(!is.na(trait1), !is.na(trait2), !is.na(diff_value), !is.na(overlap_use))

if (nrow(pairwise_data_viz) == 0) {
  warning("No valid non-NA rows available for pairwise visualization")
}

# Save results
write.csv(pairwise_data, file.path(ci_dir, "pairwise_data.csv"), 
          quote = FALSE, row.names = FALSE)

# Display preview
head(pairwise_data)
```

## 4. Difference Matrix Creation

This section transforms pairwise data into matrix format for visualization.

```{r diff_matrix, paged.print=TRUE, message=TRUE, warning=TRUE}
# Helper function to build difference matrix
build_diff_matrix <- function(pairwise_data_input) {
  pairwise_data_input %>%
    select(species1, species2, diff_value) %>%
    pivot_wider(names_from = species2, values_from = diff_value) %>%
    column_to_rownames("species1") %>%
    as.matrix()
}

# Create matrices
matrix <- build_diff_matrix(pairwise_data_viz)

# Save matrices
write.csv(matrix, file.path(ci_dir, "diff_matrix-CI.csv"), 
          quote = FALSE, row.names = TRUE)

# Display preview
head(matrix)
```

## 5. Heatmap Visualization

This section creates heatmaps to visualize pairwise differences in prevalence. Species labels are colored based on prevalence quantiles.

```{r plot_CI, fig.height=12, fig.width=15, message=TRUE, warning=TRUE, dev="png", dpi=320}
# Simplified function to create colored labels based on trait quantiles
create_colored_labels <- function(trait_data) {
  # Calculate quantile thresholds
  q75 <- quantile(trait_data$trait, 0.75, na.rm = TRUE)
  q25 <- quantile(trait_data$trait, 0.25, na.rm = TRUE)
  
  # Create a lookup table with species and their colors
  label_df <- trait_data %>%
    mutate(
      color = case_when(
        trait >= q75 ~ as.character(tableau_10["vermillion"]),  # Red for top quartile
        trait <= q25 ~ as.character(tableau_10["bluish_green"]),  # Green for bottom quartile
        TRUE ~ "#000000"           # Black for middle
      )
    ) %>%
    select(species, color)
  
  # Create named vector of colored HTML labels
  colored_labels <- setNames(
    paste0("<span style='color:", label_df$color, ";'>", label_df$species, "</span>"),
    label_df$species
  )
  
  return(colored_labels)
}

# Helper function to create legend plot
create_legend_plot <- function() {
  legend_df <- tibble(
    category = c("Top", "Bottom"),
    color = c(tableau_10["vermillion"], tableau_10["bluish_green"])
  )
  
  ggplot(legend_df, aes(x = 1, y = category, color = category)) +
    geom_point(size = 0) +
    scale_color_manual(
      values = setNames(legend_df$color, legend_df$category),
      guide = guide_legend(override.aes = list(size = 5))
    ) +
    labs(color = "Trait Categories") +
    theme_void() +
    theme(
      legend.position = "right",
      legend.title = element_text(size = 12),
      legend.text = element_text(size = 10)
    )
}

# Helper function to create heatmap
create_heatmap <- function(pairwise_data, colored_labels, 
                          highlight_nonoverlap = TRUE) {
  # Ensure label mapping is stable across axes
  if (is.null(names(colored_labels))) {
    stop("colored_labels must be a named vector with species names.")
  }
  pairwise_data <- pairwise_data %>%
    mutate(
      species1 = factor(species1, levels = names(colored_labels)),
      species2 = factor(species2, levels = names(colored_labels))
    )

  p <- ggplot(pairwise_data, aes(x = species2, y = species1, fill = abs(diff_value))) +
    geom_tile(color = "black", size = 0.25) +
    geom_text(aes(label = sprintf("%.2f", abs(diff_value))), size = 3) +
    scale_fill_gradient(low = "white", high = tableau_10["vermillion"], name = "Difference") +
    scale_x_discrete(labels = function(x) colored_labels[x]) +
    scale_y_discrete(labels = function(y) colored_labels[y]) +
    theme_minimal(base_size = 16) +
    theme(
      axis.text.x = element_markdown(angle = 45, hjust = 1),
      axis.text.y = element_markdown(angle = 0),
      panel.grid = element_blank(),
      legend.position = "none"
    )
  
  # Add bold border for non-overlapping CIs if requested
  if (highlight_nonoverlap) {
    p <- p + 
      geom_tile(data = filter(pairwise_data, !overlap_use), 
                fill = NA, color = "black", size = 1.5) +
      labs(
        x = "Species", y = "Species",
        title = paste("Pairwise Differences in trait: ", capitalized_trait, " with Non-overlapping CIs", sep = ""),
        caption = "Bold border: non-overlapping CIs."
      )
  } else {
    p <- p +
      labs(
        x = "Species", y = "Species",
        title = paste("Pairwise Differences in trait: ", capitalized_trait, sep = ""),
        caption = "Bold border: non-overlapping CIs."
      )
  }
  
  return(p)
}

# Create colored labels directly from trait data
my_labels_colored <- create_colored_labels(trait_ci)

# Save colored labels
write.csv(my_labels_colored, file.path(ci_dir, "my_labels_colored.csv"), 
          quote = FALSE, row.names = TRUE)

# Create legend
legend_plot <- create_legend_plot()

# Create heatmaps with non-overlapping CI highlights
p_overlap <- create_heatmap(pairwise_data_viz, my_labels_colored, highlight_nonoverlap = TRUE) + 
  legend_plot + plot_layout(ncol = 2, widths = c(1000, 1))

p_flat <- create_heatmap(pairwise_data_viz, my_labels_colored, highlight_nonoverlap = FALSE) + 
  legend_plot + plot_layout(ncol = 2, widths = c(1000, 1))

# Save heatmaps
ggsave(file.path(ci_dir, "visual_matrix_no_overlap.png"), p_overlap, 
       width = 15, height = 12, dpi = "retina")
ggsave(file.path(ci_dir, "visual_matrix_flat.png"), p_flat, 
       width = 15, height = 12, dpi = "retina")

# Display plots
p_overlap
p_flat
```

## 6. Polished Tables Preparation

This section creates readable summary tables for the trait and pairwise comparisons.

```{r polished_tables, message=FALSE, warning=FALSE}
# Conditionally include count columns (only present in Jeffreys CI path)
count_cols_select  <- if (use_count_data) c("n_population", "n_cases") else character(0)

# Polished table for trait values
trait_ci_polished <- trait_ci %>%
  mutate(
    trait = round(trait, 3),
  ) %>%
  dplyr::select(species, all_of(count_cols_select), category,
                trait, trait_ci_lb, trait_ci_ub, trait_ci) %>%
  dplyr::rename(
    "Category"           = category,
    "Trait"              = trait,
    "Trait CI Lower Bound" = trait_ci_lb,
    "Trait CI Upper Bound" = trait_ci_ub,
    "Trait CI / Group"   = trait_ci
  )

write.csv(trait_ci_polished, 
          file.path(ci_dir, "trait_CI_polished.csv"), 
          quote = FALSE, row.names = FALSE)

# Polished pairwise table - recreate from original data with available CI columns
trait_diff <- trait_ci %>%
  dplyr::select(species, all_of(count_cols_select), category, trait,
                trait_ci_lb, trait_ci_ub, trait_ci)

# Generate all pairwise species combinations
pairwise_data_full <- expand.grid(species1 = trait_diff$species,
                                   species2 = trait_diff$species,
                                   stringsAsFactors = FALSE) %>%
  left_join(trait_diff %>% rename_with(~paste0(.x, "1"), .cols = -species),
            by = c("species1" = "species")) %>%
  left_join(trait_diff %>% rename_with(~paste0(.x, "2"), .cols = -species),
            by = c("species2" = "species"))

# Create polished pairwise table
pairwise_data_end_polished <- pairwise_data_full %>%
  dplyr::mutate(
    trait_diff = abs(trait1 - trait2),
    trait_diff = round(trait_diff, 3),
    # Format CIs for both Wilson and Jeffreys methods
    trait_CI1 = sprintf("[%.2f-%.2f]", 
                                 as.numeric(trait_ci_lb1), 
                                 as.numeric(trait_ci_ub1)),
    trait_CI2 = sprintf("[%.2f-%.2f]", 
                                 as.numeric(trait_ci_lb2), 
                                 as.numeric(trait_ci_ub2)),
    diff_value = trait_diff
  ) %>%
  dplyr::select(species1, species2, 
         trait_diff, trait_CI1, trait_CI2) %>%
  dplyr::rename(
    "Trait Difference" = trait_diff,
    "Trait CI1" = trait_CI1,        
    "Trait CI2" = trait_CI2
  )

write.csv(pairwise_data_end_polished, 
          file.path(ci_dir, "pairwise_data_end_polished.csv"), 
          quote = FALSE, row.names = FALSE)

# Display preview
head(trait_ci_polished)

```

## Session Information

```{r debug_dump, results='asis'}

if (length(phylo_debug_log) > 0) {
  cat("### Debug log\n")
  cat(paste0("[DEBUG] ", phylo_debug_log, "\n"))
}

```

```{r session_info}
sessionInfo()
```
