---
title: "Independent Contrast Selection for Trait Analysis"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
params:
  trait_file: NULL
  tree_file: NULL
  output_dir: !r getwd()
  seed: ""
  clade_name: "clade"
  taxon_of_interest: "family"
  traitname: "trait"
  n_trait: ""
  c_trait: ""
  tax_id: ""
  secondary_trait: ""
  branch_trait: ""
  discrete_method: "decile"
  top_quantile: "0.90"
  bottom_quantile: "0.10"
  trait_type: ""
  contrast_max_iter: "3"
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

```{r header}

#  ██████╗ ██╗  ██╗██╗   ██╗██╗      ██████╗ ██████╗ ██╗  ██╗███████╗██████╗ ███████╗
#  ██╔══██╗██║  ██║╚██╗ ██╔╝██║     ██╔═══██╗██╔══██╗██║  ██║██╔════╝██╔══██╗██╔════╝
#  ██████╔╝███████║ ╚████╔╝ ██║     ██║   ██║██████╔╝███████║█████╗  ██████╔╝█████╗
#  ██╔═══╝ ██╔══██║  ╚██╔╝  ██║     ██║   ██║██╔═══╝ ██╔══██║██╔══╝  ██╔══██╗██╔══╝
#  ██║     ██║  ██║   ██║   ███████╗╚██████╔╝██║     ██║  ██║███████╗██║  ██║███████╗
#  ╚═╝     ╚═╝  ╚═╝   ╚═╝   ╚══════╝ ╚═════╝ ╚═╝     ╚═╝  ╚═╝╚══════╝╚═╝  ╚═╝╚══════╝

## PHYLOPHERE: A Nextflow pipeline including a complete set
## of phylogenetic comparative tools and analyses for Phenome-Genome studies

### Github: https://github.com/nozerorma/caastools/nf-phylophere
### Author: Miguel Ramon (miguel.ramon@upf.edu)

#### File: Independent_contrasts.Rmd

```

# Independent Contrast Selection for Neoplasia Analysis in Primates
This document outlines the process of selecting phylogenetically independent species pairs for trait analysis using CAASTools-paired. 


## Setup and Directory Configuration

This section configures the working environment, sets directories, and loads necessary functions and libraries.

```{r setup, message=FALSE, warning=FALSE}

# Make params available in environment before sourcing commons.R
if (!exists("params")) {
  stop("params object not found - ensure this Rmd is rendered with rmarkdown::render()")
}
assign("params", params, envir = .GlobalEnv)

# Call the setup function from commons.R
source("./src/commons.R")
setup_rmd()

# Debug helper (prints into HTML output)
if (is.null(getOption("phylo_debug"))) {
  options(phylo_debug = TRUE)
}
if (!exists("phylo_debug_log", envir = .GlobalEnv)) {
  phylo_debug_log <- character()
}
if (!exists("debug_log", inherits = TRUE)) {
  debug_log <- function(...) {
    msg <- sprintf(...)
    phylo_debug_log <<- c(phylo_debug_log, msg)
    cat("[DEBUG] ", msg, "\n", sep = "")
  }
}

# Load additional libraries
library(dplyr)
library(purrr)
library(tibble)
library(ggplot2)
library(readr)
library(reshape2)
library(ggrepel)
library(ggtree)
library(ggnewscale)
library(ggstar)
library(colorspace)
library(treeio)
library(tidytree)
library(phylolm)
library(ggtreeExtra)
library(geiger)
library(rphylopic)
library(paletteer)

# Load selection algorithm
source("./src/selection_algorithm.R")

color_palette <- paste0(clade_name, "_palette") # Color palette for the clade
capitalized_taxon <- tools::toTitleCase(taxon_of_interest) # Capitalized taxon name
capitalized_trait <- tools::toTitleCase(gsub("_", " ", trait)) # Capitalized and deconvoluted trait name
capitalized_clade <- tools::toTitleCase(gsub("_", " ", clade_name)) # Capitalized and deconvoluted clade name

createDir(traitfile_dir)
createDir(bootstrap_traitfile_dir)

```

## 2. Patristic Distance Calculation

This section calculates evolutionary distances between species using the phylogenetic tree. Patristic distance represents the sum of branch lengths separating two species.

```{r patristic_distances, message=FALSE, warning=FALSE}

# Calculate patristic distances using pruned tree
distance_matrix <- calculate_patristic_distances(
  tree = pruned_tree, 
  df = trait_df
)

# Display distance matrix preview
cat("Distance matrix dimensions:", dim(distance_matrix), "\n")
head(distance_matrix[, 1:5])

```

## 3. Independent Contrast Pair Selection

This section selects phylogenetically independent species pairs for CAAS analysis. The algorithm prioritizes pairs with small evolutionary distances but large trait differences, while ensuring phylogenetic independence between pairs using the Dunn index.

When the input `traits_df` contains a numeric `p` column (population size), tie-breaking now follows: highest Dunn index → highest absolute trait difference (`diff`) → highest combined `n` for the pair. If `n` is absent, behaviour is unchanged.

```{r pair_selection, message=FALSE, warning=FALSE}

# Load summary stats from Dataset Exploration
stats_path <- file.path(species_distribution_dir, "trait_stats.csv")
if (!file.exists(stats_path)) {
  stop("Missing trait stats file: ", stats_path)
}
stats_df <- read.csv(stats_path, stringsAsFactors = FALSE)

# Load pairwise overlap data from CI-composition step
pairwise_path <- file.path(ci_dir, "pairwise_data.csv")
if (!file.exists(pairwise_path)) {
  stop("Missing pairwise overlap file: ", pairwise_path)
}
pairwise_data <- read.csv(pairwise_path, stringsAsFactors = FALSE) %>%
  dplyr::filter(trait_overlap == FALSE)

# Run pair selection for both malignant and neoplasia traits
selected_pairs.f <- function(my_trait) {
  
  cat("\n=== Selecting pairs for", my_trait, "===\n")
  
  # Select pairs using phylogeny-aware algorithm
  selected_pairs <- pair_sel.f(
    distance_matrix = distance_matrix,
    overlap_df = pairwise_data,
    traits_df = stats_df,
    my_trait = my_trait
  )$selected_pairs
  
  cat("Selected", nrow(selected_pairs), "phylogenetically independent pairs\n")
  
  return(selected_pairs)
}

selected_pairs <- selected_pairs.f(trait)

# Display selected pairs
print(selected_pairs)
```


## 4. Trait File Generation

This section generates trait files in the format required for CAAS discovery, assigning contrast groups (high/low extremes) to each species.

```{r traitfile_generation, message=FALSE, warning=FALSE}

# Function to generate trait files from selected pairs
gen_traitfiles <- function(df) {
  df <- df %>%
    tibble::rownames_to_column(var = "pair_id") %>%
    dplyr::mutate(rank = dplyr::row_number()) %>%
    tidyr::pivot_longer(
      cols = c(species1, species2), 
      names_to = "global_label", 
      values_to = "species"
    ) %>%
    dplyr::mutate(
      global_label = ifelse(
        global_label == "species1", 
        "high_extreme", 
        "low_extreme"
      )
    )
  
  # Assign contrast groups
  filtered_species <- df %>%
    dplyr::group_by(global_label) %>%
    dplyr::select(species, global_label) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(
      contrast_group = case_when(
        global_label == "high_extreme" ~ 1,
        global_label == "low_extreme" ~ 0,
        TRUE ~ NA_integer_
      )
    ) %>%
    dplyr::select(-global_label) %>%
    dplyr::mutate(pair = rep(1:(nrow(df)/2), each = 2)) %>%
    dplyr::arrange(contrast_group, pair)
  
  return(filtered_species)
}

# Generate trait files for selected pairs
traits_res <- gen_traitfiles(selected_pairs)

# Save trait files
trait_path <- file.path(traitfile_dir)
createDir(trait_path)
write.table(  
  traits_res, 
  file.path(trait_path, "traitfile.tab"), 
  quote = FALSE, 
  row.names = FALSE, 
  col.names = FALSE, 
  sep = "\t"
)
cat("Trait file saved for", trait, "with", nrow(traits_res), "entries\n")

head(traits_res)

```

## 4b. Phylogenetic Tree of Selected Contrast Pairs

This section places the selected contrast pairs onto the phylogeny, highlighting the top/high extreme (red) and bottom/low extreme (green) species chosen for the trait contrast.

```{r selected_pairs_tree, fig.height=15, fig.width=15, dev="png", dpi=320, message=FALSE, warning=FALSE}

# Fan-layout tree with highlighted contrast species — mirrors the base fan
# tree in 2.Phenotype_exploration.Rmd (ggtree(..., layout="fan", open.angle=15,
# size=tree_line_width), square-sized via species_plot_profile()$diagnostic_size).
# A rectangular tree with one tip label per species doesn't scale past a
# handful of tips (labels overlap / the plot grows unreadably tall); the fan
# layout fixes that, and only the SELECTED contrast species get a text label
# (everyone else just gets a coloured tip point) to keep it legible.
n_tips <- length(pruned_tree$tip.label)
tree_profile <- species_plot_profile(n_species = n_tips)$tree

dyn_size <- tree_profile$diagnostic_size
tip_label_size <- clamp_value(6.5 - 0.015 * n_tips, 3.5, 6.5)
tree_line_size <- tree_profile$tree_line_width * 1.8
# Offset scales with total tree depth (root-to-tip branch-length units) rather
# than a fixed constant, since chronogram vs phylogram branch-length scales
# differ by orders of magnitude across traits/datasets -- a fixed offset would
# sit on the tips for one tree and float off-panel for another.
tip_label_offset <- max(ape::node.depth.edgelength(pruned_tree)) * 0.03

knitr::opts_current$set(fig.width = dyn_size, fig.height = dyn_size)

tree_tip_df <- tibble::tibble(species = pruned_tree$tip.label) %>%
  dplyr::left_join(
    traits_res %>% dplyr::select(species, contrast_group),
    by = "species"
  ) %>%
  dplyr::mutate(
    contrast_label = dplyr::case_when(
      contrast_group == 1 ~ "Top/High",
      contrast_group == 0 ~ "Bottom/Low",
      TRUE ~ "Not selected"
    ),
    # Precomputed logical (rather than comparing to a string literal inside
    # aes(subset=...)) — ggtree's aes_()/aes_string() compatibility shim mangles
    # a quoted string comparison there and throws a parse error.
    is_selected = contrast_label != "Not selected"
  )

# Attach tip metadata via tidytree::as.treedata
contrast_tree_tibble <- tidytree::as_tibble(pruned_tree) %>%
  dplyr::rename(species = label)
contrast_tree_with_labels <- dplyr::full_join(contrast_tree_tibble, tree_tip_df, by = "species")
contrast_tree_data <- tidytree::as.treedata(contrast_tree_with_labels)

selected_pairs_tree <- ggtree(contrast_tree_data, aes(color = contrast_label),
                              layout = "fan", open.angle = 15,
                              size = tree_line_size) +
  geom_tippoint(size = tree_line_size * 1.5) +
  # label = species: the tip tibble renames the tree's own `label` column to
  # `species` (matching report 2's convention) before rebuilding the treedata
  # object, so the treedata no longer has a `label` field for geom_tiplab's
  # default aes(label=label) to fall back on — it silently drew node-index
  # numbers instead of species names without this explicit mapping.
  # align=FALSE (report 2's own tiplab/cladelab usage never sets align=TRUE
  # either): the aligned-connector-line codepath (geom_segment2) errors under
  # this environment's ggtree/ggplot2 pairing (missing internal `is.waive`).
  geom_tiplab(aes(subset = is_selected, label = species),
              size = tip_label_size, offset = tip_label_offset,
              show.legend = FALSE) +
  scale_color_manual(
    name = "Contrast group",
    values = c("Top/High" = "red", "Bottom/Low" = "forestgreen", "Not selected" = "grey70"),
    breaks = c("Top/High", "Bottom/Low", "Not selected"),
    na.value = "grey70"
  ) +
  labs(title = paste("Selected contrast pairs for trait:", capitalized_trait)) +
  # Extra panel expansion (plain ggplot2 scales, not a ggtree-specific helper
  # -- ggtree's fan layout is Cartesian under the hood) so the offset tip
  # labels have room to render fully instead of being clipped at the device
  # edge; the default 5% expansion isn't enough once labels sit off the tips.
  scale_x_continuous(expand = expansion(mult = 0.18)) +
  scale_y_continuous(expand = expansion(mult = 0.18)) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 13),
    axis.title.y = element_text(margin = ggplot2::margin(l = 20, r = 20)),
    axis.title.x = element_text(margin = ggplot2::margin(t = 12)),
    plot.title = element_text(size = 20, hjust = 0, face = "bold", margin = ggplot2::margin(b = 6)),
    plot.margin = margin(20, 20, 20, 20)
  )

ggsave(
  file.path(traitfile_dir, "selected_pairs_tree.png"),
  selected_pairs_tree, device = "png", width = dyn_size, height = dyn_size, dpi = "retina"
)

selected_pairs_tree

```

## 5. Bootstrap Trait Files for Permutation Analysis

This section extracts species  for use in bootstrap/permutation testing.
```{r bootstrap_traitfiles, message=FALSE, warning=FALSE}

# Extract bootstrap traitfile, retaining count columns (n_trait/c_trait) if present
boot_cols <- c("species", trait)
if (exists("n_trait", inherits = TRUE) && nzchar(n_trait) && n_trait %in% names(trait_df)) {
  boot_cols <- c(boot_cols, n_trait)
}
if (exists("c_trait", inherits = TRUE) && nzchar(c_trait) && c_trait %in% names(trait_df)) {
  boot_cols <- c(boot_cols, c_trait)
}

bootstrap_traitfile <- trait_df %>%
  dplyr::select(all_of(boot_cols)) %>%
  dplyr::filter(!is.na(.data[[trait]]))

cat("Removed", nrow(trait_df) - nrow(bootstrap_traitfile), "species with NA trait values from bootstrap traitfile\n")

# Save bootstrap traitfile with headers so permulations.R can auto-detect count columns
boot_path <- file.path(bootstrap_traitfile_dir)
createDir(boot_path)
write.table(
  bootstrap_traitfile, 
  file.path(boot_path, "boot_traitfile.tab"), 
  quote = FALSE, 
  row.names = FALSE, 
  col.names = TRUE, 
  sep = "\t"
)

# Display summary
cat("Bootstrap traitfile:", nrow(bootstrap_traitfile), "species\n")

# Preview
head(bootstrap_traitfile)

```

---

## Summary

This report summarizes the results of the **Trait Analysis** analysis.

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — Trait Analysis Module*
*Date of generation: `r Sys.Date()`*
