---
title: "Independent Contrast Selection for Trait Analysis"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
params:
  trait_file: NULL
  tree_file: NULL
  output_dir: !r getwd()
  seed: ""
  clade_name: "clade"
  taxon_of_interest: "family"
  sp_colname: "species"
  traitname: "trait"
  max_fop: 100
  n_trait: ""
  c_trait: ""
  secondary_trait: ""
  branch_trait: ""
  pss_top_pct: "0.05"
  perm_strategy: "best_model"
  trait_type: ""
  max_contrasts: "0"
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

<!-- 4.Independent_contrasts.Rmd — Select independent contrast pairs and write the CT trait files. -->
<!-- PhyloPhere | subworkflows/TRAIT_ANALYSIS/local/ -->

```{r header}

#  ██████╗ ██╗  ██╗██╗   ██╗██╗      ██████╗ ██████╗ ██╗  ██╗███████╗██████╗ ███████╗
#  ██╔══██╗██║  ██║╚██╗ ██╔╝██║     ██╔═══██╗██╔══██╗██║  ██║██╔════╝██╔══██╗██╔════╝
#  ██████╔╝███████║ ╚████╔╝ ██║     ██║   ██║██████╔╝███████║█████╗  ██████╔╝█████╗
#  ██╔═══╝ ██╔══██║  ╚██╔╝  ██║     ██║   ██║██╔═══╝ ██╔══██║██╔══╝  ██╔══██╗██╔══╝
#  ██║     ██║  ██║   ██║   ███████╗╚██████╔╝██║     ██║  ██║███████╗██║  ██║███████╗
#  ╚═╝     ╚═╝  ╚═╝   ╚═╝   ╚══════╝ ╚═════╝ ╚═╝     ╚═╝  ╚═╝╚══════╝╚═╝  ╚═╝╚══════╝

## PHYLOPHERE: A Nextflow pipeline including a complete set
## of phylogenetic comparative tools and analyses for Phenome-Genome studies

### Github: https://github.com/nozerorma/caastools/nf-phylophere
### Author: Miguel Ramon (miguel.ramon@upf.edu)

#### File: 4.Independent_contrasts.Rmd

```

# Independent Contrast Selection for Neoplasia Analysis in Primates
This document outlines the process of selecting phylogenetically independent species pairs for trait analysis using CAASTools-paired.

Contrast selection runs through the shared selection core (`src/lean_contrast_selector.R`), the same code the CAAS permulation null uses. Candidate pairs pass the trait-type gate (count: non-overlapping Jeffreys intervals; ordinal: top level vs bottom level; continuous: the top `pss_top_pct` of hi>lo pairs by **Phylogenetic Shift Score (PSS)**, disabled when `pss_top_pct` is outside `(0,1)`) and are ranked by PSS under the AIC-selected OU/BM model (Fabio Barteri's phyloq engine, vendored in `src/pss_core.R`); patristic distance is the fallback key when no finite PSS is available. Per-species trait values and Jeffreys intervals come from `3.CI-composition.Rmd` (`pairwise_data.csv`).


## Setup and Directory Configuration

This section configures the working environment, sets directories, and loads necessary functions and libraries.

```{r setup, message=FALSE, warning=FALSE}

if (exists("params")) {
  tryCatch(assign("params", params, envir = .GlobalEnv), error = function(e) NULL)
}

# Call the setup function from commons.R
source("./src/commons.R")
setup_rmd()

# Fallback logger, used only when commons.R has not defined debug_log().
if (is.null(getOption("phylo_debug"))) {
  options(phylo_debug = TRUE)
}
if (!exists("phylo_debug_log", envir = .GlobalEnv)) {
  phylo_debug_log <- character()
}
if (!exists("debug_log", inherits = TRUE)) {
  debug_log <- function(...) {
    msg <- sprintf(...)
    phylo_debug_log <<- c(phylo_debug_log, msg)
    cat("[DEBUG] ", msg, "\n", sep = "")
  }
}

# ── Dependencies ──────────────────────────────────────────────────────────────

library(dplyr)
library(purrr)
library(tibble)
library(ggplot2)
library(readr)
library(reshape2)
library(ggrepel)
library(ggtree)
library(ggnewscale)
library(ggstar)
library(colorspace)
library(treeio)
library(tidytree)
library(phylolm)
library(ggtreeExtra)
library(geiger)
library(rphylopic)
library(paletteer)
library(patchwork)

# fop_pair_sel.f(), which sources the shared selection core (lean_contrast_selector.R).
source("./src/selection_algorithm.R")


# ── Labels and output directories ─────────────────────────────────────────────

color_palette <- paste0(clade_name, "_palette") # name of the clade palette defined in palettes.R
capitalized_taxon <- tools::toTitleCase(taxon_of_interest)
capitalized_trait <- tools::toTitleCase(gsub("_", " ", trait)) # underscores shown as spaces
capitalized_clade <- tools::toTitleCase(gsub("_", " ", clade_name))

createDir(traitfile_dir)
createDir(permulation_traitfile_dir)

```

## 2. Selection Context

The tree is pruned to the species with a trait value and dichotomised (zero-length edges set to 1e-8); patristic distances and the AIC-selected OU/BM model are computed on it once by `selection_context()`. Patristic distance is the sum of branch lengths separating two species.

```{r selection_context, message=FALSE, warning=FALSE}

# Per-species trait values, Jeffreys intervals and sample sizes written by
# 3.CI-composition.Rmd; pairwise_data.csv carries them in the columns of species1.
pairwise_path <- file.path(ci_dir, "pairwise_data.csv")
if (!file.exists(pairwise_path)) {
  stop("Missing pairwise data file: ", pairwise_path)
}
pairwise_all <- read.csv(pairwise_path, stringsAsFactors = FALSE)
sp_tab <- pairwise_all[!duplicated(pairwise_all$species1), ]
sel_trait <- setNames(as.numeric(sp_tab$trait1), sp_tab$species1)

use_count_data <- has.n && has.c
sel_ci_lb <- if (use_count_data) setNames(as.numeric(sp_tab$trait_ci_lb1), sp_tab$species1) else NULL
sel_ci_ub <- if (use_count_data) setNames(as.numeric(sp_tab$trait_ci_ub1), sp_tab$species1) else NULL
sel_n     <- if (use_count_data) setNames(as.numeric(sp_tab$n_population1), sp_tab$species1) else NULL
sel_ordinal <- !use_count_data && is_ordinal_trait(sel_trait, trait_type)

force_model <- if (tolower(perm_strategy) %in% c("bm", "ou")) toupper(perm_strategy) else NULL
sel_ctx <- selection_context(sel_trait, pruned_tree, force_model = force_model)
distance_matrix <- sel_ctx$D

cat("Evolutionary model:", sel_ctx$selected_model, "\n")
cat("Distance matrix dimensions:", dim(distance_matrix), "\n")
head(distance_matrix[, 1:5])

```

## 3. Independent Contrast Pair Selection

This section selects phylogenetically independent species pairs for CAAS analysis. Candidates are ranked by **PSS descending** (smallest patristic distance when no finite PSS is available), and `greedy_dunn_select()` walks that ranked list, admitting a pair only if it keeps the selected set phylogenetically independent (modified Dunn index ≥ 1).

Tie-breaking after the PSS (or distance) key: highest absolute trait difference (`abs_diff`) → highest combined pair `n` (count traits) → species names. PSS is compared at 4 decimals, so floating-point noise never decides a tie.

```{r pair_selection, message=FALSE, warning=FALSE}

# Per-species statistics written by 1.Dataset_exploration.Rmd.
stats_path <- file.path(species_distribution_dir, "trait_stats.csv")
if (!file.exists(stats_path)) {
  stop("Missing trait stats file: ", stats_path)
}
stats_df <- read.csv(stats_path, stringsAsFactors = FALSE)

# Canonical greedy selection (H1) and the FOP parallel hypotheses, from the shared core.
if (is.na(suppressWarnings(as.integer(seed_val)))) {
  stop("4.Independent_contrasts.Rmd: params$seed is required (the FOP harvest draws are seeded with it).")
}
fop_results <- fop_pair_sel.f(
  ctx = sel_ctx,
  ci_lb = sel_ci_lb, ci_ub = sel_ci_ub, n_vec = sel_n,
  ordinal = sel_ordinal,
  top_pct = pss_top_pct,
  max_contrasts = max_contrasts,
  max_fop = { .mf <- suppressWarnings(as.integer(params$max_fop)); if (is.na(.mf) || .mf < 1) 100L else .mf },
  seed = as.integer(seed_val)
)

selected_pairs <- fop_results$canon_pairs
hypotheses <- fop_results$hypotheses
summary_df <- fop_results$summary_df
species_domain <- fop_results$species_domain

cat("Selected", nrow(selected_pairs), "canonical independent pairs\n")
cat("Harvested", length(hypotheses), "FOP parallel independent hypotheses\n")

print(selected_pairs)
```

## 3b. Phylogenetic Shift Score (PSS) Analysis & Contrast Ranking

This section characterizes the candidate contrast pairs ranked by **Phylogenetic Shift Score (PSS)**, highlighting the pairs admitted into the canonical contrast set ($H_1$), the divergence profile of the extracted pairs, and the PSS quality across parallel hypotheses.

```{r pss_representation, fig.height=11, fig.width=16, dev="png", dpi=320, message=FALSE, warning=FALSE}
# ── PSS Candidate Ranking & Extracted Contrasts ────────────────────────────────

cand_pool <- if (exists("fop_results") && !is.null(fop_results$cand_df)) fop_results$cand_df else data.frame()
have_candidates <- nrow(cand_pool) > 0 && "pss_score" %in% names(cand_pool) && any(is.finite(cand_pool$pss_score))

if (have_candidates) {
  cand_pool <- cand_pool %>%
    dplyr::filter(is.finite(pss_score)) %>%
    dplyr::mutate(
      cand_rank = dplyr::row_number(),
      pair_name = paste0(species1, " vs ", species2)
    )

  # Match extracted canonical pairs
  selected_sigs <- paste(selected_pairs$species1, selected_pairs$species2, sep = " ~ ")
  cand_sigs <- paste(cand_pool$species1, cand_pool$species2, sep = " ~ ")
  cand_pool$is_extracted <- cand_sigs %in% selected_sigs
  cand_pool$pair_idx <- match(cand_sigs, selected_sigs)

  # Plot A: Candidate Rank vs PSS Score
  p_rank <- ggplot2::ggplot(cand_pool, ggplot2::aes(x = cand_rank, y = pss_score)) +
    ggplot2::geom_line(color = "grey65", linewidth = 0.8) +
    ggplot2::geom_point(ggplot2::aes(color = is_extracted, size = is_extracted), alpha = 0.85) +
    ggrepel::geom_text_repel(
      data = subset(cand_pool, is_extracted),
      ggplot2::aes(label = paste0("P", pair_idx, ": ", species1, "\nvs ", species2)),
      size = 3.6, fontface = "bold", family = "sans",
      box.padding = 0.6, point.padding = 0.4, max.overlaps = 25,
      segment.color = "grey40"
    ) +
    ggplot2::scale_color_manual(
      name = "Status:",
      values = c("TRUE" = "#E64B35", "FALSE" = "#4A6FA5"),
      labels = c("TRUE" = "Extracted (Canonical H1)", "FALSE" = "Candidate pool")
    ) +
    ggplot2::scale_size_manual(
      name = "Status:",
      values = c("TRUE" = 4.5, "FALSE" = 1.8),
      labels = c("TRUE" = "Extracted (Canonical H1)", "FALSE" = "Candidate pool")
    ) +
    ggplot2::labs(
      title = "A. Candidate Contrast Ranking by PSS",
      subtitle = "All eligible candidate pairs ordered by PSS descending; admitted canonical pairs highlighted",
      x = "Candidate Rank",
      y = "Phylogenetic Shift Score (PSS)"
    ) +
    ggplot2::theme_minimal(base_family = "sans") +
    ggplot2::theme(
      plot.title = ggplot2::element_text(size = 14, face = "bold"),
      plot.subtitle = ggplot2::element_text(size = 11, color = "grey30"),
      legend.position = "bottom",
      axis.title = ggplot2::element_text(size = 12, face = "bold"),
      axis.text = ggplot2::element_text(size = 10)
    )

  # Plot B: Extracted Canonical Pairs Divergence (PSS Lollipop)
  if (nrow(selected_pairs) > 0 && "pss_score" %in% names(selected_pairs)) {
    sel_plot_df <- selected_pairs %>%
      dplyr::mutate(
        pair_num = dplyr::row_number(),
        pair_label = factor(
          paste0("P", pair_num, ": ", species1, " vs ", species2),
          levels = rev(paste0("P", pair_num, ": ", species1, " vs ", species2))
        )
      )

    p_extracted <- ggplot2::ggplot(sel_plot_df, ggplot2::aes(x = pss_score, y = pair_label)) +
      ggplot2::geom_segment(ggplot2::aes(x = 0, xend = pss_score, y = pair_label, yend = pair_label),
                            color = "grey50", linewidth = 1.0) +
      ggplot2::geom_point(ggplot2::aes(fill = pss_score), size = 5.5, shape = 21, color = "black", stroke = 1.2) +
      ggplot2::geom_text(ggplot2::aes(label = sprintf("%.4f", pss_score)),
                         hjust = -0.3, fontface = "bold", size = 3.8, family = "sans") +
      ggplot2::scale_fill_gradient(low = "#FFD166", high = "#E64B35", name = "PSS:") +
      ggplot2::scale_x_continuous(expand = ggplot2::expansion(mult = c(0.02, 0.25))) +
      ggplot2::labs(
        title = "B. Extracted Canonical Contrasts (H1)",
        subtitle = "PSS of admitted phylogenetically independent pairs",
        x = "Phylogenetic Shift Score (PSS)",
        y = NULL
      ) +
      ggplot2::theme_minimal(base_family = "sans") +
      ggplot2::theme(
        plot.title = ggplot2::element_text(size = 14, face = "bold"),
        plot.subtitle = ggplot2::element_text(size = 11, color = "grey30"),
        legend.position = "none",
        axis.title = ggplot2::element_text(size = 12, face = "bold"),
        axis.text.y = ggplot2::element_text(size = 10, face = "bold", color = "black"),
        axis.text.x = ggplot2::element_text(size = 10)
      )

    # Plot C: Trait Difference vs Patristic Distance
    p_tradeoff <- ggplot2::ggplot(sel_plot_df, ggplot2::aes(x = distance, y = abs_diff)) +
      ggplot2::geom_point(ggplot2::aes(fill = pss_score), size = 5.5, shape = 21, color = "black", stroke = 1.2) +
      ggrepel::geom_text_repel(
        ggplot2::aes(label = paste0("Pair ", pair_num)),
        size = 3.8, fontface = "bold", family = "sans", box.padding = 0.5
      ) +
      ggplot2::scale_fill_gradient(low = "#FFD166", high = "#E64B35", name = "PSS:") +
      ggplot2::labs(
        title = "C. Trait Divergence vs Patristic Distance",
        subtitle = "Optimal contrasts balance phenotype difference against phylogenetic distance",
        x = "Patristic Distance",
        y = "Absolute Trait Difference (|Δ|)"
      ) +
      ggplot2::theme_minimal(base_family = "sans") +
      ggplot2::theme(
        plot.title = ggplot2::element_text(size = 14, face = "bold"),
        plot.subtitle = ggplot2::element_text(size = 11, color = "grey30"),
        legend.position = "right",
        axis.title = ggplot2::element_text(size = 12, face = "bold"),
        axis.text = ggplot2::element_text(size = 10)
      )

    p_pss_grid <- (p_rank) / (p_extracted | p_tradeoff) +
      patchwork::plot_layout(heights = c(1.1, 1.0))
  } else {
    p_pss_grid <- p_rank
  }

  ggplot2::ggsave(
    file.path(traitfile_dir, "pss_contrast_ranking.png"),
    p_pss_grid, device = "png", width = 16, height = 11, dpi = "retina"
  )
  p_pss_grid
} else {
  cat("PSS candidate scores not available for ranking visualization.\n")
}
```

```{r pss_extracted_table, message=FALSE, warning=FALSE}
# Formatted summary table of extracted canonical pairs
if (nrow(selected_pairs) > 0) {
  table_df <- selected_pairs %>%
    dplyr::mutate(
      Pair = dplyr::row_number(),
      `Species 1 (Top/High)` = species1,
      `Species 2 (Bottom/Low)` = species2,
      `Trait Difference (|Δ|)` = sprintf("%.4f", abs_diff),
      `Patristic Distance` = sprintf("%.4f", distance),
      `PSS Score` = if ("pss_score" %in% names(.)) sprintf("%.4f", pss_score) else "NA",
      `Dunn Index` = if ("Dunn_index" %in% names(.)) sprintf("%.4f", Dunn_index) else "NA"
    ) %>%
    dplyr::select(Pair, `Species 1 (Top/High)`, `Species 2 (Bottom/Low)`,
                  `Trait Difference (|Δ|)`, `Patristic Distance`, `PSS Score`, `Dunn Index`)

  knitr::kable(table_df, caption = "Summary of Canonical Independent Contrast Pairs (H1)")
}
```


## 4. Trait File Generation

This section generates trait files in the format required for CAAS discovery, assigning contrast groups (high/low extremes) to each species across all harvested parallel hypotheses.

`fop_pair_sel.f` partitions the species into one Voronoi domain per canonical pair and harvests alternative Dunn-independent K-pair configurations from those domains. When the domain product is small the space is swept exhaustively; larger spaces fall back to capped random draws seeded with the pipeline seed. The harvested hypotheses are then ranked and `H2 … Hn` are numbered by **decreasing minimum PSS across their K pairs** (then mean PSS, then overall Dunn, then species composition), so the post-`H1` order is a quality order and the `max_fop` cap is a quality truncation. `contrast_hypotheses_summary.tsv` carries both `min_pss_score` and `mean_pss_score` per hypothesis.

```{r traitfile_generation, message=FALSE, warning=FALSE}

# Traitfile rows of a pair table (species1 = high extreme, species2 = low extreme):
# species, contrast_group (1 = high, 0 = low) and pair (running pair number).
gen_traitfiles <- function(df) {
  df <- df %>%
    tibble::rownames_to_column(var = "pair_id") %>%
    dplyr::mutate(rank = dplyr::row_number()) %>%
    tidyr::pivot_longer(
      cols = c(species1, species2), 
      names_to = "global_label", 
      values_to = "species"
    ) %>%
    dplyr::mutate(
      global_label = ifelse(
        global_label == "species1", 
        "high_extreme", 
        "low_extreme"
      )
    )
  
  filtered_species <- df %>%
    dplyr::group_by(global_label) %>%
    dplyr::select(species, global_label) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(
      contrast_group = case_when(
        global_label == "high_extreme" ~ 1,
        global_label == "low_extreme" ~ 0,
        TRUE ~ NA_integer_
      )
    ) %>%
    dplyr::select(-global_label) %>%
    dplyr::mutate(pair = rep(1:(nrow(df)/2), each = 2)) %>%
    dplyr::arrange(contrast_group, pair)
  
  return(filtered_species)
}

traits_res <- gen_traitfiles(selected_pairs)

# ── Trait files ───────────────────────────────────────────────────────────────

trait_path <- file.path(traitfile_dir)
createDir(trait_path)

# traitfile.tab: the canonical pairs (H1), without header.
write.table(  
  traits_res, 
  file.path(trait_path, "traitfile.tab"), 
  quote = FALSE, 
  row.names = FALSE, 
  col.names = FALSE, 
  sep = "\t"
)

# One traitfile_H{m}.tab per hypothesis, plus traitfile_fop.tab (with header), which stacks
# all of them with a hypothesis_id column.
fop_master_rows <- list()

for (h_id in names(hypotheses)) {
  h_df <- hypotheses[[h_id]]
  h_traits_res <- gen_traitfiles(h_df)
  
  write.table(
    h_traits_res,
    file.path(trait_path, paste0("traitfile_", h_id, ".tab")),
    quote = FALSE,
    row.names = FALSE,
    col.names = FALSE,
    sep = "\t"
  )
  
  h_traits_res$hypothesis_id <- h_id
  fop_master_rows[[h_id]] <- h_traits_res
}

if (length(fop_master_rows) > 0) {
  fop_master_df <- do.call(rbind, fop_master_rows)
  write.table(
    fop_master_df,
    file.path(trait_path, "traitfile_fop.tab"),
    quote = FALSE,
    row.names = FALSE,
    col.names = TRUE,
    sep = "\t"
  )
}

# ── Pairs per hypothesis ──────────────────────────────────────────────────────

# contrast_hypotheses_pairs.tsv has one row per (hypothesis, Voronoi domain). `pair` is the
# domain id: the FOP harvest draws one pair per domain and numbers the pairs in domain order
# (the `cluster` column of lean_contrast_selector.R). The scoring steps (CT_DISAMBIGUATION,
# SCORING) read `pair` as the domain id and `pss_score` as the weight of that domain.
# H1 is the canonical baseline; H2 to Hn are the PSS-ranked alternatives.
pair_rows <- lapply(names(hypotheses), function(h_id) {
  hd <- hypotheses[[h_id]]
  if (is.null(hd) || nrow(hd) == 0) return(NULL)
  data.frame(
    hypothesis_id = h_id,
    pair          = if ("cluster" %in% names(hd)) as.integer(hd$cluster) else seq_len(nrow(hd)),
    species1      = hd$species1,
    species2      = hd$species2,
    distance      = if ("distance"  %in% names(hd)) hd$distance  else NA_real_,
    abs_diff      = if ("abs_diff"  %in% names(hd)) hd$abs_diff  else NA_real_,
    pss_score     = if ("pss_score" %in% names(hd)) hd$pss_score else NA_real_,
    stringsAsFactors = FALSE
  )
})
contrast_pairs_df <- do.call(rbind, pair_rows)
if (!is.null(contrast_pairs_df) && nrow(contrast_pairs_df) > 0) {
  write.table(
    contrast_pairs_df,
    file.path(trait_path, "contrast_hypotheses_pairs.tsv"),
    quote = FALSE,
    row.names = FALSE,
    col.names = TRUE,
    sep = "\t"
  )
  cat(sprintf("Wrote contrast_hypotheses_pairs.tsv (%d hypothesis x domain rows)\n",
              nrow(contrast_pairs_df)))
}

# Summary of the hypotheses (min_pss_score, mean_pss_score, Dunn index, ...).
if (nrow(summary_df) > 0) {
  write.csv(
    summary_df,
    file.path(trait_path, "contrast_hypotheses_summary.csv"),
    row.names = FALSE
  )
  write.table(
    summary_df,
    file.path(trait_path, "contrast_hypotheses_summary.tsv"),
    row.names = FALSE,
    sep = "\t"
  )
}

cat("Trait files saved for", trait, "with", length(hypotheses), "FOP parallel hypotheses generated\n")

head(traits_res)


```

## 4b. Phylogenetic Tree of Selected Contrast Pairs

This section places the selected contrast pairs onto the phylogeny, highlighting the top/high extreme (red) and bottom/low extreme (green) species chosen for the trait contrast.

```{r selected_pairs_tree, fig.height=18, fig.width=18, dev="png", dpi=320, message=FALSE, warning=FALSE}

# Fan-layout tree. Only the selected contrast species carry a label and a tip point; a
# rectangular tree with one label per species is unreadable beyond a few tips.
n_tips <- length(pruned_tree$tip.label)
tree_profile <- species_plot_profile(n_species = n_tips)$tree

dyn_size <- tree_profile$diagnostic_size
# Tree reduced to the species of the trait table.
viz_species <- intersect(pruned_tree$tip.label, trait_df$species)
viz_tree <- if (length(viz_species) > 0) ape::keep.tip(pruned_tree, viz_species) else pruned_tree

branch_line_size <- max(0.85, tree_profile$tree_line_width * 1.5)
tip_label_size <- 5.8
max_depth <- max(ape::node.depth.edgelength(viz_tree))
# Offset so tip marker and label text never collide
tip_label_offset <- max_depth * 0.020

contrast_df <- traits_res

tree_tip_df <- tibble::tibble(species = viz_tree$tip.label) %>%
  dplyr::left_join(contrast_df, by = "species") %>%
  dplyr::mutate(
    contrast_label = dplyr::case_when(
      contrast_group == 1 ~ "Top/High",
      contrast_group == 0 ~ "Bottom/Low",
      TRUE ~ "Not selected"
    ),
    is_selected = contrast_label != "Not selected"
  )

# Attach contrast annotation with %<+% to preserve phylo branch lengths
selected_pairs_tree <- ggtree(viz_tree, layout = "fan", open.angle = 15,
                              size = branch_line_size) %<+% tree_tip_df +
  aes(color = contrast_label) +
  geom_tippoint(aes(subset = is_selected), size = branch_line_size * 4.0) +
  geom_tiplab(aes(subset = is_selected, label = label),
              size = tip_label_size, offset = tip_label_offset,
              fontface = "bold",
              show.legend = FALSE) +
  scale_color_manual(
    name = "Contrast group:",
    values = c("Top/High" = "red", "Bottom/Low" = "forestgreen", "Not selected" = "grey75"),
    breaks = c("Top/High", "Bottom/Low", "Not selected"),
    na.value = "grey75"
  ) +
  labs(title = paste("Selected contrast pairs for trait:", capitalized_trait)) +
  # Expand radial limits so the tree is smaller relative to canvas, leaving room for labels
  expand_limits(x = max_depth * 1.30) +
  theme_tree() +
  theme(
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    legend.position = "bottom",
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    plot.title = element_text(size = 20, hjust = 0.5, face = "bold", margin = ggplot2::margin(t = 15, b = 25)),
    legend.box.margin = margin(t = 30, b = 15),
    plot.margin = margin(25, 30, 25, 30)
  )

ggsave(
  file.path(traitfile_dir, "selected_pairs_tree.png"),
  selected_pairs_tree, device = "png", width = dyn_size, height = dyn_size, dpi = "retina"
)

selected_pairs_tree

# ── Species selected across all hypotheses ────────────────────────────────────

# Labels give the number of hypotheses in which each species is selected.
if (exists("fop_master_df") && !is.null(fop_master_df) && nrow(fop_master_df) > 0) {
  total_hypotheses <- length(unique(fop_master_df$hypothesis_id))
  canon_species <- unique(c(selected_pairs$species1, selected_pairs$species2))

  sp_hyp_summary <- fop_master_df %>%
    dplyr::group_by(species) %>%
    dplyr::summarise(
      n_hyp = dplyr::n_distinct(hypothesis_id),
      grp = dplyr::first(contrast_group),
      .groups = "drop"
    )

  tree_tip_df_all <- tibble::tibble(species = viz_tree$tip.label) %>%
    dplyr::left_join(sp_hyp_summary, by = "species") %>%
    dplyr::mutate(
      contrast_label = dplyr::case_when(
        grp == 1 ~ "Top/High",
        grp == 0 ~ "Bottom/Low",
        TRUE ~ "Not selected"
      ),
      is_selected = contrast_label != "Not selected",
      is_canonical = species %in% canon_species,
      display_label = dplyr::case_when(
        is_canonical ~ paste0("[H1] ", species, " (", contrast_label, " · ", n_hyp, "/", total_hypotheses, ")"),
        is_selected  ~ paste0(species, " (", contrast_label, " · ", n_hyp, "/", total_hypotheses, ")"),
        TRUE         ~ species
      )
    )

  dyn_size_all <- tree_profile$diagnostic_size
  tip_label_size_all <- 4.8
  tip_label_offset_all <- max_depth * 0.020

  # The data are attached to the plain phylo tree with %<+%. A treedata object built with
  # tidytree::as.treedata() from a joined tibble, as in the plot above, breaks later
  # ggtreeExtra::geom_fruit() layers ("replacement has 1 row, data has 0" in
  # ggtreeExtra:::set_mapping()).
  selected_pairs_tree_all <- ggtree(viz_tree, layout = "fan", open.angle = 15,
                                   size = branch_line_size) %<+% tree_tip_df_all +
    aes(color = contrast_label) +
    # Non-canonical selected species (points)
    geom_tippoint(aes(subset = is_selected & !is_canonical), size = branch_line_size * 3.5) +
    # Canonical H1 species: highlighted with gold diamond and bold outline
    geom_tippoint(aes(subset = is_canonical), shape = 23, size = branch_line_size * 5.5,
                  stroke = 1.4, fill = "#FFD700") +
    geom_tiplab(aes(subset = is_selected, label = display_label),
                size = tip_label_size_all, offset = tip_label_offset_all,
                fontface = "bold",
                show.legend = FALSE) +
    scale_color_manual(
      name = "Contrast group:",
      values = c("Top/High" = "red", "Bottom/Low" = "forestgreen", "Not selected" = "grey75"),
      breaks = c("Top/High", "Bottom/Low", "Not selected"),
      na.value = "grey75"
    )

  # Voronoi domains: `species_domain` (from fop_pair_sel.f) assigns every species to its
  # nearest canonical pair by patristic distance, the partition from which H2 to Hn are
  # drawn. A domain is not always a single clade, so a highlight at its MRCA could cover
  # the tips of a neighboring domain. Each domain is therefore split into its maximal
  # monophyletic blocks (nodes whose descendant tips all belong to the domain, without a
  # parent that also qualifies), and each block is highlighted separately with an
  # "encircle" shape. Blocks with no species selected in any hypothesis are not drawn.
  if (exists("species_domain") && !is.null(species_domain) && length(species_domain) > 0) {
    domain_in_tree <- species_domain[names(species_domain) %in% viz_tree$tip.label]

    if (length(domain_in_tree) > 0) {
      find_maximal_domain_blocks <- function(tree, sp_domain, keep_tips) {
        n_tip <- length(tree$tip.label)
        total_nodes <- n_tip + tree$Nnode
        children_of <- lapply(seq_len(total_nodes), function(nd) tree$edge[tree$edge[, 1] == nd, 2])
        desc_cache <- vector("list", total_nodes)
        get_desc_tips <- function(node) {
          if (!is.null(desc_cache[[node]])) return(desc_cache[[node]])
          d <- if (node <= n_tip) node else unlist(lapply(children_of[[node]], get_desc_tips))
          desc_cache[[node]] <<- d
          d
        }
        parent_of <- setNames(tree$edge[, 1], tree$edge[, 2])
        domain_ids <- sort(unique(sp_domain))
        setNames(lapply(domain_ids, function(d) {
          target_tips <- match(names(sp_domain)[sp_domain == d], tree$tip.label)
          full_nodes <- Filter(function(node) {
            dt <- get_desc_tips(node)
            length(dt) > 0 && all(dt %in% target_tips)
          }, seq_len(total_nodes))
          Filter(function(nd) {
            p <- parent_of[as.character(nd)]
            (is.na(p) || !(p %in% full_nodes)) && any(get_desc_tips(nd) %in% keep_tips)
          }, full_nodes)
        }), domain_ids)
      }

      selected_tip_idx <- match(tree_tip_df_all$species[tree_tip_df_all$is_selected],
                                viz_tree$tip.label)
      domain_blocks <- find_maximal_domain_blocks(viz_tree, domain_in_tree, selected_tip_idx)
      hilight_df <- dplyr::bind_rows(lapply(names(domain_blocks), function(d) {
        nds <- domain_blocks[[d]]
        if (length(nds) == 0) return(NULL)
        tibble::tibble(node = nds, domain = paste0("Domain ", d))
      }))

      if (!is.null(hilight_df) && nrow(hilight_df) > 0) {
        domain_levels <- paste0("Domain ", sort(unique(domain_in_tree)))
        hilight_df$domain <- factor(hilight_df$domain, levels = domain_levels)
        domain_palette <- setNames(scales::hue_pal()(length(domain_levels)), domain_levels)

        selected_pairs_tree_all <- selected_pairs_tree_all +
          ggtree::geom_hilight(
            data = hilight_df,
            mapping = aes(node = node, fill = domain),
            alpha = 0.25,
            type = "encircle",
            to.bottom = TRUE
          ) +
          scale_fill_manual(name = "Voronoi domain:", values = domain_palette)
      }
    }
  }

  selected_pairs_tree_all <- selected_pairs_tree_all +
    labs(
      title = paste("All Selected Species Across", total_hypotheses, "Parallel Hypotheses for Trait:", capitalized_trait),
      subtitle = "Gold diamond [H1] indicates Canonical H1 species. Selection frequency shown across FOP non-replacement hypotheses."
    ) +
    # Expand radial limits so the tree is smaller relative to canvas, leaving room for labels
    expand_limits(x = max_depth * 1.38) +
    theme_tree() +
    theme(
      panel.background = element_rect(fill = "transparent", colour = NA),
      plot.background = element_rect(fill = "transparent", colour = NA),
      legend.position = "bottom",
      legend.title = element_text(size = 16, face = "bold"),
      legend.text = element_text(size = 14),
      plot.title = element_text(size = 20, hjust = 0.5, face = "bold", margin = ggplot2::margin(t = 15, b = 6)),
      plot.subtitle = element_text(size = 15, hjust = 0.5, margin = ggplot2::margin(b = 30)),
      legend.box.margin = margin(t = 30, b = 15),
      plot.margin = margin(25, 30, 25, 30)
    )

  ggsave(
    file.path(traitfile_dir, "selected_pairs_tree_all_hypotheses.png"),
    selected_pairs_tree_all, device = "png", width = dyn_size_all, height = dyn_size_all, dpi = "retina"
  )

  selected_pairs_tree_all
}

```

## 4c. Contrast & Violin Plots — Selected Pairs Highlighted

This reproduces the trait-distribution plots from `2.Phenotype_exploration.Rmd`, adding a highlight for the species actually chosen as Top/High or Bottom/Low contrast pairs in `4b.` above -- placing the observed distribution and the resulting contrast set side by side, rather than having to cross-reference two separate reports.

```{r contrast_violin_highlighted, fig.height=10, fig.width=15, message=FALSE, warning=FALSE, dev='png', dpi=320}

# stats_df keeps its original column names, which fop_pair_sel.f and the trait-file
# generation use. The plots work on a copy with the value and taxa columns, joined to the
# canonical contrast assignment (traits_res) and all FOP hypotheses (fop_master_df)
# so both canonical H1 and secondary FOP species can be distinguished.
fop_species_summary <- if (exists("fop_master_df") && !is.null(fop_master_df) && nrow(fop_master_df) > 0) {
  fop_master_df %>%
    dplyr::group_by(species) %>%
    dplyr::summarise(
      n_hyp = dplyr::n_distinct(hypothesis_id),
      fop_grp = dplyr::first(contrast_group),
      .groups = "drop"
    )
} else {
  data.frame(species = character(), n_hyp = integer(), fop_grp = integer(), stringsAsFactors = FALSE)
}

canon_species_h1 <- unique(c(selected_pairs$species1, selected_pairs$species2))

stats_df_hl <- stats_df %>%
  dplyr::mutate(
    value = .data[[trait]],
    taxa = .data[[taxon_of_interest]]
  ) %>%
  dplyr::left_join(
    traits_res %>% dplyr::select(species, contrast_group),
    by = "species"
  ) %>%
  dplyr::left_join(
    fop_species_summary,
    by = "species"
  ) %>%
  dplyr::mutate(
    is_canonical = species %in% canon_species_h1,
    is_secondary = !is_canonical & !is.na(n_hyp) & n_hyp > 0,
    contrast_label = dplyr::case_when(
      is_canonical & contrast_group == 1 ~ "Top/High (H1)",
      is_canonical & contrast_group == 0 ~ "Bottom/Low (H1)",
      is_secondary & fop_grp == 1        ~ "Top/High (FOP secondary)",
      is_secondary & fop_grp == 0        ~ "Bottom/Low (FOP secondary)",
      TRUE                               ~ "Not selected"
    ),
    is_selected = contrast_label != "Not selected"
  )

ordered_species_hl <- tidytree::as_tibble(pruned_tree) %>%
  dplyr::arrange(desc(.data$node)) %>%
  # Tip nodes are numbered 1..Ntip (ape convention). Filtering on `!is.na(label)` alone
  # would also admit internal-node support labels, which repeat across nodes and give
  # factor(levels =) duplicates.
  dplyr::filter(.data$node <= length(pruned_tree$tip.label)) %>%
  dplyr::pull(.data$label)
stats_df_hl$species <- factor(stats_df_hl$species, levels = ordered_species_hl)

ordered_taxa_hl <- ordered_species_hl %>%
  sapply(function(sp) {
    idx <- which(stats_df_hl$species == sp)
    if (length(idx) == 0) return(NA_character_)
    stats_df_hl$taxa[idx[1]]
  }) %>%
  unname() %>%
  na.omit() %>%
  unique()

contrast_fill_values <- c(
  "Top/High (H1)"              = "red",
  "Bottom/Low (H1)"           = "forestgreen",
  "Top/High (FOP secondary)"   = "red",
  "Bottom/Low (FOP secondary)" = "forestgreen"
)

contrast_shape_values <- c(
  "Top/High (H1)"              = 21,  # filled circle with black border
  "Bottom/Low (H1)"           = 21,  # filled circle with black border
  "Top/High (FOP secondary)"   = 1,   # open/unfilled circle
  "Bottom/Low (FOP secondary)" = 1    # open/unfilled circle
)

# Resolve family/taxa palette matching Reports 1 & 2
taxa_palette_values <- resolve_taxa_palette(ordered_taxa_hl, palette_name = color_palette)
dark_palette_values <- get_palette_values(paste0("dark_", color_palette))
if (is.null(dark_palette_values)) dark_palette_values <- taxa_palette_values

# ── Contrast plot, highlighted ────────────────────────────────────────────────

# Trait values by taxon: background points colored by family; canonical H1 species
# are filled points (red/green) with black border; secondary FOP species are drawn as
# unfilled open circles (red/green).
contrast_plot_highlighted.f <- function(df, ordered_taxa) {
  plot_profile <- species_plot_profile(
    n_species = dplyr::n_distinct(df$species),
    n_taxa = dplyr::n_distinct(df$taxa)
  )
  contrast_dims <- plot_profile$contrast
  # Make distribution points larger and more legible
  base_pt_size <- max(3.5, contrast_dims$point_size * 1.5)
  h1_pt_size   <- base_pt_size * 1.6
  fop_pt_size  <- base_pt_size * 1.5
  global_median <- df$g_median[1]

  p <- ggplot2::ggplot(df, ggplot2::aes(x = value, y = taxa)) +
    ggplot2::scale_y_discrete(limits = rev(ordered_taxa)) +
    ggplot2::geom_point(
      ggplot2::aes(color = taxa), shape = 16, size = base_pt_size, alpha = 0.35
    )

  if (!is.null(taxa_palette_values) && length(taxa_palette_values) > 0) {
    p <- p + ggplot2::scale_color_manual(values = taxa_palette_values, guide = "none")
  } else {
    p <- p + ggplot2::scale_color_discrete(guide = "none")
  }

  p <- p +
    ggnewscale::new_scale_color() +
    # Secondary FOP species: open circles (unfilled, colored border)
    ggplot2::geom_point(
      data = subset(df, is_secondary),
      ggplot2::aes(color = contrast_label),
      shape = 1, stroke = 1.3, size = fop_pt_size
    ) +
    # Canonical H1 species: filled circles (shape 21, solid fill + black stroke)
    ggplot2::geom_point(
      data = subset(df, is_canonical),
      ggplot2::aes(fill = contrast_label),
      shape = 21, color = "black", stroke = 1.2, size = h1_pt_size
    ) +
    ggplot2::scale_color_manual(
      name = "Secondary (FOP):",
      values = c("Top/High (FOP secondary)" = "red", "Bottom/Low (FOP secondary)" = "forestgreen"),
      drop = FALSE
    ) +
    ggplot2::scale_fill_manual(
      name = "Canonical (H1):",
      values = c("Top/High (H1)" = "red", "Bottom/Low (H1)" = "forestgreen"),
      drop = FALSE
    ) +
    ggplot2::geom_vline(xintercept = global_median, linetype = "longdash", linewidth = 1.2, color = "grey30") +
    ggplot2::labs(
      x = capitalized_trait,
      y = paste(capitalized_taxon, "-", capitalized_clade),
      title = paste("Trait distribution with selected contrast pairs highlighted:", capitalized_trait),
      caption = "Global median: dashed line. Filled points: Canonical H1 pairs. Open circles: FOP secondary alternatives."
    ) +
    ggplot2::theme_minimal() +
    ggplot2::theme(
      plot.title = ggplot2::element_text(size = contrast_dims$title_size, hjust = 0, margin = ggplot2::margin(t = 20, b = 6), face = "bold"),
      axis.text.y = ggplot2::element_text(size = contrast_dims$axis_text_y),
      axis.title.x = ggplot2::element_text(size = contrast_dims$axis_title, margin = ggplot2::margin(t = 12)),
      axis.title.y = ggplot2::element_text(size = contrast_dims$axis_title, angle = 90, margin = ggplot2::margin(l = 20, r = 20)),
      plot.caption = ggplot2::element_text(size = contrast_dims$caption_size, hjust = 0, margin = ggplot2::margin(t = 20)),
      legend.text = ggplot2::element_text(size = 14),
      legend.title = ggplot2::element_text(size = 14, face = "bold"),
      legend.position = "bottom",
      legend.box = "horizontal"
    )
  p
}

contrast_hl_graph <- contrast_plot_highlighted.f(stats_df_hl, ordered_taxa_hl)
contrast_hl_dims <- species_plot_profile(
  n_species = dplyr::n_distinct(stats_df_hl$species),
  n_taxa = dplyr::n_distinct(stats_df_hl$taxa)
)$contrast
knitr::opts_current$set(fig.width = contrast_hl_dims$width, fig.height = contrast_hl_dims$height)
ggplot2::ggsave(
  file.path(traitfile_dir, paste0(trait, "_contrast_plot_highlighted.png")),
  contrast_hl_graph, device = "png", width = contrast_hl_dims$width, height = contrast_hl_dims$height, dpi = "retina"
)
contrast_hl_graph

```

```{r violin_highlighted, fig.height=10, fig.width=8, message=FALSE, warning=FALSE, dev='png', dpi=320}

# ── Violin plot, highlighted ──────────────────────────────────────────────────

# Violins and background jitter points respect the family palette of Reports 1 and 2.
# Canonical H1 species are filled circles (red/green) with black outline;
# Secondary FOP species are open circles (unfilled red/green).
violin_highlighted.f <- function(df, ordered_taxa) {
  plot_profile <- species_plot_profile(
    n_species = dplyr::n_distinct(df$species),
    n_taxa = dplyr::n_distinct(df$taxa)
  )
  violin_dims <- plot_profile$violin
  base_pt_size <- max(2.2, violin_dims$point_size * 1.3)
  h1_pt_size   <- base_pt_size * 1.6
  fop_pt_size  <- base_pt_size * 1.5
  median_value <- median(df$value, na.rm = TRUE)

  p <- ggplot2::ggplot(df, ggplot2::aes(x = value, y = taxa)) +
    ggplot2::scale_y_discrete(limits = rev(ordered_taxa))

  # Violin fill with family palette
  if (!is.null(dark_palette_values) && length(dark_palette_values) > 0) {
    p <- p +
      ggplot2::geom_violin(
        ggplot2::aes(fill = taxa), color = "black", width = 0.5, trim = TRUE, scale = "width", alpha = 0.20
      ) +
      ggplot2::scale_fill_manual(values = dark_palette_values, guide = "none") +
      ggnewscale::new_scale_fill() +
      ggplot2::geom_point(
        ggplot2::aes(color = taxa), shape = 16, size = base_pt_size, alpha = 0.35,
        position = ggplot2::position_jitter(height = violin_dims$jitter_height)
      ) +
      ggplot2::scale_color_manual(values = taxa_palette_values, guide = "none")
  } else {
    p <- p +
      ggplot2::geom_violin(
        ggplot2::aes(fill = taxa), color = "black", width = 0.5, trim = TRUE, scale = "width", alpha = 0.20
      ) +
      ggplot2::scale_fill_discrete(guide = "none") +
      ggnewscale::new_scale_fill() +
      ggplot2::geom_point(
        ggplot2::aes(color = taxa), shape = 16, size = base_pt_size, alpha = 0.35,
        position = ggplot2::position_jitter(height = violin_dims$jitter_height)
      ) +
      ggplot2::scale_color_discrete(guide = "none")
  }

  p <- p +
    ggnewscale::new_scale_color() +
    # Secondary FOP species: open circles (shape 1, unfilled, jittered)
    ggplot2::geom_point(
      data = subset(df, is_secondary),
      ggplot2::aes(color = contrast_label),
      shape = 1, stroke = 1.2, size = fop_pt_size,
      position = ggplot2::position_jitter(height = violin_dims$jitter_height)
    ) +
    # Canonical H1 species: filled circles (shape 21, jittered)
    ggplot2::geom_point(
      data = subset(df, is_canonical),
      ggplot2::aes(fill = contrast_label),
      shape = 21, color = "black", stroke = 1.1, size = h1_pt_size,
      position = ggplot2::position_jitter(height = violin_dims$jitter_height)
    ) +
    ggplot2::scale_color_manual(
      name = "Secondary (FOP):",
      values = c("Top/High (FOP secondary)" = "red", "Bottom/Low (FOP secondary)" = "forestgreen"),
      drop = FALSE,
      guide = ggplot2::guide_legend(nrow = 1, byrow = TRUE, order = 2)
    ) +
    ggplot2::scale_fill_manual(
      name = "Canonical (H1):",
      values = c("Top/High (H1)" = "red", "Bottom/Low (H1)" = "forestgreen"),
      drop = FALSE,
      guide = ggplot2::guide_legend(nrow = 1, byrow = TRUE, order = 1)
    ) +
    ggplot2::geom_vline(ggplot2::aes(xintercept = median_value), linewidth = 1, color = "grey30", alpha = 0.5) +
    ggplot2::labs(x = capitalized_trait) +
    ggplot2::theme_minimal() +
    ggplot2::theme(
      axis.text.y = ggplot2::element_text(size = violin_dims$axis_text_y, hjust = 0.5, face = "bold"),
      axis.title.x = ggplot2::element_text(size = violin_dims$axis_title, face = "bold", margin = ggplot2::margin(t = 12)),
      axis.title.y = ggplot2::element_blank(),
      legend.text = ggplot2::element_text(size = 12),
      legend.title = ggplot2::element_text(size = 12, face = "bold"),
      legend.position = "bottom",
      legend.box = "vertical",
      legend.spacing.y = grid::unit(0.15, "cm")
    )
  p
}

violin_hl_graph <- violin_highlighted.f(stats_df_hl, ordered_taxa_hl)
violin_hl_dims <- species_plot_profile(
  n_species = dplyr::n_distinct(stats_df_hl$species),
  n_taxa = dplyr::n_distinct(stats_df_hl$taxa)
)$violin
knitr::opts_current$set(fig.width = violin_hl_dims$width, fig.height = violin_hl_dims$height)
ggplot2::ggsave(
  file.path(traitfile_dir, paste0(trait, "_violin_plot_highlighted.png")),
  violin_hl_graph, device = "png", width = violin_hl_dims$width, height = violin_hl_dims$height, dpi = "retina"
)
violin_hl_graph

```

```{r permulation_traitfiles, include=FALSE, message=FALSE, warning=FALSE}

# ── Permulation trait file ────────────────────────────────────────────────────

# Species with a trait value, keeping the count columns (n_trait, c_trait) when present.
perm_cols <- c("species", trait)
if (exists("n_trait", inherits = TRUE) && nzchar(n_trait) && n_trait %in% names(trait_df)) {
  perm_cols <- c(perm_cols, n_trait)
}
if (exists("c_trait", inherits = TRUE) && nzchar(c_trait) && c_trait %in% names(trait_df)) {
  perm_cols <- c(perm_cols, c_trait)
}

permulation_traitfile <- trait_df %>%
  dplyr::select(all_of(perm_cols)) %>%
  dplyr::filter(!is.na(.data[[trait]]))

# Written with a header, so permulations.R can detect the count columns.
perm_path <- file.path(permulation_traitfile_dir)
createDir(perm_path)
write.table(
  permulation_traitfile,
  file.path(perm_path, "permulation_traitfile.tab"),
  quote = FALSE,
  row.names = FALSE,
  col.names = TRUE,
  sep = "\t"
)
```

---

## Summary

This report summarizes the results of the **Trait Analysis** analysis.

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — Trait Analysis Module*
*Date of generation: `r Sys.Date()`*
