---
title: "Independent Contrast Selection for Trait Analysis"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
params:
  trait_file: NULL
  tree_file: NULL
  output_dir: !r getwd()
  seed: ""
  clade_name: "clade"
  taxon_of_interest: "family"
  sp_colname: "species"
  traitname: "trait"
  max_fop: 100
  n_trait: ""
  c_trait: ""
  tax_id: ""
  secondary_trait: ""
  branch_trait: ""
  pss_top_pct: "0.05"
  perm_strategy: "best_model"
  trait_type: ""
  max_contrasts: "0"
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

```{r header}

#  ██████╗ ██╗  ██╗██╗   ██╗██╗      ██████╗ ██████╗ ██╗  ██╗███████╗██████╗ ███████╗
#  ██╔══██╗██║  ██║╚██╗ ██╔╝██║     ██╔═══██╗██╔══██╗██║  ██║██╔════╝██╔══██╗██╔════╝
#  ██████╔╝███████║ ╚████╔╝ ██║     ██║   ██║██████╔╝███████║█████╗  ██████╔╝█████╗
#  ██╔═══╝ ██╔══██║  ╚██╔╝  ██║     ██║   ██║██╔═══╝ ██╔══██║██╔══╝  ██╔══██╗██╔══╝
#  ██║     ██║  ██║   ██║   ███████╗╚██████╔╝██║     ██║  ██║███████╗██║  ██║███████╗
#  ╚═╝     ╚═╝  ╚═╝   ╚═╝   ╚══════╝ ╚═════╝ ╚═╝     ╚═╝  ╚═╝╚══════╝╚═╝  ╚═╝╚══════╝

## PHYLOPHERE: A Nextflow pipeline including a complete set
## of phylogenetic comparative tools and analyses for Phenome-Genome studies

### Github: https://github.com/nozerorma/caastools/nf-phylophere
### Author: Miguel Ramon (miguel.ramon@upf.edu)

#### File: Independent_contrasts.Rmd

```

# Independent Contrast Selection for Neoplasia Analysis in Primates
This document outlines the process of selecting phylogenetically independent species pairs for trait analysis using CAASTools-paired.

The eligible candidate pairs and their **Phylogenetic Shift Score (PSS)** are produced upstream in `3.CI-composition.Rmd`: each pair carries a `pss_score` computed under the AIC-selected OU/BM model (Fabio Barteri's phyloq engine, vendored in `src/pss_core.R`); for continuous traits only the top `pss_top_pct` slice of hi>lo pairs is kept unless `pss_top_pct` is set outside `(0,1)`, which disables the slice and keeps every hi>lo pair (PSS then ranks only). This report consumes that `pairwise_data.csv` and uses the PSS as the **primary ranking key** for the greedy Dunn selection below; patristic distance is only the fallback when no finite PSS is available (e.g. the model fit failed upstream).


## Setup and Directory Configuration

This section configures the working environment, sets directories, and loads necessary functions and libraries.

```{r setup, message=FALSE, warning=FALSE}

if (exists("params")) {
  tryCatch(assign("params", params, envir = .GlobalEnv), error = function(e) NULL)
}

# Call the setup function from commons.R
source("./src/commons.R")
setup_rmd()

# Debug helper (prints into HTML output)
if (is.null(getOption("phylo_debug"))) {
  options(phylo_debug = TRUE)
}
if (!exists("phylo_debug_log", envir = .GlobalEnv)) {
  phylo_debug_log <- character()
}
if (!exists("debug_log", inherits = TRUE)) {
  debug_log <- function(...) {
    msg <- sprintf(...)
    phylo_debug_log <<- c(phylo_debug_log, msg)
    cat("[DEBUG] ", msg, "\n", sep = "")
  }
}

# Load additional libraries
library(dplyr)
library(purrr)
library(tibble)
library(ggplot2)
library(readr)
library(reshape2)
library(ggrepel)
library(ggtree)
library(ggnewscale)
library(ggstar)
library(colorspace)
library(treeio)
library(tidytree)
library(phylolm)
library(ggtreeExtra)
library(geiger)
library(rphylopic)
library(paletteer)

# Load selection algorithm
source("./src/selection_algorithm.R")

color_palette <- paste0(clade_name, "_palette") # Color palette for the clade
capitalized_taxon <- tools::toTitleCase(taxon_of_interest) # Capitalized taxon name
capitalized_trait <- tools::toTitleCase(gsub("_", " ", trait)) # Capitalized and deconvoluted trait name
capitalized_clade <- tools::toTitleCase(gsub("_", " ", clade_name)) # Capitalized and deconvoluted clade name

createDir(traitfile_dir)
createDir(permulation_traitfile_dir)

```

## 2. Patristic Distance Calculation

This section calculates evolutionary distances between species using the phylogenetic tree. Patristic distance represents the sum of branch lengths separating two species.

```{r patristic_distances, message=FALSE, warning=FALSE}

# Calculate patristic distances using pruned tree
distance_matrix <- calculate_patristic_distances(
  tree = pruned_tree, 
  df = trait_df
)

# Display distance matrix preview
cat("Distance matrix dimensions:", dim(distance_matrix), "\n")
head(distance_matrix[, 1:5])

```

## 3. Independent Contrast Pair Selection

This section selects phylogenetically independent species pairs for CAAS analysis. Candidates are ranked by **Phylogenetic Shift Score (PSS) descending** (the per-pair trait-divergence signal under the AIC-selected OU/BM model, carried over from `3.CI-composition.Rmd`); when no finite PSS is available the ranking falls back to smallest patristic distance. The shared `rank_candidates()` / `greedy_dunn_select()` core then walks that ranked list, admitting a pair only if it keeps the selected set phylogenetically independent (Dunn index above threshold).

Tie-breaking after the PSS (or distance) key: highest absolute trait difference (`abs_diff`) → highest combined pair `n` (population size, when a numeric `p`/`n` column is present) → incoming order.

```{r pair_selection, message=FALSE, warning=FALSE}

# Load summary stats from Dataset Exploration
stats_path <- file.path(species_distribution_dir, "trait_stats.csv")
if (!file.exists(stats_path)) {
  stop("Missing trait stats file: ", stats_path)
}
stats_df <- read.csv(stats_path, stringsAsFactors = FALSE)

# Load pairwise overlap data from CI-composition step
pairwise_path <- file.path(ci_dir, "pairwise_data.csv")
if (!file.exists(pairwise_path)) {
  stop("Missing pairwise overlap file: ", pairwise_path)
}
pairwise_data <- read.csv(pairwise_path, stringsAsFactors = FALSE) %>%
  dplyr::filter(trait_overlap == FALSE)

# Run pair selection for both canonical greedy and FOP parallel hypotheses
fop_results <- fop_pair_sel.f(
  distance_matrix = distance_matrix,
  overlap_df = pairwise_data,
  traits_df = stats_df,
  my_trait = trait,
  max_fop = { .mf <- suppressWarnings(as.integer(params$max_fop)); if (is.na(.mf) || .mf < 1) 100L else .mf }
)

selected_pairs <- fop_results$canon_pairs
hypotheses <- fop_results$hypotheses
summary_df <- fop_results$summary_df
species_domain <- fop_results$species_domain

cat("Selected", nrow(selected_pairs), "canonical independent pairs\n")
cat("Harvested", length(hypotheses), "FOP parallel independent hypotheses\n")

# Display selected pairs
print(selected_pairs)
```


## 4. Trait File Generation

This section generates trait files in the format required for CAAS discovery, assigning contrast groups (high/low extremes) to each species across all harvested parallel hypotheses.

`fop_pair_sel.f` partitions the species into one Voronoi domain per canonical pair and harvests alternative Dunn-independent K-pair configurations from those domains. When the domain product is small the space is swept exhaustively; larger spaces fall back to capped random draws. The harvested hypotheses are then ranked and `H2 … Hn` are numbered by **decreasing minimum PSS across their K pairs** (then mean PSS, then overall Dunn), so the post-`H1` order is a quality order and the `max_fop` cap is a quality truncation. `contrast_hypotheses_summary.tsv` carries both `min_pss_score` and `mean_pss_score` per hypothesis.

```{r traitfile_generation, message=FALSE, warning=FALSE}

# Function to generate trait files from selected pairs
gen_traitfiles <- function(df) {
  df <- df %>%
    tibble::rownames_to_column(var = "pair_id") %>%
    dplyr::mutate(rank = dplyr::row_number()) %>%
    tidyr::pivot_longer(
      cols = c(species1, species2), 
      names_to = "global_label", 
      values_to = "species"
    ) %>%
    dplyr::mutate(
      global_label = ifelse(
        global_label == "species1", 
        "high_extreme", 
        "low_extreme"
      )
    )
  
  # Assign contrast groups
  filtered_species <- df %>%
    dplyr::group_by(global_label) %>%
    dplyr::select(species, global_label) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(
      contrast_group = case_when(
        global_label == "high_extreme" ~ 1,
        global_label == "low_extreme" ~ 0,
        TRUE ~ NA_integer_
      )
    ) %>%
    dplyr::select(-global_label) %>%
    dplyr::mutate(pair = rep(1:(nrow(df)/2), each = 2)) %>%
    dplyr::arrange(contrast_group, pair)
  
  return(filtered_species)
}

# Generate trait files for selected pairs
traits_res <- gen_traitfiles(selected_pairs)

# Save trait files directory
trait_path <- file.path(traitfile_dir)
createDir(trait_path)

# Canonical baseline traitfile.tab
write.table(  
  traits_res, 
  file.path(trait_path, "traitfile.tab"), 
  quote = FALSE, 
  row.names = FALSE, 
  col.names = FALSE, 
  sep = "\t"
)

# Write individual traitfile_H{m}.tab files and build master traitfile_fop.tab matrix
fop_master_rows <- list()

for (h_id in names(hypotheses)) {
  h_df <- hypotheses[[h_id]]
  h_traits_res <- gen_traitfiles(h_df)
  
  # Save individual hypothesis traitfile_H{m}.tab
  write.table(
    h_traits_res,
    file.path(trait_path, paste0("traitfile_", h_id, ".tab")),
    quote = FALSE,
    row.names = FALSE,
    col.names = FALSE,
    sep = "\t"
  )
  
  # Accumulate into master matrix
  h_traits_res$hypothesis_id <- h_id
  fop_master_rows[[h_id]] <- h_traits_res
}

if (length(fop_master_rows) > 0) {
  fop_master_df <- do.call(rbind, fop_master_rows)
  write.table(
    fop_master_df,
    file.path(trait_path, "traitfile_fop.tab"),
    quote = FALSE,
    row.names = FALSE,
    col.names = TRUE,
    sep = "\t"
  )
}

# ── Per-pair hypothesis table (drives domain-pooled scoring downstream) ───────
# One row per (hypothesis, Voronoi domain). `pair` == the domain id: the FOP
# harvest draws exactly one pair per domain and numbers them in domain order
# (selection_algorithm.R::fop_pair_sel.f, `hdf$cluster <- seq_len(K)`), and
# disambiguation enumerates each traitfile_H{m}.tab's pairs in that same order.
# So SCORING can key `mrca_<i>_path_score` (domain i's isolation score under
# hypothesis h) back to this pair's PSS weight via (hypothesis_id, pair).
# H1 is the canonical baseline; H2..Hn are PSS-ranked alternatives.
pair_rows <- lapply(names(hypotheses), function(h_id) {
  hd <- hypotheses[[h_id]]
  if (is.null(hd) || nrow(hd) == 0) return(NULL)
  data.frame(
    hypothesis_id = h_id,
    pair          = if ("cluster" %in% names(hd)) as.integer(hd$cluster) else seq_len(nrow(hd)),
    species1      = hd$species1,
    species2      = hd$species2,
    distance      = if ("distance"  %in% names(hd)) hd$distance  else NA_real_,
    abs_diff      = if ("abs_diff"  %in% names(hd)) hd$abs_diff  else NA_real_,
    pss_score     = if ("pss_score" %in% names(hd)) hd$pss_score else NA_real_,
    stringsAsFactors = FALSE
  )
})
contrast_pairs_df <- do.call(rbind, pair_rows)
if (!is.null(contrast_pairs_df) && nrow(contrast_pairs_df) > 0) {
  write.table(
    contrast_pairs_df,
    file.path(trait_path, "contrast_hypotheses_pairs.tsv"),
    quote = FALSE,
    row.names = FALSE,
    col.names = TRUE,
    sep = "\t"
  )
  cat(sprintf("Wrote contrast_hypotheses_pairs.tsv (%d hypothesis x domain rows)\n",
              nrow(contrast_pairs_df)))
}

# Save metadata summary
if (nrow(summary_df) > 0) {
  write.csv(
    summary_df,
    file.path(trait_path, "contrast_hypotheses_summary.csv"),
    row.names = FALSE
  )
  write.table(
    summary_df,
    file.path(trait_path, "contrast_hypotheses_summary.tsv"),
    row.names = FALSE,
    sep = "\t"
  )
}

cat("Trait files saved for", trait, "with", length(hypotheses), "FOP parallel hypotheses generated\n")

head(traits_res)


```

## 4b. Phylogenetic Tree of Selected Contrast Pairs

This section places the selected contrast pairs onto the phylogeny, highlighting the top/high extreme (red) and bottom/low extreme (green) species chosen for the trait contrast.

```{r selected_pairs_tree, fig.height=15, fig.width=15, dev="png", dpi=320, message=FALSE, warning=FALSE}

# Fan-layout tree with highlighted contrast species — mirrors the base fan
# tree in 2.Phenotype_exploration.Rmd (ggtree(..., layout="fan", open.angle=15,
# size=tree_line_width), square-sized via species_plot_profile()$diagnostic_size).
# A rectangular tree with one tip label per species doesn't scale past a
# handful of tips (labels overlap / the plot grows unreadably tall); the fan
# layout fixes that, and only the SELECTED contrast species get a text label
# (everyone else just gets a coloured tip point) to keep it legible.
n_tips <- length(pruned_tree$tip.label)
tree_profile <- species_plot_profile(n_species = n_tips)$tree

dyn_size <- tree_profile$diagnostic_size
# Prune tree to trait exploration species for clear visualization
viz_species <- intersect(pruned_tree$tip.label, trait_df$species)
viz_tree <- if (length(viz_species) > 0) ape::keep.tip(pruned_tree, viz_species) else pruned_tree

dyn_size_w <- 16
dyn_size_h <- 18.5
tip_label_size <- 3.8
tree_line_size <- 0.95
max_depth <- max(ape::node.depth.edgelength(viz_tree))
tip_label_offset <- max_depth * 0.006

contrast_df <- traits_res

tree_tip_df <- tibble::tibble(species = viz_tree$tip.label) %>%
  dplyr::left_join(contrast_df, by = "species") %>%
  dplyr::mutate(
    contrast_label = dplyr::case_when(
      contrast_group == 1 ~ "Top/High",
      contrast_group == 0 ~ "Bottom/Low",
      TRUE ~ "Not selected"
    ),
    is_selected = contrast_label != "Not selected"
  )

contrast_tree_tibble <- tidytree::as_tibble(viz_tree) %>% dplyr::rename(species = label)
contrast_tree_with_labels <- dplyr::full_join(contrast_tree_tibble, tree_tip_df, by = "species")
contrast_tree_data <- tidytree::as.treedata(contrast_tree_with_labels)

selected_pairs_tree <- ggtree(contrast_tree_data, aes(color = contrast_label),
                              layout = "fan", open.angle = 25,
                              size = tree_line_size) +
  geom_tippoint(aes(subset = is_selected), size = tree_line_size * 2.8) +
  geom_tiplab(aes(subset = is_selected, label = species),
              size = tip_label_size, offset = tip_label_offset,
              fontface = "bold",
              show.legend = FALSE) +
  scale_color_manual(
    name = "Contrast group:",
    values = c("Top/High" = "red", "Bottom/Low" = "forestgreen", "Not selected" = "grey75"),
    breaks = c("Top/High", "Bottom/Low", "Not selected"),
    na.value = "grey75"
  ) +
  labs(title = paste("Selected contrast pairs for trait:", capitalized_trait)) +
  # Bumped from 0.35 -- the tree itself (not the tip labels/points) was eating
  # too much of the canvas, leaving selected species' labels cramped against
  # the panel edge. hexpand/vexpand pad the coordinate system around the
  # tree's own bounding box; since only branches+tips are in that box, more
  # padding shrinks just the branches' footprint, leaving more room for labels.
  hexpand(0.55) +
  vexpand(0.55) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 15, face = "bold"),
    legend.text = element_text(size = 13),
    plot.title = element_text(size = 18, hjust = 0.5, face = "bold", margin = ggplot2::margin(t = 15, b = 25)),
    legend.box.margin = margin(t = 35, b = 15),
    plot.margin = margin(20, 25, 20, 25)
  )

ggsave(
  file.path(traitfile_dir, "selected_pairs_tree.png"),
  selected_pairs_tree, device = "png", width = dyn_size_w, height = dyn_size_h, dpi = "retina"
)

selected_pairs_tree

# ── Multi-hypothesis tree plot: highlight ALL species selected across hypotheses ──
if (exists("fop_master_df") && !is.null(fop_master_df) && nrow(fop_master_df) > 0) {
  total_hypotheses <- length(unique(fop_master_df$hypothesis_id))
  
  sp_hyp_summary <- fop_master_df %>%
    dplyr::group_by(species) %>%
    dplyr::summarise(
      n_hyp = dplyr::n_distinct(hypothesis_id),
      grp = dplyr::first(contrast_group),
      .groups = "drop"
    )
  
  tree_tip_df_all <- tibble::tibble(species = viz_tree$tip.label) %>%
    dplyr::left_join(sp_hyp_summary, by = "species") %>%
    dplyr::mutate(
      contrast_label = dplyr::case_when(
        grp == 1 ~ "Top/High",
        grp == 0 ~ "Bottom/Low",
        TRUE ~ "Not selected"
      ),
      is_selected = contrast_label != "Not selected",
      display_label = dplyr::if_else(
        is_selected,
        paste0(species, "\n(", contrast_label, " - ", n_hyp, "/", total_hypotheses, " hyp)"),
        species
      )
    )
    
  # Building the base plot from a pre-joined treedata object (tidytree::as.treedata()
  # on a full_join'd tip+internal-node tibble, as the first `selected_pairs_tree` plot
  # above does) breaks ggtreeExtra::geom_fruit() later in this environment -- it throws
  # "replacement has 1 row, data has 0" from deep inside ggtreeExtra:::set_mapping(),
  # reproducibly, on the treedata object alone (bare, any layout, no other layers).
  # ggtree's idiomatic %<+% data-attachment onto the plain phylo tree does not trigger
  # it (verified against this same viz_tree/tree_tip_df_all), so use that instead.
  selected_pairs_tree_all <- ggtree(viz_tree, layout = "fan", open.angle = 25,
                                   size = tree_line_size) %<+% tree_tip_df_all +
    aes(color = contrast_label) +
    geom_tippoint(aes(subset = is_selected), size = tree_line_size * 2.8) +
    geom_tiplab(aes(subset = is_selected, label = display_label),
                size = tip_label_size, offset = tip_label_offset,
                fontface = "bold",
                show.legend = FALSE) +
    scale_color_manual(
      name = "Contrast group:",
      values = c("Top/High" = "red", "Bottom/Low" = "forestgreen", "Not selected" = "grey75"),
      breaks = c("Top/High", "Bottom/Low", "Not selected"),
      na.value = "grey75"
    )

  # ── Voronoi domain highlight: which canonical-pair domain each species falls into ──
  # `species_domain` (fop_pair_sel.f, selection_algorithm.R) assigns EVERY species
  # to its nearest canonical pair by patristic distance — the partition that the
  # FOP harvest draws H2..Hn from. Domains aren't guaranteed to be single clades, so
  # one geom_hilight straight off each domain's MRCA could overshoot into whatever
  # sits between two non-monophyletic members and swallow a neighboring domain's
  # tips. Instead, decompose each domain into its maximal monophyletic sub-blocks
  # (recursing from the tips up, stopping the moment a node's full descendant set
  # would spill outside the domain) and highlight those individually as translucent
  # "encircle" blobs -- native ggtree, so this sidesteps ggtreeExtra entirely.
  if (exists("species_domain") && !is.null(species_domain) && length(species_domain) > 0) {
    domain_in_tree <- species_domain[names(species_domain) %in% viz_tree$tip.label]

    if (length(domain_in_tree) > 0) {
      find_maximal_domain_blocks <- function(tree, sp_domain) {
        n_tip <- length(tree$tip.label)
        total_nodes <- n_tip + tree$Nnode
        children_of <- lapply(seq_len(total_nodes), function(nd) tree$edge[tree$edge[, 1] == nd, 2])
        desc_cache <- vector("list", total_nodes)
        get_desc_tips <- function(node) {
          if (!is.null(desc_cache[[node]])) return(desc_cache[[node]])
          d <- if (node <= n_tip) node else unlist(lapply(children_of[[node]], get_desc_tips))
          desc_cache[[node]] <<- d
          d
        }
        parent_of <- setNames(tree$edge[, 1], tree$edge[, 2])
        domain_ids <- sort(unique(sp_domain))
        setNames(lapply(domain_ids, function(d) {
          target_tips <- match(names(sp_domain)[sp_domain == d], tree$tip.label)
          full_nodes <- Filter(function(node) {
            dt <- get_desc_tips(node)
            length(dt) > 0 && all(dt %in% target_tips)
          }, seq_len(total_nodes))
          Filter(function(nd) {
            p <- parent_of[as.character(nd)]
            is.na(p) || !(p %in% full_nodes)
          }, full_nodes)
        }), domain_ids)
      }

      domain_blocks <- find_maximal_domain_blocks(viz_tree, domain_in_tree)
      hilight_df <- dplyr::bind_rows(lapply(names(domain_blocks), function(d) {
        nds <- domain_blocks[[d]]
        if (length(nds) == 0) return(NULL)
        tibble::tibble(node = nds, domain = paste0("Domain ", d))
      }))

      if (!is.null(hilight_df) && nrow(hilight_df) > 0) {
        domain_levels <- paste0("Domain ", sort(unique(domain_in_tree)))
        hilight_df$domain <- factor(hilight_df$domain, levels = domain_levels)
        domain_palette <- setNames(scales::hue_pal()(length(domain_levels)), domain_levels)

        selected_pairs_tree_all <- selected_pairs_tree_all +
          ggtree::geom_hilight(
            data = hilight_df,
            mapping = aes(node = node, fill = domain),
            alpha = 0.25,
            type = "encircle",
            to.bottom = TRUE
          ) +
          scale_fill_manual(name = "Voronoi domain:", values = domain_palette)
      }
    }
  }

  selected_pairs_tree_all <- selected_pairs_tree_all +
    labs(
      title = paste("All Selected Species Across", total_hypotheses, "Parallel Hypotheses for Trait:", capitalized_trait),
      subtitle = "Includes selection frequency across FOP non-replacement hypotheses"
    ) +
    hexpand(0.55) +
    vexpand(0.55) +
    theme(
      legend.position = "bottom",
      legend.title = element_text(size = 15, face = "bold"),
      legend.text = element_text(size = 13),
      plot.title = element_text(size = 18, hjust = 0.5, face = "bold", margin = ggplot2::margin(t = 15, b = 4)),
      plot.subtitle = element_text(size = 14, hjust = 0.5, margin = ggplot2::margin(b = 35)),
      legend.box.margin = margin(t = 35, b = 15),
      plot.margin = margin(20, 25, 20, 25)
    )

  ggsave(
    file.path(traitfile_dir, "selected_pairs_tree_all_hypotheses.png"),
    selected_pairs_tree_all, device = "png", width = dyn_size_w, height = dyn_size_h, dpi = "retina"
  )

  selected_pairs_tree_all
}

```

## 4c. Contrast & Violin Plots — Selected Pairs Highlighted

This reproduces the trait-distribution plots from `2.Phenotype_exploration.Rmd`, adding a highlight for the species actually chosen as Top/High or Bottom/Low contrast pairs in `4b.` above -- placing the observed distribution and the resulting contrast set side by side, rather than having to cross-reference two separate reports.

```{r contrast_violin_highlighted, fig.height=10, fig.width=15, message=FALSE, warning=FALSE, dev='png', dpi=320}

# stats_df above stays in its raw (un-melted) column-name form for
# fop_pair_sel.f / the trait-file generation in section 4 -- build a melted
# copy here instead of mutating it in place, mirroring 2.Phenotype_
# exploration.Rmd's own trait_import chunk, and join in this section's
# contrast assignment (traits_res) so the highlight below shows WHERE the
# selected pairs actually sit in the distribution.
stats_df_hl <- stats_df %>%
  dplyr::mutate(
    value = .data[[trait]],
    taxa = .data[[taxon_of_interest]]
  ) %>%
  dplyr::left_join(
    traits_res %>% dplyr::select(species, contrast_group),
    by = "species"
  ) %>%
  dplyr::mutate(
    contrast_label = dplyr::case_when(
      contrast_group == 1 ~ "Top/High",
      contrast_group == 0 ~ "Bottom/Low",
      TRUE ~ "Not selected"
    ),
    is_selected = contrast_label != "Not selected"
  )

ordered_species_hl <- tidytree::as_tibble(pruned_tree) %>%
  dplyr::arrange(desc(.data$node)) %>%
  # Tip nodes are numbered 1..Ntip (ape convention); `!is.na(label)` alone
  # would also admit internal-node bootstrap-support labels, which repeat
  # across nodes on a real ML tree and can hand factor(levels=) duplicates
  # (see 2.Phenotype_exploration.Rmd's identical fix).
  dplyr::filter(.data$node <= length(pruned_tree$tip.label)) %>%
  dplyr::pull(.data$label)
stats_df_hl$species <- factor(stats_df_hl$species, levels = ordered_species_hl)

ordered_taxa_hl <- ordered_species_hl %>%
  sapply(function(sp) {
    idx <- which(stats_df_hl$species == sp)
    if (length(idx) == 0) return(NA_character_)
    stats_df_hl$taxa[idx[1]]
  }) %>%
  unname() %>%
  na.omit() %>%
  unique()

contrast_fill_values <- c("Top/High" = "red", "Bottom/Low" = "forestgreen")

### 1. Contrast plot, highlighted --------------------------------------------
contrast_plot_highlighted.f <- function(df, ordered_taxa) {
  plot_profile <- species_plot_profile(
    n_species = dplyr::n_distinct(df$species),
    n_taxa = dplyr::n_distinct(df$taxa)
  )
  contrast_dims <- plot_profile$contrast
  global_median <- df$g_median[1]

  ggplot2::ggplot(df, ggplot2::aes(x = value, y = taxa)) +
    ggplot2::scale_y_discrete(limits = rev(ordered_taxa)) +
    ggplot2::geom_point(ggplot2::aes(color = taxa), shape = 16, size = contrast_dims$point_size, alpha = 0.30) +
    ggplot2::scale_color_discrete(guide = "none") +
    ggplot2::geom_point(
      data = subset(df, is_selected),
      ggplot2::aes(fill = contrast_label),
      shape = 21, color = "black", stroke = 1.1, size = contrast_dims$point_size * 1.8
    ) +
    ggplot2::scale_fill_manual(name = "Contrast group:", values = contrast_fill_values) +
    ggplot2::geom_vline(xintercept = global_median, linetype = "longdash", linewidth = 1.2, color = "grey30") +
    ggplot2::labs(x = capitalized_trait,
                  y = paste(capitalized_taxon, "-", capitalized_clade),
                  title = paste("Trait distribution with selected contrast pairs highlighted:", capitalized_trait),
                  caption = "Global median: longdash grey line. Outlined points: species selected as Top/High or Bottom/Low contrast pairs.") +
    ggplot2::theme_minimal() +
    ggplot2::theme(
      plot.title = ggplot2::element_text(size = contrast_dims$title_size, hjust = 0, margin = ggplot2::margin(t = 20, b = 6), face = "bold"),
      axis.text.y = ggplot2::element_text(size = contrast_dims$axis_text_y),
      axis.title.x = ggplot2::element_text(size = contrast_dims$axis_title, margin = ggplot2::margin(t = 12)),
      axis.title.y = ggplot2::element_text(size = contrast_dims$axis_title, angle = 90, margin = ggplot2::margin(l = 20, r = 20)),
      plot.caption = ggplot2::element_text(size = contrast_dims$caption_size, hjust = 0, margin = ggplot2::margin(t = 20)),
      legend.text = ggplot2::element_text(size = 15),
      legend.position = "bottom"
    )
}

contrast_hl_graph <- contrast_plot_highlighted.f(stats_df_hl, ordered_taxa_hl)
contrast_hl_dims <- species_plot_profile(
  n_species = dplyr::n_distinct(stats_df_hl$species),
  n_taxa = dplyr::n_distinct(stats_df_hl$taxa)
)$contrast
knitr::opts_current$set(fig.width = contrast_hl_dims$width, fig.height = contrast_hl_dims$height)
ggplot2::ggsave(
  file.path(traitfile_dir, paste0(trait, "_contrast_plot_highlighted.png")),
  contrast_hl_graph, device = "png", width = contrast_hl_dims$width, height = contrast_hl_dims$height, dpi = "retina"
)
contrast_hl_graph

```

```{r violin_highlighted, fig.height=10, fig.width=8, message=FALSE, warning=FALSE, dev='png', dpi=320}

### 2. Violin plot, highlighted -----------------------------------------------
violin_highlighted.f <- function(df, ordered_taxa) {
  plot_profile <- species_plot_profile(
    n_species = dplyr::n_distinct(df$species),
    n_taxa = dplyr::n_distinct(df$taxa)
  )
  violin_dims <- plot_profile$violin
  median_value <- median(df$value, na.rm = TRUE)

  ggplot2::ggplot(df, ggplot2::aes(x = value, y = taxa)) +
    ggplot2::scale_y_discrete(limits = rev(ordered_taxa)) +
    ggplot2::geom_violin(ggplot2::aes(fill = taxa), color = "black", width = 0.5, trim = TRUE, scale = "width", alpha = 0.15) +
    ggplot2::scale_fill_discrete(guide = "none") +
    ggplot2::geom_point(
      ggplot2::aes(color = taxa), shape = 16, size = violin_dims$point_size, alpha = 0.30,
      position = ggplot2::position_jitter(height = violin_dims$jitter_height)
    ) +
    ggplot2::scale_color_discrete(guide = "none") +
    ggnewscale::new_scale_fill() +
    ggplot2::geom_point(
      data = subset(df, is_selected),
      ggplot2::aes(fill = contrast_label),
      shape = 21, color = "black", stroke = 1.1, size = violin_dims$point_size * 1.8,
      position = ggplot2::position_jitter(height = violin_dims$jitter_height)
    ) +
    ggplot2::scale_fill_manual(name = "Contrast group:", values = contrast_fill_values) +
    ggplot2::geom_vline(ggplot2::aes(xintercept = median_value), linewidth = 1, color = "grey30", alpha = 0.5) +
    ggplot2::labs(x = capitalized_trait) +
    ggplot2::theme_minimal() +
    ggplot2::theme(
      axis.text.y = ggplot2::element_text(size = violin_dims$axis_text_y, hjust = 0.5, face = "bold"),
      axis.title.x = ggplot2::element_text(size = violin_dims$axis_title, face = "bold", margin = ggplot2::margin(t = 12)),
      axis.title.y = ggplot2::element_blank(),
      legend.text = ggplot2::element_text(size = 15),
      legend.position = "bottom"
    )
}

violin_hl_graph <- violin_highlighted.f(stats_df_hl, ordered_taxa_hl)
violin_hl_dims <- species_plot_profile(
  n_species = dplyr::n_distinct(stats_df_hl$species),
  n_taxa = dplyr::n_distinct(stats_df_hl$taxa)
)$violin
knitr::opts_current$set(fig.width = violin_hl_dims$width, fig.height = violin_hl_dims$height)
ggplot2::ggsave(
  file.path(traitfile_dir, paste0(trait, "_violin_plot_highlighted.png")),
  violin_hl_graph, device = "png", width = violin_hl_dims$width, height = violin_hl_dims$height, dpi = "retina"
)
violin_hl_graph

```

## 5. Permulation Trait Files for Permutation Analysis


This section extracts species  for use in permulation testing.
```{r permulation_traitfiles, message=FALSE, warning=FALSE}

# Extract permulation traitfile, retaining count columns (n_trait/c_trait) if present
perm_cols <- c("species", trait)
if (exists("n_trait", inherits = TRUE) && nzchar(n_trait) && n_trait %in% names(trait_df)) {
  perm_cols <- c(perm_cols, n_trait)
}
if (exists("c_trait", inherits = TRUE) && nzchar(c_trait) && c_trait %in% names(trait_df)) {
  perm_cols <- c(perm_cols, c_trait)
}

permulation_traitfile <- trait_df %>%
  dplyr::select(all_of(perm_cols)) %>%
  dplyr::filter(!is.na(.data[[trait]]))

cat("Removed", nrow(trait_df) - nrow(permulation_traitfile), "species with NA trait values from permulation traitfile\n")

# Save permulation traitfile with headers so permulations.R can auto-detect count columns
perm_path <- file.path(permulation_traitfile_dir)
createDir(perm_path)
write.table(
  permulation_traitfile,
  file.path(perm_path, "permulation_traitfile.tab"),
  quote = FALSE,
  row.names = FALSE,
  col.names = TRUE,
  sep = "\t"
)

# Display summary
cat("Permulation traitfile:", nrow(permulation_traitfile), "species\n")

# Preview
head(permulation_traitfile)

```

---

## Summary

This report summarizes the results of the **Trait Analysis** analysis.

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — Trait Analysis Module*
*Date of generation: `r Sys.Date()`*
