---
title: "Independent Contrast Selection for Trait Analysis"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
params:
  trait_file: NULL
  tree_file: NULL
  output_dir: !r getwd()
  seed: ""
  clade_name: "clade"
  taxon_of_interest: "family"
  traitname: "trait"
  n_trait: ""
  c_trait: ""
  tax_id: ""
  secondary_trait: ""
  branch_trait: ""
  pss_top_pct: "0.05"
  perm_strategy: "best_model"
  trait_type: ""
  max_contrasts: "0"
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

```{r header}

#  ██████╗ ██╗  ██╗██╗   ██╗██╗      ██████╗ ██████╗ ██╗  ██╗███████╗██████╗ ███████╗
#  ██╔══██╗██║  ██║╚██╗ ██╔╝██║     ██╔═══██╗██╔══██╗██║  ██║██╔════╝██╔══██╗██╔════╝
#  ██████╔╝███████║ ╚████╔╝ ██║     ██║   ██║██████╔╝███████║█████╗  ██████╔╝█████╗
#  ██╔═══╝ ██╔══██║  ╚██╔╝  ██║     ██║   ██║██╔═══╝ ██╔══██║██╔══╝  ██╔══██╗██╔══╝
#  ██║     ██║  ██║   ██║   ███████╗╚██████╔╝██║     ██║  ██║███████╗██║  ██║███████╗
#  ╚═╝     ╚═╝  ╚═╝   ╚═╝   ╚══════╝ ╚═════╝ ╚═╝     ╚═╝  ╚═╝╚══════╝╚═╝  ╚═╝╚══════╝

## PHYLOPHERE: A Nextflow pipeline including a complete set
## of phylogenetic comparative tools and analyses for Phenome-Genome studies

### Github: https://github.com/nozerorma/caastools/nf-phylophere
### Author: Miguel Ramon (miguel.ramon@upf.edu)

#### File: Independent_contrasts.Rmd

```

# Independent Contrast Selection for Neoplasia Analysis in Primates
This document outlines the process of selecting phylogenetically independent species pairs for trait analysis using CAASTools-paired. 


## Setup and Directory Configuration

This section configures the working environment, sets directories, and loads necessary functions and libraries.

```{r setup, message=FALSE, warning=FALSE}

if (exists("params")) {
  tryCatch(assign("params", params, envir = .GlobalEnv), error = function(e) NULL)
}

# Call the setup function from commons.R
source("./src/commons.R")
setup_rmd()

# Debug helper (prints into HTML output)
if (is.null(getOption("phylo_debug"))) {
  options(phylo_debug = TRUE)
}
if (!exists("phylo_debug_log", envir = .GlobalEnv)) {
  phylo_debug_log <- character()
}
if (!exists("debug_log", inherits = TRUE)) {
  debug_log <- function(...) {
    msg <- sprintf(...)
    phylo_debug_log <<- c(phylo_debug_log, msg)
    cat("[DEBUG] ", msg, "\n", sep = "")
  }
}

# Load additional libraries
library(dplyr)
library(purrr)
library(tibble)
library(ggplot2)
library(readr)
library(reshape2)
library(ggrepel)
library(ggtree)
library(ggnewscale)
library(ggstar)
library(colorspace)
library(treeio)
library(tidytree)
library(phylolm)
library(ggtreeExtra)
library(geiger)
library(rphylopic)
library(paletteer)

# Load selection algorithm
source("./src/selection_algorithm.R")

color_palette <- paste0(clade_name, "_palette") # Color palette for the clade
capitalized_taxon <- tools::toTitleCase(taxon_of_interest) # Capitalized taxon name
capitalized_trait <- tools::toTitleCase(gsub("_", " ", trait)) # Capitalized and deconvoluted trait name
capitalized_clade <- tools::toTitleCase(gsub("_", " ", clade_name)) # Capitalized and deconvoluted clade name

createDir(traitfile_dir)
createDir(bootstrap_traitfile_dir)

```

## 2. Patristic Distance Calculation

This section calculates evolutionary distances between species using the phylogenetic tree. Patristic distance represents the sum of branch lengths separating two species.

```{r patristic_distances, message=FALSE, warning=FALSE}

# Calculate patristic distances using pruned tree
distance_matrix <- calculate_patristic_distances(
  tree = pruned_tree, 
  df = trait_df
)

# Display distance matrix preview
cat("Distance matrix dimensions:", dim(distance_matrix), "\n")
head(distance_matrix[, 1:5])

```

## 3. Independent Contrast Pair Selection

This section selects phylogenetically independent species pairs for CAAS analysis. The algorithm prioritizes pairs with small evolutionary distances but large trait differences, while ensuring phylogenetic independence between pairs using the Dunn index.

When the input `traits_df` contains a numeric `p` column (population size), tie-breaking now follows: highest Dunn index → highest absolute trait difference (`diff`) → highest combined `n` for the pair. If `n` is absent, behaviour is unchanged.

```{r pair_selection, message=FALSE, warning=FALSE}

# Load summary stats from Dataset Exploration
stats_path <- file.path(species_distribution_dir, "trait_stats.csv")
if (!file.exists(stats_path)) {
  stop("Missing trait stats file: ", stats_path)
}
stats_df <- read.csv(stats_path, stringsAsFactors = FALSE)

# Load pairwise overlap data from CI-composition step
pairwise_path <- file.path(ci_dir, "pairwise_data.csv")
if (!file.exists(pairwise_path)) {
  stop("Missing pairwise overlap file: ", pairwise_path)
}
pairwise_data <- read.csv(pairwise_path, stringsAsFactors = FALSE) %>%
  dplyr::filter(trait_overlap == FALSE)

# Run pair selection for both canonical greedy and FOP parallel hypotheses
fop_results <- fop_pair_sel.f(
  distance_matrix = distance_matrix,
  overlap_df = pairwise_data,
  traits_df = stats_df,
  my_trait = trait,
  max_fop = 100
)

selected_pairs <- fop_results$canon_pairs
hypotheses <- fop_results$hypotheses
summary_df <- fop_results$summary_df

cat("Selected", nrow(selected_pairs), "canonical independent pairs\n")
cat("Harvested", length(hypotheses), "FOP parallel independent hypotheses\n")

# Display selected pairs
print(selected_pairs)
```


## 4. Trait File Generation

This section generates trait files in the format required for CAAS discovery, assigning contrast groups (high/low extremes) to each species across all harvested parallel hypotheses.

```{r traitfile_generation, message=FALSE, warning=FALSE}

# Function to generate trait files from selected pairs
gen_traitfiles <- function(df) {
  df <- df %>%
    tibble::rownames_to_column(var = "pair_id") %>%
    dplyr::mutate(rank = dplyr::row_number()) %>%
    tidyr::pivot_longer(
      cols = c(species1, species2), 
      names_to = "global_label", 
      values_to = "species"
    ) %>%
    dplyr::mutate(
      global_label = ifelse(
        global_label == "species1", 
        "high_extreme", 
        "low_extreme"
      )
    )
  
  # Assign contrast groups
  filtered_species <- df %>%
    dplyr::group_by(global_label) %>%
    dplyr::select(species, global_label) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(
      contrast_group = case_when(
        global_label == "high_extreme" ~ 1,
        global_label == "low_extreme" ~ 0,
        TRUE ~ NA_integer_
      )
    ) %>%
    dplyr::select(-global_label) %>%
    dplyr::mutate(pair = rep(1:(nrow(df)/2), each = 2)) %>%
    dplyr::arrange(contrast_group, pair)
  
  return(filtered_species)
}

# Generate trait files for selected pairs
traits_res <- gen_traitfiles(selected_pairs)

# Save trait files directory
trait_path <- file.path(traitfile_dir)
createDir(trait_path)

# Canonical baseline traitfile.tab
write.table(  
  traits_res, 
  file.path(trait_path, "traitfile.tab"), 
  quote = FALSE, 
  row.names = FALSE, 
  col.names = FALSE, 
  sep = "\t"
)

# Write individual traitfile_H{m}.tab files and build master traitfile_fop.tab matrix
fop_master_rows <- list()

for (h_id in names(hypotheses)) {
  h_df <- hypotheses[[h_id]]
  h_traits_res <- gen_traitfiles(h_df)
  
  # Save individual hypothesis traitfile_H{m}.tab
  write.table(
    h_traits_res,
    file.path(trait_path, paste0("traitfile_", h_id, ".tab")),
    quote = FALSE,
    row.names = FALSE,
    col.names = FALSE,
    sep = "\t"
  )
  
  # Accumulate into master matrix
  h_traits_res$hypothesis_id <- h_id
  fop_master_rows[[h_id]] <- h_traits_res
}

if (length(fop_master_rows) > 0) {
  fop_master_df <- do.call(rbind, fop_master_rows)
  write.table(
    fop_master_df,
    file.path(trait_path, "traitfile_fop.tab"),
    quote = FALSE,
    row.names = FALSE,
    col.names = TRUE,
    sep = "\t"
  )
}

# Save metadata summary
if (nrow(summary_df) > 0) {
  write.csv(
    summary_df,
    file.path(trait_path, "contrast_hypotheses_summary.csv"),
    row.names = FALSE
  )
  write.table(
    summary_df,
    file.path(trait_path, "contrast_hypotheses_summary.tsv"),
    row.names = FALSE,
    sep = "\t"
  )
}

cat("Trait files saved for", trait, "with", length(hypotheses), "FOP parallel hypotheses generated\n")

head(traits_res)


```

## 4b. Phylogenetic Tree of Selected Contrast Pairs

This section places the selected contrast pairs onto the phylogeny, highlighting the top/high extreme (red) and bottom/low extreme (green) species chosen for the trait contrast.

```{r selected_pairs_tree, fig.height=15, fig.width=15, dev="png", dpi=320, message=FALSE, warning=FALSE}

# Fan-layout tree with highlighted contrast species — mirrors the base fan
# tree in 2.Phenotype_exploration.Rmd (ggtree(..., layout="fan", open.angle=15,
# size=tree_line_width), square-sized via species_plot_profile()$diagnostic_size).
# A rectangular tree with one tip label per species doesn't scale past a
# handful of tips (labels overlap / the plot grows unreadably tall); the fan
# layout fixes that, and only the SELECTED contrast species get a text label
# (everyone else just gets a coloured tip point) to keep it legible.
n_tips <- length(pruned_tree$tip.label)
tree_profile <- species_plot_profile(n_species = n_tips)$tree

dyn_size <- tree_profile$diagnostic_size
# Prune tree to trait exploration species for clear visualization
viz_species <- intersect(pruned_tree$tip.label, trait_df$species)
viz_tree <- if (length(viz_species) > 0) ape::keep.tip(pruned_tree, viz_species) else pruned_tree

dyn_size_w <- 16
dyn_size_h <- 18.5
tip_label_size <- 3.8
tree_line_size <- 0.95
max_depth <- max(ape::node.depth.edgelength(viz_tree))
tip_label_offset <- max_depth * 0.006

contrast_df <- traits_res

tree_tip_df <- tibble::tibble(species = viz_tree$tip.label) %>%
  dplyr::left_join(contrast_df, by = "species") %>%
  dplyr::mutate(
    contrast_label = dplyr::case_when(
      contrast_group == 1 ~ "Top/High",
      contrast_group == 0 ~ "Bottom/Low",
      TRUE ~ "Not selected"
    ),
    is_selected = contrast_label != "Not selected"
  )

contrast_tree_tibble <- tidytree::as_tibble(viz_tree) %>% dplyr::rename(species = label)
contrast_tree_with_labels <- dplyr::full_join(contrast_tree_tibble, tree_tip_df, by = "species")
contrast_tree_data <- tidytree::as.treedata(contrast_tree_with_labels)

selected_pairs_tree <- ggtree(contrast_tree_data, aes(color = contrast_label),
                              layout = "fan", open.angle = 25,
                              size = tree_line_size) +
  geom_tippoint(aes(subset = is_selected), size = tree_line_size * 2.8) +
  geom_tiplab(aes(subset = is_selected, label = species),
              size = tip_label_size, offset = tip_label_offset,
              fontface = "bold",
              show.legend = FALSE) +
  scale_color_manual(
    name = "Contrast group:",
    values = c("Top/High" = "red", "Bottom/Low" = "forestgreen", "Not selected" = "grey75"),
    breaks = c("Top/High", "Bottom/Low", "Not selected"),
    na.value = "grey75"
  ) +
  labs(title = paste("Selected contrast pairs for trait:", capitalized_trait)) +
  hexpand(0.35) +
  vexpand(0.35) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 15, face = "bold"),
    legend.text = element_text(size = 13),
    plot.title = element_text(size = 18, hjust = 0.5, face = "bold", margin = ggplot2::margin(t = 15, b = 25)),
    legend.box.margin = margin(t = 35, b = 15),
    plot.margin = margin(20, 25, 20, 25)
  )

ggsave(
  file.path(traitfile_dir, "selected_pairs_tree.png"),
  selected_pairs_tree, device = "png", width = dyn_size_w, height = dyn_size_h, dpi = "retina"
)

selected_pairs_tree

# ── Multi-hypothesis tree plot: highlight ALL species selected across hypotheses ──
if (exists("fop_master_df") && !is.null(fop_master_df) && nrow(fop_master_df) > 0) {
  total_hypotheses <- length(unique(fop_master_df$hypothesis_id))
  
  sp_hyp_summary <- fop_master_df %>%
    dplyr::group_by(species) %>%
    dplyr::summarise(
      n_hyp = dplyr::n_distinct(hypothesis_id),
      grp = dplyr::first(contrast_group),
      .groups = "drop"
    )
  
  tree_tip_df_all <- tibble::tibble(species = viz_tree$tip.label) %>%
    dplyr::left_join(sp_hyp_summary, by = "species") %>%
    dplyr::mutate(
      contrast_label = dplyr::case_when(
        grp == 1 ~ "Top/High",
        grp == 0 ~ "Bottom/Low",
        TRUE ~ "Not selected"
      ),
      is_selected = contrast_label != "Not selected",
      display_label = dplyr::if_else(
        is_selected,
        paste0(species, "\n(", contrast_label, " - ", n_hyp, "/", total_hypotheses, " hyp)"),
        species
      )
    )
    
  contrast_tree_with_labels_all <- dplyr::full_join(contrast_tree_tibble, tree_tip_df_all, by = "species")
  contrast_tree_data_all <- tidytree::as.treedata(contrast_tree_with_labels_all)

  selected_pairs_tree_all <- ggtree(contrast_tree_data_all, aes(color = contrast_label),
                                   layout = "fan", open.angle = 25,
                                   size = tree_line_size) +
    geom_tippoint(aes(subset = is_selected), size = tree_line_size * 2.8) +
    geom_tiplab(aes(subset = is_selected, label = display_label),
                size = tip_label_size, offset = tip_label_offset,
                fontface = "bold",
                show.legend = FALSE) +
    scale_color_manual(
      name = "Contrast group:",
      values = c("Top/High" = "red", "Bottom/Low" = "forestgreen", "Not selected" = "grey75"),
      breaks = c("Top/High", "Bottom/Low", "Not selected"),
      na.value = "grey75"
    ) +
    labs(
      title = paste("All Selected Species Across", total_hypotheses, "Parallel Hypotheses for Trait:", capitalized_trait),
      subtitle = "Includes selection frequency across FOP non-replacement hypotheses"
    ) +
    hexpand(0.35) +
    vexpand(0.35) +
    theme(
      legend.position = "bottom",
      legend.title = element_text(size = 15, face = "bold"),
      legend.text = element_text(size = 13),
      plot.title = element_text(size = 18, hjust = 0.5, face = "bold", margin = ggplot2::margin(t = 15, b = 4)),
      plot.subtitle = element_text(size = 14, hjust = 0.5, margin = ggplot2::margin(b = 35)),
      legend.box.margin = margin(t = 35, b = 15),
      plot.margin = margin(20, 25, 20, 25)
    )

  ggsave(
    file.path(traitfile_dir, "selected_pairs_tree_all_hypotheses.png"),
    selected_pairs_tree_all, device = "png", width = dyn_size_w, height = dyn_size_h, dpi = "retina"
  )

  selected_pairs_tree_all
}

```

## 5. Bootstrap Trait Files for Permutation Analysis


This section extracts species  for use in bootstrap/permutation testing.
```{r bootstrap_traitfiles, message=FALSE, warning=FALSE}

# Extract bootstrap traitfile, retaining count columns (n_trait/c_trait) if present
boot_cols <- c("species", trait)
if (exists("n_trait", inherits = TRUE) && nzchar(n_trait) && n_trait %in% names(trait_df)) {
  boot_cols <- c(boot_cols, n_trait)
}
if (exists("c_trait", inherits = TRUE) && nzchar(c_trait) && c_trait %in% names(trait_df)) {
  boot_cols <- c(boot_cols, c_trait)
}

bootstrap_traitfile <- trait_df %>%
  dplyr::select(all_of(boot_cols)) %>%
  dplyr::filter(!is.na(.data[[trait]]))

cat("Removed", nrow(trait_df) - nrow(bootstrap_traitfile), "species with NA trait values from bootstrap traitfile\n")

# Save bootstrap traitfile with headers so permulations.R can auto-detect count columns
boot_path <- file.path(bootstrap_traitfile_dir)
createDir(boot_path)
write.table(
  bootstrap_traitfile, 
  file.path(boot_path, "boot_traitfile.tab"), 
  quote = FALSE, 
  row.names = FALSE, 
  col.names = TRUE, 
  sep = "\t"
)

# Display summary
cat("Bootstrap traitfile:", nrow(bootstrap_traitfile), "species\n")

# Preview
head(bootstrap_traitfile)

```

---

## Summary

This report summarizes the results of the **Trait Analysis** analysis.

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — Trait Analysis Module*
*Date of generation: `r Sys.Date()`*
