---
title: "ASR Path Robustness Characterisation"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
params:
  disambig_dir:        NULL   # path to ct_disambiguation/ output directory
  posterior_threshold: 0.95   # params.ct_disambig_posterior_threshold
  output_dir:          "."    # working directory (TSVs → ./tsv/, plots → ./plots/)
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

<!--
  ASR Path Robustness Characterisation — PhyloPhere Pipeline
  ======================================================
  Evaluates the continuous `asr_path_score` and its subcomponents:
  (independence, mrca_diversity, derived_agreement, conservation_gate, core)
  assigned to sites during ASR evaluation.

  Author: Miguel Ramon (miguel.ramon@upf.edu)
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 11,
  fig.height = 6,
  dpi        = 180,
  dev        = "png"
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(ggplot2)
  library(patchwork)
  library(scales)
  library(knitr)
  library(kableExtra)
})

# ── Parameter validation ──────────────────────────────────────────────────────
disambig_dir        <- params$disambig_dir
posterior_threshold <- as.numeric(params$posterior_threshold)
output_dir          <- params$output_dir

stopifnot(
  !is.null(disambig_dir)
)

# ── Output directories ────────────────────────────────────────────────────────
tsv_dir   <- file.path(output_dir, "tsv")
plots_dir <- file.path(output_dir, "plots")
dir.create(tsv_dir,   recursive = TRUE, showWarnings = FALSE)
dir.create(plots_dir, recursive = TRUE, showWarnings = FALSE)

write_tsv_safe <- function(df, fname) {
  path <- file.path(tsv_dir, fname)
  readr::write_tsv(df, path)
  invisible(path)
}

save_plot <- function(p, fname, width = 11, height = 6) {
  path <- file.path(plots_dir, fname)
  ggsave(path, plot = p, width = width, height = height, dpi = 180)
  invisible(path)
}

# Color palette definition
.tableau10 <- c(
  "#1F77B4", "#AEC7E8", "#FF7F0E", "#FFBB78",
  "#2CA02C", "#98DF8A", "#D62728", "#FF9896",
  "#9467BD", "#C5B0D5"
)
```

## Introduction

With the introduction of the continuous `asr_path_score`, the ASR evaluations have fundamentally shifted from binary cut-offs to a holistic score based on a path-to-root walk. This report details the distribution of the `asr_path_score` across evaluated sites and breaks down the final score into its constituent penalty and reward components:

*   **Replication (`core`)**: P(≥2 independent changes) — combinatorial probability that ≥2 pairs changed independently in at least one phenotype direction (top or bottom).
*   **Independence at shared ancestors (`independence`)**: The probability that the derived state was absent at shared ancestors (LAC nodes).
*   **Parallel vs. non-parallel origins (`mrca_diversity`)**: How distinct the ancestral starting states are across participating pairs.
*   **Convergence agreement (`derived_agreement`)**: Coherence of derived states per phenotype side — penalises divergent derived residues.
*   **Conserved-pair confirmation (`conservation_gate`)**: Soft gate from conserved pairs: deep conservation confirms the contrast; drifted controls weaken it.

---

```{r load-data}
master_csv <- file.path(disambig_dir, "caas_convergence_master.csv")
if (!file.exists(master_csv)) {
  stop(sprintf("[asr_robustness] caas_convergence_master.csv not found in: %s", disambig_dir))
}

master_df <- readr::read_csv(master_csv, show_col_types = FALSE, guess_max = 5000)

# Normalize column names if necessary
if ("msa_pos" %in% names(master_df)) master_df <- dplyr::rename(master_df, site = msa_pos)
if (!("gene" %in% names(master_df)) && "Gene" %in% names(master_df))
  master_df <- dplyr::rename(master_df, gene = Gene)
if (!("site" %in% names(master_df)) && "Position" %in% names(master_df))
  master_df <- dplyr::rename(master_df, site = Position)

# Isolate scored rows
score_df <- master_df %>% filter(!is.na(asr_path_score))

if (nrow(score_df) == 0) {
  stop("No ASR path scores found in the provided CSV.")
}

comp_vars <- c("independence", "mrca_diversity", "derived_agreement", "conservation_gate", "core")
available_comps <- intersect(comp_vars, colnames(score_df))
```

> **Pipeline threshold** (`params.ct_disambig_posterior_threshold`): **`r posterior_threshold`**
>
> **Disambiguation directory**: `r disambig_dir`
>
> **Total convergent sites loaded**: `r nrow(master_df)`
>
> **ASR path scored sites**: `r nrow(score_df)`

---

## 1. Overall ASR Path Score Distribution

The overall `asr_path_score` aggregates both the private walk and shared-node checks. High scores indicate independent, clean mutations that are well-supported by ancestral reconstructions.

```{r plot-score-dist, fig.cap="Distribution of the final ASR Path Score across positions."}
p_score <- ggplot(score_df, aes(x = asr_path_score)) +
  geom_histogram(binwidth = 0.02, fill = .tableau10[1], colour = "white", alpha = 0.8) +
  geom_density(aes(y = after_stat(density) * 0.02 * nrow(score_df)), color = .tableau10[7], linewidth = 1) +
  theme_bw(base_size = 12) +
  labs(
    title = "Overall ASR Path Score Distribution",
    subtitle = sprintf("%s positions across %d genes",
                       formatC(nrow(score_df), big.mark = ","),
                       n_distinct(score_df$gene)),
    x = "ASR Path Score",
    y = "Count"
  )

print(p_score)
save_plot(p_score, "asr_path_score_distribution.png", width = 8, height = 5)
```

---

## 2. Component Distributions

The final path score is composed of several continuous subcomponents. We plot their density distributions below to understand which factors act as the main limiters or contributors to the final scores in this dataset.

```{r plot-comp-dists, fig.height=8, fig.cap="Density distributions of the individual components that make up the ASR Path Score."}
comp_df <- score_df %>%
  select(gene, site, all_of(available_comps)) %>%
  pivot_longer(cols = all_of(available_comps), names_to = "component", values_to = "value") %>%
  filter(!is.na(value))

p_comps <- ggplot(comp_df, aes(x = value, fill = component)) +
  geom_density(alpha = 0.6, colour = "grey30", linewidth = 0.3) +
  facet_wrap(~ component, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = .tableau10[seq_along(available_comps)]) +
  theme_bw(base_size = 12) +
  theme(legend.position = "none",
        strip.background = element_rect(fill = "grey95", colour = "grey80"),
        strip.text = element_text(face = "bold")) +
  labs(
    title = "Distributions of ASR Path Score Components",
    x = "Component Value",
    y = "Density"
  )

print(p_comps)
save_plot(p_comps, "asr_path_components_density.png", width = 10, height = 8)
```

---

## 3. Subcomponent Influence on Final Score

To visualize how each subcomponent influences the final `asr_path_score`, we plot their pairwise scatter plots. A linear model (red dashed line) is fitted to highlight global trends.

```{r plot-comp-scatters, fig.height=9, fig.width=11, fig.cap="Scatter plots comparing individual component values against the final ASR Path Score."}
scatter_list <- lapply(available_comps, function(comp) {
  ggplot(score_df, aes(x = .data[[comp]], y = asr_path_score)) +
    geom_jitter(alpha = 0.3, color = .tableau10[1], width = 0.01, height = 0) +
    geom_smooth(method = "lm", color = .tableau10[7], se = FALSE, linetype = "dashed", formula = y ~ x) +
    theme_bw(base_size = 11) +
    labs(x = comp, y = "Final Path Score")
})

p_scatter <- wrap_plots(scatter_list, ncol = 2) +
  plot_annotation(
    title = "Influence of Components on Final ASR Path Score",
    theme = theme(plot.title = element_text(size = 14, face = "bold"))
  )

print(p_scatter)
save_plot(p_scatter, "asr_path_components_scatter.png", width = 11, height = 9)
```

---

## 4. Summary Statistics

The table below lists the summary statistics for the final path score and each of its continuous subcomponents across all scored positions.

```{r summary-table}
summary_table <- score_df %>%
  select(asr_path_score, all_of(available_comps)) %>%
  summarise(across(everything(), list(
    Mean   = ~mean(., na.rm = TRUE),
    Median = ~median(., na.rm = TRUE),
    Min    = ~min(., na.rm = TRUE),
    Max    = ~max(., na.rm = TRUE),
    SD     = ~sd(., na.rm = TRUE)
  ))) %>%
  pivot_longer(everything(), names_to = c("Variable", "Stat"), names_sep = "_(?=[A-Za-z]+$)") %>%
  pivot_wider(names_from = Stat, values_from = value)

write_tsv_safe(summary_table, "asr_path_summary_statistics.tsv")

summary_table %>%
  kable(format = "html", caption = "Summary Statistics for ASR Path Score & Components (asr_path_summary_statistics.tsv)", digits = 4) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE)
```

---

## Output Files Summary

The following generated output tables and plots have been saved under the robustness results directory:

```{r output_summary, echo=FALSE}
tsv_files   <- list.files(tsv_dir,   pattern = "\\.tsv$", full.names = FALSE)
plot_files  <- list.files(plots_dir, pattern = "\\.png$", full.names = FALSE)

tibble(
  Type = c(rep("TSV", length(tsv_files)), rep("Plot", length(plot_files))),
  File = c(tsv_files, plot_files)
) %>%
  kable(format = "html", caption = "All output files generated") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — ASR Path Robustness Module*
*Date of generation: `r Sys.Date()`*
