---
title: "ASR Path Robustness Characterisation"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
params:
  disambig_dir:        NULL   # path to ct_disambiguation/ output directory
  posterior_threshold: 0.95   # params.ct_disambig_posterior_threshold
  output_dir:          "."    # working directory (TSVs → ./tsv/, plots → ./plots/)
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

<!-- 9.ASR_robustness.Rmd — Distribution of asr_path_score and derived_agreement over the scored CAAS positions. -->
<!-- PhyloPhere | subworkflows/ASR_ROBUSTNESS/local/ -->
<!--
  Called by: ASR_ROBUSTNESS_REPORT (asr_robustness.nf), through rmarkdown::render().
  Params:    disambig_dir (ct_disambiguation/ with caas_convergence_master.csv),
             posterior_threshold (shown in the report, not applied), output_dir.
  asr_path_score is the convergence score of the position on its Voronoi domains
  (fop_pool.py, "core"); derived_agreement is a descriptor and does not enter the score.
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 11,
  fig.height = 6,
  dpi        = 180,
  dev        = "png"
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(ggplot2)
  library(patchwork)
  library(scales)
  library(knitr)
  library(kableExtra)
})

theme_set(
  theme_bw(base_size = 12) +
  theme(
    axis.title.y  = element_text(margin = ggplot2::margin(l = 20, r = 20)),
    axis.title.x  = element_text(margin = ggplot2::margin(t = 12)),
    axis.text.x   = element_text(margin = ggplot2::margin(t = 5)),
    plot.title    = element_text(face = "bold", hjust = 0, margin = ggplot2::margin(b = 6)),
    plot.subtitle = element_text(colour = "grey40", hjust = 0, margin = ggplot2::margin(b = 10))
  )
)

# ── Parameters ────────────────────────────────────────────────────────────────
disambig_dir        <- params$disambig_dir
posterior_threshold <- as.numeric(params$posterior_threshold)
output_dir          <- params$output_dir

stopifnot(
  !is.null(disambig_dir)
)

# ── Output directories ────────────────────────────────────────────────────────
# Tables go to tsv/ and figures to plots/, both published by ASR_ROBUSTNESS_REPORT.
tsv_dir   <- file.path(output_dir, "tsv")
plots_dir <- file.path(output_dir, "plots")
dir.create(tsv_dir,   recursive = TRUE, showWarnings = FALSE)
dir.create(plots_dir, recursive = TRUE, showWarnings = FALSE)

write_tsv_safe <- function(df, fname) {
  path <- file.path(tsv_dir, fname)
  readr::write_tsv(df, path)
  invisible(path)
}

save_plot <- function(p, fname, width = 11, height = 6) {
  path <- file.path(plots_dir, fname)
  ggsave(path, plot = p, width = width, height = height, dpi = 180)
  invisible(path)
}

# Categorical palette (Tableau 10)
.tableau10 <- c(
  "#1F77B4", "#AEC7E8", "#FF7F0E", "#FFBB78",
  "#2CA02C", "#98DF8A", "#D62728", "#FF9896",
  "#9467BD", "#C5B0D5"
)
```

## Introduction

With the introduction of the continuous `asr_path_score`, the ASR evaluations have fundamentally shifted from binary cut-offs to a holistic score. Core v3: `asr_path_score == core`, the CAAS convergence score on the pair's Voronoi domain. This report details the distribution of the `asr_path_score` across evaluated sites alongside its retained descriptor:

*   **Convergence score (`core`)**: PSS-weighted mean over the K fixed Voronoi domains of each domain's noisy-OR agreement with its same-residue partners, discounted by the derived state's probability at the shared LCA.
*   **Convergence agreement (`derived_agreement`)**: descriptor only — coherence of derived states per phenotype side; penalises divergent derived residues.

(Core v3 removed the pairwise isolation factor and the earlier `independence` / `mrca_diversity` / `conservation_gate` axes from the score.)

---

```{r load-data}
master_csv <- file.path(disambig_dir, "caas_convergence_master.csv")
if (!file.exists(master_csv)) {
  stop(sprintf("[asr_robustness] caas_convergence_master.csv not found in: %s", disambig_dir))
}

master_df <- readr::read_csv(master_csv, show_col_types = FALSE, guess_max = 5000)

# Column names are normalized to gene and site (the master CSV may name them Gene, msa_pos or Position)
if ("msa_pos" %in% names(master_df)) master_df <- dplyr::rename(master_df, site = msa_pos)
if (!("gene" %in% names(master_df)) && "Gene" %in% names(master_df))
  master_df <- dplyr::rename(master_df, gene = Gene)
if (!("site" %in% names(master_df)) && "Position" %in% names(master_df))
  master_df <- dplyr::rename(master_df, site = Position)

# Rows without an asr_path_score are not scored and are left out
score_df <- master_df %>% filter(!is.na(asr_path_score))

if (nrow(score_df) == 0) {
  stop("No ASR path scores found in the provided CSV.")
}

comp_vars <- c("core", "derived_agreement")
available_comps <- intersect(comp_vars, colnames(score_df))
```

> **Pipeline threshold** (`params.ct_disambig_posterior_threshold`): **`r posterior_threshold`**
>
> **Disambiguation directory**: `r disambig_dir`
>
> **Total convergent sites loaded**: `r nrow(master_df)`
>
> **ASR path scored sites**: `r nrow(score_df)`

---

## 1. Overall ASR Path Score Distribution

The overall `asr_path_score` aggregates both the private walk and shared-node checks. High scores indicate independent, clean mutations that are well-supported by ancestral reconstructions.

```{r plot-score-dist, fig.cap="Distribution of the final ASR Path Score across positions."}
p_score <- ggplot(score_df, aes(x = asr_path_score)) +
  geom_histogram(binwidth = 0.02, fill = .tableau10[1], colour = "white", alpha = 0.8) +
  geom_density(aes(y = after_stat(density) * 0.02 * nrow(score_df)), color = .tableau10[7], linewidth = 1) +
  theme_bw(base_size = 12) +
  labs(
    title = "Overall ASR Path Score Distribution",
    subtitle = sprintf("%s positions across %d genes",
                       formatC(nrow(score_df), big.mark = ","),
                       n_distinct(score_df$gene)),
    x = "ASR Path Score",
    y = "Count"
  )

print(p_score)
save_plot(p_score, "asr_path_score_distribution.png", width = 8, height = 5)
```

---

## 2. Component Distributions

The final path score is composed of several continuous subcomponents. We plot their density distributions below to understand which factors act as the main limiters or contributors to the final scores in this dataset.

```{r plot-comp-dists, fig.height=8, fig.cap="Density distributions of the individual components that make up the ASR Path Score."}
comp_df <- score_df %>%
  select(gene, site, all_of(available_comps)) %>%
  pivot_longer(cols = all_of(available_comps), names_to = "component", values_to = "value") %>%
  filter(!is.na(value))

p_comps <- ggplot(comp_df, aes(x = value, fill = component)) +
  geom_density(alpha = 0.6, colour = "grey30", linewidth = 0.3) +
  facet_wrap(~ component, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = .tableau10[seq_along(available_comps)]) +
  theme_bw(base_size = 12) +
  theme(legend.position = "none",
        strip.background = element_rect(fill = "grey95", colour = "grey80"),
        strip.text = element_text(face = "bold")) +
  labs(
    title = "Distributions of ASR Path Score Components",
    x = "Component Value",
    y = "Density"
  )

print(p_comps)
save_plot(p_comps, "asr_path_components_density.png", width = 10, height = 8)
```

---

## 3. Summary Statistics

The table below lists the summary statistics for the final path score and each of its continuous subcomponents across all scored positions.

```{r summary-table}
summary_table <- score_df %>%
  select(asr_path_score, all_of(available_comps)) %>%
  summarise(across(everything(), list(
    Mean   = ~mean(., na.rm = TRUE),
    Median = ~median(., na.rm = TRUE),
    Min    = ~min(., na.rm = TRUE),
    Max    = ~max(., na.rm = TRUE),
    SD     = ~sd(., na.rm = TRUE)
  ))) %>%
  pivot_longer(everything(), names_to = c("Variable", "Stat"), names_sep = "_(?=[A-Za-z]+$)") %>%
  pivot_wider(names_from = Stat, values_from = value)

write_tsv_safe(summary_table, "asr_path_summary_statistics.tsv")

summary_table %>%
  kable(format = "html", caption = "Summary Statistics for ASR Path Score & Components (asr_path_summary_statistics.tsv)", digits = 4) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE)
```

---

## Output Files Summary

The following generated output tables and plots have been saved under the robustness results directory:

```{r output_summary, echo=FALSE}
tsv_files   <- list.files(tsv_dir,   pattern = "\\.tsv$", full.names = FALSE)
plot_files  <- list.files(plots_dir, pattern = "\\.png$", full.names = FALSE)

tibble(
  Type = c(rep("TSV", length(tsv_files)), rep("Plot", length(plot_files))),
  File = c(tsv_files, plot_files)
) %>%
  kable(format = "html", caption = "All output files generated") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — ASR Path Robustness Module*
*Date of generation: `r Sys.Date()`*
