---
title: "ASR Robustness Characterisation"
author: "PhyloPhere Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: cosmo
    code_folding: hide
    df_print: paged
params:
  disambig_dir:        NULL   # path to ct_disambiguation/ output directory
  posterior_threshold: 0.95   # params.ct_disambig_posterior_threshold
  output_dir:          "."    # working directory (TSVs → ./tsv/, plots → ./plots/)
---

<!--
  ASR Robustness Characterisation — PhyloPhere Pipeline
  ======================================================
  Core filtering rule:
    ASR uncertainty is described globally and via a compact per-gene summary
    (asr_gene_map_summary.tsv). Filtering for convergence inference is based
    only on the focal MRCA nodes that define change polarity. A site is
    retained only if the minimum MAP posterior across all focal nodes is >=
    params.ct_disambig_posterior_threshold.
    Additional threshold values are evaluated ONLY in the sensitivity analysis
    section; they do not affect pipeline filtering decisions.

  Performance note:
    JSONL parsing is streaming — only MAP values are materialised globally;
    full rows are kept only for focal nodes. This keeps memory tractable at
    tens of thousands of genes.

  Author: Miguel Ramon (miguel.ramon@upf.edu)
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 11,
  fig.height = 6,
  dpi        = 180,
  dev        = "png"
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(ggplot2)
  library(patchwork)
  library(scales)
  library(knitr)
  library(kableExtra)
  library(jsonlite)
})

# ── Parameter validation ──────────────────────────────────────────────────────
disambig_dir        <- params$disambig_dir
posterior_threshold <- as.numeric(params$posterior_threshold)
output_dir          <- params$output_dir

stopifnot(
  !is.null(disambig_dir),
  !is.null(posterior_threshold),
  posterior_threshold > 0, posterior_threshold <= 1
)

# ── Output directories ────────────────────────────────────────────────────────
tsv_dir   <- file.path(output_dir, "tsv")
plots_dir <- file.path(output_dir, "plots")
dir.create(tsv_dir,   recursive = TRUE, showWarnings = FALSE)
dir.create(plots_dir, recursive = TRUE, showWarnings = FALSE)

write_tsv_safe <- function(df, fname) {
  # readr::write_tsv requires atomic vector columns only.
  # Guard against list/matrix/data.frame columns that can appear after
  # pivot/join edge cases (especially with duplicated keys).
  scalarize_cell <- function(x) {
    if (is.null(x) || length(x) == 0) return(NA_real_)
    x <- suppressWarnings(as.numeric(unlist(x)))
    x <- x[!is.na(x)]
    if (length(x) == 0) return(NA_real_)
    if (length(x) == 1) return(x[[1]])
    mean(x)
  }

  bad_cols <- vapply(df, function(col) {
    is.list(col) ||
      is.data.frame(col) ||
      !is.null(dim(col)) ||
      !is.atomic(col)
  }, logical(1))

  if (any(bad_cols)) {
    bad_names <- names(df)[bad_cols]
    for (nm in bad_names) {
      col <- df[[nm]]
      if (is.data.frame(col) || !is.null(dim(col))) {
        mat <- as.matrix(col)
        df[[nm]] <- vapply(seq_len(nrow(mat)), function(i) scalarize_cell(mat[i, ]), numeric(1))
      } else if (is.list(col)) {
        df[[nm]] <- vapply(col, scalarize_cell, numeric(1))
      } else {
        df[[nm]] <- vapply(as.list(col), scalarize_cell, numeric(1))
      }
    }
    warning(sprintf(
      "[write_tsv_safe] Sanitized non-atomic columns in %s: %s",
      fname, paste(bad_names, collapse = ", ")
    ))
  }

  path <- file.path(tsv_dir, fname)
  readr::write_tsv(df, path)
  invisible(path)
}

save_plot <- function(p, fname, width = 11, height = 6) {
  path <- file.path(plots_dir, fname)
  ggsave(path, plot = p, width = width, height = height, dpi = 180)
  invisible(path)
}

as_bool <- function(x) {
  tolower(trimws(as.character(x))) %in% c("true", "t", "1", "yes", "y")
}

# Sensitivity tau grid — used ONLY in 7; never drives pipeline filtering
TAU_GRID <- c(0.80, 0.90, 0.95, 0.99)

# ── Helper: entropy (nats) ────────────────────────────────────────────────────
shannon_entropy <- function(probs) {
  p <- probs[probs > 0]
  -sum(p * log(p))
}

# ── Helper: styled kable ──────────────────────────────────────────────────────
kbl_pp <- function(df, caption = NULL) {
  df |>
    kable(format = "html", caption = caption, digits = 4) |>
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE, font_size = 12)
}
```

---

```{r paths, include=FALSE}
node_dumps_dir <- file.path(disambig_dir, "diagnostics", "node_dumps")
master_csv     <- file.path(disambig_dir, "caas_convergence_master.csv")

jsonl_files_exist <- dir.exists(node_dumps_dir) &&
  length(list.files(node_dumps_dir, pattern = "jsonl$")) > 0
master_csv_exists <- file.exists(master_csv)

if (!master_csv_exists) {
  stop(sprintf(
    "[asr_robustness] caas_convergence_master.csv not found in: %s", disambig_dir
  ))
}
```

```{r load_master_early, include=FALSE}
# ── Load master CSV early to build per-gene focal-node lookup ─────────────────
# Required before parse_jsonl so the parser can skip non-focal nodes.
master_df <- readr::read_csv(master_csv, show_col_types = FALSE, guess_max = 5000)

# Normalise column names (same logic mirrored in 4 load_master_csv)
if ("msa_pos" %in% names(master_df))  master_df <- dplyr::rename(master_df, site = msa_pos)
if (!("gene" %in% names(master_df)) && "Gene" %in% names(master_df))
  master_df <- dplyr::rename(master_df, gene = Gene)
if (!("site" %in% names(master_df)) && "Position" %in% names(master_df))
  master_df <- dplyr::rename(master_df, site = Position)
if (!("is_conserved_meta" %in% names(master_df))) master_df$is_conserved_meta <- FALSE
if (!("asr_is_conserved" %in% names(master_df))) master_df$asr_is_conserved <- FALSE
if (!("asr_root_conserved" %in% names(master_df))) master_df$asr_root_conserved <- FALSE

master_df <- master_df |>
  dplyr::mutate(
    is_conserved_meta = as_bool(is_conserved_meta),
    asr_is_conserved = as_bool(asr_is_conserved),
    asr_root_conserved = as_bool(asr_root_conserved)
  )

# Build focal-node lookup: named list(gene_name → integer vector of focal node IDs)
node_cols_early <- grep("^mrca_\\d+_node$", names(master_df), value = TRUE)
if (length(node_cols_early) > 0) {
  focal_lookup_map <- master_df |>
    dplyr::select(gene, all_of(node_cols_early)) |>
    tidyr::pivot_longer(all_of(node_cols_early), values_to = "focal_node_id", names_to = NULL) |>
    dplyr::filter(!is.na(focal_node_id)) |>
    dplyr::group_by(gene) |>
    dplyr::summarise(focal_ids = list(unique(as.integer(focal_node_id))), .groups = "drop") |>
    (\(x) setNames(x$focal_ids, x$gene))()
} else {
  focal_lookup_map <- list()
}
```

> **Pipeline threshold** (`params.ct_disambig_posterior_threshold`): **`r posterior_threshold`**

> **Disambiguation directory**: `r disambig_dir`

---

## 0 – Conserved-status burden scenarios

```{r conserved_status_burden}
scenario_counts <- tibble(
  scenario = c("include_both", "exclude_root_only", "exclude_both"),
  description = c(
    "No conserved-status exclusion",
    "Exclude conserved-meta rows with asr_root_conserved = FALSE",
    "Exclude conserved-meta rows with asr_is_conserved = FALSE or asr_root_conserved = FALSE"
  )
) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    rows_total = nrow(master_df),
    rows_retained = dplyr::case_when(
      scenario == "include_both" ~ nrow(master_df),
      scenario == "exclude_root_only" ~ sum(!(master_df$is_conserved_meta & !master_df$asr_root_conserved)),
      TRUE ~ sum(!(master_df$is_conserved_meta & (!master_df$asr_is_conserved | !master_df$asr_root_conserved)))
    ),
    rows_removed = rows_total - rows_retained,
    retained_fraction = rows_retained / rows_total,
    removed_fraction = rows_removed / rows_total
  ) |>
  dplyr::ungroup()

scenario_comparison <- scenario_counts |>
  dplyr::transmute(
    scenario,
    description,
    rows_removed,
    removed_fraction,
    incremental_vs_include_both = rows_removed - rows_removed[scenario == "include_both"],
    incremental_vs_root_only = rows_removed - rows_removed[scenario == "exclude_root_only"]
  )

write_tsv_safe(scenario_counts, "asr_conserved_status_burden.tsv")
write_tsv_safe(scenario_comparison, "asr_conserved_status_comparison.tsv")

kbl_pp(scenario_counts, caption = "0 Conserved-status burden across master-table scenarios (asr_conserved_status_burden.tsv)")
```

```{r conserved_status_comparison}
kbl_pp(scenario_comparison, caption = "0 Conserved-status scenario comparison (asr_conserved_status_comparison.tsv)")
```

---

## 1 – Per-(gene, node, site) ASR confidence metrics

```{r parse_jsonl, cache=FALSE}
# ── JSONL parser: focal nodes only ───────────────────────────────────────────
# Non-focal node lines are rejected with a plain string match BEFORE any JSON
# parse — avoiding millions of fromJSON() calls at 16k-gene scale.
# Focal rows capture MAP, margin, and entropy (needed for 4 join and 8).
# global_map_vec is derived from master_df posterior columns (already loaded),
# which contain exactly the MAP posteriors at focal nodes for all convergent
# sites — no JSONL traversal required for 2.

extract_focal_jsonl <- function(path, gene_name, focal_ids) {
  if (length(focal_ids) == 0L) return(NULL)

  lines <- readLines(path, warn = FALSE)
  lines <- lines[nzchar(trimws(lines))]

  # Build a fast lookup string for each focal id (e.g. '"node_id": 42')
  # so we can reject non-focal lines with grepl before any JSON parse.
  focal_patterns <- paste0('"node_id"\\s*:\\s*', focal_ids, '\\b')
  focal_regex    <- paste(focal_patterns, collapse = "|")

  all_rows <- vector("list", length(focal_ids) * 1000L)
  row_i    <- 0L

  for (ln in lines) {
    # Fast pre-filter: skip lines that cannot contain a focal node_id
    if (!grepl(focal_regex, ln, perl = TRUE)) next

    # Extract node_id with sub() — cheaper than full JSON parse
    nid_raw <- sub('.*"node_id"\\s*:\\s*([0-9]+).*', '\\1', ln, perl = TRUE)
    node_id <- suppressWarnings(as.integer(nid_raw))
    if (is.na(node_id) || !node_id %in% focal_ids) next

    # Full JSON parse only for confirmed focal nodes
    obj <- tryCatch(jsonlite::fromJSON(ln, simplifyVector = TRUE), error = function(e) NULL)
    if (is.null(obj) || !is.null(obj[["__metadata__"]])) next

    positions <- obj[["positions"]]
    if (is.null(positions) || !is.list(positions)) next

    site_names <- names(positions)
    aa_lists   <- lapply(positions, unlist)
    valid      <- vapply(aa_lists, function(x) is.numeric(x) && length(x) > 0, logical(1))
    aa_lists   <- aa_lists[valid]
    site_names <- site_names[valid]
    if (length(site_names) == 0L) next

    # Vectorized extraction across sites for this focal node
    sorted_list <- lapply(aa_lists, sort, decreasing = TRUE)
    MAP_vec     <- vapply(sorted_list, `[[`, numeric(1L), 1L)
    P_sec_vec   <- vapply(sorted_list,
                          function(x) if (length(x) >= 2L) x[[2L]] else 0,
                          numeric(1L))
    ent_vec     <- vapply(aa_lists, function(x) {
                            p <- x[x > 0]; -sum(p * log(p))
                          }, numeric(1L))

    n_new <- length(site_names)
    row_i <- row_i + 1L
    all_rows[[row_i]] <- data.frame(
      gene    = rep(gene_name, n_new),
      node    = rep(node_id,   n_new),
      site    = as.integer(site_names),
      MAP     = MAP_vec,
      margin  = MAP_vec - P_sec_vec,
      entropy = ent_vec,
      stringsAsFactors = FALSE
    )
  }

  if (row_i == 0L) return(NULL)
  dplyr::bind_rows(all_rows[seq_len(row_i)])
}

# ── Run parser over all JSONL files ──────────────────────────────────────────
focal_jm       <- NULL
n_genes_parsed <- 0L

if (jsonl_files_exist) {
  jfiles <- list.files(node_dumps_dir, pattern = "_posteriors\\.jsonl$", full.names = TRUE)
  n_genes_parsed <- length(jfiles)

  focal_rows_list <- lapply(jfiles, function(f) {
    gene_lower <- sub("_posteriors\\.jsonl$", "", basename(f))
    # JSONL filenames are lowercase; master_df gene names may be uppercase —
    # try exact match first, then case-insensitive fallback.
    gene <- if (!is.null(focal_lookup_map[[gene_lower]])) {
      gene_lower
    } else {
      gene_upper <- toupper(gene_lower)
      if (!is.null(focal_lookup_map[[gene_upper]])) gene_upper else gene_lower
    }
    fids <- focal_lookup_map[[gene]]
    if (is.null(fids)) return(NULL)
    extract_focal_jsonl(f, gene, fids)
  })

  not_null <- !vapply(focal_rows_list, is.null, logical(1L))
  if (any(not_null)) {
    focal_jm <- dplyr::bind_rows(focal_rows_list[not_null])
    message(sprintf("[1] %d gene files | %d focal-node rows recovered",
                    n_genes_parsed, nrow(focal_jm)))
  } else {
    message("[1] No focal-node rows recovered from JSONL files.")
    jsonl_files_exist <- FALSE
  }
} else {
  message("[1] No posterior JSONL files found under ", node_dumps_dir,
          " — sections 1–3 skipped. Enable ct_disambig_run_diagnostics=true.")
}

# ── Global MAP vector from master_df posterior columns ───────────────────────
# The mrca_*_posterior columns hold the MAP posteriors at focal nodes for all
# convergent sites — the exact distribution relevant for the report.
post_cols_early <- grep("^mrca_\\d+_posterior$", names(master_df), value = TRUE)
global_map_vec <- if (length(post_cols_early) > 0) {
  unlist(master_df[, post_cols_early], use.names = FALSE)
} else {
  NULL
}
global_map_vec <- global_map_vec[!is.na(global_map_vec)]
jsonl_or_master_ok <- !is.null(global_map_vec) && length(global_map_vec) > 0L
```

`r if (!jsonl_or_master_ok) "> ⚠️ No focal-node MAP posteriors available. Sections 1–3 are skipped." `

```{r show_metrics_head, eval=jsonl_or_master_ok}
cat(sprintf(
  "Global MAP vector: %s values (focal-node posteriors from %d convergent sites × %d pairs).\nJSONL parsed for margin/entropy: %d focal-node rows.",
  formatC(length(global_map_vec), big.mark = ",", format = "d"),
  nrow(master_df),
  length(post_cols_early),
  if (!is.null(focal_jm)) nrow(focal_jm) else 0L
))
```

---

## 2 – Global MAP distribution

```{r global_stats, eval=jsonl_or_master_ok}
# ── 2a Full distribution ─────────────────────────────────────────────────────
compute_map_stats <- function(map_vec) {
  pcts <- c(0.01, 0.05, 0.10, 0.20, 0.25, 0.40, 0.50, 0.60, 0.75, 0.80, 0.90, 0.95, 0.99)
  cuts <- c(1.00, 0.99, 0.95, 0.90, 0.80, 0.60)

  q <- quantile(map_vec, probs = pcts, na.rm = TRUE)
  frac_below <- vapply(cuts, function(t) mean(map_vec < t, na.rm = TRUE), numeric(1))

  c(
    N_total = length(map_vec),
    mean    = mean(map_vec, na.rm = TRUE),
    sd      = sd(map_vec, na.rm = TRUE),
    min     = min(map_vec, na.rm = TRUE),
    max     = max(map_vec, na.rm = TRUE),
    setNames(as.numeric(q),     paste0("pct_", pcts * 100)),
    setNames(frac_below,        paste0("frac_MAP_lt_", sub("\\.", "", as.character(cuts))))
  )
}

map_vec_full <- global_map_vec
stats_full   <- compute_map_stats(map_vec_full)

global_stats_df <- tibble(statistic = names(stats_full), value = as.numeric(stats_full))
write_tsv_safe(global_stats_df, "asr_global_statistics.tsv")

# ── 2b Uncertainty tail (MAP < 1.0) ─────────────────────────────────────────
map_vec_tail  <- map_vec_full[map_vec_full < 1.0]
stats_tail    <- compute_map_stats(map_vec_tail)
tail_stats_df <- tibble(statistic = names(stats_tail), value = as.numeric(stats_tail))
write_tsv_safe(tail_stats_df, "asr_uncertain_tail.tsv")

kbl_pp(global_stats_df, caption = "2a Global MAP statistics (asr_global_statistics.tsv)")
```

```{r global_tail_summary, eval=jsonl_or_master_ok}
kbl_pp(tail_stats_df, caption = "2b Global MAP statistics — uncertainty tail MAP < 1.0 (asr_uncertain_tail.tsv)")
```

---

## 3 – Per-gene MAP summary

`r if (jsonl_files_exist) sprintf("> **%d gene** JSONL files processed. Per-gene MAP summaries (N, mean, sd, min, fraction MAP=1) written to **asr_gene_map_summary.tsv**.", n_genes_parsed)`
`r if (!jsonl_files_exist) "> ℹ️ JSONL files not found — per-gene margin/entropy summary not available. Global MAP stats are derived from master CSV."`

---

## 4 – Focal MRCA node extraction

```{r load_master_csv}
# master_df was already loaded and column-normalised in load_master_early.
# Detect focal-node pair columns dynamically.
node_col_regex   <- "^mrca_(\\d+)_node$"
post_col_regex   <- "^mrca_(\\d+)_posterior$"
margin_col_regex <- "^mrca_(\\d+)_margin$"

node_cols <- grep(node_col_regex, names(master_df), value = TRUE)
post_cols <- grep(post_col_regex, names(master_df), value = TRUE)
n_pairs   <- length(post_cols)

if (n_pairs == 0) stop("[asr_robustness 4] No mrca_*_posterior columns found in master CSV")

pair_ids <- as.integer(sub("^mrca_(\\d+)_posterior$", "\\1", post_cols))
message(sprintf("[4] Detected %d focal MRCA pairs: %s", n_pairs, paste(pair_ids, collapse = ", ")))
```

```{r focal_extraction}
# ── Build per-(gene, site) focal-node posteriors ──────────────────────────────
# Each row in master_df is one (gene, site) convergent position.
focal_df <- master_df |>
  dplyr::select(gene, site,
                all_of(c(node_cols, post_cols))) |>
  dplyr::rename_with(~ paste0("node", pair_ids), all_of(node_cols)) |>
  dplyr::rename_with(~ paste0("p",    pair_ids), all_of(post_cols))

p_cols <- paste0("p",      pair_ids)
m_cols <- paste0("margin", pair_ids)
e_cols <- paste0("entropy", pair_ids)

# ── Join margin & entropy from focal-only JSONL rows ─────────────────────────
# focal_jm has columns: gene, node, site, margin, entropy.
# Strategy: pivot focal_df node columns to long → single join → pivot back wide.
# Avoids scanning focal_jm once per pair (old sequential for-loop approach).
if (!is.null(focal_jm) && nrow(focal_jm) > 0) {
  node_long <- focal_df |>
    dplyr::select(gene, site, all_of(paste0("node", pair_ids))) |>
    tidyr::pivot_longer(
      cols      = all_of(paste0("node", pair_ids)),
      names_to  = "pair_col",
      values_to = "node_id"
    ) |>
    dplyr::mutate(
      pair_num = as.integer(sub("^node", "", pair_col)),
      node_id  = as.integer(node_id)
    ) |>
    dplyr::filter(!is.na(node_id)) |>
    dplyr::left_join(
      focal_jm |> dplyr::mutate(node = as.integer(node)),
      by = c("gene" = "gene", "site" = "site", "node_id" = "node")
    ) |>
    dplyr::select(gene, site, pair_num, margin, entropy) |>
    tidyr::pivot_wider(
      id_cols     = c(gene, site),
      names_from  = pair_num,
      values_from = c(margin, entropy),
      names_glue  = "{.value}{pair_num}",  # → margin1, entropy1, margin2, ...
      values_fn   = list(
        margin  = ~ if (all(is.na(.x))) NA_real_ else mean(.x, na.rm = TRUE),
        entropy = ~ if (all(is.na(.x))) NA_real_ else mean(.x, na.rm = TRUE)
      ),                                   # collapse duplicates (same gene/site/node in focal_jm)
      values_fill = NA_real_               # avoid list columns when a cell has no observations
    )
  focal_df <- dplyr::left_join(focal_df, node_long, by = c("gene", "site"))
}

# ── Vectorized aggregate stats across focal nodes ────────────────────────────
# pmin / pmax across columns — much faster than dplyr::rowwise() at scale.
p_mat          <- as.matrix(focal_df[, p_cols, drop = FALSE])
focal_df$p_min  <- do.call(pmin, c(asplit(p_mat, 2), list(na.rm = TRUE)))
focal_df$p_prod <- Reduce(`*`, lapply(p_cols, function(col) focal_df[[col]]))

has_margin_cols  <- all(m_cols %in% names(focal_df))
has_entropy_cols <- all(e_cols %in% names(focal_df))

if (has_margin_cols) {
  m_mat <- as.matrix(focal_df[, m_cols, drop = FALSE])
  focal_df$margin_min <- do.call(pmin, c(asplit(m_mat, 2), list(na.rm = TRUE)))
  focal_df$margin_min <- ifelse(is.infinite(focal_df$margin_min), NA_real_, focal_df$margin_min)
}

if (has_entropy_cols) {
  e_mat <- as.matrix(focal_df[, e_cols, drop = FALSE])
  focal_df$entropy_max <- do.call(pmax, c(asplit(e_mat, 2), list(na.rm = TRUE)))
  focal_df$entropy_max <- ifelse(is.infinite(focal_df$entropy_max), NA_real_, focal_df$entropy_max)
}

# Flatten any residual list columns that can arise from pivot_wider edge cases
# (e.g. tidyr producing list-type cells when values_fn is applied to mixed input).
list_col_idx <- which(vapply(focal_df, is.list, logical(1)))
if (length(list_col_idx) > 0) {
  focal_df <- dplyr::mutate(focal_df, dplyr::across(
    where(is.list),
    ~ vapply(., function(x) {
        x <- unlist(x)
        if (length(x) == 0 || all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE)
      }, numeric(1))
  ))
}

# Defensive final normalization before writing TSV:
# ensure no column keeps matrix/list-like structure under S3 wrappers.
focal_df <- as.data.frame(focal_df, stringsAsFactors = FALSE)

write_tsv_safe(focal_df, "asr_focal_node_sites.tsv")
cat(sprintf("Focal-node table: %d sites across %d genes.\n",
            nrow(focal_df), dplyr::n_distinct(focal_df$gene)))
```

---

## 5 – Site filtering

The filtering rule is:

$$\text{PASS} \iff p_{\min} \geq \tau \quad\text{where}\quad \tau = \texttt{params.ct\_disambig\_posterior\_threshold} = `r posterior_threshold`$$

```{r site_filter}
tau <- posterior_threshold

filter_df <- focal_df |>
  dplyr::select(gene, site, all_of(p_cols), p_min, p_prod) |>
  dplyr::mutate(
    PASS            = p_min >= tau,
    threshold_used  = tau
  )

write_tsv_safe(filter_df, "asr_site_filtering.tsv")

n_pass <- sum(filter_df$PASS,  na.rm = TRUE)
n_fail <- sum(!filter_df$PASS, na.rm = TRUE)

cat(sprintf(
  "τ = %.4f | Sites PASS: %d  |  Sites FAIL: %d  |  Retention: %.1f%%\n",
  tau, n_pass, n_fail, 100 * n_pass / nrow(filter_df)
))

filter_df |> head(10) |> kbl_pp(caption = "5 Site-level filtering (asr_site_filtering.tsv)")
```

---

## 6 – Retention statistics

```{r retention_stats}
# ── Global ────────────────────────────────────────────────────────────────────
retention_global <- tibble(
  N_sites_total      = nrow(filter_df),
  N_sites_retained   = n_pass,
  N_sites_excluded   = n_fail,
  retention_fraction = n_pass / nrow(filter_df),
  exclusion_fraction = n_fail / nrow(filter_df),
  threshold_used     = tau
)
write_tsv_safe(retention_global, "asr_retention_summary.tsv")

# ── Per gene ──────────────────────────────────────────────────────────────────
retention_gene <- filter_df |>
  dplyr::group_by(gene) |>
  dplyr::summarise(
    sites_total        = dplyr::n(),
    sites_retained     = sum(PASS, na.rm = TRUE),
    sites_excluded     = sum(!PASS, na.rm = TRUE),
    retention_fraction = sites_retained / sites_total,
    exclusion_fraction = sites_excluded / sites_total,
    threshold_used     = tau,
    .groups = "drop"
  )
write_tsv_safe(retention_gene, "asr_gene_retention.tsv")

kbl_pp(retention_global, caption = "6 Global retention summary (asr_retention_summary.tsv)")
```


---

## 7 – Sensitivity analysis

*The pipeline always filters using `params.ct_disambig_posterior_threshold`. The tau grid below is a diagnostic summary only and does not affect any pipeline decision.*

```{r sensitivity}
tau_grid <- TAU_GRID  # c(0.80, 0.90, 0.95, 0.99) — diagnostic only

apply_tau <- function(fd, t) {
  kept <- fd |> dplyr::filter(p_min >= t)
  tibble(
    tau                = t,
    N_sites_total      = nrow(fd),
    N_sites_retained   = nrow(kept),
    retention_fraction = nrow(kept) / nrow(fd),
    N_convergent_sites = dplyr::n_distinct(paste(kept$gene, kept$site)),
    N_convergent_genes = dplyr::n_distinct(kept$gene)
  )
}

sensitivity_df <- dplyr::bind_rows(lapply(tau_grid, function(t) apply_tau(filter_df, t)))
write_tsv_safe(sensitivity_df, "convergence_sensitivity.tsv")

# ── Callset overlap (Jaccard) ─────────────────────────────────────────────────
callset <- function(t) {
  filter_df |>
    dplyr::filter(p_min >= t) |>
    dplyr::mutate(key = paste(gene, site, sep = ":")) |>
    dplyr::pull(key)
}

pairs <- list(
  c("0.90", "0.95"),
  c("0.95", "0.99"),
  c("0.90", "0.99")
)

overlap_rows <- lapply(pairs, function(pr) {
  t1 <- as.numeric(pr[[1]]);  t2 <- as.numeric(pr[[2]])
  s1 <- callset(t1);           s2 <- callset(t2)
  inter <- length(intersect(s1, s2))
  union <- length(union(s1, s2))
  tibble(
    tau_A        = t1,
    tau_B        = t2,
    n_A          = length(s1),
    n_B          = length(s2),
    n_intersect  = inter,
    n_union      = union,
    jaccard      = if (union > 0) inter / union else NA_real_
  )
})

overlap_df <- dplyr::bind_rows(overlap_rows)
write_tsv_safe(overlap_df, "convergence_overlap.tsv")

kbl_pp(sensitivity_df, caption = "7 Sensitivity analysis (convergence_sensitivity.tsv)")
```

```{r overlap_tbl}
kbl_pp(overlap_df, caption = "7 Callset overlap between tau thresholds (convergence_overlap.tsv)")
```

---

## 8 – Diagnostic plots

### Plot 1 — Global MAP distribution

```{r plot1_global_map, eval=jsonl_or_master_ok, fig.height=5}
# Main panel: full distribution including MAP=1.0
# Use coord_cartesian so the bin at exactly 1.0 is not clipped
p_full <- ggplot(data.frame(MAP = global_map_vec), aes(x = MAP)) +
  geom_histogram(binwidth = 0.01, fill = "#4C8FC0", colour = NA, alpha = 0.8,
                 boundary = 0, closed = "right") +
  coord_cartesian(xlim = c(0, 1.01)) +
  scale_x_continuous(breaks = seq(0, 1, 0.1)) +
  labs(
    title    = "MAP posterior distribution at focal MRCA nodes (all convergent sites)",
    subtitle = sprintf("%s sites × %d focal node pairs | %s MAP values",
                       formatC(nrow(master_df), big.mark = ",", format = "d"),
                       length(post_cols_early),
                       formatC(length(global_map_vec), big.mark = ",", format = "d")),
    x = "MAP posterior",
    y = "Count"
  ) +
  theme_bw(base_size = 12)

# Inset: uncertainty tail (MAP < 1.0)
p_inset <- ggplot(data.frame(MAP = global_map_vec[global_map_vec < 1.0]), aes(x = MAP)) +
  geom_histogram(binwidth = 0.01, fill = "#E07B39", colour = NA, alpha = 0.85,
                 boundary = 0, closed = "right") +
  coord_cartesian(xlim = c(0, 1)) +
  scale_x_continuous(breaks = seq(0, 1, 0.2)) +
  labs(title = "MAP < 1.0 (uncertainty tail)", x = NULL, y = NULL) +
  theme_bw(base_size = 9) +
  theme(plot.title = element_text(size = 8))

p1 <- p_full +
  inset_element(p_inset, left = 0.02, bottom = 0.45, right = 0.48, top = 0.98)

save_plot(p1, "plot_asr_global_map_distribution.png", width = 11, height = 5)
p1
```

### Plot 2 — Focal-node p_min distribution

```{r plot2_pmin, fig.height=5}
p4 <- ggplot(focal_df, aes(x = p_min)) +
  geom_histogram(binwidth = 0.01, fill = "#4C8FC0", colour = NA, alpha = 0.85) +
  geom_vline(xintercept = posterior_threshold,
             colour = "#D62728", linewidth = 0.9, linetype = "dashed") +
  annotate("text",
           x     = posterior_threshold - 0.01,
           y     = Inf,
           label = sprintf("τ = %.2f", posterior_threshold),
           colour = "#D62728", hjust = 1, vjust = 1.5, size = 3.5) +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.1)) +
  labs(
    title    = "Distribution of p_min across focal MRCA nodes",
    subtitle = "Dashed line = params.ct_disambig_posterior_threshold",
    x        = "p_min (minimum MAP across focal nodes for each site)",
    y        = "Count"
  ) +
  theme_bw(base_size = 12)

save_plot(p4, "plot_asr_focal_pmin_distribution.png", width = 11, height = 5)
p4
```

### Plot 3 — Sensitivity: retained sites vs threshold

```{r plot3_retention_sensitivity, fig.height=5}
p6 <- ggplot(sensitivity_df, aes(x = tau, y = retention_fraction)) +
  geom_line(colour = "#4C8FC0", linewidth = 1.1) +
  geom_point(colour = "#4C8FC0", size = 3) +
  geom_vline(xintercept = posterior_threshold,
             colour = "#D62728", linewidth = 0.8, linetype = "dashed") +
  annotate("text",
           x = posterior_threshold + 0.002, y = 0.95,
           label = sprintf("pipeline τ = %.2f", posterior_threshold),
           colour = "#D62728", hjust = 0, size = 3.5) +
  scale_x_continuous(breaks = TAU_GRID) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1.05)) +
  labs(
    title    = "Sensitivity of site retention to posterior threshold",
    subtitle = "Global retention fraction across tau grid (diagnostic only)",
    x        = "Posterior threshold (τ)",
    y        = "Retention fraction"
  ) +
  theme_bw(base_size = 12)

# Add per-gene lines if not too many genes
per_gene_sens <- dplyr::bind_rows(lapply(tau_grid, function(t) {
  filter_df |>
    dplyr::group_by(gene) |>
    dplyr::summarise(
      tau                = t,
      retention_fraction = mean(p_min >= t, na.rm = TRUE),
      .groups = "drop"
    )
}))

if (dplyr::n_distinct(per_gene_sens$gene) <= 20) {
  p6 <- p6 +
    geom_line(data = per_gene_sens,
              aes(x = tau, y = retention_fraction, group = gene),
              colour = "grey70", linewidth = 0.5, alpha = 0.6)
}

save_plot(p6, "plot_asr_retention_sensitivity.png", width = 11, height = 5)
p6
```

### Plot 4 — Sensitivity of convergence calls to threshold

```{r plot4_convergence_sensitivity, fig.height=5}
p7a <- ggplot(sensitivity_df, aes(x = tau, y = N_convergent_sites)) +
  geom_line(colour = "#4C8FC0", linewidth = 1.1) +
  geom_point(colour = "#4C8FC0", size = 3) +
  geom_vline(xintercept = posterior_threshold,
             colour = "#D62728", linewidth = 0.8, linetype = "dashed") +
  scale_x_continuous(breaks = TAU_GRID) +
  labs(title = "Convergent sites", x = "Threshold (τ)", y = "N convergent sites") +
  theme_bw(base_size = 11)

p7b <- ggplot(sensitivity_df, aes(x = tau, y = N_convergent_genes)) +
  geom_line(colour = "#E07B39", linewidth = 1.1) +
  geom_point(colour = "#E07B39", size = 3) +
  geom_vline(xintercept = posterior_threshold,
             colour = "#D62728", linewidth = 0.8, linetype = "dashed") +
  scale_x_continuous(breaks = TAU_GRID) +
  labs(title = "Convergent genes", x = "Threshold (τ)", y = "N convergent genes") +
  theme_bw(base_size = 11)

p7 <- p7a + p7b +
  plot_annotation(
    title    = "Sensitivity of convergence calls to posterior threshold",
    subtitle = "Diagnostic only — pipeline always uses params.ct_disambig_posterior_threshold",
    theme    = theme(plot.title = element_text(size = 13))
  )

save_plot(p7, "plot_convergence_sensitivity.png", width = 13, height = 5)
p7
```

---

## Output file summary

```{r output_summary, echo=FALSE}
tsv_files   <- list.files(tsv_dir,   pattern = "\\.tsv$", full.names = FALSE)
plot_files  <- list.files(plots_dir, pattern = "\\.png$", full.names = FALSE)

tibble(
  type = c(rep("TSV", length(tsv_files)), rep("Plot", length(plot_files))),
  file = c(tsv_files, plot_files)
) |>
  kbl_pp(caption = "All output files generated")
```

---

*Generated by PhyloPhere — ASR Robustness Module*  
*`params.ct_disambig_posterior_threshold` = **`r posterior_threshold`***
