---
title: "ASR Robustness Characterisation"
author: "PhyloPhere Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: cosmo
    code_folding: hide
    df_print: paged
params:
  disambig_dir:        NULL   # path to ct_disambiguation/ output directory
  posterior_threshold: 0.95   # params.ct_disambig_posterior_threshold
  output_dir:          "."    # working directory (TSVs → ./tsv/, plots → ./plots/)
---

<!--
  ASR Robustness Characterisation — PhyloPhere Pipeline
  ======================================================
  Core filtering rule:
    ASR uncertainty is described globally and per gene, but filtering for
    convergence inference is based only on the focal MRCA nodes that define
    change polarity. A site is retained only if the minimum MAP posterior
    across all focal nodes is >= params.ct_disambig_posterior_threshold.
    Additional threshold values are evaluated ONLY in the sensitivity analysis
    section; they do not affect pipeline filtering decisions.

  Author: Miguel Ramon (miguel.ramon@upf.edu)
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 11,
  fig.height = 6,
  dpi        = 180,
  dev        = "png"
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(ggplot2)
  library(patchwork)
  library(scales)
  library(knitr)
  library(kableExtra)
  library(jsonlite)
})

# ── Parameter validation ──────────────────────────────────────────────────────
disambig_dir        <- params$disambig_dir
posterior_threshold <- as.numeric(params$posterior_threshold)
output_dir          <- params$output_dir

stopifnot(
  !is.null(disambig_dir),
  !is.null(posterior_threshold),
  posterior_threshold > 0, posterior_threshold <= 1
)

# ── Output directories ────────────────────────────────────────────────────────
tsv_dir   <- file.path(output_dir, "tsv")
plots_dir <- file.path(output_dir, "plots")
dir.create(tsv_dir,   recursive = TRUE, showWarnings = FALSE)
dir.create(plots_dir, recursive = TRUE, showWarnings = FALSE)

write_tsv_safe <- function(df, fname) {
  path <- file.path(tsv_dir, fname)
  readr::write_tsv(df, path)
  invisible(path)
}

save_plot <- function(p, fname, width = 11, height = 6) {
  path <- file.path(plots_dir, fname)
  ggsave(path, plot = p, width = width, height = height, dpi = 180)
  invisible(path)
}

# Sensitivity tau grid — used ONLY in §7; never drives pipeline filtering
TAU_GRID <- c(0.90, 0.95, 0.99)

# ── Helper: entropy (nats) ────────────────────────────────────────────────────
shannon_entropy <- function(probs) {
  p <- probs[probs > 0]
  -sum(p * log(p))
}

# ── Helper: styled kable ──────────────────────────────────────────────────────
kbl_pp <- function(df, caption = NULL) {
  df |>
    kable(caption = caption, digits = 4) |>
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE, font_size = 12)
}
```

---

```{r paths, include=FALSE}
node_dumps_dir <- file.path(disambig_dir, "diagnostics", "node_dumps")
master_csv     <- file.path(disambig_dir, "caas_convergence_master.csv")

jsonl_files_exist <- dir.exists(node_dumps_dir) &&
  length(list.files(node_dumps_dir, pattern = "_posteriors\\.jsonl$")) > 0
master_csv_exists <- file.exists(master_csv)

if (!master_csv_exists) {
  stop(sprintf(
    "[asr_robustness] caas_convergence_master.csv not found in: %s", disambig_dir
  ))
}
```

> **Pipeline threshold** (`params.ct_disambig_posterior_threshold`): **`r posterior_threshold`**

> **Disambiguation directory**: `r disambig_dir`

---

## §1 – Per-(gene, node, site) ASR confidence metrics

```{r parse_jsonl, cache=FALSE}
# ── Parse per-gene posterior JSONL files ──────────────────────────────────────
# Format (one JSON object per line):
#   {"node_id": <int>, "positions": {"<site>": {"A": p, "C": p, ...}}}
# Optional first line: {"__metadata__": {...}}  — skipped.
#
# Returns a data.frame with columns:
#   gene | node | site | MAP | AA_MAP | AA_second | P_second | margin | entropy

parse_single_jsonl <- function(path, gene_name) {
  lines <- readLines(path, warn = FALSE)
  lines <- lines[nzchar(trimws(lines))]

  rows <- vector("list", length(lines) * 50L)  # pre-allocate generously
  row_i <- 0L

  for (ln in lines) {
    obj <- tryCatch(jsonlite::fromJSON(ln, simplifyVector = TRUE), error = function(e) NULL)
    if (is.null(obj)) next
    if (!is.null(obj[["__metadata__"]])) next  # skip metadata header

    node_id   <- as.integer(obj[["node_id"]])
    positions <- obj[["positions"]]
    if (is.null(positions) || !is.list(positions)) next

    for (site_s in names(positions)) {
      aa_map <- positions[[site_s]]
      if (!is.numeric(aa_map) || length(aa_map) == 0) next

      site <- as.integer(site_s)
      probs <- sort(aa_map, decreasing = TRUE)

      MAP      <- probs[[1]]
      AA_MAP   <- names(probs)[[1]]
      AA_sec   <- if (length(probs) >= 2) names(probs)[[2]] else NA_character_
      P_sec    <- if (length(probs) >= 2) probs[[2]]        else 0
      margin   <- MAP - P_sec
      Hent     <- shannon_entropy(aa_map)

      row_i <- row_i + 1L
      rows[[row_i]] <- list(
        gene      = gene_name,
        node      = node_id,
        site      = site,
        MAP       = MAP,
        AA_MAP    = AA_MAP,
        AA_second = AA_sec,
        P_second  = P_sec,
        margin    = margin,
        entropy   = Hent
      )
    }
  }

  if (row_i == 0L) return(NULL)
  dplyr::bind_rows(rows[seq_len(row_i)])
}

node_site_metrics <- NULL

if (jsonl_files_exist) {
  jfiles <- list.files(node_dumps_dir, pattern = "_posteriors\\.jsonl$", full.names = TRUE)

  parsed <- lapply(jfiles, function(f) {
    gene <- sub("_posteriors\\.jsonl$", "", basename(f))
    parse_single_jsonl(f, gene)
  })

  node_site_metrics <- dplyr::bind_rows(parsed)
  message(sprintf("[§1] Parsed %d JSONL files → %d (gene, node, site) rows",
                  length(jfiles), nrow(node_site_metrics)))
  write_tsv_safe(node_site_metrics, "asr_node_site_metrics.tsv")
} else {
  message("[§1] No posterior JSONL files found under ", node_dumps_dir,
          " — sections 1-3 will be skipped. Enable ct_disambig_run_diagnostics=true.")
}
```

`r if (!jsonl_files_exist) "> ⚠️ Posterior JSONL files were not found (ct_disambig_run_diagnostics may have been disabled). Sections 1–3 are skipped." `

```{r show_metrics_head, eval=jsonl_files_exist}
node_site_metrics |> head(10) |> kbl_pp(caption = "First 10 rows of asr_node_site_metrics.tsv")
```

---

## §2 – Global MAP distribution

```{r global_stats, eval=jsonl_files_exist}
# ── §2a Full distribution ─────────────────────────────────────────────────────
compute_map_stats <- function(map_vec) {
  pcts <- c(0.01, 0.05, 0.10, 0.20, 0.25, 0.40, 0.50, 0.60, 0.75, 0.80, 0.90, 0.95, 0.99)
  cuts <- c(1.00, 0.99, 0.95, 0.90, 0.80, 0.60)

  q <- quantile(map_vec, probs = pcts, na.rm = TRUE)
  frac_below <- vapply(cuts, function(t) mean(map_vec < t, na.rm = TRUE), numeric(1))

  c(
    N_total = length(map_vec),
    mean    = mean(map_vec, na.rm = TRUE),
    sd      = sd(map_vec, na.rm = TRUE),
    min     = min(map_vec, na.rm = TRUE),
    max     = max(map_vec, na.rm = TRUE),
    setNames(as.numeric(q),     paste0("pct_", pcts * 100)),
    setNames(frac_below,        paste0("frac_MAP_lt_", sub("\\.", "", as.character(cuts))))
  )
}

map_vec_full <- node_site_metrics$MAP
stats_full   <- compute_map_stats(map_vec_full)

global_stats_df <- tibble(statistic = names(stats_full), value = as.numeric(stats_full))
write_tsv_safe(global_stats_df, "asr_global_statistics.tsv")

# ── §2b Uncertainty tail (MAP < 1.0) ─────────────────────────────────────────
map_vec_tail  <- map_vec_full[map_vec_full < 1.0]
stats_tail    <- compute_map_stats(map_vec_tail)
tail_stats_df <- tibble(statistic = names(stats_tail), value = as.numeric(stats_tail))
write_tsv_safe(tail_stats_df, "asr_uncertain_tail.tsv")

kbl_pp(global_stats_df, caption = "§2a Global MAP statistics (asr_global_statistics.tsv)")
```

```{r global_tail_summary, eval=jsonl_files_exist}
kbl_pp(tail_stats_df, caption = "§2b Global MAP statistics — uncertainty tail MAP < 1.0 (asr_uncertain_tail.tsv)")
```

---

## §3 – Per-gene MAP statistics

```{r gene_stats, eval=jsonl_files_exist}
compute_gene_stats <- function(df_gene) {
  m <- df_gene$MAP
  tibble(
    N_total            = length(m),
    fraction_MAP_1     = mean(m == 1,    na.rm = TRUE),
    frac_MAP_lt_100    = mean(m < 1.00,  na.rm = TRUE),
    frac_MAP_lt_099    = mean(m < 0.99,  na.rm = TRUE),
    frac_MAP_lt_095    = mean(m < 0.95,  na.rm = TRUE),
    frac_MAP_lt_090    = mean(m < 0.90,  na.rm = TRUE),
    frac_MAP_lt_080    = mean(m < 0.80,  na.rm = TRUE),
    frac_MAP_lt_060    = mean(m < 0.60,  na.rm = TRUE),
    min_MAP            = min(m,  na.rm = TRUE),
    mean_MAP           = mean(m, na.rm = TRUE),
    sd_MAP             = sd(m,   na.rm = TRUE)
  )
}

gene_stats_df <- node_site_metrics |>
  group_by(gene) |>
  group_modify(~ compute_gene_stats(.x)) |>
  ungroup()

write_tsv_safe(gene_stats_df, "asr_gene_statistics.tsv")

# Tail version (MAP < 1.0 per gene)
gene_tail_df <- node_site_metrics |>
  filter(MAP < 1.0) |>
  group_by(gene) |>
  group_modify(~ compute_gene_stats(.x)) |>
  ungroup()

write_tsv_safe(gene_tail_df, "asr_gene_uncertain_tail.tsv")

kbl_pp(gene_stats_df, caption = "§3 Per-gene MAP statistics (asr_gene_statistics.tsv)")
```

```{r gene_tail_tbl, eval=jsonl_files_exist}
kbl_pp(gene_tail_df, caption = "§3 Per-gene MAP statistics — uncertainty tail MAP < 1.0 (asr_gene_uncertain_tail.tsv)")
```

---

## §4 – Focal MRCA node extraction

```{r load_master_csv}
# ── Load caas_convergence_master.csv ─────────────────────────────────────────
# Focal MRCA nodes are auto-derived from mrca_*_node and mrca_*_posterior
# columns.  No user-supplied node list is required.
master_df <- readr::read_csv(master_csv, show_col_types = FALSE, guess_max = 5000)

# Normalise column names
if ("msa_pos" %in% names(master_df))  master_df <- dplyr::rename(master_df, site = msa_pos)
if (!("gene" %in% names(master_df)) && "Gene" %in% names(master_df))
  master_df <- dplyr::rename(master_df, gene = Gene)
if (!("site" %in% names(master_df)) && "Position" %in% names(master_df))
  master_df <- dplyr::rename(master_df, site = Position)

# Detect focal-node pair columns dynamically
node_col_regex     <- "^mrca_(\\d+)_node$"
post_col_regex     <- "^mrca_(\\d+)_posterior$"
margin_col_regex   <- "^mrca_(\\d+)_margin$"

node_cols   <- grep(node_col_regex,   names(master_df), value = TRUE)
post_cols   <- grep(post_col_regex,   names(master_df), value = TRUE)
n_pairs     <- length(post_cols)

if (n_pairs == 0) stop("[asr_robustness §4] No mrca_*_posterior columns found in master CSV")

pair_ids <- as.integer(sub("^mrca_(\\d+)_posterior$", "\\1", post_cols))
message(sprintf("[§4] Detected %d focal MRCA pairs: %s", n_pairs, paste(pair_ids, collapse = ", ")))
```

```{r focal_extraction}
# ── Build per-(gene, site) focal-node rows ────────────────────────────────────
# Each row in master_df corresponds to one (gene, site) convergent position.
# p_i = mrca_i_posterior (MAP at focal node i for this site).

focal_df <- master_df |>
  dplyr::select(gene, site,
                all_of(c(node_cols, post_cols))) |>
  dplyr::rename_with(~ paste0("node", pair_ids), all_of(node_cols)) |>
  dplyr::rename_with(~ paste0("p",    pair_ids), all_of(post_cols))

# Compute margin and entropy per focal node if available in node_site_metrics
# If JSONL data are present, join margin/entropy per (gene, node_id, site)
if (!is.null(node_site_metrics)) {
  # Pivot JSONL metrics for each pair's node
  # We need to match (gene, site, node_id) → margin_i, entropy_i
  jm_sel <- node_site_metrics |>
    dplyr::select(gene, node, site, margin, entropy) |>
    dplyr::rename(jm_node_id = node, jm_margin = margin, jm_entropy = entropy)

  for (i in seq_along(pair_ids)) {
    pid <- pair_ids[[i]]
    nc  <- paste0("node", pid)
    # node column may be numeric or character; cast to character for join
    focal_df <- focal_df |>
      dplyr::mutate(!!nc := as.character(.data[[nc]])) |>
      dplyr::left_join(
        jm_sel |>
          dplyr::mutate(jm_node_id = as.character(jm_node_id)),
        by = c("gene" = "gene", "site" = "site", !!nc := "jm_node_id")
      ) |>
      dplyr::rename(
        !!paste0("margin",  pid) := jm_margin,
        !!paste0("entropy", pid) := jm_entropy
      )
  }
}

# Compute aggregate focal-node stats
p_cols  <- paste0("p",      pair_ids)
m_cols  <- paste0("margin",  pair_ids)
e_cols  <- paste0("entropy", pair_ids)

focal_df <- focal_df |>
  dplyr::rowwise() |>
  dplyr::mutate(
    p_min    = min(dplyr::c_across(all_of(p_cols)),  na.rm = TRUE),
    p_prod   = prod(dplyr::c_across(all_of(p_cols)), na.rm = TRUE)
  )

if (all(m_cols %in% names(focal_df))) {
  focal_df <- focal_df |>
    dplyr::mutate(
      margin_min  = min(dplyr::c_across(all_of(m_cols)),  na.rm = TRUE),
      entropy_max = max(dplyr::c_across(all_of(e_cols)),  na.rm = TRUE)
    )
}

focal_df <- dplyr::ungroup(focal_df)
write_tsv_safe(focal_df, "asr_focal_node_sites.tsv")

focal_df |> head(10) |> kbl_pp(caption = "§4 Focal-node site posteriors (asr_focal_node_sites.tsv)")
```

---

## §5 – Site filtering

The filtering rule is:

$$\text{PASS} \iff p_{\min} \geq \tau \quad\text{where}\quad \tau = \texttt{params.ct\_disambig\_posterior\_threshold} = `r posterior_threshold`$$

```{r site_filter}
tau <- posterior_threshold

filter_df <- focal_df |>
  dplyr::select(gene, site, all_of(p_cols), p_min, p_prod) |>
  dplyr::mutate(
    PASS            = p_min >= tau,
    threshold_used  = tau
  )

write_tsv_safe(filter_df, "asr_site_filtering.tsv")

n_pass <- sum(filter_df$PASS,  na.rm = TRUE)
n_fail <- sum(!filter_df$PASS, na.rm = TRUE)

cat(sprintf(
  "τ = %.4f | Sites PASS: %d  |  Sites FAIL: %d  |  Retention: %.1f%%\n",
  tau, n_pass, n_fail, 100 * n_pass / nrow(filter_df)
))

filter_df |> head(10) |> kbl_pp(caption = "§5 Site-level filtering (asr_site_filtering.tsv)")
```

---

## §6 – Retention statistics

```{r retention_stats}
# ── Global ────────────────────────────────────────────────────────────────────
retention_global <- tibble(
  N_sites_total      = nrow(filter_df),
  N_sites_retained   = n_pass,
  N_sites_excluded   = n_fail,
  retention_fraction = n_pass / nrow(filter_df),
  exclusion_fraction = n_fail / nrow(filter_df),
  threshold_used     = tau
)
write_tsv_safe(retention_global, "asr_retention_summary.tsv")

# ── Per gene ──────────────────────────────────────────────────────────────────
retention_gene <- filter_df |>
  dplyr::group_by(gene) |>
  dplyr::summarise(
    sites_total        = dplyr::n(),
    sites_retained     = sum(PASS, na.rm = TRUE),
    sites_excluded     = sum(!PASS, na.rm = TRUE),
    retention_fraction = sites_retained / sites_total,
    exclusion_fraction = sites_excluded / sites_total,
    threshold_used     = tau,
    .groups = "drop"
  )
write_tsv_safe(retention_gene, "asr_gene_retention.tsv")

kbl_pp(retention_global, caption = "§6 Global retention summary (asr_retention_summary.tsv)")
```

```{r retention_gene_tbl}
kbl_pp(retention_gene, caption = "§6 Per-gene retention (asr_gene_retention.tsv)")
```

---

## §7 – Sensitivity analysis

*The pipeline always filters using `params.ct_disambig_posterior_threshold`. The tau grid below is a diagnostic summary only and does not affect any pipeline decision.*

```{r sensitivity}
tau_grid <- TAU_GRID  # c(0.90, 0.95, 0.99) — diagnostic only

apply_tau <- function(fd, t) {
  kept <- fd |> dplyr::filter(p_min >= t)
  tibble(
    tau                = t,
    N_sites_total      = nrow(fd),
    N_sites_retained   = nrow(kept),
    retention_fraction = nrow(kept) / nrow(fd),
    N_convergent_sites = dplyr::n_distinct(paste(kept$gene, kept$site)),
    N_convergent_genes = dplyr::n_distinct(kept$gene)
  )
}

sensitivity_df <- dplyr::bind_rows(lapply(tau_grid, function(t) apply_tau(filter_df, t)))
write_tsv_safe(sensitivity_df, "convergence_sensitivity.tsv")

# ── Callset overlap (Jaccard) ─────────────────────────────────────────────────
callset <- function(t) {
  filter_df |>
    dplyr::filter(p_min >= t) |>
    dplyr::mutate(key = paste(gene, site, sep = ":")) |>
    dplyr::pull(key)
}

pairs <- list(
  c("0.90", "0.95"),
  c("0.95", "0.99"),
  c("0.90", "0.99")
)

overlap_rows <- lapply(pairs, function(pr) {
  t1 <- as.numeric(pr[[1]]);  t2 <- as.numeric(pr[[2]])
  s1 <- callset(t1);           s2 <- callset(t2)
  inter <- length(intersect(s1, s2))
  union <- length(union(s1, s2))
  tibble(
    tau_A        = t1,
    tau_B        = t2,
    n_A          = length(s1),
    n_B          = length(s2),
    n_intersect  = inter,
    n_union      = union,
    jaccard      = if (union > 0) inter / union else NA_real_
  )
})

overlap_df <- dplyr::bind_rows(overlap_rows)
write_tsv_safe(overlap_df, "convergence_overlap.tsv")

kbl_pp(sensitivity_df, caption = "§7 Sensitivity analysis (convergence_sensitivity.tsv)")
```

```{r overlap_tbl}
kbl_pp(overlap_df, caption = "§7 Callset overlap between tau thresholds (convergence_overlap.tsv)")
```

---

## §8 – Confidence diagnostics (focal nodes)

```{r diagnostics}
# Descriptive summaries of margin, entropy, p_prod at focal nodes
diag_rows <- list()

if ("margin_min" %in% names(focal_df) && "entropy_max" %in% names(focal_df)) {
  diag_rows[["margin_min"]] <- summary(focal_df$margin_min)
  diag_rows[["entropy_max"]] <- summary(focal_df$entropy_max)
}
diag_rows[["p_prod"]]  <- summary(focal_df$p_prod)
diag_rows[["p_min"]]   <- summary(focal_df$p_min)

diag_df <- dplyr::bind_rows(
  lapply(names(diag_rows), function(nm) {
    s <- diag_rows[[nm]]
    tibble(
      metric  = nm,
      Min     = s[["Min."]],
      Q1      = s[["1st Qu."]],
      Median  = s[["Median"]],
      Mean    = s[["Mean"]],
      Q3      = s[["3rd Qu."]],
      Max     = s[["Max."]]
    )
  })
)

# Proportion of sites below margin thresholds and above entropy quantiles
extra_rows <- list()

if ("margin_min" %in% names(focal_df)) {
  extra_rows[["fraction_sites_margin_lt_0.05"]] <- mean(focal_df$margin_min < 0.05, na.rm = TRUE)
  extra_rows[["fraction_sites_margin_lt_0.10"]] <- mean(focal_df$margin_min < 0.10, na.rm = TRUE)
}

if ("entropy_max" %in% names(focal_df)) {
  ent_q <- quantile(focal_df$entropy_max, probs = c(0.75, 0.90, 0.95), na.rm = TRUE)
  extra_rows[["fraction_sites_entropy_gt_q75"]] <- mean(focal_df$entropy_max > ent_q[["75%"]], na.rm = TRUE)
  extra_rows[["fraction_sites_entropy_gt_q90"]] <- mean(focal_df$entropy_max > ent_q[["90%"]], na.rm = TRUE)
  extra_rows[["fraction_sites_entropy_gt_q95"]] <- mean(focal_df$entropy_max > ent_q[["95%"]], na.rm = TRUE)
  # Record the actual quantile values for reference
  extra_rows[["entropy_q75_value"]] <- as.numeric(ent_q[["75%"]])
  extra_rows[["entropy_q90_value"]] <- as.numeric(ent_q[["90%"]])
  extra_rows[["entropy_q95_value"]] <- as.numeric(ent_q[["95%"]])
}

if (length(extra_rows) > 0) {
  extra_df <- tibble(
    metric = names(extra_rows),
    Min    = as.numeric(extra_rows),
    Q1     = NA_real_, Median = NA_real_, Mean = NA_real_, Q3 = NA_real_, Max = NA_real_
  )
  diag_full_df <- dplyr::bind_rows(diag_df, extra_df)
} else {
  diag_full_df <- diag_df
}

write_tsv_safe(diag_full_df, "asr_confidence_diagnostics.tsv")
kbl_pp(diag_full_df, caption = "§8 Confidence diagnostics at focal nodes (asr_confidence_diagnostics.tsv)")
```

---

## §9 – Diagnostic plots

### Plot 1 — Global MAP distribution

```{r plot1_global_map, eval=jsonl_files_exist, fig.height=5}
p_full <- ggplot(node_site_metrics, aes(x = MAP)) +
  geom_histogram(binwidth = 0.01, fill = "#4C8FC0", colour = NA, alpha = 0.8) +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.1)) +
  labs(
    title    = "Global MAP posterior distribution (all nodes × sites)",
    subtitle = sprintf("%s genes | %s total observations",
                       scales::comma(dplyr::n_distinct(node_site_metrics$gene)),
                       scales::comma(nrow(node_site_metrics))),
    x = "MAP posterior",
    y = "Count"
  ) +
  theme_bw(base_size = 12)

# Inset: uncertainty tail (MAP < 1.0)
p_inset <- ggplot(node_site_metrics |> dplyr::filter(MAP < 1.0), aes(x = MAP)) +
  geom_histogram(binwidth = 0.01, fill = "#E07B39", colour = NA, alpha = 0.85) +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  labs(title = "MAP < 1.0 (uncertainty tail)", x = NULL, y = NULL) +
  theme_bw(base_size = 9) +
  theme(plot.title = element_text(size = 8))

p1 <- p_full +
  inset_element(p_inset, left = 0.02, bottom = 0.45, right = 0.48, top = 0.98)

save_plot(p1, "plot_asr_global_map_distribution.png", width = 11, height = 5)
p1
```

### Plot 2 — Uncertainty-tail distribution by gene

```{r plot2_tail_by_gene, eval=jsonl_files_exist, fig.height=6}
tail_gene_df <- node_site_metrics |> dplyr::filter(MAP < 1.0)

p2 <- ggplot(tail_gene_df, aes(x = reorder(gene, MAP, median), y = MAP)) +
  geom_violin(fill = "#4C8FC0", colour = NA, alpha = 0.7, scale = "width") +
  geom_boxplot(width = 0.15, outlier.size = 0.4, colour = "grey30", fill = "white") +
  coord_flip() +
  labs(
    title    = "MAP distribution (uncertainty tail: MAP < 1.0) by gene",
    subtitle = "Violin + boxplot; genes sorted by median MAP",
    x = NULL,
    y = "MAP posterior"
  ) +
  theme_bw(base_size = 12)

save_plot(p2, "plot_asr_uncertain_tail_by_gene.png", width = 11, height = max(6, 0.4 * dplyr::n_distinct(tail_gene_df$gene)))
p2
```

### Plot 3 — Fraction of uncertain sites by gene

```{r plot3_uncertainty_fractions, eval=jsonl_files_exist, fig.height=5}
unc_fracs <- gene_stats_df |>
  dplyr::select(gene, frac_MAP_lt_100, frac_MAP_lt_095, frac_MAP_lt_090, frac_MAP_lt_080) |>
  dplyr::rename(
    `< 1.00` = frac_MAP_lt_100,
    `< 0.95` = frac_MAP_lt_095,
    `< 0.90` = frac_MAP_lt_090,
    `< 0.80` = frac_MAP_lt_080
  ) |>
  tidyr::pivot_longer(-gene, names_to = "Threshold", values_to = "Fraction") |>
  dplyr::mutate(Threshold = factor(Threshold, levels = c("< 1.00", "< 0.95", "< 0.90", "< 0.80")))

p3 <- ggplot(unc_fracs, aes(x = reorder(gene, Fraction), y = Fraction, fill = Threshold)) +
  geom_col(position = "dodge", alpha = 0.85) +
  coord_flip() +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Blues", direction = -1) +
  labs(
    title = "Fraction of uncertain sites per gene",
    x     = NULL,
    y     = "Fraction of sites",
    fill  = "MAP threshold"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "right")

save_plot(p3, "plot_asr_uncertainty_fractions_by_gene.png", width = 11, height = 5)
p3
```

### Plot 4 — Focal-node p_min distribution

```{r plot4_pmin, fig.height=5}
p4 <- ggplot(focal_df, aes(x = p_min)) +
  geom_histogram(binwidth = 0.01, fill = "#4C8FC0", colour = NA, alpha = 0.85) +
  geom_vline(xintercept = posterior_threshold,
             colour = "#D62728", linewidth = 0.9, linetype = "dashed") +
  annotate("text",
           x     = posterior_threshold - 0.01,
           y     = Inf,
           label = sprintf("τ = %.2f", posterior_threshold),
           colour = "#D62728", hjust = 1, vjust = 1.5, size = 3.5) +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.1)) +
  labs(
    title    = "Distribution of p_min across focal MRCA nodes",
    subtitle = "Dashed line = params.ct_disambig_posterior_threshold",
    x        = "p_min (minimum MAP across focal nodes for each site)",
    y        = "Count"
  ) +
  theme_bw(base_size = 12)

save_plot(p4, "plot_asr_focal_pmin_distribution.png", width = 11, height = 5)
p4
```

### Plot 5 — Site retention by gene

```{r plot5_retention, fig.height=5}
p5 <- ggplot(
  retention_gene,
  aes(x = reorder(gene, retention_fraction), y = retention_fraction)
) +
  geom_col(fill = "#4C8FC0", alpha = 0.85) +
  geom_hline(yintercept = retention_global$retention_fraction,
             colour = "#D62728", linewidth = 0.8, linetype = "dashed") +
  coord_flip() +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  labs(
    title    = sprintf("Site retention fraction per gene (τ = %.2f)", tau),
    subtitle = "Dashed line = global retention fraction",
    x        = NULL,
    y        = "Retention fraction"
  ) +
  theme_bw(base_size = 12)

save_plot(p5, "plot_asr_gene_retention.png", width = 11, height = max(5, 0.4 * nrow(retention_gene)))
p5
```

### Plot 6 — Sensitivity: retained sites vs threshold

```{r plot6_retention_sensitivity, fig.height=5}
p6 <- ggplot(sensitivity_df, aes(x = tau, y = retention_fraction)) +
  geom_line(colour = "#4C8FC0", linewidth = 1.1) +
  geom_point(colour = "#4C8FC0", size = 3) +
  geom_vline(xintercept = posterior_threshold,
             colour = "#D62728", linewidth = 0.8, linetype = "dashed") +
  annotate("text",
           x = posterior_threshold + 0.002, y = 0.95,
           label = sprintf("pipeline τ = %.2f", posterior_threshold),
           colour = "#D62728", hjust = 0, size = 3.5) +
  scale_x_continuous(breaks = TAU_GRID) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1.05)) +
  labs(
    title    = "Sensitivity of site retention to posterior threshold",
    subtitle = "Global retention fraction across tau grid (diagnostic only)",
    x        = "Posterior threshold (τ)",
    y        = "Retention fraction"
  ) +
  theme_bw(base_size = 12)

# Add per-gene lines if not too many genes
per_gene_sens <- dplyr::bind_rows(lapply(tau_grid, function(t) {
  filter_df |>
    dplyr::group_by(gene) |>
    dplyr::summarise(
      tau                = t,
      retention_fraction = mean(p_min >= t, na.rm = TRUE),
      .groups = "drop"
    )
}))

if (dplyr::n_distinct(per_gene_sens$gene) <= 20) {
  p6 <- p6 +
    geom_line(data = per_gene_sens,
              aes(x = tau, y = retention_fraction, group = gene),
              colour = "grey70", linewidth = 0.5, alpha = 0.6)
}

save_plot(p6, "plot_asr_retention_sensitivity.png", width = 11, height = 5)
p6
```

### Plot 7 — Sensitivity of convergence calls to threshold

```{r plot7_convergence_sensitivity, fig.height=5}
p7a <- ggplot(sensitivity_df, aes(x = tau, y = N_convergent_sites)) +
  geom_line(colour = "#4C8FC0", linewidth = 1.1) +
  geom_point(colour = "#4C8FC0", size = 3) +
  geom_vline(xintercept = posterior_threshold,
             colour = "#D62728", linewidth = 0.8, linetype = "dashed") +
  scale_x_continuous(breaks = TAU_GRID) +
  labs(title = "Convergent sites", x = "Threshold (τ)", y = "N convergent sites") +
  theme_bw(base_size = 11)

p7b <- ggplot(sensitivity_df, aes(x = tau, y = N_convergent_genes)) +
  geom_line(colour = "#E07B39", linewidth = 1.1) +
  geom_point(colour = "#E07B39", size = 3) +
  geom_vline(xintercept = posterior_threshold,
             colour = "#D62728", linewidth = 0.8, linetype = "dashed") +
  scale_x_continuous(breaks = TAU_GRID) +
  labs(title = "Convergent genes", x = "Threshold (τ)", y = "N convergent genes") +
  theme_bw(base_size = 11)

p7 <- p7a + p7b +
  plot_annotation(
    title    = "Sensitivity of convergence calls to posterior threshold",
    subtitle = "Diagnostic only — pipeline always uses params.ct_disambig_posterior_threshold",
    theme    = theme(plot.title = element_text(size = 13))
  )

save_plot(p7, "plot_convergence_sensitivity.png", width = 13, height = 5)
p7
```

### Plots 8–9 — Margin and entropy by gene (focal nodes)

```{r plot8_margin, eval="margin_min" %in% names(focal_df), fig.height=5}
p8 <- ggplot(focal_df, aes(x = reorder(gene, margin_min, median), y = margin_min)) +
  geom_violin(fill = "#7BAD6E", colour = NA, alpha = 0.7, scale = "width") +
  geom_boxplot(width = 0.15, outlier.size = 0.4, colour = "grey30", fill = "white") +
  coord_flip() +
  labs(
    title = "MAP margin (MAP − P_second) at focal nodes, per gene",
    x     = NULL,
    y     = "Minimum margin across focal nodes"
  ) +
  theme_bw(base_size = 12)

save_plot(p8, "plot_asr_margin_by_gene.png", width = 11, height = max(5, 0.4 * dplyr::n_distinct(focal_df$gene)))
p8
```

```{r plot9_entropy, eval="entropy_max" %in% names(focal_df), fig.height=5}
p9 <- ggplot(focal_df, aes(x = reorder(gene, entropy_max, median), y = entropy_max)) +
  geom_violin(fill = "#C97DB0", colour = NA, alpha = 0.7, scale = "width") +
  geom_boxplot(width = 0.15, outlier.size = 0.4, colour = "grey30", fill = "white") +
  coord_flip() +
  labs(
    title = "Shannon entropy at focal nodes (nats), per gene",
    x     = NULL,
    y     = "Maximum entropy across focal nodes"
  ) +
  theme_bw(base_size = 12)

save_plot(p9, "plot_asr_entropy_by_gene.png", width = 11, height = max(5, 0.4 * dplyr::n_distinct(focal_df$gene)))
p9
```

---

## Output file summary

```{r output_summary, echo=FALSE}
tsv_files   <- list.files(tsv_dir,   pattern = "\\.tsv$", full.names = FALSE)
plot_files  <- list.files(plots_dir, pattern = "\\.png$", full.names = FALSE)

tibble(
  type = c(rep("TSV", length(tsv_files)), rep("Plot", length(plot_files))),
  file = c(tsv_files, plot_files)
) |>
  kbl_pp(caption = "All output files generated")
```

---

*Generated by PhyloPhere — ASR Robustness Module*  
*`params.ct_disambig_posterior_threshold` = **`r posterior_threshold`***
