---
title: "CT Accumulation Report"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
params:
  accum_dir:      "."       # path to accumulation/ root (contains top/, bottom/, all/, aggregation/)
  traitname:      "unknown_trait"
  fdr_threshold:  0.1
  randomization_type: "cons_decile"
  output_dir:     "."
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

<!-- 10.Accumulation_report.Rmd — Per-gene CAAS accumulation results by direction and grouping scheme. -->
<!-- PhyloPhere | subworkflows/CT_ACCUMULATION/local/ -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, out.width = "100%")

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggrepel)
  library(DT)
  library(paletteer)
  library(stringr)
  library(purrr)
  library(readr)
  library(knitr)
  library(kableExtra)
})

.tableau10 <- as.character(paletteer_d("ggthemes::Tableau_10"))
dir_colours <- c(top = .tableau10[1], bottom = .tableau10[2], all = .tableau10[3])

theme_set(
  theme_bw(base_size = 16) +
  theme(
    axis.title.y  = element_text(margin = ggplot2::margin(l = 20, r = 20)),
    axis.title.x  = element_text(margin = ggplot2::margin(t = 12)),
    axis.text.x   = element_text(margin = ggplot2::margin(t = 5)),
    plot.title    = element_text(face = "bold", hjust = 0, margin = ggplot2::margin(b = 6)),
    plot.subtitle = element_text(colour = "grey40", hjust = 0, margin = ggplot2::margin(b = 10))
  )
)
# Order of schemes: accumulation evaluates the unweighted scheme (US) only
scheme_order <- c("us")

fdr_thr   <- params$fdr_threshold
pval_thr  <- fdr_thr
accum_dir <- params$accum_dir
```

```{r load-data}
# ── Load all per-scheme CSVs across directions ────────────────────────────────
directions <- c("top", "bottom", "all")
schemes    <- scheme_order

load_scheme_csv <- function(direction, scheme) {
  path <- file.path(accum_dir, direction, "randomization",
                    sprintf("accumulation_%s_%s_aggregated_results.csv", direction, scheme))
  if (!file.exists(path)) return(NULL)
  df <- read_csv(path, show_col_types = FALSE)
  # Drop the _<scheme> suffix of the columns (ActualCount_us becomes ActualCount)
  df <- df %>%
    rename_with(~ str_remove(.x, paste0("_", scheme, "$"))) %>%
    mutate(direction = direction, scheme = scheme)
  df
}

raw <- map_dfr(directions, function(d) map_dfr(schemes, ~ load_scheme_csv(d, .x)))

has_data <- nrow(raw) > 0
if (!has_data) stop("No accumulation CSVs found under: ", accum_dir)

# Standardize the gene column name (Gene / gene)
if ("Gene" %in% names(raw) && !"gene" %in% names(raw)) raw <- rename(raw, gene = Gene)

# Accumulation evaluates the unweighted scheme (US) only; biochemical GS schemes
# describe positions and are excluded from accumulation.
raw <- raw %>% filter(scheme == "us")

# ── BH FDR per direction × scheme, excluding genes with ActualCount = 0 ───────
# Genes with ActualCount = 0 have p = 1 by construction, not by test, and must not
# enter the FDR denominator: they would inflate it and reduce power.
accum <- raw %>%
  mutate(
    scheme    = factor(scheme, levels = scheme_order),
    direction = factor(direction, levels = directions)
  ) %>%
  group_by(direction, scheme) %>%
  mutate(
    FDR = {
      tested <- !is.na(ActualCount) & ActualCount > 0 & !is.na(PValueEmpirical)
      q      <- rep(NA_real_, n())
      if (any(tested))
        q[tested] <- p.adjust(PValueEmpirical[tested], method = "BH")
      q
    }
  ) %>%
  ungroup() %>%
  mutate(significant = !is.na(FDR) & FDR < pval_thr)

# Global position table of the aggregation, if present
global_path  <- file.path(accum_dir, "aggregation", "accumulation_global.csv")
has_global   <- file.exists(global_path)
if (has_global)  global_df  <- read_csv(global_path,  show_col_types = FALSE)
```

---

```{r rand-type-caveat, results='asis'}
rand_type <- if (!is.null(params$randomization_type)) params$randomization_type else "cons_decile"
if (identical(rand_type, "permulation")) {
  cat(
    "> **Null model: permulation.** The randomised CAAS set for each gene at each ",
    "cycle is that gene's ACTUAL detections replayed from a prior CAAS permulation ",
    "run (the same phenotype-label permutations CAAS's own genome-wide null uses ",
    "elsewhere), not a synthetic draw from a conservation- or position-matched pool. ",
    "This holds the phenotype-tree confound but is **not** conservation-decile ",
    "matched -- it answers a different question than `cons_decile`, not a stricter ",
    "version of it. The per-gene CAAS count is free to vary per cycle (it is not ",
    "pinned to the observed count, as an occupancy null would pin it), so this is ",
    "an *excess* null: a gene is significant when it accumulates more CAAS across ",
    "permuted labelings than its own null replay would predict.\n\n",
    sep = ""
  )
}
```

## Overview

```{r overview-table}
summary_tbl <- accum %>%
  group_by(direction, scheme) %>%
  summarise(
    n_genes        = n_distinct(gene),
    n_significant  = sum(significant, na.rm = TRUE),
    pct_significant = round(100 * mean(significant, na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  arrange(direction, scheme)

DT::datatable(
  summary_tbl,
  caption  = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Significant genes (FDR < %g) per direction × scheme | Trait: %s",
            fdr_thr, params$traitname)
  ),
  rownames = FALSE,
  options  = list(pageLength = 21, dom = "t")
) %>%
  DT::formatStyle("n_significant",
    background = DT::styleColorBar(range(summary_tbl$n_significant), "#cce5ff"),
    backgroundSize = "98% 88%", backgroundRepeat = "no-repeat",
    backgroundPosition = "center"
  )
```

---

## Actual vs Simulated count distributions {.tabset}

```{r scatter-tabs, results='asis'}
for (dir in directions) {
  cat(sprintf("\n\n### %s\n\n", str_to_title(dir)))
  sub <- accum %>% filter(direction == dir)
  if (nrow(sub) == 0) { cat("*No data.*\n\n"); next }

  p <- ggplot(sub, aes(x = MeanSim, y = ActualCount, colour = significant)) +
    geom_point(alpha = 0.5, size = 1.2) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
    geom_text_repel(
      data = sub %>% filter(significant) %>% arrange(PValueEmpirical) %>% slice_head(n = 20),
      aes(label = gene), size = 2.5, max.overlaps = 20, show.legend = FALSE
    ) +
    scale_colour_manual(values = c("FALSE" = "grey70", "TRUE" = dir_colours[[dir]])) +
    labs(
      x      = "Mean simulated count",
      y      = "Actual count",
      colour = sprintf("FDR < %g", fdr_thr),
      title  = sprintf("Direction: %s | Trait: %s", str_to_title(dir), params$traitname)
    ) +
    theme_bw(base_size = 16) +
    theme(legend.position = "bottom")

  print(p)
  cat("\n\n")
}
```

---

## Empirical p-value distributions {.tabset}

```{r pval-tabs, results='asis'}
for (dir in directions) {
  cat(sprintf("\n\n### %s\n\n", str_to_title(dir)))
  sub <- accum %>% filter(direction == dir, !is.na(PValueEmpirical), !is.na(ActualCount), ActualCount > 0)
  if (nrow(sub) == 0) { cat("*No data.*\n\n"); next }

  p <- ggplot(sub, aes(x = PValueEmpirical, fill = scheme)) +
    geom_histogram(bins = 40, alpha = 0.7, colour = "white") +
    geom_vline(xintercept = fdr_thr, linetype = "dashed", colour = "red") +
    scale_fill_manual(values = setNames(.tableau10[seq_along(scheme_order)], scheme_order)) +
    labs(x = "Empirical p-value", y = "Gene count",
         title = sprintf("Direction: %s", str_to_title(dir))) +
    theme_bw(base_size = 16) +
    theme(legend.position = "none")

  print(p)
  cat("\n\n")
}
```

---

## Significant genes {.tabset}

> Genes with at least one observed CAAS (ActualCount > 0) are tested; significance is evaluated by empirical p-value under the Unweighted Scheme (US), adjusted by Benjamini-Hochberg FDR (`FDR < 0.05`).

```{r sig-tables, results='asis'}
for (dir in directions) {
  cat(sprintf("\n\n### %s\n\n", str_to_title(dir)))
  sig <- accum %>%
    filter(direction == dir, significant) %>%
    select(gene, scheme, ActualCount, MeanSim, SD, PValueEmpirical, FDR, RandsAbove, RandsBelow) %>%
    arrange(FDR, gene)

  if (nrow(sig) == 0) { cat("*No significant genes.*\n\n"); next }

  print(
    knitr::kable(
      sig, rownames = FALSE, digits = 4, format = "html", escape = TRUE,
      caption = sprintf("Significant genes | direction: %s | FDR < %g", str_to_title(dir), fdr_thr)
    ) %>% kableExtra::kable_styling(bootstrap_options = c("striped", "condensed", "hover"), full_width = TRUE)
  )
  cat("\n\n")
}
```

---

```{r write-summary}
# Summary TSV: one row per significant (gene, direction)
sig_summary <- accum %>%
  filter(significant) %>%
  group_by(gene, direction) %>%
  summarise(
    n_schemes_significant = n(),
    min_pval              = min(PValueEmpirical, na.rm = TRUE),
    min_fdr               = min(FDR, na.rm = TRUE),
    schemes               = paste(sort(unique(as.character(scheme))), collapse = ","),
    .groups               = "drop"
  ) %>%
  arrange(direction, min_fdr)

out_tsv <- file.path(params$output_dir, paste0("accumulation_summary_", params$traitname, ".tsv"))
write_tsv(sig_summary, out_tsv)

invisible(NULL)
```

---

## Summary

This report summarizes the results of the **CT Accumulation** analysis.

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — CT Accumulation Module*
*Date of generation: `r Sys.Date()`*
