---
title: "CT Accumulation Report"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
params:
  accum_dir:      "."       # path to accumulation/ root (contains top/, bottom/, all/, aggregation/)
  traitname:      "unknown_trait"
  pval_threshold: 0.05
  output_dir:     "."
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, out.width = "100%")

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggrepel)
  library(DT)
  library(paletteer)
  library(stringr)
  library(purrr)
  library(readr)
  library(knitr)
  library(kableExtra)
})

.tableau10 <- as.character(paletteer_d("ggthemes::Tableau_10"))
dir_colours <- c(top = .tableau10[1], bottom = .tableau10[2], all = .tableau10[3])

theme_set(
  theme_bw(base_size = 16) +
  theme(
    axis.title.y  = element_text(margin = ggplot2::margin(l = 20, r = 20)),
    axis.title.x  = element_text(margin = ggplot2::margin(t = 12)),
    axis.text.x   = element_text(margin = ggplot2::margin(t = 5)),
    plot.title    = element_text(face = "bold", hjust = 0, margin = ggplot2::margin(b = 6)),
    plot.subtitle = element_text(colour = "grey40", hjust = 0, margin = ggplot2::margin(b = 10))
  )
)
scheme_order <- c("us", "gs1", "gs2", "gs3", "gs4", "fisher")

pval_thr  <- params$pval_threshold
accum_dir <- params$accum_dir
```

```{r load-data}
# ── Load all per-scheme CSVs across directions ────────────────────────────────
directions <- c("top", "bottom", "all")
schemes    <- scheme_order

load_scheme_csv <- function(direction, scheme) {
  path <- file.path(accum_dir, direction, "randomization",
                    sprintf("accumulation_%s_%s_aggregated_results.csv", direction, scheme))
  if (!file.exists(path)) return(NULL)
  df <- read_csv(path, show_col_types = FALSE)
  # Rename scheme-suffixed columns to generic names
  df <- df %>%
    rename_with(~ str_remove(.x, paste0("_", scheme, "$"))) %>%
    mutate(direction = direction, scheme = scheme)
  df
}

raw <- map_dfr(directions, function(d) map_dfr(schemes, ~ load_scheme_csv(d, .x)))

has_data <- nrow(raw) > 0
if (!has_data) stop("No accumulation CSVs found under: ", accum_dir)

# Standardise column names (Gene / gene)
if ("Gene" %in% names(raw) && !"gene" %in% names(raw)) raw <- rename(raw, gene = Gene)

# ── Fisher's combined p-value across the 5 group schemes ─────────────────────
# Replaces full_pool: each group draws independently so their p-values are
# uncorrelated, making Fisher's χ²= -2Σln(p) ~ χ²(2k) valid.
# Groups with ActualCount=0 contribute p=1 → ln(1)=0, adding no spurious signal.
.fisher_combine <- function(ps) {
  ps <- pmax(ps, 1e-300)   # guard log(0)
  pchisq(-2 * sum(log(ps)), df = 2 * length(ps), lower.tail = FALSE)
}

group_schemes <- c("us", "gs1", "gs2", "gs3", "gs4")

fisher_df <- raw %>%
  filter(scheme %in% group_schemes, !is.na(PValueEmpirical)) %>%
  group_by(gene, direction) %>%
  summarise(
    PValueEmpirical = .fisher_combine(PValueEmpirical),
    ActualCount     = sum(ActualCount, na.rm = TRUE),
    # Pooled under independence: sums combine, variances add, RandsAbove/RandsBelow
    # (order statistics of a single group's null draws) don't carry a pooled meaning
    # across independently-drawn groups and are therefore omitted (kept NA).
    MeanSim         = sum(MeanSim, na.rm = TRUE),
    SD              = sqrt(sum(SD^2, na.rm = TRUE)),
    RandsAbove      = NA_integer_,
    RandsBelow      = NA_integer_,
    NumRands        = max(NumRands, na.rm = TRUE),
    scheme          = "fisher",
    .groups         = "drop"
  )

# Drop full_pool rows if present (legacy runs), append Fisher
raw <- bind_rows(raw %>% filter(scheme != "full_pool"), fisher_df)

# ── BH FDR — applied per direction × scheme, excluding ActualCount=0 genes ───
# Genes with ActualCount=0 have p=1 by construction (not by test) and must not
# enter the FDR denominator — they would inflate it and reduce power.
accum <- raw %>%
  mutate(
    scheme    = factor(scheme, levels = scheme_order),
    direction = factor(direction, levels = directions)
  ) %>%
  group_by(direction, scheme) %>%
  mutate(
    FDR = {
      tested <- !is.na(ActualCount) & ActualCount > 0 & !is.na(PValueEmpirical)
      q      <- rep(NA_real_, n())
      if (any(tested))
        q[tested] <- p.adjust(PValueEmpirical[tested], method = "BH")
      q
    }
  ) %>%
  ungroup() %>%
  mutate(significant = !is.na(FDR) & FDR < pval_thr)

# Load aggregation-level files if present
global_path  <- file.path(accum_dir, "aggregation", "accumulation_global.csv")
deciles_path <- file.path(accum_dir, "aggregation", "accumulation_deciles.csv")
has_global   <- file.exists(global_path)
has_deciles  <- file.exists(deciles_path)
if (has_global)  global_df  <- read_csv(global_path,  show_col_types = FALSE)
if (has_deciles) deciles_df <- read_csv(deciles_path, show_col_types = FALSE)
```

---

## Overview

```{r overview-table}
summary_tbl <- accum %>%
  group_by(direction, scheme) %>%
  summarise(
    n_genes        = n_distinct(gene),
    n_significant  = sum(significant, na.rm = TRUE),
    pct_significant = round(100 * mean(significant, na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  arrange(direction, scheme)

DT::datatable(
  summary_tbl,
  caption  = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Significant genes (p < %g) per direction × scheme | Trait: %s",
            pval_thr, params$traitname)
  ),
  rownames = FALSE,
  options  = list(pageLength = 21, dom = "t")
) %>%
  DT::formatStyle("n_significant",
    background = DT::styleColorBar(range(summary_tbl$n_significant), "#cce5ff"),
    backgroundSize = "98% 88%", backgroundRepeat = "no-repeat",
    backgroundPosition = "center"
  )
```

---

## Actual vs Simulated count distributions {.tabset}

```{r scatter-tabs, results='asis'}
for (dir in directions) {
  cat(sprintf("\n\n### %s\n\n", str_to_title(dir)))
  sub <- accum %>% filter(direction == dir, scheme != "fisher")
  if (nrow(sub) == 0) { cat("*No data.*\n\n"); next }

  p <- ggplot(sub, aes(x = MeanSim, y = ActualCount, colour = significant)) +
    geom_point(alpha = 0.5, size = 1.2) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
    geom_text_repel(
      data = sub %>% filter(significant) %>% arrange(PValueEmpirical) %>% slice_head(n = 20),
      aes(label = gene), size = 2.5, max.overlaps = 20, show.legend = FALSE
    ) +
    scale_colour_manual(values = c("FALSE" = "grey70", "TRUE" = dir_colours[[dir]])) +
    facet_wrap(~ scheme, nrow = 2) +
    labs(
      x      = "Mean simulated count",
      y      = "Actual count",
      colour = sprintf("p < %g", pval_thr),
      title  = sprintf("Direction: %s | Trait: %s", str_to_title(dir), params$traitname)
    ) +
    theme_bw(base_size = 16) +
    theme(legend.position = "bottom")

  print(p)
  cat("\n\n")
}
```

---

## Empirical p-value distributions {.tabset}

```{r pval-tabs, results='asis'}
for (dir in directions) {
  cat(sprintf("\n\n### %s\n\n", str_to_title(dir)))
  sub <- accum %>% filter(direction == dir, !is.na(PValueEmpirical), !is.na(ActualCount), ActualCount > 0)
  if (nrow(sub) == 0) { cat("*No data.*\n\n"); next }

  p <- ggplot(sub, aes(x = PValueEmpirical, fill = scheme)) +
    geom_histogram(bins = 40, alpha = 0.7, colour = "white") +
    geom_vline(xintercept = pval_thr, linetype = "dashed", colour = "red") +
    facet_wrap(~ scheme, nrow = 2, scales = "free_y") +
    scale_fill_manual(values = setNames(.tableau10[seq_along(scheme_order)], scheme_order)) +
    labs(x = "Empirical p-value", y = "Gene count",
         title = sprintf("Direction: %s", str_to_title(dir))) +
    theme_bw(base_size = 16) +
    theme(legend.position = "none")

  print(p)
  cat("\n\n")
}
```

---

## Significant genes {.tabset}

> **Note on the `fisher` scheme:** rows for the `fisher` scheme are pooled across the five
> group schemes (`us`, `gs1`-`gs4`) under the assumption that each group's randomization draws
> independently. `MeanSim` and `SD` are combined accordingly (sum of means; SD from summed
> variances), and `PValueEmpirical` is Fisher's combined p-value. `RandsAbove`/`RandsBelow` are
> per-group order statistics that have no valid pooled interpretation across independent groups
> and are therefore omitted (shown as `NA`) for `fisher` rows.

```{r sig-tables, results='asis'}
for (dir in directions) {
  cat(sprintf("\n\n### %s\n\n", str_to_title(dir)))
  sig <- accum %>%
    filter(direction == dir, significant) %>%
    select(gene, scheme, ActualCount, MeanSim, SD, PValueEmpirical, FDR, RandsAbove, RandsBelow) %>%
    arrange(FDR, gene)

  if (nrow(sig) == 0) { cat("*No significant genes.*\n\n"); next }

  print(
    knitr::kable(
      sig, rownames = FALSE, digits = 4, format = "html", escape = TRUE,
      caption = sprintf("Significant genes | direction: %s | p < %g", str_to_title(dir), pval_thr)
    ) %>% kableExtra::kable_styling(bootstrap_options = c("striped", "condensed", "hover"), full_width = TRUE)
  )
  cat("\n\n")
}
```

---

## Gene overlap across schemes {.tabset}

### Overview

```{r overlap}
sig_genes_by_scheme <- accum %>%
  filter(significant) %>%
  group_by(direction, scheme) %>%
  summarise(genes = list(unique(gene)), .groups = "drop")

overlap_tbl <- sig_genes_by_scheme %>%
  group_by(direction) %>%
  summarise(
    schemes_with_hits = sum(map_int(genes, length) > 0),
    union_genes       = n_distinct(unlist(genes)),
    intersection_genes = {
      gl <- genes[map_int(genes, length) > 0]
      if (length(gl) == 0) 0L
      else length(Reduce(intersect, gl))
    },
    .groups = "drop"
  )

knitr::kable(overlap_tbl, rownames = FALSE, caption = "Summary gene overlap counts across schemes per direction") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r overlap-details, results='asis'}
for (dir in directions) {
  cat(sprintf("\n\n### %s Details\n\n", str_to_title(dir)))
  
  sub_sig <- accum %>%
    filter(direction == dir, significant)
  
  if (nrow(sub_sig) == 0) {
    cat("*No significant genes to show overlap for.*\n\n")
    next
  }
  
  overlap_detail <- sub_sig %>%
    mutate(has_sig = "✔") %>%
    pivot_wider(
      id_cols = gene,
      names_from = scheme,
      values_from = has_sig,
      values_fill = ""
    )
  
  all_cols <- c("gene", scheme_order)
  present_cols <- intersect(all_cols, names(overlap_detail))
  overlap_detail <- overlap_detail %>% select(all_of(present_cols))
  
  print(
    knitr::kable(
      overlap_detail, rownames = FALSE, format = "html", escape = TRUE,
      caption = sprintf("Detailed gene overlap matrix across schemes | direction: %s", str_to_title(dir))
    ) %>% kableExtra::kable_styling(bootstrap_options = c("striped", "condensed", "hover"), full_width = TRUE)
  )
  cat("\n\n")
}
```


```{r write-summary}
# Write summary TSV (union of significant genes across all schemes per direction)
sig_summary <- accum %>%
  filter(significant) %>%
  group_by(gene, direction) %>%
  summarise(
    n_schemes_significant = n(),
    min_pval              = min(PValueEmpirical, na.rm = TRUE),
    min_fdr               = min(FDR, na.rm = TRUE),
    schemes               = paste(sort(unique(as.character(scheme))), collapse = ","),
    .groups               = "drop"
  ) %>%
  arrange(direction, min_fdr)

out_tsv <- file.path(params$output_dir, paste0("accumulation_summary_", params$traitname, ".tsv"))
write_tsv(sig_summary, out_tsv)

# Write independent Fisher file (all tested genes, sorted by FDR)
fisher_export <- accum %>%
  filter(scheme == "fisher") %>%
  select(gene, direction, ActualCount, PValueEmpirical, FDR, significant) %>%
  arrange(direction, FDR)

out_fisher <- file.path(params$output_dir, paste0("accumulation_summary_fisher_", params$traitname, ".tsv"))
write_tsv(fisher_export, out_fisher)

invisible(NULL)
```

---

## Summary

This report summarizes the results of the **CT Accumulation** analysis.

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — CT Accumulation Module*
*Date of generation: `r Sys.Date()`*
