---
title: "ASR Path Score Characterization"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
params:
  disambig_dir:        NULL   # path to ct_disambiguation/ output directory
  posterior_threshold: NULL   # params.ct_disambig_posterior_threshold (shown, not applied)
  output_dir:          "."    # working directory (TSVs → ./tsv/, plots → ./plots/)
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

<!-- 9.ASR_diagnostics.Rmd — Distribution of asr_path_score, derived_agreement and the per-domain ancestral posteriors of the scored CAAS rows. -->
<!-- PhyloPhere | subworkflows/ASR_DIAGNOSTICS/local/ -->
<!--
  Called by: ASR_DIAGNOSTICS_REPORT (asr_diagnostics.nf), through rmarkdown::render().
  Params:    disambig_dir (ct_disambiguation/ with caas_convergence_master.csv),
             posterior_threshold (shown in the report, not applied), output_dir.
  A row of the master is one (gene, msa_pos, caap_group, side). asr_path_score is the score of the row on the
  Voronoi domains (fop_pool.py); derived_agreement is a descriptor and does not enter the score;
  domain_<k>_posterior is the posterior of the ancestral state of domain k under the scheme of the row.
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 11,
  fig.height = 6,
  dpi        = 180,
  dev        = "png"
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(ggplot2)
  library(scales)
  library(knitr)
  library(kableExtra)
})

theme_set(
  theme_bw(base_size = 12) +
  theme(
    axis.title.y  = element_text(margin = ggplot2::margin(l = 20, r = 20)),
    axis.title.x  = element_text(margin = ggplot2::margin(t = 12)),
    axis.text.x   = element_text(margin = ggplot2::margin(t = 5)),
    plot.title    = element_text(face = "bold", hjust = 0, margin = ggplot2::margin(b = 6)),
    plot.subtitle = element_text(colour = "grey40", hjust = 0, margin = ggplot2::margin(b = 10)),
    strip.background = element_rect(fill = "grey95", colour = "grey80"),
    strip.text    = element_text(face = "bold")
  )
)

# ── Parameters ────────────────────────────────────────────────────────────────
disambig_dir        <- params$disambig_dir
posterior_threshold <- if (is.null(params$posterior_threshold)) NA_real_ else as.numeric(params$posterior_threshold)
output_dir          <- params$output_dir

stopifnot(!is.null(disambig_dir))

# ── Output directories ────────────────────────────────────────────────────────
# Tables go to tsv/ and figures to plots/, both published by ASR_DIAGNOSTICS_REPORT.
tsv_dir   <- file.path(output_dir, "tsv")
plots_dir <- file.path(output_dir, "plots")
dir.create(tsv_dir,   recursive = TRUE, showWarnings = FALSE)
dir.create(plots_dir, recursive = TRUE, showWarnings = FALSE)

write_tsv_safe <- function(df, fname) {
  readr::write_tsv(df, file.path(tsv_dir, fname))
  invisible(file.path(tsv_dir, fname))
}

save_plot <- function(p, fname, width = 11, height = 6) {
  ggsave(file.path(plots_dir, fname), plot = p, width = width, height = height, dpi = 180)
  invisible(file.path(plots_dir, fname))
}

# Categorical palette (Tableau 10)
.tableau10 <- c(
  "#1F77B4", "#AEC7E8", "#FF7F0E", "#FFBB78",
  "#2CA02C", "#98DF8A", "#D62728", "#FF9896",
  "#9467BD", "#C5B0D5"
)
```

## Introduction

`asr_path_score` is the CAAS convergence score of a position on the K fixed Voronoi domains of the contrast design. It is computed for each scheme (US, GS1 to GS4) and each phenotype side, so a position contributes up to ten rows to the master table.

For one hypothesis and one side, a domain is *changed* when its observed foreground residue, encoded in the scheme, differs from the modal ancestral residue at the MRCA of its pair. A changed domain scores the noisy-OR, over the changed domains that carry the same encoded residue, of `1 - P`, where `P` is the posterior mass of that residue at the LCA of the two domains. `P` is the recorded mass of the residue and, when the residue is not in the recorded distribution of the node, the whole unrecorded mass (worst case). A changed domain without a same-residue partner scores 0. The domain scores are averaged over the hypotheses and the score of the row is their mean weighted by the PSS of each domain; domains without a valid reconstruction stay in the denominator with score 0.

`derived_agreement` is `agree_num / agree_den`: the fraction of the changed domains whose modal encoded derived residue is the most frequent one. It describes the row and does not enter the score.

The posterior threshold removes from the recorded distribution of a node every residue below it, except the most probable one. Removed mass is unrecorded, so a higher threshold can only raise `P` and lower the scores. The threshold is applied when the ASR posteriors are read; this report shows its value and does not vary it.

---

```{r load-data}
master_csv <- file.path(disambig_dir, "caas_convergence_master.csv")
if (!file.exists(master_csv)) {
  stop(sprintf("[asr_diagnostics] caas_convergence_master.csv not found in: %s", disambig_dir))
}

# Only the columns used here are read; the support tallies of the master are large.
domain_re <- "^domain_[0-9]+_(posterior|score|anc_aa|top_aa|bot_aa)$"
base_cols <- c("gene", "msa_pos", "caap_group", "side", "convergence_type", "n_hypotheses",
               "asr_path_score", "derived_agreement", "agreement_ambiguous")
hdr  <- names(readr::read_csv(master_csv, n_max = 0, show_col_types = FALSE))
need <- c("gene", "msa_pos", "caap_group", "side", "asr_path_score")
if (!all(need %in% hdr)) {
  stop(sprintf("[asr_diagnostics] the master lacks the columns: %s", paste(setdiff(need, hdr), collapse = ", ")))
}
keep <- hdr[hdr %in% base_cols | grepl(domain_re, hdr)]

master_df <- readr::read_csv(
  master_csv, col_select = all_of(keep), show_col_types = FALSE, guess_max = 10000,
  col_types = readr::cols(.default = readr::col_guess(), gene = "c", caap_group = "c", side = "c",
                          convergence_type = "c")
) %>%
  mutate(across(matches("_(anc|top|bot)_aa$"), as.character))
# Optional columns are present as NA when the master lacks them
for (nm in c("derived_agreement", "convergence_type", "agreement_ambiguous", "n_hypotheses")) {
  if (!nm %in% names(master_df)) master_df[[nm]] <- NA
}
master_df <- master_df %>%
  mutate(agreement_ambiguous = as.logical(as.character(agreement_ambiguous)),
         derived_agreement   = suppressWarnings(as.numeric(derived_agreement)))

# Rows without an asr_path_score are not scored and are left out
score_df <- master_df %>% filter(!is.na(asr_path_score))
if (nrow(score_df) == 0) stop("No ASR path scores found in the provided CSV.")

scheme_levels <- intersect(c("US", "GS1", "GS2", "GS3", "GS4"), unique(score_df$caap_group))
score_df <- score_df %>% mutate(caap_group = factor(caap_group, levels = scheme_levels))

n_positions <- score_df %>% distinct(gene, msa_pos) %>% nrow()
```

> **Posterior threshold** (`params.ct_disambig_posterior_threshold`): **`r if (is.na(posterior_threshold)) "not provided" else posterior_threshold`**
>
> **Disambiguation directory**: `r disambig_dir`
>
> **Rows in the master** (gene × position × scheme × side): `r format(nrow(master_df), big.mark = ",")`
>
> **Scored rows**: `r format(nrow(score_df), big.mark = ",")`, over `r format(n_positions, big.mark = ",")` positions in `r n_distinct(score_df$gene)` genes

```{r rows-table}
score_df %>%
  count(caap_group, side) %>%
  pivot_wider(names_from = side, values_from = n, values_fill = 0) %>%
  rename(Scheme = caap_group) %>%
  kable(format = "html", caption = "Scored rows per scheme and side") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

---

## 1. Distribution of the ASR path score

The schemes GS1 to GS4 co-encode residues of the same group, so two domains with different residues can agree under them; scores are therefore not expected to be equal across schemes, and each scheme is shown separately.

```{r plot-score-dist, fig.height=9, fig.cap="Distribution of asr_path_score per scheme and side."}
p_score <- ggplot(score_df, aes(x = asr_path_score)) +
  geom_histogram(binwidth = 0.02, fill = .tableau10[1], colour = "white", alpha = 0.85) +
  facet_grid(caap_group ~ side, scales = "free_y") +
  labs(
    title = "ASR path score",
    subtitle = sprintf("%s scored rows, %s positions",
                       formatC(nrow(score_df), big.mark = ","), formatC(n_positions, big.mark = ",")),
    x = "asr_path_score",
    y = "Rows"
  )

print(p_score)
save_plot(p_score, "asr_path_score_distribution.png", width = 9, height = 10)
```

---

## 2. Derived agreement

`derived_agreement` is a descriptor of the row (see Introduction). `agreement_ambiguous` marks rows where a changed domain had two derived residues tied across hypotheses and the smallest residue was taken by convention; for those rows, `derived_agreement` and `convergence_type` rest on that convention.

```{r plot-agreement, fig.height=9, fig.cap="Distribution of derived_agreement per scheme and side."}
agree_df <- score_df %>% filter(!is.na(derived_agreement))

p_agree <- ggplot(agree_df, aes(x = derived_agreement)) +
  geom_histogram(binwidth = 0.05, fill = .tableau10[3], colour = "white", alpha = 0.85) +
  facet_grid(caap_group ~ side, scales = "free_y") +
  labs(title = "Derived agreement", x = "derived_agreement", y = "Rows")

print(p_agree)
save_plot(p_agree, "derived_agreement_distribution.png", width = 9, height = 10)
```

```{r agreement-table}
agree_tab <- score_df %>%
  group_by(Scheme = caap_group) %>%
  summarise(
    Rows = n(),
    `Convergent (%)` = 100 * mean(convergence_type == "convergent", na.rm = TRUE),
    `Divergent (%)`  = 100 * mean(convergence_type == "divergent",  na.rm = TRUE),
    `Single (%)`     = 100 * mean(convergence_type == "single",     na.rm = TRUE),
    `Ambiguous agreement (%)` = 100 * mean(agreement_ambiguous, na.rm = TRUE),
    .groups = "drop"
  )

write_tsv_safe(agree_tab, "convergence_type_by_scheme.tsv")

agree_tab %>%
  kable(format = "html", digits = 2, caption = "Convergence type and ambiguous agreement per scheme (convergence_type_by_scheme.tsv)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

The two descriptors against each other: the score depends on how many domains changed and on the posterior at their LCAs, `derived_agreement` only on which residues the changed domains carry.

```{r plot-score-vs-agreement, fig.height=8, fig.cap="asr_path_score against derived_agreement; colour is the number of rows per bin (log scale)."}
p_sa <- ggplot(agree_df, aes(x = derived_agreement, y = asr_path_score)) +
  geom_bin2d(bins = 25) +
  scale_fill_viridis_c(trans = "log10", name = "Rows") +
  facet_wrap(~ side) +
  labs(title = "Score against derived agreement", x = "derived_agreement", y = "asr_path_score")

print(p_sa)
save_plot(p_sa, "asr_path_score_vs_derived_agreement.png", width = 10, height = 6)
```

---

## 3. Ancestral-state posterior of the domains

`domain_<k>_posterior` is the posterior of the ancestral residue of domain `k` at the MRCA of its pair, summed over the residues of the scheme of the row. A domain without a reconstruction has posterior 0 and cannot be changed. The score reads the ASR at a different place (the mass of the shared residue at the LCA of two domains), so the domain posterior is a descriptor of how well the ancestral residue that defines *changed* is supported, and not a sensitivity test of the score. The tables and plots below describe an association between the two; they do not show what the score would be under another threshold.

```{r domain-long}
# One domain at a time: only the changed-domain records are kept, plus per-scheme counters of all records.
score_df$row_id <- seq_len(nrow(score_df))
dom_ids <- sub("^domain_([0-9]+)_posterior$", "\\1",
               grep("^domain_[0-9]+_posterior$", names(score_df), value = TRUE))

col_or_na <- function(k, suffix) {
  nm <- sprintf("domain_%s_%s", k, suffix)
  if (nm %in% names(score_df)) score_df[[nm]] else rep(NA_character_, nrow(score_df))
}

per_domain <- lapply(dom_ids, function(k) {
  # The derived residue of the row's side marks the domain as changed on that side
  der <- if_else(score_df$side == "top", col_or_na(k, "top_aa"), col_or_na(k, "bot_aa"))
  d <- tibble(
    row_id = score_df$row_id, caap_group = score_df$caap_group, side = score_df$side,
    asr_path_score = score_df$asr_path_score, domain = k,
    posterior = suppressWarnings(as.numeric(col_or_na(k, "posterior"))),
    changed = !is.na(der) & der != ""
  )
  list(
    counts  = d %>% group_by(caap_group) %>%
      summarise(records = n(), not_recorded = sum(is.na(posterior)), no_rec = sum(!is.na(posterior) & posterior <= 0),
                changed = sum(changed), .groups = "drop"),
    changed = d %>% filter(changed, !is.na(posterior))
  )
})

domain_counts <- bind_rows(lapply(per_domain, `[[`, "counts")) %>%
  group_by(caap_group) %>% summarise(across(c(records, not_recorded, no_rec, changed), sum), .groups = "drop")

# The posterior columns can be present and empty; the sections below then say so and stop
has_posterior <- length(dom_ids) > 0 && sum(domain_counts$records - domain_counts$not_recorded) > 0
changed_df <- bind_rows(lapply(per_domain, `[[`, "changed"))
```

```{r domain-counts}
if (!has_posterior) {
  cat("The master table carries no domain posteriors (the domain_<k>_posterior columns are absent or empty), so this section is skipped.")
} else {
  domain_counts %>%
    transmute(Scheme = caap_group,
              `Domain records` = records,
              `Posterior not recorded (%)` = 100 * not_recorded / records,
              `Without reconstruction (%)` = 100 * no_rec / records,
              `Changed on the side (%)`    = 100 * changed / records) %>%
    kable(format = "html", digits = 2, caption = "Domain records per scheme") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
}
```

```{r plot-domain-posterior, fig.height=7, fig.cap="Ancestral posterior of the changed domains, per scheme."}
if (has_posterior && nrow(changed_df) > 0) {
  p_post <- ggplot(changed_df, aes(x = posterior)) +
    geom_histogram(binwidth = 0.02, fill = .tableau10[5], colour = "white", alpha = 0.85) +
    facet_wrap(~ caap_group, ncol = 2, scales = "free_y") +
    labs(
      title = "Ancestral posterior of the changed domains",
      subtitle = sprintf("%s changed domain records", formatC(nrow(changed_df), big.mark = ",")),
      x = "domain posterior", y = "Domain records"
    )
  print(p_post)
  save_plot(p_post, "domain_posterior_distribution.png", width = 10, height = 7)

  post_summary <- changed_df %>%
    group_by(caap_group) %>%
    summarise(
      n = n(),
      Mean = mean(posterior), Median = median(posterior),
      Q05 = quantile(posterior, 0.05), Q25 = quantile(posterior, 0.25),
      `Fraction < 0.5` = mean(posterior < 0.5),
      `Fraction < 0.8` = mean(posterior < 0.8),
      .groups = "drop"
    ) %>%
    rename(Scheme = caap_group)
  write_tsv_safe(post_summary, "domain_posterior_summary.tsv")

  post_summary %>%
    kable(format = "html", digits = 3, caption = "Posterior of the changed domains (domain_posterior_summary.tsv)") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
}
```

### Score against the weakest changed domain

For each row, the lowest posterior among its changed domains is the weakest ancestral call that the row rests on.

```{r plot-score-vs-minpost, fig.height=8, fig.cap="asr_path_score by the lowest posterior among the changed domains of the row."}
if (has_posterior && nrow(changed_df) > 0) {
  row_min <- changed_df %>%
    group_by(row_id, caap_group, side, asr_path_score) %>%
    summarise(min_posterior = min(posterior), .groups = "drop") %>%
    mutate(post_bin = cut(min_posterior, breaks = c(0, 0.5, 0.8, 0.95, 1.0000001),
                          labels = c("<0.5", "0.5-0.8", "0.8-0.95", ">=0.95"), right = FALSE,
                          include.lowest = TRUE))

  p_min <- ggplot(row_min, aes(x = post_bin, y = asr_path_score)) +
    geom_boxplot(fill = .tableau10[1], alpha = 0.6, outlier.size = 0.4) +
    facet_wrap(~ caap_group, ncol = 2) +
    labs(title = "Score by weakest changed-domain posterior",
         x = "Lowest posterior among the changed domains", y = "asr_path_score")
  print(p_min)
  save_plot(p_min, "asr_path_score_by_min_posterior.png", width = 10, height = 8)

  # Rank correlation, descriptive: rows share positions and genes, so no p-value is given
  min_tab <- row_min %>%
    group_by(Scheme = caap_group) %>%
    summarise(
      Rows = n(),
      `Spearman (score, min posterior)` = suppressWarnings(cor(asr_path_score, min_posterior, method = "spearman")),
      `Median score, min posterior < 0.8` = median(asr_path_score[min_posterior < 0.8]),
      `Median score, min posterior >= 0.8` = median(asr_path_score[min_posterior >= 0.8]),
      .groups = "drop"
    )
  write_tsv_safe(min_tab, "score_vs_min_posterior.tsv")

  min_tab %>%
    kable(format = "html", digits = 3, caption = "Score against the weakest changed-domain posterior (score_vs_min_posterior.tsv)") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
}
```

---

## 4. Summary statistics

```{r summary-table}
summary_table <- score_df %>%
  select(caap_group, side, asr_path_score, derived_agreement) %>%
  pivot_longer(c(asr_path_score, derived_agreement), names_to = "Variable", values_to = "value") %>%
  filter(!is.na(value)) %>%
  group_by(Variable, Scheme = caap_group, Side = side) %>%
  summarise(
    n = n(), Mean = mean(value), Median = median(value),
    SD = sd(value), Min = min(value), Max = max(value),
    .groups = "drop"
  )

write_tsv_safe(summary_table, "asr_path_summary_statistics.tsv")

summary_table %>%
  kable(format = "html", digits = 4, caption = "Summary statistics per scheme and side (asr_path_summary_statistics.tsv)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE)
```

---

## Output Files Summary

```{r output_summary, echo=FALSE}
tsv_files  <- list.files(tsv_dir,   pattern = "\\.tsv$", full.names = FALSE)
plot_files <- list.files(plots_dir, pattern = "\\.png$", full.names = FALSE)

tibble(
  Type = c(rep("TSV", length(tsv_files)), rep("Plot", length(plot_files))),
  File = c(tsv_files, plot_files)
) %>%
  kable(format = "html", caption = "Files written under asr_diagnostics/") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — ASR Diagnostics Module*
*Date of generation: `r Sys.Date()`*
