---
title: "ASR Path Score Characterization"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
params:
  disambig_dir:        NULL   # path to ct_disambiguation/ output directory
  posterior_threshold: NULL   # params.ct_disambig_posterior_threshold (shown, not applied)
  output_dir:          "."    # working directory (TSVs → ./tsv/, plots → ./plots/)
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

<!-- 9.ASR_diagnostics.Rmd — Distribution of asr_path_score, derived_agreement of the scored CAAS rows. -->
<!-- PhyloPhere | subworkflows/ASR_DIAGNOSTICS/local/ -->
<!--
  Called by: ASR_DIAGNOSTICS_REPORT (asr_diagnostics.nf), through rmarkdown::render().
  Params:    disambig_dir (ct_disambiguation/ with caas_convergence_master.csv),
             posterior_threshold (shown in the report, not applied), output_dir.
  A row of the master is one (gene, msa_pos, caap_group, side). asr_path_score is the score of the row on the
  Voronoi domains (fop_pool.py); derived_agreement is a descriptor and does not enter the score.
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 11,
  fig.height = 6,
  dpi        = 180,
  dev        = "png"
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(ggplot2)
  library(scales)
  library(knitr)
  library(kableExtra)
})

theme_set(
  theme_bw(base_size = 12) +
  theme(
    axis.title.y  = element_text(margin = ggplot2::margin(l = 20, r = 20)),
    axis.title.x  = element_text(margin = ggplot2::margin(t = 12)),
    axis.text.x   = element_text(margin = ggplot2::margin(t = 5)),
    plot.title    = element_text(face = "bold", hjust = 0, margin = ggplot2::margin(b = 6)),
    plot.subtitle = element_text(colour = "grey40", hjust = 0, margin = ggplot2::margin(b = 10)),
    strip.background = element_rect(fill = "grey95", colour = "grey80"),
    strip.text    = element_text(face = "bold")
  )
)

# ── Parameters ────────────────────────────────────────────────────────────────
disambig_dir        <- params$disambig_dir
posterior_threshold <- if (is.null(params$posterior_threshold)) NA_real_ else as.numeric(params$posterior_threshold)
output_dir          <- params$output_dir

stopifnot(!is.null(disambig_dir))

# ── Output directories ────────────────────────────────────────────────────────
# Tables go to tsv/ and figures to plots/, both published by ASR_DIAGNOSTICS_REPORT.
tsv_dir   <- file.path(output_dir, "tsv")
plots_dir <- file.path(output_dir, "plots")
dir.create(tsv_dir,   recursive = TRUE, showWarnings = FALSE)
dir.create(plots_dir, recursive = TRUE, showWarnings = FALSE)

write_tsv_safe <- function(df, fname) {
  readr::write_tsv(df, file.path(tsv_dir, fname))
  invisible(file.path(tsv_dir, fname))
}

save_plot <- function(p, fname, width = 11, height = 6) {
  ggsave(file.path(plots_dir, fname), plot = p, width = width, height = height, dpi = 180)
  invisible(file.path(plots_dir, fname))
}

# Categorical palette (Tableau 10)
.tableau10 <- c(
  "#1F77B4", "#AEC7E8", "#FF7F0E", "#FFBB78",
  "#2CA02C", "#98DF8A", "#D62728", "#FF9896",
  "#9467BD", "#C5B0D5"
)
```

## Introduction

`asr_path_score` is the CAAS convergence score of a position on the K fixed Voronoi domains of the contrast design. It is computed for each scheme (US, GS1 to GS4) and each phenotype side, so a position contributes up to ten rows to the master table.

For one hypothesis and one side, a domain is *changed* when its observed foreground residue, encoded in the scheme, differs from the modal ancestral residue at the MRCA of its pair. A changed domain scores the noisy-OR, over the changed domains that carry the same encoded residue, of `1 - P`, where `P` is the posterior mass of that residue at the LCA of the two domains. `P` is the recorded mass of the residue and, when the residue is not in the recorded distribution of the node, the whole unrecorded mass (worst case). A changed domain without a same-residue partner scores 0. The domain scores are averaged over the hypotheses and the score of the row is their mean weighted by the PSS of each domain; domains without a valid reconstruction stay in the denominator with score 0.

`derived_agreement` is `agree_num / agree_den`: the fraction of the changed domains whose modal encoded derived residue is the most frequent one. It describes the row and does not enter the score.

The posterior threshold removes from the recorded distribution of a node every residue below it, except the most probable one. Removed mass is unrecorded, so a higher threshold can only raise `P` and lower the scores. The threshold is applied when the ASR posteriors are read; this report shows its value and does not vary it.

---

```{r load-data}
master_csv <- file.path(disambig_dir, "caas_convergence_master.csv")
if (!file.exists(master_csv)) {
  stop(sprintf("[asr_diagnostics] caas_convergence_master.csv not found in: %s", disambig_dir))
}

# Only the columns used here are read; the support tallies of the master are large.
base_cols <- c("gene", "msa_pos", "caap_group", "side", "convergence_type", "n_hypotheses",
               "asr_path_score", "derived_agreement", "agreement_ambiguous")
hdr  <- names(readr::read_csv(master_csv, n_max = 0, show_col_types = FALSE))
need <- c("gene", "msa_pos", "caap_group", "side", "asr_path_score")
if (!all(need %in% hdr)) {
  stop(sprintf("[asr_diagnostics] the master lacks the columns: %s", paste(setdiff(need, hdr), collapse = ", ")))
}
keep <- hdr[hdr %in% base_cols]

master_df <- readr::read_csv(
  master_csv, col_select = all_of(keep), show_col_types = FALSE, guess_max = 10000,
  col_types = readr::cols(.default = readr::col_guess(), gene = "c", caap_group = "c", side = "c",
                          convergence_type = "c")
)
# Optional columns are present as NA when the master lacks them
for (nm in c("derived_agreement", "convergence_type", "agreement_ambiguous", "n_hypotheses")) {
  if (!nm %in% names(master_df)) master_df[[nm]] <- NA
}
master_df <- master_df %>%
  mutate(agreement_ambiguous = as.logical(as.character(agreement_ambiguous)),
         derived_agreement   = suppressWarnings(as.numeric(derived_agreement)))

# Rows without an asr_path_score are not scored and are left out
score_df <- master_df %>% filter(!is.na(asr_path_score))
if (nrow(score_df) == 0) stop("No ASR path scores found in the provided CSV.")

scheme_levels <- intersect(c("US", "GS1", "GS2", "GS3", "GS4"), unique(score_df$caap_group))
score_df <- score_df %>% mutate(caap_group = factor(caap_group, levels = scheme_levels))

n_positions <- score_df %>% distinct(gene, msa_pos) %>% nrow()
```

> **Posterior threshold** (`params.ct_disambig_posterior_threshold`): **`r if (is.na(posterior_threshold)) "not provided" else posterior_threshold`**
>
> **Disambiguation directory**: `r disambig_dir`
>
> **Rows in the master** (gene × position × scheme × side): `r format(nrow(master_df), big.mark = ",")`
>
> **Scored rows**: `r format(nrow(score_df), big.mark = ",")`, over `r format(n_positions, big.mark = ",")` positions in `r n_distinct(score_df$gene)` genes

```{r rows-table}
score_df %>%
  count(caap_group, side) %>%
  pivot_wider(names_from = side, values_from = n, values_fill = 0) %>%
  rename(Scheme = caap_group) %>%
  kable(format = "html", caption = "Scored rows per scheme and side") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

---

## 1. Distribution of the ASR path score

The schemes GS1 to GS4 co-encode residues of the same group, so two domains with different residues can agree under them; scores are therefore not expected to be equal across schemes, and each scheme is shown separately.

```{r plot-score-dist, fig.height=9, fig.cap="Distribution of asr_path_score per scheme and side."}
p_score <- ggplot(score_df, aes(x = asr_path_score)) +
  geom_histogram(binwidth = 0.02, fill = .tableau10[1], colour = "white", alpha = 0.85) +
  facet_grid(caap_group ~ side, scales = "free_y") +
  labs(
    title = "ASR path score",
    subtitle = sprintf("%s scored rows, %s positions",
                       formatC(nrow(score_df), big.mark = ","), formatC(n_positions, big.mark = ",")),
    x = "asr_path_score",
    y = "Rows"
  )

print(p_score)
save_plot(p_score, "asr_path_score_distribution.png", width = 9, height = 10)
```

---

## 2. Derived agreement

`derived_agreement` is a descriptor of the row (see Introduction). `agreement_ambiguous` marks rows where a changed domain had two derived residues tied across hypotheses and the smallest residue was taken by convention; for those rows, `derived_agreement` and `convergence_type` rest on that convention.

```{r plot-agreement, fig.height=9, fig.cap="Distribution of derived_agreement per scheme and side."}
agree_df <- score_df %>% filter(!is.na(derived_agreement))

p_agree <- ggplot(agree_df, aes(x = derived_agreement)) +
  geom_histogram(binwidth = 0.05, fill = .tableau10[3], colour = "white", alpha = 0.85) +
  facet_grid(caap_group ~ side, scales = "free_y") +
  labs(title = "Derived agreement", x = "derived_agreement", y = "Rows")

print(p_agree)
save_plot(p_agree, "derived_agreement_distribution.png", width = 9, height = 10)
```

```{r agreement-table}
agree_tab <- score_df %>%
  group_by(Scheme = caap_group) %>%
  summarise(
    Rows = n(),
    `Convergent (%)` = 100 * mean(convergence_type == "convergent", na.rm = TRUE),
    `Divergent (%)`  = 100 * mean(convergence_type == "divergent",  na.rm = TRUE),
    `Single (%)`     = 100 * mean(convergence_type == "single",     na.rm = TRUE),
    `Ambiguous agreement (%)` = 100 * mean(agreement_ambiguous, na.rm = TRUE),
    .groups = "drop"
  )

write_tsv_safe(agree_tab, "convergence_type_by_scheme.tsv")

agree_tab %>%
  kable(format = "html", digits = 2, caption = "Convergence type and ambiguous agreement per scheme (convergence_type_by_scheme.tsv)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

The two descriptors against each other: the score depends on how many domains changed and on the posterior at their LCAs, `derived_agreement` only on which residues the changed domains carry.

```{r plot-score-vs-agreement, fig.height=8, fig.cap="asr_path_score against derived_agreement; colour is the number of rows per bin (log scale)."}
p_sa <- ggplot(agree_df, aes(x = derived_agreement, y = asr_path_score)) +
  geom_bin2d(bins = 25) +
  scale_fill_viridis_c(trans = "log10", name = "Rows") +
  facet_wrap(~ side) +
  labs(title = "Score against derived agreement", x = "derived_agreement", y = "asr_path_score")

print(p_sa)
save_plot(p_sa, "asr_path_score_vs_derived_agreement.png", width = 10, height = 6)
```

---

## 3. Summary statistics

```{r summary-table}
summary_table <- score_df %>%
  select(caap_group, side, asr_path_score, derived_agreement) %>%
  pivot_longer(c(asr_path_score, derived_agreement), names_to = "Variable", values_to = "value") %>%
  filter(!is.na(value)) %>%
  group_by(Variable, Scheme = caap_group, Side = side) %>%
  summarise(
    n = n(), Mean = mean(value), Median = median(value),
    SD = sd(value), Min = min(value), Max = max(value),
    .groups = "drop"
  )

write_tsv_safe(summary_table, "asr_path_summary_statistics.tsv")

summary_table %>%
  kable(format = "html", digits = 4, caption = "Summary statistics per scheme and side (asr_path_summary_statistics.tsv)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE)
```

---

## Output Files Summary

```{r output_summary, echo=FALSE}
tsv_files  <- list.files(tsv_dir,   pattern = "\\.tsv$", full.names = FALSE)
plot_files <- list.files(plots_dir, pattern = "\\.png$", full.names = FALSE)

tibble(
  Type = c(rep("TSV", length(tsv_files)), rep("Plot", length(plot_files))),
  File = c(tsv_files, plot_files)
) %>%
  kable(format = "html", caption = "Files written under asr_diagnostics/") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — ASR Diagnostics Module*
*Date of generation: `r Sys.Date()`*
