---
title: "MoleRate — Relative Evolutionary Rate Analysis"
subtitle: "`r paste0(params$traitname, ' | ', toupper(params$direction), ' direction')`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 11
    fig_height: 6
params:
  json_dir:         "."
  direction:        "top"
  traitname:        "unknown_trait"
  pval_threshold:   0.05
  log2fc_threshold: 0.5
  top_n_labels:     15
  output_dir:       "."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  out.width = "100%"
)

suppressPackageStartupMessages({
  library(jsonlite)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggrepel)
  library(paletteer)
  library(DT)
})

# Tableau_10 colour swatch (10 named colours for pipeline-wide consistency)
.tableau10 <- as.character(paletteer_d("ggthemes::Tableau_10"))
```

---

## Overview

This report summarises the results of **MoleRate** (relative evolutionary
molecular rate; Kosakovsky Pond _et al._, 2021–2025, hyphy-analyses) applied
to **`r params$direction`** foreground branches associated with the phenotype
**`r params$traitname`**.

MoleRate tests whether the **foreground (test) branches** evolve at a different
rate than the **background branches** relative to a shared reference tree with
given branch lengths.

Four nested models are fitted per gene:

| Model                | Test branches     | Background branches |
|----------------------|:-----------------:|:-------------------:|
| Proportional         | R_shared × ref    | R_shared × ref      |
| Proportional Part.   | R_test × ref      | R_bg × ref          |
| Unconstrained Test   | Independent       | R_bg × ref          |
| Unconstrained        | Independent       | Independent         |

The **key test** (most analogous to RERconverge) is  
`Proportional Partitioned` vs `Proportional` — it asks whether the average
relative rate differs between foreground and background branches.

P-values are Holm–Bonferroni corrected per gene (within‐gene, for branch-level
tests) and are additionally FDR-corrected across genes (between-gene, for the
gene-level tests) in this report.

---

```{r load-jsons}
json_files <- list.files(
  path       = params$json_dir,
  pattern    = "\\.molerate\\.json$",
  full.names = TRUE
)

if (length(json_files) == 0) {
  message("No *.molerate.json files found in '", params$json_dir,
          "'. No MoleRate results for the ", params$direction, " direction ",
          "(genes may all have failed or no genes were significant).")
  knitr::knit_exit()
}
cat(sprintf("Found %d MoleRate result files.\n", length(json_files)))
```

```{r parse-molerate, results='hide'}
# ── Helpers to safely extract nested elements ────────────────────────────────
safe_get <- function(x, ...) {
  keys <- list(...)
  tryCatch({
    for (k in keys) x <- x[[k]]
    x
  }, error = function(e) NULL)
}

as_num <- function(x, default = NA_real_) {
  v <- suppressWarnings(as.numeric(x))
  if (is.null(v) || length(v) == 0 || is.na(v)) default else v
}

# ── JSON parser ──────────────────────────────────────────────────────────────
parse_molerate_json <- function(path) {
  gene_id <- sub("\\.(top|bottom)\\.molerate\\.json$", "", basename(path))

  tryCatch({
    js <- fromJSON(path, simplifyVector = FALSE)

    # ── Proportional Partitioned fit → R_test, R_bg ──────────────────────────
    pp_fit  <- safe_get(js, "fits", "Proportional Partitioned")

    r_test <- as_num(
      safe_get(pp_fit, "Rate Distributions", "Test",       0, "omega") %||%
      safe_get(pp_fit, "Rate Distributions", "test",       0, "omega") %||%
      safe_get(pp_fit, "Rate Distributions", "Foreground", 0, "omega")
    )
    r_bg <- as_num(
      safe_get(pp_fit, "Rate Distributions", "Background", 0, "omega") %||%
      safe_get(pp_fit, "Rate Distributions", "background", 0, "omega")
    )
    logL_pp <- as_num(safe_get(pp_fit, "Log Likelihood") %||%
                      safe_get(pp_fit, "log likelihood"))

    prop_fit   <- safe_get(js, "fits", "Proportional")
    logL_prop  <- as_num(safe_get(prop_fit, "Log Likelihood") %||%
                         safe_get(prop_fit, "log likelihood"))

    # ── LRT p-value (Proportional Partitioned vs Proportional) ───────────────
    # MoleRate stores corrected p-values in "test results" as a nested structure
    # Try multiple possible key names used across HyPhy versions.
    pval_pp_vs_prop <- as_num(
      safe_get(js, "test results", "Proportional Partitioned", "p-value") %||%
      safe_get(js, "test results", "Proportional:Proportional Partitioned", "p-value") %||%
      safe_get(js, "LRT", "Proportional Partitioned", "p-value")
    )

    # Unconstrained Test vs Proportional Partitioned p-value
    pval_uc_vs_pp <- as_num(
      safe_get(js, "test results", "Unconstrained Test",  "p-value") %||%
      safe_get(js, "test results", "Proportional Partitioned:Unconstrained Test", "p-value")
    )

    # ── AIC-c values ─────────────────────────────────────────────────────────
    aicc_pp   <- as_num(safe_get(pp_fit, "AIC-c"))
    aicc_prop <- as_num(safe_get(prop_fit, "AIC-c"))

    # ── Branch-level results (optional) ──────────────────────────────────────
    bl_results <- safe_get(js, "branch level results")
    n_sig_branches <- if (!is.null(bl_results)) {
      sigs <- sapply(bl_results, function(br) {
        pv <- as_num(safe_get(br, "single branch", "p-value") %||%
                     safe_get(br, "p-value"))
        !is.na(pv) && pv < 0.05
      })
      sum(sigs)
    } else NA_integer_

    data.frame(
      gene               = gene_id,
      r_test             = r_test,
      r_bg               = r_bg,
      log2_ratio         = log2(r_test / r_bg),
      logL_pp            = logL_pp,
      logL_prop          = logL_prop,
      aicc_pp            = aicc_pp,
      aicc_prop          = aicc_prop,
      pval_pp_vs_prop    = pval_pp_vs_prop,
      pval_uc_vs_pp      = pval_uc_vs_pp,
      n_sig_branches     = n_sig_branches,
      stringsAsFactors   = FALSE
    )

  }, error = function(e) {
    warning("Failed to parse ", gene_id, ": ", conditionMessage(e))
    NULL
  })
}

# Null-coalescing operator (base R doesn't have one natively)
`%||%` <- function(a, b) if (!is.null(a) && length(a) > 0) a else b

all_results <- lapply(json_files, parse_molerate_json)
all_results <- Filter(Negate(is.null), all_results)

if (length(all_results) == 0) {
  message("All MoleRate JSON files failed to parse (empty or corrupt). ",
          "No valid results for the ", params$direction, " direction.")
  knitr::knit_exit()
}

mr_df <- do.call(rbind, all_results) %>%
  mutate(
    # FDR correction across genes on the main test (PP vs Prop)
    qval_pp_vs_prop = p.adjust(pval_pp_vs_prop, method = "BH"),
    signif          = !is.na(qval_pp_vs_prop) & qval_pp_vs_prop < params$pval_threshold,
    accelerated     = signif & log2_ratio >  params$log2fc_threshold,
    decelerated     = signif & log2_ratio < -params$log2fc_threshold
  ) %>%
  arrange(qval_pp_vs_prop)

cat(sprintf(
  "Parsed %d genes. Significant after FDR: %d accelerated, %d decelerated.\n",
  nrow(mr_df),
  sum(mr_df$accelerated, na.rm = TRUE),
  sum(mr_df$decelerated, na.rm = TRUE)
))
```

---

## Gene-level Summary

```{r gene-table}
display_df <- mr_df %>%
  mutate(
    r_test          = round(r_test,          4),
    r_bg            = round(r_bg,            4),
    log2_ratio      = round(log2_ratio,       3),
    pval_pp_vs_prop = signif(pval_pp_vs_prop, 3),
    qval_pp_vs_prop = signif(qval_pp_vs_prop, 3),
    call            = case_when(
      accelerated  ~ "Accelerated",
      decelerated  ~ "Decelerated",
      signif       ~ "Significant",
      TRUE         ~ ""
    )
  ) %>%
  rename(
    Gene              = gene,
    `R_test`          = r_test,
    `R_background`    = r_bg,
    `log2(R_t/R_bg)`  = log2_ratio,
    `LRT p-value`     = pval_pp_vs_prop,
    `FDR q-value`     = qval_pp_vs_prop,
    `Sig. branches`   = n_sig_branches,
    Call              = call
  ) %>%
  select(Gene, R_test, R_background, `log2(R_t/R_bg)`,
         `LRT p-value`, `FDR q-value`, `Sig. branches`, Call)

DT::datatable(
  display_df,
  filter   = "top",
  rownames = FALSE,
  options  = list(pageLength = 30, scrollX = TRUE),
  caption  = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf(
      "MoleRate gene-level results | BH-FDR < %g | |log2 ratio| ≥ %g | Direction: %s",
      params$pval_threshold, params$log2fc_threshold, toupper(params$direction)
    )
  )
) %>%
  DT::formatStyle(
    "Call",
    backgroundColor = DT::styleEqual(
      c("Accelerated", "Decelerated", "Significant", ""),
      c("#ffe0e0",     "#dff0ff",     "#fff0d0",      "white")
    )
  )
```

```{r write-tsv}
out_tsv <- file.path(params$output_dir,
                     paste0("molerate_summary_", params$direction, ".tsv"))
write.table(
  mr_df %>% select(-signif, -accelerated, -decelerated),
  out_tsv, sep = "\t", row.names = FALSE, quote = FALSE
)
invisible(NULL)
```

---

## Visualisations

### Volcano plot

```{r volcano, fig.height=7, fig.width=9}
# Prepare labels for the top significant genes
label_df <- mr_df %>%
  filter(signif) %>%
  arrange(qval_pp_vs_prop) %>%
  slice_head(n = params$top_n_labels)

ggplot(mr_df, aes(x = log2_ratio, y = -log10(qval_pp_vs_prop + 1e-300))) +
  geom_point(
    aes(colour = case_when(
      accelerated ~ "Accelerated",
      decelerated ~ "Decelerated",
      signif      ~ "Significant",
      TRUE        ~ "Non-significant"
    )),
    alpha = 0.7,
    size  = 2
  ) +
  geom_vline(xintercept = c(-params$log2fc_threshold, params$log2fc_threshold),
             linetype = "dashed", colour = "grey40", linewidth = 0.5) +
  geom_hline(yintercept = -log10(params$pval_threshold),
             linetype = "dashed", colour = "grey40", linewidth = 0.5) +
  geom_text_repel(
    data    = label_df,
    aes(label = gene),
    size    = 3,
    max.overlaps = 30
  ) +
  scale_colour_manual(
    values = c(
      "Accelerated"     = .tableau10[3],
      "Decelerated"     = .tableau10[1],
      "Significant"     = .tableau10[2],
      "Non-significant" = "grey60"
    ),
    name = NULL
  ) +
  labs(
    title    = paste("MoleRate volcano plot —", toupper(params$direction), "direction"),
    subtitle = paste0("Trait: ", params$traitname, " | n = ", nrow(mr_df), " genes"),
    x        = "log2(R_test / R_background)",
    y        = expression(-log[10](q-value))
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "top")
```

### Rate ratio distribution

```{r rate-dist, fig.height=5}
ggplot(mr_df %>% filter(!is.na(log2_ratio) & is.finite(log2_ratio)),
       aes(x = log2_ratio)) +
  geom_histogram(aes(fill = after_stat(x) > 0), bins = 60,
                 colour = "white", alpha = 0.85) +
  scale_fill_manual(
    values = c("TRUE" = .tableau10[3], "FALSE" = .tableau10[1]),
    labels = c("TRUE" = "Faster (FG)", "FALSE" = "Slower (FG)"),
    name   = NULL
  ) +
  geom_vline(xintercept = 0, linewidth = 0.7, colour = "black") +
  labs(
    title    = "Distribution of log2(R_test / R_background) across genes",
    subtitle = paste0("Direction: ", toupper(params$direction),
                     " | Trait: ", params$traitname),
    x = "log2(R_test / R_background)",
    y = "Number of genes"
  ) +
  theme_bw(base_size = 13)
```

### Scatter: R_test vs R_background

```{r scatter, fig.height=6, fig.width=7}
ggplot(mr_df, aes(x = r_bg, y = r_test)) +
  geom_point(
    aes(colour = case_when(
      accelerated ~ "Accelerated",
      decelerated ~ "Decelerated",
      TRUE        ~ "Other"
    )),
    alpha = 0.65, size = 2
  ) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey40", linewidth = 0.6) +
  geom_text_repel(
    data    = label_df,
    aes(label = gene),
    size    = 2.6,
    max.overlaps = 20
  ) +
  scale_colour_manual(
    values = c("Accelerated" = .tableau10[3], "Decelerated" = .tableau10[1], "Other" = "grey60"),
    name   = NULL
  ) +
  scale_x_log10() +
  scale_y_log10() +
  annotation_logticks(sides = "bl") +
  labs(
    title    = "R_test versus R_background per gene",
    subtitle = paste0("Direction: ", toupper(params$direction),
                     " | Dashed diagonal = equal rates"),
    x = "R_background (log10)",
    y = "R_test (log10)"
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "top")
```

---

## Accelerated Genes (top `r params$top_n_labels`)

```{r top-accelerated}
top_acc <- mr_df %>%
  filter(accelerated) %>%
  arrange(qval_pp_vs_prop, desc(log2_ratio)) %>%
  slice_head(n = params$top_n_labels) %>%
  mutate(
    r_test       = round(r_test,       4),
    r_bg         = round(r_bg,         4),
    log2_ratio   = round(log2_ratio,   3),
    qval         = signif(qval_pp_vs_prop, 3)
  ) %>%
  select(gene, r_test, r_bg, log2_ratio, qval, n_sig_branches)

if (nrow(top_acc) == 0) {
  cat("No significantly accelerated genes at the given thresholds.\n")
} else {
  knitr::kable(
    top_acc,
    col.names = c("Gene", "R_test", "R_bg", "log2 ratio", "BH q-value", "Sig. branches"),
    caption   = "Top accelerated genes (FG branches evolve faster than background)"
  )
}
```

## Decelerated Genes (top `r params$top_n_labels`)

```{r top-decelerated}
top_dec <- mr_df %>%
  filter(decelerated) %>%
  arrange(qval_pp_vs_prop, log2_ratio) %>%
  slice_head(n = params$top_n_labels) %>%
  mutate(
    r_test     = round(r_test,     4),
    r_bg       = round(r_bg,       4),
    log2_ratio = round(log2_ratio, 3),
    qval       = signif(qval_pp_vs_prop, 3)
  ) %>%
  select(gene, r_test, r_bg, log2_ratio, qval, n_sig_branches)

if (nrow(top_dec) == 0) {
  cat("No significantly decelerated genes at the given thresholds.\n")
} else {
  knitr::kable(
    top_dec,
    col.names = c("Gene", "R_test", "R_bg", "log2 ratio", "BH q-value", "Sig. branches"),
    caption   = "Top decelerated genes (FG branches evolve slower than background)"
  )
}
```

---

## Methods

**Tool:** MoleRate (relative evolutionary molecular rate) via HyPhy hyphy-analyses
(https://github.com/veg/hyphy-analyses/tree/master/molerate).

**Branch labelling:** Foreground (test) branches correspond to species in the
**`r params$direction`** contrast group of the caastools traitfile for trait
`r params$traitname`. For a `r params$direction` analysis, contrast_group
`r ifelse(params$direction == "top", "= 1 (high-trait extreme)", "= 0 (low-trait extreme after FG/BG flip)")` species were designated as test branches.

**Reference tree:** The shared primate phylogeny with given branch lengths was
used as the reference for computing relative rates (all four models scale
branches relative to this reference).

**Internal labelling strategy:** `r params$.GlobalEnv$mr_labeling %||% "all-descendants"` —
an internal branch is labelled as test if all (or any, depending on strategy)
of its descendants are test branches.

**Statistical inference:** The key hypothesis test is
_Proportional Partitioned_ vs _Proportional_ (LRT, 1 d.f.), asking whether
the average rate of test branches differs from background. P-values are
Benjamini–Hochberg FDR-corrected across all genes (q-value). A gene is called
**accelerated** if q < `r params$pval_threshold` and
log2(R_test/R_bg) > `r params$log2fc_threshold`, and **decelerated** if
log2(R_test/R_bg) < `r -params$log2fc_threshold`.

---

## Session Information

```{r session-info}
sessionInfo()
```
