---
title: "CT Post-Processing and Disambiguation Characterization"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
params:
  discovery_file: NULL
  filter_summary_file: NULL
  filter_dir: NULL
  gene_ensembl_file: NULL
  gene_stats_file: NULL
  processing_mode: "exploratory"
  output_dir: "ct_postproc"
  extreme_threshold: 0.99
  iqr_multiplier: 3.0
  alpha_threshold: 0.1
  gene_filter_mode: "both"
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7,
  dev = 'png',
  dpi = 320
)

suppressPackageStartupMessages({
  library(stringr)
  library(purrr)
  library(tibble)
  library(patchwork)
  library(scales)
  library(ggplot2)
  library(forcats)
  library(knitr)
  library(kableExtra)
  library(paletteer)
  library(tidyverse)
})

createDir <- function(directory) {
  if (!dir.exists(directory)) {
    dir.create(directory, recursive = TRUE, showWarnings = FALSE)
  }
}

find_first_existing <- function(candidates) {
  hits <- candidates[file.exists(candidates)]
  if (length(hits) == 0) return(NA_character_)
  hits[[1]]
}

find_recursive_file <- function(root_dir, pattern) {
  if (is.null(root_dir) || is.na(root_dir) || !dir.exists(root_dir)) return(NA_character_)
  files <- list.files(root_dir, pattern = pattern, full.names = TRUE, recursive = TRUE)
  if (length(files) == 0) return(NA_character_)
  files[[1]]
}

as_bool <- function(x) {
  tolower(as.character(x)) %in% c("true", "t", "1", "yes", "y")
}

format_threshold <- function(x) {
  formatC(as.numeric(x), format = "f", digits = 2)
}

make_significance_class <- function(sig_hyp, sig_perm) {
  case_when(
    sig_hyp & sig_perm ~ "both",
    sig_hyp ~ "hyper_only",
    sig_perm ~ "perm_only",
    TRUE ~ "none"
  )
}

collapse_change_side <- function(x) {
  x <- tolower(as.character(x))
  x <- x[x %in% c("top", "bottom", "both")]
  if (length(x) == 0) return("none")
  if ("both" %in% x) return("both")
  ux <- unique(x)
  if (all(c("top", "bottom") %in% ux)) return("both")
  ux[[1]]
}

combo_from_flags <- function(df, cols) {
  apply(df[, cols, drop = FALSE], 1, function(v) {
    labs <- cols[as.logical(v)]
    if (length(labs) == 0) "none" else paste(labs, collapse = "+")
  })
}

# The parallel_* axis was retired upstream (superseded by the continuous
# mrca_diversity axis of the ASR path score), so the convergence label is now the
# convergence_type alone.
make_pattern_parallel_label <- function(convergence_type) {
  pattern_chr <- as.character(convergence_type)
  pattern_chr[is.na(pattern_chr) | pattern_chr == ""] <- "unknown"
  pattern_chr
}

format_pattern_parallel_axis <- function(x) {
  stringr::str_replace_all(as.character(x), "\\s*/\\s*", "\n/ ")
}

derive_plot_zoom_upper <- function(plot_obj, zoom_fraction = 0.10, min_upper = 3) {
  built <- ggplot_build(plot_obj)
  layer_max <- vapply(built$data, function(layer_df) {
    metric_cols <- intersect(c("count", "y", "density"), names(layer_df))
    if (length(metric_cols) == 0) return(0)
    max(suppressWarnings(as.numeric(layer_df[[metric_cols[[1]]]])), na.rm = TRUE)
  }, numeric(1))

  max_val <- suppressWarnings(max(layer_max, na.rm = TRUE))
  if (!is.finite(max_val) || max_val <= 0) return(min_upper)
  max(min_upper, ceiling(max_val * zoom_fraction))
}

compute_barplot_dims <- function(n_labels, n_facets = 1, facet_cols = 1,
                                 base_width = 8, base_height = 5.5,
                                 per_label = 0.55, facet_col_bonus = 1.4,
                                 facet_row_height = 2.2,
                                 min_width = 8, max_width = 24,
                                 min_height = 5, max_height = 18) {
  n_labels <- max(1, suppressWarnings(as.integer(n_labels)))
  n_facets <- max(1, suppressWarnings(as.integer(n_facets)))
  facet_cols <- max(1, suppressWarnings(as.integer(facet_cols)))
  facet_rows <- ceiling(n_facets / facet_cols)

  width <- base_width + max(0, n_labels - 3) * per_label + max(0, min(n_facets, facet_cols) - 1) * facet_col_bonus
  height <- base_height + max(0, facet_rows - 1) * facet_row_height

  list(
    width = min(max_width, max(min_width, width)),
    height = min(max_height, max(min_height, height))
  )
}

save_barplot <- function(filename, plot, n_labels, n_facets = 1, facet_cols = 1,
                         dpi = 320, base_width = 8, base_height = 5.5,
                         per_label = 0.55, facet_row_height = 2.2) {
  dims <- compute_barplot_dims(
    n_labels = n_labels,
    n_facets = n_facets,
    facet_cols = facet_cols,
    base_width = base_width,
    base_height = base_height,
    per_label = per_label,
    facet_row_height = facet_row_height
  )
  ggsave(filename, plot, width = dims$width, height = dims$height, dpi = dpi)
}

build_orientation_rows <- function(df) {
  bind_rows(
    df %>% mutate(orientation = "global"),
    df %>% filter(change_side %in% c("top", "both")) %>% mutate(orientation = "top"),
    df %>% filter(change_side %in% c("bottom", "both")) %>% mutate(orientation = "bottom"),
    df %>% filter(change_side == "both") %>% mutate(orientation = "both")
  ) %>%
    mutate(
      orientation = factor(orientation, levels = c("global", "top", "bottom", "both"))
    )
}

build_group_flag_table <- function(df) {
  df %>%
    mutate(caap_group = toupper(caap_group)) %>%
    group_by(orientation, site_id, gene, position) %>%
    summarise(
      us = any(caap_group == "US"),
      gs1 = any(caap_group == "GS1"),
      gs2 = any(caap_group == "GS2"),
      gs3 = any(caap_group == "GS3"),
      gs4 = any(caap_group == "GS4"),
      .groups = "drop"
    )
}

build_threshold_line_df <- function(alpha_threshold) {
  tibble(
    line_key = c("configured", "comparison_010", "comparison_005"),
    threshold = c(alpha_threshold, 0.10, 0.05),
    line_label = c(
      paste0("Configured alpha = ", format_threshold(alpha_threshold)),
      "Comparison threshold = 0.10",
      "Comparison threshold = 0.05"
    )
  ) %>%
    group_by(threshold) %>%
    summarise(
      line_key = if ("configured" %in% line_key) "configured" else first(line_key),
      line_label = if ("configured" %in% line_key && n() > 1) {
        paste0("Configured alpha = Comparison threshold = ", format_threshold(first(threshold)))
      } else {
        paste(unique(line_label), collapse = " = ")
      },
      .groups = "drop"
    ) %>%
    arrange(desc(line_key == "configured"), desc(threshold))
}

resolve_filtered_cluster_file <- function(minlen, maxcaas_int, filter_root = NULL, postproc_root = NULL) {
  search_roots <- unique(c(filter_root, postproc_root, getwd()))
  search_roots <- search_roots[!is.na(search_roots) & nzchar(search_roots) & dir.exists(search_roots)]
  filter_root_norm <- if (!is.null(filter_root) && length(filter_root) == 1 && !is.na(filter_root) && nzchar(filter_root)) {
    normalizePath(filter_root, winslash = "/", mustWork = FALSE)
  } else {
    NA_character_
  }
  postproc_root_norm <- if (!is.null(postproc_root) && length(postproc_root) == 1 && !is.na(postproc_root) && nzchar(postproc_root)) {
    normalizePath(postproc_root, winslash = "/", mustWork = FALSE)
  } else {
    NA_character_
  }

  candidate_files <- map(search_roots, ~ list.files(
    path = .x,
    pattern = "\\.filtered\\.minlen[0-9]+\\.maxcaas[0-9]+\\.tsv$",
    recursive = TRUE,
    full.names = TRUE
  )) %>%
    unlist() %>%
    unique()

  if (length(candidate_files) == 0) {
    return(list(
      path = NA_character_,
      status = "missing",
      message = "No filtered cluster TSV files were found in the report workspace or filter output directories."
    ))
  }

  file_index <- tibble(file_path = candidate_files) %>%
    mutate(
      normalized_path = normalizePath(file_path, winslash = "/", mustWork = FALSE),
      file_name = basename(file_path),
      minlen_match = str_match(file_name, "minlen([0-9]+)\\.maxcaas([0-9]+)"),
      minlen_file = suppressWarnings(as.integer(minlen_match[, 2])),
      maxcaas_file = suppressWarnings(as.integer(minlen_match[, 3])),
      priority = case_when(
        !is.na(filter_root_norm) & str_starts(normalized_path, fixed(filter_root_norm)) ~ 1L,
        !is.na(postproc_root_norm) & str_starts(normalized_path, fixed(postproc_root_norm)) ~ 2L,
        TRUE ~ 3L
      )
    ) %>%
    filter(!is.na(minlen_file), !is.na(maxcaas_file)) %>%
    distinct(normalized_path, .keep_all = TRUE)

  matches <- file_index %>%
    filter(minlen_file == as.integer(minlen), maxcaas_file == as.integer(maxcaas_int)) %>%
    arrange(priority, file_name)

  if (nrow(matches) == 0) {
    return(list(
      path = NA_character_,
      status = "missing",
      message = sprintf(
        "No filtered file matched minlen=%d and maxcaas=%d. Expected any '*.filtered.minlen%d.maxcaas%d.tsv'.",
        as.integer(minlen), as.integer(maxcaas_int), as.integer(minlen), as.integer(maxcaas_int)
      )
    ))
  }

  if (nrow(matches) > 1) {
    return(list(
      path = matches$normalized_path[[1]],
      status = "multiple",
      message = sprintf(
        "Multiple matches found for minlen=%d and maxcaas=%d; using preferred file `%s`.",
        as.integer(minlen), as.integer(maxcaas_int), basename(matches$normalized_path[[1]])
      )
    ))
  }

  list(
    path = matches$normalized_path[[1]],
    status = "ok",
    message = sprintf("Using filtered file `%s`.", basename(matches$normalized_path[[1]]))
  )
}

normalize_disambig_schema <- function(df) {
  # This normalises the DISAMBIGUATION output
  if ('Gene' %in% names(df) && !'gene' %in% names(df)) df <- df %>% rename(gene = Gene)
  if ('Position' %in% names(df) && !'position' %in% names(df)) df <- df %>% rename(position = Position)
  if (!'caap_group' %in% names(df)) df$caap_group <- 'US'
  if (!'convergence_type' %in% names(df)) df$convergence_type <- 'unknown'
  if (!'change_side' %in% names(df)) df$change_side <- 'none'
  if (!'change_top' %in% names(df)) df$change_top <- 'no_change'
  if (!'change_bottom' %in% names(df)) df$change_bottom <- 'no_change'
  if (!'is_conserved_meta' %in% names(df)) df$is_conserved_meta <- FALSE
  if (!'pvalue' %in% names(df)) df$pvalue <- NA_real_
  if (!'pvalue_boot' %in% names(df)) df$pvalue_boot <- NA_real_

  df %>%
    mutate(
      gene = as.character(gene),
      position = suppressWarnings(as.integer(position)),
      caap_group = as.character(caap_group),
      convergence_type = as.character(convergence_type),
      change_side = tolower(as.character(change_side)),
      change_top = tolower(as.character(change_top)),
      change_bottom = tolower(as.character(change_bottom)),
      is_conserved_meta = as_bool(is_conserved_meta),
      pvalue = suppressWarnings(as.numeric(pvalue)),
      pvalue_boot = suppressWarnings(as.numeric(pvalue_boot)),
      sig_hyp = !is.na(pvalue) & pvalue <= alpha_threshold,
      sig_perm = !is.na(pvalue_boot) & pvalue_boot <= alpha_threshold,
      is_significant = sig_hyp & sig_perm,
      sig_hyp_010 = !is.na(pvalue) & pvalue <= 0.10,
      sig_perm_010 = !is.na(pvalue_boot) & pvalue_boot <= 0.10,
      is_significant_010 = sig_hyp_010 & sig_perm_010,
      sig_hyp_005 = !is.na(pvalue) & pvalue <= 0.05,
      sig_perm_005 = !is.na(pvalue_boot) & pvalue_boot <= 0.05,
      is_significant_005 = sig_hyp_005 & sig_perm_005,
      significance_class = make_significance_class(sig_hyp, sig_perm),
      significance_class_010 = make_significance_class(sig_hyp_010, sig_perm_010),
      significance_class_005 = make_significance_class(sig_hyp_005, sig_perm_005),
      site_id = paste(gene, position, sep = '@')
    )
}

seed_val <- 12345
set.seed(seed_val)

discovery_file <- params$discovery_file
filter_summary_file <- params$filter_summary_file
gene_ensembl_file <- params$gene_ensembl_file
gene_stats_file <- params$gene_stats_file
processing_mode <- params$processing_mode
extreme_threshold <- params$extreme_threshold
iqr_multiplier <- params$iqr_multiplier
alpha_threshold <- suppressWarnings(as.numeric(params$alpha_threshold))
if (is.na(alpha_threshold) || alpha_threshold <= 0 || alpha_threshold > 1) alpha_threshold <- 0.10
alpha_threshold_label <- format_threshold(alpha_threshold)
threshold_line_df <- build_threshold_line_df(alpha_threshold)
gene_filter_mode <- params$gene_filter_mode

if (is.null(discovery_file)) stop("Error: discovery_file parameter is required")
if (is.null(filter_summary_file)) stop("Error: filter_summary_file parameter is required")
if (is.null(gene_ensembl_file)) stop("Error: gene_ensembl_file parameter is required")

stopifnot(file.exists(discovery_file), file.exists(filter_summary_file), file.exists(gene_ensembl_file))

ct_postproc_dir <- "."
outliers_dir <- "outliers"
cluster_dir <- "clusters"
summary_stats_dir <- "summary_statistics"
disambig_dir <- "disambiguation_characterization"
overlap_out_dir <- file.path(disambig_dir, "overlap_figures")

createDir(ct_postproc_dir)
createDir(outliers_dir)
createDir(cluster_dir)
createDir(summary_stats_dir)
createDir(disambig_dir)
createDir(overlap_out_dir)

filter_root <- params$filter_dir
postproc_root <- if (!is.null(filter_root) && !is.na(filter_root)) dirname(filter_root) else NA_character_
removed_precluster_file <- find_first_existing(c(
  file.path(filter_root, "preprocessed", "removed_patterns_precluster.tsv"),
  file.path(postproc_root, "preprocessed", "removed_patterns_precluster.tsv"),
  find_recursive_file(getwd(), "removed_patterns_precluster\\.tsv$")
))
removed_genes_file <- find_first_existing(c(
  file.path(filter_root, "gene_filtering", "removed_genes_summary.tsv"),
  file.path(postproc_root, "gene_filtering", "removed_genes_summary.tsv"),
  find_recursive_file(getwd(), "removed_genes_summary\\.tsv$")
))

tableau_10 <- paletteer_d("ggthemes::Tableau_10")
names(tableau_10) <- c(
  "blue", "orange", "vermillion", "bluish_green",
  "green", "yellow", "reddish_purple", "pink",
  "brown", "gray"
)

threshold_line_colours <- c(
  configured = tableau_10[["orange"]],
  comparison_010 = tableau_10[["blue"]],
  comparison_005 = tableau_10[["gray"]]
)
threshold_line_types <- c(
  configured = "solid",
  comparison_010 = "dashed",
  comparison_005 = "dotdash"
)

category_palette <- c(
  "Normal" = tableau_10[["gray"]],
  "Dubious" = tableau_10[["blue"]],
  "Extreme" = tableau_10[["vermillion"]],
  "Both" = tableau_10[["reddish_purple"]]
)

removal_palette <- c(
  Discarded = tableau_10[["vermillion"]],
  Good = tableau_10[["green"]]
)
```

# A) Post-processing characterization

## Data Loading and Gene length Annotation

```{r data_loading}
raw_discovery <- read_delim(discovery_file, delim = "\t", show_col_types = FALSE)
discovery_std <- normalize_disambig_schema(raw_discovery)
convergence_plot_df <- discovery_std %>%
  mutate(
    change_side = if_else(change_side %in% c("top", "bottom", "both"), change_side, "none"),
    pattern_parallel = make_pattern_parallel_label(convergence_type),
    pattern_parallel = forcats::fct_infreq(pattern_parallel)
  )
orientation_rows <- build_orientation_rows(convergence_plot_df) %>%
  mutate(caap_group = toupper(caap_group))
orientation_flags <- build_group_flag_table(orientation_rows)

discovery_df <- discovery_std %>%
  transmute(
    caap_group = caap_group,
    Gene = gene,
    Position = position,
    convergence_type = convergence_type,
    change_side,
    is_conserved_meta,
    pvalue = pvalue,
    pvalue_boot = pvalue_boot,
    sig_hyp,
    sig_perm,
    is_significant,
    significance_class,
    site_id
  )

if (!"caap_group" %in% names(discovery_df)) discovery_df <- discovery_df %>% mutate(caap_group = "US")

gene_ensembl <- read_tsv(gene_ensembl_file, show_col_types = FALSE)
names(gene_ensembl) <- tolower(names(gene_ensembl))
if (!"gene" %in% names(gene_ensembl) || !"length" %in% names(gene_ensembl)) {
  stop("Gene ensembl file must have 'gene' and 'length' columns")
}

gene_length <- gene_ensembl %>% select(Gene = gene, length = length)
discovery_df <- discovery_df %>% left_join(gene_length, by = "Gene") %>% filter(!is.na(length))

cat(
  "=== CT Post-Processing Configuration ===\n",
  "Processing mode:", processing_mode, "\n",
  "Gene filtering:", gene_filter_mode, "\n",
  "Selected alpha threshold:", alpha_threshold_label, "\n",
  "Extreme threshold:", extreme_threshold, paste0("(top ", 100*(1-extreme_threshold), "%)"), "\n",
  "IQR multiplier:", iqr_multiplier, "\n",
  "\nInput files:\n",
  "  Discovery:", discovery_file, "\n",
  "  Filter summary:", filter_summary_file, "\n",
  "  Gene ensembl:", gene_ensembl_file, "\n",
  "  Gene stats:", ifelse(is.null(gene_stats_file) || is.na(gene_stats_file), "Not provided (will recompute)", gene_stats_file), "\n",
  "\n  Loaded CT positions:", nrow(discovery_df), "\n",
  "  Unique genes with length annotation:", n_distinct(discovery_df$Gene), "\n"
)
```

## Conserved metadata composition after precluster pruning

```{r conserved_meta_composition}
precluster_removed_burden <- tibble()
if (!is.na(removed_precluster_file)) {
  precluster_removed_burden <- read_delim(removed_precluster_file, delim = "\t", show_col_types = FALSE) %>%
    normalize_disambig_schema() %>%
    mutate(removal_reason = if_else(is.na(removal_reason), "unknown", removal_reason))
}

rows_retained <- nrow(discovery_std)
rows_removed_mrca <- sum(str_detect(precluster_removed_burden$removal_reason, "^mrca_posterior_below_"), na.rm = TRUE)
rows_total <- rows_retained + rows_removed_mrca

conserved_status_burden <- tibble(
  scenario = c("all_positions", "retained_positions", "retained_conserved_meta", "retained_meta_not_conserved", "removed_low_mrca"),
  description = c(
    "Retained plus precluster-removed positions tracked by CT postproc",
    "Positions retained after precluster hard filters",
    "Retained positions where is_conserved_meta = TRUE",
    "Retained positions where is_conserved_meta = FALSE",
    "Precluster removals due to MRCA posterior threshold"
  ),
  rows = c(
    rows_total,
    rows_retained,
    sum(discovery_std$is_conserved_meta, na.rm = TRUE),
    sum(!discovery_std$is_conserved_meta, na.rm = TRUE),
    rows_removed_mrca
  )
) %>%
  mutate(fraction_of_total = if (rows_total > 0) rows / rows_total else NA_real_)

write_tsv(conserved_status_burden, file.path(disambig_dir, "conserved_status_burden.tsv"))

kable(
  conserved_status_burden %>% mutate(fraction_of_total = percent(fraction_of_total, accuracy = 0.1)),
  caption = "Conserved-status burden across retained and precluster-removed CT positions"
) %>% kable_styling()
```

## Outlier Gene Detection

```{r outlier_detection, fig.height=10, fig.width=14}
# Load gene statistics from precomputed file or compute fallback
gene_summary_annotated <- NULL

if (!is.null(gene_stats_file) && !is.na(gene_stats_file) && file.exists(gene_stats_file)) {
  cat("Loading precomputed gene statistics from:", gene_stats_file, "\n")
  gene_summary_annotated <- read_tsv(gene_stats_file, show_col_types = FALSE)

  cat("Loaded gene statistics for", nrow(gene_summary_annotated), "genes\n")

  # Ensure all downstream-required columns exist (file may lack some)
  if (!"caap_group" %in% names(gene_summary_annotated)) {
    gene_summary_annotated$caap_group <- "US"
  }
  if (!"n_CAAS_per_length" %in% names(gene_summary_annotated)) {
    if ("n_CAAS" %in% names(gene_summary_annotated) && "length" %in% names(gene_summary_annotated)) {
      gene_summary_annotated$n_CAAS_per_length <- gene_summary_annotated$n_CAAS / gene_summary_annotated$length * 100
    } else {
      gene_summary_annotated$n_CAAS_per_length <- NA_real_
    }
  }
  if (!"Extreme" %in% names(gene_summary_annotated))          gene_summary_annotated$Extreme          <- FALSE
  if (!"Dubious" %in% names(gene_summary_annotated))          gene_summary_annotated$Dubious          <- FALSE
  if (!"category" %in% names(gene_summary_annotated))         gene_summary_annotated$category         <- "Normal"
  if (!"threshold_extreme" %in% names(gene_summary_annotated)) gene_summary_annotated$threshold_extreme <- NA_real_
  if (!"threshold_dubious" %in% names(gene_summary_annotated)) gene_summary_annotated$threshold_dubious <- NA_real_
  
} else {
  cat("Gene statistics file not found or not provided\n")
}

# Ensure caap_group column exists for plotting
if (!"caap_group" %in% names(gene_summary_annotated)) {
  gene_summary_annotated$caap_group <- "US"
}

cat("Groups found:", paste(sort(unique(gene_summary_annotated$caap_group)), collapse=", "), "\n")

if (nrow(gene_summary_annotated) == 0) {
  cat("WARNING: gene_summary_annotated is empty — skipping outlier distribution plots.\n")
} else {

# Pre-compute vline data — guard against empty frames that break facet_wrap
vline_dubious_data  <- gene_summary_annotated %>% filter(!is.na(threshold_dubious))  %>% distinct(caap_group, threshold_dubious)
vline_extreme_data  <- gene_summary_annotated %>% filter(!is.na(threshold_extreme)) %>% distinct(caap_group, threshold_extreme)

# Generate histograms with increased font sizes for slides
p_gross_dist <- ggplot(gene_summary_annotated, aes(x = n_CAAS, fill = category)) +
  geom_histogram(binwidth = 0.05, color = "black", alpha = 0.8) +
  scale_fill_manual(values = category_palette, breaks = c("Extreme", "Dubious", "Both", "Normal")) +
  scale_x_log10(breaks = c(1, 2, 5, 10, 20, 50, 100, 200)) +
  facet_wrap(~caap_group, scales = "free_y") +
  labs(title = "CAAP Count Distribution by Group",
       subtitle = "Dubious threshold (Q3 + 3×IQR) shown as dashed line",
       x = "Number of CAAP Positions", y = "Gene Count") +
  theme_minimal(base_size = 16) +
  theme(
    legend.position = "bottom",
    axis.title.y = element_text(size = 15, margin = ggplot2::margin(l = 20, r = 20)),
    axis.title.x = element_text(size = 15, margin = ggplot2::margin(t = 12)),
    plot.title = element_text(face = "bold", size = 18, hjust = 0, margin = ggplot2::margin(b = 6)),
    plot.subtitle = element_text(size = 14, hjust = 0, color = "grey40", margin = ggplot2::margin(b = 10)),
    strip.text = element_text(face = "bold", size = 15)
  )
if (nrow(vline_dubious_data) > 0) {
  p_gross_dist <- p_gross_dist +
    geom_vline(data = vline_dubious_data, aes(xintercept = threshold_dubious),
               color = tableau_10[["blue"]], linetype = "dashed", linewidth = 1)
}

gross_zoom_upper <- derive_plot_zoom_upper(p_gross_dist, zoom_fraction = 0.10, min_upper = 5)
p_zoom_gross <- p_gross_dist + coord_cartesian(ylim = c(0, gross_zoom_upper)) +
  labs(title = paste0("Zoomed View (bottom 10% of y-range; y ≤ ", gross_zoom_upper, ")"))

p_density_dist <- ggplot(gene_summary_annotated, aes(x = n_CAAS_per_length, fill = category)) +
  geom_histogram(binwidth = 0.05, color = "black", alpha = 0.8) +
  scale_fill_manual(values = category_palette, breaks = c("Extreme", "Dubious", "Both", "Normal")) +
  scale_x_log10(breaks = c(0.01, 0.02, 0.05, 0.1, 0.2, 0.5, 1, 2, 5, 10, 20, 50, 100)) +
  facet_wrap(~ caap_group, scales = "free_y") +
  labs(title = "CAAP Density Distribution by Group",
       subtitle = paste0("Top ", 100*(1-extreme_threshold), "% density threshold shown as dashed line"),
       x = "CAAP Density (positions / gene length)", y = "Gene Count") +
  theme_minimal(base_size = 16) +
  theme(
    legend.position = "bottom",
    axis.title.y = element_text(size = 15, margin = ggplot2::margin(l = 20, r = 20)),
    axis.title.x = element_text(size = 15, margin = ggplot2::margin(t = 12)),
    plot.title = element_text(face = "bold", size = 18, hjust = 0, margin = ggplot2::margin(b = 6)),
    plot.subtitle = element_text(size = 14, hjust = 0, color = "grey40", margin = ggplot2::margin(b = 10)),
    strip.text = element_text(face = "bold", size = 15)
  )
if (nrow(vline_extreme_data) > 0) {
  p_density_dist <- p_density_dist +
    geom_vline(data = vline_extreme_data, aes(xintercept = threshold_extreme),
               color = tableau_10[["vermillion"]], linetype = "dashed", linewidth = 1)
}

density_zoom_upper <- derive_plot_zoom_upper(p_density_dist, zoom_fraction = 0.10, min_upper = 5)
p_zoom_density <- p_density_dist + coord_cartesian(ylim = c(0, density_zoom_upper)) +
  labs(title = paste0("Zoomed View (bottom 10% of y-range; y ≤ ", density_zoom_upper, ")"))

combined_outlier_plot <- p_gross_dist / p_density_dist
combined_zoomed_plot <- p_zoom_gross / p_zoom_density

print(combined_outlier_plot)
print(combined_zoomed_plot)

facet_count_outliers <- n_distinct(gene_summary_annotated$caap_group)
outlier_dims <- compute_barplot_dims(
  n_labels = 8,
  n_facets = facet_count_outliers,
  facet_cols = min(3, facet_count_outliers),
  base_width = 12,
  base_height = 8,
  per_label = 0.35,
  facet_row_height = 1.8,
  max_width = 20,
  max_height = 14
)
ggsave(file.path(outliers_dir, "outlier_distributions.png"), combined_outlier_plot,
       width = outlier_dims$width, height = outlier_dims$height, dpi = 320)
ggsave(file.path(outliers_dir, "outlier_distributions_zoomed.png"), combined_zoomed_plot,
       width = outlier_dims$width, height = outlier_dims$height, dpi = 320)

# Dedicated bar chart: removed gene counts per category and caap_group
# Clearly separates Extreme-only, Dubious-only, and Both (genes failing both criteria).
removed_cat_summary <- gene_summary_annotated %>%
  filter(category != "Normal") %>%
  count(caap_group, category)

if (nrow(removed_cat_summary) > 0) {
  p_removed_cats <- ggplot(removed_cat_summary,
                           aes(x = category, y = n, fill = category)) +
    geom_col(color = "black", linewidth = 0.3, width = 0.7) +
    facet_wrap(~ caap_group, scales = "free_y", ncol = min(3, n_distinct(removed_cat_summary$caap_group))) +
    scale_fill_manual(
      values = category_palette,
      breaks = c("Extreme", "Dubious", "Both")
    ) +
    labs(
      title    = "Removed Genes by category and Group",
      subtitle = paste0(
        "Extreme: top ", 100 * (1 - extreme_threshold), "% density  |  ",
        "Dubious: Q3 + ", iqr_multiplier, "\u00d7IQR count  |  ",
        "Both: both criteria"
      ),
      x = "Removal category",
      y = "Gene Count"
    ) +
    theme_minimal(base_size = 16) +
    theme(
      legend.position  = "none",
      axis.title.y     = element_text(size = 15, margin = ggplot2::margin(l = 20, r = 20)),
      axis.title.x     = element_text(size = 15, margin = ggplot2::margin(t = 12)),
      plot.title       = element_text(face = "bold", size = 18, hjust = 0, margin = ggplot2::margin(b = 6)),
      plot.subtitle    = element_text(size = 14, hjust = 0, color = "grey40", margin = ggplot2::margin(b = 10)),
      axis.text.x      = element_text(angle = 25, hjust = 1),
      strip.text       = element_text(face = "bold", size = 15)
    )

  print(p_removed_cats)
  save_barplot(
    file.path(outliers_dir, "removed_genes_by_category.png"),
    p_removed_cats,
    n_labels = n_distinct(removed_cat_summary$category),
    n_facets = n_distinct(removed_cat_summary$caap_group),
    facet_cols = min(3, n_distinct(removed_cat_summary$caap_group)),
    base_width = 9,
    base_height = 5.5
  )
}

thresholds_summary <- gene_summary_annotated %>% distinct(caap_group, threshold_extreme, threshold_dubious) %>% arrange(caap_group)
thresh_lines <- vapply(seq_len(nrow(thresholds_summary)), function(i) {
  sprintf(
    "  %s:\n    Extreme (top %.0f%% density): %.4f\n    Dubious (Q3 + %.1f×IQR): %.0f positions\n",
    thresholds_summary$caap_group[i],
    100 * (1 - extreme_threshold),
    thresholds_summary$threshold_extreme[i],
    iqr_multiplier,
    thresholds_summary$threshold_dubious[i]
  )
}, character(1))

outlier_counts <- gene_summary_annotated %>% count(caap_group, category) %>% pivot_wider(names_from = category, values_from = n, values_fill = 0)

cat(
  "\n=== Outlier Detection Summary ===\n",
  "CAAP/US unified mode: Group-aware thresholds calculated\n\n",
  "Thresholds by group:\n",
  paste(thresh_lines, collapse = "\n"), "\n",
  "Gene counts by group and category:\n"
)
print(outlier_counts)

write_tsv(gene_summary_annotated, file.path(outliers_dir, "gene_summary_annotated.tsv"))

} # end if (nrow(gene_summary_annotated) > 0)
```

## Cluster Filtering Analysis

```{r cluster_analysis_load}
filter_summary <- read_tsv(filter_summary_file, show_col_types = FALSE)
load_msgs <- c(
  "Loading cluster filter summary...\n",
  sprintf("  Parameter combinations tested: %d\n", nrow(filter_summary))
)

if (!"Minlen" %in% names(filter_summary) || !"Maxcaas" %in% names(filter_summary)) {
  if ("Parameter" %in% names(filter_summary)) {
    parsed_params <- stringr::str_split_fixed(as.character(filter_summary$Parameter), "/", 2)
    if (!"Minlen" %in% names(filter_summary)) filter_summary$Minlen <- suppressWarnings(as.integer(parsed_params[, 1]))
    if (!"Maxcaas" %in% names(filter_summary)) filter_summary$Maxcaas <- suppressWarnings(as.numeric(parsed_params[, 2]))
    load_msgs <- c(load_msgs, "  Parsed missing Minlen/Maxcaas from Parameter column\n")
  }
}

if (!"DiscardedCount" %in% names(filter_summary)) {
  protected_cols <- c("Parameter", "Minlen", "Maxcaas", "caap_groups", "Groups_List", "N_Groups")
  candidate_cols <- setdiff(names(filter_summary), protected_cols)
  numeric_candidate_cols <- candidate_cols[vapply(candidate_cols, function(col) is.numeric(filter_summary[[col]]), logical(1))]
  if (length(numeric_candidate_cols) > 0) {
    filter_summary <- filter_summary %>% mutate(DiscardedCount = rowSums(across(all_of(numeric_candidate_cols)), na.rm = TRUE))
    load_msgs <- c(load_msgs, sprintf("  Built DiscardedCount from columns: %s\n", paste(numeric_candidate_cols, collapse = ", ")))
  } else {
    filter_summary$DiscardedCount <- 0L
    load_msgs <- c(load_msgs, "  Warning: Could not infer DiscardedCount from summary columns; defaulting to 0\n")
  }
} else {
  filter_summary$DiscardedCount <- suppressWarnings(as.numeric(filter_summary$DiscardedCount))
  filter_summary$DiscardedCount[is.na(filter_summary$DiscardedCount)] <- 0
}

has_caap_groups_in_summary <- 'caap_groups' %in% names(filter_summary) && any(filter_summary$caap_groups != 'N/A' & filter_summary$caap_groups != '')
if (has_caap_groups_in_summary) {
  load_msgs <- c(load_msgs, "  CAAP mode detected in filter summary\n")
  filter_summary <- filter_summary %>% mutate(Groups_List = strsplit(as.character(caap_groups), ","), N_Groups = lengths(Groups_List))
}

cat(paste(load_msgs, collapse = ""))

if (has_caap_groups_in_summary) {
  group_cols <- setdiff(names(filter_summary), c('Parameter', 'Minlen', 'Maxcaas'))
  filter_table <- filter_summary %>% select(Parameter, all_of(sort(group_cols)))
} else {
  filter_table <- filter_summary %>% select(Parameter, DiscardedCount)
  names(filter_table)[2] <- "Discarded"
}

kable(filter_table, caption = "Cluster Filtering Parameter Sweep Results", align = c("l", rep("r", ncol(filter_table) - 1))) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

if (nrow(filter_summary) > 1) {
  if (has_caap_groups_in_summary) {
    group_cols <- setdiff(names(filter_summary), c('Parameter', 'Minlen', 'Maxcaas'))
    plot_data <- filter_summary %>%
      pivot_longer(cols = all_of(group_cols), names_to = "caap_group", values_to = "DiscardedCount") %>%
      mutate(param_label = sprintf("%d/%d", Minlen, as.integer(Maxcaas)), param_label = factor(param_label, levels = unique(param_label)))

    p_param_sweep <- ggplot(plot_data, aes(x = param_label, y = DiscardedCount, fill = caap_group)) +
      geom_col(position = "dodge", color = "black", linewidth = 0.3) +
      scale_fill_paletteer_d("ggthemes::Tableau_10") +
      labs(title = "Discarded CAAS Positions by Parameter Combination",
           subtitle = "Grouped by CAAP group - minlen/maxcaas combinations",
           x = "Parameter Combination (minlen/maxcaas)", y = "Positions Discarded", fill = "CAAP Group") +
      theme_minimal(base_size = 16) +
      theme(legend.position = "right", axis.title.y = element_text(size = 15, margin = ggplot2::margin(l = 20, r = 20)), axis.title.x = element_text(size = 15, margin = ggplot2::margin(t = 12)), plot.title = element_text(face = "bold", size = 18, hjust = 0, margin = ggplot2::margin(b = 6)), plot.subtitle = element_text(size = 14, hjust = 0, color = "grey40", margin = ggplot2::margin(b = 10)), axis.text.x = element_text(angle = 55, hjust = 1))
  } else {
    p_param_sweep <- ggplot(filter_summary, aes(x = factor(Minlen), y = DiscardedCount, color = factor(Maxcaas), group = factor(Maxcaas))) +
      geom_point(size = 4) +
      geom_line(linewidth = 1.2) +
      scale_color_paletteer_d("ggthemes::Tableau_10", labels = function(x) sprintf("%.0f%%", as.numeric(x))) +
      labs(title = "Discarded CAAS Positions by Parameter Combination",
           subtitle = "Effect of minimum cluster length and maximum density threshold",
           x = "Minimum Cluster length (minlen)", y = "Positions Discarded", color = "Max Density") +
      theme_minimal(base_size = 16) +
      theme(legend.position = "right", axis.title.y = element_text(size = 15, margin = ggplot2::margin(l = 20, r = 20)), axis.title.x = element_text(size = 15, margin = ggplot2::margin(t = 12)), plot.title = element_text(face = "bold", size = 18, hjust = 0, margin = ggplot2::margin(b = 6)), plot.subtitle = element_text(size = 14, hjust = 0, color = "grey40", margin = ggplot2::margin(b = 10)))
  }

  print(p_param_sweep)
  if (has_caap_groups_in_summary) {
    save_barplot(
      file.path(cluster_dir, "parameter_sweep_overview.png"),
      p_param_sweep,
      n_labels = n_distinct(plot_data$param_label),
      n_facets = 1,
      facet_cols = 1,
      base_width = 10,
      base_height = 6,
      per_label = 0.65
    )
  } else {
    ggsave(file.path(cluster_dir, "parameter_sweep_overview.png"), p_param_sweep, width = 12, height = 6, dpi = 320)
  }
}

write_tsv(filter_summary, file.path(cluster_dir, "filter_summary.tsv"))
```

```{r cluster_detailed_analysis, results='asis', fig.height=8, fig.width=12, eval=(processing_mode == 'exploratory' && nrow(filter_summary) > 1)}
if (processing_mode == 'exploratory' && nrow(filter_summary) > 1) {
  cat("\n## Detailed Parameter Analysis\n\n")
  detail_filter_root <- params$filter_dir
  detail_postproc_root <- if (!is.null(detail_filter_root) && !is.na(detail_filter_root)) dirname(detail_filter_root) else NA_character_

  for (i in 1:nrow(filter_summary)) {
    param_row <- filter_summary[i, ]
    minlen_i <- suppressWarnings(as.integer(param_row$Minlen))
    maxcaas_i <- suppressWarnings(as.integer(param_row$Maxcaas))
    param_label <- sprintf("minlen=%d, maxcaas=%d", minlen_i, maxcaas_i)
    cat(sprintf("\n### Parameter Set %d: %s {.tabset}\n\n", i, param_label))

    resolved_file <- resolve_filtered_cluster_file(
      minlen_i,
      maxcaas_i,
      filter_root = detail_filter_root,
      postproc_root = detail_postproc_root
    )
    cat(sprintf("Resolved filtered file status: %s\n\n", resolved_file$message))

    if (is.na(resolved_file$path) || !file.exists(resolved_file$path)) {
      cat("No per-parameter file was available for this combination, so no spatial summary could be rendered.\n")
      cat("\n---\n\n")
      next
    }

    filtered_data <- read_tsv(resolved_file$path, show_col_types = FALSE)
    if (!"clustering_flag" %in% names(filtered_data)) {
      cat("Matched file does not contain `clustering_flag`; skipping this parameter set.\n")
      cat("\n---\n\n")
      next
    }

    discarded <- filtered_data %>% filter(clustering_flag == "Discarded")
    if (nrow(discarded) == 0) {
      cat("No discarded clusters were found for this parameter set.\n")
      cat("\n---\n\n")
      next
    }

    if (!"caap_group" %in% names(discarded)) {
      discarded <- discarded %>% mutate(caap_group = "US")
    }

    detect_runs <- function(positions) {
      positions <- sort(unique(positions))
      if (length(positions) == 0) return(integer(0))
      breaks <- c(0, which(diff(positions) > 1), length(positions))
      diff(breaks)
    }

    spatial_analysis <- discarded %>%
      group_by(caap_group, Gene) %>%
      summarise(positions = list(Position), .groups = "drop") %>%
      mutate(run_lengths = map(positions, detect_runs)) %>%
      unnest(run_lengths, keep_empty = TRUE) %>%
      mutate(
        cluster_type = case_when(
          run_lengths == 1 ~ "Singlet (1)",
          run_lengths == 2 ~ "Doublet (2)",
          run_lengths == 3 ~ "Triplet (3)",
          run_lengths == 4 ~ "Quad (4)",
          run_lengths == 5 ~ "Quint (5)",
          run_lengths >= 6 ~ "6+ Long",
          TRUE ~ "Other"
        )
      )

    spatial_summary <- spatial_analysis %>%
      count(caap_group, cluster_type) %>%
      complete(caap_group, cluster_type, fill = list(n = 0)) %>%
      group_by(caap_group) %>%
      mutate(percentage = round(100 * n / sum(n), 1)) %>%
      ungroup()

    p_spatial <- ggplot(spatial_summary, aes(x = caap_group, y = percentage / 100, fill = cluster_type)) +
      geom_col(position = "fill", color = "black", size = 0.2) +
      scale_fill_paletteer_d("ggthemes::Tableau_10") +
      scale_y_continuous(labels = scales::percent) +
      labs(
        title = sprintf("Spatial Clustering: %s", param_label),
        subtitle = sprintf("Discarded positions only | %s", resolved_file$message),
        x = "Group",
        y = "Percentage",
        fill = "Cluster Type"
      ) +
      theme_minimal(base_size = 16) +
      theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1))

    print(p_spatial)
    cat("\n---\n\n")
  }
}
```

```{r disambig_unified_data, eval=(processing_mode != 'exploratory'), results='asis'}
cat("\n# B) Characterizing CAAS disambiguation\n\n")
cat("## B.1 Build unified removal + retention dataset\n\n")
cluster_files <- list.files(path = params$filter_dir, pattern = "\\.filtered\\.minlen[0-9]+\\.maxcaas[0-9]+\\.tsv$", full.names = TRUE, recursive = TRUE)
if (length(cluster_files) == 0) {
  cluster_files <- list.files(path = ifelse(is.na(postproc_root), getwd(), postproc_root), pattern = "\\.filtered\\.minlen[0-9]+\\.maxcaas[0-9]+\\.tsv$", full.names = TRUE, recursive = TRUE)
}

precluster_events <- tibble()
if (!is.na(removed_precluster_file)) {
  precluster_df <- normalize_disambig_schema(read_delim(removed_precluster_file, delim = "\t", show_col_types = FALSE))
  precluster_events <- precluster_df %>% transmute(
    site_id, gene, position, caap_group, convergence_type, pvalue, pvalue_boot,
    change_top, change_bottom, change_side,
    sig_hyp, sig_perm, is_significant, significance_class,
    removal_stage = case_when(
      str_detect(removal_reason, "^mrca_posterior_below_") ~ "precluster_mrca_threshold",
      TRUE ~ "precluster_other"
    )
  )
}

cluster_events <- tibble()
if (length(cluster_files) > 0) {
  cluster_df <- map_dfr(cluster_files, ~ normalize_disambig_schema(read_tsv(.x, show_col_types = FALSE)))
  cluster_hits <- cluster_df %>% filter(clustering_flag == "Discarded") %>% transmute(gene = as.character(gene), position = as.integer(position), caap_group = as.character(caap_group)) %>% distinct()
  cluster_events <- discovery_std %>%
    inner_join(cluster_hits, by = c("gene", "position", "caap_group")) %>%
    transmute(
      site_id, gene, position, caap_group, convergence_type, pvalue, pvalue_boot,
      change_top, change_bottom, change_side,
      sig_hyp, sig_perm, is_significant, significance_class,
      removal_stage = "cluster"
    ) %>%
    distinct()
}

gene_events <- tibble()
if (!is.na(removed_genes_file)) {
  rg <- read_tsv(removed_genes_file, show_col_types = FALSE)
  names(rg) <- tolower(names(rg))
  if (!"caap_group" %in% names(rg)) rg$caap_group <- "US"
  rg_norm <- rg %>% transmute(gene = as.character(gene), caap_group = as.character(caap_group), category = as.character(category)) %>% distinct()

  gene_events <- discovery_std %>%
    inner_join(rg_norm, by = c("gene", "caap_group")) %>%
    mutate(removal_stage = case_when(
      tolower(category) == "extreme" ~ "extreme_gene",
      tolower(category) == "dubious" ~ "dubious_with_clusters",
      tolower(category) == "both" ~ "extreme_and_dubious",
      TRUE ~ "gene_filter"
    )) %>%
    transmute(
      site_id, gene, position, caap_group, convergence_type, pvalue, pvalue_boot,
      change_top, change_bottom, change_side,
      sig_hyp, sig_perm, is_significant, significance_class, removal_stage
    ) %>%
    distinct()
}

removal_events <- bind_rows(precluster_events, cluster_events, gene_events)
write_tsv(removal_events, file.path(disambig_dir, "removal_events.tsv"))

```

```{r filter_overlap_gene_table, eval=(processing_mode != 'exploratory'), results='asis'}
cat("\n### B.1a Gene-Level Filter Overlap Diagnostics {.tabset}\n\n")
# ─────────────────────────────────────────────────────────────────────────────
# Build a gene-level membership table that records for each gene whether it
# was flagged by each of the three independent filters:
#   • Extreme   – top 1% CAAS density (gene_summary_annotated$Extreme)
#   • Dubious   – IQR-outlier CAAS count (gene_summary_annotated$Dubious)
#   • Cluster   – at least one position removed by the cluster filter
# ─────────────────────────────────────────────────────────────────────────────

# --- 1. Gene-level filter flags -------------------------------------------

# Extreme / Dubious come straight from the already-computed gene_summary_annotated
gene_filter_flags <- gene_summary_annotated %>%
  transmute(
    Gene        = Gene,
    caap_group  = caap_group,
    n_CAAS      = n_CAAS,
    length      = length,
    n_CAAS_per_length = n_CAAS_per_length,
    Extreme     = as.logical(Extreme),
    Dubious     = as.logical(Dubious)
  )

# Cluster_removed: any gene that contributed at least one site to cluster_events
cluster_removed_genes <- if (nrow(cluster_events) > 0) {
  cluster_events %>%
    transmute(Gene = gene, caap_group = caap_group) %>%
    distinct() %>%
    mutate(Cluster_removed = TRUE)
} else {
  tibble(Gene = character(), caap_group = character(), Cluster_removed = logical())
}

gene_filter_flags <- gene_filter_flags %>%
  left_join(cluster_removed_genes, by = c("Gene", "caap_group")) %>%
  mutate(
    Cluster_removed = coalesce(Cluster_removed, FALSE),
    # Composite label
    filter_combo = case_when(
      Extreme & Dubious & Cluster_removed ~ "Extreme+Dubious+Cluster",
      Extreme & Dubious                  ~ "Extreme+Dubious",
      Extreme & Cluster_removed          ~ "Extreme+Cluster",
      Dubious & Cluster_removed          ~ "Dubious+Cluster",
      Extreme                            ~ "Extreme",
      Dubious                            ~ "Dubious",
      Cluster_removed                    ~ "Cluster",
      TRUE                               ~ "None"
    ),
    filter_combo = factor(filter_combo, levels = c(
      "None", "Extreme", "Dubious", "Cluster",
      "Extreme+Dubious", "Extreme+Cluster", "Dubious+Cluster",
      "Extreme+Dubious+Cluster"
    ))
  )

# --- 2. Per-gene significance descriptors from full discovery data ----------

gene_sig_stats <- discovery_std %>%
  group_by(gene, caap_group) %>%
  summarise(
    frac_is_significant = mean(is_significant, na.rm = TRUE),
    median_pval_hyp  = median(pvalue,      na.rm = TRUE),
    min_pval_hyp     = min(pvalue,         na.rm = TRUE),
    median_pval_boot = median(pvalue_boot, na.rm = TRUE),
    min_pval_boot    = min(pvalue_boot,    na.rm = TRUE),
    .groups = "drop"
  ) %>%
  rename(Gene = gene, caap_group = caap_group)

gene_filter_flags <- gene_filter_flags %>%
  left_join(gene_sig_stats, by = c("Gene", "caap_group"))

# --- 3. Quick membership summary printed to console ------------------------

membership_counts <- gene_filter_flags %>%
  count(caap_group, filter_combo, .drop = FALSE) %>%
  arrange(caap_group, filter_combo)

filter_totals_msg <- sprintf(
  "\nFilter-specific totals:\n  Extreme only or combined : %d genes\n  Dubious only or combined : %d genes\n  Cluster only or combined : %d genes\n  Any filter               : %d / %d genes (%.1f%%)\n",
  sum(gene_filter_flags$Extreme, na.rm = TRUE),
  sum(gene_filter_flags$Dubious, na.rm = TRUE),
  sum(gene_filter_flags$Cluster_removed, na.rm = TRUE),
  sum(gene_filter_flags$filter_combo != "None"),
  nrow(gene_filter_flags),
  100 * mean(gene_filter_flags$filter_combo != "None")
)

sig_sites_msg <- ""
if (nrow(removal_events) > 0) {
  total_is_significant  <- sum(discovery_std$is_significant, na.rm = TRUE)
  removed_extreme <- discovery_std %>%
    filter(is_significant, gene %in% filter(gene_filter_flags, Extreme)$Gene,
           caap_group %in% filter(gene_filter_flags, Extreme)$caap_group)
  removed_dubious <- discovery_std %>%
    filter(is_significant, gene %in% filter(gene_filter_flags, Dubious)$Gene,
           caap_group %in% filter(gene_filter_flags, Dubious)$caap_group)
  removed_cluster <- discovery_std %>%
    filter(is_significant, site_id %in% cluster_events$site_id)
  
  sig_sites_msg <- sprintf(
    "  Significant sites in total  : %d\n  Significant in Extreme genes: %d (%.1f%%)\n  Significant in Dubious genes: %d (%.1f%%)\n  Significant in Cluster sites: %d (%.1f%%)\n",
    total_is_significant,
    nrow(removed_extreme), 100 * nrow(removed_extreme) / max(total_is_significant, 1),
    nrow(removed_dubious), 100 * nrow(removed_dubious) / max(total_is_significant, 1),
    nrow(removed_cluster), 100 * nrow(removed_cluster) / max(total_is_significant, 1)
  )
}

cat("\n=== Gene-Level Filter Overlap Summary ===\n")
print(membership_counts)
cat(filter_totals_msg, sig_sites_msg, "\n")
```

```{r filter_overlap_stats, fig.width=14, fig.height=16, eval=(processing_mode != 'exploratory'), results='asis'}
cat("\n#### Descriptor distributions by filter membership\n\n")
# Ordered factor for x-axis: None → single → double → triple
combo_levels <- c(
  "None", "Extreme", "Dubious", "Cluster",
  "Extreme+Dubious", "Extreme+Cluster", "Dubious+Cluster",
  "Extreme+Dubious+Cluster"
)

gene_filter_plot <- gene_filter_flags %>%
  mutate(filter_combo = factor(as.character(filter_combo), levels = combo_levels))

# Colour palette for combos
combo_palette <- setNames(
  c(
    tableau_10[["gray"]],
    tableau_10[["vermillion"]], tableau_10[["blue"]], tableau_10[["green"]],
    tableau_10[["reddish_purple"]], tableau_10[["orange"]], tableau_10[["bluish_green"]],
    tableau_10[["pink"]]
  ),
  combo_levels
)

make_descriptor_plot <- function(data, y_var, y_label, log_scale = FALSE) {
  plot_data <- data %>% filter(!is.na(.data[[y_var]]), !is.na(caap_group))
  if (nrow(plot_data) == 0 || length(unique(plot_data$caap_group)) == 0) {
    return(
      ggplot() +
        annotate("text", x = 0.5, y = 0.5, label = paste("No data for:", y_label),
                 size = 5, hjust = 0.5) +
        theme_void()
    )
  }
  p <- ggplot(plot_data,
              aes(x = filter_combo, y = .data[[y_var]],
                  fill = filter_combo, colour = filter_combo)) +
    geom_boxplot(alpha = 0.5, outlier.shape = NA, linewidth = 0.6) +
    geom_jitter(width = 0.18, alpha = 0.35, size = 0.9, shape = 16) +
    facet_wrap(~ caap_group, scales = "free_y") +
    scale_fill_manual(values   = combo_palette, drop = TRUE) +
    scale_colour_manual(values = combo_palette, drop = TRUE) +
    labs(x = NULL, y = y_label) +
    theme_minimal(base_size = 12) +
    theme(
      legend.position  = "none",
      axis.text.x      = element_text(angle = 40, hjust = 1, size = 9),
      strip.text       = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
  if (log_scale) p <- p + scale_y_log10()
  p
}

p_length     <- make_descriptor_plot(gene_filter_plot, "length",           "Gene length (aa)",              log_scale = TRUE)
p_ncaas      <- make_descriptor_plot(gene_filter_plot, "n_CAAS",           "Total CAAS count",              log_scale = TRUE)
p_density    <- make_descriptor_plot(gene_filter_plot, "n_CAAS_per_length","CAAS density (per 100 aa)",     log_scale = TRUE)
p_fracsig    <- make_descriptor_plot(gene_filter_plot, "frac_is_significant","Fraction significant sites",    log_scale = FALSE)
p_medpval    <- make_descriptor_plot(gene_filter_plot, "median_pval_hyp",  "Median hyperg. p-value",        log_scale = FALSE)
p_minpval    <- make_descriptor_plot(gene_filter_plot, "min_pval_hyp",     "Min hyperg. p-value",           log_scale = FALSE)

desc_grid <- (
  p_length  | p_ncaas  | p_density
) / (
  p_fracsig | p_medpval | p_minpval
) +
  plot_annotation(
    title    = "Gene descriptor distributions by filter membership",
    subtitle = sprintf(
      "Extreme = top %.0f%% density  |  Dubious = Q3 + %.1f×IQR count  |  Cluster = cluster-filtered",
      100 * (1 - extreme_threshold), iqr_multiplier
    ),
    theme = theme(plot.title = element_text(face = "bold", size = 16),
                  plot.subtitle = element_text(size = 12))
  )

print(desc_grid)
ggsave(file.path(outliers_dir, "filter_overlap_descriptors.png"),
       desc_grid, width = 14, height = 10, dpi = 280)
```

```{r filter_overlap_summary_table, eval=(processing_mode != 'exploratory'), results='asis'}
cat("\n#### Filter overlap summary table\n\n")
# Per-combo × per-group mean ± SD for all five descriptors
overlap_summary <- gene_filter_flags %>%
  group_by(caap_group, filter_combo) %>%
  summarise(
    n_genes                  = n(),
    mean_Length              = round(mean(length,            na.rm = TRUE), 1),
    sd_Length                = round(sd(length,              na.rm = TRUE), 1),
    mean_n_CAAS              = round(mean(n_CAAS,            na.rm = TRUE), 2),
    sd_n_CAAS                = round(sd(n_CAAS,              na.rm = TRUE), 2),
    mean_density             = round(mean(n_CAAS_per_length, na.rm = TRUE), 4),
    sd_density               = round(sd(n_CAAS_per_length,   na.rm = TRUE), 4),
    mean_frac_is_significant = round(mean(frac_is_significant, na.rm = TRUE), 4),
    sd_frac_is_significant   = round(sd(frac_is_significant,   na.rm = TRUE), 4),
    mean_median_pval         = round(mean(median_pval_hyp,   na.rm = TRUE), 4),
    sd_median_pval           = round(sd(median_pval_hyp,     na.rm = TRUE), 4),
    .groups = "drop"
  ) %>%
  arrange(caap_group, filter_combo)

write_tsv(gene_filter_flags,  file.path(outliers_dir, "filter_overlap_gene_table.tsv"))
write_tsv(overlap_summary,    file.path(outliers_dir, "filter_overlap_summary.tsv"))

kable(
  overlap_summary,
  caption = "Filter overlap summary: per-combo x per-group mean +/- SD across descriptors"
) %>% kable_styling()
```

```{r removed_pattern_sig, eval=(processing_mode != 'exploratory'), results='asis'}
cat("\n## B.2 Removed positions by `convergence_type` and significance\n\n")
if (nrow(removal_events) > 0) {
  removal_events_plot <- removal_events %>%
    mutate(
      change_side = if_else(change_side %in% c("top", "bottom", "both"), change_side, "none"),
      pattern_parallel = make_pattern_parallel_label(convergence_type)
    )

  by_pattern <- removal_events_plot %>%
    count(removal_stage, change_side, pattern_parallel, sort = TRUE)
  by_sig <- removal_events %>%
    count(removal_stage, significance_class, sort = TRUE)

  kable(by_pattern, caption = "Removed positions by stage, change_side, and pattern/parallel label") %>% kable_styling()
  kable(by_sig, caption = "Removed positions by stage and descriptive significance class") %>% kable_styling()

  p_rm <- ggplot(
    by_pattern,
    aes(y = forcats::fct_rev(pattern_parallel), x = n, fill = removal_stage)
  ) +
    geom_col(position = position_dodge(width = 0.75), width = 0.7) +
    facet_wrap(~change_side, scales = "free_y", ncol = 2) +
    scale_fill_paletteer_d("ggthemes::Tableau_10") +
    scale_y_discrete(labels = format_pattern_parallel_axis) +
    theme_minimal(base_size = 16) +
    labs(title = "Removed positions by convergence label", x = "n", y = "convergence label (convergence_type)") +
    theme(
      axis.text.y = element_text(size = 10, lineheight = 0.95),
      strip.text = element_text(face = "bold")
    )
  print(p_rm)
  save_barplot(
    file.path(disambig_dir, "removed_pattern_parallel_all.png"),
    p_rm,
    n_labels = n_distinct(by_pattern$pattern_parallel),
    n_facets = n_distinct(by_pattern$change_side),
    facet_cols = 2,
    base_width = 11,
    base_height = 7.5
  )
}
```

```{r pvalue_removed_vs_retained, fig.width=12, fig.height=9, eval=(processing_mode != 'exploratory'), results='asis'}
cat("\n## B.2b P-value distributions: removed vs retained positions\n\n")
if (nrow(discovery_std) > 0) {
  removed_site_ids <- if (nrow(removal_events) > 0) unique(removal_events$site_id) else character(0)

  pval_df <- discovery_std %>%
    mutate(
      filter_status = if_else(site_id %in% removed_site_ids, "removed", "retained"),
      filter_status = factor(filter_status, levels = c("retained", "removed"))
    ) %>%
    filter(!is.na(pvalue) | !is.na(pvalue_boot)) %>%
    pivot_longer(
      cols      = c(pvalue, pvalue_boot),
      names_to  = "pval_type",
      values_to = "p"
    ) %>%
    filter(!is.na(p)) %>%
    mutate(
      pval_type = recode(pval_type,
        pvalue      = "Hypothesis p-value",
        pvalue_boot = "Permutation p-value (boot)"
      )
    )

  status_colours <- c(
    retained = tableau_10[["blue"]],
    removed = tableau_10[["vermillion"]]
  )
  threshold_scope_levels <- c("all", "<= 0.10", "<= 0.05")

  pvalue_burden_summary <- pval_df %>%
    group_by(pval_type, filter_status) %>%
    summarise(
      total = n(),
      `all` = n(),
      `<= 0.10` = sum(p <= 0.10, na.rm = TRUE),
      `<= 0.05` = sum(p <= 0.05, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    pivot_longer(cols = c("all", "<= 0.10", "<= 0.05"), names_to = "threshold_scope", values_to = "count") %>%
    mutate(
      threshold_scope = factor(threshold_scope, levels = threshold_scope_levels),
      proportion = count / total
    ) %>%
    arrange(pval_type, filter_status, threshold_scope)

  kable(
    pvalue_burden_summary %>% mutate(proportion = percent(proportion, accuracy = 0.1)),
    caption = paste0("Removed vs retained burden by p-value threshold (configured alpha = ", alpha_threshold_label, ")")
  ) %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

  write_tsv(
    pvalue_burden_summary %>% mutate(threshold_scope = as.character(threshold_scope)),
    file.path(disambig_dir, "pvalue_burden_summary.tsv")
  )

  p_burden <- ggplot(
    pvalue_burden_summary,
    aes(x = threshold_scope, y = proportion, fill = filter_status)
  ) +
    geom_col(position = position_dodge(width = 0.7), width = 0.65, colour = "black", linewidth = 0.2) +
    geom_text(
      aes(label = sprintf("%d", count)),
      position = position_dodge(width = 0.7),
      vjust = -0.25,
      size = 3
    ) +
    facet_wrap(~pval_type, ncol = 2) +
    scale_fill_manual(values = status_colours, name = "Status") +
    scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1.05)) +
    labs(
      title = "P-value burden comparison: all vs <= 0.10 vs <= 0.05",
      subtitle = paste0("Configured alpha = ", alpha_threshold_label, " | labels show counts"),
      x = "Threshold scope",
      y = "Proportion of positions"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 25, hjust = 1)
    )
  print(p_burden)
  save_barplot(
    file.path(disambig_dir, "pvalue_burden_summary.png"),
    p_burden,
    n_labels = n_distinct(pvalue_burden_summary$threshold_scope),
    n_facets = n_distinct(pvalue_burden_summary$pval_type),
    facet_cols = 2,
    dpi = 160,
    base_width = 9,
    base_height = 5
  )

  # ── Plot 1: global double-layered histogram ─────────────────────────────────
  p_glob <- ggplot(pval_df, aes(x = p, fill = filter_status)) +
    geom_histogram(
      aes(y = after_stat(density)),
      bins      = 40,
      alpha     = 0.55,
      position  = "identity",
      colour    = "black",
      linewidth = 0.25
    ) +
    geom_vline(
      data = threshold_line_df,
      aes(xintercept = threshold, colour = line_label, linetype = line_label),
      linewidth = 0.6,
      show.legend = TRUE,
      inherit.aes = FALSE
    ) +
    facet_wrap(~pval_type, scales = "free_y", ncol = 2) +
    scale_fill_manual(values = status_colours, name = "Status") +
    scale_colour_manual(
      values = setNames(threshold_line_colours[threshold_line_df$line_key], threshold_line_df$line_label),
      guide = "none"
    ) +
    scale_linetype_manual(
      values = setNames(threshold_line_types[threshold_line_df$line_key], threshold_line_df$line_label),
      name = "Threshold lines"
    ) +
    scale_x_continuous(breaks = c(0, 0.05, 0.10, 0.25, 0.5, 0.75, 1), limits = c(0, 1)) +
    labs(
      title    = "P-value distribution: removed vs retained positions",
      subtitle = sprintf(
        "%d retained · %d removed | configured alpha = %s | comparison lines = 0.10 and 0.05",
        sum(pval_df$pval_type == "Hypothesis p-value" & pval_df$filter_status == "retained"),
        sum(pval_df$pval_type == "Hypothesis p-value" & pval_df$filter_status == "removed"),
        alpha_threshold_label
      ),
      x = "p-value", y = "Density"
    ) +
    theme_minimal(base_size = 16) +
    theme(
      legend.position  = "bottom",
      strip.text       = element_text(face = "bold"),
      panel.grid.minor = element_blank(),
      axis.text.x      = element_text(angle = 60, hjust = 1, vjust = 1)
    )
  print(p_glob)
  ggsave(file.path(disambig_dir, "pvalue_removed_vs_retained_global.png"),
         p_glob, width = 12, height = 5, dpi = 160)

  # ── Plot 2: same, faceted by convergence_type ───────────────────────────────────
  pt_counts <- pval_df %>%
    filter(pval_type == "Hypothesis p-value") %>%
    count(convergence_type, filter_status) %>%
    pivot_wider(names_from = filter_status, values_from = n, values_fill = 0)

  pval_df_pt <- pval_df %>%
    filter(convergence_type %in% (
      pval_df %>%
        filter(pval_type == "Hypothesis p-value") %>%
        count(convergence_type) %>%
        filter(n >= 5) %>%
        pull(convergence_type)
    ))

  if (nrow(pval_df_pt) > 0 && n_distinct(pval_df_pt$convergence_type) >= 1) {
    p_pt <- ggplot(pval_df_pt, aes(x = p, fill = filter_status)) +
      geom_histogram(
        aes(y = after_stat(density)),
        bins      = 30,
        alpha     = 0.55,
        position  = "identity",
        colour    = "black",
        linewidth = 0.25
      ) +
      geom_vline(
        data = threshold_line_df,
        aes(xintercept = threshold, colour = line_label, linetype = line_label),
        linewidth = 0.55,
        show.legend = TRUE,
        inherit.aes = FALSE
      ) +
      facet_grid(convergence_type ~ pval_type, scales = "free_y") +
      scale_fill_manual(values = status_colours, name = "Status") +
      scale_colour_manual(
        values = setNames(threshold_line_colours[threshold_line_df$line_key], threshold_line_df$line_label),
        guide = "none"
      ) +
      scale_linetype_manual(
        values = setNames(threshold_line_types[threshold_line_df$line_key], threshold_line_df$line_label),
        name = "Threshold lines"
      ) +
      scale_x_continuous(breaks = c(0, 0.05, 0.10, 0.5, 1), limits = c(0, 1)) +
      labs(
        title = "P-value distribution by convergence_type: removed vs retained",
        subtitle = paste0("Configured alpha = ", alpha_threshold_label, " | comparison lines = 0.10 and 0.05"),
        x = "p-value", y = "Density"
      ) +
      theme_minimal(base_size = 11) +
      theme(
        legend.position  = "bottom",
        strip.text       = element_text(face = "bold", size = 9),
        panel.grid.minor = element_blank(),
        axis.text.x      = element_text(angle = 60, hjust = 1, vjust = 1)
      )
    print(p_pt)
    ggsave(file.path(disambig_dir, "pvalue_removed_vs_retained_by_pattern.png"),
           p_pt, width = 12, height = max(4, 2.5 * n_distinct(pval_df_pt$convergence_type)), dpi = 160)
  } else {
    cat("Skipping per-convergence_type p-value plot: no convergence_type with \u2265 5 observations.\n")
  }
} else {
  cat("No discovery data available — skipping p-value removal histograms.\n")
}

```{r excluded_gene_lists, eval=(processing_mode != 'exploratory'), results='asis'}
cat("\n## B.2c Excluded gene list exports\n\n")
# ── Build per-category excluded gene lists ────────────────────────────────────────────
# Six lists are exported to
#   disambiguation_characterization/gene_relation_analysis/txt/excluded/
# for manual inspection of which genes were excluded and why.
#
# Categories (mutually exclusive by design for excluded_nonsig):
#   excluded_preproc             – genes with ≥1 position removed during pre-processing
#                                  (pattern-type filter / MRCA-posterior threshold)
#   excluded_postproc_clusters   – genes with ≥1 position discarded by cluster filter
#   excluded_postproc_genefilter – genes flagged Extreme and/or Dubious (gene-level filter)
#   excluded_postproc            – union of the two postproc exclusion categories
#   excluded_nonsig              – genes that survived ALL filters but never achieved
#                                  sig_both == TRUE; does NOT overlap with preproc/postproc
#   excluded_overall             – union of all four categories

# 1. Pre-processing exclusions (from precluster_events, loaded in B.1)
excluded_preproc <- sort(unique(precluster_events$gene))

# 2. Post-processing: cluster-filter genes
excluded_postproc_clusters <- sort(unique(
  filter(gene_filter_flags, Cluster_removed)$Gene
))

# 3. Post-processing: gene-level filter (Extreme / Dubious / Both)
excluded_postproc_genefilter <- sort(unique(
  filter(gene_filter_flags, Extreme | Dubious)$Gene
))

# 4. All post-processing exclusions (union of 2 and 3)
excluded_postproc <- sort(unique(c(excluded_postproc_clusters, excluded_postproc_genefilter)))

# 5. Non-significant genes: survived all filters, but zero sig_both positions;
#    explicitly does NOT overlap with preproc / postproc categories
any_filtered_genes <- sort(unique(filter(gene_filter_flags, filter_combo != "None")$Gene))
genes_ever_sig     <- sort(unique(filter(discovery_std, is_significant)$gene))
survived_genes     <- sort(setdiff(unique(gene_filter_flags$Gene), any_filtered_genes))
excluded_nonsig    <- sort(setdiff(survived_genes, c(genes_ever_sig, excluded_preproc)))

# 6. Overall: union of all exclusion categories
excluded_overall <- sort(unique(c(excluded_preproc, excluded_postproc, excluded_nonsig)))

# ── Write to gene_relation_analysis/txt/excluded/ ────────────────────────────────────
gr_excl_dir <- file.path(disambig_dir, "gene_relation_analysis", "txt", "excluded")
createDir(gr_excl_dir)

readr::write_lines(excluded_preproc,             file.path(gr_excl_dir, "excluded_preproc.txt"))
readr::write_lines(excluded_postproc_clusters,   file.path(gr_excl_dir, "excluded_postproc_clusters.txt"))
readr::write_lines(excluded_postproc_genefilter, file.path(gr_excl_dir, "excluded_postproc_genefilter.txt"))
readr::write_lines(excluded_postproc,            file.path(gr_excl_dir, "excluded_postproc.txt"))
readr::write_lines(excluded_nonsig,              file.path(gr_excl_dir, "excluded_nonsig.txt"))
readr::write_lines(excluded_overall,             file.path(gr_excl_dir, "excluded_overall.txt"))

cat(
  "\n=== Excluded Gene Lists ===\n",
  sprintf("  excluded_preproc             : %d genes\n", length(excluded_preproc)),
  sprintf("  excluded_postproc_clusters   : %d genes\n", length(excluded_postproc_clusters)),
  sprintf("  excluded_postproc_genefilter : %d genes\n", length(excluded_postproc_genefilter)),
  sprintf("  excluded_postproc (union)    : %d genes\n", length(excluded_postproc)),
  sprintf("  excluded_nonsig              : %d genes\n", length(excluded_nonsig)),
  sprintf("  excluded_overall             : %d genes\n", length(excluded_overall)),
  sprintf("  Written to: %s\n\n", gr_excl_dir)
)
```

```{r global_pattern_sig, eval=(processing_mode != 'exploratory'), results='asis'}
cat("\n## B.3 Global side-aware convergence characterization\n\n")
pattern_parallel_by_group <- convergence_plot_df %>%
  count(change_side, caap_group, pattern_parallel, sort = TRUE)

write_tsv(pattern_parallel_by_group, file.path(disambig_dir, "pattern_parallel_by_caap_group.tsv"))

pg <- ggplot(
  pattern_parallel_by_group,
  aes(y = forcats::fct_rev(pattern_parallel), x = n, fill = caap_group)
) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  facet_wrap(~change_side, scales = "free_y", ncol = 2) +
  scale_fill_paletteer_d("ggthemes::Tableau_10") +
  scale_y_discrete(labels = format_pattern_parallel_axis) +
  theme_minimal(base_size = 14) +
  labs(title = "Convergence labels across CAAP groups", x = "n", y = "convergence label") +
  theme(
    axis.title.y = element_text(margin = ggplot2::margin(l = 20, r = 20)),
    axis.title.x = element_text(margin = ggplot2::margin(t = 12)),
    plot.title   = element_text(face = "bold", hjust = 0, margin = ggplot2::margin(b = 6)),
    axis.text.y = element_text(size = 10, lineheight = 0.95),
    strip.text = element_text(face = "bold")
  )
print(pg)
save_barplot(
  file.path(disambig_dir, "pattern_parallel_by_caap_group.png"),
  pg,
  n_labels = n_distinct(pattern_parallel_by_group$pattern_parallel),
  n_facets = n_distinct(pattern_parallel_by_group$change_side),
  facet_cols = 2,
  base_width = 11,
  base_height = 7.5
)
```

```{r pattern_sig_descriptor, eval=(processing_mode != 'exploratory'), results='asis'}
cat("\n## B.3b Convergence descriptors by significance class\n\n")
pattern_parallel_sig <- convergence_plot_df %>%
  count(change_side, significance_class, pattern_parallel, sort = TRUE)

# Long count table dropped — barplot below conveys the same; counts exported.
write_tsv(pattern_parallel_sig, file.path(disambig_dir, "pattern_parallel_significance_descriptor.tsv"))

if (nrow(pattern_parallel_sig) > 0) {
  pg_sig <- ggplot(
    pattern_parallel_sig,
    aes(y = forcats::fct_rev(pattern_parallel), x = n, fill = significance_class)
  ) +
    geom_col(position = position_dodge(width = 0.75), width = 0.7) +
    facet_wrap(~change_side, scales = "free_y", ncol = 2) +
    scale_fill_paletteer_d("ggthemes::Tableau_10") +
    scale_y_discrete(labels = format_pattern_parallel_axis) +
    theme_minimal(base_size = 14) +
    labs(title = "Descriptive significance classes across convergence labels", x = "n", y = "convergence label") +
    theme(
      axis.title.y = element_text(margin = ggplot2::margin(l = 20, r = 20)),
      axis.title.x = element_text(margin = ggplot2::margin(t = 12)),
      plot.title   = element_text(face = "bold", hjust = 0, margin = ggplot2::margin(b = 6)),
      axis.text.y = element_text(size = 10, lineheight = 0.95),
      strip.text = element_text(face = "bold")
    )
  print(pg_sig)
  save_barplot(
    file.path(disambig_dir, "pattern_parallel_significance_descriptor.png"),
    pg_sig,
    n_labels = n_distinct(pattern_parallel_sig$pattern_parallel),
    n_facets = n_distinct(pattern_parallel_sig$change_side),
    facet_cols = 2,
    base_width = 11,
    base_height = 7.5
  )
}
```

```{r manhattan_disambiguation, fig.height=8, fig.width=16, eval=(processing_mode != 'exploratory'), results='asis'}
cat("\n## B.5 Manhattan: genomic distribution of genes (CAAS density per gene length)\n\n")
if (!"chr" %in% names(gene_ensembl)) {
  cat("⚠ Chromosome information not available in gene ensembl file, skipping Manhattan plot\n")
} else {
  # Calculate retained CAAS density per gene across all CT positions.
  gene_caas_density <- discovery_std %>%
    group_by(gene) %>%
    summarise(
      n_caas_positions = n_distinct(site_id),
      frac_is_significant = mean(is_significant, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    left_join(gene_ensembl %>% 
              select(gene, chr, start, end) %>%
              mutate(gene_length = end - start + 1), 
              by = "gene") %>%
    filter(!is.na(chr), gene_length > 0) %>%
    mutate(
      caas_density = n_caas_positions / gene_length,
      log_caas_density = log10(pmax(caas_density, 1e-10))  # Avoid log10(0)
    )

  # Visualise all genes with at least one retained CT position.
  manhattan_data <- gene_caas_density %>%
    filter(caas_density > 0) %>%
    mutate(
      chr_num   = str_replace(chr, "chr", ""),
      chr_order = case_when(
        chr_num %in% as.character(1:22) ~ as.numeric(chr_num),
        chr_num == "X"                  ~ 23,
        chr_num == "Y"                  ~ 24,
        TRUE                            ~ 25
      ),
      pos = (start + end) / 2
    ) %>%
    arrange(chr_order, pos)

  if (nrow(manhattan_data) > 0) {
    # Cumulative chromosome positions and boundaries
    chr_lengths <- manhattan_data %>%
      group_by(chr_num, chr_order) %>%
      summarise(chr_max = max(pos, na.rm = TRUE), .groups = "drop") %>%
      arrange(chr_order) %>%
      mutate(
        chr_start  = lag(cumsum(chr_max), default = 0),
        chr_end    = chr_start + chr_max,
        chr_center = chr_start + chr_max / 2
      )

    manhattan_data <- manhattan_data %>%
      left_join(chr_lengths %>% select(chr_num, chr_start, chr_end), by = "chr_num") %>%
      mutate(BPcum = pos + chr_start)

    n_genes_plotted <- n_distinct(manhattan_data$gene)
    n_chr_plotted <- n_distinct(chr_lengths$chr_num)
    manhattan_width <- min(22, max(16, 10 + 0.55 * n_chr_plotted))
    manhattan_height <- min(12, max(9, 8 + 0.015 * n_genes_plotted))

    p_manhattan <- ggplot(manhattan_data,
        aes(x = BPcum, y = caas_density)) +
      # Discontinuous chromosome boundary lines
      geom_vline(
        data        = chr_lengths,
        aes(xintercept = chr_end),
        linetype    = "dotted", color = "grey60", alpha = 0.7, linewidth = 0.4,
        inherit.aes = FALSE
      ) +
      geom_point(alpha = 0.85, size = 3.1, color = "#2E86AB") +
      scale_x_continuous(
        labels = chr_lengths$chr_num,
        breaks = chr_lengths$chr_center,
        expand = expansion(mult = c(0.005, 0.005))
      ) +
      scale_y_continuous(
        labels = scales::label_number(scale_cut = cut_short_scale()),
        expand = expansion(mult = c(0.02, 0.08))
      ) +
      labs(
        title    = "Genomic distribution of retained CAAS density",
        subtitle = sprintf(
          "n = %d genes with retained CT positions | y = retained positions / gene length (bp)",
          n_genes_plotted
        ),
        x     = "Chromosome",
        y     = "CAAS density (retained positions / gene length)"
      ) +
      theme_minimal(base_size = 24) +
      theme(
        plot.title = element_text(face = "bold", size = 28),
        plot.subtitle = element_text(size = 20),
        axis.title = element_text(size = 22),
        axis.text = element_text(size = 18),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank()
      )

    print(p_manhattan)
    ggsave(
      filename = file.path(disambig_dir, "manhattan_caas_density.png"),
      plot     = p_manhattan, width = manhattan_width, height = manhattan_height, dpi = 320, dev = "png"
    )
  } else {
    cat("No genes with retained CT positions found — Manhattan plot skipped.\n")
  }
}
```

## 5. Impact Summary

```{r impact_summary}
cluster_info <- ""
if (nrow(filter_summary) > 0) {
  cluster_lines <- vapply(seq_len(nrow(filter_summary)), function(i) {
    sprintf("    minlen=%d, maxcaas=%.2f: %d positions", filter_summary$Minlen[i], filter_summary$Maxcaas[i], filter_summary$DiscardedCount[i])
  }, character(1))
  cluster_info <- paste0("  Positions flagged by cluster filtering:\n", paste(cluster_lines, collapse = "\n"), "\n")
}

cat(
  "\n=== CT Post-Processing Impact Summary ===\n\n",
  "Gene-level statistics:\n",
  sprintf("  Total genes analyzed: %d\n\n", n_distinct(discovery_df$Gene)),
  "Position-level statistics:\n",
  sprintf("  Total CT positions: %d\n", nrow(discovery_df)),
  cluster_info
)

summary_table <- tibble(
  Metric = c("Total CT Positions", "Unique Genes"),
  Value  = c(
    as.character(nrow(discovery_df)),
    as.character(n_distinct(discovery_df$Gene))
  )
)

kable(summary_table, caption = "CT Post-Processing Summary Statistics", align = c("l", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

write_tsv(summary_table, file.path(summary_stats_dir, "summary_statistics.tsv"))
```

---

## Summary

This report summarizes the results of the **CT Post-Processing** analysis.

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — CT Post-Processing Module*
*Date of generation: `r Sys.Date()`*
