---
title: "CT Post-Processing and Disambiguation Characterization"
author: "PhyloPhere Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    code_folding: hide
    df_print: paged
params:
  discovery_file: NULL
  filter_summary_file: NULL
  filter_dir: NULL
  gene_ensembl_file: NULL
  processing_mode: "exploratory"
  output_dir: "ct_postproc"
  extreme_threshold: 0.99
  iqr_multiplier: 3.0
  gene_filter_mode: "both"
  generate_manhattan: TRUE
  manhattan_min_density: 0.008
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7,
  dev = 'png',
  dpi = 320
)

suppressPackageStartupMessages({
  library(stringr)
  library(purrr)
  library(tibble)
  library(patchwork)
  library(scales)
  library(ggplot2)
  library(forcats)
  library(ggvenn)
  library(knitr)
  library(kableExtra)
  library(paletteer)
  library(tidyverse)
})

createDir <- function(directory) {
  if (!dir.exists(directory)) {
    dir.create(directory, recursive = TRUE, showWarnings = FALSE)
  }
}

find_first_existing <- function(candidates) {
  hits <- candidates[file.exists(candidates)]
  if (length(hits) == 0) return(NA_character_)
  hits[[1]]
}

find_recursive_file <- function(root_dir, pattern) {
  if (is.null(root_dir) || is.na(root_dir) || !dir.exists(root_dir)) return(NA_character_)
  files <- list.files(root_dir, pattern = pattern, full.names = TRUE, recursive = TRUE)
  if (length(files) == 0) return(NA_character_)
  files[[1]]
}

as_bool <- function(x) {
  tolower(as.character(x)) %in% c("true", "t", "1", "yes", "y")
}

normalize_disambig_schema <- function(df) {
  if ('Pattern' %in% names(df) && !'pattern_type' %in% names(df)) df <- df %>% rename(pattern_type = Pattern)
  if ('Pvalue' %in% names(df) && !'pvalue' %in% names(df)) df <- df %>% rename(pvalue = Pvalue)
  if ('Pvalue.boot' %in% names(df) && !'pvalue_boot' %in% names(df)) df <- df %>% rename(pvalue_boot = `Pvalue.boot`)
  if ('ChangeSide' %in% names(df) && !'change_side' %in% names(df)) df <- df %>% rename(change_side = ChangeSide)
  if ('Gene' %in% names(df) && !'gene' %in% names(df)) df <- df %>% rename(gene = Gene)
  if ('Position' %in% names(df) && !'position' %in% names(df)) df <- df %>% rename(position = Position)
  if ('CAAP_Group' %in% names(df) && !'caap_group' %in% names(df)) df <- df %>% rename(caap_group = CAAP_Group)
  if ('IsConserved' %in% names(df) && !'is_conserved_meta' %in% names(df)) df <- df %>% rename(is_conserved_meta = IsConserved)
  if ('ASR_ConsAntiCAAS' %in% names(df) && !'asr_is_conserved' %in% names(df)) df <- df %>% rename(asr_is_conserved = ASR_ConsAntiCAAS)
  if (!'caap_group' %in% names(df)) df$caap_group <- 'US'
  if (!'change_side' %in% names(df)) df$change_side <- 'none'
  if (!'is_conserved_meta' %in% names(df)) df$is_conserved_meta <- FALSE
  if (!'asr_is_conserved' %in% names(df)) df$asr_is_conserved <- FALSE

  df %>%
    mutate(
      gene = as.character(gene),
      position = suppressWarnings(as.integer(position)),
      caap_group = as.character(caap_group),
      pattern_type = as.character(pattern_type),
      change_side = tolower(as.character(change_side)),
      is_conserved_meta = as_bool(is_conserved_meta),
      asr_is_conserved = as_bool(asr_is_conserved),
      pvalue = suppressWarnings(as.numeric(pvalue)),
      pvalue_boot = suppressWarnings(as.numeric(pvalue_boot)),
      sig_hyp = !is.na(pvalue) & pvalue <= 0.05,
      sig_perm = !is.na(pvalue_boot) & pvalue_boot <= 0.05,
      sig_both = sig_hyp & sig_perm,
      significance_class = case_when(
        sig_both ~ 'both',
        sig_hyp ~ 'hyper_only',
        sig_perm ~ 'perm_only',
        TRUE ~ 'none'
      ),
      site_id = paste(gene, position, sep = '@')
    )
}

seed_val <- 12345
set.seed(seed_val)

discovery_file <- params$discovery_file
filter_summary_file <- params$filter_summary_file
gene_ensembl_file <- params$gene_ensembl_file
processing_mode <- params$processing_mode
extreme_threshold <- params$extreme_threshold
iqr_multiplier <- params$iqr_multiplier
gene_filter_mode <- params$gene_filter_mode
generate_manhattan <- params$generate_manhattan
manhattan_min_density <- params$manhattan_min_density

if (is.null(discovery_file)) stop("Error: discovery_file parameter is required")
if (is.null(filter_summary_file)) stop("Error: filter_summary_file parameter is required")
if (is.null(gene_ensembl_file)) stop("Error: gene_ensembl_file parameter is required")

stopifnot(file.exists(discovery_file), file.exists(filter_summary_file), file.exists(gene_ensembl_file))

ct_postproc_dir <- "."
outliers_dir <- "outliers"
cluster_dir <- "clusters"
summary_stats_dir <- "summary_statistics"
disambig_dir <- "disambiguation_characterization"

createDir(ct_postproc_dir)
createDir(outliers_dir)
createDir(cluster_dir)
createDir(summary_stats_dir)
createDir(disambig_dir)

tableau_10 <- paletteer_d("ggthemes::Tableau_10")
names(tableau_10) <- c(
  "blue", "orange", "vermillion", "bluish_green",
  "green", "yellow", "reddish_purple", "pink",
  "brown", "gray"
)

category_palette <- c(
  "Normal" = tableau_10[["gray"]],
  "Dubious" = tableau_10[["blue"]],
  "Extreme" = tableau_10[["vermillion"]],
  "Both" = tableau_10[["reddish_purple"]]
)

removal_palette <- c(
  Discarded = tableau_10[["vermillion"]],
  Good = tableau_10[["green"]]
)
```

# A) Post-processing characterization

## Data Loading and Gene Length Annotation

```{r data_loading}
cat("=== CT Post-Processing Configuration ===\n")
cat("Processing mode:", processing_mode, "\n")
cat("Gene filtering:", gene_filter_mode, "\n")
cat("Extreme threshold:", extreme_threshold, paste0("(top ", 100*(1-extreme_threshold), "%)"), "\n")
cat("IQR multiplier:", iqr_multiplier, "\n")
cat("Generate Manhattan plots:", generate_manhattan, "\n")
if (generate_manhattan) cat("Manhattan min density:", manhattan_min_density, "\n")

cat("\nInput files:\n")
cat("  Discovery:", discovery_file, "\n")
cat("  Filter summary:", filter_summary_file, "\n")
cat("  Gene ensembl:", gene_ensembl_file, "\n\n")

raw_discovery <- read_delim(discovery_file, delim = "\t", show_col_types = FALSE)
discovery_std <- normalize_disambig_schema(raw_discovery)

discovery_df <- discovery_std %>%
  transmute(
    CAAP_Group = caap_group,
    Gene = gene,
    Position = position,
    Pattern = pattern_type,
    is_conserved_meta,
    asr_is_conserved,
    Pvalue = pvalue,
    `Pvalue.boot` = pvalue_boot,
    sig_hyp,
    sig_perm,
    sig_both,
    significance_class,
    site_id
  )

if (!"CAAP_Group" %in% names(discovery_df)) discovery_df <- discovery_df %>% mutate(CAAP_Group = "US")

cat("  Loaded CT positions:", nrow(discovery_df), "\n")
cat("  Unique genes:", n_distinct(discovery_df$Gene), "\n")

gene_ensembl <- read_tsv(gene_ensembl_file, show_col_types = FALSE)
names(gene_ensembl) <- tolower(names(gene_ensembl))
if (!"gene" %in% names(gene_ensembl) || !"length" %in% names(gene_ensembl)) {
  stop("Gene ensembl file must have 'gene' and 'length' columns")
}

gene_length <- gene_ensembl %>% select(Gene = gene, Length = length)
discovery_df <- discovery_df %>% left_join(gene_length, by = "Gene") %>% filter(!is.na(Length))

cat("  Genes with length annotation:", n_distinct(discovery_df$Gene), "\n\n")
```

## Conserved metadata composition after precluster pruning

```{r conserved_meta_composition}
conserved_meta_comp <- discovery_df %>%
  mutate(is_conserved_meta = if_else(is_conserved_meta, "TRUE", "FALSE")) %>%
  count(is_conserved_meta, name = "n") %>%
  mutate(prop = n / sum(n))

kable(
  conserved_meta_comp,
  caption = "is_conserved_meta composition after precluster pruning"
) %>% kable_styling()

p_conserved_meta_comp <- ggplot(conserved_meta_comp, aes(x = is_conserved_meta, y = prop, fill = is_conserved_meta)) +
  geom_col(color = "black", linewidth = 0.25) +
  geom_text(aes(label = scales::percent(prop, accuracy = 0.1)), vjust = -0.3, size = 4) +
  scale_fill_paletteer_d("ggthemes::Tableau_10") +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title = "Proportion of is_conserved_meta after pruning",
    x = "is_conserved_meta",
    y = "Proportion of CAAS"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

print(p_conserved_meta_comp)
ggsave(file.path(disambig_dir, "is_conserved_meta_proportion_after_pruning.png"), p_conserved_meta_comp, width = 8, height = 5, dpi = 320)
```

## Outlier Gene Detection

```{r outlier_detection, fig.height=10, fig.width=14}
has_caap_group <- TRUE
cat("Unified grouping mode enabled (CAAP_Group always present)\n")
cat("Groups found:", paste(sort(unique(discovery_df$CAAP_Group)), collapse=", "), "\n")

gene_summary <- discovery_df %>%
  group_by(CAAP_Group, Gene) %>%
  summarise(
    n_CAAS = n(),
    Length = first(Length),
    n_CAAS_per_length = n() / first(Length) * 100,
    .groups = 'drop'
  )

iqr_thresholds <- gene_summary %>%
  group_by(CAAP_Group) %>%
  summarise(
    Q1 = quantile(n_CAAS, 0.25, na.rm = TRUE),
    Q3 = quantile(n_CAAS, 0.75, na.rm = TRUE),
    IQR_val = Q3 - Q1,
    threshold_dubious = Q3 + iqr_multiplier * IQR_val,
    .groups = 'drop'
  )

extreme_thresholds <- gene_summary %>%
  group_by(CAAP_Group) %>%
  summarise(
    threshold_extreme = quantile(n_CAAS_per_length, extreme_threshold, na.rm = TRUE),
    .groups = 'drop'
  )

gene_summary_annotated <- gene_summary %>%
  left_join(extreme_thresholds, by = 'CAAP_Group') %>%
  left_join(iqr_thresholds, by = 'CAAP_Group') %>%
  mutate(
    Extreme = n_CAAS_per_length >= threshold_extreme,
    Dubious = n_CAAS >= threshold_dubious,
    Category = case_when(
      Extreme & Dubious ~ "Both",
      Extreme ~ "Extreme",
      Dubious ~ "Dubious",
      TRUE ~ "Normal"
    )
  ) %>%
  arrange(desc(n_CAAS))

p_gross_dist <- ggplot(gene_summary_annotated, aes(x = n_CAAS, fill = Category)) +
  geom_histogram(binwidth = 0.05, color = "black", alpha = 0.8) +
  geom_vline(
    data = gene_summary_annotated %>% filter(!is.na(threshold_dubious)) %>% distinct(CAAP_Group, threshold_dubious),
    aes(xintercept = threshold_dubious),
    color = tableau_10[["blue"]], linetype = "dashed", linewidth = 1
  ) +
  scale_fill_manual(values = category_palette, breaks = c("Extreme", "Dubious", "Both", "Normal")) +
  scale_x_log10(breaks = c(1, 2, 5, 10, 20, 50, 100, 200)) +
  facet_wrap(~CAAP_Group, scales = "free_y") +
  labs(title = "CAAP Count Distribution by Group",
       subtitle = "Dubious threshold (Q3 + 3×IQR) shown as dashed line",
       x = "Number of CAAP Positions", y = "Gene Count") +
  theme_minimal() +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold", size = 14), axis.title = element_text(size = 12))

p_zoom_gross <- p_gross_dist + coord_cartesian(ylim = c(0, 100)) + labs(title = "Zoomed View (y ≤ 100)")

p_density_dist <- ggplot(gene_summary_annotated, aes(x = n_CAAS_per_length, fill = Category)) +
  geom_histogram(binwidth = 0.05, color = "black", alpha = 0.8) +
  geom_vline(
    data = gene_summary_annotated %>% filter(!is.na(threshold_extreme)) %>% distinct(CAAP_Group, threshold_extreme),
    aes(xintercept = threshold_extreme),
    color = tableau_10[["vermillion"]], linetype = "dashed", linewidth = 1
  ) +
  scale_fill_manual(values = category_palette, breaks = c("Extreme", "Dubious", "Both", "Normal")) +
  scale_x_log10(breaks = c(0.01, 0.02, 0.05, 0.1, 0.2, 0.5, 1, 2, 5, 10, 20, 50, 100)) +
  facet_wrap(~ CAAP_Group, scales = "free_y") +
  labs(title = "CAAP Density Distribution by Group",
       subtitle = paste0("Top ", 100*(1-extreme_threshold), "% density threshold shown as dashed line"),
       x = "CAAP Density (positions / gene length)", y = "Gene Count") +
  theme_minimal() +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold", size = 14), axis.title = element_text(size = 12))

p_zoom_density <- p_density_dist + coord_cartesian(ylim = c(0, 50)) + labs(title = "Zoomed View (y ≤ 50)")

combined_outlier_plot <- p_gross_dist / p_density_dist
combined_zoomed_plot <- p_zoom_gross / p_zoom_density

print(combined_outlier_plot)
print(combined_zoomed_plot)

ggsave(file.path(outliers_dir, "outlier_distributions.png"), combined_outlier_plot, width = 14, height = 10, dpi = 320)
ggsave(file.path(outliers_dir, "outlier_distributions_zoomed.png"), combined_zoomed_plot, width = 14, height = 10, dpi = 320)

# Dedicated bar chart: removed gene counts per category and CAAP_Group
# Clearly separates Extreme-only, Dubious-only, and Both (genes failing both criteria).
removed_cat_summary <- gene_summary_annotated %>%
  filter(Category != "Normal") %>%
  count(CAAP_Group, Category)

if (nrow(removed_cat_summary) > 0) {
  p_removed_cats <- ggplot(removed_cat_summary,
                           aes(x = Category, y = n, fill = Category)) +
    geom_col(color = "black", linewidth = 0.3, width = 0.7) +
    facet_wrap(~ CAAP_Group, scales = "free_y") +
    scale_fill_manual(
      values = category_palette,
      breaks = c("Extreme", "Dubious", "Both")
    ) +
    labs(
      title    = "Removed Genes by Category and Group",
      subtitle = paste0(
        "Extreme: top ", 100 * (1 - extreme_threshold), "% density  |  ",
        "Dubious: Q3 + ", iqr_multiplier, "\u00d7IQR count  |  ",
        "Both: both criteria"
      ),
      x = "Removal Category",
      y = "Gene Count"
    ) +
    theme_minimal() +
    theme(
      legend.position  = "none",
      plot.title       = element_text(face = "bold", size = 14),
      plot.subtitle    = element_text(size = 10),
      axis.title       = element_text(size = 12),
      strip.text       = element_text(face = "bold")
    )

  print(p_removed_cats)
  ggsave(file.path(outliers_dir, "removed_genes_by_category.png"),
         p_removed_cats, width = 10, height = 6, dpi = 320)
}

cat("\n=== Outlier Detection Summary ===\n")
cat("CAAP/US unified mode: Group-aware thresholds calculated\n\n")

thresholds_summary <- gene_summary_annotated %>% distinct(CAAP_Group, threshold_extreme, threshold_dubious) %>% arrange(CAAP_Group)
cat("Thresholds by group:\n")
for (i in 1:nrow(thresholds_summary)) {
  cat(sprintf("  %s:\n", thresholds_summary$CAAP_Group[i]))
  cat(sprintf("    Extreme (top %.0f%% density): %.4f\n", 100*(1-extreme_threshold), thresholds_summary$threshold_extreme[i]))
  cat(sprintf("    Dubious (Q3 + %.1f×IQR): %.0f positions\n\n", iqr_multiplier, thresholds_summary$threshold_dubious[i]))
}

outlier_counts <- gene_summary_annotated %>% count(CAAP_Group, Category) %>% pivot_wider(names_from = Category, values_from = n, values_fill = 0)
cat("Gene counts by group and category:\n")
print(outlier_counts)

write_tsv(gene_summary_annotated, file.path(outliers_dir, "gene_summary_annotated.tsv"))
cat("\n✓ Gene summary exported\n\n")
```

## Cluster Filtering Analysis

```{r cluster_analysis_load}
cat("Loading cluster filter summary...\n")
filter_summary <- read_tsv(filter_summary_file, show_col_types = FALSE)
cat("  Parameter combinations tested:", nrow(filter_summary), "\n")

if (!"Minlen" %in% names(filter_summary) || !"Maxcaas" %in% names(filter_summary)) {
  if ("Parameter" %in% names(filter_summary)) {
    parsed_params <- stringr::str_split_fixed(as.character(filter_summary$Parameter), "/", 2)
    if (!"Minlen" %in% names(filter_summary)) filter_summary$Minlen <- suppressWarnings(as.integer(parsed_params[, 1]))
    if (!"Maxcaas" %in% names(filter_summary)) filter_summary$Maxcaas <- suppressWarnings(as.numeric(parsed_params[, 2]))
    cat("  Parsed missing Minlen/Maxcaas from Parameter column\n")
  }
}

if (!"DiscardedCount" %in% names(filter_summary)) {
  protected_cols <- c("Parameter", "Minlen", "Maxcaas", "CAAP_Groups", "Groups_List", "N_Groups")
  candidate_cols <- setdiff(names(filter_summary), protected_cols)
  numeric_candidate_cols <- candidate_cols[vapply(candidate_cols, function(col) is.numeric(filter_summary[[col]]), logical(1))]
  if (length(numeric_candidate_cols) > 0) {
    filter_summary <- filter_summary %>% mutate(DiscardedCount = rowSums(across(all_of(numeric_candidate_cols)), na.rm = TRUE))
    cat("  Built DiscardedCount from columns:", paste(numeric_candidate_cols, collapse = ", "), "\n")
  } else {
    filter_summary$DiscardedCount <- 0L
    cat("  Warning: Could not infer DiscardedCount from summary columns; defaulting to 0\n")
  }
} else {
  filter_summary$DiscardedCount <- suppressWarnings(as.numeric(filter_summary$DiscardedCount))
  filter_summary$DiscardedCount[is.na(filter_summary$DiscardedCount)] <- 0
}

has_caap_groups_in_summary <- 'CAAP_Groups' %in% names(filter_summary) && any(filter_summary$CAAP_Groups != 'N/A' & filter_summary$CAAP_Groups != '')
if (has_caap_groups_in_summary) {
  cat("  CAAP mode detected in filter summary\n")
  filter_summary <- filter_summary %>% mutate(Groups_List = strsplit(as.character(CAAP_Groups), ","), N_Groups = lengths(Groups_List))
}

if (has_caap_groups_in_summary) {
  group_cols <- setdiff(names(filter_summary), c('Parameter', 'Minlen', 'Maxcaas'))
  filter_table <- filter_summary %>% select(Parameter, all_of(sort(group_cols)))
} else {
  filter_table <- filter_summary %>% select(Parameter, DiscardedCount)
  names(filter_table)[2] <- "Discarded"
}

kable(filter_table, caption = "Cluster Filtering Parameter Sweep Results", align = c("l", rep("r", ncol(filter_table) - 1))) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

if (nrow(filter_summary) > 1) {
  if (has_caap_groups_in_summary) {
    group_cols <- setdiff(names(filter_summary), c('Parameter', 'Minlen', 'Maxcaas'))
    plot_data <- filter_summary %>%
      pivot_longer(cols = all_of(group_cols), names_to = "CAAP_Group", values_to = "DiscardedCount") %>%
      mutate(param_label = sprintf("%d/%d", Minlen, as.integer(Maxcaas)), param_label = factor(param_label, levels = unique(param_label)))

    p_param_sweep <- ggplot(plot_data, aes(x = param_label, y = DiscardedCount, fill = CAAP_Group)) +
      geom_col(position = "dodge", color = "black", linewidth = 0.3) +
      scale_fill_paletteer_d("ggthemes::Tableau_10") +
      labs(title = "Discarded CAAS Positions by Parameter Combination",
           subtitle = "Grouped by CAAP group - minlen/maxcaas combinations",
           x = "Parameter Combination (minlen/maxcaas)", y = "Positions Discarded", fill = "CAAP Group") +
      theme_minimal() +
      theme(legend.position = "right", plot.title = element_text(face = "bold", size = 14), axis.title = element_text(size = 12), axis.text.x = element_text(angle = 45, hjust = 1))
  } else {
    p_param_sweep <- ggplot(filter_summary, aes(x = factor(Minlen), y = DiscardedCount, color = factor(Maxcaas), group = factor(Maxcaas))) +
      geom_point(size = 4) +
      geom_line(linewidth = 1.2) +
      scale_color_paletteer_d("ggthemes::Tableau_10", labels = function(x) sprintf("%.0f%%", as.numeric(x))) +
      labs(title = "Discarded CAAS Positions by Parameter Combination",
           subtitle = "Effect of minimum cluster length and maximum density threshold",
           x = "Minimum Cluster Length (minlen)", y = "Positions Discarded", color = "Max Density") +
      theme_minimal() +
      theme(legend.position = "right", plot.title = element_text(face = "bold", size = 14), axis.title = element_text(size = 12))
  }

  print(p_param_sweep)
  ggsave(file.path(cluster_dir, "parameter_sweep_overview.png"), p_param_sweep, width = 12, height = 6, dpi = 320)
}

write_tsv(filter_summary, file.path(cluster_dir, "filter_summary.tsv"))
cat("✓ Filter summary exported\n\n")
```

```{r cluster_detailed_analysis, results='asis', fig.height=8, fig.width=12, eval=(processing_mode == 'exploratory' && nrow(filter_summary) > 1)}
if (processing_mode == 'exploratory' && nrow(filter_summary) > 1) {
  cat("\n## Detailed Parameter Analysis\n\n")

  available_filtered_files <- list.files(
    path = getwd(),
    pattern = "^discovery\\.filtered\\.minlen[0-9]+\\.maxcaas[0-9]+\\.tsv$",
    recursive = TRUE,
    full.names = TRUE
  )

  for (i in 1:nrow(filter_summary)) {
    param_row <- filter_summary[i, ]
    param_label <- sprintf("minlen=%d, maxcaas=%.2f", param_row$Minlen, param_row$Maxcaas)
    cat(sprintf("\n### Parameter Set %d: %s {.tabset}\n\n", i, param_label))

    pattern_str <- sprintf("discovery.filtered.minlen%d.maxcaas%d.tsv", param_row$Minlen, as.integer(param_row$Maxcaas))
    matching_file <- available_filtered_files[basename(available_filtered_files) == pattern_str]

    if (length(matching_file) > 0 && file.exists(matching_file[1])) {
      filtered_data <- read_tsv(matching_file[1], show_col_types = FALSE)
      if ("ClusteringFlag" %in% names(filtered_data)) {
        discarded <- filtered_data %>% filter(ClusteringFlag == "Discarded")
        if (nrow(discarded) > 0 && "CAAP_Group" %in% names(discarded)) {
          detect_runs <- function(positions) {
            positions <- sort(unique(positions))
            if (length(positions) == 0) return(integer(0))
            breaks <- c(0, which(diff(positions) > 1), length(positions))
            diff(breaks)
          }

          spatial_analysis <- discarded %>%
            group_by(CAAP_Group, Gene) %>%
            summarise(positions = list(Position), .groups = "drop") %>%
            mutate(run_lengths = map(positions, detect_runs)) %>%
            unnest(run_lengths, keep_empty = TRUE) %>%
            mutate(
              cluster_type = case_when(
                run_lengths == 1 ~ "Singlet (1)",
                run_lengths == 2 ~ "Doublet (2)",
                run_lengths == 3 ~ "Triplet (3)",
                run_lengths == 4 ~ "Quad (4)",
                run_lengths == 5 ~ "Quint (5)",
                run_lengths >= 6 ~ "6+ Long",
                TRUE ~ "Other"
              )
            )

          spatial_summary <- spatial_analysis %>%
            count(CAAP_Group, cluster_type) %>%
            complete(CAAP_Group, cluster_type, fill = list(n = 0)) %>%
            group_by(CAAP_Group) %>%
            mutate(percentage = round(100 * n / sum(n), 1)) %>%
            ungroup()

          p_spatial <- ggplot(spatial_summary, aes(x = CAAP_Group, y = percentage/100, fill = cluster_type)) +
            geom_col(position = "fill", color = "black", size = 0.2) +
            scale_fill_paletteer_d("ggthemes::Tableau_10") +
            scale_y_continuous(labels = scales::percent) +
            labs(title = sprintf("Spatial Clustering: %s", param_label), x = "Group", y = "Percentage", fill = "Cluster Type") +
            theme_minimal(base_size = 12) +
            theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1))

          print(p_spatial)
        }
      }
    }
    cat("\n---\n\n")
  }
}
```

# B) Characterizing CAAS disambiguation

## B.1 Build unified removal + retention dataset

```{r disambig_unified_data}
filter_root <- params$filter_dir
postproc_root <- if (!is.null(filter_root) && !is.na(filter_root)) dirname(filter_root) else NA_character_

removed_precluster_file <- find_first_existing(c(
  file.path(filter_root, "preprocessed", "removed_patterns_precluster.tsv"),
  file.path(postproc_root, "preprocessed", "removed_patterns_precluster.tsv"),
  find_recursive_file(getwd(), "removed_patterns_precluster\\.tsv$")
))

removed_genes_file <- find_first_existing(c(
  file.path(filter_root, "gene_filtering", "removed_genes_summary.tsv"),
  file.path(postproc_root, "gene_filtering", "removed_genes_summary.tsv"),
  find_recursive_file(getwd(), "removed_genes_summary\\.tsv$")
))

cluster_files <- list.files(path = params$filter_dir, pattern = "\\.filtered\\.minlen[0-9]+\\.maxcaas[0-9]+\\.tsv$", full.names = TRUE, recursive = TRUE)
if (length(cluster_files) == 0) {
  cluster_files <- list.files(path = ifelse(is.na(postproc_root), getwd(), postproc_root), pattern = "\\.filtered\\.minlen[0-9]+\\.maxcaas[0-9]+\\.tsv$", full.names = TRUE, recursive = TRUE)
}

precluster_events <- tibble()
if (!is.na(removed_precluster_file)) {
  precluster_df <- normalize_disambig_schema(read_delim(removed_precluster_file, delim = "\t", show_col_types = FALSE))
  precluster_events <- precluster_df %>% transmute(site_id, gene, position, caap_group, pattern_type, pvalue, pvalue_boot, sig_hyp, sig_perm, sig_both, significance_class, RemovalStage = "precluster_pattern")
}

cluster_events <- tibble()
if (length(cluster_files) > 0) {
  cluster_df <- map_dfr(cluster_files, ~ read_tsv(.x, show_col_types = FALSE))
  cluster_hits <- cluster_df %>% filter(ClusteringFlag == "Discarded") %>% transmute(gene = as.character(Gene), position = as.integer(Position), caap_group = as.character(CAAP_Group)) %>% distinct()
  cluster_events <- discovery_std %>% inner_join(cluster_hits, by = c("gene", "position", "caap_group")) %>% transmute(site_id, gene, position, caap_group, pattern_type, pvalue, pvalue_boot, sig_hyp, sig_perm, sig_both, significance_class, RemovalStage = "cluster") %>% distinct()
}

gene_events <- tibble()
if (!is.na(removed_genes_file)) {
  rg <- read_tsv(removed_genes_file, show_col_types = FALSE)
  names(rg) <- tolower(names(rg))
  if (!"caap_group" %in% names(rg)) rg$caap_group <- "US"
  rg_norm <- rg %>% transmute(gene = as.character(gene), caap_group = as.character(caap_group), category = as.character(category)) %>% distinct()

  gene_events <- discovery_std %>%
    inner_join(rg_norm, by = c("gene", "caap_group")) %>%
    mutate(RemovalStage = case_when(
      tolower(category) == "extreme" ~ "extreme_gene",
      tolower(category) == "dubious" ~ "dubious_with_clusters",
      tolower(category) == "both" ~ "extreme_and_dubious",
      TRUE ~ "gene_filter"
    )) %>%
    transmute(site_id, gene, position, caap_group, pattern_type, pvalue, pvalue_boot, sig_hyp, sig_perm, sig_both, significance_class, RemovalStage) %>%
    distinct()
}

removal_events <- bind_rows(precluster_events, cluster_events, gene_events)
write_tsv(removal_events, file.path(disambig_dir, "removal_events.tsv"))
```

## B.2 Removed positions by `pattern_type` and significance

```{r removed_pattern_sig}
if (nrow(removal_events) > 0) {
  by_pattern <- removal_events %>% count(RemovalStage, pattern_type, sort = TRUE)
  by_sig <- removal_events %>%
    mutate(sig_label = case_when(sig_both ~ "both<=0.05", sig_hyp ~ "hyp<=0.05", sig_perm ~ "perm<=0.05", TRUE ~ "not_sig")) %>%
    count(RemovalStage, sig_label, sort = TRUE)

  kable(by_pattern, caption = "Removed positions by stage and pattern_type") %>% kable_styling()
  kable(by_sig, caption = "Removed positions by stage and significance") %>% kable_styling()

  p_rm <- ggplot(by_pattern, aes(pattern_type, n, fill = RemovalStage)) +
    geom_col(position = "dodge") +
    scale_fill_paletteer_d("ggthemes::Tableau_10") +
    theme_minimal() +
    labs(title = "Removed positions: pattern_type composition", x = "pattern_type") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  print(p_rm)
  ggsave(file.path(disambig_dir, "removed_pattern_type.png"), p_rm, width = 10, height = 5)
}
```

## B.2b P-value distributions: removed vs retained positions

```{r pvalue_removed_vs_retained, fig.width=12, fig.height=9}
# Build a status-annotated dataset:
# every position from discovery_std gets "removed" if it appears in removal_events,
# "retained" otherwise.  We keep the most severe removal stage per site to avoid
# duplicates when a position was caught by more than one filter.
if (nrow(discovery_std) > 0) {
  removed_site_ids <- if (nrow(removal_events) > 0) unique(removal_events$site_id) else character(0)

  pval_df <- discovery_std %>%
    mutate(
      filter_status = if_else(site_id %in% removed_site_ids, "removed", "retained"),
      filter_status = factor(filter_status, levels = c("retained", "removed"))
    ) %>%
    filter(!is.na(pvalue) | !is.na(pvalue_boot)) %>%
    pivot_longer(
      cols      = c(pvalue, pvalue_boot),
      names_to  = "pval_type",
      values_to = "p"
    ) %>%
    filter(!is.na(p)) %>%
    mutate(
      pval_type = recode(pval_type,
        pvalue      = "Hypothesis p-value",
        pvalue_boot = "Permutation p-value (boot)"
      )
    )

  status_colours <- c(retained = "#2196F3", removed = "#F44336")

  # ── Plot 1: global double-layered histogram ─────────────────────────────────
  p_glob <- ggplot(pval_df, aes(x = p, fill = filter_status, colour = filter_status)) +
    geom_histogram(
      aes(y = after_stat(density)),
      bins      = 40,
      alpha     = 0.55,
      position  = "identity",
      linewidth = 0.25
    ) +
    geom_vline(xintercept = 0.05, linetype = "dashed", colour = "grey30", linewidth = 0.4) +
    facet_wrap(~pval_type, scales = "free_y", ncol = 2) +
    scale_fill_manual(values = status_colours, name = "Status") +
    scale_colour_manual(values = status_colours, name = "Status") +
    scale_x_continuous(breaks = c(0, 0.05, 0.25, 0.5, 0.75, 1), limits = c(0, 1)) +
    labs(
      title    = "P-value distribution: removed vs retained positions (all)",
      subtitle = sprintf(
        "%d retained · %d removed | dashed line = 0.05",
        sum(pval_df$pval_type == "Hypothesis p-value" & pval_df$filter_status == "retained"),
        sum(pval_df$pval_type == "Hypothesis p-value" & pval_df$filter_status == "removed")
      ),
      x = "p-value", y = "Density"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      legend.position  = "bottom",
      strip.text       = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
  print(p_glob)
  ggsave(file.path(disambig_dir, "pvalue_removed_vs_retained_global.png"),
         p_glob, width = 12, height = 5, dpi = 160)

  # ── Plot 2: same, faceted by pattern_type ───────────────────────────────────
  pt_counts <- pval_df %>%
    filter(pval_type == "Hypothesis p-value") %>%
    count(pattern_type, filter_status) %>%
    pivot_wider(names_from = filter_status, values_from = n, values_fill = 0)

  pval_df_pt <- pval_df %>%
    filter(pattern_type %in% (
      pval_df %>%
        filter(pval_type == "Hypothesis p-value") %>%
        count(pattern_type) %>%
        filter(n >= 5) %>%
        pull(pattern_type)
    ))

  if (nrow(pval_df_pt) > 0 && n_distinct(pval_df_pt$pattern_type) >= 1) {
    p_pt <- ggplot(pval_df_pt, aes(x = p, fill = filter_status, colour = filter_status)) +
      geom_histogram(
        aes(y = after_stat(density)),
        bins      = 30,
        alpha     = 0.55,
        position  = "identity",
        linewidth = 0.25
      ) +
      geom_vline(xintercept = 0.05, linetype = "dashed", colour = "grey30", linewidth = 0.4) +
      facet_grid(pattern_type ~ pval_type, scales = "free_y") +
      scale_fill_manual(values = status_colours, name = "Status") +
      scale_colour_manual(values = status_colours, name = "Status") +
      scale_x_continuous(breaks = c(0, 0.05, 0.5, 1), limits = c(0, 1)) +
      labs(
        title = "P-value distribution by pattern_type: removed vs retained",
        x = "p-value", y = "Density"
      ) +
      theme_minimal(base_size = 11) +
      theme(
        legend.position  = "bottom",
        strip.text       = element_text(face = "bold", size = 9),
        panel.grid.minor = element_blank(),
        axis.text.x      = element_text(angle = 45, hjust = 1)
      )
    print(p_pt)
    ggsave(file.path(disambig_dir, "pvalue_removed_vs_retained_by_pattern.png"),
           p_pt, width = 12, height = max(4, 2.5 * n_distinct(pval_df_pt$pattern_type)), dpi = 160)
  } else {
    cat("Skipping per-pattern_type p-value plot: no pattern_type with \u2265 5 observations.\n")
  }
} else {
  cat("No discovery data available — skipping p-value removal histograms.\n")
}
```

## B.3 Global change-pattern and significance characterization

```{r global_pattern_sig}
pattern_sig <- discovery_std %>%
  group_by(pattern_type) %>%
  summarise(n = n(), sig_hyp = sum(sig_hyp, na.rm = TRUE), sig_perm = sum(sig_perm, na.rm = TRUE), sig_both = sum(sig_both, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(n))

pattern_by_group <- discovery_std %>% count(caap_group, pattern_type, sort = TRUE)

kable(pattern_sig, caption = "pattern_type counts and significance (all three criteria)") %>% kable_styling()
kable(pattern_by_group, caption = "pattern_type by CAAP group") %>% kable_styling()

pg <- ggplot(pattern_by_group, aes(pattern_type, n, fill = caap_group)) +
  geom_col(position = "dodge") +
  scale_fill_paletteer_d("ggthemes::Tableau_10") +
  theme_minimal() +
  labs(title = "pattern_type across CAAP groups", x = "pattern_type", y = "n") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(pg)
```

## B.4 Convergent overlap across groups and genes

```{r convergent_overlap}
conv_sites <- discovery_std %>% filter(pattern_type == "convergent") %>% distinct(site_id, gene, caap_group)

if (nrow(conv_sites) > 0) {
  overlap_site_pairs <- expand_grid(g1 = sort(unique(conv_sites$caap_group)), g2 = sort(unique(conv_sites$caap_group))) %>%
    rowwise() %>%
    mutate(jaccard = {
      s1 <- conv_sites %>% filter(caap_group == g1) %>% pull(site_id) %>% unique()
      s2 <- conv_sites %>% filter(caap_group == g2) %>% pull(site_id) %>% unique()
      if (length(union(s1, s2)) == 0) 0 else length(intersect(s1, s2)) / length(union(s1, s2))
    }) %>% ungroup()

  overlap_gene_pairs <- expand_grid(g1 = sort(unique(conv_sites$caap_group)), g2 = sort(unique(conv_sites$caap_group))) %>%
    rowwise() %>%
    mutate(jaccard = {
      s1 <- conv_sites %>% filter(caap_group == g1) %>% pull(gene) %>% unique()
      s2 <- conv_sites %>% filter(caap_group == g2) %>% pull(gene) %>% unique()
      if (length(union(s1, s2)) == 0) 0 else length(intersect(s1, s2)) / length(union(s1, s2))
    }) %>% ungroup()

  p_site_hm <- ggplot(overlap_site_pairs, aes(g1, g2, fill = jaccard)) + geom_tile() + scale_fill_viridis_c(limits = c(0,1)) + theme_minimal() + labs(title = "Convergent-site overlap (Jaccard)")
  p_gene_hm <- ggplot(overlap_gene_pairs, aes(g1, g2, fill = jaccard)) + geom_tile() + scale_fill_viridis_c(limits = c(0,1)) + theme_minimal() + labs(title = "Convergent-gene overlap (Jaccard)")
  print(p_site_hm + p_gene_hm)
}
```

## B.5 US/GS non-divergence/divergence relations (upset-like + structured exports)

```{r us_gs_relation_analysis, fig.height=9, fig.width=14}
collapse_change_side <- function(x) {
  x <- tolower(as.character(x))
  x <- x[x %in% c("top", "bottom", "both")]
  if (length(x) == 0) return("none")
  if ("both" %in% x) return("both")
  ux <- unique(x)
  if (all(c("top", "bottom") %in% ux)) return("both")
  ux[[1]]
}

combo_from_flags <- function(df, cols) {
  apply(df[, cols, drop = FALSE], 1, function(v) {
    labs <- cols[as.logical(v)]
    if (length(labs) == 0) "none" else paste(labs, collapse = "+")
  })
}

plot_upset_like <- function(intersections_df, set_cols, title_txt, out_file) {
  if (nrow(intersections_df) == 0) return(invisible(NULL))

  plot_df <- intersections_df %>%
    arrange(desc(n), combo_label) %>%
    mutate(combo_id = factor(seq_len(n())))

  p_bar <- ggplot(plot_df, aes(x = combo_id, y = n)) +
    geom_col(fill = tableau_10[["blue"]], color = "black", linewidth = 0.2) +
    labs(title = title_txt, x = NULL, y = "Intersection size") +
    theme_minimal(base_size = 10) +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

  mat_long <- plot_df %>%
    select(combo_id, all_of(set_cols)) %>%
    pivot_longer(cols = all_of(set_cols), names_to = "set_name", values_to = "present") %>%
    mutate(
      set_name = case_when(
        set_name == "us_nondiv" ~ "US_nondiv",
        set_name == "us_divergent" ~ "US_div",
        set_name == "gs_div_grp" ~ "GS(1-4)_div",
        set_name == "gs_nondiv_grp" ~ "GS(1-4)_nondiv",
        set_name == "gs0_div" ~ "GS0_div",
        set_name == "gs0_nondiv" ~ "GS0_nondiv",
        str_detect(set_name, "^nondiv_GS") ~ str_replace(set_name, "^nondiv_", "") %>% paste0("_nondiv"),
        str_detect(set_name, "^div_GS") ~ str_replace(set_name, "^div_", "") %>% paste0("_div"),
        TRUE ~ set_name
      )
    )

  p_mat <- ggplot(mat_long, aes(x = combo_id, y = set_name, fill = present)) +
    geom_tile(color = "white", linewidth = 0.3) +
    scale_fill_manual(values = c(`TRUE` = tableau_10[["vermillion"]], `FALSE` = "#eeeeee")) +
    labs(x = "Intersections", y = "Set") +
    theme_minimal(base_size = 10) +
    theme(legend.position = "none", axis.text.x = element_blank(), axis.ticks.x = element_blank())

  p <- p_bar / p_mat + plot_layout(heights = c(2.2, 1.3))
  print(p)
  ggsave(out_file, p, width = 13, height = 7, dpi = 320)
}

write_export_pair <- function(df, key, txt_dir, tsv_dir) {
  createDir(txt_dir)
  createDir(tsv_dir)
  base_cols <- c("gene", "position", "CAAP_group", "change_pattern", "change_side")

  out_df <- df %>% select(all_of(base_cols)) %>% distinct() %>% arrange(gene, position, CAAP_group)
  genes <- out_df %>% distinct(gene) %>% arrange(gene)

  readr::write_lines(genes$gene, file.path(txt_dir, paste0(key, ".txt")))
  write_tsv(out_df, file.path(tsv_dir, paste0(key, ".tsv")))
}

us_gs_dir <- file.path(disambig_dir, "us_gs_relations")
us_gs_plot_dir <- file.path(us_gs_dir, "plots")
us_gs_export_dir <- file.path(us_gs_dir, "exports")
us_gs_txt_dir <- file.path(us_gs_export_dir, "txt")
us_gs_tsv_dir <- file.path(us_gs_export_dir, "tsv")
createDir(us_gs_dir)
createDir(us_gs_plot_dir)
createDir(us_gs_export_dir)
createDir(us_gs_txt_dir)
createDir(us_gs_tsv_dir)

gs_all <- c("GS0", "GS1", "GS2", "GS3", "GS4")
gs_core <- c("GS1", "GS2", "GS3", "GS4")

site_group <- discovery_std %>%
  mutate(
    caap_group = toupper(caap_group),
    is_nondiv = pattern_type != "divergent"
  ) %>%
  group_by(site_id, gene, position, caap_group) %>%
  summarise(
    nondiv = any(is_nondiv, na.rm = TRUE),
    sig_both = any(sig_both, na.rm = TRUE),
    is_conserved_meta = any(is_conserved_meta, na.rm = TRUE),
    change_side = collapse_change_side(change_side),
    .groups = "drop"
  )

site_nondiv_wide <- site_group %>%
  filter(caap_group %in% c("US", gs_all)) %>%
  select(site_id, gene, position, caap_group, nondiv) %>%
  distinct() %>%
  pivot_wider(names_from = caap_group, values_from = nondiv, values_fill = FALSE, names_prefix = "nondiv_")

site_present_wide <- site_group %>%
  filter(caap_group %in% c("US", gs_all)) %>%
  mutate(present = TRUE) %>%
  select(site_id, gene, position, caap_group, present) %>%
  distinct() %>%
  pivot_wider(names_from = caap_group, values_from = present, values_fill = FALSE, names_prefix = "present_")

site_meta <- site_group %>%
  filter(caap_group %in% c("US", gs_all)) %>%
  group_by(site_id, gene, position) %>%
  summarise(
    us_present = any(caap_group == "US"),
    us_sig_both = any(sig_both[caap_group == "US"], na.rm = TRUE),
    us_change_side = collapse_change_side(change_side[caap_group == "US"]),
    gs_sig_both = any(sig_both[caap_group %in% gs_all], na.rm = TRUE),
    gs_change_side = collapse_change_side(change_side[caap_group %in% gs_all]),
    is_conserved_meta = any(is_conserved_meta, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    sig_both_ref = if_else(us_present, us_sig_both, gs_sig_both),
    change_side_ref = case_when(
      us_present & us_change_side %in% c("top", "bottom", "both") ~ us_change_side,
      !us_present & gs_change_side %in% c("top", "bottom", "both") ~ gs_change_side,
      TRUE ~ "none"
    )
  )

site_matrix <- site_nondiv_wide %>%
  left_join(site_present_wide, by = c("site_id", "gene", "position")) %>%
  left_join(site_meta, by = c("site_id", "gene", "position"))

for (g in c("US", gs_all)) {
  nd <- paste0("nondiv_", g)
  pr <- paste0("present_", g)
  if (!nd %in% names(site_matrix)) site_matrix[[nd]] <- FALSE
  if (!pr %in% names(site_matrix)) site_matrix[[pr]] <- FALSE
}

site_matrix <- site_matrix %>%
  mutate(
    us_nondiv = nondiv_US,
    us_divergent = us_present & !nondiv_US,
    div_GS0 = present_GS0 & !nondiv_GS0,
    div_GS1 = present_GS1 & !nondiv_GS1,
    div_GS2 = present_GS2 & !nondiv_GS2,
    div_GS3 = present_GS3 & !nondiv_GS3,
    div_GS4 = present_GS4 & !nondiv_GS4,
    gs_div_grp = rowSums(across(all_of(paste0("div_GS", 1:4)), as.integer), na.rm = TRUE) > 0,
    gs_nondiv_grp = rowSums(across(all_of(paste0("nondiv_GS", 1:4)), as.integer), na.rm = TRUE) > 0,
    gs0_div = div_GS0,
    gs0_nondiv = nondiv_GS0,
    gs_any_nondiv = rowSums(across(all_of(paste0("nondiv_", gs_all)), as.integer), na.rm = TRUE) > 0,
    gs_any_nondiv_core = rowSums(across(all_of(paste0("nondiv_", gs_core)), as.integer), na.rm = TRUE) > 0,
    gs0_poisoned = present_GS0,
    case_us_nondiv_total = us_nondiv,
    case_us_nondiv_and_gs_nondiv = us_nondiv & gs_any_nondiv,
    case_us_div_and_gs_nondiv = us_divergent & gs_any_nondiv,
    case_gs_nondiv_no_us_excl_gs0 = !us_present & gs_any_nondiv_core & !gs0_poisoned
  )

# New plot: number of is_conserved_meta=TRUE CAAS per case
conserved_case_counts <- site_matrix %>%
  filter(is_conserved_meta) %>%
  summarise(
    case_us_nondiv_total = sum(case_us_nondiv_total, na.rm = TRUE),
    case_us_nondiv_and_gs_nondiv = sum(case_us_nondiv_and_gs_nondiv, na.rm = TRUE),
    case_us_div_and_gs_nondiv = sum(case_us_div_and_gs_nondiv, na.rm = TRUE),
    case_gs_nondiv_no_us_excl_gs0 = sum(case_gs_nondiv_no_us_excl_gs0, na.rm = TRUE)
  ) %>%
  pivot_longer(everything(), names_to = "case", values_to = "n_conserved_meta") %>%
  mutate(
    case = recode(
      case,
      case_us_nondiv_total = "US nondiv total",
      case_us_nondiv_and_gs_nondiv = "US nondiv + GS nondiv",
      case_us_div_and_gs_nondiv = "US div + GS nondiv",
      case_gs_nondiv_no_us_excl_gs0 = "GS nondiv no US (excl GS0)"
    ),
    case = forcats::fct_reorder(case, n_conserved_meta, .desc = TRUE)
  )

p_conserved_case_counts <- ggplot(conserved_case_counts, aes(x = case, y = n_conserved_meta, fill = case)) +
  geom_col(color = "black", linewidth = 0.25) +
  geom_text(aes(label = n_conserved_meta), vjust = -0.3, size = 3.5) +
  scale_fill_paletteer_d("ggthemes::Tableau_10") +
  labs(
    title = "Number of is_conserved_meta = TRUE CAAS per US/GS case",
    x = "Case",
    y = "Count"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1))

print(p_conserved_case_counts)
ggsave(file.path(us_gs_plot_dir, "conserved_meta_true_counts_by_case.png"), p_conserved_case_counts, width = 10, height = 6, dpi = 320)

# Upset 1: GS1..GS4 divergent grouped, GS0 divergent separate
upset_sets <- c("us_nondiv", "us_divergent", paste0("nondiv_", gs_all), "gs_div_grp", "gs0_div")

upset_intersections <- site_matrix %>%
  count(across(all_of(upset_sets)), name = "n") %>%
  mutate(combo_label = combo_from_flags(cur_data(), upset_sets)) %>%
  filter(n > 0)

plot_upset_like(
  intersections_df = upset_intersections,
  set_cols = upset_sets,
  title_txt = "US/GS upset-like intersections (GS1..GS4 div grouped; GS0 div separate)",
  out_file = file.path(us_gs_plot_dir, "us_gs_upset_like_global.png")
)

# Upset 2: grouped GS1..GS4 non-div and div, plus GS0 non-div/div
upset_sets_grouped <- c("us_divergent", "us_nondiv", "gs_div_grp", "gs_nondiv_grp", "gs0_div", "gs0_nondiv")
upset_intersections_grouped <- site_matrix %>%
  count(across(all_of(upset_sets_grouped)), name = "n") %>%
  mutate(combo_label = combo_from_flags(cur_data(), upset_sets_grouped)) %>%
  filter(n > 0)

plot_upset_like(
  intersections_df = upset_intersections_grouped,
  set_cols = upset_sets_grouped,
  title_txt = "US/GS upset-like intersections (US_div, US_nondiv, GS_div, GS_nondiv, GS0_div, GS0_nondiv)",
  out_file = file.path(us_gs_plot_dir, "us_gs_upset_like_grouped_gs14.png")
)

for (cs in c("top", "bottom", "both")) {
  cs_intersections <- site_matrix %>%
    filter(change_side_ref == cs) %>%
    count(across(all_of(upset_sets)), name = "n") %>%
    mutate(combo_label = combo_from_flags(cur_data(), upset_sets)) %>%
    filter(n > 0)

  if (nrow(cs_intersections) > 0) {
    plot_upset_like(
      intersections_df = cs_intersections,
      set_cols = upset_sets,
      title_txt = paste0("US/GS upset-like intersections (change_side=", cs, ")"),
      out_file = file.path(us_gs_plot_dir, paste0("us_gs_upset_like_change_side_", cs, ".png"))
    )
  }
}

detail_rows <- discovery_std %>%
  transmute(
    site_id,
    gene,
    position,
    CAAP_group = toupper(caap_group),
    change_pattern = pattern_type,
    change_side = if_else(change_side %in% c("top", "bottom", "both"), change_side, "none")
  ) %>%
  distinct()

site_flags <- site_matrix %>%
  transmute(
    site_id,
    sig_both_ref,
    change_side_ref,
    case_us_nondiv_total,
    case_us_nondiv_and_gs_nondiv,
    case_us_div_and_gs_nondiv,
    case_gs_nondiv_no_us_excl_gs0
  )

detail_with_flags <- detail_rows %>%
  left_join(site_flags, by = "site_id") %>%
  mutate(
    sig_both_ref = replace_na(sig_both_ref, FALSE),
    change_side_ref = replace_na(change_side_ref, "none"),
    case_us_nondiv_total = replace_na(case_us_nondiv_total, FALSE),
    case_us_nondiv_and_gs_nondiv = replace_na(case_us_nondiv_and_gs_nondiv, FALSE),
    case_us_div_and_gs_nondiv = replace_na(case_us_div_and_gs_nondiv, FALSE),
    case_gs_nondiv_no_us_excl_gs0 = replace_na(case_gs_nondiv_no_us_excl_gs0, FALSE)
  )

write_export_pair(detail_with_flags, "global", us_gs_txt_dir, us_gs_tsv_dir)
write_export_pair(detail_with_flags %>% filter(sig_both_ref), "global_significant", us_gs_txt_dir, us_gs_tsv_dir)

for (cs in c("top", "bottom", "both")) {
  write_export_pair(
    detail_with_flags %>% filter(change_side_ref == cs),
    paste0("change_side_", cs),
    us_gs_txt_dir,
    us_gs_tsv_dir
  )
  write_export_pair(
    detail_with_flags %>% filter(change_side_ref == cs, sig_both_ref),
    paste0("change_side_", cs, "_significant"),
    us_gs_txt_dir,
    us_gs_tsv_dir
  )
}

special1 <- detail_with_flags %>%
  filter(case_us_nondiv_total | case_us_nondiv_and_gs_nondiv | case_us_div_and_gs_nondiv)

write_export_pair(special1, "special_union_us_nondiv_and_us_gs_cases", us_gs_txt_dir, us_gs_tsv_dir)
write_export_pair(special1 %>% filter(sig_both_ref), "special_union_us_nondiv_and_us_gs_cases_significant", us_gs_txt_dir, us_gs_tsv_dir)
write_export_pair(special1 %>% filter(!sig_both_ref), "special_union_us_nondiv_and_us_gs_cases_nonsignificant", us_gs_txt_dir, us_gs_tsv_dir)

for (cs in c("top", "bottom", "both")) {
  s1_cs <- special1 %>% filter(change_side_ref == cs)
  write_export_pair(s1_cs, paste0("special_union_us_nondiv_and_us_gs_cases_change_side_", cs), us_gs_txt_dir, us_gs_tsv_dir)
  write_export_pair(s1_cs %>% filter(sig_both_ref), paste0("special_union_us_nondiv_and_us_gs_cases_change_side_", cs, "_significant"), us_gs_txt_dir, us_gs_tsv_dir)
}

special2 <- detail_with_flags %>%
  filter(case_gs_nondiv_no_us_excl_gs0)

write_export_pair(special2, "special_gs_nondiv_no_us_excluding_gs0", us_gs_txt_dir, us_gs_tsv_dir)
write_export_pair(special2 %>% filter(sig_both_ref), "special_gs_nondiv_no_us_excluding_gs0_significant", us_gs_txt_dir, us_gs_tsv_dir)
write_export_pair(special2 %>% filter(!sig_both_ref), "special_gs_nondiv_no_us_excluding_gs0_nonsignificant", us_gs_txt_dir, us_gs_tsv_dir)

for (cs in c("top", "bottom", "both")) {
  s2_cs <- special2 %>% filter(change_side_ref == cs)
  write_export_pair(s2_cs, paste0("special_gs_nondiv_no_us_excluding_gs0_change_side_", cs), us_gs_txt_dir, us_gs_tsv_dir)
  write_export_pair(s2_cs %>% filter(sig_both_ref), paste0("special_gs_nondiv_no_us_excluding_gs0_change_side_", cs, "_significant"), us_gs_txt_dir, us_gs_tsv_dir)
}

cat("
✓ US/GS upset-like plots and exports generated:
")
cat("  Plots:", us_gs_plot_dir, "
")
cat("  TXT:", us_gs_txt_dir, "
")
cat("  TSV:", us_gs_tsv_dir, "

")
```

## B.6 Manhattan: genomic distribution of removed genes (−log10 permutation p-value)

```{r manhattan_disambiguation, fig.height=8, fig.width=16, eval=generate_manhattan}
if (!"chr" %in% names(gene_ensembl)) {
  cat("⚠ Chromosome information not available in gene ensembl file, skipping Manhattan plot\n")
} else {
  # Aggregate per-gene best (minimum) bootstrap p-value from position-level data
  gene_boot_pval <- discovery_std %>%
    group_by(gene) %>%
    summarise(
      min_boot_pval = min(pvalue_boot, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      min_boot_pval = pmax(min_boot_pval, .Machine$double.xmin, na.rm = TRUE),
      neglog10_boot = -log10(min_boot_pval)
    )

  # Only visualise removed genes (Extreme, Dubious, Both); Normal genes are not plotted
  manhattan_data <- gene_summary_annotated %>%
    filter(Category != "Normal") %>%
    left_join(gene_ensembl %>% select(gene, chr, start, end), by = c("Gene" = "gene")) %>%
    left_join(gene_boot_pval, by = c("Gene" = "gene")) %>%
    filter(!is.na(chr), !is.na(neglog10_boot)) %>%
    mutate(
      chr_num   = str_replace(chr, "chr", ""),
      chr_order = case_when(
        chr_num %in% as.character(1:22) ~ as.numeric(chr_num),
        chr_num == "X"                  ~ 23,
        chr_num == "Y"                  ~ 24,
        TRUE                            ~ 25
      ),
      pos = (start + end) / 2,
      Category = factor(Category, levels = c("Extreme", "Dubious", "Both"))
    ) %>%
    arrange(chr_order, pos)

  if (nrow(manhattan_data) > 0) {
    # Cumulative chromosome positions and boundaries
    chr_lengths <- manhattan_data %>%
      group_by(chr_num, chr_order) %>%
      summarise(chr_max = max(pos, na.rm = TRUE), .groups = "drop") %>%
      arrange(chr_order) %>%
      mutate(
        chr_start  = lag(cumsum(chr_max), default = 0),
        chr_end    = chr_start + chr_max,
        chr_center = chr_start + chr_max / 2
      )

    manhattan_data <- manhattan_data %>%
      left_join(chr_lengths %>% select(chr_num, chr_start, chr_end), by = "chr_num") %>%
      mutate(BPcum = pos + chr_start)

    # Bonferroni threshold across all genes with a genomic position
    n_genes_tested  <- n_distinct(manhattan_data$Gene)
    bonferroni_thr  <- -log10(0.05 / max(n_genes_tested, 1L))

    removed_palette <- category_palette[intersect(c("Extreme", "Dubious", "Both"), levels(manhattan_data$Category))]
    removed_shapes  <- c(Extreme = 17L, Dubious = 16L, Both = 15L)

    p_manhattan <- ggplot(manhattan_data,
        aes(x = BPcum, y = neglog10_boot, color = Category, shape = Category)) +
      # Discontinuous chromosome boundary lines
      geom_vline(
        data        = chr_lengths,
        aes(xintercept = chr_end),
        linetype    = "dotted", color = "grey60", alpha = 0.7, linewidth = 0.4,
        inherit.aes = FALSE
      ) +
      geom_point(alpha = 0.80, size = 2.5) +
      # Bonferroni threshold
      geom_hline(yintercept = bonferroni_thr, linetype = "dashed",
                 color = "black", linewidth = 0.6) +
      annotate(
        "text",
        x     = min(manhattan_data$BPcum, na.rm = TRUE),
        y     = bonferroni_thr + max(manhattan_data$neglog10_boot, na.rm = TRUE) * 0.03,
        label = sprintf("Bonferroni (α = 0.05, n = %d)", n_genes_tested),
        hjust = 0, size = 3, color = "black"
      ) +
      scale_color_manual(values = removed_palette, drop = FALSE) +
      scale_shape_manual(values = removed_shapes,  drop = FALSE) +
      scale_x_continuous(
        labels = chr_lengths$chr_num,
        breaks = chr_lengths$chr_center,
        expand = expansion(mult = c(0.005, 0.005))
      ) +
      labs(
        title    = "Genomic distribution of removed genes after disambiguation",
        subtitle = sprintf(
          "Removed genes only (n = %d) | y = −log10(min permutation p-value per gene) | dashed line = Bonferroni threshold",
          n_genes_tested
        ),
        x     = "Chromosome",
        y     = "−log10(min permutation p-value)",
        color = "Removal category",
        shape = "Removal category"
      ) +
      theme_minimal(base_size = 11) +
      theme(
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()
      )

    print(p_manhattan)
    ggsave(
      filename = file.path(disambig_dir, "manhattan_disambiguation_removed.png"),
      plot     = p_manhattan, width = 16, height = 8, dpi = 320, dev = "png"
    )
  } else {
    cat("No removed genes with genomic coordinates available — Manhattan plot skipped.\n")
  }
}
```

## 5. Impact Summary

```{r impact_summary}
cat("\n=== CT Post-Processing Impact Summary ===\n\n")

threshold_extreme <- gene_summary_annotated$threshold_extreme[1]
threshold_dubious <- gene_summary_annotated$threshold_dubious[1]

cat("Gene-level statistics:\n")
cat("  Total genes analyzed:", n_distinct(discovery_df$Gene), "\n")
cat("  Extreme genes (top 1% density):", sum(gene_summary_annotated$Extreme), "\n")
cat("  Dubious genes (IQR outliers):", sum(gene_summary_annotated$Dubious), "\n")
cat("  Both categories:", sum(gene_summary_annotated$Extreme & gene_summary_annotated$Dubious), "\n")
cat("  Normal genes:", sum(gene_summary_annotated$Category == "Normal"), "\n\n")

cat("Position-level statistics:\n")
cat("  Total CT positions:", nrow(discovery_df), "\n")
if (nrow(filter_summary) > 0) {
  cat("  Positions flagged by filtering:\n")
  for (i in 1:nrow(filter_summary)) {
    cat(sprintf("    minlen=%d, maxcaas=%.2f: %d positions\n", filter_summary$Minlen[i], filter_summary$Maxcaas[i], filter_summary$DiscardedCount[i]))
  }
}

summary_table <- tibble(
  Metric = c("Total CT Positions", "Unique Genes", "Extreme Genes (top 1% density)", "Dubious Genes (IQR outliers)", "Both Extreme and Dubious", "Normal Genes", "Extreme Threshold (density)", "Dubious Threshold (count)"),
  Value = c(
    nrow(discovery_df),
    n_distinct(discovery_df$Gene),
    sum(gene_summary_annotated$Extreme),
    sum(gene_summary_annotated$Dubious),
    sum(gene_summary_annotated$Extreme & gene_summary_annotated$Dubious),
    sum(gene_summary_annotated$Category == "Normal"),
    sprintf("%.4f", threshold_extreme),
    as.character(threshold_dubious)
  )
)

kable(summary_table, caption = "CT Post-Processing Summary Statistics", align = c("l", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

write_tsv(summary_table, file.path(summary_stats_dir, "summary_statistics.tsv"))
cat("\n✓ Summary statistics exported\n")
cat("\n=== Analysis Complete ===\n")
```

## Session Info

```{r session_info}
sessionInfo()
```
