---
title: "CT Post-Processing and Disambiguation Characterization"
author: "PhyloPhere Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    code_folding: hide
    df_print: paged
params:
  discovery_file: NULL
  filter_summary_file: NULL
  filter_dir: NULL
  gene_ensembl_file: NULL
  gene_stats_file: NULL
  processing_mode: "exploratory"
  output_dir: "ct_postproc"
  extreme_threshold: 0.99
  iqr_multiplier: 3.0
  alpha_threshold: 0.1
  gene_filter_mode: "both"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7,
  dev = 'png',
  dpi = 320
)

suppressPackageStartupMessages({
  library(stringr)
  library(purrr)
  library(tibble)
  library(patchwork)
  library(scales)
  library(ggplot2)
  library(forcats)
  library(knitr)
  library(kableExtra)
  library(paletteer)
  library(tidyverse)
  if (requireNamespace("UpSetR", quietly = TRUE)) library(UpSetR)
})

createDir <- function(directory) {
  if (!dir.exists(directory)) {
    dir.create(directory, recursive = TRUE, showWarnings = FALSE)
  }
}

find_first_existing <- function(candidates) {
  hits <- candidates[file.exists(candidates)]
  if (length(hits) == 0) return(NA_character_)
  hits[[1]]
}

find_recursive_file <- function(root_dir, pattern) {
  if (is.null(root_dir) || is.na(root_dir) || !dir.exists(root_dir)) return(NA_character_)
  files <- list.files(root_dir, pattern = pattern, full.names = TRUE, recursive = TRUE)
  if (length(files) == 0) return(NA_character_)
  files[[1]]
}

as_bool <- function(x) {
  tolower(as.character(x)) %in% c("true", "t", "1", "yes", "y")
}

format_threshold <- function(x) {
  formatC(as.numeric(x), format = "f", digits = 2)
}

make_significance_class <- function(sig_hyp, sig_perm) {
  case_when(
    sig_hyp & sig_perm ~ "both",
    sig_hyp ~ "hyper_only",
    sig_perm ~ "perm_only",
    TRUE ~ "none"
  )
}

collapse_change_side <- function(x) {
  x <- tolower(as.character(x))
  x <- x[x %in% c("top", "bottom", "both")]
  if (length(x) == 0) return("none")
  if ("both" %in% x) return("both")
  ux <- unique(x)
  if (all(c("top", "bottom") %in% ux)) return("both")
  ux[[1]]
}

combo_from_flags <- function(df, cols) {
  apply(df[, cols, drop = FALSE], 1, function(v) {
    labs <- cols[as.logical(v)]
    if (length(labs) == 0) "none" else paste(labs, collapse = "+")
  })
}

normalize_parallel_type <- function(x) {
  x_chr <- tolower(trimws(as.character(x)))
  x_chr[is.na(x_chr) | x_chr %in% c("", "na", "nan", "null")] <- "none"
  x_chr
}

make_pattern_parallel_label <- function(pattern_type, parallel_type) {
  pattern_chr <- as.character(pattern_type)
  pattern_chr[is.na(pattern_chr) | pattern_chr == ""] <- "unknown"
  parallel_chr <- normalize_parallel_type(parallel_type)
  if_else(parallel_chr == "none", pattern_chr, paste(pattern_chr, parallel_chr, sep = " / "))
}

format_pattern_parallel_axis <- function(x) {
  stringr::str_replace_all(as.character(x), "\\s*/\\s*", "\n/ ")
}

derive_plot_zoom_upper <- function(plot_obj, zoom_fraction = 0.10, min_upper = 3) {
  built <- ggplot_build(plot_obj)
  layer_max <- vapply(built$data, function(layer_df) {
    metric_cols <- intersect(c("count", "y", "density"), names(layer_df))
    if (length(metric_cols) == 0) return(0)
    max(suppressWarnings(as.numeric(layer_df[[metric_cols[[1]]]])), na.rm = TRUE)
  }, numeric(1))

  max_val <- suppressWarnings(max(layer_max, na.rm = TRUE))
  if (!is.finite(max_val) || max_val <= 0) return(min_upper)
  max(min_upper, ceiling(max_val * zoom_fraction))
}

compute_barplot_dims <- function(n_labels, n_facets = 1, facet_cols = 1,
                                 base_width = 8, base_height = 5.5,
                                 per_label = 0.55, facet_col_bonus = 1.4,
                                 facet_row_height = 2.2,
                                 min_width = 8, max_width = 24,
                                 min_height = 5, max_height = 18) {
  n_labels <- max(1, suppressWarnings(as.integer(n_labels)))
  n_facets <- max(1, suppressWarnings(as.integer(n_facets)))
  facet_cols <- max(1, suppressWarnings(as.integer(facet_cols)))
  facet_rows <- ceiling(n_facets / facet_cols)

  width <- base_width + max(0, n_labels - 3) * per_label + max(0, min(n_facets, facet_cols) - 1) * facet_col_bonus
  height <- base_height + max(0, facet_rows - 1) * facet_row_height

  list(
    width = min(max_width, max(min_width, width)),
    height = min(max_height, max(min_height, height))
  )
}

save_barplot <- function(filename, plot, n_labels, n_facets = 1, facet_cols = 1,
                         dpi = 320, base_width = 8, base_height = 5.5,
                         per_label = 0.55, facet_row_height = 2.2) {
  dims <- compute_barplot_dims(
    n_labels = n_labels,
    n_facets = n_facets,
    facet_cols = facet_cols,
    base_width = base_width,
    base_height = base_height,
    per_label = per_label,
    facet_row_height = facet_row_height
  )
  ggsave(filename, plot, width = dims$width, height = dims$height, dpi = dpi)
}

build_pairwise_overlap_plot <- function(set_list, level_label, subset_label) {
  set_names <- names(set_list)
  pairwise_df <- expand.grid(group_a = set_names, group_b = set_names, stringsAsFactors = FALSE) %>%
    mutate(
      overlap_n = purrr::map2_int(group_a, group_b, ~ length(intersect(set_list[[.x]], set_list[[.y]]))),
      union_n = purrr::map2_int(group_a, group_b, ~ length(union(set_list[[.x]], set_list[[.y]]))),
      jaccard = if_else(union_n > 0, overlap_n / union_n, 0)
    )

  ggplot(pairwise_df, aes(group_a, group_b, fill = jaccard)) +
    geom_tile(color = "white", linewidth = 0.4) +
    geom_text(aes(label = overlap_n), size = 4.5, fontface = "bold") +
    scale_fill_gradient(low = "#F1F5F9", high = tableau_10[["blue"]], limits = c(0, 1), name = "Jaccard") +
    labs(
      title = paste0(level_label, "-level overlap [", subset_label, "]"),
      subtitle = "Fallback pairwise overlap heatmap; cell labels show intersection size",
      x = NULL,
      y = NULL
    ) +
    theme_minimal(base_size = 15) +
    theme(
      plot.title = element_text(face = "bold", size = 18),
      plot.subtitle = element_text(size = 12),
      axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
      axis.text.y = element_text(face = "bold"),
      panel.grid = element_blank()
    )
}

build_orientation_rows <- function(df) {
  bind_rows(
    df %>% mutate(orientation = "global"),
    df %>% filter(change_side %in% c("top", "both")) %>% mutate(orientation = "top"),
    df %>% filter(change_side %in% c("bottom", "both")) %>% mutate(orientation = "bottom"),
    df %>% filter(change_side == "both") %>% mutate(orientation = "both")
  ) %>%
    mutate(
      orientation = factor(orientation, levels = c("global", "top", "bottom", "both"))
    )
}

build_group_flag_table <- function(df) {
  df %>%
    mutate(caap_group = toupper(caap_group)) %>%
    group_by(orientation, site_id, gene, position) %>%
    summarise(
      us = any(caap_group == "US"),
      gs0 = any(caap_group == "GS0"),
      gs1 = any(caap_group == "GS1"),
      gs2 = any(caap_group == "GS2"),
      gs3 = any(caap_group == "GS3"),
      gs4 = any(caap_group == "GS4"),
      .groups = "drop"
    )
}

build_threshold_line_df <- function(alpha_threshold) {
  tibble(
    line_key = c("configured", "comparison_010", "comparison_005"),
    threshold = c(alpha_threshold, 0.10, 0.05),
    line_label = c(
      paste0("Configured alpha = ", format_threshold(alpha_threshold)),
      "Comparison threshold = 0.10",
      "Comparison threshold = 0.05"
    )
  ) %>%
    group_by(threshold) %>%
    summarise(
      line_key = if ("configured" %in% line_key) "configured" else first(line_key),
      line_label = paste(unique(line_label), collapse = " = "),
      .groups = "drop"
    ) %>%
    arrange(desc(line_key == "configured"), desc(threshold))
}

resolve_filtered_cluster_file <- function(minlen, maxcaas_int, filter_root = NULL, postproc_root = NULL) {
  search_roots <- unique(c(filter_root, postproc_root, getwd()))
  search_roots <- search_roots[!is.na(search_roots) & nzchar(search_roots) & dir.exists(search_roots)]
  filter_root_norm <- if (!is.null(filter_root) && length(filter_root) == 1 && !is.na(filter_root) && nzchar(filter_root)) {
    normalizePath(filter_root, winslash = "/", mustWork = FALSE)
  } else {
    NA_character_
  }
  postproc_root_norm <- if (!is.null(postproc_root) && length(postproc_root) == 1 && !is.na(postproc_root) && nzchar(postproc_root)) {
    normalizePath(postproc_root, winslash = "/", mustWork = FALSE)
  } else {
    NA_character_
  }

  candidate_files <- map(search_roots, ~ list.files(
    path = .x,
    pattern = "\\.filtered\\.minlen[0-9]+\\.maxcaas[0-9]+\\.tsv$",
    recursive = TRUE,
    full.names = TRUE
  )) %>%
    unlist() %>%
    unique()

  if (length(candidate_files) == 0) {
    return(list(
      path = NA_character_,
      status = "missing",
      message = "No filtered cluster TSV files were found in the report workspace or filter output directories."
    ))
  }

  file_index <- tibble(file_path = candidate_files) %>%
    mutate(
      normalized_path = normalizePath(file_path, winslash = "/", mustWork = FALSE),
      file_name = basename(file_path),
      minlen_match = str_match(file_name, "minlen([0-9]+)\\.maxcaas([0-9]+)"),
      minlen_file = suppressWarnings(as.integer(minlen_match[, 2])),
      maxcaas_file = suppressWarnings(as.integer(minlen_match[, 3])),
      priority = case_when(
        !is.na(filter_root_norm) & str_starts(normalized_path, fixed(filter_root_norm)) ~ 1L,
        !is.na(postproc_root_norm) & str_starts(normalized_path, fixed(postproc_root_norm)) ~ 2L,
        TRUE ~ 3L
      )
    ) %>%
    filter(!is.na(minlen_file), !is.na(maxcaas_file)) %>%
    distinct(normalized_path, .keep_all = TRUE)

  matches <- file_index %>%
    filter(minlen_file == as.integer(minlen), maxcaas_file == as.integer(maxcaas_int)) %>%
    arrange(priority, file_name)

  if (nrow(matches) == 0) {
    return(list(
      path = NA_character_,
      status = "missing",
      message = sprintf(
        "No filtered file matched minlen=%d and maxcaas=%d. Expected any '*.filtered.minlen%d.maxcaas%d.tsv'.",
        as.integer(minlen), as.integer(maxcaas_int), as.integer(minlen), as.integer(maxcaas_int)
      )
    ))
  }

  if (nrow(matches) > 1) {
    return(list(
      path = matches$normalized_path[[1]],
      status = "multiple",
      message = sprintf(
        "Multiple matches found for minlen=%d and maxcaas=%d; using preferred file `%s`.",
        as.integer(minlen), as.integer(maxcaas_int), basename(matches$normalized_path[[1]])
      )
    ))
  }

  list(
    path = matches$normalized_path[[1]],
    status = "ok",
    message = sprintf("Using filtered file `%s`.", basename(matches$normalized_path[[1]]))
  )
}

normalize_disambig_schema <- function(df) {
  if ('Pattern' %in% names(df) && !'pattern_type' %in% names(df)) df <- df %>% rename(pattern_type = Pattern)
  if ('Pvalue' %in% names(df) && !'pvalue' %in% names(df)) df <- df %>% rename(pvalue = Pvalue)
  if ('Pvalue.boot' %in% names(df) && !'pvalue_boot' %in% names(df)) df <- df %>% rename(pvalue_boot = `Pvalue.boot`)
  if ('ChangeSide' %in% names(df) && !'change_side' %in% names(df)) df <- df %>% rename(change_side = ChangeSide)
  if ('ChangeTop' %in% names(df) && !'change_top' %in% names(df)) df <- df %>% rename(change_top = ChangeTop)
  if ('ChangeBottom' %in% names(df) && !'change_bottom' %in% names(df)) df <- df %>% rename(change_bottom = ChangeBottom)
  if ('ParallelTop' %in% names(df) && !'parallel_top' %in% names(df)) df <- df %>% rename(parallel_top = ParallelTop)
  if ('ParallelBottom' %in% names(df) && !'parallel_bottom' %in% names(df)) df <- df %>% rename(parallel_bottom = ParallelBottom)
  if ('ParallelType' %in% names(df) && !'parallel_type' %in% names(df)) df <- df %>% rename(parallel_type = ParallelType)
  if ('Gene' %in% names(df) && !'gene' %in% names(df)) df <- df %>% rename(gene = Gene)
  if ('Position' %in% names(df) && !'position' %in% names(df)) df <- df %>% rename(position = Position)
  if ('CAAP_Group' %in% names(df) && !'caap_group' %in% names(df)) df <- df %>% rename(caap_group = CAAP_Group)
  if ('ASR_ConsAntiCAAS' %in% names(df) && !'asr_is_conserved' %in% names(df)) df <- df %>% rename(asr_is_conserved = ASR_ConsAntiCAAS)
  if (!'caap_group' %in% names(df)) df$caap_group <- 'US'
  if (!'pattern_type' %in% names(df)) df$pattern_type <- 'unknown'
  if (!'change_side' %in% names(df)) df$change_side <- 'none'
  if (!'change_top' %in% names(df)) df$change_top <- 'no_change'
  if (!'change_bottom' %in% names(df)) df$change_bottom <- 'no_change'
  if (!'parallel_top' %in% names(df)) df$parallel_top <- 'none'
  if (!'parallel_bottom' %in% names(df)) df$parallel_bottom <- 'none'
  if (!'parallel_type' %in% names(df)) df$parallel_type <- 'none'
  if (!'is_conserved_meta' %in% names(df)) df$is_conserved_meta <- FALSE
  if (!'asr_is_conserved' %in% names(df)) df$asr_is_conserved <- FALSE
  if (!'asr_root_conserved' %in% names(df)) df$asr_root_conserved <- FALSE

  df %>%
    mutate(
      gene = as.character(gene),
      position = suppressWarnings(as.integer(position)),
      caap_group = as.character(caap_group),
      pattern_type = as.character(pattern_type),
      change_side = tolower(as.character(change_side)),
      change_top = tolower(as.character(change_top)),
      change_bottom = tolower(as.character(change_bottom)),
      parallel_top = normalize_parallel_type(parallel_top),
      parallel_bottom = normalize_parallel_type(parallel_bottom),
      parallel_type = normalize_parallel_type(parallel_type),
      is_conserved_meta = as_bool(is_conserved_meta),
      asr_is_conserved = as_bool(asr_is_conserved),
      asr_root_conserved = as_bool(asr_root_conserved),
      pvalue = suppressWarnings(as.numeric(pvalue)),
      pvalue_boot = suppressWarnings(as.numeric(pvalue_boot)),
      sig_hyp = !is.na(pvalue) & pvalue <= alpha_threshold,
      sig_perm = !is.na(pvalue_boot) & pvalue_boot <= alpha_threshold,
      sig_both = sig_hyp & sig_perm,
      sig_hyp_010 = !is.na(pvalue) & pvalue <= 0.10,
      sig_perm_010 = !is.na(pvalue_boot) & pvalue_boot <= 0.10,
      sig_both_010 = sig_hyp_010 & sig_perm_010,
      sig_hyp_005 = !is.na(pvalue) & pvalue <= 0.05,
      sig_perm_005 = !is.na(pvalue_boot) & pvalue_boot <= 0.05,
      sig_both_005 = sig_hyp_005 & sig_perm_005,
      significance_class = make_significance_class(sig_hyp, sig_perm),
      significance_class_010 = make_significance_class(sig_hyp_010, sig_perm_010),
      significance_class_005 = make_significance_class(sig_hyp_005, sig_perm_005),
      site_id = paste(gene, position, sep = '@')
    )
}

seed_val <- 12345
set.seed(seed_val)

discovery_file <- params$discovery_file
filter_summary_file <- params$filter_summary_file
gene_ensembl_file <- params$gene_ensembl_file
gene_stats_file <- params$gene_stats_file
processing_mode <- params$processing_mode
extreme_threshold <- params$extreme_threshold
iqr_multiplier <- params$iqr_multiplier
alpha_threshold <- suppressWarnings(as.numeric(params$alpha_threshold))
if (is.na(alpha_threshold) || alpha_threshold <= 0 || alpha_threshold > 1) alpha_threshold <- 0.10
alpha_threshold_label <- format_threshold(alpha_threshold)
threshold_line_df <- build_threshold_line_df(alpha_threshold)
gene_filter_mode <- params$gene_filter_mode

if (is.null(discovery_file)) stop("Error: discovery_file parameter is required")
if (is.null(filter_summary_file)) stop("Error: filter_summary_file parameter is required")
if (is.null(gene_ensembl_file)) stop("Error: gene_ensembl_file parameter is required")

stopifnot(file.exists(discovery_file), file.exists(filter_summary_file), file.exists(gene_ensembl_file))

ct_postproc_dir <- "."
outliers_dir <- "outliers"
cluster_dir <- "clusters"
summary_stats_dir <- "summary_statistics"
disambig_dir <- "disambiguation_characterization"
overlap_out_dir <- file.path(disambig_dir, "overlap_figures")

createDir(ct_postproc_dir)
createDir(outliers_dir)
createDir(cluster_dir)
createDir(summary_stats_dir)
createDir(disambig_dir)
createDir(overlap_out_dir)

filter_root <- params$filter_dir
postproc_root <- if (!is.null(filter_root) && !is.na(filter_root)) dirname(filter_root) else NA_character_
removed_precluster_file <- find_first_existing(c(
  file.path(filter_root, "preprocessed", "removed_patterns_precluster.tsv"),
  file.path(postproc_root, "preprocessed", "removed_patterns_precluster.tsv"),
  find_recursive_file(getwd(), "removed_patterns_precluster\\.tsv$")
))
removed_genes_file <- find_first_existing(c(
  file.path(filter_root, "gene_filtering", "removed_genes_summary.tsv"),
  file.path(postproc_root, "gene_filtering", "removed_genes_summary.tsv"),
  find_recursive_file(getwd(), "removed_genes_summary\\.tsv$")
))

tableau_10 <- paletteer_d("ggthemes::Tableau_10")
names(tableau_10) <- c(
  "blue", "orange", "vermillion", "bluish_green",
  "green", "yellow", "reddish_purple", "pink",
  "brown", "gray"
)

threshold_line_colours <- c(
  configured = tableau_10[["orange"]],
  comparison_010 = tableau_10[["blue"]],
  comparison_005 = tableau_10[["gray"]]
)
threshold_line_types <- c(
  configured = "solid",
  comparison_010 = "dashed",
  comparison_005 = "dotdash"
)

category_palette <- c(
  "Normal" = tableau_10[["gray"]],
  "Dubious" = tableau_10[["blue"]],
  "Extreme" = tableau_10[["vermillion"]],
  "Both" = tableau_10[["reddish_purple"]]
)

removal_palette <- c(
  Discarded = tableau_10[["vermillion"]],
  Good = tableau_10[["green"]]
)
```

# A) Post-processing characterization

## Data Loading and Gene Length Annotation

```{r data_loading}
cat("=== CT Post-Processing Configuration ===\n")
cat("Processing mode:", processing_mode, "\n")
cat("Gene filtering:", gene_filter_mode, "\n")
cat("Selected alpha threshold:", alpha_threshold_label, "\n")
cat("Extreme threshold:", extreme_threshold, paste0("(top ", 100*(1-extreme_threshold), "%)"), "\n")
cat("IQR multiplier:", iqr_multiplier, "\n")

cat("\nInput files:\n")
cat("  Discovery:", discovery_file, "\n")
cat("  Filter summary:", filter_summary_file, "\n")
cat("  Gene ensembl:", gene_ensembl_file, "\n")
cat("  Gene stats:", ifelse(is.null(gene_stats_file) || is.na(gene_stats_file), "Not provided (will recompute)", gene_stats_file), "\n\n")

raw_discovery <- read_delim(discovery_file, delim = "\t", show_col_types = FALSE)
discovery_std <- normalize_disambig_schema(raw_discovery)
convergence_plot_df <- discovery_std %>%
  mutate(
    change_side = if_else(change_side %in% c("top", "bottom", "both"), change_side, "none"),
    pattern_parallel = make_pattern_parallel_label(pattern_type, parallel_type),
    pattern_parallel = forcats::fct_infreq(pattern_parallel)
  )
orientation_rows <- build_orientation_rows(convergence_plot_df) %>%
  mutate(caap_group = toupper(caap_group))
orientation_flags <- build_group_flag_table(orientation_rows)

discovery_df <- discovery_std %>%
  transmute(
    CAAP_Group = caap_group,
    Gene = gene,
    Position = position,
    Pattern = pattern_type,
    parallel_type,
    change_side,
    is_conserved_meta,
    asr_is_conserved,
    asr_root_conserved,
    Pvalue = pvalue,
    `Pvalue.boot` = pvalue_boot,
    sig_hyp,
    sig_perm,
    sig_both,
    significance_class,
    site_id
  )

if (!"CAAP_Group" %in% names(discovery_df)) discovery_df <- discovery_df %>% mutate(CAAP_Group = "US")

cat("  Loaded CT positions:", nrow(discovery_df), "\n")
cat("  Unique genes:", n_distinct(discovery_df$Gene), "\n")

gene_ensembl <- read_tsv(gene_ensembl_file, show_col_types = FALSE)
names(gene_ensembl) <- tolower(names(gene_ensembl))
if (!"gene" %in% names(gene_ensembl) || !"length" %in% names(gene_ensembl)) {
  stop("Gene ensembl file must have 'gene' and 'length' columns")
}

gene_length <- gene_ensembl %>% select(Gene = gene, Length = length)
discovery_df <- discovery_df %>% left_join(gene_length, by = "Gene") %>% filter(!is.na(Length))

cat("  Genes with length annotation:", n_distinct(discovery_df$Gene), "\n\n")
```

## Conserved metadata composition after precluster pruning

```{r conserved_meta_composition}
precluster_removed_burden <- tibble()
if (!is.na(removed_precluster_file)) {
  precluster_removed_burden <- read_delim(removed_precluster_file, delim = "\t", show_col_types = FALSE) %>%
    normalize_disambig_schema() %>%
    mutate(RemovalReason = if_else(is.na(RemovalReason), "unknown", RemovalReason))
}

rows_retained <- nrow(discovery_std)
rows_removed_mrca <- sum(str_detect(precluster_removed_burden$RemovalReason, "^mrca_posterior_below_"), na.rm = TRUE)
rows_total <- rows_retained + rows_removed_mrca

conserved_status_burden <- tibble(
  scenario = c("all_positions", "retained_positions", "retained_conserved_meta", "retained_meta_not_conserved", "removed_low_mrca"),
  description = c(
    "Retained plus precluster-removed positions tracked by CT postproc",
    "Positions retained after precluster hard filters",
    "Retained positions where is_conserved_meta = TRUE",
    "Retained positions where is_conserved_meta = FALSE",
    "Precluster removals due to MRCA posterior threshold"
  ),
  rows = c(
    rows_total,
    rows_retained,
    sum(discovery_std$is_conserved_meta, na.rm = TRUE),
    sum(!discovery_std$is_conserved_meta, na.rm = TRUE),
    rows_removed_mrca
  )
) %>%
  mutate(fraction_of_total = if (rows_total > 0) rows / rows_total else NA_real_)

write_tsv(conserved_status_burden, file.path(disambig_dir, "conserved_status_burden.tsv"))

kable(
  conserved_status_burden %>% mutate(fraction_of_total = percent(fraction_of_total, accuracy = 0.1)),
  caption = "Conserved-status burden across retained and precluster-removed CT positions"
) %>% kable_styling()
```

## Outlier Gene Detection

```{r outlier_detection, fig.height=10, fig.width=14}
# Load gene statistics from precomputed file or compute fallback
gene_summary_annotated <- NULL

if (!is.null(gene_stats_file) && !is.na(gene_stats_file) && file.exists(gene_stats_file)) {
  cat("Loading precomputed gene statistics from:", gene_stats_file, "\n")
  gene_summary_annotated <- read_tsv(gene_stats_file, show_col_types = FALSE)
  
  # Ensure required columns exist and rename for compatibility
  if ("CAAP_Group" %in% names(gene_summary_annotated)) {
    colnames(gene_summary_annotated)[colnames(gene_summary_annotated) == "CAAP_Group"] <- "CAAP_Group"
  } else if ("Trait" %in% names(gene_summary_annotated)) {
    gene_summary_annotated$CAAP_Group <- "US"  # default for non-CAAP mode
  }
  
  cat("Loaded gene statistics for", nrow(gene_summary_annotated), "genes\n")

  # Ensure all downstream-required columns exist (file may lack some)
  if (!"CAAP_Group" %in% names(gene_summary_annotated)) {
    gene_summary_annotated$CAAP_Group <- "US"
  }
  if (!"n_CAAS_per_length" %in% names(gene_summary_annotated)) {
    if ("n_CAAS" %in% names(gene_summary_annotated) && "Length" %in% names(gene_summary_annotated)) {
      gene_summary_annotated$n_CAAS_per_length <- gene_summary_annotated$n_CAAS / gene_summary_annotated$Length * 100
    } else {
      gene_summary_annotated$n_CAAS_per_length <- NA_real_
    }
  }
  if (!"Extreme" %in% names(gene_summary_annotated))          gene_summary_annotated$Extreme          <- FALSE
  if (!"Dubious" %in% names(gene_summary_annotated))          gene_summary_annotated$Dubious          <- FALSE
  if (!"Category" %in% names(gene_summary_annotated))         gene_summary_annotated$Category         <- "Normal"
  if (!"threshold_extreme" %in% names(gene_summary_annotated)) gene_summary_annotated$threshold_extreme <- NA_real_
  if (!"threshold_dubious" %in% names(gene_summary_annotated)) gene_summary_annotated$threshold_dubious <- NA_real_
  
} else {
  cat("Gene statistics file not found or not provided\n")
}

# Ensure CAAP_Group column exists for plotting
if (!"CAAP_Group" %in% names(gene_summary_annotated)) {
  gene_summary_annotated$CAAP_Group <- "US"
}

cat("Groups found:", paste(sort(unique(gene_summary_annotated$CAAP_Group)), collapse=", "), "\n")

if (nrow(gene_summary_annotated) == 0) {
  cat("WARNING: gene_summary_annotated is empty — skipping outlier distribution plots.\n")
} else {

# Pre-compute vline data — guard against empty frames that break facet_wrap
vline_dubious_data  <- gene_summary_annotated %>% filter(!is.na(threshold_dubious))  %>% distinct(CAAP_Group, threshold_dubious)
vline_extreme_data  <- gene_summary_annotated %>% filter(!is.na(threshold_extreme)) %>% distinct(CAAP_Group, threshold_extreme)

# Generate histograms with increased font sizes for slides
p_gross_dist <- ggplot(gene_summary_annotated, aes(x = n_CAAS, fill = Category)) +
  geom_histogram(binwidth = 0.05, color = "black", alpha = 0.8) +
  scale_fill_manual(values = category_palette, breaks = c("Extreme", "Dubious", "Both", "Normal")) +
  scale_x_log10(breaks = c(1, 2, 5, 10, 20, 50, 100, 200)) +
  facet_wrap(~CAAP_Group, scales = "free_y") +
  labs(title = "CAAP Count Distribution by Group",
       subtitle = "Dubious threshold (Q3 + 3×IQR) shown as dashed line",
       x = "Number of CAAP Positions", y = "Gene Count") +
  theme_minimal(base_size = 16) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 14),
    axis.title = element_text(size = 15),
    strip.text = element_text(face = "bold", size = 15)
  )
if (nrow(vline_dubious_data) > 0) {
  p_gross_dist <- p_gross_dist +
    geom_vline(data = vline_dubious_data, aes(xintercept = threshold_dubious),
               color = tableau_10[["blue"]], linetype = "dashed", linewidth = 1)
}

gross_zoom_upper <- derive_plot_zoom_upper(p_gross_dist, zoom_fraction = 0.10, min_upper = 5)
p_zoom_gross <- p_gross_dist + coord_cartesian(ylim = c(0, gross_zoom_upper)) +
  labs(title = paste0("Zoomed View (bottom 10% of y-range; y ≤ ", gross_zoom_upper, ")"))

p_density_dist <- ggplot(gene_summary_annotated, aes(x = n_CAAS_per_length, fill = Category)) +
  geom_histogram(binwidth = 0.05, color = "black", alpha = 0.8) +
  scale_fill_manual(values = category_palette, breaks = c("Extreme", "Dubious", "Both", "Normal")) +
  scale_x_log10(breaks = c(0.01, 0.02, 0.05, 0.1, 0.2, 0.5, 1, 2, 5, 10, 20, 50, 100)) +
  facet_wrap(~ CAAP_Group, scales = "free_y") +
  labs(title = "CAAP Density Distribution by Group",
       subtitle = paste0("Top ", 100*(1-extreme_threshold), "% density threshold shown as dashed line"),
       x = "CAAP Density (positions / gene length)", y = "Gene Count") +
  theme_minimal(base_size = 16) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 14),
    axis.title = element_text(size = 15),
    strip.text = element_text(face = "bold", size = 15)
  )
if (nrow(vline_extreme_data) > 0) {
  p_density_dist <- p_density_dist +
    geom_vline(data = vline_extreme_data, aes(xintercept = threshold_extreme),
               color = tableau_10[["vermillion"]], linetype = "dashed", linewidth = 1)
}

density_zoom_upper <- derive_plot_zoom_upper(p_density_dist, zoom_fraction = 0.10, min_upper = 5)
p_zoom_density <- p_density_dist + coord_cartesian(ylim = c(0, density_zoom_upper)) +
  labs(title = paste0("Zoomed View (bottom 10% of y-range; y ≤ ", density_zoom_upper, ")"))

combined_outlier_plot <- p_gross_dist / p_density_dist
combined_zoomed_plot <- p_zoom_gross / p_zoom_density

print(combined_outlier_plot)
print(combined_zoomed_plot)

facet_count_outliers <- n_distinct(gene_summary_annotated$CAAP_Group)
outlier_dims <- compute_barplot_dims(
  n_labels = 8,
  n_facets = facet_count_outliers,
  facet_cols = min(3, facet_count_outliers),
  base_width = 12,
  base_height = 8,
  per_label = 0.35,
  facet_row_height = 1.8,
  max_width = 20,
  max_height = 14
)
ggsave(file.path(outliers_dir, "outlier_distributions.png"), combined_outlier_plot,
       width = outlier_dims$width, height = outlier_dims$height, dpi = 320)
ggsave(file.path(outliers_dir, "outlier_distributions_zoomed.png"), combined_zoomed_plot,
       width = outlier_dims$width, height = outlier_dims$height, dpi = 320)

# Dedicated bar chart: removed gene counts per category and CAAP_Group
# Clearly separates Extreme-only, Dubious-only, and Both (genes failing both criteria).
removed_cat_summary <- gene_summary_annotated %>%
  filter(Category != "Normal") %>%
  count(CAAP_Group, Category)

if (nrow(removed_cat_summary) > 0) {
  p_removed_cats <- ggplot(removed_cat_summary,
                           aes(x = Category, y = n, fill = Category)) +
    geom_col(color = "black", linewidth = 0.3, width = 0.7) +
    facet_wrap(~ CAAP_Group, scales = "free_y", ncol = min(3, n_distinct(removed_cat_summary$CAAP_Group))) +
    scale_fill_manual(
      values = category_palette,
      breaks = c("Extreme", "Dubious", "Both")
    ) +
    labs(
      title    = "Removed Genes by Category and Group",
      subtitle = paste0(
        "Extreme: top ", 100 * (1 - extreme_threshold), "% density  |  ",
        "Dubious: Q3 + ", iqr_multiplier, "\u00d7IQR count  |  ",
        "Both: both criteria"
      ),
      x = "Removal Category",
      y = "Gene Count"
    ) +
    theme_minimal(base_size = 16) +
    theme(
      legend.position  = "none",
      plot.title       = element_text(face = "bold", size = 18),
      plot.subtitle    = element_text(size = 14),
      axis.title       = element_text(size = 15),
      axis.text.x      = element_text(angle = 25, hjust = 1),
      strip.text       = element_text(face = "bold", size = 15)
    )

  print(p_removed_cats)
  save_barplot(
    file.path(outliers_dir, "removed_genes_by_category.png"),
    p_removed_cats,
    n_labels = n_distinct(removed_cat_summary$Category),
    n_facets = n_distinct(removed_cat_summary$CAAP_Group),
    facet_cols = min(3, n_distinct(removed_cat_summary$CAAP_Group)),
    base_width = 9,
    base_height = 5.5
  )
}

cat("\n=== Outlier Detection Summary ===\n")
cat("CAAP/US unified mode: Group-aware thresholds calculated\n\n")

thresholds_summary <- gene_summary_annotated %>% distinct(CAAP_Group, threshold_extreme, threshold_dubious) %>% arrange(CAAP_Group)
cat("Thresholds by group:\n")
for (i in 1:nrow(thresholds_summary)) {
  cat(sprintf("  %s:\n", thresholds_summary$CAAP_Group[i]))
  cat(sprintf("    Extreme (top %.0f%% density): %.4f\n", 100*(1-extreme_threshold), thresholds_summary$threshold_extreme[i]))
  cat(sprintf("    Dubious (Q3 + %.1f×IQR): %.0f positions\n\n", iqr_multiplier, thresholds_summary$threshold_dubious[i]))
}

outlier_counts <- gene_summary_annotated %>% count(CAAP_Group, Category) %>% pivot_wider(names_from = Category, values_from = n, values_fill = 0)
cat("Gene counts by group and category:\n")
print(outlier_counts)

write_tsv(gene_summary_annotated, file.path(outliers_dir, "gene_summary_annotated.tsv"))
cat("\n✓ Gene summary exported\n\n")

} # end if (nrow(gene_summary_annotated) > 0)
```

## Cluster Filtering Analysis

```{r cluster_analysis_load}
cat("Loading cluster filter summary...\n")
filter_summary <- read_tsv(filter_summary_file, show_col_types = FALSE)
cat("  Parameter combinations tested:", nrow(filter_summary), "\n")

if (!"Minlen" %in% names(filter_summary) || !"Maxcaas" %in% names(filter_summary)) {
  if ("Parameter" %in% names(filter_summary)) {
    parsed_params <- stringr::str_split_fixed(as.character(filter_summary$Parameter), "/", 2)
    if (!"Minlen" %in% names(filter_summary)) filter_summary$Minlen <- suppressWarnings(as.integer(parsed_params[, 1]))
    if (!"Maxcaas" %in% names(filter_summary)) filter_summary$Maxcaas <- suppressWarnings(as.numeric(parsed_params[, 2]))
    cat("  Parsed missing Minlen/Maxcaas from Parameter column\n")
  }
}

if (!"DiscardedCount" %in% names(filter_summary)) {
  protected_cols <- c("Parameter", "Minlen", "Maxcaas", "CAAP_Groups", "Groups_List", "N_Groups")
  candidate_cols <- setdiff(names(filter_summary), protected_cols)
  numeric_candidate_cols <- candidate_cols[vapply(candidate_cols, function(col) is.numeric(filter_summary[[col]]), logical(1))]
  if (length(numeric_candidate_cols) > 0) {
    filter_summary <- filter_summary %>% mutate(DiscardedCount = rowSums(across(all_of(numeric_candidate_cols)), na.rm = TRUE))
    cat("  Built DiscardedCount from columns:", paste(numeric_candidate_cols, collapse = ", "), "\n")
  } else {
    filter_summary$DiscardedCount <- 0L
    cat("  Warning: Could not infer DiscardedCount from summary columns; defaulting to 0\n")
  }
} else {
  filter_summary$DiscardedCount <- suppressWarnings(as.numeric(filter_summary$DiscardedCount))
  filter_summary$DiscardedCount[is.na(filter_summary$DiscardedCount)] <- 0
}

has_caap_groups_in_summary <- 'CAAP_Groups' %in% names(filter_summary) && any(filter_summary$CAAP_Groups != 'N/A' & filter_summary$CAAP_Groups != '')
if (has_caap_groups_in_summary) {
  cat("  CAAP mode detected in filter summary\n")
  filter_summary <- filter_summary %>% mutate(Groups_List = strsplit(as.character(CAAP_Groups), ","), N_Groups = lengths(Groups_List))
}

if (has_caap_groups_in_summary) {
  group_cols <- setdiff(names(filter_summary), c('Parameter', 'Minlen', 'Maxcaas'))
  filter_table <- filter_summary %>% select(Parameter, all_of(sort(group_cols)))
} else {
  filter_table <- filter_summary %>% select(Parameter, DiscardedCount)
  names(filter_table)[2] <- "Discarded"
}

kable(filter_table, caption = "Cluster Filtering Parameter Sweep Results", align = c("l", rep("r", ncol(filter_table) - 1))) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

if (nrow(filter_summary) > 1) {
  if (has_caap_groups_in_summary) {
    group_cols <- setdiff(names(filter_summary), c('Parameter', 'Minlen', 'Maxcaas'))
    plot_data <- filter_summary %>%
      pivot_longer(cols = all_of(group_cols), names_to = "CAAP_Group", values_to = "DiscardedCount") %>%
      mutate(param_label = sprintf("%d/%d", Minlen, as.integer(Maxcaas)), param_label = factor(param_label, levels = unique(param_label)))

    p_param_sweep <- ggplot(plot_data, aes(x = param_label, y = DiscardedCount, fill = CAAP_Group)) +
      geom_col(position = "dodge", color = "black", linewidth = 0.3) +
      scale_fill_paletteer_d("ggthemes::Tableau_10") +
      labs(title = "Discarded CAAS Positions by Parameter Combination",
           subtitle = "Grouped by CAAP group - minlen/maxcaas combinations",
           x = "Parameter Combination (minlen/maxcaas)", y = "Positions Discarded", fill = "CAAP Group") +
      theme_minimal(base_size = 16) +
      theme(legend.position = "right", plot.title = element_text(face = "bold", size = 18), axis.title = element_text(size = 15), axis.text.x = element_text(angle = 55, hjust = 1))
  } else {
    p_param_sweep <- ggplot(filter_summary, aes(x = factor(Minlen), y = DiscardedCount, color = factor(Maxcaas), group = factor(Maxcaas))) +
      geom_point(size = 4) +
      geom_line(linewidth = 1.2) +
      scale_color_paletteer_d("ggthemes::Tableau_10", labels = function(x) sprintf("%.0f%%", as.numeric(x))) +
      labs(title = "Discarded CAAS Positions by Parameter Combination",
           subtitle = "Effect of minimum cluster length and maximum density threshold",
           x = "Minimum Cluster Length (minlen)", y = "Positions Discarded", color = "Max Density") +
      theme_minimal(base_size = 16) +
      theme(legend.position = "right", plot.title = element_text(face = "bold", size = 18), axis.title = element_text(size = 15))
  }

  print(p_param_sweep)
  if (has_caap_groups_in_summary) {
    save_barplot(
      file.path(cluster_dir, "parameter_sweep_overview.png"),
      p_param_sweep,
      n_labels = n_distinct(plot_data$param_label),
      n_facets = 1,
      facet_cols = 1,
      base_width = 10,
      base_height = 6,
      per_label = 0.65
    )
  } else {
    ggsave(file.path(cluster_dir, "parameter_sweep_overview.png"), p_param_sweep, width = 12, height = 6, dpi = 320)
  }
}

write_tsv(filter_summary, file.path(cluster_dir, "filter_summary.tsv"))
cat("✓ Filter summary exported\n\n")
```

```{r cluster_detailed_analysis, results='asis', fig.height=8, fig.width=12, eval=(processing_mode == 'exploratory' && nrow(filter_summary) > 1)}
if (processing_mode == 'exploratory' && nrow(filter_summary) > 1) {
  cat("\n## Detailed Parameter Analysis\n\n")
  detail_filter_root <- params$filter_dir
  detail_postproc_root <- if (!is.null(detail_filter_root) && !is.na(detail_filter_root)) dirname(detail_filter_root) else NA_character_

  for (i in 1:nrow(filter_summary)) {
    param_row <- filter_summary[i, ]
    minlen_i <- suppressWarnings(as.integer(param_row$Minlen))
    maxcaas_i <- suppressWarnings(as.integer(param_row$Maxcaas))
    param_label <- sprintf("minlen=%d, maxcaas=%d", minlen_i, maxcaas_i)
    cat(sprintf("\n### Parameter Set %d: %s {.tabset}\n\n", i, param_label))

    resolved_file <- resolve_filtered_cluster_file(
      minlen_i,
      maxcaas_i,
      filter_root = detail_filter_root,
      postproc_root = detail_postproc_root
    )
    cat(sprintf("Resolved filtered file status: %s\n\n", resolved_file$message))

    if (is.na(resolved_file$path) || !file.exists(resolved_file$path)) {
      cat("No per-parameter file was available for this combination, so no spatial summary could be rendered.\n")
      cat("\n---\n\n")
      next
    }

    filtered_data <- read_tsv(resolved_file$path, show_col_types = FALSE)
    if (!"ClusteringFlag" %in% names(filtered_data)) {
      cat("Matched file does not contain `ClusteringFlag`; skipping this parameter set.\n")
      cat("\n---\n\n")
      next
    }

    discarded <- filtered_data %>% filter(ClusteringFlag == "Discarded")
    if (nrow(discarded) == 0) {
      cat("No discarded clusters were found for this parameter set.\n")
      cat("\n---\n\n")
      next
    }

    if (!"CAAP_Group" %in% names(discarded)) {
      discarded <- discarded %>% mutate(CAAP_Group = "US")
    }

    detect_runs <- function(positions) {
      positions <- sort(unique(positions))
      if (length(positions) == 0) return(integer(0))
      breaks <- c(0, which(diff(positions) > 1), length(positions))
      diff(breaks)
    }

    spatial_analysis <- discarded %>%
      group_by(CAAP_Group, Gene) %>%
      summarise(positions = list(Position), .groups = "drop") %>%
      mutate(run_lengths = map(positions, detect_runs)) %>%
      unnest(run_lengths, keep_empty = TRUE) %>%
      mutate(
        cluster_type = case_when(
          run_lengths == 1 ~ "Singlet (1)",
          run_lengths == 2 ~ "Doublet (2)",
          run_lengths == 3 ~ "Triplet (3)",
          run_lengths == 4 ~ "Quad (4)",
          run_lengths == 5 ~ "Quint (5)",
          run_lengths >= 6 ~ "6+ Long",
          TRUE ~ "Other"
        )
      )

    spatial_summary <- spatial_analysis %>%
      count(CAAP_Group, cluster_type) %>%
      complete(CAAP_Group, cluster_type, fill = list(n = 0)) %>%
      group_by(CAAP_Group) %>%
      mutate(percentage = round(100 * n / sum(n), 1)) %>%
      ungroup()

    p_spatial <- ggplot(spatial_summary, aes(x = CAAP_Group, y = percentage / 100, fill = cluster_type)) +
      geom_col(position = "fill", color = "black", size = 0.2) +
      scale_fill_paletteer_d("ggthemes::Tableau_10") +
      scale_y_continuous(labels = scales::percent) +
      labs(
        title = sprintf("Spatial Clustering: %s", param_label),
        subtitle = sprintf("Discarded positions only | %s", resolved_file$message),
        x = "Group",
        y = "Percentage",
        fill = "Cluster Type"
      ) +
      theme_minimal(base_size = 16) +
      theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1))

    print(p_spatial)
    cat("\n---\n\n")
  }
}
```

# B) Characterizing CAAS disambiguation

## B.1 Build unified removal + retention dataset

```{r disambig_unified_data}
cluster_files <- list.files(path = params$filter_dir, pattern = "\\.filtered\\.minlen[0-9]+\\.maxcaas[0-9]+\\.tsv$", full.names = TRUE, recursive = TRUE)
if (length(cluster_files) == 0) {
  cluster_files <- list.files(path = ifelse(is.na(postproc_root), getwd(), postproc_root), pattern = "\\.filtered\\.minlen[0-9]+\\.maxcaas[0-9]+\\.tsv$", full.names = TRUE, recursive = TRUE)
}

precluster_events <- tibble()
if (!is.na(removed_precluster_file)) {
  precluster_df <- normalize_disambig_schema(read_delim(removed_precluster_file, delim = "\t", show_col_types = FALSE))
  precluster_events <- precluster_df %>% transmute(
    site_id, gene, position, caap_group, pattern_type, pvalue, pvalue_boot,
    change_top, change_bottom, change_side, parallel_top, parallel_bottom, parallel_type,
    sig_hyp, sig_perm, sig_both, significance_class,
    RemovalStage = case_when(
      str_detect(RemovalReason, "^mrca_posterior_below_") ~ "precluster_mrca_threshold",
      TRUE ~ "precluster_other"
    )
  )
}

cluster_events <- tibble()
if (length(cluster_files) > 0) {
  cluster_df <- map_dfr(cluster_files, ~ read_tsv(.x, show_col_types = FALSE))
  cluster_hits <- cluster_df %>% filter(ClusteringFlag == "Discarded") %>% transmute(gene = as.character(Gene), position = as.integer(Position), caap_group = as.character(CAAP_Group)) %>% distinct()
  cluster_events <- discovery_std %>%
    inner_join(cluster_hits, by = c("gene", "position", "caap_group")) %>%
    transmute(
      site_id, gene, position, caap_group, pattern_type, pvalue, pvalue_boot,
      change_top, change_bottom, change_side, parallel_top, parallel_bottom, parallel_type,
      sig_hyp, sig_perm, sig_both, significance_class,
      RemovalStage = "cluster"
    ) %>%
    distinct()
}

gene_events <- tibble()
if (!is.na(removed_genes_file)) {
  rg <- read_tsv(removed_genes_file, show_col_types = FALSE)
  names(rg) <- tolower(names(rg))
  if (!"caap_group" %in% names(rg)) rg$caap_group <- "US"
  rg_norm <- rg %>% transmute(gene = as.character(gene), caap_group = as.character(caap_group), category = as.character(category)) %>% distinct()

  gene_events <- discovery_std %>%
    inner_join(rg_norm, by = c("gene", "caap_group")) %>%
    mutate(RemovalStage = case_when(
      tolower(category) == "extreme" ~ "extreme_gene",
      tolower(category) == "dubious" ~ "dubious_with_clusters",
      tolower(category) == "both" ~ "extreme_and_dubious",
      TRUE ~ "gene_filter"
    )) %>%
    transmute(
      site_id, gene, position, caap_group, pattern_type, pvalue, pvalue_boot,
      change_top, change_bottom, change_side, parallel_top, parallel_bottom, parallel_type,
      sig_hyp, sig_perm, sig_both, significance_class, RemovalStage
    ) %>%
    distinct()
}

removal_events <- bind_rows(precluster_events, cluster_events, gene_events)
write_tsv(removal_events, file.path(disambig_dir, "removal_events.tsv"))

```

## B.1b Focal relationships

```{r focal_sets, results='asis'}
focal_source <- orientation_rows %>%
  transmute(
    orientation = as.character(orientation),
    site_id,
    gene,
    position,
    caap_group,
    change_side,
    pattern_type,
    parallel_type,
    pattern_parallel
  ) %>%
  distinct()

build_focal_table <- function(df, orientation_name) {
  orientation_df <- df %>% filter(orientation == orientation_name)
  if (nrow(orientation_df) == 0) {
    return(tibble(
      Scope = character(),
      CAAP_Group = character(),
      Pattern = character(),
      Parallel = character(),
      PatternParallel = character(),
      Positions_total = integer(),
      Genes_total = integer()
    ))
  }

  overall_tbl <- orientation_df %>%
    group_by(pattern_type, parallel_type, pattern_parallel) %>%
    summarise(
      Positions_total = n_distinct(site_id),
      Genes_total = n_distinct(gene),
      .groups = "drop"
    ) %>%
    mutate(Scope = "overall", CAAP_Group = "ALL")

  per_group_tbl <- orientation_df %>%
    group_by(caap_group, pattern_type, parallel_type, pattern_parallel) %>%
    summarise(
      Positions_total = n_distinct(site_id),
      Genes_total = n_distinct(gene),
      .groups = "drop"
    ) %>%
    mutate(Scope = "per_group", CAAP_Group = caap_group)

  bind_rows(overall_tbl, per_group_tbl) %>%
    transmute(
      Scope,
      CAAP_Group,
      Pattern = pattern_type,
      Parallel = parallel_type,
      PatternParallel = pattern_parallel,
      Positions_total,
      Genes_total
    ) %>%
    arrange(factor(Scope, levels = c("overall", "per_group")), CAAP_Group, desc(Positions_total), PatternParallel)
}

for (orientation_name in c("global", "top", "bottom", "both")) {
  tbl <- build_focal_table(focal_source, orientation_name)
  cat(sprintf("\n### %s\n\n", tools::toTitleCase(orientation_name)))
  print(kable(tbl, caption = paste("Focal sets —", orientation_name)) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed")))
  write_tsv(tbl, file.path(disambig_dir, paste0("focal_sets_", orientation_name, ".tsv")))
}

cat("✓ Focal sets tables exported\n")

```

## B.1c Gene-Level Filter Overlap Diagnostics

```{r filter_overlap_gene_table}
# ─────────────────────────────────────────────────────────────────────────────
# Build a gene-level membership table that records for each gene whether it
# was flagged by each of the three independent filters:
#   • Extreme   – top 1% CAAS density (gene_summary_annotated$Extreme)
#   • Dubious   – IQR-outlier CAAS count (gene_summary_annotated$Dubious)
#   • Cluster   – at least one position removed by the cluster filter
# ─────────────────────────────────────────────────────────────────────────────

# --- 1. Gene-level filter flags -------------------------------------------

# Extreme / Dubious come straight from the already-computed gene_summary_annotated
gene_filter_flags <- gene_summary_annotated %>%
  transmute(
    Gene        = Gene,
    CAAP_Group  = CAAP_Group,
    n_CAAS      = n_CAAS,
    Length      = Length,
    n_CAAS_per_length = n_CAAS_per_length,
    Extreme     = as.logical(Extreme),
    Dubious     = as.logical(Dubious)
  )

# Cluster_removed: any gene that contributed at least one site to cluster_events
cluster_removed_genes <- if (nrow(cluster_events) > 0) {
  cluster_events %>%
    transmute(Gene = gene, CAAP_Group = caap_group) %>%
    distinct() %>%
    mutate(Cluster_removed = TRUE)
} else {
  tibble(Gene = character(), CAAP_Group = character(), Cluster_removed = logical())
}

gene_filter_flags <- gene_filter_flags %>%
  left_join(cluster_removed_genes, by = c("Gene", "CAAP_Group")) %>%
  mutate(
    Cluster_removed = coalesce(Cluster_removed, FALSE),
    # Composite label
    filter_combo = case_when(
      Extreme & Dubious & Cluster_removed ~ "Extreme+Dubious+Cluster",
      Extreme & Dubious                  ~ "Extreme+Dubious",
      Extreme & Cluster_removed          ~ "Extreme+Cluster",
      Dubious & Cluster_removed          ~ "Dubious+Cluster",
      Extreme                            ~ "Extreme",
      Dubious                            ~ "Dubious",
      Cluster_removed                    ~ "Cluster",
      TRUE                               ~ "None"
    ),
    filter_combo = factor(filter_combo, levels = c(
      "None", "Extreme", "Dubious", "Cluster",
      "Extreme+Dubious", "Extreme+Cluster", "Dubious+Cluster",
      "Extreme+Dubious+Cluster"
    ))
  )

# --- 2. Per-gene significance descriptors from full discovery data ----------

gene_sig_stats <- discovery_std %>%
  group_by(gene, caap_group) %>%
  summarise(
    frac_sig_both    = mean(sig_both, na.rm = TRUE),
    median_pval_hyp  = median(pvalue,      na.rm = TRUE),
    min_pval_hyp     = min(pvalue,         na.rm = TRUE),
    median_pval_boot = median(pvalue_boot, na.rm = TRUE),
    min_pval_boot    = min(pvalue_boot,    na.rm = TRUE),
    .groups = "drop"
  ) %>%
  rename(Gene = gene, CAAP_Group = caap_group)

gene_filter_flags <- gene_filter_flags %>%
  left_join(gene_sig_stats, by = c("Gene", "CAAP_Group"))

# --- 3. Quick membership summary printed to console ------------------------

cat("\n=== Gene-Level Filter Overlap Summary ===\n")
membership_counts <- gene_filter_flags %>%
  count(CAAP_Group, filter_combo, .drop = FALSE) %>%
  arrange(CAAP_Group, filter_combo)
print(membership_counts)

cat("\nFilter-specific totals:\n")
cat(sprintf("  Extreme only or combined : %d genes\n",
    sum(gene_filter_flags$Extreme, na.rm = TRUE)))
cat(sprintf("  Dubious only or combined : %d genes\n",
    sum(gene_filter_flags$Dubious, na.rm = TRUE)))
cat(sprintf("  Cluster only or combined : %d genes\n",
    sum(gene_filter_flags$Cluster_removed, na.rm = TRUE)))
cat(sprintf("  Any filter               : %d / %d genes (%.1f%%)\n",
    sum(gene_filter_flags$filter_combo != "None"),
    nrow(gene_filter_flags),
    100 * mean(gene_filter_flags$filter_combo != "None")))

# Fraction of sig-both sites removed by each filter pathway
if (nrow(removal_events) > 0) {
  total_sig_both  <- sum(discovery_std$sig_both, na.rm = TRUE)
  removed_extreme <- discovery_std %>%
    filter(sig_both, gene %in% filter(gene_filter_flags, Extreme)$Gene,
           caap_group %in% filter(gene_filter_flags, Extreme)$CAAP_Group)
  removed_dubious <- discovery_std %>%
    filter(sig_both, gene %in% filter(gene_filter_flags, Dubious)$Gene,
           caap_group %in% filter(gene_filter_flags, Dubious)$CAAP_Group)
  removed_cluster <- discovery_std %>%
    filter(sig_both, site_id %in% cluster_events$site_id)
  cat(sprintf("\n  Sig-both sites in total  : %d\n", total_sig_both))
  cat(sprintf("  Sig-both in Extreme genes: %d (%.1f%%)\n",
      nrow(removed_extreme), 100 * nrow(removed_extreme) / max(total_sig_both, 1)))
  cat(sprintf("  Sig-both in Dubious genes: %d (%.1f%%)\n",
      nrow(removed_dubious), 100 * nrow(removed_dubious) / max(total_sig_both, 1)))
  cat(sprintf("  Sig-both in Cluster sites: %d (%.1f%%)\n",
      nrow(removed_cluster), 100 * nrow(removed_cluster) / max(total_sig_both, 1)))
}

cat("\n")
```

## B.1c-i UpSet plot of filter intersections

```{r filter_overlap_upset, fig.width=10, fig.height=6}
# Build a binary incidence data frame suitable for UpSetR
upset_df <- gene_filter_flags %>%
  transmute(
    Extreme  = as.integer(Extreme),
    Dubious  = as.integer(Dubious),
    Cluster  = as.integer(Cluster_removed)
  ) %>%
  as.data.frame()

# Guard: only proceed if at least two filters have any hits
n_active_filters <- sum(c(
  any(upset_df$Extreme == 1),
  any(upset_df$Dubious == 1),
  any(upset_df$Cluster == 1)
))

if (n_active_filters < 1) {
  cat("No genes flagged by any filter — UpSet plot skipped.\n")
} else if (!requireNamespace("UpSetR", quietly = TRUE)) {
  cat("Package 'UpSetR' not available — falling back to membership table.\n")
  kable(membership_counts, caption = "Gene counts per filter combination") %>% kable_styling()
} else {
  # One UpSet per CAAP group (or single global plot when only one group)
  groups_present <- sort(unique(gene_filter_flags$CAAP_Group))

  for (grp in groups_present) {
    grp_df <- gene_filter_flags %>%
      filter(CAAP_Group == grp) %>%
      transmute(
        Extreme = as.integer(Extreme),
        Dubious = as.integer(Dubious),
        Cluster = as.integer(Cluster_removed)
      ) %>%
      as.data.frame()

    # Drop columns with zero variance (no genes hit by that filter in this group)
    active_cols <- names(grp_df)[colSums(grp_df, na.rm = TRUE) > 0]
    if (length(active_cols) < 1) {
      cat(sprintf("  Group %s: no filter hits — skipping.\n", grp))
      next
    }
    grp_df <- grp_df[, active_cols, drop = FALSE]

    plot_title <- sprintf("Filter overlap — %s (n = %d genes)",
                          grp, nrow(grp_df))
    cat(sprintf("\n**%s**\n\n", plot_title))

    # UpSetR::upset returns a grid object; use grid::grid.draw inside a device call
    png_path <- file.path(outliers_dir,
                          sprintf("filter_overlap_upset_%s.png", tolower(grp)))

    # Render to file
    png(png_path, width = 2800, height = 1600, res = 320)
    tryCatch(
      print(UpSetR::upset(
        grp_df,
        sets            = rev(active_cols),
        order.by        = "freq",
        decreasing      = TRUE,
        text.scale      = c(1.4, 1.2, 1.2, 1.0, 1.2, 1.0),
        mainbar.y.label = "Gene Intersections",
        sets.x.label    = "Genes per filter",
        mb.ratio        = c(0.60, 0.40),
        main.bar.color  = "#4472C4",
        sets.bar.color  = "#ED7D31"
      )),
      error = function(e) cat("UpSet render error:", conditionMessage(e), "\n")
    )
    dev.off()

    # Also render inline
    tryCatch(
      print(UpSetR::upset(
        grp_df,
        sets            = rev(active_cols),
        order.by        = "freq",
        decreasing      = TRUE,
        text.scale      = c(1.4, 1.2, 1.2, 1.0, 1.2, 1.0),
        mainbar.y.label = "Gene Intersections",
        sets.x.label    = "Genes per filter",
        mb.ratio        = c(0.60, 0.40),
        main.bar.color  = "#4472C4",
        sets.bar.color  = "#ED7D31"
      )),
      error = function(e) cat("UpSet inline error:", conditionMessage(e), "\n")
    )
  }
}
```

## B.1c-ii Descriptor distributions by filter membership

```{r filter_overlap_stats, fig.width=14, fig.height=16}
# Ordered factor for x-axis: None → single → double → triple
combo_levels <- c(
  "None", "Extreme", "Dubious", "Cluster",
  "Extreme+Dubious", "Extreme+Cluster", "Dubious+Cluster",
  "Extreme+Dubious+Cluster"
)

gene_filter_plot <- gene_filter_flags %>%
  mutate(filter_combo = factor(as.character(filter_combo), levels = combo_levels))

# Colour palette for combos
combo_palette <- setNames(
  c(
    tableau_10[["gray"]],
    tableau_10[["vermillion"]], tableau_10[["blue"]], tableau_10[["green"]],
    tableau_10[["reddish_purple"]], tableau_10[["orange"]], tableau_10[["bluish_green"]],
    tableau_10[["pink"]]
  ),
  combo_levels
)

make_descriptor_plot <- function(data, y_var, y_label, log_scale = FALSE) {
  plot_data <- data %>% filter(!is.na(.data[[y_var]]), !is.na(CAAP_Group))
  if (nrow(plot_data) == 0 || length(unique(plot_data$CAAP_Group)) == 0) {
    return(
      ggplot() +
        annotate("text", x = 0.5, y = 0.5, label = paste("No data for:", y_label),
                 size = 5, hjust = 0.5) +
        theme_void()
    )
  }
  p <- ggplot(plot_data,
              aes(x = filter_combo, y = .data[[y_var]],
                  fill = filter_combo, colour = filter_combo)) +
    geom_boxplot(alpha = 0.5, outlier.shape = NA, linewidth = 0.6) +
    geom_jitter(width = 0.18, alpha = 0.35, size = 0.9, shape = 16) +
    facet_wrap(~ CAAP_Group, scales = "free_y") +
    scale_fill_manual(values   = combo_palette, drop = TRUE) +
    scale_colour_manual(values = combo_palette, drop = TRUE) +
    labs(x = NULL, y = y_label) +
    theme_minimal(base_size = 12) +
    theme(
      legend.position  = "none",
      axis.text.x      = element_text(angle = 40, hjust = 1, size = 9),
      strip.text       = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
  if (log_scale) p <- p + scale_y_log10()
  p
}

p_length     <- make_descriptor_plot(gene_filter_plot, "Length",           "Gene length (aa)",              log_scale = TRUE)
p_ncaas      <- make_descriptor_plot(gene_filter_plot, "n_CAAS",           "Total CAAS count",              log_scale = TRUE)
p_density    <- make_descriptor_plot(gene_filter_plot, "n_CAAS_per_length","CAAS density (per 100 aa)",     log_scale = TRUE)
p_fracsig    <- make_descriptor_plot(gene_filter_plot, "frac_sig_both",    "Fraction sig-both sites",       log_scale = FALSE)
p_medpval    <- make_descriptor_plot(gene_filter_plot, "median_pval_hyp",  "Median hyperg. p-value",        log_scale = FALSE)
p_minpval    <- make_descriptor_plot(gene_filter_plot, "min_pval_hyp",     "Min hyperg. p-value",           log_scale = FALSE)

desc_grid <- (
  p_length  | p_ncaas  | p_density
) / (
  p_fracsig | p_medpval | p_minpval
) +
  plot_annotation(
    title    = "Gene descriptor distributions by filter membership",
    subtitle = sprintf(
      "Extreme = top %.0f%% density  |  Dubious = Q3 + %.1f×IQR count  |  Cluster = cluster-filtered",
      100 * (1 - extreme_threshold), iqr_multiplier
    ),
    theme = theme(plot.title = element_text(face = "bold", size = 16),
                  plot.subtitle = element_text(size = 12))
  )

print(desc_grid)
ggsave(file.path(outliers_dir, "filter_overlap_descriptors.png"),
       desc_grid, width = 14, height = 10, dpi = 280)
```

## B.1c-iii Filter overlap summary table

```{r filter_overlap_summary_table}
# Per-combo × per-group mean ± SD for all five descriptors
overlap_summary <- gene_filter_flags %>%
  group_by(CAAP_Group, filter_combo) %>%
  summarise(
    n_genes            = n(),
    mean_Length        = round(mean(Length,            na.rm = TRUE), 1),
    sd_Length          = round(sd(Length,              na.rm = TRUE), 1),
    mean_n_CAAS        = round(mean(n_CAAS,            na.rm = TRUE), 2),
    sd_n_CAAS          = round(sd(n_CAAS,              na.rm = TRUE), 2),
    mean_density       = round(mean(n_CAAS_per_length, na.rm = TRUE), 4),
    sd_density         = round(sd(n_CAAS_per_length,   na.rm = TRUE), 4),
    mean_frac_sig_both = round(mean(frac_sig_both,     na.rm = TRUE), 4),
    sd_frac_sig_both   = round(sd(frac_sig_both,       na.rm = TRUE), 4),
    mean_median_pval   = round(mean(median_pval_hyp,   na.rm = TRUE), 4),
    sd_median_pval     = round(sd(median_pval_hyp,     na.rm = TRUE), 4),
    .groups = "drop"
  ) %>%
  arrange(CAAP_Group, filter_combo)

kable(
  overlap_summary,
  caption = "Gene-level filter overlap: descriptive statistics per membership group"
) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 11)

write_tsv(gene_filter_flags,  file.path(outliers_dir, "filter_overlap_gene_table.tsv"))
write_tsv(overlap_summary,    file.path(outliers_dir, "filter_overlap_summary.tsv"))
cat("✓ Filter overlap tables exported\n")
```

## B.2 Removed positions by `pattern_type` and significance

```{r removed_pattern_sig}
if (nrow(removal_events) > 0) {
  removal_events_plot <- removal_events %>%
    mutate(
      change_side = if_else(change_side %in% c("top", "bottom", "both"), change_side, "none"),
      parallel_type = normalize_parallel_type(parallel_type),
      pattern_parallel = make_pattern_parallel_label(pattern_type, parallel_type)
    )

  by_pattern <- removal_events_plot %>%
    count(RemovalStage, change_side, pattern_parallel, sort = TRUE)
  by_sig <- removal_events %>%
    count(RemovalStage, significance_class, sort = TRUE)

  kable(by_pattern, caption = "Removed positions by stage, change_side, and pattern/parallel label") %>% kable_styling()
  kable(by_sig, caption = "Removed positions by stage and descriptive significance class") %>% kable_styling()

  p_rm <- ggplot(
    by_pattern,
    aes(y = forcats::fct_rev(pattern_parallel), x = n, fill = RemovalStage)
  ) +
    geom_col(position = position_dodge(width = 0.75), width = 0.7) +
    facet_wrap(~change_side, scales = "free_y", ncol = 2) +
    scale_fill_paletteer_d("ggthemes::Tableau_10") +
    scale_y_discrete(labels = format_pattern_parallel_axis) +
    theme_minimal(base_size = 16) +
    labs(title = "Removed positions by convergence label", x = "n", y = "pattern_type / parallel_type") +
    theme(
      axis.text.y = element_text(size = 10, lineheight = 0.95),
      strip.text = element_text(face = "bold")
    )
  print(p_rm)
  save_barplot(
    file.path(disambig_dir, "removed_pattern_parallel_all.png"),
    p_rm,
    n_labels = n_distinct(by_pattern$pattern_parallel),
    n_facets = n_distinct(by_pattern$change_side),
    facet_cols = 2,
    base_width = 11,
    base_height = 7.5
  )
}
```

## B.2b P-value distributions: removed vs retained positions

```{r pvalue_removed_vs_retained, fig.width=12, fig.height=9}
if (nrow(discovery_std) > 0) {
  removed_site_ids <- if (nrow(removal_events) > 0) unique(removal_events$site_id) else character(0)

  pval_df <- discovery_std %>%
    mutate(
      filter_status = if_else(site_id %in% removed_site_ids, "removed", "retained"),
      filter_status = factor(filter_status, levels = c("retained", "removed"))
    ) %>%
    filter(!is.na(pvalue) | !is.na(pvalue_boot)) %>%
    pivot_longer(
      cols      = c(pvalue, pvalue_boot),
      names_to  = "pval_type",
      values_to = "p"
    ) %>%
    filter(!is.na(p)) %>%
    mutate(
      pval_type = recode(pval_type,
        pvalue      = "Hypothesis p-value",
        pvalue_boot = "Permutation p-value (boot)"
      )
    )

  status_colours <- c(
    retained = tableau_10[["blue"]],
    removed = tableau_10[["vermillion"]]
  )
  threshold_scope_levels <- c("all", "<= 0.10", "<= 0.05")

  pvalue_burden_summary <- pval_df %>%
    group_by(pval_type, filter_status) %>%
    summarise(
      total = n(),
      `all` = n(),
      `<= 0.10` = sum(p <= 0.10, na.rm = TRUE),
      `<= 0.05` = sum(p <= 0.05, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    pivot_longer(cols = c("all", "<= 0.10", "<= 0.05"), names_to = "threshold_scope", values_to = "count") %>%
    mutate(
      threshold_scope = factor(threshold_scope, levels = threshold_scope_levels),
      proportion = count / total
    ) %>%
    arrange(pval_type, filter_status, threshold_scope)

  kable(
    pvalue_burden_summary %>% mutate(proportion = percent(proportion, accuracy = 0.1)),
    caption = paste0("Removed vs retained burden by p-value threshold (configured alpha = ", alpha_threshold_label, ")")
  ) %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

  write_tsv(
    pvalue_burden_summary %>% mutate(threshold_scope = as.character(threshold_scope)),
    file.path(disambig_dir, "pvalue_burden_summary.tsv")
  )

  p_burden <- ggplot(
    pvalue_burden_summary,
    aes(x = threshold_scope, y = proportion, fill = filter_status)
  ) +
    geom_col(position = position_dodge(width = 0.7), width = 0.65, colour = "black", linewidth = 0.2) +
    geom_text(
      aes(label = sprintf("%d", count)),
      position = position_dodge(width = 0.7),
      vjust = -0.25,
      size = 3
    ) +
    facet_wrap(~pval_type, ncol = 2) +
    scale_fill_manual(values = status_colours, name = "Status") +
    scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1.05)) +
    labs(
      title = "P-value burden comparison: all vs <= 0.10 vs <= 0.05",
      subtitle = paste0("Configured alpha = ", alpha_threshold_label, " | labels show counts"),
      x = "Threshold scope",
      y = "Proportion of positions"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 25, hjust = 1)
    )
  print(p_burden)
  save_barplot(
    file.path(disambig_dir, "pvalue_burden_summary.png"),
    p_burden,
    n_labels = n_distinct(pvalue_burden_summary$threshold_scope),
    n_facets = n_distinct(pvalue_burden_summary$pval_type),
    facet_cols = 2,
    dpi = 160,
    base_width = 9,
    base_height = 5
  )

  # ── Plot 1: global double-layered histogram ─────────────────────────────────
  p_glob <- ggplot(pval_df, aes(x = p, fill = filter_status)) +
    geom_histogram(
      aes(y = after_stat(density)),
      bins      = 40,
      alpha     = 0.55,
      position  = "identity",
      colour    = "black",
      linewidth = 0.25
    ) +
    geom_vline(
      data = threshold_line_df,
      aes(xintercept = threshold, colour = line_label, linetype = line_label),
      linewidth = 0.6,
      show.legend = TRUE,
      inherit.aes = FALSE
    ) +
    facet_wrap(~pval_type, scales = "free_y", ncol = 2) +
    scale_fill_manual(values = status_colours, name = "Status") +
    scale_colour_manual(
      values = setNames(threshold_line_colours[threshold_line_df$line_key], threshold_line_df$line_label),
      guide = "none"
    ) +
    scale_linetype_manual(
      values = setNames(threshold_line_types[threshold_line_df$line_key], threshold_line_df$line_label),
      name = "Threshold lines"
    ) +
    scale_x_continuous(breaks = c(0, 0.05, 0.10, 0.25, 0.5, 0.75, 1), limits = c(0, 1)) +
    labs(
      title    = "P-value distribution: removed vs retained positions",
      subtitle = sprintf(
        "%d retained · %d removed | configured alpha = %s | comparison lines = 0.10 and 0.05",
        sum(pval_df$pval_type == "Hypothesis p-value" & pval_df$filter_status == "retained"),
        sum(pval_df$pval_type == "Hypothesis p-value" & pval_df$filter_status == "removed"),
        alpha_threshold_label
      ),
      x = "p-value", y = "Density"
    ) +
    theme_minimal(base_size = 16) +
    theme(
      legend.position  = "bottom",
      strip.text       = element_text(face = "bold"),
      panel.grid.minor = element_blank(),
      axis.text.x      = element_text(angle = 60, hjust = 1, vjust = 1)
    )
  print(p_glob)
  ggsave(file.path(disambig_dir, "pvalue_removed_vs_retained_global.png"),
         p_glob, width = 12, height = 5, dpi = 160)

  # ── Plot 2: same, faceted by pattern_type ───────────────────────────────────
  pt_counts <- pval_df %>%
    filter(pval_type == "Hypothesis p-value") %>%
    count(pattern_type, filter_status) %>%
    pivot_wider(names_from = filter_status, values_from = n, values_fill = 0)

  pval_df_pt <- pval_df %>%
    filter(pattern_type %in% (
      pval_df %>%
        filter(pval_type == "Hypothesis p-value") %>%
        count(pattern_type) %>%
        filter(n >= 5) %>%
        pull(pattern_type)
    ))

  if (nrow(pval_df_pt) > 0 && n_distinct(pval_df_pt$pattern_type) >= 1) {
    p_pt <- ggplot(pval_df_pt, aes(x = p, fill = filter_status)) +
      geom_histogram(
        aes(y = after_stat(density)),
        bins      = 30,
        alpha     = 0.55,
        position  = "identity",
        colour    = "black",
        linewidth = 0.25
      ) +
      geom_vline(
        data = threshold_line_df,
        aes(xintercept = threshold, colour = line_label, linetype = line_label),
        linewidth = 0.55,
        show.legend = TRUE,
        inherit.aes = FALSE
      ) +
      facet_grid(pattern_type ~ pval_type, scales = "free_y") +
      scale_fill_manual(values = status_colours, name = "Status") +
      scale_colour_manual(
        values = setNames(threshold_line_colours[threshold_line_df$line_key], threshold_line_df$line_label),
        guide = "none"
      ) +
      scale_linetype_manual(
        values = setNames(threshold_line_types[threshold_line_df$line_key], threshold_line_df$line_label),
        name = "Threshold lines"
      ) +
      scale_x_continuous(breaks = c(0, 0.05, 0.10, 0.5, 1), limits = c(0, 1)) +
      labs(
        title = "P-value distribution by pattern_type: removed vs retained",
        subtitle = paste0("Configured alpha = ", alpha_threshold_label, " | comparison lines = 0.10 and 0.05"),
        x = "p-value", y = "Density"
      ) +
      theme_minimal(base_size = 11) +
      theme(
        legend.position  = "bottom",
        strip.text       = element_text(face = "bold", size = 9),
        panel.grid.minor = element_blank(),
        axis.text.x      = element_text(angle = 60, hjust = 1, vjust = 1)
      )
    print(p_pt)
    ggsave(file.path(disambig_dir, "pvalue_removed_vs_retained_by_pattern.png"),
           p_pt, width = 12, height = max(4, 2.5 * n_distinct(pval_df_pt$pattern_type)), dpi = 160)
  } else {
    cat("Skipping per-pattern_type p-value plot: no pattern_type with \u2265 5 observations.\n")
  }
} else {
  cat("No discovery data available — skipping p-value removal histograms.\n")
}
```

## B.3 Global side-aware convergence characterization

```{r global_pattern_sig}
pattern_parallel_by_group <- convergence_plot_df %>%
  count(change_side, caap_group, pattern_parallel, sort = TRUE)

kable(pattern_parallel_by_group, caption = "Convergence labels by CAAP group and change_side") %>% kable_styling()

pg <- ggplot(
  pattern_parallel_by_group,
  aes(y = forcats::fct_rev(pattern_parallel), x = n, fill = caap_group)
) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  facet_wrap(~change_side, scales = "free_y", ncol = 2) +
  scale_fill_paletteer_d("ggthemes::Tableau_10") +
  scale_y_discrete(labels = format_pattern_parallel_axis) +
  theme_minimal(base_size = 14) +
  labs(title = "Convergence labels across CAAP groups", x = "n", y = "pattern_type / parallel_type") +
  theme(
    axis.text.y = element_text(size = 10, lineheight = 0.95),
    strip.text = element_text(face = "bold")
  )
print(pg)
save_barplot(
  file.path(disambig_dir, "pattern_parallel_by_caap_group.png"),
  pg,
  n_labels = n_distinct(pattern_parallel_by_group$pattern_parallel),
  n_facets = n_distinct(pattern_parallel_by_group$change_side),
  facet_cols = 2,
  base_width = 11,
  base_height = 7.5
)
```

## B.3b Convergence descriptors by significance class

```{r pattern_sig_descriptor}
pattern_parallel_sig <- convergence_plot_df %>%
  count(change_side, significance_class, pattern_parallel, sort = TRUE)

kable(pattern_parallel_sig, caption = "Descriptive significance classes across convergence labels") %>% kable_styling()

if (nrow(pattern_parallel_sig) > 0) {
  pg_sig <- ggplot(
    pattern_parallel_sig,
    aes(y = forcats::fct_rev(pattern_parallel), x = n, fill = significance_class)
  ) +
    geom_col(position = position_dodge(width = 0.75), width = 0.7) +
    facet_wrap(~change_side, scales = "free_y", ncol = 2) +
    scale_fill_paletteer_d("ggthemes::Tableau_10") +
    scale_y_discrete(labels = format_pattern_parallel_axis) +
    theme_minimal(base_size = 14) +
    labs(title = "Descriptive significance classes across convergence labels", x = "n", y = "pattern_type / parallel_type") +
    theme(
      axis.text.y = element_text(size = 10, lineheight = 0.95),
      strip.text = element_text(face = "bold")
    )
  print(pg_sig)
  save_barplot(
    file.path(disambig_dir, "pattern_parallel_significance_descriptor.png"),
    pg_sig,
    n_labels = n_distinct(pattern_parallel_sig$pattern_parallel),
    n_facets = n_distinct(pattern_parallel_sig$change_side),
    facet_cols = 2,
    base_width = 11,
    base_height = 7.5
  )
}
```

## B.4 Overlap across CAAP groupings (gene & position level)

Venn diagrams are rendered per orientation (`global`, `top`, `bottom`, `both`) using the
same `change_side` logic as the focal summaries.

```{r overlap_helper_functions, include=FALSE}
has_ggvenn <- requireNamespace("ggVennDiagram", quietly = TRUE)
has_venn_diagram <- requireNamespace("VennDiagram", quietly = TRUE)
if (has_ggvenn) library(ggVennDiagram)

create_group_sets <- function(df, id_col) {
  groups <- sort(unique(df$caap_group))
  setNames(lapply(groups, function(g) unique(df[[id_col]][df$caap_group == g])), groups)
}

make_overlap_plot <- function(df, id_col, level_label, subset_label, out_dir) {
  if (nrow(df) == 0) return(NULL)
  groups <- sort(unique(df$caap_group))
  if (length(groups) < 2 || length(groups) > 5) return(NULL)

  png_name <- paste0("overlap_", tolower(level_label), "_", subset_label, ".png")
  png_path <- file.path(out_dir, png_name)
  set_list <- create_group_sets(df, id_col)

  if (has_ggvenn) {
    venn_plot <- tryCatch({
      ggVennDiagram(set_list, label_alpha = 0.8, set_size = 5) +
        scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Count") +
        labs(
          title    = paste0(level_label, "-level overlap [", subset_label, "]"),
          subtitle = paste0("n = ", n_distinct(df[[id_col]]), " unique ", tolower(level_label), "s")
        ) +
        theme(
          plot.title    = element_text(face = "bold", size = 20),
          plot.subtitle = element_text(size = 15)
        )
    }, error = function(e) NULL)

    if (!is.null(venn_plot)) {
      ggsave(png_path, venn_plot, width = 14, height = 10, dpi = 320)
      return(png_path)
    }
  }

  if (has_venn_diagram) {
    venn_fill <- rep(unname(as.character(tableau_10)), length.out = length(set_list))
    venn_path <- tryCatch({
      venn_grob <- VennDiagram::venn.diagram(
        x = set_list,
        filename = NULL,
        fill = venn_fill,
        alpha = 0.45,
        cex = 1.4,
        cat.cex = 1.2,
        cat.fontface = "bold",
        margin = 0.08,
        main = paste0(level_label, "-level overlap [", subset_label, "]"),
        sub = paste0("n = ", n_distinct(df[[id_col]]), " unique ", tolower(level_label), "s")
      )
      png(png_path, width = 14, height = 10, units = "in", res = 320)
      grid::grid.newpage()
      grid::grid.draw(venn_grob)
      dev.off()
      png_path
    }, error = function(e) NULL)

    if (!is.null(venn_path) && file.exists(venn_path)) {
      return(venn_path)
    }
  }

  tryCatch({
    fallback_plot <- build_pairwise_overlap_plot(set_list, level_label, subset_label)
    ggsave(png_path, fallback_plot, width = 10, height = 8, dpi = 320)
    png_path
  }, error = function(e) NULL)
}
render_overlap_scenario <- function(df, orientation_name, id_col, level_label) {
  scenario_defs <- list(
    scenario1 = df %>% filter(toupper(caap_group) != "GS0"),
    scenario2 = df %>% filter(toupper(caap_group) != "US"),
    scenario3 = df %>% filter(toupper(caap_group) %in% c("GS0", "US"))
  )

  for (scenario_name in names(scenario_defs)) {
    scenario_df <- scenario_defs[[scenario_name]]
    cat(sprintf("**%s-level overlap: %s / %s**\n\n", level_label, tools::toTitleCase(orientation_name), scenario_name))
    plot_path <- make_overlap_plot(scenario_df, id_col, level_label, paste0(orientation_name, "_", scenario_name), overlap_out_dir)
    if (!is.null(plot_path) && file.exists(plot_path)) {
      print(knitr::include_graphics(plot_path))
      cat("\n")
    } else {
      cat("_No overlap plot rendered for this subset._\n\n")
    }
  }
}
```

```{r overlap_orientation, results='asis', fig.width=14, fig.height=10}
overlap_subsets <- orientation_rows %>%
  distinct(orientation, site_id, gene, caap_group) %>%
  split(.$orientation)

for (orientation_name in c("global", "top", "bottom", "both")) {
  subset_df <- overlap_subsets[[orientation_name]]
  if (is.null(subset_df) || nrow(subset_df) == 0) next
  cat(sprintf("\n### %s\n\n", tools::toTitleCase(orientation_name)))
  render_overlap_scenario(subset_df, orientation_name, "site_id", "Position")
  render_overlap_scenario(subset_df, orientation_name, "gene", "Gene")
}
```

## B.5 US/GS group intersections (upset plots + structured exports)

```{r us_gs_relation_analysis, fig.height=18, fig.width=16}
# collapse_change_side and combo_from_flags are defined in setup.

# ── Upset helper: build bar + membership-matrix composite and return it ────────
plot_upset_like <- function(intersections_df, set_cols, title_txt, out_file = NULL) {
  if (is.null(intersections_df) || nrow(intersections_df) == 0) return(invisible(NULL))

  label_map <- c(us = "US", gs0 = "GS0", gs1 = "GS1", gs2 = "GS2", gs3 = "GS3", gs4 = "GS4")

  plot_df <- intersections_df %>%
    arrange(desc(n), combo_label) %>%
    mutate(combo_id = factor(seq_len(n())))

  p_bar <- ggplot(plot_df, aes(x = combo_id, y = n)) +
    geom_col(fill = tableau_10[["blue"]], color = "black", linewidth = 0.2) +
    geom_text(aes(label = n), vjust = -0.35, size = 5.2, fontface = "bold") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
    labs(title = title_txt, x = NULL, y = "Intersection size") +
    theme_minimal(base_size = 15) +
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      plot.title = element_text(face = "bold", size = 16)
    )

  mat_long <- plot_df %>%
    select(combo_id, all_of(set_cols)) %>%
    pivot_longer(cols = all_of(set_cols), names_to = "set_name", values_to = "present") %>%
    mutate(set_label = recode(set_name, !!!label_map),
           set_label = factor(set_label, levels = rev(unname(label_map[set_cols]))))

  p_mat <- ggplot(mat_long, aes(x = combo_id, y = set_label, fill = present)) +
    geom_tile(color = "white", linewidth = 0.3) +
    scale_fill_manual(values = c(`TRUE` = tableau_10[["vermillion"]], `FALSE` = "#eeeeee")) +
    labs(x = "Intersections", y = NULL) +
    theme_minimal(base_size = 14) +
    theme(
      legend.position = "none",
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.text.y = element_text(face = "bold")
    )

  p <- p_bar / p_mat + plot_layout(heights = c(1.7, 0.9))
  if (!is.null(out_file)) {
    upset_width <- min(20, max(12, 6 + 0.8 * nrow(plot_df)))
    ggsave(out_file, p, width = upset_width, height = 6.5, dpi = 320)
  }
  invisible(p)
}

# ── Export helper ──────────────────────────────────────────────────────────────
write_export_pair <- function(df, key, txt_d, tsv_d) {
  createDir(txt_d)
  createDir(tsv_d)
  base_cols <- intersect(c("gene", "position", "caap_group", "pattern_type", "parallel_type", "change_side"), names(df))
  out_df <- df %>% select(all_of(base_cols)) %>% distinct() %>% arrange(gene, position)
  readr::write_lines(sort(unique(out_df$gene)), file.path(txt_d, paste0(key, ".txt")))
  write_tsv(out_df, file.path(tsv_d, paste0(key, ".tsv")))
}

# ── Output directories ─────────────────────────────────────────────────────────
gene_rel_dir  <- file.path(disambig_dir, "gene_relation_analysis")
gr_plot_dir   <- file.path(gene_rel_dir, "plots")
gr_txt_dir    <- file.path(gene_rel_dir, "txt")
gr_tsv_dir    <- file.path(gene_rel_dir, "tsv")
createDir(gr_plot_dir)
createDir(gr_txt_dir)
createDir(gr_tsv_dir)

grp_sets_global <- c("us", "gs0", "gs1", "gs2", "gs3", "gs4")

make_intersections <- function(sf, set_cols) {
  sf %>%
    filter(if_any(all_of(set_cols))) %>%
    count(across(all_of(set_cols)), name = "n") %>%
    mutate(combo_label = combo_from_flags(cur_data(), set_cols)) %>%
    filter(n > 0)
}

detail_rows <- orientation_rows %>%
  transmute(
    orientation = as.character(orientation),
    site_id,
    gene,
    position,
    caap_group = toupper(caap_group),
    pattern_type,
    parallel_type,
    change_side
  ) %>%
  distinct()

for (ori_name in c("global", "top", "bottom", "both")) {
  flags_ori <- orientation_flags %>% filter(as.character(orientation) == ori_name)
  p_all <- plot_upset_like(
    make_intersections(flags_ori, grp_sets_global),
    grp_sets_global,
    paste0("US/GS group intersections — ", tools::toTitleCase(ori_name)),
    out_file = file.path(gr_plot_dir, paste0("us_gs_upset_", ori_name, ".png"))
  )
  if (!is.null(p_all)) print(p_all)

  base_df <- detail_rows %>% filter(orientation == ori_name)
  write_export_pair(base_df, paste0("all_", ori_name), gr_txt_dir, gr_tsv_dir)
  for (grp_name in c("US", "GS0", "GS1", "GS2", "GS3", "GS4")) {
    write_export_pair(
      base_df %>% filter(caap_group == grp_name),
      paste0(tolower(grp_name), "_", ori_name),
      gr_txt_dir,
      gr_tsv_dir
    )
  }
  cat("✓ Upset and exports:", ori_name, "\n")
}

cat("\n✓ All B.5 upset plots and exports generated\n")
cat("  Plots:", gr_plot_dir, "\n")
cat("  TXT:  ", gr_txt_dir,  "\n")
cat("  TSV:  ", gr_tsv_dir,  "\n\n")
```

## B.5b Excluded gene list exports

```{r excluded_gene_lists}
# ── Build per-category excluded gene lists ────────────────────────────────────────────
# Six lists are exported to
#   disambiguation_characterization/gene_relation_analysis/txt/excluded/
# for a separate ORA run against the original (pre-cleanup) background from discovery.
#
# Categories (mutually exclusive by design for excluded_nonsig):
#   excluded_preproc             – genes with ≥1 position removed during pre-processing
#                                  (pattern-type filter / MRCA-posterior threshold)
#   excluded_postproc_clusters   – genes with ≥1 position discarded by cluster filter
#   excluded_postproc_genefilter – genes flagged Extreme and/or Dubious (gene-level filter)
#   excluded_postproc            – union of the two postproc exclusion categories
#   excluded_nonsig              – genes that survived ALL filters but never achieved
#                                  sig_both == TRUE; does NOT overlap with preproc/postproc
#   excluded_overall             – union of all four categories

# 1. Pre-processing exclusions (from precluster_events, loaded in B.1)
excluded_preproc <- sort(unique(precluster_events$gene))

# 2. Post-processing: cluster-filter genes
excluded_postproc_clusters <- sort(unique(
  filter(gene_filter_flags, Cluster_removed)$Gene
))

# 3. Post-processing: gene-level filter (Extreme / Dubious / Both)
excluded_postproc_genefilter <- sort(unique(
  filter(gene_filter_flags, Extreme | Dubious)$Gene
))

# 4. All post-processing exclusions (union of 2 and 3)
excluded_postproc <- sort(unique(c(excluded_postproc_clusters, excluded_postproc_genefilter)))

# 5. Non-significant genes: survived all filters, but zero sig_both positions;
#    explicitly does NOT overlap with preproc / postproc categories
any_filtered_genes <- sort(unique(filter(gene_filter_flags, filter_combo != "None")$Gene))
genes_ever_sig     <- sort(unique(filter(discovery_std, sig_both)$gene))
survived_genes     <- sort(setdiff(unique(gene_filter_flags$Gene), any_filtered_genes))
excluded_nonsig    <- sort(setdiff(survived_genes, c(genes_ever_sig, excluded_preproc)))

# 6. Overall: union of all exclusion categories
excluded_overall <- sort(unique(c(excluded_preproc, excluded_postproc, excluded_nonsig)))

# ── Write to gene_relation_analysis/txt/excluded/ ────────────────────────────────────
gr_excl_dir <- file.path(gr_txt_dir, "excluded")
createDir(gr_excl_dir)

readr::write_lines(excluded_preproc,             file.path(gr_excl_dir, "excluded_preproc.txt"))
readr::write_lines(excluded_postproc_clusters,   file.path(gr_excl_dir, "excluded_postproc_clusters.txt"))
readr::write_lines(excluded_postproc_genefilter, file.path(gr_excl_dir, "excluded_postproc_genefilter.txt"))
readr::write_lines(excluded_postproc,            file.path(gr_excl_dir, "excluded_postproc.txt"))
readr::write_lines(excluded_nonsig,              file.path(gr_excl_dir, "excluded_nonsig.txt"))
readr::write_lines(excluded_overall,             file.path(gr_excl_dir, "excluded_overall.txt"))

cat("\n=== Excluded Gene Lists ===\n")
cat(sprintf("  excluded_preproc             : %d genes\n", length(excluded_preproc)))
cat(sprintf("  excluded_postproc_clusters   : %d genes\n", length(excluded_postproc_clusters)))
cat(sprintf("  excluded_postproc_genefilter : %d genes\n", length(excluded_postproc_genefilter)))
cat(sprintf("  excluded_postproc (union)    : %d genes\n", length(excluded_postproc)))
cat(sprintf("  excluded_nonsig              : %d genes\n", length(excluded_nonsig)))
cat(sprintf("  excluded_overall             : %d genes\n", length(excluded_overall)))
cat(sprintf("  Written to: %s\n\n", gr_excl_dir))
```

## B.6 Manhattan: genomic distribution of genes (CAAS density per gene length)

```{r manhattan_disambiguation, fig.height=8, fig.width=16}
if (!"chr" %in% names(gene_ensembl)) {
  cat("⚠ Chromosome information not available in gene ensembl file, skipping Manhattan plot\n")
} else {
  # Calculate retained CAAS density per gene across all CT positions.
  gene_caas_density <- discovery_std %>%
    group_by(gene) %>%
    summarise(
      n_caas_positions = n_distinct(site_id),
      frac_sig_both = mean(sig_both, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    left_join(gene_ensembl %>% 
              select(gene, chr, start, end) %>%
              mutate(gene_length = end - start + 1), 
              by = "gene") %>%
    filter(!is.na(chr), gene_length > 0) %>%
    mutate(
      caas_density = n_caas_positions / gene_length,
      log_caas_density = log10(pmax(caas_density, 1e-10))  # Avoid log10(0)
    )

  # Visualise all genes with at least one retained CT position.
  manhattan_data <- gene_caas_density %>%
    filter(caas_density > 0) %>%
    mutate(
      chr_num   = str_replace(chr, "chr", ""),
      chr_order = case_when(
        chr_num %in% as.character(1:22) ~ as.numeric(chr_num),
        chr_num == "X"                  ~ 23,
        chr_num == "Y"                  ~ 24,
        TRUE                            ~ 25
      ),
      pos = (start + end) / 2
    ) %>%
    arrange(chr_order, pos)

  if (nrow(manhattan_data) > 0) {
    # Cumulative chromosome positions and boundaries
    chr_lengths <- manhattan_data %>%
      group_by(chr_num, chr_order) %>%
      summarise(chr_max = max(pos, na.rm = TRUE), .groups = "drop") %>%
      arrange(chr_order) %>%
      mutate(
        chr_start  = lag(cumsum(chr_max), default = 0),
        chr_end    = chr_start + chr_max,
        chr_center = chr_start + chr_max / 2
      )

    manhattan_data <- manhattan_data %>%
      left_join(chr_lengths %>% select(chr_num, chr_start, chr_end), by = "chr_num") %>%
      mutate(BPcum = pos + chr_start)

    n_genes_plotted <- n_distinct(manhattan_data$gene)
    n_chr_plotted <- n_distinct(chr_lengths$chr_num)
    manhattan_width <- min(22, max(16, 10 + 0.55 * n_chr_plotted))
    manhattan_height <- min(12, max(9, 8 + 0.015 * n_genes_plotted))

    p_manhattan <- ggplot(manhattan_data,
        aes(x = BPcum, y = caas_density)) +
      # Discontinuous chromosome boundary lines
      geom_vline(
        data        = chr_lengths,
        aes(xintercept = chr_end),
        linetype    = "dotted", color = "grey60", alpha = 0.7, linewidth = 0.4,
        inherit.aes = FALSE
      ) +
      geom_point(alpha = 0.85, size = 3.1, color = "#2E86AB") +
      scale_x_continuous(
        labels = chr_lengths$chr_num,
        breaks = chr_lengths$chr_center,
        expand = expansion(mult = c(0.005, 0.005))
      ) +
      scale_y_continuous(
        labels = scales::label_number(scale_cut = cut_short_scale()),
        expand = expansion(mult = c(0.02, 0.08))
      ) +
      labs(
        title    = "Genomic distribution of retained CAAS density",
        subtitle = sprintf(
          "n = %d genes with retained CT positions | y = retained positions / gene length (bp)",
          n_genes_plotted
        ),
        x     = "Chromosome",
        y     = "CAAS density (retained positions / gene length)"
      ) +
      theme_minimal(base_size = 16) +
      theme(
        plot.title = element_text(face = "bold", size = 20),
        plot.subtitle = element_text(size = 14),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 13),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank()
      )

    print(p_manhattan)
    ggsave(
      filename = file.path(disambig_dir, "manhattan_caas_density.png"),
      plot     = p_manhattan, width = manhattan_width, height = manhattan_height, dpi = 320, dev = "png"
    )
  } else {
    cat("No genes with retained CT positions found — Manhattan plot skipped.\n")
  }
}
```

## 5. Impact Summary

```{r impact_summary}
cat("\n=== CT Post-Processing Impact Summary ===\n\n")

cat("Gene-level statistics:\n")
cat("  Total genes analyzed:", n_distinct(discovery_df$Gene), "\n\n")

cat("Position-level statistics:\n")
cat("  Total CT positions:", nrow(discovery_df), "\n")
if (nrow(filter_summary) > 0) {
  cat("  Positions flagged by cluster filtering:\n")
  for (i in 1:nrow(filter_summary)) {
    cat(sprintf("    minlen=%d, maxcaas=%.2f: %d positions\n", filter_summary$Minlen[i], filter_summary$Maxcaas[i], filter_summary$DiscardedCount[i]))
  }
}

summary_table <- tibble(
  Metric = c("Total CT Positions", "Unique Genes"),
  Value  = c(
    as.character(nrow(discovery_df)),
    as.character(n_distinct(discovery_df$Gene))
  )
)

kable(summary_table, caption = "CT Post-Processing Summary Statistics", align = c("l", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

write_tsv(summary_table, file.path(summary_stats_dir, "summary_statistics.tsv"))
cat("\n✓ Summary statistics exported\n")
cat("\n=== Analysis Complete ===\n")
```

## Session Info

```{r session_info}
sessionInfo()
```
