---
title: "CT Post-Processing and Disambiguation Characterization"
author: "PhyloPhere Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    code_folding: hide
    df_print: paged
params:
  discovery_file: NULL
  filter_summary_file: NULL
  filter_dir: NULL
  gene_ensembl_file: NULL
  gene_stats_file: NULL
  processing_mode: "exploratory"
  output_dir: "ct_postproc"
  extreme_threshold: 0.99
  iqr_multiplier: 3.0
  gene_filter_mode: "both"
  generate_manhattan: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7,
  dev = 'png',
  dpi = 320
)

suppressPackageStartupMessages({
  library(stringr)
  library(purrr)
  library(tibble)
  library(patchwork)
  library(scales)
  library(ggplot2)
  library(forcats)
  library(knitr)
  library(kableExtra)
  library(paletteer)
  library(tidyverse)
  if (requireNamespace("UpSetR", quietly = TRUE)) library(UpSetR)
})

createDir <- function(directory) {
  if (!dir.exists(directory)) {
    dir.create(directory, recursive = TRUE, showWarnings = FALSE)
  }
}

find_first_existing <- function(candidates) {
  hits <- candidates[file.exists(candidates)]
  if (length(hits) == 0) return(NA_character_)
  hits[[1]]
}

find_recursive_file <- function(root_dir, pattern) {
  if (is.null(root_dir) || is.na(root_dir) || !dir.exists(root_dir)) return(NA_character_)
  files <- list.files(root_dir, pattern = pattern, full.names = TRUE, recursive = TRUE)
  if (length(files) == 0) return(NA_character_)
  files[[1]]
}

as_bool <- function(x) {
  tolower(as.character(x)) %in% c("true", "t", "1", "yes", "y")
}

collapse_change_side <- function(x) {
  x <- tolower(as.character(x))
  x <- x[x %in% c("top", "bottom", "both")]
  if (length(x) == 0) return("none")
  if ("both" %in% x) return("both")
  ux <- unique(x)
  if (all(c("top", "bottom") %in% ux)) return("both")
  ux[[1]]
}

combo_from_flags <- function(df, cols) {
  apply(df[, cols, drop = FALSE], 1, function(v) {
    labs <- cols[as.logical(v)]
    if (length(labs) == 0) "none" else paste(labs, collapse = "+")
  })
}

normalize_disambig_schema <- function(df) {
  if ('Pattern' %in% names(df) && !'pattern_type' %in% names(df)) df <- df %>% rename(pattern_type = Pattern)
  if ('Pvalue' %in% names(df) && !'pvalue' %in% names(df)) df <- df %>% rename(pvalue = Pvalue)
  if ('Pvalue.boot' %in% names(df) && !'pvalue_boot' %in% names(df)) df <- df %>% rename(pvalue_boot = `Pvalue.boot`)
  if ('ChangeSide' %in% names(df) && !'change_side' %in% names(df)) df <- df %>% rename(change_side = ChangeSide)
  if ('Gene' %in% names(df) && !'gene' %in% names(df)) df <- df %>% rename(gene = Gene)
  if ('Position' %in% names(df) && !'position' %in% names(df)) df <- df %>% rename(position = Position)
  if ('CAAP_Group' %in% names(df) && !'caap_group' %in% names(df)) df <- df %>% rename(caap_group = CAAP_Group)
  if ('IsConserved' %in% names(df) && !'is_conserved_meta' %in% names(df)) df <- df %>% rename(is_conserved_meta = IsConserved)
  if ('ASR_ConsAntiCAAS' %in% names(df) && !'asr_is_conserved' %in% names(df)) df <- df %>% rename(asr_is_conserved = ASR_ConsAntiCAAS)
  if (!'caap_group' %in% names(df)) df$caap_group <- 'US'
  if (!'change_side' %in% names(df)) df$change_side <- 'none'
  if (!'is_conserved_meta' %in% names(df)) df$is_conserved_meta <- FALSE
  if (!'asr_is_conserved' %in% names(df)) df$asr_is_conserved <- FALSE

  df %>%
    mutate(
      gene = as.character(gene),
      position = suppressWarnings(as.integer(position)),
      caap_group = as.character(caap_group),
      pattern_type = as.character(pattern_type),
      change_side = tolower(as.character(change_side)),
      is_conserved_meta = as_bool(is_conserved_meta),
      asr_is_conserved = as_bool(asr_is_conserved),
      pvalue = suppressWarnings(as.numeric(pvalue)),
      pvalue_boot = suppressWarnings(as.numeric(pvalue_boot)),
      sig_hyp = !is.na(pvalue) & pvalue <= 0.05,
      sig_perm = !is.na(pvalue_boot) & pvalue_boot <= 0.05,
      sig_both = sig_hyp & sig_perm,
      significance_class = case_when(
        sig_both ~ 'both',
        sig_hyp ~ 'hyper_only',
        sig_perm ~ 'perm_only',
        TRUE ~ 'none'
      ),
      site_id = paste(gene, position, sep = '@')
    )
}

seed_val <- 12345
set.seed(seed_val)

discovery_file <- params$discovery_file
filter_summary_file <- params$filter_summary_file
gene_ensembl_file <- params$gene_ensembl_file
gene_stats_file <- params$gene_stats_file
processing_mode <- params$processing_mode
extreme_threshold <- params$extreme_threshold
iqr_multiplier <- params$iqr_multiplier
gene_filter_mode <- params$gene_filter_mode
generate_manhattan <- params$generate_manhattan
manhattan_min_density <- (params$extreme_threshold + (params$extreme_threshold) / 2)

if (is.null(discovery_file)) stop("Error: discovery_file parameter is required")
if (is.null(filter_summary_file)) stop("Error: filter_summary_file parameter is required")
if (is.null(gene_ensembl_file)) stop("Error: gene_ensembl_file parameter is required")

stopifnot(file.exists(discovery_file), file.exists(filter_summary_file), file.exists(gene_ensembl_file))

ct_postproc_dir <- "."
outliers_dir <- "outliers"
cluster_dir <- "clusters"
summary_stats_dir <- "summary_statistics"
disambig_dir <- "disambiguation_characterization"
overlap_out_dir <- file.path(disambig_dir, "overlap_figures")

createDir(ct_postproc_dir)
createDir(outliers_dir)
createDir(cluster_dir)
createDir(summary_stats_dir)
createDir(disambig_dir)
createDir(overlap_out_dir)

tableau_10 <- paletteer_d("ggthemes::Tableau_10")
names(tableau_10) <- c(
  "blue", "orange", "vermillion", "bluish_green",
  "green", "yellow", "reddish_purple", "pink",
  "brown", "gray"
)

category_palette <- c(
  "Normal" = tableau_10[["gray"]],
  "Dubious" = tableau_10[["blue"]],
  "Extreme" = tableau_10[["vermillion"]],
  "Both" = tableau_10[["reddish_purple"]]
)

removal_palette <- c(
  Discarded = tableau_10[["vermillion"]],
  Good = tableau_10[["green"]]
)
```

# A) Post-processing characterization

## Data Loading and Gene Length Annotation

```{r data_loading}
cat("=== CT Post-Processing Configuration ===\n")
cat("Processing mode:", processing_mode, "\n")
cat("Gene filtering:", gene_filter_mode, "\n")
cat("Extreme threshold:", extreme_threshold, paste0("(top ", 100*(1-extreme_threshold), "%)"), "\n")
cat("IQR multiplier:", iqr_multiplier, "\n")
cat("Generate Manhattan plots:", generate_manhattan, "\n")

cat("\nInput files:\n")
cat("  Discovery:", discovery_file, "\n")
cat("  Filter summary:", filter_summary_file, "\n")
cat("  Gene ensembl:", gene_ensembl_file, "\n")
cat("  Gene stats:", ifelse(is.null(gene_stats_file) || is.na(gene_stats_file), "Not provided (will recompute)", gene_stats_file), "\n\n")

raw_discovery <- read_delim(discovery_file, delim = "\t", show_col_types = FALSE)
discovery_std <- normalize_disambig_schema(raw_discovery)

discovery_df <- discovery_std %>%
  transmute(
    CAAP_Group = caap_group,
    Gene = gene,
    Position = position,
    Pattern = pattern_type,
    is_conserved_meta,
    asr_is_conserved,
    Pvalue = pvalue,
    `Pvalue.boot` = pvalue_boot,
    sig_hyp,
    sig_perm,
    sig_both,
    significance_class,
    site_id
  )

if (!"CAAP_Group" %in% names(discovery_df)) discovery_df <- discovery_df %>% mutate(CAAP_Group = "US")

cat("  Loaded CT positions:", nrow(discovery_df), "\n")
cat("  Unique genes:", n_distinct(discovery_df$Gene), "\n")

gene_ensembl <- read_tsv(gene_ensembl_file, show_col_types = FALSE)
names(gene_ensembl) <- tolower(names(gene_ensembl))
if (!"gene" %in% names(gene_ensembl) || !"length" %in% names(gene_ensembl)) {
  stop("Gene ensembl file must have 'gene' and 'length' columns")
}

gene_length <- gene_ensembl %>% select(Gene = gene, Length = length)
discovery_df <- discovery_df %>% left_join(gene_length, by = "Gene") %>% filter(!is.na(Length))

cat("  Genes with length annotation:", n_distinct(discovery_df$Gene), "\n\n")
```

## Conserved metadata composition after precluster pruning

```{r conserved_meta_composition}

# Relation of genes that do seem to present conserved positions but they aren't supported by ASR (i.e. they are not conserved in the ASR node but they are in the meta-analysis) 
#These could be interesting to explore as potential false positives or cases where ASR is missing signal. Removed downstream.
non_asr_conserved <- discovery_df %>% dplyr::filter(is_conserved_meta == TRUE, asr_is_conserved == FALSE)
count_non_asr_conserved <- nrow(non_asr_conserved)

cat("  CAAS with conserved meta but not ASR support:", count_non_asr_conserved, "\n")

discovery_df <- discovery_df %>%
  # Remove non-ASR-supported conserved sites to focus on the more robustly supported ones
  dplyr::filter(!(is_conserved_meta == TRUE & asr_is_conserved == FALSE))

conserved_meta_comp <- discovery_df %>%
  mutate(is_conserved_meta = if_else(is_conserved_meta, "TRUE", "FALSE")) %>%
  count(is_conserved_meta, name = "n") %>%
  mutate(prop = n / sum(n))

kable(
  conserved_meta_comp,
  caption = "is_conserved_meta composition after precluster pruning"
) %>% kable_styling()
```

## Outlier Gene Detection

```{r outlier_detection, fig.height=10, fig.width=14}
# Load gene statistics from precomputed file or compute fallback
gene_summary_annotated <- NULL

if (!is.null(gene_stats_file) && !is.na(gene_stats_file) && file.exists(gene_stats_file)) {
  cat("Loading precomputed gene statistics from:", gene_stats_file, "\n")
  gene_summary_annotated <- read_tsv(gene_stats_file, show_col_types = FALSE)
  
  # Ensure required columns exist and rename for compatibility
  if ("CAAP_Group" %in% names(gene_summary_annotated)) {
    colnames(gene_summary_annotated)[colnames(gene_summary_annotated) == "CAAP_Group"] <- "CAAP_Group"
  } else if ("Trait" %in% names(gene_summary_annotated)) {
    gene_summary_annotated$CAAP_Group <- "US"  # default for non-CAAP mode
  }
  
  cat("Loaded gene statistics for", nrow(gene_summary_annotated), "genes\n")

  # Ensure all downstream-required columns exist (file may lack some)
  if (!"CAAP_Group" %in% names(gene_summary_annotated)) {
    gene_summary_annotated$CAAP_Group <- "US"
  }
  if (!"n_CAAS_per_length" %in% names(gene_summary_annotated)) {
    if ("n_CAAS" %in% names(gene_summary_annotated) && "Length" %in% names(gene_summary_annotated)) {
      gene_summary_annotated$n_CAAS_per_length <- gene_summary_annotated$n_CAAS / gene_summary_annotated$Length * 100
    } else {
      gene_summary_annotated$n_CAAS_per_length <- NA_real_
    }
  }
  if (!"Extreme" %in% names(gene_summary_annotated))          gene_summary_annotated$Extreme          <- FALSE
  if (!"Dubious" %in% names(gene_summary_annotated))          gene_summary_annotated$Dubious          <- FALSE
  if (!"Category" %in% names(gene_summary_annotated))         gene_summary_annotated$Category         <- "Normal"
  if (!"threshold_extreme" %in% names(gene_summary_annotated)) gene_summary_annotated$threshold_extreme <- NA_real_
  if (!"threshold_dubious" %in% names(gene_summary_annotated)) gene_summary_annotated$threshold_dubious <- NA_real_
  
} else {
  cat("Warning: Gene stats file not provided or not found. Computing fallback statistics...\n")
  
  # Fallback: recompute (but this may have the original bugs)
  has_caap_group <- TRUE
  cat("Unified grouping mode enabled (CAAP_Group always present)\n")
  cat("Groups found:", paste(sort(unique(discovery_df$CAAP_Group)), collapse=", "), "\n")
  
  gene_summary <- discovery_df %>%
    group_by(CAAP_Group, Gene) %>%
    summarise(
      n_CAAS = n(),
      Length = first(Length),
      n_CAAS_per_length = n() / first(Length) * 100,
      .groups = 'drop'
    )
  
  iqr_thresholds <- gene_summary %>%
    group_by(CAAP_Group) %>%
    summarise(
      Q1 = quantile(n_CAAS, 0.25, na.rm = TRUE),
      Q3 = quantile(n_CAAS, 0.75, na.rm = TRUE),
      IQR_val = Q3 - Q1,
      threshold_dubious = Q3 + iqr_multiplier * IQR_val,
      .groups = 'drop'
    )
  
  extreme_thresholds <- gene_summary %>%
    group_by(CAAP_Group) %>%
    summarise(
      threshold_extreme = quantile(n_CAAS_per_length, extreme_threshold, na.rm = TRUE),
      .groups = 'drop'
    )
  
  gene_summary_annotated <- gene_summary %>%
    left_join(extreme_thresholds, by = 'CAAP_Group') %>%
    left_join(iqr_thresholds, by = 'CAAP_Group') %>%
    mutate(
      Extreme = n_CAAS_per_length >= threshold_extreme,
      Dubious = n_CAAS >= threshold_dubious,
      Category = case_when(
        Extreme & Dubious ~ "Both",
        Extreme ~ "Extreme",
        Dubious ~ "Dubious",
        TRUE ~ "Normal"
      )
    ) %>%
    arrange(desc(n_CAAS))
}

# Ensure CAAP_Group column exists for plotting
if (!"CAAP_Group" %in% names(gene_summary_annotated)) {
  gene_summary_annotated$CAAP_Group <- "US"
}

cat("Groups found:", paste(sort(unique(gene_summary_annotated$CAAP_Group)), collapse=", "), "\n")

if (nrow(gene_summary_annotated) == 0) {
  cat("WARNING: gene_summary_annotated is empty — skipping outlier distribution plots.\n")
} else {

# Pre-compute vline data — guard against empty frames that break facet_wrap
vline_dubious_data  <- gene_summary_annotated %>% filter(!is.na(threshold_dubious))  %>% distinct(CAAP_Group, threshold_dubious)
vline_extreme_data  <- gene_summary_annotated %>% filter(!is.na(threshold_extreme)) %>% distinct(CAAP_Group, threshold_extreme)

# Generate histograms with increased font sizes for slides
p_gross_dist <- ggplot(gene_summary_annotated, aes(x = n_CAAS, fill = Category)) +
  geom_histogram(binwidth = 0.05, color = "black", alpha = 0.8) +
  scale_fill_manual(values = category_palette, breaks = c("Extreme", "Dubious", "Both", "Normal")) +
  scale_x_log10(breaks = c(1, 2, 5, 10, 20, 50, 100, 200)) +
  facet_wrap(~CAAP_Group, scales = "free_y") +
  labs(title = "CAAP Count Distribution by Group",
       subtitle = "Dubious threshold (Q3 + 3×IQR) shown as dashed line",
       x = "Number of CAAP Positions", y = "Gene Count") +
  theme_minimal(base_size = 16) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 14),
    axis.title = element_text(size = 15),
    strip.text = element_text(face = "bold", size = 15)
  )
if (nrow(vline_dubious_data) > 0) {
  p_gross_dist <- p_gross_dist +
    geom_vline(data = vline_dubious_data, aes(xintercept = threshold_dubious),
               color = tableau_10[["blue"]], linetype = "dashed", linewidth = 1)
}

p_zoom_gross <- p_gross_dist + coord_cartesian(ylim = c(0, 100)) + 
  labs(title = "Zoomed View (y ≤ 100)")

p_density_dist <- ggplot(gene_summary_annotated, aes(x = n_CAAS_per_length, fill = Category)) +
  geom_histogram(binwidth = 0.05, color = "black", alpha = 0.8) +
  scale_fill_manual(values = category_palette, breaks = c("Extreme", "Dubious", "Both", "Normal")) +
  scale_x_log10(breaks = c(0.01, 0.02, 0.05, 0.1, 0.2, 0.5, 1, 2, 5, 10, 20, 50, 100)) +
  facet_wrap(~ CAAP_Group, scales = "free_y") +
  labs(title = "CAAP Density Distribution by Group",
       subtitle = paste0("Top ", 100*(1-extreme_threshold), "% density threshold shown as dashed line"),
       x = "CAAP Density (positions / gene length)", y = "Gene Count") +
  theme_minimal(base_size = 16) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 14),
    axis.title = element_text(size = 15),
    strip.text = element_text(face = "bold", size = 15)
  )
if (nrow(vline_extreme_data) > 0) {
  p_density_dist <- p_density_dist +
    geom_vline(data = vline_extreme_data, aes(xintercept = threshold_extreme),
               color = tableau_10[["vermillion"]], linetype = "dashed", linewidth = 1)
}

p_zoom_density <- p_density_dist + coord_cartesian(ylim = c(0, 50)) + 
  labs(title = "Zoomed View (y ≤ 50)")

combined_outlier_plot <- p_gross_dist / p_density_dist
combined_zoomed_plot <- p_zoom_gross / p_zoom_density

print(combined_outlier_plot)
print(combined_zoomed_plot)

ggsave(file.path(outliers_dir, "outlier_distributions.png"), combined_outlier_plot, width = 14, height = 10, dpi = 320)
ggsave(file.path(outliers_dir, "outlier_distributions_zoomed.png"), combined_zoomed_plot, width = 14, height = 10, dpi = 320)

# Dedicated bar chart: removed gene counts per category and CAAP_Group
# Clearly separates Extreme-only, Dubious-only, and Both (genes failing both criteria).
removed_cat_summary <- gene_summary_annotated %>%
  filter(Category != "Normal") %>%
  count(CAAP_Group, Category)

if (nrow(removed_cat_summary) > 0) {
  p_removed_cats <- ggplot(removed_cat_summary,
                           aes(x = Category, y = n, fill = Category)) +
    geom_col(color = "black", linewidth = 0.3, width = 0.7) +
    facet_wrap(~ CAAP_Group, scales = "free_y") +
    scale_fill_manual(
      values = category_palette,
      breaks = c("Extreme", "Dubious", "Both")
    ) +
    labs(
      title    = "Removed Genes by Category and Group",
      subtitle = paste0(
        "Extreme: top ", 100 * (1 - extreme_threshold), "% density  |  ",
        "Dubious: Q3 + ", iqr_multiplier, "\u00d7IQR count  |  ",
        "Both: both criteria"
      ),
      x = "Removal Category",
      y = "Gene Count"
    ) +
    theme_minimal(base_size = 16) +
    theme(
      legend.position  = "none",
      plot.title       = element_text(face = "bold", size = 18),
      plot.subtitle    = element_text(size = 14),
      axis.title       = element_text(size = 15),
      strip.text       = element_text(face = "bold", size = 15)
    )

  print(p_removed_cats)
  ggsave(file.path(outliers_dir, "removed_genes_by_category.png"),
         p_removed_cats, width = 10, height = 6, dpi = 320)
}

cat("\n=== Outlier Detection Summary ===\n")
cat("CAAP/US unified mode: Group-aware thresholds calculated\n\n")

thresholds_summary <- gene_summary_annotated %>% distinct(CAAP_Group, threshold_extreme, threshold_dubious) %>% arrange(CAAP_Group)
cat("Thresholds by group:\n")
for (i in 1:nrow(thresholds_summary)) {
  cat(sprintf("  %s:\n", thresholds_summary$CAAP_Group[i]))
  cat(sprintf("    Extreme (top %.0f%% density): %.4f\n", 100*(1-extreme_threshold), thresholds_summary$threshold_extreme[i]))
  cat(sprintf("    Dubious (Q3 + %.1f×IQR): %.0f positions\n\n", iqr_multiplier, thresholds_summary$threshold_dubious[i]))
}

outlier_counts <- gene_summary_annotated %>% count(CAAP_Group, Category) %>% pivot_wider(names_from = Category, values_from = n, values_fill = 0)
cat("Gene counts by group and category:\n")
print(outlier_counts)

write_tsv(gene_summary_annotated, file.path(outliers_dir, "gene_summary_annotated.tsv"))
cat("\n✓ Gene summary exported\n\n")

} # end if (nrow(gene_summary_annotated) > 0)
```

## Cluster Filtering Analysis

```{r cluster_analysis_load}
cat("Loading cluster filter summary...\n")
filter_summary <- read_tsv(filter_summary_file, show_col_types = FALSE)
cat("  Parameter combinations tested:", nrow(filter_summary), "\n")

if (!"Minlen" %in% names(filter_summary) || !"Maxcaas" %in% names(filter_summary)) {
  if ("Parameter" %in% names(filter_summary)) {
    parsed_params <- stringr::str_split_fixed(as.character(filter_summary$Parameter), "/", 2)
    if (!"Minlen" %in% names(filter_summary)) filter_summary$Minlen <- suppressWarnings(as.integer(parsed_params[, 1]))
    if (!"Maxcaas" %in% names(filter_summary)) filter_summary$Maxcaas <- suppressWarnings(as.numeric(parsed_params[, 2]))
    cat("  Parsed missing Minlen/Maxcaas from Parameter column\n")
  }
}

if (!"DiscardedCount" %in% names(filter_summary)) {
  protected_cols <- c("Parameter", "Minlen", "Maxcaas", "CAAP_Groups", "Groups_List", "N_Groups")
  candidate_cols <- setdiff(names(filter_summary), protected_cols)
  numeric_candidate_cols <- candidate_cols[vapply(candidate_cols, function(col) is.numeric(filter_summary[[col]]), logical(1))]
  if (length(numeric_candidate_cols) > 0) {
    filter_summary <- filter_summary %>% mutate(DiscardedCount = rowSums(across(all_of(numeric_candidate_cols)), na.rm = TRUE))
    cat("  Built DiscardedCount from columns:", paste(numeric_candidate_cols, collapse = ", "), "\n")
  } else {
    filter_summary$DiscardedCount <- 0L
    cat("  Warning: Could not infer DiscardedCount from summary columns; defaulting to 0\n")
  }
} else {
  filter_summary$DiscardedCount <- suppressWarnings(as.numeric(filter_summary$DiscardedCount))
  filter_summary$DiscardedCount[is.na(filter_summary$DiscardedCount)] <- 0
}

has_caap_groups_in_summary <- 'CAAP_Groups' %in% names(filter_summary) && any(filter_summary$CAAP_Groups != 'N/A' & filter_summary$CAAP_Groups != '')
if (has_caap_groups_in_summary) {
  cat("  CAAP mode detected in filter summary\n")
  filter_summary <- filter_summary %>% mutate(Groups_List = strsplit(as.character(CAAP_Groups), ","), N_Groups = lengths(Groups_List))
}

if (has_caap_groups_in_summary) {
  group_cols <- setdiff(names(filter_summary), c('Parameter', 'Minlen', 'Maxcaas'))
  filter_table <- filter_summary %>% select(Parameter, all_of(sort(group_cols)))
} else {
  filter_table <- filter_summary %>% select(Parameter, DiscardedCount)
  names(filter_table)[2] <- "Discarded"
}

kable(filter_table, caption = "Cluster Filtering Parameter Sweep Results", align = c("l", rep("r", ncol(filter_table) - 1))) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

if (nrow(filter_summary) > 1) {
  if (has_caap_groups_in_summary) {
    group_cols <- setdiff(names(filter_summary), c('Parameter', 'Minlen', 'Maxcaas'))
    plot_data <- filter_summary %>%
      pivot_longer(cols = all_of(group_cols), names_to = "CAAP_Group", values_to = "DiscardedCount") %>%
      mutate(param_label = sprintf("%d/%d", Minlen, as.integer(Maxcaas)), param_label = factor(param_label, levels = unique(param_label)))

    p_param_sweep <- ggplot(plot_data, aes(x = param_label, y = DiscardedCount, fill = CAAP_Group)) +
      geom_col(position = "dodge", color = "black", linewidth = 0.3) +
      scale_fill_paletteer_d("ggthemes::Tableau_10") +
      labs(title = "Discarded CAAS Positions by Parameter Combination",
           subtitle = "Grouped by CAAP group - minlen/maxcaas combinations",
           x = "Parameter Combination (minlen/maxcaas)", y = "Positions Discarded", fill = "CAAP Group") +
      theme_minimal(base_size = 16) +
      theme(legend.position = "right", plot.title = element_text(face = "bold", size = 18), axis.title = element_text(size = 15), axis.text.x = element_text(angle = 45, hjust = 1))
  } else {
    p_param_sweep <- ggplot(filter_summary, aes(x = factor(Minlen), y = DiscardedCount, color = factor(Maxcaas), group = factor(Maxcaas))) +
      geom_point(size = 4) +
      geom_line(linewidth = 1.2) +
      scale_color_paletteer_d("ggthemes::Tableau_10", labels = function(x) sprintf("%.0f%%", as.numeric(x))) +
      labs(title = "Discarded CAAS Positions by Parameter Combination",
           subtitle = "Effect of minimum cluster length and maximum density threshold",
           x = "Minimum Cluster Length (minlen)", y = "Positions Discarded", color = "Max Density") +
      theme_minimal(base_size = 16) +
      theme(legend.position = "right", plot.title = element_text(face = "bold", size = 18), axis.title = element_text(size = 15))
  }

  print(p_param_sweep)
  ggsave(file.path(cluster_dir, "parameter_sweep_overview.png"), p_param_sweep, width = 12, height = 6, dpi = 320)
}

write_tsv(filter_summary, file.path(cluster_dir, "filter_summary.tsv"))
cat("✓ Filter summary exported\n\n")
```

```{r cluster_detailed_analysis, results='asis', fig.height=8, fig.width=12, eval=(processing_mode == 'exploratory' && nrow(filter_summary) > 1)}
if (processing_mode == 'exploratory' && nrow(filter_summary) > 1) {
  cat("\n## Detailed Parameter Analysis\n\n")

  available_filtered_files <- list.files(
    path = getwd(),
    pattern = "^discovery\\.filtered\\.minlen[0-9]+\\.maxcaas[0-9]+\\.tsv$",
    recursive = TRUE,
    full.names = TRUE
  )

  for (i in 1:nrow(filter_summary)) {
    param_row <- filter_summary[i, ]
    param_label <- sprintf("minlen=%d, maxcaas=%.2f", param_row$Minlen, param_row$Maxcaas)
    cat(sprintf("\n### Parameter Set %d: %s {.tabset}\n\n", i, param_label))

    pattern_str <- sprintf("discovery.filtered.minlen%d.maxcaas%d.tsv", param_row$Minlen, as.integer(param_row$Maxcaas))
    matching_file <- available_filtered_files[basename(available_filtered_files) == pattern_str]

    if (length(matching_file) > 0 && file.exists(matching_file[1])) {
      filtered_data <- read_tsv(matching_file[1], show_col_types = FALSE)
      if ("ClusteringFlag" %in% names(filtered_data)) {
        discarded <- filtered_data %>% filter(ClusteringFlag == "Discarded")
        if (nrow(discarded) > 0 && "CAAP_Group" %in% names(discarded)) {
          detect_runs <- function(positions) {
            positions <- sort(unique(positions))
            if (length(positions) == 0) return(integer(0))
            breaks <- c(0, which(diff(positions) > 1), length(positions))
            diff(breaks)
          }

          spatial_analysis <- discarded %>%
            group_by(CAAP_Group, Gene) %>%
            summarise(positions = list(Position), .groups = "drop") %>%
            mutate(run_lengths = map(positions, detect_runs)) %>%
            unnest(run_lengths, keep_empty = TRUE) %>%
            mutate(
              cluster_type = case_when(
                run_lengths == 1 ~ "Singlet (1)",
                run_lengths == 2 ~ "Doublet (2)",
                run_lengths == 3 ~ "Triplet (3)",
                run_lengths == 4 ~ "Quad (4)",
                run_lengths == 5 ~ "Quint (5)",
                run_lengths >= 6 ~ "6+ Long",
                TRUE ~ "Other"
              )
            )

          spatial_summary <- spatial_analysis %>%
            count(CAAP_Group, cluster_type) %>%
            complete(CAAP_Group, cluster_type, fill = list(n = 0)) %>%
            group_by(CAAP_Group) %>%
            mutate(percentage = round(100 * n / sum(n), 1)) %>%
            ungroup()

          p_spatial <- ggplot(spatial_summary, aes(x = CAAP_Group, y = percentage/100, fill = cluster_type)) +
            geom_col(position = "fill", color = "black", size = 0.2) +
            scale_fill_paletteer_d("ggthemes::Tableau_10") +
            scale_y_continuous(labels = scales::percent) +
            labs(title = sprintf("Spatial Clustering: %s", param_label), x = "Group", y = "Percentage", fill = "Cluster Type") +
            theme_minimal(base_size = 16) +
            theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1))

          print(p_spatial)
        }
      }
    }
    cat("\n---\n\n")
  }
}
```

# B) Characterizing CAAS disambiguation

## B.1 Build unified removal + retention dataset

```{r disambig_unified_data}
filter_root <- params$filter_dir
postproc_root <- if (!is.null(filter_root) && !is.na(filter_root)) dirname(filter_root) else NA_character_

removed_precluster_file <- find_first_existing(c(
  file.path(filter_root, "preprocessed", "removed_patterns_precluster.tsv"),
  file.path(postproc_root, "preprocessed", "removed_patterns_precluster.tsv"),
  find_recursive_file(getwd(), "removed_patterns_precluster\\.tsv$")
))

removed_genes_file <- find_first_existing(c(
  file.path(filter_root, "gene_filtering", "removed_genes_summary.tsv"),
  file.path(postproc_root, "gene_filtering", "removed_genes_summary.tsv"),
  find_recursive_file(getwd(), "removed_genes_summary\\.tsv$")
))

cluster_files <- list.files(path = params$filter_dir, pattern = "\\.filtered\\.minlen[0-9]+\\.maxcaas[0-9]+\\.tsv$", full.names = TRUE, recursive = TRUE)
if (length(cluster_files) == 0) {
  cluster_files <- list.files(path = ifelse(is.na(postproc_root), getwd(), postproc_root), pattern = "\\.filtered\\.minlen[0-9]+\\.maxcaas[0-9]+\\.tsv$", full.names = TRUE, recursive = TRUE)
}

precluster_events <- tibble()
if (!is.na(removed_precluster_file)) {
  precluster_df <- normalize_disambig_schema(read_delim(removed_precluster_file, delim = "\t", show_col_types = FALSE))
  precluster_events <- precluster_df %>% transmute(site_id, gene, position, caap_group, pattern_type, pvalue, pvalue_boot, sig_hyp, sig_perm, sig_both, significance_class, RemovalStage = "precluster_pattern")
}

cluster_events <- tibble()
if (length(cluster_files) > 0) {
  cluster_df <- map_dfr(cluster_files, ~ read_tsv(.x, show_col_types = FALSE))
  cluster_hits <- cluster_df %>% filter(ClusteringFlag == "Discarded") %>% transmute(gene = as.character(Gene), position = as.integer(Position), caap_group = as.character(CAAP_Group)) %>% distinct()
  cluster_events <- discovery_std %>% inner_join(cluster_hits, by = c("gene", "position", "caap_group")) %>% transmute(site_id, gene, position, caap_group, pattern_type, pvalue, pvalue_boot, sig_hyp, sig_perm, sig_both, significance_class, RemovalStage = "cluster") %>% distinct()
}

gene_events <- tibble()
if (!is.na(removed_genes_file)) {
  rg <- read_tsv(removed_genes_file, show_col_types = FALSE)
  names(rg) <- tolower(names(rg))
  if (!"caap_group" %in% names(rg)) rg$caap_group <- "US"
  rg_norm <- rg %>% transmute(gene = as.character(gene), caap_group = as.character(caap_group), category = as.character(category)) %>% distinct()

  gene_events <- discovery_std %>%
    inner_join(rg_norm, by = c("gene", "caap_group")) %>%
    mutate(RemovalStage = case_when(
      tolower(category) == "extreme" ~ "extreme_gene",
      tolower(category) == "dubious" ~ "dubious_with_clusters",
      tolower(category) == "both" ~ "extreme_and_dubious",
      TRUE ~ "gene_filter"
    )) %>%
    transmute(site_id, gene, position, caap_group, pattern_type, pvalue, pvalue_boot, sig_hyp, sig_perm, sig_both, significance_class, RemovalStage) %>%
    distinct()
}

removal_events <- bind_rows(precluster_events, cluster_events, gene_events)
write_tsv(removal_events, file.path(disambig_dir, "removal_events.tsv"))

# Diagnostic counts for reporting
n_gs0 <- sum(toupper(discovery_std$caap_group) == "GS0")

cat("  GS0 rows:", n_gs0, "\n")
```

## B.1b Focal relationships

```{r focal_sets, results='asis'}

# Helper: summarise one row per (site_id, gene, position) with per-group sig flags.
# Flags: global (any presence), global_sig (sig_both in any group),
#        per group: <g> = sig_both, <g>_nondiv = sig_both & nondivergent, <g>_div = sig_both & divergent.
# Also derives change_side per site and top/bottom indicators.
compute_site_flags <- function(df) {
  df %>%
    mutate(
      caap_group_up = toupper(caap_group),
      is_nondiv     = pattern_type != "divergent",
      is_div        = pattern_type == "divergent"
    ) %>%
    group_by(site_id, gene, position) %>%
    summarise(
      global        = any(!is.na(caap_group_up), na.rm = TRUE),
      global_sig    = any(sig_both, na.rm = TRUE),
      us            = any(caap_group_up == "US"  & sig_both, na.rm = TRUE),
      us_nondiv     = any(caap_group_up == "US"  & is_nondiv & sig_both, na.rm = TRUE),
      us_div        = any(caap_group_up == "US"  & is_div    & sig_both, na.rm = TRUE),
      gs            = any(caap_group_up != "US"  & sig_both, na.rm = TRUE),
      gs_nondiv     = any(caap_group_up != "US"  & is_nondiv & sig_both, na.rm = TRUE),
      gs_div        = any(caap_group_up != "US"  & is_div    & sig_both, na.rm = TRUE),
      gs0           = any(caap_group_up == "GS0" & sig_both, na.rm = TRUE),
      gs0_nondiv    = any(caap_group_up == "GS0" & is_nondiv & sig_both, na.rm = TRUE),
      gs0_div       = any(caap_group_up == "GS0" & is_div    & sig_both, na.rm = TRUE),
      gs1           = any(caap_group_up == "GS1" & sig_both, na.rm = TRUE),
      gs1_nondiv    = any(caap_group_up == "GS1" & is_nondiv & sig_both, na.rm = TRUE),
      gs1_div       = any(caap_group_up == "GS1" & is_div    & sig_both, na.rm = TRUE),
      gs2           = any(caap_group_up == "GS2" & sig_both, na.rm = TRUE),
      gs2_nondiv    = any(caap_group_up == "GS2" & is_nondiv & sig_both, na.rm = TRUE),
      gs2_div       = any(caap_group_up == "GS2" & is_div    & sig_both, na.rm = TRUE),
      gs3           = any(caap_group_up == "GS3" & sig_both, na.rm = TRUE),
      gs3_nondiv    = any(caap_group_up == "GS3" & is_nondiv & sig_both, na.rm = TRUE),
      gs3_div       = any(caap_group_up == "GS3" & is_div    & sig_both, na.rm = TRUE),
      gs4           = any(caap_group_up == "GS4" & sig_both, na.rm = TRUE),
      gs4_nondiv    = any(caap_group_up == "GS4" & is_nondiv & sig_both, na.rm = TRUE),
      gs4_div       = any(caap_group_up == "GS4" & is_div    & sig_both, na.rm = TRUE),
      change_side   = collapse_change_side(change_side),
      .groups = "drop"
    ) %>%
    mutate(
      convergent_pool = us_nondiv | gs_nondiv,
      divergent_pool  = us_div    | gs_div,
      is_top    = change_side %in% c("top",    "both"),
      is_bottom = change_side %in% c("bottom", "both")
    )
}

# Build site flags with GS0 included and without
site_flags       <- compute_site_flags(discovery_std)
site_flags_no_gs0 <- compute_site_flags(discovery_std %>% filter(toupper(caap_group) != "GS0"))

# Helper: build a tidy summary table from a site_flags data frame.
# include_gs0 controls whether GS0 rows are included.
# filter_expr is a logical vector or NULL (NULL = all rows).
build_focal_table <- function(sf, include_gs0 = TRUE, filter_mask = NULL) {
  if (!is.null(filter_mask)) sf <- sf[filter_mask, ]

  grp_rows <- function(g, label) {
    gl <- g; nd <- paste0(g, "_nondiv"); dv <- paste0(g, "_div")
    tibble(
      Set             = c(paste0(label, " — global significant"),
                          paste0(label, " — convergent (nondivergent)"),
                          paste0(label, " — divergent")),
      Description     = c(paste0("Significant (sig_both) positions in ", label),
                          paste0("Significant nondivergent positions in ",  label),
                          paste0("Significant divergent positions in ",     label)),
      Positions_total = c(sum(sf[[gl]]),              sum(sf[[nd]]),              sum(sf[[dv]])),
      Genes_total     = c(n_distinct(sf$gene[sf[[gl]]]), n_distinct(sf$gene[sf[[nd]]]), n_distinct(sf$gene[sf[[dv]]]))
    )
  }

  pool_rows <- tibble(
    Set             = c("Global pool", "Global significant pool",
                        "Convergent pool (nondivergent)", "Divergent pool"),
    Description     = c("All positions in any group",
                        "Positions significant (sig_both) in any group",
                        "Positions with sig nondivergent change in US or GS",
                        "Positions with sig divergent change in US or GS"),
    Positions_total = c(sum(sf$global), sum(sf$global_sig),
                        sum(sf$convergent_pool), sum(sf$divergent_pool)),
    Genes_total     = c(n_distinct(sf$gene[sf$global]),         n_distinct(sf$gene[sf$global_sig]),
                        n_distinct(sf$gene[sf$convergent_pool]), n_distinct(sf$gene[sf$divergent_pool]))
  )

  gs_groups <- if (include_gs0) c("gs0","gs1","gs2","gs3","gs4") else c("gs1","gs2","gs3","gs4")
  gs_labels <- if (include_gs0) c("GS0","GS1","GS2","GS3","GS4") else c("GS1","GS2","GS3","GS4")

  bind_rows(pool_rows, grp_rows("us", "US"), map2_dfr(gs_groups, gs_labels, grp_rows))
}

# ── Six summary tables: {global | top | bottom} × {GS0 included | excluded} ───────

make_and_print_focal_tables <- function(sf, label_suffix, include_gs0) {
  # global (all significant positions)
  tbl_g <- build_focal_table(sf, include_gs0 = include_gs0)
  cat(sprintf("\n### %s — Global (all sig_both positions)\n\n", label_suffix))
  print(kable(tbl_g, caption = paste("Focal sets — global —", label_suffix)) %>%
    kable_styling(bootstrap_options = c("striped","hover","condensed")))
  fname <- tolower(paste0("focal_sets_global_", gsub(" ", "_", label_suffix), ".tsv"))
  write_tsv(tbl_g, file.path(disambig_dir, fname))

  # top (change_side == top or both)
  tbl_t <- build_focal_table(sf, include_gs0 = include_gs0, filter_mask = sf$is_top)
  cat(sprintf("\n### %s — Top (change_side = top/both)\n\n", label_suffix))
  print(kable(tbl_t, caption = paste("Focal sets — top —", label_suffix)) %>%
    kable_styling(bootstrap_options = c("striped","hover","condensed")))
  fname <- tolower(paste0("focal_sets_top_", gsub(" ", "_", label_suffix), ".tsv"))
  write_tsv(tbl_t, file.path(disambig_dir, fname))

  # bottom (change_side == bottom or both)
  tbl_b <- build_focal_table(sf, include_gs0 = include_gs0, filter_mask = sf$is_bottom)
  cat(sprintf("\n### %s — Bottom (change_side = bottom/both)\n\n", label_suffix))
  print(kable(tbl_b, caption = paste("Focal sets — bottom —", label_suffix)) %>%
    kable_styling(bootstrap_options = c("striped","hover","condensed")))
  fname <- tolower(paste0("focal_sets_bottom_", gsub(" ", "_", label_suffix), ".tsv"))
  write_tsv(tbl_b, file.path(disambig_dir, fname))

  invisible(list(global = tbl_g, top = tbl_t, bottom = tbl_b))
}

# With GS0
make_and_print_focal_tables(site_flags,        "GS0 included", include_gs0 = TRUE)
# Without GS0
make_and_print_focal_tables(site_flags_no_gs0, "GS0 excluded", include_gs0 = FALSE)

cat("✓ Focal sets tables exported\n")

```

## B.1c Gene-Level Filter Overlap Diagnostics

```{r filter_overlap_gene_table}
# ─────────────────────────────────────────────────────────────────────────────
# Build a gene-level membership table that records for each gene whether it
# was flagged by each of the three independent filters:
#   • Extreme   – top 1% CAAS density (gene_summary_annotated$Extreme)
#   • Dubious   – IQR-outlier CAAS count (gene_summary_annotated$Dubious)
#   • Cluster   – at least one position removed by the cluster filter
# ─────────────────────────────────────────────────────────────────────────────

# --- 1. Gene-level filter flags -------------------------------------------

# Extreme / Dubious come straight from the already-computed gene_summary_annotated
gene_filter_flags <- gene_summary_annotated %>%
  transmute(
    Gene        = Gene,
    CAAP_Group  = CAAP_Group,
    n_CAAS      = n_CAAS,
    Length      = Length,
    n_CAAS_per_length = n_CAAS_per_length,
    Extreme     = as.logical(Extreme),
    Dubious     = as.logical(Dubious)
  )

# Cluster_removed: any gene that contributed at least one site to cluster_events
cluster_removed_genes <- if (nrow(cluster_events) > 0) {
  cluster_events %>%
    transmute(Gene = gene, CAAP_Group = caap_group) %>%
    distinct() %>%
    mutate(Cluster_removed = TRUE)
} else {
  tibble(Gene = character(), CAAP_Group = character(), Cluster_removed = logical())
}

gene_filter_flags <- gene_filter_flags %>%
  left_join(cluster_removed_genes, by = c("Gene", "CAAP_Group")) %>%
  mutate(
    Cluster_removed = coalesce(Cluster_removed, FALSE),
    # Composite label
    filter_combo = case_when(
      Extreme & Dubious & Cluster_removed ~ "Extreme+Dubious+Cluster",
      Extreme & Dubious                  ~ "Extreme+Dubious",
      Extreme & Cluster_removed          ~ "Extreme+Cluster",
      Dubious & Cluster_removed          ~ "Dubious+Cluster",
      Extreme                            ~ "Extreme",
      Dubious                            ~ "Dubious",
      Cluster_removed                    ~ "Cluster",
      TRUE                               ~ "None"
    ),
    filter_combo = factor(filter_combo, levels = c(
      "None", "Extreme", "Dubious", "Cluster",
      "Extreme+Dubious", "Extreme+Cluster", "Dubious+Cluster",
      "Extreme+Dubious+Cluster"
    ))
  )

# --- 2. Per-gene significance descriptors from full discovery data ----------

gene_sig_stats <- discovery_std %>%
  group_by(gene, caap_group) %>%
  summarise(
    frac_sig_both    = mean(sig_both, na.rm = TRUE),
    median_pval_hyp  = median(pvalue,      na.rm = TRUE),
    min_pval_hyp     = min(pvalue,         na.rm = TRUE),
    median_pval_boot = median(pvalue_boot, na.rm = TRUE),
    min_pval_boot    = min(pvalue_boot,    na.rm = TRUE),
    .groups = "drop"
  ) %>%
  rename(Gene = gene, CAAP_Group = caap_group)

gene_filter_flags <- gene_filter_flags %>%
  left_join(gene_sig_stats, by = c("Gene", "CAAP_Group"))

# --- 3. Quick membership summary printed to console ------------------------

cat("\n=== Gene-Level Filter Overlap Summary ===\n")
membership_counts <- gene_filter_flags %>%
  count(CAAP_Group, filter_combo, .drop = FALSE) %>%
  arrange(CAAP_Group, filter_combo)
print(membership_counts)

cat("\nFilter-specific totals:\n")
cat(sprintf("  Extreme only or combined : %d genes\n",
    sum(gene_filter_flags$Extreme, na.rm = TRUE)))
cat(sprintf("  Dubious only or combined : %d genes\n",
    sum(gene_filter_flags$Dubious, na.rm = TRUE)))
cat(sprintf("  Cluster only or combined : %d genes\n",
    sum(gene_filter_flags$Cluster_removed, na.rm = TRUE)))
cat(sprintf("  Any filter               : %d / %d genes (%.1f%%)\n",
    sum(gene_filter_flags$filter_combo != "None"),
    nrow(gene_filter_flags),
    100 * mean(gene_filter_flags$filter_combo != "None")))

# Fraction of sig-both sites removed by each filter pathway
if (nrow(removal_events) > 0) {
  total_sig_both  <- sum(discovery_std$sig_both, na.rm = TRUE)
  removed_extreme <- discovery_std %>%
    filter(sig_both, gene %in% filter(gene_filter_flags, Extreme)$Gene,
           caap_group %in% filter(gene_filter_flags, Extreme)$CAAP_Group)
  removed_dubious <- discovery_std %>%
    filter(sig_both, gene %in% filter(gene_filter_flags, Dubious)$Gene,
           caap_group %in% filter(gene_filter_flags, Dubious)$CAAP_Group)
  removed_cluster <- discovery_std %>%
    filter(sig_both, site_id %in% cluster_events$site_id)
  cat(sprintf("\n  Sig-both sites in total  : %d\n", total_sig_both))
  cat(sprintf("  Sig-both in Extreme genes: %d (%.1f%%)\n",
      nrow(removed_extreme), 100 * nrow(removed_extreme) / max(total_sig_both, 1)))
  cat(sprintf("  Sig-both in Dubious genes: %d (%.1f%%)\n",
      nrow(removed_dubious), 100 * nrow(removed_dubious) / max(total_sig_both, 1)))
  cat(sprintf("  Sig-both in Cluster sites: %d (%.1f%%)\n",
      nrow(removed_cluster), 100 * nrow(removed_cluster) / max(total_sig_both, 1)))
}

cat("\n")
```

## B.1c-i UpSet plot of filter intersections

```{r filter_overlap_upset, fig.width=10, fig.height=6}
# Build a binary incidence data frame suitable for UpSetR
upset_df <- gene_filter_flags %>%
  transmute(
    Extreme  = as.integer(Extreme),
    Dubious  = as.integer(Dubious),
    Cluster  = as.integer(Cluster_removed)
  ) %>%
  as.data.frame()

# Guard: only proceed if at least two filters have any hits
n_active_filters <- sum(c(
  any(upset_df$Extreme == 1),
  any(upset_df$Dubious == 1),
  any(upset_df$Cluster == 1)
))

if (n_active_filters < 1) {
  cat("No genes flagged by any filter — UpSet plot skipped.\n")
} else if (!requireNamespace("UpSetR", quietly = TRUE)) {
  cat("Package 'UpSetR' not available — falling back to membership table.\n")
  kable(membership_counts, caption = "Gene counts per filter combination") %>% kable_styling()
} else {
  # One UpSet per CAAP group (or single global plot when only one group)
  groups_present <- sort(unique(gene_filter_flags$CAAP_Group))

  for (grp in groups_present) {
    grp_df <- gene_filter_flags %>%
      filter(CAAP_Group == grp) %>%
      transmute(
        Extreme = as.integer(Extreme),
        Dubious = as.integer(Dubious),
        Cluster = as.integer(Cluster_removed)
      ) %>%
      as.data.frame()

    # Drop columns with zero variance (no genes hit by that filter in this group)
    active_cols <- names(grp_df)[colSums(grp_df, na.rm = TRUE) > 0]
    if (length(active_cols) < 1) {
      cat(sprintf("  Group %s: no filter hits — skipping.\n", grp))
      next
    }
    grp_df <- grp_df[, active_cols, drop = FALSE]

    plot_title <- sprintf("Filter overlap — %s (n = %d genes)",
                          grp, nrow(grp_df))
    cat(sprintf("\n**%s**\n\n", plot_title))

    # UpSetR::upset returns a grid object; use grid::grid.draw inside a device call
    png_path <- file.path(outliers_dir,
                          sprintf("filter_overlap_upset_%s.png", tolower(grp)))

    # Render to file
    png(png_path, width = 2800, height = 1600, res = 320)
    tryCatch(
      print(UpSetR::upset(
        grp_df,
        sets            = rev(active_cols),
        order.by        = "freq",
        decreasing      = TRUE,
        text.scale      = c(1.4, 1.2, 1.2, 1.0, 1.2, 1.0),
        mainbar.y.label = "Gene Intersections",
        sets.x.label    = "Genes per filter",
        mb.ratio        = c(0.60, 0.40),
        main.bar.color  = "#4472C4",
        sets.bar.color  = "#ED7D31"
      )),
      error = function(e) cat("UpSet render error:", conditionMessage(e), "\n")
    )
    dev.off()

    # Also render inline
    tryCatch(
      print(UpSetR::upset(
        grp_df,
        sets            = rev(active_cols),
        order.by        = "freq",
        decreasing      = TRUE,
        text.scale      = c(1.4, 1.2, 1.2, 1.0, 1.2, 1.0),
        mainbar.y.label = "Gene Intersections",
        sets.x.label    = "Genes per filter",
        mb.ratio        = c(0.60, 0.40),
        main.bar.color  = "#4472C4",
        sets.bar.color  = "#ED7D31"
      )),
      error = function(e) cat("UpSet inline error:", conditionMessage(e), "\n")
    )
  }
}
```

## B.1c-ii Descriptor distributions by filter membership

```{r filter_overlap_stats, fig.width=14, fig.height=16}
# Ordered factor for x-axis: None → single → double → triple
combo_levels <- c(
  "None", "Extreme", "Dubious", "Cluster",
  "Extreme+Dubious", "Extreme+Cluster", "Dubious+Cluster",
  "Extreme+Dubious+Cluster"
)

gene_filter_plot <- gene_filter_flags %>%
  mutate(filter_combo = factor(as.character(filter_combo), levels = combo_levels))

# Colour palette for combos
combo_palette <- setNames(
  c(
    tableau_10[["gray"]],
    tableau_10[["vermillion"]], tableau_10[["blue"]], tableau_10[["green"]],
    tableau_10[["reddish_purple"]], tableau_10[["orange"]], tableau_10[["bluish_green"]],
    tableau_10[["pink"]]
  ),
  combo_levels
)

make_descriptor_plot <- function(data, y_var, y_label, log_scale = FALSE) {
  plot_data <- data %>% filter(!is.na(.data[[y_var]]), !is.na(CAAP_Group))
  if (nrow(plot_data) == 0 || length(unique(plot_data$CAAP_Group)) == 0) {
    return(
      ggplot() +
        annotate("text", x = 0.5, y = 0.5, label = paste("No data for:", y_label),
                 size = 5, hjust = 0.5) +
        theme_void()
    )
  }
  p <- ggplot(plot_data,
              aes(x = filter_combo, y = .data[[y_var]],
                  fill = filter_combo, colour = filter_combo)) +
    geom_boxplot(alpha = 0.5, outlier.shape = NA, linewidth = 0.6) +
    geom_jitter(width = 0.18, alpha = 0.35, size = 0.9, shape = 16) +
    facet_wrap(~ CAAP_Group, scales = "free_y") +
    scale_fill_manual(values   = combo_palette, drop = TRUE) +
    scale_colour_manual(values = combo_palette, drop = TRUE) +
    labs(x = NULL, y = y_label) +
    theme_minimal(base_size = 12) +
    theme(
      legend.position  = "none",
      axis.text.x      = element_text(angle = 40, hjust = 1, size = 9),
      strip.text       = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
  if (log_scale) p <- p + scale_y_log10()
  p
}

p_length     <- make_descriptor_plot(gene_filter_plot, "Length",           "Gene length (aa)",              log_scale = TRUE)
p_ncaas      <- make_descriptor_plot(gene_filter_plot, "n_CAAS",           "Total CAAS count",              log_scale = TRUE)
p_density    <- make_descriptor_plot(gene_filter_plot, "n_CAAS_per_length","CAAS density (per 100 aa)",     log_scale = TRUE)
p_fracsig    <- make_descriptor_plot(gene_filter_plot, "frac_sig_both",    "Fraction sig-both sites",       log_scale = FALSE)
p_medpval    <- make_descriptor_plot(gene_filter_plot, "median_pval_hyp",  "Median hyperg. p-value",        log_scale = FALSE)
p_minpval    <- make_descriptor_plot(gene_filter_plot, "min_pval_hyp",     "Min hyperg. p-value",           log_scale = FALSE)

desc_grid <- (
  p_length  | p_ncaas  | p_density
) / (
  p_fracsig | p_medpval | p_minpval
) +
  plot_annotation(
    title    = "Gene descriptor distributions by filter membership",
    subtitle = sprintf(
      "Extreme = top %.0f%% density  |  Dubious = Q3 + %.1f×IQR count  |  Cluster = cluster-filtered",
      100 * (1 - extreme_threshold), iqr_multiplier
    ),
    theme = theme(plot.title = element_text(face = "bold", size = 16),
                  plot.subtitle = element_text(size = 12))
  )

print(desc_grid)
ggsave(file.path(outliers_dir, "filter_overlap_descriptors.png"),
       desc_grid, width = 14, height = 10, dpi = 280)
```

## B.1c-iii Filter overlap summary table

```{r filter_overlap_summary_table}
# Per-combo × per-group mean ± SD for all five descriptors
overlap_summary <- gene_filter_flags %>%
  group_by(CAAP_Group, filter_combo) %>%
  summarise(
    n_genes            = n(),
    mean_Length        = round(mean(Length,            na.rm = TRUE), 1),
    sd_Length          = round(sd(Length,              na.rm = TRUE), 1),
    mean_n_CAAS        = round(mean(n_CAAS,            na.rm = TRUE), 2),
    sd_n_CAAS          = round(sd(n_CAAS,              na.rm = TRUE), 2),
    mean_density       = round(mean(n_CAAS_per_length, na.rm = TRUE), 4),
    sd_density         = round(sd(n_CAAS_per_length,   na.rm = TRUE), 4),
    mean_frac_sig_both = round(mean(frac_sig_both,     na.rm = TRUE), 4),
    sd_frac_sig_both   = round(sd(frac_sig_both,       na.rm = TRUE), 4),
    mean_median_pval   = round(mean(median_pval_hyp,   na.rm = TRUE), 4),
    sd_median_pval     = round(sd(median_pval_hyp,     na.rm = TRUE), 4),
    .groups = "drop"
  ) %>%
  arrange(CAAP_Group, filter_combo)

kable(
  overlap_summary,
  caption = "Gene-level filter overlap: descriptive statistics per membership group"
) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 11)

write_tsv(gene_filter_flags,  file.path(outliers_dir, "filter_overlap_gene_table.tsv"))
write_tsv(overlap_summary,    file.path(outliers_dir, "filter_overlap_summary.tsv"))
cat("✓ Filter overlap tables exported\n")
```

## B.2 Removed positions by `pattern_type` and significance

```{r removed_pattern_sig}
if (nrow(removal_events) > 0) {
  by_pattern <- removal_events %>% count(RemovalStage, pattern_type, sort = TRUE)
  by_sig <- removal_events %>%
    mutate(sig_label = case_when(sig_both ~ "both<=0.05", sig_hyp ~ "hyp<=0.05", sig_perm ~ "perm<=0.05", TRUE ~ "not_sig")) %>%
    count(RemovalStage, sig_label, sort = TRUE)

  kable(by_pattern, caption = "Removed positions by stage and pattern_type") %>% kable_styling()
  kable(by_sig, caption = "Removed positions by stage and significance") %>% kable_styling()

  p_rm <- ggplot(by_pattern, aes(pattern_type, n, fill = RemovalStage)) +
    geom_col(position = "dodge") +
    scale_fill_paletteer_d("ggthemes::Tableau_10") +
    theme_minimal(base_size = 16) +
    labs(title = "Removed positions: pattern_type composition (all)", x = "pattern_type") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  print(p_rm)
  ggsave(file.path(disambig_dir, "removed_pattern_type_all.png"), p_rm, width = 10, height = 5)
  
  # Focus on significant-both positions only
  removal_sig_both <- removal_events %>% filter(sig_both)
  
  if (nrow(removal_sig_both) > 0) {
    by_pattern_sig_both <- removal_sig_both %>% count(RemovalStage, pattern_type, sort = TRUE)
    
    kable(by_pattern_sig_both, caption = "Removed positions by stage and pattern_type (SIGNIFICANT-BOTH only)") %>% kable_styling()
    
    p_rm_sig_both <- ggplot(by_pattern_sig_both, aes(pattern_type, n, fill = RemovalStage)) +
      geom_col(position = "dodge") +
      scale_fill_paletteer_d("ggthemes::Tableau_10") +
      theme_minimal(base_size = 16) +
      labs(title = "Removed positions: pattern_type composition (SIGNIFICANT-BOTH only)", 
           subtitle = paste0("n = ", sum(by_pattern_sig_both$n), " significant-both removed positions"),
           x = "pattern_type") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
    print(p_rm_sig_both)
    ggsave(file.path(disambig_dir, "removed_pattern_type_sig_both.png"), p_rm_sig_both, width = 10, height = 5)
  } else {
    cat("No significant-both positions in removal events.\n")
  }
}
```

## B.2b P-value distributions: removed vs retained positions

```{r pvalue_removed_vs_retained, fig.width=12, fig.height=9}
if (nrow(discovery_std) > 0) {
  removed_site_ids <- if (nrow(removal_events) > 0) unique(removal_events$site_id) else character(0)

  pval_df <- discovery_std %>%
    mutate(
      filter_status = if_else(site_id %in% removed_site_ids, "removed", "retained"),
      filter_status = factor(filter_status, levels = c("retained", "removed"))
    ) %>%
    filter(!is.na(pvalue) | !is.na(pvalue_boot)) %>%
    pivot_longer(
      cols      = c(pvalue, pvalue_boot),
      names_to  = "pval_type",
      values_to = "p"
    ) %>%
    filter(!is.na(p)) %>%
    mutate(
      pval_type = recode(pval_type,
        pvalue      = "Hypothesis p-value",
        pvalue_boot = "Permutation p-value (boot)"
      )
    )

  status_colours <- c(retained = "#2196F3", removed = "#F44336")

  # ── Plot 1: global double-layered histogram ─────────────────────────────────
  p_glob <- ggplot(pval_df, aes(x = p, fill = filter_status, colour = filter_status)) +
    geom_histogram(
      aes(y = after_stat(density)),
      bins      = 40,
      alpha     = 0.55,
      position  = "identity",
      linewidth = 0.25
    ) +
    geom_vline(xintercept = 0.05, linetype = "dashed", colour = "grey30", linewidth = 0.4) +
    facet_wrap(~pval_type, scales = "free_y", ncol = 2) +
    scale_fill_manual(values = status_colours, name = "Status") +
    scale_colour_manual(values = status_colours, name = "Status") +
    scale_x_continuous(breaks = c(0, 0.05, 0.25, 0.5, 0.75, 1), limits = c(0, 1)) +
    labs(
      title    = "P-value distribution: removed vs retained positions (all)",
      subtitle = sprintf(
        "%d retained · %d removed | dashed line = 0.05",
        sum(pval_df$pval_type == "Hypothesis p-value" & pval_df$filter_status == "retained"),
        sum(pval_df$pval_type == "Hypothesis p-value" & pval_df$filter_status == "removed")
      ),
      x = "p-value", y = "Density"
    ) +
    theme_minimal(base_size = 16) +
    theme(
      legend.position  = "bottom",
      strip.text       = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
  print(p_glob)
  ggsave(file.path(disambig_dir, "pvalue_removed_vs_retained_global.png"),
         p_glob, width = 12, height = 5, dpi = 160)

  # ── Plot 2: same, faceted by pattern_type ───────────────────────────────────
  pt_counts <- pval_df %>%
    filter(pval_type == "Hypothesis p-value") %>%
    count(pattern_type, filter_status) %>%
    pivot_wider(names_from = filter_status, values_from = n, values_fill = 0)

  pval_df_pt <- pval_df %>%
    filter(pattern_type %in% (
      pval_df %>%
        filter(pval_type == "Hypothesis p-value") %>%
        count(pattern_type) %>%
        filter(n >= 5) %>%
        pull(pattern_type)
    ))

  if (nrow(pval_df_pt) > 0 && n_distinct(pval_df_pt$pattern_type) >= 1) {
    p_pt <- ggplot(pval_df_pt, aes(x = p, fill = filter_status, colour = filter_status)) +
      geom_histogram(
        aes(y = after_stat(density)),
        bins      = 30,
        alpha     = 0.55,
        position  = "identity",
        linewidth = 0.25
      ) +
      geom_vline(xintercept = 0.05, linetype = "dashed", colour = "grey30", linewidth = 0.4) +
      facet_grid(pattern_type ~ pval_type, scales = "free_y") +
      scale_fill_manual(values = status_colours, name = "Status") +
      scale_colour_manual(values = status_colours, name = "Status") +
      scale_x_continuous(breaks = c(0, 0.05, 0.5, 1), limits = c(0, 1)) +
      labs(
        title = "P-value distribution by pattern_type: removed vs retained",
        x = "p-value", y = "Density"
      ) +
      theme_minimal(base_size = 11) +
      theme(
        legend.position  = "bottom",
        strip.text       = element_text(face = "bold", size = 9),
        panel.grid.minor = element_blank(),
        axis.text.x      = element_text(angle = 45, hjust = 1)
      )
    print(p_pt)
    ggsave(file.path(disambig_dir, "pvalue_removed_vs_retained_by_pattern.png"),
           p_pt, width = 12, height = max(4, 2.5 * n_distinct(pval_df_pt$pattern_type)), dpi = 160)
  } else {
    cat("Skipping per-pattern_type p-value plot: no pattern_type with \u2265 5 observations.\n")
  }
} else {
  cat("No discovery data available — skipping p-value removal histograms.\n")
}
```

## B.3 Global change-pattern and significance characterization

```{r global_pattern_sig}
# Focus on pattern_type distribution across CAAP groups, first globally then with significant-both focus
pattern_by_group <- discovery_std %>% count(caap_group, pattern_type, sort = TRUE)
kable(pattern_by_group, caption = "Convergence pattern by CAAP group") %>% kable_styling()

pg <- ggplot(pattern_by_group, aes(pattern_type, n, fill = caap_group)) +
  geom_col(position = "dodge") +
  scale_fill_paletteer_d("ggthemes::Tableau_10") +
  theme_minimal() +
  labs(title = "Convergence pattern across CAAP groups (all)", x = "pattern_type", y = "n") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(pg)
ggsave(file.path(disambig_dir, "pattern_by_caap_group_all.png"), pg, width = 10, height = 6)
```

## B.3b Enhanced pattern_type analysis: Significant-both focus

```{r pattern_sig_both_focus}
# Focus exclusively on significant-both cases
pattern_by_group_sig_both <- discovery_std %>%
  filter(sig_both) %>%
  count(caap_group, pattern_type, sort = TRUE)

kable(pattern_by_group_sig_both, caption = "Convergence pattern by CAAP group (SIGNIFICANT-BOTH only)") %>% kable_styling()

if (nrow(pattern_by_group_sig_both) > 0) {
  pg_sig_both <- ggplot(pattern_by_group_sig_both, aes(pattern_type, n, fill = caap_group)) +
    geom_col(position = "dodge") +
    scale_fill_paletteer_d("ggthemes::Tableau_10") +
    theme_minimal() +
    labs(title = "Convergence pattern across CAAP groups (SIGNIFICANT-BOTH only)", 
         subtitle = paste0("n = ", sum(pattern_by_group_sig_both$n), " significant-both positions"),
         x = "pattern_type", y = "n") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  print(pg_sig_both)
  ggsave(file.path(disambig_dir, "pattern_by_caap_group_sig_both.png"), pg_sig_both, width = 10, height = 6)
} else {
  cat("No significant-both positions found for pattern_type analysis.\n")
}
```

## B.4 Overlap across CAAP groupings (gene & position level)

For each of two significance subsets (**all**, **sig_both**)
two composite figures are produced: one at the **position** level (site_id = gene@position) and one at
the **gene** level. Each figure shows a Venn diagram of group membership across US and GS0–GS4.

All figures are also saved to `disambiguation_characterization/overlap_figures/`.

```{r overlap_helper_functions, include=FALSE}
# Load additional packages for Venn diagrams
has_ggvenn <- requireNamespace("ggVennDiagram", quietly = TRUE)
if (has_ggvenn) library(ggVennDiagram)

# ---- Helper: create group sets for Venn diagrams ------------------------------------
create_group_sets <- function(df, id_col) {
  groups <- sort(unique(df$caap_group))
  setNames(lapply(groups, function(g) unique(df[[id_col]][df$caap_group == g])), groups)
}

# ---- Helper: build + save Venn diagram ------------------------------------------
make_overlap_plot <- function(df, id_col, level_label, subset_label, out_dir) {
  if (!has_ggvenn) {
    message("  [skip] ggVennDiagram not installed — skipping Venn diagram for subset='",
            subset_label, "' level='", level_label, "'")
    return(invisible(NULL))
  }
  if (nrow(df) == 0) {
    message("  [skip] No rows for subset='", subset_label, "' level='", level_label, "'")
    return(invisible(NULL))
  }
  groups <- sort(unique(df$caap_group))
  if (length(groups) < 2) {
    message("  [skip] Only one group for subset='", subset_label, "' level='", level_label, "'")
    return(invisible(NULL))
  }
  if (length(groups) > 5) {
    message("  [skip Venn] ", length(groups), " groups — ggVennDiagram requires 2-5.")
    return(invisible(NULL))
  }

  group_sets <- create_group_sets(df, id_col)

  venn_plot <- tryCatch({
    ggVennDiagram(group_sets, label_alpha = 0.8, set_size = 5) +
      scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Count") +
      labs(
        title    = paste0(level_label, "-level overlap  [", subset_label, "]"),
        subtitle = paste0("n = ", n_distinct(df[[id_col]]), " unique ", tolower(level_label), "s")
      ) +
      theme(
        plot.title    = element_text(face = "bold", size = 20),
        plot.subtitle = element_text(size = 15)
      )
  }, error = function(e) {
    message("Venn diagram generation failed: ", e$message)
    NULL
  })

  if (is.null(venn_plot)) return(invisible(NULL))

  png_name <- paste0("overlap_", tolower(level_label), "_", subset_label, ".png")
  ggsave(file.path(out_dir, png_name), venn_plot, width = 14, height = 10, dpi = 320)
  invisible(venn_plot)
}
```

```{r group_overlap_analysis_setup}
# Four significance subsets — GS0 excluded (discovery_std)
overlap_subsets <- list(
  all      = discovery_std,
  sig_both = discovery_std %>% filter(sig_both)
)
```

### Subset: all

```{r overlap_all, fig.width=14, fig.height=10}
subset_name <- "all"
sub_df_wo0 <- overlap_subsets[[subset_name]] %>% distinct(site_id, gene, caap_group) %>%
    filter(toupper(caap_group) != "GS0")

sub_df_gs <- overlap_subsets[[subset_name]] %>% distinct(site_id, gene, caap_group) %>%
    filter(toupper(caap_group) != "US")

sub_df_us0 <- overlap_subsets[[subset_name]] %>% distinct(site_id, gene, caap_group) %>%
    filter(toupper(caap_group) %in% c("GS0", "US"))

cat("Scenario 1 - US and GS groups (no GS0). Positions:", n_distinct(sub_df_wo0$site_id), " | Genes:", n_distinct(sub_df_wo0$gene),
    " | Groups:", paste(sort(unique(sub_df_wo0$caap_group)), collapse = ", "), "\n\n")
cat("Scenario 2 - GS groups only (US excluded). Positions:", n_distinct(sub_df_gs$site_id), " | Genes:", n_distinct(sub_df_gs$gene),
    " | Groups:", paste(sort(unique(sub_df_gs$caap_group)), collapse = ", "), "\n\n")
cat("Scenario 3 - US and GS0 only. Positions:", n_distinct(sub_df_us0$site_id), " | Genes:", n_distinct(sub_df_us0$gene),
    " | Groups:", paste(sort(unique(sub_df_us0$caap_group)), collapse = ", "), "\n\n")

cat("**Position-level overlap in Scenario 1 (US and GS groups, no GS0)**\n\n")
p_pos <- make_overlap_plot(sub_df_wo0, "site_id", "Position", subset_name, overlap_out_dir)
if (!is.null(p_pos)) print(p_pos)

cat("**Position-level overlap in Scenario 2 (GS groups only, US excluded)**\n\n")
p_pos <- make_overlap_plot(sub_df_gs, "site_id", "Position", subset_name, overlap_out_dir)
if (!is.null(p_pos)) print(p_pos)

cat("**Position-level overlap in Scenario 3 (US and GS0 only)**\n\n")
p_pos <- make_overlap_plot(sub_df_us0, "site_id", "Position", subset_name, overlap_out_dir)
if (!is.null(p_pos)) print(p_pos)
```

```{r overlap_all_gene, fig.width=14, fig.height=10}
cat("**Gene-level overlap in Scenario 1 (US and GS groups, no GS0)**\n\n")
p_gene <- make_overlap_plot(sub_df_wo0, "gene", "Gene", subset_name, overlap_out_dir)
if (!is.null(p_gene)) print(p_gene)
cat("**Gene-level overlap in Scenario 2 (GS groups only, US excluded)**\n\n")
p_gene <- make_overlap_plot(sub_df_gs, "gene", "Gene", subset_name, overlap_out_dir)
if (!is.null(p_gene)) print(p_gene)
cat("**Gene-level overlap in Scenario 3 (US and GS0 only)**\n\n")
p_gene <- make_overlap_plot(sub_df_us0, "gene", "Gene", subset_name, overlap_out_dir)
if (!is.null(p_gene)) print(p_gene)
```

### Subset: sig_both

```{r overlap_sig_both, fig.width=14, fig.height=10}
subset_name <- "sig_both"
sub_df_wo0 <- overlap_subsets[[subset_name]] %>% distinct(site_id, gene, caap_group) %>%
    filter(toupper(caap_group) != "GS0")

sub_df_gs <- overlap_subsets[[subset_name]] %>% distinct(site_id, gene, caap_group) %>%
    filter(toupper(caap_group) != "US")

sub_df_us0 <- overlap_subsets[[subset_name]] %>% distinct(site_id, gene, caap_group) %>%
    filter(toupper(caap_group) %in% c("GS0", "US"))



cat("Scenario 1 - US and GS groups (no GS0). Positions:", n_distinct(sub_df_wo0$site_id), " | Genes:", n_distinct(sub_df_wo0$gene),
    " | Groups:", paste(sort(unique(sub_df_wo0$caap_group)), collapse = ", "), "\n\n")
cat("Scenario 2 - GS groups only (US excluded). Positions:", n_distinct(sub_df_gs$site_id), " | Genes:", n_distinct(sub_df_gs$gene),
    " | Groups:", paste(sort(unique(sub_df_gs$caap_group)), collapse = ", "), "\n\n")
cat("Scenario 3 - US and GS0 only. Positions:", n_distinct(sub_df_us0$site_id), " | Genes:", n_distinct(sub_df_us0$gene),
    " | Groups:", paste(sort(unique(sub_df_us0$caap_group)), collapse = ", "), "\n\n")

cat("**Position-level overlap in Scenario 1 (US and GS groups, no GS0)**\n\n")
p_pos <- make_overlap_plot(sub_df_wo0, "site_id", "Position", subset_name, overlap_out_dir)
if (!is.null(p_pos)) print(p_pos)

cat("**Position-level overlap in Scenario 2 (GS groups only, US excluded)**\n\n")
p_pos <- make_overlap_plot(sub_df_gs, "site_id", "Position", subset_name, overlap_out_dir)
if (!is.null(p_pos)) print(p_pos)

cat("**Position-level overlap in Scenario 3 (US and GS0 only)**\n\n")
p_pos <- make_overlap_plot(sub_df_us0, "site_id", "Position", subset_name, overlap_out_dir)
if (!is.null(p_pos)) print(p_pos)
```

```{r overlap_sig_both_gene, fig.width=14, fig.height=10}
cat("**Gene-level overlap in Scenario 1 (US and GS groups, no GS0)**\n\n")
p_gene <- make_overlap_plot(sub_df_wo0, "gene", "Gene", subset_name, overlap_out_dir)
if (!is.null(p_gene)) print(p_gene)
cat("**Gene-level overlap in Scenario 2 (GS groups only, US excluded)**\n\n")
p_gene <- make_overlap_plot(sub_df_gs, "gene", "Gene", subset_name, overlap_out_dir)
if (!is.null(p_gene)) print(p_gene)
cat("**Gene-level overlap in Scenario 3 (US and GS0 only)**\n\n")
p_gene <- make_overlap_plot(sub_df_us0, "gene", "Gene", subset_name, overlap_out_dir)
if (!is.null(p_gene)) print(p_gene)
```

## B.5 US/GS group intersections (upset plots + structured exports)

```{r us_gs_relation_analysis, fig.height=27, fig.width=16}
# collapse_change_side and combo_from_flags are defined in setup.

# ── Upset helper: build bar + membership-matrix composite and return it ────────
plot_upset_like <- function(intersections_df, set_cols, title_txt, out_file = NULL) {
  if (is.null(intersections_df) || nrow(intersections_df) == 0) return(invisible(NULL))

  label_map <- c(
    us = "US", gs0 = "GS0", gs1 = "GS1", gs2 = "GS2", gs3 = "GS3", gs4 = "GS4",
    us_nondiv = "US nondiv",   gs0_nondiv = "GS0 nondiv", gs1_nondiv = "GS1 nondiv",
    gs2_nondiv = "GS2 nondiv", gs3_nondiv = "GS3 nondiv", gs4_nondiv = "GS4 nondiv",
    us_div = "US div",   gs0_div = "GS0 div", gs1_div = "GS1 div",
    gs2_div = "GS2 div", gs3_div = "GS3 div", gs4_div = "GS4 div"
  )

  plot_df <- intersections_df %>%
    arrange(desc(n), combo_label) %>%
    mutate(combo_id = factor(seq_len(n())))

  p_bar <- ggplot(plot_df, aes(x = combo_id, y = n)) +
    geom_col(fill = tableau_10[["blue"]], color = "black", linewidth = 0.2) +
    geom_text(aes(label = n), vjust = -0.4, size = 3.5, angle = 90) +
    labs(title = title_txt, x = NULL, y = "Intersection size") +
    theme_minimal(base_size = 13) +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          plot.title = element_text(face = "bold", size = 13), axis.title = element_text(size = 12))

  mat_long <- plot_df %>%
    select(combo_id, all_of(set_cols)) %>%
    pivot_longer(cols = all_of(set_cols), names_to = "set_name", values_to = "present") %>%
    mutate(set_label = recode(set_name, !!!label_map, .default = set_name),
           set_label = factor(set_label, levels = rev(unname(label_map[set_cols]))))

  p_mat <- ggplot(mat_long, aes(x = combo_id, y = set_label, fill = present)) +
    geom_tile(color = "white", linewidth = 0.3) +
    scale_fill_manual(values = c(`TRUE` = tableau_10[["vermillion"]], `FALSE` = "#eeeeee")) +
    labs(x = "Intersections", y = NULL) +
    theme_minimal(base_size = 13) +
    theme(legend.position = "none", axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          axis.text.y = element_text(size = 11), axis.title = element_text(size = 12))

  p <- p_bar / p_mat + plot_layout(heights = c(2.2, 1.3))
  if (!is.null(out_file)) ggsave(out_file, p, width = 16, height = 9, dpi = 320)
  invisible(p)
}

# ── Export helper ──────────────────────────────────────────────────────────────
write_export_pair <- function(df, key, txt_d, tsv_d) {
  createDir(txt_d); createDir(tsv_d)
  base_cols <- intersect(c("gene", "position", "caap_group", "pattern_type", "change_side"), names(df))
  out_df <- df %>% select(all_of(base_cols)) %>% distinct() %>% arrange(gene, position)
  readr::write_lines(sort(unique(out_df$gene)), file.path(txt_d, paste0(key, ".txt")))
  write_tsv(out_df, file.path(tsv_d, paste0(key, ".tsv")))
}

# ── Output directories ─────────────────────────────────────────────────────────
gene_rel_dir  <- file.path(disambig_dir, "gene_relation_analysis")
gr_plot_dir   <- file.path(gene_rel_dir,  "plots")
gr_txt_dir    <- file.path(gene_rel_dir,  "txt")
gr_tsv_dir    <- file.path(gene_rel_dir,  "tsv")
createDir(gr_plot_dir); createDir(gr_txt_dir); createDir(gr_tsv_dir)

# ── Set definitions ────────────────────────────────────────────────────────────
grp_sets_global <- c("us", "gs0", "gs1", "gs2", "gs3", "gs4")
grp_sets_nondiv <- c("us_nondiv", "gs0_nondiv", "gs1_nondiv", "gs2_nondiv", "gs3_nondiv", "gs4_nondiv")
grp_sets_div    <- c("us_div",    "gs0_div",    "gs1_div",    "gs2_div",    "gs3_div",    "gs4_div")

make_intersections <- function(sf, set_cols) {
  sf %>%
    filter(if_any(all_of(set_cols))) %>%
    count(across(all_of(set_cols)), name = "n") %>%
    mutate(combo_label = combo_from_flags(cur_data(), set_cols)) %>%
    filter(n > 0)
}

# ── Three upset figures: global / top / bottom ─────────────────────────────────
upset_orientations <- list(
  global = site_flags,
  top    = site_flags[site_flags$is_top, ],
  bottom = site_flags[site_flags$is_bottom, ]
)

for (ori_name in names(upset_orientations)) {
  sf_ori <- upset_orientations[[ori_name]]
  ori_label <- tools::toTitleCase(ori_name)

  p_all   <- plot_upset_like(make_intersections(sf_ori, grp_sets_global),
                             grp_sets_global, paste0("Global significant (", ori_label, ")"),
                             out_file = file.path(gr_plot_dir, paste0("upset_global_", ori_name, ".png")))
  p_nondiv <- plot_upset_like(make_intersections(sf_ori, grp_sets_nondiv),
                              grp_sets_nondiv, paste0("Nondivergent / convergent (", ori_label, ")"),
                              out_file = file.path(gr_plot_dir, paste0("upset_nondiv_", ori_name, ".png")))
  p_div   <- plot_upset_like(make_intersections(sf_ori, grp_sets_div),
                             grp_sets_div, paste0("Divergent (", ori_label, ")"),
                             out_file = file.path(gr_plot_dir, paste0("upset_div_", ori_name, ".png")))

  plots_valid <- Filter(Negate(is.null), list(p_all, p_nondiv, p_div))
  if (length(plots_valid) > 0) {
    combined <- wrap_plots(plots_valid, ncol = 1) +
      plot_annotation(
        title = paste0("US/GS group intersections — ", ori_label),
        theme = theme(plot.title = element_text(face = "bold", size = 16))
      )
    print(combined)
    ggsave(file.path(gr_plot_dir, paste0("us_gs_upset_", ori_name, ".png")),
           combined, width = 16, height = 9 * length(plots_valid), dpi = 320)
  }

  cat("✓ Upset figure:", ori_name, "| n =", nrow(sf_ori), "sites\n")
}

# ── Structured exports: per group × global / top / bottom × sig_both ──────────
# detail_rows: one row per position × CAAP group (for export files)
detail_rows <- discovery_std %>%
  transmute(
    site_id,
    gene,
    position,
    caap_group = toupper(caap_group),
    pattern_type,
    change_side = if_else(change_side %in% c("top", "bottom", "both"), change_side, "none"),
    sig_both
  ) %>%
  distinct()

# Join site-level flags back to detail rows for filtering
detail_with_flags <- detail_rows %>%
  left_join(
    site_flags %>% select(site_id, is_top, is_bottom,
                          global_sig, convergent_pool, divergent_pool,
                          us, gs0, gs1, gs2, gs3, gs4,
                          us_nondiv, gs0_nondiv, gs1_nondiv, gs2_nondiv, gs3_nondiv, gs4_nondiv,
                          us_div,    gs0_div,    gs1_div,    gs2_div,    gs3_div,    gs4_div),
    by = "site_id"
  ) %>%
  mutate(across(where(is.logical), ~ replace_na(.x, FALSE)))

# Export orientations with unified direction labels
orientations <- list(
  nodir  = rep(TRUE, nrow(detail_with_flags)),
  top    = detail_with_flags$is_top,
  bottom = detail_with_flags$is_bottom
)

# Export global scenario pools (general/convergent/divergent) across orientations
scenario_exports <- list(
  global_sig      = "global_sig",
  convergent_pool = "convergent_pool",
  divergent_pool  = "divergent_pool"
)

for (sc_name in names(scenario_exports)) {
  sc_col <- scenario_exports[[sc_name]]

  for (ori_name in names(orientations)) {
    ori_mask <- orientations[[ori_name]]
    base_df  <- detail_with_flags[ori_mask & detail_with_flags$sig_both, ]

    write_export_pair(
      base_df[base_df[[sc_col]], ],
      paste0(sc_name, "_", ori_name),
      gr_txt_dir,
      gr_tsv_dir
    )
  }
  cat("✓ Scenario exports:", sc_name, "\n")
}

# Export per group (US, GS0–GS4) across orientations
export_groups <- list(
  us  = "us",  gs0 = "gs0",
  gs1 = "gs1", gs2 = "gs2", gs3 = "gs3", gs4 = "gs4"
)

for (grp_name in names(export_groups)) {
  gc  <- export_groups[[grp_name]]
  gc_nd <- paste0(gc, "_nondiv")
  gc_dv <- paste0(gc, "_div")

  for (ori_name in names(orientations)) {
    ori_mask <- orientations[[ori_name]]
    base_df  <- detail_with_flags[ori_mask & detail_with_flags$sig_both, ]

    write_export_pair(base_df[base_df[[gc]],    ], paste0(grp_name, "_sig_",  ori_name), gr_txt_dir, gr_tsv_dir)
    write_export_pair(base_df[base_df[[gc_nd]], ], paste0(grp_name, "_nondiv_", ori_name), gr_txt_dir, gr_tsv_dir)
    write_export_pair(base_df[base_df[[gc_dv]], ], paste0(grp_name, "_div_",  ori_name), gr_txt_dir, gr_tsv_dir)
  }
  cat("✓ Exports:", grp_name, "\n")
}

cat("\n✓ All B.5 upset plots and exports generated\n")
cat("  Plots:", gr_plot_dir, "\n")
cat("  TXT:  ", gr_txt_dir,  "\n")
cat("  TSV:  ", gr_tsv_dir,  "\n\n")
```

## B.5b Excluded gene list exports

```{r excluded_gene_lists}
# ── Build per-category excluded gene lists ────────────────────────────────────────────
# Six lists are exported to
#   disambiguation_characterization/gene_relation_analysis/txt/excluded/
# for a separate ORA run against the original (pre-cleanup) background from discovery.
#
# Categories (mutually exclusive by design for excluded_nonsig):
#   excluded_preproc             – genes with ≥1 position removed during pre-processing
#                                  (pattern-type filter / MRCA-posterior threshold)
#   excluded_postproc_clusters   – genes with ≥1 position discarded by cluster filter
#   excluded_postproc_genefilter – genes flagged Extreme and/or Dubious (gene-level filter)
#   excluded_postproc            – union of the two postproc exclusion categories
#   excluded_nonsig              – genes that survived ALL filters but never achieved
#                                  sig_both == TRUE; does NOT overlap with preproc/postproc
#   excluded_overall             – union of all four categories

# 1. Pre-processing exclusions (from precluster_events, loaded in B.1)
excluded_preproc <- sort(unique(precluster_events$gene))

# 2. Post-processing: cluster-filter genes
excluded_postproc_clusters <- sort(unique(
  filter(gene_filter_flags, Cluster_removed)$Gene
))

# 3. Post-processing: gene-level filter (Extreme / Dubious / Both)
excluded_postproc_genefilter <- sort(unique(
  filter(gene_filter_flags, Extreme | Dubious)$Gene
))

# 4. All post-processing exclusions (union of 2 and 3)
excluded_postproc <- sort(unique(c(excluded_postproc_clusters, excluded_postproc_genefilter)))

# 5. Non-significant genes: survived all filters, but zero sig_both positions;
#    explicitly does NOT overlap with preproc / postproc categories
any_filtered_genes <- sort(unique(filter(gene_filter_flags, filter_combo != "None")$Gene))
genes_ever_sig     <- sort(unique(filter(discovery_std, sig_both)$gene))
survived_genes     <- sort(setdiff(unique(gene_filter_flags$Gene), any_filtered_genes))
excluded_nonsig    <- sort(setdiff(survived_genes, c(genes_ever_sig, excluded_preproc)))

# 6. Overall: union of all exclusion categories
excluded_overall <- sort(unique(c(excluded_preproc, excluded_postproc, excluded_nonsig)))

# ── Write to gene_relation_analysis/txt/excluded/ ────────────────────────────────────
gr_excl_dir <- file.path(gr_txt_dir, "excluded")
createDir(gr_excl_dir)

readr::write_lines(excluded_preproc,             file.path(gr_excl_dir, "excluded_preproc.txt"))
readr::write_lines(excluded_postproc_clusters,   file.path(gr_excl_dir, "excluded_postproc_clusters.txt"))
readr::write_lines(excluded_postproc_genefilter, file.path(gr_excl_dir, "excluded_postproc_genefilter.txt"))
readr::write_lines(excluded_postproc,            file.path(gr_excl_dir, "excluded_postproc.txt"))
readr::write_lines(excluded_nonsig,              file.path(gr_excl_dir, "excluded_nonsig.txt"))
readr::write_lines(excluded_overall,             file.path(gr_excl_dir, "excluded_overall.txt"))

cat("\n=== Excluded Gene Lists ===\n")
cat(sprintf("  excluded_preproc             : %d genes\n", length(excluded_preproc)))
cat(sprintf("  excluded_postproc_clusters   : %d genes\n", length(excluded_postproc_clusters)))
cat(sprintf("  excluded_postproc_genefilter : %d genes\n", length(excluded_postproc_genefilter)))
cat(sprintf("  excluded_postproc (union)    : %d genes\n", length(excluded_postproc)))
cat(sprintf("  excluded_nonsig              : %d genes\n", length(excluded_nonsig)))
cat(sprintf("  excluded_overall             : %d genes\n", length(excluded_overall)))
cat(sprintf("  Written to: %s\n\n", gr_excl_dir))
```

## B.6 Manhattan: genomic distribution of genes (CAAS density per gene length)

```{r manhattan_disambiguation, fig.height=8, fig.width=16, eval=generate_manhattan}
if (!"chr" %in% names(gene_ensembl)) {
  cat("⚠ Chromosome information not available in gene ensembl file, skipping Manhattan plot\n")
} else {
  # Calculate CAAS density per gene (significant-both positions / gene length).
  # Uses all discovery data (GS0 included here: genomic position is gene-level).
  gene_caas_density <- discovery_std %>%
    filter(sig_both) %>%  # Focus on significant-both positions only
    group_by(gene) %>%
    summarise(
      n_caas_positions = n_distinct(site_id),
      .groups = "drop"
    ) %>%
    left_join(gene_ensembl %>% 
              select(gene, chr, start, end) %>%
              mutate(gene_length = end - start + 1), 
              by = "gene") %>%
    filter(!is.na(chr), gene_length > 0) %>%
    mutate(
      caas_density = n_caas_positions / gene_length,
      log_caas_density = log10(pmax(caas_density, 1e-10))  # Avoid log10(0)
    )

  # Visualise all genes with at least one significant-both CAAS position.
  # manhattan_min_density is used only as a reference line, not a hard filter.
  manhattan_data <- gene_caas_density %>%
    filter(caas_density > 0) %>%
    mutate(
      chr_num   = str_replace(chr, "chr", ""),
      chr_order = case_when(
        chr_num %in% as.character(1:22) ~ as.numeric(chr_num),
        chr_num == "X"                  ~ 23,
        chr_num == "Y"                  ~ 24,
        TRUE                            ~ 25
      ),
      pos = (start + end) / 2
    ) %>%
    arrange(chr_order, pos)

  if (nrow(manhattan_data) > 0) {
    # Cumulative chromosome positions and boundaries
    chr_lengths <- manhattan_data %>%
      group_by(chr_num, chr_order) %>%
      summarise(chr_max = max(pos, na.rm = TRUE), .groups = "drop") %>%
      arrange(chr_order) %>%
      mutate(
        chr_start  = lag(cumsum(chr_max), default = 0),
        chr_end    = chr_start + chr_max,
        chr_center = chr_start + chr_max / 2
      )

    manhattan_data <- manhattan_data %>%
      left_join(chr_lengths %>% select(chr_num, chr_start, chr_end), by = "chr_num") %>%
      mutate(BPcum = pos + chr_start)

    # No Bonferroni threshold for CAAS density plot
    n_genes_plotted <- n_distinct(manhattan_data$gene)

    ref_density <- max(manhattan_data$caas_density, na.rm = TRUE) * 1.5

    p_manhattan <- ggplot(manhattan_data,
        aes(x = BPcum, y = caas_density)) +
      # Discontinuous chromosome boundary lines
      geom_vline(
        data        = chr_lengths,
        aes(xintercept = chr_end),
        linetype    = "dotted", color = "grey60", alpha = 0.7, linewidth = 0.4,
        inherit.aes = FALSE
      ) +
      geom_point(alpha = 0.80, size = 2.5, color = "#2E86AB") +
      geom_hline(
        yintercept = ref_density,
        linetype   = "dashed", colour = "#E74C3C", linewidth = 0.6
      ) +
      annotate(
        "text",
        x     = Inf, y = ref_density,
        label = sprintf("ref: %.3g", ref_density),
        hjust = 1.05, vjust = -0.4, size = 3.5, colour = "#E74C3C"
      ) +
      scale_x_continuous(
        labels = chr_lengths$chr_num,
        breaks = chr_lengths$chr_center,
        expand = expansion(mult = c(0.005, 0.005))
      ) +
      scale_y_continuous(
        labels = scales::label_number(scale_cut = cut_short_scale()),
        expand = expansion(mult = c(0.02, 0.08))
      ) +
      labs(
        title    = "Genomic distribution of CAAS density (significant-both positions only)",
        subtitle = sprintf(
          "n = %d genes with \u2265 1 sig-both CAAS position | y = sig-both positions / gene length (bp) | dashed line = reference density %.3g",
          n_genes_plotted, ref_density
        ),
        x     = "Chromosome",
        y     = "CAAS density (sig-both positions / gene length)"
      ) +
      theme_minimal(base_size = 11) +
      theme(
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()
      )

    print(p_manhattan)
    ggsave(
      filename = file.path(disambig_dir, "manhattan_caas_density.png"),
      plot     = p_manhattan, width = 16, height = 8, dpi = 320, dev = "png"
    )
  } else {
    cat("No genes with sig-both CAAS positions found — Manhattan plot skipped.\n")
  }
}
```

## 5. Impact Summary

```{r impact_summary}
cat("\n=== CT Post-Processing Impact Summary ===\n\n")

cat("Gene-level statistics:\n")
cat("  Total genes analyzed:", n_distinct(discovery_df$Gene), "\n\n")

cat("Position-level statistics:\n")
cat("  Total CT positions:", nrow(discovery_df), "\n")
if (nrow(filter_summary) > 0) {
  cat("  Positions flagged by cluster filtering:\n")
  for (i in 1:nrow(filter_summary)) {
    cat(sprintf("    minlen=%d, maxcaas=%.2f: %d positions\n", filter_summary$Minlen[i], filter_summary$Maxcaas[i], filter_summary$DiscardedCount[i]))
  }
}

summary_table <- tibble(
  Metric = c("Total CT Positions", "Unique Genes"),
  Value  = c(
    as.character(nrow(discovery_df)),
    as.character(n_distinct(discovery_df$Gene))
  )
)

kable(summary_table, caption = "CT Post-Processing Summary Statistics", align = c("l", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

write_tsv(summary_table, file.path(summary_stats_dir, "summary_statistics.tsv"))
cat("\n✓ Summary statistics exported\n")
cat("\n=== Analysis Complete ===\n")
```

## Session Info

```{r session_info}
sessionInfo()
```
