---
title: "RERconverge — Relative Evolutionary Rate Convergence Analysis"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
params:
  continuous_rds:    "."
  traitname:         "unknown_trait"
  pval_threshold:    0.05
  top_n_labels:      15
  output_dir:        "."
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  out.width = "100%"
)

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggrepel)
  library(paletteer)
  library(DT)
  library(gridExtra)
  library(stringr)
})

# Tableau_10 colour swatch (10 named colours for pipeline-wide consistency)
.tableau10 <- as.character(paletteer_d("ggthemes::Tableau_10"))

theme_set(
  theme_bw(base_size = 13) +
  theme(
    axis.title.y  = element_text(margin = ggplot2::margin(l = 20, r = 20)),
    axis.title.x  = element_text(margin = ggplot2::margin(t = 12)),
    axis.text.x   = element_text(margin = ggplot2::margin(t = 5)),
    plot.title    = element_text(face = "bold", hjust = 0, margin = ggplot2::margin(b = 6)),
    plot.subtitle = element_text(colour = "grey40", hjust = 0, margin = ggplot2::margin(b = 10))
  )
)
```

---

## Overview

This report summarises the results of **RERconverge** (relative evolutionary rate convergence;
Clark _et al._, 2012–2019) applied to the continuous phenotype **`r params$traitname`**.

RERconverge tests whether genes show correlated **relative evolutionary rates (RERs)** across
species with differences in a continuous trait. A positive correlation (Rho > 0) means that
lineages with higher trait values tend to show **faster** evolution in that gene; a negative
correlation (Rho < 0) means **slower** evolution.

The analysis uses the following key statistics:

| Statistic | Meaning |
|-----------|---------|
| **Rho**     | Pearson correlation between trait path values and gene RER path values |
| **N**       | Number of branches (species) with data for the gene |
| **P**       | Raw **parametric** p-value of the Pearson correlation test (it treats the branches as independent observations) |
| **p.adj**   | Benjamini–Hochberg FDR-adjusted **parametric** p-value over all genes (computed internally by `correlateWithContinuousPhenotype()`, Kowalczyk _et al._ 2019) |
| **p.perm**  | Empirical **permulation** p-value (if run): per gene, the null correlations of Brownian-Motion null phenotypes are compared with the observed Rho, centred on the median of the gene's own null (`permpvalcor()`, Saputra _et al._ 2021). Its minimum is about 1/(N/2 + 1) for N permulations, not 1/(N + 1) (the observed minimum is shown below). |
| **p.perm.adj** | Benjamini–Hochberg adjustment of `p.perm` over all genes |

> **Note:** The parametric `P` and `p.adj` can be miscalibrated in either direction depending on the phenotype
> (Saputra _et al._ 2021, Fig. 1A): anti-conservative for a binary marine phenotype and conservative for a
> continuous long-lived large-bodied one. They rank genes, but they are not a calibrated FDR.
> Genes called **significant** here (the `Call` column and the gene lists) are those with a raw permulation p-value
> `p.perm < pval_threshold`, which is **not corrected for multiple testing**; the BH-adjusted `p.perm.adj` is
> shown next to it. Saputra _et al._ correct the permulation p-values (Storey, or BH) and use up to ~1,000
> permulations; with fewer than ~125 the permulation does not improve on the parametric p (their Fig. 5C).
> Without permulations, significance falls back to `p.adj < pval_threshold`.
> Trait transformations (e.g. logit, arcsin, log10, or none) can be configured via the `rer_transform` parameter.

---

```{r load-rds}
if (!file.exists(params$continuous_rds)) {
  message("RDS file not found: '", params$continuous_rds,
          "'. Cannot generate report.")
  knitr::knit_exit()
}

res <- readRDS(params$continuous_rds)

if (is.null(res) || nrow(res) == 0) {
  message("RDS file loaded but contains no results. Cannot generate report.")
  knitr::knit_exit()
}

# Total permutation count stored as attribute by continuous_rer.R.
# Falls back to NA if running the report against an older RDS.
n_perms_total <- attr(res, "n_perms")

cat(sprintf("Loaded RERconverge results: %d genes.\n", nrow(res)))
```

```{r process-results, results='hide'}
has_perm <- "p.perm" %in% colnames(res) && any(!is.na(res$p.perm))

# p.perm carries a pseudo-count, so it is never 0
fmt_pperm <- function(x) ifelse(is.na(x), NA_character_, as.character(signif(x, 3)))

rer_df <- res %>%
  as.data.frame() %>%
  # Ensure p.perm column exists (as NA if not present) to avoid downstream errors
  {if (!("p.perm" %in% colnames(.))) mutate(., p.perm = NA_real_) else .} %>%
  mutate(gene = rownames(res)) %>%   # row names ARE gene IDs
  relocate(gene)

# BH adjustment of p.perm over all genes (continuous_rer.R writes it; an older RDS gets it here)
if (has_perm && !("p.perm.adj" %in% colnames(rer_df))) {
  rer_df$p.perm.adj <- p.adjust(rer_df$p.perm, method = "BH")
}

# Compute significance flags based on whether permutation p-values exist
if (has_perm) {
  rer_df <- rer_df %>%
    mutate(
      # Parametric significance (always available)
      signif_param = !is.na(p.adj) & p.adj < params$pval_threshold,
      # Significance is defined directly by raw p.perm when permutations are available
      signif       = !is.na(p.perm) & p.perm < params$pval_threshold,
      # FDR-based call: BH-adjusted permulation p-value
      signif_perm_fdr = !is.na(p.perm.adj) & p.perm.adj < params$pval_threshold,
      pos_corr     = signif &  Rho >  0,
      neg_corr     = signif &  Rho <  0
    ) %>%
    arrange(p.perm)
} else {
  rer_df <- rer_df %>%
    mutate(
      signif_param = !is.na(p.adj) & p.adj < params$pval_threshold,
      signif       = signif_param,
      pos_corr     = signif &  Rho >  0,
      neg_corr     = signif &  Rho <  0
    ) %>%
    arrange(p.adj)
}

# Define the human-readable Call classification column
rer_df <- rer_df %>%
  mutate(
    Call = case_when(
      pos_corr ~ "Positively correlated",
      neg_corr ~ "Negatively correlated",
      signif   ~ "Significant",
      TRUE     ~ ""
    )
  )

# ── Prioritisation: top X% per direction by the signed RER statistic ──────────
# Ranking uses the RER-standard signed statistic sign(Rho) * -log10(P) —
# continuous and tie-free, the same quantity RERconverge ranks by for enrichment.
.dir_top <- function(stat, frac, direction) {
  flag <- rep(FALSE, length(stat))
  pool <- if (direction == "acc") stat > 0 & is.finite(stat) else stat < 0 & is.finite(stat)
  if (any(pool, na.rm = TRUE)) {
    thr <- quantile(stat[pool], if (direction == "acc") 1 - frac else frac, na.rm = TRUE)
    flag[pool] <- if (direction == "acc") stat[pool] >= thr else stat[pool] <= thr
  }
  flag
}

top_pct  <- 0.10
top5_pct <- 0.05
top1_pct <- 0.01

rer_df <- rer_df %>%
  mutate(
    rer_stat      = sign(Rho) * -log10(pmax(P, 1e-300)),
    is_top_10_acc = .dir_top(rer_stat, top_pct,  "acc"),
    is_top_5_acc  = .dir_top(rer_stat, top5_pct, "acc"),
    is_top_1_acc  = .dir_top(rer_stat, top1_pct, "acc"),
    is_top_10_dec = .dir_top(rer_stat, top_pct,  "dec"),
    is_top_5_dec  = .dir_top(rer_stat, top5_pct, "dec"),
    is_top_1_dec  = .dir_top(rer_stat, top1_pct, "dec")
  )

cat(sprintf(
  "Significant after BH-FDR (p.adj < %g): %d genes.\n",
  params$pval_threshold,
  sum(rer_df$signif_param, na.rm = TRUE)
))
cat(sprintf(
  "Prioritised (top %.0f%%/%.0f%%/%.0f%% by signed stat): accelerated %d/%d/%d, decelerated %d/%d/%d.\n",
  100 * top_pct, 100 * top5_pct, 100 * top1_pct,
  sum(rer_df$is_top_10_acc), sum(rer_df$is_top_5_acc), sum(rer_df$is_top_1_acc),
  sum(rer_df$is_top_10_dec), sum(rer_df$is_top_5_dec), sum(rer_df$is_top_1_dec)
))
if (has_perm) {
  cat(sprintf(
    "Significant by raw permutation p-value (p.perm < %g, not corrected for multiple testing): %d genes.\n",
    params$pval_threshold,
    sum(rer_df$signif, na.rm = TRUE)
  ))
  cat(sprintf(
    "Significant by BH-adjusted permutation p-value (p.perm.adj < %g): %d genes.\n",
    params$pval_threshold,
    sum(rer_df$signif_perm_fdr, na.rm = TRUE)
  ))
}
```

---

## Gene-level Summary

```{r gene-table}
display_df <- rer_df %>%
  mutate(
    Rho   = round(Rho, 4),
    P     = signif(P,    3),
    p.adj = signif(p.adj, 3),
    p.perm = if (has_perm) signif(p.perm, 3) else NA_real_,
    p.perm.adj = if (has_perm) signif(p.perm.adj, 3) else NA_real_
  )

# Include p.perm column only when permutations were run
if (has_perm) {
  display_df <- display_df %>%
    select(gene, Rho, N, P, p.adj, p.perm, p.perm.adj, Call) %>%
    rename(Gene = gene, `Raw P (parametric)` = P, `p.adj (parametric BH)` = p.adj,
           `p.perm (permulation)` = p.perm, `p.perm.adj (permulation BH)` = p.perm.adj)
} else {
  display_df <- display_df %>%
    select(gene, Rho, N, P, p.adj, Call) %>%
    rename(Gene = gene, `Raw P (parametric)` = P, `p.adj (parametric BH)` = p.adj)
}

DT::datatable(
  display_df,
  filter   = "top",
  rownames = FALSE,
  options  = list(pageLength = 30, scrollX = TRUE),
  caption  = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf(
      "RERconverge gene-level results | Trait: %s",
      params$traitname
    )
  )
) %>%
  DT::formatStyle(
    "Call",
    backgroundColor = DT::styleEqual(
      c("Positively correlated", "Negatively correlated", "Significant", ""),
      c("#ffe0e0",               "#dff0ff",               "#fff0d0",      "white")
    )
  )
```

```{r write-tsv}
# Write to the Nextflow work directory (current dir) so the optional emit: can
# capture it.  A copy is also placed in params$output_dir for human inspection.
local_tsv <- paste0("rerconverge_summary_", params$traitname, ".tsv")
write.table(
  rer_df,
  local_tsv, sep = "\t", row.names = FALSE, quote = FALSE
)
if (!is.null(params$output_dir) && nzchar(params$output_dir)) {
  out_tsv <- file.path(params$output_dir, local_tsv)
  file.copy(local_tsv, out_tsv, overwrite = TRUE)
}
invisible(NULL)
```

---

## Visualisations

### Correlation plot

```{r corr-plot, fig.height=7, fig.width=9}
label_df <- rer_df %>%
  filter(signif) %>%
  arrange(if (has_perm) p.perm else p.adj) %>%
  slice_head(n = params$top_n_labels)

ggplot(rer_df, aes(x = Rho, y = -log10(if (has_perm) p.perm else (P + 1e-300)))) +
  geom_point(
    aes(colour = case_when(
      pos_corr ~ "Positively correlated",
      neg_corr ~ "Negatively correlated",
      signif   ~ "Significant",
      TRUE     ~ "Non-significant"
    )),
    alpha = 0.7,
    size  = 2
  ) +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey60", linewidth = 0.5) +
  geom_hline(yintercept = -log10(params$pval_threshold),
             linetype = "dashed", colour = "grey40", linewidth = 0.5) +
  geom_text_repel(
    data    = label_df,
    aes(label = gene),
    size    = 3,
    max.overlaps = 30
  ) +
  scale_colour_manual(
    values = c(
      "Positively correlated" = .tableau10[3],
      "Negatively correlated" = .tableau10[1],
      "Significant"           = .tableau10[2],
      "Non-significant"       = "grey60"
    ),
    name = NULL
  ) +
  labs(
    title    = paste("RERconverge correlation plot — Trait:", params$traitname),
    subtitle = paste0("n = ", nrow(rer_df), " genes"),
    x        = expression(rho ~ "(Pearson correlation with trait)"),
    y        = if (has_perm) expression(-log[10](p.perm)) else expression(-log[10](P))
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "top")
```

### Rho distribution

```{r rho-dist, fig.height=5}
ggplot(rer_df %>% filter(!is.na(Rho) & is.finite(Rho)),
       aes(x = Rho)) +
  geom_histogram(aes(fill = after_stat(x) > 0), bins = 60,
                 colour = "white", alpha = 0.85) +
  scale_fill_manual(
    values = c("TRUE" = .tableau10[3], "FALSE" = .tableau10[1]),
    labels = c("TRUE" = "Positive correlation", "FALSE" = "Negative correlation"),
    name   = NULL
  ) +
  geom_vline(xintercept = 0, linewidth = 0.7, colour = "black") +
  labs(
    title    = "Distribution of Rho across genes",
    subtitle = paste0("Trait: ", params$traitname),
    x        = expression(rho ~ "(Pearson correlation)"),
    y        = "Number of genes"
  ) +
  theme_bw(base_size = 13)
```

### Sample size vs correlation strength

```{r n-vs-rho, fig.height=6, fig.width=7}
ggplot(rer_df, aes(x = N, y = Rho)) +
  geom_point(
    aes(colour = case_when(
      pos_corr ~ "Positively correlated",
      neg_corr ~ "Negatively correlated",
      signif   ~ "Significant",
      TRUE     ~ "Non-significant"
    )),
    alpha = 0.65, size = 2
  ) +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = "grey40", linewidth = 0.6) +
  geom_text_repel(
    data    = label_df,
    aes(label = gene),
    size    = 2.6,
    max.overlaps = 20
  ) +
  scale_colour_manual(
    values = c(
      "Positively correlated" = .tableau10[3],
      "Negatively correlated" = .tableau10[1],
      "Significant"           = .tableau10[2],
      "Non-significant"       = "grey60"
    ),
    name = NULL
  ) +
  labs(
    title    = "Number of species (N) vs correlation strength (Rho)",
    subtitle = paste0("Trait: ", params$traitname,
                     " | Dashed line = zero correlation"),
    x = "N (branches with data)",
    y = expression(rho ~ "(Pearson correlation)")
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "top")
```

### Permutation p-values

```{r perm-plot, fig.height=6, fig.width=10, eval=has_perm}
# P vs p.perm scatter — genes on the diagonal have consistent parametric
# and empirical evidence; off-diagonal genes may be driven by phylogenetic structure.
perm_label_df <- rer_df %>%
  filter(signif) %>%
  arrange(p.perm) %>%
  slice_head(n = params$top_n_labels)

# Permutation floor: permpvalcor() centres each gene's null on its median and divides by the null values on the
# observed side, about N/2, so the smallest p.perm is about 1 / (N/2 + 1), not 1 / (N + 1). The floor drawn is the
# smallest p.perm of the table, which does not depend on that rule.
perm_floor_val <- if (has_perm) min(rer_df$p.perm, na.rm = TRUE) else NA_real_
perm_floor_log10 <- if (!is.na(perm_floor_val)) -log10(perm_floor_val) else NA_real_
n_floor_genes <- if (!is.na(perm_floor_val)) sum(!is.na(rer_df$p.perm) & rer_df$p.perm <= perm_floor_val * 1.01) else 0

p1 <- ggplot(rer_df, aes(x = -log10(P + 1e-300), y = -log10(p.perm))) +
  geom_point(aes(colour = case_when(
    pos_corr    ~ "Positively correlated",
    neg_corr    ~ "Negatively correlated",
    signif      ~ "Significant",
    TRUE        ~ "Non-significant"
  )), alpha = 0.7, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey40", linewidth = 0.5) +
  # Permutation significance threshold line
  geom_hline(yintercept = -log10(params$pval_threshold),
             linetype = "dotted", colour = "grey50", linewidth = 0.5) +
  # Floor line: p.perm floor is shown
  {if (!is.na(perm_floor_log10))
     geom_hline(yintercept = perm_floor_log10,
                linetype = "solid", colour = "#d62728", linewidth = 0.5, alpha = 0.6)
   else list()} +
  {if (!is.na(perm_floor_log10) && n_floor_genes > 0)
     annotate("text", x = Inf, y = perm_floor_log10,
              label = sprintf(" floor: p = %.4f  (%d genes)", perm_floor_val, n_floor_genes),
              hjust = 1.05, vjust = -0.4, size = 3, colour = "#d62728")
   else list()} +
  geom_text_repel(data = perm_label_df, aes(label = gene),
                  size = 2.8, max.overlaps = 25) +
  scale_colour_manual(
    values = c(
      "Positively correlated" = .tableau10[3],
      "Negatively correlated" = .tableau10[1],
      "Significant"           = .tableau10[2],
      "Non-significant"       = "grey60"
    ), name = NULL
  ) +
  labs(
    title    = "Parametric (raw P) vs empirical (p.perm) significance",
    subtitle = paste0("Trait: ", params$traitname,
                     " | Dashed diagonal = equal evidence"),
    x = expression(-log[10](P)),
    y = expression(-log[10](p.perm))
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "top")

# Histogram: separate the floor bin in a distinct colour so that
# genes at the permutation resolution limit are clearly visible.
perm_df_hist <- rer_df %>%
  filter(!is.na(p.perm)) %>%
  mutate(at_floor = !is.na(perm_floor_val) & p.perm <= perm_floor_val * 1.01)

p2 <- ggplot(perm_df_hist, aes(x = p.perm, fill = at_floor)) +
  geom_histogram(bins = 50, colour = "white", alpha = 0.85) +
  geom_vline(xintercept = params$pval_threshold,
             linetype = "dashed", colour = "grey40", linewidth = 0.7) +
  scale_fill_manual(
    values = c(`FALSE` = .tableau10[5], `TRUE` = "#d62728"),
    labels = c(`FALSE` = "Observed p.perm", `TRUE` = sprintf("Floor (p = %.4f)", if (!is.na(perm_floor_val)) perm_floor_val else 0)),
    name   = NULL
  ) +
  labs(
    title    = "Distribution of permutation p-values",
    subtitle = if (n_floor_genes > 0) sprintf("%d genes at the permutation floor (p.perm = %.4f)", n_floor_genes, perm_floor_val) else NULL,
    x = "p.perm",
    y = "Number of genes"
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "top")

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

```{r perm-missing, eval=!has_perm, results='asis'}
cat("> **Permutation p-values not available.** Re-run with `rer_perm_batches > 0` to enable.\n")
```

---

## Parametric adjusted p-value (BH p.adj) distribution

```{r padj-dist, fig.height=5}
ggplot(rer_df %>% filter(!is.na(p.adj)), aes(x = p.adj)) +
  geom_histogram(bins = 50, colour = "white", fill = .tableau10[2], alpha = 0.85) +
  geom_vline(xintercept = params$pval_threshold, linetype = "dashed", colour = "grey40", linewidth = 0.7) +
  labs(
    title    = "Distribution of parametric adjusted p-values (BH p.adj)",
    subtitle = paste0("Trait: ", params$traitname),
    x        = "BH p.adj",
    y        = "Number of genes"
  ) +
  theme_bw(base_size = 13)
```

---

## Positively Correlated Genes (top `r params$top_n_labels`)

```{r top-positive}
sort_col <- if (has_perm) "p.perm" else "p.adj"

top_pos <- rer_df %>%
  filter(pos_corr) %>%
  arrange(.data[[sort_col]], desc(Rho)) %>%
  slice_head(n = params$top_n_labels) %>%
  mutate(Rho = round(Rho, 4), p.adj = signif(p.adj, 3))

if (has_perm) {
  top_pos <- top_pos %>%
    mutate(p.perm = fmt_pperm(p.perm), p.perm.adj = signif(p.perm.adj, 3)) %>%
    select(gene, Rho, N, p.adj, p.perm, p.perm.adj)
  col_names <- c("Gene", "Rho", "N", "p.adj (parametric BH)", "p.perm", "p.perm.adj")
} else {
  top_pos <- top_pos %>% select(gene, Rho, N, p.adj)
  col_names <- c("Gene", "Rho", "N", "p.adj (parametric BH)")
}

if (nrow(top_pos) == 0) {
  cat("No positively correlated genes at the given thresholds.\n")
} else {
  knitr::kable(
    top_pos,
    col.names = col_names,
    caption   = "Top positively correlated genes (faster evolution with higher trait value)"
  )
}
```

## Negatively Correlated Genes (top `r params$top_n_labels`)

```{r top-negative}
top_neg <- rer_df %>%
  filter(neg_corr) %>%
  arrange(.data[[sort_col]], Rho) %>%
  slice_head(n = params$top_n_labels) %>%
  mutate(Rho = round(Rho, 4), p.adj = signif(p.adj, 3))

if (has_perm) {
  top_neg <- top_neg %>%
    mutate(p.perm = fmt_pperm(p.perm), p.perm.adj = signif(p.perm.adj, 3)) %>%
    select(gene, Rho, N, p.adj, p.perm, p.perm.adj)
} else {
  top_neg <- top_neg %>% select(gene, Rho, N, p.adj)
}

if (nrow(top_neg) == 0) {
  cat("No negatively correlated genes at the given thresholds.\n")
} else {
  knitr::kable(
    top_neg,
    col.names = col_names,
    caption   = "Top negatively correlated genes (slower evolution with higher trait value)"
  )
}
```



---

## Methods

**Tool:** RERconverge (Kowalczyk _et al._, 2019, _Bioinformatics_ 35:4815;
https://github.com/nclark-lab/RERconverge) via the `correlateWithContinuousPhenotype()`
function. Permulations: Saputra _et al._ 2021, _Mol. Biol. Evol._ 38:3004, doi:10.1093/molbev/msab068.

**Trait:** `r params$traitname`

**RER computation:** Relative evolutionary rates were calculated with `getAllResiduals()` using
a square-root branch-length transform, weighted calculation, and scaling (`transform="sqrt"`,
`weighted=TRUE`, `scale=TRUE`). Genes with insufficient species were excluded (`min.sp` parameter).

**Winsorization:** RER and trait values were optionally winsorized to reduce the influence of extreme values
(via `winsorizeRER` and `winsorizetrait` parameters).

**Trait normalisation:** Prior to correlation, trait values were tested for normality with
Shapiro-Wilk. If the null hypothesis of normality was rejected (p < 0.05), a log10 transform
was applied (log10(x + ε), ε = min non-zero value / 10), following Valenzuela _et al._ (2024).

**Statistical inference:** Pearson correlation between trait paths and RER paths per gene.
The parametric p-values are Benjamini–Hochberg corrected within `correlateWithContinuousPhenotype()`
and reported as `p.adj` (Kowalczyk _et al._ 2019); they treat the branches as independent and can be miscalibrated
in either direction (Saputra _et al._ 2021). Significance is assessed against `pval_threshold`, on the raw permulation
p-value `p.perm` when permulations were run, and on `p.adj` otherwise. A gene is called **positively correlated** if
it is significant and Rho > 0, and **negatively correlated** if it is significant and Rho < 0.

**Permutation statistics:** `r if (has_perm) { n_str <- if (!is.null(n_perms_total) && !is.na(n_perms_total)) paste0(n_perms_total, " total permulations; the smallest p.perm is ", signif(perm_floor_val, 3), " (permpvalcor() divides by the null values on the observed side of the median, about N/2, so the floor is about 1/(N/2 + 1))") else "N permulations (count not stored in this RDS)"; paste0("Empirical p-values (p.perm) were computed by generating null phenotypes under a Brownian Motion model of evolution using getPermsContinuous() (permulations, Saputra et al. 2021). Per gene, permpvalcor() compares the observed correlation with the gene's own null correlations, centred on their median, with a pseudo-count (num + 1) / (denom + 1). ", n_str, ". Batches combined with combinePermData(). p.perm.adj is the Benjamini-Hochberg adjustment of p.perm over all genes. Saputra et al. typically use 1,000 permulations and correct the permulation p-values (Storey, or BH) before calling genes; with fewer than about 125 permulations the permulation does not improve on the parametric p (their Fig. 5C).") } else "Not run (rer_perm_batches = 0). Re-run with rer_perm_batches > 0 to enable (Saputra et al. 2021 typically use 1,000 permulations, here 10 batches × 100)."`

---

## Summary

This report summarizes the results of the **RERconverge** analysis.

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — RERconverge Module*
*Date of generation: `r Sys.Date()`*
