---
title: "CAAS Scoring Report"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
    mathjax: null                       # no math in this report
    md_extensions: "-tex_math_dollars"  # gene/pathway names can contain '$' — never parse as TeX math
params:
  position_scores_file: NULL
  gene_scores_file: NULL
  gene_corr_file: NULL
  traitname: "unknown_trait"
  output_dir: "."
  top_pct: 0.10
  gene_top_pct: 0.10
  stress_summary_file: NULL
  stress_corr_file: NULL
  stress_rank_file: NULL
  stress_overlap_file: NULL
  stress_variants_file: NULL
  stress_loadings_file: NULL
  fade_site_top_file: NULL
  fade_site_bot_file: NULL
  vep_primateai_file: NULL
  vep_cosmic_file: NULL
  genomic_info_file: NULL
  caas_perms_file: NULL
  caas_pos_pval_file: NULL
  caas_pos_sample_file: NULL
  filtered_discovery_file: NULL
  window_size_bp: 1000000
  direction: "combined"
  background_file: NULL
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

<!--
  Scoring Report — PhyloPhere Pipeline
  ======================================================

  Author: Miguel Ramon (miguel.ramon@upf.edu)
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  out.width = "100%"
)

suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(ggplot2)
  library(DT)
  library(scales)
  library(ggrepel)
})

# Tableau_10 colour swatch for pipeline-wide consistency
if (requireNamespace("paletteer", quietly = TRUE)) {
  .tableau10 <- as.character(paletteer::paletteer_d("ggthemes::Tableau_10"))
} else {
  .tableau10 <- c(
    "#4E79A7", "#F28E2B", "#E15759", "#76B7B2", "#59A14F",
    "#EDC948", "#B07AA1", "#FF9DA7", "#9C755F", "#BAB0AC"
  )
}
theme_set(theme_minimal(base_size = 13))

# Safe correlation helper: returns NA (not an error) when inputs have fewer
# than 3 complete pairs — avoids crashes in sparse optional-module sections.
safe_cor <- function(x, y, method = "pearson") {
  ok <- complete.cases(x, y)
  if (sum(ok) < 3) return(NA_real_)
  cor(x[ok], y[ok], method = method)
}

format_bound <- function(x, digits = 2) {
  ifelse(
    is.na(x),
    NA_character_,
    ifelse(abs(x) < 1e-3 | abs(x) >= 1e3,
           formatC(x, format = "e", digits = digits),
           format(round(x, digits), nsmall = digits, trim = TRUE))
  )
}

decile_boundary_tbl <- function(values, score_name,
                                lower_better = TRUE,
                                value_label = "Value",
                                digits = 2) {
  vals <- values[is.finite(values) & !is.na(values)]
  if (length(vals) < 10) return(NULL)
  breaks <- quantile(vals, probs = seq(0, 1, 0.1), na.rm = TRUE, names = FALSE)
  deciles <- if (lower_better) 10:1 else 1:10
  # For lower-better metrics (e.g. p-values), decile 10 should map to the
  # smallest interval [q0, q10%], decile 9 to [q10%, q20%], ..., decile 1 to
  # the largest interval [q90%, q100%].
  lower_bounds <- breaks[1:10]
  upper_bounds <- c(breaks[2:10], max(vals, na.rm = TRUE))
  score_vals <- (deciles - 1) / 9
  tibble(
    Decile = deciles,
    `Lower bound` = format_bound(lower_bounds, digits = digits),
    `Upper bound` = format_bound(upper_bounds, digits = digits),
    !!score_name := round(score_vals, 3)
  ) %>%
    rename(
      !!paste(value_label, "lower bound") := `Lower bound`,
      !!paste(value_label, "upper bound") := `Upper bound`
    )
}
```

```{r load-data}
pos_df  <- read_tsv(params$position_scores_file, show_col_types = FALSE)
gene_df <- read_tsv(params$gene_scores_file, show_col_types = FALSE)
corr_df <- read_tsv(params$gene_corr_file, show_col_types = FALSE)

# Some runs export raw p-values; others export only the derived decile scores.
# Pick the available measure for each distribution section.
hyper_measure_col <- if ("pvalue" %in% names(pos_df)) {
  "pvalue"
} else if ("variability_score" %in% names(pos_df)) {
  "variability_score"
} else {
  stop("position_scores.tsv must contain either pvalue or variability_score")
}
boot_measure_col <- if ("pvalue_boot" %in% names(pos_df)) {
  "pvalue_boot"
} else if ("pattern_score" %in% names(pos_df)) {
  "pattern_score"
} else {
  stop("position_scores.tsv must contain either pvalue_boot or pattern_score")
}
hyper_measure_is_raw <- identical(hyper_measure_col, "pvalue")
boot_measure_is_raw  <- identical(boot_measure_col,  "pvalue_boot")

# FDR-corrected hypergeometric p-value columns
fdr_measure_col <- if ("pvalue_hyp_fdr" %in% names(pos_df)) {
  "pvalue_hyp_fdr"
} else {
  NULL
}
has_fdr_measure <- !is.null(fdr_measure_col)

# Provisional detection — overridden in the load-data block after directional scores are computed
has_rand <- "accum_fisher_p" %in% names(gene_df) && any(!is.na(gene_df$accum_fisher_p))
has_rer  <- "rer_min_pval"    %in% names(gene_df) && any(!is.na(gene_df$rer_min_pval)) &&
            "rer_rho"         %in% names(gene_df)

# ── VEP optional inputs ───────────────────────────────────────────────────────
.file_ok <- function(p) !is.null(p) && !is.na(p) && nchar(p) > 0 && file.exists(p)

has_scoring_stress <- .file_ok(params$stress_summary_file) &&
  .file_ok(params$stress_corr_file) &&
  .file_ok(params$stress_rank_file) &&
  .file_ok(params$stress_overlap_file) &&
  .file_ok(params$stress_variants_file)
has_vep_primateai <- .file_ok(params$vep_primateai_file)
has_genomic       <- .file_ok(params$genomic_info_file)
has_perms         <- .file_ok(params$caas_perms_file)
has_pos_pval      <- .file_ok(params$caas_pos_pval_file)      # lean recovery p-value per (gene,position,scheme)
has_pos_sample    <- .file_ok(params$caas_pos_sample_file)    # lean capped null sample (asr + row_caas)
has_filtered      <- .file_ok(params$filtered_discovery_file) # observed per-scheme asr_path_score + pvalue_boot
# FADE site data: use top direction if available, otherwise try bottom
has_fade_pos      <- .file_ok(params$fade_site_top_file) || .file_ok(params$fade_site_bot_file)
# Initialized FALSE here so later chunks with eval=has_fade_any don't crash when
# the fade-pos-setup chunk is skipped (has_fade_pos == FALSE).
has_fade_top <- FALSE
has_fade_bot <- FALSE
has_fade_any <- FALSE
has_vep_cosmic     <- .file_ok(params$vep_cosmic_file)
has_pos_char       <- has_vep_primateai || has_vep_cosmic || has_genomic

if (has_vep_primateai) {
  vep_pai_df <- read_tsv(params$vep_primateai_file, show_col_types = FALSE)
  # read_tsv may infer Position as character (e.g. empty/NA cells); the scored
  # positions it is joined to are numeric, so coerce to a common integer type.
  if ("Position" %in% names(vep_pai_df)) vep_pai_df$Position <- suppressWarnings(as.integer(vep_pai_df$Position))
} else {
  vep_pai_df <- NULL
}

if (has_vep_cosmic) {
  vep_cosmic_df <- read_tsv(params$vep_cosmic_file, show_col_types = FALSE)
  if ("Position" %in% names(vep_cosmic_df)) vep_cosmic_df$Position <- suppressWarnings(as.integer(vep_cosmic_df$Position))
} else {
  vep_cosmic_df <- NULL
}

if (has_genomic) {
  # Format: gene, chr, start, end, length  (no header assumed; detect if header present)
  genomic_raw <- read_tsv(params$genomic_info_file,
                          col_names = c("Gene", "chr", "gene_start", "gene_end", "gene_length"),
                          show_col_types = FALSE)
  # Drop header row if first cell is literally "gene" or "Gene"
  if (tolower(genomic_raw$Gene[1]) == "gene") genomic_raw <- genomic_raw[-1, ]
  genomic_raw <- genomic_raw %>%
    mutate(
      gene_start  = as.integer(gene_start),
      gene_end    = as.integer(gene_end),
      gene_length = as.integer(gene_length),
      gene_mid    = as.integer((gene_start + gene_end) / 2L)
    )
} else {
  genomic_raw <- NULL
}

# ── Gene-level permulation null (rds: asr corStat + NEW caas corStat) ──────────
# Tidy both byrank lists into one long frame (gene × cycle × direction × metric)
# for the gene-level tabs. asr names are *_asr; strip the suffix → global/top/bottom.
if (has_perms) {
  caas_perms_obj <- tryCatch(readRDS(params$caas_perms_file), error = function(e) NULL)
  .tidy_byrank <- function(cs, metric) {
    purrr::imap_dfr(cs, function(mat, rank_name) {
      mat <- as.matrix(mat)
      val <- as.numeric(mat)
      if (length(val) > 100000) {
        set.seed(42)
        val <- sample(val, 100000)
      }
      tibble::tibble(metric = metric, direction = sub("_asr$", "", rank_name),
                     null_score = val)
    })
  }
  asr_cs  <- if (!is.null(caas_perms_obj)) caas_perms_obj$corStat_byrank      else NULL
  caas_cs <- if (!is.null(caas_perms_obj)) caas_perms_obj$caas_corStat_byrank else NULL
  if (is.null(asr_cs) || length(asr_cs) == 0) {
    has_perms <- FALSE; perm_null_gene <- NULL
  } else {
    perm_null_gene <- dplyr::bind_rows(
        .tidy_byrank(asr_cs, "asr_path_score"),
        if (!is.null(caas_cs) && length(caas_cs)) .tidy_byrank(caas_cs, "caas_score") else NULL) %>%
      mutate(direction = factor(direction, levels = c("global", "top", "bottom")),
             metric    = factor(metric, levels = c("caas_score", "asr_path_score")))
    perm_n_cycles <- ncol(as.matrix(asr_cs[[1]]))
    perm_n_genes  <- nrow(as.matrix(asr_cs[[1]]))
  }
} else {
  caas_perms_obj <- NULL; perm_null_gene <- NULL
}

# ── Position-level null (lean sample) + observed (filtered_discovery) per scheme ──
# Null sample carries asr_path_score, recovery_pval, row_caas per (scheme,cycle).
# Observed row_caas mirrors scoring_compute.R: (1 - pvalue_boot) * asr * 0.2.
scheme_levels <- c("US", "GS1", "GS2", "GS3", "GS4")
if (has_pos_sample && has_filtered) {
  null_pos <- readr::read_tsv(params$caas_pos_sample_file, show_col_types = FALSE) %>%
    transmute(caap_group,
              asr_path_score = suppressWarnings(as.numeric(asr_path_score)),
              caas_row       = suppressWarnings(as.numeric(row_caas)),
              pval           = suppressWarnings(as.numeric(recovery_pval)),
              source = "null")
  obs_pos <- readr::read_tsv(params$filtered_discovery_file, show_col_types = FALSE) %>%
    transmute(caap_group,
              asr_path_score = suppressWarnings(as.numeric(asr_path_score)),
              pval           = suppressWarnings(as.numeric(pvalue_boot)),
              caas_row       = (1 - pmin(pmax(pval, 0), 1)) * asr_path_score * 0.2,
              source = "observed")
  scheme_overlay <- bind_rows(null_pos, obs_pos) %>%
    filter(caap_group %in% scheme_levels) %>%
    mutate(caap_group = factor(caap_group, levels = scheme_levels),
           source     = factor(source, levels = c("null", "observed")))
  has_scheme_overlay <- nrow(scheme_overlay) > 0
} else {
  scheme_overlay <- NULL
  has_scheme_overlay <- FALSE
}

# Full recovery p-value table (small) — for the position-level p-value distribution.
if (has_pos_pval) {
  perm_pos_pval <- readr::read_tsv(params$caas_pos_pval_file, show_col_types = FALSE) %>%
    filter(caap_group %in% scheme_levels) %>%
    mutate(caap_group = factor(caap_group, levels = scheme_levels))
} else {
  perm_pos_pval <- NULL
}

accum_scheme_info <- tibble(
  scheme = c("us", "gs4", "gs3", "gs2", "gs1"),
  label = c("US", "GS4", "GS3", "GS2", "GS1"),
  weight = c(0.5, 0.2, 0.1, 0.1, 0.1),
  pval_col = paste0("accum_pval_", c("us", "gs4", "gs3", "gs2", "gs1")),
  score_col = paste0("accum_score_", c("us", "gs4", "gs3", "gs2", "gs1")),
  weighted_col = paste0("accum_weighted_", c("us", "gs4", "gs3", "gs2", "gs1"))
) %>%
  filter(pval_col %in% names(gene_df))

if (has_scoring_stress) {
  stress_summary_df <- read_tsv(params$stress_summary_file, show_col_types = FALSE)
  stress_corr_df <- read_tsv(params$stress_corr_file, show_col_types = FALSE)
  stress_rank_df <- read_tsv(params$stress_rank_file, show_col_types = FALSE)
  stress_overlap_df <- read_tsv(params$stress_overlap_file, show_col_types = FALSE)
  stress_variants_df <- read_tsv(params$stress_variants_file, show_col_types = FALSE)
  stress_loadings_df <- if (.file_ok(params$stress_loadings_file)) {
    read_tsv(params$stress_loadings_file, show_col_types = FALSE)
  } else {
    NULL
  }
} else {
  stress_summary_df <- NULL
  stress_corr_df <- NULL
  stress_rank_df <- NULL
  stress_overlap_df <- NULL
  stress_variants_df <- NULL
  stress_loadings_df <- NULL
}

# Thresholds
top_pct   <- params$top_pct
top25_pct <- 0.25
top5_pct  <- 0.05
top1_pct  <- 0.01

# ── RERConverge availability ────────────────────────────────────────────────
has_rer <- "rer_min_pval" %in% names(gene_df) &&
           any(!is.na(gene_df$rer_min_pval)) &&
           "rer_rho" %in% names(gene_df)

# ── Directional position subsets (change_side from discovery) ──────────────
pos_top    <- pos_df %>% filter(change_side %in% c("top",    "both"))
pos_bottom <- pos_df %>% filter(change_side %in% c("bottom", "both"))

# ── Directional gene CAAS scores ────────────────────────────────────────────
# 90th percentile of CAAS_score per gene, computed separately for TOP and BOTTOM
# positions. This mirrors how gene_caas_score is computed but direction-aware.
gene_caas_top_df <- pos_top %>%
  group_by(Gene) %>%
  summarise(gene_caas_top = quantile(CAAS_score, 0.9, na.rm = TRUE), .groups = "drop")

gene_caas_bot_df <- pos_bottom %>%
  group_by(Gene) %>%
  summarise(gene_caas_bot = quantile(CAAS_score, 0.9, na.rm = TRUE), .groups = "drop")

gene_df <- gene_df %>%
  left_join(gene_caas_top_df, by = "Gene") %>%
  left_join(gene_caas_bot_df, by = "Gene")

# ── Gene-level accumulation significance ─────────────────────────────────────
# Accumulation is represented by its Fisher combined p-value (accum_fisher_p) across
# the per-scheme accum p-values, with BH FDR (accum_fdr) and accum_significant
# (accum_fdr < 0.05). The old -log10 "score" (gene_rand_score) is deprecated.
has_accum_score <- "accum_fisher_p" %in% names(gene_df) && any(!is.na(gene_df$accum_fisher_p))
has_rand <- has_accum_score

# ── Gene-level direction assignment ─────────────────────────────────────────
# Assign direction to each gene from its top-scoring position
gene_direction_df <- pos_df %>%
  group_by(Gene) %>%
  slice_max(CAAS_score, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(Gene, gene_direction = change_side)

gene_df <- gene_df %>% left_join(gene_direction_df, by = "Gene")

# ── Gate-based gene lists (global / p-val sig / FDR sig) ─────────────────────
# Derived from position-level hypergeometric significance flags so gene membership
# is tied to the test outcome rather than an arbitrary top-% rank cut.
.gate_genes <- function(pos_sub, gate_col = NULL) {
  if (!is.null(gate_col) && gate_col %in% names(pos_sub))
    pos_sub %>% filter(.data[[gate_col]] %in% TRUE) %>% pull(Gene) %>% unique()
  else
    pos_sub %>% pull(Gene) %>% unique()
}

gene_global_top <- .gate_genes(pos_top)
gene_global_bot <- .gate_genes(pos_bottom)
gene_sig_top    <- .gate_genes(pos_top,    "gate_sig")
gene_sig_bot    <- .gate_genes(pos_bottom, "gate_sig")
gene_fdr_top    <- .gate_genes(pos_top,    "gate_fdr")
gene_fdr_bot    <- .gate_genes(pos_bottom, "gate_fdr")

# Primary aliases used in tables — p-val sig gate is most informative default
gene_top_list <- gene_sig_top
gene_bot_list <- gene_sig_bot

# Named list for looping over all three gates in overlap sections
.caas_gates <- list(
  list(lbl = "Global",        top = gene_global_top, bot = gene_global_bot),
  list(lbl = "Sig (p<0.05)", top = gene_sig_top,    bot = gene_sig_bot),
  list(lbl = "FDR sig",      top = gene_fdr_top,    bot = gene_fdr_bot)
)

# Background universe for gene-set overlap Fisher tests (RER / FADE / Accumulation
# Venn panels) — every gene that received a CAAS gene score in this run.
gene_universe_n <- nrow(gene_df)

# ── Normalisation helper ──────────────────────────────────────────────────────
rank_norm <- function(x) {
  r <- rank(x, na.last = "keep", ties.method = "average")
  (r - 1) / (max(r, na.rm = TRUE) - 1)
}

# ── Venn diagram helpers (used across RER / FADE / Accumulation tabs) ─────────
# Overlap enrichment is tested with a Fisher exact test (mathematically the
# hypergeometric test on a 2x2 table) against the full gene universe passed in
# `universe_n` — for gene-level overlaps this is nrow(gene_df), i.e. every gene
# that received a CAAS gene score in this run.
.venn_counts <- function(set_a, set_b, label_a, label_b, universe_n) {
  n_a  <- length(set_a); n_b <- length(set_b)
  n_ab <- length(intersect(set_a, set_b))
  n_only_a  <- n_a - n_ab
  n_only_b  <- n_b - n_ab
  n_neither <- universe_n - n_a - n_b + n_ab
  ft <- tryCatch(
    fisher.test(matrix(c(n_ab, n_only_b, n_only_a, n_neither), nrow = 2)),
    error = function(e) list(estimate = NA_real_, p.value = NA_real_)
  )
  tibble(label_a, label_b, n_a, n_b, n_ab,
         n_only_a = n_only_a, n_only_b = n_only_b,
         odds_ratio = unname(ft$estimate), fisher_p = ft$p.value)
}
.draw_venn <- function(vc, col_a, col_b, title) {
  or_txt <- if (is.na(vc$odds_ratio)) {
    "OR = NA"
  } else {
    sprintf("OR = %.2f (Fisher p = %s)", vc$odds_ratio,
            formatC(vc$fisher_p, format = "e", digits = 1))
  }
  ggplot() +
    geom_rect(aes(xmin=0.05, xmax=0.65, ymin=0.22, ymax=0.85),
              fill=col_a, alpha=0.35, colour=col_a, linewidth=1) +
    geom_rect(aes(xmin=0.35, xmax=0.95, ymin=0.22, ymax=0.85),
              fill=col_b, alpha=0.35, colour=col_b, linewidth=1) +
    annotate("text", x=0.20, y=0.535, label=vc$n_only_a, size=7, fontface="bold") +
    annotate("text", x=0.50, y=0.535, label=vc$n_ab,     size=7, fontface="bold", colour="grey20") +
    annotate("text", x=0.80, y=0.535, label=vc$n_only_b, size=7, fontface="bold") +
    annotate("text", x=0.20, y=0.15, label=vc$label_a, size=3.5, colour=col_a) +
    annotate("text", x=0.80, y=0.15, label=vc$label_b, size=3.5, colour=col_b) +
    annotate("text", x=0.50, y=0.93,
             label=sprintf("Shared: %d genes (%.0f%% of smaller set)",
                           vc$n_ab, 100 * vc$n_ab / max(min(vc$n_a, vc$n_b), 1)),
             size=3, colour="grey40") +
    annotate("text", x=0.50, y=0.02, label=or_txt, size=3, fontface="bold", colour="grey20") +
    ggtitle(title) + theme_void() +
    coord_cartesian(xlim = c(0, 1), ylim = c(-0.06, 1.02), clip = "off") +
    theme(plot.title  = element_text(face="bold", hjust=0.5, size=11),
          plot.margin = margin(t = 4, r = 4, b = 14, l = 4))
}
```

---

## Introduction

The **CAAS Scoring** module integrates signals from multiple upstream analyses
into a unified, interpretable metric for prioritising convergent amino acid
substitutions (CAAS).  The scoring operates at two levels:

1. **Position-level CAAS Score** — each alignment position receives a composite
   score that combines:
   - **Phenotype evolution** — permulation partition evidence ($1 - \text{pvalue\_boot}$).
   - **Position evolution** — continuous ASR path score ($\text{asr\_path\_score}$ / $\text{asr\_score}$), aggregating isolation, independence, diversity, agreement, and conservation elements.

2. **Gene-level scores** — each gene receives directional CAAS scores and is
   compared against independent evolutionary signals:
   - **gene_caas_top / gene_caas_bot** — 90th percentile of position-level CAAS
     scores for TOP and BOTTOM direction positions respectively.
   `r if (has_rand) "- **accum_fisher_p** — Fisher combined p-value across per-scheme accumulation p-values (+ BH FDR accum_fdr)."`
   `r if (has_rer) "- RERConverge p-values and direction (accelerating / decelerating)."`
   - FADE gene-level BF ≥ 100 significance, tested separately per direction.
   `r if (has_vep_primateai) "- **PrimateAI-3D Validation** — PAI3D pathogenicity scores collapsed to position-level (maximum per CAAP group, averaged across schemes)."`

All comparisons use rank-normalised variables for correlations.

## Companion reports

This scoring report is the hub of the downstream enrichment suite. The reports below
are rendered as sibling HTML files in the same `html_reports/` directory and are
auto-annotated with the cross-module scoring signals (CAAS gates, top/bottom
percentiles, FADE, accumulation) produced here. A link is live only if the
corresponding module was part of this run. FADE and Accumulation no longer run
their own FCS ranking (unreliable for genome-wide gene-set enrichment — see the
Comparison report's methods note); they remain available there as cross-module
corroboration flags. Position Characterisation (PrimateAI-3D + COSMIC validation)
has moved into the Position-level FCS report below.

```{r companion-reports, results='asis'}
.trait <- if (is.null(params$traitname) || !nzchar(params$traitname)) "unknown_trait" else params$traitname
.companions <- tibble::tribble(
  ~Report,                              ~File,
  "FCS — CAAS (this scoring run)",      sprintf("13.FCS_scoring_%s.html",    .trait),
  "FCS — RERconverge (+ p.perm)",       sprintf("13.FCS_rer_%s.html",        .trait),
  "Position-level FCS (+ Position Characterisation)", sprintf("16.Position_enrichment_report_%s.html", .trait),
  "STRING (network / PPI / clusters)",  sprintf("15.STRING_report_%s.html", .trait),
  "COMPARE — cross-module overview",    sprintf("12.Comparison_report_%s.html",.trait)
)
cat(paste0("- [", .companions$Report, "](", .companions$File, ")", collapse = "\n"))
cat("\n")
```

---

## Scoring Methodology

This section documents the exact weighting and combination logic used to
produce every score in this report, so results can be interpreted and
reproduced independently of the pipeline code.

<details>
<summary><strong>Click to expand full methodology</strong></summary>

---

### Position-Level Score (two-way)

The position-level **`CAAS_score`** is the product of **two orthogonal evidence
axes**, each already on a raw **[0, 1]** scale (no decile transforms), computed
**per scheme row** and then scheme-weight-aggregated. The hypergeometric p-value is
*not* part of the score — it is a separate three-way significance **gate** (below).

#### 1 · Phenotype evolution — `phen_score`

Permulation evidence: one minus the **permutation p-value** (`pvalue_boot`), which
permutes the phenotype on the tree. Low permutation p-value → strong, phylogenetically
honest phenotype-partition signal → high evidence.

$$\text{phen\_score} = 1 - P_{\text{boot}}$$

#### 2 · Position evolution — `asr_path_score` (`asr_score`)

The unified ASR path score, built from the per-node ASR posteriors as the product of
**five factors** (full definitions in
[`docs/ASR_PATH_SCORE.md`](../../docs/ASR_PATH_SCORE.md)):

| Factor | What it rewards |
|--------|-----------------|
| `core` | P(≥2 independent changes): combinatorial replication across directions, built from per-pair isolation (the change is made above the MRCA, not inherited) |
| `independence` | derived state **absent** at the pairs' shared ancestors (LAC nodes) |
| `mrca_diversity` | non-parallel origins (distinct ancestral starting posteriors) |
| `derived_agreement` | changed pairs landing on the **same** derived residue per side |
| `conservation_gate` | conserved pairs holding the ancestral state deeply |

It folds in the old binary ASR gate, convergence, and parallel signals; those legacy
columns survive only as display diagnostics.

#### 3 · Biochemical weighting & aggregation

Each scoring scheme (US, GS1–GS4) carries **equal weight of 0.2**. The per-row two-way
score is multiplied by its scheme weight:

$$\text{caas\_row} = \text{phen\_score} \times \text{asr\_path\_score}$$

$$\text{row\_caas} = \text{caas\_row} \times w_s, \qquad w_s = 0.2$$

$$\text{CAAS\_score} = \sum_{s \in \text{detected schemes}} \text{row\_caas}$$

Each scheme contributes at most **0.2**; the maximum achievable `CAAS_score` = **1.0**
(all five schemes, both axes = 1.0). `biochem_weight_sum` is the sum of detected
scheme weights (max 1.0), a diagnostic of how many schemes flagged the position.

#### 4 · Significance gate (hypergeometric `pvalue`)

The hypergeometric CAAS p-value gates positions into three nested tiers, so callers
rank by `CAAS_score` and then filter to a defensible significance level:

| Tier | Column | Condition |
|------|--------|-----------|
| All | `gate_all` | every scored position (the full ranked pool) |
| Significant | `gate_sig` | `pvalue < 0.05` |
| FDR-significant | `gate_fdr` | BH-adjusted `pvalue < alpha_fdr` (`pvalue_hyp_fdr`) (threshold `r if ("alpha_fdr" %in% names(pos_df)) paste(unique(pos_df$alpha_fdr), collapse = ", ") else "0.05"` in this run) |

---

### Gene-Level Scores

#### gene\_caas\_top / gene\_caas\_bot

90th percentile of `CAAS_score` computed separately for TOP-direction and
BOTTOM-direction positions within each gene:

$$\text{gene\_caas\_top} = Q_{0.90}(\text{CAAS\_score} \mid \text{change\_side} \in \{\text{top, both}\})$$
$$\text{gene\_caas\_bot} = Q_{0.90}(\text{CAAS\_score} \mid \text{change\_side} \in \{\text{bottom, both}\})$$

#### gene\_accum\_score  *(optional)*

Equally weighted sum of accumulation randomization p-values.
Weights are equal at 0.2 for each of the active schemes (US, GS1-GS4):

$$\text{gene\_accum\_score} = 1 - \frac{\sum_{s \in \{US, GS4, GS3, GS2, GS1\}} 0.2 \times P_{\text{accum},s}}{n_{\text{available schemes}}}$$

Maximum = 1.0 (all schemes p-value = 0).

#### PrimateAI-3D validation  *(optional)*

The PrimateAI-3D characterization score (`pai_char_score`) is calculated per protein position as the average of the maximum PAI3D score per CAAP group across detected biochemically grouped schemes (US, GS1-GS4):
1. **Stance collapse**: Take the maximum PAI3D score per (Gene × Position × CAAP group) to collapse multiple alternative amino acids per stance.
2. **Stance average**: Average these maximums across the CAAP groups where the scheme weight is positive (`scheme_weight > 0`).

#### RERConverge characterisation  *(optional)*

Raw permutation p-value (`p.perm`) used directly. Direction determined by sign
of `Rho`: positive = accelerating (high-phenotype lineages have elevated rate),
negative = decelerating. Significance threshold: p.perm ≤ 0.05.

#### FADE gene-level characterisation  *(optional)*

Sites are aggregated to gene level. A gene is FADE-significant if it contains
at least one position with `max_site_bf ≥ 100`, evaluated separately for
TOP and BOTTOM directions.

---

### ORA Gene List Extraction

Gene lists for Over-Representation Analysis are extracted at three thresholds
(configurable via `--scoring_position_top_pct` / `--scoring_position_top5_pct` /
`--scoring_position_top1_pct` and their gene-level equivalents) and three
directional subsets:

| Subset | Filter |
|--------|--------|
| **top** | Positions / genes where `change_side == "top"` |
| **bottom** | Positions / genes where `change_side == "bottom"` |
| **both** | All top-scoring positions / genes regardless of direction |

For **position-level** lists, the threshold is applied to `CAAS_score` and
the gene name is extracted from the qualifying positions.

For **gene-level** lists, the threshold is applied to `gene_caas_top` (TOP
direction) and `gene_caas_bot` (BOTTOM direction) independently.

</details>

---

## Input Distributions {.tabset}

This section displays the distributions of the primary input significance measures (Hypergeometric P-values, FDR-adjusted hypergeometric P-values, and Permutation P-values) along with joint comparisons to show how they behave relative to each other and to the final CAAS score.

### Hypergeometric P-value

```{r hyper-pval-dist, fig.height=8}
# Compute decile boundaries from the selected hypergeometric measure.
hyper_values <- pos_df[[hyper_measure_col]]
pv_breaks <- quantile(hyper_values, probs = seq(0, 1, by = 0.1), na.rm = TRUE)
decile_df_h <- pos_df %>%
  filter(!is.na(.data[[hyper_measure_col]])) %>%
  mutate(
    measure_value = .data[[hyper_measure_col]],
    plot_value    = if (hyper_measure_is_raw) -log10(pmax(measure_value, 1e-300)) else measure_value,
    decile_raw    = dplyr::ntile(measure_value, 10),
    decile        = if (hyper_measure_is_raw) 11 - decile_raw else decile_raw,
    score_val     = (decile - 1) / 9
  )

# Panel 1: density on the selected measure scale with decile boundaries
p_hyp_density <- ggplot(decile_df_h, aes(x = plot_value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60,
                 fill = .tableau10[1], alpha = 0.6) +
  geom_density(linewidth = 0.8, colour = .tableau10[2]) +
  geom_vline(
    xintercept = if (hyper_measure_is_raw) -log10(pmax(pv_breaks[2:10], 1e-300)) else pv_breaks[2:10],
    linetype = "dotted", colour = "grey50", linewidth = 0.4
  ) +
  labs(
    title    = if (hyper_measure_is_raw) "Hypergeometric P-value Distribution (-log10 scale)" else "Variability Score Distribution",
    subtitle = if (hyper_measure_is_raw) {
      "Dotted lines: decile boundaries (D10 = smallest p-values, highest score; D1 = largest p-values)"
    } else {
      "Dotted lines: decile boundaries of the exported variability_score values"
    },
    x        = if (hyper_measure_is_raw) "-log10(pvalue)" else "variability_score",
    y        = "Density"
  ) +
  scale_x_continuous(
    breaks = scales::pretty_breaks(n = 14),
    minor_breaks = scales::pretty_breaks(n = 28)
  ) +
  theme(plot.title = element_text(face = "bold"))

print(p_hyp_density)

# Decile boundary table: what p-value range does each decile cover?
if (hyper_measure_is_raw) {
  .hyp_lower <- pv_breaks[1:10]
  .hyp_upper <- c(pv_breaks[2:10], max(hyper_values, na.rm = TRUE))
  hyp_boundary_tbl <- tibble(
    Decile                = 10:1,
    `P-value lower bound` = formatC(.hyp_lower, format = "e", digits = 2),
    `P-value upper bound` = formatC(.hyp_upper, format = "e", digits = 2),
    variability_score     = round((10:1 - 1) / 9, 3)
  )
  knitr::kable(
    hyp_boundary_tbl,
    caption = "Hypergeometric p-value range per decile (decile 10 = smallest p-values, score = 1.0)"
  )
}
```

### FDR-adjusted P-value

```{r fdr-pval-dist, fig.height=8, eval=has_fdr_measure}
# Compute decile boundaries from the FDR hypergeometric measure.
fdr_values <- pos_df[[fdr_measure_col]]
pv_breaks_f <- quantile(fdr_values, probs = seq(0, 1, by = 0.1), na.rm = TRUE)
decile_df_f <- pos_df %>%
  filter(!is.na(.data[[fdr_measure_col]])) %>%
  mutate(
    measure_value = .data[[fdr_measure_col]],
    plot_value    = -log10(pmax(measure_value, 1e-300)),
    decile_raw    = dplyr::ntile(measure_value, 10),
    decile        = 11 - decile_raw,
    score_val     = (decile - 1) / 9
  )

# Panel 1: density on the FDR measure scale with decile boundaries
p_fdr_density <- ggplot(decile_df_f, aes(x = plot_value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60,
                 fill = .tableau10[4], alpha = 0.6) +
  geom_density(linewidth = 0.8, colour = .tableau10[3]) +
  geom_vline(
    xintercept = -log10(pmax(pv_breaks_f[2:10], 1e-300)),
    linetype = "dotted", colour = "grey50", linewidth = 0.4
  ) +
  labs(
    title    = "FDR-adjusted Hypergeometric P-value Distribution (-log10 scale)",
    subtitle = "Dotted lines: decile boundaries (D10 = smallest p-values, D1 = largest p-values)",
    x        = "-log10(pvalue_hyp_fdr)",
    y        = "Density"
  ) +
  scale_x_continuous(
    breaks = scales::pretty_breaks(n = 14),
    minor_breaks = scales::pretty_breaks(n = 28)
  ) +
  theme(plot.title = element_text(face = "bold"))

print(p_fdr_density)

# Decile boundary table
.fdr_lower <- pv_breaks_f[1:10]
.fdr_upper <- c(pv_breaks_f[2:10], max(fdr_values, na.rm = TRUE))
fdr_boundary_tbl <- tibble(
  Decile                = 10:1,
  `P-value lower bound` = formatC(.fdr_lower, format = "e", digits = 2),
  `P-value upper bound` = formatC(.fdr_upper, format = "e", digits = 2),
  score                 = round((10:1 - 1) / 9, 3)
)
knitr::kable(
  fdr_boundary_tbl,
  caption = "FDR p-value range per decile (decile 10 = smallest p-values)"
)
```

`r if (!has_fdr_measure) "_FDR hypergeometric p-values not available in this run._"`

### Significance Gate

The hypergeometric p-value is not part of `CAAS_score`; it tiers positions into
nested significance levels. Rank by `CAAS_score`, then keep whichever tier you need.

```{r gate-summary}
if (all(c("gate_all", "gate_sig", "gate_fdr") %in% names(pos_df))) {
  gate_tbl <- tibble::tibble(
    Tier = c("All (ranked pool)", "Significant (pvalue < 0.05)",
             "FDR-significant (BH < 0.05)"),
    Positions = c(sum(pos_df$gate_all, na.rm = TRUE),
                  sum(pos_df$gate_sig, na.rm = TRUE),
                  sum(pos_df$gate_fdr, na.rm = TRUE))
  ) %>% mutate(`% of pool` = round(100 * Positions / max(Positions, 1), 1))
  knitr::kable(gate_tbl,
               caption = "Hypergeometric significance gate — positions per tier")
} else {
  cat("_Significance gate columns (gate_all / gate_sig / gate_fdr) not present in this run._\n")
}
```

### Joint Decile Summary

```{r joint-decile-summary, fig.height=5, fig.width=10}
# Scatter of hypergeometric decile vs pattern decile, and FDR decile vs pattern decile, colored by CAAS_score.
joint_df <- pos_df %>%
  filter(!is.na(.data[[hyper_measure_col]]), !is.na(.data[[boot_measure_col]])) %>%
  mutate(
    hyp_decile_raw  = dplyr::ntile(.data[[hyper_measure_col]], 10),
    boot_decile_raw = dplyr::ntile(.data[[boot_measure_col]], 10),
    hyp_decile      = if (hyper_measure_is_raw) 11 - hyp_decile_raw else hyp_decile_raw,
    boot_decile     = if (boot_measure_is_raw) 11 - boot_decile_raw else boot_decile_raw
  )

.r_hyp_caas  <- round(cor(joint_df$hyp_decile,  joint_df$CAAS_score, use = "complete.obs"), 2)
.r_boot_caas <- round(cor(joint_df$boot_decile, joint_df$CAAS_score, use = "complete.obs"), 2)
.r_hyp_boot  <- round(cor(joint_df$hyp_decile,  joint_df$boot_decile, use = "complete.obs"), 2)

p1 <- ggplot(joint_df, aes(x = hyp_decile, y = boot_decile, colour = CAAS_score)) +
  geom_jitter(alpha = 0.45, size = 1.8, width = 0.35, height = 0.35) +
  scale_colour_gradientn(
    colours = c("#4E79A7", "#76B7B2", "#EDC948", "#F28E2B", "#E15759"),
    name    = "CAAS\nscore"
  ) +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(breaks = 1:10) +
  annotate("text", x = 1, y = 10,
           label = sprintf("r(hyp, CAAS) = %.2f\nr(perm, CAAS) = %.2f\nr(hyp, perm) = %.2f",
                           .r_hyp_caas, .r_boot_caas, .r_hyp_boot),
           hjust = 0, vjust = 1, size = 3.2, colour = "grey30") +
  labs(
    title    = "Hypergeometric vs Permulation Decile",
    subtitle = "High CAAS scores cluster toward 10x10",
    x        = "Hypergeometric decile (10 = most significant)",
    y        = "Permulation decile (10 = most significant)"
  ) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "none")

if (has_fdr_measure) {
  joint_df_f <- joint_df %>%
    filter(!is.na(.data[[fdr_measure_col]])) %>%
    mutate(
      fdr_decile_raw = dplyr::ntile(.data[[fdr_measure_col]], 10),
      fdr_decile     = 11 - fdr_decile_raw
    )

  .r_fdr_caas  <- round(cor(joint_df_f$fdr_decile,  joint_df_f$CAAS_score, use = "complete.obs"), 2)
  .r_fdr_boot  <- round(cor(joint_df_f$fdr_decile,  joint_df_f$boot_decile, use = "complete.obs"), 2)

  p2 <- ggplot(joint_df_f, aes(x = fdr_decile, y = boot_decile, colour = CAAS_score)) +
    geom_jitter(alpha = 0.45, size = 1.8, width = 0.35, height = 0.35) +
    scale_colour_gradientn(
      colours = c("#4E79A7", "#76B7B2", "#EDC948", "#F28E2B", "#E15759"),
      name    = "CAAS\nscore"
    ) +
    scale_x_continuous(breaks = 1:10) +
    scale_y_continuous(breaks = 1:10) +
    annotate("text", x = 1, y = 10,
             label = sprintf("r(fdr, CAAS) = %.2f\nr(perm, CAAS) = %.2f\nr(fdr, perm) = %.2f",
                             .r_fdr_caas, .r_boot_caas, .r_fdr_boot),
             hjust = 0, vjust = 1, size = 3.2, colour = "grey30") +
    labs(
      title    = "FDR vs Permulation Decile",
      subtitle = "High CAAS scores cluster toward 10x10",
      x        = "FDR decile (10 = most significant)",
      y        = "Permulation decile (10 = most significant)"
    ) +
    theme(plot.title = element_text(face = "bold"),
          legend.position = "right")

  gridExtra::grid.arrange(p1, p2, ncol = 2)
} else {
  p1 <- p1 + theme(legend.position = "right")
  print(p1)
}
```

---

## CAAS Permulation Null {.tabset}

The permulation re-labels the phylogeny `N` times and recomputes the per-position
`asr_path_score` under each null labeling — **the only re-permuted component**. Two
null quantities are derived from it, mirroring the observed pipeline:

- **`asr_path_score`** — the permuted position-evolution signal (gene null = 0.90-quantile over the gene's positions).
- **`CAAS_score`** = Σ_scheme `(1 − recovery_pvalue) × asr_path_score × 0.2`, where
  `recovery_pvalue = (k + 1)/(N + 1)` (`k` = cycles a position is recovered; floor `1/(N+1)`)
  is the null analogue of the observed `pvalue_boot`. `pheno_score = 1 − recovery_pvalue`
  is a per-position aggregate over all `N` rounds (no per-cycle value), so it appears here
  only as the **recovery p-value distribution** (Position Level).

The formal per-pathway `p.perm` (asr axis) is in the companion **FCS** report.

```{r perm-gene-obs-setup}
dir_pal <- c(global = "#4E79A7", top = "#E15759", bottom = "#59A14F")
# Observed gene-level distributions by direction (always computed: used both by the
# full null-overlay plots when a permulation null exists, and by the standalone
# summary table below when it doesn't).
.obs_gene_caas_dist <- dplyr::bind_rows(
  tibble::tibble(direction = "global", value = gene_df$gene_caas_score),
  tibble::tibble(direction = "top",    value = gene_df$gene_caas_score_top),
  tibble::tibble(direction = "bottom", value = gene_df$gene_caas_score_bottom)
)
.obs_gene_caas <- .obs_gene_caas_dist %>%
  group_by(direction) %>% summarise(obs_median = median(value, na.rm = TRUE), .groups = "drop")

.q90 <- function(v) if (length(v)) as.numeric(quantile(v, .9, na.rm = TRUE)) else NA_real_
.obs_gene_asr_dist <- pos_df %>%
  group_by(Gene) %>%
  summarise(global = .q90(asr_score),
            top    = .q90(asr_score[change_side %in% c("top", "both")]),
            bottom = .q90(asr_score[change_side %in% c("bottom", "both")]),
            .groups = "drop") %>%
  tidyr::pivot_longer(c(global, top, bottom), names_to = "direction", values_to = "value")
.obs_gene_asr <- .obs_gene_asr_dist %>%
  group_by(direction) %>% summarise(obs_median = median(value, na.rm = TRUE), .groups = "drop")

# caas_cs (caas_corStat_byrank) can be missing from caas_perms.rds even when the asr
# null (asr_cs) is present — flag that separately so the CAAS-score tab can show a
# warning instead of an empty plot.
has_caas_null <- has_perms && !is.null(perm_null_gene) && any(perm_null_gene$metric == "caas_score", na.rm = TRUE)

.gene_null_table <- function(metric_name, obs_df) {
  perm_null_gene %>% filter(metric == metric_name) %>% group_by(direction) %>%
    summarise(`n (gene×cycle)` = dplyr::n(), null_mean = mean(null_score, na.rm = TRUE),
              null_median = median(null_score, na.rm = TRUE), null_sd = sd(null_score, na.rm = TRUE),
              null_IQR = IQR(null_score, na.rm = TRUE), .groups = "drop") %>%
    mutate(direction = as.character(direction)) %>%
    left_join(obs_df, by = "direction") %>%
    arrange(factor(direction, levels = c("global", "top", "bottom")))
}
.gene_null_plot <- function(metric_name, obs_dist_df, ylab, title) {
  nd <- perm_null_gene %>% filter(metric == metric_name, is.finite(null_score)) %>%
    transmute(direction = as.character(direction), value = null_score, source = "Null")
  od <- obs_dist_df %>% transmute(direction = as.character(direction), value, source = "Observed") %>%
    filter(is.finite(value))
  combined <- dplyr::bind_rows(nd, od) %>%
    mutate(direction = factor(direction, levels = c("global", "top", "bottom")),
           source = factor(source, levels = c("Null", "Observed")))
  ggplot(combined, aes(direction, value, fill = source)) +
    geom_violin(scale = "width", alpha = .7, colour = "grey30",
                position = position_dodge(width = 0.8)) +
    scale_fill_manual(values = c(Null = "grey60", Observed = "#E15759"), name = NULL) +
    labs(title = title, subtitle = "Null (permulation, 100k random sample) vs observed gene-level distribution, by direction",
         x = "Direction", y = ylab) +
    theme_minimal(base_size = 12)
}
```

```{r perm-na, eval=!has_perms, results='asis'}
cat("_No CAAS permulation null available for this run",
    " (`--caas_permulation_enrichment` off or `caas_perms.rds` absent)._\n\n", sep = "")
cat("**Observed statistics summary** (no permulation null to compare against):\n\n")
print(knitr::kable(
  dplyr::bind_rows(
    .obs_gene_caas %>% mutate(metric = "gene CAAS_score"),
    .obs_gene_asr  %>% mutate(metric = "gene asr_path_score (q90 over positions)")
  ) %>% dplyr::select(metric, direction, obs_median),
  digits = 4, caption = "Observed gene-level medians by direction"
))
```

```{r gene-level-header, eval=has_perms, results='asis'}
cat(r"(### Gene Level {.tabset}

> [!NOTE]
> **Performance Optimization:** To ensure memory efficiency during report compilation, the gene-level null distributions (spanning up to 20 million points per direction) are randomly downsampled to a representative subset of 100,000 points per direction. This downsampling maintains high statistical precision while keeping rendering fast and lightweight.

)")
```

```{r caas-score-header, eval=has_perms, results='asis'}
if (has_caas_null) cat("#### CAAS score\n\n")
```

```{r perm-gene-caas, eval=has_caas_null, fig.height=5, fig.width=10}
knitr::kable(.gene_null_table("caas_score", .obs_gene_caas), digits = 4,
  caption = sprintf("Gene-level null CAAS_score by direction (q90 over positions; %d genes × %d cycles). obs_median = observed gene CAAS_score.",
                    perm_n_genes, perm_n_cycles))
print(.gene_null_plot("caas_score", .obs_gene_caas_dist, "CAAS_score", "Gene-level null vs observed CAAS_score"))
```

```{r caas-null-missing, eval=has_perms && !has_caas_null, results='asis'}
cat("#### CAAS score\n\n")
cat("_CAAS score permulation null unavailable for this run (`caas_corStat_byrank` missing from `caas_perms.rds`)._\n\n")
```

```{r asr-score-header, eval=has_perms, results='asis'}
cat("#### asr_path_score\n\n")
```

```{r perm-gene-asr, eval=has_perms, fig.height=5, fig.width=10}
knitr::kable(.gene_null_table("asr_path_score", .obs_gene_asr), digits = 4,
  caption = "Gene-level null asr_path_score by direction (q90 over positions). obs_median = observed q90(asr_score).")
print(.gene_null_plot("asr_path_score", .obs_gene_asr_dist, "asr_path_score (q90)", "Gene-level null vs observed asr_path_score"))
```

### Position Level {.tabset}

Per CAAP grouping scheme (`US`, `GS1`–`GS4`): observed overlaid on the permulation null (H0).

```{r perm-pos-na, eval=!has_scheme_overlay, results='asis'}
cat("_Position-level null overlay unavailable (needs the lean perm sample + filtered_discovery)._\n")
```

```{r perm-pos-setup, eval=has_scheme_overlay}
src_pal <- c(null = "#BAB0AC", observed = "#E15759")
.pos_density <- function(value_col, xlab, title) {
  set.seed(1)
  nd <- scheme_overlay %>% filter(source == "null") %>% group_by(caap_group) %>%
        slice_sample(n = 20000) %>% ungroup()
  od <- scheme_overlay %>% filter(source == "observed")
  d  <- bind_rows(nd, od) %>% filter(is.finite(.data[[value_col]]))
  ggplot(d, aes(.data[[value_col]], fill = source, colour = source)) +
    geom_density(alpha = .45, adjust = 1.1) +
    facet_wrap(~ caap_group, scales = "free_y", ncol = 3) +
    scale_fill_manual(values = src_pal) + scale_colour_manual(values = src_pal) +
    labs(title = title, subtitle = "observed (red) overlaid on permulation null (grey)",
         x = xlab, y = "density") + theme_minimal(base_size = 12) + theme(legend.position = "top")
}
.pos_table <- function(value_col) {
  scheme_overlay %>% group_by(caap_group, source) %>%
    summarise(n = dplyr::n(), mean = mean(.data[[value_col]], na.rm = TRUE),
              median = median(.data[[value_col]], na.rm = TRUE), .groups = "drop") %>%
    arrange(caap_group, source)
}
```

#### CAAS score

```{r perm-pos-caas, eval=has_scheme_overlay, fig.height=6, fig.width=10}
knitr::kable(.pos_table("caas_row"), digits = 4,
  caption = "Per-scheme CAAS contribution (1−pval)×asr×0.2: permulation null vs observed.")
print(.pos_density("caas_row", "per-scheme CAAS contribution", "CAAS contribution by functional group: observed vs null"))
```

#### asr_path_score

```{r perm-pos-asr, eval=has_scheme_overlay, fig.height=6, fig.width=10}
knitr::kable(.pos_table("asr_path_score"), digits = 4,
  caption = "Per-scheme asr_path_score: permulation null vs observed.")
print(.pos_density("asr_path_score", "asr_path_score", "asr_path_score by functional group: observed vs null"))
```

#### Recovery p-value

```{r perm-pos-pval, eval=has_scheme_overlay, fig.height=6, fig.width=10}
if (!is.null(perm_pos_pval)) {
  knitr::kable(
    perm_pos_pval %>% group_by(caap_group) %>%
      summarise(n = dplyr::n(), median = median(recovery_pval, na.rm = TRUE),
                mean = mean(recovery_pval, na.rm = TRUE),
                `% < 0.05` = round(100 * mean(recovery_pval < 0.05, na.rm = TRUE), 1), .groups = "drop"),
    digits = 4,
    caption = sprintf("Null recovery p-value = (k+1)/(N+1) per scheme (floor 1/(N+1), N=%d).", perm_pos_pval$n_cycles[1]))
  print(ggplot(perm_pos_pval, aes(recovery_pval)) +
    geom_histogram(bins = 30, fill = "#4E79A7", colour = "white") +
    facet_wrap(~ caap_group, scales = "free_y", ncol = 3) +
    labs(title = "Null recovery p-value distribution by functional group",
         x = "recovery_pvalue = (k+1)/(N+1)", y = "count") + theme_minimal(base_size = 12))
}
```

---

## Position-Level Scores {.tabset}

### Significance gate & p-values (directional)

```{r pos-obs-pval-gate}
# Observed hypergeometric p-values, FDR p-values and significance-gate pass-rates by
# direction (these are observed quantities — the permulation does not permute them).
.dir_subsets <- list(
  global = pos_df,
  top    = pos_df %>% filter(change_side %in% c("top", "both")),
  bottom = pos_df %>% filter(change_side %in% c("bottom", "both"))
)
obs_tbl <- purrr::imap_dfr(.dir_subsets, function(d, dir) {
  tibble::tibble(
    direction           = dir,
    positions           = nrow(d),
    `median pvalue`     = if ("pvalue" %in% names(d)) median(d$pvalue, na.rm = TRUE) else NA_real_,
    `mean pvalue`       = if ("pvalue" %in% names(d)) mean(d$pvalue, na.rm = TRUE) else NA_real_,
    `median FDR pvalue` = if ("pvalue_hyp_fdr" %in% names(d)) median(d$pvalue_hyp_fdr, na.rm = TRUE) else NA_real_,
    `% gate_sig`        = if ("gate_sig" %in% names(d)) round(100 * mean(d$gate_sig, na.rm = TRUE), 1) else NA_real_,
    `% gate_fdr`        = if ("gate_fdr" %in% names(d)) round(100 * mean(d$gate_fdr, na.rm = TRUE), 1) else NA_real_,
    `median CAAS score` = if ("CAAS_score" %in% names(d)) median(d$CAAS_score, na.rm = TRUE) else NA_real_
  )
})
knitr::kable(obs_tbl, digits = 4,
  caption = "Observed hypergeometric p-values, FDR p-values and significance-gate pass-rates by direction.")
```

### CAAS Score Distribution

```{r caas-score-dist, fig.height=9}
# Global thresholds (used throughout the report for CAAS filters)
thr_10 <- quantile(pos_df$CAAS_score, 1 - top_pct,  na.rm = TRUE)
thr_5  <- quantile(pos_df$CAAS_score, 1 - top5_pct, na.rm = TRUE)
thr_1  <- quantile(pos_df$CAAS_score, 1 - top1_pct, na.rm = TRUE)

# Direction-specific thresholds (used for is_Xpct flags)
.thr_top_10 <- quantile(pos_top$CAAS_score, 1 - top_pct,  na.rm = TRUE)
.thr_top_5  <- quantile(pos_top$CAAS_score, 1 - top5_pct, na.rm = TRUE)
.thr_top_1  <- quantile(pos_top$CAAS_score, 1 - top1_pct, na.rm = TRUE)
.thr_bot_10 <- quantile(pos_bottom$CAAS_score, 1 - top_pct,  na.rm = TRUE)
.thr_bot_5  <- quantile(pos_bottom$CAAS_score, 1 - top5_pct, na.rm = TRUE)
.thr_bot_1  <- quantile(pos_bottom$CAAS_score, 1 - top1_pct, na.rm = TRUE)

p_combined <- ggplot(pos_df, aes(x = CAAS_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50,
                 fill = .tableau10[1], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[2]) +
  geom_vline(xintercept = thr_10, linetype = "dashed",
             colour = .tableau10[3], linewidth = 0.7) +
  geom_vline(xintercept = thr_5,  linetype = "longdash",
             colour = .tableau10[5], linewidth = 0.7) +
  geom_vline(xintercept = thr_1,  linetype = "dotted",
             colour = .tableau10[4], linewidth = 0.7) +
  annotate("text", x = thr_10, y = Inf, label = sprintf("Top %g%%", top_pct * 100),
           vjust = 2,    hjust = -0.1, colour = .tableau10[3], size = 3.5) +
  annotate("text", x = thr_5,  y = Inf, label = sprintf("Top %g%%", top5_pct * 100),
           vjust = 2.75, hjust = -0.1, colour = .tableau10[5], size = 3.5) +
  annotate("text", x = thr_1,  y = Inf, label = sprintf("Top %g%%", top1_pct * 100),
           vjust = 3.5,  hjust = -0.1, colour = .tableau10[4], size = 3.5) +
  labs(title = "CAAS Score Distribution — Combined",
       subtitle = sprintf("n = %s positions | Trait: %s",
                          format(nrow(pos_df), big.mark = ","), params$traitname),
       x = "CAAS Score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))

print(p_combined)
```

### Score axis

These are the two axes whose product is `CAAS_score` (`phen_score` × `asr_score`),
and the biochemical weight sum showing how many schemes flagged each position.
(`asr_score` is itself a five-factor product — core × independence × diversity ×
agreement × conservation — but those sub-factors are not shown here.)

```{r score-axes-dist, fig.height=9}
.axis_plot <- function(col, title, subtitle, fill, line) {
  if (!col %in% names(pos_df)) return(NULL)
  ggplot(pos_df, aes(x = .data[[col]])) +
    geom_histogram(aes(y = after_stat(density)), bins = 30, fill = fill, alpha = 0.7) +
    geom_density(linewidth = 0.8, colour = line) +
    labs(title = title, subtitle = subtitle, x = col, y = "Density") +
    theme(plot.title = element_text(face = "bold"))
}
.axis_plots <- Filter(Negate(is.null), list(
  .axis_plot("asr_score", "ASR Path Score (position evolution)",
             "Unified five-factor ASR path score (core × independence × diversity × agreement × conservation)",
             .tableau10[3], .tableau10[4]),
  .axis_plot("phen_score", "Phenotype Evolution (1 - pvalue_boot)",
             "Permulation evidence — the phenotype-partition axis of CAAS_score",
             .tableau10[1], .tableau10[2])
))
if (length(.axis_plots)) gridExtra::grid.arrange(grobs = .axis_plots, ncol = 1)

p_bwsum <- ggplot(pos_df, aes(x = factor(biochem_weight_sum))) +
  geom_bar(fill = .tableau10[4], alpha = 0.8) +
  labs(title = "Biochemical Weight Sum Distribution",
       subtitle = "Sum of scheme weights across detected schemes per position (max = 1.0). Reflects breadth of biochemical evidence.",
       x = "biochem_weight_sum", y = "Count") +
  theme(plot.title = element_text(face = "bold"))

# Scheme combination plot: which schemes co-occur most often?
scheme_combo_cols <- grep("^scheme_weight_", names(pos_df), value = TRUE)
if (length(scheme_combo_cols) > 0) {
  scheme_combo_df <- pos_df %>%
    select(all_of(scheme_combo_cols)) %>%
    mutate(across(everything(), ~as.integer(.x > 0))) %>%
    unite("combo", everything(), sep = "/", remove = FALSE) %>%
    count(combo, sort = TRUE) %>%
    slice_head(n = 15)
  p_scheme_combo <- ggplot(scheme_combo_df, aes(x = reorder(combo, n), y = n)) +
    geom_col(fill = .tableau10[6], alpha = 0.85) +
    coord_flip() +
    labs(title = "Most common scheme combinations",
         subtitle = "Top 15 scheme flag patterns (1 = scheme detected, 0 = absent)",
         x = "Scheme combination", y = "Count") +
    theme(plot.title = element_text(face = "bold"))
  gridExtra::grid.arrange(p_bwsum, p_scheme_combo, ncol = 2)
} else {
  print(p_bwsum)
}
```

### Stress Test

`r if (!has_scoring_stress) "> **Note:** scoring stress-test outputs are not available for this run."`

| Variant | What it represents |
|---------|---------------------|
| `CAAS_current` | The production `CAAS_score` used throughout this report (`phen_score` x `asr_score`, scaled by the biochemical scheme weight). |
| `CAAS_unweighted` | `phen_score` x `asr_score` without the biochemical multiplier -- isolates the effect of biochem scaling. |
| `CAAS_no_significance` | `asr_decile` x `biochem_weight_sum` / 5 -- drops the significance (p-value) axis, keeping ASR + biochem only. |
| `CAAS_no_asr` | `significance_score` x `biochem_weight_sum` / 5 -- drops the ASR axis, keeping significance + biochem only. |
| `PCA_all_components` | First principal component of the raw inputs (`variability_score`, `pattern_score`, `asr_decile`) -- a data-driven, unsupervised weighting of the same axes. |
| `Factor1_all_components` | Single-factor exploratory factor analysis over the same raw inputs -- separates shared variance (the factor) from each input's unique variance. |

A robust run shows consistent Spearman rank correlation with `CAAS_current` across all variants.

```{r stress-summary-table, eval=has_scoring_stress}
DT::datatable(
  stress_summary_df %>%
    arrange(desc(spearman_to_current), mean_abs_rank_shift) %>%
    mutate(across(where(is.numeric), ~round(., 4))),
  rownames = FALSE,
  options = list(pageLength = 20, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "Assumption sensitivity summary for alternative position-level score variants"
  )
)
```

### Top Positions

```{r top-positions-table}
# Two-way score axes first (phen_score × asr_score), then the gate tiers, then
# legacy diagnostics. any_of() keeps this robust to absent columns.
display_cols <- c("Gene", "Position", "CAAS_score",
                  "phen_score", "asr_score", "core",
                  "gate_sig", "gate_fdr",
                  "pvalue", "pvalue_boot", "pvalue_hyp_fdr",
                  "biochem_weight_sum", "change_side",
                  "pre_score",
                  "variability_score", "pattern_score",
                  "is_top_10pct", "is_top_5pct", "is_top_1pct",
                  "is_bot_10pct", "is_bot_5pct", "is_bot_1pct")

DT::datatable(
  pos_df %>%
    arrange(desc(CAAS_score)) %>%
    select(any_of(display_cols)) %>%
    mutate(across(where(is.numeric), ~round(., 4))),
  filter    = "top",
  rownames  = FALSE,
  options   = list(pageLength = 25, scrollX = TRUE),
  caption   = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Position-level CAAS scores | Trait: %s", params$traitname)
  )
)
```

### CAAS Score Manhattan (Position Level)

`r if (!has_genomic) "> **Note:** Genomic coordinates not available — position-level Manhattan skipped."`

```{r pos-caas-manhattan-setup, eval=has_genomic}
# Join all scored positions with genomic coordinates via their gene.
# Each position inherits its gene's genomic midpoint as the x-coordinate.
# We build the chr offsets independently here (rather than relying on the
# genomic-setup chunk, which runs later in the document).
pos_genomic <- pos_df %>%
  inner_join(genomic_raw, by = "Gene") %>%
  mutate(
    chr_label = sub("^chr", "", chr),
    chr_num   = suppressWarnings(as.integer(chr_label)),
    chr_order = dplyr::case_when(
      !is.na(chr_num)              ~ chr_num,
      chr_label == "X"             ~ 23L,
      chr_label == "Y"             ~ 24L,
      chr_label %in% c("MT","M")   ~ 25L,
      TRUE                         ~ 26L
    )
  ) %>%
  arrange(chr_order, gene_mid)

pos_chr_offsets <- pos_genomic %>%
  group_by(chr_order, chr_label) %>%
  summarise(chr_max = max(gene_end, na.rm = TRUE), .groups = "drop") %>%
  arrange(chr_order) %>%
  mutate(offset = cumsum(lag(chr_max, default = 0L)) + (chr_order - 1L) * 5e6L)

pos_genomic <- pos_genomic %>%
  left_join(pos_chr_offsets %>% select(chr_order, offset), by = "chr_order") %>%
  mutate(
    manhattan_x = gene_mid + offset,
    chr_even    = chr_order %% 2 == 0
  )

pos_chr_label_positions <- pos_chr_offsets %>%
  left_join(
    pos_genomic %>% group_by(chr_order) %>%
      summarise(mid_x = mean(manhattan_x, na.rm = TRUE), .groups = "drop"),
    by = "chr_order"
  )
```

```{r pos-caas-manhattan-plot, eval=has_genomic, fig.height=5, fig.width=14}
pos_genomic <- pos_genomic %>%
  mutate(caas_log = -log10(1 - pmin(CAAS_score, 0.9999)))
pos_thr_log <- -log10(1 - pmin(thr_10, 0.9999))

ggplot(pos_genomic, aes(x = manhattan_x, y = caas_log, colour = chr_even)) +
  geom_point(alpha = 0.45, size = 0.6, show.legend = FALSE) +
  geom_hline(yintercept = pos_thr_log, linetype = "dashed",
             colour = "grey40", linewidth = 0.6) +
  scale_colour_manual(
    values = c("FALSE" = .tableau10[1], "TRUE" = .tableau10[2]),
    guide = "none"
  ) +
  scale_x_continuous(
    breaks = pos_chr_label_positions$mid_x,
    labels = pos_chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  labs(
    title    = "Position-Level CAAS Score — Manhattan View",
    subtitle = sprintf("%d positions with genomic coordinates | dashed = top %g%% threshold",
                       nrow(pos_genomic), params$top_pct * 100),
    x = "Chromosome", y = "-log10(1 - CAAS Score)"
  ) +
  theme(
    plot.title         = element_text(face = "bold"),
    axis.text.x        = element_text(size = 7),
    panel.grid.major.x = element_blank()
  )
```

---

## Direction Score Profiles {.tabset}

Each tab shows distribution plots for TOP-direction and BOTTOM-direction positions
side by side, allowing direct comparison of how each scoring component behaves
across the two phenotypic directions. These tabs mirror the global Position-Level
Scores section, faceted by direction.

### CAAS Score

```{r dir-caas-dist, fig.height=9}
make_dir_dist <- function(df, dir_label, thr10, thr5, thr1, col) {
  ggplot(df, aes(x = CAAS_score)) +
    geom_histogram(aes(y = after_stat(density)), bins = 50,
                   fill = col, alpha = 0.7) +
    geom_density(linewidth = 0.8, colour = col) +
    geom_vline(xintercept = thr10, linetype = "dashed",  colour = "grey30", linewidth = 0.7) +
    geom_vline(xintercept = thr5,  linetype = "longdash", colour = "grey30", linewidth = 0.7) +
    geom_vline(xintercept = thr1,  linetype = "dotted",  colour = "grey30", linewidth = 0.7) +
    annotate("text", x = thr10, y = Inf,
             label = sprintf("Top %g%%\n(%.4f)", top_pct * 100,  thr10),
             vjust = 2,    hjust = -0.05, colour = "grey20", size = 3) +
    annotate("text", x = thr5,  y = Inf,
             label = sprintf("Top %g%%\n(%.4f)", top5_pct * 100, thr5),
             vjust = 3.8,  hjust = -0.05, colour = "grey20", size = 3) +
    annotate("text", x = thr1,  y = Inf,
             label = sprintf("Top %g%%\n(%.4f)", top1_pct * 100, thr1),
             vjust = 5.6,  hjust = -0.05, colour = "grey20", size = 3) +
    labs(title = sprintf("CAAS Score Distribution — %s positions", dir_label),
         subtitle = sprintf("n = %s | thresholds computed within direction",
                            format(nrow(df), big.mark = ",")),
         x = "CAAS Score", y = "Density") +
    theme(plot.title = element_text(face = "bold"))
}

gridExtra::grid.arrange(
  make_dir_dist(pos_top,    "TOP",    .thr_top_10, .thr_top_5, .thr_top_1, .tableau10[1]),
  make_dir_dist(pos_bottom, "BOTTOM", .thr_bot_10, .thr_bot_5, .thr_bot_1, .tableau10[3]),
  ncol = 1
)
```

### P-Value Distributions

```{r dir-pval-dist, fig.height=9}
make_dir_pval_panel <- function(df, measure_col, is_raw, dir_label, col) {
  vals   <- df[[measure_col]]
  breaks <- quantile(vals, probs = seq(0, 1, 0.1), na.rm = TRUE)
  plot_x <- if (is_raw) -log10(pmax(vals, 1e-300)) else vals
  xlab   <- if (is_raw) sprintf("-log10(%s)", measure_col) else measure_col
  title  <- sprintf("%s — %s", ifelse(is_raw, toupper(measure_col), measure_col), dir_label)
  df2 <- data.frame(x = plot_x)
  ggplot(df2, aes(x = x)) +
    geom_histogram(aes(y = after_stat(density)), bins = 50, fill = col, alpha = 0.65) +
    geom_density(linewidth = 0.8, colour = col) +
    geom_vline(
      xintercept = if (is_raw) -log10(pmax(breaks[2:10], 1e-300)) else breaks[2:10],
      linetype = "dotted", colour = "grey50", linewidth = 0.3
    ) +
    labs(title = title,
         subtitle = sprintf("n = %s positions | dotted = decile boundaries",
                            format(nrow(df), big.mark = ",")),
         x = xlab, y = "Density") +
    theme(plot.title = element_text(face = "bold", size = 10))
}

p_hyp_top <- make_dir_pval_panel(pos_top,    hyper_measure_col, hyper_measure_is_raw, "TOP",    .tableau10[1])
p_hyp_bot <- make_dir_pval_panel(pos_bottom, hyper_measure_col, hyper_measure_is_raw, "BOTTOM", .tableau10[3])

# Permulation (pvalue_boot) distribution omitted here — it is the phen_score axis,
# already shown in the Score axis tab below.
gridExtra::grid.arrange(
  p_hyp_top, p_hyp_bot,
  ncol = 2,
  top  = grid::textGrob(
    "Hypergeometric p-value distribution by direction",
    gp = grid::gpar(fontface = "bold", fontsize = 12)
  )
)
```

### Significance Gate

```{r dir-gate-summary}
if (all(c("gate_all", "gate_sig", "gate_fdr") %in% names(pos_df))) {
  gate_dir_tbl <- bind_rows(
    pos_top    %>% summarise(Direction = "TOP",
                              `All positions`        = sum(gate_all, na.rm = TRUE),
                              `Significant (p<0.05)` = sum(gate_sig, na.rm = TRUE),
                              `FDR-significant`      = sum(gate_fdr, na.rm = TRUE)),
    pos_bottom %>% summarise(Direction = "BOTTOM",
                              `All positions`        = sum(gate_all, na.rm = TRUE),
                              `Significant (p<0.05)` = sum(gate_sig, na.rm = TRUE),
                              `FDR-significant`      = sum(gate_fdr, na.rm = TRUE))
  ) %>%
    mutate(
      `% sig`     = round(100 * `Significant (p<0.05)` / pmax(`All positions`, 1), 1),
      `% FDR-sig` = round(100 * `FDR-significant`      / pmax(`All positions`, 1), 1)
    )
  knitr::kable(gate_dir_tbl,
               caption = "Hypergeometric significance gate by direction — positions per tier")
} else {
  cat("_Significance gate columns not present in this run._\n")
}
```

### Score axis

```{r dir-score-axis-dist, fig.height=9}
make_dir_component_plot <- function(df, col, dir_label, colour) {
  if (!col %in% names(df)) return(NULL)
  ggplot(df, aes(x = .data[[col]])) +
    geom_histogram(aes(y = after_stat(density)), bins = 40, fill = colour, alpha = 0.65) +
    geom_density(linewidth = 0.8, colour = colour) +
    labs(title    = sprintf("%s — %s", col, dir_label),
         subtitle = sprintf("n = %s positions", format(nrow(df), big.mark = ",")),
         x = col, y = "Density") +
    theme(plot.title = element_text(face = "bold", size = 10))
}

dir_axis_plots <- Filter(Negate(is.null), list(
  make_dir_component_plot(pos_top,    "phen_score", "TOP",    .tableau10[1]),
  make_dir_component_plot(pos_bottom, "phen_score", "BOTTOM", .tableau10[3]),
  make_dir_component_plot(pos_top,    "asr_score",  "TOP",    .tableau10[5]),
  make_dir_component_plot(pos_bottom, "asr_score",  "BOTTOM", .tableau10[7])
))
if (length(dir_axis_plots)) {
  gridExtra::grid.arrange(
    grobs = dir_axis_plots, ncol = 2,
    top = grid::textGrob(
      "Score axis distributions by direction: phenotype evolution (row 1) and ASR path score (row 2)",
      gp = grid::gpar(fontface = "bold", fontsize = 12)
    )
  )
}
```

### Top Positions by Direction

```{r dir-top-positions-table}
has_caas_col <- "caas" %in% names(pos_df)

dir_display_cols <- c("Gene", "Position", "CAAS_score",
                      "phen_score", "asr_score", "core",
                      "gate_sig", "gate_fdr",
                      if (has_caas_col) "caas",
                      "biochem_weight_sum",
                      "variability_score", "pattern_score",
                      "pre_score",
                      "pvalue", "pvalue_boot", "pvalue_hyp_fdr",
                      "change_side",
                      "is_top_10pct", "is_top_5pct", "is_top_1pct",
                      "is_bot_10pct", "is_bot_5pct", "is_bot_1pct")

dir_pos_tbl <- bind_rows(
  pos_top    %>% mutate(.direction = "TOP"),
  pos_bottom %>% mutate(.direction = "BOTTOM")
) %>%
  arrange(.direction, desc(CAAS_score)) %>%
  select(Direction = .direction, any_of(dir_display_cols)) %>%
  mutate(across(where(is.numeric), ~round(., 4)))

DT::datatable(
  dir_pos_tbl,
  filter   = "top",
  rownames = FALSE,
  options  = list(pageLength = 25, scrollX = TRUE),
  caption  = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Top-scoring positions by direction | Trait: %s", params$traitname)
  )
)
```

```{r fade-pos-char-header, results='asis'}
if (has_fade_pos) cat(r"(
---

## FADE Position Characterisation {.tabset}

> **FADE** (FUBAR Approach to Directional Evolution) tests whether foreground branches show
> elevated substitution bias toward a particular amino acid at each site.
> Top-direction FADE characterises positions where the phenotypic group shows accelerated
> amino acid evolution; bottom-direction shows the opposite.
> CAAS positions (BF ≥ 100) overlap reveals sites that are both convergent and under directional selection.

)")
```

```{r fade-pos-setup, eval=has_fade_pos}
# Load FADE site tables for each direction separately.
# FADE_report.Rmd emits site = seq_along(...) making site numbers 1-based, while
# CAAS/CT uses 0-based positions (alimport.py range(0, length), caas_id.py as-is).
# Subtract 1 here to align coordinate systems before joining.
.load_fade_sites <- function(path) {
  if (!.file_ok(path)) return(NULL)
  df <- read_tsv(path, show_col_types = FALSE)
  names(df) <- tolower(names(df))
  
  if (!"gene" %in% names(df)) names(df)[1] <- "gene"
  if ("site" %in% names(df)) {
    df$position <- as.integer(df$site) - 1L
  } else if ("position" %in% names(df)) {
    df$position <- as.integer(df$position)
  } else {
    df$position <- as.integer(df[[2]]) - 1L
  }
  
  bf_col <- intersect(c("max_site_bf", "max_bf", "bayes_factor", "bf"), names(df))[1]
  if (is.na(bf_col)) {
    df$max_site_bf <- 0
  } else {
    df$max_site_bf <- as.numeric(df[[bf_col]])
  }
  
  aa_col <- intersect(c("top_target_aa", "target_aa", "aa"), names(df))[1]
  if (!is.na(aa_col)) {
    df$top_target_aa <- as.character(df[[aa_col]])
  } else {
    df$top_target_aa <- NA_character_
  }
  
  bias_col <- intersect(c("fade_biased", "biased"), names(df))[1]
  if (!is.na(bias_col)) {
    df$fade_biased <- as.logical(df[[bias_col]])
  } else {
    df$fade_biased <- df$max_site_bf >= 100
  }
  
  df %>%
    select(Gene = gene, Position = position, max_site_bf, top_target_aa, fade_biased)
}
fade_top_df <- .load_fade_sites(params$fade_site_top_file)
fade_bot_df <- .load_fade_sites(params$fade_site_bot_file)

# Join FADE BFs to directional CAAS position subsets
.join_fade <- function(pos_subset, fade_df) {
  if (is.null(fade_df) || nrow(pos_subset) == 0) return(NULL)
  pos_subset %>%
    left_join(
      fade_df %>% select(Gene, Position, max_site_bf, top_target_aa, fade_biased),
      by = c("Gene", "Position")
    ) %>%
    mutate(
      fade_biased = replace_na(fade_biased, FALSE),
      fade_sig    = !is.na(max_site_bf) & max_site_bf >= 100,
      caas_decile = dplyr::ntile(CAAS_score, 10),
      norm_caas   = rank(CAAS_score,   ties.method = "average", na.last = "keep") /
                    sum(!is.na(CAAS_score)),
      norm_log_bf = {
        lbf <- log10(pmax(max_site_bf, 0.01))
        rank(lbf, ties.method = "average", na.last = "keep") / sum(!is.na(lbf))
      }
    )
}
pos_fade_top <- .join_fade(pos_top,    fade_top_df)
pos_fade_bot <- .join_fade(pos_bottom, fade_bot_df)

has_fade_top <- !is.null(pos_fade_top) && any(pos_fade_top$fade_sig, na.rm = TRUE)
has_fade_bot <- !is.null(pos_fade_bot) && any(pos_fade_bot$fade_sig, na.rm = TRUE)
has_fade_any <- has_fade_top || has_fade_bot

# Helper: draw 2-set Venn diagram using base ggplot2 (no extra packages).
# Circle labels are placed outside and above each circle to avoid overlapping
# with the count annotations inside.
.draw_venn2 <- function(n_only_A, n_only_B, n_AB, n_universe,
                         label_A, label_B, title_str) {
  theta <- seq(0, 2 * pi, length.out = 300)
  r <- 1.7; dx <- 1.05
  ciA <- data.frame(x = -dx + r * cos(theta), y = r * sin(theta))
  ciB <- data.frame(x =  dx + r * cos(theta), y = r * sin(theta))
  ggplot() +
    geom_polygon(data = ciA, aes(x, y), fill = .tableau10[1], alpha = 0.22, colour = NA) +
    geom_polygon(data = ciB, aes(x, y), fill = .tableau10[3], alpha = 0.22, colour = NA) +
    geom_path(data = ciA, aes(x, y), colour = .tableau10[1], linewidth = 1.4) +
    geom_path(data = ciB, aes(x, y), colour = .tableau10[3], linewidth = 1.4) +
    annotate("text", x = -2.6, y = 0,
             label = paste0(label_A, " only\n", scales::comma(n_only_A)),
             size = 4.2, colour = .tableau10[1], fontface = "bold") +
    annotate("text", x = 0, y = 0,
             label = paste0("Both\n", scales::comma(n_AB)),
             size = 4.2, colour = "grey20", fontface = "bold") +
    annotate("text", x = 2.6, y = 0,
             label = paste0(label_B, " only\n", scales::comma(n_only_B)),
             size = 4.2, colour = .tableau10[3], fontface = "bold") +
    # Circle titles: pushed to x = -3.5 and +3.5 (outside the overlap zone)
    annotate("text", x = -3.5, y = 2.8, label = label_A, size = 4.5,
             colour = .tableau10[1], fontface = "bold", hjust = 0.5) +
    annotate("text", x =  3.5, y = 2.8, label = label_B, size = 4.5,
             colour = .tableau10[3], fontface = "bold", hjust = 0.5) +
    annotate("text", x = 0, y = -2.6,
             label = sprintf("Universe: %s positions", scales::comma(n_universe)),
             size = 3.5, colour = "grey50") +
    coord_fixed(xlim = c(-5.5, 5.5), ylim = c(-3.2, 3.5)) +
    labs(title = title_str) +
    theme_void() +
    theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 12))
}

# Helper: compute Venn counts (CAAS dir positions vs full FADE sig set)
.fade_venn_counts <- function(pos_dir, fade_df) {
  if (is.null(pos_dir) || is.null(fade_df)) return(NULL)
  caas_keys     <- pos_dir  %>% select(Gene, Position) %>% distinct()
  fade_sig_keys <- fade_df  %>%
    filter(!is.na(max_site_bf), max_site_bf >= 100) %>%
    select(Gene, Position) %>% distinct()
  n_A   <- nrow(caas_keys)
  n_B   <- nrow(fade_sig_keys)
  n_AB  <- nrow(inner_join(caas_keys, fade_sig_keys, by = c("Gene", "Position")))
  list(n_only_A = n_A - n_AB, n_only_B = n_B - n_AB, n_AB = n_AB,
       n_A = n_A, n_B = n_B, n_universe = n_A + max(n_B - n_AB, 0))
}

vc_top <- .fade_venn_counts(pos_top,    fade_top_df)
vc_bot <- .fade_venn_counts(pos_bottom, fade_bot_df)
```

```{r fade-venn-subsec, results='asis'}
if (has_fade_pos) {
  cat("### Venn Diagrams\n\n")
  if (!has_fade_any)
    cat("> **Note:** No CAAS positions found with FADE BF ≥ 100 in this run.",
        "Venn diagrams shown with zero overlap — all CAAS positions fall below the BF threshold.\n\n")
}
```

```{r fade-venn-plots, eval=has_fade_pos, fig.height=5, fig.width=14}
venn_plots <- list()
if (!is.null(vc_top)) {
  venn_plots[["top"]] <- .draw_venn2(
    vc_top$n_only_A, vc_top$n_only_B, vc_top$n_AB, vc_top$n_universe,
    "CAAS (top)", "FADE BF\u2265100 (top)",
    "TOP direction: CAAS vs FADE-significant positions"
  )
}
if (!is.null(vc_bot)) {
  venn_plots[["bot"]] <- .draw_venn2(
    vc_bot$n_only_A, vc_bot$n_only_B, vc_bot$n_AB, vc_bot$n_universe,
    "CAAS (bottom)", "FADE BF\u2265100 (bottom)",
    "BOTTOM direction: CAAS vs FADE-significant positions"
  )
}
if (length(venn_plots) == 2) {
  gridExtra::grid.arrange(grobs = venn_plots, ncol = 2)
} else if (length(venn_plots) == 1) {
  print(venn_plots[[1]])
}
```

```{r fade-venn-overlap-table, eval=has_fade_pos}
# Table of positions in the intersection (CAAS + FADE BF>=100), by direction
.venn_overlap_positions <- function(pos_dir, fade_df, dir_label, thr10, thr5, thr1) {
  if (is.null(pos_dir) || is.null(fade_df)) return(NULL)
  caas_sel <- if ("caas" %in% names(pos_dir))
    c("Gene", "Position", "CAAS_score", "caas", "change_side") else
    c("Gene", "Position", "CAAS_score", "change_side")
  inner_join(
    pos_dir  %>% select(any_of(caas_sel)),
    fade_df  %>% filter(!is.na(max_site_bf), max_site_bf >= 100) %>%
                 select(Gene, Position, max_site_bf, top_target_aa),
    by = c("Gene", "Position")
  ) %>%
    mutate(
      Direction = dir_label,
      IsTop10   = CAAS_score >= thr10,
      IsTop5    = CAAS_score >= thr5,
      IsTop1    = CAAS_score >= thr1
    ) %>%
    arrange(desc(max_site_bf))
}

overlap_pos_tbl <- bind_rows(
  .venn_overlap_positions(pos_top,    fade_top_df, "TOP",    .thr_top_10, .thr_top_5, .thr_top_1),
  .venn_overlap_positions(pos_bottom, fade_bot_df, "BOTTOM", .thr_bot_10, .thr_bot_5, .thr_bot_1)
)

if (nrow(overlap_pos_tbl) > 0) {
  DT::datatable(
    overlap_pos_tbl %>% mutate(across(where(is.numeric), ~round(., 4))),
    filter   = "top",
    rownames = FALSE,
    options  = list(pageLength = 25, scrollX = TRUE),
    caption  = htmltools::tags$caption(
      style = "caption-side: top; text-align: left; font-weight: bold;",
      "Positions in the CAAS \u2229 FADE (BF \u2265 100) intersection, by direction"
    )
  )
} else {
  cat("> No positions found in the CAAS \u2229 FADE intersection.")
}
```

```{r fade-overlap-subsec, results='asis'}
if (has_fade_pos) cat("### Overlap Table\n\n")
```

```{r fade-overlap-table, eval=has_fade_pos}
.fade_overlap_row <- function(pos_fade_dir, dir_label, pct_val, pct_label) {
  if (is.null(pos_fade_dir)) return(NULL)
  thr  <- quantile(pos_fade_dir$CAAS_score, 1 - pct_val, na.rm = TRUE)
  n_caas <- sum(pos_fade_dir$CAAS_score >= thr, na.rm = TRUE)
  n_sig  <- sum(pos_fade_dir$fade_sig, na.rm = TRUE)
  n_both <- sum(pos_fade_dir$CAAS_score >= thr & pos_fade_dir$fade_sig, na.rm = TRUE)
  tibble(
    Direction           = dir_label,
    `CAAS threshold`    = pct_label,
    `N CAAS positions`  = n_caas,
    `N FADE BF >= 100`  = n_sig,
    `Overlap`           = n_both,
    `% CAAS in FADE`    = sprintf("%.1f%%", 100 * n_both / max(n_caas, 1)),
    `% FADE in CAAS`    = sprintf("%.1f%%", 100 * n_both / max(n_sig,  1))
  )
}

pct_list <- list(
  list(val = params$top_pct,  lbl = sprintf("Top %g%%", params$top_pct  * 100)),
  list(val = 0.05, lbl = sprintf("Top %g%%", 0.05 * 100)),
  list(val = 0.01, lbl = sprintf("Top %g%%", 0.01 * 100))
)

fade_overlap_tbl <- bind_rows(
  purrr::map_dfr(pct_list, ~.fade_overlap_row(pos_fade_top, "TOP",    .x$val, .x$lbl)),
  purrr::map_dfr(pct_list, ~.fade_overlap_row(pos_fade_bot, "BOTTOM", .x$val, .x$lbl))
)

knitr::kable(
  fade_overlap_tbl,
  caption = "Overlap between top-CAAS positions and FADE-significant positions (BF \u2265 100), by direction and CAAS threshold"
)
```

```{r fade-corr-subsec, results='asis'}
if (has_fade_pos) cat("### CAAS vs FADE Correlation\n\n")
```

```{r fade-caas-corr, eval=has_fade_any, fig.height=5}
# Combine top and bottom joined datasets; plot CAAS_score vs log10(BF)
# with rank-normalised axes so Pearson r is not dominated by BF outliers.
.fade_corr_df <- bind_rows(
  if (!is.null(pos_fade_top)) pos_fade_top %>% mutate(fade_direction = "TOP")    else NULL,
  if (!is.null(pos_fade_bot)) pos_fade_bot %>% mutate(fade_direction = "BOTTOM") else NULL
) %>%
  filter(!is.na(max_site_bf)) %>%
  mutate(log10_bf = log10(pmax(max_site_bf, 0.01)))

.corr_label <- .fade_corr_df %>%
  group_by(fade_direction) %>%
  summarise(
    r_pearson  = cor(norm_caas, norm_log_bf, use = "complete.obs"),
    r_spearman = cor(CAAS_score, log10_bf, method = "spearman", use = "complete.obs"),
    .groups = "drop"
  ) %>%
  mutate(label = sprintf("%s: r=%.3f (rank-norm Pearson) / \u03c1=%.3f (Spearman)",
                         fade_direction, r_pearson, r_spearman))

ggplot(.fade_corr_df, aes(x = CAAS_score, y = log10_bf, colour = fade_direction)) +
  geom_point(alpha = 0.35, size = 0.9) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8, alpha = 0.12) +
  geom_hline(yintercept = 2, linetype = "dashed", colour = "grey50", linewidth = 0.5) +
  scale_colour_manual(values = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[3]),
                      name = "Direction") +
  annotate("text", x = -Inf, y = 2.05, hjust = -0.05, vjust = 0,
           label = "BF = 100", size = 3, colour = "grey40") +
  labs(
    title    = "CAAS Score vs FADE max Bayes Factor",
    subtitle = paste(.corr_label$label, collapse = "\n"),
    x = "CAAS Score", y = "log10(max site BF)"
  ) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(size = 9, colour = "grey30"))
```

---

## Gene-Level Characterisation {.tabset}

Directional gene sets (TOP and BOTTOM) are characterised against independent
evolutionary signals. All correlations use rank-normalised variables.

### Gene CAAS Score

```{r gene-caas-dir-dist, fig.height=5}
# Directional thresholds
g_top_thr_10 <- quantile(gene_df$gene_caas_top, 1 - params$gene_top_pct, na.rm = TRUE)
g_top_thr_5  <- quantile(gene_df$gene_caas_top, 1 - 0.05,                na.rm = TRUE)
g_top_thr_1  <- quantile(gene_df$gene_caas_top, 1 - 0.01,                na.rm = TRUE)
g_bot_thr_10 <- quantile(gene_df$gene_caas_bot, 1 - params$gene_top_pct, na.rm = TRUE)
g_bot_thr_5  <- quantile(gene_df$gene_caas_bot, 1 - 0.05,                na.rm = TRUE)
g_bot_thr_1  <- quantile(gene_df$gene_caas_bot, 1 - 0.01,                na.rm = TRUE)
g_thr_10 <- quantile(gene_df$gene_caas_score, 1 - params$gene_top_pct, na.rm = TRUE)

long_dir_scores <- bind_rows(
  gene_df %>% filter(!is.na(gene_caas_top)) %>%
    transmute(Gene, score = gene_caas_top, direction = "TOP"),
  gene_df %>% filter(!is.na(gene_caas_bot)) %>%
    transmute(Gene, score = gene_caas_bot, direction = "BOTTOM")
)

ggplot(long_dir_scores, aes(x = score, fill = direction, colour = direction)) +
  geom_density(alpha = 0.45, linewidth = 0.7) +
  geom_vline(xintercept = g_top_thr_10, linetype = "dashed",
             colour = .tableau10[1], linewidth = 0.6) +
  geom_vline(xintercept = g_bot_thr_10, linetype = "dashed",
             colour = .tableau10[2], linewidth = 0.6) +
  scale_fill_manual(values   = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[2])) +
  scale_colour_manual(values = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[2])) +
  labs(title = "Directional Gene CAAS Score Distribution",
       subtitle = sprintf("90th pctile of directional positions | dashed = top %g%% threshold per direction",
                          params$gene_top_pct * 100),
       x = "Gene CAAS Score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### RERConverge

`r if (!has_rer) "> **Note:** RERConverge results not available for this run."`

> Each panel below compares a CAAS gene gate (left circle) against an
> RERConverge-significant set (right circle). The **shared region** is the
> count of RERConverge-significant genes that are *also* members of the CAAS
> gate -- i.e. how much of the alternative (RER) evidence is corroborated by
> CAAS. **OR** is a Fisher exact test (equivalent to a hypergeometric test on
> the 2x2 table) against the full scored-gene universe (n = `r nrow(gene_df)`
> genes); OR > 1 means the overlap is larger than expected by chance.

```{r gene-rer-venn, eval=has_rer, fig.height=14, fig.width=12}
rer_accel_sig <- gene_df %>%
  filter(!is.na(rer_min_pval), rer_min_pval <= 0.05,
         !is.na(rer_rho), rer_rho > 0) %>% pull(Gene)
rer_decel_sig <- gene_df %>%
  filter(!is.na(rer_min_pval), rer_min_pval <= 0.05,
         !is.na(rer_rho), rer_rho < 0) %>% pull(Gene)

# 3 rows (gate) x 4 columns (TOP/BOT x Accel/Decel)
rer_venn_plots <- list()
for (g in .caas_gates) {
  rer_venn_plots <- c(rer_venn_plots, list(
    .draw_venn(.venn_counts(g$top, rer_accel_sig, sprintf("CAAS TOP\n%s", g$lbl), "RER Accel.", gene_universe_n),
               .tableau10[1], .tableau10[3], sprintf("%s | TOP vs Accel.", g$lbl)),
    .draw_venn(.venn_counts(g$top, rer_decel_sig, sprintf("CAAS TOP\n%s", g$lbl), "RER Decel.", gene_universe_n),
               .tableau10[1], .tableau10[4], sprintf("%s | TOP vs Decel.", g$lbl)),
    .draw_venn(.venn_counts(g$bot, rer_accel_sig, sprintf("CAAS BOT\n%s", g$lbl), "RER Accel.", gene_universe_n),
               .tableau10[2], .tableau10[3], sprintf("%s | BOT vs Accel.", g$lbl)),
    .draw_venn(.venn_counts(g$bot, rer_decel_sig, sprintf("CAAS BOT\n%s", g$lbl), "RER Decel.", gene_universe_n),
               .tableau10[2], .tableau10[4], sprintf("%s | BOT vs Decel.", g$lbl))
  ))
}
gridExtra::grid.arrange(
  grobs = rer_venn_plots, ncol = 4,
  top = grid::textGrob(
    "RERConverge Gene Overlap \u2014 CAAS gates (rows) \u00d7 direction\u00d7RER (cols) | RER p \u2264 0.05",
    gp = grid::gpar(fontface = "bold", fontsize = 12)))
```

```{r gene-rer-table, eval=has_rer}
rer_tbl <- gene_df %>%
  filter(!is.na(rer_min_pval)) %>%
  mutate(
    rer_direction = case_when(rer_rho > 0 ~ "accelerating",
                               rer_rho < 0 ~ "decelerating",
                               TRUE        ~ "neutral"),
    in_top_global = Gene %in% gene_global_top, in_top_sig = Gene %in% gene_sig_top,
    in_top_fdr    = Gene %in% gene_fdr_top,
    in_bot_global = Gene %in% gene_global_bot, in_bot_sig = Gene %in% gene_sig_bot,
    in_bot_fdr    = Gene %in% gene_fdr_bot
  ) %>%
  arrange(rer_min_pval) %>%
  select(Gene, rer_min_pval, rer_rho, rer_direction,
         gene_caas_score, gene_caas_top, gene_caas_bot,
         in_top_global, in_top_sig, in_top_fdr,
         in_bot_global, in_bot_sig, in_bot_fdr) %>%
  mutate(across(where(is.numeric), ~round(., 4)))

DT::datatable(
  rer_tbl, filter="top", rownames=FALSE,
  options = list(pageLength=20, scrollX=TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "RERConverge results ranked by p-value — direction (rho sign) and CAAS gene list membership"
  )
)
```

### FADE Gene

`r if (!has_fade_pos) "> **Note:** FADE site data not available — gene-level FADE analysis skipped."`

```{r gene-fade-setup, eval=has_fade_pos}
if (!is.null(fade_top_df) && nrow(fade_top_df) > 0) {
  gene_fade_top_agg <- fade_top_df %>%
    group_by(Gene) %>%
    summarise(fade_top_sig    = any(max_site_bf >= 100, na.rm=TRUE),
              fade_top_max_bf = max(max_site_bf, na.rm=TRUE), .groups="drop")
} else {
  gene_fade_top_agg <- tibble(Gene=character(), fade_top_sig=logical(), fade_top_max_bf=numeric())
}
if (!is.null(fade_bot_df) && nrow(fade_bot_df) > 0) {
  gene_fade_bot_agg <- fade_bot_df %>%
    group_by(Gene) %>%
    summarise(fade_bot_sig    = any(max_site_bf >= 100, na.rm=TRUE),
              fade_bot_max_bf = max(max_site_bf, na.rm=TRUE), .groups="drop")
} else {
  gene_fade_bot_agg <- tibble(Gene=character(), fade_bot_sig=logical(), fade_bot_max_bf=numeric())
}
gene_fade_df <- gene_df %>%
  select(Gene, gene_caas_score, gene_caas_top, gene_caas_bot, n_positions) %>%
  left_join(gene_fade_top_agg, by="Gene") %>%
  left_join(gene_fade_bot_agg, by="Gene") %>%
  mutate(fade_top_sig = replace_na(fade_top_sig, FALSE),
         fade_bot_sig = replace_na(fade_bot_sig, FALSE))
has_gene_fade_any <- any(gene_fade_df$fade_top_sig) || any(gene_fade_df$fade_bot_sig)
```

> Shared region = FADE-significant genes that are also in the CAAS gate.
> OR = Fisher exact test vs. the full scored-gene universe (n = `r nrow(gene_df)` genes).

```{r gene-fade-venn, eval=has_fade_pos && exists("has_gene_fade_any") && has_gene_fade_any, fig.height=11, fig.width=12}
fade_top_genes <- gene_fade_df %>% filter(fade_top_sig) %>% pull(Gene)
fade_bot_genes <- gene_fade_df %>% filter(fade_bot_sig) %>% pull(Gene)
fade_venn_plots <- list()
for (g in .caas_gates) {
  fade_venn_plots <- c(fade_venn_plots, list(
    .draw_venn(.venn_counts(g$top, fade_top_genes, sprintf("CAAS TOP\n%s", g$lbl), "FADE TOP sig", gene_universe_n),
               .tableau10[1], .tableau10[5], sprintf("%s | TOP vs FADE TOP (BF\u2265100)", g$lbl)),
    .draw_venn(.venn_counts(g$bot, fade_bot_genes, sprintf("CAAS BOT\n%s", g$lbl), "FADE BOT sig", gene_universe_n),
               .tableau10[2], .tableau10[6], sprintf("%s | BOT vs FADE BOT (BF\u2265100)", g$lbl))
  ))
}
gridExtra::grid.arrange(
  grobs = fade_venn_plots, ncol = 2,
  top = grid::textGrob("FADE Gene Overlap \u2014 CAAS gates (rows) \u00d7 direction (cols) | BF \u2265 100",
                        gp = grid::gpar(fontface = "bold", fontsize = 12)))
```

```{r gene-fade-correlation, eval=has_fade_pos && exists("has_gene_fade_any") && has_gene_fade_any, fig.height=5, fig.width=12}
fade_cor_top <- gene_fade_df %>%
  filter(!is.na(gene_caas_top), !is.na(fade_top_max_bf), is.finite(fade_top_max_bf)) %>%
  mutate(x_norm=rank_norm(gene_caas_top), y_norm=rank_norm(fade_top_max_bf))
fade_cor_bot <- gene_fade_df %>%
  filter(!is.na(gene_caas_bot), !is.na(fade_bot_max_bf), is.finite(fade_bot_max_bf)) %>%
  mutate(x_norm=rank_norm(gene_caas_bot), y_norm=rank_norm(fade_bot_max_bf))

.fade_scatter <- function(df, col, title, xlab, ylab) {
  ggplot(df, aes(x=x_norm, y=y_norm)) +
    geom_point(alpha=0.5, size=1.2, colour=col) +
    geom_smooth(method="lm", se=TRUE, colour="grey30", linewidth=0.7) +
    ggrepel::geom_text_repel(
      data = df %>% filter(y_norm > quantile(y_norm, 0.95, na.rm=TRUE)),
      aes(label=Gene), size=2.5, max.overlaps=10, show.legend=FALSE
    ) +
    labs(title=title,
         subtitle=sprintf("r = %.3f", safe_cor(df$x_norm, df$y_norm)),
         x=xlab, y=ylab) +
    theme(plot.title = element_text(face="bold", size=11))
}
gridExtra::grid.arrange(
  .fade_scatter(fade_cor_top, .tableau10[1],
                "TOP: CAAS vs FADE BF (rank-norm)",
                "gene_caas_top (rank-norm)", "FADE top max BF (rank-norm)"),
  .fade_scatter(fade_cor_bot, .tableau10[2],
                "BOTTOM: CAAS vs FADE BF (rank-norm)",
                "gene_caas_bot (rank-norm)", "FADE bot max BF (rank-norm)"),
  ncol=2,
  top=grid::textGrob("Gene CAAS Score vs FADE BF — rank-normalised",
                      gp=grid::gpar(fontface="bold", fontsize=13)))
```

```{r gene-fade-table, eval=has_fade_pos}
fade_tbl <- gene_fade_df %>%
  mutate(
    in_top_global = Gene %in% gene_global_top, in_top_sig = Gene %in% gene_sig_top,
    in_top_fdr    = Gene %in% gene_fdr_top,
    in_bot_global = Gene %in% gene_global_bot, in_bot_sig = Gene %in% gene_sig_bot,
    in_bot_fdr    = Gene %in% gene_fdr_bot
  ) %>%
  arrange(desc(pmax(replace_na(fade_top_max_bf, 0), replace_na(fade_bot_max_bf, 0)))) %>%
  select(Gene,
         fade_top_max_bf, fade_top_sig,
         fade_bot_max_bf, fade_bot_sig,
         gene_caas_score, gene_caas_top, gene_caas_bot,
         in_top_global, in_top_sig, in_top_fdr,
         in_bot_global, in_bot_sig, in_bot_fdr) %>%
  mutate(across(where(is.numeric), ~round(., 4)))

DT::datatable(
  fade_tbl, filter = "top", rownames = FALSE,
  options = list(pageLength = 20, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "FADE gene-level results ranked by max BF — direction and CAAS gene list membership"
  )
)
```

### Accumulation

`r if (!has_accum_score) "> **Note:** Accumulation scores not available for this run."`

> Shared region = Accumulation-significant genes that are also in the CAAS gate.
> OR = Fisher exact test vs. the full scored-gene universe (n = `r nrow(gene_df)` genes).

```{r gene-accum-venn, eval=has_accum_score, fig.height=11, fig.width=12}
accum_sig_genes <- gene_df %>%
  filter(!is.na(accum_fisher_p), accum_significant %in% TRUE) %>% pull(Gene)

accum_venn_plots <- list()
accum_overlap_exports <- list()
for (g in .caas_gates) {
  for (dir_lbl in c("TOP", "BOTTOM")) {
    caas_genes <- if (dir_lbl == "TOP") g$top else g$bot
    col_a      <- if (dir_lbl == "TOP") .tableau10[1] else .tableau10[2]
    dir_slug   <- tolower(dir_lbl)
    gate_slug  <- gsub("[^a-z0-9]", "_", tolower(g$lbl))
    overlap_genes <- intersect(caas_genes, accum_sig_genes)
    accum_overlap_exports[[sprintf("gene_caas_%s_%s_x_accum.txt", dir_slug, gate_slug)]] <- overlap_genes
    vc <- .venn_counts(caas_genes, accum_sig_genes,
                       sprintf("CAAS %s\n%s", dir_lbl, g$lbl), "Accum. sig", gene_universe_n)
    accum_venn_plots[[length(accum_venn_plots) + 1]] <-
      .draw_venn(vc, col_a, .tableau10[5], sprintf("%s | %s vs Accum.", g$lbl, dir_lbl))
  }
}
# Note: Overlap text file exports (gene_caas_*_x_accum.txt) removed as requested.

gridExtra::grid.arrange(
  grobs = accum_venn_plots, ncol = 2,
  top = grid::textGrob(
    "Accumulation Gene Overlap — CAAS gates (rows) × direction (cols) | accum_significant",
    gp = grid::gpar(fontface = "bold", fontsize = 12)))
```

```{r gene-accum-table, eval=has_accum_score}
accum_pval_cols <- grep("^accum_pval_", names(gene_df), value = TRUE)
accum_tbl <- gene_df %>%
  filter(!is.na(accum_fisher_p)) %>%
  mutate(
    in_top_global = Gene %in% gene_global_top, in_top_sig = Gene %in% gene_sig_top,
    in_top_fdr    = Gene %in% gene_fdr_top,
    in_bot_global = Gene %in% gene_global_bot, in_bot_sig = Gene %in% gene_sig_bot,
    in_bot_fdr    = Gene %in% gene_fdr_bot
  ) %>%
  arrange(accum_fisher_p) %>%
  select(Gene, accum_fisher_p, any_of("accum_fdr"), accum_significant, any_of(accum_pval_cols),
         gene_caas_score, gene_caas_top, gene_caas_bot,
         in_top_global, in_top_sig, in_top_fdr,
         in_bot_global, in_bot_sig, in_bot_fdr) %>%
  mutate(across(where(is.numeric), ~round(., 4)))

DT::datatable(
  accum_tbl, filter = "top", rownames = FALSE,
  options = list(pageLength = 20, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "Accumulation genes ranked by Fisher combined p-value (accum_fisher_p) — per-scheme p-values and CAAS membership"
  )
)
```

### PrimateAI-3D

`r if (!has_vep_primateai) "> **Note:** PrimateAI-3D scores are not available for this run."`

> **Note:** PrimateAI-3D scores require an access request to Illumina —
> see illumina.com/science/genomics-research/articles/primateai-3d.html

```{r gene-pai-restore, eval=has_vep_primateai}
# Position Characterisation (the PAI3D/COSMIC report section) moved to
# posenrich — but this gene-level PAI3D rollup stays here, so it needs its
# own minimal reconstruction of pos_vep_pai (position -> pai_char_score),
# same logic as the moved section's vep-pai-setup/vep-pai-join chunks.
.pai_score_col <- intersect(
  c("score_PAI3D", "score", "primateai_score", "pathogenicity_score", "PrimateAI_score"),
  names(vep_pai_df)
)[1]
if (is.na(.pai_score_col)) {
  warning("PrimateAI: no recognised score column found — skipping.")
  has_vep_primateai <- FALSE
} else {
  pos_score_meta <- pos_df %>%
    select(Gene, Position, CAAS_score, biochem_weight_sum, change_side, gate_sig, gate_fdr)
  bg_genes <- unique(gene_df$Gene)
  shared_genes_pai <- intersect(bg_genes, unique(vep_pai_df$Gene))
  pos_pai_shared <- pos_score_meta %>% filter(Gene %in% shared_genes_pai)

  pos_vep_pai <- vep_pai_df %>%
    filter(caap_group == "US", Gene %in% shared_genes_pai) %>%
    mutate(pai_score = as.numeric(.data[[.pai_score_col]])) %>%
    filter(is.finite(pai_score)) %>%
    group_by(Gene, Position) %>%
    summarise(pai_char_score = max(pai_score, na.rm = TRUE),
              n_pai_schemes  = 1L, .groups = "drop") %>%
    inner_join(pos_pai_shared, by = c("Gene", "Position")) %>%
    distinct(Gene, Position, .keep_all = TRUE)
  has_vep_primateai <- nrow(pos_vep_pai) > 0
}
```

```{r gene-pai-directional, eval=has_vep_primateai, fig.height=5, fig.width=12}
gene_vep_pai <- pos_vep_pai %>%
  group_by(Gene) %>%
  summarise(n_pai_positions = n(),
            mean_pai_char   = mean(pai_char_score, na.rm=TRUE),
            sum_pai_char    = sum(pai_char_score,  na.rm=TRUE),
            .groups = "drop") %>%
  inner_join(gene_df %>% select(Gene, gene_caas_score, gene_caas_top,
                                gene_caas_bot, n_positions, gene_direction),
             by = "Gene")

pai_top_df <- gene_vep_pai %>%
  filter(gene_direction %in% c("top","both"), !is.na(gene_caas_top)) %>%
  mutate(x_norm=rank_norm(gene_caas_top), y_norm=rank_norm(mean_pai_char))
pai_bot_df <- gene_vep_pai %>%
  filter(gene_direction %in% c("bottom","both"), !is.na(gene_caas_bot)) %>%
  mutate(x_norm=rank_norm(gene_caas_bot), y_norm=rank_norm(mean_pai_char))

gridExtra::grid.arrange(
  ggplot(pai_top_df, aes(x=x_norm, y=y_norm, size=n_positions)) +
    geom_point(alpha=0.65, colour=.tableau10[1]) +
    geom_smooth(method="lm", se=TRUE, colour="grey30", linewidth=0.7) +
    ggrepel::geom_text_repel(
      data=pai_top_df %>% slice_max(y_norm, n=15, with_ties=FALSE),
      aes(label=Gene), size=2.5, max.overlaps=15, show.legend=FALSE) +
    scale_size_continuous(name="N pos", range=c(1,6)) +
    labs(title="TOP genes: CAAS vs PrimateAI-3D",
         subtitle=sprintf("r = %.3f | rank-normalised",
                          safe_cor(pai_top_df$x_norm, pai_top_df$y_norm)),
         x="gene_caas_top (rank-norm)", y="Mean PAI3D char score (rank-norm)") +
    theme(plot.title=element_text(face="bold", size=11)),
  ggplot(pai_bot_df, aes(x=x_norm, y=y_norm, size=n_positions)) +
    geom_point(alpha=0.65, colour=.tableau10[2]) +
    geom_smooth(method="lm", se=TRUE, colour="grey30", linewidth=0.7) +
    ggrepel::geom_text_repel(
      data=pai_bot_df %>% slice_max(y_norm, n=15, with_ties=FALSE),
      aes(label=Gene), size=2.5, max.overlaps=15, show.legend=FALSE) +
    scale_size_continuous(name="N pos", range=c(1,6)) +
    labs(title="BOTTOM genes: CAAS vs PrimateAI-3D",
         subtitle=sprintf("r = %.3f | rank-normalised",
                          safe_cor(pai_bot_df$x_norm, pai_bot_df$y_norm)),
         x="gene_caas_bot (rank-norm)", y="Mean PAI3D char score (rank-norm)") +
    theme(plot.title=element_text(face="bold", size=11)),
  ncol=2,
  top=grid::textGrob("Gene CAAS vs PrimateAI-3D — Direction-aware (rank-normalised)",
                      gp=grid::gpar(fontface="bold", fontsize=13)))
```

```{r gene-pai-top-bar, eval=has_vep_primateai, fig.height=6}
top_pai_genes <- gene_vep_pai %>%
  filter(n_pai_positions >= 2, mean_pai_char > 0) %>%
  slice_max(mean_pai_char, n=30, with_ties=FALSE) %>%
  arrange(mean_pai_char) %>%
  mutate(Gene = factor(Gene, levels=Gene))

ggplot(top_pai_genes, aes(x=mean_pai_char, y=Gene, fill=gene_caas_score)) +
  geom_col() +
  scale_fill_viridis_c(name="Gene\nCAAS score", option="plasma") +
  scale_x_continuous(expand=expansion(mult=c(0, 0.05))) +
  labs(title = "Top 30 Genes by Mean PrimateAI-3D Characterisation Score",
       subtitle = paste0("Genes with \u2265 2 positions with PAI3D coverage | colour = gene_caas_score"),
       x = "Mean pai_char_score", y = NULL) +
  theme(plot.title=element_text(face="bold"), axis.text.y=element_text(size=8))
```

```{r gene-pai-table, eval=has_vep_primateai}
pai_tbl <- gene_vep_pai %>%
  mutate(
    in_top_global = Gene %in% gene_global_top, in_top_sig = Gene %in% gene_sig_top,
    in_top_fdr    = Gene %in% gene_fdr_top,
    in_bot_global = Gene %in% gene_global_bot, in_bot_sig = Gene %in% gene_sig_bot,
    in_bot_fdr    = Gene %in% gene_fdr_bot
  ) %>%
  arrange(desc(mean_pai_char)) %>%
  select(Gene, n_pai_positions, mean_pai_char, sum_pai_char,
         gene_caas_score, gene_caas_top, gene_caas_bot, gene_direction,
         in_top_global, in_top_sig, in_top_fdr,
         in_bot_global, in_bot_sig, in_bot_fdr) %>%
  mutate(across(where(is.numeric), ~round(., 4)))

DT::datatable(
  pai_tbl, filter = "top", rownames = FALSE,
  options = list(pageLength = 20, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "PrimateAI-3D gene scores ranked by mean PAI characterisation score"
  )
)
```

### Top Genes

```{r top-genes-table}
top_genes_df <- gene_df %>%
  arrange(desc(gene_caas_score))

# ── Direction assignment ──────────────────────────────────────────────────────
if (has_rer) {
  top_genes_df <- top_genes_df %>%
    mutate(rer_direction = case_when(
      !is.na(rer_rho) & rer_rho > 0 ~ "accelerating",
      !is.na(rer_rho) & rer_rho < 0 ~ "decelerating",
      TRUE ~ NA_character_
    ))
}

# ── FADE: max BF and BF>=100 significance flag per direction ─────────────────
if (has_fade_pos && exists("gene_fade_df")) {
  top_genes_df <- top_genes_df %>%
    left_join(
      gene_fade_df %>% select(Gene, fade_top_max_bf, fade_bot_max_bf,
                               fade_top_sig, fade_bot_sig),
      by = "Gene"
    )
}

# ── Accumulation score ────────────────────────────────────────────────────────
# accum_fisher_p already on top_genes_df; no extra join needed.

# ── PrimateAI-3D mean characterisation score per gene ────────────────────────
if (has_vep_primateai && exists("gene_vep_pai")) {
  top_genes_df <- top_genes_df %>%
    left_join(gene_vep_pai %>% select(Gene, mean_pai_char, n_pai_positions),
              by = "Gene")
}

# ── CAAS-threshold indicators (10%, 50% and 100%) per direction ──────────────
g_top_thr_50 <- quantile(gene_df$gene_caas_top, 0.50, na.rm = TRUE)
g_bot_thr_50 <- quantile(gene_df$gene_caas_bot, 0.50, na.rm = TRUE)

top_genes_df <- top_genes_df %>%
  mutate(
    in_top_10pct = replace_na(gene_caas_top >= g_top_thr_10, FALSE),
    in_top_50pct = replace_na(gene_caas_top >= g_top_thr_50, FALSE),
    in_bot_10pct = replace_na(gene_caas_bot >= g_bot_thr_10, FALSE),
    in_bot_50pct = replace_na(gene_caas_bot >= g_bot_thr_50, FALSE)
  )

# ── Build display column list ─────────────────────────────────────────────────
gene_display_cols <- c("Gene", "n_positions",
                       "gene_caas_score", "gene_caas_top", "gene_caas_bot",
                       "in_top_10pct", "in_top_50pct",
                       "in_bot_10pct", "in_bot_50pct")
if (has_accum_score)
  gene_display_cols <- c(gene_display_cols, "accum_fisher_p")
if (has_rer)
  gene_display_cols <- c(gene_display_cols, "rer_min_pval", "rer_direction")
if (has_fade_pos && "fade_top_max_bf" %in% names(top_genes_df))
  gene_display_cols <- c(gene_display_cols,
                         "fade_top_max_bf", "fade_top_sig",
                         "fade_bot_max_bf", "fade_bot_sig")
if (has_vep_primateai && "mean_pai_char" %in% names(top_genes_df))
  gene_display_cols <- c(gene_display_cols, "n_pai_positions", "mean_pai_char")

DT::datatable(
  top_genes_df %>%
    select(any_of(gene_display_cols)) %>%
    mutate(across(where(is.numeric) & !matches("n_positions|n_pai"), ~round(., 4))),
  filter = "top", rownames = FALSE,
  options = list(pageLength = 25, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Gene-level scores | Trait: %s | sorted by gene_caas_score", params$traitname)
  )
)
```

---

## Genomic Accumulation {.tabset}

`r if (!has_genomic) "> **Note:** Genomic coordinates (gene_ensembl_file) are not available for this run — genomic accumulation skipped."`

```{r genomic-setup, eval=has_genomic}
# Join gene_df with genomic coordinates, then annotate with VEP
gene_genomic <- gene_df %>%
  select(Gene, gene_caas_score, gene_caas_top, gene_caas_bot, n_positions) %>%
  inner_join(genomic_raw, by = "Gene") %>%
  # Standardise chromosome labels (remove "chr" prefix if present)
  mutate(
    chr_label = sub("^chr", "", chr),
    # Order chromosomes numerically then X, Y, MT
    chr_num   = suppressWarnings(as.integer(chr_label)),
    chr_order = dplyr::case_when(
      !is.na(chr_num)       ~ chr_num,
      chr_label == "X"      ~ 23L,
      chr_label == "Y"      ~ 24L,
      chr_label %in% c("MT","M") ~ 25L,
      TRUE                  ~ 26L
    )
  ) %>%
  arrange(chr_order, gene_mid)

# Optionally add mean PAI characterisation score
if (has_vep_primateai && exists("gene_vep_pai")) {
  gene_genomic <- gene_genomic %>%
    left_join(gene_vep_pai %>% select(Gene, mean_pai_char),
              by = "Gene")
} else {
  gene_genomic$mean_pai_char <- NA_real_
}

# Build a cumulative x-axis offset per chromosome for a stitched Manhattan view
chr_offsets <- gene_genomic %>%
  group_by(chr_order, chr_label) %>%
  summarise(chr_max = max(gene_end, na.rm = TRUE), .groups = "drop") %>%
  arrange(chr_order) %>%
  mutate(offset = cumsum(lag(chr_max, default = 0L)) + (chr_order - 1L) * 5e6L)

gene_genomic <- gene_genomic %>%
  left_join(chr_offsets %>% select(chr_order, offset, chr_label), by = c("chr_order", "chr_label")) %>%
  mutate(manhattan_x = gene_mid + offset)

chr_label_positions <- chr_offsets %>%
  left_join(gene_genomic %>% group_by(chr_order) %>%
              summarise(mid_x = mean(manhattan_x, na.rm = TRUE), .groups = "drop"),
            by = "chr_order")
```

```{r window-helper, eval=has_genomic}
# Helper: tile chromosomes into non-overlapping windows of win_size bp.
# Each gene/position is assigned to the window containing its genomic midpoint.
# Returns a dataframe with one row per window, summarising mean and max score.
make_window_accum <- function(df, score_col,
                               win_size = params$window_size_bp) {
  df %>%
    mutate(
      win_id    = floor(gene_mid / win_size),
      win_start = win_id * win_size,
      win_end   = win_start + win_size - 1L,
      win_mid   = win_start + win_size / 2L
    ) %>%
    group_by(chr_order, chr_label, win_id, win_start, win_end, win_mid) %>%
    summarise(
      n_genes    = dplyr::n(),
      mean_score = mean(.data[[score_col]], na.rm = TRUE),
      max_score  = max(.data[[score_col]],  na.rm = TRUE),
      genes      = paste(unique(Gene), collapse = ", "),
      .groups    = "drop"
    )
}

# Stitch window midpoints onto the cumulative chromosome x-axis (same offsets
# as gene_genomic so windows align with the per-gene manhattans above).
stitch_windows <- function(win_df) {
  win_df %>%
    left_join(chr_offsets %>% select(chr_order, offset), by = "chr_order") %>%
    mutate(win_x = win_mid + offset,
           chr_even = chr_order %% 2 == 0)
}
```

### CAAS Score Manhattan

```{r genomic-manhattan-caas, eval=has_genomic, fig.height=5, fig.width=14}
gene_genomic <- gene_genomic %>%
  mutate(
    chr_even      = chr_order %% 2 == 0,
    caas_log      = -log10(1 - pmin(gene_caas_score, 0.9999)),
    caas_top_log  = -log10(1 - pmin(replace_na(gene_caas_top, 0), 0.9999)),
    caas_bot_log  = -log10(1 - pmin(replace_na(gene_caas_bot, 0), 0.9999))
  )
caas_log_thr <- -log10(1 - pmin(g_thr_10, 0.9999))

ggplot(gene_genomic,
       aes(x = manhattan_x, y = caas_log,
           colour = chr_even, size = n_positions)) +
  geom_point(alpha = 0.65, show.legend = FALSE) +
  geom_hline(yintercept = caas_log_thr, linetype = "dashed",
             colour = "grey40", linewidth = 0.6) +
  ggrepel::geom_text_repel(
    data = gene_genomic %>% slice_max(caas_log, n = 20, with_ties = FALSE),
    aes(label = Gene), size = 2.2, max.overlaps = 15,
    show.legend = FALSE, colour = "black"
  ) +
  scale_colour_manual(values = c("FALSE"=.tableau10[1], "TRUE"=.tableau10[2]), guide="none") +
  scale_size_continuous(name = "N positions", range = c(0.5, 5)) +
  scale_x_continuous(
    breaks = chr_label_positions$mid_x,
    labels = chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  labs(
    title    = "Gene CAAS Score — Manhattan View",
    subtitle = sprintf("%d genes | y = -log10(1 - gene_caas_score) | dashed = top %g%% threshold",
                       nrow(gene_genomic), params$gene_top_pct * 100),
    x = "Chromosome", y = "-log10(1 - gene_caas_score)"
  ) +
  theme(plot.title=element_text(face="bold"),
        axis.text.x=element_text(size=14), axis.text.y=element_text(size=14),
        axis.title=element_text(size=14), legend.text=element_text(size=12),
        panel.grid.major.x=element_blank())
```

### PAI3D Signal Manhattan

`r if (!has_vep_primateai) "> **Note:** PrimateAI-3D scores not available — PAI3D Manhattan skipped."`

```{r genomic-manhattan-pai, eval=has_genomic && has_vep_primateai && exists("gene_vep_pai"), fig.height=5, fig.width=14}
gene_genomic_pai <- gene_genomic
if (!"mean_pai_char" %in% names(gene_genomic_pai)) {
  gene_genomic_pai <- gene_genomic_pai %>%
    left_join(gene_vep_pai %>% select(Gene, mean_pai_char), by = "Gene")
}

ggplot(gene_genomic_pai %>% filter(!is.na(mean_pai_char)),
       aes(x=manhattan_x, y=caas_log, size=mean_pai_char, colour=mean_pai_char)) +
  geom_point(alpha=0.70) +
  geom_hline(yintercept=caas_log_thr, linetype="dashed",
             colour="grey60", linewidth=0.5) +
  ggrepel::geom_text_repel(
    data = gene_genomic_pai %>% filter(!is.na(mean_pai_char)) %>%
      slice_max(mean_pai_char, n=20, with_ties=FALSE),
    aes(label=Gene), size=2.2, max.overlaps=12, show.legend=FALSE, colour="black"
  ) +
  scale_colour_viridis_c(name="Mean PAI3D\nchar score", option="inferno") +
  scale_size_continuous(name="Mean PAI3D\nchar score", range=c(0.5, 6)) +
  scale_x_continuous(breaks=chr_label_positions$mid_x, labels=chr_label_positions$chr_label,
                     expand=expansion(mult=0.01)) +
  guides(colour=guide_legend(), size=guide_legend()) +
  labs(title    = "Gene CAAS Score + PrimateAI-3D — Manhattan View",
      subtitle = "Bubble size and colour: mean PAI3D characterisation score | y = -log10(1 - gene_caas_score)",
       x = "Chromosome", y = "-log10(1 - gene_caas_score)") +
  theme(plot.title=element_text(face="bold"),
        axis.text.x=element_text(size=14), axis.text.y=element_text(size=14),
        axis.title=element_text(size=14), legend.text=element_text(size=12),
        panel.grid.major.x=element_blank())
```

### CAAS Window Accumulation (Positions)

`r if (!has_genomic) "> **Note:** Genomic coordinates not available — position window accumulation skipped."`

```{r pos-window-accum, eval=has_genomic, fig.height=5, fig.width=14}
# Tile position-level CAAS scores into genomic windows.
# pos_genomic is built in the pos-caas-manhattan-setup chunk above.
# We use the same pos_chr_offsets to align x-coordinates with the gene-level plots.
win_size_label <- if (params$window_size_bp >= 1e6) {
  sprintf("%g Mb", params$window_size_bp / 1e6)
} else {
  sprintf("%g kb", params$window_size_bp / 1e3)
}

pos_wins <- pos_genomic %>%
  rename(Gene = Gene) %>%
  make_window_accum("CAAS_score", win_size = params$window_size_bp) %>%
  left_join(pos_chr_offsets %>% select(chr_order, offset), by = "chr_order") %>%
  mutate(n_positions = n_genes,
         win_x       = win_mid + offset,
         chr_even    = chr_order %% 2 == 0,
         mean_log    = -log10(1 - pmin(mean_score, 0.9999)),
         max_log     = -log10(1 - pmin(max_score,  0.9999)))

# Label the top windows as genomic regions, not gene names.
# Ranked by n_positions (window density) rather than mean_log, to surface the
# regions with the most positions rather than the highest mean score.
top_pos_wins <- pos_wins %>%
  slice_max(n_positions, n = 5, with_ties = FALSE) %>%
  mutate(region_label = sprintf("chr%s:%0.1f-%0.1f Mb", chr_label, win_start / 1e6, win_end / 1e6))

ggplot(pos_wins, aes(x = win_x, y = mean_log, fill = chr_even)) +
  geom_col(width = params$window_size_bp * 0.9, alpha = 0.75) +
  geom_point(aes(y = max_log, colour = chr_even), size = 1.2, alpha = 0.85, show.legend = FALSE) +
  ggrepel::geom_text_repel(
    data = top_pos_wins,
    aes(y = max_log, label = sprintf("%s\n(%d positions)", region_label, n_positions)),
    size = 2.2, max.overlaps = 10, colour = "black",
    show.legend = FALSE
  ) +
  scale_fill_manual(
    values = c("FALSE" = .tableau10[1], "TRUE" = .tableau10[2]),
    guide = "none"
  ) +
  scale_colour_manual(
    values = c("FALSE" = .tableau10[3], "TRUE" = .tableau10[4]),
    guide = "none"
  ) +
  scale_x_continuous(
    breaks = pos_chr_label_positions$mid_x,
    labels = pos_chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  labs(
    title    = sprintf("Position-Level CAAS Score — %s Window Accumulation", win_size_label),
    subtitle = sprintf("Bars = mean | Points = max | %d windows across %d chromosomes",
                       nrow(pos_wins), length(unique(pos_wins$chr_order))),
    x = "Chromosome", y = "-log10(1 - mean CAAS score)"
  ) +
  theme(
    plot.title         = element_text(face = "bold"),
    axis.text.x        = element_text(size = 14),
    axis.text.y        = element_text(size = 14),
    axis.title         = element_text(size = 14),
    legend.text        = element_text(size = 12),
    panel.grid.major.x = element_blank()
  )
```

```{r pos-window-accum-table, eval=has_genomic && exists("pos_wins") && nrow(pos_wins) > 0}
DT::datatable(
  pos_wins %>%
    select(chr_label, win_start, win_end, n_positions, mean_score, max_score, mean_log) %>%
    mutate(
      win_start  = format(win_start, big.mark = ",", scientific = FALSE),
      win_end    = format(win_end,   big.mark = ",", scientific = FALSE),
      mean_score = round(mean_score, 4),
      max_score  = round(max_score,  4),
      mean_log   = round(mean_log,   4)
    ) %>%
    arrange(desc(n_positions)) %>%
    rename(chr = chr_label, start = win_start, end = win_end,
           n_pos = n_positions, mean_CAAS = mean_score, max_CAAS = max_score,
           `-log10(1-mean)` = mean_log),
  rownames = FALSE,
  options = list(pageLength = 15, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Top genomic windows by position-level CAAS accumulation (%s windows)", win_size_label)
  )
)
```

### CAAS Window Accumulation (Genes)

`r if (!has_genomic) "> **Note:** Genomic coordinates not available — gene-level CAAS window accumulation skipped."`

```{r gene-caas-window-accum, eval=has_genomic, fig.height=5, fig.width=14}
gene_caas_wins <- gene_genomic %>%
  make_window_accum("gene_caas_score", win_size = params$window_size_bp) %>%
  stitch_windows() %>%
  mutate(mean_log = -log10(1 - pmin(mean_score, 0.9999)),
         max_log  = -log10(1 - pmin(max_score,  0.9999)))

# Ranked by n_genes (window density) rather than mean_log, to surface the
# regions with the most genes rather than the highest mean score.
top_caas_wins <- gene_caas_wins %>%
  slice_max(n_genes, n = 5, with_ties = FALSE) %>%
  mutate(region_label = sprintf("chr%s:%0.1f-%0.1f Mb", chr_label, win_start / 1e6, win_end / 1e6))

ggplot(gene_caas_wins, aes(x = win_x, y = mean_log, fill = chr_even)) +
  geom_col(width = params$window_size_bp * 0.9, alpha = 0.75) +
  geom_point(aes(y = max_log, colour = chr_even), size = 1.5, alpha = 0.85, show.legend = FALSE) +
  ggrepel::geom_text_repel(
    data = top_caas_wins,
    aes(y = max_log, label = sprintf("%s\n(%d genes)", region_label, n_genes)),
    size = 2.2, max.overlaps = 10, colour = "black",
    show.legend = FALSE
  ) +
  scale_fill_manual(
    values = c("FALSE" = .tableau10[1], "TRUE" = .tableau10[2]),
    guide = "none"
  ) +
  scale_colour_manual(
    values = c("FALSE" = .tableau10[3], "TRUE" = .tableau10[4]),
    guide = "none"
  ) +
  scale_x_continuous(
    breaks = chr_label_positions$mid_x,
    labels = chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  labs(
    title    = sprintf("Gene CAAS Score — %s Window Accumulation", win_size_label),
    subtitle = sprintf("Bars = mean | Points = max | %d windows", nrow(gene_caas_wins)),
    x = "Chromosome", y = "-log10(1 - mean gene_caas_score)"
  ) +
  theme(
    plot.title         = element_text(face = "bold"),
    axis.text.x        = element_text(size = 14),
    axis.text.y        = element_text(size = 14),
    axis.title         = element_text(size = 14),
    legend.text        = element_text(size = 12),
    panel.grid.major.x = element_blank()
  )
```

```{r gene-caas-window-accum-table, eval=has_genomic && exists("gene_caas_wins") && nrow(gene_caas_wins) > 0}
DT::datatable(
  gene_caas_wins %>%
    select(chr_label, win_start, win_end, n_genes, mean_score, max_score, mean_log) %>%
    mutate(
      win_start  = format(win_start, big.mark = ",", scientific = FALSE),
      win_end    = format(win_end,   big.mark = ",", scientific = FALSE),
      mean_score = round(mean_score, 4),
      max_score  = round(max_score,  4),
      mean_log   = round(mean_log,   4)
    ) %>%
    arrange(desc(n_genes)) %>%
    rename(chr = chr_label, start = win_start, end = win_end,
           n_genes = n_genes, mean_CAAS = mean_score, max_CAAS = max_score,
           `-log10(1-mean)` = mean_log),
  rownames = FALSE,
  options = list(pageLength = 15, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Top genomic windows by gene CAAS score accumulation (%s windows)", win_size_label)
  )
)
```

```{r genomic-window-tsv-export, eval=has_genomic}
if (exists("pos_wins") && nrow(pos_wins) > 0) {
  readr::write_tsv(
    pos_wins %>% select(-chr_even, -mean_log, -max_log),
    file.path(params$output_dir, "genomic_window_accum_positions.tsv"))
}
if (exists("gene_caas_wins") && nrow(gene_caas_wins) > 0) {
  readr::write_tsv(
    gene_caas_wins %>% select(-chr_even, -mean_log, -max_log),
    file.path(params$output_dir, "genomic_window_accum_gene_caas.tsv"))
}
cat("Exported genomic window accumulation TSVs\n")
```

---

## Supplementary Exports

```{r supplementary-tsv-export}
# Build a wide supplementary table joining position_scores with FADE and PAI data.
# Exported as position_scores_annotated.tsv alongside the report outputs.

supp_df <- pos_df

# Join FADE data: use the direction-matched FADE table for each position
if (has_fade_pos) {
  fade_combined <- bind_rows(
    if (!is.null(fade_top_df)) fade_top_df %>% mutate(fade_direction = "TOP")    else NULL,
    if (!is.null(fade_bot_df)) fade_bot_df %>% mutate(fade_direction = "BOTTOM") else NULL
  )
  if (nrow(fade_combined) > 0) {
    # For positions in both directions (change_side == "both"), take TOP FADE by priority
    fade_for_join <- fade_combined %>%
      mutate(fade_direction_priority = if_else(fade_direction == "TOP", 1L, 2L)) %>%
      group_by(Gene, Position) %>%
      slice_min(fade_direction_priority, n = 1, with_ties = FALSE) %>%
      ungroup() %>%
      select(Gene, Position,
             fade_max_site_bf     = max_site_bf,
             fade_top_target_aa   = top_target_aa,
             fade_biased,
             fade_direction)
    supp_df <- supp_df %>% left_join(fade_for_join, by = c("Gene", "Position"))
  }
}

# Join PAI data
if (has_vep_primateai && exists("pos_vep_pai") && nrow(pos_vep_pai) > 0) {
  pai_for_join <- pos_vep_pai %>%
    select(Gene, Position, pai_char_score, n_pai_schemes)
  supp_df <- supp_df %>% left_join(pai_for_join, by = c("Gene", "Position"))
}

readr::write_tsv(supp_df,
  file.path(params$output_dir, "position_scores_annotated.tsv"))

cat(sprintf("Exported position_scores_annotated.tsv: %d rows, %d columns\n",
            nrow(supp_df), ncol(supp_df)))
cat(sprintf("Columns: %s\n", paste(names(supp_df), collapse = ", ")))
```

---

## Summary

```{r summary-stats}
n_components <- 5L  # variability, pattern, asr, convergence, parallel

n_gene_components <- 1L  # gene_caas always present
if (has_accum_score) n_gene_components <- n_gene_components + 1L
if (has_rer)         n_gene_components <- n_gene_components + 1L

summary_tbl <- tibble(
  Metric = c(
    "Positions scored", "Genes scored",
    "Position-level components", "Gene-level components",
    "CAAS Score (median)", "CAAS Score (top 10% threshold)",
    "Gene CAAS TOP (median)", "Gene CAAS TOP (top 10% threshold)",
    "FADE site-level analysis available",
    "Accumulation score available", "RERConverge available",
    "VEP PrimateAI-3D available",
    "Genomic coordinates available"
  ),
  Value = c(
    format(nrow(pos_df), big.mark = ","),
    format(nrow(gene_df), big.mark = ","),
    as.character(n_components),
    as.character(n_gene_components),
    sprintf("%.4f", median(pos_df$CAAS_score, na.rm = TRUE)),
    sprintf("%.4f", thr_10),
    sprintf("%.4f", median(gene_df$gene_caas_top, na.rm = TRUE)),
    sprintf("%.4f", g_top_thr_10),
    ifelse(has_fade_pos,     "Yes", "No"),
    ifelse(has_accum_score,  "Yes", "No"),
    ifelse(has_rer,          "Yes", "No"),
    ifelse(has_vep_primateai,"Yes", "No"),
    ifelse(has_genomic,      "Yes", "No")
  )
)

knitr::kable(summary_tbl, align = c("l", "r"))
```

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere — Scoring Module*
*Date of generation: `r Sys.Date()`*
