---
title: "CAAS Scoring Report"
subtitle: "`r if (exists('params') && !is.null(params$traitname)) paste0('Trait: ', params$traitname) else if (exists('params') && !is.null(params$trait)) paste0('Trait: ', params$trait) else if (exists('params') && !is.null(params$project_name)) paste0('Project: ', params$project_name) else ''`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
    mathjax: null                       # no math in this report
    md_extensions: "-tex_math_dollars"  # gene/pathway names can contain '$' - never parse as TeX math
params:
  position_scores_file: NULL
  gene_scores_file: NULL
  gene_corr_file: NULL
  traitname: "unknown_trait"
  output_dir: "."
  top_pct: 0.10
  gene_top_pct: 0.10
  stress_summary_file: NULL
  stress_corr_file: NULL
  stress_rank_file: NULL
  stress_overlap_file: NULL
  stress_variants_file: NULL
  fade_site_top_file: NULL
  fade_site_bot_file: NULL
  genomic_info_file: NULL
  caas_perms_file: NULL
  caas_pos_pval_file: NULL
  caas_pos_sample_file: NULL
  caas_pos_quantiles_file: NULL
  filtered_discovery_file: NULL
  window_size_bp: 1000000
  direction: "combined"
  background_file: NULL
  caas_pos_detail_dir: NULL
  caas_gene_cycle_scores_file: NULL
  seed: "1998"
---

<style type="text/css">
  body {
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .main-container {
    max-width: 95% !important;
    padding-left: 40px !important;
    padding-right: 40px !important;
    margin-left: auto;
    margin-right: auto;
  }
  .title {
    font-size: 38px;
    font-weight: bold;
    color: #2c3e50;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .subtitle {
    font-size: 24px;
    font-weight: normal;
    color: #7f8c8d;
    margin-top: 10px;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
  .author, .date {
    font-size: 16px;
    color: #7f8c8d;
    font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  }
</style>

<!--
  Scoring Report - PhyloPhere Pipeline
  ======================================================

  Author: Miguel Ramon (miguel.ramon@upf.edu)
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  out.width = "100%"
)

suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(ggplot2)
  library(DT)
  library(scales)
  library(ggrepel)
})

# Tableau_10 colour swatch for pipeline-wide consistency
if (requireNamespace("paletteer", quietly = TRUE)) {
  .tableau10 <- as.character(paletteer::paletteer_d("ggthemes::Tableau_10"))
} else {
  .tableau10 <- c(
    "#4E79A7", "#F28E2B", "#E15759", "#76B7B2", "#59A14F",
    "#EDC948", "#B07AA1", "#FF9DA7", "#9C755F", "#BAB0AC"
  )
}
theme_set(
  theme_minimal(base_size = 13) +
  theme(
    axis.title.y  = element_text(margin = ggplot2::margin(l = 20, r = 20)),
    axis.title.x  = element_text(margin = ggplot2::margin(t = 12)),
    axis.text.x   = element_text(margin = ggplot2::margin(t = 5)),
    plot.title    = element_text(face = "bold", hjust = 0, margin = ggplot2::margin(b = 6)),
    plot.subtitle = element_text(colour = "grey40", hjust = 0, margin = ggplot2::margin(b = 10))
  )
)

seed_val <- if (exists("params") && !is.null(params$seed) && nzchar(as.character(params$seed))) as.integer(params$seed) else 1998
set.seed(seed_val)

# Safe correlation helper: returns NA (not an error) when inputs have fewer
# than 3 complete pairs - avoids crashes in sparse optional-module sections.
safe_cor <- function(x, y, method = "pearson") {
  ok <- complete.cases(x, y)
  if (sum(ok) < 3) return(NA_real_)
  cor(x[ok], y[ok], method = method)
}

# Safe quantile: base quantile(x, p, na.rm = TRUE) still ERRORS ("missing values
# and NaN's not allowed if 'na.rm' is FALSE") when na.rm drops every element -
# which happens on small/degenerate permulation pools (e.g. strict
# min_divergent_fraction with few positions). Return NA there instead of halting.
.safe_q <- function(v, p) {
  v <- v[is.finite(v)]
  if (!length(v)) return(NA_real_)
  as.numeric(stats::quantile(v, p, names = FALSE))
}

```

```{r load-data}
pos_df  <- read_tsv(params$position_scores_file, show_col_types = FALSE)
gene_df <- read_tsv(params$gene_scores_file, show_col_types = FALSE)
corr_df <- read_tsv(params$gene_corr_file, show_col_types = FALSE)

# CT p-value measures
boot_measure_col <- if ("recovery_boot" %in% names(pos_df)) {
  "recovery_boot"
} else {
  stop("position_scores.tsv must contain recovery_boot column")
}
boot_measure_is_raw  <- identical(boot_measure_col,  "recovery_boot")

# Provisional detection - overridden in the load-data block after directional scores are computed
# NOTE: what for then?
has_rand <- "accum_cct_p" %in% names(gene_df) && any(!is.na(gene_df$accum_cct_p))
has_rer  <- "rer_min_pval"    %in% names(gene_df) && any(!is.na(gene_df$rer_min_pval)) &&
            "rer_rho"         %in% names(gene_df)

.file_ok <- function(p) !is.null(p) && !is.na(p) && nchar(p) > 0 && file.exists(p)

has_scoring_stress <- .file_ok(params$stress_summary_file) &&
  .file_ok(params$stress_corr_file) &&
  .file_ok(params$stress_rank_file) &&
  .file_ok(params$stress_overlap_file) &&
  .file_ok(params$stress_variants_file)
has_genomic       <- .file_ok(params$genomic_info_file)
has_perms         <- .file_ok(params$caas_perms_file)
has_pos_pval      <- .file_ok(params$caas_pos_pval_file)      # LOO null_pvalue_boot per (gene,position,scheme)
has_pos_sample    <- .file_ok(params$caas_pos_sample_file)    # cycle-stratified null sample (asr + phen + row_caas)
has_filtered      <- .file_ok(params$filtered_discovery_file) # observed per-scheme asr_path_score + recovery_boot
# FADE site (position-level) data: use top direction if available, otherwise try bottom
has_fade_pos      <- .file_ok(params$fade_site_top_file) || .file_ok(params$fade_site_bot_file)
# FADE gene-level data: computed independently by scoring_compute.R from
# fade_summary_top/bottom.tsv.
has_fade_gene     <- any(c("fade_max_bf_top", "fade_max_bf_bottom") %in% names(gene_df))
# Initialized FALSE here so later chunks with eval=has_fade_any don't crash when
# the fade-pos-setup chunk is skipped (has_fade_pos == FALSE).
has_fade_top <- FALSE
has_fade_bot <- FALSE
has_fade_any <- FALSE

if (has_genomic) {
  # Format: gene, chr, start, end, length  (no header assumed; detect if header present)
  genomic_raw <- read_tsv(params$genomic_info_file,
                          col_names = c("Gene", "chr", "gene_start", "gene_end", "gene_length"),
                          show_col_types = FALSE)
  # Drop header row if first cell is literally "gene" or "Gene"
  if (tolower(genomic_raw$Gene[1]) == "gene") genomic_raw <- genomic_raw[-1, ]
  genomic_raw <- genomic_raw %>%
    mutate(
      gene_start  = as.integer(gene_start),
      gene_end    = as.integer(gene_end),
      gene_length = as.integer(gene_length),
      gene_mid    = as.integer((gene_start + gene_end) / 2L)
    )
} else {
  genomic_raw <- NULL
}

# ── Gene-level permulation null (rds: asr corStat + NEW caas corStat) ──────────
# Tidy both byrank lists into one long frame (gene × cycle × direction × metric)
# for the gene-level tabs. asr names are *_asr; strip the suffix → global/top/bottom.
if (has_perms) {
  caas_perms_obj <- tryCatch(readRDS(params$caas_perms_file), error = function(e) NULL)
  .tidy_byrank <- function(cs, metric) {
    purrr::imap_dfr(cs, function(mat, rank_name) {
      mat <- as.matrix(mat)
      val <- as.numeric(mat)
      if (length(val) > 100000) {
        set.seed(seed_val)
        val <- sample(val, 100000)
      }
      tibble::tibble(metric = metric, direction = sub("_asr$", "", rank_name),
                     null_score = val)
    })
  }
  asr_cs  <- if (!is.null(caas_perms_obj)) caas_perms_obj$corStat_byrank      else NULL
  caas_cs <- if (!is.null(caas_perms_obj)) caas_perms_obj$caas_corStat_byrank else NULL
  if (is.null(asr_cs) || length(asr_cs) == 0) {
    has_perms <- FALSE; perm_null_gene <- NULL
  } else {
    perm_null_gene <- dplyr::bind_rows(
        .tidy_byrank(asr_cs, "asr_path_score"),
        if (!is.null(caas_cs) && length(caas_cs)) .tidy_byrank(caas_cs, "caas_score") else NULL) %>%
      mutate(direction = factor(direction, levels = c("global", "top", "bottom")),
             metric    = factor(metric, levels = c("caas_score", "asr_path_score")))
    perm_n_cycles <- ncol(as.matrix(asr_cs[[1]]))
    perm_n_genes  <- nrow(as.matrix(asr_cs[[1]]))
  }
} else {
  caas_perms_obj <- NULL; perm_null_gene <- NULL
}

# Position-level null (cycle-stratified sample) + observed (filtered_discovery).
# The observed side re-derives caas_row exactly as scoring_compute.R section 2f
# does, so the values plotted here are what the pipeline actually scores on:
#     phen_score = 1 - percent_rank(recovery_boot)   ranked over filtered_discovery
#     caas_row   = phen_score * asr_path_score
# The null side arrives pre-ranked from gene_wrapper.py pass B: null_phen_score is
# 1 - percent_rank of the leave-one-out null_pvalue_boot within that cycle's
# genome-wide candidate pool, the null's analogue of the line above.
scheme_levels <- c("US", "GS1", "GS2", "GS3", "GS4")

# Schema guard for the permulation sample. The position-level overlays require a
# null whose phenotype axis is a percent_rank, which the file advertises by
# carrying a `null_phen_score` column. A sample file without it holds an unranked
# phenotype axis (mean ~0.97 rather than ~0.50) and is not on the same scale as
# the ranked observed axis, so overlaying the two would compare quantities that
# only look comparable. Presence of the column is the schema marker; when it is
# absent the dependent sections are suppressed and a message is emitted instead.
.unranked_perm_sample <- FALSE
if (has_pos_sample) {
  .sample_cols <- names(readr::read_tsv(params$caas_pos_sample_file, n_max = 0,
                                        show_col_types = FALSE))
  .unranked_perm_sample <- !("null_phen_score" %in% .sample_cols)
  if (.unranked_perm_sample) has_pos_sample <- FALSE
}

if (has_pos_sample && has_filtered) {
  null_pos <- readr::read_tsv(params$caas_pos_sample_file, show_col_types = FALSE) %>%
    transmute(caap_group,
              cycle          = as.character(cycle),
              asr_path_score = suppressWarnings(as.numeric(asr_path_score)),
              caas_row       = suppressWarnings(as.numeric(null_row_caas)),
              phen_score     = suppressWarnings(as.numeric(null_phen_score)),
              source = "null")
  obs_pos <- readr::read_tsv(params$filtered_discovery_file, show_col_types = FALSE) %>%
    transmute(caap_group,
              cycle          = NA_character_,
              asr_path_score = suppressWarnings(as.numeric(asr_path_score)),
              recovery_boot    = suppressWarnings(as.numeric(recovery_boot)),
              phen_score     = 1 - dplyr::percent_rank(recovery_boot),
              caas_row       = phen_score * asr_path_score,
              source = "observed") %>%
    select(-recovery_boot)
  scheme_overlay <- bind_rows(null_pos, obs_pos) %>%
    filter(caap_group %in% scheme_levels) %>%
    mutate(caap_group = factor(caap_group, levels = scheme_levels),
           source     = factor(source, levels = c("null", "observed")))
  has_scheme_overlay <- nrow(scheme_overlay) > 0
} else {
  scheme_overlay <- NULL
  has_scheme_overlay <- FALSE
}

# Per-(cycle, scheme) null distribution shape. Each cycle is one complete null
# realisation whose candidate pool is the same order of magnitude as the observed
# discovered set, so a cycle's distribution is directly comparable to the observed
# one -- which the pooled union across all cycles is not.
has_pos_quantiles <- .file_ok(params$caas_pos_quantiles_file)
if (has_pos_quantiles) {
  perm_pos_quant <- readr::read_tsv(params$caas_pos_quantiles_file, show_col_types = FALSE) %>%
    filter(caap_group %in% scheme_levels) %>%
    mutate(caap_group = factor(caap_group, levels = scheme_levels))
} else {
  perm_pos_quant <- NULL
}

# Full LOO null_pvalue_boot table (small) - the direct observed-vs-null crosscheck
# for the phenotype axis BEFORE ranking. null_pvalue_boot = (n_detected - 1)/(N - 1)
# is the null's analogue of the observed recovery_boot = count/cycles: same statistic
# ("what fraction of resampling cycles recovered this position"), different
# resampling axis (phenotype relabeling vs species bootstrap).
.unranked_perm_pval <- FALSE
if (has_pos_pval) {
  .pval_cols <- names(readr::read_tsv(params$caas_pos_pval_file, n_max = 0,
                                      show_col_types = FALSE))
  .unranked_perm_pval <- !("null_pvalue_boot" %in% .pval_cols)
  if (.unranked_perm_pval) has_pos_pval <- FALSE   # same schema marker as above
}
if (has_pos_pval) {
  perm_pos_pval <- readr::read_tsv(params$caas_pos_pval_file, show_col_types = FALSE) %>%
    filter(caap_group %in% scheme_levels) %>%
    mutate(caap_group = factor(caap_group, levels = scheme_levels))
} else {
  perm_pos_pval <- NULL
}

# Observed recovery_boot per scheme, paired against the null table above.
if (has_filtered) {
  obs_pos_pval <- readr::read_tsv(params$filtered_discovery_file, show_col_types = FALSE) %>%
    transmute(caap_group,
              recovery_boot = suppressWarnings(as.numeric(recovery_boot))) %>%
    filter(caap_group %in% scheme_levels, is.finite(recovery_boot)) %>%
    mutate(caap_group = factor(caap_group, levels = scheme_levels))
} else {
  obs_pos_pval <- NULL
}
has_pval_pair <- !is.null(perm_pos_pval) && !is.null(obs_pos_pval) &&
                 nrow(perm_pos_pval) > 0 && nrow(obs_pos_pval) > 0

accum_scheme_info <- tibble(
  scheme = c("us", "gs4", "gs3", "gs2", "gs1"),
  label = c("US", "GS4", "GS3", "GS2", "GS1"),
  weight = c(0.5, 0.2, 0.1, 0.1, 0.1),
  pval_col = paste0("accum_pval_", c("us", "gs4", "gs3", "gs2", "gs1")),
  score_col = paste0("accum_score_", c("us", "gs4", "gs3", "gs2", "gs1")),
  weighted_col = paste0("accum_weighted_", c("us", "gs4", "gs3", "gs2", "gs1"))
) %>%
  filter(pval_col %in% names(gene_df))

if (has_scoring_stress) {
  stress_summary_df <- read_tsv(params$stress_summary_file, show_col_types = FALSE)
  stress_corr_df <- read_tsv(params$stress_corr_file, show_col_types = FALSE)
  stress_rank_df <- read_tsv(params$stress_rank_file, show_col_types = FALSE)
  stress_overlap_df <- read_tsv(params$stress_overlap_file, show_col_types = FALSE)
  stress_variants_df <- read_tsv(params$stress_variants_file, show_col_types = FALSE)
} else {
  stress_summary_df <- NULL
  stress_corr_df <- NULL
  stress_rank_df <- NULL
  stress_overlap_df <- NULL
  stress_variants_df <- NULL
}

# Thresholds
top_pct   <- params$top_pct
top25_pct <- 0.25
top5_pct  <- 0.05
top1_pct  <- 0.01

# ── RERConverge availability ────────────────────────────────────────────────
has_rer <- "rer_min_pval" %in% names(gene_df) &&
           any(!is.na(gene_df$rer_min_pval)) &&
           "rer_rho" %in% names(gene_df)

# ── Directional position subsets (change_side from discovery) ──────────────
pos_top    <- pos_df %>% filter(change_side %in% c("top",    "both"))
pos_bottom <- pos_df %>% filter(change_side %in% c("bottom", "both"))

# ── Directional gene CAAS scores ────────────────────────────────────────────
# Read from gene_scores.tsv (written by scoring_compute.R section 4a), never
# recomputed here, so the report cannot disagree with the pipeline's own output.
# gene_caas_top / gene_caas_bot survive only as local aliases: ~49 downstream
# call sites in this file already use those shorter names.
#
# Prefer the canonical gene_caas_score_{top,bottom}; fall back to the *_all
# columns they are copied from, so a gene_scores.tsv written by an older run
# still renders instead of erroring.
.pick_dir <- function(primary, fallback) {
  v <- if (primary %in% names(gene_df)) gene_df[[primary]] else NULL
  if (is.null(v) || all(is.na(v))) {
    v <- if (fallback %in% names(gene_df)) gene_df[[fallback]] else rep(NA_real_, nrow(gene_df))
  }
  suppressWarnings(as.numeric(v))
}
gene_df$gene_caas_top <- .pick_dir("gene_caas_score_top", "gene_caas_score_top_all")
gene_df$gene_caas_bot <- .pick_dir("gene_caas_score_bottom", "gene_caas_score_bottom_all")

# ── Gene-level accumulation significance ─────────────────────────────────────
# Accumulation is represented by its CCT combined p-value (accum_cct_p) across
# the per-scheme accum p-values, with BH FDR (accum_fdr) and accum_significant
# (accum_fdr < 0.05).
has_accum_score <- "accum_cct_p" %in% names(gene_df) && any(!is.na(gene_df$accum_cct_p))
has_rand <- has_accum_score

# ── Gene-level direction assignment ─────────────────────────────────────────
# Assign direction to each gene from its top-scoring position
gene_direction_df <- pos_df %>%
  group_by(Gene) %>%
  slice_max(CAAS_score, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(Gene, gene_direction = change_side)

gene_df <- gene_df %>% left_join(gene_direction_df, by = "Gene")

# ── Gate-based gene lists (global / p-val sig) ────────────────────────────────
# Derived from position-level hypergeometric significance flags so gene membership
# is tied to the test outcome rather than an arbitrary top-% rank cut.
.gate_genes <- function(pos_sub, gate_col = NULL) {
  if (!is.null(gate_col) && gate_col %in% names(pos_sub))
    pos_sub %>% filter(.data[[gate_col]] %in% TRUE) %>% pull(Gene) %>% unique()
  else
    pos_sub %>% pull(Gene) %>% unique()
}

gene_top_list <- .gate_genes(pos_top)
gene_bot_list <- .gate_genes(pos_bottom)

# Named list for looping in overlap sections
.caas_gates <- list(
  list(lbl = "All Positions", top = gene_top_list, bot = gene_bot_list)
)

# Background universe for gene-set overlap Fisher tests (RER / FADE / Accumulation
# Venn panels) - every gene that received a CAAS gene score in this run.
gene_universe_n <- nrow(gene_df)

# ── Normalisation helper ──────────────────────────────────────────────────────
rank_norm <- function(x) {
  r <- rank(x, na.last = "keep", ties.method = "average")
  (r - 1) / (max(r, na.rm = TRUE) - 1)
}

# ── Venn diagram helpers (used across RER / FADE / Accumulation tabs) ─────────
# Overlap enrichment is tested with a Fisher exact test (mathematically the
# hypergeometric test on a 2x2 table) against the full gene universe passed in
# `universe_n` - for gene-level overlaps this is nrow(gene_df), i.e. every gene
# that received a CAAS gene score in this run.
.venn_counts <- function(set_a, set_b, label_a, label_b, universe_n) {
  n_a  <- length(set_a); n_b <- length(set_b)
  n_ab <- length(intersect(set_a, set_b))
  n_only_a  <- n_a - n_ab
  n_only_b  <- n_b - n_ab
  n_neither <- universe_n - n_a - n_b + n_ab
  ft <- tryCatch(
    fisher.test(matrix(c(n_ab, n_only_b, n_only_a, n_neither), nrow = 2)),
    error = function(e) list(estimate = NA_real_, p.value = NA_real_)
  )
  tibble(label_a, label_b, n_a, n_b, n_ab,
         n_only_a = n_only_a, n_only_b = n_only_b,
         odds_ratio = unname(ft$estimate), fisher_p = ft$p.value)
}
.draw_venn <- function(vc, col_a, col_b, title) {
  or_txt <- if (is.na(vc$odds_ratio)) {
    "OR = NA"
  } else {
    sprintf("OR = %.2f (Fisher p = %s)", vc$odds_ratio,
            formatC(vc$fisher_p, format = "e", digits = 1))
  }
  ggplot() +
    geom_rect(aes(xmin=0.05, xmax=0.65, ymin=0.22, ymax=0.85),
              fill=col_a, alpha=0.35, colour=col_a, linewidth=1) +
    geom_rect(aes(xmin=0.35, xmax=0.95, ymin=0.22, ymax=0.85),
              fill=col_b, alpha=0.35, colour=col_b, linewidth=1) +
    annotate("text", x=0.20, y=0.535, label=vc$n_only_a, size=7, fontface="bold") +
    annotate("text", x=0.50, y=0.535, label=vc$n_ab,     size=7, fontface="bold", colour="grey20") +
    annotate("text", x=0.80, y=0.535, label=vc$n_only_b, size=7, fontface="bold") +
    annotate("text", x=0.20, y=0.15, label=vc$label_a, size=3.5, colour=col_a) +
    annotate("text", x=0.80, y=0.15, label=vc$label_b, size=3.5, colour=col_b) +
    annotate("text", x=0.50, y=0.93,
             label=sprintf("Shared: %d genes (%.0f%% of smaller set)",
                           vc$n_ab, 100 * vc$n_ab / max(min(vc$n_a, vc$n_b), 1)),
             size=3, colour="grey40") +
    annotate("text", x=0.50, y=0.02, label=or_txt, size=3, fontface="bold", colour="grey20") +
    ggtitle(title) + theme_void() +
    coord_cartesian(xlim = c(0, 1), ylim = c(-0.06, 1.02), clip = "off") +
    theme(plot.title  = element_text(face="bold", hjust=0, size=11, margin = ggplot2::margin(b = 6)),
          plot.margin = margin(t = 4, r = 4, b = 14, l = 4))
}
```

---

## Introduction

CAAS Scoring combines a position's convergence and biochemical evidence into one
ranked score, then rolls that up to a per-gene score comparable against
independent evolutionary evidence.

1. **Position score (`CAAS_score`)**: the mean of `phen_score x asr_score` over
   the biochemical schemes that detected the position. `phen_score` is a
   permutation-based confidence that the pattern is not noise; `asr_score` is a
   five-factor ancestral-state signal that the convergence reflects independent
   origins, not shared ancestry. See Scoring Methodology below and the
   Biochemistry tab for how the schemes factor in.

2. **Gene score (`gene_caas_score_top` / `_bottom`)**: the gene's strongest
   position, adjusted for how many positions the gene has, so genes are not
   rewarded simply for having more of them. Compared against:
   `r if (has_rand) "- **accum_cct_p**: Cauchy-combined p-value across per-scheme accumulation p-values (+ BH FDR accum_fdr)."`
   `r if (has_rer) "- RERConverge p-value and direction (accelerating / decelerating)."`
   - FADE: gene is significant per direction if any position has BF >= 100.

## Companion reports

This scoring report is the hub of the downstream enrichment suite. The reports below
are rendered as sibling HTML files in the same `html_reports/` directory and are
auto-annotated with the cross-module scoring signals (CAAS gates, top/bottom
percentiles, FADE, accumulation) produced here. A link is live only if the
corresponding module was part of this run. FADE and Accumulation do not run
their own FCS ranking; they appear there as cross-module corroboration flags.
Position Characterisation (PrimateAI-3D + COSMIC validation) is part of the
Position-level FCS report below.

```{r companion-reports, results='asis'}
.trait <- if (is.null(params$traitname) || !nzchar(params$traitname)) "unknown_trait" else params$traitname
.companions <- tibble::tribble(
  ~Report,                              ~File,
  "FCS - CAAS (this scoring run)",      sprintf("12.FCS_scoring_%s.html",    .trait),
  "FCS - RERconverge (+ p.perm)",       sprintf("12.FCS_rer_%s.html",        .trait),
  "AMI (DOMINO active modules)",        sprintf("13.AMI_analysis_%s.html", .trait),
  "Position-level FCS (+ Position Characterisation)", sprintf("14.Position_enrichment_report_%s.html", .trait),
  "COMPARE - cross-module overview",    sprintf("15.Comparison_report_%s.html",.trait)
)
cat(paste0("- [", .companions$Report, "](", .companions$File, ")", collapse = "\n"))
cat("\n")
```

---

## Scoring Methodology

This section documents the exact weighting and combination logic used to
produce every score in this report, so results can be interpreted and
reproduced independently of the pipeline code.

<details>
<summary><strong>Click to expand full methodology</strong></summary>

---

### Position-Level Score

`CAAS_score` combines two orthogonal evidence axes, each on a continuous [0, 1] scale:

#### 1. Phenotype evolution: `phen_score`

One minus the permutation p-value (`recovery_boot`), which permutes the phenotype
on the tree:

$$\text{phen\_score} = 1 - P_{\text{boot}}$$

#### 2. Position evolution: `asr_path_score` (`asr_score`)

A product of five factors built from the per-node ASR posteriors (full
definitions in [`docs/ASR_PATH_SCORE.md`](../../docs/ASR_PATH_SCORE.md)):

| Factor | What it rewards |
|--------|-----------------|
| `core` | P(>=2 independent changes): combinatorial replication across directions, from per-pair isolation (the change occurs above the MRCA, not inherited) |
| `independence` | derived state absent at the pairs' shared ancestors (LAC nodes) |
| `mrca_diversity` | non-parallel origins (distinct ancestral starting posteriors) |
| `derived_agreement` | changed pairs landing on the same derived residue per side |
| `conservation_gate` | conserved pairs holding the ancestral state deeply |

#### 3. Combining schemes

A substitution is tested against five biochemical partitions (US, GS1-GS4); a
scheme "detects" the position if the substitution crosses one of its group
boundaries. `CAAS_score` is the mean of `phen_score x asr_score` over the
schemes that detected the position:

$$\text{caas\_row} = \text{phen\_score} \times \text{asr\_path\_score}$$
$$\text{CAAS\_score} = \operatorname{mean}_{s \in \text{detected schemes}} \text{caas\_row}_s$$

`n_schemes` (how many partitions detected the position) and `scheme_set`
(which ones) are reported as descriptors and do not scale the score. See the
Biochemistry tab for what the five schemes are and why breadth is not treated
as extra evidence.



---

### Gene-Level Scores

#### gene_caas_top / gene_caas_bot

The gene's strongest position (`max(CAAS_score)`), calibrated against how many
positions the gene has: `F` is the genome-wide ECDF of `CAAS_score`
(direction-matched pool for the top/bottom variants) and `n` the gene's
position count.

$$\text{gene\_caas\_top} = F_{\text{top}}\big(\max(\text{CAAS\_score} \mid \text{change\_side} \in \{\text{top, both}\})\big)^{\,n_{\text{top}}}$$
$$\text{gene\_caas\_bot} = F_{\text{bottom}}\big(\max(\text{CAAS\_score} \mid \text{change\_side} \in \{\text{bottom, both}\})\big)^{\,n_{\text{bottom}}}$$

Without the exponent, a gene with more positions would score higher purely
from having more chances at a high maximum; raising the pool percentile to the
`n`th power removes that.

#### gene_accum_score (`accum_cct_p`) *(optional)*

Cauchy Combination Test across the per-scheme accumulation p-values, valid
under their positive dependence (the schemes test overlapping positions).
Reported with BH FDR (`accum_fdr`); `accum_significant` is `accum_fdr < 0.05`.

#### RERConverge characterisation *(optional)*

`p.perm` when present, falling back to `p.adj` / `p.value` / `pvalue` / `pval`.
Direction from the sign of `Rho`: positive is accelerating, negative is
decelerating. Significance threshold `p.perm <= 0.05`.

#### FADE gene-level characterisation *(optional)*

A gene is FADE-significant per direction if it contains at least one position
with `max_site_bf >= 100`.

---

### Gene List Extraction

Gene lists for downstream STRING/DOMINO active-module analysis are extracted
at three thresholds
(configurable via `--scoring_position_top_pct` / `--scoring_position_top5_pct` /
`--scoring_position_top1_pct` and their gene-level equivalents) and three
directional subsets:

| Subset | Filter |
|--------|--------|
| **top** | Positions / genes where `change_side == "top"` |
| **bottom** | Positions / genes where `change_side == "bottom"` |
| **both** | All top-scoring positions / genes regardless of direction |

For **position-level** lists, the threshold is applied to `CAAS_score` and
the gene name is extracted from the qualifying positions.

For **gene-level** lists, the threshold is applied to `gene_caas_top` (TOP
direction) and `gene_caas_bot` (BOTTOM direction) independently.

</details>

---

## Input Distributions {.tabset}

This section displays the distributions of the primary input significance measures (Permutation P-values and Phenotype Evolution scores) along with a joint comparison to show how they behave relative to the ancestral reconstruction axis and the final CAAS score.

### Phenotype Evolution vs Ancestral Reconstruction

```{r phen-vs-asr-scatter, fig.height=7, fig.width=8}
joint_df <- pos_df %>%
  filter(!is.na(phen_score), !is.na(asr_score))

# safe_cor() returns NA when < 3 complete pairs instead of erroring out
.r_phen_caas <- round(safe_cor(joint_df$phen_score, joint_df$CAAS_score), 2)
.r_asr_caas  <- round(safe_cor(joint_df$asr_score,  joint_df$CAAS_score), 2)
.r_phen_asr  <- round(safe_cor(joint_df$phen_score, joint_df$asr_score), 2)

ggplot(joint_df, aes(x = asr_score, y = phen_score, colour = CAAS_score)) +
  geom_point(alpha = 0.45, size = 1.6) +
  scale_colour_gradientn(
    colours = c("#4E79A7", "#76B7B2", "#EDC948", "#F28E2B", "#E15759"),
    name    = "CAAS\nscore"
  ) +
  annotate("text", x = -Inf, y = Inf,
           label = sprintf("r(phen, CAAS) = %.2f\nr(asr, CAAS) = %.2f\nr(phen, asr) = %.2f",
                           .r_phen_caas, .r_asr_caas, .r_phen_asr),
           hjust = -0.05, vjust = 1.3, size = 3.2, colour = "grey30") +
  labs(
    title    = "Orthogonal Evidence Axes: ASR Path Score vs Phenotype Evolution Score",
    x        = "asr_score (Ancestral State Path Confidence)",
    y        = "phen_score (1 - percent_rank(recovery_boot))"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

---

## CAAS Permulation Null {.tabset}

NOTE: This must be overhawled

The permulation re-labels the phylogeny `N` times and re-runs discovery + ASR scoring
under each null labeling. Each cycle is **one complete null realisation**: the positions
that cycle discovered are its candidate set, exactly as the real run's discovered set is
the observed candidate set. Both null quantities are built term-for-term like the
observed pipeline:

- **`asr_path_score`** - the permuted position-evolution signal (gene null = 0.90-quantile
  over the gene's positions). This axis is not currently wired into any ranking; the
  observed `gene_caas_score` now uses a size-adjusted max instead (see below).
- **`CAAS_score`** = Σ_scheme `null_phen_score × asr_path_score × 0.2`, where
  `null_phen_score = 1 − percent_rank(null_pvalue_boot)` and
  `null_pvalue_boot = (n_detected − 1)/(N − 1)` is the leave-one-out share of the *other*
  cycles that also recovered that position - the null analogue of the observed
  `recovery_boot = count/cycles`.

> [!IMPORTANT]
> **Both sides are `percent_rank`ed, and they must be.** The observed
> `phen_score = 1 − percent_rank(recovery_boot)` is ranked across the whole scored pool;
> the null's is ranked across that cycle's genome-wide candidate pool. The rank is what
> escapes the `1/N` resolution floor - the raw statistic is heavily concentrated near 0
> (observed mean ≈ 0.05), so without ranking the axis collapses to a near-constant ≈ 0.97
> and the null score inflates well past the observed. Ranking makes both axes uniform on
> [0, 1] and therefore comparable.
>
> Because these are **within-run percentile ranks**, absolute score values are *not*
> comparable across runs or traits - only ranks within a run are meaningful.

The formal per-pathway `p.perm` (asr axis) is in the companion **FCS** report.

```{r perm-gene-obs-setup}
dir_pal <- c(global = "#4E79A7", top = "#E15759", bottom = "#59A14F")
# Observed gene-level distributions by direction (always computed: used both by the
# full null-overlay plots when a permulation null exists, and by the standalone
# summary table below when it doesn't).
.obs_gene_caas_dist <- dplyr::bind_rows(
  tibble::tibble(direction = "global", value = gene_df$gene_caas_score),
  tibble::tibble(direction = "top",    value = gene_df$gene_caas_score_top),
  tibble::tibble(direction = "bottom", value = gene_df$gene_caas_score_bottom)
)
.obs_gene_caas <- .obs_gene_caas_dist %>%
  group_by(direction) %>% summarise(obs_median = median(value, na.rm = TRUE), .groups = "drop")

.q90 <- function(v) if (length(v)) as.numeric(quantile(v, .9, na.rm = TRUE)) else NA_real_
.obs_gene_asr_dist <- pos_df %>%
  group_by(Gene) %>%
  summarise(global = .q90(asr_score),
            top    = .q90(asr_score[change_side %in% c("top", "both")]),
            bottom = .q90(asr_score[change_side %in% c("bottom", "both")]),
            .groups = "drop") %>%
  tidyr::pivot_longer(c(global, top, bottom), names_to = "direction", values_to = "value")
.obs_gene_asr <- .obs_gene_asr_dist %>%
  group_by(direction) %>% summarise(obs_median = median(value, na.rm = TRUE), .groups = "drop")

# caas_cs (caas_corStat_byrank) can be missing from caas_perms.rds even when the asr
# null (asr_cs) is present - flag that separately so the CAAS-score tab can show a
# warning instead of an empty plot.
has_caas_null <- has_perms && !is.null(perm_null_gene) && any(perm_null_gene$metric == "caas_score", na.rm = TRUE)

.gene_null_table <- function(metric_name, obs_df) {
  # Both metrics are exactly 0 whenever a permulated labeling produced zero
  # CAAS/CAAP hits anywhere in the gene that cycle (gene_wrapper.py writes an
  # explicit 0.0 row, not NA, when there is nothing to aggregate over) - a
  # structural point mass, not sampling noise. At >=75% zero (typical here) the
  # unconditional median/IQR are both 0 and say nothing; p_zero plus the
  # nonzero-conditional median/IQR are what actually separates directions/runs.
  # null_q90 is kept because a tail quantile stays informative even under heavy
  # zero-inflation, as long as p_zero < 90%.
  perm_null_gene %>% filter(metric == metric_name) %>% group_by(direction) %>%
    summarise(`n (gene×cycle)` = dplyr::n(),
              p_zero          = mean(null_score == 0, na.rm = TRUE),
              null_median_nz  = median(null_score[null_score > 0], na.rm = TRUE),
              null_IQR_nz     = IQR(null_score[null_score > 0], na.rm = TRUE),
              null_q90        = quantile(null_score, .90, na.rm = TRUE),
              null_median     = median(null_score, na.rm = TRUE),
              null_IQR        = IQR(null_score, na.rm = TRUE),
              .groups = "drop") %>%
    mutate(direction = as.character(direction)) %>%
    left_join(obs_df, by = "direction") %>%
    arrange(factor(direction, levels = c("global", "top", "bottom")))
}
# Shared paired observed-vs-null violin. Used by the gene-level tabs and by the
# position-level phenotype-axis tab, so the two never drift apart visually.
# `combined` must carry columns: grp (x axis), value, source in {"Null","Observed"}.
.paired_violin <- function(combined, xlab, ylab, title, subtitle,
                           drop_zeros = TRUE) {
  combined <- combined %>%
    mutate(source = factor(source, levels = c("Null", "Observed"))) %>%
    filter(is.finite(value))

  # A violin's density estimate is dominated by any point mass at a boundary:
  # when most of the null sits at exactly 0 the KDE puts almost all its area in a
  # spike at y=0 and squashes the actual shape of "what happens when there IS a
  # signal" into an invisible sliver. Plot the conditional (>0) distribution,
  # which a density can actually represent, and report the discarded zero
  # fraction as a label under each violin instead, since the two pieces of
  # information don't fit in one geometry.
  if (drop_zeros) {
    zero_frac <- combined %>% group_by(grp, source) %>%
      summarise(pct_zero = sprintf("0: %.0f%%", 100 * mean(value == 0)), .groups = "drop")
    dat <- combined %>% filter(value > 0)
  } else {
    zero_frac <- NULL
    dat <- combined
  }

  p <- ggplot(dat, aes(grp, value, fill = source)) +
    geom_violin(scale = "width", alpha = .7, colour = "grey30",
                position = position_dodge(width = 0.8)) +
    scale_fill_manual(values = c(Null = "grey60", Observed = "#E15759"), name = NULL) +
    labs(title = title, subtitle = subtitle, x = xlab, y = ylab) +
    theme_minimal(base_size = 12) +
    theme(
      axis.title.y = element_text(margin = ggplot2::margin(l = 20, r = 20)),
      axis.title.x = element_text(margin = ggplot2::margin(t = 12)),
      axis.text.x  = element_text(margin = ggplot2::margin(t = 5))
    )
  if (!is.null(zero_frac)) {
    p <- p + geom_text(data = zero_frac,
                       aes(x = grp, y = -Inf, label = pct_zero, group = source),
                       position = position_dodge(width = 0.8), vjust = -0.6,
                       size = 3, colour = "grey40")
  }
  p
}

.gene_null_plot <- function(metric_name, obs_dist_df, ylab, title) {
  nd <- perm_null_gene %>% filter(metric == metric_name, is.finite(null_score)) %>%
    transmute(grp = as.character(direction), value = null_score, source = "Null")
  od <- obs_dist_df %>% transmute(grp = as.character(direction), value, source = "Observed") %>%
    filter(is.finite(value))
  combined <- dplyr::bind_rows(nd, od) %>%
    mutate(grp = factor(grp, levels = c("global", "top", "bottom")))

  .paired_violin(
    combined, xlab = "Direction", ylab = ylab, title = title,
    subtitle = paste0(
      "Distribution CONDITIONAL on a nonzero score - null: permulation, 100k random sample.\n",
      "Exact-zero fraction labelled below each violin (not part of the density)."))
}
```

```{r perm-na, eval=!has_perms, results='asis'}
cat("_No CAAS permulation null available for this run",
    " (`--caas_permulation_enrichment` off or `caas_perms.rds` absent)._\n\n", sep = "")
cat("**Observed statistics summary** (no permulation null to compare against):\n\n")
print(knitr::kable(
  dplyr::bind_rows(
    .obs_gene_caas %>% mutate(metric = "gene CAAS_score"),
    .obs_gene_asr  %>% mutate(metric = "gene asr_path_score (q90 over positions)")
  ) %>% dplyr::select(metric, direction, obs_median),
  digits = 4, caption = "Observed gene-level medians by direction"
))
```

```{r gene-level-header, eval=has_perms, results='asis'}
cat(r"(### Gene Level {.tabset}
)")
```

```{r caas-score-header, eval=has_perms, results='asis'}
if (has_caas_null) cat("#### CAAS score\n\n")
```

```{r perm-gene-caas, eval=has_caas_null, fig.height=5, fig.width=10}
knitr::kable(.gene_null_table("caas_score", .obs_gene_caas), digits = 4,
  caption = sprintf("Gene-level null CAAS_score by direction (q90 over positions; %d genes × %d cycles). obs_median = observed gene CAAS_score.",
                    perm_n_genes, perm_n_cycles))
print(.gene_null_plot("caas_score", .obs_gene_caas_dist, "CAAS_score", "Gene-level null vs observed CAAS_score"))
```

```{r caas-null-missing, eval=has_perms && !has_caas_null, results='asis'}
cat("#### CAAS score\n\n")
cat("_CAAS score permulation null unavailable for this run (`caas_corStat_byrank` missing from `caas_perms.rds`)._\n\n")
```

```{r asr-score-header, eval=has_perms, results='asis'}
cat("#### asr_path_score\n\n")
```

```{r perm-gene-asr, eval=has_perms, fig.height=5, fig.width=10}
knitr::kable(.gene_null_table("asr_path_score", .obs_gene_asr), digits = 4,
  caption = "Gene-level null asr_path_score by direction (q90 over positions). obs_median = observed q90(asr_score).")
print(.gene_null_plot("asr_path_score", .obs_gene_asr_dist, "asr_path_score (q90)", "Gene-level null vs observed asr_path_score"))
```

```{r fpr-calib-setup, include=FALSE}
# ── FPR / threshold calibration (Tier 1C) ──────────────────────────────────
# Two genome-wide count calibrations against the SAME permulation cycles as
# the gene/position p.perm above, at NO extra ASR cost (reuses the already-
# staged detail shards / gene_cycle_scores.tsv). This is a check on the
# CURRENT thresholds only (caas_perms_postproc / gap_fraction / etc. as they
# stood when the null was built) -- not a threshold sweep; a changed
# postproc setting needs the null rebuilt, not just re-plotted.
has_gcs         <- .file_ok(params$caas_gene_cycle_scores_file)
has_pos_detail  <- .file_ok(params$caas_pos_detail_dir)

# Position-level: per cycle, genome-wide count of distinct (Gene, Position,
# caap_group) candidate triples. One shard == one gene, so "distinct
# (Position, caap_group)" WITHIN a shard is already gene-disambiguated --
# summing each shard's per-cycle counts gives the true genome-wide total
# without ever holding more than one shard in memory at a time.
.count_per_cycle_from_detail <- function(detail_path) {
  shards <- if (dir.exists(detail_path)) {
    sort(list.files(detail_path, pattern = "\\.tsv\\.gz$", full.names = TRUE))
  } else {
    detail_path
  }
  if (length(shards) == 0) return(tibble::tibble(cycle = character(), n_triples = integer()))
  per_shard <- vector("list", length(shards))
  for (i in seq_along(shards)) {
    df <- tryCatch(
      readr::read_tsv(shards[i],
        col_types = readr::cols_only(cycle = "c", Position = "i", caap_group = "c"),
        show_col_types = FALSE),
      error = function(e) NULL)
    per_shard[[i]] <- if (is.null(df) || nrow(df) == 0) NULL else
      df %>% dplyr::distinct(cycle, Position, caap_group) %>% dplyr::count(cycle, name = "n")
  }
  dplyr::bind_rows(per_shard) %>% dplyr::group_by(cycle) %>%
    dplyr::summarise(n_triples = sum(n), .groups = "drop")
}

if (has_pos_detail) {
  pos_cycle_counts <- .count_per_cycle_from_detail(params$caas_pos_detail_dir)
  obs_pos_triples <- if (has_filtered) {
    readr::read_tsv(params$filtered_discovery_file, show_col_types = FALSE) %>%
      dplyr::distinct(Gene, Position, caap_group) %>% nrow()
  } else NA_integer_
  pos_calib_p <- if (nrow(pos_cycle_counts) > 0 && !is.na(obs_pos_triples)) {
    (sum(pos_cycle_counts$n_triples >= obs_pos_triples) + 1) / (nrow(pos_cycle_counts) + 1)
  } else NA_real_
} else {
  pos_cycle_counts <- tibble::tibble(cycle = character(), n_triples = integer())
  obs_pos_triples <- NA_integer_
  pos_calib_p <- NA_real_
}

# Gene-level: per cycle, count of genes with a nonzero global_caas hit
# (gene_wrapper.py writes an explicit 0.0 row for a cycle that scored
# nothing, so this is a real "no hit" count, not missing data).
if (has_gcs) {
  gcs_raw <- readr::read_tsv(params$caas_gene_cycle_scores_file, show_col_types = FALSE)
  gene_cycle_counts <- gcs_raw %>% dplyr::group_by(cycle) %>%
    dplyr::summarise(n_genes_hit = sum(global_caas > 0, na.rm = TRUE), .groups = "drop")
  obs_genes_hit <- sum(gene_df$gene_caas_score > 0, na.rm = TRUE)
  gene_calib_p <- if (nrow(gene_cycle_counts) > 0) {
    (sum(gene_cycle_counts$n_genes_hit >= obs_genes_hit) + 1) / (nrow(gene_cycle_counts) + 1)
  } else NA_real_
} else {
  gene_cycle_counts <- tibble::tibble(cycle = character(), n_genes_hit = integer())
  obs_genes_hit <- NA_integer_
  gene_calib_p <- NA_real_
}

.calib_hist <- function(counts_df, count_col, obs_val, xlab, title) {
  ggplot(counts_df, aes(x = .data[[count_col]])) +
    geom_histogram(bins = min(40, max(5, length(unique(counts_df[[count_col]])))),
                    fill = "grey60", colour = "grey30") +
    { if (is.finite(obs_val)) geom_vline(xintercept = obs_val, colour = "#E15759", linewidth = 1) } +
    labs(title = title, x = xlab, y = sprintf("cycles (N=%d)", nrow(counts_df)),
         subtitle = if (is.finite(obs_val)) sprintf("Observed = %s (red line)", format(obs_val, big.mark = ",")) else NULL) +
    theme_minimal(base_size = 12)
}
```

```{r fpr-calib-header, eval=has_gcs || has_pos_detail, results='asis'}
cat("#### FPR calibration\n\n")
cat("How large would the observed genome-wide CAAS discovery look if it were pure",
    "phenotype-tree-confound noise? Each histogram is the genome-wide count over",
    sprintf("N=%d permuted-labeling cycles", max(nrow(pos_cycle_counts), nrow(gene_cycle_counts))),
    "at the SAME thresholds this run used; the red line is the observed count.",
    "This calibrates the current thresholds only -- it is not a threshold sweep,",
    "and if `--caas_perms_postproc` differed between when this null was built and",
    "this run's `filtered_discovery.tsv`, the two sides are not distribution-matched",
    "and this figure should not be read literally (rebuild the null instead).\n\n")
```

```{r fpr-calib-pos, eval=has_pos_detail, results='asis', fig.height=4.5, fig.width=9}
if (nrow(pos_cycle_counts) > 0) {
  cat(sprintf("**Position-level** (Gene, Position, scheme) candidate triples per cycle. p = %s\n\n",
              if (is.na(pos_calib_p)) "NA (no `filtered_discovery.tsv` to compare against)" else sprintf("%.4g", pos_calib_p)))
  print(.calib_hist(pos_cycle_counts, "n_triples", obs_pos_triples,
                     "Genome-wide candidate triples / cycle",
                     "Position-level FPR calibration"))
} else {
  cat("_No cycles recovered from `perm_pos_detail`._\n\n")
}
```

```{r fpr-calib-gene, eval=has_gcs, results='asis', fig.height=4.5, fig.width=9}
if (nrow(gene_cycle_counts) > 0) {
  cat(sprintf("**Gene-level** genes with a nonzero CAAS hit per cycle. p = %s\n\n",
              if (is.na(gene_calib_p)) "NA" else sprintf("%.4g", gene_calib_p)))
  print(.calib_hist(gene_cycle_counts, "n_genes_hit", obs_genes_hit,
                     "Genes with CAAS_score > 0 / cycle",
                     "Gene-level FPR calibration"))
} else {
  cat("_No cycles recovered from `gene_cycle_scores.tsv`._\n\n")
}
```

```{r fpr-calib-missing, eval=!has_gcs && !has_pos_detail, results='asis'}
cat("#### FPR calibration\n\n")
cat("_Neither `perm_pos_detail` nor `gene_cycle_scores.tsv` was supplied to this run",
    "(pass `--caas_pos_detail_file` / `--caas_gene_cycle_scores_file`, or rerun with",
    "`--caas_permulation_enrichment`)._\n\n")
```

### Position Level {.tabset}

Per CAAP grouping scheme (`US`, `GS1`–`GS4`): observed overlaid on the permulation null (H0).

```{r perm-pos-scale-mismatch, eval=(.unranked_perm_sample || .unranked_perm_pval), results='asis'}
cat(paste0(
  "> [!WARNING]\n",
  "> **The permulation files for this run are not on the same scale as the observed\n",
  "> side, so the position-level comparisons below are suppressed.**\n",
  ">\n",
  "> These sections require a null whose phenotype axis is a percent_rank, uniform on\n",
  "> [0, 1] with mean ~0.50 - the scale the observed `phen_score` occupies. The files\n",
  "> supplied here carry an unranked phenotype axis (`recovery_pval` / `row_caas`,\n",
  "> mean ~0.97 rather than ~0.50) and no `null_phen_score` column. Plotting the two\n",
  "> together would show a null sitting above the observed purely because of the scale\n",
  "> difference, so the panels are withheld rather than shown misleadingly.\n",
  ">\n",
  "> Run `CAAS_PERMULATION` for this trait to produce `perm_pos_pval.tsv`,\n",
  "> `perm_pos_sample.tsv` and `perm_pos_quantiles.tsv` on the ranked scale\n",
  "> (`caas_perms.rds`, which feeds the FCS `p.perm`, is regenerated alongside them).\n\n"))
```

```{r perm-pos-na, eval=!has_scheme_overlay && !.unranked_perm_sample && !.unranked_perm_pval, results='asis'}
cat("_Position-level null overlay unavailable (needs the cycle-stratified perm sample + filtered_discovery)._\n")
```

```{r perm-pos-setup, eval=has_scheme_overlay}
src_pal <- c(null = "#BAB0AC", observed = "#E15759")
.pos_density <- function(value_col, xlab, title) {
  set.seed(seed_val)
  nd <- scheme_overlay %>% filter(source == "null") %>% group_by(caap_group) %>%
        slice_sample(n = 20000) %>% ungroup()
  od <- scheme_overlay %>% filter(source == "observed")
  d  <- bind_rows(nd, od) %>% filter(is.finite(.data[[value_col]]))
  ggplot(d, aes(.data[[value_col]], fill = source, colour = source)) +
    geom_density(alpha = .45, adjust = 1.1) +
    facet_wrap(~ caap_group, scales = "free_y", ncol = 3) +
    scale_fill_manual(values = src_pal) + scale_colour_manual(values = src_pal) +
    labs(title = title, subtitle = "observed (red) overlaid on permulation null (grey)",
         x = xlab, y = "density") +
    theme_minimal(base_size = 12) +
    theme(
      legend.position = "top",
      axis.title.y = element_text(margin = ggplot2::margin(l = 20, r = 20)),
      axis.title.x = element_text(margin = ggplot2::margin(t = 12)),
      axis.text.x  = element_text(margin = ggplot2::margin(t = 5))
    )
}
.pos_table <- function(value_col) {
  scheme_overlay %>% group_by(caap_group, source) %>%
    summarise(
      n = n(),
      median = median(.data[[value_col]], na.rm = TRUE),
      q90    = quantile(.data[[value_col]], .90, na.rm = TRUE),
      max    = max(.data[[value_col]], na.rm = TRUE),
      .groups = "drop"
    )
}
```

#### CAAS score

```{r perm-pos-caas, eval=has_scheme_overlay, fig.height=6, fig.width=10}
knitr::kable(.pos_table("caas_row"), digits = 4,
  caption = "Per-scheme CAAS contribution phen_score×asr×0.2: permulation null vs observed. Both sides use 1−percent_rank of their own bootstrap/recovery statistic.")
print(.pos_density("caas_row", "per-scheme CAAS contribution", "CAAS contribution by functional group: observed vs null"))
```

#### asr_path_score

```{r perm-pos-asr, eval=has_scheme_overlay, fig.height=6, fig.width=10}
knitr::kable(.pos_table("asr_path_score"), digits = 4,
  caption = "Per-scheme asr_path_score: permulation null vs observed.")
print(.pos_density("asr_path_score", "asr_path_score", "asr_path_score by functional group: observed vs null"))
```

#### Phenotype axis (raw + ranked)

The phenotype axis is the component most at risk of the two sides drifting apart, so it
gets the same paired-violin treatment as the gene-level scores rather than a null-only
histogram. **Raw** compares the two bootstrap/recovery statistics before ranking;
**ranked** compares what actually enters the score. If the ranked pair does not overlap
substantially, the two sides are not being built the same way.

```{r perm-pos-pval, eval=has_pval_pair, fig.height=6, fig.width=10}
.pval_pair <- dplyr::bind_rows(
  perm_pos_pval %>% transmute(grp = caap_group, value = null_pvalue_boot, source = "Null"),
  obs_pos_pval  %>% transmute(grp = caap_group, value = recovery_boot,      source = "Observed")
)
knitr::kable(
  .pval_pair %>% group_by(grp, source) %>%
    summarise(n = dplyr::n(), median = median(value, na.rm = TRUE),
              mean = mean(value, na.rm = TRUE),
              q90 = quantile(value, .90, na.rm = TRUE), .groups = "drop"),
  digits = 4,
  caption = sprintf(paste0("RAW phenotype statistic per scheme. Null `null_pvalue_boot` = ",
                           "(n_detected-1)/(N-1), leave-one-out over N=%d cycles; observed ",
                           "`recovery_boot` = count/cycles from the species bootstrap."),
                    perm_pos_pval$n_cycles[1]))
print(.paired_violin(
  .pval_pair, xlab = "CAAP scheme", ylab = "raw p-value",
  title = "Phenotype statistic BEFORE ranking: observed vs null",
  subtitle = "Null: leave-one-out per-cycle statistic. Observed: species-bootstrap statistic.",
  drop_zeros = FALSE))
```

```{r perm-pos-phen, eval=has_scheme_overlay, fig.height=6, fig.width=10}
.phen_pair <- scheme_overlay %>%
  transmute(grp = caap_group, value = phen_score,
            source = ifelse(source == "null", "Null", "Observed"))
knitr::kable(.pos_table("phen_score"), digits = 4,
  caption = paste0("RANKED phenotype axis (phen_score = 1 - percent_rank) per scheme. ",
                    "Null is ranked within each cycle's candidate pool; observed is ranked across the whole scored pool."))
print(.paired_violin(
  .phen_pair, xlab = "CAAP scheme", ylab = "phen_score = 1 - percent_rank",
  title = "Phenotype axis AFTER ranking: observed vs null",
  subtitle = "Null: ranked within each cycle's candidate pool. Observed: ranked across the whole scored pool.",
  drop_zeros = FALSE))
```

#### Per-cycle null envelope

```{r perm-pos-cycle-na, eval=!has_pos_quantiles, results='asis'}
cat("_Per-cycle null envelope unavailable (needs `perm_pos_quantiles.tsv` from CAAS_PERMULATION)._\n")
```

```{r perm-pos-cycle, eval=has_pos_quantiles && has_scheme_overlay, fig.height=6, fig.width=10}
# Each permulation cycle is a self-contained null realisation whose candidate pool is
# the same order of magnitude as the observed discovered set, so a cycle's distribution
# is directly comparable to the observed one. Pooling all cycles into one union (as the
# older summaries did) is NOT comparable: the union is ~N times larger than any single
# realisation and mixes positions that no single null run ever discovered together.
.cyc <- perm_pos_quant %>% filter(metric == "null_row_caas")
.obs_cyc <- scheme_overlay %>% filter(source == "observed") %>%
  group_by(caap_group) %>%
  summarise(q50 = median(caas_row, na.rm = TRUE),
            q90 = .safe_q(caas_row, .90), .groups = "drop")

if (nrow(.cyc) == 0 || nrow(.obs_cyc) == 0) {
  cat("_Per-cycle null envelope unavailable (empty null or observed pool ",
      "- degenerate permulation).\n_")
} else {
print(
  ggplot(.cyc, aes(caap_group)) +
    geom_violin(aes(y = q50, fill = "Null cycles"), scale = "width",
                alpha = .65, colour = "grey30") +
    geom_point(data = .obs_cyc, aes(y = q50, colour = "Observed"), size = 3) +
    scale_fill_manual(values = c(`Null cycles` = "grey60"), name = NULL) +
    scale_colour_manual(values = c(Observed = "#E15759"), name = NULL) +
    labs(title = "Per-cycle null median CAAS contribution vs observed",
         subtitle = paste0("Each null violin is the spread of per-cycle MEDIANS across the ",
                           "permulation cycles (one point per cycle);\nthe red point is the ",
                           "observed median. Observed should sit in or above the null's upper tail."),
         x = "CAAP scheme", y = "median per-scheme CAAS contribution") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "top")
)

knitr::kable(
  .cyc %>% group_by(caap_group) %>%
    summarise(n_cycles = dplyr::n(),
              `null median (across cycles)` = median(q50, na.rm = TRUE),
              `null q90 (across cycles)`    = .safe_q(q50, .90),
              .groups = "drop") %>%
    left_join(.obs_cyc %>% select(caap_group, `observed median` = q50), by = "caap_group"),
  digits = 5,
  caption = "Per-cycle null medians summarised across cycles, against the single observed median.")
}
```

---

## Position-Level Scores {.tabset}

### Significance gate & p-values (directional)

```{r pos-obs-pval-gate}
# Observed hypergeometric p-values and significance-gate pass-rates by
# direction (these are observed quantities - the permulation does not permute them).
.dir_subsets <- list(
  global = pos_df,
  top    = pos_df %>% filter(change_side %in% c("top", "both")),
  bottom = pos_df %>% filter(change_side %in% c("bottom", "both"))
)
obs_tbl <- purrr::imap_dfr(.dir_subsets, function(d, dir) {
  tibble::tibble(
    direction               = dir,
    positions               = nrow(d),
    `median permutation p`  = if ("recovery_boot" %in% names(d)) median(d$recovery_boot, na.rm = TRUE) else NA_real_,
    `mean permutation p`    = if ("recovery_boot" %in% names(d)) mean(d$recovery_boot,   na.rm = TRUE) else NA_real_,
    `median CAAS score`     = if ("CAAS_score" %in% names(d))  median(d$CAAS_score,  na.rm = TRUE) else NA_real_
  )
})
knitr::kable(obs_tbl, digits = 4,
  caption = "Observed permutation p-values and CAAS scores by direction.")
```

### CAAS Score Distribution

```{r caas-score-dist, fig.height=9}
# Global thresholds (used throughout the report for CAAS filters)
thr_10 <- quantile(pos_df$CAAS_score, 1 - top_pct,  na.rm = TRUE)
thr_5  <- quantile(pos_df$CAAS_score, 1 - top5_pct, na.rm = TRUE)
thr_1  <- quantile(pos_df$CAAS_score, 1 - top1_pct, na.rm = TRUE)

# Direction-specific thresholds (used for is_Xpct flags)
.thr_top_10 <- quantile(pos_top$CAAS_score, 1 - top_pct,  na.rm = TRUE)
.thr_top_5  <- quantile(pos_top$CAAS_score, 1 - top5_pct, na.rm = TRUE)
.thr_top_1  <- quantile(pos_top$CAAS_score, 1 - top1_pct, na.rm = TRUE)
.thr_bot_10 <- quantile(pos_bottom$CAAS_score, 1 - top_pct,  na.rm = TRUE)
.thr_bot_5  <- quantile(pos_bottom$CAAS_score, 1 - top5_pct, na.rm = TRUE)
.thr_bot_1  <- quantile(pos_bottom$CAAS_score, 1 - top1_pct, na.rm = TRUE)

p_combined <- ggplot(pos_df, aes(x = CAAS_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50,
                 fill = .tableau10[1], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[2]) +
  geom_vline(xintercept = thr_10, linetype = "dashed",
             colour = .tableau10[3], linewidth = 0.7) +
  geom_vline(xintercept = thr_5,  linetype = "longdash",
             colour = .tableau10[5], linewidth = 0.7) +
  geom_vline(xintercept = thr_1,  linetype = "dotted",
             colour = .tableau10[4], linewidth = 0.7) +
  annotate("text", x = thr_10, y = Inf, label = sprintf("Top %g%%", top_pct * 100),
           vjust = 2,    hjust = -0.1, colour = .tableau10[3], size = 3.5) +
  annotate("text", x = thr_5,  y = Inf, label = sprintf("Top %g%%", top5_pct * 100),
           vjust = 2.75, hjust = -0.1, colour = .tableau10[5], size = 3.5) +
  annotate("text", x = thr_1,  y = Inf, label = sprintf("Top %g%%", top1_pct * 100),
           vjust = 3.5,  hjust = -0.1, colour = .tableau10[4], size = 3.5) +
  labs(title = "CAAS Score Distribution - Combined",
       subtitle = sprintf("n = %s positions | Trait: %s",
                          format(nrow(pos_df), big.mark = ","), params$traitname),
       x = "CAAS Score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))

print(p_combined)
```

### Score axis

The two axes whose product is `CAAS_score` (`phen_score` x `asr_score`).
`asr_score` is itself a five-factor product (core x independence x diversity x
agreement x conservation); those sub-factors are not shown here.

```{r score-axes-dist, fig.height=9}
.axis_plot <- function(col, title, subtitle, fill, line) {
  if (!col %in% names(pos_df)) return(NULL)
  ggplot(pos_df, aes(x = .data[[col]])) +
    geom_histogram(aes(y = after_stat(density)), bins = 30, fill = fill, alpha = 0.7) +
    geom_density(linewidth = 0.8, colour = line) +
    labs(title = title, subtitle = subtitle, x = col, y = "Density") +
    theme(plot.title = element_text(face = "bold"))
}
.axis_plots <- Filter(Negate(is.null), list(
  .axis_plot("asr_score", "ASR Path Score (position evolution)",
             "Unified five-factor ASR path score (core x independence x diversity x agreement x conservation)",
             .tableau10[3], .tableau10[4]),
  .axis_plot("phen_score", "Phenotype Evolution (1 - recovery_boot)",
             "Permulation evidence",
             .tableau10[1], .tableau10[2])
))
if (length(.axis_plots)) gridExtra::grid.arrange(grobs = .axis_plots, ncol = 1)
```

### Biochemistry

Every candidate substitution is tested against five schemes: `US` (exact amino
acid identity) and four biochemical groupings, `GS1`-`GS4`, each partitioning the
20 amino acids by a different combination of physicochemical criteria (charge,
polarity, volume, thiol chemistry). A scheme "detects" a position when the
substitution crosses one of its group boundaries. The schemes are not nested;
they classify amino acids differently from each other, so which ones detect a
given position identifies what kind of physicochemical property changed, not
merely how much evidence exists.

`n_schemes` and `scheme_set` are descriptors of the substitution's chemistry,
not inputs to `CAAS_score` (mean over detected schemes, Scoring Methodology
above). How many schemes fire is largely a property of where the ancestral
residue sits in each partition: an amino acid alone in its group (e.g.
threonine under `GS1`) is displaced by almost any substitution, while one in a
five-member group needs a larger change to cross out. The group-size plot below
shows this directly.

```{r biochem-scheme-breadth, fig.height=9}
p_bwsum <- ggplot(pos_df, aes(x = factor(n_schemes))) +
  geom_bar(fill = .tableau10[4], alpha = 0.8) +
  labs(title = "Schemes detecting each position",
       x = "n_schemes", y = "Count") +
  theme(plot.title = element_text(face = "bold"))

# scheme_set: which combination of schemes fired, not just how many.
if ("scheme_set" %in% names(pos_df) && any(!is.na(pos_df$scheme_set))) {
  scheme_combo_df <- pos_df %>%
    filter(!is.na(scheme_set), scheme_set != "") %>%
    count(scheme_set, sort = TRUE) %>%
    slice_head(n = 15)
  p_scheme_combo <- ggplot(scheme_combo_df, aes(x = reorder(scheme_set, n), y = n)) +
    geom_col(fill = .tableau10[6], alpha = 0.85) +
    coord_flip() +
    labs(title = "Most common scheme combinations",
         x = "scheme_set", y = "Positions") +
    theme(plot.title = element_text(face = "bold"))
  gridExtra::grid.arrange(p_bwsum, p_scheme_combo, ncol = 1, nrow = 2)
} else {
  print(p_bwsum)
}

# Marginal firing rate per scheme.
if ("scheme_set" %in% names(pos_df) && any(!is.na(pos_df$scheme_set))) {
  .sets <- pos_df$scheme_set[!is.na(pos_df$scheme_set) & pos_df$scheme_set != ""]
  scheme_rate_df <- tibble::tibble(
    scheme = scheme_levels,
    rate   = vapply(scheme_levels,
                    function(s) mean(grepl(paste0("(^|\\+)", s, "($|\\+)"), .sets)),
                    numeric(1))
  )
  p_scheme_rate <- ggplot(scheme_rate_df, aes(x = reorder(scheme, -rate), y = rate)) +
    geom_col(fill = .tableau10[2], alpha = 0.85) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
    labs(title = "Per-scheme firing rate",
         x = "Scheme", y = "% of positions") +
    theme(plot.title = element_text(face = "bold"))

  print(p_scheme_rate)
}
```

```{r biochem-group-size, echo=FALSE, fig.width=11, fig.height=4.5, warning=FALSE, message=FALSE}
# Group definitions mirror src/biochem/grouping.py / CT/local/modules/caap_id.py.
.aa_groups <- list(
  GS1 = list("CV", "AGPS", "NDQE", "RHK", "ILMFWY", "T"),
  GS2 = list("C", "AGV", "DE", "NQHW", "RK", "ILFP", "YMTS"),
  GS3 = list("C", "AGPST", "NDQE", "RHK", "ILMV", "FWY"),
  GS4 = list("C", "AILV", "ST", "NQ", "DE", "RH", "G", "P", "K", "M", "F", "WY")
)
.group_size_of <- function(aa, scheme) {
  gs <- .aa_groups[[scheme]]
  hit <- Filter(function(g) grepl(aa, g, fixed = TRUE), gs)
  if (length(hit) == 0) return(NA_integer_)
  nchar(hit[[1]])
}

if (isTRUE(has_filtered)) {
  .fd <- tryCatch(readr::read_tsv(params$filtered_discovery_file, show_col_types = FALSE),
                  error = function(e) NULL)
} else .fd <- NULL

if (!is.null(.fd) && all(c("caap_group", "all_mrca_state", "Gene", "Position") %in% names(.fd))) {
  .fired <- .fd %>%
    filter(caap_group %in% scheme_levels,
           !is.na(all_mrca_state), nchar(as.character(all_mrca_state)) == 1) %>%
    group_by(Gene, Position) %>%
    summarise(anc = first(as.character(all_mrca_state)),
              fired = list(unique(as.character(caap_group))), .groups = "drop")

  gsz <- lapply(names(.aa_groups), function(sch) {
    tibble::tibble(
      scheme     = sch,
      group_size = vapply(.fired$anc, .group_size_of, integer(1), scheme = sch),
      fires      = vapply(.fired$fired, function(f) sch %in% f, logical(1))
    )
  }) %>% bind_rows() %>% filter(!is.na(group_size))

  gsz_summary <- gsz %>%
    group_by(scheme, group_size) %>%
    summarise(rate = mean(fires), n = dplyr::n(), .groups = "drop") %>%
    filter(n >= 20)

  if (nrow(gsz_summary) > 0) {
    print(
      ggplot(gsz_summary, aes(x = factor(group_size), y = rate, fill = scheme)) +
        geom_col(position = position_dodge(width = 0.8), alpha = 0.85) +
        scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
        labs(title = "Scheme firing rate vs ancestral residue's group size",
             x = "Group size of the ancestral residue under that scheme",
             y = "% of positions where the scheme fires", fill = "Scheme") +
        theme(plot.title = element_text(face = "bold"))
    )
  } else {
    cat("_Not enough positions per (scheme, group-size) cell to plot firing rate._\n")
  }
} else {
  cat("_filtered_discovery with `all_mrca_state` not available; skipping the group-size diagnostic._\n")
}
```

### Stress Test

`r if (!has_scoring_stress) "> **Note:** scoring stress-test outputs are not available for this run."`

| Variant | What it represents |
|---------|---------------------|
| `CAAS_current` | The production `CAAS_score`: mean of `phen_score` x `asr_score` over the schemes that detected the position. |
| `CAAS_phen_only` | `phen_score` alone -- drops the ASR axis. |
| `CAAS_asr_only` | `asr_score` alone -- drops the phenotype axis. |
| `CAAS_scheme_max` | `max(caas_row)` instead of the mean over detected schemes -- tests the mean-vs-max scheme-aggregation choice (section 2g). |

A robust run shows consistent Spearman rank correlation with `CAAS_current` across all variants.

```{r stress-summary-table, eval=has_scoring_stress}
DT::datatable(
  stress_summary_df %>%
    arrange(desc(spearman_to_current), mean_abs_rank_shift) %>%
    mutate(across(where(is.numeric), ~round(., 4))),
  rownames = FALSE,
  options = list(pageLength = 20, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "Assumption sensitivity summary for alternative position-level score variants"
  )
)
```

### Top Positions

```{r top-positions-table}
# Two-way score axes first (phen_score × asr_score), then the gate tiers, then
# supplementary diagnostics. any_of() keeps this robust to absent columns.
display_cols <- c("Gene", "Position", "CAAS_score",
                  "phen_score", "asr_score", "core",
                  "recovery_boot",
                  "n_schemes", "scheme_set", "change_side",
                  # raw-AA descriptors (position-level, from position_scores.tsv)
                  "derived_residues", "top_residue_support", "bottom_residue_support",
                  "n_conserved_pairs", "convergence_schemes",
                  "is_top_10pct", "is_top_5pct", "is_top_1pct",
                  "is_bot_10pct", "is_bot_5pct", "is_bot_1pct")

DT::datatable(
  pos_df %>%
    arrange(desc(CAAS_score)) %>%
    select(any_of(display_cols)) %>%
    mutate(across(where(is.numeric), ~round(., 4))),
  filter    = "top",
  rownames  = FALSE,
  options   = list(pageLength = 25, scrollX = TRUE),
  caption   = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Position-level CAAS scores | Trait: %s", params$traitname)
  )
)
```

### CAAS Score Manhattan (Position Level)

`r if (!has_genomic) "> **Note:** Genomic coordinates not available - position-level Manhattan skipped."`

```{r pos-caas-manhattan-setup, eval=has_genomic}
# Join all scored positions with genomic coordinates via their gene.
# Each position inherits its gene's genomic midpoint as the x-coordinate.
# We build the chr offsets independently here (rather than relying on the
# genomic-setup chunk, which runs later in the document).
pos_genomic <- pos_df %>%
  inner_join(genomic_raw, by = "Gene") %>%
  mutate(
    chr_label = sub("^chr", "", chr),
    chr_num   = suppressWarnings(as.integer(chr_label)),
    chr_order = dplyr::case_when(
      !is.na(chr_num)              ~ chr_num,
      chr_label == "X"             ~ 23L,
      chr_label == "Y"             ~ 24L,
      chr_label %in% c("MT","M")   ~ 25L,
      TRUE                         ~ 26L
    )
  ) %>%
  arrange(chr_order, gene_mid)

pos_chr_offsets <- pos_genomic %>%
  group_by(chr_order, chr_label) %>%
  summarise(chr_max = max(gene_end, na.rm = TRUE), .groups = "drop") %>%
  arrange(chr_order) %>%
  mutate(offset = cumsum(lag(chr_max, default = 0L)) + (chr_order - 1L) * 5e6L)

pos_genomic <- pos_genomic %>%
  left_join(pos_chr_offsets %>% select(chr_order, offset), by = "chr_order") %>%
  mutate(
    manhattan_x = gene_mid + offset,
    chr_even    = chr_order %% 2 == 0
  )

pos_chr_label_positions <- pos_chr_offsets %>%
  left_join(
    pos_genomic %>% group_by(chr_order) %>%
      summarise(mid_x = mean(manhattan_x, na.rm = TRUE), .groups = "drop"),
    by = "chr_order"
  )
```

```{r pos-caas-manhattan-plot, eval=has_genomic, fig.height=5, fig.width=14}
pos_genomic <- pos_genomic %>%
  mutate(caas_log = -log10(1 - pmin(CAAS_score, 0.9999)))
pos_thr_log <- -log10(1 - pmin(thr_10, 0.9999))

ggplot(pos_genomic, aes(x = manhattan_x, y = caas_log, colour = chr_even)) +
  geom_point(alpha = 0.45, size = 0.6, show.legend = FALSE) +
  geom_hline(yintercept = pos_thr_log, linetype = "dashed",
             colour = "grey40", linewidth = 0.6) +
  scale_colour_manual(
    values = c("FALSE" = .tableau10[1], "TRUE" = .tableau10[2]),
    guide = "none"
  ) +
  scale_x_continuous(
    breaks = pos_chr_label_positions$mid_x,
    labels = pos_chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  labs(
    title    = "Position-Level CAAS Score - Manhattan View",
    subtitle = sprintf("%d positions with genomic coordinates | dashed = top %g%% threshold",
                       nrow(pos_genomic), params$top_pct * 100),
    x = "Chromosome", y = "-log10(1 - CAAS Score)"
  ) +
  theme(
    plot.title         = element_text(face = "bold"),
    axis.text.x        = element_text(size = 7),
    panel.grid.major.x = element_blank()
  )
```

---

## Direction Score Profiles {.tabset}

Each tab shows distribution plots for TOP-direction and BOTTOM-direction positions
side by side, allowing direct comparison of how each scoring component behaves
across the two phenotypic directions. These tabs mirror the global Position-Level
Scores section, faceted by direction.

### CAAS Score

```{r dir-caas-dist, fig.height=9}
make_dir_dist <- function(df, dir_label, thr10, thr5, thr1, col) {
  ggplot(df, aes(x = CAAS_score)) +
    geom_histogram(aes(y = after_stat(density)), bins = 50,
                   fill = col, alpha = 0.7) +
    geom_density(linewidth = 0.8, colour = col) +
    geom_vline(xintercept = thr10, linetype = "dashed",  colour = "grey30", linewidth = 0.7) +
    geom_vline(xintercept = thr5,  linetype = "longdash", colour = "grey30", linewidth = 0.7) +
    geom_vline(xintercept = thr1,  linetype = "dotted",  colour = "grey30", linewidth = 0.7) +
    annotate("text", x = thr10, y = Inf,
             label = sprintf("Top %g%%\n(%.4f)", top_pct * 100,  thr10),
             vjust = 2,    hjust = -0.05, colour = "grey20", size = 3) +
    annotate("text", x = thr5,  y = Inf,
             label = sprintf("Top %g%%\n(%.4f)", top5_pct * 100, thr5),
             vjust = 3.8,  hjust = -0.05, colour = "grey20", size = 3) +
    annotate("text", x = thr1,  y = Inf,
             label = sprintf("Top %g%%\n(%.4f)", top1_pct * 100, thr1),
             vjust = 5.6,  hjust = -0.05, colour = "grey20", size = 3) +
    labs(title = sprintf("CAAS Score Distribution - %s positions", dir_label),
         subtitle = sprintf("n = %s | thresholds computed within direction",
                            format(nrow(df), big.mark = ",")),
         x = "CAAS Score", y = "Density") +
    theme(plot.title = element_text(face = "bold"))
}

gridExtra::grid.arrange(
  make_dir_dist(pos_top,    "TOP",    .thr_top_10, .thr_top_5, .thr_top_1, .tableau10[1]),
  make_dir_dist(pos_bottom, "BOTTOM", .thr_bot_10, .thr_bot_5, .thr_bot_1, .tableau10[3]),
  ncol = 1
)
```

### P-Value Distributions

```{r dir-pval-dist, fig.height=9}
make_dir_pval_panel <- function(df, measure_col, is_raw, dir_label, col) {
  vals   <- df[[measure_col]]
  breaks <- quantile(vals, probs = seq(0, 1, 0.1), na.rm = TRUE)
  plot_x <- if (is_raw) -log10(pmax(vals, 1e-300)) else vals
  xlab   <- if (is_raw) sprintf("-log10(%s)", measure_col) else measure_col
  title  <- sprintf("%s - %s", ifelse(is_raw, toupper(measure_col), measure_col), dir_label)
  df2 <- data.frame(x = plot_x)
  ggplot(df2, aes(x = x)) +
    geom_histogram(aes(y = after_stat(density)), bins = 50, fill = col, alpha = 0.65) +
    geom_density(linewidth = 0.8, colour = col) +
    geom_vline(
      xintercept = if (is_raw) -log10(pmax(breaks[2:10], 1e-300)) else breaks[2:10],
      linetype = "dotted", colour = "grey50", linewidth = 0.3
    ) +
    labs(title = title,
         subtitle = sprintf("n = %s positions | dotted = decile boundaries",
                            format(nrow(df), big.mark = ",")),
         x = xlab, y = "Density") +
    theme(plot.title = element_text(face = "bold", size = 10))
}

p_boot_top <- make_dir_pval_panel(pos_top,    boot_measure_col, boot_measure_is_raw, "TOP",    .tableau10[1])
p_boot_bot <- make_dir_pval_panel(pos_bottom, boot_measure_col, boot_measure_is_raw, "BOTTOM", .tableau10[3])

gridExtra::grid.arrange(
  p_boot_top, p_boot_bot,
  ncol = 2,
  top  = grid::textGrob(
    "Permutation p-value (recovery_boot) distribution by direction",
    gp = grid::gpar(fontface = "bold", fontsize = 12)
  )
)
```



### Score axis

```{r dir-score-axis-dist, fig.height=9}
make_dir_component_plot <- function(df, col, dir_label, colour) {
  if (!col %in% names(df)) return(NULL)
  ggplot(df, aes(x = .data[[col]])) +
    geom_histogram(aes(y = after_stat(density)), bins = 40, fill = colour, alpha = 0.65) +
    geom_density(linewidth = 0.8, colour = colour) +
    labs(title    = sprintf("%s - %s", col, dir_label),
         subtitle = sprintf("n = %s positions", format(nrow(df), big.mark = ",")),
         x = col, y = "Density") +
    theme(plot.title = element_text(face = "bold", size = 10))
}

dir_axis_plots <- Filter(Negate(is.null), list(
  make_dir_component_plot(pos_top,    "phen_score", "TOP",    .tableau10[1]),
  make_dir_component_plot(pos_bottom, "phen_score", "BOTTOM", .tableau10[3]),
  make_dir_component_plot(pos_top,    "asr_score",  "TOP",    .tableau10[5]),
  make_dir_component_plot(pos_bottom, "asr_score",  "BOTTOM", .tableau10[7])
))
if (length(dir_axis_plots)) {
  gridExtra::grid.arrange(
    grobs = dir_axis_plots, ncol = 2,
    top = grid::textGrob(
      "Score axis distributions by direction: phenotype evolution (row 1) and ASR path score (row 2)",
      gp = grid::gpar(fontface = "bold", fontsize = 12)
    )
  )
}
```

### Top Positions by Direction

```{r dir-top-positions-table}
has_caas_col <- "caas" %in% names(pos_df)

dir_display_cols <- c("Gene", "Position", "CAAS_score",
                      "phen_score", "asr_score", "core",
                      if (has_caas_col) "caas",
                      "n_schemes", "scheme_set",
                      "recovery_boot",
                      "change_side",
                      "derived_residues", "top_residue_support", "bottom_residue_support",
                      "convergence_schemes",
                      "is_top_10pct", "is_top_5pct", "is_top_1pct",
                      "is_bot_10pct", "is_bot_5pct", "is_bot_1pct")

dir_pos_tbl <- bind_rows(
  pos_top    %>% mutate(.direction = "TOP"),
  pos_bottom %>% mutate(.direction = "BOTTOM")
) %>%
  arrange(.direction, desc(CAAS_score)) %>%
  select(Direction = .direction, any_of(dir_display_cols)) %>%
  mutate(across(where(is.numeric), ~round(., 4)))

DT::datatable(
  dir_pos_tbl,
  filter   = "top",
  rownames = FALSE,
  options  = list(pageLength = 25, scrollX = TRUE),
  caption  = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Top-scoring positions by direction | Trait: %s", params$traitname)
  )
)
```

---

## Geometric convergence vs score thresholds {.tabset}

`convergence_type` (from CT_DISAMBIGUATION) is a **geometric** label - read from tip
states alone, independent of any score. This section asks: as the CAAS / ASR /
phenotype score threshold tightens, does the geometric convergence-type composition
of the retained positions shift? Same six cumulative thresholds SCORING uses
everywhere else (`top100` = every scored position, tightening to `top1` = the top
percentile by that score), cut three ways - `CAAS_score`, `asr_path_score`,
`phen_score` - plus a combined heatmap against all five ASR path-score axes.

```{r convtype-setup, include=FALSE}
.conv_thr <- tibble::tibble(
  label = factor(c("top100", "top50", "top25", "top10", "top5", "top1"),
                 levels = c("top100", "top50", "top25", "top10", "top5", "top1")),
  q     = c(0.00, 0.50, 0.75, 0.90, 0.95, 0.99)
)
.axis5_cols <- c("core", "independence", "mrca_diversity", "derived_agreement", "conservation_gate")

has_convtype <- .file_ok(params$filtered_discovery_file)
if (has_convtype) {
  .disc_df <- readr::read_tsv(params$filtered_discovery_file, show_col_types = FALSE) %>%
    dplyr::select(Gene, Position, convergence_type, any_of(.axis5_cols), any_of("asr_path_score"))
  .conv_df <- pos_df %>%
    dplyr::select(Gene, Position, CAAS_score, phen_score, any_of("change_side")) %>%
    dplyr::inner_join(.disc_df, by = c("Gene", "Position"))
  has_convtype <- nrow(.conv_df) > 0 && "convergence_type" %in% names(.conv_df)
}

# Composition of convergence_type among positions retained at each cumulative
# threshold (top100 >= top50 >= ... >= top1 by construction -- thresholds are
# computed and applied independently per row of .conv_thr, not as mutually
# exclusive bins, so this never double-counts across threshold levels).
.threshold_composition <- function(df, score_col, dir_label) {
  sub <- df %>% dplyr::filter(!is.na(.data[[score_col]]))
  if (dir_label != "Global") {
    side <- tolower(dir_label)
    sub <- sub %>% dplyr::filter(change_side %in% c(side, "both"))
  }
  if (nrow(sub) == 0) return(NULL)
  purrr::map_dfr(seq_len(nrow(.conv_thr)), function(i) {
    thr <- stats::quantile(sub[[score_col]], .conv_thr$q[i], na.rm = TRUE)
    sel <- sub %>% dplyr::filter(.data[[score_col]] >= thr)
    if (nrow(sel) == 0) return(tibble::tibble())
    sel %>% dplyr::count(convergence_type) %>%
      dplyr::mutate(threshold = .conv_thr$label[i], n_total = nrow(sel), pct = round(100 * n / n_total, 2))
  })
}

.plot_threshold_composition <- function(comp, title) {
  if (is.null(comp) || nrow(comp) == 0) return(NULL)
  ggplot(comp, aes(x = threshold, y = pct, fill = convergence_type)) +
    geom_col(position = "stack") +
    scale_fill_viridis_d(option = "turbo", name = "convergence_type") +
    labs(title = title, x = "score threshold (cumulative, tightening →)",
         y = "% of positions retained at that threshold") +
    theme_minimal(base_size = 12)
}

.render_convtype_score_tab <- function(score_col, score_title) {
  for (dr in c("Global", "Top", "Bottom")) {
    cat(sprintf("\n#### %s\n\n", dr))
    if (!has_convtype) { cat("_No `filtered_discovery_file` / `convergence_type` available._\n\n"); next }
    if (!score_col %in% names(.conv_df)) { cat(sprintf("_`%s` not available in this run._\n\n", score_col)); next }
    comp <- .threshold_composition(.conv_df, score_col, dr)
    p <- .plot_threshold_composition(comp, sprintf("%s thresholds vs convergence_type (%s)", score_title, dr))
    if (is.null(p)) { cat("_No data for this direction._\n\n"); next }
    print(p)
    cat("\n\n")
    print(knitr::kable(
      comp %>% dplyr::arrange(threshold, dplyr::desc(pct)) %>%
        dplyr::select(threshold, convergence_type, n, n_total, `%` = pct),
      caption = sprintf("%s threshold composition - %s", score_title, dr)
    ))
    cat("\n\n")
  }
}
```

### CAAS score {.tabset}

```{r convtype-caas, results='asis', fig.height=5}
.render_convtype_score_tab("CAAS_score", "CAAS score")
```

### ASR path score {.tabset}

```{r convtype-asr, results='asis', fig.height=5}
.render_convtype_score_tab("asr_path_score", "ASR path score")
```

### Phenotype score {.tabset}

```{r convtype-phen, results='asis', fig.height=5}
.render_convtype_score_tab("phen_score", "Phenotype score")
```

### Axes + scores heatmap {.tabset}

Mean value of each of the five ASR path-score axes, plus `CAAS_score` /
`asr_path_score` / `phen_score` themselves, per `convergence_type` - shows which
axis (or score) drives or sinks each geometric class, e.g. parallel labels should
score low on `mrca_diversity`, divergent ones low on `derived_agreement`.

```{r convtype-heatmap, results='asis', fig.height=6}
for (dr in c("Global", "Top", "Bottom")) {
  cat(sprintf("\n#### %s\n\n", dr))
  if (!has_convtype) { cat("_No `filtered_discovery_file` / `convergence_type` available._\n\n"); next }
  sub <- .conv_df
  if (dr != "Global") sub <- sub %>% dplyr::filter(change_side %in% c(tolower(dr), "both"))
  value_cols <- intersect(c(.axis5_cols, "CAAS_score", "asr_path_score", "phen_score"), names(sub))
  if (length(value_cols) == 0 || nrow(sub) == 0) { cat("_No data for this direction._\n\n"); next }

  axis_long <- sub %>%
    dplyr::select(convergence_type, dplyr::all_of(value_cols)) %>%
    tidyr::pivot_longer(dplyr::all_of(value_cols), names_to = "axis", values_to = "value") %>%
    dplyr::group_by(convergence_type, axis) %>%
    dplyr::summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop")

  p <- ggplot(axis_long, aes(x = axis, y = convergence_type, fill = mean_value)) +
    geom_tile(colour = "grey90") +
    geom_text(aes(label = sprintf("%.2f", mean_value)), size = 3) +
    scale_fill_viridis_c(option = "C", name = "mean") +
    theme_minimal(base_size = 12) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
    labs(title = sprintf("Axes + scores by convergence_type (%s)", dr), x = NULL, y = NULL)
  print(p)
  cat("\n\n")

  print(knitr::kable(
    axis_long %>% dplyr::mutate(mean_value = round(mean_value, 3)) %>%
      tidyr::pivot_wider(names_from = axis, values_from = mean_value),
    caption = sprintf("Mean axis/score value by convergence_type - %s", dr)
  ))
  cat("\n\n")
}
```

---

```{r fade-pos-char-header, results='asis'}
if (has_fade_pos) cat(r"(
---

## FADE Position Characterisation {.tabset}

> **FADE** (FUBAR Approach to Directional Evolution) tests whether foreground branches show
> elevated substitution bias toward a particular amino acid at each site.
> Top-direction FADE characterises positions where the phenotypic group shows accelerated
> amino acid evolution; bottom-direction shows the opposite.
> CAAS positions (BF ≥ 100) overlap reveals sites that are both convergent and under directional selection.

)")
```

```{r fade-pos-setup, eval=has_fade_pos}
# Load FADE site tables for each direction separately.
# FADE_report.Rmd emits site = seq_along(...) making site numbers 1-based, while
# CAAS/CT uses 0-based positions (alimport.py range(0, length), caas_id.py as-is).
# Subtract 1 here to align coordinate systems before joining.
.load_fade_sites <- function(path) {
  if (!.file_ok(path)) return(NULL)
  # FADE_JSON_TO_CSV emits a comma-separated file (fade_sites_*.csv); older
  # paths handed a .tsv. Pick the reader by the delimiter actually present on
  # the header line rather than trusting the extension.
  hdr <- tryCatch(readLines(path, n = 1L, warn = FALSE), error = function(e) "")
  df <- if (length(hdr) && grepl(",", hdr) && !grepl("\t", hdr)) {
    read_csv(path, show_col_types = FALSE)
  } else {
    read_tsv(path, show_col_types = FALSE)
  }
  names(df) <- tolower(names(df))
  if (ncol(df) < 2) return(NULL)  # malformed / single-column: nothing to join
  
  if (!"gene" %in% names(df)) names(df)[1] <- "gene"
  if ("site" %in% names(df)) {
    df$position <- as.integer(df$site) - 1L
  } else if ("position" %in% names(df)) {
    df$position <- as.integer(df$position)
  } else {
    df$position <- as.integer(df[[2]]) - 1L
  }
  
  bf_col <- intersect(c("max_site_bf", "max_bf", "bayes_factor", "bf"), names(df))[1]
  if (is.na(bf_col)) {
    df$max_site_bf <- 0
  } else {
    df$max_site_bf <- as.numeric(df[[bf_col]])
  }
  
  aa_col <- intersect(c("top_target_aa", "target_aa", "aa"), names(df))[1]
  if (!is.na(aa_col)) {
    df$top_target_aa <- as.character(df[[aa_col]])
  } else {
    df$top_target_aa <- NA_character_
  }
  
  bias_col <- intersect(c("fade_biased", "biased"), names(df))[1]
  if (!is.na(bias_col)) {
    df$fade_biased <- as.logical(df[[bias_col]])
  } else {
    df$fade_biased <- df$max_site_bf >= 100
  }
  
  df %>%
    select(Gene = gene, Position = position, max_site_bf, top_target_aa, fade_biased)
}
fade_top_df <- .load_fade_sites(params$fade_site_top_file)
fade_bot_df <- .load_fade_sites(params$fade_site_bot_file)

# Join FADE BFs to directional CAAS position subsets
.join_fade <- function(pos_subset, fade_df) {
  if (is.null(fade_df) || nrow(pos_subset) == 0) return(NULL)
  pos_subset %>%
    left_join(
      fade_df %>% select(Gene, Position, max_site_bf, top_target_aa, fade_biased),
      by = c("Gene", "Position")
    ) %>%
    mutate(
      fade_biased = replace_na(fade_biased, FALSE),
      fade_sig    = !is.na(max_site_bf) & max_site_bf >= 100,
      caas_decile = dplyr::ntile(CAAS_score, 10),
      norm_caas   = rank(CAAS_score,   ties.method = "average", na.last = "keep") /
                    sum(!is.na(CAAS_score)),
      norm_log_bf = {
        lbf <- log10(pmax(max_site_bf, 0.01))
        rank(lbf, ties.method = "average", na.last = "keep") / sum(!is.na(lbf))
      }
    )
}
pos_fade_top <- .join_fade(pos_top,    fade_top_df)
pos_fade_bot <- .join_fade(pos_bottom, fade_bot_df)

has_fade_top <- !is.null(pos_fade_top) && any(pos_fade_top$fade_sig, na.rm = TRUE)
has_fade_bot <- !is.null(pos_fade_bot) && any(pos_fade_bot$fade_sig, na.rm = TRUE)
has_fade_any <- has_fade_top || has_fade_bot

# Helper: draw 2-set Venn diagram using base ggplot2 (no extra packages).
# Circle labels are placed outside and above each circle to avoid overlapping
# with the count annotations inside.
.draw_venn2 <- function(n_only_A, n_only_B, n_AB, n_universe,
                         label_A, label_B, title_str) {
  theta <- seq(0, 2 * pi, length.out = 300)
  r <- 1.7; dx <- 1.05
  ciA <- data.frame(x = -dx + r * cos(theta), y = r * sin(theta))
  ciB <- data.frame(x =  dx + r * cos(theta), y = r * sin(theta))
  ggplot() +
    geom_polygon(data = ciA, aes(x, y), fill = .tableau10[1], alpha = 0.22, colour = NA) +
    geom_polygon(data = ciB, aes(x, y), fill = .tableau10[3], alpha = 0.22, colour = NA) +
    geom_path(data = ciA, aes(x, y), colour = .tableau10[1], linewidth = 1.4) +
    geom_path(data = ciB, aes(x, y), colour = .tableau10[3], linewidth = 1.4) +
    annotate("text", x = -2.6, y = 0,
             label = paste0(label_A, " only\n", scales::comma(n_only_A)),
             size = 4.2, colour = .tableau10[1], fontface = "bold") +
    annotate("text", x = 0, y = 0,
             label = paste0("Both\n", scales::comma(n_AB)),
             size = 4.2, colour = "grey20", fontface = "bold") +
    annotate("text", x = 2.6, y = 0,
             label = paste0(label_B, " only\n", scales::comma(n_only_B)),
             size = 4.2, colour = .tableau10[3], fontface = "bold") +
    # Circle titles: pushed to x = -3.5 and +3.5 (outside the overlap zone)
    annotate("text", x = -3.5, y = 2.8, label = label_A, size = 4.5,
             colour = .tableau10[1], fontface = "bold", hjust = 0.5) +
    annotate("text", x =  3.5, y = 2.8, label = label_B, size = 4.5,
             colour = .tableau10[3], fontface = "bold", hjust = 0.5) +
    annotate("text", x = 0, y = -2.6,
             label = sprintf("Universe: %s positions", scales::comma(n_universe)),
             size = 3.5, colour = "grey50") +
    coord_fixed(xlim = c(-5.5, 5.5), ylim = c(-3.2, 3.5)) +
    labs(title = title_str) +
    theme_void() +
    theme(plot.title = element_text(face = "bold", hjust = 0, size = 12, margin = ggplot2::margin(b = 6)))
}

# Helper: compute Venn counts (CAAS dir positions vs full FADE sig set)
.fade_venn_counts <- function(pos_dir, fade_df) {
  if (is.null(pos_dir) || is.null(fade_df)) return(NULL)
  caas_keys     <- pos_dir  %>% select(Gene, Position) %>% distinct()
  fade_sig_keys <- fade_df  %>%
    filter(!is.na(max_site_bf), max_site_bf >= 100) %>%
    select(Gene, Position) %>% distinct()
  n_A   <- nrow(caas_keys)
  n_B   <- nrow(fade_sig_keys)
  n_AB  <- nrow(inner_join(caas_keys, fade_sig_keys, by = c("Gene", "Position")))
  list(n_only_A = n_A - n_AB, n_only_B = n_B - n_AB, n_AB = n_AB,
       n_A = n_A, n_B = n_B, n_universe = n_A + max(n_B - n_AB, 0))
}

vc_top <- .fade_venn_counts(pos_top,    fade_top_df)
vc_bot <- .fade_venn_counts(pos_bottom, fade_bot_df)
```

```{r fade-venn-subsec, results='asis'}
if (has_fade_pos) {
  cat("### Venn Diagrams\n\n")
  if (!has_fade_any)
    cat("> **Note:** No CAAS positions found with FADE BF ≥ 100 in this run.",
        "Venn diagrams shown with zero overlap - all CAAS positions fall below the BF threshold.\n\n")
}
```

```{r fade-venn-plots, eval=has_fade_pos, fig.height=5, fig.width=14}
venn_plots <- list()
if (!is.null(vc_top)) {
  venn_plots[["top"]] <- .draw_venn2(
    vc_top$n_only_A, vc_top$n_only_B, vc_top$n_AB, vc_top$n_universe,
    "CAAS (top)", "FADE BF\u2265100 (top)",
    "TOP direction: CAAS vs FADE-significant positions"
  )
}
if (!is.null(vc_bot)) {
  venn_plots[["bot"]] <- .draw_venn2(
    vc_bot$n_only_A, vc_bot$n_only_B, vc_bot$n_AB, vc_bot$n_universe,
    "CAAS (bottom)", "FADE BF\u2265100 (bottom)",
    "BOTTOM direction: CAAS vs FADE-significant positions"
  )
}
if (length(venn_plots) == 2) {
  gridExtra::grid.arrange(grobs = venn_plots, ncol = 2)
} else if (length(venn_plots) == 1) {
  print(venn_plots[[1]])
}
```

```{r fade-venn-overlap-table, eval=has_fade_pos}
# Table of positions in the intersection (CAAS + FADE BF>=100), by direction
.venn_overlap_positions <- function(pos_dir, fade_df, dir_label, thr10, thr5, thr1) {
  if (is.null(pos_dir) || is.null(fade_df)) return(NULL)
  caas_sel <- if ("caas" %in% names(pos_dir))
    c("Gene", "Position", "CAAS_score", "caas", "change_side") else
    c("Gene", "Position", "CAAS_score", "change_side")
  inner_join(
    pos_dir  %>% select(any_of(caas_sel)),
    fade_df  %>% filter(!is.na(max_site_bf), max_site_bf >= 100) %>%
                 select(Gene, Position, max_site_bf, top_target_aa),
    by = c("Gene", "Position")
  ) %>%
    mutate(
      Direction = dir_label,
      IsTop10   = CAAS_score >= thr10,
      IsTop5    = CAAS_score >= thr5,
      IsTop1    = CAAS_score >= thr1
    ) %>%
    arrange(desc(max_site_bf))
}

overlap_pos_tbl <- bind_rows(
  .venn_overlap_positions(pos_top,    fade_top_df, "TOP",    .thr_top_10, .thr_top_5, .thr_top_1),
  .venn_overlap_positions(pos_bottom, fade_bot_df, "BOTTOM", .thr_bot_10, .thr_bot_5, .thr_bot_1)
)

if (nrow(overlap_pos_tbl) > 0) {
  DT::datatable(
    overlap_pos_tbl %>% mutate(across(where(is.numeric), ~round(., 4))),
    filter   = "top",
    rownames = FALSE,
    options  = list(pageLength = 25, scrollX = TRUE),
    caption  = htmltools::tags$caption(
      style = "caption-side: top; text-align: left; font-weight: bold;",
      "Positions in the CAAS \u2229 FADE (BF \u2265 100) intersection, by direction"
    )
  )
} else {
  cat("> No positions found in the CAAS \u2229 FADE intersection.")
}
```

```{r fade-overlap-subsec, results='asis'}
if (has_fade_pos) cat("### Overlap Table\n\n")
```

```{r fade-overlap-table, eval=has_fade_pos}
.fade_overlap_row <- function(pos_fade_dir, dir_label, pct_val, pct_label) {
  if (is.null(pos_fade_dir)) return(NULL)
  thr  <- quantile(pos_fade_dir$CAAS_score, 1 - pct_val, na.rm = TRUE)
  n_caas <- sum(pos_fade_dir$CAAS_score >= thr, na.rm = TRUE)
  n_sig  <- sum(pos_fade_dir$fade_sig, na.rm = TRUE)
  n_both <- sum(pos_fade_dir$CAAS_score >= thr & pos_fade_dir$fade_sig, na.rm = TRUE)
  tibble(
    Direction           = dir_label,
    `CAAS threshold`    = pct_label,
    `N CAAS positions`  = n_caas,
    `N FADE BF >= 100`  = n_sig,
    `Overlap`           = n_both,
    `% CAAS in FADE`    = sprintf("%.1f%%", 100 * n_both / max(n_caas, 1)),
    `% FADE in CAAS`    = sprintf("%.1f%%", 100 * n_both / max(n_sig,  1))
  )
}

pct_list <- list(
  list(val = params$top_pct,  lbl = sprintf("Top %g%%", params$top_pct  * 100)),
  list(val = 0.05, lbl = sprintf("Top %g%%", 0.05 * 100)),
  list(val = 0.01, lbl = sprintf("Top %g%%", 0.01 * 100))
)

fade_overlap_tbl <- bind_rows(
  purrr::map_dfr(pct_list, ~.fade_overlap_row(pos_fade_top, "TOP",    .x$val, .x$lbl)),
  purrr::map_dfr(pct_list, ~.fade_overlap_row(pos_fade_bot, "BOTTOM", .x$val, .x$lbl))
)

knitr::kable(
  fade_overlap_tbl,
  caption = "Overlap between top-CAAS positions and FADE-significant positions (BF \u2265 100), by direction and CAAS threshold"
)
```

```{r fade-corr-subsec, results='asis'}
if (has_fade_pos) cat("### CAAS vs FADE Correlation\n\n")
```

```{r fade-caas-corr, eval=has_fade_any, fig.height=5}
# Combine top and bottom joined datasets; plot CAAS_score vs log10(BF)
# with rank-normalised axes so Pearson r is not dominated by BF outliers.
.fade_corr_df <- bind_rows(
  if (!is.null(pos_fade_top)) pos_fade_top %>% mutate(fade_direction = "TOP")    else NULL,
  if (!is.null(pos_fade_bot)) pos_fade_bot %>% mutate(fade_direction = "BOTTOM") else NULL
) %>%
  filter(!is.na(max_site_bf)) %>%
  mutate(log10_bf = log10(pmax(max_site_bf, 0.01)))

.corr_label <- .fade_corr_df %>%
  group_by(fade_direction) %>%
  summarise(
    r_pearson  = safe_cor(norm_caas, norm_log_bf),
    r_spearman = safe_cor(CAAS_score, log10_bf, method = "spearman"),
    .groups = "drop"
  ) %>%
  mutate(label = sprintf("%s: r=%.3f (rank-norm Pearson) / \u03c1=%.3f (Spearman)",
                         fade_direction, r_pearson, r_spearman))

ggplot(.fade_corr_df, aes(x = CAAS_score, y = log10_bf, colour = fade_direction)) +
  geom_point(alpha = 0.35, size = 0.9) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8, alpha = 0.12) +
  geom_hline(yintercept = 2, linetype = "dashed", colour = "grey50", linewidth = 0.5) +
  scale_colour_manual(values = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[3]),
                      name = "Direction") +
  annotate("text", x = -Inf, y = 2.05, hjust = -0.05, vjust = 0,
           label = "BF = 100", size = 3, colour = "grey40") +
  labs(
    title    = "CAAS Score vs FADE max Bayes Factor",
    subtitle = paste(.corr_label$label, collapse = "\n"),
    x = "CAAS Score", y = "log10(max site BF)"
  ) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(size = 9, colour = "grey30"))
```

---

## Gene-Level Characterisation {.tabset}

Directional gene sets (TOP and BOTTOM) are characterised against independent
evolutionary signals. All correlations use rank-normalised variables.

### Gene CAAS Score

```{r gene-caas-dir-dist, fig.height=5}
# Directional thresholds
g_top_thr_10 <- quantile(gene_df$gene_caas_top, 1 - params$gene_top_pct, na.rm = TRUE)
g_top_thr_5  <- quantile(gene_df$gene_caas_top, 1 - 0.05,                na.rm = TRUE)
g_top_thr_1  <- quantile(gene_df$gene_caas_top, 1 - 0.01,                na.rm = TRUE)
g_bot_thr_10 <- quantile(gene_df$gene_caas_bot, 1 - params$gene_top_pct, na.rm = TRUE)
g_bot_thr_5  <- quantile(gene_df$gene_caas_bot, 1 - 0.05,                na.rm = TRUE)
g_bot_thr_1  <- quantile(gene_df$gene_caas_bot, 1 - 0.01,                na.rm = TRUE)
g_thr_10 <- quantile(gene_df$gene_caas_score, 1 - params$gene_top_pct, na.rm = TRUE)

long_dir_scores <- bind_rows(
  gene_df %>% filter(!is.na(gene_caas_top)) %>%
    transmute(Gene, score = gene_caas_top, direction = "TOP"),
  gene_df %>% filter(!is.na(gene_caas_bot)) %>%
    transmute(Gene, score = gene_caas_bot, direction = "BOTTOM")
)

ggplot(long_dir_scores, aes(x = score, fill = direction, colour = direction)) +
  geom_density(alpha = 0.45, linewidth = 0.7) +
  geom_vline(xintercept = g_top_thr_10, linetype = "dashed",
             colour = .tableau10[1], linewidth = 0.6) +
  geom_vline(xintercept = g_bot_thr_10, linetype = "dashed",
             colour = .tableau10[2], linewidth = 0.6) +
  scale_fill_manual(values   = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[2])) +
  scale_colour_manual(values = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[2])) +
  labs(title = "Directional Gene CAAS Score Distribution",
       subtitle = sprintf("size-adjusted max of directional positions | dashed = top %g%% threshold per direction",
                          params$gene_top_pct * 100),
       x = "Gene CAAS Score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### RERConverge

`r if (!has_rer) "> **Note:** RERConverge results not available for this run."`

> Each panel below compares a CAAS gene gate (left circle) against an
> RERConverge-significant set (right circle). The **shared region** is the
> count of RERConverge-significant genes that are *also* members of the CAAS
> gate -- i.e. how much of the alternative (RER) evidence is corroborated by
> CAAS. **OR** is a Fisher exact test (equivalent to a hypergeometric test on
> the 2x2 table) against the full scored-gene universe (n = `r nrow(gene_df)`
> genes); OR > 1 means the overlap is larger than expected by chance.

```{r gene-rer-venn, eval=has_rer, fig.height=14, fig.width=12}
rer_accel_sig <- gene_df %>%
  filter(!is.na(rer_min_pval), rer_min_pval <= 0.05,
         !is.na(rer_rho), rer_rho > 0) %>% pull(Gene)
rer_decel_sig <- gene_df %>%
  filter(!is.na(rer_min_pval), rer_min_pval <= 0.05,
         !is.na(rer_rho), rer_rho < 0) %>% pull(Gene)

# 3 rows (gate) x 4 columns (TOP/BOT x Accel/Decel)
rer_venn_plots <- list()
for (g in .caas_gates) {
  rer_venn_plots <- c(rer_venn_plots, list(
    .draw_venn(.venn_counts(g$top, rer_accel_sig, sprintf("CAAS TOP\n%s", g$lbl), "RER Accel.", gene_universe_n),
               .tableau10[1], .tableau10[3], sprintf("%s | TOP vs Accel.", g$lbl)),
    .draw_venn(.venn_counts(g$top, rer_decel_sig, sprintf("CAAS TOP\n%s", g$lbl), "RER Decel.", gene_universe_n),
               .tableau10[1], .tableau10[4], sprintf("%s | TOP vs Decel.", g$lbl)),
    .draw_venn(.venn_counts(g$bot, rer_accel_sig, sprintf("CAAS BOT\n%s", g$lbl), "RER Accel.", gene_universe_n),
               .tableau10[2], .tableau10[3], sprintf("%s | BOT vs Accel.", g$lbl)),
    .draw_venn(.venn_counts(g$bot, rer_decel_sig, sprintf("CAAS BOT\n%s", g$lbl), "RER Decel.", gene_universe_n),
               .tableau10[2], .tableau10[4], sprintf("%s | BOT vs Decel.", g$lbl))
  ))
}
gridExtra::grid.arrange(
  grobs = rer_venn_plots, ncol = 4,
  top = grid::textGrob(
    "RERConverge Gene Overlap \u2014 CAAS gates (rows) \u00d7 direction\u00d7RER (cols) | RER p \u2264 0.05",
    gp = grid::gpar(fontface = "bold", fontsize = 12)))
```

```{r gene-rer-table, eval=has_rer}
rer_tbl <- gene_df %>%
  filter(!is.na(rer_min_pval)) %>%
  mutate(
    rer_direction = case_when(rer_rho > 0 ~ "accelerating",
                               rer_rho < 0 ~ "decelerating",
                               TRUE        ~ "neutral"),
    in_top = Gene %in% gene_top_list,
    in_bot = Gene %in% gene_bot_list
  ) %>%
  arrange(rer_min_pval) %>%
  select(Gene, rer_min_pval, rer_rho, rer_direction,
         gene_caas_score, gene_caas_top, gene_caas_bot,
         in_top, in_bot) %>%
  mutate(across(where(is.numeric), ~round(., 4)))

DT::datatable(
  rer_tbl, filter="top", rownames=FALSE,
  options = list(pageLength=20, scrollX=TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "RERConverge results ranked by p-value - direction (rho sign) and CAAS gene list membership"
  )
)
```

### FADE Gene

`r if (!has_fade_gene) "> **Note:** No gene-level FADE data available for this run (neither fade_summary_top/bottom.tsv nor site-level FADE files were supplied)."`

```{r gene-fade-setup, eval=has_fade_gene}
# Prefer aggregating the site-level (position) FADE data when it's available --
# it's the same information scoring_compute.R already rolled up into gene_df's
# fade_max_bf_top/fade_significant_top/fade_max_bf_bottom/fade_significant_bottom,
# but keeping this path lets the Manhattan/venn plots below reuse fade_top_df/
# fade_bot_df where they exist. When the site-level files are absent or broken
# (has_fade_pos == FALSE, e.g. a dangling symlink to a renamed results dir),
# fall back directly to gene_df's own columns -- those are computed
# independently from fade_summary_top/bottom.tsv and don't depend on the
# site-level files at all, so this section still renders correctly.
if (has_fade_pos && exists("fade_top_df") && !is.null(fade_top_df) && nrow(fade_top_df) > 0) {
  gene_fade_top_agg <- fade_top_df %>%
    group_by(Gene) %>%
    summarise(fade_top_sig    = any(max_site_bf >= 100, na.rm=TRUE),
              fade_top_max_bf = max(max_site_bf, na.rm=TRUE), .groups="drop")
} else {
  gene_fade_top_agg <- tibble(Gene=character(), fade_top_sig=logical(), fade_top_max_bf=numeric())
}
if (has_fade_pos && exists("fade_bot_df") && !is.null(fade_bot_df) && nrow(fade_bot_df) > 0) {
  gene_fade_bot_agg <- fade_bot_df %>%
    group_by(Gene) %>%
    summarise(fade_bot_sig    = any(max_site_bf >= 100, na.rm=TRUE),
              fade_bot_max_bf = max(max_site_bf, na.rm=TRUE), .groups="drop")
} else {
  gene_fade_bot_agg <- tibble(Gene=character(), fade_bot_sig=logical(), fade_bot_max_bf=numeric())
}
# Existence-safe column pull -- gene_df may have only the top or only the
# bottom gene-level FADE columns (fade_summary_top.tsv and fade_summary_bottom.tsv
# are supplied independently), so an unconditional gene_df$col reference isn't safe.
.gcol_fade <- function(df, col) if (col %in% names(df)) df[[col]] else rep(NA, nrow(df))

gene_fade_df <- gene_df %>%
  select(Gene, gene_caas_score, gene_caas_top, gene_caas_bot, n_positions) %>%
  mutate(
    .gene_max_bf_top    = as.numeric(.gcol_fade(gene_df, "fade_max_bf_top")),
    .gene_max_bf_bottom = as.numeric(.gcol_fade(gene_df, "fade_max_bf_bottom")),
    # gene_scores.tsv round-trips logical columns as "TRUE"/"FALSE" text, which
    # readr already parses back into logical -- no extra coercion needed here.
    .gene_sig_top       = as.logical(.gcol_fade(gene_df, "fade_significant_top")),
    .gene_sig_bottom    = as.logical(.gcol_fade(gene_df, "fade_significant_bottom"))
  ) %>%
  left_join(gene_fade_top_agg, by="Gene") %>%
  left_join(gene_fade_bot_agg, by="Gene") %>%
  mutate(
    # Site-level aggregate wins when present; otherwise fall back to gene_df's
    # own gene-level FADE columns (from fade_summary_top/bottom.tsv).
    fade_top_max_bf = dplyr::coalesce(fade_top_max_bf, .gene_max_bf_top),
    fade_bot_max_bf = dplyr::coalesce(fade_bot_max_bf, .gene_max_bf_bottom),
    fade_top_sig    = dplyr::coalesce(fade_top_sig, .gene_sig_top),
    fade_bot_sig    = dplyr::coalesce(fade_bot_sig, .gene_sig_bottom),
    fade_top_sig    = replace_na(fade_top_sig, FALSE),
    fade_bot_sig    = replace_na(fade_bot_sig, FALSE)
  ) %>%
  select(-.gene_max_bf_top, -.gene_max_bf_bottom, -.gene_sig_top, -.gene_sig_bottom)
has_gene_fade_any <- any(gene_fade_df$fade_top_sig) || any(gene_fade_df$fade_bot_sig)
```

> Shared region = FADE-significant genes that are also in the CAAS gate.
> OR = Fisher exact test vs. the full scored-gene universe (n = `r nrow(gene_df)` genes).

```{r gene-fade-venn, eval=has_fade_gene && exists("has_gene_fade_any") && has_gene_fade_any, fig.height=11, fig.width=12}
fade_top_genes <- gene_fade_df %>% filter(fade_top_sig) %>% pull(Gene)
fade_bot_genes <- gene_fade_df %>% filter(fade_bot_sig) %>% pull(Gene)
fade_venn_plots <- list()
for (g in .caas_gates) {
  fade_venn_plots <- c(fade_venn_plots, list(
    .draw_venn(.venn_counts(g$top, fade_top_genes, sprintf("CAAS TOP\n%s", g$lbl), "FADE TOP sig", gene_universe_n),
               .tableau10[1], .tableau10[5], sprintf("%s | TOP vs FADE TOP (BF\u2265100)", g$lbl)),
    .draw_venn(.venn_counts(g$bot, fade_bot_genes, sprintf("CAAS BOT\n%s", g$lbl), "FADE BOT sig", gene_universe_n),
               .tableau10[2], .tableau10[6], sprintf("%s | BOT vs FADE BOT (BF\u2265100)", g$lbl))
  ))
}
gridExtra::grid.arrange(
  grobs = fade_venn_plots, ncol = 2,
  top = grid::textGrob("FADE Gene Overlap \u2014 CAAS gates (rows) \u00d7 direction (cols) | BF \u2265 100",
                        gp = grid::gpar(fontface = "bold", fontsize = 12)))
```

```{r gene-fade-correlation, eval=has_fade_gene && exists("has_gene_fade_any") && has_gene_fade_any, fig.height=5, fig.width=12}
fade_cor_top <- gene_fade_df %>%
  filter(!is.na(gene_caas_top), !is.na(fade_top_max_bf), is.finite(fade_top_max_bf)) %>%
  mutate(x_norm=rank_norm(gene_caas_top), y_norm=rank_norm(fade_top_max_bf))
fade_cor_bot <- gene_fade_df %>%
  filter(!is.na(gene_caas_bot), !is.na(fade_bot_max_bf), is.finite(fade_bot_max_bf)) %>%
  mutate(x_norm=rank_norm(gene_caas_bot), y_norm=rank_norm(fade_bot_max_bf))

.fade_scatter <- function(df, col, title, xlab, ylab) {
  ggplot(df, aes(x=x_norm, y=y_norm)) +
    geom_point(alpha=0.5, size=1.2, colour=col) +
    geom_smooth(method="lm", se=TRUE, colour="grey30", linewidth=0.7) +
    ggrepel::geom_text_repel(
      data = df %>% filter(y_norm > quantile(y_norm, 0.95, na.rm=TRUE)),
      aes(label=Gene), size=2.5, max.overlaps=10, show.legend=FALSE
    ) +
    labs(title=title,
         subtitle=sprintf("r = %.3f", safe_cor(df$x_norm, df$y_norm)),
         x=xlab, y=ylab) +
    theme(plot.title = element_text(face="bold", size=11))
}
gridExtra::grid.arrange(
  .fade_scatter(fade_cor_top, .tableau10[1],
                "TOP: CAAS vs FADE BF (rank-norm)",
                "gene_caas_top (rank-norm)", "FADE top max BF (rank-norm)"),
  .fade_scatter(fade_cor_bot, .tableau10[2],
                "BOTTOM: CAAS vs FADE BF (rank-norm)",
                "gene_caas_bot (rank-norm)", "FADE bot max BF (rank-norm)"),
  ncol=2,
  top=grid::textGrob("Gene CAAS Score vs FADE BF - rank-normalised",
                      gp=grid::gpar(fontface="bold", fontsize=13)))
```

```{r gene-fade-table, eval=has_fade_gene}
fade_tbl <- gene_fade_df %>%
  mutate(
    in_top = Gene %in% gene_top_list,
    in_bot = Gene %in% gene_bot_list
  ) %>%
  arrange(desc(pmax(replace_na(fade_top_max_bf, 0), replace_na(fade_bot_max_bf, 0)))) %>%
  select(Gene,
         fade_top_max_bf, fade_top_sig,
         fade_bot_max_bf, fade_bot_sig,
         gene_caas_score, gene_caas_top, gene_caas_bot,
         in_top, in_bot) %>%
  mutate(across(where(is.numeric), ~round(., 4)))

DT::datatable(
  fade_tbl, filter = "top", rownames = FALSE,
  options = list(pageLength = 20, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "FADE gene-level results ranked by max BF - direction and CAAS gene list membership"
  )
)
```

### Accumulation

`r if (!has_accum_score) "> **Note:** Accumulation scores not available for this run."`

> Shared region = Accumulation-significant genes that are also in the CAAS gate.
> OR = Fisher exact test vs. the full scored-gene universe (n = `r nrow(gene_df)` genes).

```{r gene-accum-venn, eval=has_accum_score, fig.height=11, fig.width=12}
accum_sig_genes <- gene_df %>%
  filter(!is.na(accum_cct_p), accum_significant %in% TRUE) %>% pull(Gene)

accum_venn_plots <- list()
accum_overlap_exports <- list()
for (g in .caas_gates) {
  for (dir_lbl in c("TOP", "BOTTOM")) {
    caas_genes <- if (dir_lbl == "TOP") g$top else g$bot
    col_a      <- if (dir_lbl == "TOP") .tableau10[1] else .tableau10[2]
    dir_slug   <- tolower(dir_lbl)
    gate_slug  <- gsub("[^a-z0-9]", "_", tolower(g$lbl))
    overlap_genes <- intersect(caas_genes, accum_sig_genes)
    accum_overlap_exports[[sprintf("gene_caas_%s_%s_x_accum.txt", dir_slug, gate_slug)]] <- overlap_genes
    vc <- .venn_counts(caas_genes, accum_sig_genes,
                       sprintf("CAAS %s\n%s", dir_lbl, g$lbl), "Accum. sig", gene_universe_n)
    accum_venn_plots[[length(accum_venn_plots) + 1]] <-
      .draw_venn(vc, col_a, .tableau10[5], sprintf("%s | %s vs Accum.", g$lbl, dir_lbl))
  }
}

gridExtra::grid.arrange(
  grobs = accum_venn_plots, ncol = 2,
  top = grid::textGrob(
    "Accumulation Gene Overlap - CAAS gates (rows) × direction (cols) | accum_significant",
    gp = grid::gpar(fontface = "bold", fontsize = 12)))
```

```{r gene-accum-table, eval=has_accum_score}
accum_pval_cols <- grep("^accum_pval_", names(gene_df), value = TRUE)
accum_tbl <- gene_df %>%
  filter(!is.na(accum_cct_p)) %>%
  mutate(
    in_top = Gene %in% gene_top_list,
    in_bot = Gene %in% gene_bot_list
  ) %>%
  arrange(accum_cct_p) %>%
  select(Gene, accum_cct_p, any_of("accum_fdr"), accum_significant, any_of(accum_pval_cols),
         gene_caas_score, gene_caas_top, gene_caas_bot,
         in_top, in_bot) %>%
  mutate(across(where(is.numeric), ~round(., 4)))

DT::datatable(
  accum_tbl, filter = "top", rownames = FALSE,
  options = list(pageLength = 20, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "Accumulation genes ranked by CCT combined p-value (accum_cct_p) - per-scheme p-values and CAAS membership"
  )
)
```

### Top Genes

`gene_caas_pperm` is the CAAS permulation-excess null's per-gene right-tailed
p-value (fraction of permuted-labeling cycles whose genome-wide gene score
matches or exceeds this gene's observed `gene_caas_score`), `gene_caas_pperm_adj`
its BH adjustment across genes. At the pipeline's typical
`caas_full_perms` this floors around 1/(N+1) (~1e-3 at N=1000), so treat it as a
**prioritisation** signal with FDR context, not by itself a genome-wide-significant
call — a gene at the floor is "no cycle beat it", which does not distinguish
between genes that all cleared every cycle. Blank when no CAAS permulation null
was supplied to this run.

```{r top-genes-table}
top_genes_df <- gene_df %>%
  arrange(desc(gene_caas_score))

# ── Direction assignment ──────────────────────────────────────────────────────
if (has_rer) {
  top_genes_df <- top_genes_df %>%
    mutate(rer_direction = case_when(
      !is.na(rer_rho) & rer_rho > 0 ~ "accelerating",
      !is.na(rer_rho) & rer_rho < 0 ~ "decelerating",
      TRUE ~ NA_character_
    ))
}

# ── FADE: max BF and BF>=100 significance flag per direction ─────────────────
if (has_fade_gene && exists("gene_fade_df")) {
  top_genes_df <- top_genes_df %>%
    left_join(
      gene_fade_df %>% select(Gene, fade_top_max_bf, fade_bot_max_bf,
                               fade_top_sig, fade_bot_sig),
      by = "Gene"
    )
}

# ── Accumulation score ────────────────────────────────────────────────────────
# accum_cct_p already on top_genes_df; no extra join needed.

# ── CAAS-threshold indicators (10%, 50% and 100%) per direction ──────────────
g_top_thr_50 <- quantile(gene_df$gene_caas_top, 0.50, na.rm = TRUE)
g_bot_thr_50 <- quantile(gene_df$gene_caas_bot, 0.50, na.rm = TRUE)

top_genes_df <- top_genes_df %>%
  mutate(
    in_top_10pct = replace_na(gene_caas_top >= g_top_thr_10, FALSE),
    in_top_50pct = replace_na(gene_caas_top >= g_top_thr_50, FALSE),
    in_bot_10pct = replace_na(gene_caas_bot >= g_bot_thr_10, FALSE),
    in_bot_50pct = replace_na(gene_caas_bot >= g_bot_thr_50, FALSE)
  )

# ── Build display column list ─────────────────────────────────────────────────
gene_display_cols <- c("Gene", "n_positions",
                       "gene_caas_score", "gene_caas_top", "gene_caas_bot",
                       "in_top_10pct", "in_top_50pct",
                       "in_bot_10pct", "in_bot_50pct",
                       "gene_caas_pperm", "gene_caas_pperm_adj")
if (has_accum_score)
  gene_display_cols <- c(gene_display_cols, "accum_cct_p")
if (has_rer)
  gene_display_cols <- c(gene_display_cols, "rer_min_pval", "rer_direction")
if (has_fade_gene && "fade_top_max_bf" %in% names(top_genes_df))
  gene_display_cols <- c(gene_display_cols,
                         "fade_top_max_bf", "fade_top_sig",
                         "fade_bot_max_bf", "fade_bot_sig")

DT::datatable(
  top_genes_df %>%
    select(any_of(gene_display_cols)) %>%
    mutate(across(where(is.numeric) & !matches("n_positions"), ~round(., 4))),
  filter = "top", rownames = FALSE,
  options = list(pageLength = 25, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Gene-level scores | Trait: %s | sorted by gene_caas_score", params$traitname)
  )
)
```

---

## Genomic Accumulation {.tabset}

`r if (!has_genomic) "> **Note:** Genomic coordinates (gene_ensembl_file) are not available for this run - genomic accumulation skipped."`

```{r genomic-setup, eval=has_genomic}
# Join gene_df with genomic coordinates.
gene_genomic <- gene_df %>%
  select(Gene, gene_caas_score, gene_caas_top, gene_caas_bot, n_positions) %>%
  inner_join(genomic_raw, by = "Gene") %>%
  # Standardise chromosome labels (remove "chr" prefix if present)
  mutate(
    chr_label = sub("^chr", "", chr),
    # Order chromosomes numerically then X, Y, MT
    chr_num   = suppressWarnings(as.integer(chr_label)),
    chr_order = dplyr::case_when(
      !is.na(chr_num)       ~ chr_num,
      chr_label == "X"      ~ 23L,
      chr_label == "Y"      ~ 24L,
      chr_label %in% c("MT","M") ~ 25L,
      TRUE                  ~ 26L
    )
  ) %>%
  arrange(chr_order, gene_mid)

# Build a cumulative x-axis offset per chromosome for a stitched Manhattan view
chr_offsets <- gene_genomic %>%
  group_by(chr_order, chr_label) %>%
  summarise(chr_max = max(gene_end, na.rm = TRUE), .groups = "drop") %>%
  arrange(chr_order) %>%
  mutate(offset = cumsum(lag(chr_max, default = 0L)) + (chr_order - 1L) * 5e6L)

gene_genomic <- gene_genomic %>%
  left_join(chr_offsets %>% select(chr_order, offset, chr_label), by = c("chr_order", "chr_label")) %>%
  mutate(manhattan_x = gene_mid + offset)

chr_label_positions <- chr_offsets %>%
  left_join(gene_genomic %>% group_by(chr_order) %>%
              summarise(mid_x = mean(manhattan_x, na.rm = TRUE), .groups = "drop"),
            by = "chr_order")
```

```{r window-helper, eval=has_genomic}
# Helper: tile chromosomes into non-overlapping windows of win_size bp.
# Each gene/position is assigned to the window containing its genomic midpoint.
# Returns a dataframe with one row per window, summarising mean and max score.
make_window_accum <- function(df, score_col,
                               win_size = params$window_size_bp) {
  df %>%
    mutate(
      win_id    = floor(gene_mid / win_size),
      win_start = win_id * win_size,
      win_end   = win_start + win_size - 1L,
      win_mid   = win_start + win_size / 2L
    ) %>%
    group_by(chr_order, chr_label, win_id, win_start, win_end, win_mid) %>%
    summarise(
      n_genes    = dplyr::n(),
      mean_score = mean(.data[[score_col]], na.rm = TRUE),
      max_score  = max(.data[[score_col]],  na.rm = TRUE),
      genes      = paste(unique(Gene), collapse = ", "),
      .groups    = "drop"
    )
}

# Stitch window midpoints onto the cumulative chromosome x-axis (same offsets
# as gene_genomic so windows align with the per-gene manhattans above).
stitch_windows <- function(win_df) {
  win_df %>%
    left_join(chr_offsets %>% select(chr_order, offset), by = "chr_order") %>%
    mutate(win_x = win_mid + offset,
           chr_even = chr_order %% 2 == 0)
}
```

### CAAS Score Manhattan

```{r genomic-manhattan-caas, eval=has_genomic, fig.height=5, fig.width=14}
gene_genomic <- gene_genomic %>%
  mutate(
    chr_even      = chr_order %% 2 == 0,
    caas_log      = -log10(1 - pmin(gene_caas_score, 0.9999)),
    caas_top_log  = -log10(1 - pmin(replace_na(gene_caas_top, 0), 0.9999)),
    caas_bot_log  = -log10(1 - pmin(replace_na(gene_caas_bot, 0), 0.9999))
  )
caas_log_thr <- -log10(1 - pmin(g_thr_10, 0.9999))

ggplot(gene_genomic,
       aes(x = manhattan_x, y = caas_log,
           colour = chr_even, size = n_positions)) +
  geom_point(alpha = 0.65, show.legend = FALSE) +
  geom_hline(yintercept = caas_log_thr, linetype = "dashed",
             colour = "grey40", linewidth = 0.6) +
  ggrepel::geom_text_repel(
    data = gene_genomic %>% slice_max(caas_log, n = 20, with_ties = FALSE),
    aes(label = Gene), size = 2.2, max.overlaps = 15,
    show.legend = FALSE, colour = "black"
  ) +
  scale_colour_manual(values = c("FALSE"=.tableau10[1], "TRUE"=.tableau10[2]), guide="none") +
  scale_size_continuous(name = "N positions", range = c(0.5, 5)) +
  scale_x_continuous(
    breaks = chr_label_positions$mid_x,
    labels = chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  labs(
    title    = "Gene CAAS Score - Manhattan View",
    subtitle = sprintf("%d genes | y = -log10(1 - gene_caas_score) | dashed = top %g%% threshold",
                       nrow(gene_genomic), params$gene_top_pct * 100),
    x = "Chromosome", y = "-log10(1 - gene_caas_score)"
  ) +
  theme(plot.title=element_text(face="bold"),
        axis.text.x=element_text(size=14), axis.text.y=element_text(size=14),
        axis.title=element_text(size=14), legend.text=element_text(size=12),
        panel.grid.major.x=element_blank())
```

### CAAS Window Accumulation (Positions)

`r if (!has_genomic) "> **Note:** Genomic coordinates not available - position window accumulation skipped."`

```{r pos-window-accum, eval=has_genomic, fig.height=5, fig.width=14}
# Tile position-level CAAS scores into genomic windows.
# pos_genomic is built in the pos-caas-manhattan-setup chunk above.
# We use the same pos_chr_offsets to align x-coordinates with the gene-level plots.
win_size_label <- if (params$window_size_bp >= 1e6) {
  sprintf("%g Mb", params$window_size_bp / 1e6)
} else {
  sprintf("%g kb", params$window_size_bp / 1e3)
}

pos_wins <- pos_genomic %>%
  rename(Gene = Gene) %>%
  make_window_accum("CAAS_score", win_size = params$window_size_bp) %>%
  left_join(pos_chr_offsets %>% select(chr_order, offset), by = "chr_order") %>%
  mutate(n_positions = n_genes,
         win_x       = win_mid + offset,
         chr_even    = chr_order %% 2 == 0,
         mean_log    = -log10(1 - pmin(mean_score, 0.9999)),
         max_log     = -log10(1 - pmin(max_score,  0.9999)))

# Label the top windows as genomic regions, not gene names.
# Ranked by n_positions (window density) rather than mean_log, to surface the
# regions with the most positions rather than the highest mean score.
top_pos_wins <- pos_wins %>%
  slice_max(n_positions, n = 5, with_ties = FALSE) %>%
  mutate(region_label = sprintf("chr%s:%0.1f-%0.1f Mb", chr_label, win_start / 1e6, win_end / 1e6))

ggplot(pos_wins, aes(x = win_x, y = mean_log, fill = chr_even)) +
  geom_col(width = params$window_size_bp * 0.9, alpha = 0.75) +
  geom_point(aes(y = max_log, colour = chr_even), size = 1.2, alpha = 0.85, show.legend = FALSE) +
  ggrepel::geom_text_repel(
    data = top_pos_wins,
    aes(y = max_log, label = sprintf("%s\n(%d positions)", region_label, n_positions)),
    size = 2.2, max.overlaps = 10, colour = "black",
    show.legend = FALSE
  ) +
  scale_fill_manual(
    values = c("FALSE" = .tableau10[1], "TRUE" = .tableau10[2]),
    guide = "none"
  ) +
  scale_colour_manual(
    values = c("FALSE" = .tableau10[3], "TRUE" = .tableau10[4]),
    guide = "none"
  ) +
  scale_x_continuous(
    breaks = pos_chr_label_positions$mid_x,
    labels = pos_chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  labs(
    title    = sprintf("Position-Level CAAS Score - %s Window Accumulation", win_size_label),
    subtitle = sprintf("Bars = mean | Points = max | %d windows across %d chromosomes",
                       nrow(pos_wins), length(unique(pos_wins$chr_order))),
    x = "Chromosome", y = "-log10(1 - mean CAAS score)"
  ) +
  theme(
    plot.title         = element_text(face = "bold"),
    axis.text.x        = element_text(size = 14),
    axis.text.y        = element_text(size = 14),
    axis.title         = element_text(size = 14),
    legend.text        = element_text(size = 12),
    panel.grid.major.x = element_blank()
  )
```

```{r pos-window-accum-table, eval=has_genomic && exists("pos_wins") && nrow(pos_wins) > 0}
DT::datatable(
  pos_wins %>%
    select(chr_label, win_start, win_end, n_positions, mean_score, max_score, mean_log) %>%
    mutate(
      win_start  = format(win_start, big.mark = ",", scientific = FALSE),
      win_end    = format(win_end,   big.mark = ",", scientific = FALSE),
      mean_score = round(mean_score, 4),
      max_score  = round(max_score,  4),
      mean_log   = round(mean_log,   4)
    ) %>%
    arrange(desc(n_positions)) %>%
    rename(chr = chr_label, start = win_start, end = win_end,
           n_pos = n_positions, mean_CAAS = mean_score, max_CAAS = max_score,
           `-log10(1-mean)` = mean_log),
  rownames = FALSE,
  options = list(pageLength = 15, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Top genomic windows by position-level CAAS accumulation (%s windows)", win_size_label)
  )
)
```

### CAAS Window Accumulation (Genes)

`r if (!has_genomic) "> **Note:** Genomic coordinates not available - gene-level CAAS window accumulation skipped."`

```{r gene-caas-window-accum, eval=has_genomic, fig.height=5, fig.width=14}
gene_caas_wins <- gene_genomic %>%
  make_window_accum("gene_caas_score", win_size = params$window_size_bp) %>%
  stitch_windows() %>%
  mutate(mean_log = -log10(1 - pmin(mean_score, 0.9999)),
         max_log  = -log10(1 - pmin(max_score,  0.9999)))

# Ranked by n_genes (window density) rather than mean_log, to surface the
# regions with the most genes rather than the highest mean score.
top_caas_wins <- gene_caas_wins %>%
  slice_max(n_genes, n = 5, with_ties = FALSE) %>%
  mutate(region_label = sprintf("chr%s:%0.1f-%0.1f Mb", chr_label, win_start / 1e6, win_end / 1e6))

ggplot(gene_caas_wins, aes(x = win_x, y = mean_log, fill = chr_even)) +
  geom_col(width = params$window_size_bp * 0.9, alpha = 0.75) +
  geom_point(aes(y = max_log, colour = chr_even), size = 1.5, alpha = 0.85, show.legend = FALSE) +
  ggrepel::geom_text_repel(
    data = top_caas_wins,
    aes(y = max_log, label = sprintf("%s\n(%d genes)", region_label, n_genes)),
    size = 2.2, max.overlaps = 10, colour = "black",
    show.legend = FALSE
  ) +
  scale_fill_manual(
    values = c("FALSE" = .tableau10[1], "TRUE" = .tableau10[2]),
    guide = "none"
  ) +
  scale_colour_manual(
    values = c("FALSE" = .tableau10[3], "TRUE" = .tableau10[4]),
    guide = "none"
  ) +
  scale_x_continuous(
    breaks = chr_label_positions$mid_x,
    labels = chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  labs(
    title    = sprintf("Gene CAAS Score - %s Window Accumulation", win_size_label),
    subtitle = sprintf("Bars = mean | Points = max | %d windows", nrow(gene_caas_wins)),
    x = "Chromosome", y = "-log10(1 - mean gene_caas_score)"
  ) +
  theme(
    plot.title         = element_text(face = "bold"),
    axis.text.x        = element_text(size = 14),
    axis.text.y        = element_text(size = 14),
    axis.title         = element_text(size = 14),
    legend.text        = element_text(size = 12),
    panel.grid.major.x = element_blank()
  )
```

```{r gene-caas-window-accum-table, eval=has_genomic && exists("gene_caas_wins") && nrow(gene_caas_wins) > 0}
DT::datatable(
  gene_caas_wins %>%
    select(chr_label, win_start, win_end, n_genes, mean_score, max_score, mean_log) %>%
    mutate(
      win_start  = format(win_start, big.mark = ",", scientific = FALSE),
      win_end    = format(win_end,   big.mark = ",", scientific = FALSE),
      mean_score = round(mean_score, 4),
      max_score  = round(max_score,  4),
      mean_log   = round(mean_log,   4)
    ) %>%
    arrange(desc(n_genes)) %>%
    rename(chr = chr_label, start = win_start, end = win_end,
           n_genes = n_genes, mean_CAAS = mean_score, max_CAAS = max_score,
           `-log10(1-mean)` = mean_log),
  rownames = FALSE,
  options = list(pageLength = 15, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Top genomic windows by gene CAAS score accumulation (%s windows)", win_size_label)
  )
)
```

```{r genomic-window-tsv-export, eval=has_genomic}
if (exists("pos_wins") && nrow(pos_wins) > 0) {
  readr::write_tsv(
    pos_wins %>% select(-chr_even, -mean_log, -max_log),
    file.path(params$output_dir, "genomic_window_accum_positions.tsv"))
}
if (exists("gene_caas_wins") && nrow(gene_caas_wins) > 0) {
  readr::write_tsv(
    gene_caas_wins %>% select(-chr_even, -mean_log, -max_log),
    file.path(params$output_dir, "genomic_window_accum_gene_caas.tsv"))
}
cat("Exported genomic window accumulation TSVs\n")
```

---

## Supplementary Exports

```{r supplementary-tsv-export}
# Build a wide supplementary table joining position_scores with FADE data.
# Exported as position_scores_annotated.tsv alongside the report outputs.

supp_df <- pos_df

# Join FADE data: use the direction-matched FADE table for each position
if (has_fade_pos) {
  fade_combined <- bind_rows(
    if (!is.null(fade_top_df)) fade_top_df %>% mutate(fade_direction = "TOP")    else NULL,
    if (!is.null(fade_bot_df)) fade_bot_df %>% mutate(fade_direction = "BOTTOM") else NULL
  )
  if (nrow(fade_combined) > 0) {
    # For positions in both directions (change_side == "both"), take TOP FADE by priority
    fade_for_join <- fade_combined %>%
      mutate(fade_direction_priority = if_else(fade_direction == "TOP", 1L, 2L)) %>%
      group_by(Gene, Position) %>%
      slice_min(fade_direction_priority, n = 1, with_ties = FALSE) %>%
      ungroup() %>%
      select(Gene, Position,
             fade_max_site_bf     = max_site_bf,
             fade_top_target_aa   = top_target_aa,
             fade_biased,
             fade_direction)
    supp_df <- supp_df %>% left_join(fade_for_join, by = c("Gene", "Position"))
  }
}

readr::write_tsv(supp_df,
  file.path(params$output_dir, "position_scores_annotated.tsv"))

cat(sprintf("Exported position_scores_annotated.tsv: %d rows, %d columns\n",
            nrow(supp_df), ncol(supp_df)))
cat(sprintf("Columns: %s\n", paste(names(supp_df), collapse = ", ")))
```

---

## Summary

```{r summary-stats}
n_components <- 5L  # variability, pattern, asr, convergence, parallel

n_gene_components <- 1L  # gene_caas always present
if (has_accum_score) n_gene_components <- n_gene_components + 1L
if (has_rer)         n_gene_components <- n_gene_components + 1L

summary_tbl <- tibble(
  Metric = c(
    "Positions scored", "Genes scored",
    "Position-level components", "Gene-level components",
    "CAAS Score (median)", "CAAS Score (top 10% threshold)",
    "Gene CAAS TOP (median)", "Gene CAAS TOP (top 10% threshold)",
    "FADE site-level analysis available",
    "FADE gene-level analysis available",
    "Accumulation score available", "RERConverge available",
    "Genomic coordinates available"
  ),
  Value = c(
    format(nrow(pos_df), big.mark = ","),
    format(nrow(gene_df), big.mark = ","),
    as.character(n_components),
    as.character(n_gene_components),
    sprintf("%.4f", median(pos_df$CAAS_score, na.rm = TRUE)),
    sprintf("%.4f", thr_10),
    sprintf("%.4f", median(gene_df$gene_caas_top, na.rm = TRUE)),
    sprintf("%.4f", g_top_thr_10),
    ifelse(has_fade_pos,     "Yes", "No"),
    ifelse(has_fade_gene,    "Yes", "No"),
    ifelse(has_accum_score,  "Yes", "No"),
    ifelse(has_rer,          "Yes", "No"),
    ifelse(has_genomic,      "Yes", "No")
  )
)

knitr::kable(summary_tbl, align = c("l", "r"))
```

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>

*Generated by PhyloPhere - Scoring Module*
*Date of generation: `r Sys.Date()`*
