---
title: "CAAS Scoring Report"
subtitle: "`r paste0(params$traitname, ' — ', toupper(params$direction), '-phenotype direction')`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
params:
  position_scores_file: NULL
  gene_scores_file: NULL
  gene_corr_file: NULL
  traitname: "unknown_trait"
  output_dir: "."
  top_pct: 0.10
  top5_pct: 0.05
  top1_pct: 0.01
  gene_top_pct: 0.10
  gene_top5_pct: 0.05
  gene_top1_pct: 0.01
  stress_summary_file: NULL
  stress_corr_file: NULL
  stress_rank_file: NULL
  stress_overlap_file: NULL
  stress_variants_file: NULL
  stress_loadings_file: NULL
  vep_transvar_file: NULL
  vep_primateai_file: NULL
  vep_aa2prot_file: NULL
  genomic_info_file: NULL
  direction: "both"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  out.width = "100%"
)

suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(DT)
  library(scales)
  library(ggrepel)
})

# Tableau_10 colour swatch for pipeline-wide consistency
if (requireNamespace("paletteer", quietly = TRUE)) {
  .tableau10 <- as.character(paletteer::paletteer_d("ggthemes::Tableau_10"))
} else {
  .tableau10 <- c(
    "#4E79A7", "#F28E2B", "#E15759", "#76B7B2", "#59A14F",
    "#EDC948", "#B07AA1", "#FF9DA7", "#9C755F", "#BAB0AC"
  )
}
theme_set(theme_minimal(base_size = 13))
```

```{r load-data}
pos_df  <- read_tsv(params$position_scores_file, show_col_types = FALSE)
gene_df <- read_tsv(params$gene_scores_file, show_col_types = FALSE)
corr_df <- read_tsv(params$gene_corr_file, show_col_types = FALSE)

# Some runs export raw p-values; others export only the derived decile scores.
# Pick the available measure for each distribution section.
hyper_measure_col <- if ("Pvalue" %in% names(pos_df)) {
  "Pvalue"
} else if ("variability_score" %in% names(pos_df)) {
  "variability_score"
} else {
  stop("position_scores.tsv must contain either Pvalue or variability_score")
}
boot_measure_col <- if ("pvalue_boot" %in% names(pos_df)) {
  "pvalue_boot"
} else if ("pattern_score" %in% names(pos_df)) {
  "pattern_score"
} else {
  stop("position_scores.tsv must contain either pvalue_boot or pattern_score")
}
hyper_measure_is_raw <- identical(hyper_measure_col, "Pvalue")
boot_measure_is_raw  <- identical(boot_measure_col,  "pvalue_boot")

# Detect which optional components are available
has_rand     <- any(!is.na(gene_df$gene_rand_score))
has_rer      <- any(!is.na(gene_df$gene_rer_score))
has_fade     <- "gene_fade_score"     %in% names(gene_df) && any(!is.na(gene_df$gene_fade_score))
has_molerate <- "gene_molerate_score" %in% names(gene_df) && any(!is.na(gene_df$gene_molerate_score))

# ── VEP optional inputs ───────────────────────────────────────────────────────
.file_ok <- function(p) !is.null(p) && !is.na(p) && nchar(p) > 0 && file.exists(p)

has_scoring_stress <- .file_ok(params$stress_summary_file) &&
  .file_ok(params$stress_corr_file) &&
  .file_ok(params$stress_rank_file) &&
  .file_ok(params$stress_overlap_file) &&
  .file_ok(params$stress_variants_file)
has_vep_primateai <- .file_ok(params$vep_primateai_file)
has_aa2prot       <- .file_ok(params$vep_aa2prot_file)
has_genomic       <- .file_ok(params$genomic_info_file)
# TransVar is no longer used in the scoring report (removed — it added no actionable signal)
has_vep_transvar  <- FALSE

if (has_vep_primateai) {
  vep_pai_df <- read_tsv(params$vep_primateai_file, show_col_types = FALSE)
} else {
  vep_pai_df <- NULL
}

# aa2prot: tab-separated, no header — Gene | aln_pos (0-based) | caas_id | Gene:p.N
# Used to bridge alignment positions in pos_df to 1-based protein positions used by
# PrimateAI-3D lookup (map_to_primateai.py outputs transvar_query in this format).
if (has_aa2prot) {
  aa2prot_raw <- read_tsv(params$vep_aa2prot_file,
                          col_names = c("Gene", "aln_pos", "caas_id", "prot_query"),
                          show_col_types = FALSE)
  pos_prot_map <- aa2prot_raw %>%
    distinct(Gene, aln_pos, prot_query) %>%
    rename(Position = aln_pos)
} else {
  pos_prot_map <- NULL
}

if (has_genomic) {
  # Format: gene, chr, start, end, length  (no header assumed; detect if header present)
  genomic_raw <- read_tsv(params$genomic_info_file,
                          col_names = c("Gene", "chr", "gene_start", "gene_end", "gene_length"),
                          show_col_types = FALSE)
  # Drop header row if first cell is literally "gene" or "Gene"
  if (tolower(genomic_raw$Gene[1]) == "gene") genomic_raw <- genomic_raw[-1, ]
  genomic_raw <- genomic_raw %>%
    mutate(
      gene_start  = as.integer(gene_start),
      gene_end    = as.integer(gene_end),
      gene_length = as.integer(gene_length),
      gene_mid    = as.integer((gene_start + gene_end) / 2L)
    )
} else {
  genomic_raw <- NULL
}

if (has_scoring_stress) {
  stress_summary_df <- read_tsv(params$stress_summary_file, show_col_types = FALSE)
  stress_corr_df <- read_tsv(params$stress_corr_file, show_col_types = FALSE)
  stress_rank_df <- read_tsv(params$stress_rank_file, show_col_types = FALSE)
  stress_overlap_df <- read_tsv(params$stress_overlap_file, show_col_types = FALSE)
  stress_variants_df <- read_tsv(params$stress_variants_file, show_col_types = FALSE)
  stress_loadings_df <- if (.file_ok(params$stress_loadings_file)) {
    read_tsv(params$stress_loadings_file, show_col_types = FALSE)
  } else {
    NULL
  }
} else {
  stress_summary_df <- NULL
  stress_corr_df <- NULL
  stress_rank_df <- NULL
  stress_overlap_df <- NULL
  stress_variants_df <- NULL
  stress_loadings_df <- NULL
}

# Thresholds
top_pct  <- params$top_pct
top5_pct <- params$top5_pct
top1_pct <- params$top1_pct
```

---

## Introduction

The **CAAS Scoring** module integrates signals from multiple upstream analyses
into a unified, interpretable metric for prioritising convergent amino acid
substitutions (CAAS).  The scoring operates at two levels:

1. **Position-level CAAS Score** — each alignment position receives a composite
   score that combines:
   - **Biochemical score** — amino acid variability and pattern association,
     weighted by grouping scheme specificity.
   - **ASR score** — ancestral state reconstruction confidence: novel changes
     score 1.0; contradicted conserved cases score 0; ASR-supported conserved
     cases receive pair-weighted posterior support capped below novelty.
   - **Convergence score** — ordinal change-type support with a mild
     completeness penalty when only one side is assessable.
   - **Parallel score** — ordinal parallelism support weighted by MRCA
     posterior confidence and penalised mildly when only one side is assessable.

2. **Gene-level scores** — each gene receives up to five independent scores:
   - **gene_caas_score** — 90th percentile of position-level CAAS scores.
   `r if (has_rand) "- **gene_rand_score** — randomization enrichment across grouping schemes."`
   `r if (has_rer) "- **gene_rer_score** — RERConverge evolutionary rate correlation."`
   `r if (has_fade) "- **gene_fade_score** — directional selection signal (FADE max Bayes Factor, gene-level)."`
   `r if (has_molerate) "- **gene_molerate_score** — differential evolutionary rate (MoleRate, min p-value across top/bottom directions)."`
   - **gene_composite_score** — unweighted mean of available gene scores.

All component scores are mapped to a 0--1 scale using decile ranking, making
them directly comparable and suitable for downstream enrichment analysis.

---

## Scoring Methodology

This section documents the exact weighting and combination logic used to
produce every score in this report, so results can be interpreted and
reproduced independently of the pipeline code.

<details>
<summary><strong>Click to expand full methodology</strong></summary>

---

### Position-Level Components

All position-level component scores are mapped to the **[0, 1]** interval.
The four core components are always computed; FADE is included only
when the upstream module has been run.

#### 1 · Variability score

Decile rank of the **hypergeometric p-value** (`Pvalue`) across all
Gene × Position × Scheme rows, inverted so that smaller p-values receive
higher scores.

$$\text{variability\_score} = 1 - \frac{\text{ntile}(P_{\text{hyp}}, 10) - 1}{9}$$

*Range: 0.0 (least significant decile) → 1.0 (most significant decile).*

> This score does **not** enter `CAAS_score` directly — it feeds into
> `biochem_score` as a multiplicative factor.

---

#### 2 · Pattern score

Decile rank of the **permutation p-value** (`pvalue_boot`), inverted in the
same way.

$$\text{pattern\_score} = 1 - \frac{\text{ntile}(P_{\text{boot}}, 10) - 1}{9}$$

> Same role as variability\_score: multiplicative input into `biochem_score`.

---

#### 3 · Biochemical score

Sums the contribution of each grouping **scheme** in which the position was
detected, weighted by scheme specificity:

| Scheme | Weight |
|--------|-------:|
| US     | 0.50   |
| GS4    | 0.20   |
| GS3    | 0.10   |
| GS2    | 0.10   |
| GS1    | 0.10   |
| GS0    | 0.00   |

Each scheme contributes:

$$\text{contribution}_s = w_s \times \text{variability\_score} \times \text{pattern\_score}$$

The final score is the **cumulative sum** over all schemes where the position
was detected:

$$\text{biochem\_score} = \sum_{s \in \text{detected schemes}} w_s \times \text{variability\_score} \times \text{pattern\_score}$$

Maximum value when detected in all weighted schemes (US + GS4 + GS3 + GS2 + GS1)
with variability = pattern = 1.0 is **1.0** (0.5 + 0.2 + 0.1 + 0.1 + 0.1).

---

#### 4 · ASR score

Measures ancestral-state-reconstruction confidence. Three mutually exclusive cases:

| Condition | Score |
|-----------|------:|
| `is_conserved_meta = FALSE` (novel change) | **1.0** |
| `is_conserved_meta = TRUE` **and** `asr_is_conserved = TRUE` | see formula below |
| `is_conserved_meta = TRUE` **and** `asr_is_conserved = FALSE` | **0.0** |

For ASR-confirmed conserved positions, the score is pair-weighted across the
detected foreground-pair configuration. Let $w = 1 / n_{\text{pairs}}$. Then:

$$
\text{asr\_score} =
\sum_{\text{divergent pairs}} w +
\sum_{\text{conserved pairs}} \min\left(
\frac{w \times P_{\text{MRCA}(i)} + w \times P_{\text{root}} \times I(\texttt{asr\_root\_conserved})}{2},
0.9w
\right)
$$

This keeps supported conserved explanations informative while ensuring they do
not outrank genuinely novel changes, which always score **1.0**.

---

#### 5 · Convergence score

Scores the **type and directionality** of the amino acid change on each
assessed side (top / bottom foreground lineages).

Side values:

| Change type | Value |
|-------------|------:|
| convergent  | 1.0   |
| codivergent | 0.5   |
| divergent   | 0.25  |
| ambiguous / no change | excluded (NA) |

$$\text{mean\_change} = \text{mean of non-NA side values}$$

$$\text{completeness penalty} = \begin{cases} 1.0 & \text{both sides assessed} \\ 0.75 & \text{one side only} \end{cases}$$

$$\text{convergence\_score} = \text{mean\_change} \times \text{completeness penalty}$$

Range: 0.1875 (one side, divergent: 0.25 × 0.75) → 1.0 (both sides convergent).

---

#### 6 · Parallel score

Scores whether the amino acid change is **parallel** (same substitution in
independent lineages) or **non-parallel**, weighted by MRCA posterior
confidence.

Parallelism coefficients (note: non-parallel receives the *highest* weight
because independent parallel changes are less informative as a signal of
convergence):

| Parallel type | Coefficient |
|---------------|------------:|
| non-parallel  | 1.0         |
| mixed         | 0.75        |
| parallel      | 0.5         |

For each assessable side (where `change_top`/`change_bottom` is
convergent / codivergent / divergent):

$$\text{side\_score} = \text{coeff} \times \overline{P_{\text{MRCA}}}$$

where $\overline{P_{\text{MRCA}}}$ is the mean of `mrca_1_posterior`,
`mrca_2_posterior`, `mrca_3_posterior` (NA excluded).

$$\text{mean\_parallel} = \text{mean of non-NA side scores}$$

$$\text{completeness penalty} = \begin{cases} 1.0 & \text{both sides assessed} \\ 0.75 & \text{one side only} \end{cases}$$

$$\text{parallel\_score} = \text{mean\_parallel} \times \text{completeness penalty}$$

---

#### Composite CAAS score

Unweighted mean of the four core components. FADE is **not**
included — it is a gene-level signal (see below).

$$\text{CAAS\_score} = \frac{\text{biochem} + \text{asr} + \text{convergence} + \text{parallel}}{4}$$

---

### Gene-Level Scores

#### gene\_caas\_score

90th percentile of `CAAS_score` across all scored positions in that gene:

$$\text{gene\_caas\_score} = Q_{0.90}(\text{CAAS\_score})$$

#### gene\_rand\_score  *(optional)*

Scheme-weighted sum of decile-ranked accumulation randomization p-values.
The same scheme weights as `biochem_score` are used (GS0 excluded):

$$\text{gene\_rand\_score} = \sum_{s \in \{US, GS4, GS3, GS2, GS1\}} w_s \times \left(1 - \frac{\text{ntile}(P_{\text{empirical},s}, 10) - 1}{9}\right)$$

Maximum = 1.0 (all schemes p-value in best decile).

#### gene\_rer\_score  *(optional)*

Decile rank of the RERConverge adjusted p-value (`p.adj`; or `p.perm` if
available and non-zero), inverted:

$$\text{gene\_rer\_score} = 1 - \frac{\text{ntile}(p_{\text{RER}}, 10) - 1}{9}$$

#### gene\_fade\_score  *(optional)*

Gene-level maximum Bayes Factor from FADE directional selection analysis,
taking the best of the top/bottom foreground directions. Decile rank,
higher BF = higher score:

$$\text{gene\_fade\_score} = \frac{\text{ntile}(\max_{d \in \{\text{top,bottom}\}} \text{BF}_{d}, 10) - 1}{9}$$

#### gene\_composite\_score

Unweighted mean of all available gene-level sub-scores:

$$\text{gene\_composite\_score} = \frac{\text{gene\_caas} + [\text{gene\_rand}] + [\text{gene\_rer}] + [\text{gene\_fade}] + [\text{gene\_molerate}]}{N_{\text{available}}}$$

---

### ORA Gene List Extraction

Gene lists for Over-Representation Analysis are extracted at three thresholds
(configurable via `--scoring_position_top_pct` / `--scoring_position_top5_pct` /
`--scoring_position_top1_pct` and their gene-level equivalents) and three
directional subsets:

| Subset | Filter |
|--------|--------|
| **top** | Positions / genes where `change_side == "top"` |
| **bottom** | Positions / genes where `change_side == "bottom"` |
| **both** | All top-scoring positions / genes regardless of direction |

For **position-level** lists, the threshold is applied to `CAAS_score` and
the gene name is extracted from the qualifying positions.

For **gene-level** lists, the threshold is applied to each gene-level score
independently (`gene_caas_score`, `gene_rand_score`, `gene_rer_score`,
`gene_composite_score`), and direction is assigned from the gene's
highest-scoring position.

</details>

---

## Input Distributions {.tabset}

The `variability_score` and `pattern_score` components entering `biochem_score` are
derived from the **hypergeometric** and **permutation** signals used by the
scoring model. If the input table still contains raw p-values, those are shown
directly; otherwise the report falls back to the exported decile scores, which
preserve the same ranking but not the original p-value scale.

### `r if (hyper_measure_is_raw) "Hypergeometric P-value" else "Variability Score"`

```{r hyper-pval-dist, fig.height=8}
# Compute decile boundaries from the selected hypergeometric measure.
hyper_values <- pos_df[[hyper_measure_col]]
pv_breaks <- quantile(hyper_values, probs = seq(0, 1, by = 0.1), na.rm = TRUE)
decile_df_h <- pos_df %>%
  filter(!is.na(.data[[hyper_measure_col]])) %>%
  mutate(
    measure_value = .data[[hyper_measure_col]],
    plot_value    = if (hyper_measure_is_raw) -log10(pmax(measure_value, 1e-300)) else measure_value,
    decile        = dplyr::ntile(measure_value, 10),
    # Raw p-value: smaller = better → ntile rank 1 = best → invert so best = 1.0
    # Pre-computed score: higher = better → ntile rank 10 = best → direct mapping
    score_val     = if (hyper_measure_is_raw) 1 - (decile - 1) / 9 else (decile - 1) / 9
  )

# Panel 1: density on the selected measure scale with decile boundaries
p_hyp_density <- ggplot(decile_df_h, aes(x = plot_value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60,
                 fill = .tableau10[1], alpha = 0.6) +
  geom_density(linewidth = 0.8, colour = .tableau10[2]) +
  geom_vline(
    xintercept = if (hyper_measure_is_raw) -log10(pmax(pv_breaks[2:10], 1e-300)) else pv_breaks[2:10],
    linetype = "dotted", colour = "grey50", linewidth = 0.4
  ) +
  labs(
    title    = if (hyper_measure_is_raw) "Hypergeometric P-value Distribution (-log10 scale)" else "Variability Score Distribution",
    subtitle = if (hyper_measure_is_raw) {
      "Dotted lines: decile boundaries (D1 = largest p-values → score 1.0; D10 = smallest → score 1.0 inverted)"
    } else {
      "Dotted lines: decile boundaries of the exported variability_score values"
    },
    x        = if (hyper_measure_is_raw) "-log10(Pvalue)" else "variability_score",
    y        = "Density"
  ) +
  theme(plot.title = element_text(face = "bold"))

# Panel 2: count per decile, coloured by variability_score
decile_counts_h <- decile_df_h %>%
  group_by(decile, score_val) %>%
  summarise(n = n(), .groups = "drop")

p_hyp_decile <- ggplot(decile_counts_h,
                        aes(x = factor(decile), y = n, fill = score_val)) +
  geom_col(colour = "white", linewidth = 0.3) +
  scale_fill_gradient(low = .tableau10[3], high = .tableau10[1],
                      name = "variability_score") +
  labs(
    title    = "Positions per Hypergeometric Decile",
    subtitle = if (hyper_measure_is_raw) {
      "Decile 1 = highest p-values; Decile 10 = smallest p-values (variability_score = 1.0)"
    } else {
      "Decile 1 = lowest variability_score; Decile 10 = highest variability_score"
    },
    x        = "Decile",
    y        = "Number of rows"
  ) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "right")

gridExtra::grid.arrange(p_hyp_density, p_hyp_decile, ncol = 1)
```

### `r if (boot_measure_is_raw) "Permutation P-value" else "Pattern Score"`

```{r perm-pval-dist, fig.height=8}
boot_values <- pos_df[[boot_measure_col]]
pv_breaks_b <- quantile(boot_values, probs = seq(0, 1, by = 0.1), na.rm = TRUE)
decile_df_b <- pos_df %>%
  filter(!is.na(.data[[boot_measure_col]])) %>%
  mutate(
    measure_value = .data[[boot_measure_col]],
    plot_value    = if (boot_measure_is_raw) -log10(pmax(measure_value, 1e-300)) else measure_value,
    decile        = dplyr::ntile(measure_value, 10),
    score_val     = if (boot_measure_is_raw) 1 - (decile - 1) / 9 else (decile - 1) / 9
  )

p_boot_density <- ggplot(decile_df_b, aes(x = plot_value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60,
                 fill = .tableau10[5], alpha = 0.6) +
  geom_density(linewidth = 0.8, colour = .tableau10[6]) +
  geom_vline(
    xintercept = if (boot_measure_is_raw) -log10(pmax(pv_breaks_b[2:10], 1e-300)) else pv_breaks_b[2:10],
    linetype = "dotted", colour = "grey50", linewidth = 0.4
  ) +
  labs(
    title    = if (boot_measure_is_raw) "Permutation Bootstrap P-value Distribution (-log10 scale)" else "Pattern Score Distribution",
    subtitle = if (boot_measure_is_raw) {
      "Dotted lines: decile boundaries (D1 = least significant; D10 = most significant → pattern_score = 1.0)"
    } else {
      "Dotted lines: decile boundaries of the exported pattern_score values"
    },
    x        = if (boot_measure_is_raw) "-log10(pvalue_boot)" else "pattern_score",
    y        = "Density"
  ) +
  theme(plot.title = element_text(face = "bold"))

decile_counts_b <- decile_df_b %>%
  group_by(decile, score_val) %>%
  summarise(n = n(), .groups = "drop")

p_boot_decile <- ggplot(decile_counts_b,
                         aes(x = factor(decile), y = n, fill = score_val)) +
  geom_col(colour = "white", linewidth = 0.3) +
  scale_fill_gradient(low = .tableau10[7], high = .tableau10[5],
                      name = "pattern_score") +
  labs(
    title    = "Positions per Permutation Decile",
    subtitle = if (boot_measure_is_raw) {
      "Decile 10 = most significant permutation p-values (pattern_score = 1.0)"
    } else {
      "Decile 1 = lowest pattern_score; Decile 10 = highest pattern_score"
    },
    x        = "Decile",
    y        = "Number of rows"
  ) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "right")

gridExtra::grid.arrange(p_boot_density, p_boot_decile, ncol = 1)
```

### Joint Decile Summary

```{r joint-decile-summary, fig.height=5}
# Scatter of variability_score vs pattern_score coloured by CAAS_score
joint_df <- pos_df %>%
  filter(!is.na(variability_score), !is.na(pattern_score)) %>%
  mutate(
    hyp_decile  = dplyr::ntile(.data[[hyper_measure_col]], 10),
    boot_decile = dplyr::ntile(.data[[boot_measure_col]], 10)
  )

.r_hyp_caas  <- round(cor(joint_df$hyp_decile,  joint_df$CAAS_score, use = "complete.obs"), 2)
.r_boot_caas <- round(cor(joint_df$boot_decile, joint_df$CAAS_score, use = "complete.obs"), 2)
.r_hyp_boot  <- round(cor(joint_df$hyp_decile,  joint_df$boot_decile, use = "complete.obs"), 2)

ggplot(joint_df, aes(x = hyp_decile, y = boot_decile, colour = CAAS_score)) +
  geom_jitter(alpha = 0.45, size = 1.8, width = 0.35, height = 0.35) +
  scale_colour_gradientn(
    colours = c("#4E79A7", "#76B7B2", "#EDC948", "#F28E2B", "#E15759"),
    name    = "CAAS\nscore"
  ) +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(breaks = 1:10) +
  annotate("text", x = 1, y = 10,
           label = sprintf("r(hyp, CAAS) = %.2f\nr(perm, CAAS) = %.2f\nr(hyp, perm) = %.2f",
                           .r_hyp_caas, .r_boot_caas, .r_hyp_boot),
           hjust = 0, vjust = 1, size = 3.2, colour = "grey30") +
  labs(
    title    = "Hypergeometric vs Permutation Decile Assignment per Position",
    subtitle = "High CAAS scores (warm colours) cluster toward decile 10 × 10. Pearson r computed on decile ranks.",
    x        = "Hypergeometric decile (variability)",
    y        = "Permutation decile (pattern)"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

---

## Position-Level Scores {.tabset}

### CAAS Score Distribution

```{r caas-score-dist, fig.height=5}
thr_10 <- quantile(pos_df$CAAS_score, 1 - top_pct,  na.rm = TRUE)
thr_5  <- quantile(pos_df$CAAS_score, 1 - top5_pct, na.rm = TRUE)
thr_1  <- quantile(pos_df$CAAS_score, 1 - top1_pct, na.rm = TRUE)

ggplot(pos_df, aes(x = CAAS_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50,
                 fill = .tableau10[1], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[2]) +
  geom_vline(xintercept = thr_10, linetype = "dashed",
             colour = .tableau10[3], linewidth = 0.7) +
  geom_vline(xintercept = thr_5,  linetype = "longdash",
             colour = .tableau10[5], linewidth = 0.7) +
  geom_vline(xintercept = thr_1,  linetype = "dotted",
             colour = .tableau10[4], linewidth = 0.7) +
  annotate("text", x = thr_10, y = Inf, label = sprintf("Top %g%%", top_pct * 100),
           vjust = 2,    hjust = -0.1, colour = .tableau10[3], size = 3.5) +
  annotate("text", x = thr_5,  y = Inf, label = sprintf("Top %g%%", top5_pct * 100),
           vjust = 2.75, hjust = -0.1, colour = .tableau10[5], size = 3.5) +
  annotate("text", x = thr_1,  y = Inf, label = sprintf("Top %g%%", top1_pct * 100),
           vjust = 3.5,  hjust = -0.1, colour = .tableau10[4], size = 3.5) +
  labs(title = "CAAS Score Distribution",
       subtitle = sprintf("n = %s positions | Trait: %s",
                          format(nrow(pos_df), big.mark = ","), params$traitname),
       x = "CAAS Score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### Biochemical Score

```{r biochem-dist, fig.height=4.5}
ggplot(pos_df, aes(x = biochem_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = .tableau10[1], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[2]) +
  labs(title = "Biochemical Score Distribution",
       subtitle = "Weighted sum of scheme-specific variability x pattern contributions",
       x = "biochem_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### ASR Score

```{r asr-dist, fig.height=4.5}
ggplot(pos_df, aes(x = asr_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = .tableau10[3], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[4]) +
  labs(title = "ASR Score Distribution",
       subtitle = "Novel = 1.0 | contradicted conserved = 0 | supported conserved = pair-weighted posterior support below novelty",
       x = "asr_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### Convergence Score

```{r convergence-dist, fig.height=4.5}
ggplot(pos_df, aes(x = convergence_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = .tableau10[5], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[6]) +
  labs(title = "Convergence Score Distribution",
       subtitle = "Ordinal change type (1 / 0.5 / 0.25) with a 1.0 / 0.75 completeness penalty",
       x = "convergence_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### Parallel Score

```{r parallel-dist, fig.height=4.5}
ggplot(pos_df, aes(x = parallel_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = .tableau10[7], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[8]) +
  labs(title = "Parallel Score Distribution",
       subtitle = "Non-parallel / mixed / parallel = 1 / 0.75 / 0.5, weighted by MRCA confidence and completeness",
       x = "parallel_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### Component Overview

```{r component-overview, fig.height=6}
# Faceted density comparison of all position-level components
component_cols <- c("biochem_score", "asr_score", "convergence_score", "parallel_score")

long_df <- pos_df %>%
  select(all_of(component_cols)) %>%
  pivot_longer(everything(), names_to = "Component", values_to = "Score") %>%
  filter(!is.na(Score)) %>%
  mutate(Component = factor(Component, levels = component_cols))

ggplot(long_df, aes(x = Score, fill = Component)) +
  geom_density(alpha = 0.6, linewidth = 0.3) +
  scale_fill_manual(values = .tableau10[seq_along(component_cols)]) +
  facet_wrap(~Component, scales = "free_y", ncol = 2) +
  labs(title = "Position-Level Component Score Distributions",
       x = "Score (0-1)", y = "Density") +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "none")
```

### Threshold Profiles

```{r threshold-profiles, fig.height=14, fig.width=12}
# Component score distributions for positions inside each ORA gene-list cutoff.
# Each panel shows overlaid densities of all position-level components so you
# can read off which signal drives positions into the top-N% × direction bucket.

component_cols_tp <- c("biochem_score", "asr_score", "convergence_score", "parallel_score")

thr_10_tp <- quantile(pos_df$CAAS_score, 1 - top_pct,  na.rm = TRUE)
thr_5_tp  <- quantile(pos_df$CAAS_score, 1 - top5_pct, na.rm = TRUE)
thr_1_tp  <- quantile(pos_df$CAAS_score, 1 - top1_pct, na.rm = TRUE)

make_threshold_panel <- function(df, threshold, pct_label, direction) {
  subset_df <- if (direction == "both") {
    df %>% filter(CAAS_score >= threshold)
  } else {
    df %>% filter(CAAS_score >= threshold, change_side == direction)
  }

  if (nrow(subset_df) == 0) {
    return(
      ggplot() +
        annotate("text", x = 0.5, y = 0.5, label = "No positions",
                 size = 5, colour = "grey50") +
        labs(title = sprintf("Top %s — %s", pct_label, direction),
             x = "Score (0–1)", y = "Density") +
        xlim(0, 1) + theme_minimal()
    )
  }

  long_df <- subset_df %>%
    select(all_of(component_cols_tp)) %>%
    pivot_longer(everything(), names_to = "Component", values_to = "Score") %>%
    filter(!is.na(Score)) %>%
    mutate(Component = factor(Component, levels = component_cols_tp))

  ggplot(long_df, aes(x = Score, fill = Component, colour = Component)) +
    geom_density(alpha = 0.35, linewidth = 0.6) +
    scale_fill_manual(values   = .tableau10[seq_along(component_cols_tp)]) +
    scale_colour_manual(values = .tableau10[seq_along(component_cols_tp)]) +
    scale_x_continuous(limits = c(0, 1), expand = c(0.01, 0)) +
    labs(
      title    = sprintf("Top %s — %s", pct_label, direction),
      subtitle = sprintf("n = %d positions", nrow(subset_df)),
      x = "Score (0–1)", y = "Density",
      fill = NULL, colour = NULL
    ) +
    theme(
      plot.title    = element_text(face = "bold", size = 11),
      plot.subtitle = element_text(size = 9, colour = "grey40"),
      legend.position = "bottom",
      legend.text     = element_text(size = 8)
    )
}

panels <- list()
for (direction in c("top", "bottom", "both")) {
  panels[[paste0("10pct_", direction)]] <-
    make_threshold_panel(pos_df, thr_10_tp, "10%", direction)
  panels[[paste0("5pct_",  direction)]] <-
    make_threshold_panel(pos_df, thr_5_tp,  "5%",  direction)
  panels[[paste0("1pct_",  direction)]] <-
    make_threshold_panel(pos_df, thr_1_tp,  "1%",  direction)
}

# Layout: rows = direction (top / bottom / both), cols = threshold (10% | 5% | 1%)
gridExtra::grid.arrange(
  panels[["10pct_top"]],    panels[["5pct_top"]],    panels[["1pct_top"]],
  panels[["10pct_bottom"]], panels[["5pct_bottom"]], panels[["1pct_bottom"]],
  panels[["10pct_both"]],   panels[["5pct_both"]],   panels[["1pct_both"]],
  ncol = 3,
  top  = grid::textGrob(
    "Component score profiles for top-scoring position subsets",
    gp = grid::gpar(fontface = "bold", fontsize = 13)
  ),
  left = grid::textGrob(
    "Direction:  top                          bottom                          both",
    rot = 90, gp = grid::gpar(fontsize = 9, col = "grey40")
  )
)
```

### Correlation Analysis

```{r position-correlation-setup}
pos_score_cols <- c(
  "variability_score",
  "pattern_score",
  "biochem_score",
  "asr_score",
  "convergence_score",
  "parallel_score"
)
pos_corr_df <- if (length(pos_score_cols) >= 2) {
  expand.grid(
    score_a = pos_score_cols, score_b = pos_score_cols,
    stringsAsFactors = FALSE
  ) %>%
    filter(score_a < score_b) %>%
    rowwise() %>%
    mutate(
      pearson_r  = cor(pos_df[[score_a]], pos_df[[score_b]],
                       method = "pearson", use = "complete.obs"),
      spearman_r = cor(pos_df[[score_a]], pos_df[[score_b]],
                       method = "spearman", use = "complete.obs"),
      n_positions = sum(complete.cases(pos_df[[score_a]], pos_df[[score_b]]))
    ) %>%
    ungroup()
} else {
  tibble(
    score_a = character(), score_b = character(),
    pearson_r = numeric(), spearman_r = numeric(), n_positions = integer()
  )
}
```

```{r position-correlation-scatter, eval=(nrow(pos_corr_df) > 0), fig.height=14, fig.width=10}
pos_pairs_df <- pos_df %>%
  select(all_of(pos_score_cols)) %>%
  drop_na()

pos_pairs_long <- expand.grid(
  x_var = pos_score_cols, y_var = pos_score_cols,
  stringsAsFactors = FALSE
) %>%
  filter(x_var != y_var)

pos_plots <- list()
for (i in seq_len(nrow(pos_pairs_long))) {
  xv <- pos_pairs_long$x_var[i]
  yv <- pos_pairs_long$y_var[i]
  r_info <- pos_corr_df %>%
    filter((score_a == xv & score_b == yv) | (score_a == yv & score_b == xv))
  label <- if (nrow(r_info) > 0) {
    sprintf("r=%.2f / rho=%.2f", r_info$pearson_r[1], r_info$spearman_r[1])
  } else ""

  pos_plots[[length(pos_plots) + 1]] <- ggplot(pos_pairs_df, aes(x = .data[[xv]], y = .data[[yv]])) +
    geom_point(alpha = 0.18, size = 0.5, colour = .tableau10[1]) +
    geom_smooth(method = "lm", se = FALSE, colour = .tableau10[4], linewidth = 0.6) +
    annotate("text", x = Inf, y = -Inf, label = label,
             hjust = 1.1, vjust = -0.5, size = 2.8, colour = .tableau10[4]) +
    labs(
      x = sub("_score$", "", xv),
      y = sub("_score$", "", yv)
    ) +
    theme(axis.title = element_text(size = 8))
}

if (length(pos_plots) > 0) {
  gridExtra::grid.arrange(
    grobs = pos_plots,
    ncol = min(length(pos_score_cols) - 1, 3),
    top = grid::textGrob(
      "Position-Level Score Correlations (r = Pearson; rho = Spearman rank correlation)",
      gp = grid::gpar(fontface = "bold", fontsize = 13)
    )
  )
}
```

```{r position-correlation-table, eval=(nrow(pos_corr_df) > 0)}
DT::datatable(
  pos_corr_df %>%
    mutate(across(where(is.numeric), ~round(., 4))),
  rownames = FALSE,
  options  = list(pageLength = 15, dom = "t"),
  caption  = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "Pairwise correlations between position-level scores"
  )
)
```

### Stress Test

`r if (!has_scoring_stress) "> **Note:** scoring stress-test outputs are not available for this run."`

```{r stress-summary-table, eval=has_scoring_stress}
DT::datatable(
  stress_summary_df %>%
    arrange(desc(spearman_to_current), mean_abs_rank_shift) %>%
    mutate(across(where(is.numeric), ~round(., 4))),
  rownames = FALSE,
  options = list(pageLength = 20, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "Assumption sensitivity summary for alternative position-level score variants"
  )
)
```

```{r stress-component-overview, eval=has_scoring_stress, fig.height=6}
stress_component_cols <- c("variability_score", "pattern_score", "caap_score", "biochem_score")
stress_components_long <- stress_variants_df %>%
  select(any_of(stress_component_cols)) %>%
  pivot_longer(everything(), names_to = "Component", values_to = "Score") %>%
  filter(!is.na(Score))

ggplot(stress_components_long, aes(x = Score, fill = Component)) +
  geom_density(alpha = 0.55, linewidth = 0.4) +
  scale_fill_manual(values = .tableau10[seq_along(unique(stress_components_long$Component))]) +
  facet_wrap(~Component, scales = "free_y", ncol = 2) +
  labs(
    title = "Biochemical Decomposition Overview",
    subtitle = paste0("variability_score and pattern_score: per-row decile inputs\n",
                      "caap_score: sum of active scheme weights (scheme coverage); ",
                      "biochem_score: weighted product of the above"),
    x = "Score", y = "Density"
  ) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "none")
```

```{r stress-correlation-heatmap, eval=has_scoring_stress, fig.height=12, fig.width=10}
stress_heatmap_df <- stress_corr_df %>%
  pivot_longer(c(pearson_r, spearman_r), names_to = "metric", values_to = "value") %>%
  mutate(metric = recode(metric, pearson_r = "Pearson", spearman_r = "Spearman"))

ggplot(stress_heatmap_df, aes(x = score_a, y = score_b, fill = value)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.2f", value)), size = 2.8) +
  scale_fill_gradient2(low = .tableau10[3], mid = "white", high = .tableau10[1], midpoint = 0, limits = c(-1, 1)) +
  facet_wrap(~metric, ncol = 1) +
  labs(
    title = "Position-Level Correlation Heatmap",
    subtitle = paste0("r = Pearson correlation; rho = Spearman rank correlation\n",
                      "Values near \u00b11 indicate strong linear/monotone association between scores"),
    x = NULL, y = NULL, fill = "r"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold")
  )
```

```{r stress-rank-heatmap, eval=has_scoring_stress, fig.height=10, fig.width=9}
ggplot(stress_rank_df, aes(x = score_a, y = score_b, fill = spearman_rank)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.2f", spearman_rank)), size = 2.8) +
  scale_fill_gradient2(low = .tableau10[3], mid = "white", high = .tableau10[1], midpoint = 0, limits = c(-1, 1)) +
  labs(
    title = "Composite Rank-Agreement Heatmap",
    subtitle = paste0("rho = Spearman rank correlation between composite score rankings\n",
                      "Values near 1 mean both scoring variants rank positions nearly identically"),
    x = NULL, y = NULL, fill = "rho"
  ) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r stress-top-overlap, eval=has_scoring_stress}
DT::datatable(
  stress_overlap_df %>%
    arrange(threshold, desc(jaccard)) %>%
    mutate(across(where(is.numeric), ~round(., 4))),
  rownames = FALSE,
  options = list(pageLength = 20, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "Top-hit overlap across alternative position-level composite variants"
  )
)
```

```{r stress-instability, eval=has_scoring_stress}
rank_cols <- grep("^rank_", names(stress_variants_df), value = TRUE)
instability_df <- if (length(rank_cols) > 0) {
  stress_variants_df %>%
    select(Gene, Position, CAAS_current, any_of(rank_cols), rank_spread) %>%
    arrange(desc(rank_spread), desc(CAAS_current)) %>%
    slice_head(n = 25)
} else {
  tibble(Gene = character(), Position = integer(), rank_spread = numeric())
}

DT::datatable(
  instability_df %>% mutate(across(where(is.numeric), ~round(., 4))),
  rownames = FALSE,
  options = list(pageLength = 25, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "Top positions by rank instability across stress-test variants"
  )
)
```

Latent models ask: *if we let the data decide the weights, which component scores drive
the most variance?*  Two approaches are computed:

- **PCA_all_components** / **PCA_no_biochem** — First principal component of all
  position-level scores (or their decomposed biochemical sub-scores). The loadings
  show how much each input score contributes to PC1. A positive loading means the
  component co-varies with the main source of variation across positions.
- **Factor1_all_components** — Exploratory factor analysis with a single latent
  factor. Unlike PCA, factor analysis separates shared variance (the factor) from
  unique variance per variable; loadings quantify each component's contribution to
  the shared signal.
- **CAAS_residualized_biochem** — Biochemical score after regressing out ASR,
  convergence, and parallel scores, isolating the fraction of biochemical signal
  not explained by the structural components.

All latent scores are converted to 0–1 decile scale for comparison. A high
correlation with `CAAS_current` (see summary table) means the latent model and
the hand-engineered composite agree on which positions are most interesting.

```{r stress-latent, eval=has_scoring_stress, fig.height=5}
if (!is.null(stress_loadings_df) && nrow(stress_loadings_df) > 0) {
  ggplot(stress_loadings_df, aes(x = reorder(variable, loading), y = loading, fill = model)) +
    geom_col(position = "dodge") +
    coord_flip() +
    labs(
      title    = "Latent-Model Component Loadings",
      subtitle = paste0("How much each score component drives each latent model's first dimension\n",
                        "Positive = co-varies with PC1/Factor1; compare models to spot dominant components"),
      x = NULL, y = "Loading"
    ) +
    theme(plot.title    = element_text(face = "bold"),
          plot.subtitle = element_text(size = 9, colour = "grey40"))
} else {
  plot.new()
  text(0.5, 0.5, "No latent-model loadings available for this run")
}
```

`r if (has_scoring_stress) '> **How to read the stress test:**\n> \n> | Output | What to look for | What it means |\n> |--------|-----------------|---------------|\n> | **Biochemical decomposition** | Distributions of variability/pattern/caap/biochem | Checks that per-row inputs span the full 0–1 range; heavy skew may indicate low diversity |\n> | **Correlation heatmap** | r / rho close to 1 between a variant and CAAS_current | Scoring assumptions barely change position rankings |\n> | **Rank-agreement heatmap** | rho close to 1 | The two variants order positions the same way |\n> | **Top-overlap table** | High Jaccard at top-10% / top-5% thresholds | The same genes appear in gene lists regardless of assumptions |\n> | **Instability table** | High rank_spread | These positions are assumption-sensitive; examine their component scores manually |\n> | **Latent models** | PC1 / Factor1 loadings; correlation with CAAS_current | Which components drive shared variance; if latent models agree with CAAS_current the hand-engineered weights are data-consistent |\n> \n> **Score variant naming:** <code>CAAS_no_*</code> — drops one component; <code>CAAS_biochem_alt_*</code> — swaps the biochemical aggregation function; <code>PCA_*</code> / <code>Factor1_*</code> — data-driven weighting via PCA or factor analysis; <code>CAAS_residualized_biochem</code> — biochemical signal after removing correlation with structural components.\n> A robust run shows consistent rankings across all variants.'`

### Top Positions

```{r top-positions-table}
display_cols <- c("Gene", "Position", "CAAS_score", "biochem_score",
                  "asr_score", "convergence_score", "parallel_score", "change_side")

DT::datatable(
  pos_df %>%
    arrange(desc(CAAS_score)) %>%
    select(all_of(display_cols)) %>%
    mutate(across(where(is.numeric), ~round(., 4))),
  filter    = "top",
  rownames  = FALSE,
  options   = list(pageLength = 25, scrollX = TRUE),
  caption   = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Position-level CAAS scores | Trait: %s", params$traitname)
  )
)
```

---

## Position Characterisation {.tabset #position-characterisation}

This section examines how ranked positions behave in relation to external
signals that do **not** contribute to the `CAAS_score`. PrimateAI-3D is treated
as an independent characterisation layer: its signal is scheme-weighted exactly
like the biochemical score (US = 0.5 → GS4 = 0.2 → GS3/2/1 = 0.1 each → GS0 = 0),
producing a continuous 0–1 characterisation score.

- **`pai_char_score`** = Σ scheme\_weight × max(PAI3D\_score per CAAP group)
  per Gene × Position — captures the scheme-weighted pathogenicity of the
  CAAS-specific substitution.

This score ranges from 0 (no support in any scheme) to ≤ 1.0 (strong support
across all schemes). It is **not** included in `CAAS_score`.

### VEP: PrimateAI by CAAS Decile

`r if (!has_vep_primateai) "> **Note:** PrimateAI-3D scores are not available for this run."`

```{r vep-pai-setup, eval=has_vep_primateai}
# Detect PrimateAI score column (varies by database version)
.pai_score_col <- intersect(
  c("score_PAI3D", "score", "primateai_score", "pathogenicity_score", "PrimateAI_score"),
  names(vep_pai_df)
)[1]
if (is.na(.pai_score_col)) {
  warning("PrimateAI: no recognised score column found — skipping.")
  has_vep_primateai <- FALSE
}
```

```{r vep-pai-join, eval=has_vep_primateai}
pos_score_meta <- pos_df %>%
  select(Gene, Position, CAAS_score) %>%
  mutate(
    caas_decile  = dplyr::ntile(CAAS_score, 10),
    in_top_10pct = CAAS_score >= thr_10,
    in_top_5pct  = CAAS_score >= thr_5,
    in_top_1pct  = CAAS_score >= thr_1
  )

# pai_char_score = Σ scheme_weight × max(PAI3D per CAAP group) per protein position.
# This mirrors the biochem_score formula: cumulative scheme-weighted pathogenicity.
# map_to_primateai.py outputs caap_group and scheme_weight columns when available.
has_pai_schemes <- all(c("caap_group", "scheme_weight") %in% names(vep_pai_df))

if (has_pai_schemes) {
  # Two-step: max per (query × CAAP group) to collapse multiple alt AAs per scheme,
  # then weighted sum across CAAP groups (scheme_weight > 0 only).
  pai_by_prot <- vep_pai_df %>%
    mutate(
      pai_score     = as.numeric(.data[[.pai_score_col]]),
      scheme_weight = as.numeric(scheme_weight)
    ) %>%
    filter(is.finite(pai_score), is.finite(scheme_weight), scheme_weight > 0) %>%
    group_by(transvar_query, caap_group, scheme_weight) %>%
    summarise(max_pai = max(pai_score, na.rm = TRUE), .groups = "drop") %>%
    group_by(transvar_query) %>%
    summarise(
      pai_char_score = sum(max_pai * scheme_weight, na.rm = TRUE),
      n_pai_schemes  = n(),
      .groups = "drop"
    )
} else {
  # Legacy fallback: no scheme information — treat as US-only (weight = 0.5)
  pai_by_prot <- vep_pai_df %>%
    mutate(pai_score = as.numeric(.data[[.pai_score_col]])) %>%
    filter(is.finite(pai_score)) %>%
    group_by(transvar_query) %>%
    summarise(
      pai_char_score = 0.5 * max(pai_score, na.rm = TRUE),  # assume US weight
      n_pai_schemes  = 1L,
      .groups = "drop"
    )
}

if (!is.null(pos_prot_map)) {
  pos_vep_pai <- pos_prot_map %>%
    filter(!is.na(prot_query)) %>%
    inner_join(pai_by_prot, by = c("prot_query" = "transvar_query")) %>%
    inner_join(pos_score_meta, by = c("Gene", "Position")) %>%
    distinct(Gene, Position, .keep_all = TRUE)
} else {
  # fallback: parse GENE:p.N from transvar_query — will mostly miss without aa2prot
  pos_vep_pai <- pai_by_prot %>%
    mutate(
      Gene     = sub(":p\\..*", "", transvar_query),
      Position = as.integer(sub(".*:p\\.", "", transvar_query))
    ) %>%
    select(Gene, Position, pai_char_score, n_pai_schemes) %>%
    group_by(Gene, Position) %>%
    summarise(
      pai_char_score = max(pai_char_score, na.rm = TRUE),
      n_pai_schemes  = max(n_pai_schemes),
      .groups = "drop"
    ) %>%
    inner_join(pos_score_meta, by = c("Gene", "Position"))
}
```

```{r vep-pai-violin, eval=has_vep_primateai, fig.height=5}
# pai_char_score violin + boxplot per CAAS decile.
# Higher deciles should trend toward higher pai_char_score if CAAS and pathogenicity agree.
ggplot(pos_vep_pai, aes(x = factor(caas_decile), y = pai_char_score)) +
  geom_violin(fill = .tableau10[1], alpha = 0.65, trim = TRUE, scale = "width") +
  geom_boxplot(width = 0.14, outlier.size = 0.4, alpha = 0.9,
               fill = "white", colour = "grey30") +
  geom_hline(yintercept = median(pos_vep_pai$pai_char_score, na.rm = TRUE),
             linetype = "dashed", colour = .tableau10[3], linewidth = 0.7) +
  annotate("text", x = 0.4,
           y = median(pos_vep_pai$pai_char_score, na.rm = TRUE) + 0.005,
           hjust = 0, label = "Overall median",
           colour = .tableau10[3], size = 3.2) +
  scale_x_discrete(labels = c("1\n(low)", as.character(2:9), "10\n(high)")) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  labs(
    title    = "PrimateAI-3D Scheme-Weighted Score (pai_char_score) by CAAS Decile",
    subtitle = sprintf(
      "%d positions with PAI3D coverage (of %d total) | score = \u03a3 scheme_weight \u00d7 max(PAI3D per CAAP group)",
      nrow(pos_vep_pai), nrow(pos_df)
    ),
    x = "CAAS Score Decile", y = "pai_char_score"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r vep-pai-scatter, eval=has_vep_primateai, fig.height=5}
# Scatter: CAAS_score vs pai_char_score, coloured by CAAS decile.
ggplot(pos_vep_pai, aes(x = CAAS_score, y = pai_char_score, colour = caas_decile)) +
  geom_point(alpha = 0.45, size = 0.9) +
  geom_smooth(method = "lm", se = TRUE, colour = .tableau10[1],
              linewidth = 0.8, fill = .tableau10[1], alpha = 0.15) +
  geom_vline(xintercept = thr_10, linetype = "dashed",
             colour = .tableau10[3], linewidth = 0.6) +
  geom_vline(xintercept = thr_5,  linetype = "longdash",
             colour = .tableau10[5], linewidth = 0.6) +
  geom_vline(xintercept = thr_1,  linetype = "dotted",
             colour = .tableau10[4], linewidth = 0.6) +
  scale_colour_viridis_c(name = "CAAS\nDecile", option = "plasma") +
  labs(
    title    = "CAAS Score vs PrimateAI-3D Scheme-Weighted Score",
    subtitle = "Dashed verticals: CAAS top-10%, top-5%, top-1% thresholds",
    x = "CAAS Score", y = "pai_char_score"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r vep-pai-fisher, eval=has_vep_primateai}
# Fisher's exact: are positions with pai_char_score above the median enriched
# among the top CAAS positions?
# Using the median as a data-driven split rather than a fixed raw-score threshold.
.pai_med <- median(pos_vep_pai$pai_char_score, na.rm = TRUE)

run_pai_fisher <- function(df, flag_col, label) {
  df <- df %>% mutate(high_pai = !is.na(pai_char_score) & pai_char_score > .pai_med)
  n_top      <- sum( df[[flag_col]], na.rm = TRUE)
  n_top_h    <- sum( df[[flag_col]] & df$high_pai, na.rm = TRUE)
  n_rest     <- sum(!df[[flag_col]], na.rm = TRUE)
  n_rest_h   <- sum(!df[[flag_col]] & df$high_pai, na.rm = TRUE)
  mat <- matrix(
    c(n_top - n_top_h, n_rest - n_rest_h, n_top_h, n_rest_h),
    nrow = 2,
    dimnames = list(In_top = c("FALSE","TRUE"), High_PAI = c("FALSE","TRUE"))
  )
  ft <- fisher.test(mat)
  tibble(
    Threshold        = label,
    N_top            = n_top,
    N_high_pai_top   = n_top_h,
    Pct_top          = sprintf("%.1f%%", 100 * n_top_h  / max(n_top,  1)),
    N_rest           = n_rest,
    N_high_pai_rest  = n_rest_h,
    Pct_rest         = sprintf("%.1f%%", 100 * n_rest_h / max(n_rest, 1)),
    Odds_ratio       = sprintf("%.2f", ft$estimate),
    CI_95            = sprintf("[%.2f, %.2f]", ft$conf.int[1], ft$conf.int[2]),
    P_value          = formatC(ft$p.value, format = "e", digits = 2)
  )
}

pai_fisher <- bind_rows(
  run_pai_fisher(pos_vep_pai, "in_top_10pct",
                 sprintf("Top %g%% CAAS (decile)",    top_pct  * 100)),
  run_pai_fisher(pos_vep_pai, "in_top_5pct",
                 sprintf("Top %g%% CAAS",             top5_pct * 100)),
  run_pai_fisher(pos_vep_pai, "in_top_1pct",
                 sprintf("Top %g%% CAAS (percentile)", top1_pct * 100))
)

knitr::kable(
  pai_fisher,
  caption = sprintf(
    "Fisher's exact test — enrichment of positions with pai_char_score > median (%.4f) in top-ranking CAAS positions.",
    .pai_med
  ),
  align = c("l", rep("r", ncol(pai_fisher) - 1))
)
```


---

## Gene-Level Scores {.tabset}

### Gene CAAS Score

```{r gene-caas-dist, fig.height=4.5}
g_thr_10 <- quantile(gene_df$gene_caas_score, 1 - params$gene_top_pct,  na.rm = TRUE)
g_thr_5  <- quantile(gene_df$gene_caas_score, 1 - params$gene_top5_pct, na.rm = TRUE)
g_thr_1  <- quantile(gene_df$gene_caas_score, 1 - params$gene_top1_pct, na.rm = TRUE)

ggplot(gene_df, aes(x = gene_caas_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = .tableau10[1], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[2]) +
  geom_vline(xintercept = g_thr_10, linetype = "dashed",
             colour = .tableau10[3], linewidth = 0.7) +
  geom_vline(xintercept = g_thr_5,  linetype = "longdash",
             colour = .tableau10[5], linewidth = 0.7) +
  geom_vline(xintercept = g_thr_1,  linetype = "dotted",
             colour = .tableau10[4], linewidth = 0.7) +
  labs(title = "Gene CAAS Score Distribution",
       subtitle = "90th percentile of position-level CAAS scores per gene",
       x = "gene_caas_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

```{r gene-rand-dist, eval=has_rand, fig.height=4.5}
ggplot(gene_df %>% filter(!is.na(gene_rand_score)), aes(x = gene_rand_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = .tableau10[5], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[6]) +
  labs(title = "Gene Randomization Score Distribution",
       subtitle = "Scheme-weighted deciles of accumulation randomization p-values",
       x = "gene_rand_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

```{r gene-rer-dist, eval=has_rer, fig.height=4.5}
ggplot(gene_df %>% filter(!is.na(gene_rer_score)), aes(x = gene_rer_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = .tableau10[7], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[8]) +
  labs(title = "Gene RERConverge Score Distribution",
       subtitle = "Deciles of evolutionary rate correlation p-values",
       x = "gene_rer_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

```{r gene-fade-dist, eval=has_fade, fig.height=4.5}
ggplot(gene_df %>% filter(!is.na(gene_fade_score)), aes(x = gene_fade_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = .tableau10[9], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[10]) +
  labs(title = "Gene FADE Score Distribution",
       subtitle = "Deciles of max directional selection Bayes Factor per gene",
       x = "gene_fade_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

```{r gene-molerate-dist, eval=has_molerate, fig.height=4.5}
ggplot(gene_df %>% filter(!is.na(gene_molerate_score)), aes(x = gene_molerate_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = .tableau10[3], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[4]) +
  labs(title = "Gene MoleRate Score Distribution",
       subtitle = "Deciles of min(pval_pp_vs_prop) across top/bottom directions per gene",
       x = "gene_molerate_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

```{r gene-molerate-log2ratio, eval=has_molerate, fig.height=5}
# Visualise log2_ratio (acceleration vs deceleration) per direction.
# mol_log2ratio_top    > 0  → foreground accelerated in top-phenotype lineages
# mol_log2ratio_bottom > 0  → foreground accelerated in bottom-phenotype lineages
has_mol_log2_top    <- "mol_log2ratio_top"    %in% names(gene_df) && any(!is.na(gene_df$mol_log2ratio_top))
has_mol_log2_bottom <- "mol_log2ratio_bottom" %in% names(gene_df) && any(!is.na(gene_df$mol_log2ratio_bottom))

if (has_mol_log2_top || has_mol_log2_bottom) {
  mol_long <- gene_df %>%
    select(Gene,
           any_of(c("mol_log2ratio_top", "mol_log2ratio_bottom"))) %>%
    pivot_longer(-Gene, names_to = "Direction", values_to = "log2_ratio") %>%
    filter(!is.na(log2_ratio)) %>%
    mutate(Direction = recode(Direction,
                              mol_log2ratio_top    = "top",
                              mol_log2ratio_bottom = "bottom"))

  ggplot(mol_long, aes(x = log2_ratio, fill = Direction, colour = Direction)) +
    geom_histogram(aes(y = after_stat(density)), bins = 40, alpha = 0.5,
                   position = "identity") +
    geom_density(linewidth = 0.8, alpha = 0) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey30", linewidth = 0.7) +
    scale_fill_manual(values   = c(top = .tableau10[1], bottom = .tableau10[2])) +
    scale_colour_manual(values = c(top = .tableau10[1], bottom = .tableau10[2])) +
    facet_wrap(~Direction, ncol = 2) +
    labs(
      title    = "MoleRate log2-ratio: Acceleration vs. Deceleration",
      subtitle = "log2(fg rate / bg rate) — positive = acceleration in foreground lineages",
      x = "log2_ratio", y = "Density", fill = NULL, colour = NULL
    ) +
    theme(plot.title      = element_text(face = "bold"),
          legend.position = "none",
          strip.text      = element_text(face = "bold"))
}
```

### Gene Composite Score

```{r gene-composite-dist, fig.height=5}
gc_thr_10 <- quantile(gene_df$gene_composite_score, 1 - params$gene_top_pct,  na.rm = TRUE)
gc_thr_5  <- quantile(gene_df$gene_composite_score, 1 - params$gene_top5_pct, na.rm = TRUE)
gc_thr_1  <- quantile(gene_df$gene_composite_score, 1 - params$gene_top1_pct, na.rm = TRUE)

ggplot(gene_df, aes(x = gene_composite_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = .tableau10[1], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[2]) +
  geom_vline(xintercept = gc_thr_10, linetype = "dashed",
             colour = .tableau10[3], linewidth = 0.7) +
  geom_vline(xintercept = gc_thr_5,  linetype = "longdash",
             colour = .tableau10[5], linewidth = 0.7) +
  geom_vline(xintercept = gc_thr_1,  linetype = "dotted",
             colour = .tableau10[4], linewidth = 0.7) +
  annotate("text", x = gc_thr_10, y = Inf,
           label = sprintf("Top %g%%", params$gene_top_pct * 100),
           vjust = 2,    hjust = -0.1, colour = .tableau10[3], size = 3.5) +
  annotate("text", x = gc_thr_5,  y = Inf,
           label = sprintf("Top %g%%", params$gene_top5_pct * 100),
           vjust = 2.75, hjust = -0.1, colour = .tableau10[5], size = 3.5) +
  annotate("text", x = gc_thr_1,  y = Inf,
           label = sprintf("Top %g%%", params$gene_top1_pct * 100),
           vjust = 3.5,  hjust = -0.1, colour = .tableau10[4], size = 3.5) +
  labs(title = "Gene Composite Score Distribution",
       subtitle = sprintf("Unweighted mean of available gene scores | n = %s genes",
                          format(nrow(gene_df), big.mark = ",")),
       x = "gene_composite_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### Gene Threshold Profiles

```{r gene-threshold-profiles, fig.height=14, fig.width=12}
# Same idea as position-level threshold profiles but for gene-level scores.
# For each top-N% × direction bucket, show overlaid densities of gene sub-scores.

gene_score_cols_tp <- "gene_caas_score"
if (has_rand)     gene_score_cols_tp <- c(gene_score_cols_tp, "gene_rand_score")
if (has_rer)      gene_score_cols_tp <- c(gene_score_cols_tp, "gene_rer_score")
if (has_fade)     gene_score_cols_tp <- c(gene_score_cols_tp, "gene_fade_score")
if (has_molerate) gene_score_cols_tp <- c(gene_score_cols_tp, "gene_molerate_score")

# Direction per gene is inherited from its top-scoring position
gene_direction_tp <- pos_df %>%
  group_by(Gene) %>%
  slice_max(CAAS_score, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(Gene, gene_direction = change_side)

gene_df_dir <- gene_df %>% left_join(gene_direction_tp, by = "Gene")

# Directional subsets: "all" = no filter; "top" = top or both; "bottom" = bottom or both
gene_dir_subsets <- list(
  all    = gene_df_dir,
  top    = gene_df_dir %>% filter(gene_direction %in% c("top",    "both")),
  bottom = gene_df_dir %>% filter(gene_direction %in% c("bottom", "both"))
)

make_gene_threshold_panel <- function(df_dir, pct_label, pct_val, direction) {
  # Threshold is computed within the directional subset, not globally
  thr <- quantile(df_dir$gene_composite_score, 1 - pct_val, na.rm = TRUE)
  subset_df <- df_dir %>% filter(gene_composite_score >= thr)

  if (nrow(subset_df) == 0 || length(gene_score_cols_tp) < 2) {
    return(
      ggplot() +
        annotate("text", x = 0.5, y = 0.5,
                 label = ifelse(nrow(subset_df) == 0, "No genes", "Single score only"),
                 size = 5, colour = "grey50") +
        labs(title = sprintf("Top %s — %s", pct_label, direction),
             x = "Score (0–1)", y = "Density") +
        xlim(0, 1) + theme_minimal()
    )
  }

  long_df <- subset_df %>%
    select(all_of(gene_score_cols_tp)) %>%
    pivot_longer(everything(), names_to = "Score_type", values_to = "Score") %>%
    filter(!is.na(Score)) %>%
    mutate(Score_type = factor(Score_type, levels = gene_score_cols_tp))

  ggplot(long_df, aes(x = Score, fill = Score_type, colour = Score_type)) +
    geom_density(alpha = 0.35, linewidth = 0.6) +
    scale_fill_manual(values   = .tableau10[seq_along(gene_score_cols_tp)],
                      labels   = sub("gene_", "", sub("_score", "", gene_score_cols_tp))) +
    scale_colour_manual(values = .tableau10[seq_along(gene_score_cols_tp)],
                        labels = sub("gene_", "", sub("_score", "", gene_score_cols_tp))) +
    scale_x_continuous(limits = c(0, 1), expand = c(0.01, 0)) +
    labs(
      title    = sprintf("Top %s — %s", pct_label, direction),
      subtitle = sprintf("n = %d genes (thr = %.3f)", nrow(subset_df), thr),
      x = "Score (0–1)", y = "Density",
      fill = NULL, colour = NULL
    ) +
    theme(
      plot.title      = element_text(face = "bold", size = 11),
      plot.subtitle   = element_text(size = 9, colour = "grey40"),
      legend.position = "bottom",
      legend.text     = element_text(size = 8)
    )
}

gene_panels <- list()
for (direction in names(gene_dir_subsets)) {
  df_d <- gene_dir_subsets[[direction]]
  gene_panels[[paste0("10pct_", direction)]] <-
    make_gene_threshold_panel(df_d, "10%", params$gene_top_pct,  direction)
  gene_panels[[paste0("5pct_",  direction)]] <-
    make_gene_threshold_panel(df_d, "5%",  params$gene_top5_pct, direction)
  gene_panels[[paste0("1pct_",  direction)]] <-
    make_gene_threshold_panel(df_d, "1%",  params$gene_top1_pct, direction)
}

gridExtra::grid.arrange(
  gene_panels[["10pct_all"]],    gene_panels[["5pct_all"]],    gene_panels[["1pct_all"]],
  gene_panels[["10pct_top"]],    gene_panels[["5pct_top"]],    gene_panels[["1pct_top"]],
  gene_panels[["10pct_bottom"]], gene_panels[["5pct_bottom"]], gene_panels[["1pct_bottom"]],
  ncol = 3,
  top  = grid::textGrob(
    "Gene score profiles for top-scoring gene subsets (by gene_composite_score)",
    gp = grid::gpar(fontface = "bold", fontsize = 13)
  ),
  left = grid::textGrob(
    "Direction:  all                            top                          bottom",
    rot = 90, gp = grid::gpar(fontsize = 9, col = "grey40")
  )
)
```

### Correlation Analysis

```{r correlation-scatter, eval=(nrow(corr_df) > 0), fig.height=7}
# Build scatter matrix of gene-level scores
score_cols <- "gene_caas_score"
if (has_rand)     score_cols <- c(score_cols, "gene_rand_score")
if (has_rer)      score_cols <- c(score_cols, "gene_rer_score")
if (has_fade)     score_cols <- c(score_cols, "gene_fade_score")
if (has_molerate) score_cols <- c(score_cols, "gene_molerate_score")

if (length(score_cols) >= 2) {
  pairs_df <- gene_df %>%
    select(all_of(score_cols)) %>%
    drop_na()

  # Create pairs plot
  pairs_long <- expand.grid(
    x_var = score_cols, y_var = score_cols,
    stringsAsFactors = FALSE
  ) %>%
    filter(x_var != y_var)

  plots <- list()
  for (i in seq_len(nrow(pairs_long))) {
    xv <- pairs_long$x_var[i]
    yv <- pairs_long$y_var[i]
    r_info <- corr_df %>%
      filter((score_a == xv & score_b == yv) | (score_a == yv & score_b == xv))
    label <- if (nrow(r_info) > 0) {
      sprintf("r=%.2f / rho=%.2f", r_info$pearson_r[1], r_info$spearman_r[1])
    } else ""

    plots[[length(plots) + 1]] <- ggplot(pairs_df, aes(x = .data[[xv]], y = .data[[yv]])) +
      geom_point(alpha = 0.3, size = 0.8, colour = .tableau10[1]) +
      geom_smooth(method = "lm", se = FALSE, colour = .tableau10[4], linewidth = 0.6) +
      annotate("text", x = Inf, y = -Inf, label = label,
               hjust = 1.1, vjust = -0.5, size = 3, colour = .tableau10[4]) +
      labs(x = sub("gene_", "", sub("_score", "", xv)),
           y = sub("gene_", "", sub("_score", "", yv))) +
      theme(axis.title = element_text(size = 9))
  }

  if (length(plots) > 0) {
    gridExtra::grid.arrange(grobs = plots,
                            ncol = min(length(score_cols) - 1, 3),
                            top = grid::textGrob(
                              "Gene Score Correlations (r = Pearson; rho = Spearman rank correlation)",
                              gp = grid::gpar(fontface = "bold", fontsize = 13)))
  }
}
```

```{r correlation-table, eval=(nrow(corr_df) > 0)}
DT::datatable(
  corr_df %>%
    mutate(across(where(is.numeric), ~round(., 4))),
  rownames = FALSE,
  options  = list(pageLength = 10, dom = "t"),
  caption  = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "Pairwise correlations between gene-level scores"
  )
)
```

### Top Genes

```{r top-genes-table}
gene_display_cols <- c("Gene", "n_positions", "gene_caas_score")
if (has_rand)     gene_display_cols <- c(gene_display_cols, "gene_rand_score")
if (has_rer)      gene_display_cols <- c(gene_display_cols, "gene_rer_score")
if (has_fade)     gene_display_cols <- c(gene_display_cols, "gene_fade_score")
if (has_molerate) gene_display_cols <- c(gene_display_cols, "gene_molerate_score")
gene_display_cols <- c(gene_display_cols, "gene_composite_score")

DT::datatable(
  gene_df %>%
    arrange(desc(gene_composite_score)) %>%
    select(all_of(gene_display_cols)) %>%
    mutate(across(where(is.numeric) & !matches("n_positions"), ~round(., 4))),
  filter    = "top",
  rownames  = FALSE,
  options   = list(pageLength = 25, scrollX = TRUE),
  caption   = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Gene-level scores | Trait: %s", params$traitname)
  )
)
```

---

## Gene-Level Characterisation {.tabset}

This section accumulates PrimateAI-3D signals at the gene level, testing
whether genes with high composite scores tend to concentrate pathogenic
PrimateAI-3D annotations.

### Gene VEP Accumulation

`r if (!has_vep_primateai) "> **Note:** PrimateAI-3D scores are not available — gene-level VEP accumulation skipped."`

```{r gene-vep-pai-accum, eval=has_vep_primateai, fig.height=5}
# Per-gene: mean pai_char_score (scheme-weighted pathogenicity).
gene_vep_pai <- pos_vep_pai %>%
  group_by(Gene) %>%
  summarise(
    n_pai_positions  = n(),
    mean_pai_char    = mean(pai_char_score, na.rm = TRUE),
    sum_pai_char     = sum(pai_char_score,  na.rm = TRUE),
    n_pai_any_signal = sum(!is.na(pai_char_score) & pai_char_score > 0, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  inner_join(gene_df %>% select(Gene, gene_caas_score, gene_composite_score,
                                n_positions),
             by = "Gene")

ggplot(gene_vep_pai, aes(x = gene_caas_score, y = mean_pai_char,
                          size = n_positions, colour = gene_composite_score)) +
  geom_point(alpha = 0.65) +
  ggrepel::geom_text_repel(
    data = gene_vep_pai %>% slice_max(mean_pai_char, n = 15, with_ties = FALSE),
    aes(label = Gene), size = 2.8, max.overlaps = 20,
    show.legend = FALSE
  ) +
  geom_vline(xintercept = g_thr_10, linetype = "dashed",
             colour = .tableau10[3], linewidth = 0.7) +
  scale_colour_viridis_c(name = "Gene\nComposite", option = "plasma") +
  scale_size_continuous(name = "N positions", range = c(1, 8)) +
  labs(
    title    = "Gene CAAS Score vs Mean PrimateAI-3D Scheme-Weighted Score",
    subtitle = "Labels: top-15 genes by mean pai_char_score; dashed = CAAS top 10% threshold",
    x = "Gene CAAS Score", y = "Mean pai_char_score per gene"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r gene-vep-pai-top-bar, eval=has_vep_primateai, fig.height=6}
# Top 30 genes by mean pai_char_score, coloured by gene_caas_score
top_pai_genes <- gene_vep_pai %>%
  filter(n_pai_positions >= 2, mean_pai_char > 0) %>%
  slice_max(mean_pai_char, n = 30, with_ties = FALSE) %>%
  arrange(mean_pai_char) %>%
  mutate(Gene = factor(Gene, levels = Gene))

ggplot(top_pai_genes, aes(x = mean_pai_char, y = Gene,
                           fill = gene_caas_score)) +
  geom_col() +
  scale_fill_viridis_c(name = "Gene CAAS\nScore", option = "plasma") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Top 30 Genes by Mean PrimateAI-3D Scheme-Weighted Score",
    subtitle = paste0("Genes with \u2265 2 positions with PAI3D coverage\n",
                      "pai_char_score = \u03a3 scheme_weight \u00d7 max(PAI3D per CAAP group)"),
    x = "Mean pai_char_score", y = NULL
  ) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.y = element_text(size = 8))
```


---

## Genomic Accumulation {.tabset}

`r if (!has_genomic) "> **Note:** Genomic coordinates (gene_ensembl_file) are not available for this run — genomic accumulation skipped."`

```{r genomic-setup, eval=has_genomic}
# Join gene_df with genomic coordinates, then annotate with VEP
gene_genomic <- gene_df %>%
  select(Gene, gene_caas_score, gene_composite_score, n_positions) %>%
  inner_join(genomic_raw, by = "Gene") %>%
  # Standardise chromosome labels (remove "chr" prefix if present)
  mutate(
    chr_label = sub("^chr", "", chr),
    # Order chromosomes numerically then X, Y, MT
    chr_num   = suppressWarnings(as.integer(chr_label)),
    chr_order = dplyr::case_when(
      !is.na(chr_num)       ~ chr_num,
      chr_label == "X"      ~ 23L,
      chr_label == "Y"      ~ 24L,
      chr_label %in% c("MT","M") ~ 25L,
      TRUE                  ~ 26L
    )
  ) %>%
  arrange(chr_order, gene_mid)

# Optionally add mean PAI characterisation score
if (has_vep_primateai && exists("gene_vep_pai")) {
  gene_genomic <- gene_genomic %>%
    left_join(gene_vep_pai %>% select(Gene, mean_pai_char),
              by = "Gene")
} else {
  gene_genomic$mean_pai_char <- NA_real_
}

# Build a cumulative x-axis offset per chromosome for a stitched Manhattan view
chr_offsets <- gene_genomic %>%
  group_by(chr_order, chr_label) %>%
  summarise(chr_max = max(gene_end, na.rm = TRUE), .groups = "drop") %>%
  arrange(chr_order) %>%
  mutate(offset = cumsum(lag(chr_max, default = 0L)) + (chr_order - 1L) * 5e6L)

gene_genomic <- gene_genomic %>%
  left_join(chr_offsets %>% select(chr_order, offset, chr_label), by = c("chr_order", "chr_label")) %>%
  mutate(manhattan_x = gene_mid + offset)

chr_label_positions <- chr_offsets %>%
  left_join(gene_genomic %>% group_by(chr_order) %>%
              summarise(mid_x = mean(manhattan_x, na.rm = TRUE), .groups = "drop"),
            by = "chr_order")
```

### CAAS Score Manhattan

```{r genomic-manhattan-caas, eval=has_genomic, fig.height=5, fig.width=14}
# Manhattan-style plot: x = genomic position (stitched), y = gene_caas_score
# Alternating chromosome colours for readability.
gene_genomic <- gene_genomic %>%
  mutate(chr_even = chr_order %% 2 == 0)

ggplot(gene_genomic,
       aes(x = manhattan_x, y = gene_caas_score,
           colour = chr_even, size = n_positions)) +
  geom_point(alpha = 0.65) +
  geom_hline(yintercept = g_thr_10, linetype = "dashed",
             colour = "grey40", linewidth = 0.6) +
  ggrepel::geom_text_repel(
    data = gene_genomic %>% slice_max(gene_caas_score, n = 20, with_ties = FALSE),
    aes(label = Gene), size = 2.2, max.overlaps = 15,
    show.legend = FALSE, colour = "black"
  ) +
  scale_colour_manual(values = c(.tableau10[1], .tableau10[2]), guide = "none") +
  scale_size_continuous(name = "N positions", range = c(0.5, 5)) +
  scale_x_continuous(
    breaks = chr_label_positions$mid_x,
    labels = chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  labs(
    title    = "Gene CAAS Score — Manhattan View",
    subtitle = sprintf("%d genes with genomic coordinates | dashed = top %g%% threshold",
                       nrow(gene_genomic), params$gene_top_pct * 100),
    x = "Chromosome", y = "Gene CAAS Score"
  ) +
  theme(
    plot.title      = element_text(face = "bold"),
    axis.text.x     = element_text(size = 7, angle = 0),
    panel.grid.major.x = element_blank()
  )
```

### Composite Score Manhattan

```{r genomic-manhattan-composite, eval=has_genomic, fig.height=5, fig.width=14}
ggplot(gene_genomic,
       aes(x = manhattan_x, y = gene_composite_score,
           colour = chr_even, size = n_positions)) +
  geom_point(alpha = 0.65) +
  geom_hline(yintercept = gc_thr_10, linetype = "dashed",
             colour = "grey40", linewidth = 0.6) +
  ggrepel::geom_text_repel(
    data = gene_genomic %>% slice_max(gene_composite_score, n = 20, with_ties = FALSE),
    aes(label = Gene), size = 2.2, max.overlaps = 15,
    show.legend = FALSE, colour = "black"
  ) +
  scale_colour_manual(values = c(.tableau10[1], .tableau10[2]), guide = "none") +
  scale_size_continuous(name = "N positions", range = c(0.5, 5)) +
  scale_x_continuous(
    breaks = chr_label_positions$mid_x,
    labels = chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  labs(
    title    = "Gene Composite Score — Manhattan View",
    subtitle = sprintf("dashed = top %g%% threshold", params$gene_top_pct * 100),
    x = "Chromosome", y = "Gene Composite Score"
  ) +
  theme(
    plot.title      = element_text(face = "bold"),
    axis.text.x     = element_text(size = 7, angle = 0),
    panel.grid.major.x = element_blank()
  )
```

### PAI3D Signal Manhattan

`r if (!has_vep_primateai) "> **Note:** PrimateAI-3D scores not available — PAI3D Manhattan skipped."`

```{r genomic-manhattan-pai, eval=has_genomic && has_vep_primateai && any(!is.na(gene_genomic$mean_pai_char)), fig.height=5, fig.width=14}
ggplot(gene_genomic %>% filter(!is.na(mean_pai_char)),
       aes(x = manhattan_x, y = gene_caas_score,
           size   = mean_pai_char,
           colour = mean_pai_char)) +
  geom_point(alpha = 0.70) +
  geom_hline(yintercept = g_thr_10, linetype = "dashed",
             colour = "grey60", linewidth = 0.5) +
  ggrepel::geom_text_repel(
    data = gene_genomic %>%
      filter(!is.na(mean_pai_char)) %>%
      slice_max(mean_pai_char, n = 20, with_ties = FALSE),
    aes(label = Gene), size = 2.2, max.overlaps = 12,
    show.legend = FALSE, colour = "black"
  ) +
  scale_colour_viridis_c(name = "Mean PAI3D\nchar score", option = "inferno") +
  scale_size_continuous(name = "Mean PAI3D\nchar score", range = c(0.5, 6)) +
  scale_x_continuous(
    breaks = chr_label_positions$mid_x,
    labels = chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  guides(colour = guide_legend(), size = guide_legend()) +
  labs(
    title    = "Gene CAAS Score + PrimateAI-3D Characterisation — Manhattan View",
    subtitle = "Bubble size and colour: mean scheme-weighted PAI3D characterisation score across positions",
    x = "Chromosome", y = "Gene CAAS Score"
  ) +
  theme(
    plot.title      = element_text(face = "bold"),
    axis.text.x     = element_text(size = 7),
    panel.grid.major.x = element_blank()
  )
```

---

## Summary

```{r summary-stats}
n_components <- 4L  # biochem, asr, convergence, parallel

n_gene_components <- 1L  # gene_caas always present
if (has_rand)     n_gene_components <- n_gene_components + 1L
if (has_rer)      n_gene_components <- n_gene_components + 1L
if (has_fade)     n_gene_components <- n_gene_components + 1L
if (has_molerate) n_gene_components <- n_gene_components + 1L

summary_tbl <- tibble(
  Metric = c(
    "Positions scored", "Genes scored",
    "Position-level components", "Gene-level components",
    "CAAS Score (median)", "CAAS Score (top 10% threshold)",
    "Gene Composite (median)", "Gene Composite (top 10% threshold)",
    "FADE available",
    "Randomization available", "RERConverge available",
    "MoleRate available", "VEP PrimateAI-3D available",
    "Genomic coordinates available"
  ),
  Value = c(
    format(nrow(pos_df), big.mark = ","),
    format(nrow(gene_df), big.mark = ","),
    as.character(n_components),
    as.character(n_gene_components),
    sprintf("%.4f", median(pos_df$CAAS_score, na.rm = TRUE)),
    sprintf("%.4f", thr_10),
    sprintf("%.4f", median(gene_df$gene_composite_score, na.rm = TRUE)),
    sprintf("%.4f", gc_thr_10),
    ifelse(has_fade,         "Yes", "No"),
    ifelse(has_rand,         "Yes", "No"),
    ifelse(has_rer,          "Yes", "No"),
    ifelse(has_molerate,     "Yes", "No"),
    ifelse(has_vep_primateai,"Yes", "No"),
    ifelse(has_genomic,      "Yes", "No")
  )
)

knitr::kable(summary_tbl, align = c("l", "r"))
```

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>
