---
title: "CAAS Scoring Report"
subtitle: "`r params$traitname`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
params:
  position_scores_file: NULL
  gene_scores_file: NULL
  gene_corr_file: NULL
  traitname: "unknown_trait"
  output_dir: "."
  top_pct: 0.10
  top1_pct: 0.01
  gene_top_pct: 0.10
  gene_top1_pct: 0.01
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  out.width = "100%"
)

suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(paletteer)
  library(DT)
})

# Tableau_10 colour swatch for pipeline-wide consistency
.tableau10 <- as.character(paletteer_d("ggthemes::Tableau_10"))
theme_set(theme_minimal(base_size = 13))
```

```{r load-data}
pos_df  <- read_tsv(params$position_scores_file, show_col_types = FALSE)
gene_df <- read_tsv(params$gene_scores_file, show_col_types = FALSE)
corr_df <- read_tsv(params$gene_corr_file, show_col_types = FALSE)

# Detect which optional components are available
has_pgls <- any(!is.na(pos_df$pgls_score))
has_rand <- any(!is.na(gene_df$gene_rand_score))
has_rer  <- any(!is.na(gene_df$gene_rer_score))
has_fade <- "gene_fade_score" %in% names(gene_df) && any(!is.na(gene_df$gene_fade_score))

# Thresholds
top_pct  <- params$top_pct
top1_pct <- params$top1_pct
```

---

## Introduction

The **CAAS Scoring** module integrates signals from multiple upstream analyses
into a unified, interpretable metric for prioritising convergent amino acid
substitutions (CAAS).  The scoring operates at two levels:

1. **Position-level CAAS Score** — each alignment position receives a composite
   score that combines:
   - **Biochemical score** — amino acid variability and pattern association,
     weighted by grouping scheme specificity.
   - **ASR score** — ancestral state reconstruction confidence: novel changes
     score 1.0; ASR-confirmed conserved changes are weighted by MRCA posterior
     (max 0.70); contradicted cases score 0.
   - **Convergence score** — type and directionality of the convergent change.
   - **Parallel score** — parallelism pattern weighted by MRCA posterior
     confidence.
   `r if (has_pgls) "- **PGLS score** — significance of phylogenetic least squares association."`

2. **Gene-level scores** — each gene receives up to four independent scores:
   - **gene_caas_score** — 90th percentile of position-level CAAS scores.
   `r if (has_rand) "- **gene_rand_score** — randomization enrichment across grouping schemes."`
   `r if (has_rer) "- **gene_rer_score** — RERConverge evolutionary rate correlation."`
   `r if (has_fade) "- **gene_fade_score** — directional selection signal (FADE max Bayes Factor, gene-level)."`
   - **gene_composite_score** — unweighted mean of available gene scores.

All component scores are mapped to a 0--1 scale using decile ranking, making
them directly comparable and suitable for downstream enrichment analysis.

---

## Scoring Methodology

This section documents the exact weighting and combination logic used to
produce every score in this report, so results can be interpreted and
reproduced independently of the pipeline code.

<details>
<summary><strong>Click to expand full methodology</strong></summary>

---

### Position-Level Components

All position-level component scores are mapped to the **[0, 1]** interval.
The four core components are always computed; PGLS and FADE are included only
when the corresponding upstream module has been run.

#### 1 · Variability score

Decile rank of the **hypergeometric p-value** (`Pvalue`) across all
Gene × Position × Scheme rows, inverted so that smaller p-values receive
higher scores.

$$\text{variability\_score} = 1 - \frac{\text{ntile}(P_{\text{hyp}}, 10) - 1}{9}$$

*Range: 0.0 (least significant decile) → 1.0 (most significant decile).*

> This score does **not** enter `CAAS_score` directly — it feeds into
> `biochem_score` as a multiplicative factor.

---

#### 2 · Pattern score

Decile rank of the **permutation p-value** (`pvalue_boot`), inverted in the
same way.

$$\text{pattern\_score} = 1 - \frac{\text{ntile}(P_{\text{boot}}, 10) - 1}{9}$$

> Same role as variability\_score: multiplicative input into `biochem_score`.

---

#### 3 · Biochemical score

Sums the contribution of each grouping **scheme** in which the position was
detected, weighted by scheme specificity:

| Scheme | Weight |
|--------|-------:|
| US     | 0.50   |
| GS4    | 0.20   |
| GS3    | 0.10   |
| GS2    | 0.10   |
| GS1    | 0.10   |
| GS0    | 0.00   |

Each scheme contributes:

$$\text{contribution}_s = w_s \times \text{variability\_score} \times \text{pattern\_score}$$

The final score is the **cumulative sum** over all schemes where the position
was detected:

$$\text{biochem\_score} = \sum_{s \in \text{detected schemes}} w_s \times \text{variability\_score} \times \text{pattern\_score}$$

Maximum value when detected in all weighted schemes (US + GS4 + GS3 + GS2 + GS1)
with variability = pattern = 1.0 is **1.0** (0.5 + 0.2 + 0.1 + 0.1 + 0.1).

---

#### 4 · ASR score

Measures ancestral-state-reconstruction confidence. Three mutually exclusive cases:

| Condition | Score |
|-----------|------:|
| `is_conserved_meta = FALSE` (novel change) | **1.0** |
| `is_conserved_meta = TRUE` **and** `asr_is_conserved = TRUE` | see formula below |
| `is_conserved_meta = TRUE` **and** `asr_is_conserved = FALSE` | **0.0** |

For ASR-confirmed conserved positions, the score uses posterior probabilities
from the MRCA node identified by `conserved_pair`:

$$\text{asr\_score} = 0.35 \times P_{\text{MRCA}(\text{conserved\_pair})} + \begin{cases} 0.35 \times P_{\text{root}} & \text{if } \texttt{asr\_root\_conserved} = \text{TRUE} \\ 0 & \text{otherwise} \end{cases}$$

Maximum = **0.70** (when both posteriors = 1.0). Novel changes score 1.0,
making them always preferred over conserved changes.

---

#### 5 · Convergence score

Scores the **type and directionality** of the amino acid change on each
assessed side (top / bottom foreground lineages).

Side values:

| Change type | Value |
|-------------|------:|
| convergent  | 1.0   |
| codivergent | 0.7   |
| divergent   | 0.3   |
| ambiguous / no change | excluded (NA) |

$$\text{mean\_change} = \text{mean of non-NA side values}$$

$$\text{bidirectional} = \begin{cases} 1.0 & \text{both sides assessed} \\ 0.7 & \text{one side only} \end{cases}$$

$$\text{convergence\_score} = \frac{\text{mean\_change} + \text{bidirectional}}{2}$$

Range: 0.5 (one side, divergent: (0.3 + 0.7) / 2) → 1.0 (both convergent:
(1.0 + 1.0) / 2).

---

#### 6 · Parallel score

Scores whether the amino acid change is **parallel** (same substitution in
independent lineages) or **non-parallel**, weighted by MRCA posterior
confidence.

Parallelism coefficients (note: non-parallel receives the *highest* weight
because independent parallel changes are less informative as a signal of
convergence):

| Parallel type | Coefficient |
|---------------|------------:|
| non-parallel  | 1.0         |
| mixed         | 0.7         |
| parallel      | 0.5         |

For each assessable side (where `change_top`/`change_bottom` is
convergent / codivergent / divergent):

$$\text{side\_score} = \text{coeff} \times \overline{P_{\text{MRCA}}}$$

where $\overline{P_{\text{MRCA}}}$ is the mean of `mrca_1_posterior`,
`mrca_2_posterior`, `mrca_3_posterior` (NA excluded).

$$\text{mean\_parallel} = \text{mean of non-NA side scores}$$

$$\text{bidirectional} = \begin{cases} 1.0 & \text{both sides assessed} \\ 0.7 & \text{one side only} \end{cases}$$

$$\text{parallel\_score} = \frac{\text{mean\_parallel} + \text{bidirectional}}{2}$$

---

#### 7 · PGLS score  *(optional)*

Decile rank of the site-level phylogenetic GLS p-value (`p_pgls`), inverted:

$$\text{pgls\_score} = 1 - \frac{\text{ntile}(p_{\text{pgls}}, 10) - 1}{9}$$

Positions from genes not tested by PGLS receive `NA` and are excluded from
`CAAS_score` when PGLS is not run globally.

---

#### Composite CAAS score

Unweighted mean of the four core components plus PGLS when available.
FADE is **not** included here — it is a gene-level signal (see below).

$$\text{CAAS\_score} = \frac{\text{biochem} + \text{asr} + \text{convergence} + \text{parallel} + [\text{pgls}]}{N_{\text{components}}}$$

---

### Gene-Level Scores

#### gene\_caas\_score

90th percentile of `CAAS_score` across all scored positions in that gene:

$$\text{gene\_caas\_score} = Q_{0.90}(\text{CAAS\_score})$$

#### gene\_rand\_score  *(optional)*

Scheme-weighted sum of decile-ranked accumulation randomization p-values.
The same scheme weights as `biochem_score` are used (GS0 excluded):

$$\text{gene\_rand\_score} = \sum_{s \in \{US, GS4, GS3, GS2, GS1\}} w_s \times \left(1 - \frac{\text{ntile}(P_{\text{empirical},s}, 10) - 1}{9}\right)$$

Maximum = 1.0 (all schemes p-value in best decile).

#### gene\_rer\_score  *(optional)*

Decile rank of the RERConverge adjusted p-value (`p.adj`; or `p.perm` if
available and non-zero), inverted:

$$\text{gene\_rer\_score} = 1 - \frac{\text{ntile}(p_{\text{RER}}, 10) - 1}{9}$$

#### gene\_fade\_score  *(optional)*

Gene-level maximum Bayes Factor from FADE directional selection analysis,
taking the best of the top/bottom foreground directions. Decile rank,
higher BF = higher score:

$$\text{gene\_fade\_score} = \frac{\text{ntile}(\max_{d \in \{\text{top,bottom}\}} \text{BF}_{d}, 10) - 1}{9}$$

#### gene\_composite\_score

Unweighted mean of all available gene-level sub-scores:

$$\text{gene\_composite\_score} = \frac{\text{gene\_caas} + [\text{gene\_rand}] + [\text{gene\_rer}] + [\text{gene\_fade}]}{N_{\text{available}}}$$

---

### ORA Gene List Extraction

Gene lists for Over-Representation Analysis are extracted at two thresholds
(configurable via `--scoring_position_top_pct` / `--scoring_position_top1_pct`
and their gene-level equivalents) and three directional subsets:

| Subset | Filter |
|--------|--------|
| **top** | Positions / genes where `change_side == "top"` |
| **bottom** | Positions / genes where `change_side == "bottom"` |
| **both** | All top-scoring positions / genes regardless of direction |

For **position-level** lists, the threshold is applied to `CAAS_score` and
the gene name is extracted from the qualifying positions.

For **gene-level** lists, the threshold is applied to each gene-level score
independently (`gene_caas_score`, `gene_rand_score`, `gene_rer_score`,
`gene_composite_score`), and direction is assigned from the gene's
highest-scoring position.

</details>

---

## Position-Level Scores {.tabset}

### CAAS Score Distribution

```{r caas-score-dist, fig.height=5}
thr_10 <- quantile(pos_df$CAAS_score, 1 - top_pct,  na.rm = TRUE)
thr_1  <- quantile(pos_df$CAAS_score, 1 - top1_pct, na.rm = TRUE)

ggplot(pos_df, aes(x = CAAS_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50,
                 fill = .tableau10[1], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[2]) +
  geom_vline(xintercept = thr_10, linetype = "dashed",
             colour = .tableau10[3], linewidth = 0.7) +
  geom_vline(xintercept = thr_1,  linetype = "dotted",
             colour = .tableau10[4], linewidth = 0.7) +
  annotate("text", x = thr_10, y = Inf, label = sprintf("Top %g%%", top_pct * 100),
           vjust = 2, hjust = -0.1, colour = .tableau10[3], size = 3.5) +
  annotate("text", x = thr_1,  y = Inf, label = sprintf("Top %g%%", top1_pct * 100),
           vjust = 3.5, hjust = -0.1, colour = .tableau10[4], size = 3.5) +
  labs(title = "CAAS Score Distribution",
       subtitle = sprintf("n = %s positions | Trait: %s",
                          format(nrow(pos_df), big.mark = ","), params$traitname),
       x = "CAAS Score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### Biochemical Score

```{r biochem-dist, fig.height=4.5}
ggplot(pos_df, aes(x = biochem_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = .tableau10[1], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[2]) +
  labs(title = "Biochemical Score Distribution",
       subtitle = "Weighted sum of scheme-specific variability x pattern contributions",
       x = "biochem_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### ASR Score

```{r asr-dist, fig.height=4.5}
ggplot(pos_df, aes(x = asr_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = .tableau10[3], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[4]) +
  labs(title = "ASR Score Distribution",
       subtitle = "Novel change = 1.0 | ASR-confirmed conserved = 0.35×posterior (max 0.70) | Contradicted = 0",
       x = "asr_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### Convergence Score

```{r convergence-dist, fig.height=4.5}
ggplot(pos_df, aes(x = convergence_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = .tableau10[5], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[6]) +
  labs(title = "Convergence Score Distribution",
       subtitle = "Convergence type + bidirectional bonus",
       x = "convergence_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### Parallel Score

```{r parallel-dist, fig.height=4.5}
ggplot(pos_df, aes(x = parallel_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = .tableau10[7], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[8]) +
  labs(title = "Parallel Score Distribution",
       subtitle = "Parallelism coefficient x MRCA posterior + bidirectional bonus",
       x = "parallel_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

`r if (has_pgls) "### PGLS Score"`

```{r pgls-dist, eval=has_pgls, fig.height=4.5}
ggplot(pos_df %>% filter(!is.na(pgls_score)), aes(x = pgls_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = .tableau10[9], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[10]) +
  labs(title = "PGLS Score Distribution",
       subtitle = "Deciles of phylogenetic GLS p-values",
       x = "pgls_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### Component Overview

```{r component-overview, fig.height=6}
# Faceted density comparison of all position-level components
component_cols <- c("biochem_score", "asr_score", "convergence_score", "parallel_score")
if (has_pgls) component_cols <- c(component_cols, "pgls_score")

long_df <- pos_df %>%
  select(all_of(component_cols)) %>%
  pivot_longer(everything(), names_to = "Component", values_to = "Score") %>%
  filter(!is.na(Score)) %>%
  mutate(Component = factor(Component, levels = component_cols))

ggplot(long_df, aes(x = Score, fill = Component)) +
  geom_density(alpha = 0.6, linewidth = 0.3) +
  scale_fill_manual(values = .tableau10[seq_along(component_cols)]) +
  facet_wrap(~Component, scales = "free_y", ncol = 2) +
  labs(title = "Position-Level Component Score Distributions",
       x = "Score (0-1)", y = "Density") +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "none")
```

### Threshold Profiles

```{r threshold-profiles, fig.height=14, fig.width=12}
# Component score distributions for positions inside each ORA gene-list cutoff.
# Each panel shows overlaid densities of all position-level components so you
# can read off which signal drives positions into the top-N% × direction bucket.

component_cols_tp <- c("biochem_score", "asr_score", "convergence_score", "parallel_score")
if (has_pgls) component_cols_tp <- c(component_cols_tp, "pgls_score")

thr_10_tp <- quantile(pos_df$CAAS_score, 1 - top_pct,  na.rm = TRUE)
thr_1_tp  <- quantile(pos_df$CAAS_score, 1 - top1_pct, na.rm = TRUE)

make_threshold_panel <- function(df, threshold, pct_label, direction) {
  subset_df <- if (direction == "both") {
    df %>% filter(CAAS_score >= threshold)
  } else {
    df %>% filter(CAAS_score >= threshold, change_side == direction)
  }

  if (nrow(subset_df) == 0) {
    return(
      ggplot() +
        annotate("text", x = 0.5, y = 0.5, label = "No positions",
                 size = 5, colour = "grey50") +
        labs(title = sprintf("Top %s — %s", pct_label, direction),
             x = "Score (0–1)", y = "Density") +
        xlim(0, 1) + theme_minimal()
    )
  }

  long_df <- subset_df %>%
    select(all_of(component_cols_tp)) %>%
    pivot_longer(everything(), names_to = "Component", values_to = "Score") %>%
    filter(!is.na(Score)) %>%
    mutate(Component = factor(Component, levels = component_cols_tp))

  ggplot(long_df, aes(x = Score, fill = Component, colour = Component)) +
    geom_density(alpha = 0.35, linewidth = 0.6) +
    scale_fill_manual(values   = .tableau10[seq_along(component_cols_tp)]) +
    scale_colour_manual(values = .tableau10[seq_along(component_cols_tp)]) +
    scale_x_continuous(limits = c(0, 1), expand = c(0.01, 0)) +
    labs(
      title    = sprintf("Top %s — %s", pct_label, direction),
      subtitle = sprintf("n = %d positions", nrow(subset_df)),
      x = "Score (0–1)", y = "Density",
      fill = NULL, colour = NULL
    ) +
    theme(
      plot.title    = element_text(face = "bold", size = 11),
      plot.subtitle = element_text(size = 9, colour = "grey40"),
      legend.position = "bottom",
      legend.text     = element_text(size = 8)
    )
}

panels <- list()
for (direction in c("top", "bottom", "both")) {
  panels[[paste0("10pct_", direction)]] <-
    make_threshold_panel(pos_df, thr_10_tp, "10%", direction)
  panels[[paste0("1pct_",  direction)]] <-
    make_threshold_panel(pos_df, thr_1_tp,  "1%",  direction)
}

# Layout: rows = direction (top / bottom / both), cols = threshold (10% | 1%)
gridExtra::grid.arrange(
  panels[["10pct_top"]],    panels[["1pct_top"]],
  panels[["10pct_bottom"]], panels[["1pct_bottom"]],
  panels[["10pct_both"]],   panels[["1pct_both"]],
  ncol = 2,
  top  = grid::textGrob(
    "Component score profiles for top-scoring position subsets",
    gp = grid::gpar(fontface = "bold", fontsize = 13)
  ),
  left = grid::textGrob(
    "Direction:  top                          bottom                          both",
    rot = 90, gp = grid::gpar(fontsize = 9, col = "grey40")
  )
)
```

### Top Positions

```{r top-positions-table}
display_cols <- c("Gene", "Position", "CAAS_score", "biochem_score",
                  "asr_score", "convergence_score", "parallel_score")
if (has_pgls) display_cols <- c(display_cols, "pgls_score")
display_cols <- c(display_cols, "change_side")

DT::datatable(
  pos_df %>%
    arrange(desc(CAAS_score)) %>%
    select(all_of(display_cols)) %>%
    mutate(across(where(is.numeric), ~round(., 4))),
  filter    = "top",
  rownames  = FALSE,
  options   = list(pageLength = 25, scrollX = TRUE),
  caption   = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Position-level CAAS scores | Trait: %s", params$traitname)
  )
)
```

---

## Gene-Level Scores {.tabset}

### Gene CAAS Score

```{r gene-caas-dist, fig.height=4.5}
g_thr_10 <- quantile(gene_df$gene_caas_score, 1 - params$gene_top_pct, na.rm = TRUE)
g_thr_1  <- quantile(gene_df$gene_caas_score, 1 - params$gene_top1_pct, na.rm = TRUE)

ggplot(gene_df, aes(x = gene_caas_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = .tableau10[1], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[2]) +
  geom_vline(xintercept = g_thr_10, linetype = "dashed",
             colour = .tableau10[3], linewidth = 0.7) +
  geom_vline(xintercept = g_thr_1,  linetype = "dotted",
             colour = .tableau10[4], linewidth = 0.7) +
  labs(title = "Gene CAAS Score Distribution",
       subtitle = "90th percentile of position-level CAAS scores per gene",
       x = "gene_caas_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

```{r gene-rand-dist, eval=has_rand, fig.height=4.5}
ggplot(gene_df %>% filter(!is.na(gene_rand_score)), aes(x = gene_rand_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = .tableau10[5], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[6]) +
  labs(title = "Gene Randomization Score Distribution",
       subtitle = "Scheme-weighted deciles of accumulation randomization p-values",
       x = "gene_rand_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

```{r gene-rer-dist, eval=has_rer, fig.height=4.5}
ggplot(gene_df %>% filter(!is.na(gene_rer_score)), aes(x = gene_rer_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = .tableau10[7], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[8]) +
  labs(title = "Gene RERConverge Score Distribution",
       subtitle = "Deciles of evolutionary rate correlation p-values",
       x = "gene_rer_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

```{r gene-fade-dist, eval=has_fade, fig.height=4.5}
ggplot(gene_df %>% filter(!is.na(gene_fade_score)), aes(x = gene_fade_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = .tableau10[9], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[10]) +
  labs(title = "Gene FADE Score Distribution",
       subtitle = "Deciles of max directional selection Bayes Factor per gene",
       x = "gene_fade_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### Gene Composite Score

```{r gene-composite-dist, fig.height=5}
gc_thr_10 <- quantile(gene_df$gene_composite_score, 1 - params$gene_top_pct, na.rm = TRUE)
gc_thr_1  <- quantile(gene_df$gene_composite_score, 1 - params$gene_top1_pct, na.rm = TRUE)

ggplot(gene_df, aes(x = gene_composite_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = .tableau10[1], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[2]) +
  geom_vline(xintercept = gc_thr_10, linetype = "dashed",
             colour = .tableau10[3], linewidth = 0.7) +
  geom_vline(xintercept = gc_thr_1,  linetype = "dotted",
             colour = .tableau10[4], linewidth = 0.7) +
  annotate("text", x = gc_thr_10, y = Inf,
           label = sprintf("Top %g%%", params$gene_top_pct * 100),
           vjust = 2, hjust = -0.1, colour = .tableau10[3], size = 3.5) +
  annotate("text", x = gc_thr_1,  y = Inf,
           label = sprintf("Top %g%%", params$gene_top1_pct * 100),
           vjust = 3.5, hjust = -0.1, colour = .tableau10[4], size = 3.5) +
  labs(title = "Gene Composite Score Distribution",
       subtitle = sprintf("Unweighted mean of available gene scores | n = %s genes",
                          format(nrow(gene_df), big.mark = ",")),
       x = "gene_composite_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### Gene Threshold Profiles

```{r gene-threshold-profiles, fig.height=14, fig.width=12}
# Same idea as position-level threshold profiles but for gene-level scores.
# For each top-N% × direction bucket, show overlaid densities of gene sub-scores.

gene_score_cols_tp <- "gene_caas_score"
if (has_rand) gene_score_cols_tp <- c(gene_score_cols_tp, "gene_rand_score")
if (has_rer)  gene_score_cols_tp <- c(gene_score_cols_tp, "gene_rer_score")
if (has_fade) gene_score_cols_tp <- c(gene_score_cols_tp, "gene_fade_score")

# Direction per gene is inherited from its top-scoring position
gene_direction_tp <- pos_df %>%
  group_by(Gene) %>%
  slice_max(CAAS_score, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(Gene, gene_direction = change_side)

gene_df_dir <- gene_df %>% left_join(gene_direction_tp, by = "Gene")

gc_thr_10_tp <- quantile(gene_df$gene_composite_score, 1 - params$gene_top_pct,  na.rm = TRUE)
gc_thr_1_tp  <- quantile(gene_df$gene_composite_score, 1 - params$gene_top1_pct, na.rm = TRUE)

make_gene_threshold_panel <- function(df, threshold, pct_label, direction) {
  subset_df <- if (direction == "both") {
    df %>% filter(gene_composite_score >= threshold)
  } else {
    df %>% filter(gene_composite_score >= threshold, gene_direction == direction)
  }

  if (nrow(subset_df) == 0 || length(gene_score_cols_tp) < 2) {
    return(
      ggplot() +
        annotate("text", x = 0.5, y = 0.5,
                 label = ifelse(nrow(subset_df) == 0, "No genes", "Single score only"),
                 size = 5, colour = "grey50") +
        labs(title = sprintf("Top %s — %s", pct_label, direction),
             x = "Score (0–1)", y = "Density") +
        xlim(0, 1) + theme_minimal()
    )
  }

  long_df <- subset_df %>%
    select(all_of(gene_score_cols_tp)) %>%
    pivot_longer(everything(), names_to = "Score_type", values_to = "Score") %>%
    filter(!is.na(Score)) %>%
    mutate(Score_type = factor(Score_type, levels = gene_score_cols_tp))

  ggplot(long_df, aes(x = Score, fill = Score_type, colour = Score_type)) +
    geom_density(alpha = 0.35, linewidth = 0.6) +
    scale_fill_manual(values   = .tableau10[seq_along(gene_score_cols_tp)],
                      labels   = sub("gene_", "", sub("_score", "", gene_score_cols_tp))) +
    scale_colour_manual(values = .tableau10[seq_along(gene_score_cols_tp)],
                        labels = sub("gene_", "", sub("_score", "", gene_score_cols_tp))) +
    scale_x_continuous(limits = c(0, 1), expand = c(0.01, 0)) +
    labs(
      title    = sprintf("Top %s — %s", pct_label, direction),
      subtitle = sprintf("n = %d genes", nrow(subset_df)),
      x = "Score (0–1)", y = "Density",
      fill = NULL, colour = NULL
    ) +
    theme(
      plot.title      = element_text(face = "bold", size = 11),
      plot.subtitle   = element_text(size = 9, colour = "grey40"),
      legend.position = "bottom",
      legend.text     = element_text(size = 8)
    )
}

gene_panels <- list()
for (direction in c("top", "bottom", "both")) {
  gene_panels[[paste0("10pct_", direction)]] <-
    make_gene_threshold_panel(gene_df_dir, gc_thr_10_tp, "10%", direction)
  gene_panels[[paste0("1pct_",  direction)]] <-
    make_gene_threshold_panel(gene_df_dir, gc_thr_1_tp,  "1%",  direction)
}

gridExtra::grid.arrange(
  gene_panels[["10pct_top"]],    gene_panels[["1pct_top"]],
  gene_panels[["10pct_bottom"]], gene_panels[["1pct_bottom"]],
  gene_panels[["10pct_both"]],   gene_panels[["1pct_both"]],
  ncol = 2,
  top  = grid::textGrob(
    "Gene score profiles for top-scoring gene subsets (by gene_composite_score)",
    gp = grid::gpar(fontface = "bold", fontsize = 13)
  ),
  left = grid::textGrob(
    "Direction:  top                          bottom                          both",
    rot = 90, gp = grid::gpar(fontsize = 9, col = "grey40")
  )
)
```

### Correlation Analysis

```{r correlation-scatter, eval=(nrow(corr_df) > 0), fig.height=7}
# Build scatter matrix of gene-level scores
score_cols <- "gene_caas_score"
if (has_rand) score_cols <- c(score_cols, "gene_rand_score")
if (has_rer)  score_cols <- c(score_cols, "gene_rer_score")
if (has_fade) score_cols <- c(score_cols, "gene_fade_score")

if (length(score_cols) >= 2) {
  pairs_df <- gene_df %>%
    select(all_of(score_cols)) %>%
    drop_na()

  # Create pairs plot
  pairs_long <- expand.grid(
    x_var = score_cols, y_var = score_cols,
    stringsAsFactors = FALSE
  ) %>%
    filter(x_var != y_var)

  plots <- list()
  for (i in seq_len(nrow(pairs_long))) {
    xv <- pairs_long$x_var[i]
    yv <- pairs_long$y_var[i]
    r_info <- corr_df %>%
      filter((score_a == xv & score_b == yv) | (score_a == yv & score_b == xv))
    label <- if (nrow(r_info) > 0) {
      sprintf("r=%.2f / rho=%.2f", r_info$pearson_r[1], r_info$spearman_r[1])
    } else ""

    plots[[length(plots) + 1]] <- ggplot(pairs_df, aes(x = .data[[xv]], y = .data[[yv]])) +
      geom_point(alpha = 0.3, size = 0.8, colour = .tableau10[1]) +
      geom_smooth(method = "lm", se = FALSE, colour = .tableau10[4], linewidth = 0.6) +
      annotate("text", x = Inf, y = -Inf, label = label,
               hjust = 1.1, vjust = -0.5, size = 3, colour = .tableau10[4]) +
      labs(x = sub("gene_", "", sub("_score", "", xv)),
           y = sub("gene_", "", sub("_score", "", yv))) +
      theme(axis.title = element_text(size = 9))
  }

  if (length(plots) > 0) {
    gridExtra::grid.arrange(grobs = plots,
                            ncol = length(score_cols) - 1,
                            top = grid::textGrob("Gene Score Correlations",
                                                 gp = grid::gpar(fontface = "bold", fontsize = 14)))
  }
}
```

```{r correlation-table, eval=(nrow(corr_df) > 0)}
DT::datatable(
  corr_df %>%
    mutate(across(where(is.numeric), ~round(., 4))),
  rownames = FALSE,
  options  = list(pageLength = 10, dom = "t"),
  caption  = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "Pairwise correlations between gene-level scores"
  )
)
```

### Top Genes

```{r top-genes-table}
gene_display_cols <- c("Gene", "n_positions", "gene_caas_score")
if (has_rand) gene_display_cols <- c(gene_display_cols, "gene_rand_score")
if (has_rer)  gene_display_cols <- c(gene_display_cols, "gene_rer_score")
if (has_fade) gene_display_cols <- c(gene_display_cols, "gene_fade_score")
gene_display_cols <- c(gene_display_cols, "gene_composite_score")

DT::datatable(
  gene_df %>%
    arrange(desc(gene_composite_score)) %>%
    select(all_of(gene_display_cols)) %>%
    mutate(across(where(is.numeric) & !matches("n_positions"), ~round(., 4))),
  filter    = "top",
  rownames  = FALSE,
  options   = list(pageLength = 25, scrollX = TRUE),
  caption   = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Gene-level scores | Trait: %s", params$traitname)
  )
)
```

---

## Summary

```{r summary-stats}
n_components <- 4L  # biochem, asr, convergence, parallel
if (has_pgls) n_components <- n_components + 1L

n_gene_components <- 1L  # gene_caas always present
if (has_rand) n_gene_components <- n_gene_components + 1L
if (has_rer)  n_gene_components <- n_gene_components + 1L
if (has_fade) n_gene_components <- n_gene_components + 1L

summary_tbl <- tibble(
  Metric = c(
    "Positions scored", "Genes scored",
    "Position-level components", "Gene-level components",
    "CAAS Score (median)", "CAAS Score (top 10% threshold)",
    "Gene Composite (median)", "Gene Composite (top 10% threshold)",
    "PGLS available", "FADE available",
    "Randomization available", "RERConverge available"
  ),
  Value = c(
    format(nrow(pos_df), big.mark = ","),
    format(nrow(gene_df), big.mark = ","),
    as.character(n_components),
    as.character(n_gene_components),
    sprintf("%.4f", median(pos_df$CAAS_score, na.rm = TRUE)),
    sprintf("%.4f", thr_10),
    sprintf("%.4f", median(gene_df$gene_composite_score, na.rm = TRUE)),
    sprintf("%.4f", gc_thr_10),
    ifelse(has_pgls, "Yes", "No"),
    ifelse(has_fade, "Yes", "No"),
    ifelse(has_rand, "Yes", "No"),
    ifelse(has_rer,  "Yes", "No")
  )
)

knitr::kable(summary_tbl, align = c("l", "r"))
```

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>
