---
title: "CAAS Scoring Report"
subtitle: "`r paste0(params$traitname, ' — ', toupper(params$direction), '-phenotype direction')`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: false
params:
  position_scores_file: NULL
  gene_scores_file: NULL
  gene_corr_file: NULL
  traitname: "unknown_trait"
  output_dir: "."
  top_pct: 0.10
  top5_pct: 0.05
  top1_pct: 0.01
  gene_top_pct: 0.10
  gene_top5_pct: 0.05
  gene_top1_pct: 0.01
  stress_summary_file: NULL
  stress_corr_file: NULL
  stress_rank_file: NULL
  stress_overlap_file: NULL
  stress_variants_file: NULL
  stress_loadings_file: NULL
  fade_site_top_file: NULL
  fade_site_bot_file: NULL
  vep_transvar_file: NULL
  vep_primateai_file: NULL
  vep_aa2prot_file: NULL
  genomic_info_file: NULL
  window_size_bp: 1000000
  direction: "both"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  out.width = "100%"
)

suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(DT)
  library(scales)
  library(ggrepel)
})

# Tableau_10 colour swatch for pipeline-wide consistency
if (requireNamespace("paletteer", quietly = TRUE)) {
  .tableau10 <- as.character(paletteer::paletteer_d("ggthemes::Tableau_10"))
} else {
  .tableau10 <- c(
    "#4E79A7", "#F28E2B", "#E15759", "#76B7B2", "#59A14F",
    "#EDC948", "#B07AA1", "#FF9DA7", "#9C755F", "#BAB0AC"
  )
}
theme_set(theme_minimal(base_size = 13))

# Safe correlation helper: returns NA (not an error) when inputs have fewer
# than 3 complete pairs — avoids crashes in sparse optional-module sections.
safe_cor <- function(x, y, method = "pearson") {
  ok <- complete.cases(x, y)
  if (sum(ok) < 3) return(NA_real_)
  cor(x[ok], y[ok], method = method)
}

format_bound <- function(x, digits = 2) {
  ifelse(
    is.na(x),
    NA_character_,
    ifelse(abs(x) < 1e-3 | abs(x) >= 1e3,
           formatC(x, format = "e", digits = digits),
           format(round(x, digits), nsmall = digits, trim = TRUE))
  )
}

decile_boundary_tbl <- function(values, score_name,
                                lower_better = TRUE,
                                value_label = "Value",
                                digits = 2) {
  vals <- values[is.finite(values) & !is.na(values)]
  if (length(vals) < 10) return(NULL)
  breaks <- quantile(vals, probs = seq(0, 1, 0.1), na.rm = TRUE, names = FALSE)
  deciles <- if (lower_better) 10:1 else 1:10
  # For lower-better metrics (e.g. p-values), decile 10 should map to the
  # smallest interval [q0, q10%], decile 9 to [q10%, q20%], ..., decile 1 to
  # the largest interval [q90%, q100%].
  lower_bounds <- breaks[1:10]
  upper_bounds <- c(breaks[2:10], max(vals, na.rm = TRUE))
  score_vals <- (deciles - 1) / 9
  tibble(
    Decile = deciles,
    `Lower bound` = format_bound(lower_bounds, digits = digits),
    `Upper bound` = format_bound(upper_bounds, digits = digits),
    !!score_name := round(score_vals, 3)
  ) %>%
    rename(
      !!paste(value_label, "lower bound") := `Lower bound`,
      !!paste(value_label, "upper bound") := `Upper bound`
    )
}
```

```{r load-data}
pos_df  <- read_tsv(params$position_scores_file, show_col_types = FALSE)
gene_df <- read_tsv(params$gene_scores_file, show_col_types = FALSE)
corr_df <- read_tsv(params$gene_corr_file, show_col_types = FALSE)

# Some runs export raw p-values; others export only the derived decile scores.
# Pick the available measure for each distribution section.
hyper_measure_col <- if ("Pvalue" %in% names(pos_df)) {
  "Pvalue"
} else if ("variability_score" %in% names(pos_df)) {
  "variability_score"
} else {
  stop("position_scores.tsv must contain either Pvalue or variability_score")
}
boot_measure_col <- if ("pvalue_boot" %in% names(pos_df)) {
  "pvalue_boot"
} else if ("pattern_score" %in% names(pos_df)) {
  "pattern_score"
} else {
  stop("position_scores.tsv must contain either pvalue_boot or pattern_score")
}
hyper_measure_is_raw <- identical(hyper_measure_col, "Pvalue")
boot_measure_is_raw  <- identical(boot_measure_col,  "pvalue_boot")

# Detect which optional components are available
has_rand     <- "gene_rand_score"     %in% names(gene_df) && any(!is.na(gene_df$gene_rand_score))
has_rer      <- "gene_rer_score"      %in% names(gene_df) && any(!is.na(gene_df$gene_rer_score))
has_fade     <- "gene_fade_score"     %in% names(gene_df) && any(!is.na(gene_df$gene_fade_score))

# ── VEP optional inputs ───────────────────────────────────────────────────────
.file_ok <- function(p) !is.null(p) && !is.na(p) && nchar(p) > 0 && file.exists(p)

has_scoring_stress <- .file_ok(params$stress_summary_file) &&
  .file_ok(params$stress_corr_file) &&
  .file_ok(params$stress_rank_file) &&
  .file_ok(params$stress_overlap_file) &&
  .file_ok(params$stress_variants_file)
has_vep_primateai <- .file_ok(params$vep_primateai_file)
has_aa2prot       <- .file_ok(params$vep_aa2prot_file)
has_genomic       <- .file_ok(params$genomic_info_file)
# FADE site data: use top direction if available, otherwise try bottom
has_fade_pos      <- .file_ok(params$fade_site_top_file) || .file_ok(params$fade_site_bot_file)
# TransVar is no longer used in the scoring report (removed — it added no actionable signal)
has_vep_transvar  <- FALSE

if (has_vep_primateai) {
  vep_pai_df <- read_tsv(params$vep_primateai_file, show_col_types = FALSE)
} else {
  vep_pai_df <- NULL
}

# aa2prot: tab-separated, no header — Gene | aln_pos (0-based) | caas_id | Gene:p.N
# Used to bridge alignment positions in pos_df to 1-based protein positions used by
# PrimateAI-3D lookup (map_to_primateai.py outputs transvar_query in this format).
if (has_aa2prot) {
  aa2prot_raw <- read_tsv(params$vep_aa2prot_file,
                          col_names = c("Gene", "aln_pos", "caas_id", "prot_query"),
                          show_col_types = FALSE)
  pos_prot_map <- aa2prot_raw %>%
    distinct(Gene, aln_pos, prot_query) %>%
    rename(Position = aln_pos)
} else {
  pos_prot_map <- NULL
}

if (has_genomic) {
  # Format: gene, chr, start, end, length  (no header assumed; detect if header present)
  genomic_raw <- read_tsv(params$genomic_info_file,
                          col_names = c("Gene", "chr", "gene_start", "gene_end", "gene_length"),
                          show_col_types = FALSE)
  # Drop header row if first cell is literally "gene" or "Gene"
  if (tolower(genomic_raw$Gene[1]) == "gene") genomic_raw <- genomic_raw[-1, ]
  genomic_raw <- genomic_raw %>%
    mutate(
      gene_start  = as.integer(gene_start),
      gene_end    = as.integer(gene_end),
      gene_length = as.integer(gene_length),
      gene_mid    = as.integer((gene_start + gene_end) / 2L)
    )
} else {
  genomic_raw <- NULL
}

accum_scheme_info <- tibble(
  scheme = c("us", "gs4", "gs3", "gs2", "gs1"),
  label = c("US", "GS4", "GS3", "GS2", "GS1"),
  weight = c(0.5, 0.2, 0.1, 0.1, 0.1),
  pval_col = paste0("accum_pval_", c("us", "gs4", "gs3", "gs2", "gs1")),
  score_col = paste0("accum_score_", c("us", "gs4", "gs3", "gs2", "gs1")),
  weighted_col = paste0("accum_weighted_", c("us", "gs4", "gs3", "gs2", "gs1"))
) %>%
  filter(pval_col %in% names(gene_df))

if (has_scoring_stress) {
  stress_summary_df <- read_tsv(params$stress_summary_file, show_col_types = FALSE)
  stress_corr_df <- read_tsv(params$stress_corr_file, show_col_types = FALSE)
  stress_rank_df <- read_tsv(params$stress_rank_file, show_col_types = FALSE)
  stress_overlap_df <- read_tsv(params$stress_overlap_file, show_col_types = FALSE)
  stress_variants_df <- read_tsv(params$stress_variants_file, show_col_types = FALSE)
  stress_loadings_df <- if (.file_ok(params$stress_loadings_file)) {
    read_tsv(params$stress_loadings_file, show_col_types = FALSE)
  } else {
    NULL
  }
} else {
  stress_summary_df <- NULL
  stress_corr_df <- NULL
  stress_rank_df <- NULL
  stress_overlap_df <- NULL
  stress_variants_df <- NULL
  stress_loadings_df <- NULL
}

# Thresholds
top_pct  <- params$top_pct
top5_pct <- params$top5_pct
top1_pct <- params$top1_pct

# ── Create derived gene scores from raw values ──────────────────────────────
# Convert raw RER p-values to decile scores (lower p-value = higher score)
if ("rer_min_pval" %in% names(gene_df) && any(!is.na(gene_df$rer_min_pval))) {
  gene_df <- gene_df %>%
    mutate(
      gene_rer_score = {
        vals <- gene_df$rer_min_pval
        ranks <- dplyr::ntile(vals, 10)
        # Invert: lower p-values get higher scores
        1 - (ranks - 1) / 9
      }
    )
  has_rer <- any(!is.na(gene_df$gene_rer_score))
} else {
  has_rer <- FALSE
}

# Convert FADE max Bayes Factor to decile scores (higher BF = higher score)
has_fade_raw <- ("fade_max_bf_top" %in% names(gene_df) && any(!is.na(gene_df$fade_max_bf_top))) ||
                ("fade_max_bf_bottom" %in% names(gene_df) && any(!is.na(gene_df$fade_max_bf_bottom)))
if (has_fade_raw) {
  gene_df <- gene_df %>%
    mutate(
      fade_bf_for_scoring = case_when(
        !is.na(fade_max_bf_top)    ~ fade_max_bf_top,
        !is.na(fade_max_bf_bottom) ~ fade_max_bf_bottom,
        TRUE                       ~ NA_real_
      ),
      gene_fade_score = {
        vals <- gene_df$fade_bf_for_scoring
        ranks <- dplyr::ntile(vals, 10)
        # Direct mapping: higher BF gets higher score
        (ranks - 1) / 9
      }
    ) %>%
    select(-fade_bf_for_scoring)
  has_fade <- any(!is.na(gene_df$gene_fade_score))
} else {
  has_fade <- FALSE
}

# Create composite gene score: unweighted mean of available gene-level scores
score_cols <- c("gene_caas_score")
if ("gene_rand_score" %in% names(gene_df)) score_cols <- c(score_cols, "gene_rand_score")
if (has_rer && "gene_rer_score" %in% names(gene_df)) score_cols <- c(score_cols, "gene_rer_score")
if (has_fade && "gene_fade_score" %in% names(gene_df)) score_cols <- c(score_cols, "gene_fade_score")

if (length(score_cols) > 1) {
  gene_df <- gene_df %>%
    mutate(
      gene_composite_score = rowMeans(select(cur_data(), all_of(score_cols)), na.rm = TRUE)
    )
} else {
  # Only CAAS available
  gene_df <- gene_df %>%
    mutate(gene_composite_score = gene_caas_score)
}
```

---

## Introduction

The **CAAS Scoring** module integrates signals from multiple upstream analyses
into a unified, interpretable metric for prioritising convergent amino acid
substitutions (CAAS).  The scoring operates at two levels:

1. **Position-level CAAS Score** — each alignment position receives a composite
   score that combines:
   - **Biochemical score** — amino acid variability and pattern association,
     weighted by grouping scheme specificity.
   - **ASR score** — ancestral state reconstruction confidence: novel changes
     score 1.0; contradicted conserved cases score 0; ASR-supported conserved
     cases receive pair-weighted posterior support capped below novelty.
   - **Convergence score** — ordinal change-type support with a mild
     completeness penalty when only one side is assessable.
   - **Parallel score** — ordinal parallelism support weighted by MRCA
     posterior confidence and penalised mildly when only one side is assessable.

2. **Gene-level scores** — each gene receives up to five independent scores:
   - **gene_caas_score** — 90th percentile of position-level CAAS scores.
   `r if (has_rand) "- **gene_rand_score** — randomization enrichment across grouping schemes."`
   `r if (has_rer) "- **gene_rer_score** — RERConverge evolutionary rate correlation."`
   `r if (has_fade) "- **gene_fade_score** — directional selection signal (FADE max Bayes Factor, gene-level)."`
   - **gene_composite_score** — unweighted mean of available gene scores.

All component scores are mapped to a 0--1 scale using decile ranking, making
them directly comparable and suitable for downstream enrichment analysis.

---

## Scoring Methodology

This section documents the exact weighting and combination logic used to
produce every score in this report, so results can be interpreted and
reproduced independently of the pipeline code.

<details>
<summary><strong>Click to expand full methodology</strong></summary>

---

### Position-Level Components

All position-level component scores are mapped to the **[0, 1]** interval.
The four core components are always computed; FADE is included only
when the upstream module has been run.

#### 1 · Variability score

Decile rank of the **hypergeometric p-value** (`Pvalue`) across all
Gene × Position × Scheme rows, inverted so that smaller p-values receive
higher scores.

$$\text{variability\_score} = 1 - \frac{\text{ntile}(P_{\text{hyp}}, 10) - 1}{9}$$

*Range: 0.0 (least significant decile) → 1.0 (most significant decile).*

> This score does **not** enter `CAAS_score` directly — it feeds into
> `biochem_score` as a multiplicative factor.

---

#### 2 · Pattern score

Decile rank of the **permulation p-value** (`pvalue_boot`), inverted in the
same way.

$$\text{pattern\_score} = 1 - \frac{\text{ntile}(P_{\text{boot}}, 10) - 1}{9}$$

> Same role as variability\_score: multiplicative input into `biochem_score`.

---

#### 3 · Biochemical score

Sums the contribution of each grouping **scheme** in which the position was
detected, weighted by scheme specificity:

| Scheme | Weight |
|--------|-------:|
| US     | 0.50   |
| GS4    | 0.20   |
| GS3    | 0.10   |
| GS2    | 0.10   |
| GS1    | 0.10   |
| GS0    | 0.00   |

Each scheme contributes:

$$\text{contribution}_s = w_s \times \text{variability\_score} \times \text{pattern\_score}$$

The final score is the **cumulative sum** over all schemes where the position
was detected:

$$\text{biochem\_score} = \sum_{s \in \text{detected schemes}} w_s \times \text{variability\_score} \times \text{pattern\_score}$$

Maximum value when detected in all weighted schemes (US + GS4 + GS3 + GS2 + GS1)
with variability = pattern = 1.0 is **1.0** (0.5 + 0.2 + 0.1 + 0.1 + 0.1).

---

#### 4 · ASR score

Measures ancestral-state-reconstruction confidence. Three mutually exclusive cases:

| Condition | Score |
|-----------|------:|
| `is_conserved_meta = FALSE` (novel change) | **1.0** |
| `is_conserved_meta = TRUE` **and** `asr_is_conserved = TRUE` | see formula below |
| `is_conserved_meta = TRUE` **and** `asr_is_conserved = FALSE` | **0.0** |

For ASR-confirmed conserved positions, the score is pair-weighted across the
detected foreground-pair configuration. Let $w = 1 / n_{\text{pairs}}$. Then:

$$
\text{asr\_score} =
\sum_{\text{divergent pairs}} w +
\sum_{\text{conserved pairs}} \min\left(
\frac{w \times P_{\text{MRCA}(i)} + w \times P_{\text{root}} \times I(\texttt{asr\_root\_conserved})}{2},
0.9w
\right)
$$

This keeps supported conserved explanations informative while ensuring they do
not outrank genuinely novel changes, which always score **1.0**.

---

#### 5 · Convergence score

Scores the **type and directionality** of the amino acid change on each
assessed side (top / bottom foreground lineages).

Side values:

| Change type | Value |
|-------------|------:|
| convergent  | 1.0   |
| codivergent | 0.5   |
| divergent   | 0.25  |
| ambiguous / no change | excluded (NA) |

$$\text{mean\_change} = \text{mean of non-NA side values}$$

$$\text{completeness penalty} = \begin{cases} 1.0 & \text{both sides assessed} \\ 0.75 & \text{one side only} \end{cases}$$

$$\text{convergence\_score} = \text{mean\_change} \times \text{completeness penalty}$$

Range: 0.1875 (one side, divergent: 0.25 × 0.75) → 1.0 (both sides convergent).

---

#### 6 · Parallel score

Scores whether the amino acid change is **parallel** (same substitution in
independent lineages) or **non-parallel**, weighted by MRCA posterior
confidence.

Parallelism coefficients (note: non-parallel receives the *highest* weight
because independent parallel changes are less informative as a signal of
convergence):

| Parallel type | Coefficient |
|---------------|------------:|
| non-parallel  | 1.0         |
| mixed         | 0.75        |
| parallel      | 0.5         |

For each assessable side (where `change_top`/`change_bottom` is
convergent / codivergent / divergent):

$$\text{side\_score} = \text{coeff} \times \overline{P_{\text{MRCA}}}$$

where $\overline{P_{\text{MRCA}}}$ is the mean of `mrca_1_posterior`,
`mrca_2_posterior`, `mrca_3_posterior` (NA excluded).

$$\text{mean\_parallel} = \text{mean of non-NA side scores}$$

$$\text{completeness penalty} = \begin{cases} 1.0 & \text{both sides assessed} \\ 0.75 & \text{one side only} \end{cases}$$

$$\text{parallel\_score} = \text{mean\_parallel} \times \text{completeness penalty}$$

---

#### Composite CAAS score

Unweighted mean of the four core components. FADE is **not**
included — it is a gene-level signal (see below).

$$\text{CAAS\_score} = \frac{\text{biochem} + \text{asr} + \text{convergence} + \text{parallel}}{4}$$

---

### Gene-Level Scores

#### gene\_caas\_score

90th percentile of `CAAS_score` across all scored positions in that gene:

$$\text{gene\_caas\_score} = Q_{0.90}(\text{CAAS\_score})$$

#### gene\_rand\_score  *(optional)*

Scheme-weighted sum of decile-ranked accumulation randomization p-values.
The same scheme weights as `biochem_score` are used (GS0 excluded):

$$\text{gene\_rand\_score} = \sum_{s \in \{US, GS4, GS3, GS2, GS1\}} w_s \times \left(1 - \frac{\text{ntile}(P_{\text{empirical},s}, 10) - 1}{9}\right)$$

Maximum = 1.0 (all schemes p-value in best decile).

#### gene\_rer\_score  *(optional)*

Decile rank of the RERConverge adjusted p-value (`p.adj`; or `p.perm` if
available and non-zero), inverted:

$$\text{gene\_rer\_score} = 1 - \frac{\text{ntile}(p_{\text{RER}}, 10) - 1}{9}$$

#### gene\_fade\_score  *(optional)*

Gene-level maximum Bayes Factor from FADE directional selection analysis,
taking the best of the top/bottom foreground directions. Decile rank,
higher BF = higher score:

$$\text{gene\_fade\_score} = \frac{\text{ntile}(\max_{d \in \{\text{top,bottom}\}} \text{BF}_{d}, 10) - 1}{9}$$

#### gene\_composite\_score

Unweighted mean of all available gene-level sub-scores:

$$\text{gene\_composite\_score} = \frac{\text{gene\_caas} + [\text{gene\_rand}] + [\text{gene\_rer}] + [\text{gene\_fade}]}{N_{\text{available}}}$$

---

### ORA Gene List Extraction

Gene lists for Over-Representation Analysis are extracted at three thresholds
(configurable via `--scoring_position_top_pct` / `--scoring_position_top5_pct` /
`--scoring_position_top1_pct` and their gene-level equivalents) and three
directional subsets:

| Subset | Filter |
|--------|--------|
| **top** | Positions / genes where `change_side == "top"` |
| **bottom** | Positions / genes where `change_side == "bottom"` |
| **both** | All top-scoring positions / genes regardless of direction |

For **position-level** lists, the threshold is applied to `CAAS_score` and
the gene name is extracted from the qualifying positions.

For **gene-level** lists, the threshold is applied to each gene-level score
independently (`gene_caas_score`, `gene_rand_score`, `gene_rer_score`,
`gene_composite_score`), and direction is assigned from the gene's
highest-scoring position.

</details>

---

## Input Distributions {.tabset}

The `variability_score` and `pattern_score` components entering `biochem_score` are
derived from the **hypergeometric** and **permulation** signals used by the
scoring model. If the input table still contains raw p-values, those are shown
directly; otherwise the report falls back to the exported decile scores, which
preserve the same ranking but not the original p-value scale.

### `r if (hyper_measure_is_raw) "Hypergeometric P-value" else "Variability Score"`

```{r hyper-pval-dist, fig.height=8}
# Compute decile boundaries from the selected hypergeometric measure.
hyper_values <- pos_df[[hyper_measure_col]]
pv_breaks <- quantile(hyper_values, probs = seq(0, 1, by = 0.1), na.rm = TRUE)
decile_df_h <- pos_df %>%
  filter(!is.na(.data[[hyper_measure_col]])) %>%
  mutate(
    measure_value = .data[[hyper_measure_col]],
    plot_value    = if (hyper_measure_is_raw) -log10(pmax(measure_value, 1e-300)) else measure_value,
    decile_raw    = dplyr::ntile(measure_value, 10),
    # Keep decile direction aligned with score semantics:
    # decile 10 = strongest signal (smallest p-value or highest score).
    decile        = if (hyper_measure_is_raw) 11 - decile_raw else decile_raw,
    score_val     = (decile - 1) / 9
  )

# Panel 1: density on the selected measure scale with decile boundaries
p_hyp_density <- ggplot(decile_df_h, aes(x = plot_value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60,
                 fill = .tableau10[1], alpha = 0.6) +
  geom_density(linewidth = 0.8, colour = .tableau10[2]) +
  geom_vline(
    xintercept = if (hyper_measure_is_raw) -log10(pmax(pv_breaks[2:10], 1e-300)) else pv_breaks[2:10],
    linetype = "dotted", colour = "grey50", linewidth = 0.4
  ) +
  labs(
    title    = if (hyper_measure_is_raw) "Hypergeometric P-value Distribution (-log10 scale)" else "Variability Score Distribution",
    subtitle = if (hyper_measure_is_raw) {
      "Dotted lines: decile boundaries (D10 = smallest p-values, highest variability_score; D1 = largest p-values)"
    } else {
      "Dotted lines: decile boundaries of the exported variability_score values"
    },
    x        = if (hyper_measure_is_raw) "-log10(Pvalue)" else "variability_score",
    y        = "Density"
  ) +
  scale_x_continuous(
    breaks = scales::pretty_breaks(n = 14),
    minor_breaks = scales::pretty_breaks(n = 28)
  ) +
  theme(plot.title = element_text(face = "bold"))

# Panel 2: count per decile, coloured by variability_score
decile_counts_h <- decile_df_h %>%
  group_by(decile, score_val) %>%
  summarise(n = n(), .groups = "drop")

p_hyp_decile <- ggplot(decile_counts_h,
                        aes(x = factor(decile), y = n, fill = score_val)) +
  geom_col(colour = "white", linewidth = 0.3) +
  scale_fill_gradient(low = .tableau10[1], high = .tableau10[3],
                      name = "variability_score") +
  labs(
    title    = "Positions per Hypergeometric Decile",
    subtitle = if (hyper_measure_is_raw) {
      "Decile 10 = smallest p-values (variability_score = 1.0); Decile 1 = largest p-values"
    } else {
      "Decile 1 = lowest variability_score; Decile 10 = highest variability_score"
    },
    x        = "Decile (10 = most significant)",
    y        = "Number of rows"
  ) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "right")

gridExtra::grid.arrange(p_hyp_density, p_hyp_decile, ncol = 1)

# Decile boundary table: what p-value range does each decile cover?
if (hyper_measure_is_raw) {
  .hyp_lower <- pv_breaks[1:10]
  .hyp_upper <- c(pv_breaks[2:10], max(hyper_values, na.rm = TRUE))
  hyp_boundary_tbl <- tibble(
    Decile                = 10:1,
    `P-value lower bound` = formatC(.hyp_lower, format = "e", digits = 2),
    `P-value upper bound` = formatC(.hyp_upper, format = "e", digits = 2),
    variability_score     = round((10:1 - 1) / 9, 3)
  )
  knitr::kable(
    hyp_boundary_tbl,
    caption = "Hypergeometric p-value range per decile (decile 10 = smallest p-values, variability_score = 1.0)"
  )
}
```

### `r if (boot_measure_is_raw) "Permulation P-value" else "Pattern Score"`

```{r perm-pval-dist, fig.height=8}
boot_values <- pos_df[[boot_measure_col]]
pv_breaks_b <- quantile(boot_values, probs = seq(0, 1, by = 0.1), na.rm = TRUE)
decile_df_b <- pos_df %>%
  filter(!is.na(.data[[boot_measure_col]])) %>%
  mutate(
    measure_value = .data[[boot_measure_col]],
    plot_value    = if (boot_measure_is_raw) -log10(pmax(measure_value, 1e-300)) else measure_value,
    decile_raw    = dplyr::ntile(measure_value, 10),
    decile        = if (boot_measure_is_raw) 11 - decile_raw else decile_raw,
    score_val     = (decile - 1) / 9
  )

p_boot_density <- ggplot(decile_df_b, aes(x = plot_value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60,
                 fill = .tableau10[5], alpha = 0.6) +
  geom_density(linewidth = 0.8, colour = .tableau10[6]) +
  geom_vline(
    xintercept = if (boot_measure_is_raw) -log10(pmax(pv_breaks_b[2:10], 1e-300)) else pv_breaks_b[2:10],
    linetype = "dotted", colour = "grey50", linewidth = 0.4
  ) +
  labs(
    title    = if (boot_measure_is_raw) "Permulation Bootstrap P-value Distribution (-log10 scale)" else "Pattern Score Distribution",
    subtitle = if (boot_measure_is_raw) {
      "Dotted lines: decile boundaries (D10 = smallest p-values, highest pattern_score; D1 = largest p-values)"
    } else {
      "Dotted lines: decile boundaries of the exported pattern_score values"
    },
    x        = if (boot_measure_is_raw) "-log10(pvalue_boot)" else "pattern_score",
    y        = "Density"
  ) +
  scale_x_continuous(
    breaks = scales::pretty_breaks(n = 14),
    minor_breaks = scales::pretty_breaks(n = 28)
  ) +
  theme(plot.title = element_text(face = "bold"))

decile_counts_b <- decile_df_b %>%
  group_by(decile, score_val) %>%
  summarise(n = n(), .groups = "drop")

p_boot_decile <- ggplot(decile_counts_b,
                         aes(x = factor(decile), y = n, fill = score_val)) +
  geom_col(colour = "white", linewidth = 0.3) +
  scale_fill_gradient(low = .tableau10[7], high = .tableau10[5],
                      name = "pattern_score") +
  labs(
    title    = "Positions per Permulation Decile",
    subtitle = if (boot_measure_is_raw) {
      "Decile 10 = smallest permulation p-values (pattern_score = 1.0); Decile 1 = largest p-values"
    } else {
      "Decile 1 = lowest pattern_score; Decile 10 = highest pattern_score"
    },
    x        = "Decile (10 = most significant)",
    y        = "Number of rows"
  ) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "right")

gridExtra::grid.arrange(p_boot_density, p_boot_decile, ncol = 1)

# Decile boundary table for permulation p-value
if (boot_measure_is_raw) {
  .boot_lower <- pv_breaks_b[1:10]
  .boot_upper <- c(pv_breaks_b[2:10], max(boot_values, na.rm = TRUE))
  boot_boundary_tbl <- tibble(
    Decile                = 10:1,
    `P-value lower bound` = formatC(.boot_lower, format = "e", digits = 2),
    `P-value upper bound` = formatC(.boot_upper, format = "e", digits = 2),
    pattern_score         = round((10:1 - 1) / 9, 3)
  )
  knitr::kable(
    boot_boundary_tbl,
    caption = "Permulation p-value range per decile (decile 10 = smallest p-values, pattern_score = 1.0)"
  )
}
```

### Joint Decile Summary

```{r joint-decile-summary, fig.height=5}
# Scatter of variability_score vs pattern_score coloured by CAAS_score
joint_df <- pos_df %>%
  filter(!is.na(variability_score), !is.na(pattern_score)) %>%
  mutate(
    hyp_decile_raw  = dplyr::ntile(.data[[hyper_measure_col]], 10),
    boot_decile_raw = dplyr::ntile(.data[[boot_measure_col]], 10),
    hyp_decile      = if (hyper_measure_is_raw) 11 - hyp_decile_raw else hyp_decile_raw,
    boot_decile     = if (boot_measure_is_raw) 11 - boot_decile_raw else boot_decile_raw
  )

.r_hyp_caas  <- round(cor(joint_df$hyp_decile,  joint_df$CAAS_score, use = "complete.obs"), 2)
.r_boot_caas <- round(cor(joint_df$boot_decile, joint_df$CAAS_score, use = "complete.obs"), 2)
.r_hyp_boot  <- round(cor(joint_df$hyp_decile,  joint_df$boot_decile, use = "complete.obs"), 2)

ggplot(joint_df, aes(x = hyp_decile, y = boot_decile, colour = CAAS_score)) +
  geom_jitter(alpha = 0.45, size = 1.8, width = 0.35, height = 0.35) +
  scale_colour_gradientn(
    colours = c("#4E79A7", "#76B7B2", "#EDC948", "#F28E2B", "#E15759"),
    name    = "CAAS\nscore"
  ) +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(breaks = 1:10) +
  annotate("text", x = 1, y = 10,
           label = sprintf("r(hyp, CAAS) = %.2f\nr(perm, CAAS) = %.2f\nr(hyp, perm) = %.2f",
                           .r_hyp_caas, .r_boot_caas, .r_hyp_boot),
           hjust = 0, vjust = 1, size = 3.2, colour = "grey30") +
  labs(
    title    = "Hypergeometric vs Permulation Decile Assignment per Position",
    subtitle = "High CAAS scores (warm colours) cluster toward decile 10 × 10. Pearson r computed on decile ranks.",
    x        = "Hypergeometric decile (variability; 10 = most significant)",
    y        = "Permulation decile (pattern; 10 = most significant)"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

---

## Position-Level Scores {.tabset}

### CAAS Score Distribution

```{r caas-score-dist, fig.height=5}
thr_10 <- quantile(pos_df$CAAS_score, 1 - top_pct,  na.rm = TRUE)
thr_5  <- quantile(pos_df$CAAS_score, 1 - top5_pct, na.rm = TRUE)
thr_1  <- quantile(pos_df$CAAS_score, 1 - top1_pct, na.rm = TRUE)

ggplot(pos_df, aes(x = CAAS_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50,
                 fill = .tableau10[1], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[2]) +
  geom_vline(xintercept = thr_10, linetype = "dashed",
             colour = .tableau10[3], linewidth = 0.7) +
  geom_vline(xintercept = thr_5,  linetype = "longdash",
             colour = .tableau10[5], linewidth = 0.7) +
  geom_vline(xintercept = thr_1,  linetype = "dotted",
             colour = .tableau10[4], linewidth = 0.7) +
  annotate("text", x = thr_10, y = Inf, label = sprintf("Top %g%%", top_pct * 100),
           vjust = 2,    hjust = -0.1, colour = .tableau10[3], size = 3.5) +
  annotate("text", x = thr_5,  y = Inf, label = sprintf("Top %g%%", top5_pct * 100),
           vjust = 2.75, hjust = -0.1, colour = .tableau10[5], size = 3.5) +
  annotate("text", x = thr_1,  y = Inf, label = sprintf("Top %g%%", top1_pct * 100),
           vjust = 3.5,  hjust = -0.1, colour = .tableau10[4], size = 3.5) +
  labs(title = "CAAS Score Distribution",
       subtitle = sprintf("n = %s positions | Trait: %s",
                          format(nrow(pos_df), big.mark = ","), params$traitname),
       x = "CAAS Score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### Biochemical Score

```{r biochem-dist, fig.height=4.5}
ggplot(pos_df, aes(x = biochem_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = .tableau10[1], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[2]) +
  labs(title = "Biochemical Score Distribution",
       subtitle = "Weighted sum of scheme-specific variability x pattern contributions",
       x = "biochem_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### ASR Score

```{r asr-dist, fig.height=4.5}
ggplot(pos_df, aes(x = asr_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = .tableau10[3], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[4]) +
  labs(title = "ASR Score Distribution",
       subtitle = "Novel = 1.0 | contradicted conserved = 0 | supported conserved = pair-weighted posterior support below novelty",
       x = "asr_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### Convergence Score

```{r convergence-dist, fig.height=4.5}
ggplot(pos_df, aes(x = convergence_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = .tableau10[5], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[6]) +
  labs(title = "Convergence Score Distribution",
       subtitle = "Ordinal change type (1 / 0.5 / 0.25) with a 1.0 / 0.75 completeness penalty",
       x = "convergence_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### Parallel Score

```{r parallel-dist, fig.height=4.5}
ggplot(pos_df, aes(x = parallel_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = .tableau10[7], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[8]) +
  labs(title = "Parallel Score Distribution",
       subtitle = "Non-parallel / mixed / parallel = 1 / 0.75 / 0.5, weighted by MRCA confidence and completeness",
       x = "parallel_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### Component Overview

```{r component-overview, fig.height=6}
# Faceted density comparison of all position-level components
component_cols <- c("biochem_score", "asr_score", "convergence_score", "parallel_score")

long_df <- pos_df %>%
  select(all_of(component_cols)) %>%
  pivot_longer(everything(), names_to = "Component", values_to = "Score") %>%
  filter(!is.na(Score)) %>%
  mutate(Component = factor(Component, levels = component_cols))

ggplot(long_df, aes(x = Score, fill = Component)) +
  geom_density(alpha = 0.6, linewidth = 0.3) +
  scale_fill_manual(values = .tableau10[seq_along(component_cols)]) +
  facet_wrap(~Component, scales = "free_y", ncol = 2) +
  labs(title = "Position-Level Component Score Distributions",
       x = "Score (0-1)", y = "Density") +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "none")
```

### Threshold Profiles

```{r threshold-profiles, fig.height=14, fig.width=12}
# Component score distributions for positions inside each ORA gene-list cutoff.
# Each panel shows overlaid densities of all position-level components so you
# can read off which signal drives positions into the top-N% × direction bucket.

component_cols_tp <- c("biochem_score", "asr_score", "convergence_score", "parallel_score")

thr_10_tp <- quantile(pos_df$CAAS_score, 1 - top_pct,  na.rm = TRUE)
thr_5_tp  <- quantile(pos_df$CAAS_score, 1 - top5_pct, na.rm = TRUE)
thr_1_tp  <- quantile(pos_df$CAAS_score, 1 - top1_pct, na.rm = TRUE)

make_threshold_panel <- function(df, threshold, pct_label, direction) {
  subset_df <- if (direction == "both") {
    df %>% filter(CAAS_score >= threshold)
  } else {
    df %>% filter(CAAS_score >= threshold, change_side == direction)
  }

  if (nrow(subset_df) == 0) {
    return(
      ggplot() +
        annotate("text", x = 0.5, y = 0.5, label = "No positions",
                 size = 5, colour = "grey50") +
        labs(title = sprintf("Top %s — %s", pct_label, direction),
             x = "Score (0–1)", y = "Density") +
        xlim(0, 1) + theme_minimal()
    )
  }

  long_df <- subset_df %>%
    select(all_of(component_cols_tp)) %>%
    pivot_longer(everything(), names_to = "Component", values_to = "Score") %>%
    filter(!is.na(Score)) %>%
    mutate(Component = factor(Component, levels = component_cols_tp))

  ggplot(long_df, aes(x = Score, fill = Component, colour = Component)) +
    geom_density(alpha = 0.35, linewidth = 0.6) +
    scale_fill_manual(values   = .tableau10[seq_along(component_cols_tp)]) +
    scale_colour_manual(values = .tableau10[seq_along(component_cols_tp)]) +
    scale_x_continuous(limits = c(0, 1), expand = c(0.01, 0)) +
    labs(
      title    = sprintf("Top %s — %s", pct_label, direction),
      subtitle = sprintf("n = %d positions", nrow(subset_df)),
      x = "Score (0–1)", y = "Density",
      fill = NULL, colour = NULL
    ) +
    theme(
      plot.title    = element_text(face = "bold", size = 11),
      plot.subtitle = element_text(size = 9, colour = "grey40"),
      legend.position = "bottom",
      legend.text     = element_text(size = 8)
    )
}

panels <- list()
for (direction in c("top", "bottom", "both")) {
  panels[[paste0("10pct_", direction)]] <-
    make_threshold_panel(pos_df, thr_10_tp, "10%", direction)
  panels[[paste0("5pct_",  direction)]] <-
    make_threshold_panel(pos_df, thr_5_tp,  "5%",  direction)
  panels[[paste0("1pct_",  direction)]] <-
    make_threshold_panel(pos_df, thr_1_tp,  "1%",  direction)
}

# Layout: rows = direction (top / bottom / both), cols = threshold (10% | 5% | 1%)
gridExtra::grid.arrange(
  panels[["10pct_top"]],    panels[["5pct_top"]],    panels[["1pct_top"]],
  panels[["10pct_bottom"]], panels[["5pct_bottom"]], panels[["1pct_bottom"]],
  panels[["10pct_both"]],   panels[["5pct_both"]],   panels[["1pct_both"]],
  ncol = 3,
  top  = grid::textGrob(
    "Component score profiles for top-scoring position subsets",
    gp = grid::gpar(fontface = "bold", fontsize = 13)
  ),
  left = grid::textGrob(
    "Direction:  top                          bottom                          both",
    rot = 90, gp = grid::gpar(fontsize = 9, col = "grey40")
  )
)
```

### Correlation Analysis

```{r position-correlation-setup}
pos_score_cols <- c(
  "variability_score",
  "pattern_score",
  "biochem_score",
  "asr_score",
  "convergence_score",
  "parallel_score"
)
pos_corr_df <- if (length(pos_score_cols) >= 2) {
  expand.grid(
    score_a = pos_score_cols, score_b = pos_score_cols,
    stringsAsFactors = FALSE
  ) %>%
    filter(score_a < score_b) %>%
    rowwise() %>%
    mutate(
      pearson_r  = cor(pos_df[[score_a]], pos_df[[score_b]],
                       method = "pearson", use = "complete.obs"),
      spearman_r = cor(pos_df[[score_a]], pos_df[[score_b]],
                       method = "spearman", use = "complete.obs"),
      n_positions = sum(complete.cases(pos_df[[score_a]], pos_df[[score_b]]))
    ) %>%
    ungroup()
} else {
  tibble(
    score_a = character(), score_b = character(),
    pearson_r = numeric(), spearman_r = numeric(), n_positions = integer()
  )
}
```

```{r position-correlation-scatter, eval=(nrow(pos_corr_df) > 0), fig.height=20, fig.width=10}
pos_pairs_df <- pos_df %>%
  select(all_of(pos_score_cols)) %>%
  drop_na()

pos_pairs_long <- expand.grid(
  x_var = pos_score_cols, y_var = pos_score_cols,
  stringsAsFactors = FALSE
) %>%
  filter(x_var != y_var)

pos_plots <- list()
for (i in seq_len(nrow(pos_pairs_long))) {
  xv <- pos_pairs_long$x_var[i]
  yv <- pos_pairs_long$y_var[i]
  r_info <- pos_corr_df %>%
    filter((score_a == xv & score_b == yv) | (score_a == yv & score_b == xv))
  label <- if (nrow(r_info) > 0) {
    sprintf("r=%.2f / rho=%.2f", r_info$pearson_r[1], r_info$spearman_r[1])
  } else ""

  pos_plots[[length(pos_plots) + 1]] <- ggplot(pos_pairs_df, aes(x = .data[[xv]], y = .data[[yv]])) +
    geom_point(alpha = 0.18, size = 0.5, colour = .tableau10[1]) +
    geom_smooth(method = "lm", se = FALSE, colour = .tableau10[4], linewidth = 0.6) +
    annotate("text", x = Inf, y = -Inf, label = label,
             hjust = 1.1, vjust = -0.5, size = 2.8, colour = .tableau10[4]) +
    labs(
      x = sub("_score$", "", xv),
      y = sub("_score$", "", yv)
    ) +
    theme(axis.title = element_text(size = 8))
}

if (length(pos_plots) > 0) {
  gridExtra::grid.arrange(
    grobs = pos_plots,
    ncol = min(length(pos_score_cols) - 1, 3),
    top = grid::textGrob(
      "Position-Level Score Correlations (r = Pearson; rho = Spearman rank correlation)",
      gp = grid::gpar(fontface = "bold", fontsize = 13)
    )
  )
}
```

```{r position-correlation-table, eval=(nrow(pos_corr_df) > 0)}
DT::datatable(
  pos_corr_df %>%
    mutate(across(where(is.numeric), ~round(., 4))),
  rownames = FALSE,
  options  = list(pageLength = 15, dom = "t"),
  caption  = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "Pairwise correlations between position-level scores"
  )
)
```

### Stress Test

`r if (!has_scoring_stress) "> **Note:** scoring stress-test outputs are not available for this run."`

```{r stress-summary-table, eval=has_scoring_stress}
DT::datatable(
  stress_summary_df %>%
    arrange(desc(spearman_to_current), mean_abs_rank_shift) %>%
    mutate(across(where(is.numeric), ~round(., 4))),
  rownames = FALSE,
  options = list(pageLength = 20, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "Assumption sensitivity summary for alternative position-level score variants"
  )
)
```

```{r stress-component-overview, eval=has_scoring_stress, fig.height=6}
stress_component_cols <- c("variability_score", "pattern_score", "scheme_coverage", "biochem_score")
stress_components_long <- stress_variants_df %>%
  select(any_of(stress_component_cols)) %>%
  pivot_longer(everything(), names_to = "Component", values_to = "Score") %>%
  filter(!is.na(Score))

ggplot(stress_components_long, aes(x = Score, fill = Component)) +
  geom_density(alpha = 0.55, linewidth = 0.4) +
  scale_fill_manual(values = .tableau10[seq_along(unique(stress_components_long$Component))]) +
  facet_wrap(~Component, scales = "free_y", ncol = 2) +
  labs(
    title = "Biochemical Decomposition Overview",
    subtitle = paste0("variability_score and pattern_score: per-row decile inputs\n",
                      "scheme_coverage: sum of active scheme weights (how many GS schemes detected a CAAS); ",
                      "biochem_score: weighted product of the above"),
    x = "Score", y = "Density"
  ) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "none")
```

```{r stress-correlation-heatmap, eval=has_scoring_stress, fig.height=12, fig.width=10}
stress_heatmap_df <- stress_corr_df %>%
  pivot_longer(c(pearson_r, spearman_r), names_to = "metric", values_to = "value") %>%
  mutate(metric = recode(metric, pearson_r = "Pearson", spearman_r = "Spearman"))

ggplot(stress_heatmap_df, aes(x = score_a, y = score_b, fill = value)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.2f", value)), size = 2.8) +
  scale_fill_gradient2(low = .tableau10[3], mid = "white", high = .tableau10[1], midpoint = 0, limits = c(-1, 1)) +
  facet_wrap(~metric, ncol = 1) +
  labs(
    title = "Position-Level Correlation Heatmap",
    subtitle = paste0("r = Pearson correlation; rho = Spearman rank correlation\n",
                      "Values near \u00b11 indicate strong linear/monotone association between scores"),
    x = NULL, y = NULL, fill = "r"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold")
  )
```

```{r stress-rank-heatmap, eval=has_scoring_stress, fig.height=10, fig.width=9}
ggplot(stress_rank_df, aes(x = score_a, y = score_b, fill = spearman_rank)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.2f", spearman_rank)), size = 2.8) +
  scale_fill_gradient2(low = .tableau10[3], mid = "white", high = .tableau10[1], midpoint = 0, limits = c(-1, 1)) +
  labs(
    title = "Composite Rank-Agreement Heatmap",
    subtitle = paste0("rho = Spearman rank correlation between composite score rankings\n",
                      "Values near 1 mean both scoring variants rank positions nearly identically"),
    x = NULL, y = NULL, fill = "rho"
  ) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r stress-top-overlap, eval=has_scoring_stress}
DT::datatable(
  stress_overlap_df %>%
    arrange(threshold, desc(jaccard)) %>%
    mutate(across(where(is.numeric), ~round(., 4))),
  rownames = FALSE,
  options = list(pageLength = 20, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "Top-hit overlap across alternative position-level composite variants"
  )
)
```

```{r stress-instability, eval=has_scoring_stress}
rank_cols <- grep("^rank_", names(stress_variants_df), value = TRUE)
instability_df <- if (length(rank_cols) > 0) {
  stress_variants_df %>%
    select(Gene, Position, CAAS_current, any_of(rank_cols), rank_spread) %>%
    arrange(desc(rank_spread), desc(CAAS_current)) %>%
    slice_head(n = 25)
} else {
  tibble(Gene = character(), Position = integer(), rank_spread = numeric())
}

DT::datatable(
  instability_df %>% mutate(across(where(is.numeric), ~round(., 4))),
  rownames = FALSE,
  options = list(pageLength = 25, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "Top positions by rank instability across stress-test variants"
  )
)
```

Latent models ask: *if we let the data decide the weights, which component scores drive
the most variance?*  Two approaches are computed:

- **PCA_all_components** / **PCA_no_biochem** — First principal component of all
  position-level scores (or their decomposed biochemical sub-scores). The loadings
  show how much each input score contributes to PC1. A positive loading means the
  component co-varies with the main source of variation across positions.
- **Factor1_all_components** — Exploratory factor analysis with a single latent
  factor. Unlike PCA, factor analysis separates shared variance (the factor) from
  unique variance per variable; loadings quantify each component's contribution to
  the shared signal.
- **CAAS_residualized_biochem** — Biochemical score after regressing out ASR,
  convergence, and parallel scores, isolating the fraction of biochemical signal
  not explained by the structural components.

All latent scores are converted to 0–1 decile scale for comparison. A high
correlation with `CAAS_current` (see summary table) means the latent model and
the hand-engineered composite agree on which positions are most interesting.

```{r stress-latent, eval=has_scoring_stress, fig.height=5}
if (!is.null(stress_loadings_df) && nrow(stress_loadings_df) > 0) {
  ggplot(stress_loadings_df, aes(x = reorder(variable, loading), y = loading, fill = model)) +
    geom_col(position = "dodge") +
    coord_flip() +
    labs(
      title    = "Latent-Model Component Loadings",
      subtitle = paste0("How much each score component drives each latent model's first dimension\n",
                        "Positive = co-varies with PC1/Factor1; compare models to spot dominant components"),
      x = NULL, y = "Loading"
    ) +
    theme(plot.title    = element_text(face = "bold"),
          plot.subtitle = element_text(size = 9, colour = "grey40"))
} else {
  plot.new()
  text(0.5, 0.5, "No latent-model loadings available for this run")
}
```

`r if (has_scoring_stress) '> **How to read the stress test:**\n> \n> | Output | What to look for | What it means |\n> |--------|-----------------|---------------|\n> | **Biochemical decomposition** | Distributions of variability/pattern/caap/biochem | Checks that per-row inputs span the full 0–1 range; heavy skew may indicate low diversity |\n> | **Correlation heatmap** | r / rho close to 1 between a variant and CAAS_current | Scoring assumptions barely change position rankings |\n> | **Rank-agreement heatmap** | rho close to 1 | The two variants order positions the same way |\n> | **Top-overlap table** | High Jaccard at top-10% / top-5% thresholds | The same genes appear in gene lists regardless of assumptions |\n> | **Instability table** | High rank_spread | These positions are assumption-sensitive; examine their component scores manually |\n> | **Latent models** | PC1 / Factor1 loadings; correlation with CAAS_current | Which components drive shared variance; if latent models agree with CAAS_current the hand-engineered weights are data-consistent |\n> \n> **Score variant naming:** <code>CAAS_no_*</code> — drops one component; <code>CAAS_biochem_alt_*</code> — swaps the biochemical aggregation function; <code>PCA_*</code> / <code>Factor1_*</code> — data-driven weighting via PCA or factor analysis; <code>CAAS_residualized_biochem</code> — biochemical signal after removing correlation with structural components.\n> A robust run shows consistent rankings across all variants.'`

### Top Positions

```{r top-positions-table}
display_cols <- c("Gene", "Position", "CAAS_score", "biochem_score",
                  "asr_score", "convergence_score", "parallel_score", "change_side")

DT::datatable(
  pos_df %>%
    arrange(desc(CAAS_score)) %>%
    select(all_of(display_cols)) %>%
    mutate(across(where(is.numeric), ~round(., 4))),
  filter    = "top",
  rownames  = FALSE,
  options   = list(pageLength = 25, scrollX = TRUE),
  caption   = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Position-level CAAS scores | Trait: %s", params$traitname)
  )
)
```

### CAAS Score Manhattan (Position Level)

`r if (!has_genomic) "> **Note:** Genomic coordinates not available — position-level Manhattan skipped."`

```{r pos-caas-manhattan-setup, eval=has_genomic}
# Join all scored positions with genomic coordinates via their gene.
# Each position inherits its gene's genomic midpoint as the x-coordinate.
# We build the chr offsets independently here (rather than relying on the
# genomic-setup chunk, which runs later in the document).
pos_genomic <- pos_df %>%
  inner_join(genomic_raw, by = "Gene") %>%
  mutate(
    chr_label = sub("^chr", "", chr),
    chr_num   = suppressWarnings(as.integer(chr_label)),
    chr_order = dplyr::case_when(
      !is.na(chr_num)              ~ chr_num,
      chr_label == "X"             ~ 23L,
      chr_label == "Y"             ~ 24L,
      chr_label %in% c("MT","M")   ~ 25L,
      TRUE                         ~ 26L
    )
  ) %>%
  arrange(chr_order, gene_mid)

pos_chr_offsets <- pos_genomic %>%
  group_by(chr_order, chr_label) %>%
  summarise(chr_max = max(gene_end, na.rm = TRUE), .groups = "drop") %>%
  arrange(chr_order) %>%
  mutate(offset = cumsum(lag(chr_max, default = 0L)) + (chr_order - 1L) * 5e6L)

pos_genomic <- pos_genomic %>%
  left_join(pos_chr_offsets %>% select(chr_order, offset), by = "chr_order") %>%
  mutate(
    manhattan_x = gene_mid + offset,
    chr_even    = chr_order %% 2 == 0
  )

pos_chr_label_positions <- pos_chr_offsets %>%
  left_join(
    pos_genomic %>% group_by(chr_order) %>%
      summarise(mid_x = mean(manhattan_x, na.rm = TRUE), .groups = "drop"),
    by = "chr_order"
  )
```

```{r pos-caas-manhattan-plot, eval=has_genomic, fig.height=5, fig.width=14}
ggplot(pos_genomic, aes(x = manhattan_x, y = CAAS_score, colour = chr_even)) +
  geom_point(alpha = 0.45, size = 0.6, show.legend = FALSE) +
  geom_hline(yintercept = thr_10, linetype = "dashed",
             colour = "grey40", linewidth = 0.6) +
  scale_colour_manual(
    values = c("FALSE" = .tableau10[1], "TRUE" = .tableau10[2]),
    guide = "none"
  ) +
  scale_x_continuous(
    breaks = pos_chr_label_positions$mid_x,
    labels = pos_chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  labs(
    title    = "Position-Level CAAS Score — Manhattan View",
    subtitle = sprintf("%d positions with genomic coordinates | dashed = top %g%% threshold",
                       nrow(pos_genomic), params$top_pct * 100),
    x = "Chromosome", y = "CAAS Score"
  ) +
  theme(
    plot.title         = element_text(face = "bold"),
    axis.text.x        = element_text(size = 7),
    panel.grid.major.x = element_blank()
  )
```

---

## Position Characterisation {.tabset #position-characterisation}

This section examines how ranked positions behave in relation to external
signals that do **not** contribute to the `CAAS_score`. PrimateAI-3D is treated
as an independent characterisation layer: its signal is scheme-weighted exactly
like the biochemical score (US = 0.5 → GS4 = 0.2 → GS3/2/1 = 0.1 each → GS0 = 0),
producing a continuous 0–1 characterisation score.

- **`pai_char_score`** = Σ scheme\_weight × max(PAI3D\_score per CAAP group)
  per Gene × Position — captures the scheme-weighted pathogenicity of the
  CAAS-specific substitution.

This score ranges from 0 (no support in any scheme) to ≤ 1.0 (strong support
across all schemes). It is **not** included in `CAAS_score`.

> **Note:** PrimateAI-3D scores require an access request to Illumina —
> see illumina.com/science/genomics-research/articles/primateai-3d.html

### VEP: PrimateAI by CAAS Decile

`r if (!has_vep_primateai) "> **Note:** PrimateAI-3D scores are not available for this run."`

```{r vep-pai-setup, eval=has_vep_primateai}
# Detect PrimateAI score column (varies by database version)
.pai_score_col <- intersect(
  c("score_PAI3D", "score", "primateai_score", "pathogenicity_score", "PrimateAI_score"),
  names(vep_pai_df)
)[1]
if (is.na(.pai_score_col)) {
  warning("PrimateAI: no recognised score column found — skipping.")
  has_vep_primateai <- FALSE
}
```

```{r vep-pai-join, eval=has_vep_primateai}
pos_score_meta <- pos_df %>%
  select(Gene, Position, CAAS_score, biochem_score) %>%
  mutate(
    caas_decile  = dplyr::ntile(CAAS_score, 10),
    in_top_10pct = CAAS_score >= thr_10,
    in_top_5pct  = CAAS_score >= thr_5,
    in_top_1pct  = CAAS_score >= thr_1
  )

# pai_char_score = Σ scheme_weight × max(PAI3D per CAAP group) per protein position.
# This mirrors the biochem_score formula: cumulative scheme-weighted pathogenicity.
# map_to_primateai.py outputs caap_group and scheme_weight columns when available.
has_pai_schemes <- all(c("caap_group", "scheme_weight") %in% names(vep_pai_df))

if (has_pai_schemes) {
  # Two-step: max per (query × CAAP group) to collapse multiple alt AAs per scheme,
  # then weighted sum across CAAP groups (scheme_weight > 0 only).
  pai_by_prot <- vep_pai_df %>%
    mutate(
      pai_score     = as.numeric(.data[[.pai_score_col]]),
      scheme_weight = as.numeric(scheme_weight)
    ) %>%
    filter(is.finite(pai_score), is.finite(scheme_weight), scheme_weight > 0) %>%
    group_by(transvar_query, caap_group, scheme_weight) %>%
    summarise(max_pai = max(pai_score, na.rm = TRUE), .groups = "drop") %>%
    group_by(transvar_query) %>%
    summarise(
      pai_char_score = sum(max_pai * scheme_weight, na.rm = TRUE),
      n_pai_schemes  = n(),
      .groups = "drop"
    )
} else {
  # Legacy fallback: no scheme information — treat as US-only (weight = 0.5)
  pai_by_prot <- vep_pai_df %>%
    mutate(pai_score = as.numeric(.data[[.pai_score_col]])) %>%
    filter(is.finite(pai_score)) %>%
    group_by(transvar_query) %>%
    summarise(
      pai_char_score = 0.5 * max(pai_score, na.rm = TRUE),  # assume US weight
      n_pai_schemes  = 1L,
      .groups = "drop"
    )
}

if (!is.null(pos_prot_map)) {
  pos_vep_pai <- pos_prot_map %>%
    filter(!is.na(prot_query)) %>%
    inner_join(pai_by_prot, by = c("prot_query" = "transvar_query")) %>%
    inner_join(pos_score_meta, by = c("Gene", "Position")) %>%
    distinct(Gene, Position, .keep_all = TRUE)
} else {
  # fallback: parse GENE:p.N from transvar_query — will mostly miss without aa2prot
  pos_vep_pai <- pai_by_prot %>%
    mutate(
      Gene     = sub(":p\\..*", "", transvar_query),
      Position = as.integer(sub(".*:p\\.", "", transvar_query))
    ) %>%
    select(Gene, Position, pai_char_score, n_pai_schemes) %>%
    group_by(Gene, Position) %>%
    summarise(
      pai_char_score = max(pai_char_score, na.rm = TRUE),
      n_pai_schemes  = max(n_pai_schemes),
      .groups = "drop"
    ) %>%
    inner_join(pos_score_meta, by = c("Gene", "Position"))
}

has_vep_primateai <- nrow(pos_vep_pai) > 0
```

`r if (exists("pos_vep_pai") && nrow(pos_vep_pai) == 0) "> **Note:** PrimateAI-3D inputs were provided, but no scored positions could be matched to the VEP/PAI3D annotations for this direction."`

```{r vep-pai-decile-table, eval=has_vep_primateai}
pai_boundary_tbl <- decile_boundary_tbl(
  pos_vep_pai$pai_char_score,
  score_name = "PAI3D characterisation decile score",
  lower_better = FALSE,
  value_label = "PAI3D characterisation",
  digits = 3
)

if (!is.null(pai_boundary_tbl)) {
  knitr::kable(
    pai_boundary_tbl,
    caption = "PrimateAI-3D scheme-weighted characterisation score range per decile"
  )
}
```

```{r vep-pai-violin, eval=has_vep_primateai, fig.height=5}
# pai_char_score violin + boxplot per CAAS decile.
# Higher deciles should trend toward higher pai_char_score if CAAS and pathogenicity agree.
ggplot(pos_vep_pai, aes(x = factor(caas_decile), y = pai_char_score)) +
  geom_violin(fill = .tableau10[1], alpha = 0.65, trim = TRUE, scale = "width") +
  geom_boxplot(width = 0.14, outlier.size = 0.4, alpha = 0.9,
               fill = "white", colour = "grey30") +
  geom_hline(yintercept = median(pos_vep_pai$pai_char_score, na.rm = TRUE),
             linetype = "dashed", colour = .tableau10[3], linewidth = 0.7) +
  annotate("text", x = 0.4,
           y = median(pos_vep_pai$pai_char_score, na.rm = TRUE) + 0.005,
           hjust = 0, label = "Overall median",
           colour = .tableau10[3], size = 3.2) +
  scale_x_discrete(labels = c("1\n(low)", as.character(2:9), "10\n(high)")) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  labs(
    title    = "PrimateAI-3D Scheme-Weighted Score (pai_char_score) by CAAS Decile",
    subtitle = sprintf(
      "%d positions with PAI3D coverage (of %d total) | score = \u03a3 scheme_weight \u00d7 max(PAI3D per CAAP group)",
      nrow(pos_vep_pai), nrow(pos_df)
    ),
    x = "CAAS Score Decile", y = "pai_char_score"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r vep-pai-scatter, eval=has_vep_primateai, fig.height=5}
# Scatter: CAAS_score vs pai_char_score, coloured by CAAS decile.
ggplot(pos_vep_pai, aes(x = CAAS_score, y = pai_char_score, colour = caas_decile)) +
  geom_point(alpha = 0.45, size = 0.9) +
  geom_smooth(method = "lm", se = TRUE, colour = .tableau10[1],
              linewidth = 0.8, fill = .tableau10[1], alpha = 0.15) +
  geom_vline(xintercept = thr_10, linetype = "dashed",
             colour = .tableau10[3], linewidth = 0.6) +
  geom_vline(xintercept = thr_5,  linetype = "longdash",
             colour = .tableau10[5], linewidth = 0.6) +
  geom_vline(xintercept = thr_1,  linetype = "dotted",
             colour = .tableau10[4], linewidth = 0.6) +
  scale_colour_viridis_c(name = "CAAS\nDecile", option = "plasma") +
  labs(
    title    = "CAAS Score vs PrimateAI-3D Scheme-Weighted Score",
    subtitle = "Dashed verticals: CAAS top-10%, top-5%, top-1% thresholds",
    x = "CAAS Score", y = "pai_char_score"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r vep-pai-biochem-scatter, eval=has_vep_primateai, fig.height=5}
# Biochem score vs PrimateAI-3D: both are scheme-weighted sums — correlation
# shows whether positions with high biochemical conservation signal also
# concentrate pathogenic PrimateAI-3D annotations.
.pai_biochem_r <- safe_cor(pos_vep_pai$biochem_score, pos_vep_pai$pai_char_score)
ggplot(pos_vep_pai, aes(x = biochem_score, y = pai_char_score,
                         colour = caas_decile)) +
  geom_point(alpha = 0.45, size = 0.9) +
  geom_smooth(method = "lm", se = TRUE, colour = .tableau10[1],
              linewidth = 0.8, fill = .tableau10[1], alpha = 0.15) +
  scale_colour_viridis_c(name = "CAAS\nDecile", option = "plasma") +
  labs(
    title    = "Biochem Score vs PrimateAI-3D Scheme-Weighted Score",
    subtitle = sprintf("r = %.3f (Pearson) | Both scores use identical scheme-weight structure",
                       .pai_biochem_r),
    x = "biochem_score", y = "pai_char_score"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r vep-pai-fisher, eval=has_vep_primateai}
# Fisher's exact: are positions with pai_char_score above the median enriched
# among the top CAAS positions?
# Using the median as a data-driven split rather than a fixed raw-score threshold.
.pai_med <- median(pos_vep_pai$pai_char_score, na.rm = TRUE)

run_pai_fisher <- function(df, flag_col, label) {
  df <- df %>% mutate(high_pai = !is.na(pai_char_score) & pai_char_score > .pai_med)
  n_top      <- sum( df[[flag_col]], na.rm = TRUE)
  n_top_h    <- sum( df[[flag_col]] & df$high_pai, na.rm = TRUE)
  n_rest     <- sum(!df[[flag_col]], na.rm = TRUE)
  n_rest_h   <- sum(!df[[flag_col]] & df$high_pai, na.rm = TRUE)
  mat <- matrix(
    c(n_top - n_top_h, n_rest - n_rest_h, n_top_h, n_rest_h),
    nrow = 2,
    dimnames = list(In_top = c("FALSE","TRUE"), High_PAI = c("FALSE","TRUE"))
  )
  ft <- fisher.test(mat)
  tibble(
    Threshold        = label,
    N_top            = n_top,
    N_high_pai_top   = n_top_h,
    Pct_top          = sprintf("%.1f%%", 100 * n_top_h  / max(n_top,  1)),
    N_rest           = n_rest,
    N_high_pai_rest  = n_rest_h,
    Pct_rest         = sprintf("%.1f%%", 100 * n_rest_h / max(n_rest, 1)),
    Odds_ratio       = sprintf("%.2f", ft$estimate),
    CI_95            = sprintf("[%.2f, %.2f]", ft$conf.int[1], ft$conf.int[2]),
    P_value          = formatC(ft$p.value, format = "e", digits = 2)
  )
}

pai_fisher <- bind_rows(
  run_pai_fisher(pos_vep_pai, "in_top_10pct",
                 sprintf("Top %g%% CAAS (decile)",    top_pct  * 100)),
  run_pai_fisher(pos_vep_pai, "in_top_5pct",
                 sprintf("Top %g%% CAAS",             top5_pct * 100)),
  run_pai_fisher(pos_vep_pai, "in_top_1pct",
                 sprintf("Top %g%% CAAS (percentile)", top1_pct * 100))
)

knitr::kable(
  pai_fisher,
  caption = sprintf(
    "Fisher's exact test — enrichment of positions with pai_char_score > median (%.4f) in top-ranking CAAS positions.",
    .pai_med
  ),
  align = c("l", rep("r", ncol(pai_fisher) - 1))
)
```

### Regional Accumulation

`r if (!has_genomic) "> **Note:** Genomic coordinates not available — regional position accumulation skipped."`

```{r position-region-accum, eval=has_genomic, fig.height=5, fig.width=14}
pos_region_wins <- pos_genomic %>%
  mutate(
    win_id = floor(gene_mid / params$window_size_bp),
    win_start = win_id * params$window_size_bp,
    win_end = win_start + params$window_size_bp - 1L,
    win_mid = win_start + params$window_size_bp / 2L,
    in_top_caas = CAAS_score >= thr_10
  ) %>%
  group_by(chr_order, chr_label, win_id, win_start, win_end, win_mid) %>%
  summarise(
    n_positions = n(),
    n_top_positions = sum(in_top_caas, na.rm = TRUE),
    frac_top_positions = n_top_positions / n_positions,
    mean_caas = mean(CAAS_score, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(pos_chr_offsets %>% select(chr_order, offset), by = "chr_order") %>%
  mutate(
    win_x = win_mid + offset,
    region_label = sprintf("chr%s:%0.1f-%0.1f Mb", chr_label, win_start / 1e6, win_end / 1e6)
  )

top_region_wins <- pos_region_wins %>%
  filter(n_top_positions > 0) %>%
  slice_max(frac_top_positions, n = 8, with_ties = FALSE)

ggplot(pos_region_wins, aes(x = win_x, y = frac_top_positions)) +
  geom_col(aes(fill = mean_caas), width = params$window_size_bp * 0.9, alpha = 0.9) +
  geom_point(aes(size = n_top_positions), colour = "black", alpha = 0.8) +
  ggrepel::geom_text_repel(
    data = top_region_wins,
    aes(label = region_label),
    size = 2.3, max.overlaps = 20, colour = "black",
    show.legend = FALSE
  ) +
  scale_fill_viridis_c(name = "Mean\nCAAS", option = "magma") +
  scale_size_continuous(name = "Top-10%\npositions", range = c(1, 6)) +
  scale_x_continuous(
    breaks = pos_chr_label_positions$mid_x,
    labels = pos_chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Regional Accumulation of High-CAAS Positions",
    subtitle = "Bars show the fraction of top-10% CAAS positions per genomic window; labels mark the strongest regions",
    x = "Chromosome", y = "Fraction of positions in top 10% CAAS"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(size = 7),
    panel.grid.major.x = element_blank()
  )
```


---

## FADE Position Characterisation {.tabset}

> **FADE** (FUBAR Approach to Directional Evolution) tests whether foreground branches show
> elevated substitution bias toward a particular amino acid at each protein site.
> Here FADE characterises positions independently of their CAAS status — overlap between
> FADE-biased positions and top-CAAS positions indicates that convergent positions are also
> under directional selection in the foreground lineages.

`r if (!has_fade_pos) "> **Note:** FADE per-site data are not available for this run (fade_site_top_file and fade_site_bot_file not provided)."`

```{r fade-pos-setup, eval=has_fade_pos}
# Use top direction if available, otherwise bottom
fade_site_file <- if (.file_ok(params$fade_site_top_file)) {
  params$fade_site_top_file
} else if (.file_ok(params$fade_site_bot_file)) {
  params$fade_site_bot_file
} else {
  NULL
}

if (!is.null(fade_site_file)) {
  fade_site_df <- read_tsv(fade_site_file, show_col_types = FALSE) %>%
    rename(Gene = gene, Position = site)

  # Join FADE site data to position scores; compute caas_decile and fade_char_score
  pos_fade <- pos_df %>%
    left_join(
      fade_site_df %>% select(Gene, Position, max_site_bf, top_target_aa, fade_biased),
      by = c("Gene", "Position")
    ) %>%
    mutate(
      fade_biased     = replace_na(fade_biased, FALSE),
      caas_decile     = dplyr::ntile(CAAS_score, 10),
      # fade_char_score: scheme_coverage-weighted FADE signal
      # scheme_coverage = sum of active GS-scheme weights at this position
      fade_char_score = ifelse(fade_biased & !is.na(scheme_coverage), scheme_coverage, 0)
    )
  has_fade_pos <- nrow(pos_fade) > 0 && any(pos_fade$fade_biased, na.rm = TRUE)
} else {
  has_fade_pos <- FALSE
}
```

`r if (exists("pos_fade") && nrow(pos_fade) > 0 && !any(pos_fade$fade_biased, na.rm = TRUE)) "> **Note:** FADE site data were loaded, but no positions passed the FADE biased-site threshold for this direction."`

### Overlap with CAAS Positions

```{r fade-pos-overlap, eval=has_fade_pos}
# Overlap: how many FADE-biased positions also rank in the top CAAS tiers?
thr_10 <- quantile(pos_fade$CAAS_score, 1 - params$top_pct,  na.rm = TRUE)
thr_5  <- quantile(pos_fade$CAAS_score, 1 - params$top5_pct, na.rm = TRUE)
thr_1  <- quantile(pos_fade$CAAS_score, 1 - params$top1_pct, na.rm = TRUE)

n_total   <- nrow(pos_fade)
n_fade    <- sum(pos_fade$fade_biased, na.rm = TRUE)
n_top10   <- sum(pos_fade$CAAS_score >= thr_10, na.rm = TRUE)
n_top5    <- sum(pos_fade$CAAS_score >= thr_5,  na.rm = TRUE)
n_top1    <- sum(pos_fade$CAAS_score >= thr_1,  na.rm = TRUE)
n_fade_top10 <- sum(pos_fade$fade_biased & pos_fade$CAAS_score >= thr_10, na.rm = TRUE)
n_fade_top5  <- sum(pos_fade$fade_biased & pos_fade$CAAS_score >= thr_5,  na.rm = TRUE)
n_fade_top1  <- sum(pos_fade$fade_biased & pos_fade$CAAS_score >= thr_1,  na.rm = TRUE)

run_fade_fisher <- function(n_fade_in_top, n_top, n_fade_total, n_total, label) {
  n_fade_rest     <- n_fade_total - n_fade_in_top
  n_nonfade_top   <- n_top - n_fade_in_top
  n_nonfade_rest  <- n_total - n_top - n_fade_rest
  mat <- matrix(
    c(n_nonfade_rest, n_fade_rest, n_nonfade_top, n_fade_in_top),
    nrow = 2,
    dimnames = list(In_top = c("FALSE", "TRUE"), FADE_biased = c("FALSE", "TRUE"))
  )
  ft <- tryCatch(fisher.test(mat), error = function(e) list(estimate=NA, conf.int=c(NA,NA), p.value=NA))
  tibble(
    CAAS_threshold = label,
    N_total = n_total, N_FADE_biased = n_fade_total,
    N_in_CAAS_top = n_top, N_FADE_in_CAAS_top = n_fade_in_top,
    Pct_FADE_in_top = sprintf("%.1f%%", 100 * n_fade_in_top / max(n_top, 1)),
    Odds_ratio = sprintf("%.2f", ft$estimate),
    P_value = formatC(ft$p.value, format = "e", digits = 2)
  )
}

fade_overlap_tbl <- bind_rows(
  run_fade_fisher(n_fade_top10, n_top10, n_fade, n_total,
                  sprintf("Top %g%% CAAS", params$top_pct  * 100)),
  run_fade_fisher(n_fade_top5,  n_top5,  n_fade, n_total,
                  sprintf("Top %g%% CAAS", params$top5_pct * 100)),
  run_fade_fisher(n_fade_top1,  n_top1,  n_fade, n_total,
                  sprintf("Top %g%% CAAS", params$top1_pct * 100))
)
knitr::kable(fade_overlap_tbl,
  caption = sprintf("FADE-biased positions (BF ≥ %s) enrichment in top-CAAS positions",
                    as.character(if (!is.null(params$fade_bf_threshold)) params$fade_bf_threshold else 100))
)
```

```{r fade-pos-decile-barplot, eval=has_fade_pos, fig.height=4.5}
# Fraction of FADE-biased positions per CAAS decile
pos_fade %>%
  filter(!is.na(caas_decile)) %>%
  group_by(caas_decile) %>%
  summarise(
    n         = n(),
    n_fade    = sum(fade_biased, na.rm = TRUE),
    frac_fade = n_fade / n,
    .groups   = "drop"
  ) %>%
  ggplot(aes(x = caas_decile, y = frac_fade, fill = caas_decile)) +
  geom_col(show.legend = FALSE) +
  scale_fill_gradient(low = .tableau10[1], high = .tableau10[3]) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title    = "Fraction of FADE-Biased Positions per CAAS Decile",
    subtitle = "Decile 10 = highest-scoring CAAS positions",
    x        = "CAAS Decile", y = "Fraction FADE-biased"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

### FADE Score vs CAAS Score

```{r fade-pos-caas-scatter, eval=has_fade_pos, fig.height=5}
# Compute per-position CAAS decile for colour coding
.fade_pos_with_decile <- pos_fade %>%
  filter(!is.na(CAAS_score), !is.na(max_site_bf)) %>%
  mutate(log10_bf = log10(pmax(max_site_bf, 0.01)))

.fade_caas_r  <- cor(.fade_pos_with_decile$CAAS_score,
                      .fade_pos_with_decile$log10_bf, use = "complete.obs")
.fade_caas_rs <- cor(.fade_pos_with_decile$CAAS_score,
                      .fade_pos_with_decile$log10_bf,
                      method = "spearman", use = "complete.obs")

ggplot(.fade_pos_with_decile,
       aes(x = CAAS_score, y = log10_bf, colour = fade_biased)) +
  geom_point(alpha = 0.5, size = 1.2) +
  geom_smooth(method = "lm", se = FALSE, colour = "black", linewidth = 0.7) +
  scale_colour_manual(values = c("FALSE" = .tableau10[1], "TRUE" = .tableau10[3]),
                      name = "FADE biased") +
  labs(
    title    = "CAAS Score vs FADE max Bayes Factor (log10)",
    subtitle = sprintf("Pearson r = %.3f | Spearman ρ = %.3f",
                       .fade_caas_r, .fade_caas_rs),
    x = "CAAS_score", y = "log10(max site BF)"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r fade-pos-char-scatter, eval=has_fade_pos, fig.height=5}
.fade_char_pos <- pos_fade %>%
  filter(!is.na(CAAS_score), fade_char_score > 0)

.fade_char_r  <- cor(pos_fade$CAAS_score, pos_fade$fade_char_score,
                     use = "complete.obs")
.fade_char_rs <- cor(pos_fade$CAAS_score, pos_fade$fade_char_score,
                     method = "spearman", use = "complete.obs")

ggplot(pos_fade, aes(x = CAAS_score, y = fade_char_score)) +
  geom_point(aes(colour = fade_biased), alpha = 0.4, size = 1.2) +
  geom_smooth(data = .fade_char_pos, method = "lm", se = FALSE,
              colour = "black", linewidth = 0.7) +
  scale_colour_manual(values = c("FALSE" = .tableau10[1], "TRUE" = .tableau10[3]),
                      name = "FADE biased") +
  labs(
    title    = "CAAS Score vs FADE Scheme-Coverage-Weighted Score",
    subtitle = sprintf("Pearson r = %.3f | Spearman ρ = %.3f",
                       .fade_char_r, .fade_char_rs),
    x = "CAAS_score", y = "fade_char_score (scheme_coverage × FADE biased)"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

### FADE Score vs Biochem Score

```{r fade-pos-biochem-scatter, eval=has_fade_pos, fig.height=5}
.fade_biochem <- pos_fade %>%
  filter(!is.na(biochem_score), !is.na(max_site_bf)) %>%
  mutate(log10_bf = log10(pmax(max_site_bf, 0.01)))

.fade_biochem_r  <- cor(.fade_biochem$biochem_score, .fade_biochem$log10_bf,
                         use = "complete.obs")
.fade_biochem_rs <- cor(.fade_biochem$biochem_score, .fade_biochem$log10_bf,
                         method = "spearman", use = "complete.obs")

ggplot(.fade_biochem, aes(x = biochem_score, y = log10_bf, colour = fade_biased)) +
  geom_point(alpha = 0.5, size = 1.2) +
  geom_smooth(method = "lm", se = FALSE, colour = "black", linewidth = 0.7) +
  scale_colour_manual(values = c("FALSE" = .tableau10[1], "TRUE" = .tableau10[3]),
                      name = "FADE biased") +
  labs(
    title    = "Biochem Score vs FADE max Bayes Factor (log10)",
    subtitle = sprintf("Pearson r = %.3f | Spearman ρ = %.3f",
                       .fade_biochem_r, .fade_biochem_rs),
    x = "biochem_score", y = "log10(max site BF)"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r fade-pos-biochem-char-scatter, eval=has_fade_pos, fig.height=5}
.fade_biochem_char_r  <- cor(pos_fade$biochem_score, pos_fade$fade_char_score,
                              use = "complete.obs")
.fade_biochem_char_rs <- cor(pos_fade$biochem_score, pos_fade$fade_char_score,
                              method = "spearman", use = "complete.obs")

ggplot(pos_fade, aes(x = biochem_score, y = fade_char_score, colour = fade_biased)) +
  geom_point(alpha = 0.4, size = 1.2) +
  scale_colour_manual(values = c("FALSE" = .tableau10[1], "TRUE" = .tableau10[3]),
                      name = "FADE biased") +
  labs(
    title    = "Biochem Score vs FADE Scheme-Coverage-Weighted Score",
    subtitle = sprintf("Pearson r = %.3f | Spearman ρ = %.3f",
                       .fade_biochem_char_r, .fade_biochem_char_rs),
    x = "biochem_score", y = "fade_char_score"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

---

## Gene-Level Scores {.tabset}

### Gene CAAS Score

```{r gene-caas-dist, fig.height=4.5}
g_thr_10 <- quantile(gene_df$gene_caas_score, 1 - params$gene_top_pct,  na.rm = TRUE)
g_thr_5  <- quantile(gene_df$gene_caas_score, 1 - params$gene_top5_pct, na.rm = TRUE)
g_thr_1  <- quantile(gene_df$gene_caas_score, 1 - params$gene_top1_pct, na.rm = TRUE)

ggplot(gene_df, aes(x = gene_caas_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = .tableau10[1], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[2]) +
  geom_vline(xintercept = g_thr_10, linetype = "dashed",
             colour = .tableau10[3], linewidth = 0.7) +
  geom_vline(xintercept = g_thr_5,  linetype = "longdash",
             colour = .tableau10[5], linewidth = 0.7) +
  geom_vline(xintercept = g_thr_1,  linetype = "dotted",
             colour = .tableau10[4], linewidth = 0.7) +
  labs(title = "Gene CAAS Score Distribution",
       subtitle = "90th percentile of position-level CAAS scores per gene. gene_caas_score is not a direct p-value transform — it aggregates position-level composite scores.",
       x = "gene_caas_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### Accumulation

`r if (!has_rand) "> **Note:** Accumulation-derived gene scores are not available for this run."`

```{r gene-rand-dist, eval=has_rand, fig.height=4.5}
ggplot(gene_df %>% filter(!is.na(gene_rand_score)), aes(x = gene_rand_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = .tableau10[5], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[6]) +
  labs(title = "Gene Randomization Score Distribution",
       subtitle = "Scheme-weighted deciles of accumulation randomization p-values",
       x = "gene_rand_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

```{r gene-rand-decile-table, eval=has_rand && nrow(accum_scheme_info) > 0}
accum_boundary_tbl <- dplyr::bind_rows(lapply(seq_len(nrow(accum_scheme_info)), function(i) {
  info <- accum_scheme_info[i, ]
  tbl <- decile_boundary_tbl(
    gene_df[[info$pval_col]],
    score_name = "Scheme score",
    lower_better = TRUE,
    value_label = "P-value",
    digits = 2
  )
  if (is.null(tbl)) return(NULL)
  tbl %>%
    mutate(
      Scheme = info$label,
      Weight = info$weight,
      `Weighted contribution` = round(`Scheme score` * Weight, 3),
      .before = 1
    )
}))

if (nrow(accum_boundary_tbl) > 0) {
  knitr::kable(
    accum_boundary_tbl,
    caption = "Accumulation empirical p-value range per decile (decile 10 = smallest p-values) and its contribution to gene_rand_score, by scheme"
  )
}
```

### FADE

`r if (!has_fade) "> **Note:** FADE results are not available for this run."`

```{r gene-fade-dist, eval=has_fade, fig.height=4.5}
ggplot(gene_df %>% filter(!is.na(gene_fade_score)), aes(x = gene_fade_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = .tableau10[9], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[10]) +
  labs(title = "Gene FADE Score Distribution",
       subtitle = "Deciles of max directional selection Bayes Factor per gene",
       x = "gene_fade_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

```{r gene-fade-bf-decile, eval=has_fade && "fade_max_bf" %in% names(gene_df), fig.height=3.5}
fade_boundary_tbl <- decile_boundary_tbl(
  gene_df$fade_max_bf,
  score_name = "gene_fade_score",
  lower_better = FALSE,
  value_label = "BF",
  digits = 1
)
if (!is.null(fade_boundary_tbl)) {
  knitr::kable(fade_boundary_tbl,
               caption = "FADE max Bayes Factor range per decile (decile 10 = highest BF, gene_fade_score = 1.0)")
}
```

### RERConverge

`r if (!has_rer) "> **Note:** RERConverge results not available for this run."`

```{r gene-rer-dist, eval=has_rer, fig.height=4.5}
ggplot(gene_df %>% filter(!is.na(gene_rer_score)), aes(x = gene_rer_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = .tableau10[7], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[8]) +
  labs(title = "Gene RERConverge Score Distribution",
       subtitle = "Deciles of evolutionary rate correlation p-values (lower p-value = higher score)",
       x = "gene_rer_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

```{r gene-rer-decile-table, eval=has_rer && "rer_min_pval" %in% names(gene_df)}
rer_boundary_tbl <- decile_boundary_tbl(
  gene_df$rer_min_pval,
  score_name = "gene_rer_score",
  lower_better = TRUE,
  value_label = "P-value",
  digits = 2
)
if (!is.null(rer_boundary_tbl)) {
  knitr::kable(rer_boundary_tbl,
               caption = "RERConverge p-value range per decile (decile 10 = smallest p-values, gene_rer_score = 1.0)")
}
```

```{r gene-rer-caas-scatter, eval=has_rer, fig.height=5}
# How does RER signal relate to CAAS signal?
ggplot(gene_df %>% filter(!is.na(gene_rer_score), !is.na(gene_caas_score)),
       aes(x = gene_caas_score, y = gene_rer_score)) +
  geom_point(alpha = 0.4, size = 0.9, colour = .tableau10[7]) +
  geom_smooth(method = "lm", se = TRUE, colour = .tableau10[8], linewidth = 0.7) +
  ggrepel::geom_text_repel(
    data = gene_df %>%
      filter(!is.na(gene_rer_score), !is.na(gene_caas_score)) %>%
      filter(gene_rer_score > quantile(gene_rer_score, 0.90, na.rm=TRUE) |
             gene_caas_score > quantile(gene_caas_score, 0.90, na.rm=TRUE)),
    aes(label = Gene), size = 2.5, max.overlaps = 15
  ) +
  labs(title = "Gene CAAS Score vs RERConverge Score",
       subtitle = sprintf("r = %.3f | Genes in top 10%% of either score labelled",
                          safe_cor(gene_df$gene_caas_score, gene_df$gene_rer_score)),
       x = "gene_caas_score", y = "gene_rer_score") +
  theme(plot.title = element_text(face = "bold"))
```

### Gene Composite Score

```{r gene-composite-dist, fig.height=5}
gc_thr_10 <- quantile(gene_df$gene_composite_score, 1 - params$gene_top_pct,  na.rm = TRUE)
gc_thr_5  <- quantile(gene_df$gene_composite_score, 1 - params$gene_top5_pct, na.rm = TRUE)
gc_thr_1  <- quantile(gene_df$gene_composite_score, 1 - params$gene_top1_pct, na.rm = TRUE)

ggplot(gene_df, aes(x = gene_composite_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = .tableau10[1], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[2]) +
  geom_vline(xintercept = gc_thr_10, linetype = "dashed",
             colour = .tableau10[3], linewidth = 0.7) +
  geom_vline(xintercept = gc_thr_5,  linetype = "longdash",
             colour = .tableau10[5], linewidth = 0.7) +
  geom_vline(xintercept = gc_thr_1,  linetype = "dotted",
             colour = .tableau10[4], linewidth = 0.7) +
  annotate("text", x = gc_thr_10, y = Inf,
           label = sprintf("Top %g%%", params$gene_top_pct * 100),
           vjust = 2,    hjust = -0.1, colour = .tableau10[3], size = 3.5) +
  annotate("text", x = gc_thr_5,  y = Inf,
           label = sprintf("Top %g%%", params$gene_top5_pct * 100),
           vjust = 2.75, hjust = -0.1, colour = .tableau10[5], size = 3.5) +
  annotate("text", x = gc_thr_1,  y = Inf,
           label = sprintf("Top %g%%", params$gene_top1_pct * 100),
           vjust = 3.5,  hjust = -0.1, colour = .tableau10[4], size = 3.5) +
  labs(title = "Gene Composite Score Distribution",
       subtitle = sprintf("Unweighted mean of available gene scores | n = %s genes",
                          format(nrow(gene_df), big.mark = ",")),
       x = "gene_composite_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### Gene Threshold Profiles

```{r gene-threshold-profiles, fig.height=14, fig.width=12}
# Same idea as position-level threshold profiles but for gene-level scores.
# For each top-N% × direction bucket, show overlaid densities of gene sub-scores.

gene_score_cols_tp <- "gene_caas_score"
if (has_rand)     gene_score_cols_tp <- c(gene_score_cols_tp, "gene_rand_score")
if (has_rer)      gene_score_cols_tp <- c(gene_score_cols_tp, "gene_rer_score")
if (has_fade)     gene_score_cols_tp <- c(gene_score_cols_tp, "gene_fade_score")

# Direction per gene is inherited from its top-scoring position
gene_direction_tp <- pos_df %>%
  group_by(Gene) %>%
  slice_max(CAAS_score, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(Gene, gene_direction = change_side)

gene_df_dir <- gene_df %>% left_join(gene_direction_tp, by = "Gene")

# Directional subsets: "all" = no filter; "top" = top or both; "bottom" = bottom or both
gene_dir_subsets <- list(
  all    = gene_df_dir,
  top    = gene_df_dir %>% filter(gene_direction %in% c("top",    "both")),
  bottom = gene_df_dir %>% filter(gene_direction %in% c("bottom", "both"))
)

make_gene_threshold_panel <- function(df_dir, pct_label, pct_val, direction) {
  # Threshold is computed within the directional subset, not globally
  thr <- quantile(df_dir$gene_composite_score, 1 - pct_val, na.rm = TRUE)
  subset_df <- df_dir %>% filter(gene_composite_score >= thr)

  if (nrow(subset_df) == 0 || length(gene_score_cols_tp) < 2) {
    return(
      ggplot() +
        annotate("text", x = 0.5, y = 0.5,
                 label = ifelse(nrow(subset_df) == 0, "No genes", "Single score only"),
                 size = 5, colour = "grey50") +
        labs(title = sprintf("Top %s — %s", pct_label, direction),
             x = "Score (0–1)", y = "Density") +
        xlim(0, 1) + theme_minimal()
    )
  }

  long_df <- subset_df %>%
    select(all_of(gene_score_cols_tp)) %>%
    pivot_longer(everything(), names_to = "Score_type", values_to = "Score") %>%
    filter(!is.na(Score)) %>%
    mutate(Score_type = factor(Score_type, levels = gene_score_cols_tp))

  ggplot(long_df, aes(x = Score, fill = Score_type, colour = Score_type)) +
    geom_density(alpha = 0.35, linewidth = 0.6) +
    scale_fill_manual(values   = .tableau10[seq_along(gene_score_cols_tp)],
                      labels   = sub("gene_", "", sub("_score", "", gene_score_cols_tp))) +
    scale_colour_manual(values = .tableau10[seq_along(gene_score_cols_tp)],
                        labels = sub("gene_", "", sub("_score", "", gene_score_cols_tp))) +
    scale_x_continuous(limits = c(0, 1), expand = c(0.01, 0)) +
    labs(
      title    = sprintf("Top %s — %s", pct_label, direction),
      subtitle = sprintf("n = %d genes (thr = %.3f)", nrow(subset_df), thr),
      x = "Score (0–1)", y = "Density",
      fill = NULL, colour = NULL
    ) +
    theme(
      plot.title      = element_text(face = "bold", size = 11),
      plot.subtitle   = element_text(size = 9, colour = "grey40"),
      legend.position = "bottom",
      legend.text     = element_text(size = 8)
    )
}

gene_panels <- list()
for (direction in names(gene_dir_subsets)) {
  df_d <- gene_dir_subsets[[direction]]
  gene_panels[[paste0("10pct_", direction)]] <-
    make_gene_threshold_panel(df_d, "10%", params$gene_top_pct,  direction)
  gene_panels[[paste0("5pct_",  direction)]] <-
    make_gene_threshold_panel(df_d, "5%",  params$gene_top5_pct, direction)
  gene_panels[[paste0("1pct_",  direction)]] <-
    make_gene_threshold_panel(df_d, "1%",  params$gene_top1_pct, direction)
}

gridExtra::grid.arrange(
  gene_panels[["10pct_all"]],    gene_panels[["5pct_all"]],    gene_panels[["1pct_all"]],
  gene_panels[["10pct_top"]],    gene_panels[["5pct_top"]],    gene_panels[["1pct_top"]],
  gene_panels[["10pct_bottom"]], gene_panels[["5pct_bottom"]], gene_panels[["1pct_bottom"]],
  ncol = 3,
  top  = grid::textGrob(
    "Gene score profiles for top-scoring gene subsets (by gene_composite_score)",
    gp = grid::gpar(fontface = "bold", fontsize = 13)
  ),
  left = grid::textGrob(
    "Direction:  all                            top                          bottom",
    rot = 90, gp = grid::gpar(fontsize = 9, col = "grey40")
  )
)
```

### Correlation Analysis

```{r correlation-scatter, eval=(nrow(corr_df) > 0), fig.height=14}
# Build scatter matrix of gene-level scores
score_cols <- "gene_caas_score"
if (has_rand)     score_cols <- c(score_cols, "gene_rand_score")
if (has_rer)      score_cols <- c(score_cols, "gene_rer_score")
if (has_fade)     score_cols <- c(score_cols, "gene_fade_score")

if (length(score_cols) >= 2) {
  pairs_df <- gene_df %>%
    select(all_of(score_cols)) %>%
    drop_na()

  # Create pairs plot
  pairs_long <- expand.grid(
    x_var = score_cols, y_var = score_cols,
    stringsAsFactors = FALSE
  ) %>%
    filter(x_var != y_var)

  plots <- list()
  for (i in seq_len(nrow(pairs_long))) {
    xv <- pairs_long$x_var[i]
    yv <- pairs_long$y_var[i]
    r_info <- corr_df %>%
      filter((score_a == xv & score_b == yv) | (score_a == yv & score_b == xv))
    label <- if (nrow(r_info) > 0) {
      sprintf("r=%.2f / rho=%.2f", r_info$pearson_r[1], r_info$spearman_r[1])
    } else ""

    plots[[length(plots) + 1]] <- ggplot(pairs_df, aes(x = .data[[xv]], y = .data[[yv]])) +
      geom_point(alpha = 0.3, size = 0.8, colour = .tableau10[1]) +
      geom_smooth(method = "lm", se = FALSE, colour = .tableau10[4], linewidth = 0.6) +
      annotate("text", x = Inf, y = -Inf, label = label,
               hjust = 1.1, vjust = -0.5, size = 3, colour = .tableau10[4]) +
      labs(x = sub("gene_", "", sub("_score", "", xv)),
           y = sub("gene_", "", sub("_score", "", yv))) +
      theme(axis.title = element_text(size = 9))
  }

  if (length(plots) > 0) {
    gridExtra::grid.arrange(grobs = plots,
                            ncol = min(length(score_cols) - 1, 3),
                            top = grid::textGrob(
                              "Gene Score Correlations (r = Pearson; rho = Spearman rank correlation)",
                              gp = grid::gpar(fontface = "bold", fontsize = 13)))
  }
}
```

```{r correlation-table, eval=(nrow(corr_df) > 0)}
DT::datatable(
  corr_df %>%
    mutate(across(where(is.numeric), ~round(., 4))),
  rownames = FALSE,
  options  = list(pageLength = 10, dom = "t"),
  caption  = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "Pairwise correlations between gene-level scores"
  )
)
```

### Top Genes

```{r top-genes-table}
gene_display_cols <- c("Gene", "n_positions", "gene_caas_score")
if (has_rand)     gene_display_cols <- c(gene_display_cols, "gene_rand_score")
if (has_rer)      gene_display_cols <- c(gene_display_cols, "gene_rer_score")
if (has_fade)     gene_display_cols <- c(gene_display_cols, "gene_fade_score")
gene_display_cols <- c(gene_display_cols, "gene_composite_score")

DT::datatable(
  gene_df %>%
    arrange(desc(gene_composite_score)) %>%
    select(all_of(gene_display_cols)) %>%
    mutate(across(where(is.numeric) & !matches("n_positions"), ~round(., 4))),
  filter    = "top",
  rownames  = FALSE,
  options   = list(pageLength = 25, scrollX = TRUE),
  caption   = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Gene-level scores | Trait: %s", params$traitname)
  )
)
```

---

## Gene-Level Characterisation {.tabset}

This section accumulates PrimateAI-3D signals at the gene level, testing
whether genes with high composite scores tend to concentrate pathogenic
PrimateAI-3D annotations.

> **Note:** PrimateAI-3D scores require an access request to Illumina —
> see illumina.com/science/genomics-research/articles/primateai-3d.html

### Gene VEP Accumulation

`r if (!has_vep_primateai) "> **Note:** PrimateAI-3D scores are not available — gene-level VEP accumulation skipped."`

```{r gene-vep-pai-accum, eval=has_vep_primateai, fig.height=5}
# Per-gene: mean pai_char_score (scheme-weighted pathogenicity).
gene_vep_pai <- pos_vep_pai %>%
  group_by(Gene) %>%
  summarise(
    n_pai_positions  = n(),
    mean_pai_char    = mean(pai_char_score, na.rm = TRUE),
    sum_pai_char     = sum(pai_char_score,  na.rm = TRUE),
    n_pai_any_signal = sum(!is.na(pai_char_score) & pai_char_score > 0, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  inner_join(gene_df %>% select(Gene, gene_caas_score, gene_composite_score,
                                n_positions),
             by = "Gene")

ggplot(gene_vep_pai, aes(x = gene_composite_score, y = mean_pai_char,
                          size = n_positions, colour = gene_composite_score)) +
  geom_point(alpha = 0.65) +
  ggrepel::geom_text_repel(
    data = gene_vep_pai %>% slice_max(mean_pai_char, n = 15, with_ties = FALSE),
    aes(label = Gene), size = 2.8, max.overlaps = 20,
    show.legend = FALSE
  ) +
  geom_vline(xintercept = gc_thr_10, linetype = "dashed",
             colour = .tableau10[3], linewidth = 0.7) +
  scale_colour_viridis_c(name = "Gene\nComposite", option = "plasma") +
  scale_size_continuous(name = "N positions", range = c(1, 8)) +
  labs(
    title    = "Gene Composite Score vs Mean PrimateAI-3D Scheme-Weighted Score",
    subtitle = "Labels: top-15 genes by mean pai_char_score; dashed = composite top 10% threshold",
    x = "Gene Composite Score", y = "Mean pai_char_score per gene"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r gene-vep-pai-top-bar, eval=has_vep_primateai, fig.height=6}
# Top 30 genes by mean pai_char_score, coloured by gene_composite_score
top_pai_genes <- gene_vep_pai %>%
  filter(n_pai_positions >= 2, mean_pai_char > 0) %>%
  slice_max(mean_pai_char, n = 30, with_ties = FALSE) %>%
  arrange(mean_pai_char) %>%
  mutate(Gene = factor(Gene, levels = Gene))

ggplot(top_pai_genes, aes(x = mean_pai_char, y = Gene,
                           fill = gene_composite_score)) +
  geom_col() +
  scale_fill_viridis_c(name = "Gene Composite\nScore", option = "plasma") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Top 30 Genes by Mean PrimateAI-3D Scheme-Weighted Score",
    subtitle = paste0("Genes with \u2265 2 positions with PAI3D coverage\n",
                      "pai_char_score = \u03a3 scheme_weight \u00d7 max(PAI3D per CAAP group)"),
    x = "Mean pai_char_score", y = NULL
  ) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.y = element_text(size = 8))
```

### Gene FADE Characterisation

`r if (!has_fade) "> **Note:** FADE results are not available — gene-level FADE characterisation skipped."`

```{r gene-fade-characterisation, eval=has_fade, fig.height=5}
gene_fade_df <- gene_df %>%
  filter(!is.na(gene_fade_score), !is.na(gene_composite_score))

ggplot(gene_fade_df, aes(x = gene_composite_score, y = gene_fade_score,
                         colour = gene_caas_score, size = n_positions)) +
  geom_point(alpha = 0.65) +
  geom_smooth(method = "lm", se = TRUE, colour = "grey35", linewidth = 0.7) +
  ggrepel::geom_text_repel(
    data = gene_fade_df %>% slice_max(gene_fade_score, n = 15, with_ties = FALSE),
    aes(label = Gene), size = 2.5, max.overlaps = 20, show.legend = FALSE
  ) +
  scale_colour_viridis_c(name = "Gene\nCAAS", option = "plasma") +
  scale_size_continuous(name = "N positions", range = c(1, 7)) +
  labs(
    title = "Gene Composite Score vs FADE Score",
    subtitle = "Labels mark genes with the strongest FADE signal",
    x = "Gene Composite Score", y = "gene_fade_score"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

### Gene Accumulation Characterisation

`r if (!has_rand) "> **Note:** Accumulation-derived gene scores are not available — gene-level accumulation characterisation skipped."`

```{r gene-accum-characterisation, eval=has_rand, fig.height=5}
gene_accum_df <- gene_df %>%
  filter(!is.na(gene_rand_score), !is.na(gene_composite_score))

ggplot(gene_accum_df, aes(x = gene_composite_score, y = gene_rand_score,
                          colour = gene_caas_score, size = n_positions)) +
  geom_point(alpha = 0.65) +
  geom_smooth(method = "lm", se = TRUE, colour = "grey35", linewidth = 0.7) +
  ggrepel::geom_text_repel(
    data = gene_accum_df %>% slice_max(gene_rand_score, n = 15, with_ties = FALSE),
    aes(label = Gene), size = 2.5, max.overlaps = 20, show.legend = FALSE
  ) +
  scale_colour_viridis_c(name = "Gene\nCAAS", option = "plasma") +
  scale_size_continuous(name = "N positions", range = c(1, 7)) +
  labs(
    title = "Gene Composite Score vs Accumulation Score",
    subtitle = "Labels mark genes with the strongest accumulation-derived randomisation score",
    x = "Gene Composite Score", y = "gene_rand_score"
  ) +
  theme(plot.title = element_text(face = "bold"))
```


---

## Genomic Accumulation {.tabset}

`r if (!has_genomic) "> **Note:** Genomic coordinates (gene_ensembl_file) are not available for this run — genomic accumulation skipped."`

```{r genomic-setup, eval=has_genomic}
# Join gene_df with genomic coordinates, then annotate with VEP
gene_genomic <- gene_df %>%
  select(Gene, gene_caas_score, gene_composite_score, n_positions) %>%
  inner_join(genomic_raw, by = "Gene") %>%
  # Standardise chromosome labels (remove "chr" prefix if present)
  mutate(
    chr_label = sub("^chr", "", chr),
    # Order chromosomes numerically then X, Y, MT
    chr_num   = suppressWarnings(as.integer(chr_label)),
    chr_order = dplyr::case_when(
      !is.na(chr_num)       ~ chr_num,
      chr_label == "X"      ~ 23L,
      chr_label == "Y"      ~ 24L,
      chr_label %in% c("MT","M") ~ 25L,
      TRUE                  ~ 26L
    )
  ) %>%
  arrange(chr_order, gene_mid)

# Optionally add mean PAI characterisation score
if (has_vep_primateai && exists("gene_vep_pai")) {
  gene_genomic <- gene_genomic %>%
    left_join(gene_vep_pai %>% select(Gene, mean_pai_char),
              by = "Gene")
} else {
  gene_genomic$mean_pai_char <- NA_real_
}

# Build a cumulative x-axis offset per chromosome for a stitched Manhattan view
chr_offsets <- gene_genomic %>%
  group_by(chr_order, chr_label) %>%
  summarise(chr_max = max(gene_end, na.rm = TRUE), .groups = "drop") %>%
  arrange(chr_order) %>%
  mutate(offset = cumsum(lag(chr_max, default = 0L)) + (chr_order - 1L) * 5e6L)

gene_genomic <- gene_genomic %>%
  left_join(chr_offsets %>% select(chr_order, offset, chr_label), by = c("chr_order", "chr_label")) %>%
  mutate(manhattan_x = gene_mid + offset)

chr_label_positions <- chr_offsets %>%
  left_join(gene_genomic %>% group_by(chr_order) %>%
              summarise(mid_x = mean(manhattan_x, na.rm = TRUE), .groups = "drop"),
            by = "chr_order")
```

```{r window-helper, eval=has_genomic}
# Helper: tile chromosomes into non-overlapping windows of win_size bp.
# Each gene/position is assigned to the window containing its genomic midpoint.
# Returns a dataframe with one row per window, summarising mean and max score.
make_window_accum <- function(df, score_col,
                               win_size = params$window_size_bp) {
  df %>%
    mutate(
      win_id    = floor(gene_mid / win_size),
      win_start = win_id * win_size,
      win_end   = win_start + win_size - 1L,
      win_mid   = win_start + win_size / 2L
    ) %>%
    group_by(chr_order, chr_label, win_id, win_start, win_end, win_mid) %>%
    summarise(
      n_genes    = dplyr::n(),
      mean_score = mean(.data[[score_col]], na.rm = TRUE),
      max_score  = max(.data[[score_col]],  na.rm = TRUE),
      genes      = paste(unique(Gene), collapse = ", "),
      .groups    = "drop"
    )
}

# Stitch window midpoints onto the cumulative chromosome x-axis (same offsets
# as gene_genomic so windows align with the per-gene manhattans above).
stitch_windows <- function(win_df) {
  win_df %>%
    left_join(chr_offsets %>% select(chr_order, offset), by = "chr_order") %>%
    mutate(win_x = win_mid + offset,
           chr_even = chr_order %% 2 == 0)
}
```

### CAAS Score Manhattan

```{r genomic-manhattan-caas, eval=has_genomic, fig.height=5, fig.width=14}
# Manhattan-style plot: x = genomic position (stitched), y = gene_caas_score
# Alternating chromosome colours for readability.
gene_genomic <- gene_genomic %>%
  mutate(chr_even = chr_order %% 2 == 0)

ggplot(gene_genomic,
       aes(x = manhattan_x, y = gene_caas_score,
           colour = chr_even, size = n_positions)) +
  geom_point(alpha = 0.65, show.legend = FALSE) +
  geom_hline(yintercept = g_thr_10, linetype = "dashed",
             colour = "grey40", linewidth = 0.6) +
  ggrepel::geom_text_repel(
    data = gene_genomic %>% slice_max(gene_caas_score, n = 20, with_ties = FALSE),
    aes(label = Gene), size = 2.2, max.overlaps = 15,
    show.legend = FALSE, colour = "black"
  ) +
  scale_colour_manual(
    values = c("FALSE" = .tableau10[1], "TRUE" = .tableau10[2]),
    guide = "none"
  ) +
  scale_size_continuous(name = "N positions", range = c(0.5, 5)) +
  scale_x_continuous(
    breaks = chr_label_positions$mid_x,
    labels = chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  labs(
    title    = "Gene CAAS Score — Manhattan View",
    subtitle = sprintf("%d genes with genomic coordinates | dashed = top %g%% threshold",
                       nrow(gene_genomic), params$gene_top_pct * 100),
    x = "Chromosome", y = "Gene CAAS Score"
  ) +
  theme(
    plot.title      = element_text(face = "bold"),
    axis.text.x     = element_text(size = 7, angle = 0),
    panel.grid.major.x = element_blank()
  )
```

### Composite Score Manhattan

```{r genomic-manhattan-composite, eval=has_genomic, fig.height=5, fig.width=14}
ggplot(gene_genomic,
       aes(x = manhattan_x, y = gene_composite_score,
           colour = chr_even, size = n_positions)) +
  geom_point(alpha = 0.65, show.legend = FALSE) +
  geom_hline(yintercept = gc_thr_10, linetype = "dashed",
             colour = "grey40", linewidth = 0.6) +
  ggrepel::geom_text_repel(
    data = gene_genomic %>% slice_max(gene_composite_score, n = 20, with_ties = FALSE),
    aes(label = Gene), size = 2.2, max.overlaps = 15,
    show.legend = FALSE, colour = "black"
  ) +
  scale_colour_manual(
    values = c("FALSE" = .tableau10[1], "TRUE" = .tableau10[2]),
    guide = "none"
  ) +
  scale_size_continuous(name = "N positions", range = c(0.5, 5)) +
  scale_x_continuous(
    breaks = chr_label_positions$mid_x,
    labels = chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  labs(
    title    = "Gene Composite Score — Manhattan View",
    subtitle = sprintf("dashed = top %g%% threshold", params$gene_top_pct * 100),
    x = "Chromosome", y = "Gene Composite Score"
  ) +
  theme(
    plot.title      = element_text(face = "bold"),
    axis.text.x     = element_text(size = 7, angle = 0),
    panel.grid.major.x = element_blank()
  )
```

### PAI3D Signal Manhattan

`r if (!has_vep_primateai) "> **Note:** PrimateAI-3D scores not available — PAI3D Manhattan skipped."`

```{r genomic-manhattan-pai, eval=has_genomic && has_vep_primateai && any(!is.na(gene_genomic$mean_pai_char)), fig.height=5, fig.width=14}
ggplot(gene_genomic %>% filter(!is.na(mean_pai_char)),
       aes(x = manhattan_x, y = gene_composite_score,
           size   = mean_pai_char,
           colour = mean_pai_char)) +
  geom_point(alpha = 0.70) +
  geom_hline(yintercept = gc_thr_10, linetype = "dashed",
             colour = "grey60", linewidth = 0.5) +
  ggrepel::geom_text_repel(
    data = gene_genomic %>%
      filter(!is.na(mean_pai_char)) %>%
      slice_max(mean_pai_char, n = 20, with_ties = FALSE),
    aes(label = Gene), size = 2.2, max.overlaps = 12,
    show.legend = FALSE, colour = "black"
  ) +
  scale_colour_viridis_c(name = "Mean PAI3D\nchar score", option = "inferno") +
  scale_size_continuous(name = "Mean PAI3D\nchar score", range = c(0.5, 6)) +
  scale_x_continuous(
    breaks = chr_label_positions$mid_x,
    labels = chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  guides(colour = guide_legend(), size = guide_legend()) +
  labs(
    title    = "Gene Composite Score + PrimateAI-3D Characterisation — Manhattan View",
    subtitle = "Bubble size and colour: mean scheme-weighted PAI3D characterisation score across positions",
    x = "Chromosome", y = "Gene Composite Score"
  ) +
  theme(
    plot.title      = element_text(face = "bold"),
    axis.text.x     = element_text(size = 7),
    panel.grid.major.x = element_blank()
  )
```

### CAAS Window Accumulation (Positions)

`r if (!has_genomic) "> **Note:** Genomic coordinates not available — position window accumulation skipped."`

```{r pos-window-accum, eval=has_genomic, fig.height=5, fig.width=14}
# Tile position-level CAAS scores into genomic windows.
# pos_genomic is built in the pos-caas-manhattan-setup chunk above.
# We use the same pos_chr_offsets to align x-coordinates with the gene-level plots.
win_size_label <- if (params$window_size_bp >= 1e6) {
  sprintf("%g Mb", params$window_size_bp / 1e6)
} else {
  sprintf("%g kb", params$window_size_bp / 1e3)
}

pos_wins <- pos_genomic %>%
  rename(Gene = Gene) %>%
  make_window_accum("CAAS_score", win_size = params$window_size_bp) %>%
  left_join(pos_chr_offsets %>% select(chr_order, offset), by = "chr_order") %>%
  mutate(n_positions = n_genes,
         win_x    = win_mid + offset,
         chr_even = chr_order %% 2 == 0)

# Label the top windows as genomic regions, not gene names
top_pos_wins <- pos_wins %>%
  slice_max(mean_score, n = 5, with_ties = FALSE) %>%
  mutate(region_label = sprintf("chr%s:%0.1f-%0.1f Mb", chr_label, win_start / 1e6, win_end / 1e6))

ggplot(pos_wins, aes(x = win_x, y = mean_score, fill = chr_even)) +
  geom_col(width = params$window_size_bp * 0.9, alpha = 0.75) +
  geom_point(aes(y = max_score, colour = chr_even), size = 1.2, alpha = 0.85, show.legend = FALSE) +
  ggrepel::geom_text_repel(
    data = top_pos_wins,
    aes(y = max_score, label = sprintf("%s\n(%d positions)", region_label, n_positions)),
    size = 2.2, max.overlaps = 10, colour = "black",
    show.legend = FALSE
  ) +
  scale_fill_manual(
    values = c("FALSE" = .tableau10[1], "TRUE" = .tableau10[2]),
    guide = "none"
  ) +
  scale_colour_manual(
    values = c("FALSE" = .tableau10[3], "TRUE" = .tableau10[4]),
    guide = "none"
  ) +
  scale_x_continuous(
    breaks = pos_chr_label_positions$mid_x,
    labels = pos_chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  labs(
    title    = sprintf("Position-Level CAAS Score — %s Window Accumulation", win_size_label),
    subtitle = sprintf("Bars = mean CAAS score per region | Points = max score | %d windows across %d chromosomes",
                       nrow(pos_wins), length(unique(pos_wins$chr_order))),
    x = "Chromosome", y = "CAAS Score"
  ) +
  theme(
    plot.title         = element_text(face = "bold"),
    axis.text.x        = element_text(size = 7),
    panel.grid.major.x = element_blank()
  )
```

### CAAS Window Accumulation (Genes)

`r if (!has_genomic) "> **Note:** Genomic coordinates not available — gene-level CAAS window accumulation skipped."`

```{r gene-caas-window-accum, eval=has_genomic, fig.height=5, fig.width=14}
gene_caas_wins <- gene_genomic %>%
  make_window_accum("gene_caas_score", win_size = params$window_size_bp) %>%
  stitch_windows()

top_caas_wins <- gene_caas_wins %>%
  slice_max(mean_score, n = 5, with_ties = FALSE) %>%
  mutate(region_label = sprintf("chr%s:%0.1f-%0.1f Mb", chr_label, win_start / 1e6, win_end / 1e6))

ggplot(gene_caas_wins, aes(x = win_x, y = mean_score, fill = chr_even)) +
  geom_col(width = params$window_size_bp * 0.9, alpha = 0.75) +
  geom_point(aes(y = max_score, colour = chr_even), size = 1.5, alpha = 0.85, show.legend = FALSE) +
  ggrepel::geom_text_repel(
    data = top_caas_wins,
    aes(y = max_score, label = sprintf("%s\n(%d genes)", region_label, n_genes)),
    size = 2.2, max.overlaps = 10, colour = "black",
    show.legend = FALSE
  ) +
  scale_fill_manual(
    values = c("FALSE" = .tableau10[1], "TRUE" = .tableau10[2]),
    guide = "none"
  ) +
  scale_colour_manual(
    values = c("FALSE" = .tableau10[3], "TRUE" = .tableau10[4]),
    guide = "none"
  ) +
  scale_x_continuous(
    breaks = chr_label_positions$mid_x,
    labels = chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  labs(
    title    = sprintf("Gene CAAS Score — %s Window Accumulation", win_size_label),
    subtitle = sprintf("Bars = mean gene_caas_score per region | Points = max | %d windows",
                       nrow(gene_caas_wins)),
    x = "Chromosome", y = "Gene CAAS Score"
  ) +
  theme(
    plot.title         = element_text(face = "bold"),
    axis.text.x        = element_text(size = 7),
    panel.grid.major.x = element_blank()
  )
```

### Composite Window Accumulation (Genes)

`r if (!has_genomic) "> **Note:** Genomic coordinates not available — gene-level composite window accumulation skipped."`

```{r gene-composite-window-accum, eval=has_genomic, fig.height=5, fig.width=14}
gene_comp_wins <- gene_genomic %>%
  make_window_accum("gene_composite_score", win_size = params$window_size_bp) %>%
  stitch_windows()

top_comp_wins <- gene_comp_wins %>%
  slice_max(mean_score, n = 5, with_ties = FALSE) %>%
  mutate(region_label = sprintf("chr%s:%0.1f-%0.1f Mb", chr_label, win_start / 1e6, win_end / 1e6))

ggplot(gene_comp_wins, aes(x = win_x, y = mean_score, fill = chr_even)) +
  geom_col(width = params$window_size_bp * 0.9, alpha = 0.75) +
  geom_point(aes(y = max_score, colour = chr_even), size = 1.5, alpha = 0.85, show.legend = FALSE) +
  ggrepel::geom_text_repel(
    data = top_comp_wins,
    aes(y = max_score, label = sprintf("%s\n(%d genes)", region_label, n_genes)),
    size = 2.2, max.overlaps = 10, colour = "black",
    show.legend = FALSE
  ) +
  scale_fill_manual(
    values = c("FALSE" = .tableau10[1], "TRUE" = .tableau10[2]),
    guide = "none"
  ) +
  scale_colour_manual(
    values = c("FALSE" = .tableau10[3], "TRUE" = .tableau10[4]),
    guide = "none"
  ) +
  scale_x_continuous(
    breaks = chr_label_positions$mid_x,
    labels = chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  labs(
    title    = sprintf("Gene Composite Score — %s Window Accumulation", win_size_label),
    subtitle = sprintf("Bars = mean gene_composite_score per region | Points = max | %d windows",
                       nrow(gene_comp_wins)),
    x = "Chromosome", y = "Gene Composite Score"
  ) +
  theme(
    plot.title         = element_text(face = "bold"),
    axis.text.x        = element_text(size = 7),
    panel.grid.major.x = element_blank()
  )
```

---

## Summary

```{r summary-stats}
n_components <- 4L  # biochem, asr, convergence, parallel

n_gene_components <- 1L  # gene_caas always present
if (has_rand)     n_gene_components <- n_gene_components + 1L
if (has_rer)      n_gene_components <- n_gene_components + 1L
if (has_fade)     n_gene_components <- n_gene_components + 1L

summary_tbl <- tibble(
  Metric = c(
    "Positions scored", "Genes scored",
    "Position-level components", "Gene-level components",
    "CAAS Score (median)", "CAAS Score (top 10% threshold)",
    "Gene Composite (median)", "Gene Composite (top 10% threshold)",
    "FADE available",
    "Randomization available", "RERConverge available",
    "VEP PrimateAI-3D available",
    "Genomic coordinates available"
  ),
  Value = c(
    format(nrow(pos_df), big.mark = ","),
    format(nrow(gene_df), big.mark = ","),
    as.character(n_components),
    as.character(n_gene_components),
    sprintf("%.4f", median(pos_df$CAAS_score, na.rm = TRUE)),
    sprintf("%.4f", thr_10),
    sprintf("%.4f", median(gene_df$gene_composite_score, na.rm = TRUE)),
    sprintf("%.4f", gc_thr_10),
    ifelse(has_fade,         "Yes", "No"),
    ifelse(has_rand,         "Yes", "No"),
    ifelse(has_rer,          "Yes", "No"),
    ifelse(has_vep_primateai,"Yes", "No"),
    ifelse(has_genomic,      "Yes", "No")
  )
)

knitr::kable(summary_tbl, align = c("l", "r"))
```

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>
