---
title: "CAAS Scoring Report"
subtitle: "`r paste0(params$traitname, ' — combined-phenotype CAAS scoring')`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
    self_contained: true
params:
  position_scores_file: NULL
  gene_scores_file: NULL
  gene_corr_file: NULL
  traitname: "unknown_trait"
  output_dir: "."
  top_pct: 0.10
  top5_pct: 0.05
  top1_pct: 0.01
  gene_top_pct: 0.10
  gene_top5_pct: 0.05
  gene_top1_pct: 0.01
  stress_summary_file: NULL
  stress_corr_file: NULL
  stress_rank_file: NULL
  stress_overlap_file: NULL
  stress_variants_file: NULL
  stress_loadings_file: NULL
  fade_site_top_file: NULL
  fade_site_bot_file: NULL
  vep_transvar_file: NULL
  vep_primateai_file: NULL
  vep_aa2prot_file: NULL
  genomic_info_file: NULL
  window_size_bp: 1000000
  direction: "combined"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  out.width = "100%"
)

suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(ggplot2)
  library(DT)
  library(scales)
  library(ggrepel)
})

# Tableau_10 colour swatch for pipeline-wide consistency
if (requireNamespace("paletteer", quietly = TRUE)) {
  .tableau10 <- as.character(paletteer::paletteer_d("ggthemes::Tableau_10"))
} else {
  .tableau10 <- c(
    "#4E79A7", "#F28E2B", "#E15759", "#76B7B2", "#59A14F",
    "#EDC948", "#B07AA1", "#FF9DA7", "#9C755F", "#BAB0AC"
  )
}
theme_set(theme_minimal(base_size = 13))

# Safe correlation helper: returns NA (not an error) when inputs have fewer
# than 3 complete pairs — avoids crashes in sparse optional-module sections.
safe_cor <- function(x, y, method = "pearson") {
  ok <- complete.cases(x, y)
  if (sum(ok) < 3) return(NA_real_)
  cor(x[ok], y[ok], method = method)
}

format_bound <- function(x, digits = 2) {
  ifelse(
    is.na(x),
    NA_character_,
    ifelse(abs(x) < 1e-3 | abs(x) >= 1e3,
           formatC(x, format = "e", digits = digits),
           format(round(x, digits), nsmall = digits, trim = TRUE))
  )
}

decile_boundary_tbl <- function(values, score_name,
                                lower_better = TRUE,
                                value_label = "Value",
                                digits = 2) {
  vals <- values[is.finite(values) & !is.na(values)]
  if (length(vals) < 10) return(NULL)
  breaks <- quantile(vals, probs = seq(0, 1, 0.1), na.rm = TRUE, names = FALSE)
  deciles <- if (lower_better) 10:1 else 1:10
  # For lower-better metrics (e.g. p-values), decile 10 should map to the
  # smallest interval [q0, q10%], decile 9 to [q10%, q20%], ..., decile 1 to
  # the largest interval [q90%, q100%].
  lower_bounds <- breaks[1:10]
  upper_bounds <- c(breaks[2:10], max(vals, na.rm = TRUE))
  score_vals <- (deciles - 1) / 9
  tibble(
    Decile = deciles,
    `Lower bound` = format_bound(lower_bounds, digits = digits),
    `Upper bound` = format_bound(upper_bounds, digits = digits),
    !!score_name := round(score_vals, 3)
  ) %>%
    rename(
      !!paste(value_label, "lower bound") := `Lower bound`,
      !!paste(value_label, "upper bound") := `Upper bound`
    )
}
```

```{r load-data}
pos_df  <- read_tsv(params$position_scores_file, show_col_types = FALSE)
gene_df <- read_tsv(params$gene_scores_file, show_col_types = FALSE)
corr_df <- read_tsv(params$gene_corr_file, show_col_types = FALSE)

# Some runs export raw p-values; others export only the derived decile scores.
# Pick the available measure for each distribution section.
hyper_measure_col <- if ("Pvalue" %in% names(pos_df)) {
  "Pvalue"
} else if ("variability_score" %in% names(pos_df)) {
  "variability_score"
} else {
  stop("position_scores.tsv must contain either Pvalue or variability_score")
}
boot_measure_col <- if ("pvalue_boot" %in% names(pos_df)) {
  "pvalue_boot"
} else if ("pattern_score" %in% names(pos_df)) {
  "pattern_score"
} else {
  stop("position_scores.tsv must contain either pvalue_boot or pattern_score")
}
hyper_measure_is_raw <- identical(hyper_measure_col, "Pvalue")
boot_measure_is_raw  <- identical(boot_measure_col,  "pvalue_boot")

# Provisional detection — overridden in the load-data block after directional scores are computed
has_rand <- "gene_rand_score" %in% names(gene_df) && any(!is.na(gene_df$gene_rand_score))
has_rer  <- "rer_min_pval"    %in% names(gene_df) && any(!is.na(gene_df$rer_min_pval)) &&
            "rer_rho"         %in% names(gene_df)

# ── VEP optional inputs ───────────────────────────────────────────────────────
.file_ok <- function(p) !is.null(p) && !is.na(p) && nchar(p) > 0 && file.exists(p)

has_scoring_stress <- .file_ok(params$stress_summary_file) &&
  .file_ok(params$stress_corr_file) &&
  .file_ok(params$stress_rank_file) &&
  .file_ok(params$stress_overlap_file) &&
  .file_ok(params$stress_variants_file)
has_vep_primateai <- .file_ok(params$vep_primateai_file)
has_aa2prot       <- .file_ok(params$vep_aa2prot_file)
has_genomic       <- .file_ok(params$genomic_info_file)
# FADE site data: use top direction if available, otherwise try bottom
has_fade_pos      <- .file_ok(params$fade_site_top_file) || .file_ok(params$fade_site_bot_file)
# TransVar is no longer used in the scoring report (removed — it added no actionable signal)
has_vep_transvar  <- FALSE

if (has_vep_primateai) {
  vep_pai_df <- read_tsv(params$vep_primateai_file, show_col_types = FALSE)
} else {
  vep_pai_df <- NULL
}

# aa2prot: tab-separated, no header — Gene | aln_pos (0-based) | caas_id | Gene:p.N
# Used to bridge alignment positions in pos_df to 1-based protein positions used by
# PrimateAI-3D lookup (map_to_primateai.py outputs transvar_query in this format).
if (has_aa2prot) {
  aa2prot_raw <- read_tsv(params$vep_aa2prot_file,
                          col_names = c("Gene", "aln_pos", "caas_id", "prot_query"),
                          show_col_types = FALSE)
  pos_prot_map <- aa2prot_raw %>%
    distinct(Gene, aln_pos, prot_query) %>%
    rename(Position = aln_pos)
} else {
  pos_prot_map <- NULL
}

if (has_genomic) {
  # Format: gene, chr, start, end, length  (no header assumed; detect if header present)
  genomic_raw <- read_tsv(params$genomic_info_file,
                          col_names = c("Gene", "chr", "gene_start", "gene_end", "gene_length"),
                          show_col_types = FALSE)
  # Drop header row if first cell is literally "gene" or "Gene"
  if (tolower(genomic_raw$Gene[1]) == "gene") genomic_raw <- genomic_raw[-1, ]
  genomic_raw <- genomic_raw %>%
    mutate(
      gene_start  = as.integer(gene_start),
      gene_end    = as.integer(gene_end),
      gene_length = as.integer(gene_length),
      gene_mid    = as.integer((gene_start + gene_end) / 2L)
    )
} else {
  genomic_raw <- NULL
}

accum_scheme_info <- tibble(
  scheme = c("us", "gs4", "gs3", "gs2", "gs1"),
  label = c("US", "GS4", "GS3", "GS2", "GS1"),
  weight = c(0.5, 0.2, 0.1, 0.1, 0.1),
  pval_col = paste0("accum_pval_", c("us", "gs4", "gs3", "gs2", "gs1")),
  score_col = paste0("accum_score_", c("us", "gs4", "gs3", "gs2", "gs1")),
  weighted_col = paste0("accum_weighted_", c("us", "gs4", "gs3", "gs2", "gs1"))
) %>%
  filter(pval_col %in% names(gene_df))

if (has_scoring_stress) {
  stress_summary_df <- read_tsv(params$stress_summary_file, show_col_types = FALSE)
  stress_corr_df <- read_tsv(params$stress_corr_file, show_col_types = FALSE)
  stress_rank_df <- read_tsv(params$stress_rank_file, show_col_types = FALSE)
  stress_overlap_df <- read_tsv(params$stress_overlap_file, show_col_types = FALSE)
  stress_variants_df <- read_tsv(params$stress_variants_file, show_col_types = FALSE)
  stress_loadings_df <- if (.file_ok(params$stress_loadings_file)) {
    read_tsv(params$stress_loadings_file, show_col_types = FALSE)
  } else {
    NULL
  }
} else {
  stress_summary_df <- NULL
  stress_corr_df <- NULL
  stress_rank_df <- NULL
  stress_overlap_df <- NULL
  stress_variants_df <- NULL
  stress_loadings_df <- NULL
}

# Thresholds
top_pct  <- params$top_pct
top5_pct <- params$top5_pct
top1_pct <- params$top1_pct

# ── RERConverge availability ────────────────────────────────────────────────
has_rer <- "rer_min_pval" %in% names(gene_df) &&
           any(!is.na(gene_df$rer_min_pval)) &&
           "rer_rho" %in% names(gene_df)

# ── Directional position subsets (change_side from discovery) ──────────────
pos_top    <- pos_df %>% filter(change_side %in% c("top",    "both"))
pos_bottom <- pos_df %>% filter(change_side %in% c("bottom", "both"))

# ── Directional gene CAAS scores ────────────────────────────────────────────
# 90th percentile of CAAS_score per gene, computed separately for TOP and BOTTOM
# positions. This mirrors how gene_caas_score is computed but direction-aware.
gene_caas_top_df <- pos_top %>%
  group_by(Gene) %>%
  summarise(gene_caas_top = quantile(CAAS_score, 0.9, na.rm = TRUE), .groups = "drop")

gene_caas_bot_df <- pos_bottom %>%
  group_by(Gene) %>%
  summarise(gene_caas_bot = quantile(CAAS_score, 0.9, na.rm = TRUE), .groups = "drop")

gene_df <- gene_df %>%
  left_join(gene_caas_top_df, by = "Gene") %>%
  left_join(gene_caas_bot_df, by = "Gene")

# ── Gene-level accumulation score ────────────────────────────────────────────
# Uses gene_rand_score from scoring_compute.R (1 - weighted mean of accum p-values).
# accum_score_* columns are not exported; gene_rand_score IS the composite.
if ("gene_rand_score" %in% names(gene_df) && any(!is.na(gene_df$gene_rand_score))) {
  gene_df <- gene_df %>% mutate(gene_accum_score = gene_rand_score)
  has_accum_score <- TRUE
} else {
  gene_df$gene_accum_score <- NA_real_
  has_accum_score <- FALSE
}
has_rand <- "gene_rand_score" %in% names(gene_df) && any(!is.na(gene_df$gene_rand_score))

# ── Gene-level direction assignment ─────────────────────────────────────────
# Assign direction to each gene from its top-scoring position
gene_direction_df <- pos_df %>%
  group_by(Gene) %>%
  slice_max(CAAS_score, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(Gene, gene_direction = change_side)

gene_df <- gene_df %>% left_join(gene_direction_df, by = "Gene")

gene_top_list <- gene_df %>%
  filter(!is.na(gene_caas_top)) %>%
  slice_max(gene_caas_top, prop = params$gene_top_pct) %>%
  pull(Gene)

gene_bot_list <- gene_df %>%
  filter(!is.na(gene_caas_bot)) %>%
  slice_max(gene_caas_bot, prop = params$gene_top_pct) %>%
  pull(Gene)

# ── Normalisation helper ──────────────────────────────────────────────────────
rank_norm <- function(x) {
  r <- rank(x, na.last = "keep", ties.method = "average")
  (r - 1) / (max(r, na.rm = TRUE) - 1)
}

# ── Venn diagram helpers (used across RER / FADE / Accumulation tabs) ─────────
.venn_counts <- function(set_a, set_b, label_a, label_b) {
  n_a  <- length(set_a); n_b <- length(set_b)
  n_ab <- length(intersect(set_a, set_b))
  tibble(label_a, label_b, n_a, n_b, n_ab,
         n_only_a = n_a - n_ab, n_only_b = n_b - n_ab)
}
.draw_venn <- function(vc, col_a, col_b, title) {
  ggplot() +
    geom_rect(aes(xmin=0.05, xmax=0.65, ymin=0.15, ymax=0.85),
              fill=col_a, alpha=0.35, colour=col_a, linewidth=1) +
    geom_rect(aes(xmin=0.35, xmax=0.95, ymin=0.15, ymax=0.85),
              fill=col_b, alpha=0.35, colour=col_b, linewidth=1) +
    annotate("text", x=0.20, y=0.5, label=vc$n_only_a, size=7, fontface="bold") +
    annotate("text", x=0.50, y=0.5, label=vc$n_ab,     size=7, fontface="bold", colour="grey20") +
    annotate("text", x=0.80, y=0.5, label=vc$n_only_b, size=7, fontface="bold") +
    annotate("text", x=0.20, y=0.12, label=vc$label_a, size=3.5, colour=col_a) +
    annotate("text", x=0.80, y=0.12, label=vc$label_b, size=3.5, colour=col_b) +
    annotate("text", x=0.50, y=0.93,
             label=sprintf("Shared: %d genes (%.0f%% of smaller set)",
                           vc$n_ab, 100 * vc$n_ab / max(min(vc$n_a, vc$n_b), 1)),
             size=3, colour="grey40") +
    ggtitle(title) + theme_void() +
    theme(plot.title = element_text(face="bold", hjust=0.5, size=11))
}
```

---

## Introduction

The **CAAS Scoring** module integrates signals from multiple upstream analyses
into a unified, interpretable metric for prioritising convergent amino acid
substitutions (CAAS).  The scoring operates at two levels:

1. **Position-level CAAS Score** — each alignment position receives a composite
   score that combines:
   - **Biochemical score** — amino acid variability and pattern association,
     weighted by grouping scheme specificity.
   - **ASR score** — ancestral state reconstruction confidence: novel changes
     score 1.0; contradicted conserved cases score 0; ASR-supported conserved
     cases receive pair-weighted posterior support capped below novelty.
   - **Convergence score** — ordinal change-type support with a mild
     completeness penalty when only one side is assessable.
   - **Parallel score** — ordinal parallelism support weighted by MRCA
     posterior confidence and penalised mildly when only one side is assessable.

2. **Gene-level scores** — each gene receives directional CAAS scores and is
   compared against independent evolutionary signals:
   - **gene_caas_top / gene_caas_bot** — 90th percentile of position-level CAAS
     scores for TOP and BOTTOM direction positions respectively.
   `r if (has_rand) "- **gene_accum_score** — weighted accumulation score (US=0.5, GS4=0.2, …)."`
   `r if (has_rer) "- RERConverge p-values and direction (accelerating / decelerating)."`
   - FADE gene-level BF ≥ 100 significance, tested separately per direction.

All comparisons use rank-normalised variables for correlations.

---

## Scoring Methodology

This section documents the exact weighting and combination logic used to
produce every score in this report, so results can be interpreted and
reproduced independently of the pipeline code.

<details>
<summary><strong>Click to expand full methodology</strong></summary>

---

### Position-Level Components

All position-level component scores are mapped to the **[0, 1]** interval.
The four core components are always computed; FADE is included only
when the upstream module has been run.

#### 1 · Variability score

Decile rank of the **hypergeometric p-value** (`Pvalue`) across all
non-GS0 Gene × Position × Scheme rows, inverted so that smaller p-values receive
higher scores.

$$\text{variability\_score} = 1 - \frac{\text{ntile}(P_{\text{hyp}}, 10) - 1}{9}$$

*Range: 0.0 (least significant decile) → 1.0 (most significant decile).*

> One of five independent components entering `pre_score`.

---

#### 2 · Pattern score

Decile rank of the **permulation p-value** (`pvalue_boot`), inverted in the
same way.

$$\text{pattern\_score} = 1 - \frac{\text{ntile}(P_{\text{boot}}, 10) - 1}{9}$$

> One of five independent components entering `pre_score`.

---

#### 3 · Biochemical weight

GS0 positions (any substitution, biochemically uninformative) are **excluded
entirely** before scoring. For each remaining position, the integer specificity
points of all satisfied schemes are summed:

| Scheme | Points |
|--------|-------:|
| US     | 5      |
| GS4    | 2      |
| GS3    | 1      |
| GS2    | 1      |
| GS1    | 1      |
| GS0    | *excluded* |

$$\text{biochem\_weight\_sum} = \sum_{s \in \text{detected schemes}} w_s$$

Maximum = **10** (US + GS4 + GS3 + GS2 + GS1 = 5 + 2 + 1 + 1 + 1).

> This is not a [0,1] score — it is a **multiplier** applied to `pre_score` at the final step. A position satisfying US only scores 5/10 = 0.5 of its signal potential; one satisfying all schemes scores 10/10 = 1.0.

---

#### 4 · ASR score

Measures ancestral-state-reconstruction confidence. Three mutually exclusive cases:

| Condition | Score |
|-----------|------:|
| `is_conserved_meta = FALSE` (novel change) | **1.0** |
| `is_conserved_meta = TRUE` **and** `asr_is_conserved = TRUE` | see formula below |
| `is_conserved_meta = TRUE` **and** `asr_is_conserved = FALSE` | **0.0** |

For ASR-confirmed conserved positions, the score is pair-weighted across the
detected foreground-pair configuration. Let $w = 1 / n_{\text{pairs}}$. Then:

$$
\text{asr\_score} =
\sum_{\text{divergent pairs}} w +
\sum_{\text{conserved pairs}} \min\left(
\frac{w \times P_{\text{MRCA}(i)} + w \times P_{\text{root}} \times I(\texttt{asr\_root\_conserved})}{2},
0.9w
\right)
$$

This keeps supported conserved explanations informative while ensuring they do
not outrank genuinely novel changes, which always score **1.0**.

---

#### 5 · Convergence score

Scores the **type and directionality** of the amino acid change on each
assessed side (top / bottom foreground lineages).

Side values:

| Change type | Value |
|-------------|------:|
| convergent  | 1.0   |
| codivergent | 0.5   |
| divergent   | 0.25  |
| ambiguous / no change | excluded (NA) |

$$\text{mean\_change} = \text{mean of non-NA side values}$$

$$\text{completeness penalty} = \begin{cases} 1.0 & \text{both sides assessed} \\ 0.75 & \text{one side only} \end{cases}$$

$$\text{convergence\_score} = \text{mean\_change} \times \text{completeness penalty}$$

Range: 0.1875 (one side, divergent: 0.25 × 0.75) → 1.0 (both sides convergent).

---

#### 6 · Parallel score

Scores whether the amino acid change is **parallel** (same substitution in
independent lineages) or **non-parallel**, weighted by MRCA posterior
confidence.

Parallelism coefficients (note: non-parallel receives the *highest* weight
because independent parallel changes are less informative as a signal of
convergence):

| Parallel type | Coefficient |
|---------------|------------:|
| non-parallel  | 1.0         |
| mixed         | 0.75        |
| parallel      | 0.5         |

For each assessable side (where `change_top`/`change_bottom` is
convergent / codivergent / divergent):

$$\text{side\_score} = \text{coeff} \times \overline{P_{\text{MRCA}}}$$

where $\overline{P_{\text{MRCA}}}$ is the mean of `mrca_1_posterior`,
`mrca_2_posterior`, `mrca_3_posterior` (NA excluded).

$$\text{mean\_parallel} = \text{mean of non-NA side scores}$$

$$\text{completeness penalty} = \begin{cases} 1.0 & \text{both sides assessed} \\ 0.75 & \text{one side only} \end{cases}$$

$$\text{parallel\_score} = \text{mean\_parallel} \times \text{completeness penalty}$$

---

#### Pre-score and Composite CAAS score

**`pre_score`** is the unweighted mean of the five independent signal components.
FADE is **not** included — it is used only as an independent gene-level
characterisation layer.

$$\text{pre\_score} = \frac{\text{variability} + \text{pattern} + \text{asr} + \text{convergence} + \text{parallel}}{5}$$

**`CAAS_score`** scales `pre_score` by biochemical specificity:

$$\text{CAAS\_score} = \text{pre\_score} \times \frac{\text{biochem\_weight\_sum}}{10}$$

This factorisation keeps *how strong is the signal* (`pre_score`) orthogonal
to *how biochemically specific is the convergence* (`biochem_weight_sum / 10`).
A GS0-only position is excluded entirely; a position satisfying only GS1 carries
10% of its signal potential regardless of how high its pre_score is.

---

### Gene-Level Scores

#### gene\_caas\_top / gene\_caas\_bot

90th percentile of `CAAS_score` computed separately for TOP-direction and
BOTTOM-direction positions within each gene:

$$\text{gene\_caas\_top} = Q_{0.90}(\text{CAAS\_score} \mid \text{change\_side} \in \{\text{top, both}\})$$
$$\text{gene\_caas\_bot} = Q_{0.90}(\text{CAAS\_score} \mid \text{change\_side} \in \{\text{bottom, both}\})$$

#### gene\_accum\_score  *(optional)*

Scheme-weighted sum of accumulation randomization p-values (GS0 excluded).
Weights are the normalized versions of the integer specificity points
(US = 0.5, GS4 = 0.2, GS3/2/1 = 0.1 each — proportional to `biochem_weight_sum`):

$$\text{gene\_accum\_score} = 1 - \frac{\sum_{s \in \{US, GS4, GS3, GS2, GS1\}} w_s \times P_{\text{accum},s}}{n_{\text{available schemes}}}$$

Maximum = 1.0 (all schemes p-value = 0).

#### RERConverge characterisation  *(optional)*

Raw permutation p-value (`p.perm`) used directly. Direction determined by sign
of `Rho`: positive = accelerating (high-phenotype lineages have elevated rate),
negative = decelerating. Significance threshold: p.perm ≤ 0.05.

#### FADE gene-level characterisation  *(optional)*

Sites are aggregated to gene level. A gene is FADE-significant if it contains
at least one position with `max_site_bf ≥ 100`, evaluated separately for
TOP and BOTTOM directions.

---

### ORA Gene List Extraction

Gene lists for Over-Representation Analysis are extracted at three thresholds
(configurable via `--scoring_position_top_pct` / `--scoring_position_top5_pct` /
`--scoring_position_top1_pct` and their gene-level equivalents) and three
directional subsets:

| Subset | Filter |
|--------|--------|
| **top** | Positions / genes where `change_side == "top"` |
| **bottom** | Positions / genes where `change_side == "bottom"` |
| **both** | All top-scoring positions / genes regardless of direction |

For **position-level** lists, the threshold is applied to `CAAS_score` and
the gene name is extracted from the qualifying positions.

For **gene-level** lists, the threshold is applied to `gene_caas_top` (TOP
direction) and `gene_caas_bot` (BOTTOM direction) independently.

</details>

---

## Input Distributions {.tabset}

The five independent pre-score components (variability, pattern, ASR, convergence, parallel)
are derived from the hypergeometric, permulation, and structural signals.
If the input table still contains raw p-values, those are shown directly; otherwise
the report falls back to the exported decile scores, which preserve the same ranking.

### `r if (hyper_measure_is_raw) "Hypergeometric P-value" else "Variability Score"`

```{r hyper-pval-dist, fig.height=8}
# Compute decile boundaries from the selected hypergeometric measure.
hyper_values <- pos_df[[hyper_measure_col]]
pv_breaks <- quantile(hyper_values, probs = seq(0, 1, by = 0.1), na.rm = TRUE)
decile_df_h <- pos_df %>%
  filter(!is.na(.data[[hyper_measure_col]])) %>%
  mutate(
    measure_value = .data[[hyper_measure_col]],
    plot_value    = if (hyper_measure_is_raw) -log10(pmax(measure_value, 1e-300)) else measure_value,
    decile_raw    = dplyr::ntile(measure_value, 10),
    # Keep decile direction aligned with score semantics:
    # decile 10 = strongest signal (smallest p-value or highest score).
    decile        = if (hyper_measure_is_raw) 11 - decile_raw else decile_raw,
    score_val     = (decile - 1) / 9
  )

# Panel 1: density on the selected measure scale with decile boundaries
p_hyp_density <- ggplot(decile_df_h, aes(x = plot_value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60,
                 fill = .tableau10[1], alpha = 0.6) +
  geom_density(linewidth = 0.8, colour = .tableau10[2]) +
  geom_vline(
    xintercept = if (hyper_measure_is_raw) -log10(pmax(pv_breaks[2:10], 1e-300)) else pv_breaks[2:10],
    linetype = "dotted", colour = "grey50", linewidth = 0.4
  ) +
  labs(
    title    = if (hyper_measure_is_raw) "Hypergeometric P-value Distribution (-log10 scale)" else "Variability Score Distribution",
    subtitle = if (hyper_measure_is_raw) {
      "Dotted lines: decile boundaries (D10 = smallest p-values, highest variability_score; D1 = largest p-values)"
    } else {
      "Dotted lines: decile boundaries of the exported variability_score values"
    },
    x        = if (hyper_measure_is_raw) "-log10(Pvalue)" else "variability_score",
    y        = "Density"
  ) +
  scale_x_continuous(
    breaks = scales::pretty_breaks(n = 14),
    minor_breaks = scales::pretty_breaks(n = 28)
  ) +
  theme(plot.title = element_text(face = "bold"))

# Panel 2: count per decile, coloured by variability_score
decile_counts_h <- decile_df_h %>%
  group_by(decile, score_val) %>%
  summarise(n = n(), .groups = "drop")

p_hyp_decile <- ggplot(decile_counts_h,
                        aes(x = factor(decile), y = n, fill = score_val)) +
  geom_col(colour = "white", linewidth = 0.3) +
  scale_fill_gradient(low = .tableau10[1], high = .tableau10[3],
                      name = "variability_score") +
  labs(
    title    = "Positions per Hypergeometric Decile",
    subtitle = if (hyper_measure_is_raw) {
      "Decile 10 = smallest p-values (variability_score = 1.0); Decile 1 = largest p-values"
    } else {
      "Decile 1 = lowest variability_score; Decile 10 = highest variability_score"
    },
    x        = "Decile (10 = most significant)",
    y        = "Number of rows"
  ) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "right")

gridExtra::grid.arrange(p_hyp_density, p_hyp_decile, ncol = 1)

# Decile boundary table: what p-value range does each decile cover?
if (hyper_measure_is_raw) {
  .hyp_lower <- pv_breaks[1:10]
  .hyp_upper <- c(pv_breaks[2:10], max(hyper_values, na.rm = TRUE))
  hyp_boundary_tbl <- tibble(
    Decile                = 10:1,
    `P-value lower bound` = formatC(.hyp_lower, format = "e", digits = 2),
    `P-value upper bound` = formatC(.hyp_upper, format = "e", digits = 2),
    variability_score     = round((10:1 - 1) / 9, 3)
  )
  knitr::kable(
    hyp_boundary_tbl,
    caption = "Hypergeometric p-value range per decile (decile 10 = smallest p-values, variability_score = 1.0)"
  )
}
```

### `r if (boot_measure_is_raw) "Permulation P-value" else "Pattern Score"`

```{r perm-pval-dist, fig.height=8}
boot_values <- pos_df[[boot_measure_col]]
pv_breaks_b <- quantile(boot_values, probs = seq(0, 1, by = 0.1), na.rm = TRUE)
decile_df_b <- pos_df %>%
  filter(!is.na(.data[[boot_measure_col]])) %>%
  mutate(
    measure_value = .data[[boot_measure_col]],
    plot_value    = if (boot_measure_is_raw) -log10(pmax(measure_value, 1e-300)) else measure_value,
    decile_raw    = dplyr::ntile(measure_value, 10),
    decile        = if (boot_measure_is_raw) 11 - decile_raw else decile_raw,
    score_val     = (decile - 1) / 9
  )

p_boot_density <- ggplot(decile_df_b, aes(x = plot_value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60,
                 fill = .tableau10[5], alpha = 0.6) +
  geom_density(linewidth = 0.8, colour = .tableau10[6]) +
  geom_vline(
    xintercept = if (boot_measure_is_raw) -log10(pmax(pv_breaks_b[2:10], 1e-300)) else pv_breaks_b[2:10],
    linetype = "dotted", colour = "grey50", linewidth = 0.4
  ) +
  labs(
    title    = if (boot_measure_is_raw) "Permulation Bootstrap P-value Distribution (-log10 scale)" else "Pattern Score Distribution",
    subtitle = if (boot_measure_is_raw) {
      "Dotted lines: decile boundaries (D10 = smallest p-values, highest pattern_score; D1 = largest p-values)"
    } else {
      "Dotted lines: decile boundaries of the exported pattern_score values"
    },
    x        = if (boot_measure_is_raw) "-log10(pvalue_boot)" else "pattern_score",
    y        = "Density"
  ) +
  scale_x_continuous(
    breaks = scales::pretty_breaks(n = 14),
    minor_breaks = scales::pretty_breaks(n = 28)
  ) +
  theme(plot.title = element_text(face = "bold"))

decile_counts_b <- decile_df_b %>%
  group_by(decile, score_val) %>%
  summarise(n = n(), .groups = "drop")

p_boot_decile <- ggplot(decile_counts_b,
                         aes(x = factor(decile), y = n, fill = score_val)) +
  geom_col(colour = "white", linewidth = 0.3) +
  scale_fill_gradient(low = .tableau10[7], high = .tableau10[5],
                      name = "pattern_score") +
  labs(
    title    = "Positions per Permulation Decile",
    subtitle = if (boot_measure_is_raw) {
      "Decile 10 = smallest permulation p-values (pattern_score = 1.0); Decile 1 = largest p-values"
    } else {
      "Decile 1 = lowest pattern_score; Decile 10 = highest pattern_score"
    },
    x        = "Decile (10 = most significant)",
    y        = "Number of rows"
  ) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "right")

gridExtra::grid.arrange(p_boot_density, p_boot_decile, ncol = 1)

# Decile boundary table for permulation p-value
if (boot_measure_is_raw) {
  .boot_lower <- pv_breaks_b[1:10]
  .boot_upper <- c(pv_breaks_b[2:10], max(boot_values, na.rm = TRUE))
  boot_boundary_tbl <- tibble(
    Decile                = 10:1,
    `P-value lower bound` = formatC(.boot_lower, format = "e", digits = 2),
    `P-value upper bound` = formatC(.boot_upper, format = "e", digits = 2),
    pattern_score         = round((10:1 - 1) / 9, 3)
  )
  knitr::kable(
    boot_boundary_tbl,
    caption = "Permulation p-value range per decile (decile 10 = smallest p-values, pattern_score = 1.0)"
  )
}
```

### Joint Decile Summary

```{r joint-decile-summary, fig.height=5}
# Scatter of variability_score vs pattern_score coloured by CAAS_score
joint_df <- pos_df %>%
  filter(!is.na(variability_score), !is.na(pattern_score)) %>%
  mutate(
    hyp_decile_raw  = dplyr::ntile(.data[[hyper_measure_col]], 10),
    boot_decile_raw = dplyr::ntile(.data[[boot_measure_col]], 10),
    hyp_decile      = if (hyper_measure_is_raw) 11 - hyp_decile_raw else hyp_decile_raw,
    boot_decile     = if (boot_measure_is_raw) 11 - boot_decile_raw else boot_decile_raw
  )

.r_hyp_caas  <- round(cor(joint_df$hyp_decile,  joint_df$CAAS_score, use = "complete.obs"), 2)
.r_boot_caas <- round(cor(joint_df$boot_decile, joint_df$CAAS_score, use = "complete.obs"), 2)
.r_hyp_boot  <- round(cor(joint_df$hyp_decile,  joint_df$boot_decile, use = "complete.obs"), 2)

ggplot(joint_df, aes(x = hyp_decile, y = boot_decile, colour = CAAS_score)) +
  geom_jitter(alpha = 0.45, size = 1.8, width = 0.35, height = 0.35) +
  scale_colour_gradientn(
    colours = c("#4E79A7", "#76B7B2", "#EDC948", "#F28E2B", "#E15759"),
    name    = "CAAS\nscore"
  ) +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(breaks = 1:10) +
  annotate("text", x = 1, y = 10,
           label = sprintf("r(hyp, CAAS) = %.2f\nr(perm, CAAS) = %.2f\nr(hyp, perm) = %.2f",
                           .r_hyp_caas, .r_boot_caas, .r_hyp_boot),
           hjust = 0, vjust = 1, size = 3.2, colour = "grey30") +
  labs(
    title    = "Hypergeometric vs Permulation Decile Assignment per Position",
    subtitle = "High CAAS scores (warm colours) cluster toward decile 10 × 10. Pearson r computed on decile ranks.",
    x        = "Hypergeometric decile (variability; 10 = most significant)",
    y        = "Permulation decile (pattern; 10 = most significant)"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

---

## Position-Level Scores {.tabset}

### CAAS Score Distribution

```{r caas-score-dist, fig.height=9}
# Global thresholds (used throughout the report for CAAS filters)
thr_10 <- quantile(pos_df$CAAS_score, 1 - top_pct,  na.rm = TRUE)
thr_5  <- quantile(pos_df$CAAS_score, 1 - top5_pct, na.rm = TRUE)
thr_1  <- quantile(pos_df$CAAS_score, 1 - top1_pct, na.rm = TRUE)

# Direction-specific thresholds (used for is_Xpct flags)
.thr_top_10 <- quantile(pos_top$CAAS_score, 1 - top_pct,  na.rm = TRUE)
.thr_top_5  <- quantile(pos_top$CAAS_score, 1 - top5_pct, na.rm = TRUE)
.thr_top_1  <- quantile(pos_top$CAAS_score, 1 - top1_pct, na.rm = TRUE)
.thr_bot_10 <- quantile(pos_bottom$CAAS_score, 1 - top_pct,  na.rm = TRUE)
.thr_bot_5  <- quantile(pos_bottom$CAAS_score, 1 - top5_pct, na.rm = TRUE)
.thr_bot_1  <- quantile(pos_bottom$CAAS_score, 1 - top1_pct, na.rm = TRUE)

p_combined <- ggplot(pos_df, aes(x = CAAS_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50,
                 fill = .tableau10[1], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[2]) +
  geom_vline(xintercept = thr_10, linetype = "dashed",
             colour = .tableau10[3], linewidth = 0.7) +
  geom_vline(xintercept = thr_5,  linetype = "longdash",
             colour = .tableau10[5], linewidth = 0.7) +
  geom_vline(xintercept = thr_1,  linetype = "dotted",
             colour = .tableau10[4], linewidth = 0.7) +
  annotate("text", x = thr_10, y = Inf, label = sprintf("Top %g%%", top_pct * 100),
           vjust = 2,    hjust = -0.1, colour = .tableau10[3], size = 3.5) +
  annotate("text", x = thr_5,  y = Inf, label = sprintf("Top %g%%", top5_pct * 100),
           vjust = 2.75, hjust = -0.1, colour = .tableau10[5], size = 3.5) +
  annotate("text", x = thr_1,  y = Inf, label = sprintf("Top %g%%", top1_pct * 100),
           vjust = 3.5,  hjust = -0.1, colour = .tableau10[4], size = 3.5) +
  labs(title = "CAAS Score Distribution — Combined",
       subtitle = sprintf("n = %s positions | Trait: %s",
                          format(nrow(pos_df), big.mark = ","), params$traitname),
       x = "CAAS Score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))

make_dir_dist <- function(df, dir_label, thr10, thr5, thr1, col) {
  ggplot(df, aes(x = CAAS_score)) +
    geom_histogram(aes(y = after_stat(density)), bins = 50,
                   fill = col, alpha = 0.7) +
    geom_density(linewidth = 0.8, colour = col) +
    geom_vline(xintercept = thr10, linetype = "dashed",  colour = "grey30", linewidth = 0.7) +
    geom_vline(xintercept = thr5,  linetype = "longdash", colour = "grey30", linewidth = 0.7) +
    geom_vline(xintercept = thr1,  linetype = "dotted",  colour = "grey30", linewidth = 0.7) +
    annotate("text", x = thr10, y = Inf,
             label = sprintf("Top %g%%\n(%.4f)", top_pct * 100,  thr10),
             vjust = 2,    hjust = -0.05, colour = "grey20", size = 3) +
    annotate("text", x = thr5,  y = Inf,
             label = sprintf("Top %g%%\n(%.4f)", top5_pct * 100, thr5),
             vjust = 3.8,  hjust = -0.05, colour = "grey20", size = 3) +
    annotate("text", x = thr1,  y = Inf,
             label = sprintf("Top %g%%\n(%.4f)", top1_pct * 100, thr1),
             vjust = 5.6,  hjust = -0.05, colour = "grey20", size = 3) +
    labs(title = sprintf("CAAS Score Distribution — %s positions", dir_label),
         subtitle = sprintf("n = %s | thresholds computed within direction",
                            format(nrow(df), big.mark = ",")),
         x = "CAAS Score", y = "Density") +
    theme(plot.title = element_text(face = "bold"))
}

p_top <- make_dir_dist(pos_top,    "TOP",    .thr_top_10, .thr_top_5, .thr_top_1, .tableau10[1])
p_bot <- make_dir_dist(pos_bottom, "BOTTOM", .thr_bot_10, .thr_bot_5, .thr_bot_1, .tableau10[3])

gridExtra::grid.arrange(p_combined, p_top, p_bot, ncol = 1)
```

### Pre-Score

```{r pre-score-dist, fig.height=7}
p_prescore <- ggplot(pos_df, aes(x = pre_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = .tableau10[1], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[2]) +
  labs(title = "Pre-Score Distribution",
       subtitle = "Unweighted mean of variability, pattern, ASR, convergence and parallel scores",
       x = "pre_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))

p_bwsum <- ggplot(pos_df, aes(x = factor(biochem_weight_sum))) +
  geom_bar(fill = .tableau10[4], alpha = 0.8) +
  labs(title = "Biochemical Weight Sum Distribution",
       subtitle = "Sum of integer scheme specificity points (max = 10 if US+GS4+GS3+GS2+GS1 all satisfied)",
       x = "biochem_weight_sum", y = "Count") +
  theme(plot.title = element_text(face = "bold"))

gridExtra::grid.arrange(p_prescore, p_bwsum, ncol = 1)
```

### ASR Score

```{r asr-dist, fig.height=4.5}
ggplot(pos_df, aes(x = asr_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = .tableau10[3], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[4]) +
  labs(title = "ASR Score Distribution",
       subtitle = "Novel = 1.0 | contradicted conserved = 0 | supported conserved = pair-weighted posterior support below novelty",
       x = "asr_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### Convergence Score

```{r convergence-dist, fig.height=4.5}
ggplot(pos_df, aes(x = convergence_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = .tableau10[5], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[6]) +
  labs(title = "Convergence Score Distribution",
       subtitle = "Ordinal change type (1 / 0.5 / 0.25) with a 1.0 / 0.75 completeness penalty",
       x = "convergence_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### Parallel Score

```{r parallel-dist, fig.height=4.5}
ggplot(pos_df, aes(x = parallel_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = .tableau10[7], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[8]) +
  labs(title = "Parallel Score Distribution",
       subtitle = "Non-parallel / mixed / parallel = 1 / 0.75 / 0.5, weighted by MRCA confidence and completeness",
       x = "parallel_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### Component Overview

```{r component-overview, fig.height=8}
# Faceted density comparison of all five independent pre-score components
component_cols <- c("variability_score", "pattern_score", "asr_score", "convergence_score", "parallel_score")

long_df <- pos_df %>%
  select(all_of(component_cols)) %>%
  pivot_longer(everything(), names_to = "Component", values_to = "Score") %>%
  filter(!is.na(Score)) %>%
  mutate(Component = factor(Component, levels = component_cols))

ggplot(long_df, aes(x = Score, fill = Component)) +
  geom_density(alpha = 0.6, linewidth = 0.3) +
  scale_fill_manual(values = .tableau10[seq_along(component_cols)]) +
  facet_wrap(~Component, scales = "free_y", ncol = 2) +
  labs(title = "Position-Level Component Score Distributions",
       x = "Score (0-1)", y = "Density") +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "none")
```

### Threshold Profiles

```{r threshold-profiles, fig.height=14, fig.width=12}
# Component score distributions for positions inside each ORA gene-list cutoff.
# Each panel shows overlaid densities of all position-level components so you
# can read off which signal drives positions into the top-N% × direction bucket.

component_cols_tp <- c("variability_score", "pattern_score", "asr_score", "convergence_score", "parallel_score")

thr_10_tp <- quantile(pos_df$CAAS_score, 1 - top_pct,  na.rm = TRUE)
thr_5_tp  <- quantile(pos_df$CAAS_score, 1 - top5_pct, na.rm = TRUE)
thr_1_tp  <- quantile(pos_df$CAAS_score, 1 - top1_pct, na.rm = TRUE)

make_threshold_panel <- function(df, threshold, pct_label, direction) {
  subset_df <- if (direction == "both") {
    df %>% filter(CAAS_score >= threshold)
  } else {
    df %>% filter(CAAS_score >= threshold, change_side == direction)
  }

  if (nrow(subset_df) == 0) {
    return(
      ggplot() +
        annotate("text", x = 0.5, y = 0.5, label = "No positions",
                 size = 5, colour = "grey50") +
        labs(title = sprintf("Top %s — %s", pct_label, direction),
             x = "Score (0–1)", y = "Density") +
        xlim(0, 1) + theme_minimal()
    )
  }

  long_df <- subset_df %>%
    select(all_of(component_cols_tp)) %>%
    pivot_longer(everything(), names_to = "Component", values_to = "Score") %>%
    filter(!is.na(Score)) %>%
    mutate(Component = factor(Component, levels = component_cols_tp))

  ggplot(long_df, aes(x = Score, fill = Component, colour = Component)) +
    geom_density(alpha = 0.35, linewidth = 0.6) +
    scale_fill_manual(values   = .tableau10[seq_along(component_cols_tp)]) +
    scale_colour_manual(values = .tableau10[seq_along(component_cols_tp)]) +
    scale_x_continuous(limits = c(0, 1), expand = c(0.01, 0)) +
    labs(
      title    = sprintf("Top %s — %s", pct_label, direction),
      subtitle = sprintf("n = %d positions", nrow(subset_df)),
      x = "Score (0–1)", y = "Density",
      fill = NULL, colour = NULL
    ) +
    theme(
      plot.title    = element_text(face = "bold", size = 11),
      plot.subtitle = element_text(size = 9, colour = "grey40"),
      legend.position = "bottom",
      legend.text     = element_text(size = 8)
    )
}

panels <- list()
for (direction in c("top", "bottom", "both")) {
  panels[[paste0("10pct_", direction)]] <-
    make_threshold_panel(pos_df, thr_10_tp, "10%", direction)
  panels[[paste0("5pct_",  direction)]] <-
    make_threshold_panel(pos_df, thr_5_tp,  "5%",  direction)
  panels[[paste0("1pct_",  direction)]] <-
    make_threshold_panel(pos_df, thr_1_tp,  "1%",  direction)
}

# Layout: rows = direction (top / bottom / both), cols = threshold (10% | 5% | 1%)
gridExtra::grid.arrange(
  panels[["10pct_top"]],    panels[["5pct_top"]],    panels[["1pct_top"]],
  panels[["10pct_bottom"]], panels[["5pct_bottom"]], panels[["1pct_bottom"]],
  panels[["10pct_both"]],   panels[["5pct_both"]],   panels[["1pct_both"]],
  ncol = 3,
  top  = grid::textGrob(
    "Component score profiles for top-scoring position subsets",
    gp = grid::gpar(fontface = "bold", fontsize = 13)
  ),
  left = grid::textGrob(
    "Direction:  top                          bottom                          both",
    rot = 90, gp = grid::gpar(fontsize = 9, col = "grey40")
  )
)
```

### Stress Test

`r if (!has_scoring_stress) "> **Note:** scoring stress-test outputs are not available for this run."`

```{r stress-summary-table, eval=has_scoring_stress}
DT::datatable(
  stress_summary_df %>%
    arrange(desc(spearman_to_current), mean_abs_rank_shift) %>%
    mutate(across(where(is.numeric), ~round(., 4))),
  rownames = FALSE,
  options = list(pageLength = 20, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "Assumption sensitivity summary for alternative position-level score variants"
  )
)
```

```{r stress-component-overview, eval=has_scoring_stress, fig.height=10}
# Five independent pre-score components — none unified before biochem weighting
stress_prescore_cols <- c("variability_score", "pattern_score", "asr_score",
                          "convergence_score", "parallel_score")
stress_components_long <- stress_variants_df %>%
  select(any_of(stress_prescore_cols)) %>%
  pivot_longer(everything(), names_to = "Component", values_to = "Score") %>%
  filter(!is.na(Score)) %>%
  mutate(Component = factor(Component, levels = stress_prescore_cols))

p_components <- ggplot(stress_components_long, aes(x = Score, fill = Component)) +
  geom_density(alpha = 0.55, linewidth = 0.4) +
  scale_fill_manual(values = .tableau10[seq_along(stress_prescore_cols)]) +
  facet_wrap(~Component, scales = "free_y", ncol = 2) +
  labs(
    title    = "Pre-Score Component Distributions",
    subtitle = "Five independent signals — none aggregated before the biochem multiplier is applied",
    x = "Score (0–1)", y = "Density"
  ) +
  theme(plot.title = element_text(face = "bold"), legend.position = "none")

# Biochem weight sum — discrete, shows how many/which schemes each position satisfies
p_bwsum_stress <- ggplot(stress_variants_df, aes(x = factor(biochem_weight_sum))) +
  geom_bar(fill = .tableau10[6], alpha = 0.8) +
  labs(
    title    = "Biochemical Weight Sum (multiplier pool)",
    subtitle = "Discrete: 1=GS1/2/3 only (×0.1) … 10=all schemes (×1.0). GS0-only positions excluded.",
    x = "biochem_weight_sum", y = "Count"
  ) +
  theme(plot.title = element_text(face = "bold"))

gridExtra::grid.arrange(p_components, p_bwsum_stress, ncol = 1, heights = c(3, 1))
```

```{r stress-correlation-heatmap, eval=has_scoring_stress, fig.height=12, fig.width=10}
stress_heatmap_df <- stress_corr_df %>%
  pivot_longer(c(pearson_r, spearman_r), names_to = "metric", values_to = "value") %>%
  mutate(metric = recode(metric, pearson_r = "Pearson", spearman_r = "Spearman"))

ggplot(stress_heatmap_df, aes(x = score_a, y = score_b, fill = value)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.2f", value)), size = 2.8) +
  scale_fill_gradient2(low = .tableau10[3], mid = "white", high = .tableau10[1], midpoint = 0, limits = c(-1, 1)) +
  facet_wrap(~metric, ncol = 1) +
  labs(
    title = "Position-Level Correlation Heatmap",
    subtitle = paste0("r = Pearson correlation; rho = Spearman rank correlation\n",
                      "Values near \u00b11 indicate strong linear/monotone association between scores"),
    x = NULL, y = NULL, fill = "r"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold")
  )
```

```{r stress-rank-heatmap, eval=has_scoring_stress, fig.height=10, fig.width=9}
ggplot(stress_rank_df, aes(x = score_a, y = score_b, fill = spearman_rank)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.2f", spearman_rank)), size = 2.8) +
  scale_fill_gradient2(low = .tableau10[3], mid = "white", high = .tableau10[1], midpoint = 0, limits = c(-1, 1)) +
  labs(
    title = "Composite Rank-Agreement Heatmap",
    subtitle = paste0("rho = Spearman rank correlation between composite score rankings\n",
                      "Values near 1 mean both scoring variants rank positions nearly identically"),
    x = NULL, y = NULL, fill = "rho"
  ) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r stress-top-overlap, eval=has_scoring_stress}
DT::datatable(
  stress_overlap_df %>%
    arrange(threshold, desc(jaccard)) %>%
    mutate(across(where(is.numeric), ~round(., 4))),
  rownames = FALSE,
  options = list(pageLength = 20, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "Top-hit overlap across alternative position-level composite variants"
  )
)
```

```{r stress-instability, eval=has_scoring_stress}
rank_cols <- grep("^rank_", names(stress_variants_df), value = TRUE)
instability_df <- if (length(rank_cols) > 0) {
  stress_variants_df %>%
    select(Gene, Position, CAAS_current, any_of(rank_cols), rank_spread) %>%
    arrange(desc(rank_spread), desc(CAAS_current)) %>%
    slice_head(n = 25)
} else {
  tibble(Gene = character(), Position = integer(), rank_spread = numeric())
}

DT::datatable(
  instability_df %>% mutate(across(where(is.numeric), ~round(., 4))),
  rownames = FALSE,
  options = list(pageLength = 25, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "Top positions by rank instability across stress-test variants"
  )
)
```

Latent models ask: *if we let the data decide the weights, which of the five
pre-score components drive the most shared variance?*  Two approaches are computed:

- **PCA_all_components** — First principal component of the five pre-score
  components (variability, pattern, ASR, convergence, parallel). The loadings
  show how much each score contributes to PC1. A positive loading means the
  component co-varies with the main source of variation across positions.
- **Factor1_all_components** — Exploratory factor analysis with a single latent
  factor. Unlike PCA, factor analysis separates shared variance (the factor) from
  unique variance per variable; loadings quantify each component's contribution to
  the shared signal.

All latent scores are converted to 0–1 decile scale for comparison. A high
correlation with `CAAS_current` means the latent model and the hand-engineered
composite agree on which positions are most interesting. Note that latent models
operate on `pre_score` components only — the biochem multiplier is intentionally
excluded because it reflects biochemical specificity rather than signal strength.

```{r stress-latent, eval=has_scoring_stress, fig.height=5}
if (!is.null(stress_loadings_df) && nrow(stress_loadings_df) > 0) {
  ggplot(stress_loadings_df, aes(x = reorder(variable, loading), y = loading, fill = model)) +
    geom_col(position = "dodge") +
    coord_flip() +
    labs(
      title    = "Latent-Model Component Loadings",
      subtitle = paste0("How much each score component drives each latent model's first dimension\n",
                        "Positive = co-varies with PC1/Factor1; compare models to spot dominant components"),
      x = NULL, y = "Loading"
    ) +
    theme(plot.title    = element_text(face = "bold"),
          plot.subtitle = element_text(size = 9, colour = "grey40"))
} else {
  plot.new()
  text(0.5, 0.5, "No latent-model loadings available for this run")
}
```

`r if (has_scoring_stress) '> **How to read the stress test:**\n> \n> | Output | What to look for | What it means |\n> |--------|-----------------|---------------|\n> | **Pre-score components** | All five distributions span 0–1 without severe collapse | Checks that each signal dimension has real variance; a collapsed component adds little to pre_score |\n> | **Biochem weight sum** | Distribution shape across discrete values 1–10 | Most positions should not sit at the extremes; a spike at 1 suggests the trait has few US/GS4 hits |\n> | **Correlation heatmap** | r / rho close to 1 between a variant and CAAS_current | Dropping that component barely changes position rankings |\n> | **Rank-agreement heatmap** | rho close to 1 | The two variants order positions the same way |\n> | **Top-overlap table** | High Jaccard at top-10% / top-5% thresholds | The same genes appear in gene lists regardless of which component is dropped |\n> | **Instability table** | High rank_spread | These positions are assumption-sensitive; examine their five component scores manually |\n> | **Latent models** | PC1 / Factor1 loadings; correlation with CAAS_current | Which pre-score components drive shared variance; if latent models agree with CAAS_current the equal-weight assumption is data-consistent |\n> \n> **Score variant naming:** <code>CAAS_unweighted</code> = pre_score without the biochem multiplier (isolates the effect of biochem scaling); <code>CAAS_no_*</code> — drops one pre-score component and recomputes pre_score × biochem_weight_sum/10; <code>PCA_all_components</code> / <code>Factor1_all_components</code> — data-driven weighting of the five pre-score components.\n> A robust run shows consistent rankings across all variants.'`

### Top Positions

```{r top-positions-table}
display_cols <- c("Gene", "Position", "CAAS_score", "pre_score", "biochem_weight_sum",
                  "variability_score", "pattern_score",
                  "asr_score", "convergence_score", "parallel_score", "change_side",
                  "Pvalue", "pvalue_boot",
                  "is_top_10pct", "is_top_5pct", "is_top_1pct",
                  "is_bot_10pct", "is_bot_5pct", "is_bot_1pct")

DT::datatable(
  pos_df %>%
    arrange(desc(CAAS_score)) %>%
    select(any_of(display_cols)) %>%
    mutate(across(where(is.numeric), ~round(., 4))),
  filter    = "top",
  rownames  = FALSE,
  options   = list(pageLength = 25, scrollX = TRUE),
  caption   = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Position-level CAAS scores | Trait: %s", params$traitname)
  )
)
```

### CAAS Score Manhattan (Position Level)

`r if (!has_genomic) "> **Note:** Genomic coordinates not available — position-level Manhattan skipped."`

```{r pos-caas-manhattan-setup, eval=has_genomic}
# Join all scored positions with genomic coordinates via their gene.
# Each position inherits its gene's genomic midpoint as the x-coordinate.
# We build the chr offsets independently here (rather than relying on the
# genomic-setup chunk, which runs later in the document).
pos_genomic <- pos_df %>%
  inner_join(genomic_raw, by = "Gene") %>%
  mutate(
    chr_label = sub("^chr", "", chr),
    chr_num   = suppressWarnings(as.integer(chr_label)),
    chr_order = dplyr::case_when(
      !is.na(chr_num)              ~ chr_num,
      chr_label == "X"             ~ 23L,
      chr_label == "Y"             ~ 24L,
      chr_label %in% c("MT","M")   ~ 25L,
      TRUE                         ~ 26L
    )
  ) %>%
  arrange(chr_order, gene_mid)

pos_chr_offsets <- pos_genomic %>%
  group_by(chr_order, chr_label) %>%
  summarise(chr_max = max(gene_end, na.rm = TRUE), .groups = "drop") %>%
  arrange(chr_order) %>%
  mutate(offset = cumsum(lag(chr_max, default = 0L)) + (chr_order - 1L) * 5e6L)

pos_genomic <- pos_genomic %>%
  left_join(pos_chr_offsets %>% select(chr_order, offset), by = "chr_order") %>%
  mutate(
    manhattan_x = gene_mid + offset,
    chr_even    = chr_order %% 2 == 0
  )

pos_chr_label_positions <- pos_chr_offsets %>%
  left_join(
    pos_genomic %>% group_by(chr_order) %>%
      summarise(mid_x = mean(manhattan_x, na.rm = TRUE), .groups = "drop"),
    by = "chr_order"
  )
```

```{r pos-caas-manhattan-plot, eval=has_genomic, fig.height=5, fig.width=14}
pos_genomic <- pos_genomic %>%
  mutate(caas_log = -log10(1 - pmin(CAAS_score, 0.9999)))
pos_thr_log <- -log10(1 - pmin(thr_10, 0.9999))

ggplot(pos_genomic, aes(x = manhattan_x, y = caas_log, colour = chr_even)) +
  geom_point(alpha = 0.45, size = 0.6, show.legend = FALSE) +
  geom_hline(yintercept = pos_thr_log, linetype = "dashed",
             colour = "grey40", linewidth = 0.6) +
  scale_colour_manual(
    values = c("FALSE" = .tableau10[1], "TRUE" = .tableau10[2]),
    guide = "none"
  ) +
  scale_x_continuous(
    breaks = pos_chr_label_positions$mid_x,
    labels = pos_chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  labs(
    title    = "Position-Level CAAS Score — Manhattan View",
    subtitle = sprintf("%d positions with genomic coordinates | dashed = top %g%% threshold",
                       nrow(pos_genomic), params$top_pct * 100),
    x = "Chromosome", y = "-log10(1 - CAAS Score)"
  ) +
  theme(
    plot.title         = element_text(face = "bold"),
    axis.text.x        = element_text(size = 7),
    panel.grid.major.x = element_blank()
  )
```

---

## Direction Score Profiles {.tabset}

Each tab shows distribution plots for TOP-direction and BOTTOM-direction positions
side by side, allowing direct comparison of how each scoring component behaves
across the two phenotypic directions.

### P-Value Distributions

```{r dir-pval-dist, fig.height=9}
make_dir_pval_panel <- function(df, measure_col, is_raw, dir_label, col) {
  vals   <- df[[measure_col]]
  breaks <- quantile(vals, probs = seq(0, 1, 0.1), na.rm = TRUE)
  plot_x <- if (is_raw) -log10(pmax(vals, 1e-300)) else vals
  xlab   <- if (is_raw) sprintf("-log10(%s)", measure_col) else measure_col
  title  <- sprintf("%s — %s", ifelse(is_raw, toupper(measure_col), measure_col), dir_label)
  df2 <- data.frame(x = plot_x)
  ggplot(df2, aes(x = x)) +
    geom_histogram(aes(y = after_stat(density)), bins = 50, fill = col, alpha = 0.65) +
    geom_density(linewidth = 0.8, colour = col) +
    geom_vline(
      xintercept = if (is_raw) -log10(pmax(breaks[2:10], 1e-300)) else breaks[2:10],
      linetype = "dotted", colour = "grey50", linewidth = 0.3
    ) +
    labs(title = title,
         subtitle = sprintf("n = %s positions | dotted = decile boundaries",
                            format(nrow(df), big.mark = ",")),
         x = xlab, y = "Density") +
    theme(plot.title = element_text(face = "bold", size = 10))
}

p_hyp_top <- make_dir_pval_panel(pos_top,    hyper_measure_col, hyper_measure_is_raw, "TOP",    .tableau10[1])
p_hyp_bot <- make_dir_pval_panel(pos_bottom, hyper_measure_col, hyper_measure_is_raw, "BOTTOM", .tableau10[3])
p_per_top <- make_dir_pval_panel(pos_top,    boot_measure_col,  boot_measure_is_raw,  "TOP",    .tableau10[5])
p_per_bot <- make_dir_pval_panel(pos_bottom, boot_measure_col,  boot_measure_is_raw,  "BOTTOM", .tableau10[7])

gridExtra::grid.arrange(
  p_hyp_top, p_hyp_bot,
  p_per_top, p_per_bot,
  ncol = 2,
  top  = grid::textGrob(
    "Hypergeometric (rows 1) and Permulation (row 2) p-value distributions by direction",
    gp = grid::gpar(fontface = "bold", fontsize = 12)
  )
)
```

### Pre-Score

```{r dir-prescore-dist, fig.height=5}
make_dir_component_plot <- function(df, col, dir_label, colour, title_suffix = "") {
  ggplot(df, aes(x = .data[[col]])) +
    geom_histogram(aes(y = after_stat(density)), bins = 40, fill = colour, alpha = 0.65) +
    geom_density(linewidth = 0.8, colour = colour) +
    labs(title    = sprintf("%s — %s%s", col, dir_label, title_suffix),
         subtitle = sprintf("n = %s positions", format(nrow(df), big.mark = ",")),
         x = col, y = "Density") +
    theme(plot.title = element_text(face = "bold", size = 10))
}

gridExtra::grid.arrange(
  make_dir_component_plot(pos_top,    "pre_score", "TOP",    .tableau10[1]),
  make_dir_component_plot(pos_bottom, "pre_score", "BOTTOM", .tableau10[3]),
  ncol = 2
)
```

### ASR Score

```{r dir-asr-dist, fig.height=5}
gridExtra::grid.arrange(
  make_dir_component_plot(pos_top,    "asr_score", "TOP",    .tableau10[1]),
  make_dir_component_plot(pos_bottom, "asr_score", "BOTTOM", .tableau10[3]),
  ncol = 2
)
```

### Convergence Score

```{r dir-convergence-dist, fig.height=5}
gridExtra::grid.arrange(
  make_dir_component_plot(pos_top,    "convergence_score", "TOP",    .tableau10[1]),
  make_dir_component_plot(pos_bottom, "convergence_score", "BOTTOM", .tableau10[3]),
  ncol = 2
)
```

### Parallel Score

```{r dir-parallel-dist, fig.height=5}
gridExtra::grid.arrange(
  make_dir_component_plot(pos_top,    "parallel_score", "TOP",    .tableau10[1]),
  make_dir_component_plot(pos_bottom, "parallel_score", "BOTTOM", .tableau10[3]),
  ncol = 2
)
```

### Top Positions by Direction

```{r dir-top-positions-table}
has_caas_col <- "caas" %in% names(pos_df)

dir_display_cols <- c("Gene", "Position", "CAAS_score", "pre_score", "biochem_weight_sum",
                      if (has_caas_col) "caas",
                      "variability_score", "pattern_score",
                      "asr_score", "convergence_score", "parallel_score",
                      "Pvalue", "pvalue_boot",
                      "change_side",
                      "is_top_10pct", "is_top_5pct", "is_top_1pct",
                      "is_bot_10pct", "is_bot_5pct", "is_bot_1pct")

dir_pos_tbl <- bind_rows(
  pos_top    %>% mutate(.direction = "TOP"),
  pos_bottom %>% mutate(.direction = "BOTTOM")
) %>%
  arrange(.direction, desc(CAAS_score)) %>%
  select(Direction = .direction, any_of(dir_display_cols)) %>%
  mutate(across(where(is.numeric), ~round(., 4)))

DT::datatable(
  dir_pos_tbl,
  filter   = "top",
  rownames = FALSE,
  options  = list(pageLength = 25, scrollX = TRUE),
  caption  = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Top-scoring positions by direction | Trait: %s", params$traitname)
  )
)
```

---

## Position Characterisation {.tabset #position-characterisation}

This section examines how ranked positions behave in relation to external
signals that do **not** contribute to the `CAAS_score`. PrimateAI-3D is treated
as an independent characterisation layer: its signal is averaged across the
detected biochemical stances (GS0 excluded), producing a continuous 0–1
characterisation score.

- **`pai_char_score`** = mean(max(PAI3D\_score per CAAP group)) per Gene ×
  Position — captures the average pathogenicity of the CAAS-specific
  substitution across biochemical stances.

This score ranges from 0 (no support in any stance) to ≤ 1.0 (strong support
across stances). It is **not** included in `CAAS_score`.

> **Note:** PrimateAI-3D scores require an access request to Illumina —
> see illumina.com/science/genomics-research/articles/primateai-3d.html

### VEP: PrimateAI by CAAS Decile

`r if (!has_vep_primateai) "> **Note:** PrimateAI-3D scores are not available for this run."`

```{r vep-pai-setup, eval=has_vep_primateai}
# Detect PrimateAI score column (varies by database version)
.pai_score_col <- intersect(
  c("score_PAI3D", "score", "primateai_score", "pathogenicity_score", "PrimateAI_score"),
  names(vep_pai_df)
)[1]
if (is.na(.pai_score_col)) {
  warning("PrimateAI: no recognised score column found — skipping.")
  has_vep_primateai <- FALSE
}
```

```{r vep-pai-join, eval=has_vep_primateai}
pos_score_meta <- pos_df %>%
  select(Gene, Position, CAAS_score, pre_score, biochem_weight_sum, change_side) %>%
  mutate(
    caas_decile  = dplyr::ntile(CAAS_score, 10),
    in_top_10pct = CAAS_score >= thr_10,
    in_top_5pct  = CAAS_score >= thr_5,
    in_top_1pct  = CAAS_score >= thr_1
  )

# pai_char_score = mean(max(PAI3D per CAAP group)) per protein position.
# Uses equal contribution from each detected biochemical stance (GS0 excluded).
# map_to_primateai.py outputs caap_group and scheme_weight columns when available.
has_pai_schemes <- all(c("caap_group", "scheme_weight") %in% names(vep_pai_df))

if (has_pai_schemes) {
  # Two-step: max per (query × CAAP group) to collapse multiple alt AAs per stance,
  # then average across CAAP groups (scheme_weight > 0 only).
  pai_by_prot <- vep_pai_df %>%
    mutate(
      pai_score     = as.numeric(.data[[.pai_score_col]]),
      scheme_weight = as.numeric(scheme_weight)
    ) %>%
    filter(is.finite(pai_score), is.finite(scheme_weight), scheme_weight > 0) %>%
    group_by(transvar_query, caap_group) %>%
    summarise(max_pai = max(pai_score, na.rm = TRUE), .groups = "drop") %>%
    group_by(transvar_query) %>%
    summarise(
      pai_char_score = mean(max_pai, na.rm = TRUE),
      n_pai_schemes  = n(),
      .groups = "drop"
    )
} else {
  # Legacy fallback: no scheme information — average available PAI scores.
  pai_by_prot <- vep_pai_df %>%
    mutate(pai_score = as.numeric(.data[[.pai_score_col]])) %>%
    filter(is.finite(pai_score)) %>%
    group_by(transvar_query) %>%
    summarise(
      pai_char_score = mean(pai_score, na.rm = TRUE),
      n_pai_schemes  = 1L,
      .groups = "drop"
    )
}

if (!is.null(pos_prot_map)) {
  pos_vep_pai <- pos_prot_map %>%
    filter(!is.na(prot_query)) %>%
    inner_join(pai_by_prot, by = c("prot_query" = "transvar_query")) %>%
    inner_join(pos_score_meta, by = c("Gene", "Position")) %>%
    distinct(Gene, Position, .keep_all = TRUE)
} else {
  # fallback: parse GENE:p.N from transvar_query — will mostly miss without aa2prot
  pos_vep_pai <- pai_by_prot %>%
    mutate(
      Gene     = sub(":p\\..*", "", transvar_query),
      Position = as.integer(sub(".*:p\\.", "", transvar_query))
    ) %>%
    select(Gene, Position, pai_char_score, n_pai_schemes) %>%
    group_by(Gene, Position) %>%
    summarise(
      pai_char_score = mean(pai_char_score, na.rm = TRUE),
      n_pai_schemes  = max(n_pai_schemes),
      .groups = "drop"
    ) %>%
    inner_join(pos_score_meta, by = c("Gene", "Position"))
}

has_vep_primateai <- nrow(pos_vep_pai) > 0

# Directional PAI subsets: top+both and bottom+both
pos_vep_pai_top    <- pos_vep_pai %>% filter(change_side %in% c("top",    "both"))
pos_vep_pai_bottom <- pos_vep_pai %>% filter(change_side %in% c("bottom", "both"))
```

`r if (exists("pos_vep_pai") && nrow(pos_vep_pai) == 0) "> **Note:** PrimateAI-3D inputs were provided, but no scored positions could be matched to the VEP/PAI3D annotations for this direction."`

```{r vep-pai-decile-table, eval=has_vep_primateai}
pai_boundary_tbl <- decile_boundary_tbl(
  pos_vep_pai$pai_char_score,
  score_name = "PAI3D characterisation decile score",
  lower_better = FALSE,
  value_label = "PAI3D characterisation",
  digits = 3
)

if (!is.null(pai_boundary_tbl)) {
  knitr::kable(
    pai_boundary_tbl,
    caption = "PrimateAI-3D mean characterisation score range per decile"
  )
}
```

```{r vep-pai-violin, eval=has_vep_primateai, fig.height=7}
# pai_char_score violin per CAAS decile, split by direction (TOP and BOTTOM separately).
# pos_vep_pai_top/bottom created in vep-pai-join; each shows positions where
# change_side is "top"/"both" or "bottom"/"both" respectively.
pai_dir_df <- bind_rows(
  pos_vep_pai_top    %>% mutate(pai_direction = "TOP (top + both)"),
  pos_vep_pai_bottom %>% mutate(pai_direction = "BOTTOM (bottom + both)")
) %>%
  filter(!is.na(caas_decile), !is.na(pai_char_score))

# Per-direction median for horizontal reference lines
pai_medians <- pai_dir_df %>%
  group_by(pai_direction) %>%
  summarise(median_pai = median(pai_char_score, na.rm = TRUE), .groups = "drop")

ggplot(pai_dir_df, aes(x = factor(caas_decile), y = pai_char_score)) +
  geom_violin(aes(fill = pai_direction), alpha = 0.55, trim = TRUE, scale = "width",
              show.legend = FALSE) +
  geom_boxplot(width = 0.14, outlier.size = 0.4, alpha = 0.9,
               fill = "white", colour = "grey30") +
  geom_hline(data = pai_medians,
             aes(yintercept = median_pai),
             linetype = "dashed", colour = "grey20", linewidth = 0.7) +
  geom_text(data = pai_medians,
            aes(x = 10.6, y = median_pai,
                label = sprintf("median = %.3f", median_pai)),
            hjust = 0, vjust = -0.4, size = 3, colour = "grey20",
            inherit.aes = FALSE) +
  scale_fill_manual(values = c("TOP (top + both)" = .tableau10[1],
                               "BOTTOM (bottom + both)" = .tableau10[3])) +
  scale_x_discrete(labels = c("1\n(low)", as.character(2:9), "10\n(high)")) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.08))) +
  facet_wrap(~pai_direction, ncol = 1) +
  labs(
    title    = "PAI_TOP and PAI_BOTTOM: PrimateAI-3D Score by CAAS Decile",
    subtitle = sprintf(
      "PAI_TOP: %d positions | PAI_BOTTOM: %d positions | score = mean(max(PAI3D per CAAP group)) | dashed = per-direction median",
      nrow(pos_vep_pai_top), nrow(pos_vep_pai_bottom)
    ),
    x = "CAAS Score Decile", y = "pai_char_score"
  ) +
  theme(plot.title = element_text(face = "bold"),
        strip.text = element_text(face = "bold", size = 11))
```

```{r vep-pai-scatter, eval=has_vep_primateai, fig.height=5}
# CAAS_score vs pai_char_score, coloured by direction.
# Normalise both axes for Pearson r (rank / n).
pos_vep_pai_dir <- bind_rows(
  pos_vep_pai_top    %>% mutate(pai_direction = "TOP"),
  pos_vep_pai_bottom %>% mutate(pai_direction = "BOTTOM")
) %>% distinct(Gene, Position, .keep_all = TRUE)

.pai_caas_r_top  <- safe_cor(
  rank(pos_vep_pai_top$CAAS_score,    na.last = "keep") / sum(!is.na(pos_vep_pai_top$CAAS_score)),
  rank(pos_vep_pai_top$pai_char_score, na.last = "keep") / sum(!is.na(pos_vep_pai_top$pai_char_score))
)
.pai_caas_r_bot  <- safe_cor(
  rank(pos_vep_pai_bottom$CAAS_score,    na.last = "keep") / sum(!is.na(pos_vep_pai_bottom$CAAS_score)),
  rank(pos_vep_pai_bottom$pai_char_score, na.last = "keep") / sum(!is.na(pos_vep_pai_bottom$pai_char_score))
)

ggplot(pos_vep_pai_dir, aes(x = CAAS_score, y = pai_char_score,
                             colour = pai_direction)) +
  geom_point(alpha = 0.35, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8, alpha = 0.15) +
  geom_vline(xintercept = thr_10, linetype = "dashed",
             colour = "grey40", linewidth = 0.5) +
  scale_colour_manual(values = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[3]),
                      name = "Direction") +
  labs(
    title    = "CAAS Score vs PrimateAI-3D Score by Direction",
    subtitle = sprintf("Pearson r (rank-norm): TOP = %.3f | BOTTOM = %.3f | dashed = top-10%% CAAS",
                       .pai_caas_r_top, .pai_caas_r_bot),
    x = "CAAS Score", y = "pai_char_score"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r vep-pai-prescore-scatter, eval=has_vep_primateai, fig.height=5}
.pai_prescore_r_top <- safe_cor(
  rank(pos_vep_pai_top$pre_score,       na.last = "keep") / sum(!is.na(pos_vep_pai_top$pre_score)),
  rank(pos_vep_pai_top$pai_char_score,  na.last = "keep") / sum(!is.na(pos_vep_pai_top$pai_char_score))
)
.pai_prescore_r_bot <- safe_cor(
  rank(pos_vep_pai_bottom$pre_score,       na.last = "keep") / sum(!is.na(pos_vep_pai_bottom$pre_score)),
  rank(pos_vep_pai_bottom$pai_char_score,  na.last = "keep") / sum(!is.na(pos_vep_pai_bottom$pai_char_score))
)

ggplot(pos_vep_pai_dir, aes(x = pre_score, y = pai_char_score,
                             colour = pai_direction)) +
  geom_point(alpha = 0.35, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8, alpha = 0.15) +
  scale_colour_manual(values = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[3]),
                      name = "Direction") +
  labs(
    title    = "Pre-Score vs PrimateAI-3D Score by Direction",
    subtitle = sprintf("Pearson r (rank-norm): TOP = %.3f | BOTTOM = %.3f",
                       .pai_prescore_r_top, .pai_prescore_r_bot),
    x = "pre_score", y = "pai_char_score"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r vep-pai-fisher, eval=has_vep_primateai}
# Helper: run PAI Fisher enrichment for one direction.
# Rows are produced for each CAAS threshold; "100%" and "50%" rows cover
# all positions and the top-50% positions respectively, providing a baseline.
run_pai_fisher_dir <- function(df, label) {
  .med <- median(df$pai_char_score, na.rm = TRUE)
  df   <- df %>% mutate(high_pai = !is.na(pai_char_score) & pai_char_score > .med)
  n_all <- nrow(df)

  # Build named threshold list: CAAS-flag-based thresholds + percentage-based rows
  # Percentage rows use in_top_Xpct flags when available; for 100% and 50% we
  # derive a logical vector from the CAAS score quantile.
  thr_50_val <- quantile(df$CAAS_score, 0.50, na.rm = TRUE)
  pct_rows <- list(
    list(col = NULL, lbl = "100% (all positions)",    flag = rep(TRUE, n_all)),
    list(col = NULL, lbl = "50% (top 50% CAAS)",      flag = df$CAAS_score >= thr_50_val),
    list(col = "in_top_10pct", lbl = sprintf("Top %g%%", top_pct  * 100), flag = NULL),
    list(col = "in_top_5pct",  lbl = sprintf("Top %g%%", top5_pct * 100), flag = NULL),
    list(col = "in_top_1pct",  lbl = sprintf("Top %g%%", top1_pct * 100), flag = NULL)
  )

  purrr::map_dfr(pct_rows, function(thr) {
    in_top   <- if (!is.null(thr$flag)) thr$flag else df[[thr$col]]
    in_top   <- replace_na(in_top, FALSE)
    n_top    <- sum(in_top)
    n_top_h  <- sum(in_top & df$high_pai, na.rm = TRUE)
    n_rest   <- sum(!in_top)
    n_rest_h <- sum(!in_top & df$high_pai, na.rm = TRUE)
    mat <- matrix(c(n_top - n_top_h, n_rest - n_rest_h, n_top_h, n_rest_h), nrow = 2)
    ft  <- tryCatch(fisher.test(mat),
                    error = function(e) list(estimate = NA, conf.int = c(NA, NA), p.value = NA))
    tibble(
      Direction            = label,
      Threshold            = thr$lbl,
      `N positions`        = n_top,
      `N high PAI (>med)`  = n_top_h,
      `% high PAI`         = sprintf("%.1f%%", 100 * n_top_h / max(n_top, 1)),
      `Odds ratio`         = sprintf("%.2f", ft$estimate),
      `P value`            = formatC(ft$p.value, format = "e", digits = 2)
    )
  })
}

pai_fisher <- bind_rows(
  run_pai_fisher_dir(pos_vep_pai_top,    "TOP"),
  run_pai_fisher_dir(pos_vep_pai_bottom, "BOTTOM")
)

knitr::kable(
  pai_fisher,
  caption = paste0(
    "PAI3D enrichment of high-PAI positions (> per-direction median) by CAAS threshold and direction. ",
    "100% row = all directional positions (baseline); 50% row = top half by CAAS score; ",
    "Fisher OR and P value compare top-subset vs remainder."
  ),
  align = c("l", "l", rep("r", ncol(pai_fisher) - 2))
)
```

### Regional Accumulation

`r if (!has_genomic) "> **Note:** Genomic coordinates not available — regional position accumulation skipped."`

```{r position-region-accum, eval=has_genomic, fig.height=10, fig.width=14}
# Build directional window summaries — TOP and BOTTOM use direction-specific thresholds
.build_region_wins <- function(pos_sub, thr10, thr5, thr1, chr_offsets_df, dir_label) {
  pos_sub %>%
    mutate(
      win_id    = floor(gene_mid / params$window_size_bp),
      win_start = win_id * params$window_size_bp,
      win_end   = win_start + params$window_size_bp - 1L,
      win_mid   = win_start + params$window_size_bp / 2L,
      in_top10  = CAAS_score >= thr10,
      in_top5   = CAAS_score >= thr5,
      in_top1   = CAAS_score >= thr1
    ) %>%
    group_by(chr_order, chr_label, win_id, win_start, win_end, win_mid) %>%
    summarise(
      n_positions        = n(),
      n_top10_positions  = sum(in_top10, na.rm = TRUE),
      n_top5_positions   = sum(in_top5,  na.rm = TRUE),
      n_top1_positions   = sum(in_top1,  na.rm = TRUE),
      frac_top_positions = n_top10_positions / n_positions,
      mean_caas          = mean(CAAS_score, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    left_join(chr_offsets_df %>% select(chr_order, offset), by = "chr_order") %>%
    mutate(
      direction    = dir_label,
      win_x        = win_mid + offset,
      region_label = sprintf("chr%s:%0.1f-%0.1f Mb",
                             chr_label, win_start / 1e6, win_end / 1e6)
    )
}

# pos_genomic already has chr_order and gene_mid; split by direction
pos_genomic_top <- pos_genomic %>% filter(change_side %in% c("top",    "both"))
pos_genomic_bot <- pos_genomic %>% filter(change_side %in% c("bottom", "both"))

pos_region_top <- .build_region_wins(pos_genomic_top, .thr_top_10, .thr_top_5, .thr_top_1,
                                      pos_chr_offsets, "TOP")
pos_region_bot <- .build_region_wins(pos_genomic_bot, .thr_bot_10, .thr_bot_5, .thr_bot_1,
                                      pos_chr_offsets, "BOTTOM")

# Combined (global thresholds) for overview panel
pos_region_wins <- pos_genomic %>%
  mutate(
    win_id    = floor(gene_mid / params$window_size_bp),
    win_start = win_id * params$window_size_bp,
    win_end   = win_start + params$window_size_bp - 1L,
    win_mid   = win_start + params$window_size_bp / 2L,
    in_top10  = CAAS_score >= thr_10,
    in_top5   = CAAS_score >= thr_5,
    in_top1   = CAAS_score >= thr_1
  ) %>%
  group_by(chr_order, chr_label, win_id, win_start, win_end, win_mid) %>%
  summarise(
    n_positions        = n(),
    n_top10_positions  = sum(in_top10, na.rm = TRUE),
    n_top5_positions   = sum(in_top5,  na.rm = TRUE),
    n_top1_positions   = sum(in_top1,  na.rm = TRUE),
    frac_top_positions = n_top10_positions / n_positions,
    mean_caas          = mean(CAAS_score, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(pos_chr_offsets %>% select(chr_order, offset), by = "chr_order") %>%
  mutate(
    win_x        = win_mid + offset,
    region_label = sprintf("chr%s:%0.1f-%0.1f Mb",
                           chr_label, win_start / 1e6, win_end / 1e6)
  )

.make_accum_plot <- function(df, title_str, subtitle_str) {
  ggplot(df, aes(x = win_x, y = frac_top_positions)) +
    geom_col(aes(fill = mean_caas), width = params$window_size_bp * 0.9, alpha = 0.9) +
    geom_point(aes(size = n_top10_positions), colour = "black", alpha = 0.8) +
    scale_fill_viridis_c(name = "Mean\nCAAS", option = "magma") +
    scale_size_continuous(name = "Top-10%\npositions", range = c(3, 14)) +
    scale_x_continuous(
      breaks = pos_chr_label_positions$mid_x,
      labels = pos_chr_label_positions$chr_label,
      expand = expansion(mult = 0.01)
    ) +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    labs(title = title_str, subtitle = subtitle_str,
         x = "Chromosome", y = "Fraction in top 10%") +
    theme(plot.title = element_text(face = "bold"),
          axis.text.x = element_text(size = 7),
          panel.grid.major.x = element_blank())
}

p_accum_all <- .make_accum_plot(pos_region_wins,
  "Regional Accumulation — All positions (global thresholds)",
  "Bars: fraction of top-10% CAAS | Dot size: count of top-10% positions")
p_accum_top <- .make_accum_plot(pos_region_top,
  "Regional Accumulation — TOP direction (direction-specific thresholds)",
  sprintf("TOP positions only | top-10%% threshold = %.4f", .thr_top_10))
p_accum_bot <- .make_accum_plot(pos_region_bot,
  "Regional Accumulation — BOTTOM direction (direction-specific thresholds)",
  sprintf("BOTTOM positions only | top-10%% threshold = %.4f", .thr_bot_10))

gridExtra::grid.arrange(p_accum_all, p_accum_top, p_accum_bot, ncol = 1)
```

```{r position-region-accum-table, eval=has_genomic}
.make_region_tbl <- function(wins_df) {
  wins_df %>%
    filter(n_top10_positions > 0) %>%
    arrange(desc(frac_top_positions), desc(n_top10_positions)) %>%
    slice_head(n = 30) %>%
    mutate(
      frac_top10 = sprintf("%.1f%%", 100 * n_top10_positions / n_positions),
      frac_top5  = sprintf("%.1f%%", 100 * n_top5_positions  / n_positions),
      frac_top1  = sprintf("%.1f%%", 100 * n_top1_positions  / n_positions)
    ) %>%
    select(
      Region       = region_label,
      `N positions` = n_positions,
      `N top-10%`  = n_top10_positions,
      `% top-10%`  = frac_top10,
      `N top-5%`   = n_top5_positions,
      `% top-5%`   = frac_top5,
      `N top-1%`   = n_top1_positions,
      `% top-1%`   = frac_top1,
      `Mean CAAS`  = mean_caas
    ) %>%
    mutate(`Mean CAAS` = round(`Mean CAAS`, 4))
}

knitr::kable(.make_region_tbl(pos_region_top),
  caption = "TOP-direction: top 30 genomic windows by fraction of high-CAAS positions (direction-specific thresholds)")
knitr::kable(.make_region_tbl(pos_region_bot),
  caption = "BOTTOM-direction: top 30 genomic windows by fraction of high-CAAS positions (direction-specific thresholds)")

# Export directional regional accumulation tables as TSV
region_export <- bind_rows(
  .make_region_tbl(pos_region_top) %>% mutate(Direction = "TOP"),
  .make_region_tbl(pos_region_bot) %>% mutate(Direction = "BOTTOM")
) %>% select(Direction, everything())
readr::write_tsv(region_export,
  file.path(params$output_dir, "regional_accumulation.tsv"))
```


---

## FADE Position Characterisation {.tabset}

> **FADE** (FUBAR Approach to Directional Evolution) tests whether foreground branches show
> elevated substitution bias toward a particular amino acid at each site.
> Top-direction FADE characterises positions where the phenotypic group shows accelerated
> amino acid evolution; bottom-direction shows the opposite.
> CAAS positions (BF ≥ 100) overlap reveals sites that are both convergent and under directional selection.

`r if (!has_fade_pos) "> **Note:** FADE per-site data are not available for this run (fade_site_top_file and fade_site_bot_file not provided)."`

```{r fade-pos-setup, eval=has_fade_pos}
# Load FADE site tables for each direction separately.
# FADE_report.Rmd emits site = seq_along(...) making site numbers 1-based, while
# CAAS/CT uses 0-based positions (alimport.py range(0, length), caas_id.py as-is).
# Subtract 1 here to align coordinate systems before joining.
.load_fade_sites <- function(path) {
  if (!.file_ok(path)) return(NULL)
  read_tsv(path, show_col_types = FALSE) %>%
    rename(Gene = gene, Position = site) %>%
    mutate(Position = Position - 1L)
}
fade_top_df <- .load_fade_sites(params$fade_site_top_file)
fade_bot_df <- .load_fade_sites(params$fade_site_bot_file)

# Join FADE BFs to directional CAAS position subsets
.join_fade <- function(pos_subset, fade_df) {
  if (is.null(fade_df) || nrow(pos_subset) == 0) return(NULL)
  pos_subset %>%
    left_join(
      fade_df %>% select(Gene, Position, max_site_bf, top_target_aa, fade_biased),
      by = c("Gene", "Position")
    ) %>%
    mutate(
      fade_biased = replace_na(fade_biased, FALSE),
      fade_sig    = !is.na(max_site_bf) & max_site_bf >= 100,
      caas_decile = dplyr::ntile(CAAS_score, 10),
      norm_caas   = rank(CAAS_score,   ties.method = "average", na.last = "keep") /
                    sum(!is.na(CAAS_score)),
      norm_log_bf = {
        lbf <- log10(pmax(max_site_bf, 0.01))
        rank(lbf, ties.method = "average", na.last = "keep") / sum(!is.na(lbf))
      }
    )
}
pos_fade_top <- .join_fade(pos_top,    fade_top_df)
pos_fade_bot <- .join_fade(pos_bottom, fade_bot_df)

has_fade_top <- !is.null(pos_fade_top) && any(pos_fade_top$fade_sig, na.rm = TRUE)
has_fade_bot <- !is.null(pos_fade_bot) && any(pos_fade_bot$fade_sig, na.rm = TRUE)
has_fade_any <- has_fade_top || has_fade_bot

# Helper: draw 2-set Venn diagram using base ggplot2 (no extra packages).
# Circle labels are placed outside and above each circle to avoid overlapping
# with the count annotations inside.
.draw_venn2 <- function(n_only_A, n_only_B, n_AB, n_universe,
                         label_A, label_B, title_str) {
  theta <- seq(0, 2 * pi, length.out = 300)
  r <- 1.7; dx <- 1.05
  ciA <- data.frame(x = -dx + r * cos(theta), y = r * sin(theta))
  ciB <- data.frame(x =  dx + r * cos(theta), y = r * sin(theta))
  ggplot() +
    geom_polygon(data = ciA, aes(x, y), fill = .tableau10[1], alpha = 0.22, colour = NA) +
    geom_polygon(data = ciB, aes(x, y), fill = .tableau10[3], alpha = 0.22, colour = NA) +
    geom_path(data = ciA, aes(x, y), colour = .tableau10[1], linewidth = 1.4) +
    geom_path(data = ciB, aes(x, y), colour = .tableau10[3], linewidth = 1.4) +
    annotate("text", x = -2.6, y = 0,
             label = paste0(label_A, " only\n", scales::comma(n_only_A)),
             size = 4.2, colour = .tableau10[1], fontface = "bold") +
    annotate("text", x = 0, y = 0,
             label = paste0("Both\n", scales::comma(n_AB)),
             size = 4.2, colour = "grey20", fontface = "bold") +
    annotate("text", x = 2.6, y = 0,
             label = paste0(label_B, " only\n", scales::comma(n_only_B)),
             size = 4.2, colour = .tableau10[3], fontface = "bold") +
    # Circle titles: pushed to x = -3.5 and +3.5 (outside the overlap zone)
    annotate("text", x = -3.5, y = 2.8, label = label_A, size = 4.5,
             colour = .tableau10[1], fontface = "bold", hjust = 0.5) +
    annotate("text", x =  3.5, y = 2.8, label = label_B, size = 4.5,
             colour = .tableau10[3], fontface = "bold", hjust = 0.5) +
    annotate("text", x = 0, y = -2.6,
             label = sprintf("Universe: %s positions", scales::comma(n_universe)),
             size = 3.5, colour = "grey50") +
    coord_fixed(xlim = c(-5.5, 5.5), ylim = c(-3.2, 3.5)) +
    labs(title = title_str) +
    theme_void() +
    theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 12))
}

# Helper: compute Venn counts (CAAS dir positions vs full FADE sig set)
.fade_venn_counts <- function(pos_dir, fade_df) {
  if (is.null(pos_dir) || is.null(fade_df)) return(NULL)
  caas_keys     <- pos_dir  %>% select(Gene, Position) %>% distinct()
  fade_sig_keys <- fade_df  %>%
    filter(!is.na(max_site_bf), max_site_bf >= 100) %>%
    select(Gene, Position) %>% distinct()
  n_A   <- nrow(caas_keys)
  n_B   <- nrow(fade_sig_keys)
  n_AB  <- nrow(inner_join(caas_keys, fade_sig_keys, by = c("Gene", "Position")))
  list(n_only_A = n_A - n_AB, n_only_B = n_B - n_AB, n_AB = n_AB,
       n_A = n_A, n_B = n_B, n_universe = n_A + max(n_B - n_AB, 0))
}

vc_top <- .fade_venn_counts(pos_top,    fade_top_df)
vc_bot <- .fade_venn_counts(pos_bottom, fade_bot_df)
```

### Venn Diagrams

```{r fade-venn-plots, eval=has_fade_any, fig.height=5, fig.width=14}
venn_plots <- list()
if (!is.null(vc_top)) {
  venn_plots[["top"]] <- .draw_venn2(
    vc_top$n_only_A, vc_top$n_only_B, vc_top$n_AB, vc_top$n_universe,
    "CAAS (top)", "FADE BF\u2265100 (top)",
    "TOP direction: CAAS vs FADE-significant positions"
  )
}
if (!is.null(vc_bot)) {
  venn_plots[["bot"]] <- .draw_venn2(
    vc_bot$n_only_A, vc_bot$n_only_B, vc_bot$n_AB, vc_bot$n_universe,
    "CAAS (bottom)", "FADE BF\u2265100 (bottom)",
    "BOTTOM direction: CAAS vs FADE-significant positions"
  )
}
if (length(venn_plots) == 2) {
  gridExtra::grid.arrange(grobs = venn_plots, ncol = 2)
} else if (length(venn_plots) == 1) {
  print(venn_plots[[1]])
}
```

```{r fade-venn-overlap-table, eval=has_fade_any}
# Table of positions in the intersection (CAAS + FADE BF>=100), by direction
.venn_overlap_positions <- function(pos_dir, fade_df, dir_label) {
  if (is.null(pos_dir) || is.null(fade_df)) return(NULL)
  caas_sel <- if ("caas" %in% names(pos_dir))
    c("Gene", "Position", "CAAS_score", "caas", "change_side") else
    c("Gene", "Position", "CAAS_score", "change_side")
  inner_join(
    pos_dir  %>% select(any_of(caas_sel)),
    fade_df  %>% filter(!is.na(max_site_bf), max_site_bf >= 100) %>%
                 select(Gene, Position, max_site_bf, top_target_aa),
    by = c("Gene", "Position")
  ) %>%
    mutate(Direction = dir_label) %>%
    arrange(desc(max_site_bf))
}

overlap_pos_tbl <- bind_rows(
  .venn_overlap_positions(pos_top,    fade_top_df, "TOP"),
  .venn_overlap_positions(pos_bottom, fade_bot_df, "BOTTOM")
)

if (nrow(overlap_pos_tbl) > 0) {
  DT::datatable(
    overlap_pos_tbl %>% mutate(across(where(is.numeric), ~round(., 4))),
    filter   = "top",
    rownames = FALSE,
    options  = list(pageLength = 25, scrollX = TRUE),
    caption  = htmltools::tags$caption(
      style = "caption-side: top; text-align: left; font-weight: bold;",
      "Positions in the CAAS \u2229 FADE (BF \u2265 100) intersection, by direction"
    )
  )
} else {
  cat("> No positions found in the CAAS \u2229 FADE intersection.")
}
```

### Overlap Table

```{r fade-overlap-table, eval=has_fade_any}
.fade_overlap_row <- function(pos_fade_dir, dir_label, pct_val, pct_label) {
  if (is.null(pos_fade_dir)) return(NULL)
  thr  <- quantile(pos_fade_dir$CAAS_score, 1 - pct_val, na.rm = TRUE)
  n_caas <- sum(pos_fade_dir$CAAS_score >= thr, na.rm = TRUE)
  n_sig  <- sum(pos_fade_dir$fade_sig, na.rm = TRUE)
  n_both <- sum(pos_fade_dir$CAAS_score >= thr & pos_fade_dir$fade_sig, na.rm = TRUE)
  tibble(
    Direction           = dir_label,
    `CAAS threshold`    = pct_label,
    `N CAAS positions`  = n_caas,
    `N FADE BF >= 100`  = n_sig,
    `Overlap`           = n_both,
    `% CAAS in FADE`    = sprintf("%.1f%%", 100 * n_both / max(n_caas, 1)),
    `% FADE in CAAS`    = sprintf("%.1f%%", 100 * n_both / max(n_sig,  1))
  )
}

pct_list <- list(
  list(val = params$top_pct,  lbl = sprintf("Top %g%%", params$top_pct  * 100)),
  list(val = params$top5_pct, lbl = sprintf("Top %g%%", params$top5_pct * 100)),
  list(val = params$top1_pct, lbl = sprintf("Top %g%%", params$top1_pct * 100))
)

fade_overlap_tbl <- bind_rows(
  purrr::map_dfr(pct_list, ~.fade_overlap_row(pos_fade_top, "TOP",    .x$val, .x$lbl)),
  purrr::map_dfr(pct_list, ~.fade_overlap_row(pos_fade_bot, "BOTTOM", .x$val, .x$lbl))
)

knitr::kable(
  fade_overlap_tbl,
  caption = "Overlap between top-CAAS positions and FADE-significant positions (BF \u2265 100), by direction and CAAS threshold"
)
```

### CAAS vs FADE Correlation

```{r fade-caas-corr, eval=has_fade_any, fig.height=5}
# Combine top and bottom joined datasets; plot CAAS_score vs log10(BF)
# with rank-normalised axes so Pearson r is not dominated by BF outliers.
.fade_corr_df <- bind_rows(
  if (!is.null(pos_fade_top)) pos_fade_top %>% mutate(fade_direction = "TOP")    else NULL,
  if (!is.null(pos_fade_bot)) pos_fade_bot %>% mutate(fade_direction = "BOTTOM") else NULL
) %>%
  filter(!is.na(max_site_bf)) %>%
  mutate(log10_bf = log10(pmax(max_site_bf, 0.01)))

.corr_label <- .fade_corr_df %>%
  group_by(fade_direction) %>%
  summarise(
    r_pearson  = cor(norm_caas, norm_log_bf, use = "complete.obs"),
    r_spearman = cor(CAAS_score, log10_bf, method = "spearman", use = "complete.obs"),
    .groups = "drop"
  ) %>%
  mutate(label = sprintf("%s: r=%.3f (rank-norm Pearson) / \u03c1=%.3f (Spearman)",
                         fade_direction, r_pearson, r_spearman))

ggplot(.fade_corr_df, aes(x = CAAS_score, y = log10_bf, colour = fade_direction)) +
  geom_point(alpha = 0.35, size = 0.9) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8, alpha = 0.12) +
  geom_hline(yintercept = 2, linetype = "dashed", colour = "grey50", linewidth = 0.5) +
  scale_colour_manual(values = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[3]),
                      name = "Direction") +
  annotate("text", x = -Inf, y = 2.05, hjust = -0.05, vjust = 0,
           label = "BF = 100", size = 3, colour = "grey40") +
  labs(
    title    = "CAAS Score vs FADE max Bayes Factor",
    subtitle = paste(.corr_label$label, collapse = "\n"),
    x = "CAAS Score", y = "log10(max site BF)"
  ) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(size = 9, colour = "grey30"))
```

---

## Gene-Level Characterisation {.tabset}

Directional gene sets (TOP and BOTTOM) are characterised against independent
evolutionary signals. All correlations use rank-normalised variables.

### Gene CAAS Score

```{r gene-caas-dir-dist, fig.height=5}
# Directional thresholds
g_top_thr_10 <- quantile(gene_df$gene_caas_top, 1 - params$gene_top_pct,  na.rm = TRUE)
g_top_thr_5  <- quantile(gene_df$gene_caas_top, 1 - params$gene_top5_pct, na.rm = TRUE)
g_top_thr_1  <- quantile(gene_df$gene_caas_top, 1 - params$gene_top1_pct, na.rm = TRUE)
g_bot_thr_10 <- quantile(gene_df$gene_caas_bot, 1 - params$gene_top_pct,  na.rm = TRUE)
g_bot_thr_5  <- quantile(gene_df$gene_caas_bot, 1 - params$gene_top5_pct, na.rm = TRUE)
g_bot_thr_1  <- quantile(gene_df$gene_caas_bot, 1 - params$gene_top1_pct, na.rm = TRUE)
g_thr_10 <- quantile(gene_df$gene_caas_score, 1 - params$gene_top_pct, na.rm = TRUE)

long_dir_scores <- bind_rows(
  gene_df %>% filter(!is.na(gene_caas_top)) %>%
    transmute(Gene, score = gene_caas_top, direction = "TOP"),
  gene_df %>% filter(!is.na(gene_caas_bot)) %>%
    transmute(Gene, score = gene_caas_bot, direction = "BOTTOM")
)

ggplot(long_dir_scores, aes(x = score, fill = direction, colour = direction)) +
  geom_density(alpha = 0.45, linewidth = 0.7) +
  geom_vline(xintercept = g_top_thr_10, linetype = "dashed",
             colour = .tableau10[1], linewidth = 0.6) +
  geom_vline(xintercept = g_bot_thr_10, linetype = "dashed",
             colour = .tableau10[2], linewidth = 0.6) +
  scale_fill_manual(values   = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[2])) +
  scale_colour_manual(values = c("TOP" = .tableau10[1], "BOTTOM" = .tableau10[2])) +
  labs(title = "Directional Gene CAAS Score Distribution",
       subtitle = sprintf("90th pctile of directional positions | dashed = top %g%% threshold per direction",
                          params$gene_top_pct * 100),
       x = "Gene CAAS Score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### RERConverge

`r if (!has_rer) "> **Note:** RERConverge results not available for this run."`

```{r gene-rer-venn, eval=has_rer, fig.height=5, fig.width=12}
rer_accel_sig <- gene_df %>%
  filter(!is.na(rer_min_pval), rer_min_pval <= 0.05,
         !is.na(rer_rho), rer_rho > 0) %>% pull(Gene)
rer_decel_sig <- gene_df %>%
  filter(!is.na(rer_min_pval), rer_min_pval <= 0.05,
         !is.na(rer_rho), rer_rho < 0) %>% pull(Gene)

vc_rer_top <- .venn_counts(gene_top_list, rer_accel_sig, "CAAS TOP", "RER Accel.")
vc_rer_bot <- .venn_counts(gene_bot_list, rer_decel_sig, "CAAS BOT", "RER Decel.")
gridExtra::grid.arrange(
  .draw_venn(vc_rer_top, .tableau10[1], .tableau10[3], "TOP genes vs RER Accelerating"),
  .draw_venn(vc_rer_bot, .tableau10[2], .tableau10[4], "BOTTOM genes vs RER Decelerating"),
  ncol = 2,
  top = grid::textGrob("RERConverge Gene Overlap (CAAS top 10% vs RER p \u2264 0.05)",
                        gp = grid::gpar(fontface="bold", fontsize=13)))
```

```{r gene-rer-table, eval=has_rer}
rer_tbl <- gene_df %>%
  filter(!is.na(rer_min_pval)) %>%
  mutate(
    rer_direction = case_when(rer_rho > 0 ~ "accelerating",
                               rer_rho < 0 ~ "decelerating",
                               TRUE        ~ "neutral"),
    in_top_list = Gene %in% gene_top_list,
    in_bot_list = Gene %in% gene_bot_list
  ) %>%
  arrange(rer_min_pval) %>%
  select(Gene, rer_min_pval, rer_rho, rer_direction,
         gene_caas_score, gene_caas_top, gene_caas_bot,
         in_top_list, in_bot_list) %>%
  mutate(across(where(is.numeric), ~round(., 4)))

DT::datatable(
  rer_tbl, filter="top", rownames=FALSE,
  options = list(pageLength=20, scrollX=TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "RERConverge results ranked by p-value — direction (rho sign) and CAAS gene list membership"
  )
)
```

```{r gene-rer-correlation, eval=has_rer, fig.height=5}
rer_cor_df <- gene_df %>%
  filter(!is.na(rer_min_pval), !is.na(gene_caas_score)) %>%
  mutate(
    x_norm  = rank_norm(gene_caas_score),
    y_norm  = rank_norm(-log10(pmax(rer_min_pval, 1e-300))),
    rer_dir = case_when(rer_rho > 0 ~ "accelerating",
                         rer_rho < 0 ~ "decelerating", TRUE ~ "neutral")
  )
ggplot(rer_cor_df, aes(x=x_norm, y=y_norm, colour=rer_dir)) +
  geom_point(alpha=0.5, size=1.2) +
  geom_smooth(method="lm", se=TRUE, colour="grey30", linewidth=0.7) +
  ggrepel::geom_text_repel(
    data = rer_cor_df %>%
      filter(y_norm > quantile(y_norm, 0.95, na.rm=TRUE) |
             x_norm > quantile(x_norm, 0.95, na.rm=TRUE)),
    aes(label=Gene), size=2.5, max.overlaps=12, show.legend=FALSE
  ) +
  scale_colour_manual(
    values = c("accelerating"=.tableau10[1], "decelerating"=.tableau10[2], "neutral"="grey60"),
    name = "RER direction") +
  labs(title = "Gene CAAS Score vs RERConverge Signal (rank-normalised)",
       subtitle = sprintf("r = %.3f | x = rank_norm(gene_caas_score), y = rank_norm(-log10(p.perm))",
                          safe_cor(rer_cor_df$x_norm, rer_cor_df$y_norm)),
       x = "Gene CAAS Score (rank-norm)", y = "RER Signal (rank-norm)") +
  theme(plot.title = element_text(face="bold"))
```

### FADE Gene

`r if (!has_fade_pos) "> **Note:** FADE site data not available — gene-level FADE analysis skipped."`

```{r gene-fade-setup, eval=has_fade_pos}
if (!is.null(fade_top_df) && nrow(fade_top_df) > 0) {
  gene_fade_top_agg <- fade_top_df %>%
    group_by(Gene) %>%
    summarise(fade_top_sig    = any(max_site_bf >= 100, na.rm=TRUE),
              fade_top_max_bf = max(max_site_bf, na.rm=TRUE), .groups="drop")
} else {
  gene_fade_top_agg <- tibble(Gene=character(), fade_top_sig=logical(), fade_top_max_bf=numeric())
}
if (!is.null(fade_bot_df) && nrow(fade_bot_df) > 0) {
  gene_fade_bot_agg <- fade_bot_df %>%
    group_by(Gene) %>%
    summarise(fade_bot_sig    = any(max_site_bf >= 100, na.rm=TRUE),
              fade_bot_max_bf = max(max_site_bf, na.rm=TRUE), .groups="drop")
} else {
  gene_fade_bot_agg <- tibble(Gene=character(), fade_bot_sig=logical(), fade_bot_max_bf=numeric())
}
gene_fade_df <- gene_df %>%
  select(Gene, gene_caas_score, gene_caas_top, gene_caas_bot, n_positions) %>%
  left_join(gene_fade_top_agg, by="Gene") %>%
  left_join(gene_fade_bot_agg, by="Gene") %>%
  mutate(fade_top_sig = replace_na(fade_top_sig, FALSE),
         fade_bot_sig = replace_na(fade_bot_sig, FALSE))
has_gene_fade_any <- any(gene_fade_df$fade_top_sig) || any(gene_fade_df$fade_bot_sig)
```

```{r gene-fade-venn, eval=has_fade_pos && exists("has_gene_fade_any") && has_gene_fade_any, fig.height=5, fig.width=12}
fade_top_genes <- gene_fade_df %>% filter(fade_top_sig) %>% pull(Gene)
fade_bot_genes <- gene_fade_df %>% filter(fade_bot_sig) %>% pull(Gene)
vc_fade_top <- .venn_counts(gene_top_list, fade_top_genes, "CAAS TOP", "FADE TOP sig")
vc_fade_bot <- .venn_counts(gene_bot_list, fade_bot_genes, "CAAS BOT", "FADE BOT sig")
gridExtra::grid.arrange(
  .draw_venn(vc_fade_top, .tableau10[1], .tableau10[5], "TOP genes vs FADE TOP (BF\u2265100)"),
  .draw_venn(vc_fade_bot, .tableau10[2], .tableau10[6], "BOTTOM genes vs FADE BOT (BF\u2265100)"),
  ncol = 2,
  top = grid::textGrob("FADE Gene Overlap (CAAS top 10% vs FADE BF \u2265 100)",
                        gp = grid::gpar(fontface="bold", fontsize=13)))
```

```{r gene-fade-correlation, eval=has_fade_pos && exists("has_gene_fade_any") && has_gene_fade_any, fig.height=5, fig.width=12}
fade_cor_top <- gene_fade_df %>%
  filter(!is.na(gene_caas_top), !is.na(fade_top_max_bf), is.finite(fade_top_max_bf)) %>%
  mutate(x_norm=rank_norm(gene_caas_top), y_norm=rank_norm(fade_top_max_bf))
fade_cor_bot <- gene_fade_df %>%
  filter(!is.na(gene_caas_bot), !is.na(fade_bot_max_bf), is.finite(fade_bot_max_bf)) %>%
  mutate(x_norm=rank_norm(gene_caas_bot), y_norm=rank_norm(fade_bot_max_bf))

.fade_scatter <- function(df, col, title, xlab, ylab) {
  ggplot(df, aes(x=x_norm, y=y_norm)) +
    geom_point(alpha=0.5, size=1.2, colour=col) +
    geom_smooth(method="lm", se=TRUE, colour="grey30", linewidth=0.7) +
    ggrepel::geom_text_repel(
      data = df %>% filter(y_norm > quantile(y_norm, 0.95, na.rm=TRUE)),
      aes(label=Gene), size=2.5, max.overlaps=10, show.legend=FALSE
    ) +
    labs(title=title,
         subtitle=sprintf("r = %.3f", safe_cor(df$x_norm, df$y_norm)),
         x=xlab, y=ylab) +
    theme(plot.title = element_text(face="bold", size=11))
}
gridExtra::grid.arrange(
  .fade_scatter(fade_cor_top, .tableau10[1],
                "TOP: CAAS vs FADE BF (rank-norm)",
                "gene_caas_top (rank-norm)", "FADE top max BF (rank-norm)"),
  .fade_scatter(fade_cor_bot, .tableau10[2],
                "BOTTOM: CAAS vs FADE BF (rank-norm)",
                "gene_caas_bot (rank-norm)", "FADE bot max BF (rank-norm)"),
  ncol=2,
  top=grid::textGrob("Gene CAAS Score vs FADE BF — rank-normalised",
                      gp=grid::gpar(fontface="bold", fontsize=13)))
```

```{r gene-fade-table, eval=has_fade_pos}
fade_tbl <- gene_fade_df %>%
  mutate(
    in_top_list = Gene %in% gene_top_list,
    in_bot_list = Gene %in% gene_bot_list
  ) %>%
  arrange(desc(pmax(replace_na(fade_top_max_bf, 0), replace_na(fade_bot_max_bf, 0)))) %>%
  select(Gene,
         fade_top_max_bf, fade_top_sig,
         fade_bot_max_bf, fade_bot_sig,
         gene_caas_score, gene_caas_top, gene_caas_bot,
         in_top_list, in_bot_list) %>%
  mutate(across(where(is.numeric), ~round(., 4)))

DT::datatable(
  fade_tbl, filter = "top", rownames = FALSE,
  options = list(pageLength = 20, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "FADE gene-level results ranked by max BF — direction and CAAS gene list membership"
  )
)
```

### Accumulation

`r if (!has_accum_score) "> **Note:** Accumulation scores not available for this run."`

```{r gene-accum-venn, eval=has_accum_score, fig.height=7, fig.width=12}
pct_list_gene <- list(
  list(val=params$gene_top_pct,  lbl=sprintf("Top %g%%", params$gene_top_pct  * 100)),
  list(val=params$gene_top5_pct, lbl=sprintf("Top %g%%", params$gene_top5_pct * 100)),
  list(val=params$gene_top1_pct, lbl=sprintf("Top %g%%", params$gene_top1_pct * 100))
)
accum_venn_plots <- list()
accum_overlap_exports <- list()  # collect overlap gene lists for ORA export
for (dir_lbl in c("TOP", "BOTTOM")) {
  caas_col <- if (dir_lbl == "TOP") "gene_caas_top" else "gene_caas_bot"
  col_a    <- if (dir_lbl == "TOP") .tableau10[1]   else .tableau10[2]
  dir_slug <- tolower(dir_lbl)
  for (pct in pct_list_gene) {
    caas_genes  <- gene_df %>% filter(!is.na(.data[[caas_col]])) %>%
      slice_max(.data[[caas_col]], prop=pct$val) %>% pull(Gene)
    accum_genes <- gene_df %>% filter(!is.na(gene_accum_score)) %>%
      slice_max(gene_accum_score, prop=pct$val) %>% pull(Gene)
    overlap_genes <- intersect(caas_genes, accum_genes)
    pct_slug <- gsub("[^0-9]", "", pct$lbl)  # e.g. "10", "5", "1"
    accum_overlap_exports[[sprintf("gene_caas_%s_top%spct_x_accum.txt", dir_slug, pct_slug)]] <- overlap_genes
    vc <- .venn_counts(caas_genes, accum_genes, sprintf("CAAS %s", dir_lbl), "Accum.")
    accum_venn_plots[[length(accum_venn_plots)+1]] <-
      .draw_venn(vc, col_a, .tableau10[5], sprintf("%s — %s", dir_lbl, pct$lbl))
  }
}

# Export accumulation overlap gene lists as .txt files for downstream ORA
for (fname in names(accum_overlap_exports)) {
  genes <- unique(accum_overlap_exports[[fname]])
  genes <- genes[!is.na(genes) & genes != ""]
  if (length(genes) > 0) {
    writeLines(genes, file.path(params$output_dir, fname))
    cat(sprintf("  Exported %s: %d genes\n", fname, length(genes)))
  }
}

gridExtra::grid.arrange(grobs=accum_venn_plots, ncol=3,
  top=grid::textGrob("Directional CAAS Gene List vs Top Accumulation Genes",
                      gp=grid::gpar(fontface="bold", fontsize=13)))
```

```{r gene-accum-correlation, eval=has_accum_score, fig.height=5, fig.width=12}
accum_cor_top <- gene_df %>%
  filter(!is.na(gene_caas_top), !is.na(gene_accum_score)) %>%
  mutate(x_norm=rank_norm(gene_caas_top), y_norm=rank_norm(gene_accum_score))
accum_cor_bot <- gene_df %>%
  filter(!is.na(gene_caas_bot), !is.na(gene_accum_score)) %>%
  mutate(x_norm=rank_norm(gene_caas_bot), y_norm=rank_norm(gene_accum_score))

gridExtra::grid.arrange(
  ggplot(accum_cor_top, aes(x=x_norm, y=y_norm)) +
    geom_point(alpha=0.5, size=1.2, colour=.tableau10[1]) +
    geom_smooth(method="lm", se=TRUE, colour="grey30", linewidth=0.7) +
    ggrepel::geom_text_repel(
      data=accum_cor_top %>% filter(x_norm>quantile(x_norm,0.95,na.rm=TRUE)|
                                      y_norm>quantile(y_norm,0.95,na.rm=TRUE)),
      aes(label=Gene), size=2.5, max.overlaps=10, show.legend=FALSE) +
    labs(title="TOP: gene_caas_top vs Accumulation",
         subtitle=sprintf("r = %.3f | rank-normalised",
                          safe_cor(accum_cor_top$x_norm, accum_cor_top$y_norm)),
         x="gene_caas_top (rank-norm)", y="gene_accum_score (rank-norm)") +
    theme(plot.title=element_text(face="bold", size=11)),
  ggplot(accum_cor_bot, aes(x=x_norm, y=y_norm)) +
    geom_point(alpha=0.5, size=1.2, colour=.tableau10[2]) +
    geom_smooth(method="lm", se=TRUE, colour="grey30", linewidth=0.7) +
    ggrepel::geom_text_repel(
      data=accum_cor_bot %>% filter(x_norm>quantile(x_norm,0.95,na.rm=TRUE)|
                                      y_norm>quantile(y_norm,0.95,na.rm=TRUE)),
      aes(label=Gene), size=2.5, max.overlaps=10, show.legend=FALSE) +
    labs(title="BOTTOM: gene_caas_bot vs Accumulation",
         subtitle=sprintf("r = %.3f | rank-normalised",
                          safe_cor(accum_cor_bot$x_norm, accum_cor_bot$y_norm)),
         x="gene_caas_bot (rank-norm)", y="gene_accum_score (rank-norm)") +
    theme(plot.title=element_text(face="bold", size=11)),
  ncol=2,
  top=grid::textGrob("Gene CAAS vs Accumulation — rank-normalised",
                      gp=grid::gpar(fontface="bold", fontsize=13)))
```

```{r gene-accum-table, eval=has_accum_score}
accum_pval_cols <- grep("^accum_pval_", names(gene_df), value = TRUE)
accum_tbl <- gene_df %>%
  filter(!is.na(gene_accum_score)) %>%
  mutate(
    in_top_list = Gene %in% gene_top_list,
    in_bot_list = Gene %in% gene_bot_list
  ) %>%
  arrange(desc(gene_accum_score)) %>%
  select(Gene, gene_accum_score, any_of(accum_pval_cols),
         gene_caas_score, gene_caas_top, gene_caas_bot,
         in_top_list, in_bot_list) %>%
  mutate(across(where(is.numeric), ~round(., 4)))

DT::datatable(
  accum_tbl, filter = "top", rownames = FALSE,
  options = list(pageLength = 20, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "Accumulation gene scores ranked by gene_accum_score — per-scheme p-values and CAAS membership"
  )
)
```

### PrimateAI-3D

`r if (!has_vep_primateai) "> **Note:** PrimateAI-3D scores are not available for this run."`

> **Note:** PrimateAI-3D scores require an access request to Illumina —
> see illumina.com/science/genomics-research/articles/primateai-3d.html

```{r gene-pai-directional, eval=has_vep_primateai, fig.height=5, fig.width=12}
gene_vep_pai <- pos_vep_pai %>%
  group_by(Gene) %>%
  summarise(n_pai_positions = n(),
            mean_pai_char   = mean(pai_char_score, na.rm=TRUE),
            sum_pai_char    = sum(pai_char_score,  na.rm=TRUE),
            .groups = "drop") %>%
  inner_join(gene_df %>% select(Gene, gene_caas_score, gene_caas_top,
                                gene_caas_bot, n_positions, gene_direction),
             by = "Gene")

pai_top_df <- gene_vep_pai %>%
  filter(gene_direction %in% c("top","both"), !is.na(gene_caas_top)) %>%
  mutate(x_norm=rank_norm(gene_caas_top), y_norm=rank_norm(mean_pai_char))
pai_bot_df <- gene_vep_pai %>%
  filter(gene_direction %in% c("bottom","both"), !is.na(gene_caas_bot)) %>%
  mutate(x_norm=rank_norm(gene_caas_bot), y_norm=rank_norm(mean_pai_char))

gridExtra::grid.arrange(
  ggplot(pai_top_df, aes(x=x_norm, y=y_norm, size=n_positions)) +
    geom_point(alpha=0.65, colour=.tableau10[1]) +
    geom_smooth(method="lm", se=TRUE, colour="grey30", linewidth=0.7) +
    ggrepel::geom_text_repel(
      data=pai_top_df %>% slice_max(y_norm, n=15, with_ties=FALSE),
      aes(label=Gene), size=2.5, max.overlaps=15, show.legend=FALSE) +
    scale_size_continuous(name="N pos", range=c(1,6)) +
    labs(title="TOP genes: CAAS vs PrimateAI-3D",
         subtitle=sprintf("r = %.3f | rank-normalised",
                          safe_cor(pai_top_df$x_norm, pai_top_df$y_norm)),
         x="gene_caas_top (rank-norm)", y="Mean PAI3D char score (rank-norm)") +
    theme(plot.title=element_text(face="bold", size=11)),
  ggplot(pai_bot_df, aes(x=x_norm, y=y_norm, size=n_positions)) +
    geom_point(alpha=0.65, colour=.tableau10[2]) +
    geom_smooth(method="lm", se=TRUE, colour="grey30", linewidth=0.7) +
    ggrepel::geom_text_repel(
      data=pai_bot_df %>% slice_max(y_norm, n=15, with_ties=FALSE),
      aes(label=Gene), size=2.5, max.overlaps=15, show.legend=FALSE) +
    scale_size_continuous(name="N pos", range=c(1,6)) +
    labs(title="BOTTOM genes: CAAS vs PrimateAI-3D",
         subtitle=sprintf("r = %.3f | rank-normalised",
                          safe_cor(pai_bot_df$x_norm, pai_bot_df$y_norm)),
         x="gene_caas_bot (rank-norm)", y="Mean PAI3D char score (rank-norm)") +
    theme(plot.title=element_text(face="bold", size=11)),
  ncol=2,
  top=grid::textGrob("Gene CAAS vs PrimateAI-3D — Direction-aware (rank-normalised)",
                      gp=grid::gpar(fontface="bold", fontsize=13)))
```

```{r gene-pai-top-bar, eval=has_vep_primateai, fig.height=6}
top_pai_genes <- gene_vep_pai %>%
  filter(n_pai_positions >= 2, mean_pai_char > 0) %>%
  slice_max(mean_pai_char, n=30, with_ties=FALSE) %>%
  arrange(mean_pai_char) %>%
  mutate(Gene = factor(Gene, levels=Gene))

ggplot(top_pai_genes, aes(x=mean_pai_char, y=Gene, fill=gene_caas_score)) +
  geom_col() +
  scale_fill_viridis_c(name="Gene\nCAAS score", option="plasma") +
  scale_x_continuous(expand=expansion(mult=c(0, 0.05))) +
  labs(title = "Top 30 Genes by Mean PrimateAI-3D Characterisation Score",
       subtitle = paste0("Genes with \u2265 2 positions with PAI3D coverage | colour = gene_caas_score"),
       x = "Mean pai_char_score", y = NULL) +
  theme(plot.title=element_text(face="bold"), axis.text.y=element_text(size=8))
```

```{r gene-pai-table, eval=has_vep_primateai}
pai_tbl <- gene_vep_pai %>%
  mutate(
    in_top_list = Gene %in% gene_top_list,
    in_bot_list = Gene %in% gene_bot_list
  ) %>%
  arrange(desc(mean_pai_char)) %>%
  select(Gene, n_pai_positions, mean_pai_char, sum_pai_char,
         gene_caas_score, gene_caas_top, gene_caas_bot, gene_direction,
         in_top_list, in_bot_list) %>%
  mutate(across(where(is.numeric), ~round(., 4)))

DT::datatable(
  pai_tbl, filter = "top", rownames = FALSE,
  options = list(pageLength = 20, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "PrimateAI-3D gene scores ranked by mean PAI characterisation score"
  )
)
```

### Top Genes

```{r top-genes-table}
top_genes_df <- gene_df %>%
  arrange(desc(gene_caas_score))

# ── Direction assignment ──────────────────────────────────────────────────────
if (has_rer) {
  top_genes_df <- top_genes_df %>%
    mutate(rer_direction = case_when(
      !is.na(rer_rho) & rer_rho > 0 ~ "accelerating",
      !is.na(rer_rho) & rer_rho < 0 ~ "decelerating",
      TRUE ~ NA_character_
    ))
}

# ── FADE: max BF and BF>=100 significance flag per direction ─────────────────
if (has_fade_pos && exists("gene_fade_df")) {
  top_genes_df <- top_genes_df %>%
    left_join(
      gene_fade_df %>% select(Gene, fade_top_max_bf, fade_bot_max_bf,
                               fade_top_sig, fade_bot_sig),
      by = "Gene"
    )
}

# ── Accumulation score ────────────────────────────────────────────────────────
# gene_accum_score already on top_genes_df; no extra join needed.

# ── PrimateAI-3D mean characterisation score per gene ────────────────────────
if (has_vep_primateai && exists("gene_vep_pai")) {
  top_genes_df <- top_genes_df %>%
    left_join(gene_vep_pai %>% select(Gene, mean_pai_char, n_pai_positions),
              by = "Gene")
}

# ── CAAS-threshold indicators (10%, 50% and 100%) per direction ──────────────
g_top_thr_50 <- quantile(gene_df$gene_caas_top, 0.50, na.rm = TRUE)
g_bot_thr_50 <- quantile(gene_df$gene_caas_bot, 0.50, na.rm = TRUE)

top_genes_df <- top_genes_df %>%
  mutate(
    in_top_10pct = replace_na(gene_caas_top >= g_top_thr_10, FALSE),
    in_top_50pct = replace_na(gene_caas_top >= g_top_thr_50, FALSE),
    in_bot_10pct = replace_na(gene_caas_bot >= g_bot_thr_10, FALSE),
    in_bot_50pct = replace_na(gene_caas_bot >= g_bot_thr_50, FALSE)
  )

# ── Build display column list ─────────────────────────────────────────────────
gene_display_cols <- c("Gene", "n_positions",
                       "gene_caas_score", "gene_caas_top", "gene_caas_bot",
                       "in_top_10pct", "in_top_50pct",
                       "in_bot_10pct", "in_bot_50pct")
if (has_accum_score)
  gene_display_cols <- c(gene_display_cols, "gene_accum_score")
if (has_rer)
  gene_display_cols <- c(gene_display_cols, "rer_min_pval", "rer_direction")
if (has_fade_pos && "fade_top_max_bf" %in% names(top_genes_df))
  gene_display_cols <- c(gene_display_cols,
                         "fade_top_max_bf", "fade_top_sig",
                         "fade_bot_max_bf", "fade_bot_sig")
if (has_vep_primateai && "mean_pai_char" %in% names(top_genes_df))
  gene_display_cols <- c(gene_display_cols, "n_pai_positions", "mean_pai_char")

DT::datatable(
  top_genes_df %>%
    select(any_of(gene_display_cols)) %>%
    mutate(across(where(is.numeric) & !matches("n_positions|n_pai"), ~round(., 4))),
  filter = "top", rownames = FALSE,
  options = list(pageLength = 25, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Gene-level scores | Trait: %s | sorted by gene_caas_score", params$traitname)
  )
)
```

---

## Genomic Accumulation {.tabset}

`r if (!has_genomic) "> **Note:** Genomic coordinates (gene_ensembl_file) are not available for this run — genomic accumulation skipped."`

```{r genomic-setup, eval=has_genomic}
# Join gene_df with genomic coordinates, then annotate with VEP
gene_genomic <- gene_df %>%
  select(Gene, gene_caas_score, gene_caas_top, gene_caas_bot, n_positions) %>%
  inner_join(genomic_raw, by = "Gene") %>%
  # Standardise chromosome labels (remove "chr" prefix if present)
  mutate(
    chr_label = sub("^chr", "", chr),
    # Order chromosomes numerically then X, Y, MT
    chr_num   = suppressWarnings(as.integer(chr_label)),
    chr_order = dplyr::case_when(
      !is.na(chr_num)       ~ chr_num,
      chr_label == "X"      ~ 23L,
      chr_label == "Y"      ~ 24L,
      chr_label %in% c("MT","M") ~ 25L,
      TRUE                  ~ 26L
    )
  ) %>%
  arrange(chr_order, gene_mid)

# Optionally add mean PAI characterisation score
if (has_vep_primateai && exists("gene_vep_pai")) {
  gene_genomic <- gene_genomic %>%
    left_join(gene_vep_pai %>% select(Gene, mean_pai_char),
              by = "Gene")
} else {
  gene_genomic$mean_pai_char <- NA_real_
}

# Build a cumulative x-axis offset per chromosome for a stitched Manhattan view
chr_offsets <- gene_genomic %>%
  group_by(chr_order, chr_label) %>%
  summarise(chr_max = max(gene_end, na.rm = TRUE), .groups = "drop") %>%
  arrange(chr_order) %>%
  mutate(offset = cumsum(lag(chr_max, default = 0L)) + (chr_order - 1L) * 5e6L)

gene_genomic <- gene_genomic %>%
  left_join(chr_offsets %>% select(chr_order, offset, chr_label), by = c("chr_order", "chr_label")) %>%
  mutate(manhattan_x = gene_mid + offset)

chr_label_positions <- chr_offsets %>%
  left_join(gene_genomic %>% group_by(chr_order) %>%
              summarise(mid_x = mean(manhattan_x, na.rm = TRUE), .groups = "drop"),
            by = "chr_order")
```

```{r window-helper, eval=has_genomic}
# Helper: tile chromosomes into non-overlapping windows of win_size bp.
# Each gene/position is assigned to the window containing its genomic midpoint.
# Returns a dataframe with one row per window, summarising mean and max score.
make_window_accum <- function(df, score_col,
                               win_size = params$window_size_bp) {
  df %>%
    mutate(
      win_id    = floor(gene_mid / win_size),
      win_start = win_id * win_size,
      win_end   = win_start + win_size - 1L,
      win_mid   = win_start + win_size / 2L
    ) %>%
    group_by(chr_order, chr_label, win_id, win_start, win_end, win_mid) %>%
    summarise(
      n_genes    = dplyr::n(),
      mean_score = mean(.data[[score_col]], na.rm = TRUE),
      max_score  = max(.data[[score_col]],  na.rm = TRUE),
      genes      = paste(unique(Gene), collapse = ", "),
      .groups    = "drop"
    )
}

# Stitch window midpoints onto the cumulative chromosome x-axis (same offsets
# as gene_genomic so windows align with the per-gene manhattans above).
stitch_windows <- function(win_df) {
  win_df %>%
    left_join(chr_offsets %>% select(chr_order, offset), by = "chr_order") %>%
    mutate(win_x = win_mid + offset,
           chr_even = chr_order %% 2 == 0)
}
```

### CAAS Score Manhattan

```{r genomic-manhattan-caas, eval=has_genomic, fig.height=5, fig.width=14}
gene_genomic <- gene_genomic %>%
  mutate(
    chr_even      = chr_order %% 2 == 0,
    caas_log      = -log10(1 - pmin(gene_caas_score, 0.9999)),
    caas_top_log  = -log10(1 - pmin(replace_na(gene_caas_top, 0), 0.9999)),
    caas_bot_log  = -log10(1 - pmin(replace_na(gene_caas_bot, 0), 0.9999))
  )
caas_log_thr <- -log10(1 - pmin(g_thr_10, 0.9999))

ggplot(gene_genomic,
       aes(x = manhattan_x, y = caas_log,
           colour = chr_even, size = n_positions)) +
  geom_point(alpha = 0.65, show.legend = FALSE) +
  geom_hline(yintercept = caas_log_thr, linetype = "dashed",
             colour = "grey40", linewidth = 0.6) +
  ggrepel::geom_text_repel(
    data = gene_genomic %>% slice_max(caas_log, n = 20, with_ties = FALSE),
    aes(label = Gene), size = 2.2, max.overlaps = 15,
    show.legend = FALSE, colour = "black"
  ) +
  scale_colour_manual(values = c("FALSE"=.tableau10[1], "TRUE"=.tableau10[2]), guide="none") +
  scale_size_continuous(name = "N positions", range = c(0.5, 5)) +
  scale_x_continuous(
    breaks = chr_label_positions$mid_x,
    labels = chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  labs(
    title    = "Gene CAAS Score — Manhattan View",
    subtitle = sprintf("%d genes | y = -log10(1 - gene_caas_score) | dashed = top %g%% threshold",
                       nrow(gene_genomic), params$gene_top_pct * 100),
    x = "Chromosome", y = "-log10(1 - gene_caas_score)"
  ) +
  theme(plot.title=element_text(face="bold"), axis.text.x=element_text(size=7),
        panel.grid.major.x=element_blank())
```

### PAI3D Signal Manhattan

`r if (!has_vep_primateai) "> **Note:** PrimateAI-3D scores not available — PAI3D Manhattan skipped."`

```{r genomic-manhattan-pai, eval=has_genomic && has_vep_primateai && exists("gene_vep_pai"), fig.height=5, fig.width=14}
gene_genomic_pai <- gene_genomic
if (!"mean_pai_char" %in% names(gene_genomic_pai)) {
  gene_genomic_pai <- gene_genomic_pai %>%
    left_join(gene_vep_pai %>% select(Gene, mean_pai_char), by = "Gene")
}

ggplot(gene_genomic_pai %>% filter(!is.na(mean_pai_char)),
       aes(x=manhattan_x, y=caas_log, size=mean_pai_char, colour=mean_pai_char)) +
  geom_point(alpha=0.70) +
  geom_hline(yintercept=caas_log_thr, linetype="dashed",
             colour="grey60", linewidth=0.5) +
  ggrepel::geom_text_repel(
    data = gene_genomic_pai %>% filter(!is.na(mean_pai_char)) %>%
      slice_max(mean_pai_char, n=20, with_ties=FALSE),
    aes(label=Gene), size=2.2, max.overlaps=12, show.legend=FALSE, colour="black"
  ) +
  scale_colour_viridis_c(name="Mean PAI3D\nchar score", option="inferno") +
  scale_size_continuous(name="Mean PAI3D\nchar score", range=c(0.5, 6)) +
  scale_x_continuous(breaks=chr_label_positions$mid_x, labels=chr_label_positions$chr_label,
                     expand=expansion(mult=0.01)) +
  guides(colour=guide_legend(), size=guide_legend()) +
  labs(title    = "Gene CAAS Score + PrimateAI-3D — Manhattan View",
      subtitle = "Bubble size and colour: mean PAI3D characterisation score | y = -log10(1 - gene_caas_score)",
       x = "Chromosome", y = "-log10(1 - gene_caas_score)") +
  theme(plot.title=element_text(face="bold"), axis.text.x=element_text(size=7),
        panel.grid.major.x=element_blank())
```

### CAAS Window Accumulation (Positions)

`r if (!has_genomic) "> **Note:** Genomic coordinates not available — position window accumulation skipped."`

```{r pos-window-accum, eval=has_genomic, fig.height=5, fig.width=14}
# Tile position-level CAAS scores into genomic windows.
# pos_genomic is built in the pos-caas-manhattan-setup chunk above.
# We use the same pos_chr_offsets to align x-coordinates with the gene-level plots.
win_size_label <- if (params$window_size_bp >= 1e6) {
  sprintf("%g Mb", params$window_size_bp / 1e6)
} else {
  sprintf("%g kb", params$window_size_bp / 1e3)
}

pos_wins <- pos_genomic %>%
  rename(Gene = Gene) %>%
  make_window_accum("CAAS_score", win_size = params$window_size_bp) %>%
  left_join(pos_chr_offsets %>% select(chr_order, offset), by = "chr_order") %>%
  mutate(n_positions = n_genes,
         win_x       = win_mid + offset,
         chr_even    = chr_order %% 2 == 0,
         mean_log    = -log10(1 - pmin(mean_score, 0.9999)),
         max_log     = -log10(1 - pmin(max_score,  0.9999)))

# Label the top windows as genomic regions, not gene names
top_pos_wins <- pos_wins %>%
  slice_max(mean_log, n = 5, with_ties = FALSE) %>%
  mutate(region_label = sprintf("chr%s:%0.1f-%0.1f Mb", chr_label, win_start / 1e6, win_end / 1e6))

ggplot(pos_wins, aes(x = win_x, y = mean_log, fill = chr_even)) +
  geom_col(width = params$window_size_bp * 0.9, alpha = 0.75) +
  geom_point(aes(y = max_log, colour = chr_even), size = 1.2, alpha = 0.85, show.legend = FALSE) +
  ggrepel::geom_text_repel(
    data = top_pos_wins,
    aes(y = max_log, label = sprintf("%s\n(%d positions)", region_label, n_positions)),
    size = 2.2, max.overlaps = 10, colour = "black",
    show.legend = FALSE
  ) +
  scale_fill_manual(
    values = c("FALSE" = .tableau10[1], "TRUE" = .tableau10[2]),
    guide = "none"
  ) +
  scale_colour_manual(
    values = c("FALSE" = .tableau10[3], "TRUE" = .tableau10[4]),
    guide = "none"
  ) +
  scale_x_continuous(
    breaks = pos_chr_label_positions$mid_x,
    labels = pos_chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  labs(
    title    = sprintf("Position-Level CAAS Score — %s Window Accumulation", win_size_label),
    subtitle = sprintf("Bars = mean | Points = max | %d windows across %d chromosomes",
                       nrow(pos_wins), length(unique(pos_wins$chr_order))),
    x = "Chromosome", y = "-log10(1 - mean CAAS score)"
  ) +
  theme(
    plot.title         = element_text(face = "bold"),
    axis.text.x        = element_text(size = 7),
    panel.grid.major.x = element_blank()
  )
```

```{r pos-window-accum-table, eval=has_genomic && exists("pos_wins") && nrow(pos_wins) > 0}
DT::datatable(
  pos_wins %>%
    select(chr_label, win_start, win_end, n_positions, mean_score, max_score, mean_log) %>%
    mutate(
      win_start  = format(win_start, big.mark = ",", scientific = FALSE),
      win_end    = format(win_end,   big.mark = ",", scientific = FALSE),
      mean_score = round(mean_score, 4),
      max_score  = round(max_score,  4),
      mean_log   = round(mean_log,   4)
    ) %>%
    arrange(desc(mean_log)) %>%
    rename(chr = chr_label, start = win_start, end = win_end,
           n_pos = n_positions, mean_CAAS = mean_score, max_CAAS = max_score,
           `-log10(1-mean)` = mean_log),
  rownames = FALSE,
  options = list(pageLength = 15, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Top genomic windows by position-level CAAS accumulation (%s windows)", win_size_label)
  )
)
```

### CAAS Window Accumulation (Genes)

`r if (!has_genomic) "> **Note:** Genomic coordinates not available — gene-level CAAS window accumulation skipped."`

```{r gene-caas-window-accum, eval=has_genomic, fig.height=5, fig.width=14}
gene_caas_wins <- gene_genomic %>%
  make_window_accum("gene_caas_score", win_size = params$window_size_bp) %>%
  stitch_windows() %>%
  mutate(mean_log = -log10(1 - pmin(mean_score, 0.9999)),
         max_log  = -log10(1 - pmin(max_score,  0.9999)))

top_caas_wins <- gene_caas_wins %>%
  slice_max(mean_log, n = 5, with_ties = FALSE) %>%
  mutate(region_label = sprintf("chr%s:%0.1f-%0.1f Mb", chr_label, win_start / 1e6, win_end / 1e6))

ggplot(gene_caas_wins, aes(x = win_x, y = mean_log, fill = chr_even)) +
  geom_col(width = params$window_size_bp * 0.9, alpha = 0.75) +
  geom_point(aes(y = max_log, colour = chr_even), size = 1.5, alpha = 0.85, show.legend = FALSE) +
  ggrepel::geom_text_repel(
    data = top_caas_wins,
    aes(y = max_log, label = sprintf("%s\n(%d genes)", region_label, n_genes)),
    size = 2.2, max.overlaps = 10, colour = "black",
    show.legend = FALSE
  ) +
  scale_fill_manual(
    values = c("FALSE" = .tableau10[1], "TRUE" = .tableau10[2]),
    guide = "none"
  ) +
  scale_colour_manual(
    values = c("FALSE" = .tableau10[3], "TRUE" = .tableau10[4]),
    guide = "none"
  ) +
  scale_x_continuous(
    breaks = chr_label_positions$mid_x,
    labels = chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  labs(
    title    = sprintf("Gene CAAS Score — %s Window Accumulation", win_size_label),
    subtitle = sprintf("Bars = mean | Points = max | %d windows", nrow(gene_caas_wins)),
    x = "Chromosome", y = "-log10(1 - mean gene_caas_score)"
  ) +
  theme(
    plot.title         = element_text(face = "bold"),
    axis.text.x        = element_text(size = 7),
    panel.grid.major.x = element_blank()
  )
```

```{r gene-caas-window-accum-table, eval=has_genomic && exists("gene_caas_wins") && nrow(gene_caas_wins) > 0}
DT::datatable(
  gene_caas_wins %>%
    select(chr_label, win_start, win_end, n_genes, mean_score, max_score, mean_log) %>%
    mutate(
      win_start  = format(win_start, big.mark = ",", scientific = FALSE),
      win_end    = format(win_end,   big.mark = ",", scientific = FALSE),
      mean_score = round(mean_score, 4),
      max_score  = round(max_score,  4),
      mean_log   = round(mean_log,   4)
    ) %>%
    arrange(desc(mean_log)) %>%
    rename(chr = chr_label, start = win_start, end = win_end,
           n_genes = n_genes, mean_CAAS = mean_score, max_CAAS = max_score,
           `-log10(1-mean)` = mean_log),
  rownames = FALSE,
  options = list(pageLength = 15, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Top genomic windows by gene CAAS score accumulation (%s windows)", win_size_label)
  )
)
```

### CAAS TOP Window Accumulation (Genes)

`r if (!has_genomic) "> **Note:** Genomic coordinates not available — gene-level CAAS TOP window accumulation skipped."`

```{r gene-caas-top-window-accum, eval=has_genomic && any(!is.na(gene_genomic$gene_caas_top)), fig.height=5, fig.width=14}
gene_top_wins <- gene_genomic %>%
  filter(!is.na(gene_caas_top)) %>%
  make_window_accum("gene_caas_top", win_size = params$window_size_bp) %>%
  stitch_windows() %>%
  mutate(mean_log = -log10(1 - pmin(mean_score, 0.9999)),
         max_log  = -log10(1 - pmin(max_score,  0.9999)))

top_top_wins <- gene_top_wins %>%
  slice_max(mean_log, n = 5, with_ties = FALSE) %>%
  mutate(region_label = sprintf("chr%s:%0.1f-%0.1f Mb", chr_label, win_start / 1e6, win_end / 1e6))

ggplot(gene_top_wins, aes(x = win_x, y = mean_log, fill = chr_even)) +
  geom_col(width = params$window_size_bp * 0.9, alpha = 0.75) +
  geom_point(aes(y = max_log, colour = chr_even), size = 1.5, alpha = 0.85, show.legend = FALSE) +
  ggrepel::geom_text_repel(
    data = top_top_wins,
    aes(y = max_log, label = sprintf("%s\n(%d genes)", region_label, n_genes)),
    size = 2.2, max.overlaps = 10, colour = "black", show.legend = FALSE
  ) +
  scale_fill_manual(values = c("FALSE"=.tableau10[1], "TRUE"=.tableau10[2]), guide="none") +
  scale_colour_manual(values = c("FALSE"=.tableau10[3], "TRUE"=.tableau10[4]), guide="none") +
  scale_x_continuous(breaks=chr_label_positions$mid_x, labels=chr_label_positions$chr_label,
                     expand=expansion(mult=0.01)) +
  labs(title    = sprintf("Gene CAAS Score TOP — %s Window Accumulation", win_size_label),
       subtitle = sprintf("Bars = mean | Points = max | %d windows", nrow(gene_top_wins)),
       x = "Chromosome", y = "-log10(1 - mean gene_caas_top)") +
  theme(plot.title=element_text(face="bold"), axis.text.x=element_text(size=7),
        panel.grid.major.x=element_blank())
```

```{r gene-caas-top-window-accum-table, eval=has_genomic && exists("gene_top_wins") && nrow(gene_top_wins) > 0}
DT::datatable(
  gene_top_wins %>%
    select(chr_label, win_start, win_end, n_genes, mean_score, max_score, mean_log) %>%
    mutate(
      win_start  = format(win_start, big.mark = ",", scientific = FALSE),
      win_end    = format(win_end,   big.mark = ",", scientific = FALSE),
      mean_score = round(mean_score, 4),
      max_score  = round(max_score,  4),
      mean_log   = round(mean_log,   4)
    ) %>%
    arrange(desc(mean_log)) %>%
    rename(chr = chr_label, start = win_start, end = win_end,
           n_genes = n_genes, mean_CAAS_top = mean_score, max_CAAS_top = max_score,
           `-log10(1-mean)` = mean_log),
  rownames = FALSE,
  options = list(pageLength = 15, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Top genomic windows by directional TOP gene CAAS accumulation (%s windows)", win_size_label)
  )
)
```

```{r genomic-window-tsv-export, eval=has_genomic}
if (exists("pos_wins") && nrow(pos_wins) > 0) {
  readr::write_tsv(
    pos_wins %>% select(-chr_even, -mean_log, -max_log),
    file.path(params$output_dir, "genomic_window_accum_positions.tsv"))
}
if (exists("gene_caas_wins") && nrow(gene_caas_wins) > 0) {
  readr::write_tsv(
    gene_caas_wins %>% select(-chr_even, -mean_log, -max_log),
    file.path(params$output_dir, "genomic_window_accum_gene_caas.tsv"))
}
if (exists("gene_top_wins") && nrow(gene_top_wins) > 0) {
  readr::write_tsv(
    gene_top_wins %>% select(-chr_even, -mean_log, -max_log),
    file.path(params$output_dir, "genomic_window_accum_gene_caas_top.tsv"))
}
cat("Exported genomic window accumulation TSVs\n")
```

---

## Supplementary Exports

```{r supplementary-tsv-export}
# Build a wide supplementary table joining position_scores with FADE and PAI data.
# Exported as position_scores_annotated.tsv alongside the report outputs.

supp_df <- pos_df

# Join FADE data: use the direction-matched FADE table for each position
if (has_fade_pos) {
  fade_combined <- bind_rows(
    if (!is.null(fade_top_df)) fade_top_df %>% mutate(fade_direction = "TOP")    else NULL,
    if (!is.null(fade_bot_df)) fade_bot_df %>% mutate(fade_direction = "BOTTOM") else NULL
  )
  if (nrow(fade_combined) > 0) {
    # For positions in both directions (change_side == "both"), take TOP FADE by priority
    fade_for_join <- fade_combined %>%
      mutate(fade_direction_priority = if_else(fade_direction == "TOP", 1L, 2L)) %>%
      group_by(Gene, Position) %>%
      slice_min(fade_direction_priority, n = 1, with_ties = FALSE) %>%
      ungroup() %>%
      select(Gene, Position,
             fade_max_site_bf     = max_site_bf,
             fade_top_target_aa   = top_target_aa,
             fade_biased,
             fade_direction)
    supp_df <- supp_df %>% left_join(fade_for_join, by = c("Gene", "Position"))
  }
}

# Join PAI data
if (has_vep_primateai && exists("pos_vep_pai") && nrow(pos_vep_pai) > 0) {
  pai_for_join <- pos_vep_pai %>%
    select(Gene, Position, pai_char_score, n_pai_schemes)
  supp_df <- supp_df %>% left_join(pai_for_join, by = c("Gene", "Position"))
}

readr::write_tsv(supp_df,
  file.path(params$output_dir, "position_scores_annotated.tsv"))

cat(sprintf("Exported position_scores_annotated.tsv: %d rows, %d columns\n",
            nrow(supp_df), ncol(supp_df)))
cat(sprintf("Columns: %s\n", paste(names(supp_df), collapse = ", ")))
```

---

## Summary

```{r summary-stats}
n_components <- 4L  # biochem, asr, convergence, parallel

n_gene_components <- 1L  # gene_caas always present
if (has_accum_score) n_gene_components <- n_gene_components + 1L
if (has_rer)         n_gene_components <- n_gene_components + 1L

summary_tbl <- tibble(
  Metric = c(
    "Positions scored", "Genes scored",
    "Position-level components", "Gene-level components",
    "CAAS Score (median)", "CAAS Score (top 10% threshold)",
    "Gene CAAS TOP (median)", "Gene CAAS TOP (top 10% threshold)",
    "FADE site-level analysis available",
    "Accumulation score available", "RERConverge available",
    "VEP PrimateAI-3D available",
    "Genomic coordinates available"
  ),
  Value = c(
    format(nrow(pos_df), big.mark = ","),
    format(nrow(gene_df), big.mark = ","),
    as.character(n_components),
    as.character(n_gene_components),
    sprintf("%.4f", median(pos_df$CAAS_score, na.rm = TRUE)),
    sprintf("%.4f", thr_10),
    sprintf("%.4f", median(gene_df$gene_caas_top, na.rm = TRUE)),
    sprintf("%.4f", g_top_thr_10),
    ifelse(has_fade_pos,     "Yes", "No"),
    ifelse(has_accum_score,  "Yes", "No"),
    ifelse(has_rer,          "Yes", "No"),
    ifelse(has_vep_primateai,"Yes", "No"),
    ifelse(has_genomic,      "Yes", "No")
  )
)

knitr::kable(summary_tbl, align = c("l", "r"))
```

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>
