---
title: "CAAS Scoring Report"
subtitle: "`r params$traitname`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
params:
  position_scores_file: NULL
  gene_scores_file: NULL
  gene_corr_file: NULL
  traitname: "unknown_trait"
  output_dir: "."
  top_pct: 0.10
  top1_pct: 0.01
  gene_top_pct: 0.10
  gene_top1_pct: 0.01
  vep_transvar_file: NULL
  vep_primateai_file: NULL
  genomic_info_file: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  out.width = "100%"
)

suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(paletteer)
  library(DT)
  library(scales)
  library(ggrepel)
})

# Tableau_10 colour swatch for pipeline-wide consistency
.tableau10 <- as.character(paletteer_d("ggthemes::Tableau_10"))
theme_set(theme_minimal(base_size = 13))
```

```{r load-data}
pos_df  <- read_tsv(params$position_scores_file, show_col_types = FALSE)
gene_df <- read_tsv(params$gene_scores_file, show_col_types = FALSE)
corr_df <- read_tsv(params$gene_corr_file, show_col_types = FALSE)

# Detect which optional components are available
has_pgls <- "p_pgls_top" %in% names(pos_df) &&
            (any(!is.na(pos_df$p_pgls_top)) || any(!is.na(pos_df$p_pgls_bottom)))
has_rand <- any(!is.na(gene_df$gene_rand_score))
has_rer  <- any(!is.na(gene_df$gene_rer_score))
has_fade <- "gene_fade_score" %in% names(gene_df) && any(!is.na(gene_df$gene_fade_score))

# ── VEP optional inputs ───────────────────────────────────────────────────────
.file_ok <- function(p) !is.null(p) && !is.na(p) && nchar(p) > 0 && file.exists(p)

has_vep_transvar  <- .file_ok(params$vep_transvar_file)
has_vep_primateai <- .file_ok(params$vep_primateai_file)
has_genomic       <- .file_ok(params$genomic_info_file)

if (has_vep_transvar) {
  vep_tv_df <- read_tsv(params$vep_transvar_file, show_col_types = FALSE)
} else {
  vep_tv_df <- NULL
}

if (has_vep_primateai) {
  vep_pai_df <- read_tsv(params$vep_primateai_file, show_col_types = FALSE)
} else {
  vep_pai_df <- NULL
}

if (has_genomic) {
  # Format: gene, chr, start, end, length  (no header assumed; detect if header present)
  genomic_raw <- read_tsv(params$genomic_info_file,
                          col_names = c("Gene", "chr", "gene_start", "gene_end", "gene_length"),
                          show_col_types = FALSE)
  # Drop header row if first cell is literally "gene" or "Gene"
  if (tolower(genomic_raw$Gene[1]) == "gene") genomic_raw <- genomic_raw[-1, ]
  genomic_raw <- genomic_raw %>%
    mutate(
      gene_start  = as.integer(gene_start),
      gene_end    = as.integer(gene_end),
      gene_length = as.integer(gene_length),
      gene_mid    = as.integer((gene_start + gene_end) / 2L)
    )
} else {
  genomic_raw <- NULL
}

# Thresholds
top_pct  <- params$top_pct
top1_pct <- params$top1_pct
```

---

## Introduction

The **CAAS Scoring** module integrates signals from multiple upstream analyses
into a unified, interpretable metric for prioritising convergent amino acid
substitutions (CAAS).  The scoring operates at two levels:

1. **Position-level CAAS Score** — each alignment position receives a composite
   score that combines:
   - **Biochemical score** — amino acid variability and pattern association,
     weighted by grouping scheme specificity.
   - **ASR score** — ancestral state reconstruction confidence: novel changes
     score 1.0; ASR-confirmed conserved changes are weighted by MRCA posterior
     (max 0.70); contradicted cases score 0.
   - **Convergence score** — type and directionality of the convergent change.
   - **Parallel score** — parallelism pattern weighted by MRCA posterior
     confidence.

2. **Gene-level scores** — each gene receives up to four independent scores:
   - **gene_caas_score** — 90th percentile of position-level CAAS scores.
   `r if (has_rand) "- **gene_rand_score** — randomization enrichment across grouping schemes."`
   `r if (has_rer) "- **gene_rer_score** — RERConverge evolutionary rate correlation."`
   `r if (has_fade) "- **gene_fade_score** — directional selection signal (FADE max Bayes Factor, gene-level)."`
   - **gene_composite_score** — unweighted mean of available gene scores.

All component scores are mapped to a 0--1 scale using decile ranking, making
them directly comparable and suitable for downstream enrichment analysis.

---

## Scoring Methodology

This section documents the exact weighting and combination logic used to
produce every score in this report, so results can be interpreted and
reproduced independently of the pipeline code.

<details>
<summary><strong>Click to expand full methodology</strong></summary>

---

### Position-Level Components

All position-level component scores are mapped to the **[0, 1]** interval.
The four core components are always computed; PGLS and FADE are included only
when the corresponding upstream module has been run.

#### 1 · Variability score

Decile rank of the **hypergeometric p-value** (`Pvalue`) across all
Gene × Position × Scheme rows, inverted so that smaller p-values receive
higher scores.

$$\text{variability\_score} = 1 - \frac{\text{ntile}(P_{\text{hyp}}, 10) - 1}{9}$$

*Range: 0.0 (least significant decile) → 1.0 (most significant decile).*

> This score does **not** enter `CAAS_score` directly — it feeds into
> `biochem_score` as a multiplicative factor.

---

#### 2 · Pattern score

Decile rank of the **permutation p-value** (`pvalue_boot`), inverted in the
same way.

$$\text{pattern\_score} = 1 - \frac{\text{ntile}(P_{\text{boot}}, 10) - 1}{9}$$

> Same role as variability\_score: multiplicative input into `biochem_score`.

---

#### 3 · Biochemical score

Sums the contribution of each grouping **scheme** in which the position was
detected, weighted by scheme specificity:

| Scheme | Weight |
|--------|-------:|
| US     | 0.50   |
| GS4    | 0.20   |
| GS3    | 0.10   |
| GS2    | 0.10   |
| GS1    | 0.10   |
| GS0    | 0.00   |

Each scheme contributes:

$$\text{contribution}_s = w_s \times \text{variability\_score} \times \text{pattern\_score}$$

The final score is the **cumulative sum** over all schemes where the position
was detected:

$$\text{biochem\_score} = \sum_{s \in \text{detected schemes}} w_s \times \text{variability\_score} \times \text{pattern\_score}$$

Maximum value when detected in all weighted schemes (US + GS4 + GS3 + GS2 + GS1)
with variability = pattern = 1.0 is **1.0** (0.5 + 0.2 + 0.1 + 0.1 + 0.1).

---

#### 4 · ASR score

Measures ancestral-state-reconstruction confidence. Three mutually exclusive cases:

| Condition | Score |
|-----------|------:|
| `is_conserved_meta = FALSE` (novel change) | **1.0** |
| `is_conserved_meta = TRUE` **and** `asr_is_conserved = TRUE` | see formula below |
| `is_conserved_meta = TRUE` **and** `asr_is_conserved = FALSE` | **0.0** |

For ASR-confirmed conserved positions, the score uses posterior probabilities
from the MRCA node identified by `conserved_pair`:

$$\text{asr\_score} = 0.35 \times P_{\text{MRCA}(\text{conserved\_pair})} + \begin{cases} 0.35 \times P_{\text{root}} & \text{if } \texttt{asr\_root\_conserved} = \text{TRUE} \\ 0 & \text{otherwise} \end{cases}$$

Maximum = **0.70** (when both posteriors = 1.0). Novel changes score 1.0,
making them always preferred over conserved changes.

---

#### 5 · Convergence score

Scores the **type and directionality** of the amino acid change on each
assessed side (top / bottom foreground lineages).

Side values:

| Change type | Value |
|-------------|------:|
| convergent  | 1.0   |
| codivergent | 0.7   |
| divergent   | 0.3   |
| ambiguous / no change | excluded (NA) |

$$\text{mean\_change} = \text{mean of non-NA side values}$$

$$\text{bidirectional} = \begin{cases} 1.0 & \text{both sides assessed} \\ 0.7 & \text{one side only} \end{cases}$$

$$\text{convergence\_score} = \frac{\text{mean\_change} + \text{bidirectional}}{2}$$

Range: 0.5 (one side, divergent: (0.3 + 0.7) / 2) → 1.0 (both convergent:
(1.0 + 1.0) / 2).

---

#### 6 · Parallel score

Scores whether the amino acid change is **parallel** (same substitution in
independent lineages) or **non-parallel**, weighted by MRCA posterior
confidence.

Parallelism coefficients (note: non-parallel receives the *highest* weight
because independent parallel changes are less informative as a signal of
convergence):

| Parallel type | Coefficient |
|---------------|------------:|
| non-parallel  | 1.0         |
| mixed         | 0.7         |
| parallel      | 0.5         |

For each assessable side (where `change_top`/`change_bottom` is
convergent / codivergent / divergent):

$$\text{side\_score} = \text{coeff} \times \overline{P_{\text{MRCA}}}$$

where $\overline{P_{\text{MRCA}}}$ is the mean of `mrca_1_posterior`,
`mrca_2_posterior`, `mrca_3_posterior` (NA excluded).

$$\text{mean\_parallel} = \text{mean of non-NA side scores}$$

$$\text{bidirectional} = \begin{cases} 1.0 & \text{both sides assessed} \\ 0.7 & \text{one side only} \end{cases}$$

$$\text{parallel\_score} = \frac{\text{mean\_parallel} + \text{bidirectional}}{2}$$

---

#### Composite CAAS score

Unweighted mean of the four core components. PGLS and FADE are **not**
included — PGLS is an independent validation signal evaluated in the
[Position Characterisation](#position-characterisation) section; FADE is a
gene-level signal (see below).

$$\text{CAAS\_score} = \frac{\text{biochem} + \text{asr} + \text{convergence} + \text{parallel}}{4}$$

---

### Gene-Level Scores

#### gene\_caas\_score

90th percentile of `CAAS_score` across all scored positions in that gene:

$$\text{gene\_caas\_score} = Q_{0.90}(\text{CAAS\_score})$$

#### gene\_rand\_score  *(optional)*

Scheme-weighted sum of decile-ranked accumulation randomization p-values.
The same scheme weights as `biochem_score` are used (GS0 excluded):

$$\text{gene\_rand\_score} = \sum_{s \in \{US, GS4, GS3, GS2, GS1\}} w_s \times \left(1 - \frac{\text{ntile}(P_{\text{empirical},s}, 10) - 1}{9}\right)$$

Maximum = 1.0 (all schemes p-value in best decile).

#### gene\_rer\_score  *(optional)*

Decile rank of the RERConverge adjusted p-value (`p.adj`; or `p.perm` if
available and non-zero), inverted:

$$\text{gene\_rer\_score} = 1 - \frac{\text{ntile}(p_{\text{RER}}, 10) - 1}{9}$$

#### gene\_fade\_score  *(optional)*

Gene-level maximum Bayes Factor from FADE directional selection analysis,
taking the best of the top/bottom foreground directions. Decile rank,
higher BF = higher score:

$$\text{gene\_fade\_score} = \frac{\text{ntile}(\max_{d \in \{\text{top,bottom}\}} \text{BF}_{d}, 10) - 1}{9}$$

#### gene\_composite\_score

Unweighted mean of all available gene-level sub-scores:

$$\text{gene\_composite\_score} = \frac{\text{gene\_caas} + [\text{gene\_rand}] + [\text{gene\_rer}] + [\text{gene\_fade}]}{N_{\text{available}}}$$

---

### ORA Gene List Extraction

Gene lists for Over-Representation Analysis are extracted at two thresholds
(configurable via `--scoring_position_top_pct` / `--scoring_position_top1_pct`
and their gene-level equivalents) and three directional subsets:

| Subset | Filter |
|--------|--------|
| **top** | Positions / genes where `change_side == "top"` |
| **bottom** | Positions / genes where `change_side == "bottom"` |
| **both** | All top-scoring positions / genes regardless of direction |

For **position-level** lists, the threshold is applied to `CAAS_score` and
the gene name is extracted from the qualifying positions.

For **gene-level** lists, the threshold is applied to each gene-level score
independently (`gene_caas_score`, `gene_rand_score`, `gene_rer_score`,
`gene_composite_score`), and direction is assigned from the gene's
highest-scoring position.

</details>

---

## Input P-value Distributions {.tabset}

The `variability_score` and `pattern_score` components entering `biochem_score` are
decile ranks of the **hypergeometric p-value** (`Pvalue`) and the
**permutation bootstrap p-value** (`pvalue_boot`) respectively.  Because the
decile transformation collapses the continuous p-value into 10 bins, the plots
below show the underlying raw distribution together with the decile boundaries,
so you can assess how sharply peaked the signal is and how well separated the
scoring bins are.

### Hypergeometric P-value

```{r hyper-pval-dist, fig.height=8}
# Compute decile boundaries from the raw p-values (same logic as scoring_compute.R)
pv_breaks <- quantile(pos_df$Pvalue, probs = seq(0, 1, by = 0.1), na.rm = TRUE)
decile_df_h <- pos_df %>%
  filter(!is.na(Pvalue)) %>%
  mutate(
    neg_log10_p = -log10(pmax(Pvalue, 1e-300)),
    decile      = dplyr::ntile(Pvalue, 10),
    score_val   = 1 - (decile - 1) / 9
  )

# Panel 1: density on -log10 scale with decile boundaries
p_hyp_density <- ggplot(decile_df_h, aes(x = neg_log10_p)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60,
                 fill = .tableau10[1], alpha = 0.6) +
  geom_density(linewidth = 0.8, colour = .tableau10[2]) +
  geom_vline(
    xintercept = -log10(pmax(pv_breaks[2:10], 1e-300)),
    linetype = "dotted", colour = "grey50", linewidth = 0.4
  ) +
  labs(
    title    = "Hypergeometric P-value Distribution (-log10 scale)",
    subtitle = "Dotted lines: decile boundaries (D1 = largest p-values → score 1.0; D10 = smallest → score 1.0 inverted)",
    x        = "-log10(Pvalue)",
    y        = "Density"
  ) +
  theme(plot.title = element_text(face = "bold"))

# Panel 2: count per decile, coloured by variability_score
decile_counts_h <- decile_df_h %>%
  group_by(decile, score_val) %>%
  summarise(n = n(), .groups = "drop")

p_hyp_decile <- ggplot(decile_counts_h,
                        aes(x = factor(decile), y = n, fill = score_val)) +
  geom_col(colour = "white", linewidth = 0.3) +
  scale_fill_gradient(low = .tableau10[3], high = .tableau10[1],
                      name = "variability_score") +
  labs(
    title    = "Positions per Hypergeometric Decile",
    subtitle = "Decile 1 = highest p-values; Decile 10 = smallest p-values (variability_score = 1.0)",
    x        = "Decile",
    y        = "Number of rows"
  ) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "right")

gridExtra::grid.arrange(p_hyp_density, p_hyp_decile, ncol = 1)
```

### Permutation P-value

```{r perm-pval-dist, fig.height=8}
pv_breaks_b <- quantile(pos_df$pvalue_boot, probs = seq(0, 1, by = 0.1), na.rm = TRUE)
decile_df_b <- pos_df %>%
  filter(!is.na(pvalue_boot)) %>%
  mutate(
    neg_log10_p = -log10(pmax(pvalue_boot, 1e-300)),
    decile      = dplyr::ntile(pvalue_boot, 10),
    score_val   = 1 - (decile - 1) / 9
  )

p_boot_density <- ggplot(decile_df_b, aes(x = neg_log10_p)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60,
                 fill = .tableau10[5], alpha = 0.6) +
  geom_density(linewidth = 0.8, colour = .tableau10[6]) +
  geom_vline(
    xintercept = -log10(pmax(pv_breaks_b[2:10], 1e-300)),
    linetype = "dotted", colour = "grey50", linewidth = 0.4
  ) +
  labs(
    title    = "Permutation Bootstrap P-value Distribution (-log10 scale)",
    subtitle = "Dotted lines: decile boundaries (D1 = least significant; D10 = most significant → pattern_score = 1.0)",
    x        = "-log10(pvalue_boot)",
    y        = "Density"
  ) +
  theme(plot.title = element_text(face = "bold"))

decile_counts_b <- decile_df_b %>%
  group_by(decile, score_val) %>%
  summarise(n = n(), .groups = "drop")

p_boot_decile <- ggplot(decile_counts_b,
                         aes(x = factor(decile), y = n, fill = score_val)) +
  geom_col(colour = "white", linewidth = 0.3) +
  scale_fill_gradient(low = .tableau10[7], high = .tableau10[5],
                      name = "pattern_score") +
  labs(
    title    = "Positions per Permutation Decile",
    subtitle = "Decile 10 = most significant permutation p-values (pattern_score = 1.0)",
    x        = "Decile",
    y        = "Number of rows"
  ) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "right")

gridExtra::grid.arrange(p_boot_density, p_boot_decile, ncol = 1)
```

### Joint Decile Summary

```{r joint-decile-summary, fig.height=5}
# Scatter of variability_score vs pattern_score coloured by CAAS_score
joint_df <- pos_df %>%
  filter(!is.na(variability_score), !is.na(pattern_score)) %>%
  mutate(
    hyp_decile  = dplyr::ntile(Pvalue,      10),
    boot_decile = dplyr::ntile(pvalue_boot, 10)
  )

ggplot(joint_df, aes(x = hyp_decile, y = boot_decile, colour = CAAS_score)) +
  geom_jitter(alpha = 0.25, size = 0.6, width = 0.35, height = 0.35) +
  scale_colour_gradient(low = .tableau10[3], high = .tableau10[1],
                        name = "CAAS\nscore") +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(breaks = 1:10) +
  labs(
    title    = "Hypergeometric vs Permutation Decile Assignment per Position",
    subtitle = "High CAAS scores (warm colours) cluster toward decile 10 × 10",
    x        = "Hypergeometric decile (variability)",
    y        = "Permutation decile (pattern)"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

---

## Position-Level Scores {.tabset}

### CAAS Score Distribution

```{r caas-score-dist, fig.height=5}
thr_10 <- quantile(pos_df$CAAS_score, 1 - top_pct,  na.rm = TRUE)
thr_1  <- quantile(pos_df$CAAS_score, 1 - top1_pct, na.rm = TRUE)

ggplot(pos_df, aes(x = CAAS_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50,
                 fill = .tableau10[1], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[2]) +
  geom_vline(xintercept = thr_10, linetype = "dashed",
             colour = .tableau10[3], linewidth = 0.7) +
  geom_vline(xintercept = thr_1,  linetype = "dotted",
             colour = .tableau10[4], linewidth = 0.7) +
  annotate("text", x = thr_10, y = Inf, label = sprintf("Top %g%%", top_pct * 100),
           vjust = 2, hjust = -0.1, colour = .tableau10[3], size = 3.5) +
  annotate("text", x = thr_1,  y = Inf, label = sprintf("Top %g%%", top1_pct * 100),
           vjust = 3.5, hjust = -0.1, colour = .tableau10[4], size = 3.5) +
  labs(title = "CAAS Score Distribution",
       subtitle = sprintf("n = %s positions | Trait: %s",
                          format(nrow(pos_df), big.mark = ","), params$traitname),
       x = "CAAS Score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### Biochemical Score

```{r biochem-dist, fig.height=4.5}
ggplot(pos_df, aes(x = biochem_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = .tableau10[1], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[2]) +
  labs(title = "Biochemical Score Distribution",
       subtitle = "Weighted sum of scheme-specific variability x pattern contributions",
       x = "biochem_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### ASR Score

```{r asr-dist, fig.height=4.5}
ggplot(pos_df, aes(x = asr_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = .tableau10[3], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[4]) +
  labs(title = "ASR Score Distribution",
       subtitle = "Novel change = 1.0 | ASR-confirmed conserved = 0.35×posterior (max 0.70) | Contradicted = 0",
       x = "asr_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### Convergence Score

```{r convergence-dist, fig.height=4.5}
ggplot(pos_df, aes(x = convergence_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = .tableau10[5], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[6]) +
  labs(title = "Convergence Score Distribution",
       subtitle = "Convergence type + bidirectional bonus",
       x = "convergence_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### Parallel Score

```{r parallel-dist, fig.height=4.5}
ggplot(pos_df, aes(x = parallel_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = .tableau10[7], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[8]) +
  labs(title = "Parallel Score Distribution",
       subtitle = "Parallelism coefficient x MRCA posterior + bidirectional bonus",
       x = "parallel_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### Component Overview

```{r component-overview, fig.height=6}
# Faceted density comparison of all position-level components
component_cols <- c("biochem_score", "asr_score", "convergence_score", "parallel_score")

long_df <- pos_df %>%
  select(all_of(component_cols)) %>%
  pivot_longer(everything(), names_to = "Component", values_to = "Score") %>%
  filter(!is.na(Score)) %>%
  mutate(Component = factor(Component, levels = component_cols))

ggplot(long_df, aes(x = Score, fill = Component)) +
  geom_density(alpha = 0.6, linewidth = 0.3) +
  scale_fill_manual(values = .tableau10[seq_along(component_cols)]) +
  facet_wrap(~Component, scales = "free_y", ncol = 2) +
  labs(title = "Position-Level Component Score Distributions",
       x = "Score (0-1)", y = "Density") +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "none")
```

### Threshold Profiles

```{r threshold-profiles, fig.height=14, fig.width=12}
# Component score distributions for positions inside each ORA gene-list cutoff.
# Each panel shows overlaid densities of all position-level components so you
# can read off which signal drives positions into the top-N% × direction bucket.

component_cols_tp <- c("biochem_score", "asr_score", "convergence_score", "parallel_score")

thr_10_tp <- quantile(pos_df$CAAS_score, 1 - top_pct,  na.rm = TRUE)
thr_1_tp  <- quantile(pos_df$CAAS_score, 1 - top1_pct, na.rm = TRUE)

make_threshold_panel <- function(df, threshold, pct_label, direction) {
  subset_df <- if (direction == "both") {
    df %>% filter(CAAS_score >= threshold)
  } else {
    df %>% filter(CAAS_score >= threshold, change_side == direction)
  }

  if (nrow(subset_df) == 0) {
    return(
      ggplot() +
        annotate("text", x = 0.5, y = 0.5, label = "No positions",
                 size = 5, colour = "grey50") +
        labs(title = sprintf("Top %s — %s", pct_label, direction),
             x = "Score (0–1)", y = "Density") +
        xlim(0, 1) + theme_minimal()
    )
  }

  long_df <- subset_df %>%
    select(all_of(component_cols_tp)) %>%
    pivot_longer(everything(), names_to = "Component", values_to = "Score") %>%
    filter(!is.na(Score)) %>%
    mutate(Component = factor(Component, levels = component_cols_tp))

  ggplot(long_df, aes(x = Score, fill = Component, colour = Component)) +
    geom_density(alpha = 0.35, linewidth = 0.6) +
    scale_fill_manual(values   = .tableau10[seq_along(component_cols_tp)]) +
    scale_colour_manual(values = .tableau10[seq_along(component_cols_tp)]) +
    scale_x_continuous(limits = c(0, 1), expand = c(0.01, 0)) +
    labs(
      title    = sprintf("Top %s — %s", pct_label, direction),
      subtitle = sprintf("n = %d positions", nrow(subset_df)),
      x = "Score (0–1)", y = "Density",
      fill = NULL, colour = NULL
    ) +
    theme(
      plot.title    = element_text(face = "bold", size = 11),
      plot.subtitle = element_text(size = 9, colour = "grey40"),
      legend.position = "bottom",
      legend.text     = element_text(size = 8)
    )
}

panels <- list()
for (direction in c("top", "bottom", "both")) {
  panels[[paste0("10pct_", direction)]] <-
    make_threshold_panel(pos_df, thr_10_tp, "10%", direction)
  panels[[paste0("1pct_",  direction)]] <-
    make_threshold_panel(pos_df, thr_1_tp,  "1%",  direction)
}

# Layout: rows = direction (top / bottom / both), cols = threshold (10% | 1%)
gridExtra::grid.arrange(
  panels[["10pct_top"]],    panels[["1pct_top"]],
  panels[["10pct_bottom"]], panels[["1pct_bottom"]],
  panels[["10pct_both"]],   panels[["1pct_both"]],
  ncol = 2,
  top  = grid::textGrob(
    "Component score profiles for top-scoring position subsets",
    gp = grid::gpar(fontface = "bold", fontsize = 13)
  ),
  left = grid::textGrob(
    "Direction:  top                          bottom                          both",
    rot = 90, gp = grid::gpar(fontsize = 9, col = "grey40")
  )
)
```

### Top Positions

```{r top-positions-table}
display_cols <- c("Gene", "Position", "CAAS_score", "biochem_score",
                  "asr_score", "convergence_score", "parallel_score", "change_side")
if (has_pgls) display_cols <- c(display_cols, "p_pgls_top", "sig_p_pgls_top",
                                              "p_pgls_bottom", "sig_p_pgls_bottom")

DT::datatable(
  pos_df %>%
    arrange(desc(CAAS_score)) %>%
    select(all_of(display_cols)) %>%
    mutate(across(where(is.numeric), ~round(., 4))),
  filter    = "top",
  rownames  = FALSE,
  options   = list(pageLength = 25, scrollX = TRUE),
  caption   = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Position-level CAAS scores | Trait: %s", params$traitname)
  )
)
```

---

## Position Characterisation {.tabset #position-characterisation}

This section examines how ranked positions behave in relation to external
signals that do **not** contribute to the `CAAS_score`. PGLS is treated as
an independent validator: we test whether PGLS-significant positions are
over-represented among the highest-scoring positions.

> **Important — PGLS directionality:** `p_pgls` is a **one-tailed** test for
> `beta_top > 0` (amino acid enriched in high-trait / *top* species). Positions
> where `beta_top ≤ 0` receive `p_pgls = 1` by construction, so `sig_p_pgls`
> can only ever be `TRUE` for changes going in the **top** direction. Enrichment
> tests are therefore stratified by `change_side` and a *directionally
> consistent* flag is computed for each position.

`r if (!has_pgls) "> **Note:** PGLS results are not available for this run — characterisation skipped."`

```{r pgls-char-setup, eval=has_pgls}
# Annotate each position that has a PGLS result.
# "directionally consistent" = the direction-appropriate sig flag is TRUE:
#   change_side == "top"    → sig_p_pgls_top    (H1: beta_top > 0)
#   change_side == "bottom" → sig_p_pgls_bottom (H1: beta_top < 0)
#   change_side == "both"   → either sig flag is TRUE

as_logical <- function(x) isTRUE(x) | (!is.na(x) & as.character(x) == "TRUE")

pgls_char <- pos_df %>%
  filter(!is.na(p_pgls_top) | !is.na(p_pgls_bottom)) %>%
  mutate(
    sig_top    = as_logical(sig_p_pgls_top),
    sig_bottom = as_logical(sig_p_pgls_bottom),
    pgls_consistent = case_when(
      change_side == "top"    ~ sig_top,
      change_side == "bottom" ~ sig_bottom,
      change_side == "both"   ~ sig_top | sig_bottom,
      TRUE                    ~ FALSE
    ),
    caas_decile   = dplyr::ntile(CAAS_score, 10),
    in_top_decile = CAAS_score >= thr_10,
    in_top_pct    = CAAS_score >= thr_1
  )
```

### PGLS Enrichment

```{r pgls-char-fisher, eval=has_pgls}
# Fisher's exact test: enrichment of directionally-consistent PGLS positions
# in the top decile / top percentile, stratified by change_side direction.

run_fisher <- function(df, flag_col, label) {
  if (nrow(df) == 0 || sum(df$pgls_consistent, na.rm = TRUE) == 0) {
    return(tibble(
      Threshold = label, Direction = unique(df$change_side)[1],
      N_top = NA_integer_, N_top_consistent = NA_integer_, Pct_top = "—",
      N_rest = NA_integer_, N_rest_consistent = NA_integer_, Pct_rest = "—",
      Odds_ratio = "—", CI_95 = "—", P_value = "—"
    ))
  }
  mat <- table(
    In_top      = df[[flag_col]],
    PGLS_consist = df$pgls_consistent
  )
  # Ensure the table is 2×2 even when one cell is 0
  mat <- matrix(
    c(sum(!df[[flag_col]] & !df$pgls_consistent, na.rm = TRUE),
      sum( df[[flag_col]] & !df$pgls_consistent, na.rm = TRUE),
      sum(!df[[flag_col]] &  df$pgls_consistent, na.rm = TRUE),
      sum( df[[flag_col]] &  df$pgls_consistent, na.rm = TRUE)),
    nrow = 2, dimnames = list(
      In_top        = c("FALSE", "TRUE"),
      PGLS_consist  = c("FALSE", "TRUE")
    )
  )
  ft         <- fisher.test(mat)
  n_top      <- sum( df[[flag_col]], na.rm = TRUE)
  n_top_c    <- sum( df[[flag_col]] & df$pgls_consistent, na.rm = TRUE)
  n_rest     <- sum(!df[[flag_col]], na.rm = TRUE)
  n_rest_c   <- sum(!df[[flag_col]] & df$pgls_consistent, na.rm = TRUE)
  tibble(
    Threshold        = label,
    Direction        = unique(df$change_side)[1],
    N_top            = n_top,
    N_top_consistent = n_top_c,
    Pct_top          = sprintf("%.1f%%", 100 * n_top_c  / max(n_top,  1)),
    N_rest           = n_rest,
    N_rest_consistent = n_rest_c,
    Pct_rest         = sprintf("%.1f%%", 100 * n_rest_c / max(n_rest, 1)),
    Odds_ratio       = sprintf("%.2f", ft$estimate),
    CI_95            = sprintf("[%.2f, %.2f]", ft$conf.int[1], ft$conf.int[2]),
    P_value          = formatC(ft$p.value, format = "e", digits = 2)
  )
}

fisher_results <- bind_rows(lapply(
  c("top", "bottom", "both"),
  function(dir) {
    df_dir <- if (dir == "both") pgls_char else pgls_char %>% filter(change_side == dir)
    df_dir <- df_dir %>% mutate(change_side = dir)
    bind_rows(
      run_fisher(df_dir, "in_top_decile",
                 sprintf("Top %g%% (decile)", top_pct  * 100)),
      run_fisher(df_dir, "in_top_pct",
                 sprintf("Top %g%% (percentile)", top1_pct * 100))
    )
  }
))

knitr::kable(
  fisher_results,
  caption = paste(
    "Fisher's exact test — enrichment of directionally-consistent PGLS",
    "positions in top-ranking groups, stratified by change direction.",
    "Note: sig_p_pgls only captures beta_top > 0 (top direction);",
    "bottom-direction consistency uses beta_top < 0."
  ),
  align = c("l", "l", rep("r", ncol(fisher_results) - 2))
)
```

### PGLS Fraction by Decile

```{r pgls-char-decile-plot, eval=has_pgls, fig.height=6}
# Fraction of directionally-consistent PGLS positions per CAAS decile,
# faceted by change_side direction.

decile_summary <- pgls_char %>%
  group_by(change_side, caas_decile) %>%
  summarise(
    n_total  = n(),
    n_consist = sum(pgls_consistent, na.rm = TRUE),
    frac      = n_consist / n_total,
    .groups   = "drop"
  )

overall_rates <- pgls_char %>%
  group_by(change_side) %>%
  summarise(
    rate = sum(pgls_consistent, na.rm = TRUE) / n(),
    .groups = "drop"
  )

ggplot(decile_summary, aes(x = caas_decile, y = frac)) +
  geom_col(fill = .tableau10[1], alpha = 0.8, width = 0.7) +
  geom_hline(data = overall_rates, aes(yintercept = rate),
             linetype = "dashed", colour = .tableau10[3], linewidth = 0.8) +
  geom_text(data = overall_rates,
            aes(x = 10.3, y = rate,
                label = sprintf("Overall\n%.1f%%", rate * 100)),
            hjust = 0, vjust = 0.5, size = 2.8, colour = .tableau10[3]) +
  scale_x_continuous(breaks = 1:10,
                     labels = c("1\n(low)", as.character(2:9), "10\n(high)")) +
  scale_y_continuous(
    labels = function(x) paste0(round(x * 100), "%"),
    expand = expansion(mult = c(0, 0.20))
  ) +
  facet_wrap(~change_side, ncol = 1) +
  labs(
    title    = "Directionally-Consistent PGLS Positions by CAAS Decile",
    subtitle = "Stratified by change direction (top / bottom / both)",
    x = "CAAS Score Decile",
    y = "Fraction directionally-consistent PGLS"
  ) +
  theme(plot.title = element_text(face = "bold"),
        strip.text = element_text(face = "bold"))
```

### PGLS p-value Distribution

```{r pgls-char-pval-dist, eval=has_pgls, fig.height=6}
# For each position show the directionally-relevant p-value:
#   change_side == "top"    → p_pgls_top
#   change_side == "bottom" → p_pgls_bottom
#   change_side == "both"   → min(p_pgls_top, p_pgls_bottom)
# Faceted by change_side × CAAS rank group.

pgls_pval_plot <- pgls_char %>%
  mutate(
    p_relevant = case_when(
      change_side == "top"    ~ p_pgls_top,
      change_side == "bottom" ~ p_pgls_bottom,
      change_side == "both"   ~ pmin(p_pgls_top, p_pgls_bottom, na.rm = TRUE),
      TRUE                    ~ NA_real_
    ),
    rank_group = case_when(
      in_top_pct    ~ sprintf("Top %g%%", top1_pct * 100),
      in_top_decile ~ sprintf("Top %g%%", top_pct  * 100),
      TRUE          ~ "Rest"
    ),
    rank_group = factor(rank_group,
                        levels = c(sprintf("Top %g%%", top1_pct * 100),
                                   sprintf("Top %g%%", top_pct  * 100),
                                   "Rest"))
  ) %>%
  filter(!is.na(p_relevant))

ggplot(pgls_pval_plot, aes(x = p_relevant, fill = rank_group)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, alpha = 0.7,
                 position = "identity") +
  scale_fill_manual(values = .tableau10[c(4, 3, 1)]) +
  facet_grid(change_side ~ rank_group, scales = "free_y") +
  labs(
    title    = "Directional PGLS p-value Distribution by Direction × CAAS Score Group",
    subtitle = "p-value shown is direction-relevant (top: p_pgls_top; bottom: p_pgls_bottom; both: min of both)",
    x = "Directional p-value", y = "Density", fill = NULL
  ) +
  theme(plot.title      = element_text(face = "bold"),
        legend.position = "none",
        strip.text      = element_text(face = "bold", size = 8))
```

### VEP: PrimateAI by CAAS Decile

`r if (!has_vep_primateai) "> **Note:** PrimateAI-3D scores are not available for this run."`

```{r vep-pai-setup, eval=has_vep_primateai}
# Detect PrimateAI score column (varies by database version)
.pai_score_col <- intersect(
  c("score", "primateai_score", "pathogenicity_score", "PrimateAI_score"),
  names(vep_pai_df)
)[1]
if (is.na(.pai_score_col)) {
  warning("PrimateAI: no recognised score column found — skipping.")
  has_vep_primateai <- FALSE
}
```

```{r vep-pai-join, eval=has_vep_primateai}
pos_vep_pai <- vep_pai_df %>%
  mutate(
    Gene     = sub(":p\\..*", "", transvar_query),
    Position = as.integer(sub(".*:p\\.", "", transvar_query)),
    pai_score = .data[[.pai_score_col]]
  ) %>%
  select(Gene, Position, pai_score) %>%
  # When multiple alt-AA rows exist for the same position, take the max score
  group_by(Gene, Position) %>%
  summarise(pai_score = max(pai_score, na.rm = TRUE), .groups = "drop") %>%
  inner_join(
    pos_df %>%
      select(Gene, Position, CAAS_score) %>%
      mutate(
        caas_decile  = dplyr::ntile(CAAS_score, 10),
        in_top_10pct = CAAS_score >= thr_10,
        in_top_1pct  = CAAS_score >= thr_1
      ),
    by = c("Gene", "Position")
  )
```

```{r vep-pai-violin, eval=has_vep_primateai, fig.height=5}
# PrimateAI score violin + boxplot per CAAS decile.
# Dashed line at 0.803 = PAI3D canonical damaging threshold.
ggplot(pos_vep_pai, aes(x = factor(caas_decile), y = pai_score)) +
  geom_violin(fill = .tableau10[1], alpha = 0.65, trim = TRUE, scale = "width") +
  geom_boxplot(width = 0.14, outlier.size = 0.4, alpha = 0.9,
               fill = "white", colour = "grey30") +
  geom_hline(yintercept = 0.803, linetype = "dashed",
             colour = .tableau10[3], linewidth = 0.8) +
  annotate("text", x = 0.5, y = 0.82, hjust = 0,
           label = "PAI3D damaging (0.803)", colour = .tableau10[3], size = 3.2) +
  scale_x_discrete(labels = c("1\n(low)", as.character(2:9), "10\n(high)")) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title    = "PrimateAI-3D Score by CAAS Decile",
    subtitle = sprintf("%d positions with PAI3D coverage (of %d total)",
                       nrow(pos_vep_pai), nrow(pos_df)),
    x = "CAAS Score Decile", y = "PrimateAI-3D Score"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r vep-pai-scatter, eval=has_vep_primateai, fig.height=5}
# Scatter: CAAS_score vs PrimateAI score, coloured by decile.
ggplot(pos_vep_pai, aes(x = CAAS_score, y = pai_score, colour = caas_decile)) +
  geom_point(alpha = 0.45, size = 0.9) +
  geom_hline(yintercept = 0.803, linetype = "dashed",
             colour = "grey40", linewidth = 0.6) +
  geom_vline(xintercept = thr_10, linetype = "dashed",
             colour = .tableau10[3], linewidth = 0.6) +
  geom_vline(xintercept = thr_1, linetype = "dotted",
             colour = .tableau10[4], linewidth = 0.6) +
  scale_colour_viridis_c(name = "CAAS\nDecile", option = "plasma") +
  labs(
    title    = "CAAS Score vs PrimateAI-3D Score",
    subtitle = "Dashed: CAAS top-10% and PAI3D damaging thresholds",
    x = "CAAS Score", y = "PrimateAI-3D Score"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r vep-pai-fisher, eval=has_vep_primateai}
# Fisher's exact: enrichment of PAI3D-damaging positions in top 1% and top 10%.
.pai_thresh <- 0.803

run_pai_fisher <- function(df, flag_col, label) {
  df <- df %>% mutate(damaging = pai_score >= .pai_thresh)
  n_top  <- sum(df[[flag_col]], na.rm = TRUE)
  n_dam_top  <- sum(df[[flag_col]] & df$damaging, na.rm = TRUE)
  n_rest <- sum(!df[[flag_col]], na.rm = TRUE)
  n_dam_rest <- sum(!df[[flag_col]] & df$damaging, na.rm = TRUE)
  mat <- matrix(c(n_top - n_dam_top, n_rest - n_dam_rest, n_dam_top, n_dam_rest),
                nrow = 2,
                dimnames = list(In_top = c("FALSE","TRUE"), Damaging = c("FALSE","TRUE")))
  ft <- fisher.test(mat)
  tibble(
    Threshold        = label,
    N_top            = n_top,
    N_damaging_top   = n_dam_top,
    Pct_damaging_top = sprintf("%.1f%%", 100 * n_dam_top  / max(n_top,  1)),
    N_rest           = n_rest,
    N_damaging_rest  = n_dam_rest,
    Pct_damaging_rest = sprintf("%.1f%%", 100 * n_dam_rest / max(n_rest, 1)),
    Odds_ratio       = sprintf("%.2f", ft$estimate),
    CI_95            = sprintf("[%.2f, %.2f]", ft$conf.int[1], ft$conf.int[2]),
    P_value          = formatC(ft$p.value, format = "e", digits = 2)
  )
}

pai_fisher <- bind_rows(
  run_pai_fisher(pos_vep_pai, "in_top_10pct",
                 sprintf("Top %g%% CAAS (decile)", top_pct  * 100)),
  run_pai_fisher(pos_vep_pai, "in_top_1pct",
                 sprintf("Top %g%% CAAS (percentile)", top1_pct * 100))
)

knitr::kable(
  pai_fisher,
  caption = paste(
    "Fisher's exact test — enrichment of PAI3D-damaging positions (score ≥ 0.803)",
    "in top-ranking CAAS positions."
  ),
  align = c("l", rep("r", ncol(pai_fisher) - 1))
)
```

### VEP: TransVar Annotations

`r if (!has_vep_transvar) "> **Note:** TransVar annotations are not available for this run."`

```{r vep-tv-setup, eval=has_vep_transvar}
# TransVar output has no header (--noheader flag).  Read with explicit column names.
# The stub generates a header row for testing; drop it if the first cell is "input".
vep_tv_raw <- read_tsv(
  params$vep_transvar_file,
  col_names = c("tv_input","tv_transcript","tv_gene","tv_strand",
                "tv_coordinates","tv_region","tv_info"),
  show_col_types = FALSE
)
if (nrow(vep_tv_raw) > 0 && identical(vep_tv_raw$tv_input[1], "input")) {
  vep_tv_raw <- vep_tv_raw[-1, ]
}

# Parse region into broad functional categories
vep_tv_parsed <- vep_tv_raw %>%
  filter(!is.na(tv_input), tv_input != ".") %>%
  mutate(
    # Normalise region labels from TransVar to tidy categories
    region_cat = case_when(
      grepl("CDS|exon",            tv_region, ignore.case = TRUE) ~ "CDS",
      grepl("5'UTR|5UTR|C5'",      tv_region, ignore.case = TRUE) ~ "5'UTR",
      grepl("3'UTR|3UTR|C3'",      tv_region, ignore.case = TRUE) ~ "3'UTR",
      grepl("intron|intronic",      tv_region, ignore.case = TRUE) ~ "Intronic",
      grepl("splice",               tv_region, ignore.case = TRUE) ~ "Splice site",
      grepl("intergenic",           tv_region, ignore.case = TRUE) ~ "Intergenic",
      is.na(tv_region) | tv_region == "." ~ "Unmapped",
      TRUE ~ tv_region
    ),
    # Extract mutation type from info field if present (TransVar panno)
    mutation_type = case_when(
      grepl("synonymous",  tv_info, ignore.case = TRUE) ~ "Synonymous",
      grepl("nonsense|stop_gained|\\*",
                           tv_info, ignore.case = TRUE) ~ "Nonsense",
      grepl("missense|nonsynonymous",
                           tv_info, ignore.case = TRUE) ~ "Missense",
      grepl("splice",      tv_info, ignore.case = TRUE) ~ "Splice",
      TRUE ~ NA_character_
    )
  )

# Join to pos_df on Gene × Position
pos_vep_tv <- pos_df %>%
  select(Gene, Position, CAAS_score) %>%
  mutate(
    tv_query    = paste0(Gene, ":p.", Position),
    caas_decile = dplyr::ntile(CAAS_score, 10)
  ) %>%
  inner_join(
    vep_tv_parsed %>% select(tv_input, region_cat, mutation_type),
    by = c("tv_query" = "tv_input")
  ) %>%
  # One TransVar row per query; keep first if duplicates exist
  distinct(Gene, Position, .keep_all = TRUE)
```

```{r vep-tv-region-bar, eval=has_vep_transvar, fig.height=5}
# Stacked bar: proportion of region categories per CAAS decile
tv_decile_region <- pos_vep_tv %>%
  count(caas_decile, region_cat) %>%
  group_by(caas_decile) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup()

region_levels <- tv_decile_region %>%
  count(region_cat, wt = n, sort = TRUE) %>%
  pull(region_cat)

tv_decile_region <- tv_decile_region %>%
  mutate(region_cat = factor(region_cat, levels = region_levels))

ggplot(tv_decile_region, aes(x = factor(caas_decile), y = pct, fill = region_cat)) +
  geom_col(width = 0.75) +
  scale_x_discrete(labels = c("1\n(low)", as.character(2:9), "10\n(high)")) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = .tableau10, name = "Genomic Region") +
  labs(
    title    = "TransVar Region Annotation by CAAS Decile",
    subtitle = sprintf("%d positions with TransVar coverage", nrow(pos_vep_tv)),
    x = "CAAS Score Decile", y = "Proportion"
  ) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "right")
```

```{r vep-tv-muttype-bar, eval=has_vep_transvar && any(!is.na(pos_vep_tv$mutation_type)), fig.height=5}
# Stacked bar: mutation type distribution per CAAS decile (when available from info field)
tv_decile_mut <- pos_vep_tv %>%
  filter(!is.na(mutation_type)) %>%
  count(caas_decile, mutation_type) %>%
  group_by(caas_decile) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup()

if (nrow(tv_decile_mut) > 0) {
  ggplot(tv_decile_mut, aes(x = factor(caas_decile), y = pct, fill = mutation_type)) +
    geom_col(width = 0.75) +
    scale_x_discrete(labels = c("1\n(low)", as.character(2:9), "10\n(high)")) +
    scale_y_continuous(labels = scales::percent_format()) +
    scale_fill_manual(values = .tableau10[c(1,3,4,6,8)], name = "Mutation Type") +
    labs(
      title    = "Functional Mutation Type by CAAS Decile",
      subtitle = "Derived from TransVar info field annotation",
      x = "CAAS Score Decile", y = "Proportion"
    ) +
    theme(plot.title = element_text(face = "bold"),
          legend.position = "right")
}
```

---

## Gene-Level Scores {.tabset}

### Gene CAAS Score

```{r gene-caas-dist, fig.height=4.5}
g_thr_10 <- quantile(gene_df$gene_caas_score, 1 - params$gene_top_pct, na.rm = TRUE)
g_thr_1  <- quantile(gene_df$gene_caas_score, 1 - params$gene_top1_pct, na.rm = TRUE)

ggplot(gene_df, aes(x = gene_caas_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = .tableau10[1], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[2]) +
  geom_vline(xintercept = g_thr_10, linetype = "dashed",
             colour = .tableau10[3], linewidth = 0.7) +
  geom_vline(xintercept = g_thr_1,  linetype = "dotted",
             colour = .tableau10[4], linewidth = 0.7) +
  labs(title = "Gene CAAS Score Distribution",
       subtitle = "90th percentile of position-level CAAS scores per gene",
       x = "gene_caas_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

```{r gene-rand-dist, eval=has_rand, fig.height=4.5}
ggplot(gene_df %>% filter(!is.na(gene_rand_score)), aes(x = gene_rand_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = .tableau10[5], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[6]) +
  labs(title = "Gene Randomization Score Distribution",
       subtitle = "Scheme-weighted deciles of accumulation randomization p-values",
       x = "gene_rand_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

```{r gene-rer-dist, eval=has_rer, fig.height=4.5}
ggplot(gene_df %>% filter(!is.na(gene_rer_score)), aes(x = gene_rer_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = .tableau10[7], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[8]) +
  labs(title = "Gene RERConverge Score Distribution",
       subtitle = "Deciles of evolutionary rate correlation p-values",
       x = "gene_rer_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

```{r gene-fade-dist, eval=has_fade, fig.height=4.5}
ggplot(gene_df %>% filter(!is.na(gene_fade_score)), aes(x = gene_fade_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = .tableau10[9], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[10]) +
  labs(title = "Gene FADE Score Distribution",
       subtitle = "Deciles of max directional selection Bayes Factor per gene",
       x = "gene_fade_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### Gene Composite Score

```{r gene-composite-dist, fig.height=5}
gc_thr_10 <- quantile(gene_df$gene_composite_score, 1 - params$gene_top_pct, na.rm = TRUE)
gc_thr_1  <- quantile(gene_df$gene_composite_score, 1 - params$gene_top1_pct, na.rm = TRUE)

ggplot(gene_df, aes(x = gene_composite_score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = .tableau10[1], alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = .tableau10[2]) +
  geom_vline(xintercept = gc_thr_10, linetype = "dashed",
             colour = .tableau10[3], linewidth = 0.7) +
  geom_vline(xintercept = gc_thr_1,  linetype = "dotted",
             colour = .tableau10[4], linewidth = 0.7) +
  annotate("text", x = gc_thr_10, y = Inf,
           label = sprintf("Top %g%%", params$gene_top_pct * 100),
           vjust = 2, hjust = -0.1, colour = .tableau10[3], size = 3.5) +
  annotate("text", x = gc_thr_1,  y = Inf,
           label = sprintf("Top %g%%", params$gene_top1_pct * 100),
           vjust = 3.5, hjust = -0.1, colour = .tableau10[4], size = 3.5) +
  labs(title = "Gene Composite Score Distribution",
       subtitle = sprintf("Unweighted mean of available gene scores | n = %s genes",
                          format(nrow(gene_df), big.mark = ",")),
       x = "gene_composite_score", y = "Density") +
  theme(plot.title = element_text(face = "bold"))
```

### Gene Threshold Profiles

```{r gene-threshold-profiles, fig.height=14, fig.width=12}
# Same idea as position-level threshold profiles but for gene-level scores.
# For each top-N% × direction bucket, show overlaid densities of gene sub-scores.

gene_score_cols_tp <- "gene_caas_score"
if (has_rand) gene_score_cols_tp <- c(gene_score_cols_tp, "gene_rand_score")
if (has_rer)  gene_score_cols_tp <- c(gene_score_cols_tp, "gene_rer_score")
if (has_fade) gene_score_cols_tp <- c(gene_score_cols_tp, "gene_fade_score")

# Direction per gene is inherited from its top-scoring position
gene_direction_tp <- pos_df %>%
  group_by(Gene) %>%
  slice_max(CAAS_score, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(Gene, gene_direction = change_side)

gene_df_dir <- gene_df %>% left_join(gene_direction_tp, by = "Gene")

gc_thr_10_tp <- quantile(gene_df$gene_composite_score, 1 - params$gene_top_pct,  na.rm = TRUE)
gc_thr_1_tp  <- quantile(gene_df$gene_composite_score, 1 - params$gene_top1_pct, na.rm = TRUE)

make_gene_threshold_panel <- function(df, threshold, pct_label, direction) {
  subset_df <- if (direction == "both") {
    df %>% filter(gene_composite_score >= threshold)
  } else {
    df %>% filter(gene_composite_score >= threshold, gene_direction == direction)
  }

  if (nrow(subset_df) == 0 || length(gene_score_cols_tp) < 2) {
    return(
      ggplot() +
        annotate("text", x = 0.5, y = 0.5,
                 label = ifelse(nrow(subset_df) == 0, "No genes", "Single score only"),
                 size = 5, colour = "grey50") +
        labs(title = sprintf("Top %s — %s", pct_label, direction),
             x = "Score (0–1)", y = "Density") +
        xlim(0, 1) + theme_minimal()
    )
  }

  long_df <- subset_df %>%
    select(all_of(gene_score_cols_tp)) %>%
    pivot_longer(everything(), names_to = "Score_type", values_to = "Score") %>%
    filter(!is.na(Score)) %>%
    mutate(Score_type = factor(Score_type, levels = gene_score_cols_tp))

  ggplot(long_df, aes(x = Score, fill = Score_type, colour = Score_type)) +
    geom_density(alpha = 0.35, linewidth = 0.6) +
    scale_fill_manual(values   = .tableau10[seq_along(gene_score_cols_tp)],
                      labels   = sub("gene_", "", sub("_score", "", gene_score_cols_tp))) +
    scale_colour_manual(values = .tableau10[seq_along(gene_score_cols_tp)],
                        labels = sub("gene_", "", sub("_score", "", gene_score_cols_tp))) +
    scale_x_continuous(limits = c(0, 1), expand = c(0.01, 0)) +
    labs(
      title    = sprintf("Top %s — %s", pct_label, direction),
      subtitle = sprintf("n = %d genes", nrow(subset_df)),
      x = "Score (0–1)", y = "Density",
      fill = NULL, colour = NULL
    ) +
    theme(
      plot.title      = element_text(face = "bold", size = 11),
      plot.subtitle   = element_text(size = 9, colour = "grey40"),
      legend.position = "bottom",
      legend.text     = element_text(size = 8)
    )
}

gene_panels <- list()
for (direction in c("top", "bottom", "both")) {
  gene_panels[[paste0("10pct_", direction)]] <-
    make_gene_threshold_panel(gene_df_dir, gc_thr_10_tp, "10%", direction)
  gene_panels[[paste0("1pct_",  direction)]] <-
    make_gene_threshold_panel(gene_df_dir, gc_thr_1_tp,  "1%",  direction)
}

gridExtra::grid.arrange(
  gene_panels[["10pct_top"]],    gene_panels[["1pct_top"]],
  gene_panels[["10pct_bottom"]], gene_panels[["1pct_bottom"]],
  gene_panels[["10pct_both"]],   gene_panels[["1pct_both"]],
  ncol = 2,
  top  = grid::textGrob(
    "Gene score profiles for top-scoring gene subsets (by gene_composite_score)",
    gp = grid::gpar(fontface = "bold", fontsize = 13)
  ),
  left = grid::textGrob(
    "Direction:  top                          bottom                          both",
    rot = 90, gp = grid::gpar(fontsize = 9, col = "grey40")
  )
)
```

### Correlation Analysis

```{r correlation-scatter, eval=(nrow(corr_df) > 0), fig.height=7}
# Build scatter matrix of gene-level scores
score_cols <- "gene_caas_score"
if (has_rand) score_cols <- c(score_cols, "gene_rand_score")
if (has_rer)  score_cols <- c(score_cols, "gene_rer_score")
if (has_fade) score_cols <- c(score_cols, "gene_fade_score")

if (length(score_cols) >= 2) {
  pairs_df <- gene_df %>%
    select(all_of(score_cols)) %>%
    drop_na()

  # Create pairs plot
  pairs_long <- expand.grid(
    x_var = score_cols, y_var = score_cols,
    stringsAsFactors = FALSE
  ) %>%
    filter(x_var != y_var)

  plots <- list()
  for (i in seq_len(nrow(pairs_long))) {
    xv <- pairs_long$x_var[i]
    yv <- pairs_long$y_var[i]
    r_info <- corr_df %>%
      filter((score_a == xv & score_b == yv) | (score_a == yv & score_b == xv))
    label <- if (nrow(r_info) > 0) {
      sprintf("r=%.2f / rho=%.2f", r_info$pearson_r[1], r_info$spearman_r[1])
    } else ""

    plots[[length(plots) + 1]] <- ggplot(pairs_df, aes(x = .data[[xv]], y = .data[[yv]])) +
      geom_point(alpha = 0.3, size = 0.8, colour = .tableau10[1]) +
      geom_smooth(method = "lm", se = FALSE, colour = .tableau10[4], linewidth = 0.6) +
      annotate("text", x = Inf, y = -Inf, label = label,
               hjust = 1.1, vjust = -0.5, size = 3, colour = .tableau10[4]) +
      labs(x = sub("gene_", "", sub("_score", "", xv)),
           y = sub("gene_", "", sub("_score", "", yv))) +
      theme(axis.title = element_text(size = 9))
  }

  if (length(plots) > 0) {
    gridExtra::grid.arrange(grobs = plots,
                            ncol = length(score_cols) - 1,
                            top = grid::textGrob("Gene Score Correlations",
                                                 gp = grid::gpar(fontface = "bold", fontsize = 14)))
  }
}
```

```{r correlation-table, eval=(nrow(corr_df) > 0)}
DT::datatable(
  corr_df %>%
    mutate(across(where(is.numeric), ~round(., 4))),
  rownames = FALSE,
  options  = list(pageLength = 10, dom = "t"),
  caption  = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "Pairwise correlations between gene-level scores"
  )
)
```

### Top Genes

```{r top-genes-table}
gene_display_cols <- c("Gene", "n_positions", "gene_caas_score")
if (has_rand) gene_display_cols <- c(gene_display_cols, "gene_rand_score")
if (has_rer)  gene_display_cols <- c(gene_display_cols, "gene_rer_score")
if (has_fade) gene_display_cols <- c(gene_display_cols, "gene_fade_score")
gene_display_cols <- c(gene_display_cols, "gene_composite_score")

DT::datatable(
  gene_df %>%
    arrange(desc(gene_composite_score)) %>%
    select(all_of(gene_display_cols)) %>%
    mutate(across(where(is.numeric) & !matches("n_positions"), ~round(., 4))),
  filter    = "top",
  rownames  = FALSE,
  options   = list(pageLength = 25, scrollX = TRUE),
  caption   = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Gene-level scores | Trait: %s", params$traitname)
  )
)
```

---

## Gene-Level Characterisation {.tabset}

This section accumulates position-level PGLS and VEP signals at the gene level,
testing whether genes with high composite scores tend to concentrate
directionally-consistent PGLS hits or pathogenic VEP annotations.

### Gene PGLS Accumulation

`r if (!has_pgls) "> **Note:** PGLS results are not available — gene-level PGLS accumulation skipped."`

```{r gene-pgls-accum, eval=has_pgls, fig.height=5}
# Per-gene: count of directionally-consistent PGLS positions and fraction.
gene_pgls <- pos_df %>%
  filter(!is.na(p_pgls_top) | !is.na(p_pgls_bottom)) %>%
  mutate(
    sig_top    = isTRUE(sig_p_pgls_top)    | (!is.na(sig_p_pgls_top)    & as.character(sig_p_pgls_top)    == "TRUE"),
    sig_bottom = isTRUE(sig_p_pgls_bottom) | (!is.na(sig_p_pgls_bottom) & as.character(sig_p_pgls_bottom) == "TRUE"),
    pgls_consistent = case_when(
      change_side == "top"    ~ sig_top,
      change_side == "bottom" ~ sig_bottom,
      change_side == "both"   ~ sig_top | sig_bottom,
      TRUE                    ~ FALSE
    )
  ) %>%
  group_by(Gene) %>%
  summarise(
    n_pgls_positions  = n(),
    n_pgls_consistent = sum(pgls_consistent, na.rm = TRUE),
    frac_pgls_consist = n_pgls_consistent / n_pgls_positions,
    .groups = "drop"
  ) %>%
  inner_join(gene_df %>% select(Gene, gene_caas_score, gene_composite_score,
                                n_positions),
             by = "Gene")

# Scatter: gene_caas_score vs fraction PGLS consistent, bubble size = n_positions
ggplot(gene_pgls, aes(x = gene_caas_score, y = frac_pgls_consist,
                      size = n_positions, colour = gene_composite_score)) +
  geom_point(alpha = 0.65) +
  ggrepel::geom_text_repel(
    data = gene_pgls %>% slice_max(frac_pgls_consist, n = 15, with_ties = FALSE),
    aes(label = Gene), size = 2.8, max.overlaps = 20,
    show.legend = FALSE
  ) +
  geom_vline(xintercept = g_thr_10, linetype = "dashed",
             colour = .tableau10[3], linewidth = 0.7) +
  scale_colour_viridis_c(name = "Gene\nComposite", option = "plasma") +
  scale_size_continuous(name = "N positions", range = c(1, 8)) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title    = "Gene CAAS Score vs Fraction of Directionally-Consistent PGLS Positions",
    subtitle = "Labels: top-15 genes by PGLS consistency; dashed = CAAS top 10% threshold",
    x = "Gene CAAS Score", y = "Fraction PGLS-consistent positions"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r gene-pgls-top-bar, eval=has_pgls, fig.height=6}
# Top 30 genes by PGLS consistency, coloured by gene_caas_score
top_pgls_genes <- gene_pgls %>%
  filter(n_pgls_consistent > 0) %>%
  slice_max(frac_pgls_consist, n = 30, with_ties = FALSE) %>%
  arrange(frac_pgls_consist) %>%
  mutate(Gene = factor(Gene, levels = Gene))

ggplot(top_pgls_genes, aes(x = frac_pgls_consist, y = Gene,
                            fill = gene_caas_score)) +
  geom_col() +
  scale_fill_viridis_c(name = "Gene CAAS\nScore", option = "plasma") +
  scale_x_continuous(labels = scales::percent_format(),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Top 30 Genes by Fraction of PGLS-Consistent Positions",
    x = "Fraction PGLS-consistent", y = NULL
  ) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.y = element_text(size = 8))
```

### Gene VEP Accumulation

`r if (!has_vep_primateai && !has_vep_transvar) "> **Note:** VEP data (PrimateAI and/or TransVar) are not available — gene-level VEP accumulation skipped."`

```{r gene-vep-pai-accum, eval=has_vep_primateai, fig.height=5}
# Per-gene mean PrimateAI score and fraction of PAI3D-damaging positions
gene_vep_pai <- pos_vep_pai %>%
  group_by(Gene) %>%
  summarise(
    n_pai_positions  = n(),
    mean_pai_score   = mean(pai_score, na.rm = TRUE),
    frac_pai_damaging = sum(pai_score >= 0.803, na.rm = TRUE) / n(),
    .groups = "drop"
  ) %>%
  inner_join(gene_df %>% select(Gene, gene_caas_score, gene_composite_score,
                                n_positions),
             by = "Gene")

ggplot(gene_vep_pai, aes(x = gene_caas_score, y = mean_pai_score,
                          size = n_pai_positions, colour = frac_pai_damaging)) +
  geom_point(alpha = 0.65) +
  ggrepel::geom_text_repel(
    data = gene_vep_pai %>% slice_max(mean_pai_score, n = 15, with_ties = FALSE),
    aes(label = Gene), size = 2.8, max.overlaps = 20,
    show.legend = FALSE
  ) +
  geom_vline(xintercept = g_thr_10, linetype = "dashed",
             colour = .tableau10[3], linewidth = 0.7) +
  scale_colour_viridis_c(name = "Fraction\nPAI3D-damaging",
                          option = "inferno", labels = scales::percent_format()) +
  scale_size_continuous(name = "N positions", range = c(1, 8)) +
  labs(
    title    = "Gene CAAS Score vs Mean PrimateAI-3D Score",
    subtitle = "Labels: top-15 genes by mean PAI3D score; dashed = CAAS top 10% threshold",
    x = "Gene CAAS Score", y = "Mean PrimateAI-3D Score"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r gene-vep-pai-top-bar, eval=has_vep_primateai, fig.height=6}
# Top 30 genes by fraction PAI3D-damaging, coloured by gene_caas_score
top_pai_genes <- gene_vep_pai %>%
  filter(n_pai_positions >= 2) %>%
  slice_max(frac_pai_damaging, n = 30, with_ties = FALSE) %>%
  arrange(frac_pai_damaging) %>%
  mutate(Gene = factor(Gene, levels = Gene))

ggplot(top_pai_genes, aes(x = frac_pai_damaging, y = Gene,
                           fill = gene_caas_score)) +
  geom_col() +
  scale_fill_viridis_c(name = "Gene CAAS\nScore", option = "plasma") +
  scale_x_continuous(labels = scales::percent_format(),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Top 30 Genes by Fraction of PAI3D-Damaging CAAS Positions",
    subtitle = "Genes with ≥ 2 positions with PAI3D coverage",
    x = "Fraction PAI3D-damaging (score ≥ 0.803)", y = NULL
  ) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.y = element_text(size = 8))
```

```{r gene-vep-tv-accum, eval=has_vep_transvar, fig.height=5}
# Per-gene region annotation summary from TransVar
gene_vep_tv <- pos_vep_tv %>%
  group_by(Gene) %>%
  summarise(
    n_tv_positions = n(),
    n_cds          = sum(region_cat == "CDS", na.rm = TRUE),
    frac_cds       = n_cds / n_tv_positions,
    .groups        = "drop"
  ) %>%
  inner_join(gene_df %>% select(Gene, gene_caas_score, n_positions),
             by = "Gene")

ggplot(gene_vep_tv, aes(x = gene_caas_score, y = frac_cds,
                         size = n_tv_positions)) +
  geom_point(alpha = 0.6, colour = .tableau10[2]) +
  ggrepel::geom_text_repel(
    data = gene_vep_tv %>%
      filter(frac_cds < 0.8) %>%
      slice_max(n_tv_positions, n = 15, with_ties = FALSE),
    aes(label = Gene), size = 2.8, max.overlaps = 15,
    show.legend = FALSE
  ) +
  geom_hline(yintercept = 1.0, linetype = "dashed",
             colour = "grey50", linewidth = 0.5) +
  scale_size_continuous(name = "N positions", range = c(1, 8)) +
  scale_y_continuous(labels = scales::percent_format(),
                     limits = c(0, NA)) +
  labs(
    title    = "TransVar CDS Mapping Rate by Gene CAAS Score",
    subtitle = "Labels: genes with < 80% CDS coverage and ≥ 15 scored positions",
    x = "Gene CAAS Score", y = "Fraction of positions mapping to CDS"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

---

## Genomic Accumulation {.tabset}

`r if (!has_genomic) "> **Note:** Genomic coordinates (gene_ensembl_file) are not available for this run — genomic accumulation skipped."`

```{r genomic-setup, eval=has_genomic}
# Join gene_df with genomic coordinates, then annotate with PGLS and VEP
gene_genomic <- gene_df %>%
  select(Gene, gene_caas_score, gene_composite_score, n_positions) %>%
  inner_join(genomic_raw, by = "Gene") %>%
  # Standardise chromosome labels (remove "chr" prefix if present)
  mutate(
    chr_label = sub("^chr", "", chr),
    # Order chromosomes numerically then X, Y, MT
    chr_num   = suppressWarnings(as.integer(chr_label)),
    chr_order = dplyr::case_when(
      !is.na(chr_num)       ~ chr_num,
      chr_label == "X"      ~ 23L,
      chr_label == "Y"      ~ 24L,
      chr_label %in% c("MT","M") ~ 25L,
      TRUE                  ~ 26L
    )
  ) %>%
  arrange(chr_order, gene_mid)

# Optionally add PGLS fraction
if (has_pgls && exists("gene_pgls")) {
  gene_genomic <- gene_genomic %>%
    left_join(gene_pgls %>% select(Gene, frac_pgls_consist, n_pgls_consistent),
              by = "Gene")
} else {
  gene_genomic$frac_pgls_consist  <- NA_real_
  gene_genomic$n_pgls_consistent  <- NA_integer_
}

# Optionally add mean PAI score
if (has_vep_primateai && exists("gene_vep_pai")) {
  gene_genomic <- gene_genomic %>%
    left_join(gene_vep_pai %>% select(Gene, mean_pai_score, frac_pai_damaging),
              by = "Gene")
} else {
  gene_genomic$mean_pai_score   <- NA_real_
  gene_genomic$frac_pai_damaging <- NA_real_
}

# Build a cumulative x-axis offset per chromosome for a stitched Manhattan view
chr_offsets <- gene_genomic %>%
  group_by(chr_order, chr_label) %>%
  summarise(chr_max = max(gene_end, na.rm = TRUE), .groups = "drop") %>%
  arrange(chr_order) %>%
  mutate(offset = cumsum(lag(chr_max, default = 0L)) + (chr_order - 1L) * 5e6L)

gene_genomic <- gene_genomic %>%
  left_join(chr_offsets %>% select(chr_order, offset, chr_label), by = c("chr_order", "chr_label")) %>%
  mutate(manhattan_x = gene_mid + offset)

chr_label_positions <- chr_offsets %>%
  left_join(gene_genomic %>% group_by(chr_order) %>%
              summarise(mid_x = mean(manhattan_x, na.rm = TRUE), .groups = "drop"),
            by = "chr_order")
```

### CAAS Score Manhattan

```{r genomic-manhattan-caas, eval=has_genomic, fig.height=5, fig.width=14}
# Manhattan-style plot: x = genomic position (stitched), y = gene_caas_score
# Alternating chromosome colours for readability.
gene_genomic <- gene_genomic %>%
  mutate(chr_even = chr_order %% 2 == 0)

ggplot(gene_genomic,
       aes(x = manhattan_x, y = gene_caas_score,
           colour = chr_even, size = n_positions)) +
  geom_point(alpha = 0.65) +
  geom_hline(yintercept = g_thr_10, linetype = "dashed",
             colour = "grey40", linewidth = 0.6) +
  ggrepel::geom_text_repel(
    data = gene_genomic %>% slice_max(gene_caas_score, n = 20, with_ties = FALSE),
    aes(label = Gene), size = 2.2, max.overlaps = 15,
    show.legend = FALSE, colour = "black"
  ) +
  scale_colour_manual(values = c(.tableau10[1], .tableau10[2]), guide = "none") +
  scale_size_continuous(name = "N positions", range = c(0.5, 5)) +
  scale_x_continuous(
    breaks = chr_label_positions$mid_x,
    labels = chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  labs(
    title    = "Gene CAAS Score — Manhattan View",
    subtitle = sprintf("%d genes with genomic coordinates | dashed = top %g%% threshold",
                       nrow(gene_genomic), params$gene_top_pct * 100),
    x = "Chromosome", y = "Gene CAAS Score"
  ) +
  theme(
    plot.title      = element_text(face = "bold"),
    axis.text.x     = element_text(size = 7, angle = 0),
    panel.grid.major.x = element_blank()
  )
```

### Composite Score Manhattan

```{r genomic-manhattan-composite, eval=has_genomic, fig.height=5, fig.width=14}
ggplot(gene_genomic,
       aes(x = manhattan_x, y = gene_composite_score,
           colour = chr_even, size = n_positions)) +
  geom_point(alpha = 0.65) +
  geom_hline(yintercept = gc_thr_10, linetype = "dashed",
             colour = "grey40", linewidth = 0.6) +
  ggrepel::geom_text_repel(
    data = gene_genomic %>% slice_max(gene_composite_score, n = 20, with_ties = FALSE),
    aes(label = Gene), size = 2.2, max.overlaps = 15,
    show.legend = FALSE, colour = "black"
  ) +
  scale_colour_manual(values = c(.tableau10[1], .tableau10[2]), guide = "none") +
  scale_size_continuous(name = "N positions", range = c(0.5, 5)) +
  scale_x_continuous(
    breaks = chr_label_positions$mid_x,
    labels = chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  labs(
    title    = "Gene Composite Score — Manhattan View",
    subtitle = sprintf("dashed = top %g%% threshold", params$gene_top_pct * 100),
    x = "Chromosome", y = "Gene Composite Score"
  ) +
  theme(
    plot.title      = element_text(face = "bold"),
    axis.text.x     = element_text(size = 7, angle = 0),
    panel.grid.major.x = element_blank()
  )
```

### PGLS Signal Manhattan

`r if (!has_pgls) "> **Note:** PGLS results not available — PGLS Manhattan skipped."`

```{r genomic-manhattan-pgls, eval=has_genomic && has_pgls && any(!is.na(gene_genomic$frac_pgls_consist)), fig.height=5, fig.width=14}
ggplot(gene_genomic %>% filter(!is.na(frac_pgls_consist)),
       aes(x = manhattan_x, y = gene_caas_score,
           size  = frac_pgls_consist,
           colour = frac_pgls_consist)) +
  geom_point(alpha = 0.70) +
  geom_hline(yintercept = g_thr_10, linetype = "dashed",
             colour = "grey60", linewidth = 0.5) +
  ggrepel::geom_text_repel(
    data = gene_genomic %>%
      filter(!is.na(frac_pgls_consist), frac_pgls_consist > 0) %>%
      slice_max(frac_pgls_consist, n = 20, with_ties = FALSE),
    aes(label = Gene), size = 2.2, max.overlaps = 12,
    show.legend = FALSE, colour = "black"
  ) +
  scale_colour_viridis_c(name = "PGLS\nfraction", option = "plasma",
                          labels = scales::percent_format()) +
  scale_size_continuous(name = "PGLS\nfraction",
                         labels = scales::percent_format(), range = c(0.5, 6)) +
  scale_x_continuous(
    breaks = chr_label_positions$mid_x,
    labels = chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  guides(colour = guide_legend(), size = guide_legend()) +
  labs(
    title    = "Gene CAAS Score + PGLS Consistency — Manhattan View",
    subtitle = "Bubble size and colour: fraction of directionally-consistent PGLS positions",
    x = "Chromosome", y = "Gene CAAS Score"
  ) +
  theme(
    plot.title      = element_text(face = "bold"),
    axis.text.x     = element_text(size = 7),
    panel.grid.major.x = element_blank()
  )
```

### PAI3D Signal Manhattan

`r if (!has_vep_primateai) "> **Note:** PrimateAI-3D scores not available — PAI3D Manhattan skipped."`

```{r genomic-manhattan-pai, eval=has_genomic && has_vep_primateai && any(!is.na(gene_genomic$mean_pai_score)), fig.height=5, fig.width=14}
ggplot(gene_genomic %>% filter(!is.na(mean_pai_score)),
       aes(x = manhattan_x, y = gene_caas_score,
           size   = mean_pai_score,
           colour = mean_pai_score)) +
  geom_point(alpha = 0.70) +
  geom_hline(yintercept = g_thr_10, linetype = "dashed",
             colour = "grey60", linewidth = 0.5) +
  ggrepel::geom_text_repel(
    data = gene_genomic %>%
      filter(!is.na(mean_pai_score)) %>%
      slice_max(mean_pai_score, n = 20, with_ties = FALSE),
    aes(label = Gene), size = 2.2, max.overlaps = 12,
    show.legend = FALSE, colour = "black"
  ) +
  scale_colour_viridis_c(name = "Mean\nPAI3D", option = "inferno") +
  scale_size_continuous(name = "Mean\nPAI3D", range = c(0.5, 6)) +
  scale_x_continuous(
    breaks = chr_label_positions$mid_x,
    labels = chr_label_positions$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  guides(colour = guide_legend(), size = guide_legend()) +
  labs(
    title    = "Gene CAAS Score + Mean PrimateAI-3D — Manhattan View",
    subtitle = "Bubble size and colour: mean PrimateAI-3D score across CAAS positions",
    x = "Chromosome", y = "Gene CAAS Score"
  ) +
  theme(
    plot.title      = element_text(face = "bold"),
    axis.text.x     = element_text(size = 7),
    panel.grid.major.x = element_blank()
  )
```

---

## Summary

```{r summary-stats}
n_components <- 4L  # biochem, asr, convergence, parallel (PGLS is characterisation only)

n_gene_components <- 1L  # gene_caas always present
if (has_rand) n_gene_components <- n_gene_components + 1L
if (has_rer)  n_gene_components <- n_gene_components + 1L
if (has_fade) n_gene_components <- n_gene_components + 1L

summary_tbl <- tibble(
  Metric = c(
    "Positions scored", "Genes scored",
    "Position-level components", "Gene-level components",
    "CAAS Score (median)", "CAAS Score (top 10% threshold)",
    "Gene Composite (median)", "Gene Composite (top 10% threshold)",
    "PGLS available", "FADE available",
    "Randomization available", "RERConverge available",
    "VEP TransVar available", "VEP PrimateAI-3D available",
    "Genomic coordinates available"
  ),
  Value = c(
    format(nrow(pos_df), big.mark = ","),
    format(nrow(gene_df), big.mark = ","),
    as.character(n_components),
    as.character(n_gene_components),
    sprintf("%.4f", median(pos_df$CAAS_score, na.rm = TRUE)),
    sprintf("%.4f", thr_10),
    sprintf("%.4f", median(gene_df$gene_composite_score, na.rm = TRUE)),
    sprintf("%.4f", gc_thr_10),
    ifelse(has_pgls, "Yes", "No"),
    ifelse(has_fade, "Yes", "No"),
    ifelse(has_rand, "Yes", "No"),
    ifelse(has_rer,  "Yes", "No"),
    ifelse(has_vep_transvar,  "Yes", "No"),
    ifelse(has_vep_primateai, "Yes", "No"),
    ifelse(has_genomic,       "Yes", "No")
  )
)

knitr::kable(summary_tbl, align = c("l", "r"))
```

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>
