---
title: "CAAS Scoring — Top/Bottom Overlap Report"
subtitle: "`r paste0(params$traitname, ' — top vs bottom phenotype directions')`"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "PhyloPhere pipeline"
output:
  html_document:
    theme: flatly
    syntax_highlighting: tango
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 6
params:
  pos_scores_top_file: NULL
  pos_scores_bottom_file: NULL
  gene_scores_top_file: NULL
  gene_scores_bottom_file: NULL
  traitname: "unknown_trait"
  top_pct: 0.10
  top5_pct: 0.05
  top1_pct: 0.01
  gene_top_pct: 0.10
  gene_top5_pct: 0.05
  gene_top1_pct: 0.01
  genomic_info_file: NULL
  window_size_bp: 1000000
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  out.width = "100%"
)

suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(DT)
  library(scales)
  library(ggrepel)
})

# Tableau_10 colour palette for pipeline-wide consistency
if (requireNamespace("paletteer", quietly = TRUE)) {
  .tableau10 <- as.character(paletteer::paletteer_d("ggthemes::Tableau_10"))
} else {
  .tableau10 <- c(
    "#4E79A7", "#F28E2B", "#E15759", "#76B7B2", "#59A14F",
    "#EDC948", "#B07AA1", "#FF9DA7", "#9C755F", "#BAB0AC"
  )
}
theme_set(theme_minimal(base_size = 13))

.file_ok <- function(p) !is.null(p) && !is.na(p) && nchar(p) > 0 && file.exists(p)
```

```{r load-data}
pos_top  <- read_tsv(params$pos_scores_top_file,    show_col_types = FALSE)
pos_bot  <- read_tsv(params$pos_scores_bottom_file, show_col_types = FALSE)
gene_top <- read_tsv(params$gene_scores_top_file,   show_col_types = FALSE)
gene_bot <- read_tsv(params$gene_scores_bottom_file, show_col_types = FALSE)

has_genomic <- .file_ok(params$genomic_info_file)

if (has_genomic) {
  genomic_raw <- read_tsv(params$genomic_info_file,
                          col_names = c("Gene", "chr", "gene_start", "gene_end", "gene_length"),
                          show_col_types = FALSE)
  if (tolower(genomic_raw$Gene[1]) == "gene") genomic_raw <- genomic_raw[-1, ]
  genomic_raw <- genomic_raw %>%
    mutate(
      gene_start  = as.integer(gene_start),
      gene_end    = as.integer(gene_end),
      gene_length = as.integer(gene_length),
      gene_mid    = as.integer((gene_start + gene_end) / 2L)
    )
} else {
  genomic_raw <- NULL
}

# ── Position-level thresholds (computed on each direction's full score vector) ──
pos_thr_top  <- list(
  t10 = quantile(pos_top$CAAS_score, 1 - params$top_pct,  na.rm = TRUE),
  t5  = quantile(pos_top$CAAS_score, 1 - params$top5_pct, na.rm = TRUE),
  t1  = quantile(pos_top$CAAS_score, 1 - params$top1_pct, na.rm = TRUE)
)
pos_thr_bot  <- list(
  t10 = quantile(pos_bot$CAAS_score, 1 - params$top_pct,  na.rm = TRUE),
  t5  = quantile(pos_bot$CAAS_score, 1 - params$top5_pct, na.rm = TRUE),
  t1  = quantile(pos_bot$CAAS_score, 1 - params$top1_pct, na.rm = TRUE)
)

# ── Gene-level thresholds ──────────────────────────────────────────────────────
gene_thr_top <- list(
  t10 = quantile(gene_top$gene_composite_score, 1 - params$gene_top_pct,  na.rm = TRUE),
  t5  = quantile(gene_top$gene_composite_score, 1 - params$gene_top5_pct, na.rm = TRUE),
  t1  = quantile(gene_top$gene_composite_score, 1 - params$gene_top1_pct, na.rm = TRUE)
)
gene_thr_bot <- list(
  t10 = quantile(gene_bot$gene_composite_score, 1 - params$gene_top_pct,  na.rm = TRUE),
  t5  = quantile(gene_bot$gene_composite_score, 1 - params$gene_top5_pct, na.rm = TRUE),
  t1  = quantile(gene_bot$gene_composite_score, 1 - params$gene_top1_pct, na.rm = TRUE)
)

# ── Available gene score types ─────────────────────────────────────────────────
gene_score_cols <- c("gene_caas_score", "gene_composite_score")
for (col in c("gene_rand_score", "gene_rer_score", "gene_fade_score", "gene_molerate_score")) {
  if (col %in% names(gene_top) && col %in% names(gene_bot) &&
      any(!is.na(gene_top[[col]])) && any(!is.na(gene_bot[[col]]))) {
    gene_score_cols <- c(gene_score_cols, col)
  }
}
```

---

## Introduction

This report examines the **overlap between the top-phenotype and bottom-phenotype
CAAS scoring results** for trait **`r params$traitname`**.

The two per-direction reports score positions and genes independently — each run
filters positions for one phenotypic extreme before computing gene-level
aggregates.  Positions or genes that score highly in *both* directions are
particularly interesting: they represent loci where convergent amino acid changes
associate with *both* ends of the trait distribution, suggesting a potential
pleiotropic or bidirectional signal.

The analysis covers three resolution levels:

1. **Position-level** — Gene × Alignment-position pairs, scored by `CAAS_score`.
2. **Gene-level** — Per-gene scores, examined individually for each available
   score type (`r paste(gene_score_cols, collapse = ", ")`).
3. **Genomic window** `r if (has_genomic) paste0("(", params$window_size_bp / 1e6, " Mb windows)")` —
   `r if (!has_genomic) "Not available (no genomic coordinates supplied)." else "Genomic regions enriched in high-scoring genes in both directions."`

---

## Position-Level Overlap {.tabset}

Positions are identified by their `Gene:Position` key.  For each threshold
(top 10%, 5%, 1% by `CAAS_score`), we extract the high-scoring positions from
each direction and compute the intersection.

```{r pos-overlap-setup}
# Create unique position keys
pos_top  <- pos_top  %>% mutate(.pos_key = paste0(Gene, ":", Position))
pos_bot  <- pos_bot  %>% mutate(.pos_key = paste0(Gene, ":", Position))

compute_pos_overlap <- function(thr_top, thr_bot, pct_label) {
  set_t <- pos_top %>% filter(CAAS_score >= thr_top) %>% pull(.pos_key)
  set_b <- pos_bot %>% filter(CAAS_score >= thr_bot) %>% pull(.pos_key)
  ov    <- intersect(set_t, set_b)
  jaccard <- if ((length(set_t) + length(set_b) - length(ov)) > 0) {
    length(ov) / (length(set_t) + length(set_b) - length(ov))
  } else NA_real_
  list(
    label   = pct_label,
    n_top   = length(set_t),
    n_bot   = length(set_b),
    n_ovlp  = length(ov),
    jaccard = jaccard,
    keys    = ov
  )
}

pos_ovlp_10 <- compute_pos_overlap(pos_thr_top$t10, pos_thr_bot$t10, "Top 10%")
pos_ovlp_5  <- compute_pos_overlap(pos_thr_top$t5,  pos_thr_bot$t5,  "Top 5%")
pos_ovlp_1  <- compute_pos_overlap(pos_thr_top$t1,  pos_thr_bot$t1,  "Top 1%")

pos_summary_tbl <- tibble(
  Threshold      = c("Top 10%", "Top 5%", "Top 1%"),
  `N top`        = c(pos_ovlp_10$n_top,  pos_ovlp_5$n_top,  pos_ovlp_1$n_top),
  `N bottom`     = c(pos_ovlp_10$n_bot,  pos_ovlp_5$n_bot,  pos_ovlp_1$n_bot),
  `N overlap`    = c(pos_ovlp_10$n_ovlp, pos_ovlp_5$n_ovlp, pos_ovlp_1$n_ovlp),
  `Jaccard index`= round(c(pos_ovlp_10$jaccard, pos_ovlp_5$jaccard, pos_ovlp_1$jaccard), 4)
)
```

### Summary

```{r pos-overlap-summary-plot, fig.height=4}
pos_bar_df <- pos_summary_tbl %>%
  mutate(
    top_only    = `N top`     - `N overlap`,
    bottom_only = `N bottom`  - `N overlap`,
    overlap     = `N overlap`
  ) %>%
  select(Threshold, top_only, bottom_only, overlap) %>%
  pivot_longer(c(top_only, bottom_only, overlap),
               names_to = "Category", values_to = "Count") %>%
  mutate(Category = factor(Category,
                           levels = c("top_only", "overlap", "bottom_only"),
                           labels = c("Top only", "Overlap", "Bottom only")))

ggplot(pos_bar_df, aes(x = Threshold, y = Count, fill = Category)) +
  geom_col(position = "stack", width = 0.55) +
  geom_text(aes(label = Count), position = position_stack(vjust = 0.5),
            size = 3.2, colour = "white", fontface = "bold") +
  scale_fill_manual(
    values = c("Top only"    = .tableau10[1],
               "Overlap"     = .tableau10[3],
               "Bottom only" = .tableau10[2]),
    name = NULL
  ) +
  labs(
    title    = "Position-Level Top/Bottom Overlap",
    subtitle = "Stacked bar: unique top-only, shared (overlap), unique bottom-only positions per threshold",
    x = NULL, y = "Number of positions"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

```{r pos-overlap-summary-table}
knitr::kable(pos_summary_tbl, align = c("l", "r", "r", "r", "r"),
             caption = "Position-level overlap statistics (Jaccard = |A ∩ B| / |A ∪ B|)")
```

### Overlap at Top 10%

```{r pos-overlap-10pct}
if (length(pos_ovlp_10$keys) > 0) {
  ovlp_df <- tibble(.pos_key = pos_ovlp_10$keys) %>%
    left_join(pos_top %>% select(.pos_key, Gene, Position,
                                  CAAS_score_top    = CAAS_score,
                                  biochem_top       = biochem_score,
                                  asr_top           = asr_score,
                                  convergence_top   = convergence_score,
                                  parallel_top      = parallel_score),
              by = ".pos_key") %>%
    left_join(pos_bot %>% select(.pos_key,
                                  CAAS_score_bottom  = CAAS_score,
                                  biochem_bottom     = biochem_score,
                                  asr_bottom         = asr_score,
                                  convergence_bottom = convergence_score,
                                  parallel_bottom    = parallel_score),
              by = ".pos_key") %>%
    select(-.pos_key) %>%
    mutate(mean_CAAS = rowMeans(cbind(CAAS_score_top, CAAS_score_bottom), na.rm = TRUE)) %>%
    arrange(desc(mean_CAAS)) %>%
    mutate(across(where(is.numeric), ~round(., 4)))

  DT::datatable(
    ovlp_df,
    rownames = FALSE,
    filter   = "top",
    options  = list(pageLength = 25, scrollX = TRUE),
    caption  = htmltools::tags$caption(
      style = "caption-side: top; text-align: left; font-weight: bold;",
      sprintf("Positions in top 10%% of BOTH directions (%d positions)", nrow(ovlp_df))
    )
  )
} else {
  cat("> No positions overlap at the top 10% threshold in both directions.")
}
```

### Overlap at Top 5%

```{r pos-overlap-5pct}
if (length(pos_ovlp_5$keys) > 0) {
  ovlp_df <- tibble(.pos_key = pos_ovlp_5$keys) %>%
    left_join(pos_top %>% select(.pos_key, Gene, Position,
                                  CAAS_score_top  = CAAS_score,
                                  biochem_top     = biochem_score,
                                  asr_top         = asr_score),
              by = ".pos_key") %>%
    left_join(pos_bot %>% select(.pos_key,
                                  CAAS_score_bottom = CAAS_score,
                                  biochem_bottom    = biochem_score,
                                  asr_bottom        = asr_score),
              by = ".pos_key") %>%
    select(-.pos_key) %>%
    mutate(mean_CAAS = rowMeans(cbind(CAAS_score_top, CAAS_score_bottom), na.rm = TRUE)) %>%
    arrange(desc(mean_CAAS)) %>%
    mutate(across(where(is.numeric), ~round(., 4)))

  DT::datatable(
    ovlp_df,
    rownames = FALSE,
    filter   = "top",
    options  = list(pageLength = 25, scrollX = TRUE),
    caption  = htmltools::tags$caption(
      style = "caption-side: top; text-align: left; font-weight: bold;",
      sprintf("Positions in top 5%% of BOTH directions (%d positions)", nrow(ovlp_df))
    )
  )
} else {
  cat("> No positions overlap at the top 5% threshold in both directions.")
}
```

### Overlap at Top 1%

```{r pos-overlap-1pct}
if (length(pos_ovlp_1$keys) > 0) {
  ovlp_df <- tibble(.pos_key = pos_ovlp_1$keys) %>%
    left_join(pos_top %>% select(.pos_key, Gene, Position,
                                  CAAS_score_top  = CAAS_score,
                                  biochem_top     = biochem_score),
              by = ".pos_key") %>%
    left_join(pos_bot %>% select(.pos_key,
                                  CAAS_score_bottom = CAAS_score,
                                  biochem_bottom    = biochem_score),
              by = ".pos_key") %>%
    select(-.pos_key) %>%
    mutate(mean_CAAS = rowMeans(cbind(CAAS_score_top, CAAS_score_bottom), na.rm = TRUE)) %>%
    arrange(desc(mean_CAAS)) %>%
    mutate(across(where(is.numeric), ~round(., 4)))

  DT::datatable(
    ovlp_df,
    rownames = FALSE,
    options  = list(pageLength = 25, scrollX = TRUE),
    caption  = htmltools::tags$caption(
      style = "caption-side: top; text-align: left; font-weight: bold;",
      sprintf("Positions in top 1%% of BOTH directions (%d positions)", nrow(ovlp_df))
    )
  )
} else {
  cat("> No positions overlap at the top 1% threshold in both directions.")
}
```

---

## Gene-Level Overlap {.tabset}

For each available gene score type, we extract the high-scoring genes from each
direction and compute the intersection.  Only score types present and non-missing
in both direction runs are included.

```{r gene-overlap-setup}
# Generic function: given a score column name and thresholds, compute
# top/bottom gene sets and their overlap at three cutoff levels.
compute_gene_overlap <- function(score_col, pct_label, thr_top, thr_bot) {
  vals_top <- gene_top[[score_col]]
  vals_bot <- gene_bot[[score_col]]

  thr_t <- quantile(vals_top, 1 - thr_top, na.rm = TRUE)
  thr_b <- quantile(vals_bot, 1 - thr_bot, na.rm = TRUE)

  set_t <- gene_top %>% filter(.data[[score_col]] >= thr_t) %>% pull(Gene)
  set_b <- gene_bot %>% filter(.data[[score_col]] >= thr_b) %>% pull(Gene)
  ov    <- intersect(set_t, set_b)

  jaccard <- if ((length(set_t) + length(set_b) - length(ov)) > 0) {
    length(ov) / (length(set_t) + length(set_b) - length(ov))
  } else NA_real_

  list(
    pct_label = pct_label,
    n_top     = length(set_t),
    n_bot     = length(set_b),
    n_ovlp    = length(ov),
    jaccard   = jaccard,
    genes     = ov
  )
}

# Helper to make an overlap bar plot for a single score type
make_gene_ovlp_bar <- function(score_col) {
  thresholds <- list(
    list(pct = params$gene_top_pct,  label = "Top 10%"),
    list(pct = params$gene_top5_pct, label = "Top 5%"),
    list(pct = params$gene_top1_pct, label = "Top 1%")
  )
  rows <- lapply(thresholds, function(t) {
    r <- compute_gene_overlap(score_col, t$label, t$pct, t$pct)
    tibble(
      Threshold   = t$label,
      top_only    = r$n_top  - r$n_ovlp,
      bottom_only = r$n_bot  - r$n_ovlp,
      overlap     = r$n_ovlp,
      jaccard     = r$jaccard
    )
  })
  bar_df <- bind_rows(rows) %>%
    pivot_longer(c(top_only, bottom_only, overlap),
                 names_to = "Category", values_to = "Count") %>%
    mutate(Category = factor(Category,
                             levels = c("top_only", "overlap", "bottom_only"),
                             labels = c("Top only", "Overlap", "Bottom only")))
  list(bar_df = bar_df, summary = bind_rows(lapply(thresholds, function(t) {
    r <- compute_gene_overlap(score_col, t$label, t$pct, t$pct)
    tibble(Threshold = t$label, `N top` = r$n_top, `N bottom` = r$n_bot,
           `N overlap` = r$n_ovlp, `Jaccard` = round(r$jaccard, 4))
  })))
}

# Helper to render overlap detail table for one threshold
make_gene_ovlp_table <- function(score_col, pct_top, pct_bot, caption_text) {
  r <- compute_gene_overlap(score_col, "", pct_top, pct_bot)
  if (length(r$genes) == 0) {
    cat(sprintf("> No genes overlap at this threshold for %s.", score_col))
    return(invisible(NULL))
  }
  tbl <- tibble(Gene = r$genes) %>%
    left_join(gene_top %>% select(Gene, score_top   = !!score_col), by = "Gene") %>%
    left_join(gene_bot %>% select(Gene, score_bottom = !!score_col), by = "Gene") %>%
    mutate(mean_score = rowMeans(cbind(score_top, score_bottom), na.rm = TRUE)) %>%
    arrange(desc(mean_score)) %>%
    mutate(across(where(is.numeric), ~round(., 4)))

  DT::datatable(
    tbl, rownames = FALSE, filter = "top",
    options = list(pageLength = 25, scrollX = TRUE),
    caption = htmltools::tags$caption(
      style = "caption-side: top; text-align: left; font-weight: bold;",
      caption_text
    )
  )
}
```

### CAAS

```{r gene-overlap-caas-score, fig.height=4}
sc <- "gene_caas_score"
ovlp_data <- make_gene_ovlp_bar(sc)
ggplot(ovlp_data$bar_df, aes(x = Threshold, y = Count, fill = Category)) +
  geom_col(position = "stack", width = 0.55) +
  geom_text(aes(label = Count), position = position_stack(vjust = 0.5),
            size = 3.2, colour = "white", fontface = "bold") +
  scale_fill_manual(
    values = c("Top only"    = .tableau10[1],
               "Overlap"     = .tableau10[3],
               "Bottom only" = .tableau10[2]),
    name = NULL
  ) +
  labs(title = "Gene CAAS Score — Top/Bottom Overlap",
       x = NULL, y = "Number of genes") +
  theme(plot.title = element_text(face = "bold"))
knitr::kable(ovlp_data$summary, caption = "Gene CAAS score overlap statistics")
```

```{r gene-overlap-caas-table-10}
make_gene_ovlp_table("gene_caas_score", params$gene_top_pct, params$gene_top_pct,
                     sprintf("Genes in top %g%% by gene_caas_score in BOTH directions",
                             params$gene_top_pct * 100))
```

### Composite

```{r gene-overlap-composite-score, fig.height=4}
sc <- "gene_composite_score"
ovlp_data <- make_gene_ovlp_bar(sc)
ggplot(ovlp_data$bar_df, aes(x = Threshold, y = Count, fill = Category)) +
  geom_col(position = "stack", width = 0.55) +
  geom_text(aes(label = Count), position = position_stack(vjust = 0.5),
            size = 3.2, colour = "white", fontface = "bold") +
  scale_fill_manual(
    values = c("Top only"    = .tableau10[1],
               "Overlap"     = .tableau10[3],
               "Bottom only" = .tableau10[2]),
    name = NULL
  ) +
  labs(title = "Gene Composite Score — Top/Bottom Overlap",
       x = NULL, y = "Number of genes") +
  theme(plot.title = element_text(face = "bold"))
knitr::kable(ovlp_data$summary, caption = "Gene composite score overlap statistics")
```

```{r gene-overlap-composite-table-10}
make_gene_ovlp_table("gene_composite_score", params$gene_top_pct, params$gene_top_pct,
                     sprintf("Genes in top %g%% by gene_composite_score in BOTH directions",
                             params$gene_top_pct * 100))
```

### Accumulation

`r if (!("gene_rand_score" %in% gene_score_cols)) "> **Note:** Accumulation score overlap is not available for both directions in this run."`

```{r gene-overlap-rand-score, eval="gene_rand_score" %in% gene_score_cols, fig.height=4}
sc <- "gene_rand_score"
ovlp_data <- make_gene_ovlp_bar(sc)
ggplot(ovlp_data$bar_df, aes(x = Threshold, y = Count, fill = Category)) +
  geom_col(position = "stack", width = 0.55) +
  geom_text(aes(label = Count), position = position_stack(vjust = 0.5),
            size = 3.2, colour = "white", fontface = "bold") +
  scale_fill_manual(
    values = c("Top only"    = .tableau10[1],
               "Overlap"     = .tableau10[3],
               "Bottom only" = .tableau10[2]),
    name = NULL
  ) +
  labs(title = "Gene Randomization Score — Top/Bottom Overlap",
       x = NULL, y = "Number of genes") +
  theme(plot.title = element_text(face = "bold"))
knitr::kable(ovlp_data$summary, caption = "Gene randomization score overlap statistics")
```

```{r gene-overlap-rand-table, eval="gene_rand_score" %in% gene_score_cols}
make_gene_ovlp_table("gene_rand_score", params$gene_top_pct, params$gene_top_pct,
                     sprintf("Genes in top %g%% by gene_rand_score in BOTH directions",
                             params$gene_top_pct * 100))
```

### RERConverge

`r if (!("gene_rer_score" %in% gene_score_cols)) "> **Note:** RERConverge overlap is not available for both directions in this run."`

```{r gene-overlap-rer-score, eval="gene_rer_score" %in% gene_score_cols, fig.height=4}
sc <- "gene_rer_score"
ovlp_data <- make_gene_ovlp_bar(sc)
ggplot(ovlp_data$bar_df, aes(x = Threshold, y = Count, fill = Category)) +
  geom_col(position = "stack", width = 0.55) +
  geom_text(aes(label = Count), position = position_stack(vjust = 0.5),
            size = 3.2, colour = "white", fontface = "bold") +
  scale_fill_manual(
    values = c("Top only"    = .tableau10[1],
               "Overlap"     = .tableau10[3],
               "Bottom only" = .tableau10[2]),
    name = NULL
  ) +
  labs(title = "Gene RERConverge Score — Top/Bottom Overlap",
       x = NULL, y = "Number of genes") +
  theme(plot.title = element_text(face = "bold"))
knitr::kable(ovlp_data$summary, caption = "Gene RER score overlap statistics")
```

```{r gene-overlap-rer-table, eval="gene_rer_score" %in% gene_score_cols}
make_gene_ovlp_table("gene_rer_score", params$gene_top_pct, params$gene_top_pct,
                     sprintf("Genes in top %g%% by gene_rer_score in BOTH directions",
                             params$gene_top_pct * 100))
```

### FADE

`r if (!("gene_fade_score" %in% gene_score_cols)) "> **Note:** FADE overlap is not available for both directions in this run."`

```{r gene-overlap-fade-score, eval="gene_fade_score" %in% gene_score_cols, fig.height=4}
sc <- "gene_fade_score"
ovlp_data <- make_gene_ovlp_bar(sc)
ggplot(ovlp_data$bar_df, aes(x = Threshold, y = Count, fill = Category)) +
  geom_col(position = "stack", width = 0.55) +
  geom_text(aes(label = Count), position = position_stack(vjust = 0.5),
            size = 3.2, colour = "white", fontface = "bold") +
  scale_fill_manual(
    values = c("Top only"    = .tableau10[1],
               "Overlap"     = .tableau10[3],
               "Bottom only" = .tableau10[2]),
    name = NULL
  ) +
  labs(title = "Gene FADE Score — Top/Bottom Overlap",
       x = NULL, y = "Number of genes") +
  theme(plot.title = element_text(face = "bold"))
knitr::kable(ovlp_data$summary, caption = "Gene FADE score overlap statistics")
```

```{r gene-overlap-fade-table, eval="gene_fade_score" %in% gene_score_cols}
make_gene_ovlp_table("gene_fade_score", params$gene_top_pct, params$gene_top_pct,
                     sprintf("Genes in top %g%% by gene_fade_score in BOTH directions",
                             params$gene_top_pct * 100))
```

### MoleRate

`r if (!("gene_molerate_score" %in% gene_score_cols)) "> **Note:** MoleRate overlap is not available for both directions in this run."`

```{r gene-overlap-molerate-score, eval="gene_molerate_score" %in% gene_score_cols, fig.height=4}
sc <- "gene_molerate_score"
ovlp_data <- make_gene_ovlp_bar(sc)
ggplot(ovlp_data$bar_df, aes(x = Threshold, y = Count, fill = Category)) +
  geom_col(position = "stack", width = 0.55) +
  geom_text(aes(label = Count), position = position_stack(vjust = 0.5),
            size = 3.2, colour = "white", fontface = "bold") +
  scale_fill_manual(
    values = c("Top only"    = .tableau10[1],
               "Overlap"     = .tableau10[3],
               "Bottom only" = .tableau10[2]),
    name = NULL
  ) +
  labs(title = "Gene MoleRate Score — Top/Bottom Overlap",
       x = NULL, y = "Number of genes") +
  theme(plot.title = element_text(face = "bold"))
knitr::kable(ovlp_data$summary, caption = "Gene MoleRate score overlap statistics")
```

```{r gene-overlap-molerate-table, eval="gene_molerate_score" %in% gene_score_cols}
make_gene_ovlp_table("gene_molerate_score", params$gene_top_pct, params$gene_top_pct,
                     sprintf("Genes in top %g%% by gene_molerate_score in BOTH directions",
                             params$gene_top_pct * 100))
```

---

## Genomic Window Overlap {.tabset}

`r if (!has_genomic) "> **Note:** Genomic coordinates not available — genomic window overlap skipped."`

For each genomic window (size: `r params$window_size_bp / 1e6` Mb), we identify
windows that appear in the top gene lists for **both** directions.  The
Manhattan-style plot colours windows by their overlap status:
- **Top only** — enriched in the top-phenotype direction
- **Bottom only** — enriched in the bottom-phenotype direction
- **Overlap** — enriched in both directions (highest interest)
- **Neither** — not enriched in either direction

```{r window-overlap-setup, eval=has_genomic}
win_size <- params$window_size_bp

# Build per-direction gene sets at top-10% composite threshold
compute_window_membership <- function(gene_df_dir, score_col, thr_pct, gi_raw) {
  thr <- quantile(gene_df_dir[[score_col]], 1 - thr_pct, na.rm = TRUE)
  top_genes <- gene_df_dir %>% filter(.data[[score_col]] >= thr) %>% pull(Gene)

  gi_raw %>%
    filter(Gene %in% top_genes) %>%
    mutate(
      chr_label = sub("^chr", "", chr),
      chr_num   = suppressWarnings(as.integer(chr_label)),
      chr_order = dplyr::case_when(
        !is.na(chr_num)              ~ chr_num,
        chr_label == "X"             ~ 23L,
        chr_label == "Y"             ~ 24L,
        chr_label %in% c("MT","M")   ~ 25L,
        TRUE                         ~ 26L
      ),
      win_id = floor(gene_mid / win_size)
    ) %>%
    distinct(chr_order, chr_label, win_id)
}

# Windows in top-10% composite for each direction
win_top <- compute_window_membership(gene_top, "gene_composite_score",
                                     params$gene_top_pct, genomic_raw)
win_bot <- compute_window_membership(gene_bot, "gene_composite_score",
                                     params$gene_top_pct, genomic_raw)

# All windows that contain at least one gene (from either direction)
all_gene_wins <- bind_rows(
    gene_top %>% select(Gene) %>% mutate(dir = "top"),
    gene_bot %>% select(Gene) %>% mutate(dir = "bottom")
  ) %>%
  inner_join(genomic_raw %>%
               mutate(
                 chr_label = sub("^chr", "", chr),
                 chr_num   = suppressWarnings(as.integer(chr_label)),
                 chr_order = dplyr::case_when(
                   !is.na(chr_num)              ~ chr_num,
                   chr_label == "X"             ~ 23L,
                   chr_label == "Y"             ~ 24L,
                   chr_label %in% c("MT","M")   ~ 25L,
                   TRUE                         ~ 26L
                 ),
                 win_id    = floor(gene_mid / win_size),
                 win_start = win_id * win_size,
                 win_end   = win_start + win_size - 1L,
                 win_mid   = win_start + win_size / 2L
               ),
             by = "Gene") %>%
  distinct(chr_order, chr_label, win_id, win_start, win_end, win_mid)

# Tag each window by overlap status
all_gene_wins <- all_gene_wins %>%
  mutate(
    in_top = paste(chr_order, win_id) %in% paste(win_top$chr_order, win_top$win_id),
    in_bot = paste(chr_order, win_id) %in% paste(win_bot$chr_order, win_bot$win_id),
    status = dplyr::case_when(
      in_top  & in_bot  ~ "Overlap",
      in_top  & !in_bot ~ "Top only",
      !in_top & in_bot  ~ "Bottom only",
      TRUE              ~ "Neither"
    ),
    status = factor(status, levels = c("Overlap", "Top only", "Bottom only", "Neither"))
  )

# Build stitched x-axis
chr_offsets_w <- all_gene_wins %>%
  group_by(chr_order, chr_label) %>%
  summarise(chr_max = max(win_end, na.rm = TRUE), .groups = "drop") %>%
  arrange(chr_order) %>%
  mutate(offset = cumsum(lag(chr_max, default = 0L)) + (chr_order - 1L) * 5e6L)

all_gene_wins <- all_gene_wins %>%
  left_join(chr_offsets_w %>% select(chr_order, offset), by = "chr_order") %>%
  mutate(win_x = win_mid + offset)

chr_label_positions_w <- chr_offsets_w %>%
  left_join(
    all_gene_wins %>% group_by(chr_order) %>%
      summarise(mid_x = mean(win_x, na.rm = TRUE), .groups = "drop"),
    by = "chr_order"
  )

n_ovlp_wins <- sum(all_gene_wins$status == "Overlap")
```

### Composite Score Windows

```{r window-overlap-manhattan, eval=has_genomic, fig.height=5, fig.width=14}
status_colours <- c(
  "Overlap"     = .tableau10[3],
  "Top only"    = .tableau10[1],
  "Bottom only" = .tableau10[2],
  "Neither"     = "grey80"
)

# Label the overlap windows with top-5 by win_id (just use position label)
ovlp_wins_label <- all_gene_wins %>%
  filter(status == "Overlap") %>%
  arrange(chr_order, win_id)

ggplot(all_gene_wins, aes(x = win_x, y = 0.5, colour = status, size = status)) +
  geom_point(alpha = 0.80) +
  ggrepel::geom_text_repel(
    data = ovlp_wins_label,
    aes(label = sprintf("chr%s:%s-%sMb",
                        chr_label,
                        round(win_start / 1e6, 1),
                        round(win_end   / 1e6, 1))),
    size = 2.4, max.overlaps = 20, colour = .tableau10[3],
    show.legend = FALSE, nudge_y = 0.15
  ) +
  scale_colour_manual(values = status_colours, name = "Window status") +
  scale_size_manual(
    values = c("Overlap" = 3.5, "Top only" = 2, "Bottom only" = 2, "Neither" = 1),
    guide  = "none"
  ) +
  scale_x_continuous(
    breaks = chr_label_positions_w$mid_x,
    labels = chr_label_positions_w$chr_label,
    expand = expansion(mult = 0.01)
  ) +
  scale_y_continuous(limits = c(0, 1), breaks = NULL) +
  labs(
    title    = sprintf("Genomic Window Overlap — %s Mb windows (Gene Composite Score, top %g%%)",
                       params$window_size_bp / 1e6, params$gene_top_pct * 100),
    subtitle = sprintf("%d overlap windows (present in both top and bottom top-%g%%) | %d top-only | %d bottom-only",
                       n_ovlp_wins,
                       params$gene_top_pct * 100,
                       sum(all_gene_wins$status == "Top only"),
                       sum(all_gene_wins$status == "Bottom only")),
    x = "Chromosome", y = NULL
  ) +
  theme(
    plot.title         = element_text(face = "bold"),
    axis.text.x        = element_text(size = 7),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_blank()
  )
```

### Overlap Region Table

```{r window-overlap-table, eval=has_genomic}
if (n_ovlp_wins > 0) {
  ovlp_win_regions <- all_gene_wins %>%
    filter(status == "Overlap") %>%
    select(chr_label, win_start, win_end, chr_order, win_id) %>%
    inner_join(
      genomic_raw %>%
        mutate(
          chr_label = sub("^chr", "", chr),
          chr_num   = suppressWarnings(as.integer(chr_label)),
          chr_order = dplyr::case_when(
            !is.na(chr_num)              ~ chr_num,
            chr_label == "X"             ~ 23L,
            chr_label == "Y"             ~ 24L,
            chr_label %in% c("MT","M")   ~ 25L,
            TRUE                         ~ 26L
          ),
          win_id = floor(gene_mid / win_size)
        ) %>%
        select(Gene, chr_order, win_id),
      by = c("chr_order", "win_id")
    ) %>%
    left_join(gene_top %>% select(Gene, composite_top    = gene_composite_score,
                                       caas_top          = gene_caas_score),
              by = "Gene") %>%
    left_join(gene_bot %>% select(Gene, composite_bottom = gene_composite_score,
                                       caas_bottom       = gene_caas_score),
              by = "Gene") %>%
    group_by(chr_order, chr_label, win_start, win_end) %>%
    summarise(
      n_genes = n_distinct(Gene),
      mean_composite_top = mean(composite_top, na.rm = TRUE),
      mean_composite_bottom = mean(composite_bottom, na.rm = TRUE),
      max_caas_top = max(caas_top, na.rm = TRUE),
      max_caas_bottom = max(caas_bottom, na.rm = TRUE),
      genes = paste(sort(unique(Gene)), collapse = ", "),
      .groups = "drop"
    ) %>%
    arrange(chr_order, win_start, desc(mean_composite_top)) %>%
    mutate(
      window = sprintf("chr%s:%s-%s Mb", chr_label,
                       round(win_start / 1e6, 1), round(win_end / 1e6, 1)),
      across(where(is.numeric), ~round(., 4))
    ) %>%
    select(window, n_genes, mean_composite_top, mean_composite_bottom,
           max_caas_top, max_caas_bottom, genes)

  DT::datatable(
    ovlp_win_regions,
    rownames = FALSE,
    filter   = "top",
    options  = list(pageLength = 30, scrollX = TRUE),
    caption  = htmltools::tags$caption(
      style = "caption-side: top; text-align: left; font-weight: bold;",
      sprintf("Overlap regions enriched in top %g%% composite score for BOTH directions",
              params$gene_top_pct * 100)
    )
  )
} else {
  cat("> No genomic windows are enriched in both top and bottom directions at this threshold.")
}
```

---

## Summary

```{r overlap-summary}
# Cross-score summary table
summary_rows <- lapply(gene_score_cols, function(sc) {
  r10 <- compute_gene_overlap(sc, "Top 10%", params$gene_top_pct,  params$gene_top_pct)
  r5  <- compute_gene_overlap(sc, "Top 5%",  params$gene_top5_pct, params$gene_top5_pct)
  r1  <- compute_gene_overlap(sc, "Top 1%",  params$gene_top1_pct, params$gene_top1_pct)
  tibble(
    `Score type`    = sc,
    `N top (10%)`   = r10$n_top,
    `N bot (10%)`   = r10$n_bot,
    `Overlap (10%)` = r10$n_ovlp,
    `Jaccard (10%)` = round(r10$jaccard, 4),
    `Overlap (5%)`  = r5$n_ovlp,
    `Jaccard (5%)`  = round(r5$jaccard, 4),
    `Overlap (1%)`  = r1$n_ovlp,
    `Jaccard (1%)`  = round(r1$jaccard, 4)
  )
})

knitr::kable(
  bind_rows(summary_rows),
  align = c("l", rep("r", 8)),
  caption = "Gene-level overlap summary across all available score types"
)
```

---

<details>
<summary>Session Info</summary>
```{r session-info}
sessionInfo()
```
</details>
