---
title: "CAAS Post-Processing and Declustering Analysis"
author: "PhyloPhere Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: bootstrap
    highlight: tango
  pdf_document: default
---

```{r header}

#  ██████╗ ██╗  ██╗██╗   ██╗██╗      ██████╗ ██████╗ ██╗  ██╗███████╗██████╗ ███████╗
#  ██╔══██╗██║  ██║╚██╗ ██╔╝██║     ██╔═══██╗██╔══██╗██║  ██║██╔════╝██╔══██╗██╔════╝
#  ██████╔╝███████║ ╚████╔╝ ██║     ██║   ██║██████╔╝███████║█████╗  ██████╔╝█████╗
#  ██╔═══╝ ██╔══██║  ╚██╔╝  ██║     ██║   ██║██╔═══╝ ██╔══██║██╔══╝  ██╔══██╗██╔══╝
#  ██║     ██║  ██║   ██║   ███████╗╚██████╔╝██║     ██║  ██║███████╗██║  ██║███████╗
#  ╚═╝     ╚═╝  ╚═╝   ╚═╝   ╚══════╝ ╚═════╝ ╚═╝     ╚═╝  ╚═╝╚══════╝╚═╝  ╚═╝╚══════╝

## PHYLOPHERE: A Nextflow pipeline including a complete set
## of phylogenetic comparative tools and analyses for Phenome-Genome studies

### Github: https://github.com/nozerorma/caastools/nf-phylophere
### Author: Miguel Ramon (miguel.ramon@upf.edu)

#### File: CAAS_postproc.Rmd

```

This script processes CAAS (Context-Aware Amino Acid Substitutions) discovery results to identify and remove false positives caused by high-density position clusters ("trains") and statistical outliers. The workflow includes outlier detection, cluster filtering analysis, and generation of summary statistics for downstream analysis.

# Setup and Directory Configuration

This section configures the working environment, sets directories, and loads necessary functions and libraries.

```{r setup, include = FALSE}
# ----------------------------------------
# R Markdown Setup
# ----------------------------------------

seed_val <- 12345  # For reproducibility

# Create a directory if it doesn't exist
createDir <- function(directory) {
  if (!file.exists(directory)) {
    dir.create(directory, recursive = TRUE)
    print(sprintf("createDir: created %s", directory))
  } else {
    print(sprintf("createDir: exists %s", directory))
  }
}

setup_rmd <- function() {
  knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, dev='png', dpi=320)
  knitr::opts_knit$set(root.dir =getwd()) # Set working directory to project root
  set.seed(as.integer(seed_val)) 
}
# Define working and results directories
workingDir <- getwd()
setup_rmd()

# Load additional libraries
suppressPackageStartupMessages({
  library(stringr)
  library(purrr)
  library(tibble)
  library(patchwork)
  library(scales)
  library(ggplot2)
  library(forcats)
  library(ggvenn)
  library(knitr)
  library(kableExtra)
  library(paletteer)
  library(tidyverse)
})

# Parse command line arguments
args <- commandArgs(trailingOnly = TRUE)
if (length(args) >= 10) {
  discovery_file <- args[1]
  filter_summary_file <- args[2]
  gene_ensembl_file <- args[3]
  processing_mode <- args[4]  # "exploratory" or "filter"
  resultsDir <- args[5]
  extreme_threshold <- as.numeric(args[6])
  iqr_multiplier <- as.numeric(args[7])
  gene_filter_mode <- args[8]  # "none", "extreme", "dubious", "both"
  generate_manhattan <- as.logical(args[9])
  manhattan_min_density <- as.numeric(args[10])
} else if (length(args) >= 3) {
  # Backward compatibility - minimal args
  discovery_file <- args[1]
  filter_summary_file <- args[2]
  gene_ensembl_file <- args[3]
  processing_mode <- "exploratory"
  resultsDir <- file.path(getwd(), "caas_postproc")
  extreme_threshold <- 0.99
  iqr_multiplier <- 3.0
  gene_filter_mode <- "both"
  generate_manhattan <- TRUE
  manhattan_min_density <- 0.008
} else {
  stop("Usage: Rscript CAAS_postproc.Rmd <discovery> <filter_summary> <gene_ensembl> [mode] [outdir] [extreme_thresh] [iqr_mult] [gene_filter] [gen_manhattan] [manhattan_min_dens]")
}

# Validate input files
stopifnot(file.exists(discovery_file), 
          file.exists(filter_summary_file),
          file.exists(gene_ensembl_file))

# Define output directories from passed parameter
caas_postproc_dir <- resultsDir
outliers_dir <- file.path(caas_postproc_dir, "outliers")
cluster_dir <- file.path(caas_postproc_dir, "clusters")

# Directory for filtered discovery files (in exploratory mode)
# These are in subdirectories: filter_exploratory/minlenX_maxcaasY/
discovery_filtered_base_dir <- file.path(caas_postproc_dir, 
                                        ifelse(processing_mode == "exploratory", 
                                               "filter_exploratory", 
                                               "filter_selected"))

# Define color palettes, including a colorblind-friendly palette for categories
tableau_10 <- paletteer_d("ggthemes::Tableau_10")
names(tableau_10) <- c(
  "blue", "orange", "vermillion", "bluish_green",
  "green", "yellow", "reddish_purple", "pink",
  "mauve", "gray"
)

# Consistent colorblind-friendly palette
category_palette <- c(
  "Normal" = tableau_10[["gray"]],
  "Dubious" = tableau_10[["blue"]],
  "Extreme" = tableau_10[["vermillion"]],
  "Both" = tableau_10[["reddish_purple"]]
)

removal_palette <- c(
  Discarded = tableau_10[["vermillion"]],
  Good = tableau_10[["green"]]
)

```

## Data Loading and Gene Length Annotation

This section loads CAAS discovery results and annotates genes with their alignment lengths.

```{r data_loading}

cat("=== CAAS Post-Processing Configuration ===\n")
cat("Processing mode:", processing_mode, "\n")
cat("Gene filtering:", gene_filter_mode, "\n")
cat("Extreme threshold:", extreme_threshold, paste0("(top ", 100*(1-extreme_threshold), "%)"), "\n")
cat("IQR multiplier:", iqr_multiplier, "\n")
cat("Generate Manhattan plots:", generate_manhattan, "\n")
if (generate_manhattan) {
  cat("Manhattan min density:", manhattan_min_density, "\n")
}

createDir(caas_postproc_dir)
createDir(outliers_dir)
createDir(cluster_dir)

cat("\nInput files:\n")
cat("  Discovery:", discovery_file, "\n")
cat("  Filter summary:", filter_summary_file, "\n")
cat("  Gene ensembl:", gene_ensembl_file, "\n\n")

# Load CAAS discovery file
cat("Loading CAAS discovery results...\n")

# Here we load the discovery, either from the pipeline or from a discovery file generated previously

# Read discovery file (tab-separated, expects Gene and Position columns)
discovery_df <- read_tsv(discovery_file, show_col_types = FALSE)

cat("  Loaded CAAS positions:", nrow(discovery_df), "\n")
cat("  Unique genes:", n_distinct(discovery_df$Gene), "\n")

# Load gene lengths (no header: Gene ↔ Length)
cat("Loading gene length annotations...\n")

# Read gene ensembl file (format: gene, chr, start, end, strand, length)
gene_ensembl <- read_tsv(gene_ensembl_file, show_col_types = FALSE)

# Normalize column names to lowercase
names(gene_ensembl) <- tolower(names(gene_ensembl))

# Validate required columns
if (!"gene" %in% names(gene_ensembl) || !"length" %in% names(gene_ensembl)) {
  stop("Gene ensembl file must have 'gene' and 'length' columns")
}

# Extract Gene and Length for backwards compatibility
gene_length <- gene_ensembl %>%
  select(Gene = gene, Length = length)

cat("  Gene length records:", nrow(gene_length), "\n")

# Annotate CAAS with gene lengths
discovery_df <- discovery_df %>%
  left_join(gene_length, by = "Gene") %>%
  filter(!is.na(Length))  # Remove genes without length info

cat("  Genes with length annotation:", n_distinct(discovery_df$Gene), "\n\n")
```

## Outlier Gene Detection

This section identifies genes with unusually high CAAS counts using two complementary methods:

- **Extreme genes**: Top percentile by CAAS density (positions per gene length)
- **Dubious genes**: IQR outliers (Tukey's fences at k×IQR)

CAAP mode: If discovery data contains CAAP_Group column, thresholds are calculated per group (us, gs1-gs4). GS0 is displayed for reference but excluded from removal criteria (null hypothesis).

Note: Actual gene removal is performed by filter_caas_genes.py module. This section visualizes thresholds.

```{r outlier_detection, fig.height=10, fig.width=14}
# Detect CAAP mode
has_caap_group <- 'CAAP_Group' %in% names(discovery_df)

if (has_caap_group) {
  cat("CAAP mode detected: Group-aware analysis enabled\n")
  cat("Groups found:", paste(unique(discovery_df$CAAP_Group), collapse=", "), "\n")
  
  # Filter for non-gs0 groups for threshold calculation
  discovery_for_thresholds <- discovery_df %>% filter(CAAP_Group != 'gs0')
  groupby_cols <- c('CAAP_Group')
  
  # Summarize CAAS counts and density by gene and group
  gene_summary <- discovery_for_thresholds %>%
    group_by(CAAP_Group, Gene) %>%
    summarise(
      n_CAAS = n(),
      Length = first(Length),
      n_CAAS_per_length = n() / first(Length) * 100,
      .groups = 'drop'
    )
  
  # Calculate IQR-based outlier thresholds per group
  iqr_thresholds <- gene_summary %>%
    group_by(CAAP_Group) %>%
    summarise(
      Q1 = quantile(n_CAAS, 0.25, na.rm = TRUE),
      Q3 = quantile(n_CAAS, 0.75, na.rm = TRUE),
      IQR_val = Q3 - Q1,
      threshold_dubious = Q3 + iqr_multiplier * IQR_val,
      .groups = 'drop'
    )
  
  # Calculate top percentile threshold per group
  extreme_thresholds <- gene_summary %>%
    group_by(CAAP_Group) %>%
    summarise(
      threshold_extreme = quantile(n_CAAS_per_length, extreme_threshold, na.rm = TRUE),
      .groups = 'drop'
    )
  
  # Annotate genes with thresholds
  gene_summary_annotated <- gene_summary %>%
    left_join(extreme_thresholds, by = 'CAAP_Group') %>%
    left_join(iqr_thresholds, by = 'CAAP_Group') %>%
    mutate(
      Extreme = n_CAAS_per_length >= threshold_extreme,
      Dubious = n_CAAS >= threshold_dubious,
      Category = case_when(
        Extreme & Dubious ~ "Both",
        Extreme ~ "Extreme",
        Dubious ~ "Dubious",
        TRUE ~ "Normal"
      )
    ) %>%
    arrange(desc(n_CAAS))
  
  # Add gs0 for reference (but mark as excluded from removal)
  # NOTA: darle una vuelta
  if ('gs0' %in% discovery_df$CAAP_Group) {
    gs0_summary <- discovery_df %>%
      filter(CAAP_Group == 'gs0') %>%
      group_by(CAAP_Group, Gene) %>%
      summarise(
        n_CAAS = n(),
        Length = first(Length),
        n_CAAS_per_length = n() / first(Length) * 100,
        .groups = 'drop'
      ) %>%
      mutate(
        Category = "Normal (gs0)",
        Extreme = FALSE,
        Dubious = FALSE,
        threshold_extreme = NA_real_,
        threshold_dubious = NA_real_,
        Q1 = NA_real_,
        Q3 = NA_real_,
        IQR_val = NA_real_
      )
    
    gene_summary_with_gs0 <- bind_rows(gene_summary_annotated, gs0_summary)
  } else {
    gene_summary_with_gs0 <- gene_summary_annotated
  }
  
} else {
  cat("Standard CAAS mode (single group)\n")
  
  # Summarize CAAS counts and density by gene
  gene_summary <- discovery_df %>%
    group_by(Gene) %>%
    summarise(
      n_CAAS = n(),
      Length = first(Length),
      n_CAAS_per_length = n() / first(Length) * 100,
      .groups = 'drop'
    )
  
  # Calculate IQR-based extreme outlier thresholds (k×IQR from config)
  Q1 <- quantile(gene_summary$n_CAAS, 0.25, na.rm = TRUE)
  Q3 <- quantile(gene_summary$n_CAAS, 0.75, na.rm = TRUE)
  IQR_val <- Q3 - Q1
  threshold_dubious <- Q3 + iqr_multiplier * IQR_val
  
  # Calculate top percentile threshold by CAAS density (from config)
  threshold_extreme <- quantile(gene_summary$n_CAAS_per_length, extreme_threshold, na.rm = TRUE)
  
  # Annotate genes as Extreme and/or Dubious
  gene_summary_annotated <- gene_summary %>%
    mutate(
      Extreme = n_CAAS_per_length >= threshold_extreme,
      Dubious = n_CAAS >= threshold_dubious,
      Category = case_when(
        Extreme & Dubious ~ "Both",
        Extreme ~ "Extreme",
        Dubious ~ "Dubious",
        TRUE ~ "Normal"
      ),
      threshold_extreme = threshold_extreme,
      threshold_dubious = threshold_dubious
    ) %>%
    arrange(desc(n_CAAS))
  
  gene_summary_with_gs0 <- gene_summary_annotated
}

# Generate plots
if (has_caap_group) {
  # Group-faceted CAAS count distribution
  p_gross_dist <- ggplot(gene_summary_with_gs0, aes(x = n_CAAS, fill = Category)) +
    geom_histogram(binwidth = 0.05, color = "black", alpha = 0.8) +
    geom_vline(
      data = gene_summary_with_gs0 %>% filter(!is.na(threshold_dubious)) %>% distinct(CAAP_Group, threshold_dubious),
      aes(xintercept = threshold_dubious),
      color = tableau_10[["blue"]],
      linetype = "dashed",
      linewidth = 1
    ) +
    scale_fill_manual(
      values = c(category_palette),
      breaks = c("Extreme", "Dubious", "Both", "Normal"),
      labels = c("Extreme", "Dubious", "Both", "Normal")
    ) +
    scale_x_log10(breaks = c(1, 2, 5, 10, 20, 50, 100, 200)) +
    facet_wrap(~CAAP_Group, scales = "free_y") +
    labs(title = "CAAP Count Distribution by Group",
         subtitle = "Dubious threshold (Q3 + 3×IQR) shown as dashed line | gs0 displayed for reference",
         x = "Number of CAAP Positions", 
         y = "Gene Count") +
    theme_minimal() +
    theme(legend.position = "bottom",
          plot.title = element_text(face = "bold", size = 14),
          axis.title = element_text(size = 12))
  
  # Zoomed view focusing on typical range
  p_zoom_gross <- p_gross_dist + 
    coord_cartesian(ylim = c(0, 100)) +
    labs(title = "Zoomed View (y ≤ 100)")

  # Group-faceted CAAS density distribution
  p_density_dist <- ggplot(gene_summary_with_gs0, aes(x = n_CAAS_per_length, fill = Category)) +
  geom_histogram(binwidth = 0.05, color = "black", alpha = 0.8) +
    geom_vline(
      data = gene_summary_with_gs0 %>% filter(!is.na(threshold_extreme)) %>% distinct(CAAP_Group, threshold_extreme),
      aes(xintercept = threshold_extreme),
      color = tableau_10[["vermillion"]],
      linetype = "dashed",
      linewidth = 1
    ) +
    scale_fill_manual(
      values = c(category_palette),
      breaks = c("Extreme", "Dubious", "Both", "Normal"),
      labels = c("Extreme", "Dubious", "Both", "Normal")
    ) +
    scale_x_log10(breaks = c(0.01, 0.02, 0.05, 0.1, 0.2, 0.5, 1, 2, 5, 10, 20, 50, 100)) +
    facet_wrap(~ CAAP_Group, scales = "free_y") +
    labs(title = "CAAP Density Distribution by Group",
         subtitle = paste0("Top ", 100*(1-extreme_threshold), "% density threshold shown as dashed line | gs0 displayed for reference"),
         x = "CAAP Density (positions / gene length)", 
         y = "Gene Count") +
    theme_minimal() +
    theme(legend.position = "bottom",
          plot.title = element_text(face = "bold", size = 14),
          axis.title = element_text(size = 12))

    # Zoomed density view
    p_zoom_density <- p_density_dist + 
      coord_cartesian(ylim = c(0, 50)) +
      labs(title = "Zoomed View (y ≤ 50)")
  
  # Combined plot
  combined_outlier_plot <- p_gross_dist / p_density_dist

  # Combined plot layout (zoomed versions side by side)
  combined_zoomed_plot <- (p_zoom_gross / p_zoom_density)

  
} else {
  # Standard single-group plots
  p_gross_dist <- ggplot(gene_summary_annotated, aes(x = n_CAAS, fill = Category)) +
    geom_histogram(binwidth = 0.05, color = "black", fill = "grey80", alpha = 0.8) +
    geom_vline(xintercept = threshold_dubious, linetype = "dashed", color = tableau_10[["blue"]], size = 1) +
    scale_x_log10(breaks = c(1, 2, 5, 10, 20, 50, 100, 200)) +
    scale_fill_manual(values = category_palette) +
    labs(title = "CAAS Count Distribution by Gene",
         subtitle = "Outliers identified by density (Extreme) and IQR (Dubious)",
         x = "Number of CAAS Positions", 
         y = "Gene Count") +
    theme_minimal() +
    theme(legend.position = "right",
          plot.title = element_text(face = "bold", size = 14),
          axis.title = element_text(size = 12))
  
  # Zoomed view focusing on typical range
  p_zoom_gross <- p_gross_dist + 
    coord_cartesian(ylim = c(0, 100)) +
    labs(title = "Zoomed View (y ≤ 100)")

  p_density_dist <- ggplot(gene_summary_annotated, aes(x = n_CAAS_per_length, fill = Category)) +
    geom_histogram(binwidth = 0.05, color = "black", fill = "grey80", alpha = 0.8) +
    scale_fill_manual(values = category_palette) +
    geom_vline(xintercept = threshold_extreme, linetype = "dashed", color = tableau_10[["vermillion"]], size = 1) +
    scale_x_log10(breaks = c(0.01, 0.02, 0.05, 0.1, 0.2, 0.5, 1, 2, 5, 10, 20, 50, 100)) +
    labs(title = "CAAS Density Distribution",
         subtitle = "Top 1% density threshold shown as dashed line",
         x = "CAAS Density (positions / gene length)", 
         y = "Gene Count") +
    theme_minimal() +
    theme(legend.position = "right",
          plot.title = element_text(face = "bold", size = 14),
          axis.title = element_text(size = 12))

  # Zoomed density view
  p_zoom_density <- p_density_dist + 
    coord_cartesian(ylim = c(0, 50)) +
    labs(title = "Zoomed View (y ≤ 50)")
  
  # Combined plot
  combined_outlier_plot <- p_gross_dist / p_density_dist

  # Combined plot layout (zoomed versions side by side)
  combined_zoomed_plot <- (p_zoom_gross / p_zoom_density)
  
}

print(combined_outlier_plot)
print(combined_zoomed_plot)

ggsave(
  file.path(outliers_dir, "outlier_distributions.png"),
  combined_outlier_plot,
  width = 14, height = 10, dpi = 320
)

ggsave(
  file.path(outliers_dir, "outlier_distributions_zoomed.png"),
  combined_zoomed_plot,
  width = 14, height = 10, dpi = 320
)

# Summary statistics
cat("\n=== Outlier Detection Summary ===\n")

if (has_caap_group) {
  cat("CAAP mode: Group-aware thresholds calculated\n")
  cat("Groups analyzed for removal: us, gs1-gs4 (gs0 excluded as null hypothesis)\n\n")
  
  # Thresholds by group
  thresholds_summary <- gene_summary_annotated %>%
    distinct(CAAP_Group, threshold_extreme, threshold_dubious) %>%
    arrange(CAAP_Group)
  
  cat("Thresholds by group:\n")
  for (i in 1:nrow(thresholds_summary)) {
    cat(sprintf("  %s:\n", thresholds_summary$CAAP_Group[i]))
    cat(sprintf("    Extreme (top %.0f%% density): %.4f\n", 
                100*(1-extreme_threshold), 
                thresholds_summary$threshold_extreme[i]))
    cat(sprintf("    Dubious (Q3 + %.1f×IQR): %.0f positions\n\n", 
                iqr_multiplier, 
                thresholds_summary$threshold_dubious[i]))
  }
  
  # Counts by group and category
  outlier_counts <- gene_summary_annotated %>%
    count(CAAP_Group, Category) %>%
    pivot_wider(names_from = Category, values_from = n, values_fill = 0)
  
  cat("Gene counts by group and category:\n")
  print(outlier_counts)
  
} else {
  cat("Thresholds:\n")
  cat("  Extreme (top 1% density):", sprintf("%.4f", threshold_extreme), "\n")
  cat("  Dubious (Q3 + 3×IQR):", threshold_dubious, "positions\n\n")
  
  outlier_counts <- gene_summary_annotated %>%
    count(Category) %>%
    mutate(Percent = sprintf("%.1f%%", 100 * n / sum(n)))
  
  print(outlier_counts)
}

# Export gene summary
write_tsv(
  gene_summary_annotated,
  file.path(outliers_dir, "gene_summary_annotated.tsv")
)

cat("\n✓ Gene summary exported\n\n")
```

# Cluster Filtering Analysis

This section analyzes the cluster filtering results from the parameter sweep or single filter run.

In exploratory mode, each parameter combination gets its own detailed analysis page.

```{r cluster_analysis_load}
# Load filter summary
cat("Loading cluster filter summary...\n")

filter_summary <- read_tsv(filter_summary_file, show_col_types = FALSE)

cat("  Parameter combinations tested:", nrow(filter_summary), "\n")

# Check if CAAP mode
has_caap_groups_in_summary <- 'CAAP_Groups' %in% names(filter_summary) && 
                               any(filter_summary$CAAP_Groups != 'N/A' & filter_summary$CAAP_Groups != '')

if (has_caap_groups_in_summary) {
  cat("  CAAP mode detected in filter summary\n")
  # Parse comma-separated groups
  filter_summary <- filter_summary %>%
    mutate(
      Groups_List = strsplit(as.character(CAAP_Groups), ","),
      N_Groups = lengths(Groups_List)
    )
}

# Create pivot table: rows = parameters, columns = groups (if CAAP)
if (has_caap_groups_in_summary) {
  # CAAP mode: create wide table with groups as columns
  filter_summary_wide <- filter_summary %>%
    mutate(param_label = sprintf("%d/%.0f", Minlen, Maxcaas)) %>%
    select(param_label, Minlen, Maxcaas, CAAP_Groups, DiscardedCount)
  
  # Parse groups and create one row per parameter-group combination
  filter_summary_long <- filter_summary_wide %>%
    separate_rows(CAAP_Groups, sep = ",") %>%
    mutate(CAAP_Groups = trimws(CAAP_Groups)) %>%
    filter(CAAP_Groups != "N/A" & CAAP_Groups != "")
  
  # Get unique groups for column ordering
  all_groups <- sort(unique(filter_summary_long$CAAP_Groups))
  
  # Pivot to wide format
  filter_table <- filter_summary_long %>%
    pivot_wider(names_from = CAAP_Groups, values_from = DiscardedCount, values_fill = 0) %>%
    select(param_label, all_of(all_groups))
  
  names(filter_table)[1] <- "Parameter"
} else {
  # Standard mode: simple table
  filter_table <- filter_summary %>%
    mutate(Parameter = sprintf("%d/%.0f", Minlen, Maxcaas)) %>%
    select(Parameter, DiscardedCount)
  names(filter_table)[2] <- "Discarded"
}

kable(filter_table, 
      caption = "Cluster Filtering Parameter Sweep Results",
      align = c("l", rep("r", ncol(filter_table) - 1))) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Plot parameter sweep results (if multiple combinations)
if (nrow(filter_summary) > 1) {
  if (has_caap_groups_in_summary) {
    # CAAP mode: use the correctly prepared long format data
    # This has the actual per-group counts, not the total duplicated across groups
    plot_data <- filter_summary_long

    print(head(plot_data))
    
    p_param_sweep <- ggplot(plot_data, 
                            aes(x = factor(Minlen), 
                                y = DiscardedCount, 
                                color = CAAP_Groups,
                                linetype = factor(Maxcaas))) +
      geom_bar(stat = "identity", position = "dodge", width = 0.7) +
      geom_line(linewidth = 1.2) +
      scale_color_paletteer_d("ggthemes::Tableau_10") +
      scale_linetype_manual(values = c("solid", "dashed", "dotted"),
                           labels = function(x) sprintf("%.0f%%", as.numeric(x))) +
      labs(title = "Discarded CAAS Positions by Parameter Combination",
           subtitle = "Effect of minimum cluster length and maximum density threshold (group-aware filtering)",
           x = "Minimum Cluster Length (minlen)", 
           y = "Positions Discarded", 
           color = "Group",
           linetype = "Max Density") +
      theme_minimal() +
      theme(legend.position = "right",
            plot.title = element_text(face = "bold", size = 14),
            axis.title = element_text(size = 12))
  } else {
    # Standard mode: single series with color for density
    p_param_sweep <- ggplot(filter_summary, 
                            aes(x = factor(Minlen), 
                                y = DiscardedCount, 
                                color = factor(Maxcaas),
                                group = factor(Maxcaas))) +
      geom_point(size = 4) +
      geom_line(linewidth = 1.2) +
      scale_color_paletteer_d("ggthemes::Tableau_10",
                             labels = function(x) sprintf("%.0f%%", as.numeric(x))) +
      labs(title = "Discarded CAAS Positions by Parameter Combination",
           subtitle = "Effect of minimum cluster length and maximum density threshold",
           x = "Minimum Cluster Length (minlen)", 
           y = "Positions Discarded", 
           color = "Max Density") +
      theme_minimal() +
      theme(legend.position = "right",
            plot.title = element_text(face = "bold", size = 14),
            axis.title = element_text(size = 12))
  }
  
  print(p_param_sweep)
  
  ggsave(
    file.path(cluster_dir, "parameter_sweep_overview.png"),
    p_param_sweep,
    width = 12, height = 6, dpi = 320
  )
}

# Export filter summary
write_tsv(
  filter_summary,
  file.path(cluster_dir, "filter_summary.tsv")
)

cat("✓ Filter summary exported\n\n")
```

```{r cluster_detailed_analysis, results='asis', fig.height=8, fig.width=12,  eval=(processing_mode == 'exploratory' && nrow(filter_summary) > 1)}
# Generate detailed pages per parameter combination in exploratory mode
if (processing_mode == 'exploratory' && nrow(filter_summary) > 1) {
  cat("\n## Detailed Parameter Analysis\n\n")
  cat("This section provides detailed analysis for each parameter combination tested.\n\n")
  
  for (i in 1:nrow(filter_summary)) {
    param_row <- filter_summary[i, ]
    param_label <- sprintf("minlen=%d, maxcaas=%.2f", param_row$Minlen, param_row$Maxcaas)
    
    cat(sprintf("\n### Parameter Set %d: %s {.tabset}\n\n", i, param_label))
    
    cat(sprintf("**Configuration:** Minimum cluster length = %d, Maximum density = %.2f\n\n", 
                param_row$Minlen, param_row$Maxcaas))
    cat(sprintf("**Positions discarded:** %d\n\n", param_row$DiscardedCount))
    
    if (has_caap_groups_in_summary && param_row$CAAP_Groups != 'N/A') {
      groups_affected <- unlist(strsplit(as.character(param_row$CAAP_Groups), ","))
      cat(sprintf("**Groups with discarded positions:** %s\n\n", paste(groups_affected, collapse=", ")))
    }
    
    cat("\n#### Summary\n\n")
    cat(sprintf("This parameter combination resulted in %d discarded CAAS positions.\n\n", 
                param_row$DiscardedCount))
    
    if (has_caap_groups_in_summary) {
      cat("In CAAP mode, cluster filtering respects group boundaries - positions are only considered clustered if they occur within the same CAAP group.\n\n")
    }
    
    # Load filtered file for detailed analysis
    # Files are in subdirectories: filter_exploratory/minlenX_maxcaasY/discovery.filtered.minlenX.maxcaasY.tsv
    filter_subdir <- sprintf("minlen%d_maxcaas%d", param_row$Minlen, as.integer(param_row$Maxcaas))
    pattern_str <- sprintf("discovery.filtered.minlen%d.maxcaas%d.tsv", 
                          param_row$Minlen, as.integer(param_row$Maxcaas))
    matching_file <- file.path(discovery_filtered_base_dir, filter_subdir, pattern_str)
    
    if (!is.na(matching_file) && file.exists(matching_file)) {
      # Load filtered data
      filtered_data <- read_tsv(matching_file, show_col_types = FALSE)
      
      if ("ClusteringFlag" %in% names(filtered_data)) {
        discarded <- filtered_data %>% filter(ClusteringFlag == "Discarded")
        
        if (nrow(discarded) > 0) {
          # Function to detect spatial clustering patterns (runs)
          detect_runs <- function(positions) {
            if (length(positions) == 0) return(integer(0))
            positions <- sort(unique(positions))
            if (length(positions) == 0) return(integer(0))
            breaks <- c(0, which(diff(positions) > 1), length(positions))
            return(diff(breaks))
          }
          
          # Detect spatial clustering by gene
          if (has_caap_groups_in_summary && "CAAP_Group" %in% names(discarded)) {
            # Group-aware analysis
            spatial_analysis <- discarded %>%
              group_by(CAAP_Group, Gene) %>%
              summarise(positions = list(Position), .groups = "drop") %>%
              mutate(run_lengths = map(positions, detect_runs)) %>%
              unnest(run_lengths, keep_empty = TRUE) %>%
              mutate(
                cluster_type = case_when(
                  run_lengths == 1 ~ "Singlet (1)",
                  run_lengths == 2 ~ "Doublet (2)",
                  run_lengths == 3 ~ "Triplet (3)",
                  run_lengths == 4 ~ "Quad (4)",
                  run_lengths == 5 ~ "Quint (5)",
                  run_lengths >= 6 ~ "6+ Long",
                  TRUE ~ "Other"
                )
              )
            
            # Summary by group
            spatial_summary <- spatial_analysis %>%
              count(CAAP_Group, cluster_type) %>%
              complete(CAAP_Group, cluster_type, fill = list(n = 0)) %>%
              group_by(CAAP_Group) %>%
              mutate(percentage = round(100 * n / sum(n), 1)) %>%
              ungroup()
            
            # Plot
            p_spatial <- ggplot(spatial_summary, 
                               aes(x = CAAP_Group, y = percentage/100, fill = cluster_type)) +
              geom_col(position = "fill", color = "black", size = 0.2) +
              scale_fill_paletteer_d("ggthemes::Tableau_10") +
              scale_y_continuous(labels = scales::percent) +
              labs(title = sprintf("Spatial Clustering: %s", param_label),
                   subtitle = "Distribution of singlets, doublets, triplets, etc. by group",
                   x = "Group",
                   y = "Percentage",
                   fill = "Cluster Type") +
              theme_minimal(base_size = 12) +
              theme(legend.position = "bottom",
                    axis.text.x = element_text(angle = 45, hjust = 1))
            
            print(p_spatial)
            
            # Export table
            cat("\\n#### Spatial Clustering Details\\n\\n")
            kable(spatial_summary %>% 
                   pivot_wider(names_from = cluster_type, values_from = n, values_fill = 0),
                 caption = sprintf("Cluster types for %s", param_label)) %>%
              kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
            
          } else {
            # Standard analysis (no groups)
            spatial_analysis <- discarded %>%
              group_by(Gene) %>%
              summarise(positions = list(Position), .groups = "drop") %>%
              mutate(run_lengths = map(positions, detect_runs)) %>%
              unnest(run_lengths, keep_empty = TRUE) %>%
              mutate(
                cluster_type = case_when(
                  run_lengths == 1 ~ "Singlet (1)",
                  run_lengths == 2 ~ "Doublet (2)",
                  run_lengths == 3 ~ "Triplet (3)",
                  run_lengths == 4 ~ "Quad (4)",
                  run_lengths == 5 ~ "Quint (5)",
                  run_lengths >= 6 ~ "6+ Long",
                  TRUE ~ "Other"
                )
              )
            
            spatial_summary <- spatial_analysis %>%
              count(cluster_type) %>%
              mutate(percentage = round(100 * n / sum(n), 1))
            
            # Plot
            p_spatial <- ggplot(spatial_summary, 
                               aes(x = cluster_type, y = n, fill = cluster_type)) +
              geom_col(color = "black", size = 0.2) +
              scale_fill_paletteer_d("ggthemes::Tableau_10") +
              geom_text(aes(label = sprintf("%d (%.1f%%)", n, percentage)), 
                       vjust = -0.5, size = 3.5) +
              labs(title = sprintf("Spatial Clustering: %s", param_label),
                   subtitle = "Distribution of singlets, doublets, triplets, etc.",
                   x = "Cluster Type",
                   y = "Count",
                   fill = "Cluster Type") +
              theme_minimal(base_size = 12) +
              theme(legend.position = "none",
                    axis.text.x = element_text(angle = 45, hjust = 1))
            
            print(p_spatial)
            
            cat("\\n#### Spatial Clustering Details\\n\\n")
            kable(spatial_summary,
                 caption = sprintf("Cluster types for %s", param_label),
                 col.names = c("Cluster Type", "Count", "Percentage (%)")) %>%
              kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
          }
          
          # Save plot
          safe_label <- gsub("[^A-Za-z0-9]", "_", param_label)
          ggsave(
            file.path(cluster_dir, sprintf("spatial_clustering_%s.png", safe_label)),
            p_spatial,
            width = 10, height = 6, dpi = 320
          )
        } else {
          cat("**Note:** No discarded positions for this parameter combination.\\n\\n")
        }
      }
    } else {
      cat("**Note:** Filtered file not found for detailed cluster type analysis.\\n\\n")
    }
    
    cat("\n---\n\n")
  }
}
```

## 4. Genomic Distribution Visualization (Manhattan Plot)

This section visualizes CAAS density across the genome if genomic coordinates are available and Manhattan plotting is enabled.

```{r manhattan_plot, fig.height=8, fig.width=16, eval=generate_manhattan}
# Check if gene_ensembl data has chromosome information
if (!"chr" %in% names(gene_ensembl)) {
  cat("⚠ Chromosome information not available in gene ensembl file, skipping Manhattan plot\n")
} else {
  cat("Preparing Manhattan plot data...\n")
  
  # Use the gene_ensembl data that's already loaded (has: gene, chr, start, end, strand, length)
  genomic_info <- gene_ensembl
  
  # Helper functions for chromosome layout
  compute_chr_layout <- function(df) {
    layout_df <- df %>%
      mutate(chr_num = str_replace(chr, "chr", "")) %>%
      mutate(chr_rank = case_when(
        chr_num %in% as.character(1:22) ~ as.numeric(chr_num),
        chr_num == "X" ~ 23,
        chr_num == "Y" ~ 24,
        chr_num %in% c("M", "MT") ~ 25,
        chr_num == "chrNA" ~ 26, # Place unknown chromosomes at the end
        TRUE ~ 26
      )) %>%
      group_by(chr_num, chr_rank) %>%
      summarise(chr_max = max(pos, na.rm = TRUE), .groups = "drop") %>%
      arrange(chr_rank) %>%
      mutate(
        chr_start = lag(cumsum(chr_max), default = 0),
        chr_center = chr_start + chr_max / 2
      )
    layout_df %>% mutate(chr_order = chr_rank)
  }
  
  apply_chr_layout <- function(df, layout_df) {
    df %>%
      mutate(chr_num = str_replace(chr, "chr", "")) %>%
      left_join(layout_df %>% select(chr_num, chr_start), by = "chr_num") %>%
      mutate(BPcum = pos + chr_start)
  }
  
  # Prepare Manhattan plot data
  manhattan_data <- gene_summary_annotated %>%
    left_join(genomic_info %>% select(gene, chr, start, end), 
              by = c("Gene" = "gene")) %>%
    filter(!is.na(chr)) %>%
    mutate(
      chr_num = str_replace(chr, "chr", ""),
      chr_order = case_when(
        chr_num %in% as.character(1:22) ~ as.numeric(chr_num),
        chr_num == "X" ~ 23,
        chr_num == "Y" ~ 24,
        chr_num == "M" ~ 25,
        chr_num == "chrNA" ~ 26, # Place unknown chromosomes at the end
        TRUE ~ 26
      ),
      pos = (start + end) / 2,
      plot_density = case_when(
        is.na(n_CAAS_per_length) ~ NA_real_,
        n_CAAS_per_length <= 0 ~ manhattan_min_density,
        TRUE ~ n_CAAS_per_length
      )
    ) %>%
    arrange(chr_order, pos)
  
  if (nrow(manhattan_data) == 0) {
    cat("⚠ No genes with genomic coordinates found\n")
  } else {
    # Calculate chromosome layout
    chr_lengths <- manhattan_data %>%
      select(chr, pos) %>%
      compute_chr_layout()
    
    # Add cumulative positions
    manhattan_data <- apply_chr_layout(manhattan_data, chr_lengths)
    
    # Axis data for chromosome labels
    axis_data <- chr_lengths %>%
      select(chr_num, chr_center) %>%
      rename(center = chr_center)
    
    # Create Manhattan plot
    p_manhattan <- ggplot(manhattan_data, aes(x = BPcum, y = plot_density, color = Category)) +
      geom_vline(
        data = chr_lengths %>% filter(chr_order > 1),
        aes(xintercept = chr_start),
        color = "gray20",
        linetype = "dashed",
        linewidth = 0.3,
        alpha = 0.6
      ) +
      geom_point(
        data = manhattan_data %>% filter(Category == "Normal"),
        alpha = 0.35, size = 1.5
      ) +
      geom_point(
        data = manhattan_data %>% filter(Category == "Dubious"),
        alpha = 0.8, size = 2.5
      ) +
      geom_point(
        data = manhattan_data %>% filter(Category == "Extreme"),
        alpha = 0.8, size = 2.5
      ) +
      geom_point(
        data = manhattan_data %>% filter(Category == "Both"),
        alpha = 1, size = 3, shape = 17
      ) +
      scale_color_manual(
        values = category_palette,
        drop = FALSE
      ) +
      scale_x_continuous(
        label = axis_data$chr_num,
        breaks = axis_data$center,
        expand = expansion(mult = c(0.005, 0.005))
      ) +
      scale_y_log10(
        breaks = c(0.01, 0.05, 0.1, 0.5, 1, 2, 5, 10, 20, 50),
        labels = c("0.01", "0.05", "0.1", "0.5", "1", "2", "5", "10", "20", "50")
      ) +
      labs(
        title = paste0("Genomic Distribution of CAAS Density (", processing_mode, " mode)"),
        subtitle = paste0("Gene filtering: ", gene_filter_mode, 
                         " | Extreme threshold: top ", 100*(1-extreme_threshold), "%",
                         " | IQR multiplier: ", iqr_multiplier, "×"),
        x = "Chromosome",
        y = "CAAS per Gene Length (%, log scale)",
        color = "Gene Category",
        caption = paste0(
          "Extreme genes (top percentile density) and Dubious genes (IQR outliers) ",
          "are flagged for potential removal. Genes in 'Both' category are flagged by both criteria."
        )
      ) +
      theme_minimal(base_size = 11) +
      theme(
        axis.text.x = element_text(angle = 0, hjust = 0.5, size = 9),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        legend.position = "top",
        legend.box = "horizontal",
        axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 10))
      )
    
    print(p_manhattan)
    
    ggsave(
      filename = file.path(outliers_dir, "manhattan_plot.png"),
      plot = p_manhattan,
      width = 16, 
      height = 8, 
      dpi = 320, 
      dev = "png"
    )
    
    cat("✓ Manhattan plot generated\n\n")
  }
}
```

## 5. Impact Summary

This section provides overall statistics on the post-processing analysis.

```{r impact_summary}
cat("\n=== CAAS Post-Processing Impact Summary ===\n\n")

# Extract thresholds from gene_summary_annotated (stored as columns)
if (has_caap_group) {
  # In CAAP mode, get thresholds from non-gs0 groups
  threshold_extreme <- gene_summary_annotated %>% 
    filter(CAAP_Group != "gs0") %>%
    pull(threshold_extreme) %>%
    first() %>%
    coalesce(NA_real_)
  threshold_dubious <- gene_summary_annotated %>% 
    filter(CAAP_Group != "gs0") %>%
    pull(threshold_dubious) %>%
    first() %>%
    coalesce(NA_real_)
} else {
  # Standard mode, extract from any row
  threshold_extreme <- gene_summary_annotated$threshold_extreme[1]
  threshold_dubious <- gene_summary_annotated$threshold_dubious[1]
}

# Gene statistics
cat("Gene-level statistics:\n")
cat("  Total genes analyzed:", n_distinct(discovery_df$Gene), "\n")
cat("  Extreme genes (top 1% density):", sum(gene_summary_annotated$Extreme), "\n")
cat("  Dubious genes (IQR outliers):", sum(gene_summary_annotated$Dubious), "\n")
cat("  Both categories:", sum(gene_summary_annotated$Extreme & gene_summary_annotated$Dubious), "\n")
cat("  Normal genes:", sum(gene_summary_annotated$Category == "Normal"), "\n\n")

# Position statistics
cat("Position-level statistics:\n")
cat("  Total CAAS positions:", nrow(discovery_df), "\n")

if (nrow(filter_summary) > 0) {
  cat("  Positions flagged by filtering:\n")
  for (i in 1:nrow(filter_summary)) {
    cat(sprintf("    minlen=%d, maxcaas=%.2f: %d positions\n",
                filter_summary$Minlen[i],
                filter_summary$Maxcaas[i],
                filter_summary$DiscardedCount[i]))
  }
}

# Create summary table for export
summary_table <- tibble(
  Metric = c(
    "Total CAAS Positions",
    "Unique Genes",
    "Extreme Genes (top 1% density)",
    "Dubious Genes (IQR outliers)",
    "Both Extreme and Dubious",
    "Normal Genes",
    "Extreme Threshold (density)",
    "Dubious Threshold (count)"
  ),
  Value = c(
    nrow(discovery_df),
    n_distinct(discovery_df$Gene),
    sum(gene_summary_annotated$Extreme),
    sum(gene_summary_annotated$Dubious),
    sum(gene_summary_annotated$Extreme & gene_summary_annotated$Dubious),
    sum(gene_summary_annotated$Category == "Normal"),
    sprintf("%.4f", threshold_extreme),
    as.character(threshold_dubious)
  )
)

kable(summary_table, 
      caption = "CAAS Post-Processing Summary Statistics",
      align = c("l", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Export summary
write_tsv(
  summary_table,
  file.path(caas_postproc_dir, "summary_statistics.tsv")
)

cat("\n✓ Summary statistics exported\n")
cat("\n=== Analysis Complete ===\n")
```

## Session Information

```{r session_info}
sessionInfo()
```
